### 2016-06-27 paper.Rnw
initial commit

\documentclass[12pt]{article}
% \usepackage{geometry}                % See geometry.pdf to learn the layout options. There are lots.
% \geometry{letterpaper}                   % ... or a4paper or a5paper or ...
%\usepackage{graphicx}
\usepackage[font=small,skip=5pt]{caption}
\usepackage{subcaption}
\usepackage{afterpage}
\usepackage{amssymb}
\usepackage{natbib}
\usepackage{amsmath}
\usepackage{amsfonts}
% \usepackage{color}
\usepackage{multirow}
\usepackage{rotating}
\usepackage[dvipsnames,svgnames,table]{xcolor}
\usepackage{hyperref}
\graphicspath{{figure/}}
% \usepackage{endfloat} % Figures to the end of the document

\DeclareGraphicsRule{.tif}{png}{.png}{`convert #1 `dirname #1`/`basename #1 .tif`.png}
%---------------------------------------------------
%                 Editing Commands
\newcommand{\hh}[1]{{\color{magenta} #1}}
\newcommand{\sct}[1]{{\color{orange} #1}}

%---------------------------------------------------
%                 Placing Figures

%---------------------------------------------------
% Define new environment
\newtheorem{theorem}{Theorem}[section]
\newtheorem{algorithm}[theorem]{Algorithm}
%---------------------------------------------------

%\pdfminorversion=4
% NOTE: To produce blinded version, replace "0" with "1" below.
\newcommand{\blind}{0}

% DON'T change margins - should be 1 inch all around.
\addtolength{\oddsidemargin}{-.5in}%
\addtolength{\evensidemargin}{-.5in}%
\addtolength{\textwidth}{1in}%
\addtolength{\textheight}{1.3in}%
\addtolength{\topmargin}{-.8in}%

\begin{document}

%\bibliographystyle{natbib}

\def\spacingset#1{\renewcommand{\baselinestretch}%
{#1}\small\normalsize} \spacingset{1}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\if0\blind
{
  \title{\bf Visual Diagnsotics of SAO Models}
  \author{Samantha Tyner\thanks{
    The authors gratefully acknowledge funding from the National Science Foundation Grant \# DMS 1007697. All data collection has been conducted with approval from the Institutional Review Board IRB 10-347}\hspace{.2cm}\\
    Department of Statistics and Statistical Laboratory, Iowa State University\\
    and \\
    Heike Hofmann\\
    Department of Statistics and Statistical Laboratory, Iowa State University}
  \maketitle
} \fi

\if1\blind
{
  \bigskip
  \bigskip
  \bigskip
  \begin{center}
    {\LARGE\bf Visual Diagnsotics of SAO Models}
\end{center}
  \medskip
} \fi

\bigskip
\begin{abstract}
 \hh{XXX just a placeholder}
\end{abstract}

\noindent%
{\it Keywords:}  Visual inference, Lineup protocol, \hh{XXX Other keywords?}.
\vfill

\newpage
\spacingset{1.45} % DON'T change the spacing!

\tableofcontents
\newpage

<<setup, echo = FALSE, message = FALSE, warning = FALSE>>=
options(replace.assign=TRUE,width=70, digits=2)
require(knitr)
opts_chunk$set(fig.path='figure/', cache.path='cache/', fig.align='center', fig.width=5, fig.height=5, par=TRUE, cache=TRUE, concordance=TRUE, autodep=TRUE, message=F, warning=F)

library(dplyr)
@

\section{Introduction} 
Graphics are an important component of big data analysis, providing a mechanism for discovering unexpected patterns in data. Pioneering research by \citet{gelman:2004}, \citet{buja:2009} and \citet{majumder:2011} provide methods to quantify the significance of discoveries made from visualizations. 
\citet{buja:2009} introduced two protocols, the Rorschach and the lineup protocol, which bridge the gulf between traditional statistical inference and exploratory data analysis. The Rorschach protocol consists of a set of $m$ (usually, $m=20$) plots (called the {\it null plots}) rendered from data that is consistent with a given null model. That way, the Rorschach protocol helps to understand the extent of randomness in the null model. Under the lineup protocol, a plot of the observed data is placed randomly among a set of $m-1$ null plots.
Human observers are then asked to  examine the lineup and to identify the most different plot. If observers identify the data plot, this is quantifiable evidence against the null hypothesis. 
The lineup protocol places a statistical plot firmly in the framework of hypothesis tests: a plot of the data is considered to be the test statistic, which is compared against the sampling distribution under the null hypothesis represented by the null plots. 
Obviously, the null generating mechanism, i.e.\ the method of obtaining the data for null plots, is crucial for both the lineup and the Rorschach protocol. 
The null hypothesis directly affects the choice of null generating method. 
Null generating methods are typically based on (a) simulation, if the null hypothesis allows us to directly specify a parametric model, (b) sampling, as for example in the case of large data sets, or (c) permutation of the original data \citep[see e.g.\ ][]{Good05}, which allows for non-parametric testing  that preserves marginal distributions  while ensuring independence in higher dimensions. 
%In the experimental data that we analyzed the null generating methods used were permutation methods and direct simulation from a null model.

The lineup protocol was formally tested in a head-to-head comparison with the equivalent conventional test in \citet{majumder:2011}. The experiment utilized human subjects from Amazon's Mechanical Turk \citep{turk} and used simulation to control conditions. The results suggest that  visual inference is comparable to conventional tests in a controlled conventional setting. This provides support for its appropriateness for testing in real exploratory situations where no conventional test exists. Interestingly, the power of a visual test increases with the number of observers engaged to evaluate lineups, and the pattern in results suggests that the power will provide results consistent with practical significance \citep{kirk:1996}.

\section{The SAO model}

\bibliographystyle{asa}
\bibliography{references}

\end{document}

### 2016-06-27 paper.Rnw
rough outline

@@
 
+\section{Visualizations}
+
+\hh{Some of the visualizations we talked about
+\begin{enumerate}
+\item visualization of the MCMC process:  movie of individual actors' step-by-step decisions; could be done with the ndtv package, as long as we can get individual steps out of RSiena
+\item global picture: 1000 simulated networks + alpha blending should give a pretty good idea of the most important edges
+\item global structure + principal components (tensor pca)
+
+\end{enumerate}}
 \bibliographystyle{asa}

### 2016-07-01 paper.Rnw

@@
 
+\hh{mostly lit review plus a detailed description of the model and fitting process.}
+
+\section{SAO Diagnostics}
+
+\hh{this needs a lit review of the methods used for diagnosing SAO models: you can start with describing the goodness-of-fit stats (`sienaGOF` function) and the meta analysis described in the RSiena manual, and then work on extending the literature cited and built on this.
+}
+
 \section{Visualizations}
+\hh{The visualizations should address most elements of structure described before, i.e.
+follow along in the setup (for descriptive and teaching purposes), the fitting (understanding the model and interpreting the results) and then diagnostics (how well does the model fit the data - where are the most differences to the actual data/what are the sensitive parameters?) }
 
@@
 \item global structure + principal components (tensor pca)
+\end{enumerate}}
 
-\end{enumerate}}
 \bibliographystyle{asa}

### 2017-01-18 VisualizingSAOMsOutline.Rmd
oultine

---
title: "Visualizing the CTMC and SAOM fitting process"
output: github_document
---

This outline exists to help me get my thoughts straight with respect to my process of visualizing the SAOM fitting process and the underlying CTMC for network change. Why am I doing this? What am I going to do? What have I already done? What do I want to accomplish?

# Outline

1. Visualize the CTMC
    a. Why? - to better understand the model. to see what's going on underneath. to see how the objective function affects tie changes.
    b. How? - write a function to simulate from time 1 to time 2 given parameter values that stores the network at each step, the time between each step, and any covariate values. visualize the one-tie-at-a-time changes in a movie, facets, any other way. use color or shape changes to see how the covariate levels affect the changes.
    c. What? - this will become part of the "removing the blindfold" chapter. introduction and first sections
2. Visualize the fitting process
    a. Why? - to better understand the updating process. to see what's influencing the parameter estimates most. 
    b. How? - write a function that mimics the `RSiena` fitting process for Method of Moments OR figure out how to extract the parameter updates from the `sienafit` object. visualize the chains of parameter changes. look at how different parameter values for starting values affect the estimations.
    c. What? - this will become part of the "removing the blindfold" chapter. Maybe model-in-data-space section.

### 2017-01-18 playingwithndtv.R

@@ -1,0 +1,35 @@
+# playing around with ndtv
+library(ndtv)
+
+wheel <- network.initialize(10) # create a toy network
+add.edges.active(wheel,tail=1:9,head=c(2:9,1),onset=1:9, terminus=11)
+add.edges.active(wheel,tail=10,head=c(1:9),onset=10, terminus=12)
+plot(wheel) # peek at the static version
+render.animation(wheel) # compute and render
+ani.replay()
+
+library(tergm)
+data("florentine") # an example network
+plot(flobusiness,displaylabels=TRUE)
+theta.diss <- log(9)
+stergm.fit.1 <- stergm(flobusiness,
+                       formation= ~edges+gwesp(0,fixed=TRUE),
+                       dissolution = ~offset(edges),
+                       targets="formation",
+                       offset.coef.diss = theta.diss,
+                       estimate = "EGMME")
+stergm.sim.1 <- simulate.stergm(stergm.fit.1,
+                                nsim=1, time.slices = 100)
+slice.par<-list(start=75,end=100,interval=1,
+                aggregate.dur=1,rule="latest")
+compute.animation(stergm.sim.1,slice.par=slice.par,
+                  animation.mode="MDSJ")
+render.par=list(tween.frames=5,show.time=TRUE,
+                how.stats="~edges+gwesp(0,fixed=TRUE)")
+render.animation(stergm.sim.1,render.par=render.par,
+                 edge.col="darkgray",displaylabels=TRUE,
+                 label.cex=.6,label.col="blue")
+render.d3movie(stergm.sim.1,render.par=render.par,
+               edge.col="darkgray",displaylabels=TRUE,
+               label.cex=.6,label.col="blue",
+               output.mode = "htmlWidget")

### 2017-01-23 VisualizingSAOMsOutline.Rmd
ideas for ch 1

@@
     c. What? - this will become part of the "removing the blindfold" chapter. Maybe model-in-data-space section.
+
+# Ideas from "Visualizing Statistical Models: Removing the Blindfold"
+
+1. Introduction
+    a. Talk about the toolbox of visualizations. We must answer the question, "What is in the network visualization toolbox?" and use those tools and more moving forward.
+2. Background 
+    a. Model family - the SAOM is the model family
+    b. Model form - which $\beta$ effects get included
+    c. Fitted model - estimates of rate parameters and $\beta$s resulting from `RSiena` fit
+    d. Question: Can we find a direct interpretation of the SAOM $\beta$ parameters? 
+    e. Question: How can we visualize these networks interactively?? 
+3. Display the Model in Data-Space
+    a. What does visualizing the model in the data space mean for SAOMs? 
+        i.
+Idea: take the `sims` output object from the SAOM model fit.
+Create adjaceny matrices from them, tile them in `geom_tile`, shading by frequency of specific ties in simulations, then overlay the squares that are in the data that was being simulated.
+        ii. Idea: focus on the probabilities of individual tie changes. How do close probabilities affect overall model behavior?? 
+4. Collections are more informative than singletons
+    a. We've already done some of this - densities of fitted estimates from `RSiena`. 
+    b. Look at distributions of fitted network statistics (corresponding to stats in model form) by the fitted parameter values. I also think we've already done some of this. 
+    c. Other ideas????
+5. Do not just look at the final result; explore how the algorithm works
+    a. Idea: We've already discussed visualizing the microsteps from current time to the next time point.
+    b. Idea: Look at the chains for updating the parameter estimates -- use multiple starting values.

### 2017-01-24 iterationsfromRSiena.Rmd
iterations from siena07. kind of nightmarish.

---
title: "Attempting to pull iterations out of RSiena"
author: "Sam Tyner"
date: "1/23/2017"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

First I set up some simple examples from the `RSiena` website.

```{r examplecode}
library(RSiena)
# ---- A. ----------------------------------------------------------------------
# First we have to create objects for the dependent variables.

# sienaDependent creates a sienaDependent object, here a network,
# from a matrix or array or list of sparse matrix of triples.
# The name of this network object (here: friendship) will be used
# in the output file.
 setwd("data/s50_data")
  friend.data.w1 <- as.matrix(read.table("s50-network1.dat"))
  friend.data.w2 <- as.matrix(read.table("s50-network2.dat"))
  friend.data.w3 <- as.matrix(read.table("s50-network3.dat"))
  drink <- as.matrix(read.table("s50-alcohol.dat"))
  smoke <- as.matrix(read.table("s50-smoke.dat"))

 friendship <- sienaDependent(
                     array( c( friend.data.w1, friend.data.w2, friend.data.w3 ),
                     dim = c( 50, 50, 3 ) ) )

# The integers in the dim() here refer to the number of nodes (senders,
# receivers) and the number of waves.
# This object is an array of dimension 50 x 50 x 3, representing
# three adjacency matrices, with a number of attributes.
# Note that this is an object of class

        class(friendship)

# with specific attributes and methods associated with it.
# You can get the detailed information by requesting

        dim( friendship )
        attributes( friendship )

# If you only are interested in the value of one particular attribute,
# you can request this by, e.g.,

        attributes( friendship )$type

# A very concise description of the friendship data is obtained by typing

        friendship

# The function sienaDependent can also be used to create a behavior variable object
# with the extra argument type = "behavior".
# (Non-mentioned attributes get the default value, and in this case
# oneMode is the default; see below.)
# (Note: only use the variable in ONE role in a given model:
#  behavior variable or changing covariate!)

# ---- B. ----------------------------------------------------------------------
# Second we construct objects for the explanatory (independent) variables.
# From the help request
#       ?sienaDataCreate
# we see that these can be of five kinds:
    # coCovar            Constant actor covariates
    # varCovar           Time-varying actor covariates
    # coDyadCovar        Constant dyadic covariates
    # varDyadCovar       Time-varying dyadic covariates
    # compositionChange  Composition change indicators

# You can get help about this by the following requests:
#       ?coCovar
#       ?varCovar
#       ?coDyadCovar
#       ?varDyadCovar
#       ?sienaCompositionChange

# The variables available for this data set all are changing actor covariates.
# For illustrative purposes, we use smoking as observed at the first wave
# as a constant covariate:

        smoke1 <- coCovar( smoke[ , 1 ] )

# We use the drinking data as a changing covariate.
# the name comes from 'varying covariate'.

        alcohol <- varCovar( drink )

# You need at least three waves in the data set to define a varying covariate
# by the function varCovar as the previous wave is used
# as a predictor of the next wave.

# The command

        attributes( alcohol )

# will tell you the information that RSiena now has added to the drink data.

# ---- C. ----------------------------------------------------------------------
# We now combine the dependent and independent variables.
# The function sienaDataCreate creates a Siena data object from input networks,
# covariates and composition change objects;
# the objects that earlier were created by sienaDependent will have the role
# of dependent variables, and similarly the other roles are predetermined
# by creation by the functions coCovar, varCovar,
# coDyadCovar, varDyadCovar, and sienaCompositionChange.

        mydata <- sienaDataCreate( friendship, smoke1, alcohol)
# You may check the result by requesting

        mydata
# ---- D. ----------------------------------------------------------------------
########################################################################## ---- DEFINING EFFECTS 
# The data set as combined in mydata implies a certain set of effects
# that can be included in the specification of the model.
# The function getEffects creates a dataframe of effects with a number of extra
# properties for use in RSiena:

        myeffnull <- getEffects( mydata )

# Let us now consider the myeff object, which is used to specify the model.
# add the transitive triples and 3-cycles effects

        myeff1addl <- includeEffects( myeffnull, transTrip )
        myeff2addl <- includeEffects( myeffnull, transTrip, cycle3 )

# fit the models to the data. 
  myalgorithm <- sienaAlgorithmCreate(projname = 's50_3')

	ansnull <- siena07( myalgorithm, data = mydata, effects = myeffnull,
	                    returnChains = TRUE, returnDataFrame = TRUE,
	                    returnDeps = TRUE, silent = TRUE, verbose = FALSE)
  ans1addl <- siena07( myalgorithm, data = mydata, effects = myeff1addl,
                       returnChains = TRUE, returnDataFrame = TRUE,
                       returnDeps = TRUE,silent = TRUE, verbose = FALSE)
  ans2addl <- siena07( myalgorithm, data = mydata, effects = myeff2addl,
                       returnChains = TRUE, returnDataFrame = TRUE,
                       returnDeps = TRUE,silent = TRUE, verbose = FALSE)

# Function siena07 produces a so-called sienaFit object, here called ans;
# and it fills in a few things in the sienaEffects object myeff,
```
Print out the results from these models for future reference: 
```{r printres}
ansnull
ans1addl
ans2addl
```

Okay, now let's look at the chains. These (supposedly) contain the microsteps from the simulation process.

```{r chains}
class(ansnull$chain)
length(ansnull$chain)
# first question: why is the length of ansnull$chain 1000? 
class(ansnull$chain[[1]])
length(ansnull$chain[[1]])
length(ansnull$chain[[1]][[1]])
length(ansnull$chain[[1]][[1]][[1]])
length(ansnull$chain[[1]][[1]][[2]])
# second question: why do these have different lengths? 
# hypothesis: these are the steps of the CTMC (microsteps), so each sim will have a different number of steps. 
class(ansnull$chain[[1]][[1]][[1]])
length(ansnull$chain[[1]][[1]][[1]][[1]])
# ok. this is the lowest level. is it the same length for the others?
length(ans1addl$chain[[1]][[1]][[1]][[1]])
length(ans2addl$chain[[1]][[1]][[1]][[1]])
# yes. Crap. so it probably doesn't contain any parameter information...
```

For the ans object, the length of `ans$chain[[1]][[1]][[m]][[n]]` is always 13 for any value of `m` and `n`. What are in these fields? Class and value are printed below.

```{r fields}
for (i in 1:13){
  print(paste(class(ansnull$chain[[1]][[1]][[1]][[1]][[i]]),
              ansnull$chain[[1]][[1]][[1]][[1]][[i]]))
}
```

My current guesses for the fields:

1. The class of the dependent variable (?)
2. No idea
3. The name of the dependent variable
4. "from" node id of microstep
5. "to" node id of microstep
6. No idea
7. $a_n$ value at parameter update (??)

for 8 - 13 I also have no idea.

When I was printing out the chains in the console, I noticed some attributes printing off. Let's see those in action.

```{r getattrs}
attributes(ansnull$chain[[1]][[1]][[1]])
```

Okay, we've got a mean (`mu`), a variance (`sigma2`), a final rate of some kind (`finalReciprocalRate`) and two empy fields, `initialStateDifferences` and `endStateDifferences`. Recall length of `ansnull$chain[[1]][[1]][[1]]` is `r length(ansnull$chain[[1]][[1]][[1]])` so these attributes exist for all microsteps in the process. That's informative-ish. Here's another set of attributes.

```{r more attrs}
attributes(ansnull$chain[[1]][[1]][[2]])
```

Only the first two attributes changed. Update guesses about what the 13 entries are:

1. The class of the dependent variable (?)
2. No idea
3. The name of the dependent variable
4. "from" node id of microstep
5. "to" node id of microstep
6. No idea
7. ~~$a_n$ value at parameter update (??)~~ `finalReciprocalRate` attribute, whatever that is.
8. No idea
9. No idea
10. The `initialStateDifferences` attribute value (?)
11. The `endStateDifferences` attribute value (?)
12. No idea
13. No idea

I'll come back to the attributes later. Next, get all of the microstep data together.
```{r collectsteps, cache=TRUE}
get_chain_info <- function(ans){

  L <- length(ans$chain)
  M <- length(ans$chain[[1]][[1]])
  bigres <- NULL
  res <- data.frame()
  for (l in 1:L){
    res <- data.frame()
    for (m in 1:M){
      sub <- data.frame(plyr::ldply(
          lapply(ansnull$chain[[l]][[1]][[m]], unlist),
        rbind), stringsAsFactors = FALSE)
      sub$rep <- l
      sub$period <- m
      res <- rbind(res, sub)
    }
    bigres[[l]] <- res  
  }
  ret <- plyr::rbind.fill(bigres)
  return(ret)
}

ansnullchains <- get_chain_info(ansnull)
```

Now that we have all these chains together, let's look at them:

```{r exploredf}
head(ansnullchains)

for (i in 1:ncol(ansnullchains)){
  print(paste(i, length(unique(ansnullchains[,i]))))
}
```

Okay, 7 of 13 columns have only 1 value. That sucks. What are they?

```{r uniques}
unique(ansnullchains[,1])
unique(ansnullchains[,2])
unique(ansnullchains[,3])
unique(ansnullchains[,6])
unique(ansnullchains[,7])
unique(ansnullchains[,8])
unique(ansnullchains[,10])
```

This has been exceedingly unhelpful. Remove the columns that only have1 value to save memory.

```{r delete}
ansnullchains <- ansnullchains[,-c(1:3, 6:8, 10)]
str(ansnullchains)
# annoyingly, a lot are factors... fix it real quick.
for (i in 1:3){
  ansnullchains[,i] <- as.numeric(as.character(ansnullchains[,i]))
}
ansnullchains[,4] <- as.logical(as.character(ansnullchains[,4]))
```

Look at the distributions of these things. 
```{r plots}
library(ggplot2)
ggplot(data = ansnullchains) + 
  geom_histogram(aes(x = X9), binwidth = .25) + 
  facet_wrap(~period) + 
  labs(title = "Period doesn't appear to make a difference in this value")
```
```{r plots2}
ggplot(data = ansnullchains) + 
  geom_histogram(aes(x = X9), binwidth = .25) + 
  facet_wrap(~X11) + 
  labs(title = "X11 makes a big difference in this value")
```
Is there a relationship between `X4`, `X5`, `X9`?
```{r plots3}
ggplot(data = ansnullchains) + 
  geom_boxplot(aes(x = X4, y = X9, group = X4)) + 
  labs(title = "Some variation in median across X4")
```
```{r plots4}
ggplot(data = ansnullchains) + 
  geom_boxplot(aes(x = X5, y = X9, group = X5)) + 
  labs(title = "Some variation in median across X5")
```
This next plot might be interesting?? There seem to me to be WAY too many on the diagonal. These are definitely the nodes in the microstep. I'm like 99% sure.  
```{r plots5}
ggplot(data = ansnullchains) + 
  geom_raster(aes(x = X4, y = X5, fill = log(..count..)), stat = "bin2d") + 
  facet_wrap(~period, labeller = 'label_both') + 
  scale_fill_continuous(low = 'white', high = 'blue') + 
  theme(aspect.ratio = 1)
```

Okay, going back to the attributes:

```{r attrs}
get_attr_info <- function(ans){

  L <- length(ans$chain)
  M <- length(ans$chain[[1]][[1]])
  bigres <- NULL
  res <- data.frame(stringsAsFactors = FALSE)
  for (l in 1:L){
    res <- data.frame(stringsAsFactors = FALSE)
    for (m in 1:M){
      sub <- data.frame(t(unlist(attributes(ans$chain[[l]][[1]][[m]]))), stringsAsFactors = FALSE)
      sub$rep <- l 
      sub$period <- m 
      res <- rbind(res, sub)
    } 
    bigres[[l]] <- res  
  }
  ret <- plyr::rbind.fill(bigres)
  return(ret)
}

ansnullattrs <- get_attr_info(ansnull) 
summary(ansnullattrs)
```

### 2017-01-25 iterationsfromRSiena.Rmd
more nightmarish unpacking

@@
 ```
+
+Plots of the attribute data:
+
+```{r plotsattr}
+ggplot(ansnullattrs) + 
+  geom_histogram(aes(x = mu), binwidth = .25)
+ggplot(ansnullattrs) + 
+  geom_histogram(aes(x = sigma2))
+ggplot(ansnullattrs) + 
+  geom_point(aes(x= mu, y = sigma2))
+cor(ansnullattrs$mu, ansnullattrs$sigma2)
+```
+
+Okay, `mu` and `sigma2` are perfectly correlated. Does this make sense??? Are they monotonic maybe? Check:
+
+```{r plotattrs2}
+ggplot(ansnullattrs) + 
+  geom_line(aes(x= rep, y = mu, color = as.factor(period)))
+ggplot(ansnullattrs) + 
+  geom_line(aes(x= rep, y = mu)) + 
+  facet_wrap(~period, nrow = 2, labeller = "label_both")
+```
+
+Okay, couple things:
+
+1. period 2 values of `mu` are pretty consistently lower than period 1 values of `mu`
+2. they're definitely no monotonic with rep, and they look like typical chains from an MCMC.
+
+I assume the `sigma2` plots have the same appearance since `mu` and `sigma` are perfectly correlated.

### 2017-01-28 iterationsfromRSiena.Rmd
can animate some microsteps

@@
 
-```{r examplecode}
+```{r examplecode, cache=TRUE}
 library(RSiena)
@@
 
-```{r delete}
+```{r delete, cache=TRUE}
 ansnullchains <- ansnullchains[,-c(1:3, 6:8, 10)]
@@
 
-```{r attrs}
+```{r attrs, cache=TRUE}
 get_attr_info <- function(ans){
@@
 I assume the `sigma2` plots have the same appearance since `mu` and `sigma` are perfectly correlated.
+
+What are you, `mu` and `sigma2`? Look at the [Siena Algorithms](http://www.stats.ox.ac.uk/~snijders/siena/Siena_algorithms.pdf)!
+
+On p. 32 of this manual, there is a defining list of variables that might be our mysterious 13 list elements!!!!
+
+1. "Aspect": 'network' or 'behavior' 
+2. No idea
+3. The name of the dependent variable
+4. "Actor i": sending actor
+5. "Actor j": receiving actor
+6. "Difference": meaningless 0 when aspect is 'network'
+7. ~~$a_n$ value at parameter update (??)~~ `finalReciprocalRate` attribute, whatever that is.
+8.
+"log OptionSet probability": log probability of making a ministep of this OptionSet.
+Okay.
+This value is `r ansnull$chain[[1]][[1]][[1]][[1]][[8]]` always.
+Now we know it's been $\log$'d.
+So I'll $\exp$ it: `r exp(ansnull$chain[[1]][[1]][[1]][[1]][[8]])`.
+This is the probabililty of randomly selecting one node out of the 50 to be changed.
+CRUSHED IT.
+9.
+"log choice probability": log probability of making a ministep of this choice.
+So, if we $\exp$ it, that's the probability of changing the tie from node $i$ to node $j$ given the current network stats.
+10. The `initialStateDifferences` attribute value (?)
+11. The `endStateDifferences` attribute value (?)
+12. No idea
+13. No idea
+
+Okay, now that I know that the `X9` variable is the log of the probability of making the tie, let's look at some plots of it when it's been exponentiated to get the probability back out.
+
+```{r newplots}
+names(ansnullchains)[1:3] <- c("actori", "actorj", "logchoiceprob")
+ggplot(data= ansnullchains) + 
+  geom_histogram(aes(x = exp(logchoiceprob)), binwidth = 0.025)
+```
+
+Now, let's color our tile plot from before by the mean probability of that tie (first requires some data manipulation):
+
+```{r newplots2}
+library(dplyr)
+ansnullchains %>% 
+  group_by(actori, actorj) %>% 
+  summarize(meanprob = mean(exp(logchoiceprob))) %>% 
+  arrange(actori, actorj, meanprob) -> ansnull_probs
+
+ggplot(data = ansnull_probs) + 
+  geom_raster(aes(x = actori, y = actorj, fill = meanprob)) + 
+  #facet_wrap(~period, labeller = 'label_both') + 
+  scale_fill_continuous(low = 'white', high = 'blue') + 
+  theme(aspect.ratio = 1)
+```
+
+Cool. That's much more informative and logical.

### 2017-01-30 visualizemicrosteps.Rmd
code to make microstep plots

---
title: "Visualizing `RSiena`'s Microsteps"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
```

## Fit a model to data

I'll start like I always do: with the friends data. This code is directly from the [`RSiena` webpage](https://www.stats.ox.ac.uk/~snijders/siena/Rscript02SienaVariableFormat.R).

```{r get_data, cache=TRUE}
library(RSiena)
setwd("data/s50_data")
friend.data.w1 <- as.matrix(read.table("s50-network1.dat"))
friend.data.w2 <- as.matrix(read.table("s50-network2.dat"))
friend.data.w3 <- as.matrix(read.table("s50-network3.dat"))
drink <- as.matrix(read.table("s50-alcohol.dat"))
smoke <- as.matrix(read.table("s50-smoke.dat"))
friendship <- sienaDependent(
                     array(c(friend.data.w1, friend.data.w2,
                             friend.data.w3),
                     dim = c(50, 50, 3)))
smoke1 <- coCovar(smoke[ , 1 ])
alcohol <- varCovar(drink)
mydata <- sienaDataCreate(friendship, smoke1, alcohol)
myeffnull <- getEffects(mydata)
myeffalt1 <- includeEffects(myeffnull, transTrip)
myalgorithm <- sienaAlgorithmCreate(projname = 's50_3')
set.seed(823746)
ansnull <- siena07(myalgorithm, data = mydata, effects = myeffnull,
	                 returnChains = TRUE, returnDataFrame = TRUE,
	                 returnDeps = TRUE, silent = TRUE, verbose = FALSE,
	                 batch = TRUE)
ansalt1 <- siena07(myalgorithm, data = mydata, effects = myeffalt1,
                    returnChains = TRUE, returnDataFrame = TRUE,
                    returnDeps = TRUE,silent = TRUE, verbose = FALSE,
                    batch = TRUE)
```

Look at a plot of the data:

```{r plot}
library(geomnet)
w1 <- fortify(as.adjmat(friend.data.w1))
w2 <- fortify(as.adjmat(friend.data.w2))
w3 <- fortify(as.adjmat(friend.data.w3))
id <- na.omit(unique(c(w1$from, w1$to, w2$from, w2$to, w3$from, w3$to)))
nodes <- data.frame(id = rep(id[order(parse_number(id))], 3),
                    wave = rep(1:3, each = length(id)),
                    drink = c(drink[-c(13,20),1],drink[-c(13,20),2],
                              drink[-c(13,20),3]), 
                    smoke = rep(smoke[-c(13,20),1],3),
                    stringsAsFactors = F)
w1$wave <- 1
w2$wave <- 2
w3$wave <- 3
fnet <- rbind(w1,w2,w3)
fnet <- merge(fnet, nodes, by.x = c('from', 'wave'),
              by.y = c('id',"wave"), all = T)
# need to change the id to be 0-49 to match the chains coding
# library(dplyr)
# fnet %>% 
 # mutate(from = parse_number(from) - 1, 
  #       to = parse_number(to) - 1) -> fnet
library(readr)

ggplot(data = fnet) + 
  geom_net(aes(from_id = from, to_id = to), fiteach = T) + 
  facet_wrap(~wave, labeller = 'label_both', nrow = 1)
```

What do the fitted models look like?

```{r fits}
summary(ansnull)
summary(ansalt1)
```

Convergence looks good in both cases!

Now, I want to visualize the individual changes, starting with wave 1 to wave 2 from the null model.

```{r getchains, cache=TRUE}
get_chain_info <- function(ans){
  L <- length(ans$chain)
  M <- length(ans$chain[[1]][[1]])
  bigres <- NULL
  res <- data.frame()
  for (l in 1:L){
    res <- data.frame()
    for (m in 1:M){
      sub <- data.frame(plyr::ldply(
          lapply(ansnull$chain[[l]][[1]][[m]], unlist),
        rbind), stringsAsFactors = FALSE)
      sub$rep <- l
      sub$period <- m
      res <- rbind(res, sub)
    }
    bigres[[l]] <- res  
  }
  ret <- plyr::rbind.fill(bigres)
  return(ret)
}
ansnullchains <- get_chain_info(ansnull)
```

```{r firstwave}
library(dplyr)
library(readr)
ansnullchains %>% 
  filter(period == 1) %>% 
  group_by(rep) %>%
  select(rep, from = X4, to = X5) %>% 
  mutate(val = as.numeric(!from == to),
         from = paste0("V", parse_number(from)+1), # make the chains
         to = paste0("V", parse_number(to)+1) # match the data vars 
         ) -> ansnullchainsw1w2

addMicrostep <- function(dat, newedge){
  N <- nrow(dat)
  selfie <- newedge$from == newedge$to
  check <- dplyr::filter(dat, from == newedge$from & to == newedge$to)
  if (nrow(check) == 0 && !selfie){
    dat2 <- tibble::add_row(dat, from = newedge$from, to = newedge$to)
    dat2$col <- FALSE
    dat2$col[N+1] <- TRUE # color the new edge. 
  } else if (nrow(check) == 1 && !selfie){
    dat2 <- dplyr::filter(dat, !(from == newedge$from & to == newedge$to))
    dat2$col <- FALSE
  } else{
    dat2 <- dat
    dat2$col <- FALSE
  }
  return(dat2)
}

listMicrosteps <- function(dat, microsteps){
  iters <- NULL
  iters[[1]] <- dat
  B <- nrow(microsteps)
  for (i in 2:(B+1)){
    iters[[i]] <- addMicrostep(iters[[(i-1)]], microsteps[(i-1),])
  }
  return(iters)
}

# test it on one of the reps from the null model

ms1 <- listMicrosteps(dat = w1, 
                      microsteps = filter(ansnullchainsw1w2, rep == 1))
ms1[[200]]
# looks legit
```

Now, I want to plot all of them! Make a movie!

```{r movie}
# layout the original wave
# get (x,y) coordinates to add to the edge lists
# merge nodes and microsteps before plotting.

library(sna)
w1.net <- as.network(friend.data.w1)
layoutxy <- gplot.layout.fruchtermanreingold(friend.data.w1, layout.par = NULL)
layoutxy <- data.frame(layoutxy)
names(layoutxy) <- c("x", "y")
layoutxy$id <- get.vertex.attribute(w1.net, "vertex.names")

w1_nodes_xy <- merge(nodes %>% filter(wave == 1), layoutxy, by = "id", all = T)

plotms <- function(dat, nodes){
  dat2 <- merge(dat, nodes, by.x = "from", by.y= "id", all=T)
  p <- ggplot(data = dat2) + 
    geom_net(aes(from_id = from, to_id = to, x = x, y = y),
             layout.alg = NULL, size = 1.5, directed = T, linewidth = .5,
             arrow = arrow(angle = 20, 
                           length = unit(5, "pt"), type = "open"),
             arrowgap = .02) + #, ecolour = c("grey40", "red")[as.factor(dat2$col)]) + 
    theme_net()
  return(p)
}

setwd("plots")
for (i in 2:length(ms1)) {
  # creating a name for each plot file with leading zeros
  print(i)
  if (i < 10) {name = paste('plots/000',i,'plot.png',sep='')}
  if (i < 100 && i >= 10) {name = paste('plots/00',i,'plot.png', sep='')}
  if (i >= 100) {name = paste('plots/0', i,'plot.png', sep='')}
  png(name)
  print(plotms(ms1[[i]], nodes = w1_nodes_xy))
  dev.off()
} 
make.mov <- function(){
     unlink("plot.mpg")
     system("convert -delay 0.5 plot*.jpg plot.mpg")
}
make.mov()
```

### 2017-02-08 paper.Rnw
start up the paper.Rnw doc again.

@@
 \section{Introduction}
-Graphics are an important component of big data analysis, providing a mechanism for discovering unexpected patterns in data.
-Pioneering research by \citet{gelman:2004}, \citet{buja:2009} and \citet{majumder:2011} provide methods to quantify the significance of discoveries made from visualizations.
-\citet{buja:2009} introduced two protocols, the Rorschach and the lineup protocol, which bridge the gulf between traditional statistical inference and exploratory data analysis.
-The Rorschach protocol consists of a set of $m$ (usually, $m=20$) plots (called the {\it null plots}) rendered from data that is consistent with a given null model.
-That way, the Rorschach protocol helps to understand the extent of randomness in the null model.
-Under the lineup protocol, a plot of the observed data is placed randomly among a set of $m-1$ null plots.
-Human observers are then asked to  examine the lineup and to identify the most different plot. If observers identify the data plot, this is quantifiable evidence against the null hypothesis. 
-The lineup protocol places a statistical plot firmly in the framework of hypothesis tests: a plot of the data is considered to be the test statistic, which is compared against the sampling distribution under the null hypothesis represented by the null plots.
-Obviously, the null generating mechanism, i.e.\ the method of obtaining the data for null plots, is crucial for both the lineup and the Rorschach protocol. 
-The null hypothesis directly affects the choice of null generating method. 
-Null generating methods are typically based on (a) simulation, if the null hypothesis allows us to directly specify a parametric model, (b) sampling, as for example in the case of large data sets, or (c) permutation of the original data \citep[see e.g.\ ][]{Good05}, which allows for non-parametric testing  that preserves marginal distributions  while ensuring independence in higher dimensions.
-%In the experimental data that we analyzed the null generating methods used were permutation methods and direct simulation from a null model.
 
-The lineup protocol was formally tested in a head-to-head comparison with the equivalent conventional test in \citet{majumder:2011}.
-The experiment utilized human subjects from Amazon's Mechanical Turk \citep{turk} and used simulation to control conditions.
-The results suggest that  visual inference is comparable to conventional tests in a controlled conventional setting.
-This provides support for its appropriateness for testing in real exploratory situations where no conventional test exists.
-Interestingly, the power of a visual test increases with the number of observers engaged to evaluate lineups, and the pattern in results suggests that the power will provide results consistent with practical significance \citep{kirk:1996}.
+The guiding principles of model visualization are (1) display the model in the data space, (2) collections of models are more informative than singletons, and (3) explore how algorithms work, instead of just looking at the final result.
+Stochastic actor-oriented models are a prime example of a set of models that could benefit greatly from application of model visualization.
+First, the models themselves include a continuous-time Markov chain (CTMC) that is completely hidden in the model fitting process.
+Bringing the CTMC out from behind-the-scenes and into the light of visualization will provide researchers with insights into the underlying features of the model.
+Secondly, SAOMs allow for a great deal of parameters that can be added to the objective function, each of which are attached to a network statistic.
+These statistics are often somewhat, if not highly, correlated, and so are the associated parameters.
+By visualizing collections of SAOMs, we can gain a better understanding of these correlations and find ways to deal with them and rectify their effects.
 
@@
 \hh{mostly lit review plus a detailed description of the model and fitting process.}
-
-\section{SAO Diagnostics}
-
-\hh{this needs a lit review of the methods used for diagnosing SAO models: you can start with describing the goodness-of-fit stats (`sienaGOF` function) and the meta analysis described in the RSiena manual, and then work on extending the literature cited and built on this.
-}
 

### 2017-02-08 visualizemicrosteps.Rmd
start up the paper.Rnw doc again.

@@
 
-plotms <- function(dat, nodes){
+plotms <- function(dat, nodes, same.layout){
   dat2 <- merge(dat, nodes, by.x = "from", by.y= "id", all=T)
+  if (same.layout){
   p <- ggplot(data = dat2) + 
@@
     theme_net() 
+  } else 
+  p <- ggplot(data = dat2) + 
+    geom_net(aes(from_id = from, to_id = to),
+            size = 1.5, directed = T, linewidth = .5,
+             arrow = arrow(angle = 20, 
+                           length = unit(5, "pt"), type = "open"),
+             arrowgap = .02) + #, ecolour = c("grey40", "red")[as.factor(dat2$col)]) + 
+    theme_net()
   return(p)
@@
 
-setwd("plots")
-for (i in 2:length(ms1)) {
+make_msplots <- function(ms, nodes, same.layout, dir, filename){
+  N <- length(ms)
+  for (i in 1:length(ms1)) {
   # creating a name for each plot file with leading zeros
   print(i)
-  if (i < 10) {name = paste('plots/000',i,'plot.png',sep='')}
-  if (i < 100 && i >= 10) {name = paste('plots/00',i,'plot.png', sep='')}
-  if (i >= 100) {name = paste('plots/0', i,'plot.png', sep='')}
+  if (i < 10) {name = paste(dir,'/000',i,filename,'.png',sep='')}
+  if (i < 100 && i >= 10) {name = paste(dir,'/00',i,filename,'.png', sep='')}
+  if (i >= 100) {name = paste(dir,'/0', i,filename, '.png', sep='')}
   png(name)
-  print(plotms(ms1[[i]], nodes = w1_nodes_xy))
+  print(plotms(ms[[i]], nodes = nodes, same.layout = same.layout) + 
+          labs(title = paste("Microstep:", i-1)))
   dev.off()
   } 
-make.mov <- function(){
-     unlink("plot.mpg")
-     system("convert -delay 0.5 plot*.jpg plot.mpg")
 }
-make.mov()
+
+make.gif <- function(dir, filename, out.name){
+    # unlink("plot.mpg")
+    command <- paste0("convert -delay 50 ", dir, "/*", filename, ".png ", dir, "/", out.name, ".gif")
+     system(command = command)
+}
 ```
+
+Now, allow the layout to change.
+
+```{r makedifflayout}
+make_msplots(ms = ms1, nodes = w1_nodes_xy, dir = "plots", filename = "w1difflayout", same.layout = FALSE)
+make.gif(dir = "plots", filename = "w1difflayout", out.name = "microsteps_difflo")
+
+```

### 2017-02-11 paper.Rnw
more intro/lit review material

@@
 \noindent%
-{\it Keywords:}  Visual inference, Lineup protocol, \hh{XXX Other keywords?}.
+{\it Keywords:} social network analysis, model visualization \hh{Other keywords?}
 \vfill
@@
 By visualizing collections of SAOMs, we can gain a better understanding of these correlations and find ways to deal with them and rectify their effects.
+Finally, estimation of the parameters in a SAOM relies on a Robins-Monro algorithm, and convergence checks rely on simulations from the fitted model.
+Again, in these steps are largely hidden from the researcher computing them with software.
+We hope that by bringing these background elements and others into the foreground, stochastic actor-oriented models will be better understood and more accessible to social network analysts.
 
@@
 
+The stochastic actor-oriented models are very aptly named: they model dynamic networks that can account for the individual node characteristics when making the edge changes.
+The object of analysis is a social network that has been observed at several discrete time points.
+The nodes in the network can have associated covariate variables that may be incorporated into modeling the tie changes in the network.
+The model itself has two primary pieces: the rate of change of ties, and the objective function of the nodes, which determines the exact change that gets made.
+
+We start with a set of network observations at $M$ discrete time points: $x(t_1), x(t_2) \dots, x(t_M)$.
+The first observation, $x(t_1)$ is always conditioned on.
+The number of nodes / actors / vertices is constant in all time points and is denoted $n$.
+The tie variables are directed and  binary, and are denoted at $x_{ij}(t_m)$ for time point $m$, and this is sometimes abbreviated to $x_{ij}$ in notation.
+If there is a tie from node $i$ to node $j$, then $x_{ij}=1$, and $x_{ij}=0$ otherwise.
+
 \hh{mostly lit review plus a detailed description of the model and fitting process.}
 
-\section{Visualizations}
+\section{Model Visualizations}
+
+\subsection{View the model in the data space}
+\subsection{Collections are more informative than singletons}
+\subsection{Explore algorithms, not just end result}
 \hh{The visualizations should address most elements of structure described before, i.e.

### 2017-02-20 paper.Rnw
change title
coulds and shoulds go out

@@
 {
-  \title{\bf Visual Diagnsotics of SAO Models}
+  \title{\bf Model Visualization Techniques for a Social Network Model}
   \author{Samantha Tyner\thanks{
@@
   \begin{center}
-    {\LARGE\bf Visual Diagnsotics of SAO Models}
+    {\LARGE\bf Model Visualization Techniques for a Social Network Model}
 \end{center}
@@
 
-The guiding principles of model visualization are (1) display the model in the data space, (2) collections of models are more informative than singletons, and (3) explore how algorithms work, instead of just looking at the final result.
-Stochastic actor-oriented models are a prime example of a set of models that could benefit greatly from application of model visualization.
+The guiding principles XXX cite XXX of model visualization are (1) display the model in the data space, (2) collections of models are more informative than singletons, and (3) explore how algorithms work, instead of just looking at the final result.
+Stochastic actor-oriented models are a prime example of a set of models that  benefit greatly from application of model visualization.
 First, the models themselves include a continuous-time Markov chain (CTMC) that is completely hidden in the model fitting process.
-Bringing the CTMC out from behind-the-scenes and into the light of visualization will provide researchers with insights into the underlying features of the model.
+Bringing the CTMC out from behind-the-scenes and into the light of visualization  provides researchers with insights into the underlying features of the model.
 Secondly, SAOMs allow for a great deal of parameters that can be added to the objective function, each of which are attached to a network statistic.
@@
 By visualizing collections of SAOMs, we can gain a better understanding of these correlations and find ways to deal with them and rectify their effects.
-Finally, estimation of the parameters in a SAOM relies on a Robins-Monro algorithm, and convergence checks rely on simulations from the fitted model.
+Finally, estimation of the parameters in a SAOM relies on a Robbins-Monro algorithm, and convergence checks rely on simulations from the fitted model.
 Again, in these steps are largely hidden from the researcher computing them with software.

### 2017-02-20 paper.Rnw
add more material for outlining the paper

--- 2017-02-11 paper.Rnw
@@
 \newcommand{\hh}[1]{{\color{magenta} #1}}
-\newcommand{\sct}[1]{{\color{orange} #1}}
+\newcommand{\st}[1]{{\color{orange} #1}}
 
@@
 
+\st{
+\begin{itemize}
+\item What? - model visualization and SAOMs. briefly dsecribe what these things are
+\item Why? - bring light to the RSiena blackbox. Need ways to visualize network models, collections of networks. 
+\item How?
+\end{itemize}
+also talk about other dynamic network models briefly to put SAOMs in context.}
+
 The guiding principles of model visualization are (1) display the model in the data space, (2) collections of models are more informative than singletons, and (3) explore how algorithms work, instead of just looking at the final result.
@@
 \section{The SAO model}
+
+\st{What do I need to communicate in this section? 
+\begin{itemize}
+\item What the models are for: dynamic social networks with actor-level covariates
+\item What are the model components? rate function, objective function (network statistics, actor covariates, interactions between covariates and network statistics), time points ($t_m = 1, \dots, M$)
+\item What is RSiena? What methods of estimation, convergence checks does it use? What are the algorithms? What kinds of models can it fit? 
+\end{itemize}}
 
@@
 
+\st{Introduce the concepts from Wickham et al. }
+
 \subsection{View the model in the data space}
+
+\st{
+\begin{itemize}
+\item WHat is the model? the network model (SAOM model is just one of many)
+\item What is the data? Edge data, node data, observed time points, unobserved timepoints (CTMC)
+\item Note: specifically dealing with dynamic networks observed at discrete time points. 
+\end{itemize}
+What are some things that could go in here? need more data examples; not just the friends data.
+Harry Potter data??? I want to find my own data, too.}
+
 \subsection{Collections are more informative than singletons}
+
+The second principle of Wickham et al states that ``collections are more informative than singletons" (p.
+210).
+They describe several different methods for producing such collections, but we focus on only a few here.
+We look at collections produced ``from exploring the space of all possible models," ``by varying model settings", ``by fitting the same model to different datasets", and ``by treating each iteration as a model." (p.
+210-11) We also fit the same model many times, resulting in distributions of fitted parameters for one data set.
+
+\st{What are some things that could go in here?:
+\begin{itemize}
+\item "average" networks -- you've already done this! Write it up, make several examples. How does the procedure for averaging generalize? What is it good for? 
+\item distributions of fitted parameters. again, you've already done this!!! just write it up and put it in here. 
+\item view correlations between model parameters under different model sets. eg 2 betas, 3 betas, 4 betas, etc.
+\end{itemize}
+ }
+
 \subsection{Explore algorithms, not just end result}
+\st{The algorithms here are many!
+\begin{enumerate}
+\item First, the microsteps / CTMC between each time point (gifs, other animations)
+\item Phases of RSiena: phase 1 = initial parameter estimate
+\item phase 2 = iterative update of parameters - trace plots
+\item phase 3 = simulations for convergence diagnostics
+\end{enumerate}
+ }
+
 \hh{The visualizations should address most elements of structure described before, i.e.
@@
 
+\section{}
+
 \bibliographystyle{asa}

### 2017-02-21 paper.Rnw
section 3.2

@@
 <<setup, echo = FALSE, message = FALSE, warning = FALSE>>=
-options(replace.assign=TRUE,width=70, digits=2)
+options(replace.assign=TRUE,width=70, digits=3)
 require(knitr)
-opts_chunk$set(fig.path='figure/', cache.path='cache/', fig.align='center', fig.width=5, fig.height=5, par=TRUE, cache=TRUE, concordance=TRUE, autodep=TRUE, message=F, warning=F)
+opts_chunk$set(fig.path='figure/', cache.path='cache/', fig.align='center', fig.width=5, fig.height=5, par=TRUE, cache=TRUE, concordance=TRUE, autodep=TRUE, message=F, warning=F, echo = FALSE)
 
@@
 
+A classical problem with statistical network models generally is the lack of an ``average network" measure.
+Statisticians frequently rely on averages and expected values, but statistical network models, especially those as complex as SAOMs, lack an intuitive expected value measure.
+How then, can we arrive at an ``average" network?
+We propose to answer this question through visualization.
+
+We first consider the 50 actor subset of the teenage friends and lifestyle data available on the \texttt{RSiena} website \citep{RSiena}.
+To this data, we fit a SAOM with a simple rate function and an objective function with two parameters, $f_i(x) = \beta_1s_{i1} + \beta_2s_{i2}$, where $s_{i1}$ is the density network statistic and $s_{i2}$ is the reciprocity network statistic for actor $i$.
+We fit this model using \texttt{RSiena} 1,000 times, then averaged the resulting parameter estimates to obtain mean estimates to use in the simulation of networks from this model.
+These values are shown in Table~\ref{tab:M1ests}
+
+<<getdata, results='asis'>>=
+library(netvizinf)
+data("M1ests_bigfriends")
+params <- data.frame(t(colMeans(M1ests_bigfriends)))
+names(params) <- c("Rate1", "Rate2", "beta1", "beta2")
+knitr::kable(params, digits = 3, align = 'c',caption = "\\label{tab:M1ests} Estimates of the four parameters from the simple SAOM fit to the 50 actor subset of the teenage friends and lifestyle data.", format = "latex")
+@
+
+To create a visualization that represents an ``average" network, we count occurences of each possible edge in the simulations, resulting in a summary network of simulations.
+Next, we subset the network to include
+
+<<averagenet>>=
+library(RSiena)
+library(dplyr)
+friend.data.w1 <- as.matrix(read.table("data/s50_data/s50-network1.dat"))
+friend.data.w2 <- as.matrix(read.table("data/s50_data/s50-network2.dat"))
+friend.data.w3 <- as.matrix(read.table("data/s50_data/s50-network3.dat"))
+drink <- as.matrix(read.table("data/s50_data/s50-alcohol.dat"))
+friendshipData <- array(c(friend.data.w1, friend.data.w2, friend.data.w3), dim = c(50, 50, 3))
+friendship <- sienaDependent(friendshipData)
+alcohol <- varCovar(drink)
+mydata <- sienaDataCreate(friendship, alcohol)
+M1eff <- getEffects(mydata)
+M1sims1000 <- saom_simulate(dat = mydata, struct = M1eff,
+                       parms = as.numeric(params), N = 1000)
+M1avg <- net_avg(M1sims1000, 1)
+@
+
 \st{What are some things that could go in here?:

### 2017-02-21 paper.Rnw
experimenting with layout of average network.

@@
 M1avg <- net_avg(M1sims1000, 1)
+library(geomnet)
+M1avg$from <- paste0("n", M1avg$from)
+M1avg$to <- paste0("n", M1avg$to)
+
+library(tidyr)
+M1avg[,-3] %>% spread(to, weight) %>% data.frame-> distmat
+rownames(distmat) <- distmat$from
+distmat <- distmat[,-1]
+distmat <- as.matrix(1/distmat)
+distmat[is.na(distmat)] <- 0
+ggplot(data = M1avg %>% filter(weight >= 51)) + 
+  geom_net(aes(from_id = from, to_id = to, linewidth = weight/500), 
+           #directed = TRUE,
+           layout.alg = "kamadakawai", 
+           layout.par = list(elen = distmat))
+ggplot(data = M1avg %>% filter(weight >= 51)) + 
+  geom_net(aes(from_id = from, to_id = to, linewidth = weight/500), 
+           #directed = TRUE,
+           layout.alg = "mds", 
+           layout.par = list(var = 'user', dist = 'none', exp = 3,vm = distmat))
+
 @

### 2017-02-24 paper.Rnw
average viz eperiments

--- 2017-02-21 paper.Rnw
@@
 To create a visualization that represents an ``average" network, we count occurences of each possible edge in the simulations, resulting in a summary network of simulations.
-Next, we subset the network to include
+Next, we subset the network to include only those edges which appear in greater than 50\% of all
 
@@
 M1eff <- getEffects(mydata)
+library(netvizinf)
 M1sims1000 <- saom_simulate(dat = mydata, struct = M1eff,
                        parms = as.numeric(params), N = 1000)
-M1avg <- net_avg(M1sims1000, 1)
+M1simsdf <- sims_to_df(M1sims1000)
+M1avg <- M1simsdf %>% filter(!is.na(to) & wave == 1) %>%
+  group_by(from, to) %>% 
+  summarise(count = n()) %>% 
+  mutate(weight = ifelse(from == to, 0, count))
+M1avg2 <- M1simsdf %>% filter(!is.na(to)) %>%
+  group_by(from, to, wave) %>% 
+  summarise(count = n()) %>% 
+  mutate(weight = ifelse(from == to, 0, count))
+
+library(geomnet)
+M1avg$from <- paste0("n", M1avg$from)
+M1avg$to <- paste0("n", M1avg$to)
+M1avg2$from <- paste0("n", M1avg2$from)
+M1avg2$to <- paste0("n", M1avg2$to)
+
+library(tidyr)
+M1avg[,-3] %>% spread(to, weight) %>% data.frame-> distmat
+rownames(distmat) <- distmat$from
+distmat <- distmat[,-1]
+distmat <- as.matrix(1/distmat)
+distmat[is.na(distmat)] <- 0
+ggplot(data = M1avg %>% filter(weight/1000 > .5)) + # edge appears > 50% of the time. (is this actually what that is????) 
+  geom_net(aes(from_id = from, to_id = to, linewidth = weight/1000), 
+           directed = TRUE, curvature = .2,
+           layout.alg = "fruchtermanreingold"
+           #layout.alg = "kamadakawai",
+           #layout.par = list(elen = distmat)
+           )
+ggplot(data = M1avg2 %>% filter(weight/1000 > .5)) + # edge appears > 50% of the time. (is this actually what that is????) 
+  geom_net(aes(from_id = from, to_id = to, linewidth = weight/1000), 
+           directed = TRUE, curvature = .2, labelon = T, labelgeom = 'label', labelcolour = 'white',
+           layout.alg = "fruchtermanreingold"
+           #layout.alg = "kamadakawai",
+           #layout.par = list(elen = distmat)
+           ) + 
+  facet_wrap(~wave) + theme_net()
+ggplot(data = M1avg %>% filter(weight >= 51)) + 
+  geom_net(aes(from_id = from, to_id = to, linewidth = weight/500), 
+           #directed = TRUE,
+           layout.alg = "mds", 
+           layout.par = list(var = 'user', dist = 'none', exp = 3,vm = distmat))
+
 @

### 2017-03-01 paper.Rnw
add section on objective and rate functions

@@
 
-\section{The SAO model}
+\section{Stochastic Actor-Oriented Models for Longitudinal Social Networks}
 
@@
 
-The stochastic actor-oriented models are very aptly named: they model dynamic networks that can account for the individual node characteristics when making the edge changes.
+%The stochastic actor-oriented models are very aptly named: they model dynamic networks that can account for the individual node characteristics when making the edge changes.
 The object of analysis is a social network that has been observed at several discrete time points.
@@
 
-We start with a set of network observations at $M$ discrete time points: $x(t_1), x(t_2) \dots, x(t_M)$.
+%We start with a set of network observations at $M$ discrete time points: $x(t_1), x(t_2) \dots, x(t_M)$.
 The first observation, $x(t_1)$ is always conditioned on.
@@
 \hh{mostly lit review plus a detailed description of the model and fitting process.}
+
+A Stochastic Actor-Oriented Model (SAOM) is a model that is changing in time in order to accomodate for observations from the same network made at different points in time and that allows for changes in network structure due to actor-level covariates.
+This model was first introduced by Snijders in 1996 \citep{saompaper}.
+The two main properties of SAOMs, modelling the changes in time and using actor covariates, are crucial to understanding networks as they exist naturally.
+Most social networks, even holding constant the set of actors over time, are ever-changing as relationships decay or grow, and most actors (or nodes) in social networks have inherent properties that could affect how they change their place within the network.
+
+\subsection{Definitions, Terminology, and Notation}
+
+In this paper, the term \textit{dynamic network} refers to a network consisting of the same set of $n$ nodes that is changing over time, and is observed at $M$ discrete time points, $t_1, \dots, t_M$.
+We denote these network observations $x(t_1), \dots, x(t_M)$, and in modelling we condition on the first observation, $x(t_1)$.
+The SAOM assumes that this longitudinal network is embedded within a continuous time Markov process (CTMP), call it $X(T)$.
+This process is almost entirely unobserved: assume that the beginning of the process, $X(0)$ is equivalent to the first network observation $x(t_1)$, while the end of the process $X(\infty)$ is equivalent to the last observation $x(t_M)$.
+All other parts of the process are unseen.
+The process $X(T)$ is a series of single tie (or edge) changes, in which one actor at a time is given the opportunity to add or remove one outgoing tie.
+Once an actor is given the chance to change a tie, it tries to maximize a sort of utility function based on the current and potential future states of the network.
+
+\subsubsection{The Rate Function}
+
+For the network $x$ and each actor $i$ in the network, the rate function dictates how often the actor $i$ gets to change its ties, $x_{ij}$, to other nodes $j \neq i$ in the network.
+For this paper, we assume a simple rate function, $\alpha_m$ that is constant for all actors between observations at time $t_m$ to $t_{m+1}$.
+In other modelling scenarios, this rate can be a function that depends on the time period of observation, some actor-level covariates or some actor-level network statistics.
+The rate parameter determines how quickly actor $i$ gets an opportunity to change one of its ties, $x_{ij}$ in the time period $t_{m} \leq T < t_{m+1}$.
+We assume that the actors $i$ are conditionally independent given their current ties, $x_{i1}, \dots, x_{in}$.
+This assumption leads to the rate for the whole network, $n\alpha$.
+In order to achieve the memorylessness property of a Markov process, for any time point, $T$, where $t_{m} \leq T < t_{m+1}$, the waiting time to the next change opportunity by actor $i$ is exponentially distributed with expected value $\frac{1}{\alpha_m}$.
+Thus, the waiting time to the next change opportunity by \textit{any} actor in the network is also exponentially distributed with expected value $\frac{1}{n\alpha_m}$
+
+\subsubsection{The Objective Function}
+
+Thanks to the conditional dependence assumptions in the model, we can consider the objective function for each node separately, since only one tie from one node is changing at a time.
+The objective function is written as $$f_i(\boldsymbol{\beta}, x) = \sum_k \beta_k s_{ik}(x, \mathbf{Z}),$$
+for $x \in \mathcal{X}$, the space of all possible directed networks with the $n$ nodes, and $\mathbf{Z}$ matrix of covariates.
+The vector $\boldsymbol{\beta}$ are the parameters of the model with corresponding network and covariate statistics, $s_{ik}(x, \mathbf{Z})$, for $k = 1,\dots, K$.
+Given the ego node, $i$, there are $n$ possible steps for the actor $i$ to take: either one of all current ties $x_{ij} = 1$ will be destroyed, a new tie will be created, or no change will occur.
+
+The parameters, $\boldsymbol{\beta}$, are attached to various actor-level network statistics, $s_{ik}(x)$.
+There are always at least two parameters, $\beta_1$ for the outdegree of a node, and $\beta_2$ for the number of reciprocal ties held by a node \citep[p.
+371]{snijders01}.
+There are many possible parameters $\beta_i$ to add to the model.
+They can be split up into two groups: first, the structural effects, which only depend on the structure of the network.
+The inclusion of these effects has origin in the classical exponential random graph model (ERGM) for networks.
+These effects are written in terms of the edge variables $x_{ij}$, for $i \neq j$.
+The second set of effects are the actor-level or covariate effects.
+These effects also depend on the structure of the network.
+They are written in terms of $x_{ij}$ but also in terms of the covariates, $\mathbf{Z}$.
+A table of some possible structural and covariate effects is given in Table~\ref{tab:effects}.
+For a complete list of the network and covariate statistics that can be included in the objective function, see \citet{RSiena}.
+
+\begin{table}
+\caption{\label{tab:effects} Some of the possible effects to be included in the stochastic actor-oriented models in RSiena.
+There are many more possible effects, but we only consider a select few here.
+For a complete list, see the RSiena manual \citep{RSiena}.}
+\centering
+\begin{tabular}{ll}
+\textbf{Structural Effects} \\
+outdegree  & $s_{i1}(x) = \sum_j x_{ij}$ \\
+reciprocity  & $s_{i2}(x) = \sum_j x_{ij}x_{ji}$ \\
+transitive triplets  & $s_{i3}(x) = \sum_{j,h} x_{ij}x_{jh}x_{ih}$ \\
+\textbf{Covariate Effects}\\
+covariate-alter  & $s_{i4}(x) = \sum_j x_{ij}z_j$ \\
+covariate-ego  & $s_{i5}(x) = z_i\sum_j x_{ij}$ \\
+same covariate & $s_{i6}(x) = \sum_j x_{ij} \mathbb{I}(z_i = z_j)$ \\
+jumping transitive triplets  & $s_{i7}(x) = \sum_{j \neq h} x_{ij}x_{ih}x_{hj} \mathbb{I}(z_i = z_h \neq z_j)$
+\end{tabular}
+\end{table}
+
+When node $i$ is given the chance to change a node, we assume that they wish to maximize the value of their objective function $f_i(\boldsymbol{\beta}, x)$ plus a random element, $U_i(x)$, where the $U_i(x)$ are from "the type 1 extreme value distribution (or Gumbel distribution) with mean 0 and scale parameter 1" \citep[p.
+368]{snijders01}.
+This distribution, which is also known as the log-Weibull distribution, has probability distribution function, using $\mu$ for the mean parameter and $\sigma$ for the scale parameter, of
+
+$$f(u|\mu, \sigma) = \frac{1}{\sigma}\exp\left\{-\left(\frac{u-\mu}{\sigma} + e^{-\frac{u-\mu}{\sigma}}\right)\right\}.$$
+
+Using this distribution is convenient because it allows the probablity the actor $i$ chooses to change its tie to actor $j$ in terms of the objective function alone.
+Let $p_{ij}(\boldsymbol{\beta}, x)$ be this probability.
+Next, write the network $x$ in its potential future state, where the tie $x_{ij}$ has changed to $1-x_{ij}$, as $x(i \leadsto j)$.
+Then, the probility that the tie $x_{ij}$ changes is
+
+$$p_{ij}(\boldsymbol{\beta}, x) = \dfrac{\exp\left\{f_i(\boldsymbol{\beta}, x(i\leadsto j))\right\}}{\sum_{h \neq i} \exp\left\{f_i(\boldsymbol{\beta}, x(i \leadsto h))\right\}}$$
+
+%### A SAOM as a CTMC {#saomctmc}
+
+%In order to fit this model definition back into the original context of the CTMC described in Section \@ref(dynamicnets), it must be written in terms of its intensity matrix, $\mathbf{Q}$.
+This matrix describes the rate of change between states of the process.
+For networks, there are a very large number of possible states, $2^{n(n-1)}$, so the intensity matrix is a square matrix of that dimension.
+But, thanks to the property of SAOMs that the states are allowed to change only one tie at a time, there are only $n$ possible states given the current state, $n-1$ of which are uniquely determined by the node $i$ that is given the opportunity to change.
+Thus, the intensity matrix $\mathbf{Q}$ is very sparse, with only $n(n-1) + 1$ non-zero entries in each row.
+Note that $n(n-1)$ of these represent the possible states that are one edge different from a given state, and the additional non-zero entry is for the state to remain the same.
+All other entries in a row are zero because those column states cannot be reached from the row state by just one change as dictated by the SAOM.
+The entries of $\mathbf{Q}$ are defined as follows: let $b \neq c \in \{1, 2, \dots, 2^{n(n-1)} \}$ be indices of two different possible states of the network, $x^b, x^c \in \mathcal{X}$.
+Then the $bc^{th}$ entry of $Q$ is:
+% \[ q(x^b, x^c) = \begin{cases} 
+%       q_{ij}(\alpha, \rho, \boldsymbol{\beta}, x^b) = \lambda_i(\alpha, \rho, x^b, m)p_{ij}(\boldsymbol{\beta}, x^b) & \text{if } x^c \in \{x^b(i \leadsto j) | \text{ any } i \neq j \in \mathcal{N}\} \\
+%       0 & \text{if } x^c \text{ differs from } x^b \text{ by more than 1 tie} \\
+%       -\sum_{i\neq j} q_{ij}(\alpha, \rho, \boldsymbol{\beta}, x^b)  & \text{if } x^b = x^c 
+%    \end{cases}
+% \]
+% 
+% Thus, the rate of change between any two states that differ by only one tie, $x_{ij}$, is the product of the rate at which actor $i$ gets to change a tie and the probability that the tie that will change is the tie to node $j$.^[Just to be clear, the change is from $x^b_{ij}$ to $x^c_{ij} = 1 - x^b_{ij}$.] Furthermore, the theory of continuous time Markov chains gives that the matrix of transition probabilities between observation times $t_{m-1}$ and $t_{m}$ is dependent only on the difference between timepoints, $t_m - t_{m-1}$.
+Following the same definition for transition probabilities in Section \@ref(dynamicnets), the matrix of transition probabilities is
+% $$e^{(t_m - t_{m-1})\mathbf{Q}},$$ 
+% where $\mathbf{Q}$ is the matrix defined above and $e^X$ for a real or complex square matrix $X$ is equal to $\sum_{k=0}^{\infty} \frac{1}{k!} X^k$.
+
+\subsection{Fitting Models to Data}
 
@@
            directed = TRUE, curvature = .2,
-           layout.alg = "fruchtermanreingold"
-           #layout.alg = "kamadakawai",
-           #layout.par = list(elen = distmat)
+           #layout.alg = "fruchtermanreingold"
+           layout.alg = "kamadakawai",
+           layout.par = list(elen = distmat)
            )
@@
   geom_net(aes(from_id = from, to_id = to, linewidth = weight/1000), 
-           directed = TRUE, curvature = .2, labelon = T, labelgeom = 'label', labelcolour = 'white',
-           layout.alg = "fruchtermanreingold"
-           #layout.alg = "kamadakawai",
-           #layout.par = list(elen = distmat)
+           directed = TRUE, curvature = .2, labelon = T,
+           #layout.alg = "fruchtermanreingold"
+           layout.alg = "kamadakawai",
+           layout.par = list(elen = distmat)
            ) + 
   facet_wrap(~wave) + theme_net()
-ggplot(data = M1avg %>% filter(weight >= 51)) + 
-  geom_net(aes(from_id = from, to_id = to, linewidth = weight/500), 
-           #directed = TRUE,
-           layout.alg = "mds", 
-           layout.par = list(var = 'user', dist = 'none', exp = 3,vm = distmat))
+# ggplot(data = M1avg %>% filter(weight >= 51)) + 
+#   geom_net(aes(from_id = from, to_id = to, linewidth = weight/500), 
+#            #directed = TRUE,
+#            layout.alg = "mds", 
+#            layout.par = list(var = 'user', dist = 'none', exp = 3,vm = distmat))
 

### 2017-03-02 animatewithtweenr.R
promising results!

@@ -1,0 +1,143 @@
+# animation
+
+library(animation)
+
+#quincunx
+
+# need to take a set of points (x1,y1) and move them to new locations (x2,y2)
+ 
+# extract layout for initial points. then use those to layout first microstep
+ 
+# Step 0: get data in and model it. 
+library(RSiena)
+setwd("data/s50_data")
+friend.data.w1 <- as.matrix(read.table("s50-network1.dat"))
+friend.data.w2 <- as.matrix(read.table("s50-network2.dat"))
+friend.data.w3 <- as.matrix(read.table("s50-network3.dat"))
+drink <- as.matrix(read.table("s50-alcohol.dat"))
+smoke <- as.matrix(read.table("s50-smoke.dat"))
+friendship <- sienaDependent(
+  array(c(friend.data.w1, friend.data.w2,
+          friend.data.w3),
+        dim = c(50, 50, 3)))
+smoke1 <- coCovar(smoke[ , 1 ])
+alcohol <- varCovar(drink)
+mydata <- sienaDataCreate(friendship, smoke1, alcohol)
+myeffnull <- getEffects(mydata)
+myeffalt1 <- includeEffects(myeffnull, transTrip)
+myalgorithm <- sienaAlgorithmCreate(projname = 's50_3')
+set.seed(823746)
+ansnull <- siena07(myalgorithm, data = mydata, effects = myeffnull,
+                   returnChains = TRUE, returnDataFrame = TRUE,
+                   returnDeps = TRUE, silent = TRUE, verbose = FALSE,
+                   batch = TRUE)
+get_chain_info <- function(ans){
+  L <- length(ans$chain)
+  M <- length(ans$chain[[1]][[1]])
+  bigres <- NULL
+  res <- data.frame()
+  for (l in 1:L){
+    res <- data.frame()
+    for (m in 1:M){
+      sub <- data.frame(plyr::ldply(
+        lapply(ansnull$chain[[l]][[1]][[m]], unlist),
+        rbind), stringsAsFactors = FALSE)
+      sub$rep <- l
+      sub$period <- m
+      res <- rbind(res, sub)
+    }
+    bigres[[l]] <- res  
+  }
+  ret <- plyr::rbind.fill(bigres)
+  return(ret)
+}
+ansnullchains <- get_chain_info(ansnull)
+library(dplyr)
+library(readr)
+ansnullchains %>% 
[…]

### 2017-03-03 animatewithtweenr.R
OMG <3

@@ -107,6 +107,10 @@ names(layout_init) <- c("x", "y")
 layout_init$colour <- 'red'
 layout_init$id <- as.factor(get.vertex.attribute(w1.net, "vertex.names"))
 data1 <- layout_init
+library(geomnet)
+wave1friends <- fortify(as.adjmat(friend.data.w1))
+netwave1 <- merge(merge(wave1friends, data1, by.x = 'from', by.y = 'id'), data1, by.x = 'to', by.y = 'id', suffixes = c(".from", ".to"))
+head(netwave1)
 # Step 2: get next network data observation
 # Microstep friend data wave 1 , step 1
 ms_fdw1.1 <- ms1[[2]]
@@ -124,10 +128,20 @@ layout_ms_fdw1.1$colour <- 'grey40'
 # as.factor() gets rid of that dumb color error. there's an internal function col_classes that makes to big of an assumption about character class columns
 layout_ms_fdw1.1$id <- as.factor(get.vertex.attribute(net_ms_fdw1.1, "vertex.names"))
 data2 <- layout_ms_fdw1.1
+
+wave1friendsms1 <- fortify(as.adjmat(ms_fdw1.1))
+netwave1ms1 <- merge(merge(wave1friendsms1, data2, by.x = 'from', by.y = 'id'), data2, by.x = 'to', by.y = 'id', suffixes = c(".from", ".to"))
+
 library(tweenr)
 # TWEEN EM
 data <- tween_states(data = list(data1, data2), tweenlength = 3, 1, "linear", 100)
-head(data)
+fulldata <- tween_states(data = list(netwave1, netwave1ms1), 3, 1, 'linear', 50)
+# doesn't work, need same # of rows. Need to deal with to NAs somehow. 
+netwave1 <- arrange(netwave1, to,from)
+netwave1ms1 <- arrange(netwave1ms1, to, from)
+head(netwave1ms1)
+n1 <- nrow(netwave1)
+n2 <- nrow(netwave1ms1)
 library(ggplot2)
 library(gganimate)
 
@@ -140,4 +154,66 @@ p <- ggplot(data=data, aes(x=x, y=y)) +
 animation::ani.options(interval = 1/15)
 gganimate(p, "dancing ball.gif", title_frame = F)
 
+# okay - here's what we need:
+  # a data frame of every possible edge
+  # from.x and from.y and to.x, to.y
+  # rules: from.x and from.y is always the point corresponding to the from node. 
[…]

### 2017-03-03 animatewithtweenr.R
weirdness

@@ -134,25 +134,25 @@ netwave1ms1 <- merge(merge(wave1friendsms1, data2, by.x = 'from', by.y = 'id'), 
 
 library(tweenr)
 # TWEEN EM
-data <- tween_states(data = list(data1, data2), tweenlength = 3, 1, "linear", 100)
-fulldata <- tween_states(data = list(netwave1, netwave1ms1), 3, 1, 'linear', 50)
-# doesn't work, need same # of rows. Need to deal with to NAs somehow. 
-netwave1 <- arrange(netwave1, to,from)
-netwave1ms1 <- arrange(netwave1ms1, to, from)
-head(netwave1ms1)
-n1 <- nrow(netwave1)
-n2 <- nrow(netwave1ms1)
+# data <- tween_states(data = list(data1, data2), tweenlength = 3, 1, "linear", 100)
+# fulldata <- tween_states(data = list(netwave1, netwave1ms1), 3, 1, 'linear', 50)
+# # doesn't work, need same # of rows. Need to deal with to NAs somehow. 
+# netwave1 <- arrange(netwave1, to,from)
+# netwave1ms1 <- arrange(netwave1ms1, to, from)
+# head(netwave1ms1)
+# n1 <- nrow(netwave1)
+# n2 <- nrow(netwave1ms1)
 library(ggplot2)
 library(gganimate)
 
 #animate em
-p <- ggplot(data=data, aes(x=x, y=y)) + 
-  geom_point(aes(frame = .frame, colour = colour), size=5) + 
-  geom_text(aes(frame = .frame, label=id), alpha = .6) + 
-  scale_colour_identity() + 
-  theme_bw()
-animation::ani.options(interval = 1/15)
-gganimate(p, "dancing ball.gif", title_frame = F)
+# p <- ggplot(data=data, aes(x=x, y=y)) + 
+#   geom_point(aes(frame = .frame, colour = colour), size=5) + 
+#   geom_text(aes(frame = .frame, label=id), alpha = .6) + 
+#   scale_colour_identity() + 
+#   theme_bw()
+# animation::ani.options(interval = 1/15)
+# gganimate(p, "dancing ball.gif", title_frame = F)
 
 # okay - here's what we need:
   # a data frame of every possible edge
@@ -205,15 +205,44 @@ big1$from <- as.factor(big1$from)
 big1$to<- as.factor(big1$to)
 big2$from <- as.factor(big2$from)
[…]

### 2017-03-12 paper.Rnw
write write write write write

@@
 
+Models for dynamic social networks are extremely important because of how realistic they are.
+Social networks do not form spontaneously; they evolve over time.
+Ties are added and removed, and new actors can join the network.
+Modeling the underlying mechanisms that create network changes over time is very complex but also has the potential to uncover hidden truths.
+
+\par One set of models for dynamic social networks are the stochastic actor-oriented models (SAOMs) of \citet{saompaper}.
+These models are uniquely different from other social network models because they incorporate network statistics and actor statistics into the model, whereas most other models just use the network statistics to determine change.
+Allowing the actor-level statistics to change the structure of the network leads to a more intuitive model for social network change.
+In the "real-world" we would expect that people with common interests would be more likely to form relationships, and that intuition is no longer ignored in modeling with SAOMs.
+
 \st{
@@
 
-The guiding principles XXX cite XXX of model visualization are (1) display the model in the data space, (2) collections of models are more informative than singletons, and (3) explore how algorithms work, instead of just looking at the final result.
+The guiding principles of model visualization from \citet{modelvis} are (1) display the model in the data space, (2) collections of models are more informative than singletons, and (3) explore how algorithms work, instead of just looking at the final result.
 Stochastic actor-oriented models are a prime example of a set of models that  benefit greatly from application of model visualization.

### 2017-03-13 paper.Rnw
beefing up section 3

@@
 \section{Introduction} 
+\st{
+\begin{itemize}
+\item What? - model visualization and SAOMs. briefly dsecribe what these things are
+\item Why? - bring light to the RSiena blackbox. Need ways to visualize network models, collections of networks. 
+\item How?
+\end{itemize}
+also talk about other dynamic network models briefly to put SAOMs in context.}
 
@@
 
-\st{
-\begin{itemize}
-\item What? - model visualization and SAOMs. briefly dsecribe what these things are
-\item Why? - bring light to the RSiena blackbox. Need ways to visualize network models, collections of networks. 
-\item How?
-\end{itemize}
-also talk about other dynamic network models briefly to put SAOMs in context.}
+\par The software SIENA, and it's \texttt{R} implementation \texttt{RSiena}, was developed by \citet{RSiena} to estimate the parameters of the SAOMs, among other social network models.
+This software is a huge contribution to the field of social network analysis, and when we began working with it we were eager to understand how it works.
+To better understand the \texttt{RSiena} software and the SAOMs themselves, we use the principals of model visualization from \citet{modelvis}.
+With the work of \citet{modelvis} guiding us, we explore RSiena and SAOMs with innovative visualizations, both static and dynamic.
+By bringing some light to the underlying methods and structures "behind the scenes" of SAOMs and RSiena, we will help other researchers working with the models better understand and better use these models.
 
-The guiding principles of model visualization from \citet{modelvis} are (1) display the model in the data space, (2) collections of models are more informative than singletons, and (3) explore how algorithms work, instead of just looking at the final result.
+\par The guiding principles of model visualization from \citet{modelvis} are (1) display the model in the data space, (2) collections of models are more informative than singletons, and (3) explore how algorithms work, instead of just looking at the final result.
 Stochastic actor-oriented models are a prime example of a set of models that  benefit greatly from application of model visualization.
@@
 
-When node $i$ is given the chance to change a node, we assume that they wish to maximize the value of their objective function $f_i(\boldsymbol{\beta}, x)$ plus a random element, $U_i(x)$, where the $U_i(x)$ are from "the type 1 extreme value distribution (or Gumbel distribution) with mean 0 and scale parameter 1" \citep[p.
+When node $i$ is given the chance to change a tie, we assume that they wish to maximize the value of their objective function $f_i(\boldsymbol{\beta}, x)$ plus a random element, $U_i(x)$, where the $U_i(x)$ are from "the type 1 extreme value distribution (or Gumbel distribution) with mean 0 and scale parameter 1" \citep[p.
 368]{snijders01}.
@@
 
+To fit a SAOM to longitudinal network data, we use the package \texttt{RSiena} \citep{RSiena}.
+This package uses simulation methods to estimate parameter values using either the method of moments or maximum likelihood estimation.
+In this paper, we focus of the method of moments estimation because the theory behind it was established in \citep{saompaper}, while the maximum likelihood estimation methods were not fully established until \citet{saomsml2010}.
+
 \section{Model Visualizations}
@@
 
+The concept of model visualization was developed to complement traditional model diagnostic tools.
+Typically, numerical summaries such as $R^2$ are used to assess model fit, and visualizations, like residual plots, are restricted to the data.
+Every good data analysis includes both numerical and visual summaries of the data, so why should model description be restricted to numerical summaries?
+\citeauthor{modelvis} outline the three different parts of a model: the model family, the model form, and the fitted model.
+The latter is primarily what one thinks of first when considering a model in a data analysis, where a model is fit to the data, and parameter estimates reported.
+In the context of SAOMs, the fitted model contains the estimated rate parameters and the estimated objective function parameters.
+The model form describes the the model before the fitting process and how variables describe the data in the model.
+In SAOMs, the model form includes description of the rate and objective functions and the variables therein that describe how the network evolves over time.
+Finally, the model family is the most broad description of the model.
+This is the type of model that you wish to fit to the data, and is chosen based on the problem and knowledge at had.
+For example, we chose to use a SAOM to model network data over an exponential random graph model (ERGM) because we believe that actor-level variables effect network dynamics.
+These three parts of a "model" are should be visualized according to the three principals of model visualization: display the model in the data space, look at a collection of models, and explore the process of fitting the model, not just the end result.
+
 \subsection{View the model in the data space}
 
+In \citeauthor{modelvis}, they define the \textit{data space} as ``the region over which we can reliably make inferences, usually a hypercube containing the data" \citep[p.~206]{modelvis}.
+What does this definition mean for network data?
+For dynamic social networks, there are a few different data ``spaces." First, we have the actors and their corresponding covariates.
+Second, we have the dyad variables that describe the relationships between the variables.
+Finally, we have the time over which the network is allowed to evolve.
+These three data pieces can be visualized together in various ways.
+The traditional network visualization uses one of many algorithms to layout the actors as points in 2D space, then draws segments connecting the points in 2D if the dyad between two nodes has value greater than zero, and draws nothing otherwise.
+The time aspect can be visualized by drawing each network observation in time and placing the observed timepoints side-by-side.
 \st{
@@
 
-The second principle of Wickham et al states that ``collections are more informative than singletons" (p.
-210).
+The second principle of Wickham et al states that ``collections are more informative than singletons" \citep[p.~210]{modelvis}.
 They describe several different methods for producing such collections, but we focus on only a few here.
-We look at collections produced ``from exploring the space of all possible models," ``by varying model settings", ``by fitting the same model to different datasets", and ``by treating each iteration as a model." (p.
-210-11) We also fit the same model many times, resulting in distributions of fitted parameters for one data set.
+We look at collections produced ``from exploring the space of all possible models," ``by varying model settings", ``by fitting the same model to different datasets", and ``by treating each iteration as a model" \citep[p.~210-11]{modelvis}.
+We also fit the same model many times, resulting in distributions of fitted parameters for one data set.
 
@@
 library(geomnet)
-M1avg$from <- paste0("n", M1avg$from)
-M1avg$to <- paste0("n", M1avg$to)
-M1avg2$from <- paste0("n", M1avg2$from)
-M1avg2$to <- paste0("n", M1avg2$to)
+M1avg$from <- paste0("V", M1avg$from)
+M1avg$to <- paste0("V", M1avg$to)
+M1avg2$from <- paste0("V", M1avg2$from)
+M1avg2$to <- paste0("V", M1avg2$to)
 

### 2017-03-13 paper.Rnw
chugging along

@@
 This package uses simulation methods to estimate parameter values using either the method of moments or maximum likelihood estimation.
-In this paper, we focus of the method of moments estimation because the theory behind it was established in \citep{saompaper}, while the maximum likelihood estimation methods were not fully established until \citet{saomsml2010}.
+In this paper, we focus of the method of moments estimation because the theory behind it was established in \citet{saompaper}, while the maximum likelihood estimation methods were not fully established until \citet{saomsml2010}.
+We also focus on the score function method for estimating the derivatives of the expected values, as opposed to the finite differences method, both of which are outlined in detail in \citet{SienaAlgs}.
 
-\section{Model Visualizations}
+\par For the score function method, the SIENA software uses a Robbins-Monro algorithm (see \citet{robbinsmonro}) to estimate the solution of the moment equation
+$$ E_{\theta}S = s_{obs}$$
+where \theta is the vector of rate and objective function parameters, and $s_{obs}$ is the observed vector of model statistics, $S$.
+The entire algorithm is provided in \citet{SienaAlgs}.
+\par There are three phases in the the SIENA algorithm, as described in \citet{RSienamanual, SienaAlgs}.
+The first phase performs initial estimation of the score functions used in the Robbins-Monro procedure.
+The second phase carries out the Robbins-Monro algorithm and obtains iterative estimates of the parameter values.
+The third phase uses the parameter vector estimated in phase two to estimate the score functions and the covariance matrix of the parameters, as well as performs convergence checks.
+In each of the the first two phases, the estimation procedure also relies on ``microsteps" that simulate from the model as it exists in its current state in order to update either the score functions or the parameter estimates.
+These microsteps become observed instances of the continuous-time Markov chain that is the backbone of the stochastic actor-oriented model.
+In Section~\ref{sec:ModelVis}, we visualize the parts of the SIENA method-of-moments algorithm, bringing them out of the ``black-box" and into the light.
+
+\section{Model Visualizations} \label{sec:ModelVis}
 
@@
 These three parts of a "model" are should be visualized according to the three principals of model visualization: display the model in the data space, look at a collection of models, and explore the process of fitting the model, not just the end result.
+
+\subsection{Explore algorithms, not just end result}
+
+The model fitting process in \texttt{RSiena} involves several simulation steps that are hidden from the user.
+This is practical and efficient if a researcher is primarily interested in fitting one model to a set of longitudinal network data and obtaining parameter estimates.
+We are more interested in how the models are fit, so we extracted and explored the different steps of that process.
+
+A key component of each step of the SIENA method of moments algorithm is the ``microstep." This step simulates from the model in its current state, $x(t_m)$ with current parameter values $\theta_{0}$, to the next state, $x(t_{m+1})$, or as close as it can get to that state.
+This simulation process follows the steps of the continuous-time Markov chain.
+First, an ``ego node'' is selected to make a change, and then it randomly makes one change in its ties according to the probability, $p_{ij}$, determined by its objective function.
+The options for change are (1) removing a current tie, (2) adding a new tie, or (3) making no change at all.
+Saving all of these steps is not efficient if you're only interested in parameter values, but they can be saved and extracted using options in \texttt{RSiena}, which is exactly what we did to create our visualizations.
+Between two network observations $x(t_m)$ and $x(t_{m+1})$, there can be hundreds or even thousands of microsteps.
+We wanted to see these in-between steps.
+
+\st{The algorithms here are many!
+\begin{enumerate}
+\item First, the microsteps / CTMC between each time point (gifs, other animations)
+\item Phases of RSiena: phase 1 = initial parameter estimate
+\item phase 2 = iterative update of parameters - trace plots
+\item phase 3 = simulations for convergence diagnostics
+\end{enumerate}
+ }
+
+\hh{The visualizations should address most elements of structure described before, i.e.
+follow along in the setup (for descriptive and teaching purposes), the fitting (understanding the model and interpreting the results) and then diagnostics (how well does the model fit the data - where are the most differences to the actual data/what are the sensitive parameters?) }
+
+\hh{Some of the visualizations we talked about
+\begin{enumerate}
+\item visualization of the MCMC process:  movie of individual actors' step-by-step decisions; could be done with the ndtv package, as long as we can get individual steps out of RSiena
+\item global picture: 1000 simulated networks + alpha blending should give a pretty good idea of the most important edges
+%\item global structure + principal components (tensor pca)
+\end{enumerate}}
 
@@
 
-\subsection{Explore algorithms, not just end result}
-\st{The algorithms here are many!
-\begin{enumerate}
-\item First, the microsteps / CTMC between each time point (gifs, other animations)
-\item Phases of RSiena: phase 1 = initial parameter estimate
-\item phase 2 = iterative update of parameters - trace plots
-\item phase 3 = simulations for convergence diagnostics
-\end{enumerate}
- }
-
-\hh{The visualizations should address most elements of structure described before, i.e.
-follow along in the setup (for descriptive and teaching purposes), the fitting (understanding the model and interpreting the results) and then diagnostics (how well does the model fit the data - where are the most differences to the actual data/what are the sensitive parameters?) }
-
-\hh{Some of the visualizations we talked about
-\begin{enumerate}
-\item visualization of the MCMC process:  movie of individual actors' step-by-step decisions; could be done with the ndtv package, as long as we can get individual steps out of RSiena
-\item global picture: 1000 simulated networks + alpha blending should give a pretty good idea of the most important edges
-\item global structure + principal components (tensor pca)
-\end{enumerate}}
-
-\section{}
+\section{Discussion}
 

### 2017-03-13 paper.Rnw
just keep swimming

@@
 We denote these network observations $x(t_1), \dots, x(t_M)$, and in modelling we condition on the first observation, $x(t_1)$.
-The SAOM assumes that this longitudinal network is embedded within a continuous time Markov process (CTMP), call it $X(T)$.
+The SAOM assumes that this longitudinal network is embedded within a continuous time Markov chain (CTMC), call it $X(T)$.
 This process is almost entirely unobserved: assume that the beginning of the process, $X(0)$ is equivalent to the first network observation $x(t_1)$, while the end of the process $X(\infty)$ is equivalent to the last observation $x(t_M)$.
@@
 \begin{table}
-\caption{\label{tab:effects} Some of the possible effects to be included in the stochastic actor-oriented models in RSiena.
-There are many more possible effects, but we only consider a select few here.
-For a complete list, see the RSiena manual \citep{RSiena}.}
 \centering
@@
 \end{tabular}
+\caption{\label{tab:effects} Some of the possible effects to be included in the stochastic actor-oriented models in RSiena.
+There are many more possible effects, but we only consider a select few here.
+For a complete list, see the RSiena manual \citep{RSiena}.}
 \end{table}
@@
 $$ E_{\theta}S = s_{obs}$$
-where \theta is the vector of rate and objective function parameters, and $s_{obs}$ is the observed vector of model statistics, $S$.
+where $\theta$ is the vector of rate and objective function parameters, and $s_{obs}$ is the observed vector of model statistics, $S$.
 The entire algorithm is provided in \citet{SienaAlgs}.
@@
 In Section~\ref{sec:ModelVis}, we visualize the parts of the SIENA method-of-moments algorithm, bringing them out of the ``black-box" and into the light.
+
+\subsection{Example Data}
+
+To guide our visual exploration of stochastic actor-oriented models, we used a subset of the 50 actor dataset from the ``Teenage Friends and Lifestyle Study" that is provided on the \texttt{RSiena} webpage.
+These data come from \citet{friendsdata}, and we chose to only work with a subset of the data to make network visualizations less busy, with changes that are more noticeable.
+The subset contained actors 20 through 35 and the ties between them, as well as the drinking behavior of each actor at each of the three waves.
+This specific subset was chosen because it showed somewhat higher connectivity than other subsets, as we've emphasized in the visualizations of the three network adjacency matrices in Figure~\ref{fig:adjmatviz}.
+For model fitting, we condition on wave 1 and estimate the parameters of the models from the second and third waves.
+We will also be working with one actor level categorical covariate, drinking behavior.
+This variable has five values in the original data: (1) does not drink, (2) drinks once or twice a year, (3) drinks once a month, (4) drinks once a week, and (5) drinks more than once a week.
+The actor covariates adn networks are visualized in Figure~\ref{fig:alldataviz}.
+<<adjmatviz, fig.align='center', fig.cap="A visualization of the adjacency matrices of the three waves of network observations in the ``Teenage Friends and Lifestyle Study'' data.
+The subset we will be using is outlined in red.", fig.width=8>>=
+library(network)
+library(sna)
+library(geomnet)
+setwd("code")
+source("00e_small_friends.R")
+view_rs2$wave <- factor(view_rs2$wave)
+levels(view_rs2$wave) <- c("Wave 1", "Wave 2", "Wave 3")
+ggplot(data= view_rs2, aes(x = x, y=y)) + 
+  geom_tile(aes(fill = as.factor(value))) + 
+  scale_fill_manual("i likes j", labels=c("no", "yes"), 
+                    values = c('white', 'black')) +
+  geom_rect(data= NULL, inherit.aes = FALSE, color = 'red',
+            aes(xmin = 19.5, xmax = 35.5, ymin = 19.5, ymax = 35.5, fill =NA)) + 
+  facet_wrap(~wave) + theme_bw() +
+  labs(x = "i", y = "j") + 
+  theme(aspect.ratio=1, axis.text = element_blank(), axis.ticks = element_blank()) 
+@
+
+<<alldataviz, fig.cap="The smaller friendship network data we will be modelling throughout the paper.", fig.width=8>>=
+actual1 <- merge(data.frame(as.edgelist(as.network(fd2.w1))), 
+                 data.frame(id = 1:16, drink = drink2[,1]),
+                 by.x = "X1", by.y = "id", all = T)
+actual1$Wave <- 1 
+actual2 <- merge(data.frame(as.edgelist(as.network(fd2.w2))),
+                 data.frame(id = 1:16,drink = drink2[,2]),
+                 by.x = "X1", by.y = "id", all = T)
+actual2$Wave <- 2
+actual3 <- merge(data.frame(as.edgelist(as.network(fd2.w3))),
+                 data.frame(id = 1:16,drink = drink2[,3]),
+                 by.x = "X1", by.y = "id", all = T)
+actual3$Wave <- 3
+alldat <- rbind(actual1, actual2, actual3)
+alldat$X1 <- paste0("V", alldat$X1)
+alldat$X2 <- ifelse(!is.na(alldat$X2), paste0("V", alldat$X2), NA)
+ggplot(data = alldat, 
+       aes(from_id = X1, to_id = X2, color = as.ordered(drink), group = as.ordered(drink))) +
+  geom_net(fiteach = FALSE, directed=TRUE, layout.alg = 'fruchtermanreingold', size = 3, arrowsize = .5, arrowgap = .03) +
+  scale_color_brewer(palette = "YlOrRd", name = "Drinking\nBehavior") + 
+  theme_net() +
+  facet_wrap(~Wave, labeller = "label_both") + 
+  theme(legend.position = 'bottom', panel.background = element_rect(fill = NA, color = "grey30"))
+@
 
@@
 Between two network observations $x(t_m)$ and $x(t_{m+1})$, there can be hundreds or even thousands of microsteps.
-We wanted to see these in-between steps.
+We wanted to see these in-between steps in order to better understand the behavior of the underlying continuous-time Markov chain.
+
+The first vizualization we present here is an animation of microsteps simulated to emulate the transition steps from wave 1 to wave 2 the subset of the friends data shown in Figure ~\ref{fig:alldataviz}.
+When each ego node is selected in a microstep, it is emphasized in the animation, then the associated edge either appears or disappears.
+If there are no changes at a particular microstep, no changes are seen.
+Some frames of the animations are shown in INSERT STATIC FIGURE, and the full movie can be viewed at INSERT LINK TO GIF.
 
@@
 \begin{enumerate}
-\item First, the microsteps / CTMC between each time point (gifs, other animations)
+%\item First, the microsteps / CTMC between each time point (gifs, other animations)
 \item Phases of RSiena: phase 1 = initial parameter estimate
@@
 Statisticians frequently rely on averages and expected values, but statistical network models, especially those as complex as SAOMs, lack an intuitive expected value measure.
+We could talk about expected values of parameters, but there is no way to talk about the expected value of the network based on the model.
 How then, can we arrive at an ``average" network?
@@
 
-We first consider the 50 actor subset of the teenage friends and lifestyle data available on the \texttt{RSiena} website \citep{RSiena}.
-To this data, we fit a SAOM with a simple rate function and an objective function with two parameters, $f_i(x) = \beta_1s_{i1} + \beta_2s_{i2}$, where $s_{i1}$ is the density network statistic and $s_{i2}$ is the reciprocity network statistic for actor $i$.
+We first consider the 16 actor subset of the teenage friends and lifestyle data available on the \texttt{RSiena} website \citep{RSiena}.
+To this data, we fit three different SAOMs.
+Each SAOM used a simple rate function, $\alpha_m$, and an objective function with two or three parameters.
+The first model, $M1$, contains the absolute minimum number of parameters in the objective function $f_i(x)$:  $$f_i(x)^{M1} = \beta_1s_{i1} + \beta_2s_{i2}$$, where $s_{i1}$ is the density network statistic and $s_{i2}$ is the reciprocity network statistic for actor $i$.
+The second and third models, $M2$ and $M3$, contain a third parameter in the objective function that was determined in the\texttt{RSienaTest} software to be significant, with $p$-values less than 0.05 \citep{RSienaTest}.
+The $M2$ model contains an actor-level covariate parameter, and the $M3$ model contains an additional strutural effect in the objective function.
+\begin{align*}
+f_i(x)^{M2} & = \beta_1s_{i1} + \beta_2s_{i2} + \beta_3s_{i3} \\
+f_i(x)^{M3} & = \beta_1s_{i1} + \beta_2s_{i2} + \beta_4s_{i4} 
+\end{align*}, where $s_{i3} = \sum_{j \neq h} x_{ij}x_{ih}x_{hj} \mathbb{I}(z_i = z_h \neq z_j)$, and $s_{i4} =  |\{j : x_{ij} = 0, \sum\limits_h x_{ih}x_{hj} \geq 2\}|$.
+These statistics are known as the number of jumping transitive triplets and the number of doubly achieved distances two effect, respectively.
+The first statistic emphasizes triad relationships that are formed between actors from different covariate groups, while the other emphasizes indirect ties between actors.
+These two effects are visually represented in Figure~\ref{fig:jtt} and ~\ref{fig:dad}, respectively.
+
+\begin{figure}\label{fig:structures}
+\begin{subfigure}[t]{.45\textwidth}
+\caption{\label{fig:jtt}Realization of a jumping transitive triplet, where $i$ is the focal actor, $j$ is the target actor, and $h$ is the intermediary.
+The group of the actors is represented by the shape of the node.}
+<<jtt, echo=FALSE, fig.width=2, fig.height=2, message = FALSE, warning = FALSE, fig.align='center'>>=
+jTTe <- data.frame(from = c('i', 'i', 'h'), to = c('h', 'j', 'j'))
+jTTn <- data.frame(id = letters[8:10], group = c(1,1,2))
+
+jTT <- merge(jTTe, jTTn, by.x = 'from', by.y = "id", all = T)
+
+set.seed(12345) 
+ggplot(data = jTT, aes(from_id = from, to_id = to)) + 
+  geom_net(aes(shape = as.factor(group)), directed = T, labelon = T, 
+           labelcolour='grey80',vjust = 0.5, hjust =0.5, arrowgap = .15, 
+           colour = 'black', size=10, 
+           ecolour = c("red", "grey40", "grey40", "grey40")) + 
+  expand_limits(x=c(-0.1,1.1), y=c(-0.1,1.1)) +
+  theme_net() +
+  theme(legend.position = "none")
+@
+\end{subfigure}\hfill
+\begin{subfigure}[t]{.45\textwidth}
+\caption{\label{fig:dad}Doubly achieved distance between actors $i$ and $k$.}
+<<dab, echo=FALSE, fig.width=2, fig.height=2, message = FALSE, warning = FALSE, fig.align='center'>>=
+dade <- data.frame(from = c('i', 'i', 'h', 'j'), to = c('h', 'j', 'k', 'k'))
+dadn <- data.frame(id = letters[8:11], group = c(1,1,1,1))
+
+dad <- merge(dade, dadn, by.x = 'from', by.y = "id", all = T)
+
+set.seed(12345) 
+ggplot(data = dad, aes(from_id = from, to_id = to)) + 
+  geom_net(aes(shape = as.factor(group)), directed = T, labelon = T, labelcolour='grey80',vjust = 0.5, hjust =0.5, arrowgap = .15, colour = 'black', size=10) + 
+  expand_limits(x=c(-0.1,1.1), y=c(-0.1,1.1)) +
+  theme_net() +
+  theme(legend.position = "none")
+@
+\end{subfigure}
+\caption{Structural newtwork effects. On the left, a jumping transitive triplet (JTT). On the right, a doubly achieved distance between $i$ and $k$.}
+\end{figure}
 We fit this model using \texttt{RSiena} 1,000 times, then averaged the resulting parameter estimates to obtain mean estimates to use in the simulation of networks from this model.
@@
 <<getdata, results='asis'>>=
-library(netvizinf)
-data("M1ests_bigfriends")
-params <- data.frame(t(colMeans(M1ests_bigfriends)))
-names(params) <- c("Rate1", "Rate2", "beta1", "beta2")
-knitr::kable(params, digits = 3, align = 'c',caption = "\\label{tab:M1ests} Estimates of the four parameters from the simple SAOM fit to the 50 actor subset of the teenage friends and lifestyle data.", format = "latex")
+setwd("data")
+nulls <- read.csv("distribution_null_model.csv")
+alt <- read.csv("distribution_jumpTT_model.csv")
+alt2nd <- read.csv("distribution_dblpairs_model.csv")
+simu2 <- read.csv("simulation-1000-M1-M2-M3.csv")
+means <- simu2 %>% group_by(Model, parameter) %>% summarize(
+  mean = mean(estimate)
+)
+library(dplyr)
+null_mod_sv <- as.numeric(nulls[,-c(1,6,7)] %>% summarise_each(funs(mean)))
+alt_mod_sv <- as.numeric(alt[,-c(1, 7,8)] %>% summarise_each(funs(mean)))
+alt_mod2_sv <- as.numeric(alt2nd[,-c(1,7,8)] %>% summarise_each(funs(mean)))
 @
@@
 library(dplyr)
-friend.data.w1 <- as.matrix(read.table("data/s50_data/s50-network1.dat"))
-friend.data.w2 <- as.matrix(read.table("data/s50_data/s50-network2.dat"))
-friend.data.w3 <- as.matrix(read.table("data/s50_data/s50-network3.dat"))
-drink <- as.matrix(read.table("data/s50_data/s50-alcohol.dat"))
+#friend.data.w1 <- as.matrix(read.table("s50_data/s50-network1.dat"))
+#friend.data.w2 <- as.matrix(read.table("s50_data/s50-network2.dat"))
+#friend.data.w3 <- as.matrix(read.table("s50_data/s50-network3.dat"))
+#drink <- as.matrix(read.table("s50_data/s50-alcohol.dat"))
 friendshipData <- array(c(friend.data.w1, friend.data.w2, friend.data.w3), dim = c(50, 50, 3))
@@
 library(netvizinf)
+data(M1ests_bigfriends)
+M1parms <- colMeans(M1ests_bigfriends)
 M1sims1000 <- saom_simulate(dat = mydata, struct = M1eff,
-                       parms = as.numeric(params), N = 1000)
+                       parms = as.numeric(M1parms), N = 1000)
 M1simsdf <- sims_to_df(M1sims1000)

### 2017-03-13 code/00e_small_friends.R
upload data

@@ -1,0 +1,170 @@
+# New model with a smaller collection of girls from the friendship data
+#deciding how to subset the girls:
+library(ggplot2)
+library(tidyr)
+library(dplyr)
+
+library(RSiena) # or RSienaTest
+# read in network data
+friend.data.w1 <- as.matrix(read.table("../data/s50_data/s50-network1.dat"))
+friend.data.w2 <- as.matrix(read.table("../data/s50_data/s50-network2.dat"))
+friend.data.w3 <- as.matrix(read.table("../data/s50_data/s50-network3.dat"))
+
+val1 <- data.frame(friend.data.w1) %>% gather(x,y, V1:V50) %>% select(y)
+val2 <- data.frame(friend.data.w2) %>% gather(x,y, V1:V50) %>% select(y)
+val3 <- data.frame(friend.data.w3) %>% gather(x,y, V1:V50) %>% select(y)
+
+all_girls <- expand.grid(x = 1:50, y= 1:50)
+
+view_rs <- data.frame(all_girls, W1 = val1, W2 = val2, W3 = val3)
+names(view_rs)[3:5] <- c('w1', 'w2','w3')
+
+view_rs %>% gather(wave, value, w1:w3) -> view_rs2
+
+# look at adjacency matrices
+# ggplot(data= view_rs2, aes(x = x, y=y)) + 
+#   geom_tile(aes(fill = as.factor(value))) + scale_fill_manual(values = c('white', 'black')) +
+#   geom_rect(data= NULL, inherit.aes = FALSE, 
+#             aes(xmin = 20, xmax = 35, ymin = 20, ymax = 35, fill =NA), color = 'red') + 
+#   facet_wrap(~wave) + theme_bw() + 
+#   theme(aspect.ratio=1)
+
+#looks like 20:35 are pretty well connected, go with those for now
+fd2.w1 <- friend.data.w1[20:35,20:35]
+fd2.w2 <- friend.data.w2[20:35,20:35]
+fd2.w3 <- friend.data.w3[20:35,20:35]
+# read in covariate data
+drink <- as.matrix(read.table("../data/s50_data/s50-alcohol.dat"))
+drink2 <- drink[20:35,]
+
+
+# siena data
+library(RSiena)
+friendData2 <- array( c( fd2.w1, fd2.w2, fd2.w3 ),
+                         dim = c( 16, 16, 3 ) )
+# and next give this the role of the dependent variable:
[…]

### 2017-03-14 paper.Rnw
most recent updates.

@@
 \begin{itemize}
-\item What? - model visualization and SAOMs. briefly dsecribe what these things are
-\item Why? - bring light to the RSiena blackbox. Need ways to visualize network models, collections of networks. 
+%\item What? - model visualization and SAOMs. briefly dsecribe what these things are
+%\item Why? - bring light to the RSiena blackbox. Need ways to visualize network models, collections of networks. 
 \item How?
 \end{itemize}
-also talk about other dynamic network models briefly to put SAOMs in context.}
+%also talk about other dynamic network models briefly to put SAOMs in context.
+}
 
@@
 
-\par The software SIENA, and it's \texttt{R} implementation \texttt{RSiena}, was developed by \citet{RSiena} to estimate the parameters of the SAOMs, among other social network models.
+\par The software SIENA, and its \texttt{R} implementation \texttt{RSiena}, was developed by \citet{RSiena} to estimate the parameters of the SAOMs, among other social network models.
 This software is a huge contribution to the field of social network analysis, and when we began working with it we were eager to understand how it works.
@@
 Again, in these steps are largely hidden from the researcher computing them with software.
-We hope that by bringing these background elements and others into the foreground, stochastic actor-oriented models will be better understood and more accessible to social network analysts.
+We hope that by bringing these background elements and others into the foregroundwith static and dynamic visualizations, stochastic actor-oriented models will be better understood and more accessible to social network analysts.
 
@@
 
-\hh{mostly lit review plus a detailed description of the model and fitting process.}
+%\hh{mostly lit review plus a detailed description of the model and fitting process.}
 

### 2017-03-15 paper.Rnw
dorying it up over here.

@@
 require(knitr)
-opts_chunk$set(fig.path='figure/', cache.path='cache/', fig.align='center', fig.width=5, fig.height=5, par=TRUE, cache=TRUE, concordance=TRUE, autodep=TRUE, message=F, warning=F, echo = FALSE)
+opts_chunk$set(fig.path='figure/', cache.path='cache/', fig.align='center', fig.width=5, fig.height=5, par=TRUE, cache=TRUE, concordance=TRUE, autodep=TRUE, message=F, warning=F, echo = FALSE, root.dir = "~/Desktop/Dissertation - LETS GO/SAOM-removing-blindfold/")
 
@@
 library(geomnet)
-setwd("code")
-source("00e_small_friends.R")
+source("code/00e_small_friends.R")
 view_rs2$wave <- factor(view_rs2$wave)
@@
 f_i(x)^{M2} & = \beta_1s_{i1} + \beta_2s_{i2} + \beta_3s_{i3} \\
-f_i(x)^{M3} & = \beta_1s_{i1} + \beta_2s_{i2} + \beta_4s_{i4} 
-\end{align*}, where $s_{i3} = \sum_{j \neq h} x_{ij}x_{ih}x_{hj} \mathbb{I}(z_i = z_h \neq z_j)$, and $s_{i4} =  |\{j : x_{ij} = 0, \sum\limits_h x_{ih}x_{hj} \geq 2\}|$.
+f_i(x)^{M3} & = \beta_1s_{i1} + \beta_2s_{i2} + \beta_4s_{i4}, 
+\end{align*} where $s_{i3} = \sum_{j \neq h} x_{ij}x_{ih}x_{hj} \mathbb{I}(z_i = z_h \neq z_j)$, and $s_{i4} =  |\{j : x_{ij} = 0, \sum\limits_h x_{ih}x_{hj} \geq 2\}|$.
 These statistics are known as the number of jumping transitive triplets and the number of doubly achieved distances two effect, respectively.
@@
 \end{figure}
-We fit this model using \texttt{RSiena} 1,000 times, then averaged the resulting parameter estimates to obtain mean estimates to use in the simulation of networks from this model.
-These values are shown in Table~\ref{tab:M1ests}
+We these three models using \texttt{RSiena} 1,000 times each.
+We then looked at the distribution of the fitted values, which are shown in Figure~\ref{fig:distests}.
+We can see from these distributions that the inclusion of the jumping transitive triplet parameter, $\beta_3$ is obviously affecting the distributions of the other four parameters included in all models, $\alpha_1$, $\alpha_2$, $\beta_1$, and $\beta2$.
+In comparison with models M1 and M3, model M2 typically has higher estimates of the rate parameter, meaning that the inclusion of the covariate statistic in the model leads to higher estimates of the number of times, on average, a node gets to change its ties.
+This seems problematic: it is not clear that the addition of a parameter to the objective function \textit{should} effect the estimation of the rate parameters.
+
+To further investigate this relationship between the parameter values, we look at correlations between each of the parameter estimates in each model.
+In Figure~\ref{fig:corplots}, we examine correlations between each of pair of parameters within each model and overall.\nocite{ggally} The strongest correlation within each model is between $\beta_1$ and $\beta_2$, with absolute value of correlation between those two parameter values greater than 0.90 in all three models.
+The $\beta_1$ parameter is also highly correlated with the $\beta_3$ parameter within model M2, but it is not as highly correlated with the $\beta_4$ parameter in model M3.
 
 <<getdata, results='asis'>>=
-setwd("data")
-nulls <- read.csv("distribution_null_model.csv")
-alt <- read.csv("distribution_jumpTT_model.csv")
-alt2nd <- read.csv("distribution_dblpairs_model.csv")
-simu2 <- read.csv("simulation-1000-M1-M2-M3.csv")
+
+nulls <- read.csv("data/distribution_null_model.csv")
+alt <- read.csv("data/distribution_jumpTT_model.csv")
+alt2nd <- read.csv("data/distribution_dblpairs_model.csv")
+simu2 <- read.csv("data/simulation-1000-M1-M2-M3.csv")
 means <- simu2 %>% group_by(Model, parameter) %>% summarize(
-  mean = mean(estimate)
+  means = mean(estimate)
 )
@@
 
-To create a visualization that represents an ``average" network, we count occurences of each possible edge in the simulations, resulting in a summary network of simulations.
-Next, we subset the network to include only those edges which appear in greater than 50\% of all
+<<distests, fig.cap="Distribution of fitted parameter values for our three SAOMs.
+The inclusion of $\\beta_3$ or $\\beta_4$ clearly has an effect on the distribution of the rate parameters, $\\alpha_1$ and $\\alpha_2$.", fig.width=8>>=
+
+#simu2 <- read.csv("data/simulation-1000-M1-M2-M3.csv")
+
+ggplot(data = simu2) + 
+  geom_density(aes(x = estimate, fill = Model), alpha = .5) + 
+  facet_wrap(~parameter, scales = 'free') + 
+  theme_bw()
+@
+
+<<corplots, fig.cap="A matrix of plots demonstrating the strong correlations between parameter estimate in our SAOMs.
+The strongest correlation within each model is between $\\beta_1$ and $\\beta_2$.", fig.width=8>>=
+library(GGally)
+library(tidyr)
+simu2 <- read.csv("data/simulation-1000-M1-M2-M3.csv")
+
+simu_spread <- simu2 %>% spread(parameter, estimate)
+ggpairs(simu_spread, columns = 3:8, ggplot2::aes(colour=Model, alpha = .5)) + theme_bw()
+@
+
+We also explore simulations from these different models given parameter values.
+Using the means from the 1,000 fitted parameter values, we simulated 1,000 observations from each of our three models.
+We condition on the first friendship network observation, and the second and third observations are simulated from the SAOM models estimated previously.
+From these simulations, we first create a visualization that represents an ``average" network, we count occurences of each possible edge in the simulations, resulting in a summary network of simulations.
 
@@
 library(RSiena)
-library(dplyr)
+library(tidyverse)
 #friend.data.w1 <- as.matrix(read.table("s50_data/s50-network1.dat"))
@@
 #drink <- as.matrix(read.table("s50_data/s50-alcohol.dat"))
-friendshipData <- array(c(friend.data.w1, friend.data.w2, friend.data.w3), dim = c(50, 50, 3))
+friendshipData <- array(c(fd2.w1, fd2.w2, fd2.w3), dim = c(16, 16, 3))
 friendship <- sienaDependent(friendshipData)
-alcohol <- varCovar(drink)
+alcohol <- varCovar(drink[20:35,])
 mydata <- sienaDataCreate(friendship, alcohol)
@@
 data(M1ests_bigfriends)
-M1parms <- colMeans(M1ests_bigfriends)
+M1parms <- (means %>% filter(Model == "M1"))$means
 M1sims1000 <- saom_simulate(dat = mydata, struct = M1eff,
@@
 
+wave2drink <- drink2[,2]
+wave2drink <- as.data.frame(wave2drink)
+wave2drink$id <- paste0("V", 1:16)
+names(wave2drink)[1] <- "drink"
+M1avgcov <- merge(M1avg, wave2drink, by.x = "from", by.y = "id", all = T)
 library(tidyr)
@@
 distmat <- distmat[,-1]
-distmat <- as.matrix(1/distmat)
+distmat <- as.matrix(sqrt(1/distmat))
 distmat[is.na(distmat)] <- 0
-ggplot(data = M1avg %>% filter(weight/1000 > .5)) + # edge appears > 50% of the time. (is this actually what that is????) 
-  geom_net(aes(from_id = from, to_id = to, linewidth = weight/1000), 
+ggplot(data = M1avgcov) + # edge appears > 50% of the time. (is this actually what that is????) 
+  geom_net(aes(from_id = from, to_id = to, linewidth = weight/251, color = as.factor(drink)), 
            directed = TRUE, curvature = .2,
@@
            layout.par = list(elen = distmat)
-           )
-ggplot(data = M1avg2 %>% filter(weight/1000 > .5)) + # edge appears > 50% of the time. (is this actually what that is????) 
-  geom_net(aes(from_id = from, to_id = to, linewidth = weight/1000), 
-           directed = TRUE, curvature = .2, labelon = T,
+           ) + 
+  scale_color_brewer(palette = "YlOrRd", name = "Drinking\nBehavior") + 
+  scale_size_continuous(name = "Weight") + 
+  theme_net()
+M1avg2$wave <- M1avg2$wave + 1
+alldrink <- data.frame(drink2[,2:3])
+alldrink <- gather(alldrink, wave,drink) %>% mutate(wave = parse_number(wave), id = rep(paste0("V",1:16),2))
+
+M1avg2cov <- merge(M1avg2, alldrink, by.x = c("from", "wave"), by.y = c("id", "wave"), all = T)
+ggplot(data = M1avg2cov) + # edge appears > 50% of the time. (is this actually what that is????) 
+  geom_net(aes(from_id = from, to_id = to, linewidth = weight/388, color = as.factor(drink)), 
+           directed = TRUE, curvature = .2, labelon = F,
            #layout.alg = "fruchtermanreingold"
-           layout.alg = "kamadakawai",
+           layout.alg = "kamadakawai", fiteach = T,
            layout.par = list(elen = distmat)
            ) + 
-  facet_wrap(~wave) + theme_net()
+  scale_color_brewer(palette = "YlOrRd", name = "Drinking\nBehavior") + 
+  facet_wrap(~wave, labeller = "label_both") + theme_net() + labs(title = "Average Wave 2 and Wave 3 Networks from Model M1")
 # ggplot(data = M1avg %>% filter(weight >= 51)) + 

### 2017-03-15 code/00e_small_friends.R

@@ -6,9 +6,9 @@ library(dplyr)
 
 library(RSiena) # or RSienaTest
 # read in network data
-friend.data.w1 <- as.matrix(read.table("../data/s50_data/s50-network1.dat"))
-friend.data.w2 <- as.matrix(read.table("../data/s50_data/s50-network2.dat"))
-friend.data.w3 <- as.matrix(read.table("../data/s50_data/s50-network3.dat"))
+friend.data.w1 <- as.matrix(read.table("data/s50_data/s50-network1.dat"))
+friend.data.w2 <- as.matrix(read.table("data/s50_data/s50-network2.dat"))
+friend.data.w3 <- as.matrix(read.table("data/s50_data/s50-network3.dat"))
 
 val1 <- data.frame(friend.data.w1) %>% gather(x,y, V1:V50) %>% select(y)
 val2 <- data.frame(friend.data.w2) %>% gather(x,y, V1:V50) %>% select(y)
@@ -34,7 +34,7 @@ fd2.w1 <- friend.data.w1[20:35,20:35]
 fd2.w2 <- friend.data.w2[20:35,20:35]
 fd2.w3 <- friend.data.w3[20:35,20:35]
 # read in covariate data
-drink <- as.matrix(read.table("../data/s50_data/s50-alcohol.dat"))
+drink <- as.matrix(read.table("data/s50_data/s50-alcohol.dat"))
 drink2 <- drink[20:35,]
 
 

### 2017-03-15 paper.Rnw
am work

@@
 alldat$X2 <- ifelse(!is.na(alldat$X2), paste0("V", alldat$X2), NA)
-ggplot(data = alldat, 
+alldataviz <- ggplot(data = alldat, 
        aes(from_id = X1, to_id = X2, color = as.ordered(drink), group = as.ordered(drink))) +
@@
   theme(legend.position = 'bottom', panel.background = element_rect(fill = NA, color = "grey30"))
+alldataviz
 @
@@
 If there are no changes at a particular microstep, no changes are seen.
-Some frames of the animations are shown in INSERT STATIC FIGURE, and the full movie can be viewed at INSERT LINK TO GIF.
+Some frames of the animations are shown in INSERT STATIC FIGURE OF THE 5-6 DIFFERENT MICROSTEPS AND ONE TRANSITION SET, and the full movie can be viewed at INSERT LINK TO GIF.
 
@@
 We condition on the first friendship network observation, and the second and third observations are simulated from the SAOM models estimated previously.
-From these simulations, we first create a visualization that represents an ``average" network, we count occurences of each possible edge in the simulations, resulting in a summary network of simulations.
+From these simulations, we first create a visualization that represents an ``average" network.
+To do this, we count occurences of each possible edge in the simulations, resulting in a summary network with weighted edges representing the number of times an edge appeared when simulating from the SAOM.
 
@@
 M1eff <- getEffects(mydata)
+M2eff <- includeEffects(M1eff, jumpXTransTrip, interaction1 = "alcohol")
+M3eff <- includeEffects(M1eff, nbrDist2twice)
 library(netvizinf)
@@
 M1parms <- (means %>% filter(Model == "M1"))$means
-M1sims1000 <- saom_simulate(dat = mydata, struct = M1eff,
-                       parms = as.numeric(M1parms), N = 1000)
+M2parms <- (means %>% filter(Model == "M2"))$means
+M3parms <- (means %>% filter(Model == "M3"))$means
+M1sims1000 <- saom_simulate(dat = mydata, 
+                            struct = M1eff,
+                       parms = as.numeric(M1parms), 
+                       N = 1000)
+M2sims1000 <- saom_simulate(dat = mydata, 
+                            struct = M2eff,
+                       parms = as.numeric(M2parms), 
+                       N = 1000)
+M3sims1000 <- saom_simulate(dat = mydata, 
+                            struct = M3eff,
+                       parms = as.numeric(M3parms), 
+                       N = 1000)
 M1simsdf <- sims_to_df(M1sims1000)
+M2simsdf <- sims_to_df(M2sims1000)
+M3simsdf <- sims_to_df(M3sims1000)
 M1avg <- M1simsdf %>% filter(!is.na(to) & wave == 1) %>%
@@
   mutate(weight = ifelse(from == to, 0, count))
+M1avgW2 <- M1simsdf %>% filter(!is.na(to) & wave == 2) %>%
+  group_by(from, to) %>% 
+  summarise(count = n()) %>% 
+  mutate(weight = ifelse(from == to, 0, count))
 M1avg2 <- M1simsdf %>% filter(!is.na(to)) %>%
+ group_by(from, to, wave) %>%
+ summarise(count = n()) %>%
+ mutate(weight = ifelse(from == to, 0, count))
+M2avg <- M2simsdf %>% filter(!is.na(to) & wave == 1) %>%
+  group_by(from, to) %>% 
+  summarise(count = n()) %>% 
+  mutate(weight = ifelse(from == to, 0, count))
+M2avgW2 <- M2simsdf %>% filter(!is.na(to) & wave == 2) %>%
+  group_by(from, to) %>% 
+  summarise(count = n()) %>% 
+  mutate(weight = ifelse(from == to, 0, count))
+M2avg2 <- M2simsdf %>% filter(!is.na(to)) %>%
+ group_by(from, to, wave) %>%
+ summarise(count = n()) %>%
+ mutate(weight = ifelse(from == to, 0, count))
+M3avg <- M3simsdf %>% filter(!is.na(to) & wave == 1) %>%
+  group_by(from, to) %>% 
+  summarise(count = n()) %>% 
+  mutate(weight = ifelse(from == to, 0, count))
+M3avgW2 <- M3simsdf %>% filter(!is.na(to) & wave == 2) %>%
+  group_by(from, to) %>% 
+  summarise(count = n()) %>% 
+  mutate(weight = ifelse(from == to, 0, count))
+M3avg2 <- M3simsdf %>% filter(!is.na(to)) %>%
  group_by(from, to, wave) %>%
@@
 M1avg2$to <- paste0("V", M1avg2$to)
+M1avgW2$from <- paste0("V", M1avgW2$from)
+M1avgW2$to <- paste0("V", M1avgW2$to)
+M2avg$from <- paste0("V", M2avg$from)
+M2avg$to <- paste0("V", M2avg$to)
+M2avgW2$from <- paste0("V", M2avgW2$from)
+M2avgW2$to <- paste0("V", M2avgW2$to)
+M2avg2$from <- paste0("V", M2avg2$from)
+M2avg2$to <- paste0("V", M2avg2$to)
+M3avg$from <- paste0("V", M3avg$from)
+M3avg$to <- paste0("V", M3avg$to)
+M3avgW2$from <- paste0("V", M3avgW2$from)
+M3avgW2$to <- paste0("V", M3avgW2$to)
+M3avg2$from <- paste0("V", M3avg2$from)
+M3avg2$to <- paste0("V", M3avg2$to)
 
@@
 names(wave2drink)[1] <- "drink"
+wave3drink <- drink2[,3]
+wave3drink <- as.data.frame(wave3drink)
+wave3drink$id <- paste0("V", 1:16)
+names(wave3drink)[1] <- "drink"
 M1avgcov <- merge(M1avg, wave2drink, by.x = "from", by.y = "id", all = T)
+M2avgcov <- merge(M2avg, wave2drink, by.x = "from", by.y = "id", all = T)
+M3avgcov <- merge(M3avg, wave2drink, by.x = "from", by.y = "id", all = T)
+M1avgW2cov <- merge(M1avgW2, wave3drink, by.x = "from", by.y = "id", all = T)
+M2avgW2cov <- merge(M2avgW2, wave3drink, by.x = "from", by.y = "id", all = T)
+M3avgW2cov <- merge(M3avgW2, wave3drink, by.x = "from", by.y = "id", all = T)
 library(tidyr)
-M1avg[,-3] %>% spread(to, weight) %>% data.frame-> distmat
-rownames(distmat) <- distmat$from
-distmat <- distmat[,-1]
-distmat <- as.matrix(sqrt(1/distmat))
-distmat[is.na(distmat)] <- 0
-ggplot(data = M1avgcov) + # edge appears > 50% of the time. (is this actually what that is????) 
+M1avg[,-3] %>% spread(to, weight) %>% data.frame-> distmatM1
+rownames(distmatM1) <- distmatM1$from
+distmatM1 <- distmatM1[,-1]
+distmatM1 <- as.matrix(sqrt(1/distmatM1))
+distmatM1[is.na(distmatM1)] <- 0
+M1avgW2[,-3] %>% spread(to, weight) %>% data.frame-> distmatM1W2
+rownames(distmatM1W2) <- distmatM1W2$from
+distmatM1W2 <- distmatM1W2[,-1]
+distmatM1W2 <- as.matrix(sqrt(1/distmatM1W2))
+distmatM1W2[is.na(distmatM1W2)] <- 0
+M2avg[,-3] %>% spread(to, weight) %>% data.frame-> distmatM2
+rownames(distmatM2) <- distmatM2$from
+distmatM2 <- distmatM2[,-1]
+distmatM2 <- as.matrix(sqrt(1/distmatM2))
+distmatM2[is.na(distmatM2)] <- 0
+M2avgW2[,-3] %>% spread(to, weight) %>% data.frame-> distmatM2W2
+rownames(distmatM2W2) <- distmatM2W2$from
+distmatM2W2 <- distmatM2W2[,-1]
+distmatM2W2 <- as.matrix(sqrt(1/distmatM2W2))
+distmatM2W2[is.na(distmatM2W2)] <- 0
+M3avg[,-3] %>% spread(to, weight) %>% data.frame-> distmatM3
+rownames(distmatM3) <- distmatM3$from
+distmatM3 <- distmatM3[,-1]
+distmatM3 <- as.matrix(sqrt(1/distmatM3))
+distmatM3[is.na(distmatM3)] <- 0
+M3avgW2[,-3] %>% spread(to, weight) %>% data.frame-> distmatM3W2
+rownames(distmatM3W2) <- distmatM3W2$from
+distmatM3W2 <- distmatM3W2[,-1]
+distmatM3W2 <- as.matrix(sqrt(1/distmatM3W2))
+distmatM3W2[is.na(distmatM3W2)] <- 0
+
+library(RColorBrewer)
+cols <- brewer.pal(n = 4, "YlOrRd")
+
+plotM1W2 <- ggplot(data = M1avgcov) + 
   geom_net(aes(from_id = from, to_id = to, linewidth = weight/251, color = as.factor(drink)), 
+           arrowgap = .03, size = 3,arrowsize = .5,
            directed = TRUE, curvature = .2,
@@
            layout.alg = "kamadakawai",
-           layout.par = list(elen = distmat)
+           layout.par = list(elen = distmatM1)
            ) + 
   scale_color_brewer(palette = "YlOrRd", name = "Drinking\nBehavior") + 
+  #labs(title = "Wave 2, M1") + 
   scale_size_continuous(name = "Weight") + 
-  theme_net()
-M1avg2$wave <- M1avg2$wave + 1
-alldrink <- data.frame(drink2[,2:3])
-alldrink <- gather(alldrink, wave,drink) %>% mutate(wave = parse_number(wave), id = rep(paste0("V",1:16),2))
-
-M1avg2cov <- merge(M1avg2, alldrink, by.x = c("from", "wave"), by.y = c("id", "wave"), all = T)
-ggplot(data = M1avg2cov) + # edge appears > 50% of the time. (is this actually what that is????) 
-  geom_net(aes(from_id = from, to_id = to, linewidth = weight/388, color = as.factor(drink)), 
-           directed = TRUE, curvature = .2, labelon = F,
+  theme_net() + 
+  theme(plot.title = element_text(hjust = .5), legend.position = 'none',plot.margin=unit(c(0,0,0,0),"mm"))
+plotM1W3 <- ggplot(data = M1avgW2cov) + 
+  geom_net(aes(from_id = from, to_id = to, linewidth = weight/388.5, color = as.factor(drink)), 
+           arrowgap = .03, size = 3,arrowsize = .5,
+           directed = TRUE, curvature = .2,
            #layout.alg = "fruchtermanreingold"
-           layout.alg = "kamadakawai", fiteach = T,
-           layout.par = list(elen = distmat)
+           layout.alg = "kamadakawai",
+           layout.par = list(elen = distmatM1W2)
+           ) + 
+  scale_color_manual(values = cols[2:4], name = "Drinking\nBehavior") + 
+  #labs(title = "Wave 3, M1") + 
+  scale_size_continuous(name = "Weight") + 
+  theme_net() + 
+  theme(plot.title = element_text(hjust = .5), legend.position = 'none',plot.margin=unit(c(0,0,0,0),"mm"))
+plotM2W2 <- ggplot(data = M2avgcov) + 
+  geom_net(aes(from_id = from, to_id = to, linewidth = weight/325, color = as.factor(drink)), 
+           arrowgap = .03, size = 3,arrowsize = .5,
+           directed = TRUE, curvature = .2,
+           #layout.alg = "fruchtermanreingold"
+           layout.alg = "kamadakawai",
+           layout.par = list(elen = distmatM2)
            ) + 
   scale_color_brewer(palette = "YlOrRd", name = "Drinking\nBehavior") + 
-  facet_wrap(~wave, labeller = "label_both") + theme_net() + labs(title = "Average Wave 2 and Wave 3 Networks from Model M1")
+  #labs(title = "Wave 2, M2") + 
+  scale_size_continuous(name = "Weight") + 
+  theme_net() + 
+  theme(plot.title = element_text(hjust = .5), legend.position = 'none',plot.margin=unit(c(0,0,0,0),"mm"))
+plotM2W3 <- ggplot(data = M2avgW2cov) + 
+  geom_net(aes(from_id = from, to_id = to, linewidth = weight/357, color = as.factor(drink)), 
+           arrowgap = .03, size = 3,arrowsize = .5,
+           directed = TRUE, curvature = .2,
+           #layout.alg = "fruchtermanreingold"
+           layout.alg = "kamadakawai",
+           layout.par = list(elen = distmatM2W2)
+           ) + 
+  scale_color_manual(values = cols[2:4], name = "Drinking\nBehavior") + 
+  #labs(title = "Wave 3, M2") + 
+  scale_size_continuous(name = "Weight") + 
+  theme_net() + 
+  theme(plot.title = element_text(hjust = .5), legend.position = 'none',plot.margin=unit(c(0,0,0,0),"mm"))
+plotM3W2 <- ggplot(data = M3avgcov) + 
+  geom_net(aes(from_id = from, to_id = to, linewidth = weight/258, color = as.factor(drink)), 
+           arrowgap = .03, size = 3,arrowsize = .5,
+           directed = TRUE, curvature = .2,
+           #layout.alg = "fruchtermanreingold"
+           layout.alg = "kamadakawai",
+           layout.par = list(elen = distmatM3)
+           ) + 
+  scale_color_brewer(palette = "YlOrRd", name = "Drinking\nBehavior") + 
+  #labs(title = "Wave 2, M3") + 
+  scale_size_continuous(name = "Weight") + 
+  theme_net() + 
+  theme(plot.title = element_text(hjust = .5), legend.position = 'none',plot.margin=unit(c(0,0,0,0),"mm"))
+plotM3W3 <- ggplot(data = M3avgW2cov) + 
+  geom_net(aes(from_id = from, to_id = to, linewidth = weight/372.5, color = as.factor(drink)), 
+           arrowgap = .03, size = 3,arrowsize = .5,
+           directed = TRUE, curvature = .2,
+           #layout.alg = "fruchtermanreingold"
+           layout.alg = "kamadakawai",
+           layout.par = list(elen = distmatM3W2)
+           ) + 
+  scale_color_manual(values = cols[2:4], name = "Drinking\nBehavior") + 
+  #labs(title = "Wave 3, M3") + 
+  scale_size_continuous(name = "Weight") + 
+  theme_net() + 
+  theme(plot.title = element_text(hjust = .5), legend.position = 'none',plot.margin=unit(c(0,0,0,0),"mm"))
+
+origW1 <- ggplot(data = alldat %>% filter(Wave==1), 
+       aes(from_id = X1, to_id = X2, color = as.ordered(drink))) +
+  geom_net(fiteach = FALSE, directed=TRUE, layout.alg = "kamadakawai", size = 3, arrowsize = .5, curvature = .2,
+           arrowgap = .03) +
+  scale_color_brewer(palette = "YlOrRd", name = "Drinking\nBehavior") + 
+  theme_net() +
+  theme(legend.position = 'none',plot.margin=unit(c(0,0,0,0),"mm"))
+origW2 <- ggplot(data = alldat %>% filter(Wave==2), 
+       aes(from_id = X1, to_id = X2, color = as.ordered(drink))) +
+  geom_net(fiteach = FALSE, directed=TRUE, layout.alg = "kamadakawai", size = 3, arrowsize = .5, curvature = .2,
+           arrowgap = .03) +
+  scale_color_brewer(palette = "YlOrRd", name = "Drinking\nBehavior") + 
+  theme_net() +
+  theme(legend.position = 'none',plot.margin=unit(c(0,0,0,0),"mm"))
+origW3 <- ggplot(data = alldat %>% filter(Wave==3), 
+       aes(from_id = X1, to_id = X2, color = as.ordered(drink))) +
+  geom_net(fiteach = FALSE, directed=TRUE, layout.alg = "kamadakawai", size = 3, arrowsize = .5, curvature = .2,
+           arrowgap = .03) +
+  scale_color_manual(values = cols[2:4], name = "Drinking\nBehavior") + 
+  theme_net() +
+  theme(legend.position = 'none',plot.margin=unit(c(0,0,0,0),"mm"))
+
+library(gridExtra)
+library(cowplot)
+plot_grid(plot.mpg, plot.diamonds, labels = c("A", "B"), nrow = 2, align = "v")
+grid.arrange(origW2, plotM1W2,
+              plotM2W2,plotM3W2, 
+              ncol=2, clip = T)
+alldataviz2 <- alldataviz + theme(legend.position = 'none')
+ggdraw() +
+  draw_plot(alldataviz2, x=0, y=.75, width=1, height=.25) + 
+  draw_plot(plotM1W2, .1, .5, .5, .25) +
+  draw_plot(plotM1W3, .3, .5, .5, .25) +
+  draw_plot(plotM2W2, .1, .25, .5, .25) +
+  draw_plot(plotM2W3, .3, .25, .5, .25) +
+  draw_plot(plotM3W2, .1, 0, .5, .25) +
+  draw_plot(plotM3W3, .3, 0, .5, .25) + 
+  draw_plot_label(label = c("Original Data", "M1: Wave 2", "M1: Wave 3", "M2: Wave 2", "M2: Wave 3", "M3: Wave 2", "M3: Wave 3"))
+
+M1avg2$Model <- "M1"
+M2avg2$Model <- "M2"
+M3avg2$Model <- "M3"
+allavg <- rbind(as.data.frame(M1avg2), 
+                as.data.frame(M2avg2),
+                as.data.frame(M3avg2))
+
+ggplot(data = allavg) + 
+  geom_histogram(aes(x = wt)) + 
+  facet_grid(wave~Model) + 
+  labs(x = "Number of times an edge occurs in 1,000 simulations") +
+  theme_bw()
+
+# ggplot(data = M3avgcov) + 
+#   geom_net(aes(from_id = from, to_id = to, linewidth = weight/251, color = as.factor(drink)), 
+#            directed = TRUE, curvature = .2,
+#            ealpha = .75,
+#            layout.alg = "kamadakawai",
+#            layout.par = list(elen = distmatM2)
+#            ) + 
+#   scale_color_brewer(palette = "YlOrRd", name = "Drinking\nBehavior") + 
+#   scale_size_continuous(name = "Weight") + 
+#   theme_net()
+# M1avg2$wave <- M1avg2$wave + 1
+# alldrink <- data.frame(drink2[,2:3])
+# alldrink <- gather(alldrink, wave,drink) %>% mutate(wave = parse_number(wave), id = rep(paste0("V",1:16),2))
+
+# M1avg2cov <- merge(M1avg2, alldrink, by.x = c("from", "wave"), by.y = c("id", "wave"), all = T)
+# ggplot(data = M1avg2cov) + # edge appears > 50% of the time. (is this actually what that is????) 
+#   geom_net(aes(from_id = from, to_id = to, linewidth = weight/388, color = as.factor(drink)), 
+#            directed = TRUE, curvature = .2, labelon = F,
+#            #layout.alg = "fruchtermanreingold"
+#            layout.alg = "kamadakawai", fiteach = T,
+#            layout.par = list(elen = distmat)
+#            ) + 
+#   scale_color_brewer(palette = "YlOrRd", name = "Drinking\nBehavior") + 
+#   facet_wrap(~wave, labeller = "label_both") + theme_net() + labs(title = "Average Wave 2 and Wave 3 Networks from Model M1")
 # ggplot(data = M1avg %>% filter(weight >= 51)) + 

### 2017-03-27 paper.Rnw
some bg on net viz

@@
 Again, in these steps are largely hidden from the researcher computing them with software.
-We hope that by bringing these background elements and others into the foregroundwith static and dynamic visualizations, stochastic actor-oriented models will be better understood and more accessible to social network analysts.
+We hope that by bringing these background elements and others into the foreground with static and dynamic visualizations, stochastic actor-oriented models will be better understood and more accessible to social network analysts.
+
+\section{Methods for Network Visualization}
+
+Network visualization, also called network mapping, is a prominent subfield of network analysis.
+Visualizing network data is uniquely difficult because of the structure of the data itself.
+Most visualizations rely on a well-defined axis inherited from the data.
+If data are numerical, histograms, scatterplots, or time series plots are straightforward to construct.
+If the data are categorical, bar charts and mosaic plots can be constructed easily.
+If the data are spatial, there is a well-defined space in which to view information.
+Network data are much less cut-and-dried.
+
+There are two primary methods used to visualize networks.
+The first method represents nodes with points in 2D Euclidean space and then represents edges by connecting the points with lines when there is an edge between the two nodes.
+But, because there is no natural placement of the points, a random placement is used, then adjusted iteratively via a layout algorithm, of which there are many.
+
+Some layout algorithms were designed to mimic physical systems, drawing the graphs based on the "forces" connecting them.
+The network's edges act as springs pushing and pulling the nodes in 2D space.
+Others use properties of the network's adjaceny matrix in order to place the nodes.
+Some commonly used layout algorithms are  XX ADD MORE DETAIL LATER XX
+\begin{itemize}
+\item Kamada-Kawai
+\item Fructerman-Reingold
+\item Multi-dimensional scaling
+\item Eigen
+\end{itemize}
+
+The second primary method for network visualization uses the adjacency matrix of the network.
+This method is particularly useful when the network is very complex, dense, or large.
+XX MORE STUFF ABOUT ADJ MAT VIZ XX
 

### 2017-03-29 paper.Rnw
forgot to push previous changes

@@
 This software is a huge contribution to the field of social network analysis, and when we began working with it we were eager to understand how it works.
-To better understand the \texttt{RSiena} software and the SAOMs themselves, we use the principals of model visualization from \citet{modelvis}.
-With the work of \citet{modelvis} guiding us, we explore RSiena and SAOMs with innovative visualizations, both static and dynamic.
+To better understand the \texttt{RSiena} software and the SAOMs themselves, we combine the principals of network visualization with those of model visualization from \citet{modelvis}.
+With the work of \citet{modelvis} guiding us, we explore RSiena and SAOMs with innovative network visualizations, both static and dynamic.
 By bringing some light to the underlying methods and structures "behind the scenes" of SAOMs and RSiena, we will help other researchers working with the models better understand and better use these models.
@@
 There are two primary methods used to visualize networks.
-The first method represents nodes with points in 2D Euclidean space and then represents edges by connecting the points with lines when there is an edge between the two nodes.
+The first method, often referred to as node-link diagrams, represents nodes with points in 2D Euclidean space and then represents edges by connecting the points with lines when there is an edge between the two nodes.
 But, because there is no natural placement of the points, a random placement is used, then adjusted iteratively via a layout algorithm, of which there are many.
@@
 <<getdata, results='asis'>>=
-
 nulls <- read.csv("data/distribution_null_model.csv")

### 2017-03-29 paper.Rnw
add ref to dyn net viz

@@
 \citeauthor{modelvis} outline the three different parts of a model: the model family, the model form, and the fitted model.
-The latter is primarily what one thinks of first when considering a model in a data analysis, where a model is fit to the data, and parameter estimates reported.
+The latter is primarily what one thinks of first when considering a model in a data analysis, where a model is fit to data, and parameter estimates are reported.
 In the context of SAOMs, the fitted model contains the estimated rate parameters and the estimated objective function parameters.
@@
 
-A key component of each step of the SIENA method of moments algorithm is the ``microstep." This step simulates from the model in its current state, $x(t_m)$ with current parameter values $\theta_{0}$, to the next state, $x(t_{m+1})$, or as close as it can get to that state.
+A key component of each step of the SIENA method of moments algorithm is the ``microstep." This step simulates from the model in its current state, $x(t_m)$ with current parameter values $\theta_{0}$, to the next state, $x(t_{m+1})$.
+This microstep process stops when the simulated network has acheived the same number of differences from $x(t_m)$ as $x(t_{m+1})$.
 This simulation process follows the steps of the continuous-time Markov chain.
@@
 Saving all of these steps is not efficient if you're only interested in parameter values, but they can be saved and extracted using options in \texttt{RSiena}, which is exactly what we did to create our visualizations.
-Between two network observations $x(t_m)$ and $x(t_{m+1})$, there can be hundreds or even thousands of microsteps.
+Between two network observations $x(t_m)$ and $x(t_{m+1})$, there can be dozens, hundreds, or even thousands of microsteps.
 We wanted to see these in-between steps in order to better understand the behavior of the underlying continuous-time Markov chain.
@@
 The first vizualization we present here is an animation of microsteps simulated to emulate the transition steps from wave 1 to wave 2 the subset of the friends data shown in Figure ~\ref{fig:alldataviz}.
+Movies similar to our animation were used to visualize the changes of dynamic networks in \citet{dynnetviz}.
 When each ego node is selected in a microstep, it is emphasized in the animation, then the associated edge either appears or disappears.

### 2017-03-30 paper.Rnw
some adj mat stuff

@@
 A key component of each step of the SIENA method of moments algorithm is the ``microstep." This step simulates from the model in its current state, $x(t_m)$ with current parameter values $\theta_{0}$, to the next state, $x(t_{m+1})$.
-This microstep process stops when the simulated network has acheived the same number of differences from $x(t_m)$ as $x(t_{m+1})$.
+This microstep process stops when the simulated network has acheived the same number of differences, $C$, from $x(t_m)$ as $x(t_{m+1})$, where
+$$C = \sum_{i \neq j} |x_{ij}(t_{m+1}) - x_{ij}(t_{m})|.$$ 
 This simulation process follows the steps of the continuous-time Markov chain.
@@
 Some frames of the animations are shown in INSERT STATIC FIGURE OF THE 5-6 DIFFERENT MICROSTEPS AND ONE TRANSITION SET, and the full movie can be viewed at INSERT LINK TO GIF.
+
+XXX HH's code XXX
+
+INSERT DESCRIPTION OF THE MOVIE - WHAT IT ALLOWS US TO SEE, ETC.
+
+We also use animation to view the changing structure of the adjacency matrix the microsteps.
+The adjacency matrices for the three waves of friendship data as shown in Figure~\ref{fig:alldataviz} are ordered by node id.
+There are 50 nodes in the data, numbered 1-50, and that order is used on the $x$ and $y$ axes for the matrix visualization.
+Viewing the adjacency matrices with this arbitrary ordering does not provide much information to the viewer about the underlying structure of the network.
+This lack of perceived structure would be exacerbated in an animation.
+So, we adjust the ordering so that the viewer can perceive more of the structure of the graph.
+This process is known as matrix seriation \citep{seriation}.
+
+To reorder the vertices for the matrix visualization, we first constructed a cumulative adjaceny matrix, $\mathbf{A}^{cum}$, for the series of microsteps simulating the network from $x(t_m)$ to $x(t_{m+1})$.
+A single entry in the cumulative adjacency matrix, $\mathbf{A}^{cum}_{ij}$, is the total number of times the edge from node $i$ to node $j$ appears in the network from the initial observation, $x(t_m) \equiv X(0)$ to the final reult of the last microstep, $X(C)$:
+$$ \mathbf{A}^{cum}_{ij} = \sum_{c = 0}^C X_{ij}(c). $$
+We then performed a principal component analysis on $\mathbf{A}^{cum}$, and used the values of the first principal component to order the vertices on the $x$ and $y$ axes for the adjacency matrix animation.
 

### 2017-04-03 paper.Rnw
add seriation ex.

@@
 The options for change are (1) removing a current tie, (2) adding a new tie, or (3) making no change at all.
-Saving all of these steps is not efficient if you're only interested in parameter values, but they can be saved and extracted using options in \texttt{RSiena}, which is exactly what we did to create our visualizations.
-Between two network observations $x(t_m)$ and $x(t_{m+1})$, there can be dozens, hundreds, or even thousands of microsteps.
+Saving all of these steps is not efficient if you're only interested in estimating parameter values, but they can be saved and extracted using options in \texttt{RSiena}, which is exactly what we did to create our visualizations.
+Between two network observations $x(t_m)$ and $x(t_{m+1})$, there can be dozens, hundreds, or even thousands of microsteps, depending on the size of the network.
 We wanted to see these in-between steps in order to better understand the behavior of the underlying continuous-time Markov chain.
@@
 
+<<makinggif>>=
+library(RSiena)
+library(tidyverse)
+library(sna)
+library(network)
+library(netvizinf)
+library(gganimate)
+friend.data.w1 <- as.matrix(read.table("../data/s50-network1.dat"))
+friend.data.w2 <- as.matrix(read.table("../data/s50-network2.dat"))
+friend.data.w3 <- as.matrix(read.table("../data/s50-network3.dat"))
+drink <- as.matrix(read.table("../data/s50-alcohol.dat"))
+fd2.w1 <- friend.data.w1[20:35,20:35]
+fd2.w2 <- friend.data.w2[20:35,20:35]
+fd2.w3 <- friend.data.w3[20:35,20:35]
+friendshipData <- array(c(fd2.w1, fd2.w2,fd2.w3), dim = c(16, 16, 3))
+
+# running Siena
+friendship <- sienaDependent(friendshipData)
+alcohol <- varCovar(drink[20:35,])
+mydata <- sienaDataCreate(friendship, alcohol)
+myeffnull <- getEffects(mydata)
+myalgorithm <- sienaAlgorithmCreate(projname = 's16_3')
+set.seed(823746)
+ansnull <- siena07(myalgorithm, data = mydata, effects = myeffnull,
+                   returnChains = TRUE, returnDataFrame = TRUE,
+                   returnDeps = TRUE, silent = TRUE, verbose = FALSE,
+                   batch = TRUE)
+ansnullchains <- get_chain_info(ansnull)
+ansnullchains %>%
+  filter(period == 1) %>%  #only look at chains from wave 1 to wave 2
+  group_by(rep) %>%
+  select(rep, from = ego, to = alter) %>%
+  mutate(val = as.numeric(!from == to),
+         from = paste0("V", parse_number(from)+1), # make the chains
+         to = paste0("V", parse_number(to)+1)) -> ansnullchainsw1w2
+library(geomnet)
+colnames(fd2.w1) <- paste0("V", 1:16)
+rownames(fd2.w1) <- paste0("V", 1:16)
+wave1friends <- fortify(as.adjmat(fd2.w1))
+ms1 <- listMicrosteps(dat = wave1friends,
+                      microsteps = filter(ansnullchainsw1w2, rep == 1))
+
+@
+
 XXX HH's code XXX
 
-INSERT DESCRIPTION OF THE MOVIE - WHAT IT ALLOWS US TO SEE, ETC.
+XX INSERT DESCRIPTION OF THE MOVIE - WHAT IT ALLOWS US TO SEE, ETC. XX
 
@@
 We then performed a principal component analysis on $\mathbf{A}^{cum}$, and used the values of the first principal component to order the vertices on the $x$ and $y$ axes for the adjacency matrix animation.
+For one such series of microsteps simulated by \texttt{RSiena}, we present the adjacency matrix ordered by the (arbitrary) vertex id alongside the seriated adjacency microsteps using the first principal component loading on the cumulative adjacency matrix, $A^{cum}$
+
+<<adjmatorder, fig.cap="On the left, the starting friendship network represented in adjacency matrix form, ordered by vertex id.
+On the right, the same adjacency matrix is presented after ordering the vertices by one repitition of the microstep simulation process from wave one to wave two.", fig.width=8>>=
+# first present the original wave 1 adj mat, then present the adj mat after being ordered by the 1st pc of the big adjmat A. 
+numordersf <- data.frame(fd2.w1) %>% 
+  mutate(from = rownames(fd2.w1)) %>% 
+  gather(key = to ,value = value, -from)
+pcaordersf <- numordersf
+
+numordersf$from <- factor(numordersf$from, levels = paste0("V", 1:16))
+numordersf$to <- factor(numordersf$to, levels = paste0("V", 1:16))
+
+p1 <- ggplot(data = numordersf, aes(x = from, y = to)) + geom_tile(color = 'black', aes(fill = as.factor(value))) + 
+  scale_fill_manual(values = c('white', 'black')) +
+  theme(legend.position = 'none', axis.text = element_text(size = 8)) + 
+  coord_fixed()
+
+sfmsall <- read_csv("data/smallfriendsmicrosteps.csv")
+
+sfmsall %>% 
+  group_by(from, to) %>% 
+  filter(rep == 1) %>% 
+  summarise(count = n()) -> sfmsall_summ
+sfmsall_summ$from <- factor(sfmsall_summ$from, levels = paste0('V', 1:16))
+sfmsall_summ$to <- factor(sfmsall_summ$to, levels = paste0('V', 1:16))
+sfmsall_summ %>% 
+  spread(to, count, fill = 0) %>% 
+  data.frame -> allmsrep1_adjmat
+allmsrep1_adjmat$V16 <- 0
+allmsrep1_adjmat <- allmsrep1_adjmat[,-1]
+allmsrep1_adjmat <- as.matrix(allmsrep1_adjmat)
+rownames(allmsrep1_adjmat) <- paste0("V", 1:16)
+
+pca1 <- prcomp(allmsrep1_adjmat)
+order1 <- names(sort(pca1$rotation[,1], decreasing = T))
+order2 <- names(sort(abs(pca1$rotation[,1]), decreasing = T))
+
+pcaordersf$from <- factor(pcaordersf$from, levels = order1)
+pcaordersf$to <- factor(pcaordersf$to, levels = order1)
+
+p2 <- ggplot(data = pcaordersf, aes(x = from, y = to)) + geom_tile(color = 'black', aes(fill = as.factor(value))) + 
+  scale_fill_manual(values = c('white', 'black')) +
+  theme(legend.position = 'none', axis.text = element_text(size = 8)) + 
+  coord_fixed()
+
+grid.arrange(p1,p2, nrow = 1)
+@
 

### 2017-04-04 paper.Rnw
add static selection of frames from animation

@@
 \noindent%
-{\it Keywords:} social network analysis, model visualization \hh{Other keywords?}
+{\it Keywords:} social network analysis, model visualization, dynamic networks, network visualization, network mapping, animation
 \vfill
@@
 opts_chunk$set(fig.path='figure/', cache.path='cache/', fig.align='center', fig.width=5, fig.height=5, par=TRUE, cache=TRUE, concordance=TRUE, autodep=TRUE, message=F, warning=F, echo = FALSE, root.dir = "~/Desktop/Dissertation - LETS GO/SAOM-removing-blindfold/")
+@
 
-library(dplyr)
+<<pkgs>>=
+library(tidyverse)
+library(RSiena)
+library(network)
+library(sna)
+library(geomnet)
+library(GGally)
+library(netvizinf)
+library(RColorBrewer)
+library(gridExtra)
+library(cowplot)
 @
@@
 The subset we will be using is outlined in red.", fig.width=8>>=
-library(network)
-library(sna)
-library(geomnet)
 source("code/00e_small_friends.R")
@@
 If there are no changes at a particular microstep, no changes are seen.
-Some frames of the animations are shown in INSERT STATIC FIGURE OF THE 5-6 DIFFERENT MICROSTEPS AND ONE TRANSITION SET, and the full movie can be viewed at INSERT LINK TO GIF.
+Some frames of the animations are shown in Figure~\ref{fig:makinggif}, and the full movie can be viewed at INSERT LINK TO GIF.
 
-<<makinggif>>=
-library(RSiena)
-library(tidyverse)
-library(sna)
-library(network)
-library(netvizinf)
-library(gganimate)
-friend.data.w1 <- as.matrix(read.table("../data/s50-network1.dat"))
-friend.data.w2 <- as.matrix(read.table("../data/s50-network2.dat"))
-friend.data.w3 <- as.matrix(read.table("../data/s50-network3.dat"))
-drink <- as.matrix(read.table("../data/s50-alcohol.dat"))
+<<makinggif, fig.cap='A selection of images in the microstep animation.
+The selected edges and nodes are emphasized by changing the size and the color, then when edges are removed, they fade out, shrinking in size, while the nodes change color and shrink to blend in with the other nodes.', fig.width=8>>=
+friend.data.w1 <- as.matrix(read.table("data/s50_data/s50-network1.dat"))
+friend.data.w2 <- as.matrix(read.table("data/s50_data/s50-network2.dat"))
+friend.data.w3 <- as.matrix(read.table("data/s50_data/s50-network3.dat"))
+drink <- as.matrix(read.table("data/s50_data/s50-alcohol.dat"))
 fd2.w1 <- friend.data.w1[20:35,20:35]
@@
          to = paste0("V", parse_number(to)+1)) -> ansnullchainsw1w2
-library(geomnet)
 colnames(fd2.w1) <- paste0("V", 1:16)
@@
                       microsteps = filter(ansnullchainsw1w2, rep == 1))
+microsteps <- ms1[1:9]
 
+microsteps_df <- getMicrostepsDF(microsteps)
+pte <- pretween_edges(microsteps = microsteps)
+ptv <- pretween_vertices(pte = pte, layoutparams = list(n = 16))
+
+# get tweened data for all microsteps
+pall <- tween_microsteps(ptv, pte, microsteps_df)
+
+static_pall <- filter(pall, .frame %in% (1:12 * 3-2))
+
+create_net_animate(static_pall) + 
+  facet_wrap(~.frame, nrow=3, labeller = 'label_both') + 
+  theme(panel.background = element_rect(color = 'black'))
 @
-
-XXX HH's code XXX
 
@@
 The adjacency matrices for the three waves of friendship data as shown in Figure~\ref{fig:alldataviz} are ordered by node id.
-There are 50 nodes in the data, numbered 1-50, and that order is used on the $x$ and $y$ axes for the matrix visualization.
+There are 16 nodes in the data, numbered 1-16, and that order is used on the $x$ and $y$ axes for the matrix visualization.
 Viewing the adjacency matrices with this arbitrary ordering does not provide much information to the viewer about the underlying structure of the network.
@@
 )
-library(dplyr)
 null_mod_sv <- as.numeric(nulls[,-c(1,6,7)] %>% summarise_each(funs(mean)))
@@
 The strongest correlation within each model is between $\\beta_1$ and $\\beta_2$.", fig.width=8>>=
-library(GGally)
-library(tidyr)
+
 simu2 <- read.csv("data/simulation-1000-M1-M2-M3.csv")
@@
 <<averagenet>>=
-library(RSiena)
-library(tidyverse)
 #friend.data.w1 <- as.matrix(read.table("s50_data/s50-network1.dat"))
@@
 M3eff <- includeEffects(M1eff, nbrDist2twice)
-library(netvizinf)
 data(M1ests_bigfriends)
@@
 
-library(geomnet)
 M1avg$from <- paste0("V", M1avg$from)
@@
 M3avgW2cov <- merge(M3avgW2, wave3drink, by.x = "from", by.y = "id", all = T)
-library(tidyr)
+
 M1avg[,-3] %>% spread(to, weight) %>% data.frame-> distmatM1
@@
 
-library(RColorBrewer)
 cols <- brewer.pal(n = 4, "YlOrRd")
@@
   theme_net() + 
-  theme(plot.title = element_text(hjust = .5), legend.position = 'none',plot.margin=unit(c(0,0,0,0),"mm"))
+  theme(plot.title = element_text(hjust = .5), legend.position = 'none',plot.margin=unit(c(0,0,0,0),"mm")) + 
+  labs(title = "Model 1 Simulated Wave 2")
 plotM1W3 <- ggplot(data = M1avgW2cov) + 
@@
   theme_net() + 
-  theme(plot.title = element_text(hjust = .5), legend.position = 'none',plot.margin=unit(c(0,0,0,0),"mm"))
+  theme(plot.title = element_text(hjust = .5), legend.position = 'none',plot.margin=unit(c(0,0,0,0),"mm")) + 
+  labs(title = "Model 2 Simulated Wave 2")
 plotM2W3 <- ggplot(data = M2avgW2cov) + 
@@
   theme_net() + 
-  theme(plot.title = element_text(hjust = .5), legend.position = 'none',plot.margin=unit(c(0,0,0,0),"mm"))
+  theme(plot.title = element_text(hjust = .5), legend.position = 'none',plot.margin=unit(c(0,0,0,0),"mm")) + 
+  labs(title = "Model 3 Simulated Wave 2")
 plotM3W3 <- ggplot(data = M3avgW2cov) + 
@@
   theme_net() +
-  theme(legend.position = 'none',plot.margin=unit(c(0,0,0,0),"mm"))
+  theme(legend.position = 'none',plot.margin=unit(c(0,0,0,0),"mm")) + 
+  labs(title = "Observed Wave 2")
 origW3 <- ggplot(data = alldat %>% filter(Wave==3), 
@@
 
-library(gridExtra)
-library(cowplot)
-plot_grid(plot.mpg, plot.diamonds, labels = c("A", "B"), nrow = 2, align = "v")
 grid.arrange(origW2, plotM1W2,
@@
 ggplot(data = allavg) + 
-  geom_histogram(aes(x = wt)) + 
-  facet_grid(wave~Model) + 
+  geom_histogram(aes(x = weight), binwidth = 25) + 
+  facet_grid(wave~Model, labeller = "label_both") + 
   labs(x = "Number of times an edge occurs in 1,000 simulations") +

### 2017-04-10 paper.Rnw
add some references and citations of layout algorithms.

@@
 Others use properties of the network's adjaceny matrix in order to place the nodes.
-Some commonly used layout algorithms are  XX ADD MORE DETAIL LATER XX
+Some commonly used layout algorithms are
 \begin{itemize}
-\item Kamada-Kawai
-\item Fructerman-Reingold
-\item Multi-dimensional scaling
+\item Kamada-Kawai: The ``graph theoretic distance'' from one node to another is represented in the 2D node placement.
+This network drawing method results in several nice properties such as ``symmetric drawings [and] a relatively small number of edge crossings'' \citep[p.
+15]{kamadakawai}.
+\item Fructerman-Reingold: This algorithm is a spring-embedder type where ``the vertices behave as [...] celestial bodies, exerting attractive and repulsive forces on one another,'' where the simple goal of drawing connected nodes near to each other \citep[p.
+1131]{fruchterman-reingold}.
+\item Multi-dimensional scaling:
 \item Eigen
@@
 grid.arrange(p1,p2, nrow = 1)
+@
+
+<<histogramalters, fig.width = 8, fig.height=11, fig.cap="Probabilities of">>=
+ansnullchains %>% group_by(period, rep) %>% 
+  mutate(ms = row_number()) %>% 
+  filter(ms == 1, period == 1) %>% 
+  mutate(alterProb = exp(logAlterProb),
+         ego = as.numeric(as.character(ego)),
+         alter = as.numeric(as.character(alter))) %>% 
+  ungroup() %>% 
+  group_by(ego, alter) %>% 
+  summarize(count = n(), alterProb = unique(alterProb)) %>% ungroup() %>% 
+  group_by(ego) %>% 
+  mutate(csum = cumsum(alterProb)) %>% 
+  arrange(ego, csum) -> cumprobs 
+cumprobs$ego <- factor(paste0("V",cumprobs$ego+1), levels = paste0("V", 1:16))
+
+cumprobs %>% ggplot() + geom_histogram(aes(x = alter, weight = alterProb), color = 'black', fill = 'white', binwidth = 1) + scale_x_continuous(labels = paste0("V",1:16), breaks = 0:15) + coord_flip() + facet_wrap(~ego, labeller = "label_both") + theme_gray()
+
+# cprobslist <- split(cumprobs, cumprobs$ego)
+# plots <- list()
+# for(i in 1:length(cprobslist)){
+#   cprobslist[[i]] %>%
+#   ggplot() + 
+#   geom_histogram(aes(x = as.numeric(as.character(alter)),
+#                      y = alterProb), stat = 'identity', fill = 'white', color = 'black') + 
+#   geom_line(aes(x = as.numeric(as.character(alter)), 
+#                 y = csum)) + 
+#   geom_point(aes(x = as.numeric(as.character(alter)), 
+#                 y = csum)) +
+#   scale_x_continuous(limits = c(-.5, 15), breaks = 0:15, labels = 0:15) +   
+#   geom_hline(yintercept = 1, color = 'red', linetype = 'dashed') + 
+#   labs(x = "alter", y = expression(p[ij]), title = paste("Ego node:", names(cprobslist)[i])) + # ylab(expression(p[ij])) + 
+#   theme_gray() + 
+#     theme(plot.title = element_text(size = 10), 
+#           axis.title = element_text(size = 8), 
+#           plot.margin = unit(rep(0,4),"pt")) -> p
+#   plots[[i]] <- p
+# }
+# grid.arrange(plots[[1]], plots[[2]], plots[[3]], plots[[4]],
+#              plots[[5]], plots[[6]], plots[[7]], plots[[8]],
+#              #plots[[9]], plots[[10]], plots[[11]], plots[[12]],
+#              #plots[[13]], plots[[14]], plots[[15]], plots[[16]], 
+#              ncol = 2, padding = unit(0, 'line'))
 @

### 2017-04-11 paper.Rnw
bit of sec 4.2

@@
 
-\section{Methods for Network Visualization}
+\section{Networks and their Visualizations}
+
+\subsection{Introduction to Network Structures}
+
+\subsection{Visualizing Network Data} \label{sec:netvizintro}
 
@@
 \item Multi-dimensional scaling: 
-\item Eigen
+\item Eigen:
 \end{itemize}
@@
 set.seed(823746)
-ansnull <- siena07(myalgorithm, data = mydata, effects = myeffnull,
-                   returnChains = TRUE, returnDataFrame = TRUE,
-                   returnDeps = TRUE, silent = TRUE, verbose = FALSE,
-                   batch = TRUE)
-ansnullchains <- get_chain_info(ansnull)
+load("data/ansnull.rda")
+ansnullchains <- get_chain_info(ansnullpaper)
 ansnullchains %>%
@@
 These three data pieces can be visualized together in various ways.
-The traditional network visualization uses one of many algorithms to layout the actors as points in 2D space, then draws segments connecting the points in 2D if the dyad between two nodes has value greater than zero, and draws nothing otherwise.
+The traditional node-link visualization uses one of many algorithms to layout the actors as points in 2D space, then draws segments connecting the points in 2D if the dyad between two nodes has value greater than zero, and draws nothing otherwise.
 The time aspect can be visualized by drawing each network observation in time and placing the observed timepoints side-by-side.
+These visual connections between the three data spaces are themselves part of the data space, because they can be viewed in different ways.
+Another way to connect the node and the edge data, as discussed in Section~\ref{sec:netvizintro}, is through the network's adjaceny matrix.
+The time dimension can also be incorporated with side-by-side adjacency matrix visualizations.
 \st{

### 2017-04-11 paper.Rnw
par about geomnet

@@
 The time aspect can be visualized by drawing each network observation in time and placing the observed timepoints side-by-side.
-These visual connections between the three data spaces are themselves part of the data space, because they can be viewed in different ways.
+
+One tool that can bring these three data spaces together in this way is the \texttt{R} package \texttt{geomnet} \citep{geomnet}.
+Different aesthesic aspects of each piece of the node-link diagram can be tied to the underlying node or edge data.
+The color, size, and shape of the points can be used to represent variables in the node data, and the color, size, and linetype of the links between points can be used to represent the edge variables.
+In a social network, node data might be age, gender, and occupation of the person in the network, and node data might be length of connection between two people, how the people first met (school, work, church, etc.), and how often they interact.
+Pulling all of this information together with \texttt{geomnet} allows the entire data space to be viewed at once.
+
+XXX INSERT AN EXAMPLE OF DATA WITH LOTS OF EDGE, NODE VARS OF INTEREST XXX
+
+These visual connections between the node, edge, and time data spaces are themselves part of the data space, because they can be viewed in different ways.
 Another way to connect the node and the edge data, as discussed in Section~\ref{sec:netvizintro}, is through the network's adjaceny matrix.

### 2017-04-12 paper.Rnw
finishing up sec 2. maybe need a simple viz example of each type.

@@
 Others use properties of the network's adjaceny matrix in order to place the nodes.
-Some commonly used layout algorithms are
-\begin{itemize}
-\item Kamada-Kawai: The ``graph theoretic distance'' from one node to another is represented in the 2D node placement.
-This network drawing method results in several nice properties such as ``symmetric drawings [and] a relatively small number of edge crossings'' \citep[p.
-15]{kamadakawai}.
-\item Fructerman-Reingold: This algorithm is a spring-embedder type where ``the vertices behave as [...] celestial bodies, exerting attractive and repulsive forces on one another,'' where the simple goal of drawing connected nodes near to each other \citep[p.
-1131]{fruchterman-reingold}.
-\item Multi-dimensional scaling: 
-\item Eigen:
-\end{itemize}
+Some commonly used layout algorithms that treat networks like physical systems are Kamada-Kawai \citep{kamadakawai}, Fructerman-Reingold  \citep{fruchterman-reingold}.
+Other algorithms inclue multi-dimensional scaling, which relies on a defined distance matrix, or an eigen value, which depends on the eigenvalues of the network's adjacency matrix.
 
@@
 This method is particularly useful when the network is very complex, dense, or large.
-XX MORE STUFF ABOUT ADJ MAT VIZ XX
+Experimental studies have shown this to be true.
+For example, with seven basic perceptual tasks on networks, including things like node and edge count, adjacency matrix visualizations outperform node-link diagrams as the size and density of the network increases \citep{adjmatviz}.
+One drawback of the adjacency matrix visualization that \citeauthor{adjmatviz} found was that edges are overrepresented for undirected graphs.
+This, however, may actually be an advantage for \emph{directed} graphs, where exactly the correct number of edges is represented in a matrix visualization, but a node-link diagram may underrepresent the edge count.
 

### 2017-04-12 paper.Rnw
network data intro

@@
 
+Network data is of frequent interest to researchers in a wide array of fields.
+There are techonological networks, like power grids or the internet, information networks, such as citation networks or the World Wide Web, biological networks, e.g.
+neural networks, and social networks, just to name a few \citep{newman}.
+Each of these examples have one thing in common: their data structure.
+There are always units of observation, which we refer to throughout as nodes, and connections of some kind between those units, which we will call edges.
+The networks can also vary over time, for example when new websites and hyperlinks are added on the World Wide Web, or when there are new people and relationships in a friendship network.
+The nodes and edges themselves also have inherent variables of interest, such as the the affiliation of authors in a coauthorship network, and the number of times two authors have collaborated.
+
+The multiple layers, so to speak, of network data structures create unique struggles to network analysts.
+Some questions that researchers aim to answer can be: How does the strength of a connection between two nodes affect the overall structure of the network?
+Do node-level differences affect the formation or destruction of edges?
+Which views of the data are most informative for communicating effects and results of statistical analyses?
+These are, of course, just a few broad questions, and we will focus on only the latter.
+
 \subsection{Visualizing Network Data} \label{sec:netvizintro}
@@
 Visualizing network data is uniquely difficult because of the structure of the data itself.
-Most visualizations rely on a well-defined axis inherited from the data.
+Most, if not all, data visualizations rely on a well-defined axis inherited from the data.
 If data are numerical, histograms, scatterplots, or time series plots are straightforward to construct.
-If the data are categorical, bar charts and mosaic plots can be constructed easily.
-If the data are spatial, there is a well-defined space in which to view information.
-Network data are much less cut-and-dried.
+If the data are categorical, bar charts and mosaic plots are available to to the researcher.
+If the data are spatial, there is a well-defined region in which to view information.
+Network data, however, are much less cut-and-dried.
 

### 2017-04-12 paper.Rnw
add small ex

@@
 \section{Introduction} 
-\st{
-\begin{itemize}
-%\item What? - model visualization and SAOMs. briefly dsecribe what these things are
-%\item Why? - bring light to the RSiena blackbox. Need ways to visualize network models, collections of networks. 
-\item How?
-\end{itemize}
-%also talk about other dynamic network models briefly to put SAOMs in context.
-}
+% \st{
+% \begin{itemize}
+% %\item What? - model visualization and SAOMs. briefly dsecribe what these things are
+% %\item Why? - bring light to the RSiena blackbox. Need ways to visualize network models, collections of networks. 
+% \item How?
+% \end{itemize}
+% %also talk about other dynamic network models briefly to put SAOMs in context.
+% }
 
@@
 
-\section{Networks and their Visualizations}
+In Section~\ref{sec:netintro}, we introduce the concepts of networks and network visualizations.
+In Section~\ref{sec:saoms}, we present the SAOM for social network analysis.
+Then, in Section~\ref{sec:ModelVis}, we combine concepts from Sections~\ref{sec:netintro} and \ref{sec:saoms} in an application of the model-vis paradigm, and conclude with a discussion in Section~\ref{sec:discussion}.
+
+\section{Networks and their Visualizations} \label{sec:netintro}
 
@@
 
-There are two primary methods used to visualize networks.
-The first method, often referred to as node-link diagrams, represents nodes with points in 2D Euclidean space and then represents edges by connecting the points with lines when there is an edge between the two nodes.
+There are two primary methods used to visualize networks: node-link diagrams and matrix visualization. We will use the toy data set show below to demonstrate these two methods.
+
+<<toyex>>=
+set.seed(81393)
+nodes <- data.frame(id = 1:5, group = sample(LETTERS[1:2], size = 5, replace = T), stringsAsFactors = F)
+g <- igraph::as_data_frame(igraph::random.graph.game(n = 5, p = .4))
+toynet <- list(nodes = nodes, edges = rbind(g, c(5,4)))
+toynet
+@
+
+The first method, the node-link diagram, represents nodes with points in 2D Euclidean space and then represents edges by connecting the points with lines when there is an edge between the two nodes.
 But, because there is no natural placement of the points, a random placement is used, then adjusted iteratively via a layout algorithm, of which there are many.
+The node-link diagram for our toy network is shown in Figure~\ref{fig:toyplots}.
 
@@
 The second primary method for network visualization uses the adjacency matrix of the network.
+An example of this type of visualization is also shown in Figure~\ref{fig:toyplots} for our toy example.
 This method is particularly useful when the network is very complex, dense, or large.
@@
 One drawback of the adjacency matrix visualization that \citeauthor{adjmatviz} found was that edges are overrepresented for undirected graphs.
-This, however, may actually be an advantage for \emph{directed} graphs, where exactly the correct number of edges is represented in a matrix visualization, but a node-link diagram may underrepresent the edge count.
-
-\section{Stochastic Actor-Oriented Models for Longitudinal Social Networks}
-
-\st{What do I need to communicate in this section? 
-\begin{itemize}
-\item What the models are for: dynamic social networks with actor-level covariates
-\item What are the model components? rate function, objective function (network statistics, actor covariates, interactions between covariates and network statistics), time points ($t_m = 1, \dots, M$)
-\item What is RSiena? What methods of estimation, convergence checks does it use? What are the algorithms? What kinds of models can it fit? 
-\end{itemize}}
+This, however, may actually be an advantage for \emph{directed} graphs, where exactly the correct number of edges is represented in a matrix visualization, but a node-link diagram may underrepresent the edge count with overlapping edges.
+
+<<toyplots, fig.show='hold', fig.cap="On the left, a node-link diagram of our undirected toy network, with nodes placed using the Kamada-Kawai algorithm.
+On the right, the adjacency matrix visualization for that same network.", out.width='.49\\linewidth'>>=
+toynetdat <- fortify(as.edgedf(toynet$edges), toynet$nodes)
+ggplot(data = toynetdat) + 
+  geom_net(aes(from_id=from_id, to_id=to_id, shape =group, color = group), labelon = T, size = 7, labelcolour = 'white', ecolour = 'black', vjust = .5, hjust = .5, arrowgap = .02) +
+  scale_color_manual(values = c("red", "blue")) + 
+  theme_net() + 
+  labs(title = "Node-Link Diagram")
+
+ggplot(data = toynet$edges) +
+  geom_tile(aes(x = from, y = to), color = 'black', fill = 'black') +
+  geom_tile(aes(x = to, y = from), color = 'black', fill = 'black') +
+  geom_vline(xintercept = 0:5+.5) + 
+  geom_hline(yintercept = 0:5+.5) + 
+  scale_x_continuous(breaks = 1:5, labels = 1:5, limits = c(0.5,5.5)) +
+  scale_y_continuous(breaks = 1:5, labels = 1:5, limits = c(0.5,5.5)) +
+  theme_bw() + 
+  labs(title = "Adjacency Matrix Visualization") + 
+  theme(aspect.ratio = 1)
+@
+
+\section{Stochastic Actor-Oriented Models for Longitudinal Social Networks} \label{sec:saoms}
+
+% \st{What do I need to communicate in this section? 
+% \begin{itemize}
+% \item What the models are for: dynamic social networks with actor-level covariates
+% \item What are the model components?
+rate function, objective function (network statistics, actor covariates, interactions between covariates and network statistics), time points ($t_m = 1, \dots, M$)
+% \item What is RSiena? What methods of estimation, convergence checks does it use? What are the algorithms? What kinds of models can it fit? 
+% \end{itemize}}
 
@@
 
-\section{Discussion}
+\section{Discussion} \label{sec:discussion}
 

### 2017-04-12 paper.Rnw
4.2

@@
 The time dimension can also be incorporated with side-by-side adjacency matrix visualizations.
-\st{
-\begin{itemize}
-\item WHat is the model? the network model (SAOM model is just one of many)
-\item What is the data? Edge data, node data, observed time points, unobserved timepoints (CTMC)
-\item Note: specifically dealing with dynamic networks observed at discrete time points. 
-\end{itemize}
-What are some things that could go in here? need more data examples; not just the friends data.
-Harry Potter data??? I want to find my own data, too.}
+
+\st{Because of the multi-level data structure of a network, viewing the model in the data space can depend on which aspect of the model you are interested in viewing.
+XXX Ideas: summary statistic dist.
+on nets simulated from model and show statistic from data; interactive plots showing objective function values on nodes; select node, select other node, show how the function values change interactively (future); XXX}
+
+% \st{
+% \begin{itemize}
+% \item WHat is the model? the network model (SAOM model is just one of many)
+% \item What is the data? Edge data, node data, observed time points, unobserved timepoints (CTMC)
+% \item Note: specifically dealing with dynamic networks observed at discrete time points. 
+% \end{itemize}
+% What are some things that could go in here? need more data examples; not just the friends data.
+% Harry Potter data??? I want to find my own data, too.}
 

### 2017-04-12 paper.Rnw
fix file read error

@@
 <<makinggif, fig.cap='A selection of images in the microstep animation.
-The selected edges and nodes are emphasized by changing the size and the color, then when edges are removed, they fade out, shrinking in size, while the nodes change color and shrink to blend in with the other nodes.', fig.width=8>>=
+The selected edges and nodes are emphasized by changing the size and the color, then when edges are removed, they fade out, shrinking in size, while the nodes change color and shrink to blend in with the other nodes.', fig.width=8, cache=TRUE>>=
 friend.data.w1 <- as.matrix(read.table("data/s50_data/s50-network1.dat"))
@@
 set.seed(823746)
-load("data/ansnull.rda")
-ansnullchains <- get_chain_info(ansnullpaper)
+load("data/ansnullpaper.rda")
+ansnullchains <- get_chain_info(ansnull)
 ansnullchains %>%

### 2017-04-13 paper.Rnw
split up animation figure by add, remove

@@
                       microsteps = filter(ansnullchainsw1w2, rep == 1))
-microsteps <- ms1[1:9]
+microsteps <- ms1[1:20]
 
@@
 ptv <- pretween_vertices(pte = pte, layoutparams = list(n = 16))
-
+# step 1 is remove, step 9 is add. 
 # get tweened data for all microsteps
@@
 
-static_pall <- filter(pall, .frame %in% (1:12 * 3-2))
-
+static_pall <- filter(pall, .frame %in% c(15,20,22,24,26:28, 71:73, 75, 79, 81, 84))
+
+static_pall$type <- NA
+static_pall$type[which(static_pall$.frame %in% c(15,20,22,24,26:28))] <- "Remove Edge"
+static_pall$type[which(static_pall$.frame %in% c(71:73, 75, 79, 81, 84))] <- "Add Edge"
+
+set.seed(34569234)
 create_net_animate(static_pall) + 
-  facet_wrap(~.frame, nrow=3, labeller = 'label_both') + 
+  facet_wrap(~.frame, nrow=2, labeller = 'label_both') + 
   theme(panel.background = element_rect(color = 'black'))
@@
 
-XX INSERT DESCRIPTION OF THE MOVIE - WHAT IT ALLOWS US TO SEE, ETC. XX
+The top row of Figure~\ref{makinggif} shows an edge being removed, and the bottom row shows one being added. In both cases
 

### 2017-04-13 paper.Rnw
just color nodes in ex data

@@
        aes(from_id = X1, to_id = X2, color = as.ordered(drink), group = as.ordered(drink))) +
-  geom_net(fiteach = FALSE, directed=TRUE, layout.alg = 'fruchtermanreingold', size = 3, arrowsize = .5, arrowgap = .03) +
+  geom_net(fiteach = FALSE, directed=TRUE, layout.alg = 'fruchtermanreingold', size = 3, arrowsize = .5, arrowgap = .03, ecolor = 'grey40') +
   scale_color_brewer(palette = "YlOrRd", name = "Drinking\nBehavior") + 

### 2017-04-13 paper.Rnw
add labels ex data viz

--- 2017-04-13 paper.Rnw
@@
 alldat <- rbind(actual1, actual2, actual3)
-alldat$X1 <- paste0("V", alldat$X1)
-alldat$X2 <- ifelse(!is.na(alldat$X2), paste0("V", alldat$X2), NA)
+#alldat$X1 <- paste0("V", alldat$X1)
+alldat$X1 <- as.factor(alldat$X1)
+#alldat$X2 <- ifelse(!is.na(alldat$X2), paste0("V", alldat$X2), NA)
+alldat$X2 <- as.factor(alldat$X2)
 alldataviz <- ggplot(data = alldat, 
        aes(from_id = X1, to_id = X2, color = as.ordered(drink), group = as.ordered(drink))) +
-  geom_net(fiteach = FALSE, directed=TRUE, layout.alg = 'fruchtermanreingold', size = 3, arrowsize = .5, arrowgap = .03) +
+  geom_net(fiteach = FALSE, directed=TRUE, layout.alg = 'fruchtermanreingold', size = 4, arrowsize = .25, arrowgap = .03, ecolour = 'grey40', labelon=T, labelcolour = 'black', fontsize = 3, vjust = .5) +
   scale_color_brewer(palette = "YlOrRd", name = "Drinking\nBehavior") + 

### 2017-04-13 paper.Rnw
add observed probabilities to histogram

@@
 
-The top row of Figure~\ref{makinggif} shows an edge being removed, and the bottom row shows one being added. In both cases
+The top row of Figure~\ref{makinggif} shows an edge being removed, and the bottom row shows one being added. In both cases, the nodes...
 
@@
   mutate(csum = cumsum(alterProb)) %>% 
-  arrange(ego, csum) -> cumprobs 
+  arrange(ego, csum) %>% 
+  mutate(totego = sum(count), 
+         empprob = count / totego) -> cumprobs 
 cumprobs$ego <- factor(paste0("V",cumprobs$ego+1), levels = paste0("V", 1:16))
 
-cumprobs %>% ggplot() + geom_histogram(aes(x = alter, weight = alterProb), color = 'black', fill = 'white', binwidth = 1) + scale_x_continuous(labels = paste0("V",1:16), breaks = 0:15) + coord_flip() + facet_wrap(~ego, labeller = "label_both") + theme_gray()
+cumprobs %>% ggplot() + 
+  geom_histogram(
+    aes(x = alter, weight = alterProb), 
+    color = 'black', fill = 'white', binwidth = 1) + 
+  geom_point(aes(x = alter, y = empprob), color = 'red') + 
+  scale_x_continuous(labels = paste0("V",1:16), breaks = 0:15) + coord_flip() + facet_wrap(~ego, labeller = "label_both") + theme_gray()
+# 
+# cumprobs %>% ggplot() + geom_histogram(aes(x = alter, weight = count), color = 'black', fill = 'white', binwidth = 1) + scale_x_continuous(labels = paste0("V",1:16), breaks = 0:15) + coord_flip() + facet_wrap(~ego, labeller = "label_both") + theme_gray()
 

### 2017-04-14 paper.Rnw
adj mat

@@
 We then performed a principal component analysis on $\mathbf{A}^{cum}$, and used the values of the first principal component to order the vertices on the $x$ and $y$ axes for the adjacency matrix animation.
-For one such series of microsteps simulated by \texttt{RSiena}, we present the adjacency matrix ordered by the (arbitrary) vertex id alongside the seriated adjacency microsteps using the first principal component loading on the cumulative adjacency matrix, $A^{cum}$
+For one such series of microsteps simulated by \texttt{RSiena}, we present the adjacency matrix ordered by the (arbitrary) vertex id alongside the seriated adjacency microsteps using the first principal component loading on the cumulative adjacency matrix, $\mathbf{A}^{cum}$, in Figure~\ref{adjmatorder}.
 
@@
 
-<<histogramalters, fig.width = 8, fig.height=11, fig.cap="Probabilities of">>=
+Using the principal component analysis on $\mathbf{A}^{cum}$ to order the rows and columns of the adjacency matrix visualization significantly improves the ability of the viewer to detect the two connected components in the first wave of the network.
+
+<<histogramalters, fig.width = 8, fig.height=11, fig.cap="Each panel shows the theoretical (as bars) and empirical (as points) probabilities of the chosen ego node changing its tie to each of the other nodes.">>=
 ansnullchains %>% group_by(period, rep) %>% 

### 2017-04-15 paper.Rnw
add plot, add some description

--- 2017-04-13 paper.Rnw
@@
 
-<<histogramalters, fig.width = 8, fig.height=11, fig.cap="Probabilities of">>=
+Using the principal component analysis on $\mathbf{A}^{cum}$ to order the rows and columns of the adjacency matrix visualization clearly shows the two distinct connected components in the first wave of the network, which are difficult to find in the arbitrarily ordered visualization.
+
+We also attempt to better understand the microstep process by visualizing all transition probabilites for the first microstep in the process.
+We only do the first step of many for now because the \texttt{RSiena} transition probabilities after the first step are only comparable for identical steps due to the conditioning on the current network state in the model.
+Thus we have one step for each of the 1,000 repititions to visualize.
+In Figure ~\ref{histogramalters}, we present each acting node, (ego node) and the resulting probabilities of change.
+The probability shown by the bars is the theoretical probability according to the objective function of the ego node changing the tie to the alter node.
+The probability shown by the points is the empirical probability of that change being made, calculated by counting all instances of the ego node first, then computing the proprotion of each change.
+In most cases, they are almost identical, showing that the algorithm is performing as expected.
+
+<<histogramalters, fig.width = 8, fig.height=11, fig.cap="Each panel shows the theoretical (as bars) and empirical (as points) probabilities of the chosen ego node changing its tie to each of the other nodes.">>=
 ansnullchains %>% group_by(period, rep) %>% 
@@
   mutate(csum = cumsum(alterProb)) %>% 
-  arrange(ego, csum) -> cumprobs 
+  arrange(ego, csum) %>% 
+  mutate(totego = sum(count), 
+         empprob = count / totego) -> cumprobs 
 cumprobs$ego <- factor(paste0("V",cumprobs$ego+1), levels = paste0("V", 1:16))
-
-cumprobs %>% ggplot() + geom_histogram(aes(x = alter, weight = alterProb), color = 'black', fill = 'white', binwidth = 1) + scale_x_continuous(labels = paste0("V",1:16), breaks = 0:15) + coord_flip() + facet_wrap(~ego, labeller = "label_both") + theme_gray()
+cumprobs$alter <- factor(paste0("V",cumprobs$alter+1), levels = paste0("V", 1:16))
+
+# is the tie that's added new, removed, or does the network stay the same
+
+changes <- cumprobs %>% select(ego, alter) %>% unique
+
+wave1friends2 <- wave1friends[,-3]
+wave1friends2$type <- "remove"
+changes2 <- left_join(changes, wave1friends2, c("ego" = "from", "alter" = "to"))
+
+changes2$type[is.na(changes2$type)] <- ifelse(changes2$ego[is.na(changes2$type)] == changes2$alter[is.na(changes2$type)], "noChange", "add")
+
+cumprobs %>% left_join(changes2) %>% ggplot() + 
+  geom_histogram(
+    aes(x = alter, weight = alterProb, fill = type), 
+    color = 'black', fill = 'white', binwidth = 1) + 
+  geom_point(aes(x = alter, y = empprob), color = 'red') + 
+  scale_x_continuous(labels = paste0("V",1:16), breaks = 0:15) + coord_flip() + facet_wrap(~ego, labeller = "label_both") + theme_gray()
+# 
+# cumprobs %>% ggplot() + geom_histogram(aes(x = alter, weight = count), color = 'black', fill = 'white', binwidth = 1) + scale_x_continuous(labels = paste0("V",1:16), breaks = 0:15) + coord_flip() + facet_wrap(~ego, labeller = "label_both") + theme_gray()
 

### 2017-04-15 paper.Rnw
more description of animation.

@@
 alldat$X2 <- as.factor(alldat$X2)
+write_csv(alldat, "smallfriends4Geomnet.csv")
 alldataviz <- ggplot(data = alldat, 
@@
 \section{Model Visualizations} \label{sec:ModelVis}
-
-\st{Introduce the concepts from Wickham et al. }
 
@@
 
-The top row of Figure~\ref{makinggif} shows an edge being removed, and the bottom row shows one being added. In both cases, the nodes...
+The top row of Figure~\ref{makinggif} shows an edge being removed, and the bottom row shows one being added.
+In both cases, the ego nodes chosen to act change color from black to red, and they also increase greatly in size.
+In the case of an edge being removed, the edge that currently exists is emphasized with the color and size change that the node also gets, and as the animation proceeds the edge shrinks to disappearing, while the node shrinks and changes color back to the original black.
+If the edge is being added, it appears colored red from nothing and grows to a large size, then changes color to match the rest of the edges, while the node shrinks and changes color to match the other nodes.
 

### 2017-04-17 paper.Rnw
light editing to intro

@@
 % }
+\hh{XXX - talk a bit about social networks first, why we want to model them, and that usually the models are not dynamic.}
 
@@
 Social networks do not form spontaneously; they evolve over time.
-Ties are added and removed, and new actors can join the network.
+Ties are added and removed, and new actors join the network.
 Modeling the underlying mechanisms that create network changes over time is very complex but also has the potential to uncover hidden truths.
 
-\par One set of models for dynamic social networks are the stochastic actor-oriented models (SAOMs) of \citet{saompaper}.
-These models are uniquely different from other social network models because they incorporate network statistics and actor statistics into the model, whereas most other models just use the network statistics to determine change.
+\par One set of models for dynamic social networks are the stochastic actor-oriented models (SAOMs) introduced by \citet{saompaper}.
+These models are fundamentally different from other social network models because they incorporate network {\emph and} actor statistics into the model, when other models only rely on  the network statistics to determine change.
 Allowing the actor-level statistics to change the structure of the network leads to a more intuitive model for social network change.
-In the "real-world" we would expect that people with common interests would be more likely to form relationships, and that intuition is no longer ignored in modeling with SAOMs.
-
-\par The software SIENA, and its \texttt{R} implementation \texttt{RSiena}, was developed by \citet{RSiena} to estimate the parameters of the SAOMs, among other social network models.
-This software is a huge contribution to the field of social network analysis, and when we began working with it we were eager to understand how it works.
-To better understand the \texttt{RSiena} software and the SAOMs themselves, we combine the principals of network visualization with those of model visualization from \citet{modelvis}.
-With the work of \citet{modelvis} guiding us, we explore RSiena and SAOMs with innovative network visualizations, both static and dynamic.
-By bringing some light to the underlying methods and structures "behind the scenes" of SAOMs and RSiena, we will help other researchers working with the models better understand and better use these models.
-
-\par The guiding principles of model visualization from \citet{modelvis} are (1) display the model in the data space, (2) collections of models are more informative than singletons, and (3) explore how algorithms work, instead of just looking at the final result.
+In the ``real" world we expect people with common interests to be more likely to form relationships.
+SAOMs allow us to incorporates this intuition  in the modeling process.
+
+\par In order to estimate the parameters of  SAOMs the software SIENA, and its \texttt{R} implementation \texttt{RSiena}, was developed \cite{RSiena}.
+This software marks a huge contribution to the field of social network analysis.%, and when we began working with it we were eager to understand how it works.
+HH: we are still interested in this.
+However, unlike other models, SAOMs are not very well understood.
+To get better insight into the \texttt{RSiena} software and the SAOMs themselves, we combine the principals of network visualization with those of model visualization as discussed in \citet{modelvis}.
+%Here, we explore RSiena and SAOMs with innovative network visualizations, both static and dynamic.
+By bringing some light to the underlying methods and structures ``behind the scenes" of SAOMs and RSiena, we  help researchers working with the models better understand and better use these models.
+
+\par The guiding principles of model visualization introduced in \citet{modelvis} are (1) display the model in the data space, (2) collections of models are more informative than singletons, and (3) explore how algorithms work, instead of just looking at the final result.
 Stochastic actor-oriented models are a prime example of a set of models that  benefit greatly from application of model visualization.
@@
 These statistics are often somewhat, if not highly, correlated, and so are the associated parameters.
-By visualizing collections of SAOMs, we can gain a better understanding of these correlations and find ways to deal with them and rectify their effects.
+By visualizing collections of SAOMs, we gain a better understanding of these correlations and find ways to deal with them and rectify their effects.
 Finally, estimation of the parameters in a SAOM relies on a Robbins-Monro algorithm, and convergence checks rely on simulations from the fitted model.
-Again, in these steps are largely hidden from the researcher computing them with software.
-We hope that by bringing these background elements and others into the foreground with static and dynamic visualizations, stochastic actor-oriented models will be better understood and more accessible to social network analysts.
+Again, these steps are largely kept in the background of the estimation process.
+With the help of static and dynamic visualizations we bring these elements and others into the foreground, leading to a better understanding and higher accessibility of stochastic actor-oriented models  for social network analysts.
 

### 2017-04-17 paper.Rnw
up to section 3

@@
 Network data is of frequent interest to researchers in a wide array of fields.
-There are techonological networks, like power grids or the internet, information networks, such as citation networks or the World Wide Web, biological networks, e.g.
+There are technological networks, like power grids or the internet, information networks, such as citation networks or the World Wide Web, biological networks, e.g.
 neural networks, and social networks, just to name a few \citep{newman}.
 Each of these examples have one thing in common: their data structure.
-There are always units of observation, which we refer to throughout as nodes, and connections of some kind between those units, which we will call edges.
-The networks can also vary over time, for example when new websites and hyperlinks are added on the World Wide Web, or when there are new people and relationships in a friendship network.
-The nodes and edges themselves also have inherent variables of interest, such as the the affiliation of authors in a coauthorship network, and the number of times two authors have collaborated.
-
-The multiple layers, so to speak, of network data structures create unique struggles to network analysts.
+There are always units of observation, which we refer to throughout this paper as {\emph nodes}, and connections of some kind between those units, which we will call {\emph edges}.
+Networks might change over time, for example when new websites and hyperlinks are added on the World Wide Web, or when there are new people and relationships in a friendship network.
+Nodes and edges themselves also have inherent variables of interest, such as the  affiliation of authors in a co-authorship network, or the number of times two authors have collaborated.
+
+The multiple layers of network data structures create unique struggles to network analysts.
 Some questions that researchers aim to answer can be: How does the strength of a connection between two nodes affect the overall structure of the network?
@@
 Which views of the data are most informative for communicating effects and results of statistical analyses?
-These are, of course, just a few broad questions, and we will focus on only the latter.
+These are, of course, just a few broad questions.
+In this paper we want to focus on the visual exploration of network data and models.
 
@@
 
-There are two primary methods used to visualize networks: node-link diagrams and matrix visualization. We will use the toy data set show below to demonstrate these two methods.
-
+There are two primary methods used to visualize networks: node-link diagrams and matrix visualization.
+%\hh{As a toy example, let us assume that we have five nodes, $\{1,2,3,4,5\}$, connected by some edges: XXX list edges as $1 \mapto 2, 2 \mapto 4, ...$ }
+We will use the toy data set shown below to demonstrate these two methods.
+
+\hh{XXX describe the dataset in words, but not with the code. }
 <<toyex>>=
@@
 The first method, the node-link diagram, represents nodes with points in 2D Euclidean space and then represents edges by connecting the points with lines when there is an edge between the two nodes.
-But, because there is no natural placement of the points, a random placement is used, then adjusted iteratively via a layout algorithm, of which there are many.
+But because there is no natural placement of the points, a random placement is used, then adjusted iteratively via a layout algorithm, of which there are many.
 The node-link diagram for our toy network is shown in Figure~\ref{fig:toyplots}.
 
-Some layout algorithms were designed to mimic physical systems, drawing the graphs based on the "forces" connecting them.
-The network's edges act as springs pushing and pulling the nodes in 2D space.
-Others use properties of the network's adjaceny matrix in order to place the nodes.
-Some commonly used layout algorithms that treat networks like physical systems are Kamada-Kawai \citep{kamadakawai}, Fructerman-Reingold  \citep{fruchterman-reingold}.
-Other algorithms inclue multi-dimensional scaling, which relies on a defined distance matrix, or an eigen value, which depends on the eigenvalues of the network's adjacency matrix.
+Some layout algorithms were designed to mimic physical systems, drawing the graphs based on the ``forces" connecting them.
+The network's edges act as springs pushing and pulling the nodes in a lower dimensional (usually two-dimensional) space.
+%Other algorithms use properties of the network's adjaceny matrix in order to place the nodes.
+\hh{XXX how does the adjacency matrix feature here?
+The adjacency matrix is  a different mathematical representation of a network, but describes the same network as a node-link diagram.}
+Some commonly used layout algorithms, such as  the Kamada-Kawai layout \citep{kamadakawai} and the Fruchterman-Reingold layout \citep{fruchterman-reingold}, treat networks like physical systems.
+Other algorithms are based on multi-dimensional scaling, which relies on a defined distance matrix, or an eigen value, which depends on the eigenvalues of the network's adjacency matrix.
 
 The second primary method for network visualization uses the adjacency matrix of the network.
-An example of this type of visualization is also shown in Figure~\ref{fig:toyplots} for our toy example.
+\hh{XXX what is an adjacency matrix?
+- explain at this point in mathematical terms} An example of this type of visualization is also shown in Figure~\ref{fig:toyplots} for our toy example.
 This method is particularly useful when the network is very complex, dense, or large.
@@
 One drawback of the adjacency matrix visualization that \citeauthor{adjmatviz} found was that edges are overrepresented for undirected graphs.
-This, however, may actually be an advantage for \emph{directed} graphs, where exactly the correct number of edges is represented in a matrix visualization, but a node-link diagram may underrepresent the edge count with overlapping edges.
+\hh{XXX What is the exact logic here that Ghoniem used to come to the conclusion that edges in undirected graphs are over-represented?
+Why would directed graphs make that better?
+wouldn't that just add to the number of edges?
+} This, however, may actually be an advantage for \emph{directed} graphs, where exactly the correct number of edges is represented in a matrix visualization, but a node-link diagram may underrepresent the edge count with overlapping edges.
 

### 2017-04-23 codeforCarson.R
i don't think crosstalk will work

@@ -1,0 +1,253 @@
+# for carson
+library(tidyverse)
+library(RSiena)
+sfmsall <- read_csv("data/smallfriendsmicrosteps.csv")
+
+sfmsall %>% filter(ms == 1) -> one_microstep_nets
+
+sfmsall %>% filter(ms == 0) %>% select(-rep) %>% unique -> initial
+
+load("data/ansnullpaper.rda")
+ansnullchains <- get_chain_info(ansnull)
+
+ansnullchains %>% 
+  filter(period == 1) %>% 
+  group_by(rep) %>%
+  mutate(ego = paste0("V", parse_number(ego)+1), # make the chains
+         alter = paste0("V", parse_number(alter)+1), 
+         ms = row_number(), 
+         alterprob = exp(logAlterProb)) %>% 
+  filter(ms == 1) -> one_microstep
+
+#join microstep data (full net) with just the one microstep data..
+
+left_join(one_microstep_nets, 
+          one_microstep, 
+          by = c("from" = "ego", "to" = "alter", "ms" = "ms", "rep" = "rep")) %>% 
+  data.frame %>% head(30)
+
+friendNet <- list(wave1 = s501[20:35, 20:35],
+                  wave2 = s502[20:35, 20:35],
+                  wave3 = s503[20:35, 20:35])
+friendAlc <- s50a[20:35,]
+colnames(friendAlc) <- paste0("wave",1:3) 
+rownames(friendAlc) <- paste0("V", 1:16)
+# compute the obj. fn. values of the models
+M1_obj_fn <- function(adjmat, egoname, parms){
+  idx <- which(rownames(adjmat) == egoname)
+  row <- adjmat[idx,]
+  col <- adjmat[,idx]
+  S_outdeg <- sum(row)
+  S_recip <- sum(row*col)
+  Ss <- c(S_outdeg, S_recip)
+  val <- sum(parms * Ss)
+  return(val)
+}
+M1_obj_fn(adjmat, egoname, pM1)
+M2_obj_fn <- function(adjmat, egoname, parms, covar){
+  N <- nrow(adjmat)
+  idx <- which(rownames(adjmat) == egoname)
+  row <- adjmat[idx,]
+  col <- adjmat[,idx]
+  S_outdeg <- sum(row)
+  S_recip <- sum(row*col)
+  # compute JTTs 
+  vec <- c()  
+  for (j in 1:N){
+    for(h in setdiff(1:N, j)){
[…]

### 2017-04-23 florentinefamily.R

@@ -1,0 +1,17 @@
+# florentine families recreate
+library(ergm)
+data(florentine)
+flobusiness
+flomarriage
+
+plot(flobusiness)
+plot(flomarriage)
+
+list.vertex.attributes(flobusiness)
+library(geomnet)
+flobusi <- fortify(flobusiness)
+flomar <- fortify(flomarriage)
+flomar$from <- paste0("V", flomar$from)
+flomar$to <- paste0("V", flomar$to)
+ggplot(data = flomar) + 
+  geom_net(aes(from_id = from, to_id = to, size = wealth))

### 2017-04-25 paper.Rnw
mini comment
heatmap not compiling but runs in console

@@
 @
+ \hh{XXX Explain each of these visualizations and either comment on the differences in the visualizations (colors) or don't show them. }
 
@@
 #              ncol = 2, padding = unit(0, 'line'))
+@
+
+Another way to view the transition probabilities is through the adjacency matrix visualization.
+In Figure~\ref{fig:heatmap}, we build on the concept of the ordered adjacency matrix of Figure~\ref{fig:adjmatorder}.
+This heatmap shows the transition probabilities of all ties that are changed in the first microstep of the 1,000 simulations.
+The heatmap is noticeably sparse, showing the lack of coverage in the model.
+Of the 256 possible steps, only 81, or about 32\%, are taken in the 1,000 simulated chains.
+XXX Check this again....
+XXX
+<<heatmap>>=
+ansnullchains %>% 
+  filter(rep == 1, period == 1) %>% 
+  mutate(ego = paste0("V",as.numeric(as.character(ego))+1),
+         alter = paste0("V",as.numeric(as.character(alter))+1), 
+         prob = exp(logAlterProb)) %>% 
+  ggplot(aes(x = factor(ego, levels = order1), 
+             y = factor(alter, levels = rev(order1)), 
+             fill = prob)) + 
+  geom_tile(color = 'black') + 
+  scale_fill_gradient(low = 'white', high = 'blue', limits = c(0,1)) +
+  theme(panel.background = element_rect(fill = 'white')) + 
+  labs(x = "ego", y = "alter")+
+  coord_fixed()
 @

### 2017-05-09 code/modified_time_plot.R
code for plot of all microsteps in "scaled time"

@@ -1,0 +1,96 @@
+# create new "all ms" visualization
+library(tidyverse)
+library(netvizinf)
+library(geomnet)
+library(RSiena)
+smallfriends <- read_csv("data/smallfriends4Geomnet.csv")
+ansnullchains <- get_chain_info(ansnull)
+chainsms <- ansnullchains %>% 
+  group_by(rep, period) %>% 
+  mutate(microstep = row_number(), 
+         prob = exp(logAlterProb),
+         from = paste0("V", as.numeric(as.character(ego))+1),
+         to = paste0("V", as.numeric(as.character(alter))+1)) %>%
+  ungroup()
+wave1friends <- fortify(as.adjmat(fd2.w1))
+chainsms_wave2 <- filter(chainsms, period == 1)
+chainsms_wave3 <- filter(chainsms, period == 2)
+wave2 <- smallfriends %>% filter(Wave == 2) %>% na.omit %>% data.frame()
+
+create_microstep_summary <- function(chains, truedat){
+  R <- max(chains$rep)
+  wave2 <- truedat 
+  names(wave2) <- c("from", "to", "drink", "Wave")
+  all_reps <- list()
+  for (i in 1:R){
+    msall <- netvizinf::listMicrosteps(dat = wave1friends,
+                                       microsteps = filter(chainsms, rep == i))
+    msall_df <- plyr::rbind.fill(lapply(X = seq_along(msall), FUN = function(i, x) {
+      x[[i]]$ms <- i - 1
+      return(x[[i]])
+    }, msall))
+    msall_df <- msall_df %>% select(from, to, ms)
+   # msall_df <- msall_df %>% add_row(from = paste0("V",wave2$from),
+   #                                  to = paste0("V",wave2$to), 
+   #                                  ms = max(msall_df$ms)+1)
+    msall_df$rep <- i 
+   all_reps[[i]] <- msall_df 
+  }
+  return(all_reps)
+}
+testing <- create_microstep_summary(chains = filter(chainsms_wave2, rep %in% 1:5), truedat = wave2)
+
+testingdf <-plyr::rbind.fill(testing)
+
+testingdf %>% 
+  group_by(from, to, ms) %>% 
+  summarize(numappears = n()) %>% 
+  arrange(ms, from, to) 
+
+# okay, it works. do it for all
[…]

### 2017-05-09 paper.Rnw
just push some stuff.

@@
 In Section~\ref{sec:ModelVis}, we visualize the parts of the SIENA method-of-moments algorithm, bringing them out of the ``black-box" and into the light.
+
+\subsection{Model Goodness-of-Fit}
+
+The \texttt{RSiena} software that fits the models to data also comes equipped with the \verbatim{sienaGOF()} function for examining model fit.
+This function ``assess[es] the fit of the model with respect to auxiliary statistics of networks" \citep[p.~53]{RSienamanual}.
+Examples of the auxiliary statistics are the out- or indegree distribution on the nodes, with the option for users to create their own.
 
@@
 # running Siena
-friendship <- sienaDependent(friendshipData)
-alcohol <- varCovar(drink[20:35,])
-mydata <- sienaDataCreate(friendship, alcohol)
-myeffnull <- getEffects(mydata)
-myalgorithm <- sienaAlgorithmCreate(projname = 's16_3')
+#friendship <- sienaDependent(friendshipData)
+#alcohol <- varCovar(drink[20:35,])
+#mydata <- sienaDataCreate(friendship, alcohol)
+#myeffnull <- getEffects(mydata)
+#myalgorithm <- sienaAlgorithmCreate(projname = 's16_3')
 set.seed(823746)

### 2017-05-09 animatewithtweenr.R

@@ -91,11 +91,17 @@ listMicrosteps <- function(dat, microsteps){
   return(iters)
 }
 
+
 # test it on one of the reps from the null model
 
 ms1 <- listMicrosteps(dat = w1, 
                       microsteps = filter(ansnullchainsw1w2, rep == 1))
-
+lapply(seq_along(mylist), 
+                  +                   function(i, x){
+                    +                       x[[i]]$id <- i
+                    +                       return (x[[i]])
+                    +                   }, mylist
+                  + )
 # Step 1: Layout the initial data
 library(sna)
 w1.net <- as.network(friend.data.w1)
@@ -134,7 +140,12 @@ netwave1ms1 <- merge(merge(wave1friendsms1, data2, by.x = 'from', by.y = 'id'), 
 
 library(tweenr)
 # TWEEN EM
-# data <- tween_states(data = list(data1, data2), tweenlength = 3, 1, "linear", 100)
+data <- tween_states(data = list(data1, data2), tweenlength = 3, 1, "linear", 100)
+
+step1 <- merge(ms1[[2]], data, by.x = "from", by.y = "id", all = TRUE)
+names(step1)[6:7] <- c("x.from", "y.from")
+step2 <- merge(step1, data, by.x = c("to", ".frame"), by.y = c("id", ".frame"), all.x = TRUE)
+names(step2)[9:10]  <- c("x.to", "y.to")
 # fulldata <- tween_states(data = list(netwave1, netwave1ms1), 3, 1, 'linear', 50)
 # # doesn't work, need same # of rows. Need to deal with to NAs somehow. 
 # netwave1 <- arrange(netwave1, to,from)
@@ -146,13 +157,14 @@ library(ggplot2)
 library(gganimate)
 
 #animate em
-# p <- ggplot(data=data, aes(x=x, y=y)) + 
-#   geom_point(aes(frame = .frame, colour = colour), size=5) + 
+ p <- ggplot(data=step2) + 
+   geom_point(aes(x = x.from, y = y.from, frame = .frame), size=5, color = 'grey40') + 
+    geom_segment(aes(x = x.from, y = y.from, frame = .frame, xend = x.to, yend = y.to))
 #   geom_text(aes(frame = .frame, label=id), alpha = .6) + 
 #   scale_colour_identity() + 
 #   theme_bw()
-# animation::ani.options(interval = 1/15)
-# gganimate(p, "dancing ball.gif", title_frame = F)
[…]

### 2017-05-09 code/modified_time_plot.R

@@ -82,6 +82,19 @@ ggplot(data = summary_allreps4 %>% filter(eid %in% 1:20)) +
 # try again
 ggplot(data = summary_allreps5) + 
   geom_segment(aes(x = scaled_time, xend = scaled_time, y = y, yend = yend))
+# another way 
+ggplot(data = summary_allreps5) +
+  geom_point(aes(x = scaled_time, y = appears_time))
+# density
+ggplot(data = summary_allreps5) +
+  geom_density(aes(x = appears_time))
+ggplot(data = summary_allreps5) +
+  geom_density(aes(x = scaled_time))
+ggplot(data = summary_allreps5) +
+  geom_density2d(aes(x = scaled_time, y = appears_time))
+
+
+
 
 max(summary_allreps$ms)
 

### 2017-05-09 codeforCarson.R

@@ -1,5 +1,6 @@
 # for carson
 library(tidyverse)
+library(netvizinf)
 library(RSiena)
 sfmsall <- read_csv("data/smallfriendsmicrosteps.csv")
 

### 2017-05-10 paper.Rnw
add siena g-o-f description and plots

@@
 This function ``assess[es] the fit of the model with respect to auxiliary statistics of networks" \citep[p.~53]{RSienamanual}.
-Examples of the auxiliary statistics are the out- or indegree distribution on the nodes, with the option for users to create their own.
+Examples of the auxiliary statistics are the out- or indegree distribution on the nodes, with the option for users to input their own statistics to examine.
+The fit is tested using Mahalanobis distance to determine a $p$-value when comparing the statistics calculated from the observed data (excluding the first observation that is conditioned on) to the distribution of the statistics when calculated on the simulated waves.
+So, good fit is acheived when $p$ is large.
+XX more detail about mahalanobis distance and calculating p-value XX
+
+The \verbatim{plot.sienaGOF()} function allows us to visualize this fit.
+Below is an example of what the goodness-of-fit plot output looks like for indegree and outdegree statistics for a small model.
+Box plots and violin plots are drawn on top of each other in Figure~\ref{fig:sienagof} in order to show the distribution of the simulated network observations compared to the true data shown in red points connected by red lines.
+If the red points lie well within the simulated values, the model is a good fit to the data.
+The model examined in Figure~\ref{fig:sienagof} appears to do a better job of capturing the outdegree distribution than the indegree distrubtion of the data.
+
+<<sienagof, fig.cap="Goodness of fit measures for two network statistics: indegree and outdegree cumulative distributions.
+On the x-axis is the degree value, and on the y-axis is the number of times that degree value occurs in the data.">>=
+load("data/ansnullpaper.rda")
+gof1 <- sienaGOF(ansnull, OutdegreeDistribution, varName ="friendship")
+gof2 <- sienaGOF(ansnull, IndegreeDistribution, varName ="friendship")
+p1 <- plot(gof1)
+p2 <- plot(gof2)
+grid.arrange(p1, p2)
+@
 
@@
 set.seed(823746)
-load("data/ansnullpaper.rda")
 ansnullchains <- get_chain_info(ansnull)

### 2017-05-10 paper.Rnw
done with gof for now. move to making lineups

@@
 If the red points lie well within the simulated values, the model is a good fit to the data.
-The model examined in Figure~\ref{fig:sienagof} appears to do a better job of capturing the outdegree distribution than the indegree distrubtion of the data.
+The model examined in Figure~\ref{fig:sienagof} appears to do a better job of capturing the outdegree distribution than the indegree distribution of the data.
 
@@
 XXX Ideas: summary statistic dist.
-on nets simulated from model and show statistic from data; interactive plots showing objective function values on nodes; select node, select other node, show how the function values change interactively (future); XXX}
+on nets simulated from model and show statistic from data; interactive plots showing objective function values on nodes; select node, select other node, show how the function values change interactively (future); goodness of fit should do here.
+XXX}
 

### 2017-05-11 code/00e_small_friends.R
creating lineups

@@ -48,7 +48,7 @@ alcohol2 <- varCovar( as.matrix(drink2) )
 # create siena data object
 mysmalldata <- sienaDataCreate( friend2, alcohol2)
 null_model_eff2 <- getEffects(mysmalldata)
-#test_model_eff2 <- includeEffects( null_model_eff2, egoX, altX, egoXaltX, interaction1 = "alcohol2")
+test_model_eff2 <- includeEffects( null_model_eff2, egoX, altX, egoXaltX, interaction1 = "alcohol2")
 #test_model_eff2 <- includeEffects( null_model_eff2, sameXTransTrip, interaction1 = "alcohol2")
 myalgorithm2 <- sienaAlgorithmCreate( projname = 's50' , n3 = 1000)
 ests_null <- siena07( myalgorithm2, data = mysmalldata, returnDeps = TRUE, effects = null_model_eff2, batch=TRUE, verbose = FALSE, silent = TRUE)

### 2017-05-11 code/making_lineups_gof.R

@@ -1,0 +1,115 @@
+# creating lineups for model goodness of fit. 
+
+load("data/ansnullpaper.rda")
+ansnull$sims
+
+# for model 1 
+set.seed(9430852)
+M6 <- sample(1000, size = 5, replace = F)
+M12 <- sample(1000, size = 11, replace = F)
+M16 <- sample(1000, size = 15, replace = F)
+M20 <- sample(1000, size = 19, replace = F)
+
+simsM6 <- ansnull$sims[M6]
+simsM12 <- ansnull$sims[M12]
+simsM16 <- ansnull$sims[M16]
+simsM20 <- ansnull$sims[M20]
+
+simsM6[[1]][[1]]$friendship$`1`
+
+# for model2 
+
+#looks like 20:35 are pretty well connected, go with those for now
+fd2.w1 <- friend.data.w1[20:35,20:35]
+fd2.w2 <- friend.data.w2[20:35,20:35]
+fd2.w3 <- friend.data.w3[20:35,20:35]
+# read in covariate data
+drink <- as.matrix(read.table("data/s50_data/s50-alcohol.dat"))
+drink2 <- drink[20:35,]
+# siena data
+library(RSiena)
+friendData2 <- array( c( fd2.w1, fd2.w2, fd2.w3 ),
+                      dim = c( 16, 16, 3 ) )
+# and next give this the role of the dependent variable:
+friend2 <- sienaDependent(friendData2)
+alcohol2 <- varCovar( as.matrix(drink2) )
+# create siena data object
+mysmalldata <- sienaDataCreate( friend2, alcohol2)
+null_model_eff2 <- getEffects(mysmalldata)
+test_model_eff2 <- includeEffects( null_model_eff2, jumpXTransTrip, interaction1 = "alcohol2")
+#test_model_eff2 <- includeEffects( null_model_eff2, sameXTransTrip, interaction1 = "alcohol2")
+myalgorithm2 <- sienaAlgorithmCreate( projname = 's50' , n3 = 1000)
+ests_test <- siena07( myalgorithm2, data = mysmalldata, returnDeps = TRUE, effects = test_model_eff2, batch=TRUE, verbose = FALSE, silent = TRUE)
+#ests_test <- siena07( myalgorithm2, data = mysmalldata, returnDeps = TRUE, effects = test_model_eff2, batch=TRUE, verbose = FALSE)
+
+set.seed(7902487)
[…]

### 2017-05-20 paper.Rnw
all plotcon plots in the paper with some description.

@@
 \centering
-\begin{tabular}{ll}
+\begin{tabular}{lll}
 \textbf{Structural Effects} \\
-outdegree  & $s_{i1}(x) = \sum_j x_{ij}$ \\
-reciprocity  & $s_{i2}(x) = \sum_j x_{ij}x_{ji}$ \\
-transitive triplets  & $s_{i3}(x) = \sum_{j,h} x_{ij}x_{jh}x_{ih}$ \\
+outdegree  & $s_{i1}(x) = \sum_j x_{ij}$ & \includegraphics[width=3cm]{img/outdegree}\\
+reciprocity  & $s_{i2}(x) = \sum_j x_{ij}x_{ji}$ & \includegraphics[width=3cm]{img/reciprocity}\\
+transitive triplets  & $s_{i3}(x) = \sum_{j,h}& \includegraphics[width=3cm]{img/transtrip} x_{ij}x_{jh}x_{ih}$ \\
 \textbf{Covariate Effects}\\
-covariate-alter  & $s_{i4}(x) = \sum_j x_{ij}z_j$ \\
-covariate-ego  & $s_{i5}(x) = z_i\sum_j x_{ij}$ \\
-same covariate & $s_{i6}(x) = \sum_j x_{ij} \mathbb{I}(z_i = z_j)$ \\
-jumping transitive triplets  & $s_{i7}(x) = \sum_{j \neq h} x_{ij}x_{ih}x_{hj} \mathbb{I}(z_i = z_h \neq z_j)$
+covariate-alter  & $s_{i4}(x) = \sum_j x_{ij}z_j$ & \includegraphics[width=3cm]{img/covaralter} \\
+covariate-ego  & $s_{i5}(x) = z_i\sum_j x_{ij}$ & \includegraphics[width=3cm]{img/covearego} \\
+same covariate & $s_{i6}(x) = \sum_j x_{ij} \mathbb{I}(z_i = z_j)$ & \includegraphics[width=3cm]{img/covarsim}%\\
+%jumping transitive triplets  & $s_{i7}(x) = \sum_{j \neq h} x_{ij}x_{ih}x_{hj} \mathbb{I}(z_i = z_h \neq z_j)$
 \end{tabular}
@@
 <<sienagof, fig.cap="Goodness of fit measures for two network statistics: indegree and outdegree cumulative distributions.
-On the x-axis is the degree value, and on the y-axis is the number of times that degree value occurs in the data.">>=
+On the x-axis is the degree value, and on the y-axis is the number of times that degree value occurs in the data.", cache = TRUE, message = FALSE, warning = FALSE>>=
 load("data/ansnullpaper.rda")
+library(lattice)
 gof1 <- sienaGOF(ansnull, OutdegreeDistribution, varName ="friendship")
@@
 p2 <- plot(gof2)
-grid.arrange(p1, p2)
+gridExtra::grid.arrange(p1, p2)
 @
@@
 alldat$X2 <- as.factor(alldat$X2)
-write_csv(alldat, "smallfriends4Geomnet.csv")
+#write_csv(alldat, "smallfriends4Geomnet.csv")
 alldataviz <- ggplot(data = alldat, 
@@
   group_by(rep) %>%
-  select(rep, from = ego, to = alter) %>%
+  dplyr::select(rep, from = ego, to = alter) %>%
   mutate(val = as.numeric(!from == to),
@@
 
-changes <- cumprobs %>% select(ego, alter) %>% unique
+changes <- cumprobs %>% dplyr::select(ego, alter) %>% unique
 
@@
 
+We also wanted to view the complete microstep process from the first wave, which we condition on, to the second wave.
+The number of steps taken from wave 1 to wave 2 varies.
+In the simulations from Model 1, the smallest number of steps taken was 58, and the longest was 248, with a mean of 106 and a median of 103.
+The standard deviation of the 1000 simulations' number of microsteps was 22.8.
+In Figure~\ref{fig:allsteps}, we see three simulations of the process from wave 1 to wave 2, with wave 1 shown on the left, and wave 2 shown on the right.
+In each of the three plots, the $y$-axis contains the edges sorted by how often they appear in the networks along the way.
+We can see that some edges are there in the beginning, but disappear and never come back, while others appear a few steps in, only to dispappear again.
+There are also some edges that exist in wave 2 that don't appear at all in the microstep process in a given repitition.
+
+<<allsteps, fig.cap="Three repetitions of the microstep process from wave 1 to wave 2.
+The $x$ axis is the microstep number, with step 0 representing the first wave of data and the final step representing the second wave of data.">>=
+smallfriends <- read_csv("data/smallfriends4Geomnet.csv")
+names(smallfriends)[1:2] <- c("from", "to")
+msall1 <- netvizinf::listMicrosteps(dat = wave1friends,
+                      microsteps = filter(ansnullchainsw1w2, rep == 1))
+msall1_df <- plyr::rbind.fill(lapply(X = seq_along(msall1), FUN = function(i, x) {
+        x[[i]]$ms <- i - 1
+        return(x[[i]])
+    }, msall1))
+msall1_df <- msall1_df %>% dplyr::select(from, to, ms)
+wave2 <- smallfriends %>% filter(Wave == 2) %>% na.omit %>% data.frame()
+msall1_df <- msall1_df %>% add_row(from = paste0("V",wave2$from),
+                      to = paste0("V",wave2$to), 
+                      ms = max(msall1_df$ms)+1)
+edges <- msall1_df %>% 
+  group_by(from, to) %>%
+  summarize(edgeappears = n()) %>% 
+  arrange(desc(edgeappears)) %>% ungroup() %>%
+  mutate(eid = row_number())
+
+msall1_df2 <- left_join(msall1_df, edges, by = c("from" = "from", "to" = "to"))
+
+p1 <- ggplot(data = msall1_df2) + 
+  geom_tile(aes(x = ms, y = -eid, fill = ms)) + 
+  geom_vline(xintercept = c(0.5,81.5)) + 
+  scale_fill_continuous(low = "red", high = "blue") + 
+  theme(legend.position = 'none', axis.ticks.y = element_blank(),
+        axis.text.y = element_blank(), axis.title = element_blank()) +
+  labs(x = "Microstep")
+
+# rep 2 
+msall2 <- netvizinf::listMicrosteps(dat = wave1friends,
+                      microsteps = filter(ansnullchainsw1w2, rep == 2))
+msall2_df <- plyr::rbind.fill(lapply(X = seq_along(msall2), FUN = function(i, x) {
+        x[[i]]$ms <- i - 1
+        return(x[[i]])
+    }, msall2))
+msall2_df <- msall2_df %>% dplyr::select(from, to, ms)
+msall2_df <- msall2_df %>% add_row(from = paste0("V",wave2$from),
+                      to = paste0("V",wave2$to), 
+                      ms = max(msall2_df$ms)+1)
+
+edges2 <- msall2_df %>% 
+  group_by(from, to) %>%
+  summarize(edgeappears = n()) %>% 
+  arrange(desc(edgeappears)) %>% ungroup() %>%
+  mutate(eid = row_number())
+
+msall2_df2 <- left_join(msall2_df, edges2, by = c("from" = "from", "to" = "to"))
+
+p2 <- ggplot(data = msall2_df2) + 
+  geom_tile(aes(x = ms, y = -eid, fill = ms)) + 
+  geom_vline(xintercept = c(0.5,49.5)) + 
+  scale_fill_continuous(low = "red", high = "blue") +
+  theme(legend.position = 'none', axis.ticks.y = element_blank(),
+        axis.text.y = element_blank(), axis.title = element_blank()) #+
+  #labs(x = "Microstep", y = "Edges (most frequent at top)")
+
+# rep 3 
+msall3 <- netvizinf::listMicrosteps(dat = wave1friends,
+                      microsteps = filter(ansnullchainsw1w2, rep == 3))
+msall3_df <- plyr::rbind.fill(lapply(X = seq_along(msall3), FUN = function(i, x) {
+        x[[i]]$ms <- i - 1
+        return(x[[i]])
+    }, msall3))
+msall3_df <- msall3_df %>% dplyr::select(from, to, ms)
+msall3_df <- msall3_df %>% add_row(from = paste0("V",wave2$from),
+                      to = paste0("V",wave2$to), 
+                      ms = max(msall3_df$ms)+1)
+
+edges3 <- msall3_df %>% 
+  group_by(from, to) %>%
+  summarize(edgeappears = n()) %>% 
+  arrange(desc(edgeappears)) %>% ungroup() %>%
+  mutate(eid = row_number())
+
+msall3_df2 <- left_join(msall3_df, edges3, by = c("from" = "from", "to" = "to"))
+
+p3 <- ggplot(data = msall3_df2) + 
+  geom_tile(aes(x = ms, y = -eid, fill = ms)) + 
+  geom_vline(xintercept = c(0.5,106.5)) + 
+  scale_fill_continuous(low = "red", high = "blue") +
+  theme(legend.position = 'none', axis.ticks.y = element_blank(),
+        axis.text.y = element_blank(), axis.title = element_blank()) #+
+  #labs(x = "Microstep", y = "Edges (most frequent at top)")
+
+grid.arrange(p1, p2, p3, nrow = 3)
+@
+
 \st{The algorithms here are many!
@@
 The time dimension can also be incorporated with side-by-side adjacency matrix visualizations.
+
+Another way to view the model in the data space is through simulation from the model.
+Since no single simulation from a model is going to look like the data, it's better to visualize many simulations together.
+For network data, one way to do this is through the traditional node-link diagram.
+In Figure~\ref{fig:summnetM1}, we show a summary network created with 1000 simulations from Model 1.
+Only edges that appear in at least 51 of the 1000 simulations are included in the visualization, with edges that appear more frequently emphasized by thicker linewidths and a bright red color.
+On either side of the summary network, we show the actual data, wave 1 on the left, and wave 2 on the right.
+Comparing the three visualizations side-by-side shows us that the summary network of simulations of wave 2 looks more like the true wave 2 than wave 1, which is a good indication that the model is capturing the true data structure.
+It's not ideal, however, because there appear to be too many reciprocated ties in the summary network, far more than there are in the data.
+This could indicate that we should remove the reciprocity parameter from the model.
+
+<<summnetM1>>=
+friendshipData <- array(c(fd2.w1, fd2.w2, fd2.w3), dim = c(16, 16, 3))
+drink1 <- s50a[20:35,1]
+drink2 <- s50a[20:35,2]
+drink3 <- s50a[20:35,3]
+friendship <- sienaDependent(friendshipData)
+alcohol <- varCovar(cbind(drink1,drink2, drink3))
+mydata <- sienaDataCreate(friendship, alcohol)
+M1eff <- getEffects(mydata)
+M2eff <- includeEffects(M1eff, jumpXTransTrip, interaction1 = "alcohol")
+M3eff <- includeEffects(M1eff, nbrDist2twice)
+means <- simu2 %>% group_by(Model, parameter) %>% summarize(
+  means = mean(estimate)
+)
+M1parms <- (means %>% filter(Model == "M1"))$means
+M2parms <- (means %>% filter(Model == "M2"))$means
+M3parms <- (means %>% filter(Model == "M3"))$means
+set.seed(4231352)
+M1sims1000 <- saom_simulate(dat = mydata, 
+                            struct = M1eff,
+                       parms = as.numeric(M1parms), 
+                       N = 1000)
+M2sims1000 <- saom_simulate(dat = mydata, 
+                            struct = M2eff,
+                       parms = as.numeric(M2parms), 
+                       N = 1000)
+M3sims1000 <- saom_simulate(dat = mydata, 
+                            struct = M3eff,
+                       parms = as.numeric(M3parms), 
+                       N = 1000)
+M1simsdf <- sims_to_df(M1sims1000)
+M2simsdf <- sims_to_df(M2sims1000)
+M3simsdf <- sims_to_df(M3sims1000)
+M1avgW2 <- M1simsdf %>% filter(!is.na(to) & wave == 2) %>%
+  group_by(from, to) %>% 
+  summarise(count = n()) %>% 
+  mutate(weight = ifelse(from == to, 0, count))
+M2avgW2 <- M2simsdf %>% filter(!is.na(to) & wave == 2) %>%
+  group_by(from, to) %>% 
+  summarise(count = n()) %>% 
+  mutate(weight = ifelse(from == to, 0, count))
+M3avgW2 <- M3simsdf %>% filter(!is.na(to) & wave == 2) %>%
+  group_by(from, to) %>% 
+  summarise(count = n()) %>% 
+  mutate(weight = ifelse(from == to, 0, count))
+
+# make a df of wave 2, wave 1, and the three averages and facet. 
+names(actual1)[1:2] <- c("from", "to")
+names(actual2)[1:2] <- c("from", "to")
+actual1$count <- 1
+actual1$weight <- 1
+actual2$count <- 1
+actual2$weight <- 1
+actual1 <- actual1 %>% dplyr::select(from,to, count, weight)
+actual2 <- actual2 %>% dplyr::select(from,to, count, weight)
+actual1$cat <- "First Observation"
+actual2$cat <- "Second Observation"
+
+avgW2M1 <- M1avgW2 %>% ungroup() %>% 
+  filter(weight > 50) %>% 
+  mutate(from = as.factor(from), to = as.factor(to),
+         cat = "Model1")
+add1 <- setdiff(as.character(1:16), unique(c(as.character(avgW2M1$from), as.character(avgW2M1$to)))) 
+avgW2M1 %>% add_row(from = add1, to = NA, count = NA, weight = NA, cat = "Model1") -> avgW2M1
+avgW2M2 <- M2avgW2 %>% ungroup() %>% 
+  filter(weight > 50) %>% 
+  mutate(from = as.factor(from), to = as.factor(to),
+         cat = "Model2")
+add2 <- setdiff(as.character(1:16), unique(c(as.character(avgW2M2$from), as.character(avgW2M2$to)))) 
+avgW2M2 %>% add_row(from = add2, to = NA, count = NA, weight = NA, cat = "Model2") -> avgW2M2
+avgW2M3 <- M3avgW2 %>% ungroup() %>% 
+  filter(weight > 50) %>% 
+  mutate(from = as.factor(from), to = as.factor(to),
+         cat = "Model3")
+add3 <- setdiff(as.character(1:16), unique(c(as.character(avgW2M3$from), as.character(avgW2M3$to)))) 
+avgW2M3 %>% add_row(from = add3, to = NA, count = NA, weight = NA, cat = "Model3") -> avgW2M3
+
+combinedavgs <- rbind(actual1, actual2, avgW2M1, avgW2M2, avgW2M3)
+combinedavgs %>% group_by(cat) %>% 
+  mutate(linewidth = weight / max(weight,na.rm = T)) -> combinedavgs
+
+combinedavgs %>% filter(cat == "Model1") %>% mutate(logweight = log(weight)) -> t1 
+colors <- tweenr::tween_color(data = c("#969696", "#67000d"), n = 26, ease = 'linear')
+t1 %>% arrange(logweight) -> t1
+t1$color <- NA
+t1$color[1:26] <- colors[[1]]
+
+p1 <- ggplot(data = t1) + 
+  geom_net(aes(from_id = from, to_id = to, linewidth= linewidth),
+           directed = T, curvature = .2, labelon=T, size = 5,
+           arrow = arrow(angle = 20, type = "open", length = unit(.1, "inches")), ecolour = t1$color,
+           color = "#252525", hjust = .5, vjust = .5, labelcolour = 'white', fontsize = 3, arrowgap = .02,
+           singletons = T) +
+  theme_net() + labs(title = "Summary Network: M1") + theme(plot.background = element_rect(color = 'black'))
+p2 <- ggplot(data = filter(combinedavgs, cat == "First Observation")) + 
+  geom_net(aes(from_id = from, to_id = to, linewidth= linewidth),
+           directed = T, curvature = .2, labelon=T, size = 5,
+           arrow = arrow(angle = 20, type = "open", length = unit(.1, "inches")), ecolour = "#969696",
+           color = "#252525", hjust = .5, vjust = .5, labelcolour = 'white', fontsize = 3, arrowgap = .02,
+           singletons = T) +
+  theme_net() + labs(title = "Wave 1") + theme(plot.background = element_rect(color = 'black'))
+p3 <- ggplot(data = filter(combinedavgs, cat == "Second Observation")) + 
+  geom_net(aes(from_id = from, to_id = to, linewidth= linewidth),
+           directed = T, curvature = .2, labelon=T, size = 5,
+           arrow = arrow(angle = 20, type = "open", length = unit(.1, "inches")), ecolour = "#969696",
+           color = "#252525", hjust = .5, vjust = .5, labelcolour = 'white', fontsize = 3, arrowgap = .02,
+           singletons = T) +
+  theme_net() + labs(title = "Wave 2")  + theme(plot.background = element_rect(color = 'black'))
+grid.arrange(p2, p1, p3, nrow = 1)
+@
 
@@
 From these simulations, we first create a visualization that represents an ``average" network.
-To do this, we count occurences of each possible edge in the simulations, resulting in a summary network with weighted edges representing the number of times an edge appeared when simulating from the SAOM.
+To do this, we count occurences of each possible edge in the simulations, resulting in a summary network with weighted edges representing the number of times an edge appeared in the simulated wave 2 when simulating from the SAOM 1000 times.
+In Figure~\ref{averagenet}, we show the first wave, the ``average'' network from the three models we fit, and the second wave.
+Comparing the three averages to waves 1 and 2, we see that they have very similar structure to wave 2.
+Model 2, which included the transitive triplet parameter, seems to have created more group strucutre overall than models 1 and 3.
+This could be an indication that, despite the highly significant nature of the parameter $\beta_3$ according to the $t$-tests in \verbatim{RSiena}, the transitive triplet parameter should not be included because it does not result in a structure similar to the data.
 
@@
 
-M1avg$from <- paste0("V", M1avg$from)
-M1avg$to <- paste0("V", M1avg$to)
-M1avg2$from <- paste0("V", M1avg2$from)
-M1avg2$to <- paste0("V", M1avg2$to)
-M1avgW2$from <- paste0("V", M1avgW2$from)
-M1avgW2$to <- paste0("V", M1avgW2$to)
-M2avg$from <- paste0("V", M2avg$from)
-M2avg$to <- paste0("V", M2avg$to)
-M2avgW2$from <- paste0("V", M2avgW2$from)
-M2avgW2$to <- paste0("V", M2avgW2$to)
-M2avg2$from <- paste0("V", M2avg2$from)
-M2avg2$to <- paste0("V", M2avg2$to)
-M3avg$from <- paste0("V", M3avg$from)
-M3avg$to <- paste0("V", M3avg$to)
-M3avgW2$from <- paste0("V", M3avgW2$from)
-M3avgW2$to <- paste0("V", M3avgW2$to)
-M3avg2$from <- paste0("V", M3avg2$from)
-M3avg2$to <- paste0("V", M3avg2$to)
-
-wave2drink <- drink2[,2]
-wave2drink <- as.data.frame(wave2drink)
-wave2drink$id <- paste0("V", 1:16)
-names(wave2drink)[1] <- "drink"
-wave3drink <- drink2[,3]
-wave3drink <- as.data.frame(wave3drink)
-wave3drink$id <- paste0("V", 1:16)
-names(wave3drink)[1] <- "drink"
-M1avgcov <- merge(M1avg, wave2drink, by.x = "from", by.y = "id", all = T)
-M2avgcov <- merge(M2avg, wave2drink, by.x = "from", by.y = "id", all = T)
-M3avgcov <- merge(M3avg, wave2drink, by.x = "from", by.y = "id", all = T)
-M1avgW2cov <- merge(M1avgW2, wave3drink, by.x = "from", by.y = "id", all = T)
-M2avgW2cov <- merge(M2avgW2, wave3drink, by.x = "from", by.y = "id", all = T)
-M3avgW2cov <- merge(M3avgW2, wave3drink, by.x = "from", by.y = "id", all = T)
-
-M1avg[,-3] %>% spread(to, weight) %>% data.frame-> distmatM1
-rownames(distmatM1) <- distmatM1$from
-distmatM1 <- distmatM1[,-1]
-distmatM1 <- as.matrix(sqrt(1/distmatM1))
-distmatM1[is.na(distmatM1)] <- 0
-M1avgW2[,-3] %>% spread(to, weight) %>% data.frame-> distmatM1W2
-rownames(distmatM1W2) <- distmatM1W2$from
-distmatM1W2 <- distmatM1W2[,-1]
-distmatM1W2 <- as.matrix(sqrt(1/distmatM1W2))
-distmatM1W2[is.na(distmatM1W2)] <- 0
-M2avg[,-3] %>% spread(to, weight) %>% data.frame-> distmatM2
-rownames(distmatM2) <- distmatM2$from
-distmatM2 <- distmatM2[,-1]
-distmatM2 <- as.matrix(sqrt(1/distmatM2))
-distmatM2[is.na(distmatM2)] <- 0
-M2avgW2[,-3] %>% spread(to, weight) %>% data.frame-> distmatM2W2
-rownames(distmatM2W2) <- distmatM2W2$from
-distmatM2W2 <- distmatM2W2[,-1]
-distmatM2W2 <- as.matrix(sqrt(1/distmatM2W2))
-distmatM2W2[is.na(distmatM2W2)] <- 0
-M3avg[,-3] %>% spread(to, weight) %>% data.frame-> distmatM3
-rownames(distmatM3) <- distmatM3$from
-distmatM3 <- distmatM3[,-1]
-distmatM3 <- as.matrix(sqrt(1/distmatM3))
-distmatM3[is.na(distmatM3)] <- 0
-M3avgW2[,-3] %>% spread(to, weight) %>% data.frame-> distmatM3W2
-rownames(distmatM3W2) <- distmatM3W2$from
-distmatM3W2 <- distmatM3W2[,-1]
-distmatM3W2 <- as.matrix(sqrt(1/distmatM3W2))
-distmatM3W2[is.na(distmatM3W2)] <- 0
-
-cols <- brewer.pal(n = 4, "YlOrRd")
-
-plotM1W2 <- ggplot(data = M1avgcov) + 
-  geom_net(aes(from_id = from, to_id = to, linewidth = weight/251, color = as.factor(drink)), 
-           arrowgap = .03, size = 3,arrowsize = .5,
-           directed = TRUE, curvature = .2,
-           #layout.alg = "fruchtermanreingold"
-           layout.alg = "kamadakawai",
-           layout.par = list(elen = distmatM1)
-           ) + 
-  scale_color_brewer(palette = "YlOrRd", name = "Drinking\nBehavior") + 
-  #labs(title = "Wave 2, M1") + 
-  scale_size_continuous(name = "Weight") + 
-  theme_net() + 
-  theme(plot.title = element_text(hjust = .5), legend.position = 'none',plot.margin=unit(c(0,0,0,0),"mm")) + 
-  labs(title = "Model 1 Simulated Wave 2")
-plotM1W3 <- ggplot(data = M1avgW2cov) + 
-  geom_net(aes(from_id = from, to_id = to, linewidth = weight/388.5, color = as.factor(drink)), 
-           arrowgap = .03, size = 3,arrowsize = .5,
-           directed = TRUE, curvature = .2,
-           #layout.alg = "fruchtermanreingold"
-           layout.alg = "kamadakawai",
-           layout.par = list(elen = distmatM1W2)
-           ) + 
-  scale_color_manual(values = cols[2:4], name = "Drinking\nBehavior") + 
-  #labs(title = "Wave 3, M1") + 
-  scale_size_continuous(name = "Weight") + 
-  theme_net() + 
-  theme(plot.title = element_text(hjust = .5), legend.position = 'none',plot.margin=unit(c(0,0,0,0),"mm"))
-plotM2W2 <- ggplot(data = M2avgcov) + 
-  geom_net(aes(from_id = from, to_id = to, linewidth = weight/325, color = as.factor(drink)), 
-           arrowgap = .03, size = 3,arrowsize = .5,
-           directed = TRUE, curvature = .2,
-           #layout.alg = "fruchtermanreingold"
-           layout.alg = "kamadakawai",
-           layout.par = list(elen = distmatM2)
-           ) + 
-  scale_color_brewer(palette = "YlOrRd", name = "Drinking\nBehavior") + 
-  #labs(title = "Wave 2, M2") + 
-  scale_size_continuous(name = "Weight") + 
-  theme_net() + 
-  theme(plot.title = element_text(hjust = .5), legend.position = 'none',plot.margin=unit(c(0,0,0,0),"mm")) + 
-  labs(title = "Model 2 Simulated Wave 2")
-plotM2W3 <- ggplot(data = M2avgW2cov) + 
-  geom_net(aes(from_id = from, to_id = to, linewidth = weight/357, color = as.factor(drink)), 
-           arrowgap = .03, size = 3,arrowsize = .5,
-           directed = TRUE, curvature = .2,
-           #layout.alg = "fruchtermanreingold"
-           layout.alg = "kamadakawai",
-           layout.par = list(elen = distmatM2W2)
-           ) + 
-  scale_color_manual(values = cols[2:4], name = "Drinking\nBehavior") + 
-  #labs(title = "Wave 3, M2") + 
-  scale_size_continuous(name = "Weight") + 
-  theme_net() + 
-  theme(plot.title = element_text(hjust = .5), legend.position = 'none',plot.margin=unit(c(0,0,0,0),"mm"))
-plotM3W2 <- ggplot(data = M3avgcov) + 
-  geom_net(aes(from_id = from, to_id = to, linewidth = weight/258, color = as.factor(drink)), 
-           arrowgap = .03, size = 3,arrowsize = .5,
-           directed = TRUE, curvature = .2,
-           #layout.alg = "fruchtermanreingold"
-           layout.alg = "kamadakawai",
-           layout.par = list(elen = distmatM3)
-           ) + 
-  scale_color_brewer(palette = "YlOrRd", name = "Drinking\nBehavior") + 
-  #labs(title = "Wave 2, M3") + 
-  scale_size_continuous(name = "Weight") + 
-  theme_net() + 
-  theme(plot.title = element_text(hjust = .5), legend.position = 'none',plot.margin=unit(c(0,0,0,0),"mm")) + 
-  labs(title = "Model 3 Simulated Wave 2")
-plotM3W3 <- ggplot(data = M3avgW2cov) + 
-  geom_net(aes(from_id = from, to_id = to, linewidth = weight/372.5, color = as.factor(drink)), 
-           arrowgap = .03, size = 3,arrowsize = .5,
-           directed = TRUE, curvature = .2,
-           #layout.alg = "fruchtermanreingold"
-           layout.alg = "kamadakawai",
-           layout.par = list(elen = distmatM3W2)
-           ) + 
-  scale_color_manual(values = cols[2:4], name = "Drinking\nBehavior") + 
-  #labs(title = "Wave 3, M3") + 
-  scale_size_continuous(name = "Weight") + 
-  theme_net() + 
-  theme(plot.title = element_text(hjust = .5), legend.position = 'none',plot.margin=unit(c(0,0,0,0),"mm"))
-
-origW1 <- ggplot(data = alldat %>% filter(Wave==1), 
-       aes(from_id = X1, to_id = X2, color = as.ordered(drink))) +
-  geom_net(fiteach = FALSE, directed=TRUE, layout.alg = "kamadakawai", size = 3, arrowsize = .5, curvature = .2,
-           arrowgap = .03) +
-  scale_color_brewer(palette = "YlOrRd", name = "Drinking\nBehavior") + 
-  theme_net() +
-  theme(legend.position = 'none',plot.margin=unit(c(0,0,0,0),"mm"))
-origW2 <- ggplot(data = alldat %>% filter(Wave==2), 
-       aes(from_id = X1, to_id = X2, color = as.ordered(drink))) +
-  geom_net(fiteach = FALSE, directed=TRUE, layout.alg = "kamadakawai", size = 3, arrowsize = .5, curvature = .2,
-           arrowgap = .03) +
-  scale_color_brewer(palette = "YlOrRd", name = "Drinking\nBehavior") + 
-  theme_net() +
-  theme(legend.position = 'none',plot.margin=unit(c(0,0,0,0),"mm")) + 
-  labs(title = "Observed Wave 2")
-origW3 <- ggplot(data = alldat %>% filter(Wave==3), 
-       aes(from_id = X1, to_id = X2, color = as.ordered(drink))) +
-  geom_net(fiteach = FALSE, directed=TRUE, layout.alg = "kamadakawai", size = 3, arrowsize = .5, curvature = .2,
-           arrowgap = .03) +
-  scale_color_manual(values = cols[2:4], name = "Drinking\nBehavior") + 
-  theme_net() +
-  theme(legend.position = 'none',plot.margin=unit(c(0,0,0,0),"mm"))
-
-grid.arrange(origW2, plotM1W2,
-              plotM2W2,plotM3W2, 
-              ncol=2, clip = T)
-alldataviz2 <- alldataviz + theme(legend.position = 'none')
-ggdraw() +
-  draw_plot(alldataviz2, x=0, y=.75, width=1, height=.25) + 
-  draw_plot(plotM1W2, .1, .5, .5, .25) +
-  draw_plot(plotM1W3, .3, .5, .5, .25) +
-  draw_plot(plotM2W2, .1, .25, .5, .25) +
-  draw_plot(plotM2W3, .3, .25, .5, .25) +
-  draw_plot(plotM3W2, .1, 0, .5, .25) +
-  draw_plot(plotM3W3, .3, 0, .5, .25) + 
-  draw_plot_label(label = c("Original Data", "M1: Wave 2", "M1: Wave 3", "M2: Wave 2", "M2: Wave 3", "M3: Wave 2", "M3: Wave 3"))
-
-M1avg2$Model <- "M1"
-M2avg2$Model <- "M2"
-M3avg2$Model <- "M3"
-allavg <- rbind(as.data.frame(M1avg2), 
-                as.data.frame(M2avg2),
-                as.data.frame(M3avg2))
-
-ggplot(data = allavg) + 
-  geom_histogram(aes(x = weight), binwidth = 25) + 
-  facet_grid(wave~Model, labeller = "label_both") + 
-  labs(x = "Number of times an edge occurs in 1,000 simulations") +
-  theme_bw()
+# M1avg$from <- paste0("V", M1avg$from)
+# M1avg$to <- paste0("V", M1avg$to)
+# M1avg2$from <- paste0("V", M1avg2$from)
+# M1avg2$to <- paste0("V", M1avg2$to)
+# M1avgW2$from <- paste0("V", M1avgW2$from)
+# M1avgW2$to <- paste0("V", M1avgW2$to)
+# M2avg$from <- paste0("V", M2avg$from)
+# M2avg$to <- paste0("V", M2avg$to)
+# M2avgW2$from <- paste0("V", M2avgW2$from)
+# M2avgW2$to <- paste0("V", M2avgW2$to)
+# M2avg2$from <- paste0("V", M2avg2$from)
+# M2avg2$to <- paste0("V", M2avg2$to)
+# M3avg$from <- paste0("V", M3avg$from)
+# M3avg$to <- paste0("V", M3avg$to)
+# M3avgW2$from <- paste0("V", M3avgW2$from)
+# M3avgW2$to <- paste0("V", M3avgW2$to)
+# M3avg2$from <- paste0("V", M3avg2$from)
+# M3avg2$to <- paste0("V", M3avg2$to)
+
+ggplot(data = combinedavgs) + 
+  geom_net(aes(from_id = from, to_id = to, linewidth= linewidth),
+           directed = T, curvature = .2, labelon=T, size = 5,
+           arrow = arrow(angle = 20, type = "open", length = unit(.1, "inches")), ecolour = "#969696", fiteach = T, 
+           color = "#252525", hjust = .5, vjust = .5, labelcolour = 'white', fontsize = 3, arrowgap = .02,
+           singletons = F) + 
+  theme_net() + theme(panel.background = element_rect(color = 'black')) + 
+  facet_wrap(~cat)
+
+# wave2drink <- drink2[,2]
+# wave2drink <- as.data.frame(wave2drink)
+# wave2drink$id <- paste0("V", 1:16)
+# names(wave2drink)[1] <- "drink"
+# wave3drink <- drink2[,3]
+# wave3drink <- as.data.frame(wave3drink)
+# wave3drink$id <- paste0("V", 1:16)
+# names(wave3drink)[1] <- "drink"
+# M1avgcov <- merge(M1avg, wave2drink, by.x = "from", by.y = "id", all = T)
+# M2avgcov <- merge(M2avg, wave2drink, by.x = "from", by.y = "id", all = T)
+# M3avgcov <- merge(M3avg, wave2drink, by.x = "from", by.y = "id", all = T)
+# M1avgW2cov <- merge(M1avgW2, wave3drink, by.x = "from", by.y = "id", all = T)
+# M2avgW2cov <- merge(M2avgW2, wave3drink, by.x = "from", by.y = "id", all = T)
+# M3avgW2cov <- merge(M3avgW2, wave3drink, by.x = "from", by.y = "id", all = T)
+# 
+# M1avg[,-3] %>% spread(to, weight) %>% data.frame-> distmatM1
+# rownames(distmatM1) <- distmatM1$from
+# distmatM1 <- distmatM1[,-1]
+# distmatM1 <- as.matrix(sqrt(1/distmatM1))
+# distmatM1[is.na(distmatM1)] <- 0
+# M1avgW2[,-3] %>% spread(to, weight) %>% data.frame-> distmatM1W2
+# rownames(distmatM1W2) <- distmatM1W2$from
+# distmatM1W2 <- distmatM1W2[,-1]
+# distmatM1W2 <- as.matrix(sqrt(1/distmatM1W2))
+# distmatM1W2[is.na(distmatM1W2)] <- 0
+# M2avg[,-3] %>% spread(to, weight) %>% data.frame-> distmatM2
+# rownames(distmatM2) <- distmatM2$from
+# distmatM2 <- distmatM2[,-1]
+# distmatM2 <- as.matrix(sqrt(1/distmatM2))
+# distmatM2[is.na(distmatM2)] <- 0
+# M2avgW2[,-3] %>% spread(to, weight) %>% data.frame-> distmatM2W2
+# rownames(distmatM2W2) <- distmatM2W2$from
+# distmatM2W2 <- distmatM2W2[,-1]
+# distmatM2W2 <- as.matrix(sqrt(1/distmatM2W2))
+# distmatM2W2[is.na(distmatM2W2)] <- 0
+# M3avg[,-3] %>% spread(to, weight) %>% data.frame-> distmatM3
+# rownames(distmatM3) <- distmatM3$from
+# distmatM3 <- distmatM3[,-1]
+# distmatM3 <- as.matrix(sqrt(1/distmatM3))
+# distmatM3[is.na(distmatM3)] <- 0
+# M3avgW2[,-3] %>% spread(to, weight) %>% data.frame-> distmatM3W2
+# rownames(distmatM3W2) <- distmatM3W2$from
+# distmatM3W2 <- distmatM3W2[,-1]
+# distmatM3W2 <- as.matrix(sqrt(1/distmatM3W2))
+# distmatM3W2[is.na(distmatM3W2)] <- 0
+
+# cols <- brewer.pal(n = 4, "YlOrRd")
+# 
+# plotM1W2 <- ggplot(data = M1avgcov) + 
+#   geom_net(aes(from_id = from, to_id = to, linewidth = weight/251, color = as.factor(drink)), 
+#            arrowgap = .03, size = 3,arrowsize = .5,
+#            directed = TRUE, curvature = .2,
+#            #layout.alg = "fruchtermanreingold"
+#            layout.alg = "kamadakawai",
+#            layout.par = list(elen = distmatM1)
+#            ) + 
+#   scale_color_brewer(palette = "YlOrRd", name = "Drinking\nBehavior") + 
+#   #labs(title = "Wave 2, M1") + 
+#   scale_size_continuous(name = "Weight") + 
+#   theme_net() + 
+#   theme(plot.title = element_text(hjust = .5), legend.position = 'none',plot.margin=unit(c(0,0,0,0),"mm")) + 
+#   labs(title = "Model 1 Simulated Wave 2")
+# plotM1W3 <- ggplot(data = M1avgW2cov) + 
+#   geom_net(aes(from_id = from, to_id = to, linewidth = weight/388.5, color = as.factor(drink)), 
+#            arrowgap = .03, size = 3,arrowsize = .5,
+#            directed = TRUE, curvature = .2,
+#            #layout.alg = "fruchtermanreingold"
+#            layout.alg = "kamadakawai",
+#            layout.par = list(elen = distmatM1W2)
+#            ) + 
+#   scale_color_manual(values = cols[2:4], name = "Drinking\nBehavior") + 
+#   #labs(title = "Wave 3, M1") + 
+#   scale_size_continuous(name = "Weight") + 
+#   theme_net() + 
+#   theme(plot.title = element_text(hjust = .5), legend.position = 'none',plot.margin=unit(c(0,0,0,0),"mm"))
+# plotM2W2 <- ggplot(data = M2avgcov) + 
+#   geom_net(aes(from_id = from, to_id = to, linewidth = weight/325, color = as.factor(drink)), 
+#            arrowgap = .03, size = 3,arrowsize = .5,
+#            directed = TRUE, curvature = .2,
+#            #layout.alg = "fruchtermanreingold"
+#            layout.alg = "kamadakawai",
+#            layout.par = list(elen = distmatM2)
+#            ) + 
+#   scale_color_brewer(palette = "YlOrRd", name = "Drinking\nBehavior") + 
+#   #labs(title = "Wave 2, M2") + 
+#   scale_size_continuous(name = "Weight") + 
+#   theme_net() + 
+#   theme(plot.title = element_text(hjust = .5), legend.position = 'none',plot.margin=unit(c(0,0,0,0),"mm")) + 
+#   labs(title = "Model 2 Simulated Wave 2")
+# plotM2W3 <- ggplot(data = M2avgW2cov) + 
+#   geom_net(aes(from_id = from, to_id = to, linewidth = weight/357, color = as.factor(drink)), 
+#            arrowgap = .03, size = 3,arrowsize = .5,
+#            directed = TRUE, curvature = .2,
+#            #layout.alg = "fruchtermanreingold"
+#            layout.alg = "kamadakawai",
+#            layout.par = list(elen = distmatM2W2)
+#            ) + 
+#   scale_color_manual(values = cols[2:4], name = "Drinking\nBehavior") + 
+#   #labs(title = "Wave 3, M2") + 
+#   scale_size_continuous(name = "Weight") + 
+#   theme_net() + 
+#   theme(plot.title = element_text(hjust = .5), legend.position = 'none',plot.margin=unit(c(0,0,0,0),"mm"))
+# plotM3W2 <- ggplot(data = M3avgcov) + 
+#   geom_net(aes(from_id = from, to_id = to, linewidth = weight/258, color = as.factor(drink)), 
+#            arrowgap = .03, size = 3,arrowsize = .5,
+#            directed = TRUE, curvature = .2,
+#            #layout.alg = "fruchtermanreingold"
+#            layout.alg = "kamadakawai",
+#            layout.par = list(elen = distmatM3)
+#            ) + 
+#   scale_color_brewer(palette = "YlOrRd", name = "Drinking\nBehavior") + 
+#   #labs(title = "Wave 2, M3") + 
+#   scale_size_continuous(name = "Weight") + 
+#   theme_net() + 
+#   theme(plot.title = element_text(hjust = .5), legend.position = 'none',plot.margin=unit(c(0,0,0,0),"mm")) + 
+#   labs(title = "Model 3 Simulated Wave 2")
+# plotM3W3 <- ggplot(data = M3avgW2cov) + 
+#   geom_net(aes(from_id = from, to_id = to, linewidth = weight/372.5, color = as.factor(drink)), 
+#            arrowgap = .03, size = 3,arrowsize = .5,
+#            directed = TRUE, curvature = .2,
+#            #layout.alg = "fruchtermanreingold"
+#            layout.alg = "kamadakawai",
+#            layout.par = list(elen = distmatM3W2)
+#            ) + 
+#   scale_color_manual(values = cols[2:4], name = "Drinking\nBehavior") + 
+#   #labs(title = "Wave 3, M3") + 
+#   scale_size_continuous(name = "Weight") + 
+#   theme_net() + 
+#   theme(plot.title = element_text(hjust = .5), legend.position = 'none',plot.margin=unit(c(0,0,0,0),"mm"))
+# 
+# origW1 <- ggplot(data = alldat %>% filter(Wave==1), 
+#        aes(from_id = X1, to_id = X2, color = as.ordered(drink))) +
+#   geom_net(fiteach = FALSE, directed=TRUE, layout.alg = "kamadakawai", size = 3, arrowsize = .5, curvature = .2,
+#            arrowgap = .03) +
+#   scale_color_brewer(palette = "YlOrRd", name = "Drinking\nBehavior") + 
+#   theme_net() +
+#   theme(legend.position = 'none',plot.margin=unit(c(0,0,0,0),"mm"))
+# origW2 <- ggplot(data = alldat %>% filter(Wave==2), 
+#        aes(from_id = X1, to_id = X2, color = as.ordered(drink))) +
+#   geom_net(fiteach = FALSE, directed=TRUE, layout.alg = "kamadakawai", size = 3, arrowsize = .5, curvature = .2,
+#            arrowgap = .03) +
+#   scale_color_brewer(palette = "YlOrRd", name = "Drinking\nBehavior") + 
+#   theme_net() +
+#   theme(legend.position = 'none',plot.margin=unit(c(0,0,0,0),"mm")) + 
+#   labs(title = "Observed Wave 2")
+# origW3 <- ggplot(data = alldat %>% filter(Wave==3), 
+#        aes(from_id = X1, to_id = X2, color = as.ordered(drink))) +
+#   geom_net(fiteach = FALSE, directed=TRUE, layout.alg = "kamadakawai", size = 3, arrowsize = .5, curvature = .2,
+#            arrowgap = .03) +
+#   scale_color_manual(values = cols[2:4], name = "Drinking\nBehavior") + 
+#   theme_net() +
+#   theme(legend.position = 'none',plot.margin=unit(c(0,0,0,0),"mm"))
+# 
+# grid.arrange(origW2, plotM1W2,
+#               plotM2W2,plotM3W2, 
+#               ncol=2, clip = T)
+# alldataviz2 <- alldataviz + theme(legend.position = 'none')
+# ggdraw() +
+#   draw_plot(alldataviz2, x=0, y=.75, width=1, height=.25) + 
+#   draw_plot(plotM1W2, .1, .5, .5, .25) +
+#   draw_plot(plotM1W3, .3, .5, .5, .25) +
+#   draw_plot(plotM2W2, .1, .25, .5, .25) +
+#   draw_plot(plotM2W3, .3, .25, .5, .25) +
+#   draw_plot(plotM3W2, .1, 0, .5, .25) +
+#   draw_plot(plotM3W3, .3, 0, .5, .25) + 
+#   draw_plot_label(label = c("Original Data", "M1: Wave 2", "M1: Wave 3", "M2: Wave 2", "M2: Wave 3", "M3: Wave 2", "M3: Wave 3"))
+# 
+# M1avg2$Model <- "M1"
+# M2avg2$Model <- "M2"
+# M3avg2$Model <- "M3"
+# allavg <- rbind(as.data.frame(M1avg2), 
+#                 as.data.frame(M2avg2),
+#                 as.data.frame(M3avg2))
+# 
+# 
+# ggplot(data = allavg) + 
+#   geom_histogram(aes(x = weight), binwidth = 25) + 
+#   facet_grid(wave~Model, labeller = "label_both") + 
+#   labs(x = "Number of times an edge occurs in 1,000 simulations") +
+#   theme_bw()
 

### 2017-05-20 code/00e_small_friends.R
add lineups. see img folder

@@ -10,9 +10,9 @@ friend.data.w1 <- as.matrix(read.table("data/s50_data/s50-network1.dat"))
 friend.data.w2 <- as.matrix(read.table("data/s50_data/s50-network2.dat"))
 friend.data.w3 <- as.matrix(read.table("data/s50_data/s50-network3.dat"))
 
-val1 <- data.frame(friend.data.w1) %>% gather(x,y, V1:V50) %>% select(y)
-val2 <- data.frame(friend.data.w2) %>% gather(x,y, V1:V50) %>% select(y)
-val3 <- data.frame(friend.data.w3) %>% gather(x,y, V1:V50) %>% select(y)
+val1 <- data.frame(friend.data.w1) %>% gather(key = x, value = y, V1:V50) %>% dplyr::select(y)
+val2 <- data.frame(friend.data.w2) %>% gather(x,y, V1:V50) %>% dplyr::select(y)
+val3 <- data.frame(friend.data.w3) %>% gather(x,y, V1:V50) %>% dplyr::select(y)
 
 all_girls <- expand.grid(x = 1:50, y= 1:50)
 

### 2017-05-20 code/modified_time_plot.R

@@ -4,6 +4,7 @@ library(netvizinf)
 library(geomnet)
 library(RSiena)
 smallfriends <- read_csv("data/smallfriends4Geomnet.csv")
+names(smallfriends)[1:2] <- c("from", "to")
 ansnullchains <- get_chain_info(ansnull)
 chainsms <- ansnullchains %>% 
   group_by(rep, period) %>% 
@@ -16,20 +17,22 @@ wave1friends <- fortify(as.adjmat(fd2.w1))
 chainsms_wave2 <- filter(chainsms, period == 1)
 chainsms_wave3 <- filter(chainsms, period == 2)
 wave2 <- smallfriends %>% filter(Wave == 2) %>% na.omit %>% data.frame()
-
+wave3 <- smallfriends %>% filter(Wave == 3) %>% na.omit %>% data.frame()
+## make more generic for wave 3. 
 create_microstep_summary <- function(chains, truedat){
+  browser()
   R <- max(chains$rep)
-  wave2 <- truedat 
-  names(wave2) <- c("from", "to", "drink", "Wave")
+  wave <- truedat 
+  names(wave) <- c("from", "to", "drink", "Wave")
   all_reps <- list()
   for (i in 1:R){
-    msall <- netvizinf::listMicrosteps(dat = wave1friends,
+    msall <- netvizinf::listMicrosteps(dat = wave,
                                        microsteps = filter(chainsms, rep == i))
     msall_df <- plyr::rbind.fill(lapply(X = seq_along(msall), FUN = function(i, x) {
       x[[i]]$ms <- i - 1
       return(x[[i]])
     }, msall))
-    msall_df <- msall_df %>% select(from, to, ms)
+    msall_df <- msall_df %>% dplyr::select(from, to, ms)
    # msall_df <- msall_df %>% add_row(from = paste0("V",wave2$from),
    #                                  to = paste0("V",wave2$to), 
    #                                  ms = max(msall_df$ms)+1)
@@ -40,7 +43,7 @@ create_microstep_summary <- function(chains, truedat){
 }
 testing <- create_microstep_summary(chains = filter(chainsms_wave2, rep %in% 1:5), truedat = wave2)
 
-testingdf <-plyr::rbind.fill(testing)
+testingdf<-plyr::rbind.fill(testing)
 
 testingdf %>% 
   group_by(from, to, ms) %>% 
[…]

### 2017-05-20 code/making_lineups_gof.R
code for lineups

@@ -1,7 +1,7 @@
 # creating lineups for model goodness of fit. 
 
 load("data/ansnullpaper.rda")
-ansnull$sims
+#ansnull$sims
 
 # for model 1 
 set.seed(9430852)
@@ -74,7 +74,7 @@ smallfriends <- read_csv("data/smallfriends4Geomnet.csv")
 smallfriends2 <- smallfriends[,c(1,2,4)]
 names(smallfriends2) <- c("from", "to", "wave")
 smallfriends2$sim = "dat"
-sims_to_lineup <- function(sims, dat, w, sd){
+sims_to_lineup <- function(sims, dat, w, sd, filename){
   dat2 <- filter(dat, wave == w) %>% dplyr::select(from, to, sim)
   sims_df <- sims_to_df(sims = sims)
   sims_df2 <- dplyr::filter(sims_df, wave == w-1) %>% 
@@ -91,14 +91,15 @@ sims_to_lineup <- function(sims, dat, w, sd){
   # copy over from create_lineup function. it's not working right
   
   dat_for_lu <- rbind(sims_df2, dat2)
-  M <- max(sims_for_plot$sim) + 1
+  M <- max(sims_df2$sim) + 1
   if (!is.null(sd)) {
     set.seed(sd)
   }
-  dat_for_lu$sim <- as.numeric(as.character(dat_for_lu$sim))
+  dat_for_lu$sim <- as.factor(dat_for_lu$sim)
+  dat_for_lu <- arrange(dat_for_lu, sim)
   dat_for_lu$plot_order <- rep(sample(M), as.vector(table(dat_for_lu$sim, 
                                                           useNA = "ifany")))
-  ans <- unique(subset(dat_for_lu, is.na(sim))$plot_order)
+  ans <- unique(subset(dat_for_lu, sim == "dat")$plot_order)
   require(geomnet)
   p <- ggplot(data = dat_for_lu) + 
     geom_net(color = 'black', directed = T, fiteach = T, 
@@ -109,7 +110,20 @@ sims_to_lineup <- function(sims, dat, w, sd){
     facet_wrap(~plot_order) + 
     theme_net() + 
     theme(panel.background = element_rect(fill = NA, color = "black") ) 
+  ggplot2::ggsave(filename, p, width = unit(11, units = "in"), height = unit(8.5, units = "in"))
[…]

### 2017-05-25 code/making_lineups_gof.R
add csvs. added node labels, other small plot changes

@@ -74,23 +74,24 @@ smallfriends <- read_csv("data/smallfriends4Geomnet.csv")
 smallfriends2 <- smallfriends[,c(1,2,4)]
 names(smallfriends2) <- c("from", "to", "wave")
 smallfriends2$sim = "dat"
-sims_to_lineup <- function(sims, dat, w, sd, filename){
+sims_to_lineup <- function(sims, dat, w, sd, modelNum){
   dat2 <- filter(dat, wave == w) %>% dplyr::select(from, to, sim)
   sims_df <- sims_to_df(sims = sims)
   sims_df2 <- dplyr::filter(sims_df, wave == w-1) %>% 
     dplyr::select(from, to, sim)
   sims_df2$from <- ifelse(is.na(sims_df2$from), NA, paste0("V", sims_df2$from))
   sims_df2$to <- ifelse(is.na(sims_df2$to), NA, paste0("V", sims_df2$to))
-  # sims_for_plot <- fortify(as.edgedf(sims_df2), group = "sim")
-  # names(sims_for_plot) <- c("from", "sim", "to")
-  # sims_for_plot <- sims_for_plot[,c("from", "to", "sim")]
+  sims_for_plot <- fortify(as.edgedf(sims_df2), data.frame(id = paste0("V", 1:16)), group = "sim")
+  names(sims_for_plot) <- c("from", "sim", "to")
+  sims_for_plot <- sims_for_plot[,c("from", "to", "sim")]
   dat2$from <- ifelse(is.na(dat2$from), NA, paste0("V", dat2$from))
   dat2$to <- ifelse(is.na(dat2$to), NA, paste0("V", dat2$to))
   
   #create_lineup(sims = sims_df2, dat = dat2, sd = 12345)
   # copy over from create_lineup function. it's not working right
   
-  dat_for_lu <- rbind(sims_df2, dat2)
+  dat_for_lu <- rbind(sims_for_plot, dat2)
+  #dat_for_lu <- rbind(sims_df2, dat2)
   M <- max(sims_df2$sim) + 1
   if (!is.null(sd)) {
     set.seed(sd)
@@ -99,31 +100,37 @@ sims_to_lineup <- function(sims, dat, w, sd, filename){
   dat_for_lu <- arrange(dat_for_lu, sim)
   dat_for_lu$plot_order <- rep(sample(M), as.vector(table(dat_for_lu$sim, 
                                                           useNA = "ifany")))
[…]

### 2017-05-25 paper.Rnw
there's some code that's not running but it compiles now!

@@
 \textbf{Structural Effects} \\
-outdegree  & $s_{i1}(x) = \sum_j x_{ij}$ & \includegraphics[width=3cm]{img/outdegree}\\
-reciprocity  & $s_{i2}(x) = \sum_j x_{ij}x_{ji}$ & \includegraphics[width=3cm]{img/reciprocity}\\
-transitive triplets  & $s_{i3}(x) = \sum_{j,h}& \includegraphics[width=3cm]{img/transtrip} x_{ij}x_{jh}x_{ih}$ \\
+outdegree  & $s_{i1}(x) = \sum_j x_{ij}$ & \includegraphics[width=2cm]{img/outdegree}\\
+reciprocity  & $s_{i2}(x) = \sum_j x_{ij}x_{ji}$ & \includegraphics[width=2cm]{img/reciprocity}\\
+transitive triplets  & $s_{i3}(x) = \sum_{j,h} x_{ij}x_{jh}x_{ih}$ & \includegraphics[width=1.5cm]{img/transtrip}  \\
 \textbf{Covariate Effects}\\
 covariate-alter  & $s_{i4}(x) = \sum_j x_{ij}z_j$ & \includegraphics[width=3cm]{img/covaralter} \\
-covariate-ego  & $s_{i5}(x) = z_i\sum_j x_{ij}$ & \includegraphics[width=3cm]{img/covearego} \\
+covariate-ego  & $s_{i5}(x) = z_i\sum_j x_{ij}$ & \includegraphics[width=3cm]{img/covarego} \\
 same covariate & $s_{i6}(x) = \sum_j x_{ij} \mathbb{I}(z_i = z_j)$ & \includegraphics[width=3cm]{img/covarsim}%\\
@@
 
-The \texttt{RSiena} software that fits the models to data also comes equipped with the \verbatim{sienaGOF()} function for examining model fit.
+The \texttt{RSiena} software that fits the models to data also comes equipped with the \texttt{sienaGOF()} function for examining model fit.
 This function ``assess[es] the fit of the model with respect to auxiliary statistics of networks" \citep[p.~53]{RSienamanual}.
@@
 
-The \verbatim{plot.sienaGOF()} function allows us to visualize this fit.
+The \texttt{plot.sienaGOF()} function allows us to visualize this fit.
 Below is an example of what the goodness-of-fit plot output looks like for indegree and outdegree statistics for a small model.
@@
 Model 2, which included the transitive triplet parameter, seems to have created more group strucutre overall than models 1 and 3.
-This could be an indication that, despite the highly significant nature of the parameter $\beta_3$ according to the $t$-tests in \verbatim{RSiena}, the transitive triplet parameter should not be included because it does not result in a structure similar to the data.
+This could be an indication that, despite the highly significant nature of the parameter $\beta_3$ according to the $t$-tests in \texttt{RSiena}, the transitive triplet parameter should not be included because it does not result in a structure similar to the data.
 

### 2017-05-30 paper.Rnw
labels bigger

@@
 ggplot(data = toynetdat) + 
-  geom_net(aes(from_id=from_id, to_id=to_id, shape =group, color = group), labelon = T, size = 7, labelcolour = 'white', ecolour = 'black', vjust = .5, hjust = .5, arrowgap = .02) +
+  geom_net(aes(from_id=from_id, to_id=to_id, shape =group, color = group), labelon = T, size = 10, labelcolour = 'white', ecolour = 'black', vjust = .5, hjust = .5, arrowgap = .02) +
   scale_color_manual(values = c("red", "blue")) + 
@@
 ggplot(data = toynet$edges) +
-  geom_tile(aes(x = from, y = to), color = 'black', fill = 'black') +
-  geom_tile(aes(x = to, y = from), color = 'black', fill = 'black') +
-  geom_vline(xintercept = 0:5+.5) + 
-  geom_hline(yintercept = 0:5+.5) + 
+  geom_tile(aes(x = from, y = to), color = 'grey20', fill = 'black') +
+  geom_tile(aes(x = to, y = from), color = 'grey20', fill = 'black') +
+  geom_segment(aes(x = x, xend = xend), y = 0.5, yend=5.5, colour="grey30", data = data.frame(x = 0:5+.5, xend= 0:5+.5)) + 
+  geom_segment(aes(y=y, yend=yend), x = 0.5, xend=5.5, colour="grey30", data = data.frame(y = 0:5+.5, yend= 0:5+.5)) + 
   scale_x_continuous(breaks = 1:5, labels = 1:5, limits = c(0.5,5.5)) +
@@
   theme_bw() + 
+  theme(panel.grid = element_blank()) +
   labs(title = "Adjacency Matrix Visualization") + 

### 2017-05-30 paper.Rnw
mini changes

@@
 This assumption leads to the rate for the whole network, $n\alpha$.
-In order to achieve the memorylessness property of a Markov process, for any time point, $T$, where $t_{m} \leq T < t_{m+1}$, the waiting time to the next change opportunity by actor $i$ is exponentially distributed with expected value $\frac{1}{\alpha_m}$.
-Thus, the waiting time to the next change opportunity by \textit{any} actor in the network is also exponentially distributed with expected value $\frac{1}{n\alpha_m}$
+In order to achieve the memorylessness property of a Markov process, for any time point, $T$, where $t_{m} \leq T < t_{m+1}$, the waiting time to the next change opportunity by actor $i$ is exponentially distributed with expected value $\alpha_m^{-1}$.
+Thus, the waiting time to the next change opportunity by \textit{any} actor in the network is also exponentially distributed with expected value $(n\alpha_m)^{-1}$
+
+\hh{XXX we tried to estimate the rate based on the simulations - you could include a histogram of the number of changes for the 1000 simulations in the viz part XXX}
 
@@
 \centering
-\begin{tabular}{lll}
-\textbf{Structural Effects} \\
-outdegree  & $s_{i1}(x) = \sum_j x_{ij}$ & \includegraphics[width=2cm]{img/outdegree}\\
-reciprocity  & $s_{i2}(x) = \sum_j x_{ij}x_{ji}$ & \includegraphics[width=2cm]{img/reciprocity}\\
+\begin{tabular}{m{4.5cm}m{4.5cm}m{3cm}}
+\multicolumn{3}{l}{\textbf{Structural Effects}} \\ \hline\hline
+outdegree  & $s_{i1}(x) = \sum_j x_{ij}$ & \includegraphics[width=2cm]{img/outdegree}\\ \hline
+reciprocity  & $s_{i2}(x) = \sum_j x_{ij}x_{ji}$ & \includegraphics[width=2cm]{img/reciprocity}\\ \hline
 transitive triplets  & $s_{i3}(x) = \sum_{j,h} x_{ij}x_{jh}x_{ih}$ & \includegraphics[width=1.5cm]{img/transtrip}  \\
-\textbf{Covariate Effects}\\
-covariate-alter  & $s_{i4}(x) = \sum_j x_{ij}z_j$ & \includegraphics[width=3cm]{img/covaralter} \\
-covariate-ego  & $s_{i5}(x) = z_i\sum_j x_{ij}$ & \includegraphics[width=3cm]{img/covarego} \\
+\\
+\multicolumn{3}{l}{\textbf{Covariate Effects}}\\  \hline \hline
+covariate-alter  & $s_{i4}(x) = \sum_j x_{ij}z_j$ & \includegraphics[width=3cm]{img/covaralter} \\  \hline
+covariate-ego  & $s_{i5}(x) = z_i\sum_j x_{ij}$ & \includegraphics[width=3cm]{img/covarego} \\  \hline
 same covariate & $s_{i6}(x) = \sum_j x_{ij} \mathbb{I}(z_i = z_j)$ & \includegraphics[width=3cm]{img/covarsim}%\\
@@
 This variable has five values in the original data: (1) does not drink, (2) drinks once or twice a year, (3) drinks once a month, (4) drinks once a week, and (5) drinks more than once a week.
-The actor covariates adn networks are visualized in Figure~\ref{fig:alldataviz}.
+The actor covariates and networks are visualized in Figures~\ref{fig:adjmatviz} and~\ref{fig:alldataviz}.
 <<adjmatviz, fig.align='center', fig.cap="A visualization of the adjacency matrices of the three waves of network observations in the ``Teenage Friends and Lifestyle Study'' data.

### 2017-05-31 paper.Rnw
page 3 edits done

--- 2017-05-25 paper.Rnw
@@
 
+Social networks have been studied for decades, beginning with a few foundational works, the most well known of which is the 1967 study, "The Small World Problem" by Stanley Milgram \citep{goldenberg09}.
+Examples of social networks include collaboration networks between academic researchers, friendship networks in a school or university, and trade networks between nations.
+In recent years, the study of social networks has grown in popularity due to an increase in the availability and access to social network data.
+There are many kinds of social networks, but there are not as many statistical models for social network data.
+Some network models that have been applied to social networks include the exponential random graph model and latent space models.
+These models, however, are only for single instance networks.
+If we only care about one network observation, that is not a problem, but if we have many observations of social network over time, these models may not be appropriate.
+When studying a network over time, which is referred to as a \textit{dynamic network}, we need a model that can take time into account.
 Models for dynamic social networks are extremely important because of how realistic they are.
@@
 
-\par One set of models for dynamic social networks are the stochastic actor-oriented models (SAOMs) introduced by \citet{saompaper}.
-These models are fundamentally different from other social network models because they incorporate network {\emph and} actor statistics into the model, when other models only rely on  the network statistics to determine change.
-Allowing the actor-level statistics to change the structure of the network leads to a more intuitive model for social network change.
-In the ``real" world we expect people with common interests to be more likely to form relationships.
-SAOMs allow us to incorporates this intuition  in the modeling process.
-
-\par In order to estimate the parameters of  SAOMs the software SIENA, and its \texttt{R} implementation \texttt{RSiena}, was developed \cite{RSiena}.
-This software marks a huge contribution to the field of social network analysis.%, and when we began working with it we were eager to understand how it works.
+\par In this paper, we concentrate on one type of model for dynamic social networks: the stochastic actor-oriented models (SAOMs), introduced by \citet{saompaper}.
+These models are fundamentally different from other social network models because they allow us to incorporate network \textit{and} actor statistics, when other models only rely on  the network statistics to determine change.
+Letting the actor-level statistics change the structure of the network leads to a more practical and relevant approach to model change in a social network: in the ``real" world we expect people with common interests to be more likely to form relationships.
+SAOMs allow us to incorporate this intuition in the modeling process.
+
+\st{
+%\par In order to estimate the parameters of  SAOMs the software SIENA, and its \texttt{R} implementation \texttt{RSiena}, was developed \cite{RSiena}.
+This software marks a huge contribution to the field of social network analysis., and when we began working with it we were eager to understand how it works.
 HH: we are still interested in this.
-However, unlike other models, SAOMs are not very well understood.
-To get better insight into the \texttt{RSiena} software and the SAOMs themselves, we combine the principals of network visualization with those of model visualization as discussed in \citet{modelvis}.
+Unlike other network models, SAOMs are not very well understood.
+Traditional modeling steps, like determining the likelihood function, are very complex.
+Because of this, SAOMs are fit to data using a series of MCMC phases for finding method of moments estimators.
+Rather than treat SAOM fits and the fitting process as black boxes, we combine the principals of network visualization with those of model visualization as discussed in \citet{modelvis}.
 %Here, we explore RSiena and SAOMs with innovative network visualizations, both static and dynamic.
-By bringing some light to the underlying methods and structures ``behind the scenes" of SAOMs and RSiena, we  help researchers working with the models better understand and better use these models.
+By bringing some light to the underlying methods and structures ``behind the scenes" of SAOMs, we help researchers working with the models better understand the implications and analysis of these models.
+}
 

### 2017-05-31 paper.Rnw
finish intro

@@
 First, the models themselves include a continuous-time Markov chain (CTMC) that is completely hidden in the model fitting process.
-Bringing the CTMC out from behind-the-scenes and into the light of visualization  provides researchers with insights into the underlying features of the model.
-Secondly, SAOMs allow for a great deal of parameters that can be added to the objective function, each of which are attached to a network statistic.
+Bringing the CTMC out of the black box and into the light of visualization  provides researchers with insights into the underlying features of the model.
+Secondly, SAOMs allow for a great deal of parameters that can be added to the objective function, each of which is attached to a network statistic.
 These statistics are often somewhat, if not highly, correlated, and so are the associated parameters.
 By visualizing collections of SAOMs, we gain a better understanding of these correlations and find ways to deal with them and rectify their effects.
-Finally, estimation of the parameters in a SAOM relies on a Robbins-Monro algorithm, and convergence checks rely on simulations from the fitted model.
+Finally, estimation of the parameters in a SAOM relies on a Robbins-Monro algorithm, and convergence checks on these estimates rely on simulations from the fitted model.
 Again, these steps are largely kept in the background of the estimation process.
@@
 In Section~\ref{sec:saoms}, we present the SAOM for social network analysis.
-Then, in Section~\ref{sec:ModelVis}, we combine concepts from Sections~\ref{sec:netintro} and \ref{sec:saoms} in an application of the model-vis paradigm, and conclude with a discussion in Section~\ref{sec:discussion}.
+In Section~\ref{sec:ModelVis}, we combine concepts from Sections~\ref{sec:netintro} and \ref{sec:saoms} in an application of the model-vis paradigm, and conclude with a discussion in Section~\ref{sec:discussion}.
 

### 2017-05-31 paper.Rnw
2.1 check

@@
 Each of these examples have one thing in common: their data structure.
-There are always units of observation, which we refer to throughout this paper as {\emph nodes}, and connections of some kind between those units, which we will call {\emph edges}.
+There are always units of observation, which we refer to throughout this paper as \textit{nodes}, and connections of some kind between those units, which we will call \textit{edges}.
 Networks might change over time, for example when new websites and hyperlinks are added on the World Wide Web, or when there are new people and relationships in a friendship network.
-Nodes and edges themselves also have inherent variables of interest, such as the  affiliation of authors in a co-authorship network, or the number of times two authors have collaborated.
+Nodes and edges themselves also have inherent variables of interest, e.g.\ the affiliation of authors in a co-authorship network, or the number of times two authors have collaborated.
 
 The multiple layers of network data structures create unique struggles to network analysts.
-Some questions that researchers aim to answer can be: How does the strength of a connection between two nodes affect the overall structure of the network?
+Some questions that researchers aim to answer are: How does the strength of a connection between two nodes affect the overall structure of the network?
 Do node-level differences affect the formation or destruction of edges?
@@
 Visualizing network data is uniquely difficult because of the structure of the data itself.
-Most, if not all, data visualizations rely on a well-defined axis inherited from the data.
-If data are numerical, histograms, scatterplots, or time series plots are straightforward to construct.
-If the data are categorical, bar charts and mosaic plots are available to to the researcher.
+Most, if not all, data visualizations rely on well-defined axes inherited from the data.
+If variables are numerical, histograms, scatterplots, or time series plots are straightforward to construct.
+If the variables are categorical, bar charts and mosaic plots are available to to the researcher.
 If the data are spatial, there is a well-defined region in which to view information.
@@
 
-There are two primary methods used to visualize networks: node-link diagrams and matrix visualization.
+There are two primary methods used to visualize networks: node-link diagrams and matrix visualization. XX INSERT REFS XX
 %\hh{As a toy example, let us assume that we have five nodes, $\{1,2,3,4,5\}$, connected by some edges: XXX list edges as $1 \mapto 2, 2 \mapto 4, ...$ }

### 2017-06-01 code/making_lineups_gof.R
updated references

@@ -134,3 +134,8 @@ sims_to_lineup(sims = simsM16.3, dat = smallfriends2, w = 2, sd = 12345, modelNu
 sims_to_lineup(sims = simsM20, dat = smallfriends2, w = 2, sd = 12345, modelNum = 1)
 sims_to_lineup(sims = simsM20.2, dat = smallfriends2, w = 2, sd = 12345, modelNum = 2)
 sims_to_lineup(sims = simsM20.3, dat = smallfriends2, w = 2, sd = 12345, modelNum = 3)
+
+
+library(netvizinf)
+
+M2sims <- sims_to_df(ests_test$sims)

### 2017-06-02 paper.Rnw
add refs to adjmat and nodelink diags

@@
 % }
-\hh{XXX - talk a bit about social networks first, why we want to model them, and that usually the models are not dynamic.}
+%\hh{XXX - talk a bit about social networks first, why we want to model them, and that usually the models are not dynamic.}
 
@@
 
-There are two primary methods used to visualize networks: node-link diagrams and matrix visualization. XX INSERT REFS XX
-%\hh{As a toy example, let us assume that we have five nodes, $\{1,2,3,4,5\}$, connected by some edges: XXX list edges as $1 \mapto 2, 2 \mapto 4, ...$ }
-We will use the toy data set shown below to demonstrate these two methods.
-
-\hh{XXX describe the dataset in words, but not with the code. }
-<<toyex>>=
+There are two primary methods used to visualize networks: node-link diagrams and matrix visualization \citep{knuth2000, adjmatpro}.
+As a toy example, let us assume that we have five nodes, $\{1,2,3,4,5\}$, connected by five edges:  $\{2 \mapto 4, 3 \mapto 4, 1 \mapto 5, 3 \mapto 5, 5 \mapto 4\}$ We use this toy data set to demonstrate the two visualization methods in Figure~\ref{fig:toyplots}.
+
+<<toyex, echo = FALSE>>=
 set.seed(81393)
@@
 toynet <- list(nodes = nodes, edges = rbind(g, c(5,4)))
-toynet
 @

### 2017-06-02 paper.Rnw
couple of mini comments

@@
 HH: we are still interested in this.
+\hh{The next sentence is very vague - make it more specific: what aspect of SAOMs is not well understood?
+I think what you want is something like ``SAOMs are not very tractable analytically with traditional methods, e.g.
+likelihood functions turn into very complex objects because of the sheer number of parameters involved.
+Therefore, computationally more tractable solutions are used to fit estimators, in particular, SAOMs are fit using ...
+}
 Unlike other network models, SAOMs are not very well understood.
@@
 
-\par The guiding principles of model visualization introduced in \citet{modelvis} are (1) display the model in the data space, (2) collections of models are more informative than singletons, and (3) explore how algorithms work, instead of just looking at the final result.
+\par The guiding principles of model visualization introduced in \citet{modelvis} are (1) display the model in the data space, (2) collections of models are more informative than singletons, and (3) explore the fitting process rather than just looking at the final result.
 Stochastic actor-oriented models are a prime example of a set of models that  benefit greatly from application of model visualization.
@@
 By visualizing collections of SAOMs, we gain a better understanding of these correlations and find ways to deal with them and rectify their effects.
-Finally, estimation of the parameters in a SAOM relies on a Robbins-Monro algorithm, and convergence checks on these estimates rely on simulations from the fitted model.
+Finally, estimation of the parameters in a SAOM relies on a Robbins-Monro algorithm, and convergence checks for these estimates rely on simulations from the fitted model.
 Again, these steps are largely kept in the background of the estimation process.

### 2017-06-02 paper.Rnw
up to bottom of page 10

@@
 
-There are two primary methods used to visualize networks: node-link diagrams and matrix visualization \citep{knuth2000, adjmatpro}.
+There are two primary methods used to visualize networks: node-link diagrams and matrix visualization \cite{knuth2000, adjmatpro}.
 As a toy example, let us assume that we have five nodes, $\{1,2,3,4,5\}$, connected by five edges:  $\{2 \mapto 4, 3 \mapto 4, 1 \mapto 5, 3 \mapto 5, 5 \mapto 4\}$ We use this toy data set to demonstrate the two visualization methods in Figure~\ref{fig:toyplots}.
@@
   labs(title = "Adjacency Matrix Visualization") + 
-  theme(aspect.ratio = 1) 
+  theme(aspect.ratio = 1) +
+  xlab("from (ego)") +
+  ylab("to (alter)")
 @
@@
 
-In this paper, the term \textit{dynamic network} refers to a network consisting of the same set of $n$ nodes that is changing over time, and is observed at $M$ discrete time points, $t_1, \dots, t_M$.
-We denote these network observations $x(t_1), \dots, x(t_M)$, and in modelling we condition on the first observation, $x(t_1)$.
-The SAOM assumes that this longitudinal network is embedded within a continuous time Markov chain (CTMC), call it $X(T)$.
+In this paper, the term \textit{dynamic network} refers to a network consisting of the same set of $n$ nodes that is changing over time, and is observed at $M$ discrete time points, $t_1, \dots, t_M$ with $t_1 \le t_2 \le \dots \le t_M$.
+We denote the network observation at timepoint $t_k$ by $x(t_k)$.
+In the modelling process we usually condition on the first observation, $x(t_1)$.
+The SAOM assumes that this longitudinal network is embedded within a continuous time Markov chain (CTMC), which we will refer to as $X(T)$.
 This process is almost entirely unobserved: assume that the beginning of the process, $X(0)$ is equivalent to the first network observation $x(t_1)$, while the end of the process $X(\infty)$ is equivalent to the last observation $x(t_M)$.
 All other parts of the process are unseen.
-The process $X(T)$ is a series of single tie (or edge) changes, in which one actor at a time is given the opportunity to add or remove one outgoing tie.
-Once an actor is given the chance to change a tie, it tries to maximize a sort of utility function based on the current and potential future states of the network.
+We model the process $X(T)$ as a series of single tie (or edge) changes, in which one actor at a time is given the opportunity to add or remove one outgoing tie.
+Once an actor is given the chance to change a tie, it tries to maximize a utility function based on the current and potential future states of the network.
 
@@
 
-For the network $x$ and each actor $i$ in the network, the rate function dictates how often the actor $i$ gets to change its ties, $x_{ij}$, to other nodes $j \neq i$ in the network.
+For the network $x$ and each actor $i$ in the network, the number of times that an actor $i$ gets to change its ties, $x_{ij}$, to other nodes $j \neq i$ in the network is dictated by a \emph{rate function} $\alpha$.
 For this paper, we assume a simple rate function, $\alpha_m$ that is constant for all actors between observations at time $t_m$ to $t_{m+1}$.
-In other modelling scenarios, this rate can be a function that depends on the time period of observation, some actor-level covariates or some actor-level network statistics.
+\hh{XXX that means that the function is essentially just a rate parameter, right?}
+Other modelling scenarios allow this rate to be more flexible,  e.g.\  be a function that depends on the time period of observation, some actor-level covariates or some actor-level network statistics.
+\hh{How is the rate parameter different from a rate parameter?}
 The rate parameter determines how quickly actor $i$ gets an opportunity to change one of its ties, $x_{ij}$ in the time period $t_{m} \leq T < t_{m+1}$.
@@
 
-Thanks to the conditional dependence assumptions in the model, we can consider the objective function for each node separately, since only one tie from one node is changing at a time.
-The objective function is written as $$f_i(\boldsymbol{\beta}, x) = \sum_k \beta_k s_{ik}(x, \mathbf{Z}),$$
+\hh{shouldn't that be independence? - make the conditional independence assumption above a labelled formula (e.g. C1) and then call it here }
+Because of the conditional dependence assumptions in the model, we can consider the objective function for each node separately, since only one tie from one node is changing at a time.
+The objective function is then written as \hh{the sum of the individual changes XXX write out the math in words XXX} $$f_i(\boldsymbol{\beta}, x) = \sum_k \beta_k s_{ik}(x, \mathbf{Z}),$$
 for $x \in \mathcal{X}$, the space of all possible directed networks with the $n$ nodes, and $\mathbf{Z}$ matrix of covariates.
 The vector $\boldsymbol{\beta}$ are the parameters of the model with corresponding network and covariate statistics, $s_{ik}(x, \mathbf{Z})$, for $k = 1,\dots, K$.
-Given the ego node, $i$, there are $n$ possible steps for the actor $i$ to take: either one of all current ties $x_{ij} = 1$ will be destroyed, a new tie will be created, or no change will occur.
+\hh{XXX ego node isn't defined yet - the description of the charts would be a good place for it.} Given the ego node, $i$, there are $n$ possible steps for the actor $i$ to take: either one of all current ties $x_{ij} = 1$ will be destroyed, a new tie will be created, or no change will occur.
 
 The parameters, $\boldsymbol{\beta}$, are attached to various actor-level network statistics, $s_{ik}(x)$.
-There are always at least two parameters, $\beta_1$ for the outdegree of a node, and $\beta_2$ for the number of reciprocal ties held by a node \citep[p.
+\hh{'There are always' ...
+that seems to come from a Snijder paper?
+Re-phrase this to be less vague: In the current implementation of RSiena (snijder) a model has minimally two parameters...} There are always at least two parameters, $\beta_1$ for the outdegree of a node, and $\beta_2$ for the number of reciprocal ties held by a node \citep[p.
 371]{snijders01}.
-There are many possible parameters $\beta_i$ to add to the model.
-They can be split up into two groups: first, the structural effects, which only depend on the structure of the network.
-The inclusion of these effects has origin in the classical exponential random graph model (ERGM) for networks.
+\hh{'There are many' another vague phrase} There are many possible parameters $\beta_i$ to add to the model.
+\hh{'They' ...
+be specific.} They can be split up into two groups: first, the structural effects, which only depend on the structure of the network.
+\sct{The inclusion of these effects has origin in the classical exponential random graph model (ERGM) for networks.
 These effects are written in terms of the edge variables $x_{ij}$, for $i \neq j$.
-The second set of effects are the actor-level or covariate effects.
-These effects also depend on the structure of the network.
+The second set of effects are the actor-level or covariate effects.} \hh{XXX move the last colored section to the front of this paragraph.
+That sets up the structure nicely, e.g.
+`In determining the covariates of a SAOM, two sources of parameters are considered: ...} These effects also depend on the structure of the network.
 They are written in terms of $x_{ij}$ but also in terms of the covariates, $\mathbf{Z}$.
 A table of some possible structural and covariate effects is given in Table~\ref{tab:effects}.
-For a complete list of the network and covariate statistics that can be included in the objective function, see \citet{RSiena}.
+For a complete list of the network and covariate statistics that can be included \hh{currently} in the objective function, see \citet{RSiena}.
 

### 2017-06-02 paper.Rnw
next paragraph

@@
 be specific.} They can be split up into two groups: first, the structural effects, which only depend on the structure of the network.
-\sct{The inclusion of these effects has origin in the classical exponential random graph model (ERGM) for networks.
+\st{The inclusion of these effects has origin in the classical exponential random graph model (ERGM) for networks.
 These effects are written in terms of the edge variables $x_{ij}$, for $i \neq j$.
@@
 
-When node $i$ is given the chance to change a tie, we assume that they wish to maximize the value of their objective function $f_i(\boldsymbol{\beta}, x)$ plus a random element, $U_i(x)$, where the $U_i(x)$ are from "the type 1 extreme value distribution (or Gumbel distribution) with mean 0 and scale parameter 1" \citep[p.
+When node $i$ is given the chance to change a tie, \hh{`we assume that they wish' sounds strange - don't assume wishes, state assumptions} we assume that they wish to maximize the value of their objective function $f_i(\boldsymbol{\beta}, x)$ plus a random element, $U_i(x)$, where the $U_i(x)$ are from \hh{XXX don't cite here - just state that Snijder suggest the use of a type 1 extreme value distribution on top of the objective function.
+XX What is the thought behind that?
+} ``the type 1 extreme value distribution (or Gumbel distribution) with mean 0 and scale parameter 1" \citep[p.
 368]{snijders01}.
@@
 $$p_{ij}(\boldsymbol{\beta}, x) = \dfrac{\exp\left\{f_i(\boldsymbol{\beta}, x(i\leadsto j))\right\}}{\sum_{h \neq i} \exp\left\{f_i(\boldsymbol{\beta}, x(i \leadsto h))\right\}}$$
+
+\hh{This paragraph ends a bit abruptly - what is the future state probability used for?}
 

### 2017-06-03 paper.Rnw
notation

@@
 
-When node $i$ is given the chance to change a tie, \hh{`we assume that they wish' sounds strange - don't assume wishes, state assumptions} we assume that they wish to maximize the value of their objective function $f_i(\boldsymbol{\beta}, x)$ plus a random element, $U_i(x)$, where the $U_i(x)$ are from \hh{XXX don't cite here - just state that Snijder suggest the use of a type 1 extreme value distribution on top of the objective function.
+When node $i$ is given the chance to change a tie, \hh{`we assume that they wish' sounds strange - don't assume wishes, state assumptions} we assume that they wish to maximize the value of their objective function $f_i(\boldsymbol{\beta}, x)$ plus a random element, $U_i(x)$, where the $U_i(x)$ are from \hh{XXX don't quote here - just state that Snijder suggest the use of a type 1 extreme value distribution on top of the objective function.
 XX What is the thought behind that?
@@
 
-Using this distribution is convenient because it allows the probablity the actor $i$ chooses to change its tie to actor $j$ in terms of the objective function alone.
+Using this distribution \hh{XXX I'm not following - which distribution do you mean by `this' distribution?
+The density above is the type 1 extreme value density.
+I got lost at the  ``objective function $f_i(\boldsymbol{\beta}, x)$ plus a random element, $U_i(x)$"} is convenient because it allows the probability the actor $i$ chooses to change its tie to actor $j$ in terms of the objective function alone.
 Let $p_{ij}(\boldsymbol{\beta}, x)$ be this probability.

### 2017-06-05 paper.Rnw
done with the edits from last week

--- 2017-06-02 paper.Rnw
@@
 HH: we are still interested in this.
-\hh{The next sentence is very vague - make it more specific: what aspect of SAOMs is not well understood?
-I think what you want is something like ``SAOMs are not very tractable analytically with traditional methods, e.g.
-likelihood functions turn into very complex objects because of the sheer number of parameters involved.
-Therefore, computationally more tractable solutions are used to fit estimators, in particular, SAOMs are fit using ...
-}
 Unlike other network models, SAOMs are not very well understood.
@@
 
-\par The guiding principles of model visualization introduced in \citet{modelvis} are (1) display the model in the data space, (2) collections of models are more informative than singletons, and (3) explore the fitting process rather than just looking at the final result.
+\par The guiding principles of model visualization introduced in \citet{modelvis} are (1) display the model in the data space, (2) collections of models are more informative than singletons, and (3) explore how algorithms work, instead of just looking at the final result.
 Stochastic actor-oriented models are a prime example of a set of models that  benefit greatly from application of model visualization.
@@
 By visualizing collections of SAOMs, we gain a better understanding of these correlations and find ways to deal with them and rectify their effects.
-Finally, estimation of the parameters in a SAOM relies on a Robbins-Monro algorithm, and convergence checks for these estimates rely on simulations from the fitted model.
+Finally, estimation of the parameters in a SAOM relies on a Robbins-Monro algorithm, and convergence checks on these estimates rely on simulations from the fitted model.
 Again, these steps are largely kept in the background of the estimation process.
@@
 
-There are two primary methods used to visualize networks: node-link diagrams and matrix visualization \cite{knuth2000, adjmatpro}.
-As a toy example, let us assume that we have five nodes, $\{1,2,3,4,5\}$, connected by five edges:  $\{2 \mapto 4, 3 \mapto 4, 1 \mapto 5, 3 \mapto 5, 5 \mapto 4\}$ We use this toy data set to demonstrate the two visualization methods in Figure~\ref{fig:toyplots}.
+There are two primary methods used to visualize networks: node-link diagrams and matrix visualization \citep{knuth2000, adjmatpro}.
+As a toy example, let us assume that we have five nodes, $\{1,2,3,4,5\}$, connected by five directed edges:  $\{2 \to 4, 3 \to 4, 1 \to 5, 3 \to 5, 5 \to 4\}$ We use this toy data set to demonstrate the two visualization methods in Figure~\ref{fig:toyplots}.
 
@@
 
-The first method, the node-link diagram, represents nodes with points in 2D Euclidean space and then represents edges by connecting the points with lines when there is an edge between the two nodes.
-But because there is no natural placement of the points, a random placement is used, then adjusted iteratively via a layout algorithm, of which there are many.
-The node-link diagram for our toy network is shown in Figure~\ref{fig:toyplots}.
+The first method, the node-link diagram, represents nodes with markers in 2D Euclidean space and then represents edges by connecting the markers with lines when there is an edge between the two nodes.
+But because there is no natural placement of the points, a random placement is used, then adjusted via a layout algorithm, of which there are many \citep{drawingalgs}.
 
@@
 Some commonly used layout algorithms, such as  the Kamada-Kawai layout \citep{kamadakawai} and the Fruchterman-Reingold layout \citep{fruchterman-reingold}, treat networks like physical systems.
-Other algorithms are based on multi-dimensional scaling, which relies on a defined distance matrix, or an eigen value, which depends on the eigenvalues of the network's adjacency matrix.
+Other algorithms are based on multi-dimensional scaling, which relies on a defined distance matrix, or an eigen value, which depends on the eigenvalues of the network's adjacency matrix \citep{drawingalgs}.
+The node-link diagram for our toy network is shown in Figure~\ref{fig:toyplots}.
 
 The second primary method for network visualization uses the adjacency matrix of the network.
-\hh{XXX what is an adjacency matrix?
-- explain at this point in mathematical terms} An example of this type of visualization is also shown in Figure~\ref{fig:toyplots} for our toy example.
-This method is particularly useful when the network is very complex, dense, or large.
-Experimental studies have shown this to be true.
-For example, with seven basic perceptual tasks on networks, including things like node and edge count, adjacency matrix visualizations outperform node-link diagrams as the size and density of the network increases \citep{adjmatviz}.
-One drawback of the adjacency matrix visualization that \citeauthor{adjmatviz} found was that edges are overrepresented for undirected graphs.
-\hh{XXX What is the exact logic here that Ghoniem used to come to the conclusion that edges in undirected graphs are over-represented?
+The adjaceny matrix of a network, $\mathbf{A}$, describes the edges of a network in matrix form.
+An entry $A_{ij}$ of $\mathbf{A}$, for two nodes $i\neq j$ in the network is defined as
+\[
+  A_{ij} =
+  \begin{cases}
+                                   1 &  \text{if an edge exists} i\to j \\
+                                   0 & otherwise
+  \end{cases}
+\]
+In an undirected network, $A_{ij} = A_{ji}$, but in a directed graph this is only true if there is an edge from $i$ to $j$ and from $j$ to $i$.
+Thus, $\mathbf{A}$ is always symmetric for undirected networks, and is symmetric for directed networks only if every edge between two nodes is reciprocated.
+%\hh{XXX what is an adjacency matrix? - explain at this point in mathematical terms} 
+This type of visualization is also shown in Figure~\ref{fig:toyplots} for our toy example.
+%\hh{XXX What is the exact logic here that Ghoniem used to come to the conclusion that edges in undirected graphs are over-represented?
 Why would directed graphs make that better?
@@
 
-<<toyplots, fig.show='hold', fig.cap="On the left, a node-link diagram of our undirected toy network, with nodes placed using the Kamada-Kawai algorithm.
+<<toyplots, fig.show='hold', fig.cap="On the left, a node-link diagram of our directed toy network, with nodes placed using the Kamada-Kawai algorithm.
 On the right, the adjacency matrix visualization for that same network.", out.width='.49\\linewidth'>>=
@@
 ggplot(data = toynetdat) + 
-  geom_net(aes(from_id=from_id, to_id=to_id, shape =group, color = group), labelon = T, size = 10, labelcolour = 'white', ecolour = 'black', vjust = .5, hjust = .5, arrowgap = .02) +
-  scale_color_manual(values = c("red", "blue")) + 
+  geom_net(aes(from_id=from_id, to_id=to_id), labelon = T, size = 10, labelcolour = 'white', ecolour = 'black', vjust = .5, hjust = .5, arrowgap = .05, directed = T) +
   theme_net() + 
@@
 ggplot(data = toynet$edges) +
-  geom_tile(aes(x = from, y = to), color = 'grey20', fill = 'black') +
-  geom_tile(aes(x = to, y = from), color = 'grey20', fill = 'black') +
-  geom_segment(aes(x = x, xend = xend), y = 0.5, yend=5.5, colour="grey30", data = data.frame(x = 0:5+.5, xend= 0:5+.5)) + 
-  geom_segment(aes(y=y, yend=yend), x = 0.5, xend=5.5, colour="grey30", data = data.frame(y = 0:5+.5, yend= 0:5+.5)) + 
-  scale_x_continuous(breaks = 1:5, labels = 1:5, limits = c(0.5,5.5)) +
-  scale_y_continuous(breaks = 1:5, labels = 1:5, limits = c(0.5,5.5)) +
+  geom_tile(aes(x = from, y = -to), color = 'grey20', fill = 'black') +
+  #geom_tile(aes(x = to, y = from), color = 'grey20', fill = 'black') +
+  geom_segment(aes(x = x, xend = xend), y = -0.5, yend=-5.5, colour="grey30", data = data.frame(x = 0:5+.5, xend= 0:5+.5)) + 
+  geom_segment(aes(y=-y, yend=-yend), x = 0.5, xend=5.5, colour="grey30", data = data.frame(y = -0:5+.5, yend= -0:5+.5)) + 
+  scale_x_continuous(breaks = 1:5, labels = 1:5, limits = c(0.5,5.5), position = "top") +
+  scale_y_continuous(breaks = (-5):(-1), labels = 5:1, limits = c(-5.5,-.5)) +
   theme_bw() + 
   theme(panel.grid = element_blank()) +
-  labs(title = "Adjacency Matrix Visualization") + 
-  theme(aspect.ratio = 1) +
-  xlab("from (ego)") +
-  ylab("to (alter)")
-@
- \hh{XXX Explain each of these visualizations and either comment on the differences in the visualizations (colors) or don't show them. }
+  labs(title = "Adjacency Matrix Visualization", y = "to") + 
+  theme(aspect.ratio = 1) 
+@
+% \hh{XXX Explain each of these visualizations and either comment on the differences in the visualizations (colors) or don't show them. }
+
+Each type of visualization comes with its own advantages and disadvantages.
+Paths between nodes are easier to determine with node-link diagrams than with adjacency matrix visualizations.
+In node-link diagrams, node information can be incorporated into the visualization by coloring or changing the shape of the points representing the nodes.
+This is not necessarily possible in matrix representation, which is more focused on edges.
+Adjacency matrix visualization is particularly useful, however, when the network is very complex, dense, or large.
+Experimental studies have shown this to be true.
+For example, with seven basic perceptual tasks on networks, including things like node and edge count, adjacency matrix visualizations outperform node-link diagrams as the size and density of the network increases \citep{adjmatviz}.
+One drawback of the adjacency matrix visualization that \citeauthor{adjmatviz} found was that edges are overrepresented for undirected graphs, due to the symmetry of $\mathbf{A}$.
+There is not one correct way to visualize network information, and we will be using both of the methods throughout this paper.
 
@@
 
-In this paper, the term \textit{dynamic network} refers to a network consisting of the same set of $n$ nodes that is changing over time, and is observed at $M$ discrete time points, $t_1, \dots, t_M$ with $t_1 \le t_2 \le \dots \le t_M$.
-We denote the network observation at timepoint $t_k$ by $x(t_k)$.
-In the modelling process we usually condition on the first observation, $x(t_1)$.
-The SAOM assumes that this longitudinal network is embedded within a continuous time Markov chain (CTMC), which we will refer to as $X(T)$.
+In this paper, the term \textit{dynamic network} refers to a network consisting of the same set of $n$ nodes that is changing over time, and is observed at $M$ discrete time points, $t_1, \dots, t_M$.
+We denote these network observations $x(t_1), \dots, x(t_M)$, and in modelling we condition on the first observation, $x(t_1)$.
+The SAOM assumes that this longitudinal network is embedded within a continuous time Markov chain (CTMC), call it $X(T)$.
 This process is almost entirely unobserved: assume that the beginning of the process, $X(0)$ is equivalent to the first network observation $x(t_1)$, while the end of the process $X(\infty)$ is equivalent to the last observation $x(t_M)$.
 All other parts of the process are unseen.
-We model the process $X(T)$ as a series of single tie (or edge) changes, in which one actor at a time is given the opportunity to add or remove one outgoing tie.
-Once an actor is given the chance to change a tie, it tries to maximize a utility function based on the current and potential future states of the network.
+The process $X(T)$ is a series of single tie (or edge) changes, in which one actor at a time is given the opportunity to add or remove one outgoing tie.
+Once an actor is given the chance to change a tie, it tries to maximize a sort of utility function based on the current and potential future states of the network.
 
@@
 
-For the network $x$ and each actor $i$ in the network, the number of times that an actor $i$ gets to change its ties, $x_{ij}$, to other nodes $j \neq i$ in the network is dictated by a \emph{rate function} $\alpha$.
+For the network $x$ and each actor $i$ in the network, the rate function dictates how often the actor $i$ gets to change its ties, $x_{ij}$, to other nodes $j \neq i$ in the network.
 For this paper, we assume a simple rate function, $\alpha_m$ that is constant for all actors between observations at time $t_m$ to $t_{m+1}$.
-\hh{XXX that means that the function is essentially just a rate parameter, right?}
-Other modelling scenarios allow this rate to be more flexible,  e.g.\  be a function that depends on the time period of observation, some actor-level covariates or some actor-level network statistics.
-\hh{How is the rate parameter different from a rate parameter?}
+In other modelling scenarios, this rate can be a function that depends on the time period of observation, some actor-level covariates or some actor-level network statistics.
 The rate parameter determines how quickly actor $i$ gets an opportunity to change one of its ties, $x_{ij}$ in the time period $t_{m} \leq T < t_{m+1}$.
@@
 
-\hh{shouldn't that be independence? - make the conditional independence assumption above a labelled formula (e.g. C1) and then call it here }
-Because of the conditional dependence assumptions in the model, we can consider the objective function for each node separately, since only one tie from one node is changing at a time.
-The objective function is then written as \hh{the sum of the individual changes XXX write out the math in words XXX} $$f_i(\boldsymbol{\beta}, x) = \sum_k \beta_k s_{ik}(x, \mathbf{Z}),$$
+Thanks to the conditional dependence assumptions in the model, we can consider the objective function for each node separately, since only one tie from one node is changing at a time.
+The objective function is written as $$f_i(\boldsymbol{\beta}, x) = \sum_k \beta_k s_{ik}(x, \mathbf{Z}),$$
 for $x \in \mathcal{X}$, the space of all possible directed networks with the $n$ nodes, and $\mathbf{Z}$ matrix of covariates.
 The vector $\boldsymbol{\beta}$ are the parameters of the model with corresponding network and covariate statistics, $s_{ik}(x, \mathbf{Z})$, for $k = 1,\dots, K$.
-\hh{XXX ego node isn't defined yet - the description of the charts would be a good place for it.} Given the ego node, $i$, there are $n$ possible steps for the actor $i$ to take: either one of all current ties $x_{ij} = 1$ will be destroyed, a new tie will be created, or no change will occur.
+Given the ego node, $i$, there are $n$ possible steps for the actor $i$ to take: either one of all current ties $x_{ij} = 1$ will be destroyed, a new tie will be created, or no change will occur.
 
 The parameters, $\boldsymbol{\beta}$, are attached to various actor-level network statistics, $s_{ik}(x)$.
-\hh{'There are always' ...
-that seems to come from a Snijder paper?
-Re-phrase this to be less vague: In the current implementation of RSiena (snijder) a model has minimally two parameters...} There are always at least two parameters, $\beta_1$ for the outdegree of a node, and $\beta_2$ for the number of reciprocal ties held by a node \citep[p.
+There are always at least two parameters, $\beta_1$ for the outdegree of a node, and $\beta_2$ for the number of reciprocal ties held by a node \citep[p.
 371]{snijders01}.
-\hh{'There are many' another vague phrase} There are many possible parameters $\beta_i$ to add to the model.
-\hh{'They' ...
-be specific.} They can be split up into two groups: first, the structural effects, which only depend on the structure of the network.
-\sct{The inclusion of these effects has origin in the classical exponential random graph model (ERGM) for networks.
+There are many possible parameters $\beta_i$ to add to the model.
+They can be split up into two groups: first, the structural effects, which only depend on the structure of the network.
+The inclusion of these effects has origin in the classical exponential random graph model (ERGM) for networks.
 These effects are written in terms of the edge variables $x_{ij}$, for $i \neq j$.
-The second set of effects are the actor-level or covariate effects.} \hh{XXX move the last colored section to the front of this paragraph.
-That sets up the structure nicely, e.g.
-`In determining the covariates of a SAOM, two sources of parameters are considered: ...} These effects also depend on the structure of the network.
+The second set of effects are the actor-level or covariate effects.
+These effects also depend on the structure of the network.
 They are written in terms of $x_{ij}$ but also in terms of the covariates, $\mathbf{Z}$.
 A table of some possible structural and covariate effects is given in Table~\ref{tab:effects}.
-For a complete list of the network and covariate statistics that can be included \hh{currently} in the objective function, see \citet{RSiena}.
+For a complete list of the network and covariate statistics that can be included in the objective function, see \citet{RSiena}.
 

### 2017-06-05 paper.Rnw

--- 2017-06-03 paper.Rnw
@@
 ggplot(data = toynet$edges) +
-  geom_tile(aes(x = from, y = to), color = 'grey20', fill = 'black') +
-  geom_tile(aes(x = to, y = from), color = 'grey20', fill = 'black') +
-  geom_segment(aes(x = x, xend = xend), y = 0.5, yend=5.5, colour="grey30", data = data.frame(x = 0:5+.5, xend= 0:5+.5)) + 
-  geom_segment(aes(y=y, yend=yend), x = 0.5, xend=5.5, colour="grey30", data = data.frame(y = 0:5+.5, yend= 0:5+.5)) + 
-  scale_x_continuous(breaks = 1:5, labels = 1:5, limits = c(0.5,5.5)) +
-  scale_y_continuous(breaks = 1:5, labels = 1:5, limits = c(0.5,5.5)) +
+  geom_tile(aes(x = from, y = -to), color = 'grey20', fill = 'black') +
+  #geom_tile(aes(x = to, y = from), color = 'grey20', fill = 'black') +
+  geom_segment(aes(x = x, xend = xend), y = -0.5, yend=-5.5, colour="grey30", data = data.frame(x = 0:5+.5, xend= 0:5+.5)) + 
+  geom_segment(aes(y=-y, yend=-yend), x = 0.5, xend=5.5, colour="grey30", data = data.frame(y = -0:5+.5, yend= -0:5+.5)) + 
+  scale_x_continuous(breaks = 1:5, labels = 1:5, limits = c(0.5,5.5), position = "top") +
+  scale_y_continuous(breaks = (-5):(-1), labels = 5:1, limits = c(-5.5,-.5)) +
   theme_bw() + 
@@
   ylab("to (alter)")
-@
- \hh{XXX Explain each of these visualizations and either comment on the differences in the visualizations (colors) or don't show them. }
+
+@
+% \hh{XXX Explain each of these visualizations and either comment on the differences in the visualizations (colors) or don't show them. }
+
+Each type of visualization comes with its own advantages and disadvantages.
+Paths between nodes are easier to determine with node-link diagrams than with adjacency matrix visualizations.
+In node-link diagrams, node information can be incorporated into the visualization by coloring or changing the shape of the points representing the nodes.
+This is not necessarily possible in matrix representation, which is more focused on edges.
+Adjacency matrix visualization is particularly useful, however, when the network is very complex, dense, or large.
+Experimental studies have shown this to be true.
+For example, with seven basic perceptual tasks on networks, including things like node and edge count, adjacency matrix visualizations outperform node-link diagrams as the size and density of the network increases \citep{adjmatviz}.
+One drawback of the adjacency matrix visualization that \citeauthor{adjmatviz} found was that edges are overrepresented for undirected graphs, due to the symmetry of $\mathbf{A}$.
+There is not one correct way to visualize network information, and we will be using both of the methods throughout this paper.
 

### 2017-06-06 paper.Rnw
add some edits. some comments i don't quite understand

@@
 HH: we are still interested in this.
-\hh{The next sentence is very vague - make it more specific: what aspect of SAOMs is not well understood?
+%\hh{The next sentence is very vague - make it more specific: what aspect of SAOMs is not well understood?
 I think what you want is something like ``SAOMs are not very tractable analytically with traditional methods, e.g.
@@
 Unlike other network models, SAOMs are not very well understood.
-Traditional modeling steps, like determining the likelihood function, are very complex.
-Because of this, SAOMs are fit to data using a series of MCMC phases for finding method of moments estimators.
+They are relatively new, especially compared to the classic exponential random graph models, and they are not very tractable analytically with traditional methods.
+Likelihood functions turn into very complex objects due to the dependency structure inherent in the data.
+Therefore, computationally more tractable solutions are used to fit estimators, in particular, % Traditional modeling steps, like determining the likelihood function, are very complex.
+Because of this,
+SAOMs are fit to data using a series of MCMC phases for finding method of moments estimators.
 Rather than treat SAOM fits and the fitting process as black boxes, we combine the principals of network visualization with those of model visualization as discussed in \citet{modelvis}.
@@
 
-For the network $x$ and each actor $i$ in the network, the number of times that an actor $i$ gets to change its ties, $x_{ij}$, to other nodes $j \neq i$ in the network is dictated by a \emph{rate function} $\alpha$.
-For this paper, we assume a simple rate function, $\alpha_m$ that is constant for all actors between observations at time $t_m$ to $t_{m+1}$.
-\hh{XXX that means that the function is essentially just a rate parameter, right?}
-Other modelling scenarios allow this rate to be more flexible,  e.g.\  be a function that depends on the time period of observation, some actor-level covariates or some actor-level network statistics.
-\hh{How is the rate parameter different from a rate parameter?}
-The rate parameter determines how quickly actor $i$ gets an opportunity to change one of its ties, $x_{ij}$ in the time period $t_{m} \leq T < t_{m+1}$.
-We assume that the actors $i$ are conditionally independent given their current ties, $x_{i1}, \dots, x_{in}$.
-This assumption leads to the rate for the whole network, $n\alpha$.
-In order to achieve the memorylessness property of a Markov process, for any time point, $T$, where $t_{m} \leq T < t_{m+1}$, the waiting time to the next change opportunity by actor $i$ is exponentially distributed with expected value $\alpha_m^{-1}$.
-Thus, the waiting time to the next change opportunity by \textit{any} actor in the network is also exponentially distributed with expected value $(n\alpha_m)^{-1}$
+For the network $x$ and each actor $i$ in the network, the number of times that an actor $i$ gets to change its ties, $x_{ij}$, to other nodes $j \neq i$ in the network is dictated by a \emph{rate function} $\rho(x, \mathbf{z}, \boldsymbol{\alpha})$.
+For this paper, we assume a simple rate function, $\rho(x, \mathbf{z}, \boldsymbol{\alpha}) = \alpha_m$ that is constant for all actors between observations at time $t_m$ to $t_{m+1}$.
+%\hh{XXX that means that the function is essentially just a rate parameter, right?}
+Other modelling scenarios allow this rate to be more flexible,  e.g.\ a function that depends on the time period of observation, some actor-level covariates or some actor-level network statistics. 
+%\hh{How is the rate parameter different from a rate parameter?}
+In our model with the simple rate function, the rate parameter determines how quickly actor $i$ gets an opportunity to change one of its ties, $x_{ij}$ in the time period from $t_{m}$ to $t_{m+1}$.
+The model also assumes that the actors $i$ are conditionally independent given their ties, $x_{i1}, \dots, x_{in}$ at the current network state.
+\st{Let $\tau_i(m)$ be the wait time until actor $i$ makes its next change.For any time point, $T$, where $t_{m} \leq T < t_{m+1}$, the waiting time to the next change opportunity by actor $i$ is exponentially distributed with expected value $\alpha_m^{-1}$.
+The conditional independence assumption is expressed in Equation~\ref{eq:c1}.
+
+\begin{equation} \label{eq:c1}
+ \tau_i(m) | x_{i1}(m), \dots, x_{in}(m) \stackrel{\text{iid}}{\sim} Exp(\alpha_m)
+\end{equation}
+
+ The waiting time to the next change opportunity by \textit{any} actor in the network is also exponentially distributed with expected value $(n\alpha_m)^{-1}$.
+The waiting time for the whole network to change, $\tau(m) = \sum_i \tau_i(m) |x_{i1}(m), \dots, x_{in}(m) $ can then be written as
+
+\begin{equation} \label{eq:c2}
+ \tau(m) \sim Exp(n\alpha_m)
+\end{equation}
+
+} The parameter for the wait time for the whole network $n\alpha_m$ is the rate at which any tie change occurs. %In order to achieve the memorylessness property of a Markov process,
 
@@
 \hh{shouldn't that be independence? - make the conditional independence assumption above a labelled formula (e.g. C1) and then call it here }
-Because of the conditional dependence assumptions in the model, we can consider the objective function for each node separately, since only one tie from one node is changing at a time.
-The objective function is then written as \hh{the sum of the individual changes XXX write out the math in words XXX} $$f_i(\boldsymbol{\beta}, x) = \sum_k \beta_k s_{ik}(x, \mathbf{Z}),$$
+Because of the conditional independence assumptions in Equation~\ref{eq:c1}, we can consider the objective function for each node separately, since only one tie from one node is changing at a time.
+The objective function is then written as
+\begin{equation} \label{eq:of1}
+f_i(\boldsymbol{\beta}, x) = \sum_k \beta_k s_{ik}(x, \mathbf{Z}),
+\end{equation}
+
 for $x \in \mathcal{X}$, the space of all possible directed networks with the $n$ nodes, and $\mathbf{Z}$ matrix of covariates.
@@
 The parameters, $\boldsymbol{\beta}$, are attached to various actor-level network statistics, $s_{ik}(x)$.
-\hh{'There are always' ...
+%\hh{'There are always' ...
 that seems to come from a Snijder paper?
-Re-phrase this to be less vague: In the current implementation of RSiena (snijder) a model has minimally two parameters...} There are always at least two parameters, $\beta_1$ for the outdegree of a node, and $\beta_2$ for the number of reciprocal ties held by a node \citep[p.
-371]{snijders01}.
-\hh{'There are many' another vague phrase} There are many possible parameters $\beta_i$ to add to the model.
-\hh{'They' ...
-be specific.} They can be split up into two groups: first, the structural effects, which only depend on the structure of the network.
-\st{The inclusion of these effects has origin in the classical exponential random graph model (ERGM) for networks.
-These effects are written in terms of the edge variables $x_{ij}$, for $i \neq j$.
-The second set of effects are the actor-level or covariate effects.} \hh{XXX move the last colored section to the front of this paragraph.
+Re-phrase this to be less vague: In the current implementation of RSiena (snijder) a model has minimally two parameters...}
+According to \citet[p.
+371]{snijders01}, there should be at least two parameters included in the model: $\beta_1$ for the outdegree of a node, and $\beta_2$ for the number of reciprocal ties held by a node \st{The inclusion of these effects has origin in the classical exponential random graph model (ERGM) for networks.
+The statistics corresponding to these effects are written in terms of the edge variables $x_{ij}$, for $i \neq j$.} In the \texttt{RSiena} software, there are over 80 possible parameters to add to the model.
+The formulas for the effects are provided in \citet{RSienamanual}.
+The parameters, $\beta_k$, in the model can be split up into two groups: first, the structural effects, whose estimation only depend on the structure of the network, like the outdegree and reciprocity parameters mentioned above.
+%\hh{XXX move the last colored section to the front of this paragraph.
 That sets up the structure nicely, e.g.
-`In determining the covariates of a SAOM, two sources of parameters are considered: ...} These effects also depend on the structure of the network.
-They are written in terms of $x_{ij}$ but also in terms of the covariates, $\mathbf{Z}$.
+`In determining the covariates of a SAOM, two sources of parameters are considered: ...}
+The second set of effects are referred to as the actor-level or covariate effects.
+These covariate effects also depend on the structure of the network, and are written in terms of the tie variables $x_{ij}$, but also in terms of the covariates, $\mathbf{Z}$.
 A table of some possible structural and covariate effects is given in Table~\ref{tab:effects}.
-For a complete list of the network and covariate statistics that can be included \hh{currently} in the objective function, see \citet{RSiena}.
+For a complete list of the network and covariate statistics that can currently be included in the objective function, see \citet{RSienamanual}.
 

### 2017-06-06 paper.Rnw
more edits.

@@
 
-When node $i$ is given the chance to change a tie, \hh{`we assume that they wish' sounds strange - don't assume wishes, state assumptions} we assume that they wish to maximize the value of their objective function $f_i(\boldsymbol{\beta}, x)$ plus a random element, $U_i(x)$, where the $U_i(x)$ are from \hh{XXX don't quote here - just state that Snijder suggest the use of a type 1 extreme value distribution on top of the objective function.
-XX What is the thought behind that?
+When node $i$ is given the chance to change a tie, %\hh{`we assume that they wish' sounds strange - don't assume wishes, state assumptions}
+it attempts to maximize the value of its objective function $f_i(\boldsymbol{\beta}, x)$ plus a random element, $U_i(x)$, to account for unknown attraction between nodes.
+In \citet{modelsSnijders}, the author recommends that the $U_i(x)$ be random draws from a type 1 extreme value distribution on top of the objective function.
+% XX What is the thought behind that?
 } ``the type 1 extreme value distribution (or Gumbel distribution) with mean 0 and scale parameter 1" \citep[p.
@@
 
-$$f(u|\mu, \sigma) = \frac{1}{\sigma}\exp\left\{-\left(\frac{u-\mu}{\sigma} + e^{-\frac{u-\mu}{\sigma}}\right)\right\}.$$
-
-Using this distribution \hh{XXX I'm not following - which distribution do you mean by `this' distribution?
+\begin{equation} \label{eq:dist1}
+f(u|\mu, \sigma) = \frac{1}{\sigma}\exp\left\{-\left(\frac{u-\mu}{\sigma} + e^{-\frac{u-\mu}{\sigma}}\right)\right\}.
+\end{equation}
+
+Using the distribution given in Equation~\ref{dist1} with mean $\mu = 0$ and scale $\sigma = 1$ as \citet{modelsSnijders} suggests %\hh{XXX I'm not following - which distribution do you mean by `this' distribution?
 The density above is the type 1 extreme value density.
-I got lost at the  ``objective function $f_i(\boldsymbol{\beta}, x)$ plus a random element, $U_i(x)$"} is convenient because it allows the probability the actor $i$ chooses to change its tie to actor $j$ in terms of the objective function alone.
+I got lost at the  ``objective function $f_i(\boldsymbol{\beta}, x)$ plus a random element, $U_i(x)$"}
+is convenient because it allows the probability the actor $i$ chooses to change its tie to actor $j$ in terms of the objective function alone.
 Let $p_{ij}(\boldsymbol{\beta}, x)$ be this probability.

### 2017-06-30 PaperToDo.Rmd
add a to-do list

---
title: "To do list for this paper"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Broadly

- [ ] Write an abstract
- [ ] Submit to journal

## Introduction

- [ ] Proof read

## Section 2

- [ ] Proof read
- [ ] Make sure I've defined ego/alter nodes

## Section 3

- [ ] Proof read
- [ ] Finish section 3.1 (elaborate on probability)

## Section 4

- [ ] Add histogram of number of tie changes from $t_1$ to $t_2$ in simulations 
- [ ] Make gif available online
- [ ] Insert link to full gif in paper
- [ ] Figure 6: reverse order of adjacency matrix y-axis
- [ ] Figure 7 not showing up??
- [ ] Fix error on p. 21 (probably related to above)
- [ ] Heatmap p 23 missing caption
- [ ] Describe Figure 8 more. What is going on??
- [ ] Add geomnet example in 4.2
- [ ] p. 26-28 figure refs & code is broken
- [ ] Get rid of repeated plot p 29
- [ ] p. 34 code is broken
- [ ] General proof reading

## Discussion

- [ ] Write up conclusions
- [ ] Discuss future work

### 2017-06-30 paper.Rnw
add a to-do list

@@
 
-\hh{shouldn't that be independence? - make the conditional independence assumption above a labelled formula (e.g. C1) and then call it here }
+%\hh{shouldn't that be independence? - make the conditional independence assumption above a labelled formula (e.g. C1) and then call it here }
 Because of the conditional independence assumptions in Equation~\ref{eq:c1}, we can consider the objective function for each node separately, since only one tie from one node is changing at a time.
@@
 
-The top row of Figure~\ref{makinggif} shows an edge being removed, and the bottom row shows one being added.
+The top row of Figure~\ref{fig:makinggif} shows an edge being removed, and the bottom row shows one being added.
 In both cases, the ego nodes chosen to act change color from black to red, and they also increase greatly in size.

### 2017-07-05 PaperToDo.Rmd
baby steps

@@
 - [ ] Insert link to full gif in paper
-- [ ] Figure 6: reverse order of adjacency matrix y-axis
+- [X] Figure 6: reverse order of adjacency matrix y-axis
 - [ ] Figure 7 not showing up??

### 2017-07-05 paper.Rnw
baby steps

@@
 numordersf$from <- factor(numordersf$from, levels = paste0("V", 1:16))
-numordersf$to <- factor(numordersf$to, levels = paste0("V", 1:16))
+numordersf$to <- factor(numordersf$to, levels = rev(paste0("V", 1:16)))
 
@@
   theme(legend.position = 'none', axis.text = element_text(size = 8)) + 
+  scale_x_discrete(position = "top") + 
   coord_fixed()
@@
 pcaordersf$from <- factor(pcaordersf$from, levels = order1)
-pcaordersf$to <- factor(pcaordersf$to, levels = order1)
+pcaordersf$to <- factor(pcaordersf$to, levels = rev(order1))
 

### 2017-07-28 paper.Rnw
uncomment all the \hh stuff that i commented out previously and start a conversation.

@@
 % }
-%\hh{XXX - talk a bit about social networks first, why we want to model them, and that usually the models are not dynamic.}
-
+\hh{XXX - talk a bit about social networks first, why we want to model them, and that usually the models are not dynamic.}
+\st{XXX - okay, here is what i have done for that:
 Social networks have been studied for decades, beginning with a few foundational works, the most well known of which is the 1967 study, "The Small World Problem" by Stanley Milgram \citep{goldenberg09}.
@@
 In recent years, the study of social networks has grown in popularity due to an increase in the availability and access to social network data.
-There are many kinds of social networks, but there are not as many statistical models for social network data.
-Some network models that have been applied to social networks include the exponential random graph model and latent space models.
+There are many kinds of social networks, but there are not as many statistical models for social network data.} Some network models that have been applied to social networks include the exponential random graph model and latent space models.
 These models, however, are only for single instance networks.
@@
 HH: we are still interested in this.
-%\hh{The next sentence is very vague - make it more specific: what aspect of SAOMs is not well understood?
+\hh{The next sentence is very vague - make it more specific: what aspect of SAOMs is not well understood?
 I think what you want is something like ``SAOMs are not very tractable analytically with traditional methods, e.g.
@@
 Unlike other network models, SAOMs are not very well understood.
-They are relatively new, especially compared to the classic exponential random graph models, and they are not very tractable analytically with traditional methods.
+\st{They are relatively new, especially compared to the classic exponential random graph models, and they are not very tractable analytically with traditional methods.
 Likelihood functions turn into very complex objects due to the dependency structure inherent in the data.
@@
 Because of this,
-SAOMs are fit to data using a series of MCMC phases for finding method of moments estimators.
-Rather than treat SAOM fits and the fitting process as black boxes, we combine the principals of network visualization with those of model visualization as discussed in \citet{modelvis}.
+SAOMs are fit to data using a series of MCMC phases for finding method of moments estimators.} Rather than treat SAOM fits and the fitting process as black boxes, we combine the principals of network visualization with those of model visualization as discussed in \citet{modelvis}.
 %Here, we explore RSiena and SAOMs with innovative network visualizations, both static and dynamic.
@@
 \hh{XXX how does the adjacency matrix feature here?
-The adjacency matrix is  a different mathematical representation of a network, but describes the same network as a node-link diagram.}
+The adjacency matrix is a different mathematical representation of a network, but describes the same network as a node-link diagram.} \st{XXX I'm taking this to mean I should describe adjacency matrices more, so I've added a discussion below.
+XXX }
 Some commonly used layout algorithms, such as  the Kamada-Kawai layout \citep{kamadakawai} and the Fruchterman-Reingold layout \citep{fruchterman-reingold}, treat networks like physical systems.
@@
 
-The second primary method for network visualization uses the adjacency matrix of the network.
+\st{The second primary method for network visualization uses the adjacency matrix of the network.
 The adjaceny matrix of a network, $\mathbf{A}$, describes the edges of a network in matrix form.
@@
 In an undirected network, $A_{ij} = A_{ji}$, but in a directed graph this is only true if there is an edge from $i$ to $j$ and from $j$ to $i$.
-Thus, $\mathbf{A}$ is always symmetric for undirected networks, and is symmetric for directed networks only if every edge between two nodes is reciprocated.
-%\hh{XXX what is an adjacency matrix? - explain at this point in mathematical terms} 
+Thus, $\mathbf{A}$ is always symmetric for undirected networks, and is symmetric for directed networks only if every edge between two nodes is reciprocated.}
+\hh{XXX what is an adjacency matrix? - explain at this point in mathematical terms} \st{XXX see above XXX}
 This type of visualization is also shown in Figure~\ref{fig:toyplots} for our toy example.  
-%\hh{XXX What is the exact logic here that Ghoniem used to come to the conclusion that edges in undirected graphs are over-represented?
-Why would directed graphs make that better?
-wouldn't that just add to the number of edges?
-} This, however, may actually be an advantage for \emph{directed} graphs, where exactly the correct number of edges is represented in a matrix visualization, but a node-link diagram may underrepresent the edge count with overlapping edges.
-
 <<toyplots, fig.show='hold', fig.cap="On the left, a node-link diagram of our directed toy network, with nodes placed using the Kamada-Kawai algorithm.
@@
 @
-% \hh{XXX Explain each of these visualizations and either comment on the differences in the visualizations (colors) or don't show them. }
+\hh{XXX Explain each of these visualizations and either comment on the differences in the visualizations (colors) or don't show them. }
 
@@
 For example, with seven basic perceptual tasks on networks, including things like node and edge count, adjacency matrix visualizations outperform node-link diagrams as the size and density of the network increases \citep{adjmatviz}.
-One drawback of the adjacency matrix visualization that \citeauthor{adjmatviz} found was that edges are overrepresented for undirected graphs, due to the symmetry of $\mathbf{A}$.
-There is not one correct way to visualize network information, and we will be using both of the methods throughout this paper.
+One drawback of the adjacency matrix visualization that \citeauthor{adjmatviz} found was that edges are overrepresented for undirected graphs, due to the symmetry of $\mathbf{A}$: the edge $x_{ij}$ for $i \neq j$ appears in $A$ twice: in $A_{ij}$ and $A_{ji}$.
+\hh{XXX What is the exact logic here that Ghoniem used to come to the conclusion that edges in undirected graphs are over-represented?
+Why would directed graphs make that better?
+wouldn't that just add to the number of edges?
+} \st{XXX in undirected nets, the adj mat is symmetric, where one edge ($x_{ij}$) is represented in entry $i,j$ in the matrix and in entry $j,i$.
+So, there are many repetitions in the matrix, which makes the network appear more dense than it really is.} \st{This, however, may actually be an advantage for \emph{directed} graphs, where exactly the correct number of edges is represented in a matrix visualization, due to the fact that the edges $x_{ij}$ and $x_{ji}$ are not interchangeable.
+A node-link diagram, however, may underrepresent the edge count if the edges $x_{ij}$ and $x_{ji}$ both exist and are drawn on top of one another.
+Ultimately, t}here is not one "correct" way to visualize network information, and we will be using both \st{the node-link and adjacency matrix visualization} methods throughout this paper.
 
@@
 
-%\hh{mostly lit review plus a detailed description of the model and fitting process.}
+\hh{mostly lit review plus a detailed description of the model and fitting process.}
 
@@
 For the network $x$ and each actor $i$ in the network, the number of times that an actor $i$ gets to change its ties, $x_{ij}$, to other nodes $j \neq i$ in the network is dictated by a \emph{rate function} $\rho(x, \mathbf{z}, \boldsymbol{\alpha})$.
-For this paper, we assume a simple rate function, $\rho(x, \mathbf{z}, \boldsymbol{\alpha}) = \alpha_m$ that is constant for all actors between observations at time $t_m$ to $t_{m+1}$.
-%\hh{XXX that means that the function is essentially just a rate parameter, right?}
+For this paper, we assume a simple rate \st{"function"}, $\rho(x, \mathbf{z}, \boldsymbol{\alpha}) = \alpha_m$ that is constant for all actors between observations at time $t_m$ to $t_{m+1}$, \st{thus our rate function is just a rate parameter in the model}.
+\hh{XXX that means that the function is essentially just a rate parameter, right?} \st{XXX yes XXX}
 Other modelling scenarios allow this rate to be more flexible,  e.g.\ a function that depends on the time period of observation, some actor-level covariates or some actor-level network statistics. 
-%\hh{How is the rate parameter different from a rate parameter?}
-In our model with the simple rate function, the rate parameter determines how quickly actor $i$ gets an opportunity to change one of its ties, $x_{ij}$ in the time period from $t_{m}$ to $t_{m+1}$.
+\hh{How is the rate parameter different from a rate parameter?} \st{I think you meant how is a rate fn different from a rate parameter?
+I updated the language XXX->}%In our model with the simple rate function,
+\st{In our simple model with a rate parameter instead of a rate function,} the rate parameter determines how quickly actor $i$ gets an opportunity to change one of its ties, $x_{ij}$ in the time period from $t_{m}$ to $t_{m+1}$.
 The model also assumes that the actors $i$ are conditionally independent given their ties, $x_{i1}, \dots, x_{in}$ at the current network state.
@@
 \hh{XXX we tried to estimate the rate based on the simulations - you could include a histogram of the number of changes for the 1000 simulations in the viz part XXX}
+\st{XXX I have the fitted models from the 1000 simulations (see Figure \ref{fig:distests}). Is that what you mean? XXX}
 
@@
 
-%\hh{shouldn't that be independence? - make the conditional independence assumption above a labelled formula (e.g. C1) and then call it here }
-Because of the conditional independence assumptions in Equation~\ref{eq:c1}, we can consider the objective function for each node separately, since only one tie from one node is changing at a time.
-The objective function is then written as
+\hh{shouldn't that be independence? - make the conditional independence assumption above a labelled formula (e.g. C1) and then call it here }
+Because of the conditional \st{in}dependence assumptions in Equation~\ref{eq:c1}, we can consider the objective function for each node separately, since only one tie from one node is \st{allow to change} at a time.
+\st{The node $i$, which is the node that is chosen to change at the current time point, is called the \textit{ego} node.
+It has the potential to interact with all other nodes in the network, $j \neq i$.
+These nodes $j$, are referred to as \textit{alters}.
+These nodes are acted upon by the ego node, and they only act when they become the ego node at a subsequent time point.} So, for the ego node, $i$, in the current network state $x$, the objective function is then written as
 \begin{equation} \label{eq:of1}
@@
 The vector $\boldsymbol{\beta}$ are the parameters of the model with corresponding network and covariate statistics, $s_{ik}(x, \mathbf{Z})$, for $k = 1,\dots, K$.
-\hh{XXX ego node isn't defined yet - the description of the charts would be a good place for it.} Given the ego node, $i$, there are $n$ possible steps for the actor $i$ to take: either one of all current ties $x_{ij} = 1$ will be destroyed, a new tie will be created, or no change will occur.
+\hh{XXX ego node isn't defined yet - the description of the charts would be a good place for it.} \st{XX I added it above the obj function equation.
+Is it ok there?} Given the ego node, $i$, there are $n$ possible steps for the actor $i$ to take: either one of all current ties $x_{ij} = 1$ will be destroyed, a new tie will be created, or no change will occur.
 
@@
 The parameters, $\beta_k$, in the model can be split up into two groups: first, the structural effects, whose estimation only depend on the structure of the network, like the outdegree and reciprocity parameters mentioned above.
-%\hh{XXX move the last colored section to the front of this paragraph.
+\hh{XXX move the last colored section to the front of this paragraph.
 That sets up the structure nicely, e.g.
-`In determining the covariates of a SAOM, two sources of parameters are considered: ...}
+`In determining the covariates of a SAOM, two sources of parameters are considered: ...} \st{XXX as I'm going through and uncommenting your comments (sorry....) this one has lost meaning a little bit.
+Could you give me more detail?
+I'm not sure which section is which at this point.
+(sorry again...)}
 The second set of effects are referred to as the actor-level or covariate effects.
@@
 
-Using the distribution given in Equation~\ref{dist1} with mean $\mu = 0$ and scale $\sigma = 1$ as \citet{modelsSnijders} suggests %\hh{XXX I'm not following - which distribution do you mean by `this' distribution?
+Using the distribution given in Equation~\ref{dist1} with mean $\mu = 0$ and scale $\sigma = 1$ as \citet{modelsSnijders} suggests \hh{XXX I'm not following - which distribution do you mean by `this' distribution?
 The density above is the type 1 extreme value density.
-I got lost at the  ``objective function $f_i(\boldsymbol{\beta}, x)$ plus a random element, $U_i(x)$"}
-is convenient because it allows the probability the actor $i$ chooses to change its tie to actor $j$ in terms of the objective function alone.
-Let $p_{ij}(\boldsymbol{\beta}, x)$ be this probability.
-Next, write the network $x$ in its potential future state, where the tie $x_{ij}$ has changed to $1-x_{ij}$, as $x(i \leadsto j)$.
+I got lost at the  ``objective function $f_i(\boldsymbol{\beta}, x)$ plus a random element, $U_i(x)$"} \st{XXX I hope it is clearer now above}
+is convenient because its \st{properties lead to a probability formula for} the probability that actor $i$ chooses to change its tie to actor $j$ \st{that can be written only} in terms of the objective function.
+Let $p_{ij}(\boldsymbol{\beta}, x)$ be the probability that actor $i$ chooses to change its tie to actor $j$.
+Next, we write the network $x$ in its potential future state, where the tie $x_{ij}$ has changed to $1-x_{ij}$, as $x(i \leadsto j)$.
 Then, the probility that the tie $x_{ij}$ changes is
@@
 
-\hh{This paragraph ends a bit abruptly - what is the future state probability used for?}
+\hh{XXX This paragraph ends a bit abruptly - what is the future state probability used for?} \st{XX The future state probability is used within the continuous time markov chain to move through the different network states.
+This needs more fleshing out.
+Adding to the to-do list.}
+
+\subsubsection{Continuous Time Markov Chain}
+
+\st{XXX TO DO XXX}
 

### 2017-07-29 paper.Rnw
need to have most recent version of netvizif package to compile paper. also need dplyr version 0.5.0 for the unexported function match_data_frame

@@
 require(knitr)
-opts_chunk$set(fig.path='figure/', cache.path='cache/', fig.align='center', fig.width=5, fig.height=5, par=TRUE, cache=TRUE, concordance=TRUE, autodep=TRUE, message=F, warning=F, echo = FALSE, root.dir = "~/Desktop/Dissertation - LETS GO/SAOM-removing-blindfold/")
+opts_chunk$set(fig.path='figure/', cache.path='cache/', fig.align='center', fig.pos='h', fig.width=5, fig.height=5, par=TRUE, cache=TRUE, concordance=TRUE, autodep=TRUE, message=F, warning=F, echo = FALSE, root.dir = "~/Desktop/Dissertation - LETS GO/SAOM-removing-blindfold/")
 @
@@
   ylab("to (alter)")
-
 @
@@
 
-Using the distribution given in Equation~\ref{dist1} with mean $\mu = 0$ and scale $\sigma = 1$ as \citet{modelsSnijders} suggests \hh{XXX I'm not following - which distribution do you mean by `this' distribution?
+Using the distribution given in Equation~\ref{eq:dist1} with mean $\mu = 0$ and scale $\sigma = 1$ as \citet{modelsSnijders} suggests \hh{XXX I'm not following - which distribution do you mean by `this' distribution?
 The density above is the type 1 extreme value density.
@@
 set.seed(823746)
+.libPaths(new = c("/Library/Frameworks/R.framework/Versions/3.4/Resources/library", "/Library/Frameworks/R.framework/Versions/3.4/Resources/oldpkgs/"))
 ansnullchains <- get_chain_info(ansnull)
@@
 
-<<histogramalters, fig.width = 8, fig.height=11, fig.cap="Each panel shows the theoretical (as bars) and empirical (as points) probabilities of the chosen ego node changing its tie to each of the other nodes.">>=
+<<histogramalters, fig.width = 8, fig.height=11, fig.cap="Each panel shows the theoretical (as lines) and empirical (as points) probabilities of the chosen ego node changing its tie to each of the other nodes.
+The color of the line indicates whether the tie from the ego to the alter node is being added, removed, or if there is no change to the network in this step.">>=
 ansnullchains %>% group_by(period, rep) %>% 
@@
 
-cumprobs %>% left_join(changes2) %>% ggplot() + 
-  geom_histogram(
-    aes(x = alter, weight = alterProb, fill = type), 
-    color = 'black', fill = 'white', binwidth = 1) + 
-  geom_point(aes(x = alter, y = empprob), color = 'red') + 
-  scale_x_continuous(labels = paste0("V",1:16), breaks = 0:15) + coord_flip() + facet_wrap(~ego, labeller = "label_both") + theme_gray()
+cumprobs %>% left_join(changes2) -> changes3
+
+changes3$ego <- factor(changes3$ego, levels = paste0("V", 1:16), labels = 1:16) 
+changes3$alter <- factor(changes3$alter, levels = paste0("V", 1:16), labels = 1:16)
+
+ggplot(data = changes3) + 
+  geom_segment(
+    aes(x = alter , xend = alter, y = 0, yend = alterProb, color=type)) + 
+  geom_point(aes(x = alter, y = empprob), shape = 4, size = 1, alpha = .5) + 
+  labs(x = "alter node", y = "probability of alter node") + 
+  coord_flip() + 
+  facet_wrap(~ego, labeller = "label_both") + 
+  theme_bw()
 # 
@@
 XXX
-<<heatmap>>=
+<<heatmap, fig.align='center', fig.cap='A heatmap showing the empirical transition probabilities for the first microstep in 1,000 simulations.
+The acting node is on the x-axis, and the receiving node is on the y-axis.
+Missing squares represent ties that did not occur.'>>=
 ansnullchains %>% 
@@
          alter = paste0("V",as.numeric(as.character(alter))+1), 
-         prob = exp(logAlterProb)) %>% 
-  ggplot(aes(x = factor(ego, levels = order1), 
+         prob = exp(logAlterProb)) -> probsDat
+
+ggplot(data = probsDat, 
+       aes(x = factor(ego, levels = order1), 
              y = factor(alter, levels = rev(order1)), 
@@
   labs(x = "ego", y = "alter")+
+  scale_x_discrete(position = 'top') + 
   coord_fixed()
@@
 
-<<summnetM1>>=
+% below chunk contains some repetitive stuff. Haven't figured out what it's purpose is yet. 
+<<summnetM1, message=FALSE>>=
 friendshipData <- array(c(fd2.w1, fd2.w2, fd2.w3), dim = c(16, 16, 3))
@@
 M3eff <- includeEffects(M1eff, nbrDist2twice)
+simu2 <- read.csv("data/simulation-1000-M1-M2-M3.csv")
 means <- simu2 %>% group_by(Model, parameter) %>% summarize(
@@
 set.seed(4231352)
-M1sims1000 <- saom_simulate(dat = mydata, 
+M1sims1000 <- netvizinf::saom_simulate(dat = mydata, 
                             struct = M1eff,
@@
                        N = 1000)
-M2sims1000 <- saom_simulate(dat = mydata, 
+M2sims1000 <- netvizinf::saom_simulate(dat = mydata, 
                             struct = M2eff,
@@
                        N = 1000)
-M3sims1000 <- saom_simulate(dat = mydata, 
+M3sims1000 <- netvizinf::saom_simulate(dat = mydata, 
                             struct = M3eff,
@@
 
-simu2 <- read.csv("data/simulation-1000-M1-M2-M3.csv")
+#simu2 <- read.csv("data/simulation-1000-M1-M2-M3.csv")
 

### 2017-08-05 PaperToDo.Rmd
uncomment all the \hh stuff that i commented out previously and start a conversation.
fix issue that was broken when dplyr got updated; find some mistakes and fix them; add some more figure descriptions

@@
 - [ ] Proof read
-- [ ] Make sure I've defined ego/alter nodes
+- [X] Make sure I've defined ego/alter nodes
+- [ ] Need to talk about the CTMC, rest of the SAOM model definition after the objective function discussion.
 
@@
 - [X] Figure 6: reverse order of adjacency matrix y-axis
-- [ ] Figure 7 not showing up??
-- [ ] Fix error on p. 21 (probably related to above)
-- [ ] Heatmap p 23 missing caption
+- [X] Figure 7 not showing up??
+- [X] Fix error on p. 21 (probably related to above)
+- [X] Heatmap p 23 missing caption
 - [ ] Describe Figure 8 more. What is going on??

### 2017-08-05 paper.Rnw
fix issue that was broken when dplyr got updated; find some mistakes and fix them; add some more figure descriptions

@@
 require(knitr)
-opts_chunk$set(fig.path='figure/', cache.path='cache/', fig.align='center', fig.pos='h', fig.width=5, fig.height=5, par=TRUE, cache=TRUE, concordance=TRUE, autodep=TRUE, message=F, warning=F, echo = FALSE, root.dir = "~/Desktop/Dissertation - LETS GO/SAOM-removing-blindfold/")
+opts_chunk$set(fig.path='figure/', cache.path='cache/', fig.align='center', fig.pos='h', out.width='.8\\linewidth', par=TRUE, cache=TRUE, concordance=TRUE, autodep=TRUE, message=F, warning=F, echo = FALSE, root.dir = "~/Desktop/Dissertation - LETS GO/SAOM-removing-blindfold/")
 @
@@
 Unlike other network models, SAOMs are not very well understood.
-\st{They are relatively new, especially compared to the classic exponential random graph models, and they are not very tractable analytically with traditional methods.
+\st{They are relatively new, especially compared to the classic exponential random graph models, and they are not very tractable analytically.
 Likelihood functions turn into very complex objects due to the dependency structure inherent in the data.
@@
 <<makinggif, fig.cap='A selection of images in the microstep animation.
-The selected edges and nodes are emphasized by changing the size and the color, then when edges are removed, they fade out, shrinking in size, while the nodes change color and shrink to blend in with the other nodes.', fig.width=8, cache=TRUE>>=
+The selected edges and nodes are emphasized by changing the size and the color, then when edges are removed, they fade out, shrinking in size, while the nodes change color and shrink to blend in with the other nodes.', fig.width=8, cache=TRUE, message = FALSE, warning = FALSE>>=
+# Need old version of dplyr to get this to work: 
+if (!require("dplyr050",character.only = TRUE)) {
+      devtools::install_github("sctyner/dplyr050")
+}
+
 friend.data.w1 <- as.matrix(read.table("data/s50_data/s50-network1.dat"))
@@
 set.seed(823746)
-.libPaths(new = c("/Library/Frameworks/R.framework/Versions/3.4/Resources/library", "/Library/Frameworks/R.framework/Versions/3.4/Resources/oldpkgs/"))
+
 ansnullchains <- get_chain_info(ansnull)
@@
 We then performed a principal component analysis on $\mathbf{A}^{cum}$, and used the values of the first principal component to order the vertices on the $x$ and $y$ axes for the adjacency matrix animation.
-For one such series of microsteps simulated by \texttt{RSiena}, we present the adjacency matrix ordered by the (arbitrary) vertex id alongside the seriated adjacency microsteps using the first principal component loading on the cumulative adjacency matrix, $\mathbf{A}^{cum}$, in Figure~\ref{adjmatorder}.
-
-<<adjmatorder, fig.cap="On the left, the starting friendship network represented in adjacency matrix form, ordered by vertex id.
+For one such series of microsteps simulated by \texttt{RSiena}, we present the adjacency matrix ordered by the (arbitrary) vertex id alongside the seriated adjacency microsteps using the first principal component loading on the cumulative adjacency matrix, $\mathbf{A}^{cum}$, in Figure~\ref{fig:adjmatorder}.
+
+<<adjmatorder, message = FALSE, warning = FALSE, fig.cap="On the left, the starting friendship network represented in adjacency matrix form, ordered by vertex id.
 On the right, the same adjacency matrix is presented after ordering the vertices by one repitition of the microstep simulation process from wave one to wave two.", fig.width=8>>=
 # first present the original wave 1 adj mat, then present the adj mat after being ordered by the 1st pc of the big adjmat A. 
+# reinstall new dplyr
 numordersf <- data.frame(fd2.w1) %>% 
@@
 For network data, one way to do this is through the traditional node-link diagram.
-In Figure~\ref{fig:summnetM1}, we show a summary network created with 1000 simulations from Model 1.
-Only edges that appear in at least 51 of the 1000 simulations are included in the visualization, with edges that appear more frequently emphasized by thicker linewidths and a bright red color.
+In Figure~\ref{fig:summnetM1}, we show a summary network created with 1,000 simulations of the second wave of the network from Model 1.
+Only edges that appear in more than 5\% of the simulations (in at least 51 of the 1000 simulations) are included in the visualization, with edges that appear more frequently emphasized by thicker linewidths and a brighter red color.
 On either side of the summary network, we show the actual data, wave 1 on the left, and wave 2 on the right.
-Comparing the three visualizations side-by-side shows us that the summary network of simulations of wave 2 looks more like the true wave 2 than wave 1, which is a good indication that the model is capturing the true data structure.
+\st{We can see that the structure of the summary network is much more similar to the first network observation than to the second network observation.
+The simulations, however, are supposed to represent the second wave of data, which is shown on the right in Figure~\ref{fig:summnetM1}.
+This is an indication that our simple model, $M1$ is doing a very poor job of capturing the change mechanism from the first to the second wave of observation.} % XXX ST - I was looking at the wrong one.
+I had wave == 2 instead of wave == 1 in the code.
+This means I was looking at the wrong simulations, so they really do look more like the first data obs than the second obs.
+XXX Comparing the three visualizations side-by-side shows us that the summary network of simulations of wave 2 looks more like the true wave 2 than wave 1, which is a good indication that the model is capturing the true data structure.
 It's not ideal, however, because there appear to be too many reciprocated ties in the summary network, far more than there are in the data.
@@
 % below chunk contains some repetitive stuff. Haven't figured out what it's purpose is yet. 
-<<summnetM1, message=FALSE>>=
+<<summnetM1, message=FALSE, fig.cap='On the left, the first wave of observed data that is conditioned on in the model.
+On the right, the second wave of observed data.
+In the middle, a summary network from the first model fit to the data.
+This summary network represents 1,000 simulations of wave 2 using the values from the simple fitted model $M1$.',results='hide'>>=
 friendshipData <- array(c(fd2.w1, fd2.w2, fd2.w3), dim = c(16, 16, 3))
@@
 M3simsdf <- sims_to_df(M3sims1000)
-M1avgW2 <- M1simsdf %>% filter(!is.na(to) & wave == 2) %>%
+M1avgW2 <- M1simsdf %>% filter(!is.na(to) & wave == 1) %>%
   group_by(from, to) %>% 
@@
   mutate(weight = ifelse(from == to, 0, count))
-M2avgW2 <- M2simsdf %>% filter(!is.na(to) & wave == 2) %>%
+M2avgW2 <- M2simsdf %>% filter(!is.na(to) & wave == 1) %>%
   group_by(from, to) %>% 
@@
   mutate(weight = ifelse(from == to, 0, count))
-M3avgW2 <- M3simsdf %>% filter(!is.na(to) & wave == 2) %>%
+M3avgW2 <- M3simsdf %>% filter(!is.na(to) & wave == 1) %>%
   group_by(from, to) %>% 
@@
   geom_net(aes(from_id = from, to_id = to, linewidth= linewidth),
-           directed = T, curvature = .2, labelon=T, size = 5,
+           directed = T, curvature = .2, labelon=T, size = 4,
            arrow = arrow(angle = 20, type = "open", length = unit(.1, "inches")), ecolour = t1$color,
-           color = "#252525", hjust = .5, vjust = .5, labelcolour = 'white', fontsize = 3, arrowgap = .02,
+           color = "#252525", hjust = .5, vjust = .5, labelcolour = 'white', fontsize = 2, arrowgap = .02,
            singletons = T) +
-  theme_net() + labs(title = "Summary Network: M1") + theme(plot.background = element_rect(color = 'black'))
+  theme_net() + labs(title = "Summary Network: M1") + 
+  theme(plot.background = element_rect(color = 'black'),
+        plot.title = element_text(size = rel(.5)))
+
 p2 <- ggplot(data = filter(combinedavgs, cat == "First Observation")) +
   geom_net(aes(from_id = from, to_id = to, linewidth= linewidth),
-           directed = T, curvature = .2, labelon=T, size = 5,
+           directed = T, curvature = .2, labelon=T, size = 4,
            arrow = arrow(angle = 20, type = "open", length = unit(.1, "inches")), ecolour = "#969696",
-           color = "#252525", hjust = .5, vjust = .5, labelcolour = 'white', fontsize = 3, arrowgap = .02,
+           color = "#252525", hjust = .5, vjust = .5, labelcolour = 'white', fontsize = 2, arrowgap = .02,
            singletons = T) +
-  theme_net() + labs(title = "Wave 1") + theme(plot.background = element_rect(color = 'black'))
+  theme_net() + labs(title = "Wave 1") + 
+    theme(plot.background = element_rect(color = 'black'),
+        plot.title = element_text(size = rel(.5)))
 p3 <- ggplot(data = filter(combinedavgs, cat == "Second Observation")) +
   geom_net(aes(from_id = from, to_id = to, linewidth= linewidth),
-           directed = T, curvature = .2, labelon=T, size = 5,
+           directed = T, curvature = .2, labelon=T, size = 4,
            arrow = arrow(angle = 20, type = "open", length = unit(.1, "inches")), ecolour = "#969696",
-           color = "#252525", hjust = .5, vjust = .5, labelcolour = 'white', fontsize = 3, arrowgap = .02,
+           color = "#252525", hjust = .5, vjust = .5, labelcolour = 'white', fontsize = 2, arrowgap = .02,
            singletons = T) +
-  theme_net() + labs(title = "Wave 2")  + theme(plot.background = element_rect(color = 'black'))
+  theme_net() + labs(title = "Wave 2")  + 
+  theme(plot.background = element_rect(color = 'black'),
+        plot.title = element_text(size = rel(.5))) 
 grid.arrange(p2, p1, p3, nrow = 1)
@@
 To do this, we count occurences of each possible edge in the simulations, resulting in a summary network with weighted edges representing the number of times an edge appeared in the simulated wave 2 when simulating from the SAOM 1000 times.
-In Figure~\ref{averagenet}, we show the first wave, the ``average'' network from the three models we fit, and the second wave.
-Comparing the three averages to waves 1 and 2, we see that they have very similar structure to wave 2.
-Model 2, which included the transitive triplet parameter, seems to have created more group strucutre overall than models 1 and 3.
-This could be an indication that, despite the highly significant nature of the parameter $\beta_3$ according to the $t$-tests in \texttt{RSiena}, the transitive triplet parameter should not be included because it does not result in a structure similar to the data.
-
-<<averagenet>>=
+In Figure~\ref{fig:averagenet}, we show the first wave, the ``average'' network from the three models we fit, and the second wave.
+\st{Comparing the three averages to waves 1 and 2, we see that they have very similar structure to wave 1.
+Model 2, which included the transitive triplet parameter, seems to have created more group structure overall than models 1 and 3.
+In particular, if we look at the group of nodes $\{10,11,14\}$, we see they are very strongly connected within the three average networks, and they are completely separate from the other nodes in the true wave 2.
+None of the three average networks show node 16 gaining ties as it does in wave two, nor do they show nodes 4 and 7 becoming isolated.
+In Model 2, however, the ties to node 7 appear much weaker than in Model 1 or Model 2, suggesting that of the three, Model 2 is the best fit for our data.}  %XXXX ST commented out after I realized I had made the same mistake as figure 7.
+XX This could be an indication that, despite the highly significant nature of the parameter $\beta_3$ according to the $t$-tests in \texttt{RSiena}, the transitive triplet parameter should not be included because it does not result in a structure similar to the data.
+
+<<averagenet, results='hide', out.width='99%',fig.cap='The three "average" networks in the middle, with the actual network observations oon either side.
+There is some difference between the three models, but overall, these three models cannot capture the structure in the true second wave of data.
+'>>=
 #friend.data.w1 <- as.matrix(read.table("s50_data/s50-network1.dat"))
@@
 M3parms <- (means %>% filter(Model == "M3"))$means
-M1sims1000 <- saom_simulate(dat = mydata, 
-                            struct = M1eff,
-                       parms = as.numeric(M1parms), 
-                       N = 1000)
-M2sims1000 <- saom_simulate(dat = mydata, 
-                            struct = M2eff,
-                       parms = as.numeric(M2parms), 
-                       N = 1000)
-M3sims1000 <- saom_simulate(dat = mydata, 
-                            struct = M3eff,
-                       parms = as.numeric(M3parms), 
-                       N = 1000)
-M1simsdf <- sims_to_df(M1sims1000)
-M2simsdf <- sims_to_df(M2sims1000)
-M3simsdf <- sims_to_df(M3sims1000)
+# M1sims1000 <- saom_simulate(dat = mydata, 
+#                             struct = M1eff,
+#                        parms = as.numeric(M1parms), 
+#                        N = 1000)
+# M2sims1000 <- saom_simulate(dat = mydata, 
+#                             struct = M2eff,
+#                        parms = as.numeric(M2parms), 
+#                        N = 1000)
+# M3sims1000 <- saom_simulate(dat = mydata, 
+#                             struct = M3eff,
+#                        parms = as.numeric(M3parms), 
+#                        N = 1000)
+# M1simsdf <- sims_to_df(M1sims1000)
+# M2simsdf <- sims_to_df(M2sims1000)
+# M3simsdf <- sims_to_df(M3sims1000)
 M1avg <- M1simsdf %>% filter(!is.na(to) & wave == 1) %>%
@@
 # M3avg2$to <- paste0("V", M3avg2$to)
-
+combinedavgs$linewidth[combinedavgs$cat %in% c("First Observation", "Second Observation")] <- .5
 ggplot(data = combinedavgs) + 
   geom_net(aes(from_id = from, to_id = to, linewidth= linewidth),
-           directed = T, curvature = .2, labelon=T, size = 5,
-           arrow = arrow(angle = 20, type = "open", length = unit(.1, "inches")), ecolour = "#969696", fiteach = T, 
-           color = "#252525", hjust = .5, vjust = .5, labelcolour = 'white', fontsize = 3, arrowgap = .02,
-           singletons = F) + 
+           directed = T, curvature = .2, labelon=T, size = 4,
+           arrow = arrow(angle = 20, type = "open", length = unit(.2, "cm")), ecolour = "#969696", fiteach = T, 
+           color = "#252525", hjust = .5, vjust = .5, labelcolour = 'white', fontsize = 2, arrowgap = .02,
+           singletons = T) + 
   theme_net() + theme(panel.background = element_rect(color = 'black')) + 
-  facet_wrap(~cat)
+  facet_wrap(~cat, nrow = 1)
 
@@
 #            layout.par = list(var = 'user', dist = 'none', exp = 3,vm = distmat))
-
 @

### 2017-08-06 PaperToDo.Rmd
add some more to-dos and some to-dones.

@@
 - [ ] Submit to journal
+- [ ] Spacing around figures WAY to large above/below captions
 
@@
 - [ ] Add histogram of number of tie changes from $t_1$ to $t_2$ in simulations 
+- [ ] Add another example of a data set for M-in-DS (Sect 4.1) (GoT?)
 - [ ] Make gif available online
@@
 - [ ] Add geomnet example in 4.2
-- [ ] p. 26-28 figure refs & code is broken
-- [ ] Get rid of repeated plot p 29
-- [ ] p. 34 code is broken
+- [X] p. 26-28 figure refs & code is broken
+- [X] Get rid of repeated plot p 29
+- [X] p. 34 code is broken
 - [ ] General proof reading

### 2017-08-06 paper.Rnw
add some more to-dos and some to-dones.

@@
 <<adjmatviz, fig.align='center', fig.cap="A visualization of the adjacency matrices of the three waves of network observations in the ``Teenage Friends and Lifestyle Study'' data.
-The subset we will be using is outlined in red.", fig.width=8>>=
+The subset we will be using is outlined in red.", fig.width=8, results='hide',fig.show='asis'>>=
 source("code/00e_small_friends.R")
@@
            directed = T, curvature = .2, labelon=T, size = 4,
-           arrow = arrow(angle = 20, type = "open", length = unit(.1, "inches")), ecolour = "#969696",
+           arrow = arrow(angle = 20, type = "open", length = unit(.2, "cm")), ecolour = "#969696",
            color = "#252525", hjust = .5, vjust = .5, labelcolour = 'white', fontsize = 2, arrowgap = .02,
@@
 
-<<getdata, results='asis'>>=
+<<getdata>>=
 nulls <- read.csv("data/distribution_null_model.csv")

### 2017-08-12 PaperToDo.Rmd
senate data example

@@
 
+Don't spend any time placing the figures!!!!! Place figure just in front of the paragraph of text describing it.
+
 ## Broadly
 
-- [ ] Write an abstract
+- [ ] Write an abstract - Gelman method/sanity check. First sentence of every paragraph of the intro & the very last sentence of intro. 
 - [ ] Submit to journal
-- [ ] Spacing around figures WAY to large above/below captions
+- [ ] Spacing around figures WAY to large above/below captions - change fig.height for coord_fixed figures 
+- [ ] Make axes labels and text larger in figures
+- [ ] Make labels of points larger
 
@@
 - [ ] Add histogram of number of tie changes from $t_1$ to $t_2$ in simulations 
-- [ ] Add another example of a data set for M-in-DS (Sect 4.1) (GoT?)
+- [ ] Add another example of a data set for M-in-DS (Sect 4.1) (Congress)
 - [ ] Make gif available online
@@
 - [X] Figure 6: reverse order of adjacency matrix y-axis
+- [ ] Figure 5: get rid of Vs on node labels, make font bigger
 - [X] Figure 7 not showing up??

### 2017-08-12 paper.Rnw
senate data example

@@
 require(knitr)
-opts_chunk$set(fig.path='figure/', cache.path='cache/', fig.align='center', fig.pos='h', out.width='.8\\linewidth', par=TRUE, cache=TRUE, concordance=TRUE, autodep=TRUE, message=F, warning=F, echo = FALSE, root.dir = "~/Desktop/Dissertation - LETS GO/SAOM-removing-blindfold/")
+opts_chunk$set(fig.path='figure/', cache.path='cache/', fig.align='center', fig.pos='h', out.width='.8\\linewidth', par=TRUE, cache=TRUE, concordance=TRUE, autodep=TRUE, message=F, warning=F, echo = FALSE, root.dir = "~/Desktop/Dissertation/SAOM-removing-blindfold/")
 @
@@
 So, good fit is acheived when $p$ is large.
-XX more detail about mahalanobis distance and calculating p-value XX
+\st{XX more detail about mahalanobis distance and calculating p-value XX}
 
@@
 Box plots and violin plots are drawn on top of each other in Figure~\ref{fig:sienagof} in order to show the distribution of the simulated network observations compared to the true data shown in red points connected by red lines.
-If the red points lie well within the simulated values, the model is a good fit to the data.
+If the red points lie "well within" the simulated values, the model is a good fit to the data.
 The model examined in Figure~\ref{fig:sienagof} appears to do a better job of capturing the outdegree distribution than the indegree distribution of the data.
@@
 <<sienagof, fig.cap="Goodness of fit measures for two network statistics: indegree and outdegree cumulative distributions.
-On the x-axis is the degree value, and on the y-axis is the number of times that degree value occurs in the data.", cache = TRUE, message = FALSE, warning = FALSE>>=
+On the x-axis is the degree value, and on the y-axis is the number of times that degree value occurs in the data.
+\st{The p-value shown on the x-axis is ....
+}", cache = TRUE, message = FALSE, warning = FALSE>>=
 load("data/ansnullpaper.rda")
@@
 <<adjmatviz, fig.align='center', fig.cap="A visualization of the adjacency matrices of the three waves of network observations in the ``Teenage Friends and Lifestyle Study'' data.
-The subset we will be using is outlined in red.", fig.width=8, results='hide',fig.show='asis'>>=
+The subset we will be using is outlined in red.", fig.width=8, results='hide',fig.show='asis', fig.width=8, fig.height=3>>=
 source("code/00e_small_friends.R")
@@
 
-<<alldataviz, fig.cap="The smaller friendship network data we will be modelling throughout the paper.", fig.width=8>>=
+<<alldataviz, fig.cap="The smaller friendship network data we will be modelling throughout the paper.", fig.width=8, fig.height=3>>=
 actual1 <- merge(data.frame(as.edgelist(as.network(fd2.w1))), 

### 2017-08-12 makesienaDataSenate.R

@@ -1,0 +1,65 @@
+# senate data to the rsiena data 
+library(network)
+library(RSiena)
+library(tidyverse)
+library(lubridate)
+# variables - nyear, party, sex
+
+seobama <- read_csv("data/congress/senateobamapres_gsw_25.csv")
+legisls <- read_csv("data/congress/legislators_us.csv")
+obamayears <- 2009:2016
+
+legisls %>% mutate(yearStart = year(start), yearEnd = year(end)) -> legisls
+
+(legisls %>% 
+    filter(mandate == "se" & (yearStart %in% obamayears | yearEnd %in% obamayears)) %>% 
+    mutate(name1 = ifelse(is.na(middle), paste(first,last), paste(first,middle,last)),
+           name = ifelse(is.na(suffix), name1, paste(name1, suffix))) %>% 
+    select(name))[[1]] %>% unique %>% sort -> allsenators
+startingsemat <- matrix(0, nrow = length(allsenators), ncol = length(allsenators))
+rownames(startingsemat) <- allsenators
+colnames(startingsemat) <- allsenators
+
+se111adj <- startingsemat
+
+se111df <- seobama %>% filter(senate == 111 & !is.na(target))
+
+make_se_adjmat <- function(se_no, adjmat=startingsemat, obamaSenates=seobama){
+  senatedf <- obamaSenates %>% filter(senate == se_no & !is.na(target))
+  N <- nrow(senatedf)
+  for(i in 1:N){
+    k <- which(rownames(adjmat) == senatedf$source[i])
+    l <- which(colnames(adjmat) == senatedf$target[i])
+    adjmat[senatedf$source[i], senatedf$target[i]] <- 1
+  }
+  return(adjmat)
+}
+
+se111adj <- make_se_adjmat(se_no = 111)
+se112adj <- make_se_adjmat(se_no = 112)
+se113adj <- make_se_adjmat(se_no = 113)
+se114adj <- make_se_adjmat(se_no = 114)
+
+
+# okay, that's the networks. now make the covariates
+senateCovars <- seobama %>% 
+  select(source,senate, nyears, party, sex) %>%
+  unique %>% 
+  spread(senate, nyears)
+
+load("data/congress/senate111.rda")
+load("data/congress/senate112.rda")
+load("data/congress/senate113.rda")
+load("data/congress/senate114.rda")
+
[…]

### 2017-08-15 PaperToDo.Rmd
add results from senate model fitting; fiddle with the axes font & labels in plots

@@
 - [ ] Submit to journal
-- [ ] Spacing around figures WAY to large above/below captions - change fig.height for coord_fixed figures 
-- [ ] Make axes labels and text larger in figures
-- [ ] Make labels of points larger
+- [X] Spacing around figures WAY to large above/below captions - change fig.height for coord_fixed figures 
+- [X] Make axes labels and text larger in figures
+- [X] Make labels of points larger
 
@@
 - [X] Figure 6: reverse order of adjacency matrix y-axis
-- [ ] Figure 5: get rid of Vs on node labels, make font bigger
+- [X] Figure 5: get rid of Vs on node labels, make font bigger
 - [X] Figure 7 not showing up??
@@
 - [X] Heatmap p 23 missing caption
-- [ ] Describe Figure 8 more. What is going on??
+- [X] Describe Figure 8 more. What is going on??
 - [ ] Add geomnet example in 4.2

### 2017-08-15 paper.Rnw
add results from senate model fitting; fiddle with the axes font & labels in plots

@@
 require(knitr)
-opts_chunk$set(fig.path='figure/', cache.path='cache/', fig.align='center', fig.pos='h', out.width='.8\\linewidth', par=TRUE, cache=TRUE, concordance=TRUE, autodep=TRUE, message=F, warning=F, echo = FALSE, root.dir = "~/Desktop/Dissertation/SAOM-removing-blindfold/")
+opts_chunk$set(fig.path='figure/', cache.path='cache/', fig.align='center', fig.pos='h', out.width='.99\\linewidth', par=TRUE, cache=FALSE, concordance=TRUE, autodep=TRUE, message=F, warning=F, echo = FALSE)#, root.dir = "~/Desktop/Dissertation/SAOM-removing-blindfold/")
 @
@@
 library(cowplot)
+
+ThemeNoNet <- theme_bw() %+replace% 
+            theme(plot.title = element_text(size = rel(2),
+                                            face = 'bold', angle = 0), 
+                  axis.title.x = element_text(size = rel(2),
+                                            face = 'plain', angle = 0),
+                  axis.title.y = element_text(size = rel(2),
+                                            face = 'plain', angle = 90),
+                  axis.text.x.top = element_text(size = rel(1.5),
+                                            face = 'plain', angle = 0), 
+                  axis.text.x.bottom = element_text(size = rel(1.5),
+                                            face = 'plain', angle = 0), 
+                  axis.text.y.left = element_text(size = rel(1.5),
+                                            face = 'plain', angle = 0),
+                  axis.text.y.right = element_text(size = rel(1.5),
+                                            face = 'plain', angle = 0),
+                  strip.text.x = element_text(size = rel(2),
+                                            face = 'plain', angle = 0, 
+                                            margin = margin(t = 3, r = 0, b = 3, l = 0, unit = "pt")),
+                  strip.text.y =element_text(size = rel(2),
+                                            face = 'plain', angle = 90,margin = margin(t = 3, r = 3, b = 3, l = 3, unit = "pt")),
+                  strip.background = element_rect(colour = "black", fill = "white")
+                    )
+ThemeNet <- theme_net() %+replace% 
+            theme(plot.title = element_text(size = rel(2),
+                                            face = 'bold', angle = 0), 
+                  strip.text.x = element_text(size = rel(2),
+                                            face = 'plain', angle = 0, 
+                                            margin = margin(t = 3, r = 0, b = 3, l = 0, unit = "pt")),
+                  strip.text.y =element_text(size = rel(2),
+                                            face = 'plain', angle = 0,margin = margin(t = 3, r = 0, b = 3, l = 0, unit = "pt")),
+                  strip.background = element_rect(colour = "black", fill = "white"))
 @
@@
 ggplot(data = toynetdat) + 
-  geom_net(aes(from_id=from_id, to_id=to_id), labelon = T, size = 10, labelcolour = 'white', ecolour = 'black', vjust = .5, hjust = .5, arrowgap = .05, directed = T) +
-  theme_net() + 
+  geom_net(aes(from_id=from_id, to_id=to_id), labelon = T, size = 11, labelcolour = 'white', ecolour = 'black', vjust = .5, hjust = .5, fontsize = 4, arrowgap = .05, directed = T) +
+  ThemeNet + 
   labs(title = "Node-Link Diagram")
@@
   scale_y_continuous(breaks = (-5):(-1), labels = 5:1, limits = c(-5.5,-.5)) +
-  theme_bw() + 
-  theme(panel.grid = element_blank()) +
+  ThemeNoNet + 
+  theme(panel.grid = element_blank(), aspect.ratio = 1) +
   labs(title = "Adjacency Matrix Visualization") + 
-  theme(aspect.ratio = 1) +
   xlab("from (ego)") +
@@
 On the x-axis is the degree value, and on the y-axis is the number of times that degree value occurs in the data.
-\st{The p-value shown on the x-axis is ....
+\\st{The p-value shown on the x-axis is ....
 }", cache = TRUE, message = FALSE, warning = FALSE>>=
@@
   labs(x = "i", y = "j") + 
+  ThemeNoNet + 
   theme(aspect.ratio=1, axis.text = element_blank(), axis.ticks = element_blank()) 
@@
        aes(from_id = X1, to_id = X2, color = as.ordered(drink), group = as.ordered(drink))) +
-  geom_net(fiteach = FALSE, directed=TRUE, layout.alg = 'fruchtermanreingold', size = 4, arrowsize = .25, arrowgap = .03, ecolour = 'grey40', labelon=T, labelcolour = 'black', fontsize = 3, vjust = .5) +
+  geom_net(fiteach = FALSE, directed=TRUE, layout.alg = 'fruchtermanreingold', size = 4, arrowsize = .25, arrowgap = .03, ecolour = 'grey40', labelon=T, labelcolour = 'black', fontsize = 4, vjust = .5) +
   scale_color_brewer(palette = "YlOrRd", name = "Drinking\nBehavior") + 
-  theme_net() +
+  ThemeNet +
   facet_wrap(~Wave, labeller = "label_both") + 
@@
   facet_wrap(~.frame, nrow=2, labeller = 'label_both') + 
+  ThemeNet + 
   theme(panel.background = element_rect(color = 'black'))
@@
 <<adjmatorder, message = FALSE, warning = FALSE, fig.cap="On the left, the starting friendship network represented in adjacency matrix form, ordered by vertex id.
-On the right, the same adjacency matrix is presented after ordering the vertices by one repitition of the microstep simulation process from wave one to wave two.", fig.width=8>>=
+On the right, the same adjacency matrix is presented after ordering the vertices by one repitition of the microstep simulation process from wave one to wave two.", fig.width=8, fig.height=3>>=
 # first present the original wave 1 adj mat, then present the adj mat after being ordered by the 1st pc of the big adjmat A. 
@@
   scale_fill_manual(values = c('white', 'black')) +
-  theme(legend.position = 'none', axis.text = element_text(size = 8)) + 
-  scale_x_discrete(position = "top") + 
+  scale_x_discrete(position = "top", labels = as.character(1:16)) +
+  scale_y_discrete(labels = rev(as.character(1:16)))  + 
+  ThemeNoNet + 
+  theme(legend.position = 'none') + 
   coord_fixed()
@@
   scale_fill_manual(values = c('white', 'black')) +
-  theme(legend.position = 'none', axis.text = element_text(size = 8)) + 
-  scale_x_discrete(position = "top") + 
+  scale_x_discrete(position = "top", labels = str_replace(order1, "V", "")) + 
+  scale_y_discrete(labels = rev(str_replace(order1, "V", ""))) + 
+  ThemeNoNet + 
+  theme(legend.position = 'none') +
   coord_fixed()
@@
   geom_segment(
-    aes(x = alter , xend = alter, y = 0, yend = alterProb, color=type)) + 
-  geom_point(aes(x = alter, y = empprob), shape = 4, size = 1, alpha = .5) + 
+    aes(x = alter , xend = alter, y = 0, yend = alterProb, color=type), size = 2) + 
+  geom_point(aes(x = alter, y = empprob), shape = 4, size = 2, alpha = .5) + 
   labs(x = "alter node", y = "probability of alter node") + 
@@
   facet_wrap(~ego, labeller = "label_both") + 
-  theme_bw()
+  ThemeNoNet
 # 
@@
 The acting node is on the x-axis, and the receiving node is on the y-axis.
-Missing squares represent ties that did not occur.'>>=
+Missing squares represent ties that did not occur.', fig.height=3>>=
 ansnullchains %>% 
@@
   scale_fill_gradient(low = 'white', high = 'blue', limits = c(0,1)) +
-  theme(panel.background = element_rect(fill = 'white')) + 
   labs(x = "ego", y = "alter")+
-  scale_x_discrete(position = 'top') + 
+  scale_x_discrete(position = 'top', labels = str_replace(order1, "V", "")) +
+  scale_y_discrete(labels = rev(str_replace(order1, "V", ""))) +
+  ThemeNoNet + theme(panel.grid = element_blank()) + 
   coord_fixed()
@@
 
-<<allsteps, fig.cap="Three repetitions of the microstep process from wave 1 to wave 2.
-The $x$ axis is the microstep number, with step 0 representing the first wave of data and the final step representing the second wave of data.">>=
+<<allsteps, fig.cap="Three simulations (out of 1,000) of the microstep process from wave 1 to wave 2.
+The $x$ axis is the microstep number, with step 0 representing the first wave of data and the final step representing the second wave of data.
+We can see that many edges are underrepresented in this process: they are in the second wave, but never appear in the microsteps.">>=
 smallfriends <- read_csv("data/smallfriends4Geomnet.csv")
@@
   scale_fill_continuous(low = "red", high = "blue") + 
+  ThemeNoNet + 
   theme(legend.position = 'none', axis.ticks.y = element_blank(),
@@
   scale_fill_continuous(low = "red", high = "blue") +
+  ThemeNoNet + 
   theme(legend.position = 'none', axis.ticks.y = element_blank(),
@@
 In the middle, a summary network from the first model fit to the data.
-This summary network represents 1,000 simulations of wave 2 using the values from the simple fitted model $M1$.',results='hide'>>=
+This summary network represents 1,000 simulations of wave 2 using the values from the simple fitted model $M1$.',results='hide', fig.height=3>>=
 friendshipData <- array(c(fd2.w1, fd2.w2, fd2.w3), dim = c(16, 16, 3))
@@
   geom_net(aes(from_id = from, to_id = to, linewidth= linewidth),
-           directed = T, curvature = .2, labelon=T, size = 4,
-           arrow = arrow(angle = 20, type = "open", length = unit(.1, "inches")), ecolour = t1$color,
-           color = "#252525", hjust = .5, vjust = .5, labelcolour = 'white', fontsize = 2, arrowgap = .02,
+           directed = T, curvature = .2, labelon=T, size = 5,
+           arrow = arrow(angle = 20, type = "open", length = unit(.2, "cm")), ecolour = t1$color,
+           color = "#252525", hjust = .5, vjust = .5, labelcolour = 'white', fontsize = 4, arrowgap = .02,
            singletons = T) +
-  theme_net() + labs(title = "Summary Network: M1") + 
-  theme(plot.background = element_rect(color = 'black'),
-        plot.title = element_text(size = rel(.5)))
+  labs(title = "Summary: M1") + ThemeNet +
+  theme(plot.background = element_rect(color = 'black'))
 
 p2 <- ggplot(data = filter(combinedavgs, cat == "First Observation")) +
-  geom_net(aes(from_id = from, to_id = to, linewidth= linewidth),
-           directed = T, curvature = .2, labelon=T, size = 4,
-           arrow = arrow(angle = 20, type = "open", length = unit(.1, "inches")), ecolour = "#969696",
-           color = "#252525", hjust = .5, vjust = .5, labelcolour = 'white', fontsize = 2, arrowgap = .02,
+  geom_net(aes(from_id = from, to_id = to, linewidth= linewidth/3),
+           directed = T, curvature = .2, labelon=T, size = 5,
+           arrow = arrow(angle = 20, type = "open", length = unit(.2, "cm")), ecolour = "#969696",
+           color = "#252525", hjust = .5, vjust = .5, labelcolour = 'white', fontsize = 4, arrowgap = .03,
            singletons = T) +
-  theme_net() + labs(title = "Wave 1") + 
-    theme(plot.background = element_rect(color = 'black'),
-        plot.title = element_text(size = rel(.5)))
+   labs(title = "Wave 1") + ThemeNet + 
+    theme(plot.background = element_rect(color = 'black'))
 p3 <- ggplot(data = filter(combinedavgs, cat == "Second Observation")) +
-  geom_net(aes(from_id = from, to_id = to, linewidth= linewidth),
-           directed = T, curvature = .2, labelon=T, size = 4,
+  geom_net(aes(from_id = from, to_id = to, linewidth= linewidth/3),
+           directed = T, curvature = .2, labelon=T, size = 5,
            arrow = arrow(angle = 20, type = "open", length = unit(.2, "cm")), ecolour = "#969696",
-           color = "#252525", hjust = .5, vjust = .5, labelcolour = 'white', fontsize = 2, arrowgap = .02,
+           color = "#252525", hjust = .5, vjust = .5, labelcolour = 'white', fontsize = 4, arrowgap = .03,
            singletons = T) +
-  theme_net() + labs(title = "Wave 2")  + 
-  theme(plot.background = element_rect(color = 'black'),
-        plot.title = element_text(size = rel(.5))) 
+  labs(title = "Wave 2")  +  ThemeNet + 
+  theme(plot.background = element_rect(color = 'black')) 
 grid.arrange(p2, p1, p3, nrow = 1)
@@
 The group of the actors is represented by the shape of the node.}
-<<jtt, echo=FALSE, fig.width=2, fig.height=2, message = FALSE, warning = FALSE, fig.align='center'>>=
+<<jtt, echo=FALSE, fig.width=2, fig.height=2, message = FALSE, warning = FALSE, fig.align='center',out.width='.49\\linewidth'>>=
 jTTe <- data.frame(from = c('i', 'i', 'h'), to = c('h', 'j', 'j'))
@@
            labelcolour='grey80',vjust = 0.5, hjust =0.5, arrowgap = .15, 
-           colour = 'black', size=10, 
+           colour = 'black', size=10, fontsize = 4,
            ecolour = c("red", "grey40", "grey40", "grey40")) + 
   expand_limits(x=c(-0.1,1.1), y=c(-0.1,1.1)) +
-  theme_net() +
+  ThemeNet +
   theme(legend.position = "none")
@@
 \caption{\label{fig:dad}Doubly achieved distance between actors $i$ and $k$.}
-<<dab, echo=FALSE, fig.width=2, fig.height=2, message = FALSE, warning = FALSE, fig.align='center'>>=
+<<dab, echo=FALSE, fig.width=2, fig.height=2, message = FALSE, warning = FALSE, fig.align='center', out.width='.49\\linewidth'>>=
 dade <- data.frame(from = c('i', 'i', 'h', 'j'), to = c('h', 'j', 'k', 'k'))
@@
 ggplot(data = dad, aes(from_id = from, to_id = to)) + 
-  geom_net(aes(shape = as.factor(group)), directed = T, labelon = T, labelcolour='grey80',vjust = 0.5, hjust =0.5, arrowgap = .15, colour = 'black', size=10) + 
+  geom_net(aes(shape = as.factor(group)), directed = T, labelon = T, labelcolour='grey80',vjust = 0.5, hjust =0.5, arrowgap = .15, colour = 'black', size=10, fontsize = 4) + 
   expand_limits(x=c(-0.1,1.1), y=c(-0.1,1.1)) +
-  theme_net() +
+  ThemeNet +
   theme(legend.position = "none")
@@
   facet_wrap(~parameter, scales = 'free') + 
-  theme_bw()
+  ThemeNoNet
 @
@@
 simu_spread <- simu2 %>% spread(parameter, estimate)
-ggpairs(simu_spread, columns = 3:8, ggplot2::aes(colour=Model, alpha = .5)) + theme_bw()
+ggpairs(simu_spread, columns = 3:8, ggplot2::aes(colour=Model, alpha = .5)) + ThemeNoNet
 @
@@
 <<averagenet, results='hide', out.width='99%',fig.cap='The three "average" networks in the middle, with the actual network observations oon either side.
-There is some difference between the three models, but overall, these three models cannot capture the structure in the true second wave of data.
-'>>=
+There is some difference between the three models, but overall, these three models cannot capture the structure in the true second wave of data.', fig.height=3>>=
 #friend.data.w1 <- as.matrix(read.table("s50_data/s50-network1.dat"))
@@
   geom_net(aes(from_id = from, to_id = to, linewidth= linewidth),
-           directed = T, curvature = .2, labelon=T, size = 4,
+           directed = T, curvature = .2, labelon=T, size = 5,
            arrow = arrow(angle = 20, type = "open", length = unit(.2, "cm")), ecolour = "#969696", fiteach = T, 
-           color = "#252525", hjust = .5, vjust = .5, labelcolour = 'white', fontsize = 2, arrowgap = .02,
+           color = "#252525", hjust = .5, vjust = .5, labelcolour = 'white', fontsize = 4, arrowgap = .02,
            singletons = T) + 
-  theme_net() + theme(panel.background = element_rect(color = 'black')) + 
+  ThemeNet + theme(panel.background = element_rect(color = 'black')) + 
   facet_wrap(~cat, nrow = 1)

### 2017-08-15 makesienaDataSenate.R

@@ -40,12 +40,19 @@ se112adj <- make_se_adjmat(se_no = 112)
 se113adj <- make_se_adjmat(se_no = 113)
 se114adj <- make_se_adjmat(se_no = 114)
 
+remove_sens <- which(!(colnames(se111adj) %in% senateCovars$source))
+
+se111adj <- se111adj[-remove_sens, -remove_sens]
+se112adj <- se112adj[-remove_sens, -remove_sens]
+se113adj <- se113adj[-remove_sens, -remove_sens]
+se114adj <- se114adj[-remove_sens, -remove_sens]
 
 # okay, that's the networks. now make the covariates
 senateCovars <- seobama %>% 
-  select(source,senate, nyears, party, sex) %>%
+  select(source,senate, n_au, party, sex) %>%
   unique %>% 
-  spread(senate, nyears)
+  spread(senate, n_au)
+
 
 load("data/congress/senate111.rda")
 load("data/congress/senate112.rda")
@@ -61,5 +68,78 @@ delete.edges(net_us_se113_hi,which(net_us_se113_hi %e% "gsw" < .25))
 net_us_se114_hi <- net_us_se114
 delete.edges(net_us_se114_hi,which(net_us_se114_hi %e% "gsw" < .25))
 
+##### MAKE SIENA DATA #######
 
+senateData <- array( c( se111adj, se112adj, se113adj, se114adj ),
+                      dim = c( 156, 156, 4 ) )
+senateData2 <- sienaDependent(senateData)
+party <- coCovar(val = as.numeric(as.factor(senateCovars$party)))
+# party 1 is dem 2 is ind 3 is rep
+sex <- coCovar(val = as.numeric(as.factor(senateCovars$sex)))
+# sex 1 is F 2 is M
+nbills_au <- senateCovars[,4:7]
+names(nbills_au) <- paste0("V",1:4) 
+nbills_au <- as.matrix(nbills_au)
+nbills_au[which(is.na(nbills_au))] <- 0
+rownames(nbills_au) <- senateCovars$source
+bills <- varCovar(nbills_au)
+# rsiena objects are party, sex, bills, senateData2
+senateSiena <- sienaDataCreate( senateData2, bills, sex, party )
+senatemodeff <- getEffects(senateSiena)
+library(netvizinf)
+senateSigEffs <- find_sig_effs(dat = senateSiena)
[…]

### 2017-08-16 PaperToDo.Rmd
add todo

@@
 - [X] Make labels of points larger
+- [ ] Reverse order of x-y axes on adjmat vizs
 

### 2017-08-17 paper.Rnw
pages 1-5

@@
 \st{XXX - okay, here is what i have done for that:
-Social networks have been studied for decades, beginning with a few foundational works, the most well known of which is the 1967 study, "The Small World Problem" by Stanley Milgram \citep{goldenberg09}.
+Social networks have been studied for decades, beginning with a few foundational works, including the 1967 study, "The Small World Problem" by Stanley Milgram \citep{goldenberg09}.
 Examples of social networks include collaboration networks between academic researchers, friendship networks in a school or university, and trade networks between nations.
@@
 These models, however, are only for single instance networks.
-If we only care about one network observation, that is not a problem, but if we have many observations of social network over time, these models may not be appropriate.
-When studying a network over time, which is referred to as a \textit{dynamic network}, we need a model that can take time into account.
-Models for dynamic social networks are extremely important because of how realistic they are.
-Social networks do not form spontaneously; they evolve over time.
-Ties are added and removed, and new actors join the network.
-Modeling the underlying mechanisms that create network changes over time is very complex but also has the potential to uncover hidden truths.
+If we only have one network observation, or only care about one state of a complex network in time, like a snapshot of the World Wide Web, using the well-established models for single network observating is not a problem.
+If, on the other hand, we have many observations of social network over time, these models may not be appropriate because they do not explicitly allow for the network to change as time passes.
+When studying a network over time, referred to as a \textit{dynamic network}, we need a model that can take the time aspect of the network into account.
+Models for dynamic social networks have a great deal of modelling potential because of how realistic their structure can be.
+A social network does not form spontaneously: it evolves over time.
+Ties are formed and dissolved, and new actors join the social structure.
+Modeling the underlying mechanisms that create network changes over time is very complex but also provides potential to uncover hidden truths.
 
 \par In this paper, we concentrate on one type of model for dynamic social networks: the stochastic actor-oriented models (SAOMs), introduced by \citet{saompaper}.
-These models are fundamentally different from other social network models because they allow us to incorporate network \textit{and} actor statistics, when other models only rely on  the network statistics to determine change.
-Letting the actor-level statistics change the structure of the network leads to a more practical and relevant approach to model change in a social network: in the ``real" world we expect people with common interests to be more likely to form relationships.
-SAOMs allow us to incorporate this intuition in the modeling process.
+These models are fundamentally different from other social network models because they allow us to incorporate network \textit{and} actor statistics, where other models only rely on the network statistics, to model the changes in the network.
+Allowing the actor-level statistics to directly effect the structure of the network leads to a more practical and relevant approach to model change in a social network.
+In the ``real" world we expect people with common interests to be more likely to form relationships, and SAOMs allow us to incorporate this intuition in the modeling process.
 
@@
 \st{They are relatively new, especially compared to the classic exponential random graph models, and they are not very tractable analytically.
-Likelihood functions turn into very complex objects due to the dependency structure inherent in the data.
-Therefore, computationally more tractable solutions are used to fit estimators, in particular, % Traditional modeling steps, like determining the likelihood function, are very complex.
+Likelihood functions quickly very complex objects to analyze due to the dependency structure inherent in the data.
+Therefore, computationally more tractable solutions are used to fit estimators, and in particular, % Traditional modeling steps, like determining the likelihood function, are very complex.
 Because of this,
-SAOMs are fit to data using a series of MCMC phases for finding method of moments estimators.} Rather than treat SAOM fits and the fitting process as black boxes, we combine the principals of network visualization with those of model visualization as discussed in \citet{modelvis}.
+SAOMs are often fit to data using a series of Markov chain Monte Carlo (MCMC) phases for finding method of moments estimators.
+In order to estimate the parameters of SAOMs, we use the software SIENA, and its \texttt{R} implementation \texttt{RSiena}, which was developed by \citet{RSiena}.
+This software marks a huge contribution to the field of social network analysis, but the many moving pieces involved in parameter estimation are largely ``behind the scenes" and hidden from the software user.} In this paper, in order to better understand the model-fitting process, we attempt to bring SAOM fits and the fitting process out of their black boxes, by combining the principals of network visualization with those of model visualization as discussed in \citet{modelvis}.
 %Here, we explore RSiena and SAOMs with innovative network visualizations, both static and dynamic.
-By bringing some light to the underlying methods and structures ``behind the scenes" of SAOMs, we help researchers working with the models better understand the implications and analysis of these models.
+By bringing some light to the underlying methods and structures that are ``behind the scenes" of when fitting SAOMs to network data, we aim to help researchers working with the models better understand the implications and analyses of these models.
 }
 
-\par The guiding principles of model visualization introduced in \citet{modelvis} are (1) display the model in the data space, (2) collections of models are more informative than singletons, and (3) explore the fitting process rather than just looking at the final result.
-Stochastic actor-oriented models are a prime example of a set of models that  benefit greatly from application of model visualization.
-First, the models themselves include a continuous-time Markov chain (CTMC) that is completely hidden in the model fitting process.
-Bringing the CTMC out of the black box and into the light of visualization  provides researchers with insights into the underlying features of the model.
-Secondly, SAOMs allow for a great deal of parameters that can be added to the objective function, each of which is attached to a network statistic.
-These statistics are often somewhat, if not highly, correlated, and so are the associated parameters.
-By visualizing collections of SAOMs, we gain a better understanding of these correlations and find ways to deal with them and rectify their effects.
-Finally, estimation of the parameters in a SAOM relies on a Robbins-Monro algorithm, and convergence checks for these estimates rely on simulations from the fitted model.
-Again, these steps are largely kept in the background of the estimation process.
-With the help of static and dynamic visualizations we bring these elements and others into the foreground, leading to a better understanding and higher accessibility of stochastic actor-oriented models  for social network analysts.
-
-In Section~\ref{sec:netintro}, we introduce the concepts of networks and network visualizations.
-In Section~\ref{sec:saoms}, we present the SAOM for social network analysis.
+\par The guiding principles of model visualization introduced in \citet{modelvis} are: 
+\begin{enumerate}
+\item Display the model in the data space.
+\item Collections of models are more informative than singletons. 
+\item Explore the fitting \textit{process} rather than just looking at the final result.
+\end{enumerate}
+Stochastic actor-oriented models are a prime example of a set of models that can benefit greatly from application of model visualization.
+For instance, the models themselves include a continuous-time Markov chain (CTMC) that is completely hidden from the analyst in the model fitting process.
+Bringing the CTMC out of the black box and into the light through model visualization can provide researchers with insights into the underlying features of the model.
+Furthermore, SAOMs can include a great deal of parameters to be added to the model structure, each of which is attached to a network statistic.
+These statistics are often somewhat, if not highly, correlated, which causes high correlation between the associated parameters in a SAOM.
+By visualizing collections of SAOMs, we gain a better understanding of these correlations and find ways to deal with them and rectify their effects in the model.
+In addition, the estimation of the parameters in a SAOM relies on a Robbins-Monro algorithm, and the convergence checks for these estimates rely on simulation from the fitted model.
+Again, each of these steps are largely kept in the background of the estimation process.
+With the help of static and dynamic visualizations we bring the hidden model fitting processes into the foreground, eventually leading to a better understanding and higher accessibility of stochastic actor-oriented models for social network analysts.
+
+In Section~\ref{sec:netintro}, we introduce basic concepts of networks and network visualizations.
+In Section~\ref{sec:saoms}, we present the family of stochastic actor-oriented models for social network analysis.
 In Section~\ref{sec:ModelVis}, we combine concepts from Sections~\ref{sec:netintro} and \ref{sec:saoms} in an application of the model-vis paradigm, and conclude with a discussion in Section~\ref{sec:discussion}.
@@
 Network data is of frequent interest to researchers in a wide array of fields.
-There are technological networks, like power grids or the internet, information networks, such as citation networks or the World Wide Web, biological networks, e.g.
-neural networks, and social networks, just to name a few \citep{newman}.
-Each of these examples have one thing in common: their data structure.
-There are always units of observation, which we refer to throughout this paper as \textit{nodes}, and connections of some kind between those units, which we will call \textit{edges}.
-Networks might change over time, for example when new websites and hyperlinks are added on the World Wide Web, or when there are new people and relationships in a friendship network.
-Nodes and edges themselves also have inherent variables of interest, e.g.\ the affiliation of authors in a co-authorship network, or the number of times two authors have collaborated.
-
-The multiple layers of network data structures create unique struggles to network analysts.
-Some questions that researchers aim to answer are: How does the strength of a connection between two nodes affect the overall structure of the network?
-Do node-level differences affect the formation or destruction of edges?
-Which views of the data are most informative for communicating effects and results of statistical analyses?
-These are, of course, just a few broad questions.
-In this paper we want to focus on the visual exploration of network data and models.
+There are technological networks, like power grids or the internet, information networks, such as citation networks or the World Wide Web, biological networks, like neural networks, and social networks, just to name a few \citep{newman}.
+Each of these examples have one thing in common: their data \textit{structure}.
+There are always units of observation: the power stations, websites, neurons, and people, which we refer to throughout this paper as \textit{nodes} or \textit{actors}.
+There are also always connections of some kind between those units: the power lines, hyperlinks, electrical signals, and relationships, which we will call \textit{edges} or \textit{ties}.
+Networks might change over time, like when new websites and hyperlinks are added on the World Wide Web, or when there are new people and relationships in a friendship network.
+The nodes and edges themselves can also have inherent variables of interest, e.g.\ the institution of authors in a co-authorship network, or the number of times two authors have collaborated.
+
+The multiple layers of network data structures pose unique problems to network analysts.
+Some questions that network researchers may aim to answer are: How does the strength of a tie between two nodes affect the overall structure of the network?
+Do node-level differences affect the formation or dissolution of edges?
+Which views of the data are most informative for communicating significant effects and other results of statistical analyses of the network of interest?
+These are, of course, just a few broad questions, and we focus here on the latter, which we aim to answer through visual exploration of network data and models.
 
@@
 
-There are two primary methods used to visualize networks: node-link diagrams and matrix visualization \citep{knuth2000, adjmatpro}.
+There are two primary methods used to visualize networks: node-link diagrams and adjacency matrix visualization \citep{knuth2000, adjmatpro}.
 As a toy example, let us assume that we have five nodes, $\{1,2,3,4,5\}$, connected by five directed edges:  $\{2 \to 4, 3 \to 4, 1 \to 5, 3 \to 5, 5 \to 4\}$ We use this toy data set to demonstrate the two visualization methods in Figure~\ref{fig:toyplots}.
@@
 
-The first method, the node-link diagram, represents nodes with markers in 2D Euclidean space and then represents edges by connecting the markers with lines when there is an edge between the two nodes.
-But because there is no natural placement of the points, a random placement is used, then adjusted via a layout algorithm, of which there are many \citep{drawingalgs}.
-
-Some layout algorithms were designed to mimic physical systems, drawing the graphs based on the ``forces" connecting them.
-The network's edges act as springs pushing and pulling the nodes in a lower dimensional (usually two-dimensional) space.
+The first method, the node-link diagram, represents nodes with points in 2D Euclidean space and then represents edges by connecting the points with lines when there is an edge between the two nodes.
+These lines can also have arrows on them indicating the direction of the edge for directed networks.
+But because there is typically no natural placement of the points unless they have important spatial locations, a random placement of the points is used, then adjusted via a layout algorithm, of which there are many \citep{drawingalgs}.
+
+Some commonly used layout algorithms, such as  the Kamada-Kawai layout \citep{kamadakawai} and the Fruchterman-Reingold layout \citep{fruchterman-reingold}, were designed to mimic physical systems, drawing the graphs based on the ``forces" connecting them.
+In these algorithms, the edges of the network act as springs pushing and pulling the nodes in a low dimensional (usually two-dimensional) space.
+Another algorithm uses multi-dimensional scaling, relying on distance metric and computing a matrix whose entries represent the ``distance" between every pair of nodes.
+Yet another layout algorithm relies on the eigenstructure of the network's adjacency matrix to place the nodes in 2D space \citep{drawingalgs}.
+The node-link diagram for our toy network is shown in Figure~\ref{fig:toyplots}.
 %Other algorithms use properties of the network's adjaceny matrix in order to place the nodes.
@@
 XXX }
-Some commonly used layout algorithms, such as  the Kamada-Kawai layout \citep{kamadakawai} and the Fruchterman-Reingold layout \citep{fruchterman-reingold}, treat networks like physical systems.
-Other algorithms are based on multi-dimensional scaling, which relies on a defined distance matrix, or an eigen value, which depends on the eigenvalues of the network's adjacency matrix \citep{drawingalgs}.
-The node-link diagram for our toy network is shown in Figure~\ref{fig:toyplots}.
 
@@
 library(lattice)
-gof1 <- sienaGOF(ansnull, OutdegreeDistribution, varName ="friendship")
-gof2 <- sienaGOF(ansnull, IndegreeDistribution, varName ="friendship")
-p1 <- plot(gof1)
-p2 <- plot(gof2)
+gof1 <- sienaGOF(ansnull, OutdegreeDistribution, varName ="friendship", join = FALSE)
+gof2 <- sienaGOF(ansnull, IndegreeDistribution, varName ="friendship", join = FALSE)
+p1 <- plot(gof1, period = 1)
+p2 <- plot(gof2, period = 1)
 gridExtra::grid.arrange(p1, p2)
+# sum(gof1$`Period 1`$SimulatedTestStat >= gof1$`Period 1`$ObservedTestStat)/1000
+# [1] 0.154
+# > sum(gof2$`Period 1`$SimulatedTestStat >= gof2$`Period 1`$ObservedTestStat)/1000
+# [1] 0.063
+
 @

### 2017-08-17 paper.Rnw
changes through sections 3.1.1

@@
 
+<<toyplots, fig.show='hold', fig.cap="On the left, a node-link diagram of our directed toy network, with nodes placed using the Kamada-Kawai algorithm.
+On the right, the adjacency matrix visualization for that same network.", out.width='.49\\linewidth'>>=
+toynetdat <- fortify(as.edgedf(toynet$edges), toynet$nodes)
+ggplot(data = toynetdat) + 
+  geom_net(aes(from_id=from_id, to_id=to_id), labelon = T, size = 11, labelcolour = 'white', ecolour = 'black', vjust = .5, hjust = .5, fontsize = 9, arrowgap = .05, directed = T) +
+  ThemeNet + 
+  labs(title = "Node-Link Diagram")
+
+ggplot(data = toynet$edges) +
+  #geom_tile(aes(x = from, y = -to), color = 'grey20', fill = 'black') +
+  geom_tile(aes(x = to, y = -from), color = 'grey20', fill = 'black') +
+  geom_segment(aes(x = x, xend = xend), y = -0.5, yend=-5.5, colour="grey30", data = data.frame(x = 0:5+.5, xend= 0:5+.5)) + 
+  geom_segment(aes(y=-y, yend=-yend), x = 0.5, xend=5.5, colour="grey30", data = data.frame(y = -0:5+.5, yend= -0:5+.5)) + 
+  scale_x_continuous(breaks = 1:5, labels = 1:5, limits = c(0.5,5.5), position = "top") +
+  scale_y_continuous(breaks = (-5):(-1), labels = 5:1, limits = c(-5.5,-.5)) +
+  ThemeNoNet + 
+  theme(panel.grid = element_blank(), aspect.ratio = 1) +
+  labs(title = "Adjacency Matrix Visualization", x = "to (alter)", y = "from (ego)")
+@
+%\hh{XXX Explain each of these visualizations and either comment on the differences in the visualizations (colors) or don't show them. }
+
 Some commonly used layout algorithms, such as  the Kamada-Kawai layout \citep{kamadakawai} and the Fruchterman-Reingold layout \citep{fruchterman-reingold}, were designed to mimic physical systems, drawing the graphs based on the ``forces" connecting them.
@@
 Another algorithm uses multi-dimensional scaling, relying on distance metric and computing a matrix whose entries represent the ``distance" between every pair of nodes.
-Yet another layout algorithm relies on the eigenstructure of the network's adjacency matrix to place the nodes in 2D space \citep{drawingalgs}.
-The node-link diagram for our toy network is shown in Figure~\ref{fig:toyplots}.
+There are also layout algorithms that use properties of the adjacency matrix, like its eigenstructure, to place the nodes in 2D space \citep{drawingalgs}.
+The node-link diagram using the Kamada-Kawai layout algorithm for our toy network is shown in Figure~\ref{fig:toyplots}.
+Unless otherwise stated, all other node-link diagrams in this paper will use the Kamada-Kawai layout.
 %Other algorithms use properties of the network's adjaceny matrix in order to place the nodes.
-\hh{XXX how does the adjacency matrix feature here?
+%hh{XXX how does the adjacency matrix feature here?
 The adjacency matrix is a different mathematical representation of a network, but describes the same network as a node-link diagram.} \st{XXX I'm taking this to mean I should describe adjacency matrices more, so I've added a discussion below.
@@
   \begin{cases}
-                                   1 &  \text{if an edge exists} i\to j \\
+                                   1 &  \text{if an edge exists } i\to j \\
                                    0 & otherwise
@@
 \]
+Note here that our edge variables are binary: we only consider the presence or absence of an edge.
+If the network has weighted edges, for example an email network where edge weights represent the number of emails sent from one person to another, the entries in the adjacency matrix are the edge weights instead of zeroes and ones.
 In an undirected network, $A_{ij} = A_{ji}$, but in a directed graph this is only true if there is an edge from $i$ to $j$ and from $j$ to $i$.
@@
 \hh{XXX what is an adjacency matrix? - explain at this point in mathematical terms} \st{XXX see above XXX}
-This type of visualization is also shown in Figure~\ref{fig:toyplots} for our toy example.  
-<<toyplots, fig.show='hold', fig.cap="On the left, a node-link diagram of our directed toy network, with nodes placed using the Kamada-Kawai algorithm.
-On the right, the adjacency matrix visualization for that same network.", out.width='.49\\linewidth'>>=
-toynetdat <- fortify(as.edgedf(toynet$edges), toynet$nodes)
-ggplot(data = toynetdat) + 
-  geom_net(aes(from_id=from_id, to_id=to_id), labelon = T, size = 11, labelcolour = 'white', ecolour = 'black', vjust = .5, hjust = .5, fontsize = 4, arrowgap = .05, directed = T) +
-  ThemeNet + 
-  labs(title = "Node-Link Diagram")
-
-ggplot(data = toynet$edges) +
-  geom_tile(aes(x = from, y = -to), color = 'grey20', fill = 'black') +
-  #geom_tile(aes(x = to, y = from), color = 'grey20', fill = 'black') +
-  geom_segment(aes(x = x, xend = xend), y = -0.5, yend=-5.5, colour="grey30", data = data.frame(x = 0:5+.5, xend= 0:5+.5)) + 
-  geom_segment(aes(y=-y, yend=-yend), x = 0.5, xend=5.5, colour="grey30", data = data.frame(y = -0:5+.5, yend= -0:5+.5)) + 
-  scale_x_continuous(breaks = 1:5, labels = 1:5, limits = c(0.5,5.5), position = "top") +
-  scale_y_continuous(breaks = (-5):(-1), labels = 5:1, limits = c(-5.5,-.5)) +
-  ThemeNoNet + 
-  theme(panel.grid = element_blank(), aspect.ratio = 1) +
-  labs(title = "Adjacency Matrix Visualization") + 
-  xlab("from (ego)") +
-  ylab("to (alter)")
-@
-\hh{XXX Explain each of these visualizations and either comment on the differences in the visualizations (colors) or don't show them. }
+An adjacency matrix visualization for our toy example is also shown in Figure~\ref{fig:toyplots}.
 
 Each type of visualization comes with its own advantages and disadvantages.
-Paths between nodes are easier to determine with node-link diagrams than with adjacency matrix visualizations.
-In node-link diagrams, node information can be incorporated into the visualization by coloring or changing the shape of the points representing the nodes.
-This is not necessarily possible in matrix representation, which is more focused on edges.
-Adjacency matrix visualization is particularly useful, however, when the network is very complex, dense, or large.
-Experimental studies have shown this to be true.
-For example, with seven basic perceptual tasks on networks, including things like node and edge count, adjacency matrix visualizations outperform node-link diagrams as the size and density of the network increases \citep{adjmatviz}.
-One drawback of the adjacency matrix visualization that \citeauthor{adjmatviz} found was that edges are overrepresented for undirected graphs, due to the symmetry of $\mathbf{A}$: the edge $x_{ij}$ for $i \neq j$ appears in $A$ twice: in $A_{ij}$ and $A_{ji}$.
+For example, paths between two nodes in a network are easier to determine with node-link diagrams than with adjacency matrix visualizations \citep{adjmatviz}.
+In node-link diagrams, node-level information can be incorporated into the visualization by coloring or changing the shape of the points representing the nodes, and edge-level information can be incorporated by coloring the lines, or changing their thickness, linetype, or color.
+Incorporating a node-level variable into an adjacency matrix visualization is not as straightforward or simple, which is more focused on edges.
+Adjacency matrix visualization has been found to be particularly useful when the network is very complex, dense, or large, and experimental studies have shown adjacency matrix visualization to be superior to node-link diagrames for large networks.
+For example, for basic perceptual tasks on networks, including node and edge count, adjacency matrix visualizations outperform node-link diagrams as the size and density of the network increases \citep{adjmatviz}.
+One drawback of the adjacency matrix visualization that \citeauthor{adjmatviz} found was that edges are overrepresented for undirected graphs, due to the symmetry of $\mathbf{A}$: the edge $x_{ij}$ for $i \neq j$ appears in $\mathbf{A}$ twice: in $A_{ij}$ and $A_{ji}$, and so it also appears twice in the adjacency matrix visualization.
 \hh{XXX What is the exact logic here that Ghoniem used to come to the conclusion that edges in undirected graphs are over-represented?
@@
 A node-link diagram, however, may underrepresent the edge count if the edges $x_{ij}$ and $x_{ji}$ both exist and are drawn on top of one another.
-Ultimately, t}here is not one "correct" way to visualize network information, and we will be using both \st{the node-link and adjacency matrix visualization} methods throughout this paper.
+Ultimately, however t}here is not one "correct" way to visualize network information, and we will be using both \st{the node-link and adjacency matrix visualization} methods throughout this paper to explore social networks and stochastic actor-oriented models.
 
@@
 
-A Stochastic Actor-Oriented Model (SAOM) is a model that is changing in time in order to accomodate for observations from the same network made at different points in time and that allows for changes in network structure due to actor-level covariates.
+A Stochastic Actor-Oriented Model (SAOM) is a model that incorporates all three components of dynamic networks: edge, node, and time information.
+It models the change of a network over time, allowing for changes in network structure due to actor-level covariates.
 This model was first introduced by Snijders in 1996 \citep{saompaper}.
-The two main properties of SAOMs, modelling the changes in time and using actor covariates, are crucial to understanding networks as they exist naturally.
-Most social networks, even holding constant the set of actors over time, are ever-changing as relationships decay or grow, and most actors (or nodes) in social networks have inherent properties that could affect how they change their place within the network.
+The two titular properties of SAOMs, stochasticity and actor-orientation, are crucial to understanding networks as they exist naturally.
+Most social networks, even holding constant the set of actors over time, are ever-changing as relationships decay or grow in seemingly random ways, and most actors (or nodes) in social networks have inherent properties that could affect how they change their role within the network, and vice versa.
 
@@
 
-In this paper, the term \textit{dynamic network} refers to a network consisting of the same set of $n$ nodes that is changing over time, and is observed at $M$ discrete time points, $t_1, \dots, t_M$ with $t_1 \le t_2 \le \dots \le t_M$.
+In this paper, the term \textit{dynamic network} refers to a network, consisting of a fixed set of $n$ nodes, that is changing over time, and is observed at $M$ discrete time points, $t_1, \dots, t_M$ with $t_1 \le t_2 \le \dots \le t_M$.
 We denote the network observation at timepoint $t_k$ by $x(t_k)$.
-In the modelling process we usually condition on the first observation, $x(t_1)$.
-The SAOM assumes that this longitudinal network is embedded within a continuous time Markov chain (CTMC), which we will refer to as $X(T)$.
-This process is almost entirely unobserved: assume that the beginning of the process, $X(0)$ is equivalent to the first network observation $x(t_1)$, while the end of the process $X(\infty)$ is equivalent to the last observation $x(t_M)$.
-All other parts of the process are unseen.
-We model the process $X(T)$ as a series of single tie (or edge) changes, in which one actor at a time is given the opportunity to add or remove one outgoing tie.
-Once an actor is given the chance to change a tie, it tries to maximize a utility function based on the current and potential future states of the network.
+In the modelling process, we condition on the first observation, $x(t_1)$.
+The SAOM assumes that this longitudinal network of discrete observations is embedded within a continuous time Markov chain (CTMC), which we will denote $X(T)$.
+This process is almost entirely unobserved: we assume that the beginning of the process, $X(0)$, is equivalent to the first network observation $x(t_1)$, while the end of the process,$X(\infty)$, is equivalent to the last observation $x(t_M)$.
+Nearly all other parts of the process are unseen, with the exception of $x(t_2), \dots, x(t_{M-1})$.
+Unlike the first and last observations of the network, these ``in-between" observations do not have direct correspondence with steps in the continuous time Markov chain.
+Thus, the ``in-between" observations are considered to be ``snapshots" of the network at some point between two steps in the CTMC.
+The whole process $X(T)$ is a series of single tie changes that happen according to some pre-defined rate function, where one actor at a time is given the opportunity to add or remove one outgoing tie, or to not make any changes.
+Once an actor is chosen at random according to the rate function, it is ``given" the chance to change a tie, and it tries to maximize a utility function based on the current and potential future states of the network.
 

### 2017-08-18 paper.Rnw
taking a break. will get back to it in the morning.

@@
 The whole process $X(T)$ is a series of single tie changes that happen according to some pre-defined rate function, where one actor at a time is given the opportunity to add or remove one outgoing tie, or to not make any changes.
-Once an actor is chosen at random according to the rate function, it is ``given" the chance to change a tie, and it tries to maximize a utility function based on the current and potential future states of the network.
+Once an actor is chosen at random according to the rate function, it is ``given" the chance to change a tie, and it tries to maximize its utility function based on the current and near future states of the network.
+We expand on the model description further in the subsequent sections.
 
@@
 
-For the network $x$ and each actor $i$ in the network, the number of times that an actor $i$ gets to change its ties, $x_{ij}$, to other nodes $j \neq i$ in the network is dictated by a \emph{rate function} $\rho(x, \mathbf{z}, \boldsymbol{\alpha})$.
-For this paper, we assume a simple rate \st{"function"}, $\rho(x, \mathbf{z}, \boldsymbol{\alpha}) = \alpha_m$ that is constant for all actors between observations at time $t_m$ to $t_{m+1}$, \st{thus our rate function is just a rate parameter in the model}.
+For the network $x$ and each actor $i$ in the network, the number of times that an actor $i$ gets to change its ties, $x_{ij}$, to other nodes $j \neq i$ in the network is dictated by a \emph{rate function} $\rho(x, \mathbf{z}, \boldsymbol{\alpha})$, where $\boldsymbol{\alpha}$ are the parameters in the function $\rho$, and $x$ is the current network state, with covariates of interest $\mathbf{z}$.
+For this paper, we assume a simple rate \st{``function"}, $\rho(x, \mathbf{z}, \boldsymbol{\alpha}) = \alpha_m$ that is constant across all actors between observations at time $t_m$ to $t_{m+1}$, \st{thus our rate function is just a rate parameter in the overall model}.
+In general, SAOMs can incorporate covariate values and network statistics into the model, so that each node will have a different rate of change.
 \hh{XXX that means that the function is essentially just a rate parameter, right?} \st{XXX yes XXX}
@@
 I updated the language XXX->}%In our model with the simple rate function,
-\st{In our simple model with a rate parameter instead of a rate function,} the rate parameter determines how quickly actor $i$ gets an opportunity to change one of its ties, $x_{ij}$ in the time period from $t_{m}$ to $t_{m+1}$.
+\st{In our simple model with a simple rate parameter instead of a rate function,} the rate parameter dictates how quickly an actor $i$ gets an opportunity to change one of its ties to the other nodes in the network, $x_{ij}$, for $j \in \{1, \dots, n\}$ in the time period from $t_{m}$ to $t_{m+1}$.
+If $j = i$, no change in the network is made.
 The model also assumes that the actors $i$ are conditionally independent given their ties, $x_{i1}, \dots, x_{in}$ at the current network state.
-\st{Let $\tau_i(m)$ be the wait time until actor $i$ makes its next change.For any time point, $T$, where $t_{m} \leq T < t_{m+1}$, the waiting time to the next change opportunity by actor $i$ is exponentially distributed with expected value $\alpha_m^{-1}$.
+\st{Let $\tau(i|x,m) $ be the wait time until actor $i$ makes its next change from its current state in the network $x$.
+Note that $m$ indicates the number of the wave that is conditioned on in the SAOM.
+For any time point, $T$, where $t_{m} \leq T < t_{m+1}$, the waiting time to the next change opportunity by actor $i$ is exponentially distributed with expected value $\alpha_m^{-1}$.
 The conditional independence assumption is expressed in Equation~\ref{eq:c1}.
@@
 \begin{equation} \label{eq:c1}
- \tau_i(m) | x_{i1}(m), \dots, x_{in}(m) \stackrel{\text{iid}}{\sim} Exp(\alpha_m)
+ \tau(i|x,m) | x_{i1}(m), \dots, x_{in}(m) \stackrel{\text{iid}}{\sim} Exp(\alpha_m)
 \end{equation}
@@
  The waiting time to the next change opportunity by \textit{any} actor in the network is also exponentially distributed with expected value $(n\alpha_m)^{-1}$.
-The waiting time for the whole network to change, $\tau(m) = \sum_i \tau_i(m) |x_{i1}(m), \dots, x_{in}(m) $ can then be written as
+The distribution of waiting time for the whole network to change, $\tau(x|m) = \sum_i \tau_i(m) |x_{i1}(m), \dots, x_{in}(m) $ can then be written as
 
 \begin{equation} \label{eq:c2}
- \tau(m) \sim Exp(n\alpha_m)
+ \tau(x|m) \sim Exp(n\alpha_m)
 \end{equation}
 
-} The parameter for the wait time for the whole network $n\alpha_m$ is the rate at which any tie change occurs. %In order to achieve the memorylessness property of a Markov process,
+} The parameter for the wait time for the whole network $n\alpha_m$ is the rate at which any tie change occurs.
+The estimation of this parameter is straightforward: a the method of moments is used to estimate the rate with the statistic $$C = \sum_i\sum_j |x_{ij}(t_{m+1}) = x_{ij}(t_m)|$$ which is the total number of changes from observation at time $t_m$ to the observation at time $t_{m+1}$.
+%In order to achieve the memorylessness property of a Markov process,
 
@@
 \hh{shouldn't that be independence? - make the conditional independence assumption above a labelled formula (e.g. C1) and then call it here }
-Because of the conditional \st{in}dependence assumptions in Equation~\ref{eq:c1}, we can consider the objective function for each node separately, since only one tie from one node is \st{allow to change} at a time.
+Because of the conditional \st{in}dependence assumptions given in Equation~\ref{eq:c1}, we can consider the objective function for each node separately, as only one tie from one node is \st{allowed to change} at a time.
 \st{The node $i$, which is the node that is chosen to change at the current time point, is called the \textit{ego} node.
 It has the potential to interact with all other nodes in the network, $j \neq i$.
-These nodes $j$, are referred to as \textit{alters}.
-These nodes are acted upon by the ego node, and they only act when they become the ego node at a subsequent time point.} So, for the ego node, $i$, in the current network state $x$, the objective function is then written as
+These nodes $j$, are referred to as \textit{alter} nodes, or simply \textit{alters}.
+These nodes are acted upon by the ego node, and they only act when they become the ego node at a subsequent time point in the CTMC.} For the ego node, $i$, in the current network state $x$, the objective function is written as
 \begin{equation} \label{eq:of1}
@@
 
-for $x \in \mathcal{X}$, the space of all possible directed networks with the $n$ nodes, and $\mathbf{Z}$ matrix of covariates.
-The vector $\boldsymbol{\beta}$ are the parameters of the model with corresponding network and covariate statistics, $s_{ik}(x, \mathbf{Z})$, for $k = 1,\dots, K$.
+for $x \in \mathcal{X}$, the space of all possible directed networks with the $n$ nodes, and $\mathbf{Z}$, the matrix of covariates.
+The vector $\boldsymbol{\beta}$ contains the parameters of the model with corresponding network and covariate statistics, $s_{ik}(x, \mathbf{Z})$, for $k = 1,\dots, K$.
 \hh{XXX ego node isn't defined yet - the description of the charts would be a good place for it.} \st{XX I added it above the obj function equation.
-Is it ok there?} Given the ego node, $i$, there are $n$ possible steps for the actor $i$ to take: either one of all current ties $x_{ij} = 1$ will be destroyed, a new tie will be created, or no change will occur.
-
-The parameters, $\boldsymbol{\beta}$, are attached to various actor-level network statistics, $s_{ik}(x)$.
-%\hh{'There are always' ...
+Is it ok there?} Given the ego node, $i$, there are $n$ possible steps for the actor $i$ to take: either one of all current ties $x_{ij} = 1$ will be destroyed, a new tie will be created thatis currently $x_{ij}=0$, or no change will occur.
+
+The parameters, $\boldsymbol{\beta}$, correspond to various actor-level network statistics, $s_{ik}(x)$.
+\hh{'There are always' ...
 that seems to come from a Snijder paper?

### 2017-08-18 paper.Rnw
through sec 3.2

@@
 These nodes $j$, are referred to as \textit{alter} nodes, or simply \textit{alters}.
-These nodes are acted upon by the ego node, and they only act when they become the ego node at a subsequent time point in the CTMC.} For the ego node, $i$, in the current network state $x$, the objective function is written as
+These nodes are acted upon by the ego node, and they only act when they become the ego node at a subsequent time point in the CTMC.} For the ego node, $i$, in the current network state $x$, its objective function, which it tries to maximize, is written as
 \begin{equation} \label{eq:of1}
@@
 According to \citet[p.
-371]{snijders01}, there should be at least two parameters included in the model: $\beta_1$ for the outdegree of a node, and $\beta_2$ for the number of reciprocal ties held by a node \st{The inclusion of these effects has origin in the classical exponential random graph model (ERGM) for networks.
-The statistics corresponding to these effects are written in terms of the edge variables $x_{ij}$, for $i \neq j$.} In the \texttt{RSiena} software, there are over 80 possible parameters to add to the model.
+371]{snijders01}, there should be at least two parameters included in the model: $\beta_1$ for the outdegree of a node, and $\beta_2$ for the number of reciprocal ties held by a node.
+\st{These effects should seem familiar to readers used to working with the classical exponential random graph model (ERGM) for networks.
+The outdegree represents the propensity of nodes with a lot of outgoing ties to form more outgoing ties (the "rich get richer" effect), and the reciprocity parameter measures the tendency of outgoing ties to be returned within a network.
+The statistics corresponding to these effects are written in terms of the edge variables $x_{ij}$, for $i \neq j$.} In the \texttt{RSiena} software that we use to fit the SAOMs, there are over 80 possible parameters to add to the model.
 The formulas for the effects are provided in \citet{RSienamanual}.
-The parameters, $\beta_k$, in the model can be split up into two groups: first, the structural effects, whose estimation only depend on the structure of the network, like the outdegree and reciprocity parameters mentioned above.
-\hh{XXX move the last colored section to the front of this paragraph.
+The parameters, $\beta_k$, in the model can be split up into two groups: first, the structural effects, whose estimation depends only on the structure of the network, like the outdegree and reciprocity parameters mentioned above.
+The parameters are included when the researcher hypothesizes that they will model underlying mechanisms of network change.
+They hope to answer questions such as, ``How does the existing network structure influence change in the network?" \hh{XXX move the last colored section to the front of this paragraph.
 That sets up the structure nicely, e.g.
@@
 (sorry again...)}
+% and ``How do the behavior and characteristics of the nodes influence change in the network?" put this sentence below somewhere
 The second set of effects are referred to as the actor-level or covariate effects.
-These covariate effects also depend on the structure of the network, and are written in terms of the tie variables $x_{ij}$, but also in terms of the covariates, $\mathbf{Z}$.
+These covariate effects also depend on the structure of the network, with the additional inclusion of node-level covariates of interest.
+The covariate effects are written in terms of the tie variables $x_{ij}$, but also in terms of the covariates, $\mathbf{Z}$.
 A table of some possible structural and covariate effects is given in Table~\ref{tab:effects}.
@@
 it attempts to maximize the value of its objective function $f_i(\boldsymbol{\beta}, x)$ plus a random element, $U_i(x)$, to account for unknown attraction between nodes.
-In \citet{modelsSnijders}, the author recommends that the $U_i(x)$ be random draws from a type 1 extreme value distribution on top of the objective function.
+In \citet{modelsSnijders}, it is recommended that the $U_i(x)$ be random draws from a type 1 extreme value distribution.
+The additional random element is included to account for any random, unexplainable change in the network ties.
 % XX What is the thought behind that?
@@
 \begin{equation} \label{eq:dist1}
-f(u|\mu, \sigma) = \frac{1}{\sigma}\exp\left\{-\left(\frac{u-\mu}{\sigma} + e^{-\frac{u-\mu}{\sigma}}\right)\right\}.
+g(u|\mu, \sigma) = \frac{1}{\sigma}\exp\left\{-\left(\frac{u-\mu}{\sigma} + \exp^{-\frac{u-\mu}{\sigma}}\right)\right\}.
 \end{equation}
 
+<<logweibull, fig.align='center', out.width = '.5\\linewidth', fig.cap='The probability distribution function for the type 1 extreme value distribution, also known as the log-weibull or Gumbel distribution with location parameter \$\\mu = 0\$ and scale parameter \$\\sigma = 0\$.'>>=
+library(reliaR)
+x <- seq(-3,10,.01)
+y <- dgumbel(x = seq(-3,10,.01), sigma = 1, mu=0)
+qplot(x = x, y= y, geom = "line") + 
+  annotate('text', x = 6, y = .25, 
+        label = "g(u) = exp(-(u + exp(-u)))",size=5)  + 
+  labs(x = "u", y = "g(u) (density)")
+@
+
+The probability density function for the type 1 extreme value distribution is shown in Figure~\ref{fig:logweibull}.
 Using the distribution given in Equation~\ref{eq:dist1} with mean $\mu = 0$ and scale $\sigma = 1$ as \citet{modelsSnijders} suggests \hh{XXX I'm not following - which distribution do you mean by `this' distribution?
@@
 I got lost at the  ``objective function $f_i(\boldsymbol{\beta}, x)$ plus a random element, $U_i(x)$"} \st{XXX I hope it is clearer now above}
-is convenient because its \st{properties lead to a probability formula for} the probability that actor $i$ chooses to change its tie to actor $j$ \st{that can be written only} in terms of the objective function.
+is convenient because it \st{leads to a simple probability formula for} the probability that actor $i$ chooses to change its tie to actor $j$ \st{that can be written \textit{only}} in terms of the objective function.
 Let $p_{ij}(\boldsymbol{\beta}, x)$ be the probability that actor $i$ chooses to change its tie to actor $j$.
-Next, we write the network $x$ in its potential future state, where the tie $x_{ij}$ has changed to $1-x_{ij}$, as $x(i \leadsto j)$.
+Next, we write the network $x$ in its potential future state, $x(i \leadsto j)$, where the tie $x_{ij}$ has changed to $1-x_{ij}$.
 Then, the probility that the tie $x_{ij}$ changes is
@@
 
+When $i = j$ in $p_{ij}$, the numerator represents the exponential of the value of the objective function when evaluated at the current network state.
+When the value of the objective function is high at the current state, the probability of not making a change in a microstep is also high.
+In the CTMC, when actor $i$ may make a change, it chooses which tie $x_{i1, \dots, x_{in}}$ to change at random according to the probabilities $p_{ij}(\boldsymbol{\beta}, x)$.
 \hh{XXX This paragraph ends a bit abruptly - what is the future state probability used for?} \st{XX The future state probability is used within the continuous time markov chain to move through the different network states.
@@
 
-\st{XXX TO DO XXX}
+In the continuous time Markov chain literature, the intensity function... see \citet{ctmc}
 

### 2017-08-18 paper.Rnw
to section 3.3. thanks, iphone hotspot for letting me work in the car! :)

@@
 
-<<logweibull, fig.align='center', out.width = '.5\\linewidth', fig.cap='The probability distribution function for the type 1 extreme value distribution, also known as the log-weibull or Gumbel distribution with location parameter \$\\mu = 0\$ and scale parameter \$\\sigma = 0\$.'>>=
+<<logweibull, fig.align='center', out.width = '.5\\linewidth', fig.cap='The probability distribution function for the type 1 extreme value distribution, also known as the log-weibull or Gumbel distribution with location parameter $ \\mu = 0$ and scale parameter $ \\sigma = 0$.'>>=
 library(reliaR)
@@
 \subsubsection{Continuous Time Markov Chain}
-
-In the continuous time Markov chain literature, the intensity function... see \citet{ctmc}
+\st{
+In the continuous-time Markov chain literature, see for instance \citet{ctmc}, the chains are characterized by their \textit{generator} or \textit{intensity} matrix $\mathbf{Q}$.
+This matrix describes the rate of change between two states of the CTMC process, and the rows of this matrix always add to zero.
+For directed networks with binary edge variables like the ones we will be working with, there are a very large number of possible states, $2^{n(n-1)}$: there are two possible states for an edge, and there are $n(n-1)$ edge relationships because self-ties are not allowed.
+The intensity matrix for a CTMC in a SAOM is then a square matrix of dimension $2^{n(n-1)} \times 2^{n(n-1)}$.
+Since only one tie changes at a time in the CTMC, there are only $n(n-1)$ reachable states from the current network state.
+Thus, the intensity matrix $\mathbf{Q}$ is very sparse, with only $n(n-1) + 1$ non-zero entries in each row.
+Note that $n(n-1)$ of these represent the possible states that are one edge different from a given state, and the additional non-zero entry is for the state to remain unchanged.
+All other entries in a row are zero because those network states cannot be reached from the current state by just one change.
+
+The two pieces of a SAOM, the rate function/parameter and the objective function, each contribute to the entries of the intensity matrix to describe the rate of change between two network states.
+The entries of $\mathbf{Q}$ are defined as follows: let $b \neq c \in \{1, 2, \dots, 2^{n(n-1)} \}$ be indices of two different possible states of the network, $x^b, x^c \in \mathcal{X}$.
+Then the $bc^{th}$ entry of $Q$ is:
+\[ q_{bc} = \begin{cases} 
+     q_{ij}= \alpha_m p_{ij}(\boldsymbol{\beta}, x^b) & \text{if } x^c \in \{x^b(i \leadsto j) | \text{ any } i \neq j \in \{1, \dots, n\}\} \\
+      0 & \text{if } \sum_i \sum_j |x_{ij}^c -x_{ij}^b | > 1 \\
+      -\sum_{i\neq j} q_{ij} & \text{if } x^b = x^c 
+   \end{cases}
+\]
+
+Thus, the rate of change between any two states, $x^b$ and $x^c$, that differ by only one tie $x_{ij}$, is the product of the rate at which actor $i$ gets to change a tie and the probability that the tie that will change is the tie to node $j$.
+
+XXX Need a transistion to next section XXX 
+}
 
@@
 
-To fit a SAOM to longitudinal network data, we use the package \texttt{RSiena} \citep{RSiena}.
+To fit a SAOM to observations of a dynamic network, we use the package \texttt{RSiena} \citep{RSiena}.
 This package uses simulation methods to estimate parameter values using either the method of moments or maximum likelihood estimation.
-In this paper, we focus of the method of moments estimation because the theory behind it was established in \citet{saompaper}, while the maximum likelihood estimation methods were not fully established until \citet{saomsml2010}.
-We also focus on the score function method for estimating the derivatives of the expected values, as opposed to the finite differences method, both of which are outlined in detail in \citet{SienaAlgs}.
+In this paper, use the method of moments estimation because the theory behind it was established in \citet{saompaper}, while the maximum likelihood estimation methods were not fully established until \citet{saomsml2010}, though \texttt{RSiena} contains capabilities to use maximum likelihood estimation.
+We also use the score function method for estimating the derivatives of the expected values, as opposed to the finite differences method, both of which are outlined in detail in \citet{SienaAlgs}.
 
@@
 $$ E_{\theta}S = s_{obs}$$
-where $\theta$ is the vector of rate and objective function parameters, and $s_{obs}$ is the observed vector of model statistics, $S$.
-The entire algorithm is provided in \citet{SienaAlgs}.
-\par There are three phases in the the SIENA algorithm, as described in \citet{RSienamanual, SienaAlgs}.
-The first phase performs initial estimation of the score functions used in the Robbins-Monro procedure.
-The second phase carries out the Robbins-Monro algorithm and obtains iterative estimates of the parameter values.
-The third phase uses the parameter vector estimated in phase two to estimate the score functions and the covariance matrix of the parameters, as well as performs convergence checks.
-In each of the the first two phases, the estimation procedure also relies on ``microsteps" that simulate from the model as it exists in its current state in order to update either the score functions or the parameter estimates.
-These microsteps become observed instances of the continuous-time Markov chain that is the backbone of the stochastic actor-oriented model.
-In Section~\ref{sec:ModelVis}, we visualize the parts of the SIENA method-of-moments algorithm, bringing them out of the ``black-box" and into the light.
+where $\theta$ is the vector of rate and objective function parameters, and $s_{obs}$ is the observed vector of model statistics, $S$. The entire algorithm is provided in \citet{SienaAlgs}.
+
+There are three phases in the the SIENA algorithm, as described in \citet{RSienamanual, SienaAlgs}.
+The first phase performs initial estimation of the score functions for use in the Robbins-Monro procedure for method-of-moments estimation.
+The second phase carries out the Robbins-Monro algorithm and obtains estimates of the parameter values through iterative updates and simulation from the CTMC at current parameter values.
+The third phase uses the parameter vector estimated in phase two to estimate the score functions and covariance matrix of the parameter estimate, and also carries out convergence checks.
+In each of the the first two phases, the estimation procedure also uses ``microsteps" that simulate from the model as it exists in its current state in order to update either the score functions or the parameter estimates.
+These simulated microsteps are observed instances of the continuous-time Markov chain that is the backbone of the stochastic actor-oriented model.
+In Section~\ref{sec:ModelVis}, we further explore these phases in the SIENA method-of-moments algorithm through visualization, bringing them out of the ``black-box" and into the light.
 

### 2017-08-19 paper.Rnw
not compiling

@@
 
-The \texttt{RSiena} software that fits the models to data also comes equipped with the \texttt{sienaGOF()} function for examining model fit.
+The \texttt{RSiena} software that fits the models to data also includes a goodnes-of-fit function for examining model fit, \texttt{sienaGOF()}.
 This function ``assess[es] the fit of the model with respect to auxiliary statistics of networks" \citep[p.~53]{RSienamanual}.
-Examples of the auxiliary statistics are the out- or indegree distribution on the nodes, with the option for users to input their own statistics to examine.
-The fit is tested using Mahalanobis distance to determine a $p$-value when comparing the statistics calculated from the observed data (excluding the first observation that is conditioned on) to the distribution of the statistics when calculated on the simulated waves.
-So, good fit is acheived when $p$ is large.
-\st{XX more detail about mahalanobis distance and calculating p-value XX}
-
-The \texttt{plot.sienaGOF()} function allows us to visualize this fit.
-Below is an example of what the goodness-of-fit plot output looks like for indegree and outdegree statistics for a small model.
-Box plots and violin plots are drawn on top of each other in Figure~\ref{fig:sienagof} in order to show the distribution of the simulated network observations compared to the true data shown in red points connected by red lines.
-If the red points lie "well within" the simulated values, the model is a good fit to the data.
-The model examined in Figure~\ref{fig:sienagof} appears to do a better job of capturing the outdegree distribution than the indegree distribution of the data.
-
-<<sienagof, fig.cap="Goodness of fit measures for two network statistics: indegree and outdegree cumulative distributions.
-On the x-axis is the degree value, and on the y-axis is the number of times that degree value occurs in the data.
-\\st{The p-value shown on the x-axis is ....
-}", cache = TRUE, message = FALSE, warning = FALSE>>=
+Examples of auxiliary statistics include the out- or indegree distribution on the nodes, with the option for users to input their own statistics to examine.
+The goodness-of-fit is evaluated as follows:
+\begin{enumerate}
+\item The auxiliary statistics are computed on the observed data and on $N$ simulated observations from the model. Typically, $N=1000$. 
+\item The mean vector and covariance matrix of the statistics on the simulations from the model are computed.
+\item The Mahalanobis distance from the observed statistics to the distribution of the simulated statistics is computed using the mean and covariance found in step 2. 
+\item The Mahalanobis distrance from each of the $N$ simulations to the same distribution is computed, and the Mahalanobis distance of the observed data is compared to this distribution of distances.
+\item A $p$-value is found by computing the proportion of simulated distances found in step 4 that are as large or larger than the Mahalanobis distance from the data.
+A SAOM is thus considered a good fit if $p$ is large.
+\end{enumerate}
+ %\st{XX more detail about mahalanobis distance and calculating p-value XX}
+
+<<sienagof, fig.cap="Goodness of fit measures for two network statisticscomputed on the first simulated wave of a SAOM: indegree and outdegree cumulative distributions.
+On the x-axis is the degree value, and on the y-axis is the number of times that degree value occurs in the network.
+\\st{The p-value shown on the x-axis is the proportion of observations simulated from the model that have observed statistic values as large or larger than the values observed in the data.}", cache = TRUE, message = FALSE, warning = FALSE, fig.height=3>>=
 load("data/ansnullpaper.rda")
@@
 
+The \texttt{plot.sienaGOF()} function allows us to visualize this fit.
+In Figure~\ref{fig:sienagof}, we provide an example of what the goodness-of-fit plot output looks like for indegree and outdegree statistics for a small model.
+Box plots and violin plots are drawn on top of each other in the figure in order to show the distribution of the simulated network observations compared to the true data shown in red points connected by red lines.
+If the red points lie "well within" the simulated values, the model is a good fit to the data.
+The model examined in Figure~\ref{fig:sienagof} appears to do a better job of capturing the outdegree distribution than the indegree distribution of the data.
+
 \subsection{Example Data}
 
-To guide our visual exploration of stochastic actor-oriented models, we used a subset of the 50 actor dataset from the ``Teenage Friends and Lifestyle Study" that is provided on the \texttt{RSiena} webpage.
-These data come from \citet{friendsdata}, and we chose to only work with a subset of the data to make network visualizations less busy, with changes that are more noticeable.
-The subset contained actors 20 through 35 and the ties between them, as well as the drinking behavior of each actor at each of the three waves.
-This specific subset was chosen because it showed somewhat higher connectivity than other subsets, as we've emphasized in the visualizations of the three network adjacency matrices in Figure~\ref{fig:adjmatviz}.
-For model fitting, we condition on wave 1 and estimate the parameters of the models from the second and third waves.
-We will also be working with one actor level categorical covariate, drinking behavior.
-This variable has five values in the original data: (1) does not drink, (2) drinks once or twice a year, (3) drinks once a month, (4) drinks once a week, and (5) drinks more than once a week.
-The actor covariates and networks are visualized in Figures~\ref{fig:adjmatviz} and~\ref{fig:alldataviz}.
 <<adjmatviz, fig.align='center', fig.cap="A visualization of the adjacency matrices of the three waves of network observations in the ``Teenage Friends and Lifestyle Study'' data.
@@
 
+To guide our visual exploration of stochastic actor-oriented models, we use two data sources.
+The first is a subset of the 50 actor dataset from the ``Teenage Friends and Lifestyle Study" that is provided on the \texttt{RSiena} webpage.
+These data come from \citet{friendsdata}, and we chose to only work with a subset of the data to make network visualizations less busy and to make any changes in the network more noticeable.
+The subset contained actors 20 through 35 and the ties between them, as well as the drinking behavior of each actor at each of the three waves.
+This specific subset was chosen because it showed somewhat higher connectivity than other subsets, as we've emphasized in the visualizations of the three network adjacency matrices in Figure~\ref{fig:adjmatviz}.
+For model fitting, we condition on wave 1 and estimate the parameters of the models from the second and third waves.
+We will also be working with one actor level categorical covariate, drinking behavior.
+This variable has five values in the original data: (1) does not drink, (2) drinks once or twice a year, (3) drinks once a month, (4) drinks once a week, and (5) drinks more than once a week.
+The network and the actor covariate values are visualized using a node-link diagram in Figure~\ref{fig:alldataviz}.
+
+\st{The second data example we use is a collaboration network in the United States Senate during the $111^{th}$ through $114^{th}$ Congresses.
+These sessions of congress correspond to the years of Barack Obama's presidency, from 2009-2016.\footnote{Details of how this data can be downloaded are provided by François Briatte at \url{https://github.com/briatte/congress}}.
+In this network, ties are directed from senator $i$ to senator $j$ when senator $i$ signs on as a cosponsor to the bill that senator $j$ authored.
+There are (somewhat surprisingly) many hundreds of ties between senators when they are connected in this way, so we simplify the network by computing a single value for each senator-senator collaboration called the \textit{weighted propensity to cosponsor} (WPC).
+This value is defined in \citet{senate} as
+\begin{equation}\label{eq:sen1}
+WPC_{ij} = \dfrac{\sum\limits_{k=1}^{n_j} \frac{Y_{ij(k)}}{c_{j(k)}}}{\sum\limits_{k=1}^{n_j} \frac{1}{c_{j(k)}}}
+\end{equation}
+
+where $n_j$ is the number of bills in a congressional session authored by senator $j$, $c_{j(k)}$ is the number of cosponsors on senator $j$'s $k^{th}$ bill, where $k \in \{1,\dots, n_j\}$, and $Y_{ij(k)}$ is a binary variable that is 1 if senator $i$ cosponsored senator $j$'s $k^{th}$ bill, and is 0 otherwise.
+This measure ranges in value from 0 to 1, where $WPC_{ij} = 1$ if senator $i$ is a cosponsor on every one of senator $j$'s bills and $WPC_{ij} = 0$ if senator $i$ is never a cosponsor any of senator $j$'s bills.
+
+<<senateplot, fig.cap='The collaboration network in the four senates during the Obama years, 2009-2016.
+Edges are shown only if the weighted propensity to cosponsor from one senator to another is greater than 0.25.', fig.height=3>>=
+seobama <- read_csv("data/congress/senateobamapres_gsw_25.csv")
+ggplot(data = seobama) + 
+  geom_net(directed = T, labelon=F, arrowsize = .25, singletons= F, fiteach = T, linewidth = .25,
+           aes(from_id = source, to_id = target, color = party),size = 1) + 
+  ThemeNet + 
+  scale_color_manual(values = c("royalblue", "forestgreen","firebrick")) + 
+  theme(legend.position = 'bottom') + 
+  facet_wrap(~senate, nrow = 1)
+@
+
+Because we require binary edges for our models, we focus only on very strong collaborations.
+The networks we constructed for the four senates during President Obama's administration are shown in Figure~\ref{fig:senateplot} so for our collaboration networks, $x$, edges are defined as
+\[
+  x_{ij} =
+  \begin{cases}
+                                   1 &  \text{if } WPC_{ij} > 0.25 \\
+                                   0 & \text{if } WPC_{in \leq 0.25}.
+  \end{cases}
+\]
+
+}
 \section{Model Visualizations} \label{sec:ModelVis}

### 2017-08-19 paper.Rnw
it compiles. Through to section 4.

@@
 On the x-axis is the degree value, and on the y-axis is the number of times that degree value occurs in the network.
-\\st{The p-value shown on the x-axis is the proportion of observations simulated from the model that have observed statistic values as large or larger than the values observed in the data.}", cache = TRUE, message = FALSE, warning = FALSE, fig.height=3>>=
+\\st{The p-value shown on the x-axis is the proportion of observations simulated from the model that have observed statistic values as large or larger than the values observed in the data.}", cache = TRUE, message = FALSE, warning = FALSE, fig.height=5>>=
 load("data/ansnullpaper.rda")
@@
 levels(view_rs2$wave) <- c("Wave 1", "Wave 2", "Wave 3")
-ggplot(data= view_rs2, aes(x = x, y=y)) + 
+ggplot(data= view_rs2, aes(x = y, y=-x)) + 
   geom_tile(aes(fill = as.factor(value))) + 
@@
   geom_rect(data= NULL, inherit.aes = FALSE, color = 'red',
-            aes(xmin = 19.5, xmax = 35.5, ymin = 19.5, ymax = 35.5, fill =NA)) + 
+            aes(xmin = 19.5, xmax = 35.5, ymax = -19.5, ymin = -35.5, fill =NA)) + 
   facet_wrap(~wave) + theme_bw() +
-  labs(x = "i", y = "j") + 
+  labs(x = "j", y = "i") + 
   ThemeNoNet + 
-  theme(aspect.ratio=1, axis.text = element_blank(), axis.ticks = element_blank()) 
+  theme(aspect.ratio=1, axis.text = element_blank(), axis.ticks = element_blank(), panel.grid = element_blank()) 
 @

### 2017-08-19 paper.Rnw
up to section 4.1

@@
 
-Because we require binary edges for our models, we focus only on very strong collaborations.
-The networks we constructed for the four senates during President Obama's administration are shown in Figure~\ref{fig:senateplot} so for our collaboration networks, $x$, edges are defined as
+Because we require binary edges for our models, we focus only on very strong collaborations. For our senate collaboration networks, $x$, edges are defined as 
 \[
@@
                                    1 &  \text{if } WPC_{ij} > 0.25 \\
-                                   0 & \text{if } WPC_{in \leq 0.25}.
+                                   0 & \text{if } WPC_{ij} \leq 0.25.
   \end{cases}
 \]
-
+The networks we constructed for the four senates during President Obama's administration are shown in Figure~\ref{fig:senateplot}. 
 }
@@
 
+Every good data analysis includes both numerical and visual summaries of the data, so why restrict model description and diagnostics to numerical summaries?
 The concept of model visualization was developed to complement traditional model diagnostic tools.
-Typically, numerical summaries such as $R^2$ are used to assess model fit, and visualizations, like residual plots, are restricted to the data.
-Every good data analysis includes both numerical and visual summaries of the data, so why should model description be restricted to numerical summaries?
-\citeauthor{modelvis} outline the three different parts of a model: the model family, the model form, and the fitted model.
-The latter is primarily what one thinks of first when considering a model in a data analysis, where a model is fit to data, and parameter estimates are reported.
-In the context of SAOMs, the fitted model contains the estimated rate parameters and the estimated objective function parameters.
-The model form describes the the model before the fitting process and how variables describe the data in the model.
+Typically, numerical summaries such as $R^2$ are used to assess model fit, and the occasional visualization, like a residual plot, are used to determine how well the model fits the data.
+\citeauthor{modelvis} outline three separate ideas, each of which are can be referred to simply as the "model": the model family, the model form, and the fitted model.
+The latter is primarily what one thinks of first when considering a model in a data analysis, where specified a model is fit to data, and parameter estimates and other numerical summaries, such as $R^2$ are reported.
+In the context of SAOMs, the fitted model contains the form of the rate and objective functions, the estimated rate parameters, and the estimated objective function parameters.
+The model form describes the the model \textit{before} the fitting process, defining which parameters are in the model within the context of the larger model family.
 In SAOMs, the model form includes description of the rate and objective functions and the variables therein that describe how the network evolves over time.
-Finally, the model family is the most broad description of the model.
-This is the type of model that you wish to fit to the data, and is chosen based on the problem and knowledge at had.
-For example, we chose to use a SAOM to model network data over an exponential random graph model (ERGM) because we believe that actor-level variables effect network dynamics.
-These three parts of a "model" are should be visualized according to the three principals of model visualization: display the model in the data space, look at a collection of models, and explore the process of fitting the model, not just the end result.
+Finally, the model family is the broadest description of the model.
+This is the type of model that you wish to fit to the data, and is chosen based on the problem, data, and knowledge at hand.
+For example, we chose to use a SAOM to model network data over an exponential random graph model (ERGM) because we believe that actor-level variables effect network structure and formation, and we wanted to model the network changes over time.
+These three uses of the term ``model" can each be visualized according to the three principals of model visualization: display the model in the data space, look at a collection of models, and explore the process of fitting the model, not just the end result.
+We turn our focus to the fitted model and the model form.
+Specifically, we want to know how the model form affects the fitted model.
+Using our example data sets and our visualization toolbox, we hope to answer this question for SAOMs.
 

### 2017-08-20 paper.Rnw
move model discussion to the front of section 4

@@
 Using our example data sets and our visualization toolbox, we hope to answer this question for SAOMs.
+
+\subsection{The Models}
+
+We first consider the 16 actor subset of the teenage friends and lifestyle data available on the \texttt{RSiena} website \citep{RSiena}.
+To this data, we fit three different SAOMs.
+Each SAOM used a simple rate function, $\alpha_m$, and an objective function with two or three parameters.
+The first model, $M1$, contains the absolute minimum number of parameters in the objective function $f_i(x)$:  $$f_i(x)^{M1} = \beta_1s_{i1} + \beta_2s_{i2},$$ where $s_{i1}$ is the density network statistic and $s_{i2}$ is the reciprocity network statistic for actor $i$ at the current network state $x$.
+The second and third models, $M2$ and $M3$, contain one additional parameter each in the objective function which were determined by a Wald-type test provided in the \texttt{RSiena} software to be significant, with $p$-values less than 0.05 \citep{RSienaTest}.
+The $M2$ model contains an actor-level covariate parameter, and the $M3$ model contains an additional strutural effect in the objective function.
+\begin{align*}
+f_i(x)^{M2} & = \beta_1s_{i1} + \beta_2s_{i2} + \beta_3s_{i3} \\
+f_i(x)^{M3} & = \beta_1s_{i1} + \beta_2s_{i2} + \beta_4s_{i4}, 
+\end{align*}
+
+\begin{figure}\label{fig:structures}
+\begin{subfigure}[t]{.45\textwidth}
+<<jtt, echo=FALSE, fig.width=2, fig.height=2, message = FALSE, warning = FALSE, fig.align='center',out.width='.49\\linewidth'>>=
+jTTe <- data.frame(from = c('i', 'i', 'h'), to = c('h', 'j', 'j'))
+jTTn <- data.frame(id = letters[8:10], group = c(1,1,2))
+
+jTT <- merge(jTTe, jTTn, by.x = 'from', by.y = "id", all = T)
+
+set.seed(12345) 
+ggplot(data = jTT, aes(from_id = from, to_id = to)) + 
+  geom_net(aes(shape = as.factor(group)), directed = T, labelon = T, 
+           labelcolour='grey80',vjust = 0.5, hjust =0.5, arrowgap = .15, 
+           colour = 'black', size=10, fontsize = 4,
+           ecolour = c("red", "grey40", "grey40", "grey40")) + 
+  expand_limits(x=c(-0.1,1.1), y=c(-0.1,1.1)) +
+  ThemeNet +
+  theme(legend.position = "none")
+@
+\caption{\label{fig:jtt}Realization of a jumping transitive triplet, where $i$ is the focal actor, $j$ is the target actor, and $h$ is the intermediary.
+The group of the actors is represented by the shape of the node.}
+\end{subfigure}\hfill
+\begin{subfigure}[t]{.45\textwidth}
+<<dab, echo=FALSE, fig.width=2, fig.height=2, message = FALSE, warning = FALSE, fig.align='center', out.width='.49\\linewidth'>>=
+dade <- data.frame(from = c('i', 'i', 'h', 'j'), to = c('h', 'j', 'k', 'k'))
+dadn <- data.frame(id = letters[8:11], group = c(1,1,1,1))
+
+dad <- merge(dade, dadn, by.x = 'from', by.y = "id", all = T)
+
+set.seed(12345) 
+ggplot(data = dad, aes(from_id = from, to_id = to)) + 
+  geom_net(aes(shape = as.factor(group)), directed = T, labelon = T, labelcolour='grey80',vjust = 0.5, hjust =0.5, arrowgap = .15, colour = 'black', size=10, fontsize = 4) + 
+  expand_limits(x=c(-0.1,1.1), y=c(-0.1,1.1)) +
+  ThemeNet +
+  theme(legend.position = "none")
+@
+\caption{\label{fig:dad}Doubly achieved distance between actors $i$ and $k$.}
+\end{subfigure}
+\caption{The additional network effects included in our models fit to the friends data. On the left, a jumping transitive triplet (JTT). On the right, a doubly achieved distance between $i$ and $k$.}
+\end{figure}
+
+where $s_{i3} = \sum_{j \neq h} x_{ij}x_{ih}x_{hj} \mathbb{I}(z_i = z_h \neq z_j)$, and $s_{i4} =  |\{j : x_{ij} = 0, \sum\limits_h x_{ih}x_{hj} \geq 2\}|$.
+These statistics are known as the number of jumping transitive triplets and the number of doubly achieved distances two effect, respectively.
+The first statistic emphasizes triad relationships that are formed between actors from different covariate groups, while the other emphasizes indirect ties between actors.
+These two effects are visually represented and further described in in Figure~\ref{fig:structures}
 
@@
 
-We first consider the 16 actor subset of the teenage friends and lifestyle data available on the \texttt{RSiena} website \citep{RSiena}.
-To this data, we fit three different SAOMs.
-Each SAOM used a simple rate function, $\alpha_m$, and an objective function with two or three parameters.
-The first model, $M1$, contains the absolute minimum number of parameters in the objective function $f_i(x)$:  $$f_i(x)^{M1} = \beta_1s_{i1} + \beta_2s_{i2}$$, where $s_{i1}$ is the density network statistic and $s_{i2}$ is the reciprocity network statistic for actor $i$.
-The second and third models, $M2$ and $M3$, contain a third parameter in the objective function that was determined in the\texttt{RSienaTest} software to be significant, with $p$-values less than 0.05 \citep{RSienaTest}.
-The $M2$ model contains an actor-level covariate parameter, and the $M3$ model contains an additional strutural effect in the objective function.
-\begin{align*}
-f_i(x)^{M2} & = \beta_1s_{i1} + \beta_2s_{i2} + \beta_3s_{i3} \\
-f_i(x)^{M3} & = \beta_1s_{i1} + \beta_2s_{i2} + \beta_4s_{i4}, 
-\end{align*} where $s_{i3} = \sum_{j \neq h} x_{ij}x_{ih}x_{hj} \mathbb{I}(z_i = z_h \neq z_j)$, and $s_{i4} =  |\{j : x_{ij} = 0, \sum\limits_h x_{ih}x_{hj} \geq 2\}|$.
-These statistics are known as the number of jumping transitive triplets and the number of doubly achieved distances two effect, respectively.
-The first statistic emphasizes triad relationships that are formed between actors from different covariate groups, while the other emphasizes indirect ties between actors.
-These two effects are visually represented in Figure~\ref{fig:jtt} and ~\ref{fig:dad}, respectively.
-
-\begin{figure}\label{fig:structures}
-\begin{subfigure}[t]{.45\textwidth}
-\caption{\label{fig:jtt}Realization of a jumping transitive triplet, where $i$ is the focal actor, $j$ is the target actor, and $h$ is the intermediary.
-The group of the actors is represented by the shape of the node.}
-<<jtt, echo=FALSE, fig.width=2, fig.height=2, message = FALSE, warning = FALSE, fig.align='center',out.width='.49\\linewidth'>>=
-jTTe <- data.frame(from = c('i', 'i', 'h'), to = c('h', 'j', 'j'))
-jTTn <- data.frame(id = letters[8:10], group = c(1,1,2))
-
-jTT <- merge(jTTe, jTTn, by.x = 'from', by.y = "id", all = T)
-
-set.seed(12345) 
-ggplot(data = jTT, aes(from_id = from, to_id = to)) + 
-  geom_net(aes(shape = as.factor(group)), directed = T, labelon = T, 
-           labelcolour='grey80',vjust = 0.5, hjust =0.5, arrowgap = .15, 
-           colour = 'black', size=10, fontsize = 4,
-           ecolour = c("red", "grey40", "grey40", "grey40")) + 
-  expand_limits(x=c(-0.1,1.1), y=c(-0.1,1.1)) +
-  ThemeNet +
-  theme(legend.position = "none")
-@
-\end{subfigure}\hfill
-\begin{subfigure}[t]{.45\textwidth}
-\caption{\label{fig:dad}Doubly achieved distance between actors $i$ and $k$.}
-<<dab, echo=FALSE, fig.width=2, fig.height=2, message = FALSE, warning = FALSE, fig.align='center', out.width='.49\\linewidth'>>=
-dade <- data.frame(from = c('i', 'i', 'h', 'j'), to = c('h', 'j', 'k', 'k'))
-dadn <- data.frame(id = letters[8:11], group = c(1,1,1,1))
-
-dad <- merge(dade, dadn, by.x = 'from', by.y = "id", all = T)
-
-set.seed(12345) 
-ggplot(data = dad, aes(from_id = from, to_id = to)) + 
-  geom_net(aes(shape = as.factor(group)), directed = T, labelon = T, labelcolour='grey80',vjust = 0.5, hjust =0.5, arrowgap = .15, colour = 'black', size=10, fontsize = 4) + 
-  expand_limits(x=c(-0.1,1.1), y=c(-0.1,1.1)) +
-  ThemeNet +
-  theme(legend.position = "none")
-@
-\end{subfigure}
-\caption{Structural newtwork effects. On the left, a jumping transitive triplet (JTT). On the right, a doubly achieved distance between $i$ and $k$.}
-\end{figure}
 We these three models using \texttt{RSiena} 1,000 times each.

### 2017-08-20 paper.Rnw
move sec 4.1 to be sec 4.3. 4.2 becomes 4.1, 4.3 becomes 4.2

@@
 
-\subsection{Explore algorithms, not just end result}
-
-The model fitting process in \texttt{RSiena} involves several simulation steps that are hidden from the user.
-This is practical and efficient if a researcher is primarily interested in fitting one model to a set of longitudinal network data and obtaining parameter estimates.
-We are more interested in how the models are fit, so we extracted and explored the different steps of that process.
-
-A key component of each step of the SIENA method of moments algorithm is the ``microstep." This step simulates from the model in its current state, $x(t_m)$ with current parameter values $\theta_{0}$, to the next state, $x(t_{m+1})$.
-This microstep process stops when the simulated network has acheived the same number of differences, $C$, from $x(t_m)$ as $x(t_{m+1})$, where
-$$C = \sum_{i \neq j} |x_{ij}(t_{m+1}) - x_{ij}(t_{m})|.$$ 
-This simulation process follows the steps of the continuous-time Markov chain.
-First, an ``ego node'' is selected to make a change, and then it randomly makes one change in its ties according to the probability, $p_{ij}$, determined by its objective function.
-The options for change are (1) removing a current tie, (2) adding a new tie, or (3) making no change at all.
-Saving all of these steps is not efficient if you're only interested in estimating parameter values, but they can be saved and extracted using options in \texttt{RSiena}, which is exactly what we did to create our visualizations.
-Between two network observations $x(t_m)$ and $x(t_{m+1})$, there can be dozens, hundreds, or even thousands of microsteps, depending on the size of the network.
-We wanted to see these in-between steps in order to better understand the behavior of the underlying continuous-time Markov chain.
-
-The first vizualization we present here is an animation of microsteps simulated to emulate the transition steps from wave 1 to wave 2 the subset of the friends data shown in Figure ~\ref{fig:alldataviz}.
-Movies similar to our animation were used to visualize the changes of dynamic networks in \citet{dynnetviz}.
-When each ego node is selected in a microstep, it is emphasized in the animation, then the associated edge either appears or disappears.
-If there are no changes at a particular microstep, no changes are seen.
-Some frames of the animations are shown in Figure~\ref{fig:makinggif}, and the full movie can be viewed at INSERT LINK TO GIF.
-
-<<makinggif, fig.cap='A selection of images in the microstep animation.
-The selected edges and nodes are emphasized by changing the size and the color, then when edges are removed, they fade out, shrinking in size, while the nodes change color and shrink to blend in with the other nodes.', fig.width=8, cache=TRUE, message = FALSE, warning = FALSE>>=
-# Need old version of dplyr to get this to work: 
-if (!require("dplyr050",character.only = TRUE)) {
-      devtools::install_github("sctyner/dplyr050")
-}
-
-friend.data.w1 <- as.matrix(read.table("data/s50_data/s50-network1.dat"))
-friend.data.w2 <- as.matrix(read.table("data/s50_data/s50-network2.dat"))
-friend.data.w3 <- as.matrix(read.table("data/s50_data/s50-network3.dat"))
-drink <- as.matrix(read.table("data/s50_data/s50-alcohol.dat"))
-fd2.w1 <- friend.data.w1[20:35,20:35]
-fd2.w2 <- friend.data.w2[20:35,20:35]
-fd2.w3 <- friend.data.w3[20:35,20:35]
-friendshipData <- array(c(fd2.w1, fd2.w2,fd2.w3), dim = c(16, 16, 3))
-
-# running Siena
-#friendship <- sienaDependent(friendshipData)
-#alcohol <- varCovar(drink[20:35,])
-#mydata <- sienaDataCreate(friendship, alcohol)
-#myeffnull <- getEffects(mydata)
-#myalgorithm <- sienaAlgorithmCreate(projname = 's16_3')
-set.seed(823746)
-
-ansnullchains <- get_chain_info(ansnull)
-ansnullchains %>%
-  filter(period == 1) %>%  #only look at chains from wave 1 to wave 2
-  group_by(rep) %>%
-  dplyr::select(rep, from = ego, to = alter) %>%
-  mutate(val = as.numeric(!from == to),
-         from = paste0("V", parse_number(from)+1), # make the chains
-         to = paste0("V", parse_number(to)+1)) -> ansnullchainsw1w2
-colnames(fd2.w1) <- paste0("V", 1:16)
-rownames(fd2.w1) <- paste0("V", 1:16)
-wave1friends <- fortify(as.adjmat(fd2.w1))
-ms1 <- listMicrosteps(dat = wave1friends,
-                      microsteps = filter(ansnullchainsw1w2, rep == 1))
-microsteps <- ms1[1:20]
-
-microsteps_df <- getMicrostepsDF(microsteps)
-pte <- pretween_edges(microsteps = microsteps)
-ptv <- pretween_vertices(pte = pte, layoutparams = list(n = 16))
-# step 1 is remove, step 9 is add. 
-# get tweened data for all microsteps
-pall <- tween_microsteps(ptv, pte, microsteps_df)
-
-static_pall <- filter(pall, .frame %in% c(15,20,22,24,26:28, 71:73, 75, 79, 81, 84))
-
-static_pall$type <- NA
-static_pall$type[which(static_pall$.frame %in% c(15,20,22,24,26:28))] <- "Remove Edge"
-static_pall$type[which(static_pall$.frame %in% c(71:73, 75, 79, 81, 84))] <- "Add Edge"
-
-set.seed(34569234)
-create_net_animate(static_pall) + 
-  facet_wrap(~.frame, nrow=2, labeller = 'label_both') + 
-  ThemeNet + 
-  theme(panel.background = element_rect(color = 'black'))
-@
-
-The top row of Figure~\ref{fig:makinggif} shows an edge being removed, and the bottom row shows one being added.
-In both cases, the ego nodes chosen to act change color from black to red, and they also increase greatly in size.
-In the case of an edge being removed, the edge that currently exists is emphasized with the color and size change that the node also gets, and as the animation proceeds the edge shrinks to disappearing, while the node shrinks and changes color back to the original black.
-If the edge is being added, it appears colored red from nothing and grows to a large size, then changes color to match the rest of the edges, while the node shrinks and changes color to match the other nodes.
-
-We also use animation to view the changing structure of the adjacency matrix the microsteps.
-The adjacency matrices for the three waves of friendship data as shown in Figure~\ref{fig:alldataviz} are ordered by node id.
-There are 16 nodes in the data, numbered 1-16, and that order is used on the $x$ and $y$ axes for the matrix visualization.
-Viewing the adjacency matrices with this arbitrary ordering does not provide much information to the viewer about the underlying structure of the network.
-This lack of perceived structure would be exacerbated in an animation.
-So, we adjust the ordering so that the viewer can perceive more of the structure of the graph.
-This process is known as matrix seriation \citep{seriation}.
-
-To reorder the vertices for the matrix visualization, we first constructed a cumulative adjaceny matrix, $\mathbf{A}^{cum}$, for the series of microsteps simulating the network from $x(t_m)$ to $x(t_{m+1})$.
-A single entry in the cumulative adjacency matrix, $\mathbf{A}^{cum}_{ij}$, is the total number of times the edge from node $i$ to node $j$ appears in the network from the initial observation, $x(t_m) \equiv X(0)$ to the final reult of the last microstep, $X(C)$:
-$$ \mathbf{A}^{cum}_{ij} = \sum_{c = 0}^C X_{ij}(c). $$
-We then performed a principal component analysis on $\mathbf{A}^{cum}$, and used the values of the first principal component to order the vertices on the $x$ and $y$ axes for the adjacency matrix animation.
-For one such series of microsteps simulated by \texttt{RSiena}, we present the adjacency matrix ordered by the (arbitrary) vertex id alongside the seriated adjacency microsteps using the first principal component loading on the cumulative adjacency matrix, $\mathbf{A}^{cum}$, in Figure~\ref{fig:adjmatorder}.
-
-<<adjmatorder, message = FALSE, warning = FALSE, fig.cap="On the left, the starting friendship network represented in adjacency matrix form, ordered by vertex id.
-On the right, the same adjacency matrix is presented after ordering the vertices by one repitition of the microstep simulation process from wave one to wave two.", fig.width=8, fig.height=3>>=
-# first present the original wave 1 adj mat, then present the adj mat after being ordered by the 1st pc of the big adjmat A. 
-# reinstall new dplyr
-numordersf <- data.frame(fd2.w1) %>% 
-  mutate(from = rownames(fd2.w1)) %>% 
-  gather(key = to ,value = value, -from)
-pcaordersf <- numordersf
-
-numordersf$from <- factor(numordersf$from, levels = paste0("V", 1:16))
-numordersf$to <- factor(numordersf$to, levels = rev(paste0("V", 1:16)))
-
-p1 <- ggplot(data = numordersf, aes(x = from, y = to)) + geom_tile(color = 'black', aes(fill = as.factor(value))) + 
-  scale_fill_manual(values = c('white', 'black')) +
-  scale_x_discrete(position = "top", labels = as.character(1:16)) +
-  scale_y_discrete(labels = rev(as.character(1:16)))  + 
-  ThemeNoNet + 
-  theme(legend.position = 'none') + 
-  coord_fixed()
-
-sfmsall <- read_csv("data/smallfriendsmicrosteps.csv")
-
-sfmsall %>% 
-  group_by(from, to) %>% 
-  filter(rep == 1) %>% 
-  summarise(count = n()) -> sfmsall_summ
-sfmsall_summ$from <- factor(sfmsall_summ$from, levels = paste0('V', 1:16))
-sfmsall_summ$to <- factor(sfmsall_summ$to, levels = paste0('V', 1:16))
-sfmsall_summ %>% 
-  spread(to, count, fill = 0) %>% 
-  data.frame -> allmsrep1_adjmat
-allmsrep1_adjmat$V16 <- 0
-allmsrep1_adjmat <- allmsrep1_adjmat[,-1]
-allmsrep1_adjmat <- as.matrix(allmsrep1_adjmat)
-rownames(allmsrep1_adjmat) <- paste0("V", 1:16)
-
-pca1 <- prcomp(allmsrep1_adjmat)
-order1 <- names(sort(pca1$rotation[,1], decreasing = T))
-order2 <- names(sort(abs(pca1$rotation[,1]), decreasing = T))
-
-pcaordersf$from <- factor(pcaordersf$from, levels = order1)
-pcaordersf$to <- factor(pcaordersf$to, levels = rev(order1))
-
-p2 <- ggplot(data = pcaordersf, aes(x = from, y = to)) + geom_tile(color = 'black', aes(fill = as.factor(value))) + 
-  scale_fill_manual(values = c('white', 'black')) +
-  scale_x_discrete(position = "top", labels = str_replace(order1, "V", "")) + 
-  scale_y_discrete(labels = rev(str_replace(order1, "V", ""))) + 
-  ThemeNoNet + 
-  theme(legend.position = 'none') +
-  coord_fixed()
-
-grid.arrange(p1,p2, nrow = 1)
-@
-
-Using the principal component analysis on $\mathbf{A}^{cum}$ to order the rows and columns of the adjacency matrix visualization clearly shows the two distinct connected components in the first wave of the network, which are difficult to find in the arbitrarily ordered visualization.
-
-We also attempt to better understand the microstep process by visualizing all transition probabilites for the first microstep in the process.
-We only do the first step of many for now because the \texttt{RSiena} transition probabilities after the first step are only comparable for identical steps due to the conditioning on the current network state in the model.
-Thus we have one step for each of the 1,000 repititions to visualize.
-In Figure ~\ref{histogramalters}, we present each acting node, (ego node) and the resulting probabilities of change.
-The probability shown by the bars is the theoretical probability according to the objective function of the ego node changing the tie to the alter node.
-The probability shown by the points is the empirical probability of that change being made, calculated by counting all instances of the ego node first, then computing the proprotion of each change.
-In most cases, they are almost identical, showing that the algorithm is performing as expected.
-
-<<histogramalters, fig.width = 8, fig.height=11, fig.cap="Each panel shows the theoretical (as lines) and empirical (as points) probabilities of the chosen ego node changing its tie to each of the other nodes.
-The color of the line indicates whether the tie from the ego to the alter node is being added, removed, or if there is no change to the network in this step.">>=
-ansnullchains %>% group_by(period, rep) %>% 
-  mutate(ms = row_number()) %>% 
-  filter(ms == 1, period == 1) %>% 
-  mutate(alterProb = exp(logAlterProb),
-         ego = as.numeric(as.character(ego)),
-         alter = as.numeric(as.character(alter))) %>% 
-  ungroup() %>% 
-  group_by(ego, alter) %>% 
-  summarize(count = n(), alterProb = unique(alterProb)) %>% ungroup() %>% 
-  group_by(ego) %>% 
-  mutate(csum = cumsum(alterProb)) %>% 
-  arrange(ego, csum) %>% 
-  mutate(totego = sum(count), 
-         empprob = count / totego) -> cumprobs 
-cumprobs$ego <- factor(paste0("V",cumprobs$ego+1), levels = paste0("V", 1:16))
-cumprobs$alter <- factor(paste0("V",cumprobs$alter+1), levels = paste0("V", 1:16))
-
-# is the tie that's added new, removed, or does the network stay the same
-
-changes <- cumprobs %>% dplyr::select(ego, alter) %>% unique
-
-wave1friends2 <- wave1friends[,-3]
-wave1friends2$type <- "remove"
-changes2 <- left_join(changes, wave1friends2, c("ego" = "from", "alter" = "to"))
-
-changes2$type[is.na(changes2$type)] <- ifelse(changes2$ego[is.na(changes2$type)] == changes2$alter[is.na(changes2$type)], "noChange", "add")
-
-cumprobs %>% left_join(changes2) -> changes3
-
-changes3$ego <- factor(changes3$ego, levels = paste0("V", 1:16), labels = 1:16) 
-changes3$alter <- factor(changes3$alter, levels = paste0("V", 1:16), labels = 1:16)
-
-ggplot(data = changes3) + 
-  geom_segment(
-    aes(x = alter , xend = alter, y = 0, yend = alterProb, color=type), size = 2) + 
-  geom_point(aes(x = alter, y = empprob), shape = 4, size = 2, alpha = .5) + 
-  labs(x = "alter node", y = "probability of alter node") + 
-  coord_flip() + 
-  facet_wrap(~ego, labeller = "label_both") + 
-  ThemeNoNet
-# 
-# cumprobs %>% ggplot() + geom_histogram(aes(x = alter, weight = count), color = 'black', fill = 'white', binwidth = 1) + scale_x_continuous(labels = paste0("V",1:16), breaks = 0:15) + coord_flip() + facet_wrap(~ego, labeller = "label_both") + theme_gray()
-
-# cprobslist <- split(cumprobs, cumprobs$ego)
-# plots <- list()
-# for(i in 1:length(cprobslist)){
-#   cprobslist[[i]] %>%
-#   ggplot() + 
-#   geom_histogram(aes(x = as.numeric(as.character(alter)),
-#                      y = alterProb), stat = 'identity', fill = 'white', color = 'black') + 
-#   geom_line(aes(x = as.numeric(as.character(alter)), 
-#                 y = csum)) + 
-#   geom_point(aes(x = as.numeric(as.character(alter)), 
-#                 y = csum)) +
-#   scale_x_continuous(limits = c(-.5, 15), breaks = 0:15, labels = 0:15) +   
-#   geom_hline(yintercept = 1, color = 'red', linetype = 'dashed') + 
-#   labs(x = "alter", y = expression(p[ij]), title = paste("Ego node:", names(cprobslist)[i])) + # ylab(expression(p[ij])) + 
-#   theme_gray() + 
-#     theme(plot.title = element_text(size = 10), 
-#           axis.title = element_text(size = 8), 
-#           plot.margin = unit(rep(0,4),"pt")) -> p
-#   plots[[i]] <- p
-# }
-# grid.arrange(plots[[1]], plots[[2]], plots[[3]], plots[[4]],
-#              plots[[5]], plots[[6]], plots[[7]], plots[[8]],
-#              #plots[[9]], plots[[10]], plots[[11]], plots[[12]],
-#              #plots[[13]], plots[[14]], plots[[15]], plots[[16]], 
-#              ncol = 2, padding = unit(0, 'line'))
-@
-
-Another way to view the transition probabilities is through the adjacency matrix visualization.
-In Figure~\ref{fig:heatmap}, we build on the concept of the ordered adjacency matrix of Figure~\ref{fig:adjmatorder}.
-This heatmap shows the transition probabilities of all ties that are changed in the first microstep of the 1,000 simulations.
-The heatmap is noticeably sparse, showing the lack of coverage in the model.
-Of the 256 possible steps, only 81, or about 32\%, are taken in the 1,000 simulated chains.
-XXX Check this again....
-XXX
-<<heatmap, fig.align='center', fig.cap='A heatmap showing the empirical transition probabilities for the first microstep in 1,000 simulations.
-The acting node is on the x-axis, and the receiving node is on the y-axis.
-Missing squares represent ties that did not occur.', fig.height=3>>=
-ansnullchains %>% 
-  filter(rep == 1, period == 1) %>% 
-  mutate(ego = paste0("V",as.numeric(as.character(ego))+1),
-         alter = paste0("V",as.numeric(as.character(alter))+1), 
-         prob = exp(logAlterProb)) -> probsDat
-
-ggplot(data = probsDat, 
-       aes(x = factor(ego, levels = order1), 
-             y = factor(alter, levels = rev(order1)), 
-             fill = prob)) + 
-  geom_tile(color = 'black') + 
-  scale_fill_gradient(low = 'white', high = 'blue', limits = c(0,1)) +
-  labs(x = "ego", y = "alter")+
-  scale_x_discrete(position = 'top', labels = str_replace(order1, "V", "")) +
-  scale_y_discrete(labels = rev(str_replace(order1, "V", ""))) +
-  ThemeNoNet + theme(panel.grid = element_blank()) + 
-  coord_fixed()
-@
-
-We also wanted to view the complete microstep process from the first wave, which we condition on, to the second wave.
-The number of steps taken from wave 1 to wave 2 varies.
-In the simulations from Model 1, the smallest number of steps taken was 58, and the longest was 248, with a mean of 106 and a median of 103.
-The standard deviation of the 1000 simulations' number of microsteps was 22.8.
-In Figure~\ref{fig:allsteps}, we see three simulations of the process from wave 1 to wave 2, with wave 1 shown on the left, and wave 2 shown on the right.
-In each of the three plots, the $y$-axis contains the edges sorted by how often they appear in the networks along the way.
-We can see that some edges are there in the beginning, but disappear and never come back, while others appear a few steps in, only to dispappear again.
-There are also some edges that exist in wave 2 that don't appear at all in the microstep process in a given repitition.
-
-<<allsteps, fig.cap="Three simulations (out of 1,000) of the microstep process from wave 1 to wave 2.
-The $x$ axis is the microstep number, with step 0 representing the first wave of data and the final step representing the second wave of data.
-We can see that many edges are underrepresented in this process: they are in the second wave, but never appear in the microsteps.">>=
-smallfriends <- read_csv("data/smallfriends4Geomnet.csv")
-names(smallfriends)[1:2] <- c("from", "to")
-msall1 <- netvizinf::listMicrosteps(dat = wave1friends,
-                      microsteps = filter(ansnullchainsw1w2, rep == 1))
-msall1_df <- plyr::rbind.fill(lapply(X = seq_along(msall1), FUN = function(i, x) {
-        x[[i]]$ms <- i - 1
-        return(x[[i]])
-    }, msall1))
-msall1_df <- msall1_df %>% dplyr::select(from, to, ms)
-wave2 <- smallfriends %>% filter(Wave == 2) %>% na.omit %>% data.frame()
-msall1_df <- msall1_df %>% add_row(from = paste0("V",wave2$from),
-                      to = paste0("V",wave2$to), 
-                      ms = max(msall1_df$ms)+1)
-edges <- msall1_df %>% 
-  group_by(from, to) %>%
-  summarize(edgeappears = n()) %>% 
-  arrange(desc(edgeappears)) %>% ungroup() %>%
-  mutate(eid = row_number())
-
-msall1_df2 <- left_join(msall1_df, edges, by = c("from" = "from", "to" = "to"))
-
-p1 <- ggplot(data = msall1_df2) + 
-  geom_tile(aes(x = ms, y = -eid, fill = ms)) + 
-  geom_vline(xintercept = c(0.5,81.5)) + 
-  scale_fill_continuous(low = "red", high = "blue") + 
-  ThemeNoNet + 
-  theme(legend.position = 'none', axis.ticks.y = element_blank(),
-        axis.text.y = element_blank(), axis.title = element_blank()) +
-  labs(x = "Microstep")
-
-# rep 2 
-msall2 <- netvizinf::listMicrosteps(dat = wave1friends,
-                      microsteps = filter(ansnullchainsw1w2, rep == 2))
-msall2_df <- plyr::rbind.fill(lapply(X = seq_along(msall2), FUN = function(i, x) {
-        x[[i]]$ms <- i - 1
-        return(x[[i]])
-    }, msall2))
-msall2_df <- msall2_df %>% dplyr::select(from, to, ms)
-msall2_df <- msall2_df %>% add_row(from = paste0("V",wave2$from),
-                      to = paste0("V",wave2$to), 
-                      ms = max(msall2_df$ms)+1)
-
-edges2 <- msall2_df %>% 
-  group_by(from, to) %>%
-  summarize(edgeappears = n()) %>% 
-  arrange(desc(edgeappears)) %>% ungroup() %>%
-  mutate(eid = row_number())
-
-msall2_df2 <- left_join(msall2_df, edges2, by = c("from" = "from", "to" = "to"))
-
-p2 <- ggplot(data = msall2_df2) + 
-  geom_tile(aes(x = ms, y = -eid, fill = ms)) + 
-  geom_vline(xintercept = c(0.5,49.5)) + 
-  scale_fill_continuous(low = "red", high = "blue") +
-  ThemeNoNet + 
-  theme(legend.position = 'none', axis.ticks.y = element_blank(),
-        axis.text.y = element_blank(), axis.title = element_blank()) #+
-  #labs(x = "Microstep", y = "Edges (most frequent at top)")
-
-# rep 3 
-msall3 <- netvizinf::listMicrosteps(dat = wave1friends,
-                      microsteps = filter(ansnullchainsw1w2, rep == 3))
-msall3_df <- plyr::rbind.fill(lapply(X = seq_along(msall3), FUN = function(i, x) {
-        x[[i]]$ms <- i - 1
-        return(x[[i]])
-    }, msall3))
-msall3_df <- msall3_df %>% dplyr::select(from, to, ms)
-msall3_df <- msall3_df %>% add_row(from = paste0("V",wave2$from),
-                      to = paste0("V",wave2$to), 
-                      ms = max(msall3_df$ms)+1)
-
-edges3 <- msall3_df %>% 
-  group_by(from, to) %>%
-  summarize(edgeappears = n()) %>% 
-  arrange(desc(edgeappears)) %>% ungroup() %>%
-  mutate(eid = row_number())
-
-msall3_df2 <- left_join(msall3_df, edges3, by = c("from" = "from", "to" = "to"))
-
-p3 <- ggplot(data = msall3_df2) + 
-  geom_tile(aes(x = ms, y = -eid, fill = ms)) + 
-  geom_vline(xintercept = c(0.5,106.5)) + 
-  scale_fill_continuous(low = "red", high = "blue") +
-  ThemeNoNet + 
-  theme(legend.position = 'none', axis.ticks.y = element_blank(),
-        axis.text.y = element_blank(), axis.title = element_blank()) #+
-  #labs(x = "Microstep", y = "Edges (most frequent at top)")
-
-grid.arrange(p1, p2, p3, nrow = 3)
-@
-
-\st{The algorithms here are many!
-\begin{enumerate}
-%\item First, the microsteps / CTMC between each time point (gifs, other animations)
-\item Phases of RSiena: phase 1 = initial parameter estimate
-\item phase 2 = iterative update of parameters - trace plots
-\item phase 3 = simulations for convergence diagnostics
-\end{enumerate}
- }
-
-\hh{The visualizations should address most elements of structure described before, i.e.
-follow along in the setup (for descriptive and teaching purposes), the fitting (understanding the model and interpreting the results) and then diagnostics (how well does the model fit the data - where are the most differences to the actual data/what are the sensitive parameters?) }
-
-\hh{Some of the visualizations we talked about
-\begin{enumerate}
-\item visualization of the MCMC process:  movie of individual actors' step-by-step decisions; could be done with the ndtv package, as long as we can get individual steps out of RSiena
-\item global picture: 1000 simulated networks + alpha blending should give a pretty good idea of the most important edges
-%\item global structure + principal components (tensor pca)
-\end{enumerate}}
+For comparison, we also fit models $M1$, $M2$, and $M3$ to the senate collaboration data.
+In addition, we fit two more models, $M4$ and $M5$ to the senate data that were found to be significant according to the Wald-type test:
+
+\begin{align*}
+f_i(x)^{M4} & = \beta_1s_{i1} + \beta_2s_{i2} + \beta_5s_{i5} \\
+f_i(x, z)^{M5} & = \beta_1s_{i1} + \beta_2s_{i2} + \beta_6s_{i6}.
+\end{align*}
+
+In $M4$, $s_{i5} = \mathbb{I}(x_{i+} = x_{+i} = 0)$ is an indicator variable for isolated nodes, called the network-isolate effect.
+This parameter attempts to measure the effect of a node being completely isolated in the network.
+If the value of the parameter $\beta_5$ is high, nodes are encouraged to remain isolated.
+In $M5$,  $s_{i6} = \sum\limits_j x_{ij}x_{ji}\mathbb{I}(z_i = z_j)$ is a measure of reciprocity between senators in the same party, $z_i, z_j \in \{\text{Democrat, Independent, Republican}\}$, generally called the same covariate reciprocity effect.
+This parameter attempts to measure the propensity for forming reciprocal ties between nodes with the same covariate value.
+If the value of the parameter $\beta_6$ is high, nodes belonging to the same covariate group are encourage to reciprocate incoming ties.
+By exploring these models more visually, we hope to understand these effects and the model fitting process better.
 
@@
 
+\subsection{Explore algorithms, not just end result}
+
+The model fitting process in \texttt{RSiena} involves several simulation steps that are hidden from the user.
+This is practical and efficient if a researcher is primarily interested in fitting one model to a set of longitudinal network data and obtaining parameter estimates.
+We are more interested in how the models are fit, so we extracted and explored the different steps of that process.
+
+A key component of each step of the SIENA method of moments algorithm is the ``microstep." This step simulates from the model in its current state, $x(t_m)$ with current parameter values $\theta_{0}$, to the next state, $x(t_{m+1})$.
+This microstep process stops when the simulated network has acheived the same number of differences, $C$, from $x(t_m)$ as $x(t_{m+1})$, where
+$$C = \sum_{i \neq j} |x_{ij}(t_{m+1}) - x_{ij}(t_{m})|.$$ 
+This simulation process follows the steps of the continuous-time Markov chain.
+First, an ``ego node'' is selected to make a change, and then it randomly makes one change in its ties according to the probability, $p_{ij}$, determined by its objective function.
+The options for change are (1) removing a current tie, (2) adding a new tie, or (3) making no change at all.
+Saving all of these steps is not efficient if you're only interested in estimating parameter values, but they can be saved and extracted using options in \texttt{RSiena}, which is exactly what we did to create our visualizations.
+Between two network observations $x(t_m)$ and $x(t_{m+1})$, there can be dozens, hundreds, or even thousands of microsteps, depending on the size of the network.
+We wanted to see these in-between steps in order to better understand the behavior of the underlying continuous-time Markov chain.
+
+The first vizualization we present here is an animation of microsteps simulated to emulate the transition steps from wave 1 to wave 2 the subset of the friends data shown in Figure ~\ref{fig:alldataviz}.
+Movies similar to our animation were used to visualize the changes of dynamic networks in \citet{dynnetviz}.
+When each ego node is selected in a microstep, it is emphasized in the animation, then the associated edge either appears or disappears.
+If there are no changes at a particular microstep, no changes are seen.
+Some frames of the animations are shown in Figure~\ref{fig:makinggif}, and the full movie can be viewed at INSERT LINK TO GIF.
+
+<<makinggif, fig.cap='A selection of images in the microstep animation.
+The selected edges and nodes are emphasized by changing the size and the color, then when edges are removed, they fade out, shrinking in size, while the nodes change color and shrink to blend in with the other nodes.', fig.width=8, cache=TRUE, message = FALSE, warning = FALSE>>=
+# Need old version of dplyr to get this to work: 
+if (!require("dplyr050",character.only = TRUE)) {
+      devtools::install_github("sctyner/dplyr050")
+}
+
+friend.data.w1 <- as.matrix(read.table("data/s50_data/s50-network1.dat"))
+friend.data.w2 <- as.matrix(read.table("data/s50_data/s50-network2.dat"))
+friend.data.w3 <- as.matrix(read.table("data/s50_data/s50-network3.dat"))
+drink <- as.matrix(read.table("data/s50_data/s50-alcohol.dat"))
+fd2.w1 <- friend.data.w1[20:35,20:35]
+fd2.w2 <- friend.data.w2[20:35,20:35]
+fd2.w3 <- friend.data.w3[20:35,20:35]
+friendshipData <- array(c(fd2.w1, fd2.w2,fd2.w3), dim = c(16, 16, 3))
+
+# running Siena
+#friendship <- sienaDependent(friendshipData)
+#alcohol <- varCovar(drink[20:35,])
+#mydata <- sienaDataCreate(friendship, alcohol)
+#myeffnull <- getEffects(mydata)
+#myalgorithm <- sienaAlgorithmCreate(projname = 's16_3')
+set.seed(823746)
+
+ansnullchains <- get_chain_info(ansnull)
+ansnullchains %>%
+  filter(period == 1) %>%  #only look at chains from wave 1 to wave 2
+  group_by(rep) %>%
+  dplyr::select(rep, from = ego, to = alter) %>%
+  mutate(val = as.numeric(!from == to),
+         from = paste0("V", parse_number(from)+1), # make the chains
+         to = paste0("V", parse_number(to)+1)) -> ansnullchainsw1w2
+colnames(fd2.w1) <- paste0("V", 1:16)
+rownames(fd2.w1) <- paste0("V", 1:16)
+wave1friends <- fortify(as.adjmat(fd2.w1))
+ms1 <- listMicrosteps(dat = wave1friends,
+                      microsteps = filter(ansnullchainsw1w2, rep == 1))
+microsteps <- ms1[1:20]
+
+microsteps_df <- getMicrostepsDF(microsteps)
+pte <- pretween_edges(microsteps = microsteps)
+ptv <- pretween_vertices(pte = pte, layoutparams = list(n = 16))
+# step 1 is remove, step 9 is add. 
+# get tweened data for all microsteps
+pall <- tween_microsteps(ptv, pte, microsteps_df)
+
+static_pall <- filter(pall, .frame %in% c(15,20,22,24,26:28, 71:73, 75, 79, 81, 84))
+
+static_pall$type <- NA
+static_pall$type[which(static_pall$.frame %in% c(15,20,22,24,26:28))] <- "Remove Edge"
+static_pall$type[which(static_pall$.frame %in% c(71:73, 75, 79, 81, 84))] <- "Add Edge"
+
+set.seed(34569234)
+create_net_animate(static_pall) + 
+  facet_wrap(~.frame, nrow=2, labeller = 'label_both') + 
+  ThemeNet + 
+  theme(panel.background = element_rect(color = 'black'))
+@
+
+The top row of Figure~\ref{fig:makinggif} shows an edge being removed, and the bottom row shows one being added.
+In both cases, the ego nodes chosen to act change color from black to red, and they also increase greatly in size.
+In the case of an edge being removed, the edge that currently exists is emphasized with the color and size change that the node also gets, and as the animation proceeds the edge shrinks to disappearing, while the node shrinks and changes color back to the original black.
+If the edge is being added, it appears colored red from nothing and grows to a large size, then changes color to match the rest of the edges, while the node shrinks and changes color to match the other nodes.
+
+We also use animation to view the changing structure of the adjacency matrix the microsteps.
+The adjacency matrices for the three waves of friendship data as shown in Figure~\ref{fig:alldataviz} are ordered by node id.
+There are 16 nodes in the data, numbered 1-16, and that order is used on the $x$ and $y$ axes for the matrix visualization.
+Viewing the adjacency matrices with this arbitrary ordering does not provide much information to the viewer about the underlying structure of the network.
+This lack of perceived structure would be exacerbated in an animation.
+So, we adjust the ordering so that the viewer can perceive more of the structure of the graph.
+This process is known as matrix seriation \citep{seriation}.
+
+To reorder the vertices for the matrix visualization, we first constructed a cumulative adjaceny matrix, $\mathbf{A}^{cum}$, for the series of microsteps simulating the network from $x(t_m)$ to $x(t_{m+1})$.
+A single entry in the cumulative adjacency matrix, $\mathbf{A}^{cum}_{ij}$, is the total number of times the edge from node $i$ to node $j$ appears in the network from the initial observation, $x(t_m) \equiv X(0)$ to the final reult of the last microstep, $X(C)$:
+$$ \mathbf{A}^{cum}_{ij} = \sum_{c = 0}^C X_{ij}(c). $$
+We then performed a principal component analysis on $\mathbf{A}^{cum}$, and used the values of the first principal component to order the vertices on the $x$ and $y$ axes for the adjacency matrix animation.
+For one such series of microsteps simulated by \texttt{RSiena}, we present the adjacency matrix ordered by the (arbitrary) vertex id alongside the seriated adjacency microsteps using the first principal component loading on the cumulative adjacency matrix, $\mathbf{A}^{cum}$, in Figure~\ref{fig:adjmatorder}.
+
+<<adjmatorder, message = FALSE, warning = FALSE, fig.cap="On the left, the starting friendship network represented in adjacency matrix form, ordered by vertex id.
+On the right, the same adjacency matrix is presented after ordering the vertices by one repitition of the microstep simulation process from wave one to wave two.", fig.width=8, fig.height=3>>=
+# first present the original wave 1 adj mat, then present the adj mat after being ordered by the 1st pc of the big adjmat A. 
+# reinstall new dplyr
+numordersf <- data.frame(fd2.w1) %>% 
+  mutate(from = rownames(fd2.w1)) %>% 
+  gather(key = to ,value = value, -from)
+pcaordersf <- numordersf
+
+numordersf$from <- factor(numordersf$from, levels = paste0("V", 1:16))
+numordersf$to <- factor(numordersf$to, levels = rev(paste0("V", 1:16)))
+
+p1 <- ggplot(data = numordersf, aes(x = from, y = to)) + geom_tile(color = 'black', aes(fill = as.factor(value))) + 
+  scale_fill_manual(values = c('white', 'black')) +
+  scale_x_discrete(position = "top", labels = as.character(1:16)) +
+  scale_y_discrete(labels = rev(as.character(1:16)))  + 
+  ThemeNoNet + 
+  theme(legend.position = 'none') + 
+  coord_fixed()
+
+sfmsall <- read_csv("data/smallfriendsmicrosteps.csv")
+
+sfmsall %>% 
+  group_by(from, to) %>% 
+  filter(rep == 1) %>% 
+  summarise(count = n()) -> sfmsall_summ
+sfmsall_summ$from <- factor(sfmsall_summ$from, levels = paste0('V', 1:16))
+sfmsall_summ$to <- factor(sfmsall_summ$to, levels = paste0('V', 1:16))
+sfmsall_summ %>% 
+  spread(to, count, fill = 0) %>% 
+  data.frame -> allmsrep1_adjmat
+allmsrep1_adjmat$V16 <- 0
+allmsrep1_adjmat <- allmsrep1_adjmat[,-1]
+allmsrep1_adjmat <- as.matrix(allmsrep1_adjmat)
+rownames(allmsrep1_adjmat) <- paste0("V", 1:16)
+
+pca1 <- prcomp(allmsrep1_adjmat)
+order1 <- names(sort(pca1$rotation[,1], decreasing = T))
+order2 <- names(sort(abs(pca1$rotation[,1]), decreasing = T))
+
+pcaordersf$from <- factor(pcaordersf$from, levels = order1)
+pcaordersf$to <- factor(pcaordersf$to, levels = rev(order1))
+
+p2 <- ggplot(data = pcaordersf, aes(x = from, y = to)) + geom_tile(color = 'black', aes(fill = as.factor(value))) + 
+  scale_fill_manual(values = c('white', 'black')) +
+  scale_x_discrete(position = "top", labels = str_replace(order1, "V", "")) + 
+  scale_y_discrete(labels = rev(str_replace(order1, "V", ""))) + 
+  ThemeNoNet + 
+  theme(legend.position = 'none') +
+  coord_fixed()
+
+grid.arrange(p1,p2, nrow = 1)
+@
+
+Using the principal component analysis on $\mathbf{A}^{cum}$ to order the rows and columns of the adjacency matrix visualization clearly shows the two distinct connected components in the first wave of the network, which are difficult to find in the arbitrarily ordered visualization.
+
+We also attempt to better understand the microstep process by visualizing all transition probabilites for the first microstep in the process.
+We only do the first step of many for now because the \texttt{RSiena} transition probabilities after the first step are only comparable for identical steps due to the conditioning on the current network state in the model.
+Thus we have one step for each of the 1,000 repititions to visualize.
+In Figure ~\ref{histogramalters}, we present each acting node, (ego node) and the resulting probabilities of change.
+The probability shown by the bars is the theoretical probability according to the objective function of the ego node changing the tie to the alter node.
+The probability shown by the points is the empirical probability of that change being made, calculated by counting all instances of the ego node first, then computing the proprotion of each change.
+In most cases, they are almost identical, showing that the algorithm is performing as expected.
+
+<<histogramalters, fig.width = 8, fig.height=11, fig.cap="Each panel shows the theoretical (as lines) and empirical (as points) probabilities of the chosen ego node changing its tie to each of the other nodes.
+The color of the line indicates whether the tie from the ego to the alter node is being added, removed, or if there is no change to the network in this step.">>=
+ansnullchains %>% group_by(period, rep) %>% 
+  mutate(ms = row_number()) %>% 
+  filter(ms == 1, period == 1) %>% 
+  mutate(alterProb = exp(logAlterProb),
+         ego = as.numeric(as.character(ego)),
+         alter = as.numeric(as.character(alter))) %>% 
+  ungroup() %>% 
+  group_by(ego, alter) %>% 
+  summarize(count = n(), alterProb = unique(alterProb)) %>% ungroup() %>% 
+  group_by(ego) %>% 
+  mutate(csum = cumsum(alterProb)) %>% 
+  arrange(ego, csum) %>% 
+  mutate(totego = sum(count), 
+         empprob = count / totego) -> cumprobs 
+cumprobs$ego <- factor(paste0("V",cumprobs$ego+1), levels = paste0("V", 1:16))
+cumprobs$alter <- factor(paste0("V",cumprobs$alter+1), levels = paste0("V", 1:16))
+
+# is the tie that's added new, removed, or does the network stay the same
+
+changes <- cumprobs %>% dplyr::select(ego, alter) %>% unique
+
+wave1friends2 <- wave1friends[,-3]
+wave1friends2$type <- "remove"
+changes2 <- left_join(changes, wave1friends2, c("ego" = "from", "alter" = "to"))
+
+changes2$type[is.na(changes2$type)] <- ifelse(changes2$ego[is.na(changes2$type)] == changes2$alter[is.na(changes2$type)], "noChange", "add")
+
+cumprobs %>% left_join(changes2) -> changes3
+
+changes3$ego <- factor(changes3$ego, levels = paste0("V", 1:16), labels = 1:16) 
+changes3$alter <- factor(changes3$alter, levels = paste0("V", 1:16), labels = 1:16)
+
+ggplot(data = changes3) + 
+  geom_segment(
+    aes(x = alter , xend = alter, y = 0, yend = alterProb, color=type), size = 2) + 
+  geom_point(aes(x = alter, y = empprob), shape = 4, size = 2, alpha = .5) + 
+  labs(x = "alter node", y = "probability of alter node") + 
+  coord_flip() + 
+  facet_wrap(~ego, labeller = "label_both") + 
+  ThemeNoNet
+# 
+# cumprobs %>% ggplot() + geom_histogram(aes(x = alter, weight = count), color = 'black', fill = 'white', binwidth = 1) + scale_x_continuous(labels = paste0("V",1:16), breaks = 0:15) + coord_flip() + facet_wrap(~ego, labeller = "label_both") + theme_gray()
+
+# cprobslist <- split(cumprobs, cumprobs$ego)
+# plots <- list()
+# for(i in 1:length(cprobslist)){
+#   cprobslist[[i]] %>%
+#   ggplot() + 
+#   geom_histogram(aes(x = as.numeric(as.character(alter)),
+#                      y = alterProb), stat = 'identity', fill = 'white', color = 'black') + 
+#   geom_line(aes(x = as.numeric(as.character(alter)), 
+#                 y = csum)) + 
+#   geom_point(aes(x = as.numeric(as.character(alter)), 
+#                 y = csum)) +
+#   scale_x_continuous(limits = c(-.5, 15), breaks = 0:15, labels = 0:15) +   
+#   geom_hline(yintercept = 1, color = 'red', linetype = 'dashed') + 
+#   labs(x = "alter", y = expression(p[ij]), title = paste("Ego node:", names(cprobslist)[i])) + # ylab(expression(p[ij])) + 
+#   theme_gray() + 
+#     theme(plot.title = element_text(size = 10), 
+#           axis.title = element_text(size = 8), 
+#           plot.margin = unit(rep(0,4),"pt")) -> p
+#   plots[[i]] <- p
+# }
+# grid.arrange(plots[[1]], plots[[2]], plots[[3]], plots[[4]],
+#              plots[[5]], plots[[6]], plots[[7]], plots[[8]],
+#              #plots[[9]], plots[[10]], plots[[11]], plots[[12]],
+#              #plots[[13]], plots[[14]], plots[[15]], plots[[16]], 
+#              ncol = 2, padding = unit(0, 'line'))
+@
+
+Another way to view the transition probabilities is through the adjacency matrix visualization.
+In Figure~\ref{fig:heatmap}, we build on the concept of the ordered adjacency matrix of Figure~\ref{fig:adjmatorder}.
+This heatmap shows the transition probabilities of all ties that are changed in the first microstep of the 1,000 simulations.
+The heatmap is noticeably sparse, showing the lack of coverage in the model.
+Of the 256 possible steps, only 81, or about 32\%, are taken in the 1,000 simulated chains.
+XXX Check this again....
+XXX
+<<heatmap, fig.align='center', fig.cap='A heatmap showing the empirical transition probabilities for the first microstep in 1,000 simulations.
+The acting node is on the x-axis, and the receiving node is on the y-axis.
+Missing squares represent ties that did not occur.', fig.height=3>>=
+ansnullchains %>% 
+  filter(rep == 1, period == 1) %>% 
+  mutate(ego = paste0("V",as.numeric(as.character(ego))+1),
+         alter = paste0("V",as.numeric(as.character(alter))+1), 
+         prob = exp(logAlterProb)) -> probsDat
+
+ggplot(data = probsDat, 
+       aes(x = factor(ego, levels = order1), 
+             y = factor(alter, levels = rev(order1)), 
+             fill = prob)) + 
+  geom_tile(color = 'black') + 
+  scale_fill_gradient(low = 'white', high = 'blue', limits = c(0,1)) +
+  labs(x = "ego", y = "alter")+
+  scale_x_discrete(position = 'top', labels = str_replace(order1, "V", "")) +
+  scale_y_discrete(labels = rev(str_replace(order1, "V", ""))) +
+  ThemeNoNet + theme(panel.grid = element_blank()) + 
+  coord_fixed()
+@
+
+We also wanted to view the complete microstep process from the first wave, which we condition on, to the second wave.
+The number of steps taken from wave 1 to wave 2 varies.
+In the simulations from Model 1, the smallest number of steps taken was 58, and the longest was 248, with a mean of 106 and a median of 103.
+The standard deviation of the 1000 simulations' number of microsteps was 22.8.
+In Figure~\ref{fig:allsteps}, we see three simulations of the process from wave 1 to wave 2, with wave 1 shown on the left, and wave 2 shown on the right.
+In each of the three plots, the $y$-axis contains the edges sorted by how often they appear in the networks along the way.
+We can see that some edges are there in the beginning, but disappear and never come back, while others appear a few steps in, only to dispappear again.
+There are also some edges that exist in wave 2 that don't appear at all in the microstep process in a given repitition.
+
+<<allsteps, fig.cap="Three simulations (out of 1,000) of the microstep process from wave 1 to wave 2.
+The $x$ axis is the microstep number, with step 0 representing the first wave of data and the final step representing the second wave of data.
+We can see that many edges are underrepresented in this process: they are in the second wave, but never appear in the microsteps.">>=
+smallfriends <- read_csv("data/smallfriends4Geomnet.csv")
+names(smallfriends)[1:2] <- c("from", "to")
+msall1 <- netvizinf::listMicrosteps(dat = wave1friends,
+                      microsteps = filter(ansnullchainsw1w2, rep == 1))
+msall1_df <- plyr::rbind.fill(lapply(X = seq_along(msall1), FUN = function(i, x) {
+        x[[i]]$ms <- i - 1
+        return(x[[i]])
+    }, msall1))
+msall1_df <- msall1_df %>% dplyr::select(from, to, ms)
+wave2 <- smallfriends %>% filter(Wave == 2) %>% na.omit %>% data.frame()
+msall1_df <- msall1_df %>% add_row(from = paste0("V",wave2$from),
+                      to = paste0("V",wave2$to), 
+                      ms = max(msall1_df$ms)+1)
+edges <- msall1_df %>% 
+  group_by(from, to) %>%
+  summarize(edgeappears = n()) %>% 
+  arrange(desc(edgeappears)) %>% ungroup() %>%
+  mutate(eid = row_number())
+
+msall1_df2 <- left_join(msall1_df, edges, by = c("from" = "from", "to" = "to"))
+
+p1 <- ggplot(data = msall1_df2) + 
+  geom_tile(aes(x = ms, y = -eid, fill = ms)) + 
+  geom_vline(xintercept = c(0.5,81.5)) + 
+  scale_fill_continuous(low = "red", high = "blue") + 
+  ThemeNoNet + 
+  theme(legend.position = 'none', axis.ticks.y = element_blank(),
+        axis.text.y = element_blank(), axis.title = element_blank()) +
+  labs(x = "Microstep")
+
+# rep 2 
+msall2 <- netvizinf::listMicrosteps(dat = wave1friends,
+                      microsteps = filter(ansnullchainsw1w2, rep == 2))
+msall2_df <- plyr::rbind.fill(lapply(X = seq_along(msall2), FUN = function(i, x) {
+        x[[i]]$ms <- i - 1
+        return(x[[i]])
+    }, msall2))
+msall2_df <- msall2_df %>% dplyr::select(from, to, ms)
+msall2_df <- msall2_df %>% add_row(from = paste0("V",wave2$from),
+                      to = paste0("V",wave2$to), 
+                      ms = max(msall2_df$ms)+1)
+
+edges2 <- msall2_df %>% 
+  group_by(from, to) %>%
+  summarize(edgeappears = n()) %>% 
+  arrange(desc(edgeappears)) %>% ungroup() %>%
+  mutate(eid = row_number())
+
+msall2_df2 <- left_join(msall2_df, edges2, by = c("from" = "from", "to" = "to"))
+
+p2 <- ggplot(data = msall2_df2) + 
+  geom_tile(aes(x = ms, y = -eid, fill = ms)) + 
+  geom_vline(xintercept = c(0.5,49.5)) + 
+  scale_fill_continuous(low = "red", high = "blue") +
+  ThemeNoNet + 
+  theme(legend.position = 'none', axis.ticks.y = element_blank(),
+        axis.text.y = element_blank(), axis.title = element_blank()) #+
+  #labs(x = "Microstep", y = "Edges (most frequent at top)")
+
+# rep 3 
+msall3 <- netvizinf::listMicrosteps(dat = wave1friends,
+                      microsteps = filter(ansnullchainsw1w2, rep == 3))
+msall3_df <- plyr::rbind.fill(lapply(X = seq_along(msall3), FUN = function(i, x) {
+        x[[i]]$ms <- i - 1
+        return(x[[i]])
+    }, msall3))
+msall3_df <- msall3_df %>% dplyr::select(from, to, ms)
+msall3_df <- msall3_df %>% add_row(from = paste0("V",wave2$from),
+                      to = paste0("V",wave2$to), 
+                      ms = max(msall3_df$ms)+1)
+
+edges3 <- msall3_df %>% 
+  group_by(from, to) %>%
+  summarize(edgeappears = n()) %>% 
+  arrange(desc(edgeappears)) %>% ungroup() %>%
+  mutate(eid = row_number())
+
+msall3_df2 <- left_join(msall3_df, edges3, by = c("from" = "from", "to" = "to"))
+
+p3 <- ggplot(data = msall3_df2) + 
+  geom_tile(aes(x = ms, y = -eid, fill = ms)) + 
+  geom_vline(xintercept = c(0.5,106.5)) + 
+  scale_fill_continuous(low = "red", high = "blue") +
+  ThemeNoNet + 
+  theme(legend.position = 'none', axis.ticks.y = element_blank(),
+        axis.text.y = element_blank(), axis.title = element_blank()) #+
+  #labs(x = "Microstep", y = "Edges (most frequent at top)")
+
+grid.arrange(p1, p2, p3, nrow = 3)
+@
+
+\st{The algorithms here are many!
+\begin{enumerate}
+%\item First, the microsteps / CTMC between each time point (gifs, other animations)
+\item Phases of RSiena: phase 1 = initial parameter estimate
+\item phase 2 = iterative update of parameters - trace plots
+\item phase 3 = simulations for convergence diagnostics
+\end{enumerate}
+ }
+
+\hh{The visualizations should address most elements of structure described before, i.e.
+follow along in the setup (for descriptive and teaching purposes), the fitting (understanding the model and interpreting the results) and then diagnostics (how well does the model fit the data - where are the most differences to the actual data/what are the sensitive parameters?) }
+
+\hh{Some of the visualizations we talked about
+\begin{enumerate}
+\item visualization of the MCMC process:  movie of individual actors' step-by-step decisions; could be done with the ndtv package, as long as we can get individual steps out of RSiena
+\item global picture: 1000 simulated networks + alpha blending should give a pretty good idea of the most important edges
+%\item global structure + principal components (tensor pca)
+\end{enumerate}}
+
 \section{Discussion} \label{sec:discussion}

### 2017-08-20 paper.Rnw
add a lineup, through new section 4.1.

@@
 <<senateplot, fig.cap='The collaboration network in the four senates during the Obama years, 2009-2016.
-Edges are shown only if the weighted propensity to cosponsor from one senator to another is greater than 0.25.', fig.height=3>>=
+Edges are shown only if the weighted propensity to cosponsor from one senator to another is greater than 0.25.
+We use the Fruchterman-Reingold algorithm to layout the node-link diagram.', fig.height=3>>=
 seobama <- read_csv("data/congress/senateobamapres_gsw_25.csv")
 ggplot(data = seobama) + 
-  geom_net(directed = T, labelon=F, arrowsize = .25, singletons= F, fiteach = T, linewidth = .25,
+  geom_net(directed = T, labelon=F, arrowsize = .25, singletons= F, fiteach = T, linewidth = .25, layout.alg = 'fruchtermanreingold',
            aes(from_id = source, to_id = target, color = party),size = 1) + 
@@
 
+XXX TO DO - working on model estimates tonight XXX
+
 In $M4$, $s_{i5} = \mathbb{I}(x_{i+} = x_{+i} = 0)$ is an indicator variable for isolated nodes, called the network-isolate effect.
@@
 
+The first way we hope to better understand stochastic actor-oriented models is by viewing the model(s) in the data space.
 In \citeauthor{modelvis}, they define the \textit{data space} as ``the region over which we can reliably make inferences, usually a hypercube containing the data" \citep[p.~206]{modelvis}.
-What does this definition mean for network data?
-For dynamic social networks, there are a few different data ``spaces." First, we have the actors and their corresponding covariates.
-Second, we have the dyad variables that describe the relationships between the variables.
-Finally, we have the time over which the network is allowed to evolve.
+But what does this definition mean for network data?
+For dynamic social networks, there are a few different data ``spaces":
+\begin{enumerate}
+\item the actors and their corresponding covariates,
+\item the edges and their correpsonding variables that describe the relationships between the nodes, and 
+\item the time over which the network evolves
+\end{enumerate}
 These three data pieces can be visualized together in various ways.
-The traditional node-link visualization uses one of many algorithms to layout the actors as points in 2D space, then draws segments connecting the points in 2D if the dyad between two nodes has value greater than zero, and draws nothing otherwise.
+The traditional node-link visualization uses one of many algorithms to layout the actors as points in 2D space, then draws segments connecting the points in 2D if there is an edge between two nodes, and draws nothing otherwise.
 The time aspect can be visualized by drawing each network observation in time and placing the observed timepoints side-by-side.
@@
 Different aesthesic aspects of each piece of the node-link diagram can be tied to the underlying node or edge data.
-The color, size, and shape of the points can be used to represent variables in the node data, and the color, size, and linetype of the links between points can be used to represent the edge variables.
-In a social network, node data might be age, gender, and occupation of the person in the network, and node data might be length of connection between two people, how the people first met (school, work, church, etc.), and how often they interact.
+The color, size, and shape of the points can be used to represent variables in the node data, while the color, linewidth, and linetype of the lines between points can be used to represent the edge variables.
+In a social network, node data might be age, gender, and occupation of the person in the network, and edge data might be length of connection between two people, how the people first met (school, work, church, etc.), and how often they interact, and we can view the network at different timepoints side-by-side to see its evolution.
 Pulling all of this information together with \texttt{geomnet} allows the entire data space to be viewed at once.
 
-XXX INSERT AN EXAMPLE OF DATA WITH LOTS OF EDGE, NODE VARS OF INTEREST XXX
-
-These visual connections between the node, edge, and time data spaces are themselves part of the data space, because they can be viewed in different ways.
+<<funsenatevis, fig.height=3, fig.cap='The 111th Senate separated into ``BC" and ``AC": before Clinton left the Senate to become Secretary of State (on the right), and after she left (on the left).' >>=
+se111clint <- filter(seobama, senate == 111)
+se111noclint <- filter(seobama, senate == 111)
+se111noclint$target[which(se111noclint$target == "Hillary Rodham Clinton")] <- NA
+se111clint$Clinton <- 'Yes'
+se111noclint$Clinton <- 'No'
+clintonSenate <- rbind(se111clint, se111noclint)
+
+ggplot(data = clintonSenate) + 
+  geom_net(directed = T, labelon=F, arrowsize = .5, singletons= F, fiteach = T, arrowgap = .01, layout.alg = 'fruchtermanreingold',
+           aes(from_id = source, to_id = target, color = party, linewidth = gsw, shape = sex,size = n_au)) + 
+  #theme_net() + 
+  ThemeNet + 
+  scale_color_manual(values = c("royalblue", "forestgreen","firebrick")) + 
+  theme(legend.position = 'bottom') + 
+  facet_wrap(~Clinton, nrow = 1, labeller = 'label_both')
+@
+
+To demonstrate, we use \texttt{geomnet} to visualize the connections in the $111^{th}$ at two different points: when Hillary Clinton was in the senate, and after she left to become Secretary of State.
+Clinton was only in the $111^{th}$ senate for 17 days, from January 3, 2009, to January 20, 2009, in the middle of her second term in the senate.
+In that time, she authored two bills and was a cosponsor on 17 other bills.
+With Clinton included in the node-link diagram, the senate looks much more collaborative than it does without her in the diagram.
+We can compare the number of bills authored throughout the senate by mapping the size of the node to the numer of bills authored by that senator.
+We also map shape to sex, and keep color mapped to party of the senator.
+In addition, we can see the strength of the tie by mapping the linewidth of the edge to the $WPC$ value between the two senators.
+% ST: I don't like this paragraph
+% The visual aspects of the node-link diagram connecting the node, edge, and time data spaces are themselves part of the data space, because they can be viewed in different ways.
 Another way to connect the node and the edge data, as discussed in Section~\ref{sec:netvizintro}, is through the network's adjaceny matrix.
 The time dimension can also be incorporated with side-by-side adjacency matrix visualizations.
-
-Another way to view the model in the data space is through simulation from the model.
-Since no single simulation from a model is going to look like the data, it's better to visualize many simulations together.
-For network data, one way to do this is through the traditional node-link diagram.
-In Figure~\ref{fig:summnetM1}, we show a summary network created with 1,000 simulations of the second wave of the network from Model 1.
-Only edges that appear in more than 5\% of the simulations (in at least 51 of the 1000 simulations) are included in the visualization, with edges that appear more frequently emphasized by thicker linewidths and a brighter red color.
-On either side of the summary network, we show the actual data, wave 1 on the left, and wave 2 on the right.
-\st{We can see that the structure of the summary network is much more similar to the first network observation than to the second network observation.
-The simulations, however, are supposed to represent the second wave of data, which is shown on the right in Figure~\ref{fig:summnetM1}.
-This is an indication that our simple model, $M1$ is doing a very poor job of capturing the change mechanism from the first to the second wave of observation.} % XXX ST - I was looking at the wrong one.
-I had wave == 2 instead of wave == 1 in the code.
-This means I was looking at the wrong simulations, so they really do look more like the first data obs than the second obs.
-XXX Comparing the three visualizations side-by-side shows us that the summary network of simulations of wave 2 looks more like the true wave 2 than wave 1, which is a good indication that the model is capturing the true data structure.
-It's not ideal, however, because there appear to be too many reciprocated ties in the summary network, far more than there are in the data.
-This could indicate that we should remove the reciprocity parameter from the model.
 
@@
 
-\st{Because of the multi-level data structure of a network, viewing the model in the data space can depend on which aspect of the model you are interested in viewing.
-XXX Ideas: summary statistic dist.
-on nets simulated from model and show statistic from data; interactive plots showing objective function values on nodes; select node, select other node, show how the function values change interactively (future); goodness of fit should do here.
-XXX}
+Another way to view the model in the data space is through simulation from the model.
+No single network alone simulated from a SAOM is going to look like the data or represent the model, just as no single value simulated from the standard normal distribution will look like a bell-shaped curve.
+It would therefore better to visualize many simulations together.
+For network data, one way we propose to do this is through a "summary" or "average" network drawn using the traditional node-link diagram.
+In Figure~\ref{fig:summnetM1}, we show a summary network created with 1,000 simulations of the second wave of the network from Model 1.
+\st{To make this summary network, we first simulated 1,000 wave 2 and wave 3 observations of our small friendship example data from model $M1$, for which parameters had previously been estimated.
+We then combine the 1,000 instances of wave 2, and count up the number of times each edge appears in a simulation.
+Then, we combine these 1,000 networks into a single network with edgeweight equal to the proportion of time that edge appears in our 1,000 simulations.
+This is the network we draw using the node-link diagram.} An edge is only drawn in the summary network if it appears in more than 5\% of the simulations (in at least 51 of the 1000 simulations), with edges that appear more frequently emphasized by thicker linewidths and a darker color.
+On either side of the summary network in Figure\ref{fig:summnetM1}, we show the actual data, wave 1 on the left, and wave 2 on the right.
+\st{We can see that the structure of the summary network is much more similar to the first network observation than to the second network observation.
+The simulations, however, are supposed to represent the second wave of data, which is shown on the right in Figure~\ref{fig:summnetM1}.
+This is an indication that our simple model, $M1$ is doing a very poor job of capturing the change mechanism from the first to the second wave of observation.} % XXX ST - I was looking at the wrong one.
+I had wave == 2 instead of wave == 1 in the code.
+This means I was looking at the wrong simulations, so they really do look more like the first data obs than the second obs.
+XXX Comparing the three visualizations side-by-side shows us that the summary network of simulations of wave 2 looks more like the true wave 2 than wave 1, which is a good indication that the model is capturing the true data structure.
+It's not ideal, however, because there appear to be too many reciprocated ties in the summary network, far more than there are in the data.
+This could indicate that we should remove the reciprocity parameter from the model.
+
+<<gof2, fig.cap="An example of viewing the ">>=
+qplot(x = log(gof1$`Period 1`$SimulatedTestStat), binwidth = .2) + 
+  geom_vline(xintercept = log(gof1$`Period 1`$ObservedTestStat)) + 
+  ThemeNoNet + 
+  labs(x = "log(Statistic Value)", title = "Distribution of log Mahalanobis Distance", subtitle = "Data shown as vertical line")
+
+@
+
+\st{Because longitudinal network data consist of three different ``spaces" of data, viewing the model in the data space can depend on which aspect of the model and the data you are interested in viewing.
+The SIENA software includes a method for modelling covariate change in the network by incorporating the network structure, assuming that the ties between nodes can affect how nodes behave over time.
+In this instance, a time series plot of predicted covariate values for the nodes as the network evolves is viewing the model in the time and node spaces.
+If, as in Figure~\ref{fig:sienagof}, summary statistics computed on the network are of interest, distributions of these statistics computed on networks simulated from the model show the model in the edge data space.
+For an example, see Figure~\ref{fig:gof2}, where the distribution of the log of the Mahalanobis distance is shown for the outdegree distribution auxiliary statistics, and the log Mahalanobis distance for the observed data is shown as a vertical line.
+
+\begin{figure}
+\includegraphics[width=.75\linewidth]{img/Model1LUSmallFriends.png}
+\caption{\label{fig:lineup} A small lineup of node-link diagrams showing the second wave of our small friendship network among five networks simulated from model $M1$.}
+\end{figure}
+
+Another potential goodness-of-fit visualization is a lineup like those proposed in \citet{buja:2009}.
+A \texttt{lineup} ``asks the witness to identify the plot of the real data from among a set of decoys, the null plots, under the veil of ignorance" \citep[p.
+4369]{buja:2009}.
+It can be thought of like a police lineup, where the "suspect" is in a lineup among several innocent lookalike fillers, and a witness picking the suspect out of the lineup is considered evidence against the suspect.
+In data and model visualization, the "suspect" is a plot of the true data, while the "filler" is composed of several plots of mock data, simulated from a hypothesized model.
+If the true data stands out among the simulated data, that is taken as evidence of poor model fit, whereas if the true data is difficult to identify among the simulated data, that is taked as evidence of good model fit.
+In Figure~\ref{fig:lineup}, the second wave of the small friendship network is shown among five simulated networks from model $M1$ using parameter values estimated from the data.
+The lineup allows us to view the model in the data space by placing instances of the model side-by-side with the data and examining the differences.}
 

### 2017-08-21 paper.Rnw
through new section 4.2

@@
 \end{enumerate}
- %\st{XX more detail about mahalanobis distance and calculating p-value XX}
 
@@
 
-\subsection{View the model in the data space}
+\subsection{View the model in the data space}\label{sec:minds}
 
@@
            singletons = T) +
-  labs(title = "Summary: M1") + ThemeNet +
+  labs(title = "Average: M1") + ThemeNet +
   theme(plot.background = element_rect(color = 'black'))
@@
 It would therefore better to visualize many simulations together.
-For network data, one way we propose to do this is through a "summary" or "average" network drawn using the traditional node-link diagram.
-In Figure~\ref{fig:summnetM1}, we show a summary network created with 1,000 simulations of the second wave of the network from Model 1.
-\st{To make this summary network, we first simulated 1,000 wave 2 and wave 3 observations of our small friendship example data from model $M1$, for which parameters had previously been estimated.
+From a statistician's point of view, a stumbling block with statistical network models generally is the lack of an ``average network" measure.
+Statisticians frequently rely on averages and expected values in data analyses, but statistical network models, especially those as complex as SAOMs, lack a single, intuitive expected value measure.
+We could talk about expected values of parameters, but the parameters can be hard to interpret.
+Furthermore, there is no way to talk about the expected value of an observation from a statistical network model.
+How then, can we arrive at an ``average" network?
+We propose to answer this question through visualization.
+For network data, one way we propose to come up with an "average" network is through a summary network drawn using the traditional node-link diagram.
+In Figure~\ref{fig:summnetM1}, we show an \textit{average network} created with 1,000 simulations of the second wave of the network from Model 1.
+\st{To make this average network, we first simulated 1,000 wave 2 and wave 3 observations of our small friendship example data from model $M1$, for which parameters had previously been estimated.
 We then combine the 1,000 instances of wave 2, and count up the number of times each edge appears in a simulation.
 Then, we combine these 1,000 networks into a single network with edgeweight equal to the proportion of time that edge appears in our 1,000 simulations.
-This is the network we draw using the node-link diagram.} An edge is only drawn in the summary network if it appears in more than 5\% of the simulations (in at least 51 of the 1000 simulations), with edges that appear more frequently emphasized by thicker linewidths and a darker color.
-On either side of the summary network in Figure\ref{fig:summnetM1}, we show the actual data, wave 1 on the left, and wave 2 on the right.
-\st{We can see that the structure of the summary network is much more similar to the first network observation than to the second network observation.
+This is the network we draw using the node-link diagram.} An edge is only drawn in the average network if it appears in more than 5\% of the simulations (in at least 51 of the 1000 simulations), with edges that appear more frequently emphasized by thicker linewidths and a darker color.
+On either side of the average network in Figure\ref{fig:summnetM1}, we show the actual data, wave 1 on the left, and wave 2 on the right.
+\st{We can see that the structure of the average network is much more similar to the first network observation than to the second network observation.
 The simulations, however, are supposed to represent the second wave of data, which is shown on the right in Figure~\ref{fig:summnetM1}.
@@
 
-The second principle of Wickham et al states that ``collections are more informative than singletons" \citep[p.~210]{modelvis}.
-They describe several different methods for producing such collections, but we focus on only a few here.
+The second principle of model visualization is that ``collections are more informative than singletons" \citep[p.~210]{modelvis}.
+The authors describe several different methods for producing such collections, but we focus on only a few here.
 We look at collections produced ``from exploring the space of all possible models," ``by varying model settings", ``by fitting the same model to different datasets", and ``by treating each iteration as a model" \citep[p.~210-11]{modelvis}.
-We also fit the same model many times, resulting in distributions of fitted parameters for one data set.
-
-A classical problem with statistical network models generally is the lack of an ``average network" measure.
-Statisticians frequently rely on averages and expected values, but statistical network models, especially those as complex as SAOMs, lack an intuitive expected value measure.
-We could talk about expected values of parameters, but there is no way to talk about the expected value of the network based on the model.
-How then, can we arrive at an ``average" network?
-We propose to answer this question through visualization.
-
-We these three models using \texttt{RSiena} 1,000 times each.
+Because fitting a SAOM requires MCMC methods, we also fit the same model many times, resulting in distributions of fitted parameters for one data set.
+
+To our small frienship network, we fit three models, $M1$, $M2$, and $M3$, using \texttt{RSiena} 1,000 times each.
 We then looked at the distribution of the fitted values, which are shown in Figure~\ref{fig:distests}.
-We can see from these distributions that the inclusion of the jumping transitive triplet parameter, $\beta_3$ is obviously affecting the distributions of the other four parameters included in all models, $\alpha_1$, $\alpha_2$, $\beta_1$, and $\beta2$.
+We can see from these distributions that the inclusion of the jumping transitive triplet parameter, $\beta_3$ is obviously affecting the distributions of the other four parameters included in all models, $\alpha_1$, $\alpha_2$, $\beta_1$, and $\beta_2$.
 In comparison with models M1 and M3, model M2 typically has higher estimates of the rate parameter, meaning that the inclusion of the covariate statistic in the model leads to higher estimates of the number of times, on average, a node gets to change its ties.
-This seems problematic: it is not clear that the addition of a parameter to the objective function \textit{should} effect the estimation of the rate parameters.
-
-To further investigate this relationship between the parameter values, we look at correlations between each of the parameter estimates in each model.
+It is not clear, however, that the addition of a parameter to the objective function \textit{should} effect the estimation of the rate parameters, so we continue to explore the collection of parameter estimates.
+
+To further investigate this potentially problematic relationship between the parameter values, we look at correlations between each of the parameter estimates in each model.
 In Figure~\ref{fig:corplots}, we examine correlations between each of pair of parameters within each model and overall.\nocite{ggally} The strongest correlation within each model is between $\beta_1$ and $\beta_2$, with absolute value of correlation between those two parameter values greater than 0.90 in all three models.
@@
 We also explore simulations from these different models given parameter values.
-Using the means from the 1,000 fitted parameter values, we simulated 1,000 observations from each of our three models.
-We condition on the first friendship network observation, and the second and third observations are simulated from the SAOM models estimated previously.
-From these simulations, we first create a visualization that represents an ``average" network.
-To do this, we count occurences of each possible edge in the simulations, resulting in a summary network with weighted edges representing the number of times an edge appeared in the simulated wave 2 when simulating from the SAOM 1000 times.
+Using the mean values of from the 1,000 fitted parameter values as the parameters in our models, we simulated 1,000 observations from each of our three models.
+In this process, we condition on the first friendship network observation, and the second and third observations are simulated from the SAOM models with the given parameter values.
+From these simulations, we first create a visualization that represents an average network.
+To do this, we follow the same procedure as in Section~\ref{sec:minds}, counting occurences of each possible edge in the simulations, resulting in a summary network with weighted edges representing the number of times an edge appeared in the simulated wave 2 when simulating from the SAOM 1,000 times.
+As in Figure~\ref{fig:summnetM1}, edges only appear in the average network if they appear more than 5\% of the time in the simulations.
 In Figure~\ref{fig:averagenet}, we show the first wave, the ``average'' network from the three models we fit, and the second wave.
 \st{Comparing the three averages to waves 1 and 2, we see that they have very similar structure to wave 1.
-Model 2, which included the transitive triplet parameter, seems to have created more group structure overall than models 1 and 3.
+Model 2, which included the transitive triplet parameter, seems to have created a larger connected component overall than models 1 and 3.
 In particular, if we look at the group of nodes $\{10,11,14\}$, we see they are very strongly connected within the three average networks, and they are completely separate from the other nodes in the true wave 2.
 None of the three average networks show node 16 gaining ties as it does in wave two, nor do they show nodes 4 and 7 becoming isolated.
-In Model 2, however, the ties to node 7 appear much weaker than in Model 1 or Model 2, suggesting that of the three, Model 2 is the best fit for our data.}  %XXXX ST commented out after I realized I had made the same mistake as figure 7.
+In Model 2, however, the ties to node 7 appear much weaker than in Model 1 or Model 2, suggesting that of the three, Model 2 may be the best fit for our data.}  %XXXX ST commented out after I realized I had made the same mistake as figure 7.
 XX This could be an indication that, despite the highly significant nature of the parameter $\beta_3$ according to the $t$-tests in \texttt{RSiena}, the transitive triplet parameter should not be included because it does not result in a structure similar to the data.

### 2017-08-21 paper.Rnw
add adjmat animation

@@
 \usepackage{subcaption}
+\usepackage{animate}
 \usepackage{afterpage}
@@
             theme(plot.title = element_text(size = rel(2),
-                                            face = 'bold', angle = 0), 
+                                            face = 'plain', angle = 0), 
                   axis.title.x = element_text(size = rel(2),
@@
             theme(plot.title = element_text(size = rel(2),
-                                            face = 'bold', angle = 0), 
+                                            face = 'plain', angle = 0), 
                   strip.text.x = element_text(size = rel(2),
@@
 
-<<gof2, fig.cap="An example of viewing the ">>=
+<<gof2, fig.cap="An example of viewing the distribution of a statistic of interest from a SAOM.
+Here, the Mahalanobis distance of the cumulative outdegree values computed on a simulated network to the mean value simulated is shown.
+The distance computed from the data is shown as a vertical line.", fig.height=3>>=
 qplot(x = log(gof1$`Period 1`$SimulatedTestStat), binwidth = .2) + 
@@
 
-The model fitting process in \texttt{RSiena} involves several simulation steps that are hidden from the user.
-This is practical and efficient if a researcher is primarily interested in fitting one model to a set of longitudinal network data and obtaining parameter estimates.
-We are more interested in how the models are fit, so we extracted and explored the different steps of that process.
-
-A key component of each step of the SIENA method of moments algorithm is the ``microstep." This step simulates from the model in its current state, $x(t_m)$ with current parameter values $\theta_{0}$, to the next state, $x(t_{m+1})$.
+\st{The last principle of model visualization is to explore the process of fitting the model, instead of just focusing on the end result.
+This principle is perhaps the most important for SAOMs because} the model fitting process in \texttt{RSiena} involves several simulation steps that are hidden from the user.
+Hiding the MCMC steps is practical and efficient if a researcher is primarily interested in fitting one model to a set of longitudinal network data, obtaining parameter estimates, \st{and drawing conclusions or making predictions}.
+We are more interested in \textit{how} the models are fit, so we extracted and explored the different steps of that process.
+
+A key component of each step of the SIENA method of moments algorithm is the ``microstep" process.
+A series of microsteps is obtained by simulating from the model in its current state, $x(t_m)$ with current parameter values $\mathbf{\theta}_{0} = \{\alpha_{1_0}, \dots \alpha_{{m-1}_0}, \beta_{1_0}, \dots, \beta_{K_0}\}$, to the next state, $x(t_{m+1})$.
 This microstep process stops when the simulated network has acheived the same number of differences, $C$, from $x(t_m)$ as $x(t_{m+1})$, where
@@
 This simulation process follows the steps of the continuous-time Markov chain.
-First, an ``ego node'' is selected to make a change, and then it randomly makes one change in its ties according to the probability, $p_{ij}$, determined by its objective function.
+\st{Each tie change in the CTMC is referred to as one ``microstep".
+At each microstep}, an ``ego node" is selected to make a change, and the chosen ego node randomly makes one change in its ties according to the probabilities, $\{p_{ij}: i\neq j \in \{1, \dots, n\}\}$, determined by its objective function.
 The options for change are (1) removing a current tie, (2) adding a new tie, or (3) making no change at all.
-Saving all of these steps is not efficient if you're only interested in estimating parameter values, but they can be saved and extracted using options in \texttt{RSiena}, which is exactly what we did to create our visualizations.
-Between two network observations $x(t_m)$ and $x(t_{m+1})$, there can be dozens, hundreds, or even thousands of microsteps, depending on the size of the network.
-We wanted to see these in-between steps in order to better understand the behavior of the underlying continuous-time Markov chain.
-
-The first vizualization we present here is an animation of microsteps simulated to emulate the transition steps from wave 1 to wave 2 the subset of the friends data shown in Figure ~\ref{fig:alldataviz}.
-Movies similar to our animation were used to visualize the changes of dynamic networks in \citet{dynnetviz}.
+Saving and exploring all of these steps is not computationally efficient if one is only interested in estimating parameter values, but they can be saved and extracted using options in \texttt{RSiena}, which is exactly what we did to create our visualizations.
+Between two network observations $x(t_m)$ and $x(t_{m+1})$, there can be dozens, hundreds, or even thousands of microsteps, depending on the size of the network and the number of changes between two network observations.
+We wanted to view these in-between steps in order to better understand the behavior of the underlying continuous-time Markov chain.
+
+The first vizualization we present here is an animation of microsteps simulated to emulate the transition steps of the CTMC from wave 1 to wave 2 of the small friendship network example shown in Figure ~\ref{fig:alldataviz}.
+Movies similar to this animation were used to visualize the changes of dynamic networks in \citet{dynnetviz}.
 When each ego node is selected in a microstep, it is emphasized in the animation, then the associated edge either appears or disappears.
@@
 In both cases, the ego nodes chosen to act change color from black to red, and they also increase greatly in size.
-In the case of an edge being removed, the edge that currently exists is emphasized with the color and size change that the node also gets, and as the animation proceeds the edge shrinks to disappearing, while the node shrinks and changes color back to the original black.
-If the edge is being added, it appears colored red from nothing and grows to a large size, then changes color to match the rest of the edges, while the node shrinks and changes color to match the other nodes.
+In the case of an edge being removed, shown in the top row, the edge that currently exists is emphasized with the same color and size change that the node gets, and as the animation proceeds the edge shrinks to nothing, as the ego node shrinks and changes color back to its original black.
+If an edge is being added, as in the bottom row of the figure, the ego node's appearances changes in the sampe way as when the edge is being removed, whil the edge appears colored red from nothing, and grows to a large size, then changes color and size to match the rest of the edges, while the node shrinks and changes color to match the other nodes.
 
@@
 Viewing the adjacency matrices with this arbitrary ordering does not provide much information to the viewer about the underlying structure of the network.
-This lack of perceived structure would be exacerbated in an animation.
-So, we adjust the ordering so that the viewer can perceive more of the structure of the graph.
+This lack of perceived structure would be exacerbated in an animation, so we adjust the ordering so that the viewer can perceive more of the structure of the graph.
 This process is known as matrix seriation \citep{seriation}.
-
-To reorder the vertices for the matrix visualization, we first constructed a cumulative adjaceny matrix, $\mathbf{A}^{cum}$, for the series of microsteps simulating the network from $x(t_m)$ to $x(t_{m+1})$.
-A single entry in the cumulative adjacency matrix, $\mathbf{A}^{cum}_{ij}$, is the total number of times the edge from node $i$ to node $j$ appears in the network from the initial observation, $x(t_m) \equiv X(0)$ to the final reult of the last microstep, $X(C)$:
-$$ \mathbf{A}^{cum}_{ij} = \sum_{c = 0}^C X_{ij}(c). $$
-We then performed a principal component analysis on $\mathbf{A}^{cum}$, and used the values of the first principal component to order the vertices on the $x$ and $y$ axes for the adjacency matrix animation.
-For one such series of microsteps simulated by \texttt{RSiena}, we present the adjacency matrix ordered by the (arbitrary) vertex id alongside the seriated adjacency microsteps using the first principal component loading on the cumulative adjacency matrix, $\mathbf{A}^{cum}$, in Figure~\ref{fig:adjmatorder}.
 
@@
 
-pcaordersf$from <- factor(pcaordersf$from, levels = order1)
-pcaordersf$to <- factor(pcaordersf$to, levels = rev(order1))
-
-p2 <- ggplot(data = pcaordersf, aes(x = from, y = to)) + geom_tile(color = 'black', aes(fill = as.factor(value))) + 
+pcaordersf$from <- factor(pcaordersf$from, levels = rev(order1))
+pcaordersf$to <- factor(pcaordersf$to, levels = order1)
+
+p2 <- ggplot(data = pcaordersf, aes(x = to, y = from)) + geom_tile(color = 'black', aes(fill = as.factor(value))) + 
   scale_fill_manual(values = c('white', 'black')) +
@@
 grid.arrange(p1,p2, nrow = 1)
+@
+
+To reorder the vertices for the matrix visualization, we first constructed a cumulative adjaceny matrix, $\mathbf{A}^{cum}$, for the series of microsteps simulating the network from $x(t_m)$ to $x(t_{m+1})$.
+A single entry in the cumulative adjacency matrix, $\mathbf{A}^{cum}_{ij}$, is the total number of times the edge from node $i$ to node $j$ appears in the network from the initial observation, $x(t_m) \equiv X(0)$ to the final result of the last microstep, $X(R)$, where $R$ is the total number of microsteps taken:
+$$ \mathbf{A}^{cum}_{ij} = \sum_{r = 0}^R X_{ij}(r). $$
+We then performed a principal component analysis on $\mathbf{A}^{cum}$, and used the values of the first principal component to order the vertices on the $x$ and $y$ axes for the adjacency matrix animation.
+For one such series of microsteps simulated by \texttt{RSiena}, we present the adjacency matrix ordered by the (arbitrary) vertex id alongside the seriated adjacency microsteps using the first principal component loading on the cumulative adjacency matrix, $\mathbf{A}^{cum}$, in Figure~\ref{fig:adjmatorder}.
+
+<<adjmatani, fig.cap='The adjacency matrix visualization animation for one series of microsteps.
+At the beginning of the animation, there are two connected components.
+By the end, the components appear to be more spread out.', fig.show='animate', warning = FALSE, message = FALSE, interval=.2, results='hide', fig.height=4>>=
+adjmatall <- matrix(0, nrow = 16, ncol = 16)
+rownames(adjmatall) <- paste0("V", 1:16)
+colnames(adjmatall) <- paste0("V", 1:16)
+dat2 <- sfmsall %>% filter(rep == 1)
+adjmatall2 <- NULL
+for(i in 0:55){
+  MS <- dat2[dat2$ms == i,c("from", "to")]
+  myadjmat <- adjmatall
+  for(j in 1:nrow(MS)){
+    myadjmat[MS[j,"from"][[1]],MS[j,"to"][[1]]] <- 1
+  }
+  adjmatall2 <- rbind(adjmatall2, myadjmat)
+}
+
+adjmatall2 <- cbind(adjmatall2, rep(0:55, each = 16))
+adjmatall2 <- data.frame(adjmatall2)
+adjmatall2$from <- rep(paste0("V", 1:16), 56)
+names(adjmatall2)[17] <- "ms"
+adjmatall2 %>% gather(to, val, V1:V16) -> adjmatall3
+
+adjmatall3$to <- factor(adjmatall3$to, levels = order1)
+adjmatall3$from <- factor(adjmatall3$from, levels = rev(order1))
+plots <- list()
+for(i in 0:55){
+plots[[i + 1]] <- ggplot(data = adjmatall3 %>% filter(ms == i)) + 
+  geom_tile(aes(x = to, y = from, fill = as.factor(val)), color = 'black') + 
+  scale_fill_manual(values = c("white", "black")) + 
+  scale_x_discrete(position = "top", labels = str_replace(order1, "V", "")) + 
+  scale_y_discrete(labels = rev(str_replace(order1, "V", ""))) + 
+  ThemeNoNet + 
+  theme(legend.position = 'none') +
+  coord_fixed() 
+}
+plots 
 @

### 2017-08-21 paper.Rnw
through section 4.3. Just need to add some more discussion of figures and write section 5. then done!

@@
 \begin{figure}
+\centering
 \includegraphics[width=.75\linewidth]{img/Model1LUSmallFriends.png}
@@
 Using the principal component analysis on $\mathbf{A}^{cum}$ to order the rows and columns of the adjacency matrix visualization clearly shows the two distinct connected components in the first wave of the network, which are difficult to find in the arbitrarily ordered visualization.
-
-We also attempt to better understand the microstep process by visualizing all transition probabilites for the first microstep in the process.
-We only do the first step of many for now because the \texttt{RSiena} transition probabilities after the first step are only comparable for identical steps due to the conditioning on the current network state in the model.
-Thus we have one step for each of the 1,000 repititions to visualize.
-In Figure ~\ref{histogramalters}, we present each acting node, (ego node) and the resulting probabilities of change.
-The probability shown by the bars is the theoretical probability according to the objective function of the ego node changing the tie to the alter node.
-The probability shown by the points is the empirical probability of that change being made, calculated by counting all instances of the ego node first, then computing the proprotion of each change.
-In most cases, they are almost identical, showing that the algorithm is performing as expected.
-
+We use this layout to fix the layout in the animation of one of the microstep process simulations.
+This animation, which is shown in Figure~\ref{fig:adjmatani} is very simple: a square appears or disappears in the animation as that edge appears or disappears in the microstep process.
+Through this animation, we can see edges appearing, and then later on disappearing.
+These in-between steps are not shown when we look at the network at our discrete observation points.
+By viewing this animation, we gain a better understanding of the dynamics of this model.
 <<histogramalters, fig.width = 8, fig.height=11, fig.cap="Each panel shows the theoretical (as lines) and empirical (as points) probabilities of the chosen ego node changing its tie to each of the other nodes.
@@
 
-Another way to view the transition probabilities is through the adjacency matrix visualization.
-In Figure~\ref{fig:heatmap}, we build on the concept of the ordered adjacency matrix of Figure~\ref{fig:adjmatorder}.
-This heatmap shows the transition probabilities of all ties that are changed in the first microstep of the 1,000 simulations.
-The heatmap is noticeably sparse, showing the lack of coverage in the model.
-Of the 256 possible steps, only 81, or about 32\%, are taken in the 1,000 simulated chains.
-XXX Check this again....
-XXX
+We also attempt to better understand the microstep process by visualizing the observed transition probabilites for the first microstep in the process.
+We only do the first step of many for now because the \texttt{RSiena} transition probabilities after the first step are only directly comparable for identical steps due to the conditioning on the current network state in the model.
+In other words, for $i \neq j \neq k \neq \ell \in \{1, \dots, n\}$, the probability of tie $x_{k\ell}$ changing in the second microstep is different in the case of $x_{ij}$ changing in the first step from the case of $x_{ij}$ not changing in the first step.
+Thus we have one transition probability for the first microstep taken for each of the 1,000 simulations to visualize.
+In Figure~\ref{fig:histogramalters}, we present each acting node, (ego node) and the resulting probabilities of tie changes.
+The probability shown by the bars is the theoretical probability, according to the objective function, of the ego node changing its tie to the alter node, while the probability shown by the points is the empirical probability of that change being made.
+The empirical probability is calculated by counting all instances of the ego node first, then computing the proportion of each different alter node change.
+In most cases, they are almost identical, which demonstrates that the algorithm is performing as expected.
+
 <<heatmap, fig.align='center', fig.cap='A heatmap showing the empirical transition probabilities for the first microstep in 1,000 simulations.
 The acting node is on the x-axis, and the receiving node is on the y-axis.
-Missing squares represent ties that did not occur.', fig.height=3>>=
+Missing squares represent ties that did not occur.', fig.height=4>>=
 ansnullchains %>% 
-  filter(rep == 1, period == 1) %>% 
+  filter(period == 1) %>% 
+  group_by(rep) %>% 
+  mutate(msno = row_number()) %>% 
+  filter(msno == 1) %>% 
   mutate(ego = paste0("V",as.numeric(as.character(ego))+1),
@@
 
+# probsDat %>% group_by(ego, alter) %>% summarize(count = n()) %>% dim
+# 103, 3
 ggplot(data = probsDat, 
@@
 
-We also wanted to view the complete microstep process from the first wave, which we condition on, to the second wave.
-The number of steps taken from wave 1 to wave 2 varies.
-In the simulations from Model 1, the smallest number of steps taken was 58, and the longest was 248, with a mean of 106 and a median of 103.
-The standard deviation of the 1000 simulations' number of microsteps was 22.8.
-In Figure~\ref{fig:allsteps}, we see three simulations of the process from wave 1 to wave 2, with wave 1 shown on the left, and wave 2 shown on the right.
-In each of the three plots, the $y$-axis contains the edges sorted by how often they appear in the networks along the way.
-We can see that some edges are there in the beginning, but disappear and never come back, while others appear a few steps in, only to dispappear again.
-There are also some edges that exist in wave 2 that don't appear at all in the microstep process in a given repitition.
+Another way to view these transition probabilities is through the adjacency matrix visualization.
+In Figure~\ref{fig:heatmap}, we build on the concept of the ordered adjacency matrix of Figure~\ref{fig:adjmatorder}.
+This heatmap shows the transition probabilities of all ties that are changed in the first microstep of the 1,000 simulations.
+The heatmap is noticeably sparse: %showing the lack of coverage in the model.
+of the 256 possible steps, only 103, or about 40\%, are taken in the 1,000 simulated chains.
 
@@
 
-p3 <- ggplot(data = msall3_df2) + 
-  geom_tile(aes(x = ms, y = -eid, fill = ms)) + 
-  geom_vline(xintercept = c(0.5,106.5)) + 
-  scale_fill_continuous(low = "red", high = "blue") +
+# p3 <- ggplot(data = msall3_df2) + 
+#   geom_tile(aes(x = ms, y = -eid, fill = ms)) + 
+#   geom_vline(xintercept = c(0.5,106.5)) + 
+#   scale_fill_continuous(low = "red", high = "blue") +
+#   ThemeNoNet + 
+#   theme(legend.position = 'none', axis.ticks.y = element_blank(),
+#         axis.text.y = element_blank(), axis.title = element_blank()) #+
+#   #labs(x = "Microstep", y = "Edges (most frequent at top)")
+
+grid.arrange(p1, p2, nrow = 2)
+@
+
+We also wanted to view the complete microstep process from the first wave, which we condition on in the model, to the second wave.
+The number of steps taken from wave 1 to wave 2 varies.
+In the simulations from Model 1, the smallest number of steps taken was 58, and the longest was 248, with a mean of 106 and a median of 103.
+In the 1,000 simulations, the standard deviation of the number of microsteps was 22.8.
+In Figure~\ref{fig:allsteps}, we see two simulations of the process from wave 1 to wave 2, with wave 1 shown on the left, and wave 2 shown on the right.
+In each of the three plots, the $y$-axis contains the edges sorted by how often they appear in the networks along the way.
+We can see that some edges are there in the beginning, but disappear and never come back, while others appear a few steps in, only to dispappear again.
+There are also some edges that were observed in wave 2 that don't appear at all in the microstep process in a given simulation.
+
+<<alledgests>>=
+# get a plot of all microsteps from w1 to w2 by edge
+alledges <- expand.grid(from = paste0("V", 1:16), to = paste0("V", 1:16))
+alledges$eid <- 1:nrow(alledges)
+
+sfmsall %>% left_join(alledges) -> sfmsall2
+
+sfmsall2 %>% group_by(rep) %>% mutate(isWave2 = (ms == max(ms))) -> sfmsall2
+
+sfmsall2 %>% group_by(from, to, eid) %>% 
+  summarize(total =n()) %>% 
+  arrange(desc(total)) %>% 
+  ungroup %>% 
+  mutate(plotorder = row_number()) -> sfmsallorder
+
+sfmsall2 %>% group_by(ms, eid) %>% 
+  summarise(count = n()) %>%
+  mutate(alpha = count/1000) %>% 
+  arrange(ms, eid) -> sfmsall3
+
+left_join(sfmsall3, sfmsallorder, by = "eid") -> sfmsall4
+
+ggplot(data = sfmsall4) + 
+  geom_tile(aes(x = ms, y = reorder(eid,-plotorder), alpha = alpha), fill = 'navy') +
+  scale_alpha_continuous("occurence %") + 
+  labs(x = "Microstep No.", y = "Edges (ordered by total # occurences)") + 
   ThemeNoNet + 
-  theme(legend.position = 'none', axis.ticks.y = element_blank(),
-        axis.text.y = element_blank(), axis.title = element_blank()) #+
-  #labs(x = "Microstep", y = "Edges (most frequent at top)")
-
-grid.arrange(p1, p2, p3, nrow = 3)
-@
-
-\st{The algorithms here are many!
-\begin{enumerate}
-%\item First, the microsteps / CTMC between each time point (gifs, other animations)
-\item Phases of RSiena: phase 1 = initial parameter estimate
-\item phase 2 = iterative update of parameters - trace plots
-\item phase 3 = simulations for convergence diagnostics
-\end{enumerate}
- }
+  theme(axis.ticks.y = element_blank(), axis.text.y = element_blank(), panel.grid = element_blank())
+@
+
+\st{We also combine all 1,000 simulations into a visualization like the one shown in Figure~\ref{fig:allsteps}.
+We first assign each possible edge an edge ID number so that we can keep track of it throughout all the microsteps and all the simlations.
+Then, we count up the total number of times each edge appears in the microstep process in each of the 1,000 simulations for use as an ordering variable later.
+We also count up the number of times an edge occurs in each microstep number in the 1,000 simulations.
+Since the number of microsteps in the process varies, the number of times an edge occurs decreases as the microstep increases.
+Next, we compute a proportion, which we call the occurence percentage, which is the number of times the edge occurred in a microstep divided by 1,000.
+Finally, we visualize all this information together in Figure ~\ref{fig:alledgeests}.
+XX more detail tomorrow XX }
+
+% \st{The algorithms here are many!
+% \begin{enumerate}
+% %\item First, the microsteps / CTMC between each time point (gifs, other animations)
+% \item Phases of RSiena: phase 1 = initial parameter estimate
+% \item phase 2 = iterative update of parameters - trace plots
+% \item phase 3 = simulations for convergence diagnostics
+% \end{enumerate}
+%  }
 
@@
 
-\hh{Some of the visualizations we talked about
-\begin{enumerate}
-\item visualization of the MCMC process:  movie of individual actors' step-by-step decisions; could be done with the ndtv package, as long as we can get individual steps out of RSiena
-\item global picture: 1000 simulated networks + alpha blending should give a pretty good idea of the most important edges
-%\item global structure + principal components (tensor pca)
-\end{enumerate}}
+% \hh{Some of the visualizations we talked about
+% \begin{enumerate}
+% \item visualization of the MCMC process:  movie of individual actors' step-by-step decisions; could be done with the ndtv package, as long as we can get individual steps out of RSiena
+% \item global picture: 1000 simulated networks + alpha blending should give a pretty good idea of the most important edges
+% %\item global structure + principal components (tensor pca)
+% \end{enumerate}}
 

### 2017-08-22 paper.Rnw
so close to being done.

@@
 From these simulations, we first create a visualization that represents an average network.
-To do this, we follow the same procedure as in Section~\ref{sec:minds}, counting occurences of each possible edge in the simulations, resulting in a summary network with weighted edges representing the number of times an edge appeared in the simulated wave 2 when simulating from the SAOM 1,000 times.
+To do this, we follow the same procedure as in Section~\ref{sec:minds}, counting occurrences of each possible edge in the simulations, resulting in a summary network with weighted edges representing the number of times an edge appeared in the simulated wave 2 when simulating from the SAOM 1,000 times.
 As in Figure~\ref{fig:summnetM1}, edges only appear in the average network if they appear more than 5\% of the time in the simulations.
@@
 The number of steps taken from wave 1 to wave 2 varies.
-In the simulations from Model 1, the smallest number of steps taken was 58, and the longest was 248, with a mean of 106 and a median of 103.
+In one set of 1,000 simulations from Model 1, the smallest number of steps taken was 58, and the longest was 248, with a mean of 106 and a median of 103.
 In the 1,000 simulations, the standard deviation of the number of microsteps was 22.8.
@@
 
-<<alledgests>>=
+<<alledgests, fig.height = 4, fig.cap="Visualizing all microsteps taken in 1,000 simulations from the model M1.
+The first wave network is shown at microstep 0, and the second wave of the network is shown as the last microstep for comparison.
+We see that it is rare for a microstep process to last longer than 150 steps, and also that the edges that appear past the 150th step tend to be in either the first wave or the second wave.">>=
 # get a plot of all microsteps from w1 to w2 by edge
@@
 
-ggplot(data = sfmsall4) + 
-  geom_tile(aes(x = ms, y = reorder(eid,-plotorder), alpha = alpha), fill = 'navy') +
-  scale_alpha_continuous("occurence %") + 
-  labs(x = "Microstep No.", y = "Edges (ordered by total # occurences)") + 
+#add true wave 2 to plot
+wave2 <- na.omit(actual2)
+wave2$from <- paste0("V", wave2$from)
+wave2$to <- paste0("V", wave2$to)
+wave2 <- left_join(wave2, alledges)
+wave2 <- left_join(wave2, sfmsallorder)
+
+sfmsall5 <- sfmsall4 %>% ungroup %>% add_row(ms = max(sfmsall4$ms) + 1, 
+        eid = wave2$eid, 
+        count = NA, 
+        alpha = 1, 
+        from = wave2$from,
+        to = wave2$to,
+        total = wave2$total, 
+        plotorder = wave2$plotorder)
+
+ggplot(data = sfmsall5) + 
+  geom_tile(aes(x = ms, y = reorder(eid,-plotorder), fill = alpha)) +
+  #scale_alpha_continuous("occurence %") + 
+  scale_fill_gradient(low = 'grey90', high = '#0868ac', name = "occurrence %") + 
+  labs(x = "Microstep No.", y = "Edges (ordered by total # occurrences)") + 
   ThemeNoNet + 
@@
 We also count up the number of times an edge occurs in each microstep number in the 1,000 simulations.
-Since the number of microsteps in the process varies, the number of times an edge occurs decreases as the microstep increases.
-Next, we compute a proportion, which we call the occurence percentage, which is the number of times the edge occurred in a microstep divided by 1,000.
-Finally, we visualize all this information together in Figure ~\ref{fig:alledgeests}.
-XX more detail tomorrow XX }
+Since the number of microsteps in the process varies, the number of times an edge occurs decreases as the microstep number increases.
+Next, we compute a proportion, which we call the occurrence percentage, which is the number of times the edge occurred in a microstep divided by 1,000.
+Finally, we visualize all this information together in Figure ~\ref{fig:alledgests}.
+In this plot, all possible edges are shown, and we see that every one of the $16 \times 15 = 240$ possible edges in the network occurs at some point in the simulation process.
+We also see, however, that the process struggles to focus in on the edges in the second wave of the data.
+Ideally, we would like to see more occurrences of the second wave of data generally.
+About half of the edges in wave two are in the bottom half when ordered by number of occurrences, meaning they aren't appearing as much as they would if the model was truly excellent at capturing the mechanisms of tie change in the network.}
 

### 2017-08-22 paper.Rnw
done with last plot description (for now). take a break, then move on to the discussion.

@@
 <<alledgests, fig.height = 4, fig.cap="Visualizing all microsteps taken in 1,000 simulations from the model M1.
+The occurrence percent is split up into groups to correspond with its distribution: only about 10\% of the edges appear more than 10\% of the time in the 1,000 simulations, while about 60\% appear less than 1\% of the time.
 The first wave network is shown at microstep 0, and the second wave of the network is shown as the last microstep for comparison.
@@
         plotorder = wave2$plotorder) 
-
+sfmsall5$alphaFctr <- cut(sfmsall5$alpha, c(0,.01, .05, .10, .25, .5, .75, 1, 1.1))
 ggplot(data = sfmsall5) + 
-  geom_tile(aes(x = ms, y = reorder(eid,-plotorder), fill = alpha)) +
-  #scale_alpha_continuous("occurence %") + 
-  scale_fill_gradient(low = 'grey90', high = '#0868ac', name = "occurrence %") + 
+  geom_tile(aes(x = ms, y = reorder(eid,-plotorder), fill = alphaFctr)) +
+  #scale_fill_gradient(low = 'grey90', high = '#0868ac', name = "occurrence %") + 
+  scale_fill_brewer(palette = "Blues", name = "Occurrence %", labels = c("(0-1]%", "(1-5]%", "(5-10]%","(10-25]%", "(25-50]%", "(50-75]%", "(75-100]%")) + 
   labs(x = "Microstep No.", y = "Edges (ordered by total # occurrences)") + 
@@
 
-\st{We also combine all 1,000 simulations into a visualization like the one shown in Figure~\ref{fig:allsteps}.
+\st{We also combine 1,000 simulations from model M1 into a visualization like the one shown in Figure~\ref{fig:allsteps}.
 We first assign each possible edge an edge ID number so that we can keep track of it throughout all the microsteps and all the simlations.
@@
 Ideally, we would like to see more occurrences of the second wave of data generally.
-About half of the edges in wave two are in the bottom half when ordered by number of occurrences, meaning they aren't appearing as much as they would if the model was truly excellent at capturing the mechanisms of tie change in the network.}
+About half of the edges in wave two are in the bottom half when ordered by number of occurrences, meaning they aren't appearing as much as they would if the model was truly excellent at capturing the mechanisms of tie change in the network.
+In addition, there is not a distinct pattern or structure that we can see in this particular edge visualization.
+The edges that the model starts with appear more often in the first 50 or so microsteps than any other set of edges.
+This makes sense, but it would be better if the model explored the space more than it currently does.
+It could be that the simple model M1 isn't complicated enough for the friendship dynamics in the data, or it could be that the chains in the MCMCs "get stuck" when exploring the parameter space.}
 

### 2017-08-22 paper.Rnw
a couple blurbs of discussion.

@@
 ThemeNoNet <- theme_bw() %+replace% 
-            theme(plot.title = element_text(size = rel(2),
+            theme(plot.title = element_text(size = rel(1.5),
                                             face = 'plain', angle = 0), 
-                  axis.title.x = element_text(size = rel(2),
+                  axis.title.x = element_text(size = rel(1.5),
                                             face = 'plain', angle = 0),
-                  axis.title.y = element_text(size = rel(2),
+                  axis.title.y = element_text(size = rel(1.5),
                                             face = 'plain', angle = 90),
@@
 % }
-\hh{XXX - talk a bit about social networks first, why we want to model them, and that usually the models are not dynamic.}
-\st{XXX - okay, here is what i have done for that:
+%\hh{XXX - talk a bit about social networks first, why we want to model them, and that usually the models are not dynamic.}
+\st{%XXX - okay, here is what i have done for that:
 Social networks have been studied for decades, beginning with a few foundational works, including the 1967 study, "The Small World Problem" by Stanley Milgram \citep{goldenberg09}.
@@
 Thus, $\mathbf{A}$ is always symmetric for undirected networks, and is symmetric for directed networks only if every edge between two nodes is reciprocated.}
-\hh{XXX what is an adjacency matrix? - explain at this point in mathematical terms} \st{XXX see above XXX}
+%\hh{XXX what is an adjacency matrix? - explain at this point in mathematical terms} \st{XXX see above XXX}
 An adjacency matrix visualization for our toy example is also shown in Figure~\ref{fig:toyplots}.
@@
 One drawback of the adjacency matrix visualization that \citeauthor{adjmatviz} found was that edges are overrepresented for undirected graphs, due to the symmetry of $\mathbf{A}$: the edge $x_{ij}$ for $i \neq j$ appears in $\mathbf{A}$ twice: in $A_{ij}$ and $A_{ji}$, and so it also appears twice in the adjacency matrix visualization.
-\hh{XXX What is the exact logic here that Ghoniem used to come to the conclusion that edges in undirected graphs are over-represented?
+%\hh{XXX What is the exact logic here that Ghoniem used to come to the conclusion that edges in undirected graphs are over-represented?
 Why would directed graphs make that better?
@@
 } \st{XXX in undirected nets, the adj mat is symmetric, where one edge ($x_{ij}$) is represented in entry $i,j$ in the matrix and in entry $j,i$.
-So, there are many repetitions in the matrix, which makes the network appear more dense than it really is.} \st{This, however, may actually be an advantage for \emph{directed} graphs, where exactly the correct number of edges is represented in a matrix visualization, due to the fact that the edges $x_{ij}$ and $x_{ji}$ are not interchangeable.
+So, there are many repetitions in the matrix, which makes the network appear more dense than it really is.}
+\st{This, however, may actually be an advantage for \emph{directed} graphs, where exactly the correct number of edges is represented in a matrix visualization, due to the fact that the edges $x_{ij}$ and $x_{ji}$ are not interchangeable.
 A node-link diagram, however, may underrepresent the edge count if the edges $x_{ij}$ and $x_{ji}$ both exist and are drawn on top of one another.
@@
 In general, SAOMs can incorporate covariate values and network statistics into the model, so that each node will have a different rate of change.
-\hh{XXX that means that the function is essentially just a rate parameter, right?} \st{XXX yes XXX}
+%\hh{XXX that means that the function is essentially just a rate parameter, right?} \st{XXX yes XXX}
 Other modelling scenarios allow this rate to be more flexible,  e.g.\ a function that depends on the time period of observation, some actor-level covariates or some actor-level network statistics. 
-\hh{How is the rate parameter different from a rate parameter?} \st{I think you meant how is a rate fn different from a rate parameter?
+%\hh{How is the rate parameter different from a rate parameter?} \st{I think you meant how is a rate fn different from a rate parameter?
 I updated the language XXX->}%In our model with the simple rate function,
@@
 
-\hh{shouldn't that be independence? - make the conditional independence assumption above a labelled formula (e.g. C1) and then call it here }
+%\hh{shouldn't that be independence? - make the conditional independence assumption above a labelled formula (e.g. C1) and then call it here }
 Because of the conditional \st{in}dependence assumptions given in Equation~\ref{eq:c1}, we can consider the objective function for each node separately, as only one tie from one node is \st{allowed to change} at a time.
@@
 The vector $\boldsymbol{\beta}$ contains the parameters of the model with corresponding network and covariate statistics, $s_{ik}(x, \mathbf{Z})$, for $k = 1,\dots, K$.
-\hh{XXX ego node isn't defined yet - the description of the charts would be a good place for it.} \st{XX I added it above the obj function equation.
-Is it ok there?} Given the ego node, $i$, there are $n$ possible steps for the actor $i$ to take: either one of all current ties $x_{ij} = 1$ will be destroyed, a new tie will be created thatis currently $x_{ij}=0$, or no change will occur.
+%\hh{XXX ego node isn't defined yet - the description of the charts would be a good place for it.} \st{XX I added it above the obj function equation.
+Is it ok there?}
+Given the ego node, $i$, there are $n$ possible steps for the actor $i$ to take: either one of all current ties $x_{ij} = 1$ will be destroyed, a new tie will be created thatis currently $x_{ij}=0$, or no change will occur.
 
 The parameters, $\boldsymbol{\beta}$, correspond to various actor-level network statistics, $s_{ik}(x)$.
-\hh{'There are always' ...
+%\hh{'There are always' ...
 that seems to come from a Snijder paper?
@@
 The probability density function for the type 1 extreme value distribution is shown in Figure~\ref{fig:logweibull}.
-Using the distribution given in Equation~\ref{eq:dist1} with mean $\mu = 0$ and scale $\sigma = 1$ as \citet{modelsSnijders} suggests \hh{XXX I'm not following - which distribution do you mean by `this' distribution?
+Using the distribution given in Equation~\ref{eq:dist1} with mean $\mu = 0$ and scale $\sigma = 1$ as \citet{modelsSnijders} suggests %\hh{XXX I'm not following - which distribution do you mean by `this' distribution?
 The density above is the type 1 extreme value density.
@@
 
-<<allsteps, fig.cap="Three simulations (out of 1,000) of the microstep process from wave 1 to wave 2.
+<<allsteps, fig.cap="Two simulations (out of 1,000) of the microstep process from wave 1 to wave 2.
 The $x$ axis is the microstep number, with step 0 representing the first wave of data and the final step representing the second wave of data.
@@
 <<alledgests, fig.height = 4, fig.cap="Visualizing all microsteps taken in 1,000 simulations from the model M1.
-The occurrence percent is split up into groups to correspond with its distribution: only about 10\% of the edges appear more than 10\% of the time in the 1,000 simulations, while about 60\% appear less than 1\% of the time.
+The occurrence percent is split up into groups to correspond with its distribution: only about 10\\% of the edges appear more than 10\\% of the time in the 1,000 simulations, while about 60\\% appear less than 1\\% of the time.
 The first wave network is shown at microstep 0, and the second wave of the network is shown as the last microstep for comparison.
@@
 
+We have used novel visualization methods in order to better understand the family of models known as stochastic actor-oriented model for social network.
+By looking at the underlying algorithms, exploring collections of these models, and viewing the model in the data space, we have been able to gain knowledge and appreciation for these complicated models and everything that goes into them.
+
+We have only just begun to scratch the surface of these complicated and multi-layered models for social networks.
+The \texttt{RSiena} software is incredibly powerful, and can fit a whole slew of much more flexible stochastic actor-oriented models than we have examined here.
+If a researcher thinks the network structure or an actor covariate effects the rate of change of the networ, there is a way to incorporate that belief into the rate function of the SAOM.
+More than one actor-level covariate can be included in the model, and way more than three parameters can be included in the objective function itself.
+In addition, \texttt{RSiena} allows the user to tell it which parameters lead to tie creation, and which parameters lead to tie endowment, or dissolution.
+We have used ``evaluation" parameters, which assume that creation and endowment are equal \citep{RSienamanual}.
+Finally, SAOMs and \texttt{RSiena} are able to also model behavior change of the actors in the network, which again is a capability we did not explore here.
+
 \bibliographystyle{asa}

### 2017-08-22 paper.Rnw
add "sanity-check" abstract

@@
 \begin{abstract}
- \hh{XXX just a placeholder}
+Social networks have been studied for decades, beginning with a few foundational works, including the 1967 study, "The Small World Problem" by Stanley Milgram.
+In this paper, we concentrate on one type of model for dynamic social networks: the stochastic actor-oriented models (SAOMs), introduced by Snijders (1996).
+Unlike other network models, SAOMs are not very well understood.
+We use model visualization techniques introduced in Wickham et al (2015) in order to make them a little less murky.
+The SAOMs are a prime example of a set of models that can benefit greatly from application of model visualization.
+With the help of static and dynamic visualizations, we bring the hidden model fitting processes into the foreground, eventually leading to a better understanding and higher accessibility of stochastic actor-oriented models for social network analysts.
 \end{abstract}
@@
 The first statistic emphasizes triad relationships that are formed between actors from different covariate groups, while the other emphasizes indirect ties between actors.
-These two effects are visually represented and further described in in Figure~\ref{fig:structures}
+These two effects are visually represented and further described in Figure~\ref{fig:jtt} and Figure~\ref{fig:dad}, respectively.
 
@@
 This summary network represents 1,000 simulations of wave 2 using the values from the simple fitted model $M1$.',results='hide', fig.height=3>>=
+library(netvizinf)
 friendshipData <- array(c(fd2.w1, fd2.w2, fd2.w3), dim = c(16, 16, 3))

### 2017-08-24 paper.Rnw
add some transitions, add a table of model estimates

@@
 \begin{abstract}
-Social networks have been studied for decades, beginning with a few foundational works, including the 1967 study, "The Small World Problem" by Stanley Milgram.
+Social networks have been studied for decades, beginning with a few foundational works, including the 1967 study, ``The Small World Problem" by Stanley Milgram.
 In this paper, we concentrate on one type of model for dynamic social networks: the stochastic actor-oriented models (SAOMs), introduced by Snijders (1996).
@@
 In the CTMC, when actor $i$ may make a change, it chooses which tie $x_{i1, \dots, x_{in}}$ to change at random according to the probabilities $p_{ij}(\boldsymbol{\beta}, x)$.
+\st{The objective function and the resulting values of $p_{ij}$ are combined with the rate function, or in our case the rate parameter, to fully describe the CTMC that is used to model network change.}
 \hh{XXX This paragraph ends a bit abruptly - what is the future state probability used for?} \st{XX The future state probability is used within the continuous time markov chain to move through the different network states.
@@
 Thus, the rate of change between any two states, $x^b$ and $x^c$, that differ by only one tie $x_{ij}$, is the product of the rate at which actor $i$ gets to change a tie and the probability that the tie that will change is the tie to node $j$.
-
-XXX Need a transistion to next section XXX 
+Estimating the parameters in these models is difficult, but thanks to the SIENA software, we have an accessible way to fit SAOMs to network observations.
+
+XXX HH can you help? I need a way better transition than that XXX
 }
@@
 
-<<sienagof, fig.cap="Goodness of fit measures for two network statisticscomputed on the first simulated wave of a SAOM: indegree and outdegree cumulative distributions.
+<<sienagof, fig.cap="Goodness of fit measures for two network statistics computed on the first simulated wave of a SAOM: indegree and outdegree cumulative distributions.
 On the x-axis is the degree value, and on the y-axis is the number of times that degree value occurs in the network.
@@
 ggplot(data = seobama) + 
-  geom_net(directed = T, labelon=F, arrowsize = .25, singletons= F, fiteach = T, linewidth = .25, layout.alg = 'fruchtermanreingold',
+  geom_net(directed = T, labelon=F, arrowsize = .25, singletons= T, fiteach = T, linewidth = .25, layout.alg = 'fruchtermanreingold',
            aes(from_id = source, to_id = target, color = party),size = 1) + 
@@
 The networks we constructed for the four senates during President Obama's administration are shown in Figure~\ref{fig:senateplot}.
+In Section~\ref{sec:ModelVis}, we fit several stochastic actor-oriented models to these data sets use those models to guide our further exploration of SAOMs.
 }
+
 \section{Model Visualizations} \label{sec:ModelVis}
@@
 
+<<tableofmeans>>=
+simu2 <- read.csv("data/simulation-1000-M1-M2-M3.csv")
+senM1est <- read.csv("data/congress/senateM1ests.csv")
+senM2est <- read.csv("data/congress/senateM2ests.csv")
+senM3est <- read.csv("data/congress/senateM3ests.csv")
+senM1est$Model <- "M1"
+senM2est$Model <- "M2"
+senM3est$Model <- "M3"
+senM1est$Sim <- 1:nrow(senM1est)
+senM2est$Sim <- 1:nrow(senM2est)
+senM3est$Sim <- 1:nrow(senM3est)
+senM1est %>% gather(parameter, estimate, Rate:recip) %>% select(Sim, Model, parameter, estimate) -> senM1est2 
+senM2est %>% gather(parameter, estimate, Rate:jumpXTransTrip) %>% select(Sim, Model, parameter, estimate) -> senM2est2 
+senM3est %>% gather(parameter, estimate, Rate:nbrDist2twice) %>% select(Sim, Model, parameter, estimate) -> senM3est2 
+senateEsts <- rbind(senM1est2, senM2est2, senM3est2)
+senateEsts$parameter <- as.factor(senateEsts$parameter)
+levels(senateEsts$parameter) <- c("beta1", "beta3", "beta4", "alpha1", "alpha2", "alpha3", "beta2")
+SSE <-
+  senateEsts %>% group_by(Model, parameter) %>% 
+  summarize(meanEst = mean(estimate), sdEst = sd(estimate)) %>% 
+  mutate(val = paste0(round(meanEst,3), " ", "(", round(sdEst,3), ")")) %>% 
+  select(Model, parameter, val) %>% 
+  spread(Model, val) %>% arrange(as.character(parameter))
+SFE <- simu2 %>% group_by(Model, parameter) %>% 
+  summarize(meanEst = mean(estimate), sdEst = sd(estimate)) %>% 
+   mutate(val = paste0(round(meanEst,3), " ", "(", round(sdEst,3), ")")) %>% 
+  select(Model, parameter, val) %>% 
+  spread(Model, val) %>% arrange(parameter)
+@
+
+\begin{table}
+\begin{tabular}{l|ccc|ccc }
+  \hline
+  & \multicolumn{3}{|c|}{Friendship Data} & \multicolumn{3}{|c|}{Senate Data} \\
+  \hline
+  $\alpha_1$ & \Sexpr{SFE$M1[1]} & \Sexpr{SFE$M2[1]} & \Sexpr{SFE$M3[1]} & \Sexpr{SSE$M1[1]} & \Sexpr{SSE$M2[1]} & \Sexpr{SSE$M3[1]} \\
+  $\alpha_2$ & \Sexpr{SFE$M1[2]} & \Sexpr{SFE$M2[2]} & \Sexpr{SFE$M3[2]} & \Sexpr{SSE$M1[2]} & \Sexpr{SSE$M2[2]} & \Sexpr{SSE$M3[2]}  \\
+  $\alpha_3$ & -- & -- & -- & \Sexpr{SSE$M1[3]} & \Sexpr{SSE$M2[3]} & \Sexpr{SSE$M3[3]} \\
+  $\beta_1$ & \Sexpr{SFE$M1[3]} & \Sexpr{SFE$M2[3]} & \Sexpr{SFE$M3[3]} & \Sexpr{SSE$M1[4]} & \Sexpr{SSE$M2[4]} & \Sexpr{SSE$M3[4]} \\
+  $\beta_2$ & \Sexpr{SFE$M1[4]} & \Sexpr{SFE$M2[4]} & \Sexpr{SFE$M3[4]} & \Sexpr{SSE$M1[5]} & \Sexpr{SSE$M2[5]} & \Sexpr{SSE$M3[5]}\\
+$\beta_3$ & -- & \Sexpr{SFE$M2[5]} & -- & -- & \Sexpr{SSE$M2[6]} & -- \\
+  $\beta_4$ & -- & --  & \Sexpr{SFE$M3[6]} & -- & -- & \Sexpr{SSE$M3[7]} \\
+  \hline
+\end{tabular}
+\caption{\label{tab:meanests} The means (standard deviations) of parameter values estimated from repeated fittings of $M1, M2, M3$ to the small friendship network and the senate collaboration network.
+Each model was fit 1,000 times to the friend data, while each model was fit 100 times to the senate data.}
+\end{table}
+
 For comparison, we also fit models $M1$, $M2$, and $M3$ to the senate collaboration data.
-In addition, we fit two more models, $M4$ and $M5$ to the senate data that were found to be significant according to the Wald-type test:
+Comparison of the parameter estimates of these models fit to the two different data sets is given in Table~\ref{tab:meanests} In addition, we fit two more models, $M4$ and $M5$ to the senate data that were found to be significant according to the Wald-type test:
 
@@
 ggplot(data = clintonSenate) + 
-  geom_net(directed = T, labelon=F, arrowsize = .5, singletons= F, fiteach = T, arrowgap = .01, layout.alg = 'fruchtermanreingold',
+  geom_net(directed = T, labelon=F, arrowsize = .3, singletons= T, fiteach = T, arrowgap = .01, layout.alg = 'fruchtermanreingold',
            aes(from_id = source, to_id = target, color = party, linewidth = gsw, shape = sex,size = n_au)) + 
@@
 To demonstrate, we use \texttt{geomnet} to visualize the connections in the $111^{th}$ at two different points: when Hillary Clinton was in the senate, and after she left to become Secretary of State.
-Clinton was only in the $111^{th}$ senate for 17 days, from January 3, 2009, to January 20, 2009, in the middle of her second term in the senate.
+Clinton was only in the $111^{th}$ senate for 17 days, from January 3, 2009, to January 20, 2009, when she was in the middle of her second term in the senate.
 In that time, she authored two bills and was a cosponsor on 17 other bills.
@@
 We can compare the number of bills authored throughout the senate by mapping the size of the node to the numer of bills authored by that senator.
-We also map shape to sex, and keep color mapped to party of the senator.
+We also map shape to gender, and keep color mapped to party of the senator.
 In addition, we can see the strength of the tie by mapping the linewidth of the edge to the $WPC$ value between the two senators.
+In this single visualization, we have viewed node information (number of bills authored by a senator and the gender of that senator), edge infortmation (the direction and strength of ties between two senators), and time (before and after Clinton left the senate).
 % ST: I don't like this paragraph
@@
 actual2 <- actual2 %>% dplyr::select(from,to, count, weight)
-actual1$cat <- "First Observation"
-actual2$cat <- "Second Observation"
+actual1$cat <- "1st Wave"
+actual2$cat <- "2nd Wave"
 
@@
 
-p2 <- ggplot(data = filter(combinedavgs, cat == "First Observation")) +
+p2 <- ggplot(data = filter(combinedavgs, cat == "1st Wave")) +
   geom_net(aes(from_id = from, to_id = to, linewidth= linewidth/3),
@@
     theme(plot.background = element_rect(color = 'black'))
-p3 <- ggplot(data = filter(combinedavgs, cat == "Second Observation")) +
+p3 <- ggplot(data = filter(combinedavgs, cat == "2nd Wave")) +
   geom_net(aes(from_id = from, to_id = to, linewidth= linewidth/3),
@@
   ThemeNoNet + 
-  labs(x = "log(Statistic Value)", title = "Distribution of log Mahalanobis Distance", subtitle = "Data shown as vertical line")
+  labs(x = "log(Statistic Value)", title = "Distribution of log Mahalanobis Distance")
 
@@
 alt2nd <- read.csv("data/distribution_dblpairs_model.csv")
-simu2 <- read.csv("data/simulation-1000-M1-M2-M3.csv")
 means <- simu2 %>% group_by(Model, parameter) %>% summarize(
@@
 #simu2 <- read.csv("data/simulation-1000-M1-M2-M3.csv")
-
 simu_spread <- simu2 %>% spread(parameter, estimate)
-ggpairs(simu_spread, columns = 3:8, ggplot2::aes(colour=Model, alpha = .5)) + ThemeNoNet
+ggpairs(simu_spread, columns = 3:8, ggplot2::aes(colour=Model, alpha = .5)) + theme(plot.title = element_text(size = rel(1.5),
+                                            face = 'plain', angle = 0), 
+                  axis.title.x = element_text(size = rel(1.5),
+                                            face = 'plain', angle = 0, inherit.blank = T),
+                  axis.title.y = element_text(size = rel(1.5),
+                                            face = 'plain', angle = 90, inherit.blank = T),
+                  axis.text.x.top = element_text(size = rel(1.5),
+                                            face = 'plain', angle = 90, inherit.blank = TRUE), 
+                  axis.text.x.bottom = element_text(size = rel(1.5),
+                                            face = 'plain', angle = 90, inherit.blank = TRUE), 
+                  axis.text.y.left = element_text(size = rel(1.5),
+                                            face = 'plain', angle = 0, inherit.blank = T),
+                  axis.text.y.right = element_text(size = rel(1.5),
+                                            face = 'plain', angle = 0, inherit.blank = T),
+                  strip.text.x = element_text(size = rel(2),
+                                            face = 'plain', angle = 0, 
+                                            margin = margin(t = 3, r = 0, b = 3, l = 0, unit = "pt"), inherit.blank = T),
+                  strip.text.y =element_text(size = rel(2),
+                                            face = 'plain', angle = 90,margin = margin(t = 3, r = 3, b = 3, l = 3, unit = "pt"),inherit.blank = T),
+                  strip.background = element_rect(colour = "black", fill = "white")
+                    )
 @
@@
 <<averagenet, results='hide', out.width='99%',fig.cap='The three "average" networks in the middle, with the actual network observations oon either side.
-There is some difference between the three models, but overall, these three models cannot capture the structure in the true second wave of data.', fig.height=3>>=
+There is some difference between the three models, but overall, these three models cannot capture the structure in the true second wave of data.', fig.height=6>>=
 #friend.data.w1 <- as.matrix(read.table("s50_data/s50-network1.dat"))
@@
 # M3avg2$to <- paste0("V", M3avg2$to)
-combinedavgs$linewidth[combinedavgs$cat %in% c("First Observation", "Second Observation")] <- .5
+combinedavgs$linewidth[combinedavgs$cat %in% c("1st Wave", "2nd Wave")] <- .5
 ggplot(data = combinedavgs) + 
@@
   ThemeNet + theme(panel.background = element_rect(color = 'black')) + 
-  facet_wrap(~cat, nrow = 1)
+  facet_wrap(~cat, nrow = 2)
 

### 2017-08-24 paper.Rnw
remove some unnecessary comments

@@
 For comparison, we also fit models $M1$, $M2$, and $M3$ to the senate collaboration data.
-Comparison of the parameter estimates of these models fit to the two different data sets is given in Table~\ref{tab:meanests} In addition, we fit two more models, $M4$ and $M5$ to the senate data that were found to be significant according to the Wald-type test:
+Comparison of the parameter estimates of these models fit to the two different data sets is given in Table~\ref{tab:meanests}.
+In addition, we fit two more models, $M4$ and $M5$ to the senate data that were found to be significant according to the Wald-type test:
 
@@
 \end{align*}
-
-XXX TO DO - working on model estimates tonight XXX
 

### 2017-08-25 paper.Rnw
add correlation plot for senate models and histogram of ms. Need to add discussion and put them somewhere appropriate.

@@
 \hh{XXX we tried to estimate the rate based on the simulations - you could include a histogram of the number of changes for the 1000 simulations in the viz part XXX}
-\st{XXX I have the fitted models from the 1000 simulations (see Figure \ref{fig:distests}). Is that what you mean? XXX}
+\st{XXX see line 1257 for plot. where should I put it? }
 
@@
                     )
+@
+
+<<corplotsSenate>>=
+senM4est <- read.csv("data/congress/senateM4ests.csv")
+senM5est <- read.csv("data/congress/senateM5ests.csv")
+names(senM4est) <- c("alpha1", "alpha2", "alpha3", "beta1", "beta2", "beta5")
+names(senM5est) <- c("alpha1", "alpha2", "alpha3", "beta1", "beta2", "beta6")
+senM4est$Model <- "M4"
+senM5est$Model <- "M5"
+senM4est$Sim <- 1:nrow(senM4est)
+senM5est$Sim <- 1:nrow(senM5est)
+senM4est2 <- senM4est %>% gather(parameter, estimate, alpha1:beta5) 
+senM5est2 <- senM5est %>% gather(parameter, estimate, alpha1:beta6) 
+senM45est <- rbind(senM4est2, senM5est2)
+senM45est %>% spread(parameter, estimate)-> senM45est2
+ggpairs(senM45est2, columns = 3:9, ggplot2::aes(colour=Model, alpha = .5))
 @
@@
 
+# histogram of number of MS in 1000 reps
+# sfmsall %>% group_by(rep) %>% summarize(noMs = max(ms)) %>% ggplot() + geom_histogram(aes(x = noMs), binwidth = 5)
+
 sfmsall <- read_csv("data/smallfriendsmicrosteps.csv")

### 2017-08-28 paper.Rnw
could you add the senate csv files?

@@
 
-Some commonly used layout algorithms, such as  the Kamada-Kawai layout \citep{kamadakawai} and the Fruchterman-Reingold layout \citep{fruchterman-reingold}, were designed to mimic physical systems, drawing the graphs based on the ``forces" connecting them.
+Some commonly used layout algorithms, such as  the Kamada-Kawai layout \citep{kamadakawai} and the Fruchterman-Reingold layout \citep{fruchterman-reingold}, are designed to mimic physical systems, drawing the graphs based on the ``forces" connecting them.
 In these algorithms, the edges of the network act as springs pushing and pulling the nodes in a low dimensional (usually two-dimensional) space.
@@
 \st{XXX see line 1257 for plot. where should I put it? }
+\hh{XXX Put a reference to the plot here.}
 
@@
 The parameters are included when the researcher hypothesizes that they will model underlying mechanisms of network change.
-They hope to answer questions such as, ``How does the existing network structure influence change in the network?" \hh{XXX move the last colored section to the front of this paragraph.
+They hope to answer questions such as, ``How does the existing network structure influence change in the network?" %\hh{XXX move the last colored section to the front of this paragraph.
 That sets up the structure nicely, e.g.
@@
 I'm not sure which section is which at this point.
-(sorry again...)}
+(sorry again...)} \hh{I guess you addressed this comment very thoroughly :)}
 % and ``How do the behavior and characteristics of the nodes influence change in the network?" put this sentence below somewhere
@@
 
-\subsubsection{Continuous Time Markov Chain}
+\subsubsection{Continuous Time Markov Chain (CTMC)}
 \st{
-In the continuous-time Markov chain literature, see for instance \citet{ctmc}, the chains are characterized by their \textit{generator} or \textit{intensity} matrix $\mathbf{Q}$.
+In the continuous-time Markov chain literature, see for instance \citet{ctmc}, chains are characterized by their \textit{generator} or \textit{intensity} matrix $\mathbf{Q}$.
 This matrix describes the rate of change between two states of the CTMC process, and the rows of this matrix always add to zero.
-For directed networks with binary edge variables like the ones we will be working with, there are a very large number of possible states, $2^{n(n-1)}$: there are two possible states for an edge, and there are $n(n-1)$ edge relationships because self-ties are not allowed.
+For directed networks with binary edge variables like the ones we will be working with, there are a very large number of possible states, $2^{n(n-1)}$: there are two possible states for an edge, and there are $n(n-1)$ edge relationships (excluding self-ties).
 The intensity matrix for a CTMC in a SAOM is then a square matrix of dimension $2^{n(n-1)} \times 2^{n(n-1)}$.
-Since only one tie changes at a time in the CTMC, there are only $n(n-1)$ reachable states from the current network state.
+Only one tie changes at a time in the CTMC, resulting in $n(n-1)$ reachable states from the current network state.
 Thus, the intensity matrix $\mathbf{Q}$ is very sparse, with only $n(n-1) + 1$ non-zero entries in each row.
-Note that $n(n-1)$ of these represent the possible states that are one edge different from a given state, and the additional non-zero entry is for the state to remain unchanged.
-All other entries in a row are zero because those network states cannot be reached from the current state by just one change.
+Note that $n(n-1)$ of these entries represent the possible states that are one edge different from a given state, while the additional non-zero entry is for the state to remain unchanged.
+All other entries in a row are structural zeroes because those network states cannot be reached from the current state in a single change.
 
@@
 Thus, the rate of change between any two states, $x^b$ and $x^c$, that differ by only one tie $x_{ij}$, is the product of the rate at which actor $i$ gets to change a tie and the probability that the tie that will change is the tie to node $j$.
-Estimating the parameters in these models is difficult, but thanks to the SIENA software, we have an accessible way to fit SAOMs to network observations.
+Estimating the parameters in these models is difficult, but thanks to the SIENA software \citep(Rsiena), we have an accessible way to fit SAOMs to network observations.
 

### 2017-08-28 paper.Rnw
dplyr::select

@@
 senM3est$Sim <- 1:nrow(senM3est)
-senM1est %>% gather(parameter, estimate, Rate:recip) %>% select(Sim, Model, parameter, estimate) -> senM1est2 
-senM2est %>% gather(parameter, estimate, Rate:jumpXTransTrip) %>% select(Sim, Model, parameter, estimate) -> senM2est2 
-senM3est %>% gather(parameter, estimate, Rate:nbrDist2twice) %>% select(Sim, Model, parameter, estimate) -> senM3est2 
+senM1est %>% gather(parameter, estimate, Rate:recip) %>% dplyr::select(Sim, Model, parameter, estimate) -> senM1est2 
+senM2est %>% gather(parameter, estimate, Rate:jumpXTransTrip) %>% dplyr::select(Sim, Model, parameter, estimate) -> senM2est2 
+senM3est %>% gather(parameter, estimate, Rate:nbrDist2twice) %>% dplyr::select(Sim, Model, parameter, estimate) -> senM3est2 
 senateEsts <- rbind(senM1est2, senM2est2, senM3est2)
@@
   mutate(val = paste0(round(meanEst,3), " ", "(", round(sdEst,3), ")")) %>% 
-  select(Model, parameter, val) %>% 
+  dplyr::select(Model, parameter, val) %>% 
   spread(Model, val) %>% arrange(as.character(parameter))
@@
    mutate(val = paste0(round(meanEst,3), " ", "(", round(sdEst,3), ")")) %>% 
-  select(Model, parameter, val) %>% 
+  dplyr::select(Model, parameter, val) %>% 
   spread(Model, val) %>% arrange(parameter)

### 2017-08-30 paper.Rnw
just about done with changes from meeting

--- 2017-08-28 paper.Rnw
@@
 
-<<logweibull, fig.align='center', out.width = '.5\\linewidth', fig.cap='The probability distribution function for the type 1 extreme value distribution, also known as the log-weibull or Gumbel distribution with location parameter $ \\mu = 0$ and scale parameter $ \\sigma = 0$.'>>=
+<<logweibull, fig.align='center', out.width = '.99\\linewidth', fig.height = 2, fig.cap='The probability distribution function for the type 1 extreme value distribution, also known as the log-weibull or Gumbel distribution with location parameter $ \\mu = 0$ and scale parameter $ \\sigma = 0$.'>>=
 library(reliaR)
@@
         label = "g(u) = exp(-(u + exp(-u)))",size=5)  + 
-  labs(x = "u", y = "g(u) (density)")
+  labs(x = "u", y = "density of g(u)")
 @
@@
 
-$$p_{ij}(\boldsymbol{\beta}, x) = \dfrac{\exp\left\{f_i(\boldsymbol{\beta}, x(i\leadsto j))\right\}}{\sum_{h \neq i} \exp\left\{f_i(\boldsymbol{\beta}, x(i \leadsto h))\right\}}$$
+\begin{equation}\label{eq:pij}
+p_{ij}(\boldsymbol{\beta}, x) = \dfrac{\exp\left\{f_i(\boldsymbol{\beta}, x(i\leadsto j))\right\}}{\sum_{h \neq i} \exp\left\{f_i(\boldsymbol{\beta}, x(i \leadsto h))\right\}}
+\end{equation}
 
@@
 gof2 <- sienaGOF(ansnull, IndegreeDistribution, varName ="friendship", join = FALSE)
-p1 <- plot(gof1, period = 1)
-p2 <- plot(gof2, period = 1)
+
+sims1 <- data.frame(gof1$`Period 1`$Simulations)
+dat1 <- gof1$`Period 1`$Observations
+dat1 %>% data.frame() %>% gather(outdegree, val) %>% 
+  mutate(outdegree = parse_number(outdegree)-1) -> dat1
+p1 <- sims1 %>% 
+  mutate(sim = row_number()) %>% 
+  gather(outdegree, val, X1:X9) %>% 
+  mutate(outdegree = parse_number(outdegree) - 1) %>% 
+  filter(outdegree <= 6) %>%  
+  ggplot(aes(x = outdegree, y = val)) + 
+  geom_boxplot(aes(group = outdegree), size = .5, outlier.shape = "x", outlier.size = 5) + 
+  geom_violin(aes(group = outdegree), bw = 2, fill = NA) + 
+  geom_point(data = dat1, color = 'red') + 
+  geom_line(data= dat1, color = 'red') + 
+  geom_text(data = dat1, aes(label = val), hjust = -.5) + 
+  labs(x = "Outdegree (p = 0.154)", y = "Statistic", 
+       title = "Goodness-of-Fit: Outdegree distribution period 1") + 
+  ThemeNoNet
+
+sims2 <- data.frame(gof2$`Period 1`$Simulations)
+dat2 <- gof2$`Period 1`$Observations
+dat2 %>% data.frame() %>% gather(indegree, val) %>% 
+  mutate(indegree = parse_number(indegree)-1) -> dat2
+p2 <- sims2 %>% 
+  mutate(sim = row_number()) %>% 
+  gather(indegree, val, X1:X9) %>% 
+  mutate(indegree = parse_number(indegree) - 1) %>% 
+  filter(indegree <= 6) %>%  
+  ggplot(aes(x = indegree, y = val)) + 
+  geom_boxplot(aes(group = indegree), outlier.shape = "x", outlier.size = 5) + 
+  geom_violin(aes(group = indegree), bw = 2, fill = NA) + 
+  geom_point(data = dat2, color = 'red') + 
+  geom_line(data= dat2, color = 'red') + 
+  geom_text(data = dat2, aes(label = val), hjust = -.5) + 
+  labs(x = "Indegree (p = 0.063)", y = "Statistic", 
+       title = "Goodness-of-Fit: Indegree distribution period 1") + 
+  ThemeNoNet
+
 gridExtra::grid.arrange(p1, p2)
@@
 
-<<alldataviz, fig.cap="The smaller friendship network data we will be modelling throughout the paper.", fig.width=8, fig.height=3>>=
+<<alldataviz, fig.cap="The smaller friendship network data we will be modelling throughout the paper.", out.width='.99\\linewidth', fig.height=3>>=
 actual1 <- merge(data.frame(as.edgelist(as.network(fd2.w1))), 
@@
        aes(from_id = X1, to_id = X2, color = as.ordered(drink), group = as.ordered(drink))) +
-  geom_net(fiteach = FALSE, directed=TRUE, layout.alg = 'fruchtermanreingold', size = 4, arrowsize = .25, arrowgap = .03, ecolour = 'grey40', labelon=T, labelcolour = 'black', fontsize = 4, vjust = .5) +
-  scale_color_brewer(palette = "YlOrRd", name = "Drinking\nBehavior") + 
+  geom_net(fiteach = FALSE, directed=TRUE, layout.alg = 'fruchtermanreingold', size = 5, arrowsize = .25, arrowgap = .03, ecolour = 'grey40', labelon=T, labelcolour = 'black', fontsize = 4, vjust = .5) +
+  scale_color_brewer(palette = "YlOrRd", name = "Drinking\nBehavior", labels = c("never", "1-2 times per year", "once per month", "once per week")) + 
   ThemeNet +
@@
   summarize(meanEst = mean(estimate), sdEst = sd(estimate)) %>% 
-  mutate(val = paste0(round(meanEst,3), " ", "(", round(sdEst,3), ")")) %>% 
+  mutate(val = paste0(sprintf("%.3f",meanEst), " ", "(", sprintf("%.3f", sdEst), ")")) %>% 
   select(Model, parameter, val) %>% 
@@
   summarize(meanEst = mean(estimate), sdEst = sd(estimate)) %>% 
-   mutate(val = paste0(round(meanEst,3), " ", "(", round(sdEst,3), ")")) %>% 
+   mutate(val = paste0(sprintf("%.3f",meanEst), " ", "(", sprintf("%.3f", sdEst), ")")) %>% 
   select(Model, parameter, val) %>% 
@@
   \hline
+  M1 & M2 & M3 & M1 & M2 & M3 \\
   $\alpha_1$ & \Sexpr{SFE$M1[1]} & \Sexpr{SFE$M2[1]} & \Sexpr{SFE$M3[1]} & \Sexpr{SSE$M1[1]} & \Sexpr{SSE$M2[1]} & \Sexpr{SSE$M3[1]} \\
@@
   ThemeNet + 
+  scale_shape_manual(values = c(17,16)) + 
   scale_color_manual(values = c("royalblue", "forestgreen","firebrick")) + 
@@
 <<distests, fig.cap="Distribution of fitted parameter values for our three SAOMs.
-The inclusion of $\\beta_3$ or $\\beta_4$ clearly has an effect on the distribution of the rate parameters, $\\alpha_1$ and $\\alpha_2$.", fig.width=8>>=
+The inclusion of $\\beta_3$ or $\\beta_4$ clearly has an effect on the distribution of the rate parameters, $\\alpha_1$ and $\\alpha_2$.", fig.height=4>>=
 
@@
   facet_wrap(~parameter, scales = 'free') + 
-  ThemeNoNet
+  ThemeNoNet + 
+  theme(legend.position = 'bottom')
 @
@@
 
-<<corplotsSenate>>=
+<<corplotsSenate, eval = FALSE>>=
 senM4est <- read.csv("data/congress/senateM4ests.csv")
@@
 <<makinggif, fig.cap='A selection of images in the microstep animation.
-The selected edges and nodes are emphasized by changing the size and the color, then when edges are removed, they fade out, shrinking in size, while the nodes change color and shrink to blend in with the other nodes.', fig.width=8, cache=TRUE, message = FALSE, warning = FALSE>>=
+The selected edges and nodes are emphasized by changing the size and the color, then when edges are removed, they fade out, shrinking in size, while the nodes change color and shrink to blend in with the other nodes.', cache=TRUE, message = FALSE, warning = FALSE, fig.height=3>>=
 # Need old version of dplyr to get this to work: 

### 2017-08-30 paper.Rnw

@@
   mutate(val = paste0(sprintf("%.3f",meanEst), " ", "(", sprintf("%.3f", sdEst), ")")) %>% 
-  select(Model, parameter, val) %>% 
+  dplyr::select(Model, parameter, val) %>% 
   spread(Model, val) %>% arrange(as.character(parameter))
@@
   select(Model, parameter, val) %>% 
+  dplyr::select(Model, parameter, val) %>% 
   spread(Model, val) %>% arrange(parameter)

### 2017-08-31 paper.Rnw
almost done :)

@@
 seobama <- read_csv("data/congress/senateobamapres_gsw_25.csv")
+to_label <- c("Joseph R. Biden Jr.", "John S. McCain", "Ted Cruz","Marco Rubio",
+  "Lindsey O. Graham","Rand Paul", "Bernard Sanders", "Jim Webb", "Mitch McConnell",
+  "Harry M. Reid", "Hillary Rodham Clinton")
+seobama$label <- as.factor(ifelse(seobama$source %in% to_label, seobama$source, ""))
+levels(seobama$label) <- c("", "Bernie Sanders", "Harry Reid", "Hillary Clinton", 
+                           "Jim Webb", "John McCain", "Joe Biden", "Lindsey Graham",
+                           "Marco Rubio", "Mitch McConnel", "Rand Paul", "Ted Cruz")
+set.seed(56049382)
 ggplot(data = seobama) + 
-  geom_net(directed = T, labelon=F, arrowsize = .25, singletons= T, fiteach = T, linewidth = .25, layout.alg = 'fruchtermanreingold',
-           aes(from_id = source, to_id = target, color = party),size = 1) + 
+  geom_net(directed = T, labelon=T, arrowsize = .25, singletons= T, fiteach = T, linewidth = .25, layout.alg = 'fruchtermanreingold', fontsize = 2, repel=T,
+           aes(from_id = source, to_id = target, color = party, label = label),
+           size = 1) + 
   ThemeNet + 
@@
   theme(legend.position = 'bottom') + 
-  facet_wrap(~senate, nrow = 1)
+  facet_wrap(~senate, nrow = 2, labeller = "label_both") + 
+  xlim(c(0,1.05)) + 
+  ylim(c(0,1.05))
+
 @

### 2017-09-13 paper.Rnw
push uncommitted changes

@@
   "Lindsey O. Graham","Rand Paul", "Bernard Sanders", "Jim Webb", "Mitch McConnell",
-  "Harry M. Reid", "Hillary Rodham Clinton")
+  "Harry M. Reid", "Hillary Rodham Clinton", "Amy Jean Klobuchar")
 seobama$label <- as.factor(ifelse(seobama$source %in% to_label, seobama$source, ""))
-levels(seobama$label) <- c("", "Bernie Sanders", "Harry Reid", "Hillary Clinton", 
-                           "Jim Webb", "John McCain", "Joe Biden", "Lindsey Graham",
-                           "Marco Rubio", "Mitch McConnel", "Rand Paul", "Ted Cruz")
+levels(seobama$label) <- c("", "Amy Klobuchar", "Bernie Sanders", "Harry Reid",
+                           "Hillary Clinton", "Jim Webb", "John McCain", "Joe Biden",
+                           "Lindsey Graham", "Marco Rubio", "Mitch McConnel", 
+                           "Rand Paul", "Ted Cruz")
 set.seed(56049382)

### 2017-09-15 paper.Rnw
updates

--- 2017-08-30 paper.Rnw
@@
 require(knitr)
-opts_chunk$set(fig.path='figure/', cache.path='cache/', fig.align='center', fig.pos='h', out.width='.99\\linewidth', par=TRUE, cache=FALSE, concordance=TRUE, autodep=TRUE, message=F, warning=F, echo = FALSE)#, root.dir = "~/Desktop/Dissertation/SAOM-removing-blindfold/")
+opts_chunk$set(fig.path='figure/', cache.path='cache/', fig.align='center', fig.pos='h', out.width='.99\\textwidth', par=TRUE, cache=FALSE, concordance=TRUE, autodep=TRUE, message=F, warning=F, echo = FALSE, dev="cairo_pdf", fig.width = 6, fig.height = 6)#, root.dir = "~/Desktop/Dissertation/SAOM-removing-blindfold/")
 @
@@
 library(cowplot)
+library(extrafont)
+loadfonts(quiet = T)
 
 ThemeNoNet <- theme_bw() %+replace% 
-            theme(plot.title = element_text(size = rel(1.5),
-                                            face = 'plain', angle = 0), 
-                  axis.title.x = element_text(size = rel(1.5),
-                                            face = 'plain', angle = 0),
-                  axis.title.y = element_text(size = rel(1.5),
-                                            face = 'plain', angle = 90),
-                  axis.text.x.top = element_text(size = rel(1.5),
-                                            face = 'plain', angle = 0), 
-                  axis.text.x.bottom = element_text(size = rel(1.5),
-                                            face = 'plain', angle = 0), 
-                  axis.text.y.left = element_text(size = rel(1.5),
-                                            face = 'plain', angle = 0),
-                  axis.text.y.right = element_text(size = rel(1.5),
-                                            face = 'plain', angle = 0),
-                  strip.text.x = element_text(size = rel(2),
-                                            face = 'plain', angle = 0, 
+            theme(plot.title = element_text(size = 12,
+                                            face = 'plain', 
+                                            angle = 0, 
+                                            family="Times New Roman"), 
+                  axis.title.x = element_text(size = 12,
+                                            face = 'plain', 
+                                            angle = 0,
+                                            family="Times New Roman"),
+                  axis.title.y = element_text(size = 12,
+                                            face = 'plain', 
+                                            angle = 90,
+                                            family="Times New Roman"),
+                  axis.text.x.top = element_text(size = 12,
+                                            face = 'plain', 
+                                            angle = 0,
+                                            family="Times New Roman"), 
+                  axis.text.x.bottom = element_text(size = 12,
+                                            face = 'plain', 
+                                            angle = 0,
+                                            family="Times New Roman"), 
+                  axis.text.y.left = element_text(size = 12,
+                                            face = 'plain', 
+                                            angle = 0,
+                                            family="Times New Roman"),
+                  axis.text.y.right = element_text(size = 12,
+                                            face = 'plain', 
+                                            angle = 0,
+                                            family="Times New Roman"),
+                  strip.text.x = element_text(size = 12,
+                                            face = 'plain', 
+                                            angle = 0, 
+                                            family="Times New Roman",
                                             margin = margin(t = 3, r = 0, b = 3, l = 0, unit = "pt")),
-                  strip.text.y =element_text(size = rel(2),
-                                            face = 'plain', angle = 90,margin = margin(t = 3, r = 3, b = 3, l = 3, unit = "pt")),
+                  strip.text.y =element_text(size = 12,
+                                            face = 'plain', 
+                                            angle = 90,
+                                            family="Times New Roman",
+                                            margin = margin(t = 3, r = 3, b = 3, l = 3, unit = "pt")),
                   strip.background = element_rect(colour = "black", fill = "white")
@@
 ThemeNet <- theme_net() %+replace% 
-            theme(plot.title = element_text(size = rel(2),
-                                            face = 'plain', angle = 0), 
-                  strip.text.x = element_text(size = rel(2),
-                                            face = 'plain', angle = 0, 
+            theme(plot.title = element_text(size = 12,
+                                            face = 'plain', angle = 0, family = "Times New Roman"), 
+                  strip.text.x = element_text(size = 12,
+                                            face = 'plain', angle = 0, family = "Times New Roman", 
                                             margin = margin(t = 3, r = 0, b = 3, l = 0, unit = "pt")),
-                  strip.text.y =element_text(size = rel(2),
-                                            face = 'plain', angle = 0,margin = margin(t = 3, r = 0, b = 3, l = 0, unit = "pt")),
+                  strip.text.y =element_text(size = 12,
+                                            face = 'plain', angle = 0, family = "Times New Roman",margin = margin(t = 3, r = 0, b = 3, l = 0, unit = "pt")),
                   strip.background = element_rect(colour = "black", fill = "white"))
@@
 Because of this,
-SAOMs are often fit to data using a series of Markov chain Monte Carlo (MCMC) phases for finding method of moments estimators.
+SAOMs are often fit to data using a series of Markov Chain Monte Carlo (MCMC) phases for finding method of moments estimators.
 In order to estimate the parameters of SAOMs, we use the software SIENA, and its \texttt{R} implementation \texttt{RSiena}, which was developed by \citet{RSiena}.
@@
 %Here, we explore RSiena and SAOMs with innovative network visualizations, both static and dynamic.
-By bringing some light to the underlying methods and structures that are ``behind the scenes" of when fitting SAOMs to network data, we aim to help researchers working with the models better understand the implications and analyses of these models.
+By bringing some light to the underlying methods and structures that are behind the scenes when fitting SAOMs to network data, we aim to help researchers working with the models better understand the implications and analyses of these models.
 }
 
-\par The guiding principles of model visualization introduced in \citet{modelvis} are: 
-\begin{enumerate}
-\item Display the model in the data space.
-\item Collections of models are more informative than singletons. 
-\item Explore the fitting \textit{process} rather than just looking at the final result.
-\end{enumerate}
-Stochastic actor-oriented models are a prime example of a set of models that can benefit greatly from application of model visualization.
+\par We get into the SAOM black box by using model visualization (see \citet{modelvis}) to display the model in the data space, view collections of models instead of single models, and visually explore the process of fitting the SAOMs as opposed to looking at only the final output.
+Stochastic actor-oriented models are a prime example of a set of models that can benefit greatly from the application of model visualization.
 For instance, the models themselves include a continuous-time Markov chain (CTMC) that is completely hidden from the analyst in the model fitting process.
@@
 <<toyplots, fig.show='hold', fig.cap="On the left, a node-link diagram of our directed toy network, with nodes placed using the Kamada-Kawai algorithm.
-On the right, the adjacency matrix visualization for that same network.", out.width='.49\\linewidth'>>=
+On the right, the adjacency matrix visualization for that same network.", out.width='.49\\textwidth'>>=
 toynetdat <- fortify(as.edgedf(toynet$edges), toynet$nodes)
@@
   ThemeNet +
+  theme(plot.title = element_text(size = 20)) + 
   labs(title = "Node-Link Diagram")
+#ggsave("figure/toynetnld.pdf",plot = nld, width = 4, height = 4)
 
@@
   ThemeNoNet + 
-  theme(panel.grid = element_blank(), aspect.ratio = 1) +
-  labs(title = "Adjacency Matrix Visualization", x = "to (alter)", y = "from (ego)")
+  theme(panel.grid = element_blank(), aspect.ratio = 1, 
+        plot.title = element_text(size = 20), 
+        axis.title.x.top = element_text(size = 18),
+        axis.title.y.left = element_text(size = 18), 
+        axis.text.x.top  = element_text(size = 18),
+        axis.text.y.left  = element_text(size = 18)) +
+  labs(title = "Adjacency Matrix Visualization", x = "\nto (alter)", y = "from (ego)\n")
 @
@@
 For the network $x$ and each actor $i$ in the network, the number of times that an actor $i$ gets to change its ties, $x_{ij}$, to other nodes $j \neq i$ in the network is dictated by a \emph{rate function} $\rho(x, \mathbf{z}, \boldsymbol{\alpha})$, where $\boldsymbol{\alpha}$ are the parameters in the function $\rho$, and $x$ is the current network state, with covariates of interest $\mathbf{z}$.
-For this paper, we assume a simple rate \st{``function"}, $\rho(x, \mathbf{z}, \boldsymbol{\alpha}) = \alpha_m$ that is constant across all actors between observations at time $t_m$ to $t_{m+1}$, \st{thus our rate function is just a rate parameter in the overall model}.
+For this paper, we assume a simple rate function, $\rho(x, \mathbf{z}, \boldsymbol{\alpha}) = \alpha_m$ that is constant across all actors between observations at time $t_m$ to $t_{m+1}$, \st{thus our rate function is just a rate parameter in the overall model}.
 In general, SAOMs can incorporate covariate values and network statistics into the model, so that each node will have a different rate of change.
@@
 
-\hh{XXX we tried to estimate the rate based on the simulations - you could include a histogram of the number of changes for the 1000 simulations in the viz part XXX}
-\st{XXX see line 1257 for plot. where should I put it? }
-\hh{XXX Put a reference to the plot here.}
+% \hh{XXX we tried to estimate the rate based on the simulations - you could include a histogram of the number of changes for the 1000 simulations in the viz part XXX}
+% \st{XXX see line 1257 for plot. where should I put it? }
+% \hh{XXX Put a reference to the plot here.}
 
@@
 
-<<logweibull, fig.align='center', out.width = '.99\\linewidth', fig.height = 2, fig.cap='The probability distribution function for the type 1 extreme value distribution, also known as the log-weibull or Gumbel distribution with location parameter $ \\mu = 0$ and scale parameter $ \\sigma = 0$.'>>=
+<<logweibull, fig.align='center', out.width = '.5\\textwidth', fig.height = 4, fig.cap='The probability distribution function for the type 1 extreme value distribution, also known as the log-weibull or Gumbel distribution with location parameter $ \\mu = 0$ and scale parameter $ \\sigma = 0$.'>>=
 library(reliaR)
@@
   annotate('text', x = 6, y = .25, 
-        label = "g(u) = exp(-(u + exp(-u)))",size=5)  + 
+        label = "g(u) = exp(-(u + exp(-u)))",size=5, family = "Times New Roman")  + 
+  theme(text = element_text(size = 20, family = "Times New Roman")) + 
   labs(x = "u", y = "density of g(u)")
@@
 In the CTMC, when actor $i$ may make a change, it chooses which tie $x_{i1, \dots, x_{in}}$ to change at random according to the probabilities $p_{ij}(\boldsymbol{\beta}, x)$.
-\st{The objective function and the resulting values of $p_{ij}$ are combined with the rate function, or in our case the rate parameter, to fully describe the CTMC that is used to model network change.}
+\st{The objective function and the resulting values of $p_{ij}$ are combined with the rate function to fully describe the CTMC that is used to model network change in a SAOM.}
+
 \hh{XXX This paragraph ends a bit abruptly - what is the future state probability used for?} \st{XX The future state probability is used within the continuous time markov chain to move through the different network states.
@@
 Thus, the rate of change between any two states, $x^b$ and $x^c$, that differ by only one tie $x_{ij}$, is the product of the rate at which actor $i$ gets to change a tie and the probability that the tie that will change is the tie to node $j$.
-Estimating the parameters in these models is difficult, but thanks to the SIENA software \citep(Rsiena), we have an accessible way to fit SAOMs to network observations.
+This matrix $\mathbf{Q}$ is the foundation for estimation of a SAOM.
+Estimating the parameters in these models is difficult, but thanks to the SIENA software \citep{RSiena}, we have an accessible way to fit SAOMs to network observations.
 
@@
 
-\subsection{Model Goodness-of-Fit}
+\subsection{Model Goodness-of-Fit}\label{sec:modgof}
 
@@
 
-<<sienagof, fig.cap="Goodness of fit measures for two network statistics computed on the first simulated wave of a SAOM: indegree and outdegree cumulative distributions.
+% <<sienagof, fig.cap="Goodness of fit measures for two network statistics computed on the first simulated wave of a SAOM: indegree and outdegree cumulative distributions.
 On the x-axis is the degree value, and on the y-axis is the number of times that degree value occurs in the network.
 \\st{The p-value shown on the x-axis is the proportion of observations simulated from the model that have observed statistic values as large or larger than the values observed in the data.}", cache = TRUE, message = FALSE, warning = FALSE, fig.height=5>>=
-load("data/ansnullpaper.rda")
-library(lattice)
-gof1 <- sienaGOF(ansnull, OutdegreeDistribution, varName ="friendship", join = FALSE)
-gof2 <- sienaGOF(ansnull, IndegreeDistribution, varName ="friendship", join = FALSE)
-
-sims1 <- data.frame(gof1$`Period 1`$Simulations)
-dat1 <- gof1$`Period 1`$Observations
-dat1 %>% data.frame() %>% gather(outdegree, val) %>% 
-  mutate(outdegree = parse_number(outdegree)-1) -> dat1
-p1 <- sims1 %>% 
-  mutate(sim = row_number()) %>% 
-  gather(outdegree, val, X1:X9) %>% 
-  mutate(outdegree = parse_number(outdegree) - 1) %>% 
-  filter(outdegree <= 6) %>%  
-  ggplot(aes(x = outdegree, y = val)) + 
-  geom_boxplot(aes(group = outdegree), size = .5, outlier.shape = "x", outlier.size = 5) + 
-  geom_violin(aes(group = outdegree), bw = 2, fill = NA) + 
-  geom_point(data = dat1, color = 'red') + 
-  geom_line(data= dat1, color = 'red') + 
-  geom_text(data = dat1, aes(label = val), hjust = -.5) + 
-  labs(x = "Outdegree (p = 0.154)", y = "Statistic", 
-       title = "Goodness-of-Fit: Outdegree distribution period 1") + 
-  ThemeNoNet
-
-sims2 <- data.frame(gof2$`Period 1`$Simulations)
-dat2 <- gof2$`Period 1`$Observations
-dat2 %>% data.frame() %>% gather(indegree, val) %>% 
-  mutate(indegree = parse_number(indegree)-1) -> dat2
-p2 <- sims2 %>% 
-  mutate(sim = row_number()) %>% 
-  gather(indegree, val, X1:X9) %>% 
-  mutate(indegree = parse_number(indegree) - 1) %>% 
-  filter(indegree <= 6) %>%  
-  ggplot(aes(x = indegree, y = val)) + 
-  geom_boxplot(aes(group = indegree), outlier.shape = "x", outlier.size = 5) + 
-  geom_violin(aes(group = indegree), bw = 2, fill = NA) + 
-  geom_point(data = dat2, color = 'red') + 
-  geom_line(data= dat2, color = 'red') + 
-  geom_text(data = dat2, aes(label = val), hjust = -.5) + 
-  labs(x = "Indegree (p = 0.063)", y = "Statistic", 
-       title = "Goodness-of-Fit: Indegree distribution period 1") + 
-  ThemeNoNet
-
-gridExtra::grid.arrange(p1, p2)
-# sum(gof1$`Period 1`$SimulatedTestStat >= gof1$`Period 1`$ObservedTestStat)/1000
-# [1] 0.154
-# > sum(gof2$`Period 1`$SimulatedTestStat >= gof2$`Period 1`$ObservedTestStat)/1000
-# [1] 0.063
-
-@
+% load("data/ansnullpaper.rda")
+% library(lattice)
+% gof1 <- sienaGOF(ansnull, OutdegreeDistribution, varName ="friendship", join = FALSE)
+% gof2 <- sienaGOF(ansnull, IndegreeDistribution, varName ="friendship", join = FALSE)
+% 
+% sims1 <- data.frame(gof1$`Period 1`$Simulations)
+% dat1 <- gof1$`Period 1`$Observations
+% dat1 %>% data.frame() %>% gather(outdegree, val) %>% 
+%   mutate(outdegree = parse_number(outdegree)-1) -> dat1
+% p1 <- sims1 %>% 
+%   mutate(sim = row_number()) %>% 
+%   gather(outdegree, val, X1:X9) %>% 
+%   mutate(outdegree = parse_number(outdegree) - 1) %>% 
+%   filter(outdegree <= 6) %>%  
+%   ggplot(aes(x = outdegree, y = val)) + 
+%   geom_boxplot(aes(group = outdegree), size = .5, outlier.shape = "x", outlier.size = 5) + 
+%   geom_violin(aes(group = outdegree), bw = 2, fill = NA) + 
+%   geom_point(data = dat1, color = 'red') + 
+%   geom_line(data= dat1, color = 'red') + 
+%   geom_text(data = dat1, aes(label = val), hjust = -.5) + 
+%   labs(x = "Outdegree (p = 0.154)", y = "Statistic", 
+%        title = "Goodness-of-Fit: Outdegree distribution period 1") + 
+%   ThemeNoNet
+% 
+% sims2 <- data.frame(gof2$`Period 1`$Simulations)
+% dat2 <- gof2$`Period 1`$Observations
+% dat2 %>% data.frame() %>% gather(indegree, val) %>% 
+%   mutate(indegree = parse_number(indegree)-1) -> dat2
+% p2 <- sims2 %>% 
+%   mutate(sim = row_number()) %>% 
+%   gather(indegree, val, X1:X9) %>% 
+%   mutate(indegree = parse_number(indegree) - 1) %>% 
+%   filter(indegree <= 6) %>%  
+%   ggplot(aes(x = indegree, y = val)) + 
+%   geom_boxplot(aes(group = indegree), outlier.shape = "x", outlier.size = 5) + 
+%   geom_violin(aes(group = indegree), bw = 2, fill = NA) + 
+%   geom_point(data = dat2, color = 'red') + 
+%   geom_line(data= dat2, color = 'red') + 
+%   geom_text(data = dat2, aes(label = val), hjust = -.5) + 
+%   labs(x = "Indegree (p = 0.063)", y = "Statistic", 
+%        title = "Goodness-of-Fit: Indegree distribution period 1") + 
+%   ThemeNoNet
+% 
+% 
+% gridExtra::grid.arrange(p1, p2)
+% # sum(gof1$`Period 1`$SimulatedTestStat >= gof1$`Period 1`$ObservedTestStat)/1000
+% # [1] 0.154
+% # > sum(gof2$`Period 1`$SimulatedTestStat >= gof2$`Period 1`$ObservedTestStat)/1000
+% # [1] 0.063
+% 
+% @
 
 The \texttt{plot.sienaGOF()} function allows us to visualize this fit.
-In Figure~\ref{fig:sienagof}, we provide an example of what the goodness-of-fit plot output looks like for indegree and outdegree statistics for a small model.
-Box plots and violin plots are drawn on top of each other in the figure in order to show the distribution of the simulated network observations compared to the true data shown in red points connected by red lines.
+%In Figure~\ref{fig:sienagof}, we provide an example of what the goodness-of-fit plot output looks like for indegree and outdegree statistics for a small model.
+This function draws a box plots and a violin plot at each value of the statistic of interest observed in the simulations.
+In order to compare the distribution of the simulated network to the true data, red points connected by red lines representing the observed values are superimposed on the boxplots.
 If the red points lie "well within" the simulated values, the model is a good fit to the data.
-The model examined in Figure~\ref{fig:sienagof} appears to do a better job of capturing the outdegree distribution than the indegree distribution of the data.
+The plot is not intuitive for understanding network data.
+%The model examined in Figure~\ref{fig:sienagof} appears to do a better job of capturing the outdegree distribution than the indegree distribution of the data.
+In Section~\ref{sec:minds}, we propose a new way of visualizing goodness-of-fit that uses the traditional node-link diagram to visualize the entire network instead of numerical summaries of the the network.
 
@@
 <<adjmatviz, fig.align='center', fig.cap="A visualization of the adjacency matrices of the three waves of network observations in the ``Teenage Friends and Lifestyle Study'' data.
-The subset we will be using is outlined in red.", fig.width=8, results='hide',fig.show='asis', fig.width=8, fig.height=3>>=
+The subset we will be using is outlined in red.", results='hide',fig.show='asis', fig.width=8, fig.height=3>>=
 source("code/00e_small_friends.R")
@@
   facet_wrap(~wave) + theme_bw() +
-  labs(x = "j", y = "i") + 
+  labs(x = "node j (alter)", y = "node i (ego)") + 
   ThemeNoNet + 
@@
 
-<<alldataviz, fig.cap="The smaller friendship network data we will be modelling throughout the paper.", out.width='.99\\linewidth', fig.height=3>>=
+<<alldataviz, fig.cap="The smaller friendship network data we will be modelling throughout the paper.
+In the first wave, we can see there are two large, separate friend group.
+By the second wave, three students with heavier drinking behavior have separated from their group, while the others become members of the other group.
+By the third wave, the large group has almost completely broken up.
+We want to capture these changes with our SAOMs.", out.width='.99\\textwidth', fig.height=3>>=
 actual1 <- merge(data.frame(as.edgelist(as.network(fd2.w1))), 
@@
   facet_wrap(~Wave, labeller = "label_both") + 
-  theme(legend.position = 'bottom', panel.background = element_rect(fill = NA, color = "grey30"))
+  theme(legend.position = 'bottom', 
+        panel.background = element_rect(fill = NA, color = "grey30"), 
+        legend.text = element_text(size = 12, family = "Times New Roman"), 
+        legend.title = element_text(size = 14, family = "Times New Roman"))
 alldataviz
@@
 Edges are shown only if the weighted propensity to cosponsor from one senator to another is greater than 0.25.
-We use the Fruchterman-Reingold algorithm to layout the node-link diagram.', fig.height=3>>=
+We use the Fruchterman-Reingold algorithm to layout the node-link diagram.', fig.width=8, fig.height=8, out.width='\\textwidth'>>=
 seobama <- read_csv("data/congress/senateobamapres_gsw_25.csv")
+to_label <- c("Joseph R. Biden Jr.", "John S. McCain", "Ted Cruz","Marco Rubio",
+  "Lindsey O. Graham","Rand Paul", "Bernard Sanders", "Jim Webb", "Mitch McConnell",
+  "Harry M. Reid", "Hillary Rodham Clinton", "Amy Jean Klobuchar")
+seobama$label <- as.factor(ifelse(seobama$source %in% to_label, seobama$source, ""))
+levels(seobama$label) <- c("", "Amy Klobuchar", "Bernie Sanders", "Harry Reid",
+                           "Hillary Clinton", "Jim Webb", "John McCain", "Joe Biden",
+                           "Lindsey Graham", "Marco Rubio", "Mitch McConnel", 
+                           "Rand Paul", "Ted Cruz")
+set.seed(56049382)
 ggplot(data = seobama) + 
-  geom_net(directed = T, labelon=F, arrowsize = .25, singletons= T, fiteach = T, linewidth = .25, layout.alg = 'fruchtermanreingold',
-           aes(from_id = source, to_id = target, color = party),size = 1) + 
+  geom_net(directed = T, labelon=T, arrowsize = .25, singletons= T, fiteach = T, linewidth = .25, layout.alg = 'fruchtermanreingold', fontsize = 3, repel=T, 
+           aes(from_id = source, to_id = target, color = party, label = label),
+           size = 1) + 
   ThemeNet + 
   scale_color_manual(values = c("royalblue", "forestgreen","firebrick")) + 
-  theme(legend.position = 'bottom') + 
-  facet_wrap(~senate, nrow = 1)
+  theme(legend.position = 'bottom', strip.text = element_text(size = 20)) + 
+  facet_wrap(~senate, nrow = 2, labeller = "label_both") + 
+  xlim(c(0,1.05)) + 
+  ylim(c(0,1.05))
 @
@@
 \begin{subfigure}[t]{.45\textwidth}
-<<jtt, echo=FALSE, fig.width=2, fig.height=2, message = FALSE, warning = FALSE, fig.align='center',out.width='.49\\linewidth'>>=
+<<jtt, echo=FALSE, fig.width=2, fig.height=2, message = FALSE, warning = FALSE, fig.align='center',out.width='.49\\textwidth'>>=
 jTTe <- data.frame(from = c('i', 'i', 'h'), to = c('h', 'j', 'j'))
@@
            labelcolour='grey80',vjust = 0.5, hjust =0.5, arrowgap = .15, 
-           colour = 'black', size=10, fontsize = 4,
-           ecolour = c("red", "grey40", "grey40", "grey40")) + 
+           colour = 'black', size=10, fontsize = 5,
+           ecolour = c("grey40", "grey40","red", "grey40")) + 
   expand_limits(x=c(-0.1,1.1), y=c(-0.1,1.1)) +
@@
 \begin{subfigure}[t]{.45\textwidth}
-<<dab, echo=FALSE, fig.width=2, fig.height=2, message = FALSE, warning = FALSE, fig.align='center', out.width='.49\\linewidth'>>=
+<<dab, echo=FALSE, fig.width=2, fig.height=2, message = FALSE, warning = FALSE, fig.align='center', out.width='.49\\textwidth'>>=
 dade <- data.frame(from = c('i', 'i', 'h', 'j'), to = c('h', 'j', 'k', 'k'))
@@
 ggplot(data = dad, aes(from_id = from, to_id = to)) + 
-  geom_net(aes(shape = as.factor(group)), directed = T, labelon = T, labelcolour='grey80',vjust = 0.5, hjust =0.5, arrowgap = .15, colour = 'black', size=10, fontsize = 4) + 
+  geom_net(aes(shape = as.factor(group)), directed = T, labelon = T, labelcolour='grey80',vjust = 0.5, hjust =0.5, arrowgap = .15, colour = 'black', size=10, fontsize = 5) + 
   expand_limits(x=c(-0.1,1.1), y=c(-0.1,1.1)) +
@@
 \begin{table}
+\scalebox{0.9}{
 \begin{tabular}{l|ccc|ccc }
@@
 \end{tabular}
+}
 \caption{\label{tab:meanests} The means (standard deviations) of parameter values estimated from repeated fittings of $M1, M2, M3$ to the small friendship network and the senate collaboration network.
@@
            aes(from_id = source, to_id = target, color = party, linewidth = gsw, shape = sex,size = n_au)) + 
+  scale_size_continuous(name="# bills\nauthored") +
   #theme_net() + 
@@
 The distance computed from the data is shown as a vertical line.", fig.height=3>>=
+load("data/ansnullpaper.rda")
+library(lattice)
+gof1 <- sienaGOF(ansnull, OutdegreeDistribution, varName ="friendship", join = FALSE)
 qplot(x = log(gof1$`Period 1`$SimulatedTestStat), binwidth = .2) + 
@@
 In this instance, a time series plot of predicted covariate values for the nodes as the network evolves is viewing the model in the time and node spaces.
-If, as in Figure~\ref{fig:sienagof}, summary statistics computed on the network are of interest, distributions of these statistics computed on networks simulated from the model show the model in the edge data space.
+If, as discussed in Section~\ref{sec:modgof},
+%Figure~\ref{fig:sienagof}, 
+summary statistics computed on the network are of interest, distributions of these statistics computed on networks simulated from the model show the model in the edge data space.
 For an example, see Figure~\ref{fig:gof2}, where the distribution of the log of the Mahalanobis distance is shown for the outdegree distribution auxiliary statistics, and the log Mahalanobis distance for the observed data is shown as a vertical line.
@@
 simu_spread <- simu2 %>% spread(parameter, estimate)
-ggpairs(simu_spread, columns = 3:8, ggplot2::aes(colour=Model, alpha = .5)) + theme(plot.title = element_text(size = rel(1.5),
-                                            face = 'plain', angle = 0), 
-                  axis.title.x = element_text(size = rel(1.5),
-                                            face = 'plain', angle = 0, inherit.blank = T),
-                  axis.title.y = element_text(size = rel(1.5),
+ggpairs(simu_spread, columns = 3:8, ggplot2::aes(colour=Model, alpha = .5)) + theme(plot.title = element_text(size = 12,
+                                            face = 'plain', angle = 0, family = "Times New Roman"), 
+                  axis.title.x = element_text(size = 12,
+                                            face = 'plain', angle = 0, family = "Times New Roman", inherit.blank = T),
+                  axis.title.y = element_text(size = 12,
                                             face = 'plain', angle = 90, inherit.blank = T),
-                  axis.text.x.top = element_text(size = rel(1.5),
+                  axis.text.x.top = element_text(size = 12,
                                             face = 'plain', angle = 90, inherit.blank = TRUE), 
-                  axis.text.x.bottom = element_text(size = rel(1.5),
+                  axis.text.x.bottom = element_text(size = 12,
                                             face = 'plain', angle = 90, inherit.blank = TRUE), 
-                  axis.text.y.left = element_text(size = rel(1.5),
-                                            face = 'plain', angle = 0, inherit.blank = T),
-                  axis.text.y.right = element_text(size = rel(1.5),
-                                            face = 'plain', angle = 0, inherit.blank = T),
-                  strip.text.x = element_text(size = rel(2),
-                                            face = 'plain', angle = 0, 
+                  axis.text.y.left = element_text(size = 12,
+                                            face = 'plain', angle = 0, family = "Times New Roman", inherit.blank = T),
+                  axis.text.y.right = element_text(size = 12,
+                                            face = 'plain', angle = 0, family = "Times New Roman", inherit.blank = T),
+                  strip.text.x = element_text(size = 14,
+                                            face = 'plain', angle = 0, family = "Times New Roman", 
                                             margin = margin(t = 3, r = 0, b = 3, l = 0, unit = "pt"), inherit.blank = T),
-                  strip.text.y =element_text(size = rel(2),
-                                            face = 'plain', angle = 90,margin = margin(t = 3, r = 3, b = 3, l = 3, unit = "pt"),inherit.blank = T),
+                  strip.text.y =element_text(size = 14,
+                                            face = 'plain', family = "Times New Roman", angle = 90,margin = margin(t = 3, r = 3, b = 3, l = 3, unit = "pt"),inherit.blank = T),
                   strip.background = element_rect(colour = "black", fill = "white")
@@
 }
-
+library(dplyr050)
 friend.data.w1 <- as.matrix(read.table("data/s50_data/s50-network1.dat"))
@@
 set.seed(823746)
-
+load("data/ansnullpaper.rda")
 ansnullchains <- get_chain_info(ansnull)
@@
 By viewing this animation, we gain a better understanding of the dynamics of this model.
-<<histogramalters, fig.width = 8, fig.height=11, fig.cap="Each panel shows the theoretical (as lines) and empirical (as points) probabilities of the chosen ego node changing its tie to each of the other nodes.
+<<histogramalters, fig.width = 8, fig.height=10, fig.cap="Each panel shows the theoretical (as lines) and empirical (as points) probabilities of the chosen ego node changing its tie to each of the other nodes.
 The color of the line indicates whether the tie from the ego to the alter node is being added, removed, or if there is no change to the network in this step.">>=
@@
 ggplot(data = changes3) + 
-  geom_segment(
-    aes(x = alter , xend = alter, y = 0, yend = alterProb, color=type), size = 2) + 
-  geom_point(aes(x = alter, y = empprob), shape = 4, size = 2, alpha = .5) + 
-  labs(x = "alter node", y = "probability of alter node") + 
+  geom_segment(aes(x = alter, xend = alter, y = 0, yend = empprob, color = type), size = 2, alpha = .75) + 
+  geom_point(
+    aes(x = alter , y = alterProb), shape = 4, size = 2, alpha = .75) + 
+  labs(x = "alter node\n", y = "\nprobability of alter node") + 
   coord_flip() + 

### 2017-09-15 paper.Rnw
talk more about lineups

@@
        aes(from_id = X1, to_id = X2, color = as.ordered(drink), group = as.ordered(drink))) +
-  geom_net(fiteach = FALSE, directed=TRUE, layout.alg = 'fruchtermanreingold', size = 5, arrowsize = .25, arrowgap = .03, ecolour = 'grey40', labelon=T, labelcolour = 'black', fontsize = 4, vjust = .5) +
+  geom_net(fiteach = FALSE, directed=TRUE, layout.alg = 'fruchtermanreingold', size = 5, arrowsize = .25, arrowgap = .05, ecolour = 'grey40', labelon=T, labelcolour = 'black', fontsize = 4, vjust = .5) +
   scale_color_brewer(palette = "YlOrRd", name = "Drinking\nBehavior", labels = c("never", "1-2 times per year", "once per month", "once per week")) + 
@@
 These data come from \citet{friendsdata}, and we chose to only work with a subset of the data to make network visualizations less busy and to make any changes in the network more noticeable.
-The subset contained actors 20 through 35 and the ties between them, as well as the drinking behavior of each actor at each of the three waves.
-This specific subset was chosen because it showed somewhat higher connectivity than other subsets, as we've emphasized in the visualizations of the three network adjacency matrices in Figure~\ref{fig:adjmatviz}.
+To determine which subset to select, we visualized all waves of the full network using the adjacency matrix visualization approach, which we show in Figure~\ref{fig:adjmatviz}.
+This adjacency matrix visualization is different from the one in Figure~\ref{fig:toyplots} because it does not show the node IDs on the axes.
+The network in Figure~\ref{fig:adjmatviz} is much larger than our toy data example, so we remove the node labels to remove clutter.
+In all three adjacency matrices, the ego and the alter nodes are ordered by node ID, 1-50, which were determined arbitrarily by the relationships in wave 1.
+The subset we selected is outlined in red in the visualization.
+This subset contained actors 20 through 35 and the ties between them, as well as the drinking behavior of each actor at each of the three waves.
+This specific subset was chosen because it showed somewhat higher connectivity than other subsets, as we've emphasized in the visualizations of the three network observations in Figure~\ref{fig:adjmatviz}.
 For model fitting, we condition on wave 1 and estimate the parameters of the models from the second and third waves.
@@
 The network and the actor covariate values are visualized using a node-link diagram in Figure~\ref{fig:alldataviz}.
+\st{In the node-link diagram, the nodes are colored according to the drinking behavior of that student.
+Over time, we can see that the students tend to drink more and become increasingly isolated into smaller groups.
+An analysis of this type of data with a SAOM should capture these dynamics in a way that allows the researcher to draw conclusions about the nature of the network and behavioral forces at play.}
 
@@
 This is the network we draw using the node-link diagram.} An edge is only drawn in the average network if it appears in more than 5\% of the simulations (in at least 51 of the 1000 simulations), with edges that appear more frequently emphasized by thicker linewidths and a darker color.
-On either side of the average network in Figure\ref{fig:summnetM1}, we show the actual data, wave 1 on the left, and wave 2 on the right.
-\st{We can see that the structure of the average network is much more similar to the first network observation than to the second network observation.
-The simulations, however, are supposed to represent the second wave of data, which is shown on the right in Figure~\ref{fig:summnetM1}.
-This is an indication that our simple model, $M1$ is doing a very poor job of capturing the change mechanism from the first to the second wave of observation.} % XXX ST - I was looking at the wrong one.
+On either side of the average network in Figure~\ref{fig:summnetM1}, we show the actual data, wave 1 on the left, and wave 2 on the right.
+\st{We can see that the structure of the average network is much more similar to the first wave than to the second wave.
+However, the simulations are supposed to represent the second wave of data, which is shown on the right in Figure~\ref{fig:summnetM1}.
+This is an indication that the simple model, $M1$, is doing a very poor job of capturing the change mechanism from the first to the second wave of observation.} % XXX ST - I was looking at the wrong one.
 I had wave == 2 instead of wave == 1 in the code.
@@
 Here, the Mahalanobis distance of the cumulative outdegree values computed on a simulated network to the mean value simulated is shown.
-The distance computed from the data is shown as a vertical line.", fig.height=3>>=
+The distance computed from the data is shown as a vertical line.", fig.height=3, out.width='.6\\textwidth'>>=
 load("data/ansnullpaper.rda")
@@
 
-\st{Because longitudinal network data consist of three different ``spaces" of data, viewing the model in the data space can depend on which aspect of the model and the data you are interested in viewing.
-The SIENA software includes a method for modelling covariate change in the network by incorporating the network structure, assuming that the ties between nodes can affect how nodes behave over time.
+\st{Because longitudinal network data consist of three different ``spaces" of data, viewing the model in the data space can depend on which aspect of the model and the data we are interested in viewing.
+%The SIENA software includes a method for modelling covariate change in the network
+Incorporating data covariates into the network structure allows us to assess whether ties between nodes are affected by how nodes behave over time.
 In this instance, a time series plot of predicted covariate values for the nodes as the network evolves is viewing the model in the time and node spaces.
@@
 4369]{buja:2009}.
-It can be thought of like a police lineup, where the "suspect" is in a lineup among several innocent lookalike fillers, and a witness picking the suspect out of the lineup is considered evidence against the suspect.
-In data and model visualization, the "suspect" is a plot of the true data, while the "filler" is composed of several plots of mock data, simulated from a hypothesized model.
+It can be thought of like a police lineup, where the ``suspect" is in a lineup among several innocent lookalike fillers, and a witness picking the suspect out of the lineup is considered evidence against the suspect.
+In data and model visualization, the ``suspect" is a plot of the true data, while the ``filler" is composed of several plots of mock data, simulated from a hypothesized model.
 If the true data stands out among the simulated data, that is taken as evidence of poor model fit, whereas if the true data is difficult to identify among the simulated data, that is taked as evidence of good model fit.
-In Figure~\ref{fig:lineup}, the second wave of the small friendship network is shown among five simulated networks from model $M1$ using parameter values estimated from the data.
-The lineup allows us to view the model in the data space by placing instances of the model side-by-side with the data and examining the differences.}
+\st{An application of the lineup protocal can be found in \citet{hofmann2012graphical}, where the authors examine, for instance, the differences between polar and cartesian coordinates for plotting categorical data, and density plots and box plots for determining distributional differences.
+They pose questions to experiment participants such as, ``Which plot is most different from the others?" for the first example and, ``In which plot is the blue group furthest to the right?" for the distributional differences.
+The data visualizations examined in \citet{hofmann2012graphical} are less complex than a node-link network diagram.
+What questions should we ask for network data visualizations?
+Asking participants to identify the most different plot may be difficult, as we show} in Figure~\ref{fig:lineup}.
+In the network lineup, the second wave of the small friendship network is shown among five simulated networks from model $M1$ using parameter values estimated from the data.
+What makes these plots ``different"?
+It seems possible to argue for any one of the six plots in Firgure~\ref{fig:lineup} as most different.
+So, we guide participants to look at the overall structure of the graphs to determine which has the most and least complex structure.
+The least complex plot, number six, has no triplets, while the most complex plot, number three, has three triplets, and in fact, plot three is the data.
+We have found in an experiment where all the node-link diagrams shown are simulated, one from an ``alternative" model and the rest from a ``null" model, the triangular shape of the triplets stands out to people the most.
+We can use the lineup protocol to help better understand SAOMs because it allows us to view the model in the data space by placing observations the model side-by-side with the data and examining the differences.}
 
@@
 
-\subsection{Collections are more informative than singletons}
+\subsection{Visualizing collections of models}
 
@@
 
-To our small frienship network, we fit three models, $M1$, $M2$, and $M3$, using \texttt{RSiena} 1,000 times each.
+To our small friendship network, we fit three models, $M1$, $M2$, and $M3$, using \texttt{RSiena} 1,000 times each.
 We then looked at the distribution of the fitted values, which are shown in Figure~\ref{fig:distests}.

### 2017-09-15 paper.Rnw
making progress

@@
 Each SAOM used a simple rate function, $\alpha_m$, and an objective function with two or three parameters.
-The first model, $M1$, contains the absolute minimum number of parameters in the objective function $f_i(x)$:  $$f_i(x)^{M1} = \beta_1s_{i1} + \beta_2s_{i2},$$ where $s_{i1}$ is the density network statistic and $s_{i2}$ is the reciprocity network statistic for actor $i$ at the current network state $x$.
-The second and third models, $M2$ and $M3$, contain one additional parameter each in the objective function which were determined by a Wald-type test provided in the \texttt{RSiena} software to be significant, with $p$-values less than 0.05 \citep{RSienaTest}.
-The $M2$ model contains an actor-level covariate parameter, and the $M3$ model contains an additional strutural effect in the objective function.
+The first model, M1, contains the absolute minimum number of parameters in the objective function $f_i(x)$:  $$f_i(x)^{M1} = \beta_1s_{i1} + \beta_2s_{i2},$$ where $s_{i1}$ is the density network statistic and $s_{i2}$ is the reciprocity network statistic for actor $i$ at the current network state $x$.
+The second and third models, M2 and M3, contain one additional parameter each in the objective function which were determined by a Wald-type test provided in the \texttt{RSiena} software to be significant, with $p$-values less than 0.05 \citep{RSienaTest}.
+The M2 model contains an actor-level covariate parameter, and the M3 model contains an additional strutural effect in the objective function.
 \begin{align*}
@@
 
-For comparison, we also fit models $M1$, $M2$, and $M3$ to the senate collaboration data.
+For comparison, we also fit models M1, M2, and M3 to the senate collaboration data.
 Comparison of the parameter estimates of these models fit to the two different data sets is given in Table~\ref{tab:meanests}.
-In addition, we fit two more models, $M4$ and $M5$ to the senate data that were found to be significant according to the Wald-type test:
+In addition, we fit two more models, M4 and M5 to the senate data that were found to be significant according to the Wald-type test:
 
@@
 
-In $M4$, $s_{i5} = \mathbb{I}(x_{i+} = x_{+i} = 0)$ is an indicator variable for isolated nodes, called the network-isolate effect.
+In M4, $s_{i5} = \mathbb{I}(x_{i+} = x_{+i} = 0)$ is an indicator variable for isolated nodes, called the network-isolate effect.
 This parameter attempts to measure the effect of a node being completely isolated in the network.
 If the value of the parameter $\beta_5$ is high, nodes are encouraged to remain isolated.
-In $M5$,  $s_{i6} = \sum\limits_j x_{ij}x_{ji}\mathbb{I}(z_i = z_j)$ is a measure of reciprocity between senators in the same party, $z_i, z_j \in \{\text{Democrat, Independent, Republican}\}$, generally called the same covariate reciprocity effect.
+In M5,  $s_{i6} = \sum\limits_j x_{ij}x_{ji}\mathbb{I}(z_i = z_j)$ is a measure of reciprocity between senators in the same party, $z_i, z_j \in \{\text{Democrat, Independent, Republican}\}$, generally called the same covariate reciprocity effect.
 This parameter attempts to measure the propensity for forming reciprocal ties between nodes with the same covariate value.
@@
 In the middle, a summary network from the first model fit to the data.
-This summary network represents 1,000 simulations of wave 2 using the values from the simple fitted model $M1$.',results='hide', fig.height=3>>=
+This summary network represents 1,000 simulations of wave 2 using the values from the simple fitted model M1.',results='hide', fig.height=3>>=
 library(netvizinf)
@@
 In Figure~\ref{fig:summnetM1}, we show an \textit{average network} created with 1,000 simulations of the second wave of the network from Model 1.
-\st{To make this average network, we first simulated 1,000 wave 2 and wave 3 observations of our small friendship example data from model $M1$, for which parameters had previously been estimated.
+\st{To make this average network, we first simulated 1,000 wave 2 and wave 3 observations of our small friendship example data from model M1, for which parameters had previously been estimated.
 We then combine the 1,000 instances of wave 2, and count up the number of times each edge appears in a simulation.
@@
 However, the simulations are supposed to represent the second wave of data, which is shown on the right in Figure~\ref{fig:summnetM1}.
-This is an indication that the simple model, $M1$, is doing a very poor job of capturing the change mechanism from the first to the second wave of observation.} % XXX ST - I was looking at the wrong one.
+This is an indication that the simple model, M1, is doing a very poor job of capturing the change mechanism from the first to the second wave of observation.} % XXX ST - I was looking at the wrong one.
 I had wave == 2 instead of wave == 1 in the code.
@@
 \includegraphics[width=.75\linewidth]{img/Model1LUSmallFriends.png}
-\caption{\label{fig:lineup} A small lineup of node-link diagrams showing the second wave of our small friendship network among five networks simulated from model $M1$.}
+\caption{\label{fig:lineup} A small lineup of node-link diagrams showing the second wave of our small friendship network among five networks simulated from model M1.}
 \end{figure}
@@
 Asking participants to identify the most different plot may be difficult, as we show} in Figure~\ref{fig:lineup}.
-In the network lineup, the second wave of the small friendship network is shown among five simulated networks from model $M1$ using parameter values estimated from the data.
+In the network lineup, the second wave of the small friendship network is shown among five simulated networks from model M1 using parameter values estimated from the data.
 What makes these plots ``different"?
@@
 
-To our small friendship network, we fit three models, $M1$, $M2$, and $M3$, using \texttt{RSiena} 1,000 times each.
+To our small friendship network, we fit three models, M1, M2, and M3, using \texttt{RSiena} 1,000 times each.
 We then looked at the distribution of the fitted values, which are shown in Figure~\ref{fig:distests}.
 We can see from these distributions that the inclusion of the jumping transitive triplet parameter, $\beta_3$ is obviously affecting the distributions of the other four parameters included in all models, $\alpha_1$, $\alpha_2$, $\beta_1$, and $\beta_2$.
-In comparison with models M1 and M3, model M2 typically has higher estimates of the rate parameter, meaning that the inclusion of the covariate statistic in the model leads to higher estimates of the number of times, on average, a node gets to change its ties.
+\st{When $\beta_3$ is included, its estimate is postitive, meaning that friendships between two girls with different drinking behaviors tend to form when there is an intermediary who is already friends with the two girls.
+The inclusion of this parameter leads to increases in the rate parameters' estimates, suggesting that encouraging the transitive triplet behavior means that the girls would also change friends more frequently.
+The outegree parameter, $\beta_1$ decreases when $\beta_3$ is included, while the reciprocity parameter, $\beta_2$ increases.
+This implies the girls in the data prefer to form closer friend groups, as indicated by reciprocated ties and jumping transitive triplet formation, as opposed to being popular and having many friends.
+Having many friends who do not reciprocate is discouraged by M2.} In comparison with models M1 and M3, model M2 typically has higher estimates of the rate parameter, meaning that the inclusion of the covariate statistic in the model leads to higher estimates of the number of times, on average, a node gets to change its ties.
 It is not clear, however, that the addition of a parameter to the objective function \textit{should} effect the estimation of the rate parameters, so we continue to explore the collection of parameter estimates.
@@
 
-<<averagenet, results='hide', out.width='99%',fig.cap='The three "average" networks in the middle, with the actual network observations oon either side.
+<<averagenet, results='hide', out.width='99%',fig.cap='The node-link diagrams from the three "average" networks that we calculated are in the top row, and the true wave 1 and wave 2 data are shown in the bottom row above.
 There is some difference between the three models, but overall, these three models cannot capture the structure in the true second wave of data.', fig.height=6>>=
@@
 combinedavgs$linewidth[combinedavgs$cat %in% c("1st Wave", "2nd Wave")] <- .5
+combinedavgs$cat <- as.factor(combinedavgs$cat)
+combinedavgs$cat <- ordered(combinedavgs$cat, levels = c("Model1", "Model2", "Model3", "1st Wave", "2nd Wave"))
+
 ggplot(data = combinedavgs) + 
@@
   ThemeNet + theme(panel.background = element_rect(color = 'black')) + 
-  facet_wrap(~cat, nrow = 2)
+  facet_wrap(~cat)
 
@@
 <<adjmatorder, message = FALSE, warning = FALSE, fig.cap="On the left, the starting friendship network represented in adjacency matrix form, ordered by vertex id.
-On the right, the same adjacency matrix is presented after ordering the vertices by one repitition of the microstep simulation process from wave one to wave two.", fig.width=8, fig.height=3>>=
+On the right, the same adjacency matrix is presented after ordering the vertices by one repetition of the microstep simulation process from wave one to wave two.", fig.width=8, fig.height=3>>=
 # first present the original wave 1 adj mat, then present the adj mat after being ordered by the 1st pc of the big adjmat A. 
@@
 
-<<adjmatani, fig.cap='The adjacency matrix visualization animation for one series of microsteps.
-At the beginning of the animation, there are two connected components.
-By the end, the components appear to be more spread out.', fig.show='animate', warning = FALSE, message = FALSE, interval=.2, results='hide', fig.height=4>>=
+<<adjmatani, fig.cap='A selection of frames from the adjacency matrix visualization animation for one series of microsteps.
+At the beginning of the animation, shown in the top row, there are two clearly connected components: one in the top left corner, and one in the bottom right corner.
+By the end, the component in the top left has spread out, while the bottom right component has shrunk, but remains closely connected.', fig.show='animate', warning = FALSE, message = FALSE, interval=.2, results='hide', fig.height=4>>=
 adjmatall <- matrix(0, nrow = 16, ncol = 16)
@@
 adjmatall3$from <- factor(adjmatall3$from, levels = rev(order1))
-plots <- list()
-for(i in 0:55){
-plots[[i + 1]] <- ggplot(data = adjmatall3 %>% filter(ms == i)) + 
+
+adjmatall3$frame <- adjmatall3$ms
+#plots <- list()
+#for(i in 0:55){
+#plots[[i + 1]] <- 
+  ggplot(data = adjmatall3 %>% filter(ms %in% c(1,3,5,7,9,11,13,43,45,47,49,51,53,55))) + 
   geom_tile(aes(x = to, y = from, fill = as.factor(val)), color = 'black') + 
   scale_fill_manual(values = c("white", "black")) + 
-  scale_x_discrete(position = "top", labels = str_replace(order1, "V", "")) + 
+  scale_x_discrete(position = "bottom", labels = str_replace(order1, "V", "")) + 
   scale_y_discrete(labels = rev(str_replace(order1, "V", ""))) + 
@@
   theme(legend.position = 'none') +
-  coord_fixed() 
-}
-plots 
+  coord_fixed() + 
+  facet_wrap(~frame, labeller = 'label_both', nrow = 2)  
+#}
+#plots 
 @
@@
 <<heatmap, fig.align='center', fig.cap='A heatmap showing the empirical transition probabilities for the first microstep in 1,000 simulations.
-The acting node is on the x-axis, and the receiving node is on the y-axis.
-Missing squares represent ties that did not occur.', fig.height=4>>=
+The acting ego node is on the y-axis, and the alter node is on the x-axis.
+There were many ties with empirical probability zero.', fig.height=4>>=
 ansnullchains %>% 
@@
 ggplot(data = probsDat, 
-       aes(x = factor(ego, levels = order1), 
-             y = factor(alter, levels = rev(order1)), 
+       aes(y = factor(ego, levels = order1), 
+             x = factor(alter, levels = rev(order1)), 
              fill = prob)) + 
@@
   scale_fill_gradient(low = 'white', high = 'blue', limits = c(0,1)) +
-  labs(x = "ego", y = "alter")+
+  labs(y = "ego", x = "alter")+
   scale_x_discrete(position = 'top', labels = str_replace(order1, "V", "")) +

### 2017-09-15 paper.Rnw
match all adjacency matrices (x,y)

@@
 
-p1 <- ggplot(data = numordersf, aes(x = from, y = to)) + geom_tile(color = 'black', aes(fill = as.factor(value))) + 
+p1 <- ggplot(data = numordersf, aes(y = from, x = to)) + geom_tile(color = 'black', aes(fill = as.factor(value))) + 
   scale_fill_manual(values = c('white', 'black')) +
@@
 At the beginning of the animation, shown in the top row, there are two clearly connected components: one in the top left corner, and one in the bottom right corner.
-By the end, the component in the top left has spread out, while the bottom right component has shrunk, but remains closely connected.', fig.show='animate', warning = FALSE, message = FALSE, interval=.2, results='hide', fig.height=4>>=
+By the end, the component in the top left has spread out, while the bottom right component has shrunk, but remains closely connected.', warning = FALSE, message = FALSE, results='hide', fig.height=4>>=
 adjmatall <- matrix(0, nrow = 16, ncol = 16)

### 2017-09-19 paper.Rnw
making progress on reorganizing section 4.

@@
 %                 Placing Figures
+\renewcommand{\topfraction}{0.99}	% max fraction of floats at top
+\renewcommand{\bottomfraction}{0.8}	% max fraction of floats at bottom
+\renewcommand{\textfraction}{0.05}	% allow minimal text w. figs
 
@@
 Unlike other network models, SAOMs are not very well understood.
-\st{They are relatively new, especially compared to the classic exponential random graph models, and they are not very tractable analytically.
+They are relatively new, especially compared to the classic exponential random graph models, and they are not very tractable analytically.
 Likelihood functions quickly very complex objects to analyze due to the dependency structure inherent in the data.
@@
 By bringing some light to the underlying methods and structures that are behind the scenes when fitting SAOMs to network data, we aim to help researchers working with the models better understand the implications and analyses of these models.
-}
 
@@
 \item The Mahalanobis distance from the observed statistics to the distribution of the simulated statistics is computed using the mean and covariance found in step 2. 
-\item The Mahalanobis distrance from each of the $N$ simulations to the same distribution is computed, and the Mahalanobis distance of the observed data is compared to this distribution of distances.
-\item A $p$-value is found by computing the proportion of simulated distances found in step 4 that are as large or larger than the Mahalanobis distance from the data.
+\item The Mahalanobis distance from each of the $N$ simulations to the same distribution is computed, and the Mahalanobis distance of the observed data is compared to this distribution of distances.
+\item An empirical $p$-value is found by computing the proportion of simulated distances found in step 4 that are as large or larger than the Mahalanobis distance from the data.
 A SAOM is thus considered a good fit if $p$ is large.
@@
 %In Figure~\ref{fig:sienagof}, we provide an example of what the goodness-of-fit plot output looks like for indegree and outdegree statistics for a small model.
-This function draws a box plots and a violin plot at each value of the statistic of interest observed in the simulations.
-In order to compare the distribution of the simulated network to the true data, red points connected by red lines representing the observed values are superimposed on the boxplots.
+This function draws a box plot and a violin plot at each value of the statistic of interest observed in the simulations: \st{on the $x$-axis, the out(in)degrees observed in the data ($0, 1, 2, 3, \dots$), and on the $y$-axis, the cumulative number of times that out(in)degree value appears in the simulations and the data.} In order to compare the distribution of the counts of nodes with the specified degree as calculated on the simulated networks to the counts observed in the true data, red points connected by red lines representing the observed values are superimposed on the boxplots.
 If the red points lie "well within" the simulated values, the model is a good fit to the data.
-The plot is not intuitive for understanding network data.
-%The model examined in Figure~\ref{fig:sienagof} appears to do a better job of capturing the outdegree distribution than the indegree distribution of the data.
+\st{This plot is not shown because it is not intuitive for understanding network data.
+Boxplots separated by the many outdegree values observed do not communicate \textit{how} the nodes are connected, just that some have more connections than others.
+In order to understand the fit of the model, we should try to understand how well the model captures the\textit{overall} structure of the network, not just one or two summary measurements of that structure.} %The model examined in Figure~\ref{fig:sienagof} appears to do a better job of capturing the outdegree distribution than the indegree distribution of the data.
 In Section~\ref{sec:minds}, we propose a new way of visualizing goodness-of-fit that uses the traditional node-link diagram to visualize the entire network instead of numerical summaries of the the network.
@@
         legend.text = element_text(size = 12, family = "Times New Roman"), 
-        legend.title = element_text(size = 14, family = "Times New Roman"))
+        legend.title = element_text(size = 12, family = "Times New Roman"))
 alldataviz
@@
 For example, we chose to use a SAOM to model network data over an exponential random graph model (ERGM) because we believe that actor-level variables effect network structure and formation, and we wanted to model the network changes over time.
-These three uses of the term ``model" can each be visualized according to the three principals of model visualization: display the model in the data space, look at a collection of models, and explore the process of fitting the model, not just the end result.
-We turn our focus to the fitted model and the model form.
-Specifically, we want to know how the model form affects the fitted model.
-Using our example data sets and our visualization toolbox, we hope to answer this question for SAOMs.
+
+%These three uses of the term ``model" 
+\st{The model family, the model form, and the fitted model can each be visualized according to the three principals of model visualization: we can }view the model in the data space, visualize collections of models, and explore the process of fitting the model, not just the end result.
+\st{Since we have already decided on our model family, SAOMs, }we now shift our focus to the fitted model and the model form.
+Specifically, we want to learn more about how the model form we choose affects the fitted model by using our example data sets and our visualization toolbox.
+We begin by introducing the five models that we fit to our example data.
+\st{Next we use the five models to guide our visual explorations of SAOMs.
+We first use novel tools and ideas to view a SAOM in the data space of a dynamic network.
+We then explore collections of the same models fit many times to the example data to see how the simulation processes in \texttt{RSiena} affect the model fits.
+Finally, we look behind the scenes and into the individual steps of the continuous time Markov chain to learn more about how this "hidden data" mechanism works and how it results in a fitted model.
+}
 
@@
 The first statistic emphasizes triad relationships that are formed between actors from different covariate groups, while the other emphasizes indirect ties between actors.
-These two effects are visually represented and further described in Figure~\ref{fig:jtt} and Figure~\ref{fig:dad}, respectively.
+The covariate groups are determined by the student's drinking behavior, the values of which are numeric and mean-centered.
+The four values in our data are $\{\text{never, 1-2 times per year, once per month, once per week}\}$, which after being converted to numeric and mean-centered become $\{-0.8125,-1.8125,0.1875,1.1875\}$.
+The two additional effects are visually represented and further described in Figure~\ref{fig:jtt} and Figure~\ref{fig:dad}, respectively.
 
@@
 
-\begin{table}
-\scalebox{0.9}{
-\begin{tabular}{l|ccc|ccc }
-  \hline
-  & \multicolumn{3}{|c|}{Friendship Data} & \multicolumn{3}{|c|}{Senate Data} \\
-  \hline
-  M1 & M2 & M3 & M1 & M2 & M3 \\
-  $\alpha_1$ & \Sexpr{SFE$M1[1]} & \Sexpr{SFE$M2[1]} & \Sexpr{SFE$M3[1]} & \Sexpr{SSE$M1[1]} & \Sexpr{SSE$M2[1]} & \Sexpr{SSE$M3[1]} \\
-  $\alpha_2$ & \Sexpr{SFE$M1[2]} & \Sexpr{SFE$M2[2]} & \Sexpr{SFE$M3[2]} & \Sexpr{SSE$M1[2]} & \Sexpr{SSE$M2[2]} & \Sexpr{SSE$M3[2]}  \\
-  $\alpha_3$ & -- & -- & -- & \Sexpr{SSE$M1[3]} & \Sexpr{SSE$M2[3]} & \Sexpr{SSE$M3[3]} \\
-  $\beta_1$ & \Sexpr{SFE$M1[3]} & \Sexpr{SFE$M2[3]} & \Sexpr{SFE$M3[3]} & \Sexpr{SSE$M1[4]} & \Sexpr{SSE$M2[4]} & \Sexpr{SSE$M3[4]} \\
-  $\beta_2$ & \Sexpr{SFE$M1[4]} & \Sexpr{SFE$M2[4]} & \Sexpr{SFE$M3[4]} & \Sexpr{SSE$M1[5]} & \Sexpr{SSE$M2[5]} & \Sexpr{SSE$M3[5]}\\
-$\beta_3$ & -- & \Sexpr{SFE$M2[5]} & -- & -- & \Sexpr{SSE$M2[6]} & -- \\
-  $\beta_4$ & -- & --  & \Sexpr{SFE$M3[6]} & -- & -- & \Sexpr{SSE$M3[7]} \\
-  \hline
-\end{tabular}
-}
-\caption{\label{tab:meanests} The means (standard deviations) of parameter values estimated from repeated fittings of $M1, M2, M3$ to the small friendship network and the senate collaboration network.
-Each model was fit 1,000 times to the friend data, while each model was fit 100 times to the senate data.}
-\end{table}
-
-For comparison, we also fit models M1, M2, and M3 to the senate collaboration data.
-Comparison of the parameter estimates of these models fit to the two different data sets is given in Table~\ref{tab:meanests}.
+In Section~\ref{sec:samemoddiffdat}, we also fit models M1, M2, and M3 to the senate collaboration data for comparison.
+%Comparison of the parameter estimates of these models fit to the two different data sets is given in Table~\ref{tab:meanests}.
 In addition, we fit two more models, M4 and M5 to the senate data that were found to be significant according to the Wald-type test:
@@
 In M5,  $s_{i6} = \sum\limits_j x_{ij}x_{ji}\mathbb{I}(z_i = z_j)$ is a measure of reciprocity between senators in the same party, $z_i, z_j \in \{\text{Democrat, Independent, Republican}\}$, generally called the same covariate reciprocity effect.
-This parameter attempts to measure the propensity for forming reciprocal ties between nodes with the same covariate value.
+The values of the covariate are converted to numeric, $\{1,2,3\}$, respectively, and mean-centered within \texttt{RSiena}.
+The mean-centered values of $z \in \{\text{Democrat, Independent, Republican}\}$ are $\{-0.96, 0.04, 1.04\}$.
+The parameter $\beta_6$ attempts to measure the propensity for forming reciprocal ties between nodes with the same covariate value.
 If the value of the parameter $\beta_6$ is high, nodes belonging to the same covariate group are encourage to reciprocate incoming ties.
-By exploring these models more visually, we hope to understand these effects and the model fitting process better.
+\st{Due to the intractability of SAOMs, it is difficult to know for certain how to interpret a fitted value of a parameter.
+We can make educated guesses based on the definition of the effect and the sign of the fitted value, but a direct interpretation is not always possible.
+But,} by exploring these models more visually, we aim to understand these effects, their interpretations, and the model fitting process better.
 
@@
 The first way we hope to better understand stochastic actor-oriented models is by viewing the model(s) in the data space.
-In \citeauthor{modelvis}, they define the \textit{data space} as ``the region over which we can reliably make inferences, usually a hypercube containing the data" \citep[p.~206]{modelvis}.
+In \citeauthor{modelvis}, they \st{happen to} define the \textit{data space} as ``the region over which we can reliably make inferences, usually a hypercube containing the data" \citep[p.~206]{modelvis}.
 But what does this definition mean for network data?
@@
 \item the edges and their correpsonding variables that describe the relationships between the nodes, and 
-\item the time over which the network evolves
+\item the time, both the continuous unobserved time and the discrete observed time points, over which the network evolves
 \end{enumerate}
@@
 
-One tool that can bring these three data spaces together in this way is the \texttt{R} package \texttt{geomnet} \citep{geomnet}.
-Different aesthesic aspects of each piece of the node-link diagram can be tied to the underlying node or edge data.
+\st{Because longitudinal network data consist of three different ``spaces" of data, viewing the model in the data space can depend on which aspect of the \textit{model} and the \textit{data} we are interested in viewing.
+%The SIENA software includes a method for modelling covariate change in the network
+Incorporating data covariates into the network structure allows us to assess whether the ties between nodes are affected by how nodes behave over time.
+In this instance, we would want to view predictions over time.
+A SAOM can also model behavior change over time, taking both the node and edge information into account.
+In this case, a plot of predicted covariate values over time would put the model into the time and node data space.
+Most likely, however, is that we would want to view all of the data spaces simultaneously.}
+
+One tool that can bring the \st{node, edge, and time} data spaces together in this way is the \texttt{R} package \texttt{geomnet} \citep{geomnet}.
+Different visual features in the node-link diagram can be tied to the underlying node or edge data.
 The color, size, and shape of the points can be used to represent variables in the node data, while the color, linewidth, and linetype of the lines between points can be used to represent the edge variables.
@@
 
-<<funsenatevis, fig.height=3, fig.cap='The 111th Senate separated into ``BC" and ``AC": before Clinton left the Senate to become Secretary of State (on the right), and after she left (on the left).' >>=
+<<funsenatevis, fig.height=3, fig.cap='The 111th Senate at two discrete time points: while Clinton was in the senate in the first few weeks of 2009 (on the left) and after she left the senate to become Secretary of State (on the right).
+We put some potential model covariates, sex and party of the senator and the number of bills they authored in the data space through the shape, color, and size of the nodes, respectively.
+We also map the strength of the tie, the $WPC$, to the width of the edge between two nodes.
+Thicker lines implies higher propensity to collaborate.
+In addition, senators with no ties higher than $WPC = 0.25$ are not shown.'>>=
 se111clint <- filter(seobama, senate == 111)
@@
 clintonSenate <- rbind(se111clint, se111noclint)
-
-ggplot(data = clintonSenate) + 
-  geom_net(directed = T, labelon=F, arrowsize = .3, singletons= T, fiteach = T, arrowgap = .01, layout.alg = 'fruchtermanreingold',
+clintonSenate$Clinton <- as.factor(clintonSenate$Clinton)
+clintonSenate$Clinton <- ordered(clintonSenate$Clinton, levels = c("Yes", "No"))
+
+clintonSenate %>% filter(source %in% target)
+
+clintonSenate %>% filter(!(source %in% c("Roland Burris", "Bernard Sanders", "Frank R. Lautenberg", "Judd A. Gregg", "Mary L. Landrieu") & is.na(target))) %>%
+ggplot() + 
+  geom_net(directed = T, labelon=F, arrowsize = .3, singletons= F, fiteach = T, arrowgap = .01, layout.alg = 'fruchtermanreingold',
            aes(from_id = source, to_id = target, color = party, linewidth = gsw, shape = sex,size = n_au)) + 
@@
 
-To demonstrate, we use \texttt{geomnet} to visualize the connections in the $111^{th}$ at two different points: when Hillary Clinton was in the senate, and after she left to become Secretary of State.
-Clinton was only in the $111^{th}$ senate for 17 days, from January 3, 2009, to January 20, 2009, when she was in the middle of her second term in the senate.
+To demonstrate, we use \texttt{geomnet} to visualize the connections in the $111^{th}$ United States Senate at two different points: when Hillary Clinton was in the senate, and after she left to become Secretary of State.
+Clinton was only in the $111^{th}$ senate for 17 days, from January 3, 2009, to January 20, 2009, when she was in the middle of her second term as senator from New York.
 In that time, she authored two bills and was a cosponsor on 17 other bills.
-With Clinton included in the node-link diagram, the senate looks much more collaborative than it does without her in the diagram.
+With Clinton included in the node-link diagram, the senate looks much more highly collaborative than it does without her in the diagram.
 We can compare the number of bills authored throughout the senate by mapping the size of the node to the numer of bills authored by that senator.
-We also map shape to gender, and keep color mapped to party of the senator.
+We also map shape to sex, and keep color mapped to party of the senator.
 In addition, we can see the strength of the tie by mapping the linewidth of the edge to the $WPC$ value between the two senators.
-In this single visualization, we have viewed node information (number of bills authored by a senator and the gender of that senator), edge infortmation (the direction and strength of ties between two senators), and time (before and after Clinton left the senate).
+In this single visualization, we have viewed node information (number of bills authored by a senator and the sex and party of that senator), edge information (the direction and strength of ties between two senators), and time (before and after Clinton left the senate).
 % ST: I don't like this paragraph
@@
 In the middle, a summary network from the first model fit to the data.
-This summary network represents 1,000 simulations of wave 2 using the values from the simple fitted model M1.',results='hide', fig.height=3>>=
+This summary network represents 1,000 simulations of wave 2 using the values from the simple fitted model M1.',results='hide', fig.height=2.5>>=
 library(netvizinf)
@@
 combinedavgs %>% filter(cat == "Model1") %>% mutate(logweight = log(weight)) -> t1 
-colors <- tweenr::tween_color(data = c("#969696", "#67000d"), n = 26, ease = 'linear')
+colors <- tweenr::tween_color(data = c("#969696", "#d73027"), n = 26, ease = 'linear')
 t1 %>% arrange(logweight) -> t1
@@
 No single network alone simulated from a SAOM is going to look like the data or represent the model, just as no single value simulated from the standard normal distribution will look like a bell-shaped curve.
-It would therefore better to visualize many simulations together.
-From a statistician's point of view, a stumbling block with statistical network models generally is the lack of an ``average network" measure.
+\st{Statisticians would prefer to look at a sample, or at least a summary statistic or two, }so it would therefore better to visualize many simulations together.
+From a statistician's point of view, a stumbling block with statistical network models generally is the lack of an ``average network" or ``expected network" value.
 Statisticians frequently rely on averages and expected values in data analyses, but statistical network models, especially those as complex as SAOMs, lack a single, intuitive expected value measure.
 We could talk about expected values of parameters, but the parameters can be hard to interpret.
-Furthermore, there is no way to talk about the expected value of an observation from a statistical network model.
+\st{Expected values of parameters are important, but if they cannot directly tell us anything about our dependent variable (networks in this case), they lose some value to us.} Furthermore, there is no way to talk about the expected value of an observation \st{simulated} from a statistical network model.
 How then, can we arrive at an ``average" network?
-We propose to answer this question through visualization.
-For network data, one way we propose to come up with an "average" network is through a summary network drawn using the traditional node-link diagram.
+We answer this question through visualization.
+For network data, one way we view an "average" network is through a summary network drawn using the traditional node-link diagram.
 In Figure~\ref{fig:summnetM1}, we show an \textit{average network} created with 1,000 simulations of the second wave of the network from Model 1.
@@
 Then, we combine these 1,000 networks into a single network with edgeweight equal to the proportion of time that edge appears in our 1,000 simulations.
-This is the network we draw using the node-link diagram.} An edge is only drawn in the average network if it appears in more than 5\% of the simulations (in at least 51 of the 1000 simulations), with edges that appear more frequently emphasized by thicker linewidths and a darker color.
+This weighted edgelist is the network we draw using the node-link diagram.} An edge is only drawn in the average network if it appears in more than 5\% of the simulations (in at least 51 of the 1000 simulations), with edges that appear more frequently emphasized by thicker linewidths and a darker color, representing the proportion of times appeared in the model simulations.
 On either side of the average network in Figure~\ref{fig:summnetM1}, we show the actual data, wave 1 on the left, and wave 2 on the right.
@@
 However, the simulations are supposed to represent the second wave of data, which is shown on the right in Figure~\ref{fig:summnetM1}.
-This is an indication that the simple model, M1, is doing a very poor job of capturing the change mechanism from the first to the second wave of observation.} % XXX ST - I was looking at the wrong one.
+This is an indication that the simple model, M1, is doing a very poor job of capturing the change mechanism from the first to the second wave of observation.
+The average network can thus be used to help determine model goodness-of-fit.
+Because the the average network looks more like the first wave than the second wave, we can use the visualization in Figure \ref{fig:summnetM1} as evidence of poor model fit.}  % XXX ST - I was looking at the wrong one.
 I had wave == 2 instead of wave == 1 in the code.
@@
 Here, the Mahalanobis distance of the cumulative outdegree values computed on a simulated network to the mean value simulated is shown.
-The distance computed from the data is shown as a vertical line.", fig.height=3, out.width='.6\\textwidth'>>=
+The distance computed from the data is shown as a vertical line.", fig.height=3, out.width='.6\\textwidth', eval = FALSE>>=
 load("data/ansnullpaper.rda")
@@
 gof1 <- sienaGOF(ansnull, OutdegreeDistribution, varName ="friendship", join = FALSE)
-qplot(x = log(gof1$`Period 1`$SimulatedTestStat), binwidth = .2) + 
-  geom_vline(xintercept = log(gof1$`Period 1`$ObservedTestStat)) + 
+histdat <- data.frame(x = log(gof1$`Period 1`$SimulatedTestStat))
+ggplot(histdat) + geom_histogram(aes(x = x), binwidth = .1, fill = 'white', color = "black") + 
+  #geom_density(aes(x = x, y = ..density..*250)) + 
+  geom_vline(xintercept = log(gof1$`Period 1`$ObservedTestStat), color = 'red') + 
   ThemeNoNet + 
@@
 
-\st{Because longitudinal network data consist of three different ``spaces" of data, viewing the model in the data space can depend on which aspect of the model and the data we are interested in viewing.
-%The SIENA software includes a method for modelling covariate change in the network
-Incorporating data covariates into the network structure allows us to assess whether ties between nodes are affected by how nodes behave over time.
-In this instance, a time series plot of predicted covariate values for the nodes as the network evolves is viewing the model in the time and node spaces.
-If, as discussed in Section~\ref{sec:modgof},
+%\st{If, as discussed in Section~\ref{sec:modgof},
 %Figure~\ref{fig:sienagof}, 
-summary statistics computed on the network are of interest, distributions of these statistics computed on networks simulated from the model show the model in the edge data space.
+%summary statistics such as outdegree and indegree distribution computed on the network are of interest, we can look at the computed distances between the data and the model simulations.
+distributions of these statistics computed on networks simulated from the model show the model in the edge data space.
 For an example, see Figure~\ref{fig:gof2}, where the distribution of the log of the Mahalanobis distance is shown for the outdegree distribution auxiliary statistics, and the log Mahalanobis distance for the observed data is shown as a vertical line.
+}
 
@@
 
-Another potential goodness-of-fit visualization is a lineup like those proposed in \citet{buja:2009}.
-A \texttt{lineup} ``asks the witness to identify the plot of the real data from among a set of decoys, the null plots, under the veil of ignorance" \citep[p.
+Another potential goodness-of-fit visualization \st{that places the model in the node and edge data space} is a lineup like those proposed in \citet{buja:2009}.
+A \textit{lineup} ``asks the witness to identify the plot of the real data from among a set of decoys, the null plots, under the veil of ignorance" \citep[p.
 4369]{buja:2009}.
@@
 What questions should we ask for network data visualizations?
-Asking participants to identify the most different plot may be difficult, as we show} in Figure~\ref{fig:lineup}.
-In the network lineup, the second wave of the small friendship network is shown among five simulated networks from model M1 using parameter values estimated from the data.
+Asking participants to identify the most different plot may be difficult.
+In the network lineup shown in Figure~\ref{fig:lineup}, the second wave of the small friendship network is shown among five simulated networks from model M1 using parameter values estimated from the data.
 What makes these plots ``different"?
-It seems possible to argue for any one of the six plots in Firgure~\ref{fig:lineup} as most different.
+It seems possible to argue for any one of the six plots in Figure~\ref{fig:lineup} as most different.
 So, we guide participants to look at the overall structure of the graphs to determine which has the most and least complex structure.
 The least complex plot, number six, has no triplets, while the most complex plot, number three, has three triplets, and in fact, plot three is the data.
-We have found in an experiment where all the node-link diagrams shown are simulated, one from an ``alternative" model and the rest from a ``null" model, the triangular shape of the triplets stands out to people the most.
-We can use the lineup protocol to help better understand SAOMs because it allows us to view the model in the data space by placing observations the model side-by-side with the data and examining the differences.}
+We have found in an experiment where all the node-link diagrams shown are based on simulated data, one observation from an ``alternative" model and the rest from a ``null" model, the triangular shape of the triplets stands out the most to participants.
+We can use the lineup protocol to help better understand SAOMs because it allows us to view the model in the data space by placing observations the model side-by-side with the data and examining the differences.
+This can also help us determine the significant structural effects, if any, of the parameters in the model on observations simulated from the model.
+The triangular shape mentioned above becomes more prevalent when a transitive triplet parameter in included in a model, but it does not always cause triplets to form in the simulated data.
+This requires more investigation with larger data sets and more complex models, but is promising for the development of additional goodness-of-fit measures for network models.}
 
@@
 
-The second principle of model visualization is that ``collections are more informative than singletons" \citep[p.~210]{modelvis}.
-The authors describe several different methods for producing such collections, but we focus on only a few here.
-We look at collections produced ``from exploring the space of all possible models," ``by varying model settings", ``by fitting the same model to different datasets", and ``by treating each iteration as a model" \citep[p.~210-11]{modelvis}.
-Because fitting a SAOM requires MCMC methods, we also fit the same model many times, resulting in distributions of fitted parameters for one data set.
-
-To our small friendship network, we fit three models, M1, M2, and M3, using \texttt{RSiena} 1,000 times each.
-We then looked at the distribution of the fitted values, which are shown in Figure~\ref{fig:distests}.
-We can see from these distributions that the inclusion of the jumping transitive triplet parameter, $\beta_3$ is obviously affecting the distributions of the other four parameters included in all models, $\alpha_1$, $\alpha_2$, $\beta_1$, and $\beta_2$.
-\st{When $\beta_3$ is included, its estimate is postitive, meaning that friendships between two girls with different drinking behaviors tend to form when there is an intermediary who is already friends with the two girls.
-The inclusion of this parameter leads to increases in the rate parameters' estimates, suggesting that encouraging the transitive triplet behavior means that the girls would also change friends more frequently.
-The outegree parameter, $\beta_1$ decreases when $\beta_3$ is included, while the reciprocity parameter, $\beta_2$ increases.
-This implies the girls in the data prefer to form closer friend groups, as indicated by reciprocated ties and jumping transitive triplet formation, as opposed to being popular and having many friends.
-Having many friends who do not reciprocate is discouraged by M2.} In comparison with models M1 and M3, model M2 typically has higher estimates of the rate parameter, meaning that the inclusion of the covariate statistic in the model leads to higher estimates of the number of times, on average, a node gets to change its ties.
-It is not clear, however, that the addition of a parameter to the objective function \textit{should} effect the estimation of the rate parameters, so we continue to explore the collection of parameter estimates.
-
-To further investigate this potentially problematic relationship between the parameter values, we look at correlations between each of the parameter estimates in each model.
-In Figure~\ref{fig:corplots}, we examine correlations between each of pair of parameters within each model and overall.\nocite{ggally} The strongest correlation within each model is between $\beta_1$ and $\beta_2$, with absolute value of correlation between those two parameter values greater than 0.90 in all three models.
-The $\beta_1$ parameter is also highly correlated with the $\beta_3$ parameter within model M2, but it is not as highly correlated with the $\beta_4$ parameter in model M3.
-
-<<getdata>>=
-nulls <- read.csv("data/distribution_null_model.csv")
-alt <- read.csv("data/distribution_jumpTT_model.csv")
-alt2nd <- read.csv("data/distribution_dblpairs_model.csv")
-means <- simu2 %>% group_by(Model, parameter) %>% summarize(
-  means = mean(estimate)
-)
-null_mod_sv <- as.numeric(nulls[,-c(1,6,7)] %>% summarise_each(funs(mean)))
-alt_mod_sv <- as.numeric(alt[,-c(1, 7,8)] %>% summarise_each(funs(mean)))
-alt_mod2_sv <- as.numeric(alt2nd[,-c(1,7,8)] %>% summarise_each(funs(mean)))
-@
-
-<<distests, fig.cap="Distribution of fitted parameter values for our three SAOMs.
-The inclusion of $\\beta_3$ or $\\beta_4$ clearly has an effect on the distribution of the rate parameters, $\\alpha_1$ and $\\alpha_2$.", fig.height=4>>=
-
-#simu2 <- read.csv("data/simulation-1000-M1-M2-M3.csv")
-
-ggplot(data = simu2) + 
-  geom_density(aes(x = estimate, fill = Model), alpha = .5) + 
-  facet_wrap(~parameter, scales = 'free') + 
-  ThemeNoNet + 
-  theme(legend.position = 'bottom')
-@
-
-<<corplots, fig.cap="A matrix of plots demonstrating the strong correlations between parameter estimate in our SAOMs.
-The strongest correlation within each model is between $\\beta_1$ and $\\beta_2$.", fig.width=8>>=
-
-#simu2 <- read.csv("data/simulation-1000-M1-M2-M3.csv")
-simu_spread <- simu2 %>% spread(parameter, estimate)
-ggpairs(simu_spread, columns = 3:8, ggplot2::aes(colour=Model, alpha = .5)) + theme(plot.title = element_text(size = 12,
-                                            face = 'plain', angle = 0, family = "Times New Roman"), 
-                  axis.title.x = element_text(size = 12,
-                                            face = 'plain', angle = 0, family = "Times New Roman", inherit.blank = T),
-                  axis.title.y = element_text(size = 12,
-                                            face = 'plain', angle = 90, inherit.blank = T),
-                  axis.text.x.top = element_text(size = 12,
-                                            face = 'plain', angle = 90, inherit.blank = TRUE), 
-                  axis.text.x.bottom = element_text(size = 12,
-                                            face = 'plain', angle = 90, inherit.blank = TRUE), 
-                  axis.text.y.left = element_text(size = 12,
-                                            face = 'plain', angle = 0, family = "Times New Roman", inherit.blank = T),
-                  axis.text.y.right = element_text(size = 12,
-                                            face = 'plain', angle = 0, family = "Times New Roman", inherit.blank = T),
-                  strip.text.x = element_text(size = 14,
-                                            face = 'plain', angle = 0, family = "Times New Roman", 
-                                            margin = margin(t = 3, r = 0, b = 3, l = 0, unit = "pt"), inherit.blank = T),
-                  strip.text.y =element_text(size = 14,
-                                            face = 'plain', family = "Times New Roman", angle = 90,margin = margin(t = 3, r = 3, b = 3, l = 3, unit = "pt"),inherit.blank = T),
-                  strip.background = element_rect(colour = "black", fill = "white")
-                    )
-@
-
-<<corplotsSenate, eval = FALSE>>=
-senM4est <- read.csv("data/congress/senateM4ests.csv")
-senM5est <- read.csv("data/congress/senateM5ests.csv")
-names(senM4est) <- c("alpha1", "alpha2", "alpha3", "beta1", "beta2", "beta5")
-names(senM5est) <- c("alpha1", "alpha2", "alpha3", "beta1", "beta2", "beta6")
-senM4est$Model <- "M4"
-senM5est$Model <- "M5"
-senM4est$Sim <- 1:nrow(senM4est)
-senM5est$Sim <- 1:nrow(senM5est)
-senM4est2 <- senM4est %>% gather(parameter, estimate, alpha1:beta5) 
-senM5est2 <- senM5est %>% gather(parameter, estimate, alpha1:beta6) 
-senM45est <- rbind(senM4est2, senM5est2)
-senM45est %>% spread(parameter, estimate)-> senM45est2
-ggpairs(senM45est2, columns = 3:9, ggplot2::aes(colour=Model, alpha = .5))
-@
-
-We also explore simulations from these different models given parameter values.
-Using the mean values of from the 1,000 fitted parameter values as the parameters in our models, we simulated 1,000 observations from each of our three models.
+\st{There are many possible ways to collect models together.
+We could look at the same models fit to different data, different models fit to the same data, or because of the nature of SAOMs, we could fit the same models to the same data many times to see how the simulations change.
+For the SAOMs, we decided that there were four collections that were most important:
+\begin{enumerate}
+\item the collections resulting from exploring the space of all possible models;
+\item the collections we get when varying model settings; 
+\item the results from fitting the same model form to different data;
+\item the results from fitting the same model to the same data many times.
+\end{enumerate}
+
+We chose these four collections because they each explore something different about SAOMs.
+The first takes the many dozens of parameters available to include in a SAOM into account, which easily translates into the second by showing how those many parameters affect the fitted models.
+Then, we look at how the same model looks when fit to different sets of dynamic network data.
+Finally, we look at the results from fitting the same model to the same data because the MCMCs and CTMCs that make up a SAOM lead to different parameter estimates every time, so it important to see how the results can vary.}
+
+\subsubsection{Exploring the space of all possible models}
+
+<<sigtesting, fig.show = 'hold', out.width='\\textwidth', fig.height=3, fig.cap="Significant effects for the two data sets, at the level of $\alpha = 0.10$ as calculated by the Wald-type test available in the \texttt{RSiena} software.">>=
+smfr_effs <- read_csv("data/effects_significance_smallFriends.csv")
+smfr_effs$inter1[is.na(smfr_effs$inter1)] <- "none"
+library(ggrepel)
+set.seed(12345678)
+smfr_effs %>% filter(type == "eval" & Waldpval <=0.10) %>% 
+ggplot(aes(x = estimate, y = Waldpval, color = inter1)) + 
+  #geom_errorbarh(aes(xmin = estimate - se, xmax = estimate + se), color = 'black') +  
+  geom_vline(xintercept = 0) + 
+  geom_point(size = 3) + 
+  geom_text_repel(aes(label = shortName)) + 
+  scale_color_brewer(palette = "Paired", name = "Interaction\nVariable") + 
+  ggtitle("Significant effects at level 0.10 for friendship data") + 
+  coord_flip() + 
+  ThemeNoNet
+sen_effs <- read_csv("data/congress/senateModelSigEffects.csv")
+#sen_effs$Waldpval <- parse_number(sen_effs$Waldpval)
+sen_effs$inter1[is.na(sen_effs$inter1)] <- "none"
+ggplot(data = sen_effs %>% filter(Waldpval <=0.10), aes(x = estimate, y = Waldpval, color = inter1)) + 
+  # geom_errorbarh(aes(xmin = estimate - se, xmax = estimate + se), color = 'black') +  
+  geom_vline(xintercept = 0) + 
+  geom_point(size = 3) + 
+  geom_text_repel(aes(label = shortName)) + 
+  scale_color_brewer(palette = "Paired", name = "Interaction\nVariable") + 
+  ggtitle("Significant effects at level 0.10 for senate data") + 
+  coord_flip() + 
+  ThemeNoNet
+@
+
+\st{The \texttt{RSiena} manual contains over eighty possible effects to include in the model.
+In order to select parameters to include in the models for our example data, we searched through the possible effects available to model given the data structure to find \textit{significant} effects.
+We tested for significance using the Wald-type tests built into \texttt{RSiena} for one-at-a-time effects testing.
+We start with the outdegree and reciprocity measures as the foundation of the models we fit, then add one evaluation effect, fit the model, test the additional effect for significance, and repeat for all possible parameters to add to the model.
+We performed this procedure for both the small friendship and the senate collaboration data.
+The results for significant effects are shown in Figure~\ref{fig:sigtesting}.
+We see in both the friendship data and the senate data results that most of the significant effects have absolute value less than ten.
+In addition, the $p$-values for the effects from the friendship data are more spread out than the $p$-values for the senate data, which are concentrated at about 0.02 or less.
+This may suggest that larger data sets tend to result in smaller $p$-values, just like a larger sample size results in smaller $p$-values in a $t$-test.}
+
+\subsubsection{Varying model settings}
+
+We have varied model settings already by choosing models M1, M2, M3 to fit to our small firendship network data set.
+In Section~\ref{sec:samemodsamedat}, we fit these models to the data 1,000 times, and in this section, we explore simulations from these three models given the mean values of from the 1,000 fitted parameter values as the parameters in our models.
+From each of these three models using the means of parameter estimates as our fixed parameter values, we simulated 1,000 observations from each of the three models.
 In this process, we condition on the first friendship network observation, and the second and third observations are simulated from the SAOM models with the given parameter values.
 From these simulations, we first create a visualization that represents an average network.
-To do this, we follow the same procedure as in Section~\ref{sec:minds}, counting occurrences of each possible edge in the simulations, resulting in a summary network with weighted edges representing the number of times an edge appeared in the simulated wave 2 when simulating from the SAOM 1,000 times.
-As in Figure~\ref{fig:summnetM1}, edges only appear in the average network if they appear more than 5\% of the time in the simulations.
-In Figure~\ref{fig:averagenet}, we show the first wave, the ``average'' network from the three models we fit, and the second wave.
-\st{Comparing the three averages to waves 1 and 2, we see that they have very similar structure to wave 1.
-Model 2, which included the transitive triplet parameter, seems to have created a larger connected component overall than models 1 and 3.
-In particular, if we look at the group of nodes $\{10,11,14\}$, we see they are very strongly connected within the three average networks, and they are completely separate from the other nodes in the true wave 2.
-None of the three average networks show node 16 gaining ties as it does in wave two, nor do they show nodes 4 and 7 becoming isolated.
-In Model 2, however, the ties to node 7 appear much weaker than in Model 1 or Model 2, suggesting that of the three, Model 2 may be the best fit for our data.}  %XXXX ST commented out after I realized I had made the same mistake as figure 7.
-XX This could be an indication that, despite the highly significant nature of the parameter $\beta_3$ according to the $t$-tests in \texttt{RSiena}, the transitive triplet parameter should not be included because it does not result in a structure similar to the data.
-
-<<averagenet, results='hide', out.width='99%',fig.cap='The node-link diagrams from the three "average" networks that we calculated are in the top row, and the true wave 1 and wave 2 data are shown in the bottom row above.
-There is some difference between the three models, but overall, these three models cannot capture the structure in the true second wave of data.', fig.height=6>>=
+
+<<averagenet, results='hide', out.width='\\textwidth',fig.cap='The node-link diagrams from the three "average" networks that we calculated are in the top row, and the true wave 1 and wave 2 data are shown in the bottom row above.
+There is some difference between the three models, but overall, these three models cannot capture the structure in the true second wave of data.', fig.height=5>>=
 #friend.data.w1 <- as.matrix(read.table("s50_data/s50-network1.dat"))
@@
 
-\st{What are some things that could go in here?:
-\begin{itemize}
-\item "average" networks -- you've already done this! Write it up, make several examples. How does the procedure for averaging generalize? What is it good for? 
-\item distributions of fitted parameters. again, you've already done this!!! just write it up and put it in here. 
-\item view correlations between model parameters under different model sets. eg 2 betas, 3 betas, 4 betas, etc.
-\end{itemize}
+To do create the average network visualization shown in Figure~\ref{fig:averagenet}, we follow the same procedure as in Section~\ref{sec:minds}, counting occurrences of each possible edge in the simulations, resulting in a summary network with weighted edges representing the number of times an edge appeared in the simulated wave 2 when simulating from the SAOM 1,000 times.
+As in Figure~\ref{fig:summnetM1}, edges only appear in the average network if they appear more than 5\% of the time in the simulations.
+In Figure~\ref{fig:averagenet}, we show the first wave, the ``average'' network from the three models we fit, and the second wave.
+\st{Comparing the three averages to waves 1 and 2, we see that they have very similar structure to wave 1.
+Model 2, which included the transitive triplet parameter, seems to have created a larger connected component overall than models 1 and 3.
+In particular, if we look at the group of nodes $\{10,11,14\}$, we see they are very strongly connected within the three average networks, and they are completely separate from the other nodes in the true wave 2.
+None of the three average networks show node 16 gaining ties as it does in wave two, nor do they show nodes 4 and 7 becoming isolated.
+In Model 2, however, the ties to node 7 appear much weaker than in Model 1 or Model 2, suggesting that of the three, Model 2 may be the best fit for our data.}  %XXXX ST commented out after I realized I had made the same mistake as figure 7.
+XX This could be an indication that, despite the highly significant nature of the parameter $\beta_3$ according to the $t$-tests in \texttt{RSiena}, the transitive triplet parameter should not be included because it does not result in a structure similar to the data.
+
+% \st{What are some things that could go in here?:
+% \begin{itemize}
+% \item "average" networks -- you've already done this! Write it up, make several examples. How does the procedure for averaging generalize? What is it good for? 
+% \item distributions of fitted parameters. again, you've already done this!!! just write it up and put it in here. 
+% \item view correlations between model parameters under different model sets. eg 2 betas, 3 betas, 4 betas, etc.
+% \end{itemize}
+%  }
+
+\subsubsection{Fitting the same model to different data}\label{sec:samemoddiffdat}
+% table of fitted M1 M2 M3
+\begin{table}
+\scalebox{0.9}{
+\begin{tabular}{l|ccc|ccc }
+  \hline
+  & \multicolumn{3}{|c|}{Friendship Data} & \multicolumn{3}{|c|}{Senate Data} \\
+  \hline
+  & M1 & M2 & M3 & M1 & M2 & M3 \\
+  $\alpha_1$ & \Sexpr{SFE$M1[1]} & \Sexpr{SFE$M2[1]} & \Sexpr{SFE$M3[1]} & \Sexpr{SSE$M1[1]} & \Sexpr{SSE$M2[1]} & \Sexpr{SSE$M3[1]} \\
+  $\alpha_2$ & \Sexpr{SFE$M1[2]} & \Sexpr{SFE$M2[2]} & \Sexpr{SFE$M3[2]} & \Sexpr{SSE$M1[2]} & \Sexpr{SSE$M2[2]} & \Sexpr{SSE$M3[2]}  \\
+  $\alpha_3$ & -- & -- & -- & \Sexpr{SSE$M1[3]} & \Sexpr{SSE$M2[3]} & \Sexpr{SSE$M3[3]} \\
+  $\beta_1$ & \Sexpr{SFE$M1[3]} & \Sexpr{SFE$M2[3]} & \Sexpr{SFE$M3[3]} & \Sexpr{SSE$M1[4]} & \Sexpr{SSE$M2[4]} & \Sexpr{SSE$M3[4]} \\
+  $\beta_2$ & \Sexpr{SFE$M1[4]} & \Sexpr{SFE$M2[4]} & \Sexpr{SFE$M3[4]} & \Sexpr{SSE$M1[5]} & \Sexpr{SSE$M2[5]} & \Sexpr{SSE$M3[5]}\\
+$\beta_3$ & -- & \Sexpr{SFE$M2[5]} & -- & -- & \Sexpr{SSE$M2[6]} & -- \\
+  $\beta_4$ & -- & --  & \Sexpr{SFE$M3[6]} & -- & -- & \Sexpr{SSE$M3[7]} \\
+  \hline
+\end{tabular}
 }
+\caption{\label{tab:meanests} The means (standard deviations) of parameter values estimated from repeated fittings of $M1, M2, M3$ to the small friendship network and the senate collaboration network.
+Each model was fit 1,000 times to the friend data, while each model was fit 100 times to the senate data.}
+\end{table}
+\subsubsection{Fitting the same model to the same data}\label{sec:samemodsamedat}
+
+%We look at collections produced ``from exploring the space of all possible models," ``by varying model settings", ``by fitting the same model to different datasets", and ``by treating each iteration as a model" \citep[p.~210-11]{modelvis}.
+Because fitting a SAOM requires MCMC methods, we also fit the same model many times, resulting in distributions of fitted parameters for one data set.
+
+To our small friendship network, we fit three models, M1, M2, and M3, using \texttt{RSiena} 1,000 times each.
+We then looked at the distribution of the fitted values, which are shown in Figure~\ref{fig:distests}.
+We can see from these distributions that the inclusion of the jumping transitive triplet parameter, $\beta_3$ is obviously affecting the distributions of the other four parameters included in all models, $\alpha_1$, $\alpha_2$, $\beta_1$, and $\beta_2$.
+\st{When $\beta_3$ is included, its estimate is postitive, meaning that friendships between two girls with different drinking behaviors tend to form when there is an intermediary who is already friends with the two girls.
+The inclusion of this parameter leads to increases in the rate parameters' estimates, suggesting that encouraging the transitive triplet behavior means that the girls would also change friends more frequently.
+The outegree parameter, $\beta_1$ decreases when $\beta_3$ is included, while the reciprocity parameter, $\beta_2$ increases.
+This implies the girls in the data prefer to form closer friend groups, as indicated by reciprocated ties and jumping transitive triplet formation, as opposed to being popular and having many friends.
+Having many friends who do not reciprocate is discouraged by M2.} In comparison with models M1 and M3, model M2 typically has higher estimates of the rate parameter, meaning that the inclusion of the covariate statistic in the model leads to higher estimates of the number of times, on average, a node gets to change its ties.
+It is not clear, however, that the addition of a parameter to the objective function \textit{should} effect the estimation of the rate parameters, so we continue to explore the collection of parameter estimates.
+
+To further investigate this potentially problematic relationship between the parameter values, we look at correlations between each of the parameter estimates in each model.
+In Figure~\ref{fig:corplots}, we examine correlations between each of pair of parameters within each model and overall.\nocite{ggally} The strongest correlation within each model is between $\beta_1$ and $\beta_2$, with absolute value of correlation between those two parameter values greater than 0.90 in all three models.
+The $\beta_1$ parameter is also highly correlated with the $\beta_3$ parameter within model M2, but it is not as highly correlated with the $\beta_4$ parameter in model M3.
+
+<<getdata>>=
+nulls <- read.csv("data/distribution_null_model.csv")
+alt <- read.csv("data/distribution_jumpTT_model.csv")
+alt2nd <- read.csv("data/distribution_dblpairs_model.csv")
+means <- simu2 %>% group_by(Model, parameter) %>% summarize(
+  means = mean(estimate)
+)
+null_mod_sv <- as.numeric(nulls[,-c(1,6,7)] %>% summarise_each(funs(mean)))
+alt_mod_sv <- as.numeric(alt[,-c(1, 7,8)] %>% summarise_each(funs(mean)))
+alt_mod2_sv <- as.numeric(alt2nd[,-c(1,7,8)] %>% summarise_each(funs(mean)))
+@
+
+<<distests, fig.cap="Distribution of fitted parameter values for our three SAOMs.
+The inclusion of $\\beta_3$ or $\\beta_4$ clearly has an effect on the distribution of the rate parameters, $\\alpha_1$ and $\\alpha_2$.", fig.height=4>>=
+
+#simu2 <- read.csv("data/simulation-1000-M1-M2-M3.csv")
+
+ggplot(data = simu2) + 
+  geom_density(aes(x = estimate, fill = Model), alpha = .5) + 
+  facet_wrap(~parameter, scales = 'free') + 
+  ThemeNoNet + 
+  theme(legend.position = 'bottom')
+@
+
+<<corplots, fig.cap="A matrix of plots demonstrating the strong correlations between parameter estimate in our SAOMs.
+The strongest correlation within each model is between $\\beta_1$ and $\\beta_2$.", fig.width=8>>=
+
+#simu2 <- read.csv("data/simulation-1000-M1-M2-M3.csv")
+simu_spread <- simu2 %>% spread(parameter, estimate)
+ggpairs(simu_spread, columns = 3:8, ggplot2::aes(colour=Model, alpha = .5)) + theme(plot.title = element_text(size = 12,
+                                            face = 'plain', angle = 0, family = "Times New Roman"), 
+                  axis.title.x = element_text(size = 12,
+                                            face = 'plain', angle = 0, family = "Times New Roman", inherit.blank = T),
+                  axis.title.y = element_text(size = 12,
+                                            face = 'plain', angle = 90, inherit.blank = T),
+                  axis.text.x.top = element_text(size = 12,
+                                            face = 'plain', angle = 90, inherit.blank = TRUE), 
+                  axis.text.x.bottom = element_text(size = 12,
+                                            face = 'plain', angle = 90, inherit.blank = TRUE), 
+                  axis.text.y.left = element_text(size = 12,
+                                            face = 'plain', angle = 0, family = "Times New Roman", inherit.blank = T),
+                  axis.text.y.right = element_text(size = 12,
+                                            face = 'plain', angle = 0, family = "Times New Roman", inherit.blank = T),
+                  strip.text.x = element_text(size = 14,
+                                            face = 'plain', angle = 0, family = "Times New Roman", 
+                                            margin = margin(t = 3, r = 0, b = 3, l = 0, unit = "pt"), inherit.blank = T),
+                  strip.text.y =element_text(size = 14,
+                                            face = 'plain', family = "Times New Roman", angle = 90,margin = margin(t = 3, r = 3, b = 3, l = 3, unit = "pt"),inherit.blank = T),
+                  strip.background = element_rect(colour = "black", fill = "white")
+                    )
+@
+
+<<corplotsSenate, eval = FALSE>>=
+senM4est <- read.csv("data/congress/senateM4ests.csv")
+senM5est <- read.csv("data/congress/senateM5ests.csv")
+names(senM4est) <- c("alpha1", "alpha2", "alpha3", "beta1", "beta2", "beta5")
+names(senM5est) <- c("alpha1", "alpha2", "alpha3", "beta1", "beta2", "beta6")
+senM4est$Model <- "M4"
+senM5est$Model <- "M5"
+senM4est$Sim <- 1:nrow(senM4est)
+senM5est$Sim <- 1:nrow(senM5est)
+senM4est2 <- senM4est %>% gather(parameter, estimate, alpha1:beta5) 
+senM5est2 <- senM5est %>% gather(parameter, estimate, alpha1:beta6) 
+senM45est <- rbind(senM4est2, senM5est2)
+senM45est %>% spread(parameter, estimate)-> senM45est2
+ggpairs(senM45est2, columns = 3:9, ggplot2::aes(colour=Model, alpha = .5))
+@
 
@@
 <<makinggif, fig.cap='A selection of images in the microstep animation.
-The selected edges and nodes are emphasized by changing the size and the color, then when edges are removed, they fade out, shrinking in size, while the nodes change color and shrink to blend in with the other nodes.', cache=TRUE, message = FALSE, warning = FALSE, fig.height=3>>=
+The selected edges and nodes are emphasized by changing the size and the color, then when edges are removed, they fade out, shrinking in size, while the nodes change color and shrink to blend in with the other nodes.', cache=TRUE, message = FALSE, warning = FALSE, fig.height=2.5>>=
 # Need old version of dplyr to get this to work: 
@@
 <<adjmatani, fig.cap='A selection of frames from the adjacency matrix visualization animation for one series of microsteps.
-At the beginning of the animation, shown in the top row, there are two clearly connected components: one in the top left corner, and one in the bottom right corner.
-By the end, the component in the top left has spread out, while the bottom right component has shrunk, but remains closely connected.', warning = FALSE, message = FALSE, results='hide', fig.height=4>>=
+(Ego and alter node labels are removed to declutter the graph.) At the beginning of the animation, shown in the top row, there are two clearly connected components: one in the top left corner, and one in the bottom right corner.
+By the end, the component in the top left has spread out, while the bottom right component has shrunk, but remains closely connected.', warning = FALSE, message = FALSE, results='hide', fig.height=2.5>>=
 adjmatall <- matrix(0, nrow = 16, ncol = 16)
@@
   scale_y_discrete(labels = rev(str_replace(order1, "V", ""))) + 
+    labs(x = "alter node (j)", y = "ego node (i)") + 
   ThemeNoNet + 
-  theme(legend.position = 'none') +
+  theme(legend.position = 'none', axis.text.x.bottom = element_blank(), 
+        axis.text.y.left = element_blank()) +
   coord_fixed() + 
@@
 p1 <- ggplot(data = msall1_df2) + 
-  geom_tile(aes(x = ms, y = -eid, fill = ms)) + 
+  geom_tile(aes(x = ms, y = -eid)) + 
+  annotate("text", x = c(-1, 84, 40), y = c(-30, -30, -60), label = c("Wave 1", "Wave 2", "Microsteps"), angle = c(90, 90, 0)) + 
   geom_vline(xintercept = c(0.5,81.5)) + 
-  scale_fill_continuous(low = "red", high = "blue") + 
+  #scale_fill_continuous(low = "red", high = "blue") + 
   ThemeNoNet + 
@@
 p2 <- ggplot(data = msall2_df2) + 
-  geom_tile(aes(x = ms, y = -eid, fill = ms)) + 
+  geom_tile(aes(x = ms, y = -eid)) + 
+  annotate("text", x = c(-1, 51, 25), y = c(-25, -25, -40), label = c("Wave 1", "Wave 2", "Microsteps"), angle = c(90, 90, 0))+ 
   geom_vline(xintercept = c(0.5,49.5)) + 
-  scale_fill_continuous(low = "red", high = "blue") +
+  #scale_fill_continuous(low = "red", high = "blue") +
   ThemeNoNet + 
@@
 # rep 3 
-msall3 <- netvizinf::listMicrosteps(dat = wave1friends,
-                      microsteps = filter(ansnullchainsw1w2, rep == 3))
-msall3_df <- plyr::rbind.fill(lapply(X = seq_along(msall3), FUN = function(i, x) {
-        x[[i]]$ms <- i - 1
-        return(x[[i]])
-    }, msall3))
-msall3_df <- msall3_df %>% dplyr::select(from, to, ms)
-msall3_df <- msall3_df %>% add_row(from = paste0("V",wave2$from),
-                      to = paste0("V",wave2$to), 
-                      ms = max(msall3_df$ms)+1)
-
-edges3 <- msall3_df %>% 
-  group_by(from, to) %>%
-  summarize(edgeappears = n()) %>% 
-  arrange(desc(edgeappears)) %>% ungroup() %>%
-  mutate(eid = row_number())
-
-msall3_df2 <- left_join(msall3_df, edges3, by = c("from" = "from", "to" = "to"))
-
+# msall3 <- netvizinf::listMicrosteps(dat = wave1friends,
+#                       microsteps = filter(ansnullchainsw1w2, rep == 3))
+# msall3_df <- plyr::rbind.fill(lapply(X = seq_along(msall3), FUN = function(i, x) {
+#         x[[i]]$ms <- i - 1
+#         return(x[[i]])
+#     }, msall3))
+# msall3_df <- msall3_df %>% dplyr::select(from, to, ms)
+# msall3_df <- msall3_df %>% add_row(from = paste0("V",wave2$from),
+#                       to = paste0("V",wave2$to), 
+#                       ms = max(msall3_df$ms)+1)
+# 
+# edges3 <- msall3_df %>% 
+#   group_by(from, to) %>%
+#   summarize(edgeappears = n()) %>% 
+#   arrange(desc(edgeappears)) %>% ungroup() %>%
+#   mutate(eid = row_number())
+# 
+# msall3_df2 <- left_join(msall3_df, edges3, by = c("from" = "from", "to" = "to"))
+# 
 # p3 <- ggplot(data = msall3_df2) + 

### 2017-09-19 paper.Rnw
fix compilation error

@@
 
-<<sigtesting, fig.show = 'hold', out.width='\\textwidth', fig.height=3, fig.cap="Significant effects for the two data sets, at the level of $\alpha = 0.10$ as calculated by the Wald-type test available in the \texttt{RSiena} software.">>=
+<<sigtesting, fig.show = 'hold', out.width='\\textwidth', fig.height=3, fig.cap="Significant effects for the two data sets, at the level of 0.10 as calculated by the Wald-type test available in the SIENA software.">>=
 smfr_effs <- read_csv("data/effects_significance_smallFriends.csv")

### 2017-09-20 paper.Rnw
so very very very close

@@
 library(GGally)
+devtools::install_github("sctyner/netvizinf")
 library(netvizinf)
@@
 
-\hh{XXX This paragraph ends a bit abruptly - what is the future state probability used for?} \st{XX The future state probability is used within the continuous time markov chain to move through the different network states.
+%\hh{XXX This paragraph ends a bit abruptly - what is the future state probability used for?} \st{XX The future state probability is used within the continuous time markov chain to move through the different network states.
 This needs more fleshing out.
@@
 This matrix describes the rate of change between two states of the CTMC process, and the rows of this matrix always add to zero.
-For directed networks with binary edge variables like the ones we will be working with, there are a very large number of possible states, $2^{n(n-1)}$: there are two possible states for an edge, and there are $n(n-1)$ edge relationships (excluding self-ties).
+For directed networks with binary edge variables like the ones we will be working with, there are a very large number of possible states for a directed network with $n$ nodes.
+We denote the state space as $\mathcal{X}$, a set which contains $2^{n(n-1)}$ states: there are two possible states for an edge, $\{0,1\}$, and there are $n(n-1)$ edge relationships because the network is directed and we exclude self-ties.
 The intensity matrix for a CTMC in a SAOM is then a square matrix of dimension $2^{n(n-1)} \times 2^{n(n-1)}$.
@@
 
-\subsection{The Models}
+\subsection{The Models}\label{sec:models}
 
@@
 
-<<tableofmeans>>=
-simu2 <- read.csv("data/simulation-1000-M1-M2-M3.csv")
-senM1est <- read.csv("data/congress/senateM1ests.csv")
-senM2est <- read.csv("data/congress/senateM2ests.csv")
-senM3est <- read.csv("data/congress/senateM3ests.csv")
-senM1est$Model <- "M1"
-senM2est$Model <- "M2"
-senM3est$Model <- "M3"
-senM1est$Sim <- 1:nrow(senM1est)
-senM2est$Sim <- 1:nrow(senM2est)
-senM3est$Sim <- 1:nrow(senM3est)
-senM1est %>% gather(parameter, estimate, Rate:recip) %>% dplyr::select(Sim, Model, parameter, estimate) -> senM1est2 
-senM2est %>% gather(parameter, estimate, Rate:jumpXTransTrip) %>% dplyr::select(Sim, Model, parameter, estimate) -> senM2est2 
-senM3est %>% gather(parameter, estimate, Rate:nbrDist2twice) %>% dplyr::select(Sim, Model, parameter, estimate) -> senM3est2 
-senateEsts <- rbind(senM1est2, senM2est2, senM3est2)
-senateEsts$parameter <- as.factor(senateEsts$parameter)
-levels(senateEsts$parameter) <- c("beta1", "beta3", "beta4", "alpha1", "alpha2", "alpha3", "beta2")
-SSE <-
-  senateEsts %>% group_by(Model, parameter) %>% 
-  summarize(meanEst = mean(estimate), sdEst = sd(estimate)) %>% 
-  mutate(val = paste0(sprintf("%.3f",meanEst), " ", "(", sprintf("%.3f", sdEst), ")")) %>% 
-  dplyr::select(Model, parameter, val) %>% 
-  spread(Model, val) %>% arrange(as.character(parameter))
-SFE <- simu2 %>% group_by(Model, parameter) %>% 
-  summarize(meanEst = mean(estimate), sdEst = sd(estimate)) %>% 
-   mutate(val = paste0(sprintf("%.3f",meanEst), " ", "(", sprintf("%.3f", sdEst), ")")) %>% 
-  select(Model, parameter, val) %>% 
-  dplyr::select(Model, parameter, val) %>% 
-  spread(Model, val) %>% arrange(parameter)
-@
-
 In Section~\ref{sec:samemoddiffdat}, we also fit models M1, M2, and M3 to the senate collaboration data for comparison.
-%Comparison of the parameter estimates of these models fit to the two different data sets is given in Table~\ref{tab:meanests}.
-In addition, we fit two more models, M4 and M5 to the senate data that were found to be significant according to the Wald-type test:
-
-\begin{align*}
-f_i(x)^{M4} & = \beta_1s_{i1} + \beta_2s_{i2} + \beta_5s_{i5} \\
-f_i(x, z)^{M5} & = \beta_1s_{i1} + \beta_2s_{i2} + \beta_6s_{i6}.
-\end{align*}
-
-In M4, $s_{i5} = \mathbb{I}(x_{i+} = x_{+i} = 0)$ is an indicator variable for isolated nodes, called the network-isolate effect.
+\st{For fitting M2 to the senate data, we use the number of bills authored by each senator as the node covariate for the jumping transitive triplet variable.
+In terms of the senate data, then, the value of $\beta_3$ should dictate how willing a senator is to coauthor a bill on which the author of the bill has a different level of authorship, assuming there is an intermediary between the two who has the same authorship level as the first.} %Comparison of the parameter estimates of these models fit to the two different data sets is given in Table~\ref{tab:meanests}.
+%In addition, we fit two more models, M4 and M5 to the senate data that were found to be significant according to the Wald-type test:
+
+%\begin{align*}
+%f_i(x)^{M4} & = \beta_1s_{i1} + \beta_2s_{i2} + \beta_5s_{i5} \\
+%f_i(x, z)^{M5} & = \beta_1s_{i1} + \beta_2s_{i2} + \beta_6s_{i6}.
+%\end{align*}
+
+%In M4, $s_{i5} = \mathbb{I}(x_{i+} = x_{+i} = 0)$ is an indicator variable for isolated nodes, called the network-isolate effect.
 This parameter attempts to measure the effect of a node being completely isolated in the network.
@@
 clintonSenate$Clinton <- ordered(clintonSenate$Clinton, levels = c("Yes", "No"))
-
-clintonSenate %>% filter(source %in% target)
 
@@
 
-<<sigtesting, fig.show = 'hold', out.width='\\textwidth', fig.height=3, fig.cap="Significant effects for the two data sets, at the level of 0.10 as calculated by the Wald-type test available in the SIENA software.">>=
+<<sigtesting, fig.show = 'hold', out.width='\\textwidth', fig.height=2, fig.cap="Significant effects for the two data sets, at a significance level of 0.10 or lower as calculated by the Wald-type test available in the SIENA software.">>=
 smfr_effs <- read_csv("data/effects_significance_smallFriends.csv")
@@
   geom_point(size = 3) + 
-  geom_text_repel(aes(label = shortName)) + 
+  #geom_text_repel(aes(label = shortName)) + 
   scale_color_brewer(palette = "Paired", name = "Interaction\nVariable") + 
@@
 In addition, the $p$-values for the effects from the friendship data are more spread out than the $p$-values for the senate data, which are concentrated at about 0.02 or less.
-This may suggest that larger data sets tend to result in smaller $p$-values, just like a larger sample size results in smaller $p$-values in a $t$-test.}
+This may suggest that larger data sets tend to result generally in smaller $p$-values, just like a larger sample size results in smaller $p$-values in a $t$-test.}
 
@@
 
-We have varied model settings already by choosing models M1, M2, M3 to fit to our small firendship network data set.
+We have varied model settings already by choosing models M1, M2, M3 to fit to our small friendship network data set.
 In Section~\ref{sec:samemodsamedat}, we fit these models to the data 1,000 times, and in this section, we explore simulations from these three models given the mean values of from the 1,000 fitted parameter values as the parameters in our models.
@@
 
-To do create the average network visualization shown in Figure~\ref{fig:averagenet}, we follow the same procedure as in Section~\ref{sec:minds}, counting occurrences of each possible edge in the simulations, resulting in a summary network with weighted edges representing the number of times an edge appeared in the simulated wave 2 when simulating from the SAOM 1,000 times.
-As in Figure~\ref{fig:summnetM1}, edges only appear in the average network if they appear more than 5\% of the time in the simulations.
-In Figure~\ref{fig:averagenet}, we show the first wave, the ``average'' network from the three models we fit, and the second wave.
+To create the average network visualization shown in Figure~\ref{fig:averagenet}, we follow the same procedure as in Section~\ref{sec:minds}, counting occurrences of each possible edge in the simulations, resulting in a summary network with weighted edges representing the number of times an edge appeared in the simulated wave 2 when simulating from the SAOM 1,000 times.
+As in Figure~\ref{fig:summnetM1}, edges only appear in the average networks if they appear more than 5\% of the time in the simulations.
+In Figure~\ref{fig:averagenet}, we show the ``average'' network from the three models we fit and the first and second waves of data.
 \st{Comparing the three averages to waves 1 and 2, we see that they have very similar structure to wave 1.
@@
 \subsubsection{Fitting the same model to different data}\label{sec:samemoddiffdat}
+
+\st{As mentioned in Section~\ref{sec:models}, we fit the models M1, M2, and M3 to both the small friendship network and the senate collaboration network.
+We fit each model to the friendship data 1,000 times, and to the senate data 100 times.
+The means and standard deviations of the parameter estimates for each combination of model and data are given in Table~\ref{tab:meanests} }
+
+<<tableofmeans>>=
+simu2 <- read.csv("data/simulation-1000-M1-M2-M3.csv")
+senM1est <- read.csv("data/congress/senateM1ests.csv")
+senM2est <- read.csv("data/congress/senateM2ests.csv")
+senM3est <- read.csv("data/congress/senateM3ests.csv")
+senM1est$Model <- "M1"
+senM2est$Model <- "M2"
+senM3est$Model <- "M3"
+senM1est$Sim <- 1:nrow(senM1est)
+senM2est$Sim <- 1:nrow(senM2est)
+senM3est$Sim <- 1:nrow(senM3est)
+senM1est %>% gather(parameter, estimate, Rate:recip) %>% dplyr::select(Sim, Model, parameter, estimate) -> senM1est2 
+senM2est %>% gather(parameter, estimate, Rate:jumpXTransTrip) %>% dplyr::select(Sim, Model, parameter, estimate) -> senM2est2 
+senM3est %>% gather(parameter, estimate, Rate:nbrDist2twice) %>% dplyr::select(Sim, Model, parameter, estimate) -> senM3est2 
+senateEsts <- rbind(senM1est2, senM2est2, senM3est2)
+senateEsts$parameter <- as.factor(senateEsts$parameter)
+levels(senateEsts$parameter) <- c("beta1", "beta3", "beta4", "alpha1", "alpha2", "alpha3", "beta2")
+SSE <-
+  senateEsts %>% group_by(Model, parameter) %>% 
+  summarize(meanEst = mean(estimate), sdEst = sd(estimate)) %>% 
+  mutate(val = paste0(sprintf("%.3f",meanEst), " ", "(", sprintf("%.3f", sdEst), ")")) %>% 
+  dplyr::select(Model, parameter, val) %>% 
+  spread(Model, val) %>% arrange(as.character(parameter))
+SFE <- simu2 %>% group_by(Model, parameter) %>% 
+  summarize(meanEst = mean(estimate), sdEst = sd(estimate)) %>% 
+   mutate(val = paste0(sprintf("%.3f",meanEst), " ", "(", sprintf("%.3f", sdEst), ")")) %>% 
+  select(Model, parameter, val) %>% 
+  dplyr::select(Model, parameter, val) %>% 
+  spread(Model, val) %>% arrange(parameter)
+@
+
 % table of fitted M1 M2 M3
@@
 \end{table}
-\subsubsection{Fitting the same model to the same data}\label{sec:samemodsamedat}
+
+<<compareModels, fig.cap="Density plots of the repeated estimates from fitting models M1, M2, and M3 to the friendship and senate example data.
+Note that the rate parameters are excluded since the two data sets have different numbers of waves.
+The first two parameters, outdegree and reciprocity, have the same relationship for both data sets: the outdegree parameter estimate is strongly negative, while the reciprocity estimate is strongly postitive.
+For the friendship data, the inclusion of the transitive triplet parameter has strong effect on the estimates of the other two parameters, while it does not affect the estimates of the other two parameters for the senate data.", fig.height = 3>>=
+simu2$data <- "friends"
+senateEsts$data <- "senate"
+allDataEsts <- rbind(senateEsts, simu2)
+allDataEsts$parameter <- as.factor(allDataEsts$parameter)
+allDataEsts$parameter <- ordered(allDataEsts$parameter, levels = c(paste0("alpha", 1:3),paste0("beta", 1:4)))
+ggplot(allDataEsts %>% filter(str_detect(parameter, "beta"))) + 
+  geom_density(aes(x = estimate, fill = Model), alpha = .5) + 
+  facet_grid(data~parameter, scales = "free", labeller = "label_both") + 
+  ThemeNoNet + 
+  theme(legend.position = 'bottom')
+@
+
+\st{Looking at Figure~\ref{fig:compareModels} and Table~\ref{tab:meanests}, we see a few patterns in the estimates from both models.
+First, we can see in both the table and the density of the estimates that the same relationship between the outdegree parameter, $\beta_1$, and the reciprocity parameter, $\beta_2$.
+In both data sets and across all three models, the estimates of $\beta_1$ are all negative and hover between -5 and -3, while the estimates of $\beta_2$ are all positive and hover between four and five.
+This suggests that in both data sets, nodes are \textit{discouraged} from forming ties that are outgoing without being reciprocated, while also being \textit{encouraged} to form outgoing ties to nodes that tie to them.
+In both data sets, people seem to want to have reciprocated relationships: teenage girls want be friends with other girls that reciprocate their friendship, and senators want to coauthor bills with senators who have also been coauthors on their bills.
+We explore the relationship between $\beta_1$ and $\beta_2$ further in Section~\ref{sec:samemodsamedat}.
+
+The inclusion of $\beta_3$, the jumping transitive triplet parameter, for the friendship data had a noticeable effect on the other parameters in the model.
+The same cannot be said for the inclusion of $\beta_3$ for the senate data.
+The covariate used in the senate data for the jumping transitive triplet calculation was the number of bills authored by the ego node in the given year.
+The number of bills authored by a senator varies wildly, from no bills to 114 bills authored in two years, so this effect could simply be nonsensical for the senate data, since senators are less likely to have the same number of bills authors than teenage girls are to have the same drinking behavior.
+Looking at the estimates of $\beta_4$, we see that the estimates for the senate data are near zero, suggesting this effect, which considers indirect ties, is not important for the senate data: indirect relationships do not describe the senate collaboration structure as much as they do the teenage friendship structure.
+}
 
@@
 Because fitting a SAOM requires MCMC methods, we also fit the same model many times, resulting in distributions of fitted parameters for one data set.
+
+\subsubsection{Fitting the same model to the same data}\label{sec:samemodsamedat}
+
+<<distests, fig.cap="Distribution of fitted parameter values for our three SAOMs.
+The inclusion of $\\beta_3$ or $\\beta_4$ clearly has an effect on the distribution of the rate parameters, $\\alpha_1$ and $\\alpha_2$.", fig.height=3>>=
+
+simu2 <- read.csv("data/simulation-1000-M1-M2-M3.csv")
+
+ggplot(data = simu2) + 
+  geom_density(aes(x = estimate, fill = Model), alpha = .5) + 
+  facet_wrap(~parameter, scales = 'free') + 
+  ThemeNoNet + 
+  theme(legend.position = 'bottom')
+@
 
@@
 
-To further investigate this potentially problematic relationship between the parameter values, we look at correlations between each of the parameter estimates in each model.
-In Figure~\ref{fig:corplots}, we examine correlations between each of pair of parameters within each model and overall.\nocite{ggally} The strongest correlation within each model is between $\beta_1$ and $\beta_2$, with absolute value of correlation between those two parameter values greater than 0.90 in all three models.
-The $\beta_1$ parameter is also highly correlated with the $\beta_3$ parameter within model M2, but it is not as highly correlated with the $\beta_4$ parameter in model M3.
-
-<<getdata>>=
+<<getdatam, eval = FALSE>>=
 nulls <- read.csv("data/distribution_null_model.csv")
@@
 alt_mod2_sv <- as.numeric(alt2nd[,-c(1,7,8)] %>% summarise_each(funs(mean)))
-@
-
-<<distests, fig.cap="Distribution of fitted parameter values for our three SAOMs.
-The inclusion of $\\beta_3$ or $\\beta_4$ clearly has an effect on the distribution of the rate parameters, $\\alpha_1$ and $\\alpha_2$.", fig.height=4>>=
-
-#simu2 <- read.csv("data/simulation-1000-M1-M2-M3.csv")
-
-ggplot(data = simu2) + 
-  geom_density(aes(x = estimate, fill = Model), alpha = .5) + 
-  facet_wrap(~parameter, scales = 'free') + 
-  ThemeNoNet + 
-  theme(legend.position = 'bottom')
 @
@@
 
+To further investigate the odd relationship between the parameter values, we look at correlations between each of the parameter estimates in each model.
+In Figure~\ref{fig:corplots}, we examine correlations between each of pair of parameters within each model and overall.\nocite{ggally} The strongest correlation within each model is between $\beta_1$ and $\beta_2$, with absolute value of correlation between those two parameter values greater than 0.90 in all three models.
+The $\beta_1$ parameter is also highly correlated with the $\beta_3$ parameter within model M2, but it is not as highly correlated with the $\beta_4$ parameter in model M3.
+\st{XX NEED TRANSITION XX }
+
 <<corplotsSenate, eval = FALSE>>=
@@
 Hiding the MCMC steps is practical and efficient if a researcher is primarily interested in fitting one model to a set of longitudinal network data, obtaining parameter estimates, \st{and drawing conclusions or making predictions}.
-We are more interested in \textit{how} the models are fit, so we extracted and explored the different steps of that process.
+We are more interested in \textit{how} the models are fit, so we extracted and explored the different steps of that process instead of allowing them stay hidden.
 
@@
 A series of microsteps is obtained by simulating from the model in its current state, $x(t_m)$ with current parameter values $\mathbf{\theta}_{0} = \{\alpha_{1_0}, \dots \alpha_{{m-1}_0}, \beta_{1_0}, \dots, \beta_{K_0}\}$, to the next state, $x(t_{m+1})$.
-This microstep process stops when the simulated network has acheived the same number of differences, $C$, from $x(t_m)$ as $x(t_{m+1})$, where
+This microstep process stops when the simulated network has achieved the same number of differences, $C$, from $x(t_m)$ as $x(t_{m+1})$, where
 $$C = \sum_{i \neq j} |x_{ij}(t_{m+1}) - x_{ij}(t_{m})|.$$ 
@@
 \st{Each tie change in the CTMC is referred to as one ``microstep".
-At each microstep}, an ``ego node" is selected to make a change, and the chosen ego node randomly makes one change in its ties according to the probabilities, $\{p_{ij}: i\neq j \in \{1, \dots, n\}\}$, determined by its objective function.
+At each microstep}, an ``ego node" is selected to make a change, and the chosen ego node randomly makes one change in its ties according to the probabilities, $\{p_{ij}: i\neq j \in \{1, \dots, n\}\}$ defined in Equation~\ref{eq:pij}, determined by its objective function.
 The options for change are (1) removing a current tie, (2) adding a new tie, or (3) making no change at all.
-Saving and exploring all of these steps is not computationally efficient if one is only interested in estimating parameter values, but they can be saved and extracted using options in \texttt{RSiena}, which is exactly what we did to create our visualizations.
+Saving and exploring all of these steps is not computationally efficient if one is only interested in estimating parameter values, but they can be saved and extracted using options in \texttt{RSiena}, which is what we did to create our visualizations.
 Between two network observations $x(t_m)$ and $x(t_{m+1})$, there can be dozens, hundreds, or even thousands of microsteps, depending on the size of the network and the number of changes between two network observations.
@@
 
-The first vizualization we present here is an animation of microsteps simulated to emulate the transition steps of the CTMC from wave 1 to wave 2 of the small friendship network example shown in Figure ~\ref{fig:alldataviz}.
+The first vizualization we present here is an animation of the simulated microsteps that form the transition steps of the CTMC from wave 1 to wave 2 of the small friendship network example shown in Figure~\ref{fig:alldataviz} when fitting model M1.
 Movies similar to this animation were used to visualize the changes of dynamic networks in \citet{dynnetviz}.
@@
 set.seed(34569234)
-create_net_animate(static_pall) + 
+froms <- static_pall %>% select(id = from, microstep, x = from.x, 
+        y = from.y, ms, .frame, size, color) %>% unique
+tos <- static_pall %>% select(id = to, microstep, x = to.x, y = to.y, 
+        ms, .frame) %>% unique
+tos <- left_join(tos, froms)
+nodedata <- rbind(froms, tos) %>% unique
+nodedata <- nodedata %>% replace_na(replace = list(size = 1, 
+        color = "#333333"))
+to_ani <- ggplot() + 
+  geom_curve(data = static_pall, aes(x = from.x, 
+        y = from.y, xend = to.x, yend = to.y, frame = .frame, 
+        colour = ecolor, size = esize), curvature = 0.1, arrow = arrow(angle = 20, 
+        length = unit(0.1, "inches"))) + 
+  geom_point(data = nodedata, 
+        aes(x = x, y = y, colour = color, size = size, frame = .frame)) +
+  scale_colour_identity() + scale_size_identity() + theme_void() + 
+        coord_fixed()
+to_ani + 
   facet_wrap(~.frame, nrow=2, labeller = 'label_both') + 
@@
 The top row of Figure~\ref{fig:makinggif} shows an edge being removed, and the bottom row shows one being added.
-In both cases, the ego nodes chosen to act change color from black to red, and they also increase greatly in size.
-In the case of an edge being removed, shown in the top row, the edge that currently exists is emphasized with the same color and size change that the node gets, and as the animation proceeds the edge shrinks to nothing, as the ego node shrinks and changes color back to its original black.
-If an edge is being added, as in the bottom row of the figure, the ego node's appearances changes in the sampe way as when the edge is being removed, whil the edge appears colored red from nothing, and grows to a large size, then changes color and size to match the rest of the edges, while the node shrinks and changes color to match the other nodes.
-
-We also use animation to view the changing structure of the adjacency matrix the microsteps.
+In both cases, the ego node chosen to act change color from black to red, and they also increase greatly in size.
+In the case of an edge being removed, in the top row, the edge that currently exists is emphasized with the same color and size change that the node gets, and as the animation proceeds the edge shrinks to nothing, as the ego node shrinks and changes color back to its original black.
+If an edge is being added, as in the bottom row of the figure, the ego node's appearance changes in the same way as when the edge is being removed, while the edge appears colored red from nothing, and grows to a large size, then changes color and size to match the rest of the edges, while the node shrinks and changes color to match the other nodes.
+
+\st{In the network animation, we see the possible steps of the unobserved CTMC process that is underlying the SAOM fit to the data.
+We see each part of the model come into play.
+First, we see the rate at which the nodes are selected to change.
+Then, we see the result of the actor maximizing its objective function by either deleting or adding a node.
+In addition, the layout of the nodes changes as edges are removed or added, which gives us a better sense of how the overall network structure changes with these individual tie changes.}
+
+We next use animation to view the changing structure of the adjacency matrix the microsteps.
 The adjacency matrices for the three waves of friendship data as shown in Figure~\ref{fig:alldataviz} are ordered by node id.
@@
 Viewing the adjacency matrices with this arbitrary ordering does not provide much information to the viewer about the underlying structure of the network.
-This lack of perceived structure would be exacerbated in an animation, so we adjust the ordering so that the viewer can perceive more of the structure of the graph.
+This lack of perceived structure would be exacerbated in an animation, so we adjust the ordering so that the viewer can better perceive the structure of the network.
 This process is known as matrix seriation \citep{seriation}.
@@
 $$ \mathbf{A}^{cum}_{ij} = \sum_{r = 0}^R X_{ij}(r). $$
-We then performed a principal component analysis on $\mathbf{A}^{cum}$, and used the values of the first principal component to order the vertices on the $x$ and $y$ axes for the adjacency matrix animation.
-For one such series of microsteps simulated by \texttt{RSiena}, we present the adjacency matrix ordered by the (arbitrary) vertex id alongside the seriated adjacency microsteps using the first principal component loading on the cumulative adjacency matrix, $\mathbf{A}^{cum}$, in Figure~\ref{fig:adjmatorder}.
+We then performed a principal component analysis (PCA) on $\mathbf{A}^{cum}$, and used the values of the first principal component to order the vertices on the $x$ and $y$ axes for the adjacency matrix animation.
+For one such series of microsteps simulated by \texttt{RSiena}, we present the adjacency matrix ordered by the (arbitrary) vertex id alongside the seriated adjacency matrix using the first principal component loading on the cumulative adjacency matrix, $\mathbf{A}^{cum}$, in Figure~\ref{fig:adjmatorder}.
 
@@
 
-Using the principal component analysis on $\mathbf{A}^{cum}$ to order the rows and columns of the adjacency matrix visualization clearly shows the two distinct connected components in the first wave of the network, which are difficult to find in the arbitrarily ordered visualization.
-We use this layout to fix the layout in the animation of one of the microstep process simulations.
-This animation, which is shown in Figure~\ref{fig:adjmatani} is very simple: a square appears or disappears in the animation as that edge appears or disappears in the microstep process.
+Using PCA on $\mathbf{A}^{cum}$ to order the rows and columns of the adjacency matrix visualization clearly shows the two distinct connected components in the first wave of the network, which are difficult to find in the arbitrarily ordered visualization.
+We also use the PCA seriated layout to fix the layout in the animation of one of the microstep process simulations.
+This animation, some frames of which are shown in Figure~\ref{fig:adjmatani} is very simple: a square appears or disappears in the animation as that edge appears or disappears in the microstep process.
 Through this animation, we can see edges appearing, and then later on disappearing.
-These in-between steps are not shown when we look at the network at our discrete observation points.
-By viewing this animation, we gain a better understanding of the dynamics of this model.
-<<histogramalters, fig.width = 8, fig.height=10, fig.cap="Each panel shows the theoretical (as lines) and empirical (as points) probabilities of the chosen ego node changing its tie to each of the other nodes.
+These in-between steps are not shown when we look at the network at our discrete observation points, so by viewing this animation we can gain a better understanding of the underlying dynamics of this model.
+
+<<histogramalters, fig.width = 7, fig.height=9, fig.cap="Each panel shows the theoretical (as lines) and empirical (as points) probabilities of the chosen ego node changing its tie to each of the other nodes.
 The color of the line indicates whether the tie from the ego to the alter node is being added, removed, or if there is no change to the network in this step.">>=
@@
 We also attempt to better understand the microstep process by visualizing the observed transition probabilites for the first microstep in the process.
-We only do the first step of many for now because the \texttt{RSiena} transition probabilities after the first step are only directly comparable for identical steps due to the conditioning on the current network state in the model.
-In other words, for $i \neq j \neq k \neq \ell \in \{1, \dots, n\}$, the probability of tie $x_{k\ell}$ changing in the second microstep is different in the case of $x_{ij}$ changing in the first step from the case of $x_{ij}$ not changing in the first step.
-Thus we have one transition probability for the first microstep taken for each of the 1,000 simulations to visualize.
-In Figure~\ref{fig:histogramalters}, we present each acting node, (ego node) and the resulting probabilities of tie changes.
+\st{We only do the first step of many because the \texttt{RSiena} transition probabilities for any two edges $i,j$ after the first step are only directly comparable for identical ego node states due to the conditioning on the current network state in the model.
+By ego node state, we mean the current set of incoming and outgoing ties to the ego node, $i$.
+Put more precisely, let the vector $X_i = (\{x_{ik}\}, \{x_{ki}\})_{k \neq i} \in \{0,1\}^{2(n-1)}$ represent the set of all tie variables involving node $i$ in the current network.
+For $p_{ij}$ to be comparable for any two network states $x$ and $x'$, the value of the vector $X_i$ must be the same in both states.
+It is possible to incorporate this information into our visualizations, but for now we look at only the first step, because we know without needing any complicated conditioning steps that the previous node state $X_i$ is identical in all cases.
+Thus we have 1,000 transition probabilities to examine: one transition probability for the first microstep taken for each of the simulations.
+}
+%Put more precisely, let the vector $X_i = (x_{ij}|_{j \neq i}, x_{ji}|_{j \neq i}$ for $i \neq j \neq k \neq \ell \in \{1, \dots, n\}$, $p_{ij}$ is not necessarily the same \st{ARGH NOT SURE THIS IS TRUE } if the current state of the network achieved from the first microstep, denoted $X(1)$, is $X(1) = x(i \leadsto \ell)$ than is different in the case of $x_{ij}$ changing in the first step from the case of $x_{ij}$ not changing in the first step.
+In Figure~\ref{fig:histogramalters}, we present each ego node and the resulting probabilities of tie changes.
 The probability shown by the bars is the theoretical probability, according to the objective function, of the ego node changing its tie to the alter node, while the probability shown by the points is the empirical probability of that change being made.
 The empirical probability is calculated by counting all instances of the ego node first, then computing the proportion of each different alter node change.
-In most cases, they are almost identical, which demonstrates that the algorithm is performing as expected.
+In most cases, they are almost identical, which demonstrates that the algorithm is performing about as expected.
+\st{There are, however, many steps that are never taken.
+Some ego nodes, like 1, 4, and 16, really explore the space of all possible outgoing ties.
+On the other hand, nodes like 13 and 15 explore very few possible outgoing ties.
+It is unclear why this would be happening; it may have something to do with the ties in wave 1, where 13 and 15 have multiple connections, while 1 and 16 have none, but there are several other counterexamples of well connected nodes that explore the space more.
+In addition, we can see that the ties changed most often are removing ties, not adding ties.
+This tracks with the overall pattern of the data: we see the most ties in the first wave of data, and the least ties in the third wave.}
 
@@
 The acting ego node is on the y-axis, and the alter node is on the x-axis.
-There were many ties with empirical probability zero.', fig.height=4>>=
+There were many ties with empirical probability zero.', fig.height=2.5>>=
 ansnullchains %>% 
@@
 The heatmap is noticeably sparse: %showing the lack of coverage in the model.
-of the 256 possible steps, only 103, or about 40\%, are taken in the 1,000 simulated chains.
+of the $16^2 = 256$ possible steps for the CTMC to take, only 103, or about 40\%, are taken in the 1,000 simulated chains.
+\st{This reinforces what we saw in Figure~\ref{fig:histogramalters}, where there are many paths not taken.
+This effect could only be exacerbated as more steps are taken in the CTMC, leading to a very large area of our network space, $\mathcal{X}$, completely unexplored by the SAOM model fitting process.}
 
@@
 The $x$ axis is the microstep number, with step 0 representing the first wave of data and the final step representing the second wave of data.
-We can see that many edges are underrepresented in this process: they are in the second wave, but never appear in the microsteps.">>=
+We can see that many edges are underrepresented in this process: they are in the second wave, but never appear in the microsteps.", fig.height=2.5>>=
 smallfriends <- read_csv("data/smallfriends4Geomnet.csv")
@@
   geom_tile(aes(x = ms, y = -eid)) + 
-  annotate("text", x = c(-1, 84, 40), y = c(-30, -30, -60), label = c("Wave 1", "Wave 2", "Microsteps"), angle = c(90, 90, 0)) + 
+  annotate("text", x = c(-1, 84, 40), y = c(-30, -30, -50), label = c("Wave 1", "Wave 2", "Microsteps"), angle = c(90, 90, 0)) + 
   geom_vline(xintercept = c(0.5,81.5)) + 
@@
 
-We also wanted to view the complete microstep process from the first wave, which we condition on in the model, to the second wave.
+We also wanted to better understand the entire microstep process from the first wave, all the way through to the second wave.
 The number of steps taken from wave 1 to wave 2 varies.
@@
 There are also some edges that were observed in wave 2 that don't appear at all in the microstep process in a given simulation.
+\st{So, even though the CTMC makes about the right \textit{number} of changes as it was designed to do, the changes it is making are not necessarily in the right direction.}
 
@@
 
-\st{We also combine 1,000 simulations from model M1 into a visualization like the one shown in Figure~\ref{fig:allsteps}.
+We also combine 1,000 simulations from model M1 into a visualization like the one shown in Figure~\ref{fig:allsteps}.
 We first assign each possible edge an edge ID number so that we can keep track of it throughout all the microsteps and all the simlations.
@@
 We also see, however, that the process struggles to focus in on the edges in the second wave of the data.
-Ideally, we would like to see more occurrences of the second wave of data generally.
-About half of the edges in wave two are in the bottom half when ordered by number of occurrences, meaning they aren't appearing as much as they would if the model was truly excellent at capturing the mechanisms of tie change in the network.
-In addition, there is not a distinct pattern or structure that we can see in this particular edge visualization.
+Ideally, we would like to see more occurrences of the edges which appear in the second wave of data.
+But, about half of the edges in wave two are in the bottom half when ordered by number of occurrences, meaning they do not appear as much as they would if the model was truly excellent at capturing the mechanisms of tie change in the network.
+\st{This solidifies what we found in Figures~\ref{fig:histogramalters} and~\ref{fig:heatmap}: the model M1 and the SAOM fitting process do not explore the data space enough to adequately capture the network change mechanism.} %In addition, there is not a distinct pattern or structure that we can see in this particular edge visualization.
 The edges that the model starts with appear more often in the first 50 or so microsteps than any other set of edges.
-This makes sense, but it would be better if the model explored the space more than it currently does.
-It could be that the simple model M1 isn't complicated enough for the friendship dynamics in the data, or it could be that the chains in the MCMCs "get stuck" when exploring the parameter space.}
+This makes sense, because we condition on the first wave, but it would be better if the model explored the space of possible edges more than it currently does.
+It could be that the simple model M1 is not complicated enough for the friendship dynamics in the data, or it could be that the chains in the MCMCs ``get stuck" when exploring the parameter space, but we cannot know this .}
 
@@
 
-We have used novel visualization methods in order to better understand the family of models known as stochastic actor-oriented model for social network.
-By looking at the underlying algorithms, exploring collections of these models, and viewing the model in the data space, we have been able to gain knowledge and appreciation for these complicated models and everything that goes into them.
+We have used novel visualization methods in order to better understand the family of models known as stochastic actor-oriented models for social network data.
+By looking at the underlying algorithms, visualizing collections of these models, and viewing the model in the data space, we have been able to gain knowledge and appreciation for these complicated models and everything that goes into them.
+
+\st{XX HH help please!}
 
@@
 The \texttt{RSiena} software is incredibly powerful, and can fit a whole slew of much more flexible stochastic actor-oriented models than we have examined here.
-If a researcher thinks the network structure or an actor covariate effects the rate of change of the networ, there is a way to incorporate that belief into the rate function of the SAOM.
+If a researcher thinks the network structure or an actor covariate effects the rate of change of the network, there is a way to incorporate that belief into the rate function of the SAOM.
 More than one actor-level covariate can be included in the model, and way more than three parameters can be included in the objective function itself.

### 2017-09-20 PaperToDo.Rmd
update to do -- cool 50 pages....

--- 2017-08-15 PaperToDo.Rmd
@@
 
-- [ ] Write an abstract - Gelman method/sanity check. First sentence of every paragraph of the intro & the very last sentence of intro. 
+- [X] Write an abstract - Gelman method/sanity check. First sentence of every paragraph of the intro & the very last sentence of intro. 
 - [ ] Submit to journal
@@
 - [X] Make labels of points larger
+- [X] Reverse order of x-y axes on adjmat vizs
 
@@
 
-- [ ] Proof read
+- [X] Proof read
 
@@
 
-- [ ] Proof read
+- [X] Proof read
 - [X] Make sure I've defined ego/alter nodes
-- [ ] Need to talk about the CTMC, rest of the SAOM model definition after the objective function discussion.
+- [X] Need to talk about the CTMC, rest of the SAOM model definition after the objective function discussion.
 
@@
 
-- [ ] Proof read
-- [ ] Finish section 3.1 (elaborate on probability)
+- [X] Proof read
+- [X] Finish section 3.1 (elaborate on probability)
+- [ ] Beef up transition after sect. 3.1.3
 
@@
 
-- [ ] Add histogram of number of tie changes from $t_1$ to $t_2$ in simulations 
-- [ ] Add another example of a data set for M-in-DS (Sect 4.1) (Congress)
-- [ ] Make gif available online
-- [ ] Insert link to full gif in paper
+- [X] Add another example of a data set for M-in-DS (Sect 4.1) (Congress)
+- [ ] Make node-link gif available online
+- [ ] Insert link to full node-link gif in paper
+- [ ] Make adjmat gif available online
+- [ ] Insert link to full adjmat gif in paper
+- [ ] Beef up transition before Sect 4.4
 - [X] Figure 6: reverse order of adjacency matrix y-axis
@@
 - [X] Describe Figure 8 more. What is going on??
-- [ ] Add geomnet example in 4.2
+- [X] Add geomnet example in 4.2
 - [X] p. 26-28 figure refs & code is broken
@@
 - [X] p. 34 code is broken
-- [ ] General proof reading
+- [X] General proof reading
 

### 2017-09-21 paper.Rnw
make HH's suggested changes

@@
 For any time point, $T$, where $t_{m} \leq T < t_{m+1}$, the waiting time to the next change opportunity by actor $i$ is exponentially distributed with expected value $\alpha_m^{-1}$.
-The conditional independence assumption is expressed in Equation~\ref{eq:c1}.
+The conditional independence of nodes in the network given their current ties is expressed in Equation~\ref{eq:c1}.
 
@@
 When node $i$ is given the chance to change a tie, %\hh{`we assume that they wish' sounds strange - don't assume wishes, state assumptions}
-it attempts to maximize the value of its objective function $f_i(\boldsymbol{\beta}, x)$ plus a random element, $U_i(x)$, to account for unknown attraction between nodes.
-In \citet{modelsSnijders}, it is recommended that the $U_i(x)$ be random draws from a type 1 extreme value distribution.
+it attempts to maximize the value of its objective function $f_i(\boldsymbol{\beta}, x)$ as well as a random element, $U_i(x)$, to account for unknown attraction between nodes.
 The additional random element is included to account for any random, unexplainable change in the network ties.
+\citet{modelsSnijders} recommends that the $U_i(x)$ be random draws from a type 1 extreme value distribution.
 % XX What is the thought behind that?
@@
 368]{snijders01}.
-This distribution, which is also known as the log-Weibull distribution, has probability distribution function, using $\mu$ for the mean parameter and $\sigma$ for the scale parameter, of
+This distribution, which is also known as the log-Weibull distribution, has probability distribution function of:
 
@@
 \end{equation}
-
-<<logweibull, fig.align='center', out.width = '.5\\textwidth', fig.height = 4, fig.cap='The probability distribution function for the type 1 extreme value distribution, also known as the log-weibull or Gumbel distribution with location parameter $ \\mu = 0$ and scale parameter $ \\sigma = 0$.'>>=
+using $\mu$ for the mean parameter and $\sigma$ for the scale parameter.
+
+<<logweibull, fig.align='center', out.width = '.5\\textwidth', fig.height = 4, fig.cap='The probability distribution function for the type 1 extreme value distribution, also known as the log-Weibull or Gumbel distribution with location parameter $ \\mu = 0$ and scale parameter $ \\sigma = 0$.'>>=
 library(reliaR)
@@
 The probability density function for the type 1 extreme value distribution is shown in Figure~\ref{fig:logweibull}.
-Using the distribution given in Equation~\ref{eq:dist1} with mean $\mu = 0$ and scale $\sigma = 1$ as \citet{modelsSnijders} suggests %\hh{XXX I'm not following - which distribution do you mean by `this' distribution?
+For mean $\mu = 0$ and scale $\sigma = 1$ as \citet{modelsSnijders} suggests %\hh{XXX I'm not following - which distribution do you mean by `this' distribution?
 The density above is the type 1 extreme value density.
@@
 This matrix $\mathbf{Q}$ is the foundation for estimation of a SAOM.
-Estimating the parameters in these models is difficult, but thanks to the SIENA software \citep{RSiena}, we have an accessible way to fit SAOMs to network observations.
-
-XXX HH can you help? I need a way better transition than that XXX
+%Estimating the parameters in these models is difficult, but thanks to the SIENA software \citep{RSiena}, we have an accessible way to fit SAOMs to network observations.
+
+%XXX HH can you help? I need a way better transition than that XXX
 }
@@
 \citeauthor{modelvis} outline three separate ideas, each of which are can be referred to simply as the "model": the model family, the model form, and the fitted model.
-The latter is primarily what one thinks of first when considering a model in a data analysis, where specified a model is fit to data, and parameter estimates and other numerical summaries, such as $R^2$ are reported.
+The latter is primarily what one thinks of first when considering a model in a data analysis, where a specified model is fit to data, and parameter estimates and other numerical summaries, such as $R^2$ are reported.
 In the context of SAOMs, the fitted model contains the form of the rate and objective functions, the estimated rate parameters, and the estimated objective function parameters.
@@
 The $\beta_1$ parameter is also highly correlated with the $\beta_3$ parameter within model M2, but it is not as highly correlated with the $\beta_4$ parameter in model M3.
-\st{XX NEED TRANSITION XX }
+\hh{It might therefore be advisable to consider only models that either allow $\beta_1$, or $\beta_2$.
+Looking at the high correlation with $\alpha$, we might swiths to a model without $\beta_1$.}
 
@@
 #plots[[i + 1]] <-
-  ggplot(data = adjmatall3 %>% filter(ms %in% c(1,3,5,7,9,11,13,43,45,47,49,51,53,55))) + 
-  geom_tile(aes(x = to, y = from, fill = as.factor(val)), color = 'black') + 
-  scale_fill_manual(values = c("white", "black")) + 
-  scale_x_discrete(position = "bottom", labels = str_replace(order1, "V", "")) + 
-  scale_y_discrete(labels = rev(str_replace(order1, "V", ""))) + 
+
+dat3 <- adjmatall3 %>% filter(ms %in% c(0,2,4,6,8,10,12,43,45,47,49,51,53,55))
+
+dat4 <- dat3 %>% select(-ms) %>% spread(frame, val) 
+dat5 <- dat4
+#p <- Sys.time()
+for(i in 4:16){
+  for(j in 1:nrow(dat5)){
+    if(dat4[j,i] != dat4[j,i-1]){
+      if(dat4[j,i] > dat4[j,i-1]){
+        dat5[j,i] <- 2
+      } else { dat5[j,i] <- 3}
+    }
+  }
+}
+#Sys.time() - p
+
+dat6 <- dat5 %>% gather(frame, val, 3:16) %>% mutate(frame = parse_number(frame))
+
+ggplot() + 
+  geom_tile(data = dat6, aes(x = to, y = from, fill = as.factor(val)), color = '#969696') +
+  scale_fill_manual(values = c("white", "black", "red", "white")) + 
+  geom_tile(data = dat6 %>% filter(val == 3), aes(x = to, y = from), fill = NA, color = 'red', size = .25) + 
+  scale_x_discrete(position = "bottom") + 
   labs(x = "alter node (j)", y = "ego node (i)") + 

### 2017-10-03 paper-child.Rnw
make the paper a child for the whole shebang

--- 2017-09-21 paper.Rnw
@@
-\documentclass[12pt]{article}
-% \usepackage{geometry}                % See geometry.pdf to learn the layout options. There are lots.
-% \geometry{letterpaper}                   % ... or a4paper or a5paper or ...
-%\usepackage{graphicx}
-\usepackage[font=small,skip=5pt]{caption}
-\usepackage{subcaption}
-\usepackage{animate}
-\usepackage{afterpage}
-\usepackage{amssymb}
-\usepackage{natbib}
-\usepackage{amsmath}
-\usepackage{amsfonts}
-% \usepackage{color}
-\usepackage{multirow}
-\usepackage{rotating}
-\usepackage[dvipsnames,svgnames,table]{xcolor}
-\usepackage{hyperref}
-\graphicspath{{figure/}}
-% \usepackage{endfloat} % Figures to the end of the document
-
-\DeclareGraphicsRule{.tif}{png}{.png}{`convert #1 `dirname #1`/`basename #1 .tif`.png}
-%---------------------------------------------------
-%                 Editing Commands
-\newcommand{\hh}[1]{{\color{magenta} #1}}
-\newcommand{\st}[1]{{\color{orange} #1}}
-
-%---------------------------------------------------
-%                 Placing Figures
-\renewcommand{\topfraction}{0.99}	% max fraction of floats at top
-\renewcommand{\bottomfraction}{0.8}	% max fraction of floats at bottom
-\renewcommand{\textfraction}{0.05}	% allow minimal text w. figs
-
-%---------------------------------------------------
-% Define new environment
-\newtheorem{theorem}{Theorem}[section]
-\newtheorem{algorithm}[theorem]{Algorithm}
-%---------------------------------------------------
-
-%\pdfminorversion=4
-% NOTE: To produce blinded version, replace "0" with "1" below.
-\newcommand{\blind}{0}
-
-% DON'T change margins - should be 1 inch all around.
-\addtolength{\oddsidemargin}{-.5in}%
-\addtolength{\evensidemargin}{-.5in}%
-\addtolength{\textwidth}{1in}%
-\addtolength{\textheight}{1.3in}%
-\addtolength{\topmargin}{-.8in}%
-
-\begin{document}
-
-%\bibliographystyle{natbib}
-
-\def\spacingset#1{\renewcommand{\baselinestretch}%
-{#1}\small\normalsize} \spacingset{1}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\if0\blind
-{
-  \title{\bf Model Visualization Techniques for a Social Network Model}
-  \author{Samantha Tyner\thanks{
-    The authors gratefully acknowledge funding from the National Science Foundation Grant \# DMS 1007697.
-All data collection has been conducted with approval from the Institutional Review Board IRB 10-347}\hspace{.2cm}\\
-    Department of Statistics and Statistical Laboratory, Iowa State University\\
-    and \\
-    Heike Hofmann\\
-    Department of Statistics and Statistical Laboratory, Iowa State University}
-  \maketitle
-} \fi
-
-\if1\blind
-{
-  \bigskip
-  \bigskip
-  \bigskip
-  \begin{center}
-    {\LARGE\bf Model Visualization Techniques for a Social Network Model}
-\end{center}
-  \medskip
-} \fi
-
-\bigskip
-\begin{abstract}
-Social networks have been studied for decades, beginning with a few foundational works, including the 1967 study, ``The Small World Problem" by Stanley Milgram.
-In this paper, we concentrate on one type of model for dynamic social networks: the stochastic actor-oriented models (SAOMs), introduced by Snijders (1996).
-Unlike other network models, SAOMs are not very well understood.
-We use model visualization techniques introduced in Wickham et al (2015) in order to make them a little less murky.
-The SAOMs are a prime example of a set of models that can benefit greatly from application of model visualization.
-With the help of static and dynamic visualizations, we bring the hidden model fitting processes into the foreground, eventually leading to a better understanding and higher accessibility of stochastic actor-oriented models for social network analysts.
-\end{abstract}
-
-\noindent%
-{\it Keywords:} social network analysis, model visualization, dynamic networks, network visualization, network mapping, animation
-\vfill
-
-\newpage
-\spacingset{1.45} % DON'T change the spacing!
-
-\tableofcontents
-\newpage
-
 <<setup, echo = FALSE, message = FALSE, warning = FALSE>>=
@@
 Finally, SAOMs and \texttt{RSiena} are able to also model behavior change of the actors in the network, which again is a capability we did not explore here.
-
-\bibliographystyle{asa}
-\bibliography{references}
-
-\end{document}

### 2017-10-03 paper.Rnw
make the paper a child for the whole shebang

--- 2017-09-19 paper.Rnw
@@
 
-<<setup, echo = FALSE, message = FALSE, warning = FALSE>>=
-options(replace.assign=TRUE,width=70, digits=3)
-require(knitr)
-opts_chunk$set(fig.path='figure/', cache.path='cache/', fig.align='center', fig.pos='h', out.width='.99\\textwidth', par=TRUE, cache=FALSE, concordance=TRUE, autodep=TRUE, message=F, warning=F, echo = FALSE, dev="cairo_pdf", fig.width = 6, fig.height = 6)#, root.dir = "~/Desktop/Dissertation/SAOM-removing-blindfold/")
+<<child-demo, child='paper-child.Rnw'>>=
 @
-
-<<pkgs>>=
-library(tidyverse)
-library(RSiena)
-library(network)
-library(sna)
-library(geomnet)
-library(GGally)
-library(netvizinf)
-library(RColorBrewer)
-library(gridExtra)
-library(cowplot)
-library(extrafont)
-loadfonts(quiet = T)
-
-ThemeNoNet <- theme_bw() %+replace% 
-            theme(plot.title = element_text(size = 12,
-                                            face = 'plain', 
-                                            angle = 0, 
-                                            family="Times New Roman"), 
-                  axis.title.x = element_text(size = 12,
-                                            face = 'plain', 
-                                            angle = 0,
-                                            family="Times New Roman"),
-                  axis.title.y = element_text(size = 12,
-                                            face = 'plain', 
-                                            angle = 90,
-                                            family="Times New Roman"),
-                  axis.text.x.top = element_text(size = 12,
-                                            face = 'plain', 
-                                            angle = 0,
-                                            family="Times New Roman"), 
-                  axis.text.x.bottom = element_text(size = 12,
-                                            face = 'plain', 
-                                            angle = 0,
-                                            family="Times New Roman"), 
-                  axis.text.y.left = element_text(size = 12,
-                                            face = 'plain', 
-                                            angle = 0,
-                                            family="Times New Roman"),
-                  axis.text.y.right = element_text(size = 12,
-                                            face = 'plain', 
-                                            angle = 0,
-                                            family="Times New Roman"),
-                  strip.text.x = element_text(size = 12,
-                                            face = 'plain', 
-                                            angle = 0, 
-                                            family="Times New Roman",
-                                            margin = margin(t = 3, r = 0, b = 3, l = 0, unit = "pt")),
-                  strip.text.y =element_text(size = 12,
-                                            face = 'plain', 
-                                            angle = 90,
-                                            family="Times New Roman",
-                                            margin = margin(t = 3, r = 3, b = 3, l = 3, unit = "pt")),
-                  strip.background = element_rect(colour = "black", fill = "white")
-                    )
-ThemeNet <- theme_net() %+replace% 
-            theme(plot.title = element_text(size = 12,
-                                            face = 'plain', angle = 0, family = "Times New Roman"), 
-                  strip.text.x = element_text(size = 12,
-                                            face = 'plain', angle = 0, family = "Times New Roman", 
-                                            margin = margin(t = 3, r = 0, b = 3, l = 0, unit = "pt")),
-                  strip.text.y =element_text(size = 12,
-                                            face = 'plain', angle = 0, family = "Times New Roman",margin = margin(t = 3, r = 0, b = 3, l = 0, unit = "pt")),
-                  strip.background = element_rect(colour = "black", fill = "white"))
-@
-
-\section{Introduction} 
-% \st{
-% \begin{itemize}
-% %\item What? - model visualization and SAOMs. briefly dsecribe what these things are
-% %\item Why? - bring light to the RSiena blackbox. Need ways to visualize network models, collections of networks. 
-% \item How?
-% \end{itemize}
-% %also talk about other dynamic network models briefly to put SAOMs in context.
-% }
-%\hh{XXX - talk a bit about social networks first, why we want to model them, and that usually the models are not dynamic.}
-\st{%XXX - okay, here is what i have done for that:
-Social networks have been studied for decades, beginning with a few foundational works, including the 1967 study, "The Small World Problem" by Stanley Milgram \citep{goldenberg09}.
-Examples of social networks include collaboration networks between academic researchers, friendship networks in a school or university, and trade networks between nations.
-In recent years, the study of social networks has grown in popularity due to an increase in the availability and access to social network data.
-There are many kinds of social networks, but there are not as many statistical models for social network data.} Some network models that have been applied to social networks include the exponential random graph model and latent space models.
-These models, however, are only for single instance networks.
-If we only have one network observation, or only care about one state of a complex network in time, like a snapshot of the World Wide Web, using the well-established models for single network observating is not a problem.
-If, on the other hand, we have many observations of social network over time, these models may not be appropriate because they do not explicitly allow for the network to change as time passes.
-When studying a network over time, referred to as a \textit{dynamic network}, we need a model that can take the time aspect of the network into account.
-Models for dynamic social networks have a great deal of modelling potential because of how realistic their structure can be.
-A social network does not form spontaneously: it evolves over time.
-Ties are formed and dissolved, and new actors join the social structure.
-Modeling the underlying mechanisms that create network changes over time is very complex but also provides potential to uncover hidden truths.
-
-\par In this paper, we concentrate on one type of model for dynamic social networks: the stochastic actor-oriented models (SAOMs), introduced by \citet{saompaper}.
-These models are fundamentally different from other social network models because they allow us to incorporate network \textit{and} actor statistics, where other models only rely on the network statistics, to model the changes in the network.
-Allowing the actor-level statistics to directly effect the structure of the network leads to a more practical and relevant approach to model change in a social network.
-In the ``real" world we expect people with common interests to be more likely to form relationships, and SAOMs allow us to incorporate this intuition in the modeling process.
-
-\st{
-%\par In order to estimate the parameters of  SAOMs the software SIENA, and its \texttt{R} implementation \texttt{RSiena}, was developed \cite{RSiena}.
-This software marks a huge contribution to the field of social network analysis., and when we began working with it we were eager to understand how it works.
-HH: we are still interested in this.
-%\hh{The next sentence is very vague - make it more specific: what aspect of SAOMs is not well understood?
-I think what you want is something like ``SAOMs are not very tractable analytically with traditional methods, e.g.
-likelihood functions turn into very complex objects because of the sheer number of parameters involved.
-Therefore, computationally more tractable solutions are used to fit estimators, in particular, SAOMs are fit using ...
-}
-Unlike other network models, SAOMs are not very well understood.
-They are relatively new, especially compared to the classic exponential random graph models, and they are not very tractable analytically.
-Likelihood functions quickly very complex objects to analyze due to the dependency structure inherent in the data.
-Therefore, computationally more tractable solutions are used to fit estimators, and in particular, % Traditional modeling steps, like determining the likelihood function, are very complex.
-Because of this,
-SAOMs are often fit to data using a series of Markov Chain Monte Carlo (MCMC) phases for finding method of moments estimators.
-In order to estimate the parameters of SAOMs, we use the software SIENA, and its \texttt{R} implementation \texttt{RSiena}, which was developed by \citet{RSiena}.
-This software marks a huge contribution to the field of social network analysis, but the many moving pieces involved in parameter estimation are largely ``behind the scenes" and hidden from the software user.} In this paper, in order to better understand the model-fitting process, we attempt to bring SAOM fits and the fitting process out of their black boxes, by combining the principals of network visualization with those of model visualization as discussed in \citet{modelvis}.
-%Here, we explore RSiena and SAOMs with innovative network visualizations, both static and dynamic.
-By bringing some light to the underlying methods and structures that are behind the scenes when fitting SAOMs to network data, we aim to help researchers working with the models better understand the implications and analyses of these models.
-
-\par We get into the SAOM black box by using model visualization (see \citet{modelvis}) to display the model in the data space, view collections of models instead of single models, and visually explore the process of fitting the SAOMs as opposed to looking at only the final output.
-Stochastic actor-oriented models are a prime example of a set of models that can benefit greatly from the application of model visualization.
-For instance, the models themselves include a continuous-time Markov chain (CTMC) that is completely hidden from the analyst in the model fitting process.
-Bringing the CTMC out of the black box and into the light through model visualization can provide researchers with insights into the underlying features of the model.
-Furthermore, SAOMs can include a great deal of parameters to be added to the model structure, each of which is attached to a network statistic.
-These statistics are often somewhat, if not highly, correlated, which causes high correlation between the associated parameters in a SAOM.
-By visualizing collections of SAOMs, we gain a better understanding of these correlations and find ways to deal with them and rectify their effects in the model.
-In addition, the estimation of the parameters in a SAOM relies on a Robbins-Monro algorithm, and the convergence checks for these estimates rely on simulation from the fitted model.
-Again, each of these steps are largely kept in the background of the estimation process.
-With the help of static and dynamic visualizations we bring the hidden model fitting processes into the foreground, eventually leading to a better understanding and higher accessibility of stochastic actor-oriented models for social network analysts.
-
-In Section~\ref{sec:netintro}, we introduce basic concepts of networks and network visualizations.
-In Section~\ref{sec:saoms}, we present the family of stochastic actor-oriented models for social network analysis.
-In Section~\ref{sec:ModelVis}, we combine concepts from Sections~\ref{sec:netintro} and \ref{sec:saoms} in an application of the model-vis paradigm, and conclude with a discussion in Section~\ref{sec:discussion}.
-
-\section{Networks and their Visualizations} \label{sec:netintro}
-
-\subsection{Introduction to Network Structures}
-
-Network data is of frequent interest to researchers in a wide array of fields.
-There are technological networks, like power grids or the internet, information networks, such as citation networks or the World Wide Web, biological networks, like neural networks, and social networks, just to name a few \citep{newman}.
-Each of these examples have one thing in common: their data \textit{structure}.
-There are always units of observation: the power stations, websites, neurons, and people, which we refer to throughout this paper as \textit{nodes} or \textit{actors}.
-There are also always connections of some kind between those units: the power lines, hyperlinks, electrical signals, and relationships, which we will call \textit{edges} or \textit{ties}.
-Networks might change over time, like when new websites and hyperlinks are added on the World Wide Web, or when there are new people and relationships in a friendship network.
-The nodes and edges themselves can also have inherent variables of interest, e.g.\ the institution of authors in a co-authorship network, or the number of times two authors have collaborated.
-
-The multiple layers of network data structures pose unique problems to network analysts.
-Some questions that network researchers may aim to answer are: How does the strength of a tie between two nodes affect the overall structure of the network?
-Do node-level differences affect the formation or dissolution of edges?
-Which views of the data are most informative for communicating significant effects and other results of statistical analyses of the network of interest?
-These are, of course, just a few broad questions, and we focus here on the latter, which we aim to answer through visual exploration of network data and models.
-
-\subsection{Visualizing Network Data} \label{sec:netvizintro}
-
-Network visualization, also called network mapping, is a prominent subfield of network analysis.
-Visualizing network data is uniquely difficult because of the structure of the data itself.
-Most, if not all, data visualizations rely on well-defined axes inherited from the data.
-If variables are numerical, histograms, scatterplots, or time series plots are straightforward to construct.
-If the variables are categorical, bar charts and mosaic plots are available to to the researcher.
-If the data are spatial, there is a well-defined region in which to view information.
-Network data, however, are much less cut-and-dried.
-
-There are two primary methods used to visualize networks: node-link diagrams and adjacency matrix visualization \citep{knuth2000, adjmatpro}.
-As a toy example, let us assume that we have five nodes, $\{1,2,3,4,5\}$, connected by five directed edges:  $\{2 \to 4, 3 \to 4, 1 \to 5, 3 \to 5, 5 \to 4\}$ We use this toy data set to demonstrate the two visualization methods in Figure~\ref{fig:toyplots}.
-
-<<toyex, echo = FALSE>>=
-set.seed(81393)
-nodes <- data.frame(id = 1:5, group = sample(LETTERS[1:2], size = 5, replace = T), stringsAsFactors = F)
-g <- igraph::as_data_frame(igraph::random.graph.game(n = 5, p = .4))
-toynet <- list(nodes = nodes, edges = rbind(g, c(5,4)))
-@
-
-The first method, the node-link diagram, represents nodes with points in 2D Euclidean space and then represents edges by connecting the points with lines when there is an edge between the two nodes.
-These lines can also have arrows on them indicating the direction of the edge for directed networks.
-But because there is typically no natural placement of the points unless they have important spatial locations, a random placement of the points is used, then adjusted via a layout algorithm, of which there are many \citep{drawingalgs}.
-
-<<toyplots, fig.show='hold', fig.cap="On the left, a node-link diagram of our directed toy network, with nodes placed using the Kamada-Kawai algorithm.
-On the right, the adjacency matrix visualization for that same network.", out.width='.49\\textwidth'>>=
-toynetdat <- fortify(as.edgedf(toynet$edges), toynet$nodes)
-ggplot(data = toynetdat) + 
-  geom_net(aes(from_id=from_id, to_id=to_id), labelon = T, size = 11, labelcolour = 'white', ecolour = 'black', vjust = .5, hjust = .5, fontsize = 9, arrowgap = .05, directed = T) +
-  ThemeNet +
-  theme(plot.title = element_text(size = 20)) + 
-  labs(title = "Node-Link Diagram")
-#ggsave("figure/toynetnld.pdf",plot = nld, width = 4, height = 4)
-
-ggplot(data = toynet$edges) +
-  #geom_tile(aes(x = from, y = -to), color = 'grey20', fill = 'black') +
-  geom_tile(aes(x = to, y = -from), color = 'grey20', fill = 'black') +
-  geom_segment(aes(x = x, xend = xend), y = -0.5, yend=-5.5, colour="grey30", data = data.frame(x = 0:5+.5, xend= 0:5+.5)) + 
-  geom_segment(aes(y=-y, yend=-yend), x = 0.5, xend=5.5, colour="grey30", data = data.frame(y = -0:5+.5, yend= -0:5+.5)) + 
-  scale_x_continuous(breaks = 1:5, labels = 1:5, limits = c(0.5,5.5), position = "top") +
-  scale_y_continuous(breaks = (-5):(-1), labels = 5:1, limits = c(-5.5,-.5)) +
-  ThemeNoNet + 
-  theme(panel.grid = element_blank(), aspect.ratio = 1, 
-        plot.title = element_text(size = 20), 
-        axis.title.x.top = element_text(size = 18),
-        axis.title.y.left = element_text(size = 18), 
-        axis.text.x.top  = element_text(size = 18),
-        axis.text.y.left  = element_text(size = 18)) +
-  labs(title = "Adjacency Matrix Visualization", x = "\nto (alter)", y = "from (ego)\n")
-@
-%\hh{XXX Explain each of these visualizations and either comment on the differences in the visualizations (colors) or don't show them. }
-
-Some commonly used layout algorithms, such as  the Kamada-Kawai layout \citep{kamadakawai} and the Fruchterman-Reingold layout \citep{fruchterman-reingold}, are designed to mimic physical systems, drawing the graphs based on the ``forces" connecting them.
-In these algorithms, the edges of the network act as springs pushing and pulling the nodes in a low dimensional (usually two-dimensional) space.
-Another algorithm uses multi-dimensional scaling, relying on distance metric and computing a matrix whose entries represent the ``distance" between every pair of nodes.
-There are also layout algorithms that use properties of the adjacency matrix, like its eigenstructure, to place the nodes in 2D space \citep{drawingalgs}.
-The node-link diagram using the Kamada-Kawai layout algorithm for our toy network is shown in Figure~\ref{fig:toyplots}.
-Unless otherwise stated, all other node-link diagrams in this paper will use the Kamada-Kawai layout.
-%Other algorithms use properties of the network's adjaceny matrix in order to place the nodes.
-%hh{XXX how does the adjacency matrix feature here?
-The adjacency matrix is a different mathematical representation of a network, but describes the same network as a node-link diagram.} \st{XXX I'm taking this to mean I should describe adjacency matrices more, so I've added a discussion below.
-XXX }
-
-\st{The second primary method for network visualization uses the adjacency matrix of the network.
-The adjaceny matrix of a network, $\mathbf{A}$, describes the edges of a network in matrix form.
-An entry $A_{ij}$ of $\mathbf{A}$, for two nodes $i\neq j$ in the network is defined as
-\[
-  A_{ij} =
-  \begin{cases}
-                                   1 &  \text{if an edge exists } i\to j \\
-                                   0 & otherwise
-  \end{cases}
-\]
-Note here that our edge variables are binary: we only consider the presence or absence of an edge.
-If the network has weighted edges, for example an email network where edge weights represent the number of emails sent from one person to another, the entries in the adjacency matrix are the edge weights instead of zeroes and ones.
-In an undirected network, $A_{ij} = A_{ji}$, but in a directed graph this is only true if there is an edge from $i$ to $j$ and from $j$ to $i$.
-Thus, $\mathbf{A}$ is always symmetric for undirected networks, and is symmetric for directed networks only if every edge between two nodes is reciprocated.}
-%\hh{XXX what is an adjacency matrix? - explain at this point in mathematical terms} \st{XXX see above XXX}
-An adjacency matrix visualization for our toy example is also shown in Figure~\ref{fig:toyplots}.
-
-Each type of visualization comes with its own advantages and disadvantages.
-For example, paths between two nodes in a network are easier to determine with node-link diagrams than with adjacency matrix visualizations \citep{adjmatviz}.
-In node-link diagrams, node-level information can be incorporated into the visualization by coloring or changing the shape of the points representing the nodes, and edge-level information can be incorporated by coloring the lines, or changing their thickness, linetype, or color.
-Incorporating a node-level variable into an adjacency matrix visualization is not as straightforward or simple, which is more focused on edges.
-Adjacency matrix visualization has been found to be particularly useful when the network is very complex, dense, or large, and experimental studies have shown adjacency matrix visualization to be superior to node-link diagrames for large networks.
-For example, for basic perceptual tasks on networks, including node and edge count, adjacency matrix visualizations outperform node-link diagrams as the size and density of the network increases \citep{adjmatviz}.
-One drawback of the adjacency matrix visualization that \citeauthor{adjmatviz} found was that edges are overrepresented for undirected graphs, due to the symmetry of $\mathbf{A}$: the edge $x_{ij}$ for $i \neq j$ appears in $\mathbf{A}$ twice: in $A_{ij}$ and $A_{ji}$, and so it also appears twice in the adjacency matrix visualization.
-%\hh{XXX What is the exact logic here that Ghoniem used to come to the conclusion that edges in undirected graphs are over-represented?
-Why would directed graphs make that better?
-wouldn't that just add to the number of edges?
-} \st{XXX in undirected nets, the adj mat is symmetric, where one edge ($x_{ij}$) is represented in entry $i,j$ in the matrix and in entry $j,i$.
-So, there are many repetitions in the matrix, which makes the network appear more dense than it really is.}
-\st{This, however, may actually be an advantage for \emph{directed} graphs, where exactly the correct number of edges is represented in a matrix visualization, due to the fact that the edges $x_{ij}$ and $x_{ji}$ are not interchangeable.
-A node-link diagram, however, may underrepresent the edge count if the edges $x_{ij}$ and $x_{ji}$ both exist and are drawn on top of one another.
-Ultimately, however t}here is not one "correct" way to visualize network information, and we will be using both \st{the node-link and adjacency matrix visualization} methods throughout this paper to explore social networks and stochastic actor-oriented models.
-
-\section{Stochastic Actor-Oriented Models for Longitudinal Social Networks} \label{sec:saoms}
-
-% \st{What do I need to communicate in this section? 
-% \begin{itemize}
-% \item What the models are for: dynamic social networks with actor-level covariates
-% \item What are the model components?
-rate function, objective function (network statistics, actor covariates, interactions between covariates and network statistics), time points ($t_m = 1, \dots, M$)
-% \item What is RSiena? What methods of estimation, convergence checks does it use? What are the algorithms? What kinds of models can it fit? 
-% \end{itemize}}
-
-%The stochastic actor-oriented models are very aptly named: they model dynamic networks that can account for the individual node characteristics when making the edge changes.
-The object of analysis is a social network that has been observed at several discrete time points.
-The nodes in the network can have associated covariate variables that may be incorporated into modeling the tie changes in the network.
-The model itself has two primary pieces: the rate of change of ties, and the objective function of the nodes, which determines the exact change that gets made.
-
-%We start with a set of network observations at $M$ discrete time points: $x(t_1), x(t_2) \dots, x(t_M)$.
-The first observation, $x(t_1)$ is always conditioned on.
-The number of nodes / actors / vertices is constant in all time points and is denoted $n$.
-The tie variables are directed and  binary, and are denoted at $x_{ij}(t_m)$ for time point $m$, and this is sometimes abbreviated to $x_{ij}$ in notation.
-If there is a tie from node $i$ to node $j$, then $x_{ij}=1$, and $x_{ij}=0$ otherwise.
-
-%\hh{mostly lit review plus a detailed description of the model and fitting process.}
-
-A Stochastic Actor-Oriented Model (SAOM) is a model that incorporates all three components of dynamic networks: edge, node, and time information.
-It models the change of a network over time, allowing for changes in network structure due to actor-level covariates.
-This model was first introduced by Snijders in 1996 \citep{saompaper}.
-The two titular properties of SAOMs, stochasticity and actor-orientation, are crucial to understanding networks as they exist naturally.
-Most social networks, even holding constant the set of actors over time, are ever-changing as relationships decay or grow in seemingly random ways, and most actors (or nodes) in social networks have inherent properties that could affect how they change their role within the network, and vice versa.
-
-\subsection{Definitions, Terminology, and Notation}
-
-In this paper, the term \textit{dynamic network} refers to a network, consisting of a fixed set of $n$ nodes, that is changing over time, and is observed at $M$ discrete time points, $t_1, \dots, t_M$ with $t_1 \le t_2 \le \dots \le t_M$.
-We denote the network observation at timepoint $t_k$ by $x(t_k)$.
-In the modelling process, we condition on the first observation, $x(t_1)$.
-The SAOM assumes that this longitudinal network of discrete observations is embedded within a continuous time Markov chain (CTMC), which we will denote $X(T)$.
-This process is almost entirely unobserved: we assume that the beginning of the process, $X(0)$, is equivalent to the first network observation $x(t_1)$, while the end of the process,$X(\infty)$, is equivalent to the last observation $x(t_M)$.
-Nearly all other parts of the process are unseen, with the exception of $x(t_2), \dots, x(t_{M-1})$.
-Unlike the first and last observations of the network, these ``in-between" observations do not have direct correspondence with steps in the continuous time Markov chain.
-Thus, the ``in-between" observations are considered to be ``snapshots" of the network at some point between two steps in the CTMC.
-The whole process $X(T)$ is a series of single tie changes that happen according to some pre-defined rate function, where one actor at a time is given the opportunity to add or remove one outgoing tie, or to not make any changes.
-Once an actor is chosen at random according to the rate function, it is ``given" the chance to change a tie, and it tries to maximize its utility function based on the current and near future states of the network.
-We expand on the model description further in the subsequent sections.
-
-\subsubsection{The Rate Function}
-
-For the network $x$ and each actor $i$ in the network, the number of times that an actor $i$ gets to change its ties, $x_{ij}$, to other nodes $j \neq i$ in the network is dictated by a \emph{rate function} $\rho(x, \mathbf{z}, \boldsymbol{\alpha})$, where $\boldsymbol{\alpha}$ are the parameters in the function $\rho$, and $x$ is the current network state, with covariates of interest $\mathbf{z}$.
-For this paper, we assume a simple rate function, $\rho(x, \mathbf{z}, \boldsymbol{\alpha}) = \alpha_m$ that is constant across all actors between observations at time $t_m$ to $t_{m+1}$, \st{thus our rate function is just a rate parameter in the overall model}.
-In general, SAOMs can incorporate covariate values and network statistics into the model, so that each node will have a different rate of change.
-%\hh{XXX that means that the function is essentially just a rate parameter, right?} \st{XXX yes XXX}
-Other modelling scenarios allow this rate to be more flexible,  e.g.\ a function that depends on the time period of observation, some actor-level covariates or some actor-level network statistics. 
-%\hh{How is the rate parameter different from a rate parameter?} \st{I think you meant how is a rate fn different from a rate parameter?
-I updated the language XXX->}%In our model with the simple rate function,
-\st{In our simple model with a simple rate parameter instead of a rate function,} the rate parameter dictates how quickly an actor $i$ gets an opportunity to change one of its ties to the other nodes in the network, $x_{ij}$, for $j \in \{1, \dots, n\}$ in the time period from $t_{m}$ to $t_{m+1}$.
-If $j = i$, no change in the network is made.
-The model also assumes that the actors $i$ are conditionally independent given their ties, $x_{i1}, \dots, x_{in}$ at the current network state.
-\st{Let $\tau(i|x,m) $ be the wait time until actor $i$ makes its next change from its current state in the network $x$.
-Note that $m$ indicates the number of the wave that is conditioned on in the SAOM.
-For any time point, $T$, where $t_{m} \leq T < t_{m+1}$, the waiting time to the next change opportunity by actor $i$ is exponentially distributed with expected value $\alpha_m^{-1}$.
-The conditional independence assumption is expressed in Equation~\ref{eq:c1}.
-
-\begin{equation} \label{eq:c1}
- \tau(i|x,m) | x_{i1}(m), \dots, x_{in}(m) \stackrel{\text{iid}}{\sim} Exp(\alpha_m)
-\end{equation}
-
- The waiting time to the next change opportunity by \textit{any} actor in the network is also exponentially distributed with expected value $(n\alpha_m)^{-1}$.
-The distribution of waiting time for the whole network to change, $\tau(x|m) = \sum_i \tau_i(m) |x_{i1}(m), \dots, x_{in}(m) $ can then be written as
-
-\begin{equation} \label{eq:c2}
- \tau(x|m) \sim Exp(n\alpha_m)
-\end{equation}
-
-} The parameter for the wait time for the whole network $n\alpha_m$ is the rate at which any tie change occurs.
-The estimation of this parameter is straightforward: a the method of moments is used to estimate the rate with the statistic $$C = \sum_i\sum_j |x_{ij}(t_{m+1}) = x_{ij}(t_m)|$$ which is the total number of changes from observation at time $t_m$ to the observation at time $t_{m+1}$.
-%In order to achieve the memorylessness property of a Markov process,
-
-% \hh{XXX we tried to estimate the rate based on the simulations - you could include a histogram of the number of changes for the 1000 simulations in the viz part XXX}
-% \st{XXX see line 1257 for plot. where should I put it? }
-% \hh{XXX Put a reference to the plot here.}
-
-\subsubsection{The Objective Function}
-
-%\hh{shouldn't that be independence? - make the conditional independence assumption above a labelled formula (e.g. C1) and then call it here }
-Because of the conditional \st{in}dependence assumptions given in Equation~\ref{eq:c1}, we can consider the objective function for each node separately, as only one tie from one node is \st{allowed to change} at a time.
-\st{The node $i$, which is the node that is chosen to change at the current time point, is called the \textit{ego} node.
-It has the potential to interact with all other nodes in the network, $j \neq i$.
-These nodes $j$, are referred to as \textit{alter} nodes, or simply \textit{alters}.
-These nodes are acted upon by the ego node, and they only act when they become the ego node at a subsequent time point in the CTMC.} For the ego node, $i$, in the current network state $x$, its objective function, which it tries to maximize, is written as
-\begin{equation} \label{eq:of1}
-f_i(\boldsymbol{\beta}, x) = \sum_k \beta_k s_{ik}(x, \mathbf{Z}),
-\end{equation}
-
-for $x \in \mathcal{X}$, the space of all possible directed networks with the $n$ nodes, and $\mathbf{Z}$, the matrix of covariates.
-The vector $\boldsymbol{\beta}$ contains the parameters of the model with corresponding network and covariate statistics, $s_{ik}(x, \mathbf{Z})$, for $k = 1,\dots, K$.
-%\hh{XXX ego node isn't defined yet - the description of the charts would be a good place for it.} \st{XX I added it above the obj function equation.
-Is it ok there?}
-Given the ego node, $i$, there are $n$ possible steps for the actor $i$ to take: either one of all current ties $x_{ij} = 1$ will be destroyed, a new tie will be created thatis currently $x_{ij}=0$, or no change will occur.
-
-The parameters, $\boldsymbol{\beta}$, correspond to various actor-level network statistics, $s_{ik}(x)$.
-%\hh{'There are always' ...
-that seems to come from a Snijder paper?
-Re-phrase this to be less vague: In the current implementation of RSiena (snijder) a model has minimally two parameters...}
-According to \citet[p.
-371]{snijders01}, there should be at least two parameters included in the model: $\beta_1$ for the outdegree of a node, and $\beta_2$ for the number of reciprocal ties held by a node.
-\st{These effects should seem familiar to readers used to working with the classical exponential random graph model (ERGM) for networks.
-The outdegree represents the propensity of nodes with a lot of outgoing ties to form more outgoing ties (the "rich get richer" effect), and the reciprocity parameter measures the tendency of outgoing ties to be returned within a network.
-The statistics corresponding to these effects are written in terms of the edge variables $x_{ij}$, for $i \neq j$.} In the \texttt{RSiena} software that we use to fit the SAOMs, there are over 80 possible parameters to add to the model.
-The formulas for the effects are provided in \citet{RSienamanual}.
-The parameters, $\beta_k$, in the model can be split up into two groups: first, the structural effects, whose estimation depends only on the structure of the network, like the outdegree and reciprocity parameters mentioned above.
-The parameters are included when the researcher hypothesizes that they will model underlying mechanisms of network change.
-They hope to answer questions such as, ``How does the existing network structure influence change in the network?" %\hh{XXX move the last colored section to the front of this paragraph.
-That sets up the structure nicely, e.g.
-`In determining the covariates of a SAOM, two sources of parameters are considered: ...} \st{XXX as I'm going through and uncommenting your comments (sorry....) this one has lost meaning a little bit.
-Could you give me more detail?
-I'm not sure which section is which at this point.
-(sorry again...)} \hh{I guess you addressed this comment very thoroughly :)}
-% and ``How do the behavior and characteristics of the nodes influence change in the network?" put this sentence below somewhere
-The second set of effects are referred to as the actor-level or covariate effects.
-These covariate effects also depend on the structure of the network, with the additional inclusion of node-level covariates of interest.
-The covariate effects are written in terms of the tie variables $x_{ij}$, but also in terms of the covariates, $\mathbf{Z}$.
-A table of some possible structural and covariate effects is given in Table~\ref{tab:effects}.
-For a complete list of the network and covariate statistics that can currently be included in the objective function, see \citet{RSienamanual}.
-
-\begin{table}
-\centering
-\begin{tabular}{m{4.5cm}m{4.5cm}m{3cm}}
-\multicolumn{3}{l}{\textbf{Structural Effects}} \\ \hline\hline
-outdegree  & $s_{i1}(x) = \sum_j x_{ij}$ & \includegraphics[width=2cm]{img/outdegree}\\ \hline
-reciprocity  & $s_{i2}(x) = \sum_j x_{ij}x_{ji}$ & \includegraphics[width=2cm]{img/reciprocity}\\ \hline
-transitive triplets  & $s_{i3}(x) = \sum_{j,h} x_{ij}x_{jh}x_{ih}$ & \includegraphics[width=1.5cm]{img/transtrip}  \\
-\\
-\multicolumn{3}{l}{\textbf{Covariate Effects}}\\  \hline \hline
-covariate-alter  & $s_{i4}(x) = \sum_j x_{ij}z_j$ & \includegraphics[width=3cm]{img/covaralter} \\  \hline
-covariate-ego  & $s_{i5}(x) = z_i\sum_j x_{ij}$ & \includegraphics[width=3cm]{img/covarego} \\  \hline
-same covariate & $s_{i6}(x) = \sum_j x_{ij} \mathbb{I}(z_i = z_j)$ & \includegraphics[width=3cm]{img/covarsim}%\\
-%jumping transitive triplets  & $s_{i7}(x) = \sum_{j \neq h} x_{ij}x_{ih}x_{hj} \mathbb{I}(z_i = z_h \neq z_j)$
-\end{tabular}
-\caption{\label{tab:effects} Some of the possible effects to be included in the stochastic actor-oriented models in RSiena.
-There are many more possible effects, but we only consider a select few here.
-For a complete list, see the RSiena manual \citep{RSiena}.}
-\end{table}
-
-When node $i$ is given the chance to change a tie, %\hh{`we assume that they wish' sounds strange - don't assume wishes, state assumptions}
-it attempts to maximize the value of its objective function $f_i(\boldsymbol{\beta}, x)$ plus a random element, $U_i(x)$, to account for unknown attraction between nodes.
-In \citet{modelsSnijders}, it is recommended that the $U_i(x)$ be random draws from a type 1 extreme value distribution.
-The additional random element is included to account for any random, unexplainable change in the network ties.
-% XX What is the thought behind that?
-} ``the type 1 extreme value distribution (or Gumbel distribution) with mean 0 and scale parameter 1" \citep[p.
-368]{snijders01}.
-This distribution, which is also known as the log-Weibull distribution, has probability distribution function, using $\mu$ for the mean parameter and $\sigma$ for the scale parameter, of
-
-\begin{equation} \label{eq:dist1}
-g(u|\mu, \sigma) = \frac{1}{\sigma}\exp\left\{-\left(\frac{u-\mu}{\sigma} + \exp^{-\frac{u-\mu}{\sigma}}\right)\right\}.
-\end{equation}
-
-<<logweibull, fig.align='center', out.width = '.5\\textwidth', fig.height = 4, fig.cap='The probability distribution function for the type 1 extreme value distribution, also known as the log-weibull or Gumbel distribution with location parameter $ \\mu = 0$ and scale parameter $ \\sigma = 0$.'>>=
-library(reliaR)
-x <- seq(-3,10,.01)
-y <- dgumbel(x = seq(-3,10,.01), sigma = 1, mu=0)
-qplot(x = x, y= y, geom = "line") + 
-  annotate('text', x = 6, y = .25, 
-        label = "g(u) = exp(-(u + exp(-u)))",size=5, family = "Times New Roman")  + 
-  theme(text = element_text(size = 20, family = "Times New Roman")) + 
-  labs(x = "u", y = "density of g(u)")
-@
-
-The probability density function for the type 1 extreme value distribution is shown in Figure~\ref{fig:logweibull}.
-Using the distribution given in Equation~\ref{eq:dist1} with mean $\mu = 0$ and scale $\sigma = 1$ as \citet{modelsSnijders} suggests %\hh{XXX I'm not following - which distribution do you mean by `this' distribution?
-The density above is the type 1 extreme value density.
-I got lost at the  ``objective function $f_i(\boldsymbol{\beta}, x)$ plus a random element, $U_i(x)$"} \st{XXX I hope it is clearer now above}
-is convenient because it \st{leads to a simple probability formula for} the probability that actor $i$ chooses to change its tie to actor $j$ \st{that can be written \textit{only}} in terms of the objective function.
-Let $p_{ij}(\boldsymbol{\beta}, x)$ be the probability that actor $i$ chooses to change its tie to actor $j$.
-Next, we write the network $x$ in its potential future state, $x(i \leadsto j)$, where the tie $x_{ij}$ has changed to $1-x_{ij}$.
-Then, the probility that the tie $x_{ij}$ changes is
-
-\begin{equation}\label{eq:pij}
-p_{ij}(\boldsymbol{\beta}, x) = \dfrac{\exp\left\{f_i(\boldsymbol{\beta}, x(i\leadsto j))\right\}}{\sum_{h \neq i} \exp\left\{f_i(\boldsymbol{\beta}, x(i \leadsto h))\right\}}
-\end{equation}
-
-When $i = j$ in $p_{ij}$, the numerator represents the exponential of the value of the objective function when evaluated at the current network state.
-When the value of the objective function is high at the current state, the probability of not making a change in a microstep is also high.
-In the CTMC, when actor $i$ may make a change, it chooses which tie $x_{i1, \dots, x_{in}}$ to change at random according to the probabilities $p_{ij}(\boldsymbol{\beta}, x)$.
-\st{The objective function and the resulting values of $p_{ij}$ are combined with the rate function to fully describe the CTMC that is used to model network change in a SAOM.}
-
-\hh{XXX This paragraph ends a bit abruptly - what is the future state probability used for?} \st{XX The future state probability is used within the continuous time markov chain to move through the different network states.
-This needs more fleshing out.
-Adding to the to-do list.}
-
-\subsubsection{Continuous Time Markov Chain (CTMC)}
-\st{
-In the continuous-time Markov chain literature, see for instance \citet{ctmc}, chains are characterized by their \textit{generator} or \textit{intensity} matrix $\mathbf{Q}$.
-This matrix describes the rate of change between two states of the CTMC process, and the rows of this matrix always add to zero.
-For directed networks with binary edge variables like the ones we will be working with, there are a very large number of possible states, $2^{n(n-1)}$: there are two possible states for an edge, and there are $n(n-1)$ edge relationships (excluding self-ties).
-The intensity matrix for a CTMC in a SAOM is then a square matrix of dimension $2^{n(n-1)} \times 2^{n(n-1)}$.
-Only one tie changes at a time in the CTMC, resulting in $n(n-1)$ reachable states from the current network state.
-Thus, the intensity matrix $\mathbf{Q}$ is very sparse, with only $n(n-1) + 1$ non-zero entries in each row.
-Note that $n(n-1)$ of these entries represent the possible states that are one edge different from a given state, while the additional non-zero entry is for the state to remain unchanged.
-All other entries in a row are structural zeroes because those network states cannot be reached from the current state in a single change.
-
-The two pieces of a SAOM, the rate function/parameter and the objective function, each contribute to the entries of the intensity matrix to describe the rate of change between two network states.
-The entries of $\mathbf{Q}$ are defined as follows: let $b \neq c \in \{1, 2, \dots, 2^{n(n-1)} \}$ be indices of two different possible states of the network, $x^b, x^c \in \mathcal{X}$.
-Then the $bc^{th}$ entry of $Q$ is:
-\[ q_{bc} = \begin{cases} 
-     q_{ij}= \alpha_m p_{ij}(\boldsymbol{\beta}, x^b) & \text{if } x^c \in \{x^b(i \leadsto j) | \text{ any } i \neq j \in \{1, \dots, n\}\} \\
-      0 & \text{if } \sum_i \sum_j |x_{ij}^c -x_{ij}^b | > 1 \\
-      -\sum_{i\neq j} q_{ij} & \text{if } x^b = x^c 
-   \end{cases}
-\]
-
-Thus, the rate of change between any two states, $x^b$ and $x^c$, that differ by only one tie $x_{ij}$, is the product of the rate at which actor $i$ gets to change a tie and the probability that the tie that will change is the tie to node $j$.
-This matrix $\mathbf{Q}$ is the foundation for estimation of a SAOM.
-Estimating the parameters in these models is difficult, but thanks to the SIENA software \citep{RSiena}, we have an accessible way to fit SAOMs to network observations.
-
-XXX HH can you help? I need a way better transition than that XXX
-}
-
-%### A SAOM as a CTMC {#saomctmc}
-
-%In order to fit this model definition back into the original context of the CTMC described in Section \@ref(dynamicnets), it must be written in terms of its intensity matrix, $\mathbf{Q}$.
-This matrix describes the rate of change between states of the process.
-For networks, there are a very large number of possible states, $2^{n(n-1)}$, so the intensity matrix is a square matrix of that dimension.
-But, thanks to the property of SAOMs that the states are allowed to change only one tie at a time, there are only $n$ possible states given the current state, $n-1$ of which are uniquely determined by the node $i$ that is given the opportunity to change.
-Thus, the intensity matrix $\mathbf{Q}$ is very sparse, with only $n(n-1) + 1$ non-zero entries in each row.
-Note that $n(n-1)$ of these represent the possible states that are one edge different from a given state, and the additional non-zero entry is for the state to remain the same.
-All other entries in a row are zero because those column states cannot be reached from the row state by just one change as dictated by the SAOM.
-The entries of $\mathbf{Q}$ are defined as follows: let $b \neq c \in \{1, 2, \dots, 2^{n(n-1)} \}$ be indices of two different possible states of the network, $x^b, x^c \in \mathcal{X}$.
-Then the $bc^{th}$ entry of $Q$ is:
-% \[ q(x^b, x^c) = \begin{cases} 
-%       q_{ij}(\alpha, \rho, \boldsymbol{\beta}, x^b) = \lambda_i(\alpha, \rho, x^b, m)p_{ij}(\boldsymbol{\beta}, x^b) & \text{if } x^c \in \{x^b(i \leadsto j) | \text{ any } i \neq j \in \mathcal{N}\} \\
-%       0 & \text{if } x^c \text{ differs from } x^b \text{ by more than 1 tie} \\
-%       -\sum_{i\neq j} q_{ij}(\alpha, \rho, \boldsymbol{\beta}, x^b)  & \text{if } x^b = x^c 
-%    \end{cases}
-% \]
-% 
-% Thus, the rate of change between any two states that differ by only one tie, $x_{ij}$, is the product of the rate at which actor $i$ gets to change a tie and the probability that the tie that will change is the tie to node $j$.^[Just to be clear, the change is from $x^b_{ij}$ to $x^c_{ij} = 1 - x^b_{ij}$.] Furthermore, the theory of continuous time Markov chains gives that the matrix of transition probabilities between observation times $t_{m-1}$ and $t_{m}$ is dependent only on the difference between timepoints, $t_m - t_{m-1}$.
-Following the same definition for transition probabilities in Section \@ref(dynamicnets), the matrix of transition probabilities is
-% $$e^{(t_m - t_{m-1})\mathbf{Q}},$$ 
-% where $\mathbf{Q}$ is the matrix defined above and $e^X$ for a real or complex square matrix $X$ is equal to $\sum_{k=0}^{\infty} \frac{1}{k!} X^k$.
-
-\subsection{Fitting Models to Data}
-
-To fit a SAOM to observations of a dynamic network, we use the package \texttt{RSiena} \citep{RSiena}.
-This package uses simulation methods to estimate parameter values using either the method of moments or maximum likelihood estimation.
-In this paper, use the method of moments estimation because the theory behind it was established in \citet{saompaper}, while the maximum likelihood estimation methods were not fully established until \citet{saomsml2010}, though \texttt{RSiena} contains capabilities to use maximum likelihood estimation.
-We also use the score function method for estimating the derivatives of the expected values, as opposed to the finite differences method, both of which are outlined in detail in \citet{SienaAlgs}.
-
-\par For the score function method, the SIENA software uses a Robbins-Monro algorithm (see \citet{robbinsmonro}) to estimate the solution of the moment equation
-$$ E_{\theta}S = s_{obs}$$
-where $\theta$ is the vector of rate and objective function parameters, and $s_{obs}$ is the observed vector of model statistics, $S$. The entire algorithm is provided in \citet{SienaAlgs}.
-
-There are three phases in the the SIENA algorithm, as described in \citet{RSienamanual, SienaAlgs}.
-The first phase performs initial estimation of the score functions for use in the Robbins-Monro procedure for method-of-moments estimation.
-The second phase carries out the Robbins-Monro algorithm and obtains estimates of the parameter values through iterative updates and simulation from the CTMC at current parameter values.
-The third phase uses the parameter vector estimated in phase two to estimate the score functions and covariance matrix of the parameter estimate, and also carries out convergence checks.
-In each of the the first two phases, the estimation procedure also uses ``microsteps" that simulate from the model as it exists in its current state in order to update either the score functions or the parameter estimates.
-These simulated microsteps are observed instances of the continuous-time Markov chain that is the backbone of the stochastic actor-oriented model.
-In Section~\ref{sec:ModelVis}, we further explore these phases in the SIENA method-of-moments algorithm through visualization, bringing them out of the ``black-box" and into the light.
-
-\subsection{Model Goodness-of-Fit}\label{sec:modgof}
-
-The \texttt{RSiena} software that fits the models to data also includes a goodnes-of-fit function for examining model fit, \texttt{sienaGOF()}.
-This function ``assess[es] the fit of the model with respect to auxiliary statistics of networks" \citep[p.~53]{RSienamanual}.
-Examples of auxiliary statistics include the out- or indegree distribution on the nodes, with the option for users to input their own statistics to examine.
-The goodness-of-fit is evaluated as follows:
-\begin{enumerate}
-\item The auxiliary statistics are computed on the observed data and on $N$ simulated observations from the model. Typically, $N=1000$. 
-\item The mean vector and covariance matrix of the statistics on the simulations from the model are computed.
-\item The Mahalanobis distance from the observed statistics to the distribution of the simulated statistics is computed using the mean and covariance found in step 2. 
-\item The Mahalanobis distance from each of the $N$ simulations to the same distribution is computed, and the Mahalanobis distance of the observed data is compared to this distribution of distances.
-\item An empirical $p$-value is found by computing the proportion of simulated distances found in step 4 that are as large or larger than the Mahalanobis distance from the data.
-A SAOM is thus considered a good fit if $p$ is large.
-\end{enumerate}
-
-% <<sienagof, fig.cap="Goodness of fit measures for two network statistics computed on the first simulated wave of a SAOM: indegree and outdegree cumulative distributions.
-On the x-axis is the degree value, and on the y-axis is the number of times that degree value occurs in the network.
-\\st{The p-value shown on the x-axis is the proportion of observations simulated from the model that have observed statistic values as large or larger than the values observed in the data.}", cache = TRUE, message = FALSE, warning = FALSE, fig.height=5>>=
-% load("data/ansnullpaper.rda")
-% library(lattice)
-% gof1 <- sienaGOF(ansnull, OutdegreeDistribution, varName ="friendship", join = FALSE)
-% gof2 <- sienaGOF(ansnull, IndegreeDistribution, varName ="friendship", join = FALSE)
-% 
-% sims1 <- data.frame(gof1$`Period 1`$Simulations)
-% dat1 <- gof1$`Period 1`$Observations
-% dat1 %>% data.frame() %>% gather(outdegree, val) %>% 
-%   mutate(outdegree = parse_number(outdegree)-1) -> dat1
-% p1 <- sims1 %>% 
-%   mutate(sim = row_number()) %>% 
-%   gather(outdegree, val, X1:X9) %>% 
-%   mutate(outdegree = parse_number(outdegree) - 1) %>% 
-%   filter(outdegree <= 6) %>%  
-%   ggplot(aes(x = outdegree, y = val)) + 
-%   geom_boxplot(aes(group = outdegree), size = .5, outlier.shape = "x", outlier.size = 5) + 
-%   geom_violin(aes(group = outdegree), bw = 2, fill = NA) + 
-%   geom_point(data = dat1, color = 'red') + 
-%   geom_line(data= dat1, color = 'red') + 
-%   geom_text(data = dat1, aes(label = val), hjust = -.5) + 
-%   labs(x = "Outdegree (p = 0.154)", y = "Statistic", 
-%        title = "Goodness-of-Fit: Outdegree distribution period 1") + 
-%   ThemeNoNet
-% 
-% sims2 <- data.frame(gof2$`Period 1`$Simulations)
-% dat2 <- gof2$`Period 1`$Observations
-% dat2 %>% data.frame() %>% gather(indegree, val) %>% 
-%   mutate(indegree = parse_number(indegree)-1) -> dat2
-% p2 <- sims2 %>% 
-%   mutate(sim = row_number()) %>% 
-%   gather(indegree, val, X1:X9) %>% 
-%   mutate(indegree = parse_number(indegree) - 1) %>% 
-%   filter(indegree <= 6) %>%  
-%   ggplot(aes(x = indegree, y = val)) + 
-%   geom_boxplot(aes(group = indegree), outlier.shape = "x", outlier.size = 5) + 
-%   geom_violin(aes(group = indegree), bw = 2, fill = NA) + 
-%   geom_point(data = dat2, color = 'red') + 
-%   geom_line(data= dat2, color = 'red') + 
-%   geom_text(data = dat2, aes(label = val), hjust = -.5) + 
-%   labs(x = "Indegree (p = 0.063)", y = "Statistic", 
-%        title = "Goodness-of-Fit: Indegree distribution period 1") + 
-%   ThemeNoNet
-% 
-% 
-% gridExtra::grid.arrange(p1, p2)
-% # sum(gof1$`Period 1`$SimulatedTestStat >= gof1$`Period 1`$ObservedTestStat)/1000
-% # [1] 0.154
-% # > sum(gof2$`Period 1`$SimulatedTestStat >= gof2$`Period 1`$ObservedTestStat)/1000
-% # [1] 0.063
-% 
-% @
-
-The \texttt{plot.sienaGOF()} function allows us to visualize this fit.
-%In Figure~\ref{fig:sienagof}, we provide an example of what the goodness-of-fit plot output looks like for indegree and outdegree statistics for a small model.
-This function draws a box plot and a violin plot at each value of the statistic of interest observed in the simulations: \st{on the $x$-axis, the out(in)degrees observed in the data ($0, 1, 2, 3, \dots$), and on the $y$-axis, the cumulative number of times that out(in)degree value appears in the simulations and the data.} In order to compare the distribution of the counts of nodes with the specified degree as calculated on the simulated networks to the counts observed in the true data, red points connected by red lines representing the observed values are superimposed on the boxplots.
-If the red points lie "well within" the simulated values, the model is a good fit to the data.
-\st{This plot is not shown because it is not intuitive for understanding network data.
-Boxplots separated by the many outdegree values observed do not communicate \textit{how} the nodes are connected, just that some have more connections than others.
-In order to understand the fit of the model, we should try to understand how well the model captures the\textit{overall} structure of the network, not just one or two summary measurements of that structure.} %The model examined in Figure~\ref{fig:sienagof} appears to do a better job of capturing the outdegree distribution than the indegree distribution of the data.
-In Section~\ref{sec:minds}, we propose a new way of visualizing goodness-of-fit that uses the traditional node-link diagram to visualize the entire network instead of numerical summaries of the the network.
-
-\subsection{Example Data}
-
-<<adjmatviz, fig.align='center', fig.cap="A visualization of the adjacency matrices of the three waves of network observations in the ``Teenage Friends and Lifestyle Study'' data.
-The subset we will be using is outlined in red.", results='hide',fig.show='asis', fig.width=8, fig.height=3>>=
-source("code/00e_small_friends.R")
-view_rs2$wave <- factor(view_rs2$wave)
-levels(view_rs2$wave) <- c("Wave 1", "Wave 2", "Wave 3")
-ggplot(data= view_rs2, aes(x = y, y=-x)) + 
-  geom_tile(aes(fill = as.factor(value))) + 
-  scale_fill_manual("i likes j", labels=c("no", "yes"), 
-                    values = c('white', 'black')) +
-  geom_rect(data= NULL, inherit.aes = FALSE, color = 'red',
-            aes(xmin = 19.5, xmax = 35.5, ymax = -19.5, ymin = -35.5, fill =NA)) + 
-  facet_wrap(~wave) + theme_bw() +
-  labs(x = "node j (alter)", y = "node i (ego)") + 
-  ThemeNoNet + 
-  theme(aspect.ratio=1, axis.text = element_blank(), axis.ticks = element_blank(), panel.grid = element_blank()) 
-@
-
-<<alldataviz, fig.cap="The smaller friendship network data we will be modelling throughout the paper.
-In the first wave, we can see there are two large, separate friend group.
-By the second wave, three students with heavier drinking behavior have separated from their group, while the others become members of the other group.
-By the third wave, the large group has almost completely broken up.
-We want to capture these changes with our SAOMs.", out.width='.99\\textwidth', fig.height=3>>=
-actual1 <- merge(data.frame(as.edgelist(as.network(fd2.w1))), 
-                 data.frame(id = 1:16, drink = drink2[,1]),
-                 by.x = "X1", by.y = "id", all = T)
-actual1$Wave <- 1 
-actual2 <- merge(data.frame(as.edgelist(as.network(fd2.w2))),
-                 data.frame(id = 1:16,drink = drink2[,2]),
-                 by.x = "X1", by.y = "id", all = T)
-actual2$Wave <- 2
-actual3 <- merge(data.frame(as.edgelist(as.network(fd2.w3))),
-                 data.frame(id = 1:16,drink = drink2[,3]),
-                 by.x = "X1", by.y = "id", all = T)
-actual3$Wave <- 3
-alldat <- rbind(actual1, actual2, actual3)
-#alldat$X1 <- paste0("V", alldat$X1)
-alldat$X1 <- as.factor(alldat$X1)
-#alldat$X2 <- ifelse(!is.na(alldat$X2), paste0("V", alldat$X2), NA)
-alldat$X2 <- as.factor(alldat$X2)
-#write_csv(alldat, "smallfriends4Geomnet.csv")
-alldataviz <- ggplot(data = alldat, 
-       aes(from_id = X1, to_id = X2, color = as.ordered(drink), group = as.ordered(drink))) +
-  geom_net(fiteach = FALSE, directed=TRUE, layout.alg = 'fruchtermanreingold', size = 5, arrowsize = .25, arrowgap = .05, ecolour = 'grey40', labelon=T, labelcolour = 'black', fontsize = 4, vjust = .5) +
-  scale_color_brewer(palette = "YlOrRd", name = "Drinking\nBehavior", labels = c("never", "1-2 times per year", "once per month", "once per week")) + 
-  ThemeNet +
-  facet_wrap(~Wave, labeller = "label_both") + 
-  theme(legend.position = 'bottom', 
-        panel.background = element_rect(fill = NA, color = "grey30"), 
-        legend.text = element_text(size = 12, family = "Times New Roman"), 
-        legend.title = element_text(size = 12, family = "Times New Roman"))
-alldataviz
-@
-
-To guide our visual exploration of stochastic actor-oriented models, we use two data sources.
-The first is a subset of the 50 actor dataset from the ``Teenage Friends and Lifestyle Study" that is provided on the \texttt{RSiena} webpage.
-These data come from \citet{friendsdata}, and we chose to only work with a subset of the data to make network visualizations less busy and to make any changes in the network more noticeable.
-To determine which subset to select, we visualized all waves of the full network using the adjacency matrix visualization approach, which we show in Figure~\ref{fig:adjmatviz}.
-This adjacency matrix visualization is different from the one in Figure~\ref{fig:toyplots} because it does not show the node IDs on the axes.
-The network in Figure~\ref{fig:adjmatviz} is much larger than our toy data example, so we remove the node labels to remove clutter.
-In all three adjacency matrices, the ego and the alter nodes are ordered by node ID, 1-50, which were determined arbitrarily by the relationships in wave 1.
-The subset we selected is outlined in red in the visualization.
-This subset contained actors 20 through 35 and the ties between them, as well as the drinking behavior of each actor at each of the three waves.
-This specific subset was chosen because it showed somewhat higher connectivity than other subsets, as we've emphasized in the visualizations of the three network observations in Figure~\ref{fig:adjmatviz}.
-For model fitting, we condition on wave 1 and estimate the parameters of the models from the second and third waves.
-We will also be working with one actor level categorical covariate, drinking behavior.
-This variable has five values in the original data: (1) does not drink, (2) drinks once or twice a year, (3) drinks once a month, (4) drinks once a week, and (5) drinks more than once a week.
-The network and the actor covariate values are visualized using a node-link diagram in Figure~\ref{fig:alldataviz}.
-\st{In the node-link diagram, the nodes are colored according to the drinking behavior of that student.
-Over time, we can see that the students tend to drink more and become increasingly isolated into smaller groups.
-An analysis of this type of data with a SAOM should capture these dynamics in a way that allows the researcher to draw conclusions about the nature of the network and behavioral forces at play.}
-
-\st{The second data example we use is a collaboration network in the United States Senate during the $111^{th}$ through $114^{th}$ Congresses.
-These sessions of congress correspond to the years of Barack Obama's presidency, from 2009-2016.\footnote{Details of how this data can be downloaded are provided by François Briatte at \url{https://github.com/briatte/congress}}.
-In this network, ties are directed from senator $i$ to senator $j$ when senator $i$ signs on as a cosponsor to the bill that senator $j$ authored.
-There are (somewhat surprisingly) many hundreds of ties between senators when they are connected in this way, so we simplify the network by computing a single value for each senator-senator collaboration called the \textit{weighted propensity to cosponsor} (WPC).
-This value is defined in \citet{senate} as
-\begin{equation}\label{eq:sen1}
-WPC_{ij} = \dfrac{\sum\limits_{k=1}^{n_j} \frac{Y_{ij(k)}}{c_{j(k)}}}{\sum\limits_{k=1}^{n_j} \frac{1}{c_{j(k)}}}
-\end{equation}
-
-where $n_j$ is the number of bills in a congressional session authored by senator $j$, $c_{j(k)}$ is the number of cosponsors on senator $j$'s $k^{th}$ bill, where $k \in \{1,\dots, n_j\}$, and $Y_{ij(k)}$ is a binary variable that is 1 if senator $i$ cosponsored senator $j$'s $k^{th}$ bill, and is 0 otherwise.
-This measure ranges in value from 0 to 1, where $WPC_{ij} = 1$ if senator $i$ is a cosponsor on every one of senator $j$'s bills and $WPC_{ij} = 0$ if senator $i$ is never a cosponsor any of senator $j$'s bills.
-
-<<senateplot, fig.cap='The collaboration network in the four senates during the Obama years, 2009-2016.
-Edges are shown only if the weighted propensity to cosponsor from one senator to another is greater than 0.25.
-We use the Fruchterman-Reingold algorithm to layout the node-link diagram.', fig.width=8, fig.height=8, out.width='\\textwidth'>>=
-seobama <- read_csv("data/congress/senateobamapres_gsw_25.csv")
-to_label <- c("Joseph R. Biden Jr.", "John S. McCain", "Ted Cruz","Marco Rubio",
-  "Lindsey O. Graham","Rand Paul", "Bernard Sanders", "Jim Webb", "Mitch McConnell",
-  "Harry M. Reid", "Hillary Rodham Clinton", "Amy Jean Klobuchar")
-seobama$label <- as.factor(ifelse(seobama$source %in% to_label, seobama$source, ""))
-levels(seobama$label) <- c("", "Amy Klobuchar", "Bernie Sanders", "Harry Reid",
-                           "Hillary Clinton", "Jim Webb", "John McCain", "Joe Biden",
-                           "Lindsey Graham", "Marco Rubio", "Mitch McConnel", 
-                           "Rand Paul", "Ted Cruz")
-set.seed(56049382)
-ggplot(data = seobama) + 
-  geom_net(directed = T, labelon=T, arrowsize = .25, singletons= T, fiteach = T, linewidth = .25, layout.alg = 'fruchtermanreingold', fontsize = 3, repel=T, 
-           aes(from_id = source, to_id = target, color = party, label = label),
-           size = 1) + 
-  ThemeNet + 
-  scale_color_manual(values = c("royalblue", "forestgreen","firebrick")) + 
-  theme(legend.position = 'bottom', strip.text = element_text(size = 20)) + 
-  facet_wrap(~senate, nrow = 2, labeller = "label_both") + 
-  xlim(c(0,1.05)) + 
-  ylim(c(0,1.05))
-@
-
-Because we require binary edges for our models, we focus only on very strong collaborations. For our senate collaboration networks, $x$, edges are defined as 
-\[
-  x_{ij} =
-  \begin{cases}
-                                   1 &  \text{if } WPC_{ij} > 0.25 \\
-                                   0 & \text{if } WPC_{ij} \leq 0.25.
-  \end{cases}
-\]
-The networks we constructed for the four senates during President Obama's administration are shown in Figure~\ref{fig:senateplot}.
-In Section~\ref{sec:ModelVis}, we fit several stochastic actor-oriented models to these data sets use those models to guide our further exploration of SAOMs.
-}
-
-\section{Model Visualizations} \label{sec:ModelVis}
-
-Every good data analysis includes both numerical and visual summaries of the data, so why restrict model description and diagnostics to numerical summaries?
-The concept of model visualization was developed to complement traditional model diagnostic tools.
-Typically, numerical summaries such as $R^2$ are used to assess model fit, and the occasional visualization, like a residual plot, are used to determine how well the model fits the data.
-\citeauthor{modelvis} outline three separate ideas, each of which are can be referred to simply as the "model": the model family, the model form, and the fitted model.
-The latter is primarily what one thinks of first when considering a model in a data analysis, where specified a model is fit to data, and parameter estimates and other numerical summaries, such as $R^2$ are reported.
-In the context of SAOMs, the fitted model contains the form of the rate and objective functions, the estimated rate parameters, and the estimated objective function parameters.
-The model form describes the the model \textit{before} the fitting process, defining which parameters are in the model within the context of the larger model family.
-In SAOMs, the model form includes description of the rate and objective functions and the variables therein that describe how the network evolves over time.
-Finally, the model family is the broadest description of the model.
-This is the type of model that you wish to fit to the data, and is chosen based on the problem, data, and knowledge at hand.
-For example, we chose to use a SAOM to model network data over an exponential random graph model (ERGM) because we believe that actor-level variables effect network structure and formation, and we wanted to model the network changes over time.
-
-%These three uses of the term ``model" 
-\st{The model family, the model form, and the fitted model can each be visualized according to the three principals of model visualization: we can }view the model in the data space, visualize collections of models, and explore the process of fitting the model, not just the end result.
-\st{Since we have already decided on our model family, SAOMs, }we now shift our focus to the fitted model and the model form.
-Specifically, we want to learn more about how the model form we choose affects the fitted model by using our example data sets and our visualization toolbox.
-We begin by introducing the five models that we fit to our example data.
-\st{Next we use the five models to guide our visual explorations of SAOMs.
-We first use novel tools and ideas to view a SAOM in the data space of a dynamic network.
-We then explore collections of the same models fit many times to the example data to see how the simulation processes in \texttt{RSiena} affect the model fits.
-Finally, we look behind the scenes and into the individual steps of the continuous time Markov chain to learn more about how this "hidden data" mechanism works and how it results in a fitted model.
-}
-
-\subsection{The Models}
-
-We first consider the 16 actor subset of the teenage friends and lifestyle data available on the \texttt{RSiena} website \citep{RSiena}.
-To this data, we fit three different SAOMs.
-Each SAOM used a simple rate function, $\alpha_m$, and an objective function with two or three parameters.
-The first model, M1, contains the absolute minimum number of parameters in the objective function $f_i(x)$:  $$f_i(x)^{M1} = \beta_1s_{i1} + \beta_2s_{i2},$$ where $s_{i1}$ is the density network statistic and $s_{i2}$ is the reciprocity network statistic for actor $i$ at the current network state $x$.
-The second and third models, M2 and M3, contain one additional parameter each in the objective function which were determined by a Wald-type test provided in the \texttt{RSiena} software to be significant, with $p$-values less than 0.05 \citep{RSienaTest}.
-The M2 model contains an actor-level covariate parameter, and the M3 model contains an additional strutural effect in the objective function.
-\begin{align*}
-f_i(x)^{M2} & = \beta_1s_{i1} + \beta_2s_{i2} + \beta_3s_{i3} \\
-f_i(x)^{M3} & = \beta_1s_{i1} + \beta_2s_{i2} + \beta_4s_{i4}, 
-\end{align*}
-
-\begin{figure}\label{fig:structures}
-\begin{subfigure}[t]{.45\textwidth}
-<<jtt, echo=FALSE, fig.width=2, fig.height=2, message = FALSE, warning = FALSE, fig.align='center',out.width='.49\\textwidth'>>=
-jTTe <- data.frame(from = c('i', 'i', 'h'), to = c('h', 'j', 'j'))
-jTTn <- data.frame(id = letters[8:10], group = c(1,1,2))
-
-jTT <- merge(jTTe, jTTn, by.x = 'from', by.y = "id", all = T)
-
-set.seed(12345) 
-ggplot(data = jTT, aes(from_id = from, to_id = to)) + 
-  geom_net(aes(shape = as.factor(group)), directed = T, labelon = T, 
-           labelcolour='grey80',vjust = 0.5, hjust =0.5, arrowgap = .15, 
-           colour = 'black', size=10, fontsize = 5,
-           ecolour = c("grey40", "grey40","red", "grey40")) + 
-  expand_limits(x=c(-0.1,1.1), y=c(-0.1,1.1)) +
-  ThemeNet +
-  theme(legend.position = "none")
-@
-\caption{\label{fig:jtt}Realization of a jumping transitive triplet, where $i$ is the focal actor, $j$ is the target actor, and $h$ is the intermediary.
-The group of the actors is represented by the shape of the node.}
-\end{subfigure}\hfill
-\begin{subfigure}[t]{.45\textwidth}
-<<dab, echo=FALSE, fig.width=2, fig.height=2, message = FALSE, warning = FALSE, fig.align='center', out.width='.49\\textwidth'>>=
-dade <- data.frame(from = c('i', 'i', 'h', 'j'), to = c('h', 'j', 'k', 'k'))
-dadn <- data.frame(id = letters[8:11], group = c(1,1,1,1))
-
-dad <- merge(dade, dadn, by.x = 'from', by.y = "id", all = T)
-
-set.seed(12345) 
-ggplot(data = dad, aes(from_id = from, to_id = to)) + 
-  geom_net(aes(shape = as.factor(group)), directed = T, labelon = T, labelcolour='grey80',vjust = 0.5, hjust =0.5, arrowgap = .15, colour = 'black', size=10, fontsize = 5) + 
-  expand_limits(x=c(-0.1,1.1), y=c(-0.1,1.1)) +
-  ThemeNet +
-  theme(legend.position = "none")
-@
-\caption{\label{fig:dad}Doubly achieved distance between actors $i$ and $k$.}
-\end{subfigure}
-\caption{The additional network effects included in our models fit to the friends data. On the left, a jumping transitive triplet (JTT). On the right, a doubly achieved distance between $i$ and $k$.}
-\end{figure}
-
-where $s_{i3} = \sum_{j \neq h} x_{ij}x_{ih}x_{hj} \mathbb{I}(z_i = z_h \neq z_j)$, and $s_{i4} =  |\{j : x_{ij} = 0, \sum\limits_h x_{ih}x_{hj} \geq 2\}|$.
-These statistics are known as the number of jumping transitive triplets and the number of doubly achieved distances two effect, respectively.
-The first statistic emphasizes triad relationships that are formed between actors from different covariate groups, while the other emphasizes indirect ties between actors.
-The covariate groups are determined by the student's drinking behavior, the values of which are numeric and mean-centered.
-The four values in our data are $\{\text{never, 1-2 times per year, once per month, once per week}\}$, which after being converted to numeric and mean-centered become $\{-0.8125,-1.8125,0.1875,1.1875\}$.
-The two additional effects are visually represented and further described in Figure~\ref{fig:jtt} and Figure~\ref{fig:dad}, respectively.
-
-<<tableofmeans>>=
-simu2 <- read.csv("data/simulation-1000-M1-M2-M3.csv")
-senM1est <- read.csv("data/congress/senateM1ests.csv")
-senM2est <- read.csv("data/congress/senateM2ests.csv")
-senM3est <- read.csv("data/congress/senateM3ests.csv")
-senM1est$Model <- "M1"
-senM2est$Model <- "M2"
-senM3est$Model <- "M3"
-senM1est$Sim <- 1:nrow(senM1est)
-senM2est$Sim <- 1:nrow(senM2est)
-senM3est$Sim <- 1:nrow(senM3est)
-senM1est %>% gather(parameter, estimate, Rate:recip) %>% dplyr::select(Sim, Model, parameter, estimate) -> senM1est2 
-senM2est %>% gather(parameter, estimate, Rate:jumpXTransTrip) %>% dplyr::select(Sim, Model, parameter, estimate) -> senM2est2 
-senM3est %>% gather(parameter, estimate, Rate:nbrDist2twice) %>% dplyr::select(Sim, Model, parameter, estimate) -> senM3est2 
-senateEsts <- rbind(senM1est2, senM2est2, senM3est2)
-senateEsts$parameter <- as.factor(senateEsts$parameter)
-levels(senateEsts$parameter) <- c("beta1", "beta3", "beta4", "alpha1", "alpha2", "alpha3", "beta2")
-SSE <-
-  senateEsts %>% group_by(Model, parameter) %>% 
-  summarize(meanEst = mean(estimate), sdEst = sd(estimate)) %>% 
-  mutate(val = paste0(sprintf("%.3f",meanEst), " ", "(", sprintf("%.3f", sdEst), ")")) %>% 
-  dplyr::select(Model, parameter, val) %>% 
-  spread(Model, val) %>% arrange(as.character(parameter))
-SFE <- simu2 %>% group_by(Model, parameter) %>% 
-  summarize(meanEst = mean(estimate), sdEst = sd(estimate)) %>% 
-   mutate(val = paste0(sprintf("%.3f",meanEst), " ", "(", sprintf("%.3f", sdEst), ")")) %>% 
-  select(Model, parameter, val) %>% 
-  dplyr::select(Model, parameter, val) %>% 
-  spread(Model, val) %>% arrange(parameter)
-@
-
-In Section~\ref{sec:samemoddiffdat}, we also fit models M1, M2, and M3 to the senate collaboration data for comparison.
-%Comparison of the parameter estimates of these models fit to the two different data sets is given in Table~\ref{tab:meanests}.
-In addition, we fit two more models, M4 and M5 to the senate data that were found to be significant according to the Wald-type test:
-
-\begin{align*}
-f_i(x)^{M4} & = \beta_1s_{i1} + \beta_2s_{i2} + \beta_5s_{i5} \\
-f_i(x, z)^{M5} & = \beta_1s_{i1} + \beta_2s_{i2} + \beta_6s_{i6}.
-\end{align*}
-
-In M4, $s_{i5} = \mathbb{I}(x_{i+} = x_{+i} = 0)$ is an indicator variable for isolated nodes, called the network-isolate effect.
-This parameter attempts to measure the effect of a node being completely isolated in the network.
-If the value of the parameter $\beta_5$ is high, nodes are encouraged to remain isolated.
-In M5,  $s_{i6} = \sum\limits_j x_{ij}x_{ji}\mathbb{I}(z_i = z_j)$ is a measure of reciprocity between senators in the same party, $z_i, z_j \in \{\text{Democrat, Independent, Republican}\}$, generally called the same covariate reciprocity effect.
-The values of the covariate are converted to numeric, $\{1,2,3\}$, respectively, and mean-centered within \texttt{RSiena}.
-The mean-centered values of $z \in \{\text{Democrat, Independent, Republican}\}$ are $\{-0.96, 0.04, 1.04\}$.
-The parameter $\beta_6$ attempts to measure the propensity for forming reciprocal ties between nodes with the same covariate value.
-If the value of the parameter $\beta_6$ is high, nodes belonging to the same covariate group are encourage to reciprocate incoming ties.
-\st{Due to the intractability of SAOMs, it is difficult to know for certain how to interpret a fitted value of a parameter.
-We can make educated guesses based on the definition of the effect and the sign of the fitted value, but a direct interpretation is not always possible.
-But,} by exploring these models more visually, we aim to understand these effects, their interpretations, and the model fitting process better.
-
-\subsection{View the model in the data space}\label{sec:minds}
-
-The first way we hope to better understand stochastic actor-oriented models is by viewing the model(s) in the data space.
-In \citeauthor{modelvis}, they \st{happen to} define the \textit{data space} as ``the region over which we can reliably make inferences, usually a hypercube containing the data" \citep[p.~206]{modelvis}.
-But what does this definition mean for network data?
-For dynamic social networks, there are a few different data ``spaces":
-\begin{enumerate}
-\item the actors and their corresponding covariates,
-\item the edges and their correpsonding variables that describe the relationships between the nodes, and 
-\item the time, both the continuous unobserved time and the discrete observed time points, over which the network evolves
-\end{enumerate}
-These three data pieces can be visualized together in various ways.
-The traditional node-link visualization uses one of many algorithms to layout the actors as points in 2D space, then draws segments connecting the points in 2D if there is an edge between two nodes, and draws nothing otherwise.
-The time aspect can be visualized by drawing each network observation in time and placing the observed timepoints side-by-side.
-
-\st{Because longitudinal network data consist of three different ``spaces" of data, viewing the model in the data space can depend on which aspect of the \textit{model} and the \textit{data} we are interested in viewing.
-%The SIENA software includes a method for modelling covariate change in the network
-Incorporating data covariates into the network structure allows us to assess whether the ties between nodes are affected by how nodes behave over time.
-In this instance, we would want to view predictions over time.
-A SAOM can also model behavior change over time, taking both the node and edge information into account.
-In this case, a plot of predicted covariate values over time would put the model into the time and node data space.
-Most likely, however, is that we would want to view all of the data spaces simultaneously.}
-
-One tool that can bring the \st{node, edge, and time} data spaces together in this way is the \texttt{R} package \texttt{geomnet} \citep{geomnet}.
-Different visual features in the node-link diagram can be tied to the underlying node or edge data.
-The color, size, and shape of the points can be used to represent variables in the node data, while the color, linewidth, and linetype of the lines between points can be used to represent the edge variables.
-In a social network, node data might be age, gender, and occupation of the person in the network, and edge data might be length of connection between two people, how the people first met (school, work, church, etc.), and how often they interact, and we can view the network at different timepoints side-by-side to see its evolution.
-Pulling all of this information together with \texttt{geomnet} allows the entire data space to be viewed at once.
-
-<<funsenatevis, fig.height=3, fig.cap='The 111th Senate at two discrete time points: while Clinton was in the senate in the first few weeks of 2009 (on the left) and after she left the senate to become Secretary of State (on the right).
-We put some potential model covariates, sex and party of the senator and the number of bills they authored in the data space through the shape, color, and size of the nodes, respectively.
-We also map the strength of the tie, the $WPC$, to the width of the edge between two nodes.
-Thicker lines implies higher propensity to collaborate.
-In addition, senators with no ties higher than $WPC = 0.25$ are not shown.'>>=
-se111clint <- filter(seobama, senate == 111)
-se111noclint <- filter(seobama, senate == 111)
-se111noclint$target[which(se111noclint$target == "Hillary Rodham Clinton")] <- NA
-se111clint$Clinton <- 'Yes'
-se111noclint$Clinton <- 'No'
-clintonSenate <- rbind(se111clint, se111noclint)
-clintonSenate$Clinton <- as.factor(clintonSenate$Clinton)
-clintonSenate$Clinton <- ordered(clintonSenate$Clinton, levels = c("Yes", "No"))
-
-clintonSenate %>% filter(source %in% target)
-
-clintonSenate %>% filter(!(source %in% c("Roland Burris", "Bernard Sanders", "Frank R. Lautenberg", "Judd A. Gregg", "Mary L. Landrieu") & is.na(target))) %>%
-ggplot() + 
-  geom_net(directed = T, labelon=F, arrowsize = .3, singletons= F, fiteach = T, arrowgap = .01, layout.alg = 'fruchtermanreingold',
-           aes(from_id = source, to_id = target, color = party, linewidth = gsw, shape = sex,size = n_au)) + 
-  scale_size_continuous(name="# bills\nauthored") +
-  #theme_net() + 
-  ThemeNet + 
-  scale_shape_manual(values = c(17,16)) + 
-  scale_color_manual(values = c("royalblue", "forestgreen","firebrick")) + 
-  theme(legend.position = 'bottom') + 
-  facet_wrap(~Clinton, nrow = 1, labeller = 'label_both')
-@
-
-To demonstrate, we use \texttt{geomnet} to visualize the connections in the $111^{th}$ United States Senate at two different points: when Hillary Clinton was in the senate, and after she left to become Secretary of State.
-Clinton was only in the $111^{th}$ senate for 17 days, from January 3, 2009, to January 20, 2009, when she was in the middle of her second term as senator from New York.
-In that time, she authored two bills and was a cosponsor on 17 other bills.
-With Clinton included in the node-link diagram, the senate looks much more highly collaborative than it does without her in the diagram.
-We can compare the number of bills authored throughout the senate by mapping the size of the node to the numer of bills authored by that senator.
-We also map shape to sex, and keep color mapped to party of the senator.
-In addition, we can see the strength of the tie by mapping the linewidth of the edge to the $WPC$ value between the two senators.
-In this single visualization, we have viewed node information (number of bills authored by a senator and the sex and party of that senator), edge information (the direction and strength of ties between two senators), and time (before and after Clinton left the senate).
-% ST: I don't like this paragraph
-% The visual aspects of the node-link diagram connecting the node, edge, and time data spaces are themselves part of the data space, because they can be viewed in different ways.
-Another way to connect the node and the edge data, as discussed in Section~\ref{sec:netvizintro}, is through the network's adjaceny matrix.
-The time dimension can also be incorporated with side-by-side adjacency matrix visualizations.
-
-% below chunk contains some repetitive stuff. Haven't figured out what it's purpose is yet. 
-<<summnetM1, message=FALSE, fig.cap='On the left, the first wave of observed data that is conditioned on in the model.
-On the right, the second wave of observed data.
-In the middle, a summary network from the first model fit to the data.
-This summary network represents 1,000 simulations of wave 2 using the values from the simple fitted model M1.',results='hide', fig.height=2.5>>=
-library(netvizinf)
-friendshipData <- array(c(fd2.w1, fd2.w2, fd2.w3), dim = c(16, 16, 3))
-drink1 <- s50a[20:35,1]
-drink2 <- s50a[20:35,2]
-drink3 <- s50a[20:35,3]
-friendship <- sienaDependent(friendshipData)
-alcohol <- varCovar(cbind(drink1,drink2, drink3))
-mydata <- sienaDataCreate(friendship, alcohol)
-M1eff <- getEffects(mydata)
-M2eff <- includeEffects(M1eff, jumpXTransTrip, interaction1 = "alcohol")
-M3eff <- includeEffects(M1eff, nbrDist2twice)
-simu2 <- read.csv("data/simulation-1000-M1-M2-M3.csv")
-means <- simu2 %>% group_by(Model, parameter) %>% summarize(
-  means = mean(estimate)
-)
-M1parms <- (means %>% filter(Model == "M1"))$means
-M2parms <- (means %>% filter(Model == "M2"))$means
-M3parms <- (means %>% filter(Model == "M3"))$means
-set.seed(4231352)
-M1sims1000 <- netvizinf::saom_simulate(dat = mydata, 
-                            struct = M1eff,
-                       parms = as.numeric(M1parms), 
-                       N = 1000)
-M2sims1000 <- netvizinf::saom_simulate(dat = mydata, 
-                            struct = M2eff,
-                       parms = as.numeric(M2parms), 
-                       N = 1000)
-M3sims1000 <- netvizinf::saom_simulate(dat = mydata, 
-                            struct = M3eff,
-                       parms = as.numeric(M3parms), 
-                       N = 1000)
-M1simsdf <- sims_to_df(M1sims1000)
-M2simsdf <- sims_to_df(M2sims1000)
-M3simsdf <- sims_to_df(M3sims1000)
-M1avgW2 <- M1simsdf %>% filter(!is.na(to) & wave == 1) %>%
-  group_by(from, to) %>% 
-  summarise(count = n()) %>% 
-  mutate(weight = ifelse(from == to, 0, count))
-M2avgW2 <- M2simsdf %>% filter(!is.na(to) & wave == 1) %>%
-  group_by(from, to) %>% 
-  summarise(count = n()) %>% 
-  mutate(weight = ifelse(from == to, 0, count))
-M3avgW2 <- M3simsdf %>% filter(!is.na(to) & wave == 1) %>%
-  group_by(from, to) %>% 
-  summarise(count = n()) %>% 
-  mutate(weight = ifelse(from == to, 0, count))
-
-# make a df of wave 2, wave 1, and the three averages and facet. 
-names(actual1)[1:2] <- c("from", "to")
-names(actual2)[1:2] <- c("from", "to")
-actual1$count <- 1
-actual1$weight <- 1
-actual2$count <- 1
-actual2$weight <- 1
-actual1 <- actual1 %>% dplyr::select(from,to, count, weight)
-actual2 <- actual2 %>% dplyr::select(from,to, count, weight)
-actual1$cat <- "1st Wave"
-actual2$cat <- "2nd Wave"
-
-avgW2M1 <- M1avgW2 %>% ungroup() %>% 
-  filter(weight > 50) %>% 
-  mutate(from = as.factor(from), to = as.factor(to),
-         cat = "Model1")
-add1 <- setdiff(as.character(1:16), unique(c(as.character(avgW2M1$from), as.character(avgW2M1$to)))) 
-avgW2M1 %>% add_row(from = add1, to = NA, count = NA, weight = NA, cat = "Model1") -> avgW2M1
-avgW2M2 <- M2avgW2 %>% ungroup() %>% 
-  filter(weight > 50) %>% 
-  mutate(from = as.factor(from), to = as.factor(to),
-         cat = "Model2")
-add2 <- setdiff(as.character(1:16), unique(c(as.character(avgW2M2$from), as.character(avgW2M2$to)))) 
-avgW2M2 %>% add_row(from = add2, to = NA, count = NA, weight = NA, cat = "Model2") -> avgW2M2
-avgW2M3 <- M3avgW2 %>% ungroup() %>% 
-  filter(weight > 50) %>% 
-  mutate(from = as.factor(from), to = as.factor(to),
-         cat = "Model3")
-add3 <- setdiff(as.character(1:16), unique(c(as.character(avgW2M3$from), as.character(avgW2M3$to)))) 
-avgW2M3 %>% add_row(from = add3, to = NA, count = NA, weight = NA, cat = "Model3") -> avgW2M3
-
-combinedavgs <- rbind(actual1, actual2, avgW2M1, avgW2M2, avgW2M3)
-combinedavgs %>% group_by(cat) %>% 
-  mutate(linewidth = weight / max(weight,na.rm = T)) -> combinedavgs
-
-combinedavgs %>% filter(cat == "Model1") %>% mutate(logweight = log(weight)) -> t1 
-colors <- tweenr::tween_color(data = c("#969696", "#d73027"), n = 26, ease = 'linear')
-t1 %>% arrange(logweight) -> t1
-t1$color <- NA
-t1$color[1:26] <- colors[[1]]
-
-p1 <- ggplot(data = t1) +
-  geom_net(aes(from_id = from, to_id = to, linewidth= linewidth),
-           directed = T, curvature = .2, labelon=T, size = 5,
-           arrow = arrow(angle = 20, type = "open", length = unit(.2, "cm")), ecolour = t1$color,
-           color = "#252525", hjust = .5, vjust = .5, labelcolour = 'white', fontsize = 4, arrowgap = .02,
-           singletons = T) +
-  labs(title = "Average: M1") + ThemeNet +
-  theme(plot.background = element_rect(color = 'black'))
-
-p2 <- ggplot(data = filter(combinedavgs, cat == "1st Wave")) +
-  geom_net(aes(from_id = from, to_id = to, linewidth= linewidth/3),
-           directed = T, curvature = .2, labelon=T, size = 5,
-           arrow = arrow(angle = 20, type = "open", length = unit(.2, "cm")), ecolour = "#969696",
-           color = "#252525", hjust = .5, vjust = .5, labelcolour = 'white', fontsize = 4, arrowgap = .03,
-           singletons = T) +
-   labs(title = "Wave 1") + ThemeNet + 
-    theme(plot.background = element_rect(color = 'black'))
-p3 <- ggplot(data = filter(combinedavgs, cat == "2nd Wave")) +
-  geom_net(aes(from_id = from, to_id = to, linewidth= linewidth/3),
-           directed = T, curvature = .2, labelon=T, size = 5,
-           arrow = arrow(angle = 20, type = "open", length = unit(.2, "cm")), ecolour = "#969696",
-           color = "#252525", hjust = .5, vjust = .5, labelcolour = 'white', fontsize = 4, arrowgap = .03,
-           singletons = T) +
-  labs(title = "Wave 2")  +  ThemeNet + 
-  theme(plot.background = element_rect(color = 'black')) 
-grid.arrange(p2, p1, p3, nrow = 1)
-@
-
-Another way to view the model in the data space is through simulation from the model.
-No single network alone simulated from a SAOM is going to look like the data or represent the model, just as no single value simulated from the standard normal distribution will look like a bell-shaped curve.
-\st{Statisticians would prefer to look at a sample, or at least a summary statistic or two, }so it would therefore better to visualize many simulations together.
-From a statistician's point of view, a stumbling block with statistical network models generally is the lack of an ``average network" or ``expected network" value.
-Statisticians frequently rely on averages and expected values in data analyses, but statistical network models, especially those as complex as SAOMs, lack a single, intuitive expected value measure.
-We could talk about expected values of parameters, but the parameters can be hard to interpret.
-\st{Expected values of parameters are important, but if they cannot directly tell us anything about our dependent variable (networks in this case), they lose some value to us.} Furthermore, there is no way to talk about the expected value of an observation \st{simulated} from a statistical network model.
-How then, can we arrive at an ``average" network?
-We answer this question through visualization.
-For network data, one way we view an "average" network is through a summary network drawn using the traditional node-link diagram.
-In Figure~\ref{fig:summnetM1}, we show an \textit{average network} created with 1,000 simulations of the second wave of the network from Model 1.
-\st{To make this average network, we first simulated 1,000 wave 2 and wave 3 observations of our small friendship example data from model M1, for which parameters had previously been estimated.
-We then combine the 1,000 instances of wave 2, and count up the number of times each edge appears in a simulation.
-Then, we combine these 1,000 networks into a single network with edgeweight equal to the proportion of time that edge appears in our 1,000 simulations.
-This weighted edgelist is the network we draw using the node-link diagram.} An edge is only drawn in the average network if it appears in more than 5\% of the simulations (in at least 51 of the 1000 simulations), with edges that appear more frequently emphasized by thicker linewidths and a darker color, representing the proportion of times appeared in the model simulations.
-On either side of the average network in Figure~\ref{fig:summnetM1}, we show the actual data, wave 1 on the left, and wave 2 on the right.
-\st{We can see that the structure of the average network is much more similar to the first wave than to the second wave.
-However, the simulations are supposed to represent the second wave of data, which is shown on the right in Figure~\ref{fig:summnetM1}.
-This is an indication that the simple model, M1, is doing a very poor job of capturing the change mechanism from the first to the second wave of observation.
-The average network can thus be used to help determine model goodness-of-fit.
-Because the the average network looks more like the first wave than the second wave, we can use the visualization in Figure \ref{fig:summnetM1} as evidence of poor model fit.}  % XXX ST - I was looking at the wrong one.
-I had wave == 2 instead of wave == 1 in the code.
-This means I was looking at the wrong simulations, so they really do look more like the first data obs than the second obs.
-XXX Comparing the three visualizations side-by-side shows us that the summary network of simulations of wave 2 looks more like the true wave 2 than wave 1, which is a good indication that the model is capturing the true data structure.
-It's not ideal, however, because there appear to be too many reciprocated ties in the summary network, far more than there are in the data.
-This could indicate that we should remove the reciprocity parameter from the model.
-
-<<gof2, fig.cap="An example of viewing the distribution of a statistic of interest from a SAOM.
-Here, the Mahalanobis distance of the cumulative outdegree values computed on a simulated network to the mean value simulated is shown.
-The distance computed from the data is shown as a vertical line.", fig.height=3, out.width='.6\\textwidth', eval = FALSE>>=
-load("data/ansnullpaper.rda")
-library(lattice)
-gof1 <- sienaGOF(ansnull, OutdegreeDistribution, varName ="friendship", join = FALSE)
-histdat <- data.frame(x = log(gof1$`Period 1`$SimulatedTestStat))
-ggplot(histdat) + geom_histogram(aes(x = x), binwidth = .1, fill = 'white', color = "black") + 
-  #geom_density(aes(x = x, y = ..density..*250)) + 
-  geom_vline(xintercept = log(gof1$`Period 1`$ObservedTestStat), color = 'red') + 
-  ThemeNoNet + 
-  labs(x = "log(Statistic Value)", title = "Distribution of log Mahalanobis Distance")
-
-@
-
-%\st{If, as discussed in Section~\ref{sec:modgof},
-%Figure~\ref{fig:sienagof}, 
-%summary statistics such as outdegree and indegree distribution computed on the network are of interest, we can look at the computed distances between the data and the model simulations.
-distributions of these statistics computed on networks simulated from the model show the model in the edge data space.
-For an example, see Figure~\ref{fig:gof2}, where the distribution of the log of the Mahalanobis distance is shown for the outdegree distribution auxiliary statistics, and the log Mahalanobis distance for the observed data is shown as a vertical line.
-}
-
-\begin{figure}
-\centering
-\includegraphics[width=.75\linewidth]{img/Model1LUSmallFriends.png}
-\caption{\label{fig:lineup} A small lineup of node-link diagrams showing the second wave of our small friendship network among five networks simulated from model M1.}
-\end{figure}
-
-Another potential goodness-of-fit visualization \st{that places the model in the node and edge data space} is a lineup like those proposed in \citet{buja:2009}.
-A \textit{lineup} ``asks the witness to identify the plot of the real data from among a set of decoys, the null plots, under the veil of ignorance" \citep[p.
-4369]{buja:2009}.
-It can be thought of like a police lineup, where the ``suspect" is in a lineup among several innocent lookalike fillers, and a witness picking the suspect out of the lineup is considered evidence against the suspect.
-In data and model visualization, the ``suspect" is a plot of the true data, while the ``filler" is composed of several plots of mock data, simulated from a hypothesized model.
-If the true data stands out among the simulated data, that is taken as evidence of poor model fit, whereas if the true data is difficult to identify among the simulated data, that is taked as evidence of good model fit.
-\st{An application of the lineup protocal can be found in \citet{hofmann2012graphical}, where the authors examine, for instance, the differences between polar and cartesian coordinates for plotting categorical data, and density plots and box plots for determining distributional differences.
-They pose questions to experiment participants such as, ``Which plot is most different from the others?" for the first example and, ``In which plot is the blue group furthest to the right?" for the distributional differences.
-The data visualizations examined in \citet{hofmann2012graphical} are less complex than a node-link network diagram.
-What questions should we ask for network data visualizations?
-Asking participants to identify the most different plot may be difficult.
-In the network lineup shown in Figure~\ref{fig:lineup}, the second wave of the small friendship network is shown among five simulated networks from model M1 using parameter values estimated from the data.
-What makes these plots ``different"?
-It seems possible to argue for any one of the six plots in Figure~\ref{fig:lineup} as most different.
-So, we guide participants to look at the overall structure of the graphs to determine which has the most and least complex structure.
-The least complex plot, number six, has no triplets, while the most complex plot, number three, has three triplets, and in fact, plot three is the data.
-We have found in an experiment where all the node-link diagrams shown are based on simulated data, one observation from an ``alternative" model and the rest from a ``null" model, the triangular shape of the triplets stands out the most to participants.
-We can use the lineup protocol to help better understand SAOMs because it allows us to view the model in the data space by placing observations the model side-by-side with the data and examining the differences.
-This can also help us determine the significant structural effects, if any, of the parameters in the model on observations simulated from the model.
-The triangular shape mentioned above becomes more prevalent when a transitive triplet parameter in included in a model, but it does not always cause triplets to form in the simulated data.
-This requires more investigation with larger data sets and more complex models, but is promising for the development of additional goodness-of-fit measures for network models.}
-
-% \st{
-% \begin{itemize}
-% \item WHat is the model? the network model (SAOM model is just one of many)
-% \item What is the data? Edge data, node data, observed time points, unobserved timepoints (CTMC)
-% \item Note: specifically dealing with dynamic networks observed at discrete time points. 
-% \end{itemize}
-% What are some things that could go in here? need more data examples; not just the friends data.
-% Harry Potter data??? I want to find my own data, too.}
-
-\subsection{Visualizing collections of models}
-
-\st{There are many possible ways to collect models together.
-We could look at the same models fit to different data, different models fit to the same data, or because of the nature of SAOMs, we could fit the same models to the same data many times to see how the simulations change.
-For the SAOMs, we decided that there were four collections that were most important:
-\begin{enumerate}
-\item the collections resulting from exploring the space of all possible models;
-\item the collections we get when varying model settings; 
-\item the results from fitting the same model form to different data;
-\item the results from fitting the same model to the same data many times.
-\end{enumerate}
-
-We chose these four collections because they each explore something different about SAOMs.
-The first takes the many dozens of parameters available to include in a SAOM into account, which easily translates into the second by showing how those many parameters affect the fitted models.
-Then, we look at how the same model looks when fit to different sets of dynamic network data.
-Finally, we look at the results from fitting the same model to the same data because the MCMCs and CTMCs that make up a SAOM lead to different parameter estimates every time, so it important to see how the results can vary.}
-
-\subsubsection{Exploring the space of all possible models}
-
-<<sigtesting, fig.show = 'hold', out.width='\\textwidth', fig.height=3, fig.cap="Significant effects for the two data sets, at the level of 0.10 as calculated by the Wald-type test available in the SIENA software.">>=
-smfr_effs <- read_csv("data/effects_significance_smallFriends.csv")
-smfr_effs$inter1[is.na(smfr_effs$inter1)] <- "none"
-library(ggrepel)
-set.seed(12345678)
-smfr_effs %>% filter(type == "eval" & Waldpval <=0.10) %>% 
-ggplot(aes(x = estimate, y = Waldpval, color = inter1)) + 
-  #geom_errorbarh(aes(xmin = estimate - se, xmax = estimate + se), color = 'black') +  
-  geom_vline(xintercept = 0) + 
-  geom_point(size = 3) + 
-  geom_text_repel(aes(label = shortName)) + 
-  scale_color_brewer(palette = "Paired", name = "Interaction\nVariable") + 
-  ggtitle("Significant effects at level 0.10 for friendship data") + 
-  coord_flip() + 
-  ThemeNoNet
-sen_effs <- read_csv("data/congress/senateModelSigEffects.csv")
-#sen_effs$Waldpval <- parse_number(sen_effs$Waldpval)
-sen_effs$inter1[is.na(sen_effs$inter1)] <- "none"
-ggplot(data = sen_effs %>% filter(Waldpval <=0.10), aes(x = estimate, y = Waldpval, color = inter1)) + 
-  # geom_errorbarh(aes(xmin = estimate - se, xmax = estimate + se), color = 'black') +  
-  geom_vline(xintercept = 0) + 
-  geom_point(size = 3) + 
-  geom_text_repel(aes(label = shortName)) + 
-  scale_color_brewer(palette = "Paired", name = "Interaction\nVariable") + 
-  ggtitle("Significant effects at level 0.10 for senate data") + 
-  coord_flip() + 
-  ThemeNoNet
-@
-
-\st{The \texttt{RSiena} manual contains over eighty possible effects to include in the model.
-In order to select parameters to include in the models for our example data, we searched through the possible effects available to model given the data structure to find \textit{significant} effects.
-We tested for significance using the Wald-type tests built into \texttt{RSiena} for one-at-a-time effects testing.
-We start with the outdegree and reciprocity measures as the foundation of the models we fit, then add one evaluation effect, fit the model, test the additional effect for significance, and repeat for all possible parameters to add to the model.
-We performed this procedure for both the small friendship and the senate collaboration data.
-The results for significant effects are shown in Figure~\ref{fig:sigtesting}.
-We see in both the friendship data and the senate data results that most of the significant effects have absolute value less than ten.
-In addition, the $p$-values for the effects from the friendship data are more spread out than the $p$-values for the senate data, which are concentrated at about 0.02 or less.
-This may suggest that larger data sets tend to result in smaller $p$-values, just like a larger sample size results in smaller $p$-values in a $t$-test.}
-
-\subsubsection{Varying model settings}
-
-We have varied model settings already by choosing models M1, M2, M3 to fit to our small firendship network data set.
-In Section~\ref{sec:samemodsamedat}, we fit these models to the data 1,000 times, and in this section, we explore simulations from these three models given the mean values of from the 1,000 fitted parameter values as the parameters in our models.
-From each of these three models using the means of parameter estimates as our fixed parameter values, we simulated 1,000 observations from each of the three models.
-In this process, we condition on the first friendship network observation, and the second and third observations are simulated from the SAOM models with the given parameter values.
-From these simulations, we first create a visualization that represents an average network.
-
-<<averagenet, results='hide', out.width='\\textwidth',fig.cap='The node-link diagrams from the three "average" networks that we calculated are in the top row, and the true wave 1 and wave 2 data are shown in the bottom row above.
-There is some difference between the three models, but overall, these three models cannot capture the structure in the true second wave of data.', fig.height=5>>=
-#friend.data.w1 <- as.matrix(read.table("s50_data/s50-network1.dat"))
-#friend.data.w2 <- as.matrix(read.table("s50_data/s50-network2.dat"))
-#friend.data.w3 <- as.matrix(read.table("s50_data/s50-network3.dat"))
-#drink <- as.matrix(read.table("s50_data/s50-alcohol.dat"))
-friendshipData <- array(c(fd2.w1, fd2.w2, fd2.w3), dim = c(16, 16, 3))
-friendship <- sienaDependent(friendshipData)
-alcohol <- varCovar(drink[20:35,])
-mydata <- sienaDataCreate(friendship, alcohol)
-M1eff <- getEffects(mydata)
-M2eff <- includeEffects(M1eff, jumpXTransTrip, interaction1 = "alcohol")
-M3eff <- includeEffects(M1eff, nbrDist2twice)
-data(M1ests_bigfriends)
-M1parms <- (means %>% filter(Model == "M1"))$means
-M2parms <- (means %>% filter(Model == "M2"))$means
-M3parms <- (means %>% filter(Model == "M3"))$means
-# M1sims1000 <- saom_simulate(dat = mydata, 
-#                             struct = M1eff,
-#                        parms = as.numeric(M1parms), 
-#                        N = 1000)
-# M2sims1000 <- saom_simulate(dat = mydata, 
-#                             struct = M2eff,
-#                        parms = as.numeric(M2parms), 
-#                        N = 1000)
-# M3sims1000 <- saom_simulate(dat = mydata, 
-#                             struct = M3eff,
-#                        parms = as.numeric(M3parms), 
-#                        N = 1000)
-# M1simsdf <- sims_to_df(M1sims1000)
-# M2simsdf <- sims_to_df(M2sims1000)
-# M3simsdf <- sims_to_df(M3sims1000)
-M1avg <- M1simsdf %>% filter(!is.na(to) & wave == 1) %>%
-  group_by(from, to) %>% 
-  summarise(count = n()) %>% 
-  mutate(weight = ifelse(from == to, 0, count))
-M1avgW2 <- M1simsdf %>% filter(!is.na(to) & wave == 2) %>%
-  group_by(from, to) %>% 
-  summarise(count = n()) %>% 
-  mutate(weight = ifelse(from == to, 0, count))
-M1avg2 <- M1simsdf %>% filter(!is.na(to)) %>%
- group_by(from, to, wave) %>%
- summarise(count = n()) %>%
- mutate(weight = ifelse(from == to, 0, count))
-M2avg <- M2simsdf %>% filter(!is.na(to) & wave == 1) %>%
-  group_by(from, to) %>% 
-  summarise(count = n()) %>% 
-  mutate(weight = ifelse(from == to, 0, count))
-M2avgW2 <- M2simsdf %>% filter(!is.na(to) & wave == 2) %>%
-  group_by(from, to) %>% 
-  summarise(count = n()) %>% 
-  mutate(weight = ifelse(from == to, 0, count))
-M2avg2 <- M2simsdf %>% filter(!is.na(to)) %>%
- group_by(from, to, wave) %>%
- summarise(count = n()) %>%
- mutate(weight = ifelse(from == to, 0, count))
-M3avg <- M3simsdf %>% filter(!is.na(to) & wave == 1) %>%
-  group_by(from, to) %>% 
-  summarise(count = n()) %>% 
-  mutate(weight = ifelse(from == to, 0, count))
-M3avgW2 <- M3simsdf %>% filter(!is.na(to) & wave == 2) %>%
-  group_by(from, to) %>% 
-  summarise(count = n()) %>% 
-  mutate(weight = ifelse(from == to, 0, count))
-M3avg2 <- M3simsdf %>% filter(!is.na(to)) %>%
- group_by(from, to, wave) %>%
- summarise(count = n()) %>%
- mutate(weight = ifelse(from == to, 0, count))
-
-# M1avg$from <- paste0("V", M1avg$from)
-# M1avg$to <- paste0("V", M1avg$to)
-# M1avg2$from <- paste0("V", M1avg2$from)
-# M1avg2$to <- paste0("V", M1avg2$to)
-# M1avgW2$from <- paste0("V", M1avgW2$from)
-# M1avgW2$to <- paste0("V", M1avgW2$to)
-# M2avg$from <- paste0("V", M2avg$from)
-# M2avg$to <- paste0("V", M2avg$to)
-# M2avgW2$from <- paste0("V", M2avgW2$from)
-# M2avgW2$to <- paste0("V", M2avgW2$to)
-# M2avg2$from <- paste0("V", M2avg2$from)
-# M2avg2$to <- paste0("V", M2avg2$to)
-# M3avg$from <- paste0("V", M3avg$from)
-# M3avg$to <- paste0("V", M3avg$to)
-# M3avgW2$from <- paste0("V", M3avgW2$from)
-# M3avgW2$to <- paste0("V", M3avgW2$to)
-# M3avg2$from <- paste0("V", M3avg2$from)
-# M3avg2$to <- paste0("V", M3avg2$to)
-combinedavgs$linewidth[combinedavgs$cat %in% c("1st Wave", "2nd Wave")] <- .5
-combinedavgs$cat <- as.factor(combinedavgs$cat)
-combinedavgs$cat <- ordered(combinedavgs$cat, levels = c("Model1", "Model2", "Model3", "1st Wave", "2nd Wave"))
-
-ggplot(data = combinedavgs) + 
-  geom_net(aes(from_id = from, to_id = to, linewidth= linewidth),
-           directed = T, curvature = .2, labelon=T, size = 5,
-           arrow = arrow(angle = 20, type = "open", length = unit(.2, "cm")), ecolour = "#969696", fiteach = T, 
-           color = "#252525", hjust = .5, vjust = .5, labelcolour = 'white', fontsize = 4, arrowgap = .02,
-           singletons = T) + 
-  ThemeNet + theme(panel.background = element_rect(color = 'black')) + 
-  facet_wrap(~cat)
-
-# wave2drink <- drink2[,2]
-# wave2drink <- as.data.frame(wave2drink)
-# wave2drink$id <- paste0("V", 1:16)
-# names(wave2drink)[1] <- "drink"
-# wave3drink <- drink2[,3]
-# wave3drink <- as.data.frame(wave3drink)
-# wave3drink$id <- paste0("V", 1:16)
-# names(wave3drink)[1] <- "drink"
-# M1avgcov <- merge(M1avg, wave2drink, by.x = "from", by.y = "id", all = T)
-# M2avgcov <- merge(M2avg, wave2drink, by.x = "from", by.y = "id", all = T)
-# M3avgcov <- merge(M3avg, wave2drink, by.x = "from", by.y = "id", all = T)
-# M1avgW2cov <- merge(M1avgW2, wave3drink, by.x = "from", by.y = "id", all = T)
-# M2avgW2cov <- merge(M2avgW2, wave3drink, by.x = "from", by.y = "id", all = T)
-# M3avgW2cov <- merge(M3avgW2, wave3drink, by.x = "from", by.y = "id", all = T)
-# 
-# M1avg[,-3] %>% spread(to, weight) %>% data.frame-> distmatM1
-# rownames(distmatM1) <- distmatM1$from
-# distmatM1 <- distmatM1[,-1]
-# distmatM1 <- as.matrix(sqrt(1/distmatM1))
-# distmatM1[is.na(distmatM1)] <- 0
-# M1avgW2[,-3] %>% spread(to, weight) %>% data.frame-> distmatM1W2
-# rownames(distmatM1W2) <- distmatM1W2$from
-# distmatM1W2 <- distmatM1W2[,-1]
-# distmatM1W2 <- as.matrix(sqrt(1/distmatM1W2))
-# distmatM1W2[is.na(distmatM1W2)] <- 0
-# M2avg[,-3] %>% spread(to, weight) %>% data.frame-> distmatM2
-# rownames(distmatM2) <- distmatM2$from
-# distmatM2 <- distmatM2[,-1]
-# distmatM2 <- as.matrix(sqrt(1/distmatM2))
-# distmatM2[is.na(distmatM2)] <- 0
-# M2avgW2[,-3] %>% spread(to, weight) %>% data.frame-> distmatM2W2
-# rownames(distmatM2W2) <- distmatM2W2$from
-# distmatM2W2 <- distmatM2W2[,-1]
-# distmatM2W2 <- as.matrix(sqrt(1/distmatM2W2))
-# distmatM2W2[is.na(distmatM2W2)] <- 0
-# M3avg[,-3] %>% spread(to, weight) %>% data.frame-> distmatM3
-# rownames(distmatM3) <- distmatM3$from
-# distmatM3 <- distmatM3[,-1]
-# distmatM3 <- as.matrix(sqrt(1/distmatM3))
-# distmatM3[is.na(distmatM3)] <- 0
-# M3avgW2[,-3] %>% spread(to, weight) %>% data.frame-> distmatM3W2
-# rownames(distmatM3W2) <- distmatM3W2$from
-# distmatM3W2 <- distmatM3W2[,-1]
-# distmatM3W2 <- as.matrix(sqrt(1/distmatM3W2))
-# distmatM3W2[is.na(distmatM3W2)] <- 0
-
-# cols <- brewer.pal(n = 4, "YlOrRd")
-# 
-# plotM1W2 <- ggplot(data = M1avgcov) + 
-#   geom_net(aes(from_id = from, to_id = to, linewidth = weight/251, color = as.factor(drink)), 
-#            arrowgap = .03, size = 3,arrowsize = .5,
-#            directed = TRUE, curvature = .2,
-#            #layout.alg = "fruchtermanreingold"
-#            layout.alg = "kamadakawai",
-#            layout.par = list(elen = distmatM1)
-#            ) + 
-#   scale_color_brewer(palette = "YlOrRd", name = "Drinking\nBehavior") + 
-#   #labs(title = "Wave 2, M1") + 
-#   scale_size_continuous(name = "Weight") + 
-#   theme_net() + 
-#   theme(plot.title = element_text(hjust = .5), legend.position = 'none',plot.margin=unit(c(0,0,0,0),"mm")) + 
-#   labs(title = "Model 1 Simulated Wave 2")
-# plotM1W3 <- ggplot(data = M1avgW2cov) + 
-#   geom_net(aes(from_id = from, to_id = to, linewidth = weight/388.5, color = as.factor(drink)), 
-#            arrowgap = .03, size = 3,arrowsize = .5,
-#            directed = TRUE, curvature = .2,
-#            #layout.alg = "fruchtermanreingold"
-#            layout.alg = "kamadakawai",
-#            layout.par = list(elen = distmatM1W2)
-#            ) + 
-#   scale_color_manual(values = cols[2:4], name = "Drinking\nBehavior") + 
-#   #labs(title = "Wave 3, M1") + 
-#   scale_size_continuous(name = "Weight") + 
-#   theme_net() + 
-#   theme(plot.title = element_text(hjust = .5), legend.position = 'none',plot.margin=unit(c(0,0,0,0),"mm"))
-# plotM2W2 <- ggplot(data = M2avgcov) + 
-#   geom_net(aes(from_id = from, to_id = to, linewidth = weight/325, color = as.factor(drink)), 
-#            arrowgap = .03, size = 3,arrowsize = .5,
-#            directed = TRUE, curvature = .2,
-#            #layout.alg = "fruchtermanreingold"
-#            layout.alg = "kamadakawai",
-#            layout.par = list(elen = distmatM2)
-#            ) + 
-#   scale_color_brewer(palette = "YlOrRd", name = "Drinking\nBehavior") + 
-#   #labs(title = "Wave 2, M2") + 
-#   scale_size_continuous(name = "Weight") + 
-#   theme_net() + 
-#   theme(plot.title = element_text(hjust = .5), legend.position = 'none',plot.margin=unit(c(0,0,0,0),"mm")) + 
-#   labs(title = "Model 2 Simulated Wave 2")
-# plotM2W3 <- ggplot(data = M2avgW2cov) + 
-#   geom_net(aes(from_id = from, to_id = to, linewidth = weight/357, color = as.factor(drink)), 
-#            arrowgap = .03, size = 3,arrowsize = .5,
-#            directed = TRUE, curvature = .2,
-#            #layout.alg = "fruchtermanreingold"
-#            layout.alg = "kamadakawai",
-#            layout.par = list(elen = distmatM2W2)
-#            ) + 
-#   scale_color_manual(values = cols[2:4], name = "Drinking\nBehavior") + 
-#   #labs(title = "Wave 3, M2") + 
-#   scale_size_continuous(name = "Weight") + 
-#   theme_net() + 
-#   theme(plot.title = element_text(hjust = .5), legend.position = 'none',plot.margin=unit(c(0,0,0,0),"mm"))
-# plotM3W2 <- ggplot(data = M3avgcov) + 
-#   geom_net(aes(from_id = from, to_id = to, linewidth = weight/258, color = as.factor(drink)), 
-#            arrowgap = .03, size = 3,arrowsize = .5,
-#            directed = TRUE, curvature = .2,
-#            #layout.alg = "fruchtermanreingold"
-#            layout.alg = "kamadakawai",
-#            layout.par = list(elen = distmatM3)
-#            ) + 
-#   scale_color_brewer(palette = "YlOrRd", name = "Drinking\nBehavior") + 
-#   #labs(title = "Wave 2, M3") + 
-#   scale_size_continuous(name = "Weight") + 
-#   theme_net() + 
-#   theme(plot.title = element_text(hjust = .5), legend.position = 'none',plot.margin=unit(c(0,0,0,0),"mm")) + 
-#   labs(title = "Model 3 Simulated Wave 2")
-# plotM3W3 <- ggplot(data = M3avgW2cov) + 
-#   geom_net(aes(from_id = from, to_id = to, linewidth = weight/372.5, color = as.factor(drink)), 
-#            arrowgap = .03, size = 3,arrowsize = .5,
-#            directed = TRUE, curvature = .2,
-#            #layout.alg = "fruchtermanreingold"
-#            layout.alg = "kamadakawai",
-#            layout.par = list(elen = distmatM3W2)
-#            ) + 
-#   scale_color_manual(values = cols[2:4], name = "Drinking\nBehavior") + 
-#   #labs(title = "Wave 3, M3") + 
-#   scale_size_continuous(name = "Weight") + 
-#   theme_net() + 
-#   theme(plot.title = element_text(hjust = .5), legend.position = 'none',plot.margin=unit(c(0,0,0,0),"mm"))
-# 
-# origW1 <- ggplot(data = alldat %>% filter(Wave==1), 
-#        aes(from_id = X1, to_id = X2, color = as.ordered(drink))) +
-#   geom_net(fiteach = FALSE, directed=TRUE, layout.alg = "kamadakawai", size = 3, arrowsize = .5, curvature = .2,
-#            arrowgap = .03) +
-#   scale_color_brewer(palette = "YlOrRd", name = "Drinking\nBehavior") + 
-#   theme_net() +
-#   theme(legend.position = 'none',plot.margin=unit(c(0,0,0,0),"mm"))
-# origW2 <- ggplot(data = alldat %>% filter(Wave==2), 
-#        aes(from_id = X1, to_id = X2, color = as.ordered(drink))) +
-#   geom_net(fiteach = FALSE, directed=TRUE, layout.alg = "kamadakawai", size = 3, arrowsize = .5, curvature = .2,
-#            arrowgap = .03) +
-#   scale_color_brewer(palette = "YlOrRd", name = "Drinking\nBehavior") + 
-#   theme_net() +
-#   theme(legend.position = 'none',plot.margin=unit(c(0,0,0,0),"mm")) + 
-#   labs(title = "Observed Wave 2")
-# origW3 <- ggplot(data = alldat %>% filter(Wave==3), 
-#        aes(from_id = X1, to_id = X2, color = as.ordered(drink))) +
-#   geom_net(fiteach = FALSE, directed=TRUE, layout.alg = "kamadakawai", size = 3, arrowsize = .5, curvature = .2,
-#            arrowgap = .03) +
-#   scale_color_manual(values = cols[2:4], name = "Drinking\nBehavior") + 
-#   theme_net() +
-#   theme(legend.position = 'none',plot.margin=unit(c(0,0,0,0),"mm"))
-# 
-# grid.arrange(origW2, plotM1W2,
-#               plotM2W2,plotM3W2, 
-#               ncol=2, clip = T)
-# alldataviz2 <- alldataviz + theme(legend.position = 'none')
-# ggdraw() +
-#   draw_plot(alldataviz2, x=0, y=.75, width=1, height=.25) + 
-#   draw_plot(plotM1W2, .1, .5, .5, .25) +
-#   draw_plot(plotM1W3, .3, .5, .5, .25) +
-#   draw_plot(plotM2W2, .1, .25, .5, .25) +
-#   draw_plot(plotM2W3, .3, .25, .5, .25) +
-#   draw_plot(plotM3W2, .1, 0, .5, .25) +
-#   draw_plot(plotM3W3, .3, 0, .5, .25) + 
-#   draw_plot_label(label = c("Original Data", "M1: Wave 2", "M1: Wave 3", "M2: Wave 2", "M2: Wave 3", "M3: Wave 2", "M3: Wave 3"))
-# 
-# M1avg2$Model <- "M1"
-# M2avg2$Model <- "M2"
-# M3avg2$Model <- "M3"
-# allavg <- rbind(as.data.frame(M1avg2), 
-#                 as.data.frame(M2avg2),
-#                 as.data.frame(M3avg2))
-# 
-# 
-# ggplot(data = allavg) + 
-#   geom_histogram(aes(x = weight), binwidth = 25) + 
-#   facet_grid(wave~Model, labeller = "label_both") + 
-#   labs(x = "Number of times an edge occurs in 1,000 simulations") +
-#   theme_bw()
-
-# ggplot(data = M3avgcov) + 
-#   geom_net(aes(from_id = from, to_id = to, linewidth = weight/251, color = as.factor(drink)), 
-#            directed = TRUE, curvature = .2,
-#            ealpha = .75,
-#            layout.alg = "kamadakawai",
-#            layout.par = list(elen = distmatM2)
-#            ) + 
-#   scale_color_brewer(palette = "YlOrRd", name = "Drinking\nBehavior") + 
-#   scale_size_continuous(name = "Weight") + 
-#   theme_net()
-# M1avg2$wave <- M1avg2$wave + 1
-# alldrink <- data.frame(drink2[,2:3])
-# alldrink <- gather(alldrink, wave,drink) %>% mutate(wave = parse_number(wave), id = rep(paste0("V",1:16),2))
-
-# M1avg2cov <- merge(M1avg2, alldrink, by.x = c("from", "wave"), by.y = c("id", "wave"), all = T)
-# ggplot(data = M1avg2cov) + # edge appears > 50% of the time. (is this actually what that is????) 
-#   geom_net(aes(from_id = from, to_id = to, linewidth = weight/388, color = as.factor(drink)), 
-#            directed = TRUE, curvature = .2, labelon = F,
-#            #layout.alg = "fruchtermanreingold"
-#            layout.alg = "kamadakawai", fiteach = T,
-#            layout.par = list(elen = distmat)
-#            ) + 
-#   scale_color_brewer(palette = "YlOrRd", name = "Drinking\nBehavior") + 
-#   facet_wrap(~wave, labeller = "label_both") + theme_net() + labs(title = "Average Wave 2 and Wave 3 Networks from Model M1")
-# ggplot(data = M1avg %>% filter(weight >= 51)) + 
-#   geom_net(aes(from_id = from, to_id = to, linewidth = weight/500), 
-#            #directed = TRUE,
-#            layout.alg = "mds", 
-#            layout.par = list(var = 'user', dist = 'none', exp = 3,vm = distmat))
-@
-
-To do create the average network visualization shown in Figure~\ref{fig:averagenet}, we follow the same procedure as in Section~\ref{sec:minds}, counting occurrences of each possible edge in the simulations, resulting in a summary network with weighted edges representing the number of times an edge appeared in the simulated wave 2 when simulating from the SAOM 1,000 times.
-As in Figure~\ref{fig:summnetM1}, edges only appear in the average network if they appear more than 5\% of the time in the simulations.
-In Figure~\ref{fig:averagenet}, we show the first wave, the ``average'' network from the three models we fit, and the second wave.
-\st{Comparing the three averages to waves 1 and 2, we see that they have very similar structure to wave 1.
-Model 2, which included the transitive triplet parameter, seems to have created a larger connected component overall than models 1 and 3.
-In particular, if we look at the group of nodes $\{10,11,14\}$, we see they are very strongly connected within the three average networks, and they are completely separate from the other nodes in the true wave 2.
-None of the three average networks show node 16 gaining ties as it does in wave two, nor do they show nodes 4 and 7 becoming isolated.
-In Model 2, however, the ties to node 7 appear much weaker than in Model 1 or Model 2, suggesting that of the three, Model 2 may be the best fit for our data.}  %XXXX ST commented out after I realized I had made the same mistake as figure 7.
-XX This could be an indication that, despite the highly significant nature of the parameter $\beta_3$ according to the $t$-tests in \texttt{RSiena}, the transitive triplet parameter should not be included because it does not result in a structure similar to the data.
-
-% \st{What are some things that could go in here?:
-% \begin{itemize}
-% \item "average" networks -- you've already done this! Write it up, make several examples. How does the procedure for averaging generalize? What is it good for? 
-% \item distributions of fitted parameters. again, you've already done this!!! just write it up and put it in here. 
-% \item view correlations between model parameters under different model sets. eg 2 betas, 3 betas, 4 betas, etc.
-% \end{itemize}
-%  }
-
-\subsubsection{Fitting the same model to different data}\label{sec:samemoddiffdat}
-% table of fitted M1 M2 M3
-\begin{table}
-\scalebox{0.9}{
-\begin{tabular}{l|ccc|ccc }
-  \hline
-  & \multicolumn{3}{|c|}{Friendship Data} & \multicolumn{3}{|c|}{Senate Data} \\
-  \hline
-  & M1 & M2 & M3 & M1 & M2 & M3 \\
-  $\alpha_1$ & \Sexpr{SFE$M1[1]} & \Sexpr{SFE$M2[1]} & \Sexpr{SFE$M3[1]} & \Sexpr{SSE$M1[1]} & \Sexpr{SSE$M2[1]} & \Sexpr{SSE$M3[1]} \\
-  $\alpha_2$ & \Sexpr{SFE$M1[2]} & \Sexpr{SFE$M2[2]} & \Sexpr{SFE$M3[2]} & \Sexpr{SSE$M1[2]} & \Sexpr{SSE$M2[2]} & \Sexpr{SSE$M3[2]}  \\
-  $\alpha_3$ & -- & -- & -- & \Sexpr{SSE$M1[3]} & \Sexpr{SSE$M2[3]} & \Sexpr{SSE$M3[3]} \\
-  $\beta_1$ & \Sexpr{SFE$M1[3]} & \Sexpr{SFE$M2[3]} & \Sexpr{SFE$M3[3]} & \Sexpr{SSE$M1[4]} & \Sexpr{SSE$M2[4]} & \Sexpr{SSE$M3[4]} \\
-  $\beta_2$ & \Sexpr{SFE$M1[4]} & \Sexpr{SFE$M2[4]} & \Sexpr{SFE$M3[4]} & \Sexpr{SSE$M1[5]} & \Sexpr{SSE$M2[5]} & \Sexpr{SSE$M3[5]}\\
-$\beta_3$ & -- & \Sexpr{SFE$M2[5]} & -- & -- & \Sexpr{SSE$M2[6]} & -- \\
-  $\beta_4$ & -- & --  & \Sexpr{SFE$M3[6]} & -- & -- & \Sexpr{SSE$M3[7]} \\
-  \hline
-\end{tabular}
-}
-\caption{\label{tab:meanests} The means (standard deviations) of parameter values estimated from repeated fittings of $M1, M2, M3$ to the small friendship network and the senate collaboration network.
-Each model was fit 1,000 times to the friend data, while each model was fit 100 times to the senate data.}
-\end{table}
-\subsubsection{Fitting the same model to the same data}\label{sec:samemodsamedat}
-
-%We look at collections produced ``from exploring the space of all possible models," ``by varying model settings", ``by fitting the same model to different datasets", and ``by treating each iteration as a model" \citep[p.~210-11]{modelvis}.
-Because fitting a SAOM requires MCMC methods, we also fit the same model many times, resulting in distributions of fitted parameters for one data set.
-
-To our small friendship network, we fit three models, M1, M2, and M3, using \texttt{RSiena} 1,000 times each.
-We then looked at the distribution of the fitted values, which are shown in Figure~\ref{fig:distests}.
-We can see from these distributions that the inclusion of the jumping transitive triplet parameter, $\beta_3$ is obviously affecting the distributions of the other four parameters included in all models, $\alpha_1$, $\alpha_2$, $\beta_1$, and $\beta_2$.
-\st{When $\beta_3$ is included, its estimate is postitive, meaning that friendships between two girls with different drinking behaviors tend to form when there is an intermediary who is already friends with the two girls.
-The inclusion of this parameter leads to increases in the rate parameters' estimates, suggesting that encouraging the transitive triplet behavior means that the girls would also change friends more frequently.
-The outegree parameter, $\beta_1$ decreases when $\beta_3$ is included, while the reciprocity parameter, $\beta_2$ increases.
-This implies the girls in the data prefer to form closer friend groups, as indicated by reciprocated ties and jumping transitive triplet formation, as opposed to being popular and having many friends.
-Having many friends who do not reciprocate is discouraged by M2.} In comparison with models M1 and M3, model M2 typically has higher estimates of the rate parameter, meaning that the inclusion of the covariate statistic in the model leads to higher estimates of the number of times, on average, a node gets to change its ties.
-It is not clear, however, that the addition of a parameter to the objective function \textit{should} effect the estimation of the rate parameters, so we continue to explore the collection of parameter estimates.
-
-To further investigate this potentially problematic relationship between the parameter values, we look at correlations between each of the parameter estimates in each model.
-In Figure~\ref{fig:corplots}, we examine correlations between each of pair of parameters within each model and overall.\nocite{ggally} The strongest correlation within each model is between $\beta_1$ and $\beta_2$, with absolute value of correlation between those two parameter values greater than 0.90 in all three models.
-The $\beta_1$ parameter is also highly correlated with the $\beta_3$ parameter within model M2, but it is not as highly correlated with the $\beta_4$ parameter in model M3.
-
-<<getdata>>=
-nulls <- read.csv("data/distribution_null_model.csv")
-alt <- read.csv("data/distribution_jumpTT_model.csv")
-alt2nd <- read.csv("data/distribution_dblpairs_model.csv")
-means <- simu2 %>% group_by(Model, parameter) %>% summarize(
-  means = mean(estimate)
-)
-null_mod_sv <- as.numeric(nulls[,-c(1,6,7)] %>% summarise_each(funs(mean)))
-alt_mod_sv <- as.numeric(alt[,-c(1, 7,8)] %>% summarise_each(funs(mean)))
-alt_mod2_sv <- as.numeric(alt2nd[,-c(1,7,8)] %>% summarise_each(funs(mean)))
-@
-
-<<distests, fig.cap="Distribution of fitted parameter values for our three SAOMs.
-The inclusion of $\\beta_3$ or $\\beta_4$ clearly has an effect on the distribution of the rate parameters, $\\alpha_1$ and $\\alpha_2$.", fig.height=4>>=
-
-#simu2 <- read.csv("data/simulation-1000-M1-M2-M3.csv")
-
-ggplot(data = simu2) + 
-  geom_density(aes(x = estimate, fill = Model), alpha = .5) + 
-  facet_wrap(~parameter, scales = 'free') + 
-  ThemeNoNet + 
-  theme(legend.position = 'bottom')
-@
-
-<<corplots, fig.cap="A matrix of plots demonstrating the strong correlations between parameter estimate in our SAOMs.
-The strongest correlation within each model is between $\\beta_1$ and $\\beta_2$.", fig.width=8>>=
-
-#simu2 <- read.csv("data/simulation-1000-M1-M2-M3.csv")
-simu_spread <- simu2 %>% spread(parameter, estimate)
-ggpairs(simu_spread, columns = 3:8, ggplot2::aes(colour=Model, alpha = .5)) + theme(plot.title = element_text(size = 12,
-                                            face = 'plain', angle = 0, family = "Times New Roman"), 
-                  axis.title.x = element_text(size = 12,
-                                            face = 'plain', angle = 0, family = "Times New Roman", inherit.blank = T),
-                  axis.title.y = element_text(size = 12,
-                                            face = 'plain', angle = 90, inherit.blank = T),
-                  axis.text.x.top = element_text(size = 12,
-                                            face = 'plain', angle = 90, inherit.blank = TRUE), 
-                  axis.text.x.bottom = element_text(size = 12,
-                                            face = 'plain', angle = 90, inherit.blank = TRUE), 
-                  axis.text.y.left = element_text(size = 12,
-                                            face = 'plain', angle = 0, family = "Times New Roman", inherit.blank = T),
-                  axis.text.y.right = element_text(size = 12,
-                                            face = 'plain', angle = 0, family = "Times New Roman", inherit.blank = T),
-                  strip.text.x = element_text(size = 14,
-                                            face = 'plain', angle = 0, family = "Times New Roman", 
-                                            margin = margin(t = 3, r = 0, b = 3, l = 0, unit = "pt"), inherit.blank = T),
-                  strip.text.y =element_text(size = 14,
-                                            face = 'plain', family = "Times New Roman", angle = 90,margin = margin(t = 3, r = 3, b = 3, l = 3, unit = "pt"),inherit.blank = T),
-                  strip.background = element_rect(colour = "black", fill = "white")
-                    )
-@
-
-<<corplotsSenate, eval = FALSE>>=
-senM4est <- read.csv("data/congress/senateM4ests.csv")
-senM5est <- read.csv("data/congress/senateM5ests.csv")
-names(senM4est) <- c("alpha1", "alpha2", "alpha3", "beta1", "beta2", "beta5")
-names(senM5est) <- c("alpha1", "alpha2", "alpha3", "beta1", "beta2", "beta6")
-senM4est$Model <- "M4"
-senM5est$Model <- "M5"
-senM4est$Sim <- 1:nrow(senM4est)
-senM5est$Sim <- 1:nrow(senM5est)
-senM4est2 <- senM4est %>% gather(parameter, estimate, alpha1:beta5) 
-senM5est2 <- senM5est %>% gather(parameter, estimate, alpha1:beta6) 
-senM45est <- rbind(senM4est2, senM5est2)
-senM45est %>% spread(parameter, estimate)-> senM45est2
-ggpairs(senM45est2, columns = 3:9, ggplot2::aes(colour=Model, alpha = .5))
-@
-
-\subsection{Explore algorithms, not just end result}
-
-\st{The last principle of model visualization is to explore the process of fitting the model, instead of just focusing on the end result.
-This principle is perhaps the most important for SAOMs because} the model fitting process in \texttt{RSiena} involves several simulation steps that are hidden from the user.
-Hiding the MCMC steps is practical and efficient if a researcher is primarily interested in fitting one model to a set of longitudinal network data, obtaining parameter estimates, \st{and drawing conclusions or making predictions}.
-We are more interested in \textit{how} the models are fit, so we extracted and explored the different steps of that process.
-
-A key component of each step of the SIENA method of moments algorithm is the ``microstep" process.
-A series of microsteps is obtained by simulating from the model in its current state, $x(t_m)$ with current parameter values $\mathbf{\theta}_{0} = \{\alpha_{1_0}, \dots \alpha_{{m-1}_0}, \beta_{1_0}, \dots, \beta_{K_0}\}$, to the next state, $x(t_{m+1})$.
-This microstep process stops when the simulated network has acheived the same number of differences, $C$, from $x(t_m)$ as $x(t_{m+1})$, where
-$$C = \sum_{i \neq j} |x_{ij}(t_{m+1}) - x_{ij}(t_{m})|.$$ 
-This simulation process follows the steps of the continuous-time Markov chain.
-\st{Each tie change in the CTMC is referred to as one ``microstep".
-At each microstep}, an ``ego node" is selected to make a change, and the chosen ego node randomly makes one change in its ties according to the probabilities, $\{p_{ij}: i\neq j \in \{1, \dots, n\}\}$, determined by its objective function.
-The options for change are (1) removing a current tie, (2) adding a new tie, or (3) making no change at all.
-Saving and exploring all of these steps is not computationally efficient if one is only interested in estimating parameter values, but they can be saved and extracted using options in \texttt{RSiena}, which is exactly what we did to create our visualizations.
-Between two network observations $x(t_m)$ and $x(t_{m+1})$, there can be dozens, hundreds, or even thousands of microsteps, depending on the size of the network and the number of changes between two network observations.
-We wanted to view these in-between steps in order to better understand the behavior of the underlying continuous-time Markov chain.
-
-The first vizualization we present here is an animation of microsteps simulated to emulate the transition steps of the CTMC from wave 1 to wave 2 of the small friendship network example shown in Figure ~\ref{fig:alldataviz}.
-Movies similar to this animation were used to visualize the changes of dynamic networks in \citet{dynnetviz}.
-When each ego node is selected in a microstep, it is emphasized in the animation, then the associated edge either appears or disappears.
-If there are no changes at a particular microstep, no changes are seen.
-Some frames of the animations are shown in Figure~\ref{fig:makinggif}, and the full movie can be viewed at INSERT LINK TO GIF.
-
-<<makinggif, fig.cap='A selection of images in the microstep animation.
-The selected edges and nodes are emphasized by changing the size and the color, then when edges are removed, they fade out, shrinking in size, while the nodes change color and shrink to blend in with the other nodes.', cache=TRUE, message = FALSE, warning = FALSE, fig.height=2.5>>=
-# Need old version of dplyr to get this to work: 
-if (!require("dplyr050",character.only = TRUE)) {
-      devtools::install_github("sctyner/dplyr050")
-}
-library(dplyr050)
-friend.data.w1 <- as.matrix(read.table("data/s50_data/s50-network1.dat"))
-friend.data.w2 <- as.matrix(read.table("data/s50_data/s50-network2.dat"))
-friend.data.w3 <- as.matrix(read.table("data/s50_data/s50-network3.dat"))
-drink <- as.matrix(read.table("data/s50_data/s50-alcohol.dat"))
-fd2.w1 <- friend.data.w1[20:35,20:35]
-fd2.w2 <- friend.data.w2[20:35,20:35]
-fd2.w3 <- friend.data.w3[20:35,20:35]
-friendshipData <- array(c(fd2.w1, fd2.w2,fd2.w3), dim = c(16, 16, 3))
-
-# running Siena
-#friendship <- sienaDependent(friendshipData)
-#alcohol <- varCovar(drink[20:35,])
-#mydata <- sienaDataCreate(friendship, alcohol)
-#myeffnull <- getEffects(mydata)
-#myalgorithm <- sienaAlgorithmCreate(projname = 's16_3')
-set.seed(823746)
-load("data/ansnullpaper.rda")
-ansnullchains <- get_chain_info(ansnull)
-ansnullchains %>%
-  filter(period == 1) %>%  #only look at chains from wave 1 to wave 2
-  group_by(rep) %>%
-  dplyr::select(rep, from = ego, to = alter) %>%
-  mutate(val = as.numeric(!from == to),
-         from = paste0("V", parse_number(from)+1), # make the chains
-         to = paste0("V", parse_number(to)+1)) -> ansnullchainsw1w2
-colnames(fd2.w1) <- paste0("V", 1:16)
-rownames(fd2.w1) <- paste0("V", 1:16)
-wave1friends <- fortify(as.adjmat(fd2.w1))
-ms1 <- listMicrosteps(dat = wave1friends,
-                      microsteps = filter(ansnullchainsw1w2, rep == 1))
-microsteps <- ms1[1:20]
-
-microsteps_df <- getMicrostepsDF(microsteps)
-pte <- pretween_edges(microsteps = microsteps)
-ptv <- pretween_vertices(pte = pte, layoutparams = list(n = 16))
-# step 1 is remove, step 9 is add. 
-# get tweened data for all microsteps
-pall <- tween_microsteps(ptv, pte, microsteps_df)
-
-static_pall <- filter(pall, .frame %in% c(15,20,22,24,26:28, 71:73, 75, 79, 81, 84))
-
-static_pall$type <- NA
-static_pall$type[which(static_pall$.frame %in% c(15,20,22,24,26:28))] <- "Remove Edge"
-static_pall$type[which(static_pall$.frame %in% c(71:73, 75, 79, 81, 84))] <- "Add Edge"
-
-set.seed(34569234)
-create_net_animate(static_pall) + 
-  facet_wrap(~.frame, nrow=2, labeller = 'label_both') + 
-  ThemeNet + 
-  theme(panel.background = element_rect(color = 'black'))
-@
-
-The top row of Figure~\ref{fig:makinggif} shows an edge being removed, and the bottom row shows one being added.
-In both cases, the ego nodes chosen to act change color from black to red, and they also increase greatly in size.
-In the case of an edge being removed, shown in the top row, the edge that currently exists is emphasized with the same color and size change that the node gets, and as the animation proceeds the edge shrinks to nothing, as the ego node shrinks and changes color back to its original black.
-If an edge is being added, as in the bottom row of the figure, the ego node's appearances changes in the sampe way as when the edge is being removed, whil the edge appears colored red from nothing, and grows to a large size, then changes color and size to match the rest of the edges, while the node shrinks and changes color to match the other nodes.
-
-We also use animation to view the changing structure of the adjacency matrix the microsteps.
-The adjacency matrices for the three waves of friendship data as shown in Figure~\ref{fig:alldataviz} are ordered by node id.
-There are 16 nodes in the data, numbered 1-16, and that order is used on the $x$ and $y$ axes for the matrix visualization.
-Viewing the adjacency matrices with this arbitrary ordering does not provide much information to the viewer about the underlying structure of the network.
-This lack of perceived structure would be exacerbated in an animation, so we adjust the ordering so that the viewer can perceive more of the structure of the graph.
-This process is known as matrix seriation \citep{seriation}.
-
-<<adjmatorder, message = FALSE, warning = FALSE, fig.cap="On the left, the starting friendship network represented in adjacency matrix form, ordered by vertex id.
-On the right, the same adjacency matrix is presented after ordering the vertices by one repetition of the microstep simulation process from wave one to wave two.", fig.width=8, fig.height=3>>=
-# first present the original wave 1 adj mat, then present the adj mat after being ordered by the 1st pc of the big adjmat A. 
-# reinstall new dplyr
-numordersf <- data.frame(fd2.w1) %>% 
-  mutate(from = rownames(fd2.w1)) %>% 
-  gather(key = to ,value = value, -from)
-pcaordersf <- numordersf
-
-numordersf$from <- factor(numordersf$from, levels = paste0("V", 1:16))
-numordersf$to <- factor(numordersf$to, levels = rev(paste0("V", 1:16)))
-
-p1 <- ggplot(data = numordersf, aes(y = from, x = to)) + geom_tile(color = 'black', aes(fill = as.factor(value))) + 
-  scale_fill_manual(values = c('white', 'black')) +
-  scale_x_discrete(position = "top", labels = as.character(1:16)) +
-  scale_y_discrete(labels = rev(as.character(1:16)))  + 
-  ThemeNoNet + 
-  theme(legend.position = 'none') + 
-  coord_fixed()
-
-# histogram of number of MS in 1000 reps
-# sfmsall %>% group_by(rep) %>% summarize(noMs = max(ms)) %>% ggplot() + geom_histogram(aes(x = noMs), binwidth = 5)
-
-sfmsall <- read_csv("data/smallfriendsmicrosteps.csv")
-
-sfmsall %>% 
-  group_by(from, to) %>% 
-  filter(rep == 1) %>% 
-  summarise(count = n()) -> sfmsall_summ
-sfmsall_summ$from <- factor(sfmsall_summ$from, levels = paste0('V', 1:16))
-sfmsall_summ$to <- factor(sfmsall_summ$to, levels = paste0('V', 1:16))
-sfmsall_summ %>% 
-  spread(to, count, fill = 0) %>% 
-  data.frame -> allmsrep1_adjmat
-allmsrep1_adjmat$V16 <- 0
-allmsrep1_adjmat <- allmsrep1_adjmat[,-1]
-allmsrep1_adjmat <- as.matrix(allmsrep1_adjmat)
-rownames(allmsrep1_adjmat) <- paste0("V", 1:16)
-
-pca1 <- prcomp(allmsrep1_adjmat)
-order1 <- names(sort(pca1$rotation[,1], decreasing = T))
-order2 <- names(sort(abs(pca1$rotation[,1]), decreasing = T))
-
-pcaordersf$from <- factor(pcaordersf$from, levels = rev(order1))
-pcaordersf$to <- factor(pcaordersf$to, levels = order1)
-
-p2 <- ggplot(data = pcaordersf, aes(x = to, y = from)) + geom_tile(color = 'black', aes(fill = as.factor(value))) + 
-  scale_fill_manual(values = c('white', 'black')) +
-  scale_x_discrete(position = "top", labels = str_replace(order1, "V", "")) + 
-  scale_y_discrete(labels = rev(str_replace(order1, "V", ""))) + 
-  ThemeNoNet + 
-  theme(legend.position = 'none') +
-  coord_fixed()
-
-grid.arrange(p1,p2, nrow = 1)
-@
-
-To reorder the vertices for the matrix visualization, we first constructed a cumulative adjaceny matrix, $\mathbf{A}^{cum}$, for the series of microsteps simulating the network from $x(t_m)$ to $x(t_{m+1})$.
-A single entry in the cumulative adjacency matrix, $\mathbf{A}^{cum}_{ij}$, is the total number of times the edge from node $i$ to node $j$ appears in the network from the initial observation, $x(t_m) \equiv X(0)$ to the final result of the last microstep, $X(R)$, where $R$ is the total number of microsteps taken:
-$$ \mathbf{A}^{cum}_{ij} = \sum_{r = 0}^R X_{ij}(r). $$
-We then performed a principal component analysis on $\mathbf{A}^{cum}$, and used the values of the first principal component to order the vertices on the $x$ and $y$ axes for the adjacency matrix animation.
-For one such series of microsteps simulated by \texttt{RSiena}, we present the adjacency matrix ordered by the (arbitrary) vertex id alongside the seriated adjacency microsteps using the first principal component loading on the cumulative adjacency matrix, $\mathbf{A}^{cum}$, in Figure~\ref{fig:adjmatorder}.
-
-<<adjmatani, fig.cap='A selection of frames from the adjacency matrix visualization animation for one series of microsteps.
-(Ego and alter node labels are removed to declutter the graph.) At the beginning of the animation, shown in the top row, there are two clearly connected components: one in the top left corner, and one in the bottom right corner.
-By the end, the component in the top left has spread out, while the bottom right component has shrunk, but remains closely connected.', warning = FALSE, message = FALSE, results='hide', fig.height=2.5>>=
-adjmatall <- matrix(0, nrow = 16, ncol = 16)
-rownames(adjmatall) <- paste0("V", 1:16)
-colnames(adjmatall) <- paste0("V", 1:16)
-dat2 <- sfmsall %>% filter(rep == 1)
-adjmatall2 <- NULL
-for(i in 0:55){
-  MS <- dat2[dat2$ms == i,c("from", "to")]
-  myadjmat <- adjmatall
-  for(j in 1:nrow(MS)){
-    myadjmat[MS[j,"from"][[1]],MS[j,"to"][[1]]] <- 1
-  }
-  adjmatall2 <- rbind(adjmatall2, myadjmat)
-}
-
-adjmatall2 <- cbind(adjmatall2, rep(0:55, each = 16))
-adjmatall2 <- data.frame(adjmatall2)
-adjmatall2$from <- rep(paste0("V", 1:16), 56)
-names(adjmatall2)[17] <- "ms"
-adjmatall2 %>% gather(to, val, V1:V16) -> adjmatall3
-
-adjmatall3$to <- factor(adjmatall3$to, levels = order1)
-adjmatall3$from <- factor(adjmatall3$from, levels = rev(order1))
-
-adjmatall3$frame <- adjmatall3$ms
-#plots <- list()
-#for(i in 0:55){
-#plots[[i + 1]] <- 
-  ggplot(data = adjmatall3 %>% filter(ms %in% c(1,3,5,7,9,11,13,43,45,47,49,51,53,55))) + 
-  geom_tile(aes(x = to, y = from, fill = as.factor(val)), color = 'black') + 
-  scale_fill_manual(values = c("white", "black")) + 
-  scale_x_discrete(position = "bottom", labels = str_replace(order1, "V", "")) + 
-  scale_y_discrete(labels = rev(str_replace(order1, "V", ""))) + 
-    labs(x = "alter node (j)", y = "ego node (i)") + 
-  ThemeNoNet + 
-  theme(legend.position = 'none', axis.text.x.bottom = element_blank(), 
-        axis.text.y.left = element_blank()) +
-  coord_fixed() + 
-  facet_wrap(~frame, labeller = 'label_both', nrow = 2)  
-#}
-#plots 
-@
-
-Using the principal component analysis on $\mathbf{A}^{cum}$ to order the rows and columns of the adjacency matrix visualization clearly shows the two distinct connected components in the first wave of the network, which are difficult to find in the arbitrarily ordered visualization.
-We use this layout to fix the layout in the animation of one of the microstep process simulations.
-This animation, which is shown in Figure~\ref{fig:adjmatani} is very simple: a square appears or disappears in the animation as that edge appears or disappears in the microstep process.
-Through this animation, we can see edges appearing, and then later on disappearing.
-These in-between steps are not shown when we look at the network at our discrete observation points.
-By viewing this animation, we gain a better understanding of the dynamics of this model.
-<<histogramalters, fig.width = 8, fig.height=10, fig.cap="Each panel shows the theoretical (as lines) and empirical (as points) probabilities of the chosen ego node changing its tie to each of the other nodes.
-The color of the line indicates whether the tie from the ego to the alter node is being added, removed, or if there is no change to the network in this step.">>=
-ansnullchains %>% group_by(period, rep) %>% 
-  mutate(ms = row_number()) %>% 
-  filter(ms == 1, period == 1) %>% 
-  mutate(alterProb = exp(logAlterProb),
-         ego = as.numeric(as.character(ego)),
-         alter = as.numeric(as.character(alter))) %>% 
-  ungroup() %>% 
-  group_by(ego, alter) %>% 
-  summarize(count = n(), alterProb = unique(alterProb)) %>% ungroup() %>% 
-  group_by(ego) %>% 
-  mutate(csum = cumsum(alterProb)) %>% 
-  arrange(ego, csum) %>% 
-  mutate(totego = sum(count), 
-         empprob = count / totego) -> cumprobs 
-cumprobs$ego <- factor(paste0("V",cumprobs$ego+1), levels = paste0("V", 1:16))
-cumprobs$alter <- factor(paste0("V",cumprobs$alter+1), levels = paste0("V", 1:16))
-
-# is the tie that's added new, removed, or does the network stay the same
-
-changes <- cumprobs %>% dplyr::select(ego, alter) %>% unique
-
-wave1friends2 <- wave1friends[,-3]
-wave1friends2$type <- "remove"
-changes2 <- left_join(changes, wave1friends2, c("ego" = "from", "alter" = "to"))
-
-changes2$type[is.na(changes2$type)] <- ifelse(changes2$ego[is.na(changes2$type)] == changes2$alter[is.na(changes2$type)], "noChange", "add")
-
-cumprobs %>% left_join(changes2) -> changes3
-
-changes3$ego <- factor(changes3$ego, levels = paste0("V", 1:16), labels = 1:16) 
-changes3$alter <- factor(changes3$alter, levels = paste0("V", 1:16), labels = 1:16)
-
-ggplot(data = changes3) + 
-  geom_segment(aes(x = alter, xend = alter, y = 0, yend = empprob, color = type), size = 2, alpha = .75) + 
-  geom_point(
-    aes(x = alter , y = alterProb), shape = 4, size = 2, alpha = .75) + 
-  labs(x = "alter node\n", y = "\nprobability of alter node") + 
-  coord_flip() + 
-  facet_wrap(~ego, labeller = "label_both") + 
-  ThemeNoNet
-# 
-# cumprobs %>% ggplot() + geom_histogram(aes(x = alter, weight = count), color = 'black', fill = 'white', binwidth = 1) + scale_x_continuous(labels = paste0("V",1:16), breaks = 0:15) + coord_flip() + facet_wrap(~ego, labeller = "label_both") + theme_gray()
-
-# cprobslist <- split(cumprobs, cumprobs$ego)
-# plots <- list()
-# for(i in 1:length(cprobslist)){
-#   cprobslist[[i]] %>%
-#   ggplot() + 
-#   geom_histogram(aes(x = as.numeric(as.character(alter)),
-#                      y = alterProb), stat = 'identity', fill = 'white', color = 'black') + 
-#   geom_line(aes(x = as.numeric(as.character(alter)), 
-#                 y = csum)) + 
-#   geom_point(aes(x = as.numeric(as.character(alter)), 
-#                 y = csum)) +
-#   scale_x_continuous(limits = c(-.5, 15), breaks = 0:15, labels = 0:15) +   
-#   geom_hline(yintercept = 1, color = 'red', linetype = 'dashed') + 
-#   labs(x = "alter", y = expression(p[ij]), title = paste("Ego node:", names(cprobslist)[i])) + # ylab(expression(p[ij])) + 
-#   theme_gray() + 
-#     theme(plot.title = element_text(size = 10), 
-#           axis.title = element_text(size = 8), 
-#           plot.margin = unit(rep(0,4),"pt")) -> p
-#   plots[[i]] <- p
-# }
-# grid.arrange(plots[[1]], plots[[2]], plots[[3]], plots[[4]],
-#              plots[[5]], plots[[6]], plots[[7]], plots[[8]],
-#              #plots[[9]], plots[[10]], plots[[11]], plots[[12]],
-#              #plots[[13]], plots[[14]], plots[[15]], plots[[16]], 
-#              ncol = 2, padding = unit(0, 'line'))
-@
-
-We also attempt to better understand the microstep process by visualizing the observed transition probabilites for the first microstep in the process.
-We only do the first step of many for now because the \texttt{RSiena} transition probabilities after the first step are only directly comparable for identical steps due to the conditioning on the current network state in the model.
-In other words, for $i \neq j \neq k \neq \ell \in \{1, \dots, n\}$, the probability of tie $x_{k\ell}$ changing in the second microstep is different in the case of $x_{ij}$ changing in the first step from the case of $x_{ij}$ not changing in the first step.
-Thus we have one transition probability for the first microstep taken for each of the 1,000 simulations to visualize.
-In Figure~\ref{fig:histogramalters}, we present each acting node, (ego node) and the resulting probabilities of tie changes.
-The probability shown by the bars is the theoretical probability, according to the objective function, of the ego node changing its tie to the alter node, while the probability shown by the points is the empirical probability of that change being made.
-The empirical probability is calculated by counting all instances of the ego node first, then computing the proportion of each different alter node change.
-In most cases, they are almost identical, which demonstrates that the algorithm is performing as expected.
-
-<<heatmap, fig.align='center', fig.cap='A heatmap showing the empirical transition probabilities for the first microstep in 1,000 simulations.
-The acting ego node is on the y-axis, and the alter node is on the x-axis.
-There were many ties with empirical probability zero.', fig.height=4>>=
-ansnullchains %>% 
-  filter(period == 1) %>% 
-  group_by(rep) %>% 
-  mutate(msno = row_number()) %>% 
-  filter(msno == 1) %>% 
-  mutate(ego = paste0("V",as.numeric(as.character(ego))+1),
-         alter = paste0("V",as.numeric(as.character(alter))+1), 
-         prob = exp(logAlterProb)) -> probsDat
-
-# probsDat %>% group_by(ego, alter) %>% summarize(count = n()) %>% dim
-# 103, 3
-ggplot(data = probsDat, 
-       aes(y = factor(ego, levels = order1), 
-             x = factor(alter, levels = rev(order1)), 
-             fill = prob)) + 
-  geom_tile(color = 'black') + 
-  scale_fill_gradient(low = 'white', high = 'blue', limits = c(0,1)) +
-  labs(y = "ego", x = "alter")+
-  scale_x_discrete(position = 'top', labels = str_replace(order1, "V", "")) +
-  scale_y_discrete(labels = rev(str_replace(order1, "V", ""))) +
-  ThemeNoNet + theme(panel.grid = element_blank()) + 
-  coord_fixed()
-@
-
-Another way to view these transition probabilities is through the adjacency matrix visualization.
-In Figure~\ref{fig:heatmap}, we build on the concept of the ordered adjacency matrix of Figure~\ref{fig:adjmatorder}.
-This heatmap shows the transition probabilities of all ties that are changed in the first microstep of the 1,000 simulations.
-The heatmap is noticeably sparse: %showing the lack of coverage in the model.
-of the 256 possible steps, only 103, or about 40\%, are taken in the 1,000 simulated chains.
-
-<<allsteps, fig.cap="Two simulations (out of 1,000) of the microstep process from wave 1 to wave 2.
-The $x$ axis is the microstep number, with step 0 representing the first wave of data and the final step representing the second wave of data.
-We can see that many edges are underrepresented in this process: they are in the second wave, but never appear in the microsteps.">>=
-smallfriends <- read_csv("data/smallfriends4Geomnet.csv")
-names(smallfriends)[1:2] <- c("from", "to")
-msall1 <- netvizinf::listMicrosteps(dat = wave1friends,
-                      microsteps = filter(ansnullchainsw1w2, rep == 1))
-msall1_df <- plyr::rbind.fill(lapply(X = seq_along(msall1), FUN = function(i, x) {
-        x[[i]]$ms <- i - 1
-        return(x[[i]])
-    }, msall1))
-msall1_df <- msall1_df %>% dplyr::select(from, to, ms)
-wave2 <- smallfriends %>% filter(Wave == 2) %>% na.omit %>% data.frame()
-msall1_df <- msall1_df %>% add_row(from = paste0("V",wave2$from),
-                      to = paste0("V",wave2$to), 
-                      ms = max(msall1_df$ms)+1)
-edges <- msall1_df %>% 
-  group_by(from, to) %>%
-  summarize(edgeappears = n()) %>% 
-  arrange(desc(edgeappears)) %>% ungroup() %>%
-  mutate(eid = row_number())
-
-msall1_df2 <- left_join(msall1_df, edges, by = c("from" = "from", "to" = "to"))
-
-p1 <- ggplot(data = msall1_df2) + 
-  geom_tile(aes(x = ms, y = -eid)) + 
-  annotate("text", x = c(-1, 84, 40), y = c(-30, -30, -60), label = c("Wave 1", "Wave 2", "Microsteps"), angle = c(90, 90, 0)) + 
-  geom_vline(xintercept = c(0.5,81.5)) + 
-  #scale_fill_continuous(low = "red", high = "blue") + 
-  ThemeNoNet + 
-  theme(legend.position = 'none', axis.ticks.y = element_blank(),
-        axis.text.y = element_blank(), axis.title = element_blank()) +
-  labs(x = "Microstep")
-
-# rep 2 
-msall2 <- netvizinf::listMicrosteps(dat = wave1friends,
-                      microsteps = filter(ansnullchainsw1w2, rep == 2))
-msall2_df <- plyr::rbind.fill(lapply(X = seq_along(msall2), FUN = function(i, x) {
-        x[[i]]$ms <- i - 1
-        return(x[[i]])
-    }, msall2))
-msall2_df <- msall2_df %>% dplyr::select(from, to, ms)
-msall2_df <- msall2_df %>% add_row(from = paste0("V",wave2$from),
-                      to = paste0("V",wave2$to), 
-                      ms = max(msall2_df$ms)+1)
-
-edges2 <- msall2_df %>% 
-  group_by(from, to) %>%
-  summarize(edgeappears = n()) %>% 
-  arrange(desc(edgeappears)) %>% ungroup() %>%
-  mutate(eid = row_number())
-
-msall2_df2 <- left_join(msall2_df, edges2, by = c("from" = "from", "to" = "to"))
-
-p2 <- ggplot(data = msall2_df2) + 
-  geom_tile(aes(x = ms, y = -eid)) + 
-  annotate("text", x = c(-1, 51, 25), y = c(-25, -25, -40), label = c("Wave 1", "Wave 2", "Microsteps"), angle = c(90, 90, 0))+ 
-  geom_vline(xintercept = c(0.5,49.5)) + 
-  #scale_fill_continuous(low = "red", high = "blue") +
-  ThemeNoNet + 
-  theme(legend.position = 'none', axis.ticks.y = element_blank(),
-        axis.text.y = element_blank(), axis.title = element_blank()) #+
-  #labs(x = "Microstep", y = "Edges (most frequent at top)")
-
-# rep 3 
-# msall3 <- netvizinf::listMicrosteps(dat = wave1friends,
-#                       microsteps = filter(ansnullchainsw1w2, rep == 3))
-# msall3_df <- plyr::rbind.fill(lapply(X = seq_along(msall3), FUN = function(i, x) {
-#         x[[i]]$ms <- i - 1
-#         return(x[[i]])
-#     }, msall3))
-# msall3_df <- msall3_df %>% dplyr::select(from, to, ms)
-# msall3_df <- msall3_df %>% add_row(from = paste0("V",wave2$from),
-#                       to = paste0("V",wave2$to), 
-#                       ms = max(msall3_df$ms)+1)
-# 
-# edges3 <- msall3_df %>% 
-#   group_by(from, to) %>%
-#   summarize(edgeappears = n()) %>% 
-#   arrange(desc(edgeappears)) %>% ungroup() %>%
-#   mutate(eid = row_number())
-# 
-# msall3_df2 <- left_join(msall3_df, edges3, by = c("from" = "from", "to" = "to"))
-# 
-# p3 <- ggplot(data = msall3_df2) + 
-#   geom_tile(aes(x = ms, y = -eid, fill = ms)) + 
-#   geom_vline(xintercept = c(0.5,106.5)) + 
-#   scale_fill_continuous(low = "red", high = "blue") +
-#   ThemeNoNet + 
-#   theme(legend.position = 'none', axis.ticks.y = element_blank(),
-#         axis.text.y = element_blank(), axis.title = element_blank()) #+
-#   #labs(x = "Microstep", y = "Edges (most frequent at top)")
-
-grid.arrange(p1, p2, nrow = 2)
-@
-
-We also wanted to view the complete microstep process from the first wave, which we condition on in the model, to the second wave.
-The number of steps taken from wave 1 to wave 2 varies.
-In one set of 1,000 simulations from Model 1, the smallest number of steps taken was 58, and the longest was 248, with a mean of 106 and a median of 103.
-In the 1,000 simulations, the standard deviation of the number of microsteps was 22.8.
-In Figure~\ref{fig:allsteps}, we see two simulations of the process from wave 1 to wave 2, with wave 1 shown on the left, and wave 2 shown on the right.
-In each of the three plots, the $y$-axis contains the edges sorted by how often they appear in the networks along the way.
-We can see that some edges are there in the beginning, but disappear and never come back, while others appear a few steps in, only to dispappear again.
-There are also some edges that were observed in wave 2 that don't appear at all in the microstep process in a given simulation.
-
-<<alledgests, fig.height = 4, fig.cap="Visualizing all microsteps taken in 1,000 simulations from the model M1.
-The occurrence percent is split up into groups to correspond with its distribution: only about 10\\% of the edges appear more than 10\\% of the time in the 1,000 simulations, while about 60\\% appear less than 1\\% of the time.
-The first wave network is shown at microstep 0, and the second wave of the network is shown as the last microstep for comparison.
-We see that it is rare for a microstep process to last longer than 150 steps, and also that the edges that appear past the 150th step tend to be in either the first wave or the second wave.">>=
-# get a plot of all microsteps from w1 to w2 by edge
-alledges <- expand.grid(from = paste0("V", 1:16), to = paste0("V", 1:16))
-alledges$eid <- 1:nrow(alledges)
-
-sfmsall %>% left_join(alledges) -> sfmsall2
-
-sfmsall2 %>% group_by(rep) %>% mutate(isWave2 = (ms == max(ms))) -> sfmsall2
-
-sfmsall2 %>% group_by(from, to, eid) %>% 
-  summarize(total =n()) %>% 
-  arrange(desc(total)) %>% 
-  ungroup %>% 
-  mutate(plotorder = row_number()) -> sfmsallorder
-
-sfmsall2 %>% group_by(ms, eid) %>% 
-  summarise(count = n()) %>%
-  mutate(alpha = count/1000) %>% 
-  arrange(ms, eid) -> sfmsall3
-
-left_join(sfmsall3, sfmsallorder, by = "eid") -> sfmsall4
-
-#add true wave 2 to plot
-wave2 <- na.omit(actual2)
-wave2$from <- paste0("V", wave2$from)
-wave2$to <- paste0("V", wave2$to)
-wave2 <- left_join(wave2, alledges)
-wave2 <- left_join(wave2, sfmsallorder)
-
-sfmsall5 <- sfmsall4 %>% ungroup %>% add_row(ms = max(sfmsall4$ms) + 1, 
-        eid = wave2$eid, 
-        count = NA, 
-        alpha = 1, 
-        from = wave2$from,
-        to = wave2$to,
-        total = wave2$total, 
-        plotorder = wave2$plotorder) 
-sfmsall5$alphaFctr <- cut(sfmsall5$alpha, c(0,.01, .05, .10, .25, .5, .75, 1, 1.1))
-ggplot(data = sfmsall5) + 
-  geom_tile(aes(x = ms, y = reorder(eid,-plotorder), fill = alphaFctr)) +
-  #scale_fill_gradient(low = 'grey90', high = '#0868ac', name = "occurrence %") + 
-  scale_fill_brewer(palette = "Blues", name = "Occurrence %", labels = c("(0-1]%", "(1-5]%", "(5-10]%","(10-25]%", "(25-50]%", "(50-75]%", "(75-100]%")) + 
-  labs(x = "Microstep No.", y = "Edges (ordered by total # occurrences)") + 
-  ThemeNoNet + 
-  theme(axis.ticks.y = element_blank(), axis.text.y = element_blank(), panel.grid = element_blank())
-@
-
-\st{We also combine 1,000 simulations from model M1 into a visualization like the one shown in Figure~\ref{fig:allsteps}.
-We first assign each possible edge an edge ID number so that we can keep track of it throughout all the microsteps and all the simlations.
-Then, we count up the total number of times each edge appears in the microstep process in each of the 1,000 simulations for use as an ordering variable later.
-We also count up the number of times an edge occurs in each microstep number in the 1,000 simulations.
-Since the number of microsteps in the process varies, the number of times an edge occurs decreases as the microstep number increases.
-Next, we compute a proportion, which we call the occurrence percentage, which is the number of times the edge occurred in a microstep divided by 1,000.
-Finally, we visualize all this information together in Figure ~\ref{fig:alledgests}.
-In this plot, all possible edges are shown, and we see that every one of the $16 \times 15 = 240$ possible edges in the network occurs at some point in the simulation process.
-We also see, however, that the process struggles to focus in on the edges in the second wave of the data.
-Ideally, we would like to see more occurrences of the second wave of data generally.
-About half of the edges in wave two are in the bottom half when ordered by number of occurrences, meaning they aren't appearing as much as they would if the model was truly excellent at capturing the mechanisms of tie change in the network.
-In addition, there is not a distinct pattern or structure that we can see in this particular edge visualization.
-The edges that the model starts with appear more often in the first 50 or so microsteps than any other set of edges.
-This makes sense, but it would be better if the model explored the space more than it currently does.
-It could be that the simple model M1 isn't complicated enough for the friendship dynamics in the data, or it could be that the chains in the MCMCs "get stuck" when exploring the parameter space.}
-
-% \st{The algorithms here are many!
-% \begin{enumerate}
-% %\item First, the microsteps / CTMC between each time point (gifs, other animations)
-% \item Phases of RSiena: phase 1 = initial parameter estimate
-% \item phase 2 = iterative update of parameters - trace plots
-% \item phase 3 = simulations for convergence diagnostics
-% \end{enumerate}
-%  }
-
-\hh{The visualizations should address most elements of structure described before, i.e.
-follow along in the setup (for descriptive and teaching purposes), the fitting (understanding the model and interpreting the results) and then diagnostics (how well does the model fit the data - where are the most differences to the actual data/what are the sensitive parameters?) }
-
-% \hh{Some of the visualizations we talked about
-% \begin{enumerate}
-% \item visualization of the MCMC process:  movie of individual actors' step-by-step decisions; could be done with the ndtv package, as long as we can get individual steps out of RSiena
-% \item global picture: 1000 simulated networks + alpha blending should give a pretty good idea of the most important edges
-% %\item global structure + principal components (tensor pca)
-% \end{enumerate}}
-
-\section{Discussion} \label{sec:discussion}
-
-We have used novel visualization methods in order to better understand the family of models known as stochastic actor-oriented model for social network.
-By looking at the underlying algorithms, exploring collections of these models, and viewing the model in the data space, we have been able to gain knowledge and appreciation for these complicated models and everything that goes into them.
-
-We have only just begun to scratch the surface of these complicated and multi-layered models for social networks.
-The \texttt{RSiena} software is incredibly powerful, and can fit a whole slew of much more flexible stochastic actor-oriented models than we have examined here.
-If a researcher thinks the network structure or an actor covariate effects the rate of change of the networ, there is a way to incorporate that belief into the rate function of the SAOM.
-More than one actor-level covariate can be included in the model, and way more than three parameters can be included in the objective function itself.
-In addition, \texttt{RSiena} allows the user to tell it which parameters lead to tie creation, and which parameters lead to tie endowment, or dissolution.
-We have used ``evaluation" parameters, which assume that creation and endowment are equal \citep{RSienamanual}.
-Finally, SAOMs and \texttt{RSiena} are able to also model behavior change of the actors in the network, which again is a capability we did not explore here.
 

### 2017-10-03 code/networkdistance.R
making the "resid" plot

@@ -1,0 +1,103 @@
+# distances between wave 1, wave2, and sims
+library(RSiena)
+library(tidyverse)
+load('data/ansnullpaper.rda')
+
+ansnullchainsw1w2 %>% head
+summary(ansnullchainsw1w2$rep)
+
+amfull <- matrix(0, 16, 16)
+
+to_adjmat <- function(empty, dat){
+  for(i in 1:nrow(dat)){
+    row <- dat$from[i]
+    coln <- dat$to[i]
+    empty[row, coln] <- dat$val[i]
+  }
+  return(empty)
+}
+
+ansnullchainsw1w2 %>% unique() %>% 
+  mutate(from = parse_number(from), to = parse_number(to)) %>% 
+  arrange(rep, from, to) %>% nest(-rep) %>% 
+  mutate(mats = map(data, to_adjmat, empty = amfull)) -> ansnullchainsw1w2nest
+
+diff1 <- function(am, wave1 = w1){
+ wave1 - am
+}
+diff2 <- function(am, wave2 = w2){
+  wave2 - am 
+}
+
+w1 <- as.matrix(read.table("data/s50_data/s50-network1.dat"))[20:35, 20:35]
+colnames(w1) <- paste0("V", 1:16)
+w2 <- as.matrix(read.table("data/s50_data/s50-network2.dat"))[20:35, 20:35]
+colnames(w2) <- paste0("V", 1:16)
+ansnullchainsw1w2nest %>% mutate(diff1 = map(mats, diff1), diff2 = map(mats, diff2)) -> ansnullchainsw1w2nest
+ansnullchainsw1w2nest %>% mutate(d1 = map_dbl(diff1, function(x){sum(abs(x))}), 
+                                 d2 = map_dbl(diff2, function(x){sum(abs(x))}), 
+                                 miss1 = map_dbl(diff1, function(x){length(which(x == 1))}), 
+                                 miss2 = map_dbl(diff2, function(x){length(which(x == 1))}),
+                                 addl1 = map_dbl(diff1, function(x){length(which(x == -1))}), 
+                                 addl2 = map_dbl(diff2, function(x){length(which(x == -1))}), 
+                                 eq1 = map_dbl(diff1, function(x){length(which(x == 0))}),
[…]

### 2017-10-21 paper-child.Rnw
works good!
fix error in clinton/no clinton figure

@@
+\newcommand{\st}[1]{{\color{orange} #1}}
+\newcommand{\hh}[1]{{\color{magenta} #1}}
+
 <<setup, echo = FALSE, message = FALSE, warning = FALSE>>=
@@
                            "Hillary Clinton", "Jim Webb", "John McCain", "Joe Biden",
-                           "Lindsey Graham", "Marco Rubio", "Mitch McConnel", 
+                           "Lindsey Graham", "Marco Rubio", "Mitch McConnell", 
                            "Rand Paul", "Ted Cruz")
@@
 
-clintonSenate %>% filter(!(source %in% c("Roland Burris", "Bernard Sanders", "Frank R. Lautenberg", "Judd A. Gregg", "Mary L. Landrieu") & is.na(target))) %>%
+clintonSenate %>% filter(!(source %in% c("Roland Burris", "Bernard Sanders", "Frank R. Lautenberg", "Mary L. Landrieu") & is.na(target))) %>%
 ggplot() + 
@@
 
-\section{Discussion} \label{sec:discussion}
+\section{Discussion}\label{sec:discussion}
 

### 2017-10-26 paper-child.Rnw
fix weird dplyr error - had something to do with cache = TRUE in knitr chunk

@@
 <<makinggif, fig.cap='A selection of images in the microstep animation.
-The selected edges and nodes are emphasized by changing the size and the color, then when edges are removed, they fade out, shrinking in size, while the nodes change color and shrink to blend in with the other nodes.', cache=TRUE, message = FALSE, warning = FALSE, fig.height=2.5>>=
+The selected edges and nodes are emphasized by changing the size and the color, then when edges are removed, they fade out, shrinking in size, while the nodes change color and shrink to blend in with the other nodes.', message = FALSE, warning = FALSE, fig.height=2.5>>=
 # Need old version of dplyr to get this to work:
-if (!require("dplyr050",character.only = TRUE)) {
-      devtools::install_github("sctyner/dplyr050")
-}
-library(dplyr050)
+# if (!require("dplyr050",character.only = TRUE)) {
+#       devtools::install_github("sctyner/dplyr050")
+# }
+# library(dplyr050)
 friend.data.w1 <- as.matrix(read.table("data/s50_data/s50-network1.dat"))
@@
 #myalgorithm <- sienaAlgorithmCreate(projname = 's16_3')
+library(dplyr)
 set.seed(823746)
@@
 ansnullchains %>%
-  filter(period == 1) %>%  #only look at chains from wave 1 to wave 2
-  group_by(rep) %>%
+  dplyr::filter(period == 1) %>%  #only look at chains from wave 1 to wave 2
+  #dplyr::group_by(rep) %>%
   dplyr::select(rep, from = ego, to = alter) %>%
-  mutate(val = as.numeric(!from == to),
+  dplyr::mutate(val = as.numeric(!from == to),
          from = paste0("V", parse_number(from)+1), # make the chains
@@
 wave1friends <- fortify(as.adjmat(fd2.w1))
-ms1 <- listMicrosteps(dat = wave1friends,
-                      microsteps = filter(ansnullchainsw1w2, rep == 1))
+ms1 <- netvizinf::listMicrosteps(dat = wave1friends,
+                      microsteps = dplyr::filter(ansnullchainsw1w2, rep == 1))
 microsteps <- ms1[1:20]
@@
 
-static_pall <- filter(pall, .frame %in% c(15,20,22,24,26:28, 71:73, 75, 79, 81, 84))
+static_pall <- dplyr::filter(pall, .frame %in% c(15,20,22,24,26:28, 71:73, 75, 79, 81, 84))
 
@@
 set.seed(34569234)
-froms <- static_pall %>% select(id = from, microstep, x = from.x, 
+froms <- static_pall %>% dplyr::select(id = from, microstep, x = from.x,
         y = from.y, ms, .frame, size, color) %>% unique
-tos <- static_pall %>% select(id = to, microstep, x = to.x, y = to.y, 
+tos <- static_pall %>% dplyr::select(id = to, microstep, x = to.x, y = to.y,
         ms, .frame) %>% unique
-tos <- left_join(tos, froms)
+tos <- dplyr::left_join(tos, froms)
 nodedata <- rbind(froms, tos) %>% unique
-nodedata <- nodedata %>% replace_na(replace = list(size = 1, 
+nodedata <- nodedata %>% tidyr::replace_na(replace = list(size = 1,
         color = "#333333"))

### 2017-10-30 paper-child.Rnw
add links to gifs and updated paper.

@@
 If there are no changes at a particular microstep, no changes are seen.
-Some frames of the animations are shown in Figure~\ref{fig:makinggif}, and the full movie can be viewed at INSERT LINK TO GIF.
+Some frames of the animations are shown in Figure~\ref{fig:makinggif}, and the full movie can be viewed at \url{https://vimeo.com/240089108}.
 
@@
 These in-between steps are not shown when we look at the network at our discrete observation points, so by viewing this animation we can gain a better understanding of the underlying dynamics of this model.
+The full movie can be viewed at \url{https://vimeo.com/240092677}.
 

### 2018-01-26 ASApaper/jcgs-template.tex
create folder for submission

%&latex
\documentclass[12pt]{article}
\usepackage{amsmath}
\usepackage{graphicx,psfrag,epsf}
\usepackage{enumerate}
\usepackage{natbib}
\usepackage{url} % not crucial - just used below for the URL

%\pdfminorversion=4
% NOTE: To produce blinded version, replace "0" with "1" below.
\newcommand{\blind}{0}

% DON'T change margins - should be 1 inch all around.
\addtolength{\oddsidemargin}{-.5in}%
\addtolength{\evensidemargin}{-.5in}%
\addtolength{\textwidth}{1in}%
\addtolength{\textheight}{1.3in}%
\addtolength{\topmargin}{-.8in}%

\begin{document}

%\bibliographystyle{natbib}

\def\spacingset#1{\renewcommand{\baselinestretch}%
{#1}\small\normalsize} \spacingset{1}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\if0\blind
{
  \title{\bf Title}
  \author{Author 1\thanks{
    The authors gratefully acknowledge \textit{please remember to list all relevant funding sources in the unblinded version}}\hspace{.2cm}\\
    Department of YYY, University of XXX\\
    and \\
    Author 2 \\
    Department of ZZZ, University of WWW}
  \maketitle
} \fi

\if1\blind
{
  \bigskip
  \bigskip
  \bigskip
  \begin{center}
    {\LARGE\bf Title}
\end{center}
  \medskip
} \fi

\bigskip
\begin{abstract}
The text of your abstract.  200 or fewer words.
\end{abstract}

\noindent%
{\it Keywords:}  3 to 6 keywords, that do not appear in the title
\vfill

\newpage
\spacingset{1.45} % DON'T change the spacing!
\section{Introduction}
\label{sec:intro}

Body of paper.  Margins in this document are roughly 0.75 inches all
around, letter size paper.

\begin{figure}
\begin{center}
\includegraphics[width=3in]{fig1.pdf}
\end{center}
\caption{Consistency comparison in fitting surrogate model in the tidal
power example. \label{fig:first}}
\end{figure}

\begin{table}
\caption{D-optimality values for design $X$ under five different scenarios.  \label{tab:tabone}}
\begin{center}
\begin{tabular}{rrrrr}
one & two & three & four & five\\\hline
1.23 & 3.45 & 5.00 & 1.21 & 3.41 \\
1.23 & 3.45 & 5.00 & 1.21 & 3.42 \\
1.23 & 3.45 & 5.00 & 1.21 & 3.43 \\
\end{tabular}
\end{center}
\end{table}

\begin{itemize}
\item Note that figures and tables (such as Figure~\ref{fig:first} and
Table~\ref{tab:tabone}) should appear in the paper, not at the end or
in separate files.
\item In the latex source, near the top of the file the command
\verb+\newcommand{\blind}{1}+ can be used to hide the authors and
acknowledgements, producing the required blinded version.
\item Remember that in the blind version, you should not identify authors
indirectly in the text.  That is, don't say ``In Smith et. al.  (2009) we
showed that ...''.  Instead, say ``Smith et. al. (2009) showed that ...''.
\item These points are only intended to remind you of some requirements.
Please refer to the instructions for authors
at \url{http://amstat.tandfonline.com/action/authorSubmission?journalCode=ubes20&page=instructions#}
\item For more about ASA\ style, please see \url{http://journals.taylorandfrancis.com/amstat/asa-style-guide/}
\item If you have supplementary material (e.g., software, data, technical
proofs), identify them in the section below.  In early stages of the
submission process, you may be unsure what to include as supplementary
material.  Don't worry---this is something that can be worked out at later stages.
\end{itemize}

\section{Methods}
\label{sec:meth}
Don't take any of these section titles seriously.  They're just for
illustration.

\section{Verifications}
\label{sec:verify}
This section will be just long enough to illustrate what a full page of
text looks like, for margins and spacing.

\cite{Campbell02}, \cite{Schubert13}, \cite{Chi81}

The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
{\bf With this spacing we have 30 lines per page.}
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.

The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.

The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.

The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.

The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.

The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.

The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.

The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.

The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.

The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.

The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.

\section{Conclusion}
\label{sec:conc}

\bigskip
\begin{center}
{\large\bf SUPPLEMENTARY MATERIAL}
\end{center}

\begin{description}

\item[Title:] Brief description. (file type)

\item[R-package for  MYNEW routine:] R-package ÒMYNEWÓ containing code to perform the diagnostic methods described in the article. The package also contains all datasets used as examples in the article. (GNU zipped tar file)

\item[HIV data set:] Data set used in the illustration of MYNEW method in Section~ 3.2. (.txt file)

\end{description}

\section{BibTeX}

We hope you've chosen to use BibTeX!\ If you have, please feel free to use the package natbib with any bibliography style you're comfortable with. The .bst file Chicago was used here, and agsm.bst has been included here for your convenience.

\bibliographystyle{Chicago}

\bibliography{Bibliography-MM-MC}
\end{document}

### 2018-01-26 ASApaper/code/00e_small_friends.R

@@ -1,0 +1,170 @@
+# New model with a smaller collection of girls from the friendship data
+#deciding how to subset the girls:
+library(ggplot2)
+library(tidyr)
+library(dplyr)
+
+library(RSiena) # or RSienaTest
+# read in network data
+friend.data.w1 <- as.matrix(read.table("data/s50_data/s50-network1.dat"))
+friend.data.w2 <- as.matrix(read.table("data/s50_data/s50-network2.dat"))
+friend.data.w3 <- as.matrix(read.table("data/s50_data/s50-network3.dat"))
+
+val1 <- data.frame(friend.data.w1) %>% gather(key = x, value = y, V1:V50) %>% dplyr::select(y)
+val2 <- data.frame(friend.data.w2) %>% gather(x,y, V1:V50) %>% dplyr::select(y)
+val3 <- data.frame(friend.data.w3) %>% gather(x,y, V1:V50) %>% dplyr::select(y)
+
+all_girls <- expand.grid(x = 1:50, y= 1:50)
+
+view_rs <- data.frame(all_girls, W1 = val1, W2 = val2, W3 = val3)
+names(view_rs)[3:5] <- c('w1', 'w2','w3')
+
+view_rs %>% gather(wave, value, w1:w3) -> view_rs2
+
+# look at adjacency matrices
+# ggplot(data= view_rs2, aes(x = x, y=y)) + 
+#   geom_tile(aes(fill = as.factor(value))) + scale_fill_manual(values = c('white', 'black')) +
+#   geom_rect(data= NULL, inherit.aes = FALSE, 
+#             aes(xmin = 20, xmax = 35, ymin = 20, ymax = 35, fill =NA), color = 'red') + 
+#   facet_wrap(~wave) + theme_bw() + 
+#   theme(aspect.ratio=1)
+
+#looks like 20:35 are pretty well connected, go with those for now
+fd2.w1 <- friend.data.w1[20:35,20:35]
+fd2.w2 <- friend.data.w2[20:35,20:35]
+fd2.w3 <- friend.data.w3[20:35,20:35]
+# read in covariate data
+drink <- as.matrix(read.table("data/s50_data/s50-alcohol.dat"))
+drink2 <- drink[20:35,]
+
+
+# siena data
+library(RSiena)
+friendData2 <- array( c( fd2.w1, fd2.w2, fd2.w3 ),
+                         dim = c( 16, 16, 3 ) )
+# and next give this the role of the dependent variable:
[…]

### 2018-01-26 ASApaper/code/fitting_models_senate.R

@@ -1,0 +1,26 @@
+# fit senate data for the same M1, M2, M3 as the small friends
+
+library(RSienaTest)
+library(RSiena)
+library(netvizinf)
+
+load("data/congress/senateSiena.RDA")
+
+
+senateM1 <- getEffects(senateSiena)
+senateM2 <- includeEffects(senateM1, jumpXTransTrip, include = TRUE, type = "eval", interaction1 = "bills")
+senateM3 <- includeEffects(senateM1, nbrDist2twice, include = TRUE, type = "eval", interaction1 = "")
+senateM4 <- includeEffects(senateM1, isolateNet, include = TRUE, type = "eval", interaction1 = "")
+senateM5 <- includeEffects(senateM1, sameXRecip, include = TRUE, type = "eval", interaction1 = "party")
+
+
+senateM1ests <- get_effects_dist(dat = senateSiena, struct = senateM1, N = 100)
+senateM2ests <- get_effects_dist(dat = senateSiena, struct = senateM2, N = 100)
+senateM3ests <- get_effects_dist(dat = senateSiena, struct = senateM3, N = 100)
+senateM4ests <- get_effects_dist(dat = senateSiena, struct = senateM4, N = 100)
+senateM5ests <- get_effects_dist(dat = senateSiena, struct = senateM5, N = 100)
+readr::write_csv(senateM1ests, "data/congress/senateM1ests.csv")
+readr::write_csv(senateM2ests, "data/congress/senateM2ests.csv")
+readr::write_csv(senateM3ests, "data/congress/senateM3ests.csv")
+readr::write_csv(senateM4ests, "data/congress/senateM4ests.csv")
+readr::write_csv(senateM5ests, "data/congress/senateM5ests.csv")

### 2018-01-26 ASApaper/code/making_am_gif.R

@@ -1,0 +1,65 @@
+# making adjmat gif
+
+adjmatall <- matrix(0, nrow = 16, ncol = 16)
+rownames(adjmatall) <- paste0("V", 1:16)
+colnames(adjmatall) <- paste0("V", 1:16)
+dat2 <- sfmsall %>% filter(rep == 1)
+adjmatall2 <- NULL
+for(i in 0:55){
+  MS <- dat2[dat2$ms == i,c("from", "to")]
+  myadjmat <- adjmatall
+  for(j in 1:nrow(MS)){
+    myadjmat[MS[j,"from"][[1]],MS[j,"to"][[1]]] <- 1
+  }
+  adjmatall2 <- rbind(adjmatall2, myadjmat)
+}
+
+adjmatall2 <- cbind(adjmatall2, rep(0:55, each = 16))
+adjmatall2 <- data.frame(adjmatall2)
+adjmatall2$from <- rep(paste0("V", 1:16), 56)
+names(adjmatall2)[17] <- "ms"
+adjmatall2 %>% gather(to, val, V1:V16) -> adjmatall3
+
+adjmatall3$to <- factor(adjmatall3$to, levels = order1)
+adjmatall3$from <- factor(adjmatall3$from, levels = rev(order1))
+
+adjmatall3$frame <- adjmatall3$ms
+#plots <- list()
+#for(i in 0:55){
+#plots[[i + 1]] <-
+
+#dat3 <- adjmatall3 %>% filter(ms %in% c(0,2,4,6,8,10,12,43,45,47,49,51,53,55))
+dat3 <- adjmatall3
+dat4 <- dat3 %>% select(-ms) %>% spread(frame, val)
+dat5 <- dat4
+#p <- Sys.time()
+for(i in 4:ncol(dat5)){
+  print(i)
+  for(j in 1:nrow(dat5)){
+    print(j)
+    if(dat4[j,i] != dat4[j,i-1]){
+      if(dat4[j,i] > dat4[j,i-1]){
+        dat5[j,i] <- 2
+      } else { dat5[j,i] <- 3}
+    }
+  }
+}
+#Sys.time() - p
+
+dat6 <- dat5 %>% gather(frame, val, 3:ncol(dat5)) %>% mutate(frame = parse_number(frame))
+
+p <- ggplot() +
+  geom_tile(data = dat6, aes(x = to, y = from, fill = as.factor(val), frame = frame), color = '#969696') +
+  scale_fill_manual(values = c("white", "black", "red", "white")) +
[…]

### 2018-01-26 ASApaper/code/making_gif.R

@@ -1,0 +1,128 @@
+# make gif of a microstep - small friends
+setwd("~/Desktop/Dissertation/SAOM-removing-blindfold/")
+library(tidyverse)
+library(geomnet)
+library(RSiena)
+library(network)
+library(netvizinf)
+library(gganimate)
+library(ggforce)
+library(tweenr)
+friend.data.w1 <- as.matrix(read.table("data/s50_data/s50-network1.dat"))
+friend.data.w2 <- as.matrix(read.table("data/s50_data/s50-network2.dat"))
+friend.data.w3 <- as.matrix(read.table("data/s50_data/s50-network3.dat"))
+drink <- as.matrix(read.table("data/s50_data/s50-alcohol.dat"))
+fd2.w1 <- friend.data.w1[20:35,20:35]
+fd2.w2 <- friend.data.w2[20:35,20:35]
+fd2.w3 <- friend.data.w3[20:35,20:35]
+friendshipData <- array(c(fd2.w1, fd2.w2,fd2.w3), dim = c(16, 16, 3))
+
+set.seed(823746)
+load("data/ansnullpaper.rda")
+ansnullchains <- get_chain_info(ansnull)
+ansnullchains %>%
+  dplyr::filter(period == 1) %>%  #only look at chains from wave 1 to wave 2
+  #dplyr::group_by(rep) %>%
+  dplyr::select(rep, from = ego, to = alter) %>%
+  dplyr::mutate(val = as.numeric(!from == to),
+                from = paste0("V", parse_number(from)+1), # make the chains
+                to = paste0("V", parse_number(to)+1)) -> ansnullchainsw1w2
+colnames(fd2.w1) <- paste0("V", 1:16)
+rownames(fd2.w1) <- paste0("V", 1:16)
+wave1friends <- fortify(as.adjmat(fd2.w1))
+ms1 <- netvizinf::listMicrosteps(dat = wave1friends,
+                                 microsteps = dplyr::filter(ansnullchainsw1w2, rep == 1))
+# all ms
+microsteps_df <- getMicrostepsDF(ms1)
+pte <- pretween_edges(microsteps = ms1)
+ptv <- pretween_vertices(pte = pte, layoutparams = list(n = 16))
+pall <- tween_microsteps(ptv, pte, microsteps_df)
+
+set.seed(34569234)
+froms <- pall %>% dplyr::select(id = from, microstep, x = from.x,
+                                y = from.y, ms, .frame, size, color) %>% unique
[…]

### 2018-01-26 ASApaper/code/making_lineups_gof.R

@@ -1,0 +1,141 @@
+# creating lineups for model goodness of fit. 
+
+load("data/ansnullpaper.rda")
+#ansnull$sims
+
+# for model 1 
+set.seed(9430852)
+M6 <- sample(1000, size = 5, replace = F)
+M12 <- sample(1000, size = 11, replace = F)
+M16 <- sample(1000, size = 15, replace = F)
+M20 <- sample(1000, size = 19, replace = F)
+
+simsM6 <- ansnull$sims[M6]
+simsM12 <- ansnull$sims[M12]
+simsM16 <- ansnull$sims[M16]
+simsM20 <- ansnull$sims[M20]
+
+simsM6[[1]][[1]]$friendship$`1`
+
+# for model2 
+
+#looks like 20:35 are pretty well connected, go with those for now
+fd2.w1 <- friend.data.w1[20:35,20:35]
+fd2.w2 <- friend.data.w2[20:35,20:35]
+fd2.w3 <- friend.data.w3[20:35,20:35]
+# read in covariate data
+drink <- as.matrix(read.table("data/s50_data/s50-alcohol.dat"))
+drink2 <- drink[20:35,]
+# siena data
+library(RSiena)
+friendData2 <- array( c( fd2.w1, fd2.w2, fd2.w3 ),
+                      dim = c( 16, 16, 3 ) )
+# and next give this the role of the dependent variable:
+friend2 <- sienaDependent(friendData2)
+alcohol2 <- varCovar( as.matrix(drink2) )
+# create siena data object
+mysmalldata <- sienaDataCreate( friend2, alcohol2)
+null_model_eff2 <- getEffects(mysmalldata)
+test_model_eff2 <- includeEffects( null_model_eff2, jumpXTransTrip, interaction1 = "alcohol2")
+#test_model_eff2 <- includeEffects( null_model_eff2, sameXTransTrip, interaction1 = "alcohol2")
+myalgorithm2 <- sienaAlgorithmCreate( projname = 's50' , n3 = 1000)
+ests_test <- siena07( myalgorithm2, data = mysmalldata, returnDeps = TRUE, effects = test_model_eff2, batch=TRUE, verbose = FALSE, silent = TRUE)
+#ests_test <- siena07( myalgorithm2, data = mysmalldata, returnDeps = TRUE, effects = test_model_eff2, batch=TRUE, verbose = FALSE)
+
+set.seed(7902487)
[…]

### 2018-01-26 ASApaper/code/modified_time_plot.R

@@ -1,0 +1,119 @@
+# create new "all ms" visualization
+library(tidyverse)
+library(netvizinf)
+library(geomnet)
+library(RSiena)
+smallfriends <- read_csv("data/smallfriends4Geomnet.csv")
+names(smallfriends)[1:2] <- c("from", "to")
+ansnullchains <- get_chain_info(ansnull)
+chainsms <- ansnullchains %>% 
+  group_by(rep, period) %>% 
+  mutate(microstep = row_number(), 
+         prob = exp(logAlterProb),
+         from = paste0("V", as.numeric(as.character(ego))+1),
+         to = paste0("V", as.numeric(as.character(alter))+1)) %>%
+  ungroup()
+wave1friends <- fortify(as.adjmat(fd2.w1))
+chainsms_wave2 <- filter(chainsms, period == 1)
+chainsms_wave3 <- filter(chainsms, period == 2)
+wave2 <- smallfriends %>% filter(Wave == 2) %>% na.omit %>% data.frame()
+wave3 <- smallfriends %>% filter(Wave == 3) %>% na.omit %>% data.frame()
+## make more generic for wave 3. 
+create_microstep_summary <- function(chains, truedat){
+  browser()
+  R <- max(chains$rep)
+  wave <- truedat 
+  names(wave) <- c("from", "to", "drink", "Wave")
+  all_reps <- list()
+  for (i in 1:R){
+    msall <- netvizinf::listMicrosteps(dat = wave,
+                                       microsteps = filter(chainsms, rep == i))
+    msall_df <- plyr::rbind.fill(lapply(X = seq_along(msall), FUN = function(i, x) {
+      x[[i]]$ms <- i - 1
+      return(x[[i]])
+    }, msall))
+    msall_df <- msall_df %>% dplyr::select(from, to, ms)
+   # msall_df <- msall_df %>% add_row(from = paste0("V",wave2$from),
+   #                                  to = paste0("V",wave2$to), 
+   #                                  ms = max(msall_df$ms)+1)
+    msall_df$rep <- i 
+   all_reps[[i]] <- msall_df 
+  }
+  return(all_reps)
+}
+testing <- create_microstep_summary(chains = filter(chainsms_wave2, rep %in% 1:5), truedat = wave2)
+
+testingdf<-plyr::rbind.fill(testing)
+
+testingdf %>% 
+  group_by(from, to, ms) %>% 
[…]

### 2018-01-26 ASApaper/code/testingfonts.R

@@ -1,0 +1,200 @@
+# Try to make font in graphs match the size of the font in paper
+
+# Source: https://gist.github.com/noamross/7576436
+
+library(ggplot2)
+sleepplot = ggplot(data = msleep, aes(x = log(bodywt), y = sleep_total))+geom_point(aes(color=vore))
+sleepplot
+#' We made a beautiful model of this relationship
+slp = lm(sleep_total ~ log(bodywt), data=msleep)
+summary(slp)
+#'Let's put the model on the plot
+sleepplot = sleepplot + geom_abline(intercept=coef(slp)[1], slope=coef(slp)[2])
+sleepplot
+
+#' First make the labels a little more useful.
+sleepplot = sleepplot + labs(x="Log body weight (Kg)", y="Time asleep (hrs/day)")
+sleepplot
+
+#' Now let's fix the legend.
+#' You would think you do this with some sort of "legend" command, but *no*,
+#' what you are looking for is "scale". 
+sleepplot + scale_color_discrete(name="Functional\nfeeding group",
+                                 labels = c("carnivore", "herbivore", "insectivore", "omnivore"))
+
+sleepplot + scale_color_brewer(name="Functional \n feeding group",
+                               labels = c("carnivore", "herbivore", "insectivore", "omnivore"),
+                               type = "qual", palette = 1)
+
+sleepplot2 = ggplot(data = msleep, aes(x = log(bodywt), y = sleep_total)) + 
+  geom_point(aes(shape=vore), size=3) + #' This time we will vary the feeding groups by shapes instead of colors
+  geom_abline(intercept=coef(slp)[1], slope=coef(slp)[2])
+sleepplot2
+#' Now to fix the labels and legend again:
+sleepplot2 = sleepplot2 + labs(x="Log body weight (Kg)", y="Time asleep (hrs/day)") +
+  #' we will use scale_shape_discrete instead of scale_color_discrete
+  scale_shape_discrete(name="Functional\nfeeding group",
+                       labels = c("carnivore", "herbivore", "insectivore", "omnivore"))
+sleepplot2
+#' Now, let's work on how the plot looks overall.
+#' 
+#' ggplot uses "themes" to adjust plot appearence without changes the actual presentation of the data.
+sleepplot2 + theme_bw(base_size=12, base_family = "Helvetica")
+#' `theme_bw()` will get rid of the background, and gives you options to 
[…]

### 2018-04-02 paper-child.Rnw
committing for safekeeping

@@
 <<dab, echo=FALSE, fig.width=2, fig.height=2, message = FALSE, warning = FALSE, fig.align='center', out.width='.49\\textwidth'>>=
-dade <- data.frame(from = c('i', 'i', 'h', 'j'), to = c('h', 'j', 'k', 'k'))
+dade <- data.frame(from = c('i', 'i', 'h', 'j', 'i'), to = c('h', 'j', 'k', 'k','k'))
 dadn <- data.frame(id = letters[8:11], group = c(1,1,1,1))
@@
 ggplot(data = dad, aes(from_id = from, to_id = to)) +
-  geom_net(aes(shape = as.factor(group)), directed = T, labelon = T, labelcolour='grey80',vjust = 0.5, hjust =0.5, arrowgap = .15, colour = 'black', size=10, fontsize = 5) +
+  geom_net(directed = T, labelon = T,
+           ecolour = c("black", "black", "black","red", "black" ,"black" ),
+           labelcolour='grey80',vjust = 0.5, hjust =0.5, arrowgap = .15, size=10, fontsize = 5) +
   expand_limits(x=c(-0.1,1.1), y=c(-0.1,1.1)) +
@@
 combinedavgs %>% filter(cat == "Model1") %>% mutate(logweight = log(weight)) -> t1
-colors <- tweenr::tween_color(data = c("#969696", "#d73027"), n = 26, ease = 'linear')
+colors <- tweenr::tween_color(data = c("#969696", "#d73027"), n = table(t1$cat) %>% as.numeric, ease = 'linear')
 t1 %>% arrange(logweight) -> t1
 t1$color <- NA
-t1$color[1:26] <- colors[[1]]
+t1$color <- colors[[1]]
 
@@
 #drink <- as.matrix(read.table("s50_data/s50-alcohol.dat"))
-friendshipData <- array(c(fd2.w1, fd2.w2, fd2.w3), dim = c(16, 16, 3))
-friendship <- sienaDependent(friendshipData)
-alcohol <- varCovar(drink[20:35,])
-mydata <- sienaDataCreate(friendship, alcohol)
-M1eff <- getEffects(mydata)
-M2eff <- includeEffects(M1eff, jumpXTransTrip, interaction1 = "alcohol")
-M3eff <- includeEffects(M1eff, nbrDist2twice)
-data(M1ests_bigfriends)
-M1parms <- (means %>% filter(Model == "M1"))$means
-M2parms <- (means %>% filter(Model == "M2"))$means
-M3parms <- (means %>% filter(Model == "M3"))$means
+# friendshipData <- array(c(fd2.w1, fd2.w2, fd2.w3), dim = c(16, 16, 3))
+# friendship <- sienaDependent(friendshipData)
+# alcohol <- varCovar(drink[20:35,])
+# mydata <- sienaDataCreate(friendship, alcohol)
+# M1eff <- getEffects(mydata)
+# M2eff <- includeEffects(M1eff, jumpXTransTrip, interaction1 = "alcohol")
+# M3eff <- includeEffects(M1eff, nbrDist2twice)
+# data(M1ests_bigfriends)
+# M1parms <- (means %>% filter(Model == "M1"))$means
+# M2parms <- (means %>% filter(Model == "M2"))$means
+# M3parms <- (means %>% filter(Model == "M3"))$means
 # M1sims1000 <- saom_simulate(dat = mydata,
@@
 # M3simsdf <- sims_to_df(M3sims1000)
-M1avg <- M1simsdf %>% filter(!is.na(to) & wave == 1) %>%
-  group_by(from, to) %>%
-  summarise(count = n()) %>%
-  mutate(weight = ifelse(from == to, 0, count))
-M1avgW2 <- M1simsdf %>% filter(!is.na(to) & wave == 2) %>%
-  group_by(from, to) %>%
-  summarise(count = n()) %>%
-  mutate(weight = ifelse(from == to, 0, count))
-M1avg2 <- M1simsdf %>% filter(!is.na(to)) %>%
- group_by(from, to, wave) %>%
- summarise(count = n()) %>%
- mutate(weight = ifelse(from == to, 0, count))
-M2avg <- M2simsdf %>% filter(!is.na(to) & wave == 1) %>%
-  group_by(from, to) %>%
-  summarise(count = n()) %>%
-  mutate(weight = ifelse(from == to, 0, count))
-M2avgW2 <- M2simsdf %>% filter(!is.na(to) & wave == 2) %>%
-  group_by(from, to) %>%
-  summarise(count = n()) %>%
-  mutate(weight = ifelse(from == to, 0, count))
-M2avg2 <- M2simsdf %>% filter(!is.na(to)) %>%
- group_by(from, to, wave) %>%
- summarise(count = n()) %>%
- mutate(weight = ifelse(from == to, 0, count))
-M3avg <- M3simsdf %>% filter(!is.na(to) & wave == 1) %>%
-  group_by(from, to) %>%
-  summarise(count = n()) %>%
-  mutate(weight = ifelse(from == to, 0, count))
-M3avgW2 <- M3simsdf %>% filter(!is.na(to) & wave == 2) %>%
-  group_by(from, to) %>%
-  summarise(count = n()) %>%
-  mutate(weight = ifelse(from == to, 0, count))
-M3avg2 <- M3simsdf %>% filter(!is.na(to)) %>%
- group_by(from, to, wave) %>%
- summarise(count = n()) %>%
- mutate(weight = ifelse(from == to, 0, count))
+# M1avg <- M1simsdf %>% filter(!is.na(to) & wave == 1) %>%
+#   group_by(from, to) %>%
+#   summarise(count = n()) %>%
+#   mutate(weight = ifelse(from == to, 0, count))
+# M1avgW2 <- M1simsdf %>% filter(!is.na(to) & wave == 2) %>%
+#   group_by(from, to) %>%
+#   summarise(count = n()) %>%
+#   mutate(weight = ifelse(from == to, 0, count))
+# M1avg2 <- M1simsdf %>% filter(!is.na(to)) %>%
+#  group_by(from, to, wave) %>%
+#  summarise(count = n()) %>%
+#  mutate(weight = ifelse(from == to, 0, count))
+# M2avg <- M2simsdf %>% filter(!is.na(to) & wave == 1) %>%
+#   group_by(from, to) %>%
+#   summarise(count = n()) %>%
+#   mutate(weight = ifelse(from == to, 0, count))
+# M2avgW2 <- M2simsdf %>% filter(!is.na(to) & wave == 2) %>%
+#   group_by(from, to) %>%
+#   summarise(count = n()) %>%
+#   mutate(weight = ifelse(from == to, 0, count))
+# M2avg2 <- M2simsdf %>% filter(!is.na(to)) %>%
+#  group_by(from, to, wave) %>%
+#  summarise(count = n()) %>%
+#  mutate(weight = ifelse(from == to, 0, count))
+# M3avg <- M3simsdf %>% filter(!is.na(to) & wave == 1) %>%
+#   group_by(from, to) %>%
+#   summarise(count = n()) %>%
+#   mutate(weight = ifelse(from == to, 0, count))
+# M3avgW2 <- M3simsdf %>% filter(!is.na(to) & wave == 2) %>%
+#   group_by(from, to) %>%
+#   summarise(count = n()) %>%
+#   mutate(weight = ifelse(from == to, 0, count))
+# M3avg2 <- M3simsdf %>% filter(!is.na(to)) %>%
+#  group_by(from, to, wave) %>%
+#  summarise(count = n()) %>%
+#  mutate(weight = ifelse(from == to, 0, count))
 

### 2018-04-03 paper-child.Rnw
edits

@@
 %\hh{XXX - talk a bit about social networks first, why we want to model them, and that usually the models are not dynamic.}
-\st{%XXX - okay, here is what i have done for that:
-Social networks have been studied for decades, beginning with a few foundational works, including the 1967 study, "The Small World Problem" by Stanley Milgram \citep{goldenberg09}.
-Examples of social networks include collaboration networks between academic researchers, friendship networks in a school or university, and trade networks between nations.
+%XXX - okay, here is what i have done for that:
+Social networks, such as collaboration networks between academic researchers, friendship networks in a school or university, and trade networks between nations have been studied for decades.  
+%beginning with a few foundational works, including the 1967 study, "The Small World Problem" by Stanley Milgram \citep{goldenberg09}. Examples of . 
 In recent years, the study of social networks has grown in popularity due to an increase in the availability and access to social network data.
-There are many kinds of social networks, but there are not as many statistical models for social network data.} Some network models that have been applied to social networks include the exponential random graph model and latent space models.
+%There are many kinds of social networks, but there are not as many statistical models for social network data.
+Some network models that have been applied to social networks include the exponential random graph model (ERGM) and latent space models.
 These models, however, are only for single instance networks.
 If we only have one network observation, or only care about one state of a complex network in time, like a snapshot of the World Wide Web, using the well-established models for single network observating is not a problem.
-If, on the other hand, we have many observations of social network over time, these models may not be appropriate because they do not explicitly allow for the network to change as time passes.
-When studying a network over time, referred to as a \textit{dynamic network}, we need a model that can take the time aspect of the network into account.
-Models for dynamic social networks have a great deal of modelling potential because of how realistic their structure can be.
-A social network does not form spontaneously: it evolves over time.
+If, on the other hand, we have many observations of social network over time, these models may not be appropriate because they are not able to model network change as time passes.
+When studying \textit{dynamic networks}, networks observed at many time points, we need a model that allows for variation in the network over time.
+Models for dynamic social networks have a great deal of potential because of how realistic they can be.
+A social network does not form spontaneously: it \textit{evolves} over time.
 Ties are formed and dissolved, and new actors join the social structure.
-Modeling the underlying mechanisms that create network changes over time is very complex but also provides potential to uncover hidden truths.
+Modeling the underlying mechanisms that create network changes in time is very complex but also provides potential to uncover hidden truths.
 
@@
 These models are fundamentally different from other social network models because they allow us to incorporate network \textit{and} actor statistics, where other models only rely on the network statistics, to model the changes in the network.
-Allowing the actor-level statistics to directly effect the structure of the network leads to a more practical and relevant approach to model change in a social network.
-In the ``real" world we expect people with common interests to be more likely to form relationships, and SAOMs allow us to incorporate this intuition in the modeling process.
-
-\st{
+Allowing the actor-level statistics to directly effect the structure of the network leads to a more practical and relevant approach to model change in a social network: we expect people with common interests to be more likely to form relationships, and SAOMs allow us to incorporate this intuition in the modeling process.
+
 %\par In order to estimate the parameters of  SAOMs the software SIENA, and its \texttt{R} implementation \texttt{RSiena}, was developed \cite{RSiena}.
@@
 Unlike other network models, SAOMs are not very well understood.
-They are relatively new, especially compared to the classic exponential random graph models, and they are not very tractable analytically.
-Likelihood functions quickly very complex objects to analyze due to the dependency structure inherent in the data.
-Therefore, computationally more tractable solutions are used to fit estimators, and in particular, % Traditional modeling steps, like determining the likelihood function, are very complex.
+They are relatively new, especially compared to the classic ERGMs, and they are not very tractable analytically.
+Likelihood functions quickly become very complex objects to analyze due to the dependency structure inherent in the data.
+Therefore, computationally more tractable solutions are used to fit models, and in particular, % Traditional modeling steps, like determining the likelihood function, are very complex.
 Because of this,
-SAOMs are often fit to data using a series of Markov Chain Monte Carlo (MCMC) phases for finding method of moments estimators.
+SAOMs are often fit to data using a series of Markov Chain Monte Carlo (MCMC) phases for finding method of moments estimates of model parameters.
 In order to estimate the parameters of SAOMs, we use the software SIENA, and its \texttt{R} implementation \texttt{RSiena}, which was developed by \citet{RSiena}.
-This software marks a huge contribution to the field of social network analysis, but the many moving pieces involved in parameter estimation are largely ``behind the scenes" and hidden from the software user.} In this paper, in order to better understand the model-fitting process, we attempt to bring SAOM fits and the fitting process out of their black boxes, by combining the principals of network visualization with those of model visualization as discussed in \citet{modelvis}.
+This software marks a huge contribution to the field of social network analysis, but the many moving pieces involved in parameter estimation are largely ``behind the scenes" and hidden from the software user.
+In this paper, in order to better understand the model-fitting process, we attempt to bring SAOM fits and the fitting process out of their black boxes.
+%, by combining the principals of network visualization with those of model visualization as discussed in \citet{modelvis}.
 %Here, we explore RSiena and SAOMs with innovative network visualizations, both static and dynamic.

### 2018-05-09 paper-child.Rnw
going to submit this week. I'm determined :muscle:

@@
 
-\par In this paper, we concentrate on one type of model for dynamic social networks: the stochastic actor-oriented models (SAOMs), introduced by \citet{saompaper}.
+\par We concentrate on one type of model for dynamic social networks: the stochastic actor-oriented models (SAOMs), introduced by \citet{saompaper}.
 These models are fundamentally different from other social network models because they allow us to incorporate network \textit{and} actor statistics, where other models only rely on the network statistics, to model the changes in the network.
@@
 This software marks a huge contribution to the field of social network analysis, but the many moving pieces involved in parameter estimation are largely ``behind the scenes" and hidden from the software user.
-In this paper, in order to better understand the model-fitting process, we attempt to bring SAOM fits and the fitting process out of their black boxes.
+In order to better understand the model-fitting process, we attempt to bring SAOM fits and the fitting process out into the light.
 %, by combining the principals of network visualization with those of model visualization as discussed in \citet{modelvis}.
 %Here, we explore RSiena and SAOMs with innovative network visualizations, both static and dynamic.
-By bringing some light to the underlying methods and structures that are behind the scenes when fitting SAOMs to network data, we aim to help researchers working with the models better understand the implications and analyses of these models.
-
-\par We get into the SAOM black box by using model visualization (see \citet{modelvis}) to display the model in the data space, view collections of models instead of single models, and visually explore the process of fitting the SAOMs as opposed to looking at only the final output.
-Stochastic actor-oriented models are a prime example of a set of models that can benefit greatly from the application of model visualization.
+By visualizing the underlying methods and structures that are behind the scenes when fitting SAOMs to network data, we aim to help researchers working with the models better understand the implications and analyses of these models.
+
+\par To get behind the scenes of the SAOM fitting process, we use model visualization techniques (see \citet{modelvis}) to display the model in the data space, view collections of models instead of single models, and visually explore the process of fitting the SAOMs as opposed to looking at only the final output.
+Stochastic actor-oriented models are an excellent example of a family of models that can benefit greatly from the application of model visualization.
 For instance, the models themselves include a continuous-time Markov chain (CTMC) that is completely hidden from the analyst in the model fitting process.
@@
 These statistics are often somewhat, if not highly, correlated, which causes high correlation between the associated parameters in a SAOM.
-By visualizing collections of SAOMs, we gain a better understanding of these correlations and find ways to deal with them and rectify their effects in the model.
+By visualizing collections of SAOMs, we gain a better understanding of these correlations and can find ways to account for their effects in the model.
 In addition, the estimation of the parameters in a SAOM relies on a Robbins-Monro algorithm, and the convergence checks for these estimates rely on simulation from the fitted model.
@@
 
+We provide a brief introduction to the structure of network data and common network visualization methods.
+
 \subsection{Introduction to Network Structures}
 
-Network data is of frequent interest to researchers in a wide array of fields.
-There are technological networks, like power grids or the internet, information networks, such as citation networks or the World Wide Web, biological networks, like neural networks, and social networks, just to name a few \citep{newman}.
-Each of these examples have one thing in common: their data \textit{structure}.
-There are always units of observation: the power stations, websites, neurons, and people, which we refer to throughout this paper as \textit{nodes} or \textit{actors}.
-There are also always connections of some kind between those units: the power lines, hyperlinks, electrical signals, and relationships, which we will call \textit{edges} or \textit{ties}.
-Networks might change over time, like when new websites and hyperlinks are added on the World Wide Web, or when there are new people and relationships in a friendship network.
-The nodes and edges themselves can also have inherent variables of interest, e.g.\ the institution of authors in a co-authorship network, or the number of times two authors have collaborated.
-
-The multiple layers of network data structures pose unique problems to network analysts.
-Some questions that network researchers may aim to answer are: How does the strength of a tie between two nodes affect the overall structure of the network?
-Do node-level differences affect the formation or dissolution of edges?
-Which views of the data are most informative for communicating significant effects and other results of statistical analyses of the network of interest?
-These are, of course, just a few broad questions, and we focus here on the latter, which we aim to answer through visual exploration of network data and models.
+Network data is of frequent interest to researchers in a wide array of fields, such as biologists studying neural networks and sociologists studying social networks.% There are technological networks, like power grids or the internet, information networks, such as citation networks or the World Wide Web, biological networks, like neural networks, and social networks, just to name a few \citep{newman}.
+% Each of these examples have one thing in common:
+Network data across fields have a similar data structure. There are always units of observation, such as neurons %the power stations, websites, neurons, 
+and people, which we refer to throughout this paper as \textit{nodes} or \textit{actors}.
+There are also always connections of some kind between those units, such as electrical signals and relationships, which we will call \textit{edges} or \textit{ties}.
+Networks often change over time, such as when there are relationships formed in a friendship network.
+The nodes and edges themselves can also have inherent variables of interest, e.g.\ the age, gender, and political affiliation of people in a friendship network and the length of a friendship.
+%the institution of authors in a co-authorship network, or the number of times two authors have collaborated.
+
+The multiple layers of network data structures pose unique problems to analysts.
+Some questions that researchers may aim to answer are: How does the strength of a tie between two nodes affect the overall structure of the network?
+Do differences in node variables affect the formation or dissolution of edges?
+Which views of the data are most informative for communicating significant model effects?
+What about other results of statistical analyses of the network?
+%These are, of course, just a few broad questions, and we focus here on the latter, which
+We aim to answer some of these questions through visual exploration of network data and models.
 
@@
 Visualizing network data is uniquely difficult because of the structure of the data itself.
-Most, if not all, data visualizations rely on well-defined axes inherited from the data.
-If variables are numerical, histograms, scatterplots, or time series plots are straightforward to construct.
+Most data visualizations rely on well-defined axes inherited from the data, such as Cartesian coordinates or spatial locations.
+%If variables are numerical, histograms, scatterplots, or time series plots are straightforward to construct.
 If the variables are categorical, bar charts and mosaic plots are available to to the researcher.
@@
 Network data, however, are much less cut-and-dried.
-
-There are two primary methods used to visualize networks: node-link diagrams and adjacency matrix visualization \citep{knuth2000, adjmatpro}.
-As a toy example, let us assume that we have five nodes, $\{1,2,3,4,5\}$, connected by five directed edges:  $\{2 \to 4, 3 \to 4, 1 \to 5, 3 \to 5, 5 \to 4\}$ We use this toy data set to demonstrate the two visualization methods in Figure~\ref{fig:toyplots}.
 
@@
 @
-
-The first method, the node-link diagram, represents nodes with points in 2D Euclidean space and then represents edges by connecting the points with lines when there is an edge between the two nodes.
-These lines can also have arrows on them indicating the direction of the edge for directed networks.
-But because there is typically no natural placement of the points unless they have important spatial locations, a random placement of the points is used, then adjusted via a layout algorithm, of which there are many \citep{drawingalgs}.
-
 <<toyplots, fig.show='hold', fig.cap="On the left, a node-link diagram of our directed toy network, with nodes placed using the Kamada-Kawai algorithm.
-On the right, the adjacency matrix visualization for that same network.", out.width='.49\\textwidth'>>=
+On the right, the adjacency matrix visualization for that same network.", out.width='.49\\textwidth', fig.height=4>>=
 toynetdat <- fortify(as.edgedf(toynet$edges), toynet$nodes)
@@
 
+There are two primary methods used to visualize networks: node-link diagrams and adjacency matrix visualization \citep{knuth2000, adjmatpro}.
+As a toy example, let us assume that we have five nodes, $\{1,2,3,4,5\}$, connected by five directed edges:  $\{2 \to 4, 3 \to 4, 1 \to 5, 3 \to 5, 5 \to 4\}$ We use this toy data set to demonstrate the two visualization methods in Figure~\ref{fig:toyplots}.
+
+The first method, the node-link diagram, represents nodes with points in two dimensions and then represents edges by connecting the points with lines when there is an edge between the two nodes.
+These lines can also have arrows on them indicating the direction of the edge for directed networks.
+But because there is typically no natural placement of the points unless they have inherent spatial locations, a random placement of the points is used, then adjusted via a layout algorithm, of which there are many \citep{drawingalgs}.
+
 Some commonly used layout algorithms, such as  the Kamada-Kawai layout \citep{kamadakawai} and the Fruchterman-Reingold layout \citep{fruchterman-reingold}, are designed to mimic physical systems, drawing the graphs based on the ``forces" connecting them.
@@
 
-\st{The second primary method for network visualization uses the adjacency matrix of the network.
-The adjaceny matrix of a network, $\mathbf{A}$, describes the edges of a network in matrix form.
-An entry $A_{ij}$ of $\mathbf{A}$, for two nodes $i\neq j$ in the network is defined as
+The second primary method for network visualization uses the adjacency matrix of the network.
+The adjaceny matrix of a network, $\mathbf{A}$ is defined as follows: an entry $A_{ij}$ of $\mathbf{A}$, for two nodes $i\neq j$ in the network is defined as
 \[
@@
 \]
-Note here that our edge variables are binary: we only consider the presence or absence of an edge.
-If the network has weighted edges, for example an email network where edge weights represent the number of emails sent from one person to another, the entries in the adjacency matrix are the edge weights instead of zeroes and ones.
-In an undirected network, $A_{ij} = A_{ji}$, but in a directed graph this is only true if there is an edge from $i$ to $j$ and from $j$ to $i$.
-Thus, $\mathbf{A}$ is always symmetric for undirected networks, and is symmetric for directed networks only if every edge between two nodes is reciprocated.}
+We only consider the presence or absence of an edge, but if the network has weighted edges, for example an email network where edge weights represent the number of emails sent from one person to another, the entries in the adjacency matrix can be the edge weights instead of zeroes and ones.
+In an undirected network, $A_{ij} = A_{ji}$ for all $i \neq j$, but in a directed graph this is only true if there is an edge from $i$ to $j$ and from $j$ to $i$.
+Thus, $\mathbf{A}$ is always symmetric for undirected networks, and is symmetric for directed networks if and only if every edge between two nodes is reciprocated.
 %\hh{XXX what is an adjacency matrix? - explain at this point in mathematical terms} \st{XXX see above XXX}

### 2018-05-10 paper-child.Rnw
to section 4

@@
                                             family="Times New Roman"),
-                  axis.text.x.top = element_text(size = 12,
+                  axis.text.x = element_text(size = 12,
                                             face = 'plain',
@@
                                             family="Times New Roman"),
-                  axis.text.x.bottom = element_text(size = 12,
+                  # axis.text.x.top = element_text(size = 12,
+                  #                           face = 'plain',
+                  #                           angle = 0,
+                  #                           family="Times New Roman"),
+                  # axis.text.x.bottom = element_text(size = 12,
+                  #                           face = 'plain',
+                  #                           angle = 0,
+                  #                           family="Times New Roman"),
+                  axis.text.y = element_text(size = 12,
                                             face = 'plain',
@@
                                             family="Times New Roman"),
-                  axis.text.y.left = element_text(size = 12,
-                                            face = 'plain',
-                                            angle = 0,
-                                            family="Times New Roman"),
-                  axis.text.y.right = element_text(size = 12,
-                                            face = 'plain',
-                                            angle = 0,
-                                            family="Times New Roman"),
+                  # axis.text.y.left = element_text(size = 12,
+                  #                           face = 'plain',
+                  #                           angle = 0,
+                  #                           family="Times New Roman"),
+                  # axis.text.y.right = element_text(size = 12,
+                  #                           face = 'plain',
+                  #                           angle = 0,
+                  #                           family="Times New Roman"),
                   strip.text.x = element_text(size = 12,
@@
 Each type of visualization comes with its own advantages and disadvantages.
-For example, paths between two nodes in a network are easier to determine with node-link diagrams than with adjacency matrix visualizations \citep{adjmatviz}.
+For instance, paths between two nodes in a network are easier to determine with node-link diagrams than with adjacency matrix visualizations \citep{adjmatviz}.
 In node-link diagrams, node-level information can be incorporated into the visualization by coloring or changing the shape of the points representing the nodes, and edge-level information can be incorporated by coloring the lines, or changing their thickness, linetype, or color.
-Incorporating a node-level variable into an adjacency matrix visualization is not as straightforward or simple, which is more focused on edges.
-Adjacency matrix visualization has been found to be particularly useful when the network is very complex, dense, or large, and experimental studies have shown adjacency matrix visualization to be superior to node-link diagrames for large networks.
+Incorporating a node-level variable into an adjacency matrix visualization is not as straightforward or simple, because this type of visualization features the edges.
+Adjacency matrix visualization can be useful when the network is very complex, dense, or large.
+Experimental studies have shown adjacency matrix visualization to be superior to node-link diagrames for large networks.
 For example, for basic perceptual tasks on networks, including node and edge count, adjacency matrix visualizations outperform node-link diagrams as the size and density of the network increases \citep{adjmatviz}.
@@
 So, there are many repetitions in the matrix, which makes the network appear more dense than it really is.}
-\st{This, however, may actually be an advantage for \emph{directed} graphs, where exactly the correct number of edges is represented in a matrix visualization, due to the fact that the edges $x_{ij}$ and $x_{ji}$ are not interchangeable.
+This is advantageous for \emph{directed} graphs, where exactly the correct number of edges is represented in a matrix visualization, due to the fact that the edges $x_{ij}$ and $x_{ji}$ are not interchangeable.
 A node-link diagram, however, may underrepresent the edge count if the edges $x_{ij}$ and $x_{ji}$ both exist and are drawn on top of one another.
-Ultimately, however t}here is not one "correct" way to visualize network information, and we will be using both \st{the node-link and adjacency matrix visualization} methods throughout this paper to explore social networks and stochastic actor-oriented models.
+Ultimately, however there is not one "correct" way to visualize network information, and we will be using both the node-link and adjacency matrix visualization methods throughout this paper to explore social networks and stochastic actor-oriented models.
 
@@
 
-A Stochastic Actor-Oriented Model (SAOM) is a model that incorporates all three components of dynamic networks: edge, node, and time information.
+A Stochastic Actor-Oriented Model (SAOM) is a model that incorporates all three data levels of dynamic networks: edge, node, and temporal information.
 It models the change of a network over time, allowing for changes in network structure due to actor-level covariates.
 This model was first introduced by Snijders in 1996 \citep{saompaper}.
-The two titular properties of SAOMs, stochasticity and actor-orientation, are crucial to understanding networks as they exist naturally.
-Most social networks, even holding constant the set of actors over time, are ever-changing as relationships decay or grow in seemingly random ways, and most actors (or nodes) in social networks have inherent properties that could affect how they change their role within the network, and vice versa.
+The two titular properties of SAOMs, stochasticity and actor-orientation, are key to understanding networks as they exist naturally.
+Most social networks, even holding constant the set of actors over time, are ever-changing as relationships decay or grow in seemingly random ways.
+Most actors in social networks have inherent properties that could affect how they change their role within the network, and ties between actors may affect formation or dissolution of ties.
 
@@
 
-In this paper, the term \textit{dynamic network} refers to a network, consisting of a fixed set of $n$ nodes, that is changing over time, and is observed at $M$ discrete time points, $t_1, \dots, t_M$ with $t_1 \le t_2 \le \dots \le t_M$.
+We use the term \textit{dynamic network} to refer to a network, consisting of a fixed set of $n$ nodes, that is changing over time, and is observed at $M$ discrete time points, $t_1, \dots, t_M$ with $t_1 \le t_2 \le \dots \le t_M$.
 We denote the network observation at timepoint $t_k$ by $x(t_k)$.
@@
 The SAOM assumes that this longitudinal network of discrete observations is embedded within a continuous time Markov chain (CTMC), which we will denote $X(T)$.
-This process is almost entirely unobserved: we assume that the beginning of the process, $X(0)$, is equivalent to the first network observation $x(t_1)$, while the end of the process,$X(\infty)$, is equivalent to the last observation $x(t_M)$.
+This process is almost entirely unobserved: we assume that the beginning of the process, $X(0)$, is equivalent to the first network observation $x(t_1)$, while the end of the process, $X(\infty)$, is equivalent to the last observation $x(t_M)$.
 Nearly all other parts of the process are unseen, with the exception of $x(t_2), \dots, x(t_{M-1})$.
@@
 Thus, the ``in-between" observations are considered to be ``snapshots" of the network at some point between two steps in the CTMC.
-The whole process $X(T)$ is a series of single tie changes that happen according to some pre-defined rate function, where one actor at a time is given the opportunity to add or remove one outgoing tie, or to not make any changes.
-Once an actor is chosen at random according to the rate function, it is ``given" the chance to change a tie, and it tries to maximize its utility function based on the current and near future states of the network.
-We expand on the model description further in the subsequent sections.
-
-\subsubsection{The Rate Function}
-
-For the network $x$ and each actor $i$ in the network, the number of times that an actor $i$ gets to change its ties, $x_{ij}$, to other nodes $j \neq i$ in the network is dictated by a \emph{rate function} $\rho(x, \mathbf{z}, \boldsymbol{\alpha})$, where $\boldsymbol{\alpha}$ are the parameters in the function $\rho$, and $x$ is the current network state, with covariates of interest $\mathbf{z}$.
-For this paper, we assume a simple rate function, $\rho(x, \mathbf{z}, \boldsymbol{\alpha}) = \alpha_m$ that is constant across all actors between observations at time $t_m$ to $t_{m+1}$, \st{thus our rate function is just a rate parameter in the overall model}.
-In general, SAOMs can incorporate covariate values and network statistics into the model, so that each node will have a different rate of change.
+
+The CTMC process $X(T)$ is a series of single tie changes that happen according to some pre-defined rate function (Section \ref{sec:saomratefn}), where one actor at a time is selected and it adds or removes one outgoing tie, or does nothing.
+Once an actor is chosen at random according to the rate function, it is ``given" the chance to change a tie, and it tries to maximize its utility function (Section \ref{sec:saomobjfn}) based on the current and possible future states of the network.
+
+\subsubsection{The Rate Function}\label{sec:saomratefn}
+
+In the broadest definition of a SAOM, each actor $i$ in the network $x$ may change its ties, $x_{ij}$, to other nodes $j \neq i$ according to the SAOM's  \emph{rate function}, $\rho(x, \mathbf{z}, \boldsymbol{\alpha})$, where $\boldsymbol{\alpha}$ are the parameters in the function $\rho$, and $x$ is the current network state, with covariates of interest $\mathbf{z}$.
+With this general formulation, it is possible that each node will have a different rate of change.
+We assume a simple rate function, $\rho(x, \mathbf{z}, \boldsymbol{\alpha}) = \alpha_m$ that is constant across all actors between observations at time $t_m$ to $t_{m+1}$, and thus our rate function is just a set of rate parameters in the overall model.
 %\hh{XXX that means that the function is essentially just a rate parameter, right?} \st{XXX yes XXX}
-Other modelling scenarios allow this rate to be more flexible,  e.g.\ a function that depends on the time period of observation, some actor-level covariates or some actor-level network statistics.
+%Other modelling scenarios allow this rate to be more flexible,  e.g.\ a function that depends on the time period of observation, some actor-level covariates or some actor-level network statistics.
 %\hh{How is the rate parameter different from a rate parameter?} \st{I think you meant how is a rate fn different from a rate parameter?
 I updated the language XXX->}%In our model with the simple rate function,
-\st{In our simple model with a simple rate parameter instead of a rate function,} the rate parameter dictates how quickly an actor $i$ gets an opportunity to change one of its ties to the other nodes in the network, $x_{ij}$, for $j \in \{1, \dots, n\}$ in the time period from $t_{m}$ to $t_{m+1}$.
-If $j = i$, no change in the network is made.
-The model also assumes that the actors $i$ are conditionally independent given their ties, $x_{i1}, \dots, x_{in}$ at the current network state.
-\st{Let $\tau(i|x,m) $ be the wait time until actor $i$ makes its next change from its current state in the network $x$.
-Note that $m$ indicates the number of the wave that is conditioned on in the SAOM.
+In the simple model with a rate parameters instead of a rate function, the rate parameters dictate how quickly an actor $i$ gets an opportunity to change one of its ties to the other nodes in the network, $x_{ij}$, for $j \in \{1, \dots, n\}$ in the time period from $t_{m}$ to $t_{m+1}$ for $m = 1, \dots, M-1$ If $j = i$, no change in the network is made.
+We also assume that the actors $i$ are conditionally independent given their ties, $x_{i1}, \dots, x_{in}$ at the current network state.
+Let $\tau(i|x,m) $ be the wait time until actor $i$ makes its next change from its current state in the network $x$.
+%Note that $m$ indicates the number of the wave that is conditioned on in the SAOM.
 For any time point, $T$, where $t_{m} \leq T < t_{m+1}$, the waiting time to the next change opportunity by actor $i$ is exponentially distributed with expected value $\alpha_m^{-1}$.
@@
 
-} The parameter for the wait time for the whole network $n\alpha_m$ is the rate at which any tie change occurs.
-The estimation of this parameter is straightforward: a the method of moments is used to estimate the rate with the statistic $$C = \sum_i\sum_j |x_{ij}(t_{m+1}) = x_{ij}(t_m)|$$ which is the total number of changes from observation at time $t_m$ to the observation at time $t_{m+1}$.
+The parameter for the wait time for the whole network $n\alpha_m$ is the rate at which any tie change occurs.
+The estimation of this parameter is straightforward: a the method of moments is used to estimate the rate with the statistic $$C = \sum_i\sum_j |x_{ij}(t_{m+1}) - x_{ij}(t_m)|$$ which is the total number of changes from observation at time $t_m$ to the observation at time $t_{m+1}$.
 %In order to achieve the memorylessness property of a Markov process,
@@
 
-\subsubsection{The Objective Function}
+\subsubsection{The Objective Function}\label{sec:saomobjfn}
 
 %\hh{shouldn't that be independence? - make the conditional independence assumption above a labelled formula (e.g. C1) and then call it here }
-Because of the conditional \st{in}dependence assumptions given in Equation~\ref{eq:c1}, we can consider the objective function for each node separately, as only one tie from one node is \st{allowed to change} at a time.
-\st{The node $i$, which is the node that is chosen to change at the current time point, is called the \textit{ego} node.
+Because of the conditional independence assumptions given in Equation~\ref{eq:c1}, we can consider the objective function for each node separately, as only one tie from one node is changing at a time.
+The node $i$, which is the node that is changing at the current time point, is called the \textit{ego} node.
 It has the potential to interact with all other nodes in the network, $j \neq i$.
-These nodes $j$, are referred to as \textit{alter} nodes, or simply \textit{alters}.
-These nodes are acted upon by the ego node, and they only act when they become the ego node at a subsequent time point in the CTMC.} For the ego node, $i$, in the current network state $x$, its objective function, which it tries to maximize, is written as
+These nodes $j$, are referred to as \textit{alter} nodes.
+These nodes are acted upon by the ego node, and they only act when they become the ego node at a subsequent time point in the CTMC.
+For the ego node, $i$, in the current network state $x$, its \emph{objective function}, which it tries to maximize, is written as
 \begin{equation} \label{eq:of1}
@@
 \end{equation}
-
 for $x \in \mathcal{X}$, the space of all possible directed networks with the $n$ nodes, and $\mathbf{Z}$, the matrix of covariates.
@@
 Is it ok there?}
-Given the ego node, $i$, there are $n$ possible steps for the actor $i$ to take: either one of all current ties $x_{ij} = 1$ will be destroyed, a new tie will be created thatis currently $x_{ij}=0$, or no change will occur.
-
-The parameters, $\boldsymbol{\beta}$, correspond to various actor-level network statistics, $s_{ik}(x)$.
+Given the ego node, $i$, there are $n$ possible steps for the actor $i$ to take: either one of all current ties $x_{ij} = 1$ will be destroyed, a new tie will be created that is currently $x_{ij}=0$, or no change will occur.
+
+The parameters, $\boldsymbol{\beta}$, correspond to various actor-level statistics, $s_{ik}(x)$.
 %\hh{'There are always' ...
@@
 371]{snijders01}, there should be at least two parameters included in the model: $\beta_1$ for the outdegree of a node, and $\beta_2$ for the number of reciprocal ties held by a node.
-\st{These effects should seem familiar to readers used to working with the classical exponential random graph model (ERGM) for networks.
+These effects are also used in ERGMs.
 The outdegree represents the propensity of nodes with a lot of outgoing ties to form more outgoing ties (the "rich get richer" effect), and the reciprocity parameter measures the tendency of outgoing ties to be returned within a network.
-The statistics corresponding to these effects are written in terms of the edge variables $x_{ij}$, for $i \neq j$.} In the \texttt{RSiena} software that we use to fit the SAOMs, there are over 80 possible parameters to add to the model.
+The statistics corresponding to these two parameters are functions of the edge variables, $x_{ij}$.
+In the \texttt{RSiena} software that we use to fit the SAOMs, there are over 80 possible parameters to add to the model.
 The formulas for the effects are provided in \citet{RSienamanual}.
-The parameters, $\beta_k$, in the model can be split up into two groups: first, the structural effects, whose estimation depends only on the structure of the network, like the outdegree and reciprocity parameters mentioned above.
-The parameters are included when the researcher hypothesizes that they will model underlying mechanisms of network change.
-They hope to answer questions such as, ``How does the existing network structure influence change in the network?" %\hh{XXX move the last colored section to the front of this paragraph.
+
+Each of the parameters in the model belongs to one of two groups: structural or covariate.
+The structural parameters, such as the outdegree and reciprocity parameters, correspond to statistics that are functions of the edge variables only.
+They model underlying mechanisms of network change, answering questions such as, ``How does the existing network structure influence change in the network?" %\hh{XXX move the last colored section to the front of this paragraph.
 That sets up the structure nicely, e.g.
@@
 % and ``How do the behavior and characteristics of the nodes influence change in the network?" put this sentence below somewhere
-The second set of effects are referred to as the actor-level or covariate effects.
-These covariate effects also depend on the structure of the network, with the additional inclusion of node-level covariates of interest.
-The covariate effects are written in terms of the tie variables $x_{ij}$, but also in terms of the covariates, $\mathbf{Z}$.
-A table of some possible structural and covariate effects is given in Table~\ref{tab:effects}.
+The covariate parameters are functions of the edge variables and additional node-level covariates of interest.
+A table of some possible structural and covariate parameters and their corresponding statistics is given in Table~\ref{tab:effects}.
 For a complete list of the network and covariate statistics that can currently be included in the objective function, see \citet{RSienamanual}.
@@
 There are many more possible effects, but we only consider a select few here.
-For a complete list, see the RSiena manual \citep{RSiena}.}
+For a complete list, see the RSiena manual \citep{RSiena}.
+The darker nodes in the small figures are the ego nodes, while the lighter are the alters.
+The dotted lines are ties being formed when the parameter corresponding to each effect is positive.
+The differently shaped nodes in the covariate effect diagrams represent different levels of the covariate $z$.}
 \end{table}
@@
 When node $i$ is given the chance to change a tie, %\hh{`we assume that they wish' sounds strange - don't assume wishes, state assumptions}
-it attempts to maximize the value of its objective function $f_i(\boldsymbol{\beta}, x)$ as well as a random element, $U_i(x)$, to account for unknown attraction between nodes.
+it attempts to maximize the value of its objective function $f_i(\boldsymbol{\beta}, x)$ %as well as a random element, $U_i(x)$, to account for unknown attraction between nodes.
 The additional random element is included to account for any random, unexplainable change in the network ties.
@@
 368]{snijders01}.
-This distribution, which is also known as the log-Weibull distribution, has probability distribution function of:
-
-\begin{equation} \label{eq:dist1}
-g(u|\mu, \sigma) = \frac{1}{\sigma}\exp\left\{-\left(\frac{u-\mu}{\sigma} + \exp^{-\frac{u-\mu}{\sigma}}\right)\right\}.
-\end{equation}
-using $\mu$ for the mean parameter and $\sigma$ for the scale parameter.
-
-<<logweibull, fig.align='center', out.width = '.5\\textwidth', fig.height = 4, fig.cap='The probability distribution function for the type 1 extreme value distribution, also known as the log-Weibull or Gumbel distribution with location parameter $ \\mu = 0$ and scale parameter $ \\sigma = 0$.'>>=
+%This distribution, which is also known as the log-Weibull distribution, has probability distribution function of:
+
+%\begin{equation} \label{eq:dist1}
+%g(u|\mu, \sigma) = \frac{1}{\sigma}\exp\left\{-\left(\frac{u-\mu}{\sigma} + \exp^{-\frac{u-\mu}{\sigma}}\right)\right\}.
+%\end{equation}
+%using $\mu$ for the mean parameter and $\sigma$ for the scale parameter.
+
+<<logweibull, eval = FALSE, fig.align='center', out.width = '.5\\textwidth', fig.height = 4, fig.cap='The probability distribution function for the type 1 extreme value distribution, also known as the log-Weibull or Gumbel distribution with location parameter $ \\mu = 0$ and scale parameter $ \\sigma = 0$.'>>=
 library(reliaR)
@@
 
-The probability density function for the type 1 extreme value distribution is shown in Figure~\ref{fig:logweibull}.
+%The probability density function for the type 1 extreme value distribution is shown in Figure~\ref{fig:logweibull}.
 For mean $\mu = 0$ and scale $\sigma = 1$ as \citet{modelsSnijders} suggests %\hh{XXX I'm not following - which distribution do you mean by `this' distribution?
@@
 I got lost at the  ``objective function $f_i(\boldsymbol{\beta}, x)$ plus a random element, $U_i(x)$"} \st{XXX I hope it is clearer now above}
-is convenient because it \st{leads to a simple probability formula for} the probability that actor $i$ chooses to change its tie to actor $j$ \st{that can be written \textit{only}} in terms of the objective function.
-Let $p_{ij}(\boldsymbol{\beta}, x)$ be the probability that actor $i$ chooses to change its tie to actor $j$.
-Next, we write the network $x$ in its potential future state, $x(i \leadsto j)$, where the tie $x_{ij}$ has changed to $1-x_{ij}$.
+%is convenient because it \st{leads to a simple probability formula for} the probability that actor $i$ chooses to change its tie to actor $j$ \st{that can be written \textit{only}} in terms of the objective function.
+We can use this objective function to define the probability that actor $i$, once selected with the rate function, chooses to change its tie to actor $j$, $p_{ij}(\boldsymbol{\beta}, x)$.
+Let $x(i \leadsto j)$, denote the network identical to $x$ with only one change: the tie $x_{ij}$ has changed to $1-x_{ij}$.
 Then, the probility that the tie $x_{ij}$ changes is
@@
 When $i = j$ in $p_{ij}$, the numerator represents the exponential of the value of the objective function when evaluated at the current network state.
-When the value of the objective function is high at the current state, the probability of not making a change in a microstep is also high.
+When the value of the objective function is high at the current state, the probability of not making a change is also high.
 In the CTMC, when actor $i$ may make a change, it chooses which tie $x_{i1, \dots, x_{in}}$ to change at random according to the probabilities $p_{ij}(\boldsymbol{\beta}, x)$.
-\st{The objective function and the resulting values of $p_{ij}$ are combined with the rate function to fully describe the CTMC that is used to model network change in a SAOM.}
+The objective function and the resulting values of $p_{ij}$ are combined with the rate function to fully describe the CTMC that is used to model network change in a SAOM.
+For more on CTMCs, see \citet{ctmc}.
 
@@
 
-\subsubsection{Continuous Time Markov Chain (CTMC)}
-\st{
-In the continuous-time Markov chain literature, see for instance \citet{ctmc}, chains are characterized by their \textit{generator} or \textit{intensity} matrix $\mathbf{Q}$.
+% \subsubsection{Continuous Time Markov Chain (CTMC)}
+% \st{
+% In the continuous-time Markov chain literature, see for instance \citet{ctmc}, chains are characterized by their \textit{generator} or \textit{intensity} matrix $\mathbf{Q}$.
 This matrix describes the rate of change between two states of the CTMC process, and the rows of this matrix always add to zero.
@@
 All other entries in a row are structural zeroes because those network states cannot be reached from the current state in a single change.
-
-The two pieces of a SAOM, the rate function/parameter and the objective function, each contribute to the entries of the intensity matrix to describe the rate of change between two network states.
+% 
+% The two pieces of a SAOM, the rate function/parameter and the objective function, each contribute to the entries of the intensity matrix to describe the rate of change between two network states.
 The entries of $\mathbf{Q}$ are defined as follows: let $b \neq c \in \{1, 2, \dots, 2^{n(n-1)} \}$ be indices of two different possible states of the network, $x^b, x^c \in \mathcal{X}$.
 Then the $bc^{th}$ entry of $Q$ is:
-\[ q_{bc} = \begin{cases}
-     q_{ij}= \alpha_m p_{ij}(\boldsymbol{\beta}, x^b) & \text{if } x^c \in \{x^b(i \leadsto j) | \text{ any } i \neq j \in \{1, \dots, n\}\} \\
-      0 & \text{if } \sum_i \sum_j |x_{ij}^c -x_{ij}^b | > 1 \\
-      -\sum_{i\neq j} q_{ij} & \text{if } x^b = x^c
-   \end{cases}
-\]
-
-Thus, the rate of change between any two states, $x^b$ and $x^c$, that differ by only one tie $x_{ij}$, is the product of the rate at which actor $i$ gets to change a tie and the probability that the tie that will change is the tie to node $j$.
+% \[ q_{bc} = \begin{cases}
+%      q_{ij}= \alpha_m p_{ij}(\boldsymbol{\beta}, x^b) & \text{if } x^c \in \{x^b(i \leadsto j) | \text{ any } i \neq j \in \{1, \dots, n\}\} \\
+%       0 & \text{if } \sum_i \sum_j |x_{ij}^c -x_{ij}^b | > 1 \\
+%       -\sum_{i\neq j} q_{ij} & \text{if } x^b = x^c
+%    \end{cases}
+% \]
+% 
+% Thus, the rate of change between any two states, $x^b$ and $x^c$, that differ by only one tie $x_{ij}$, is the product of the rate at which actor $i$ gets to change a tie and the probability that the tie that will change is the tie to node $j$.
 This matrix $\mathbf{Q}$ is the foundation for estimation of a SAOM.
 %Estimating the parameters in these models is difficult, but thanks to the SIENA software \citep{RSiena}, we have an accessible way to fit SAOMs to network observations.
-
-%XXX HH can you help? I need a way better transition than that XXX
-}
+% 
+% %XXX HH can you help? I need a way better transition than that XXX
+% }
 
@@
 This package uses simulation methods to estimate parameter values using either the method of moments or maximum likelihood estimation.
-In this paper, use the method of moments estimation because the theory behind it was established in \citet{saompaper}, while the maximum likelihood estimation methods were not fully established until \citet{saomsml2010}, though \texttt{RSiena} contains capabilities to use maximum likelihood estimation.
+We use method of moments estimation throughout.
+%because the theory behind it was established in \citet{saompaper}, while the maximum likelihood estimation methods were not fully established until \citet{saomsml2010}, though \texttt{RSiena} contains capabilities to use maximum likelihood estimation.
 We also use the score function method for estimating the derivatives of the expected values, as opposed to the finite differences method, both of which are outlined in detail in \citet{SienaAlgs}.
 
-\par For the score function method, the SIENA software uses a Robbins-Monro algorithm (see \citet{robbinsmonro}) to estimate the solution of the moment equation
+\par The underlying SIENA software uses a Robbins-Monro algorithm (see \citet{robbinsmonro}) to estimate the solution of the moment equation
 $$ E_{\theta}S = s_{obs}$$
-where $\theta$ is the vector of rate and objective function parameters, and $s_{obs}$ is the observed vector of model statistics, $S$. The entire algorithm is provided in \citet{SienaAlgs}.
+where $\theta = (\boldsymbol{alpha}, \boldsymbol{\beta})$ is the vector of rate and objective function parameters, and $s_{obs}$ is the observed vector of model statistics.
+The entire algorithm is provided in \citet{SienaAlgs}.
 
@@
 The third phase uses the parameter vector estimated in phase two to estimate the score functions and covariance matrix of the parameter estimate, and also carries out convergence checks.
-In each of the the first two phases, the estimation procedure also uses ``microsteps" that simulate from the model as it exists in its current state in order to update either the score functions or the parameter estimates.
+
+In each of the the first two phases of the algorithm, the procedure uses ``microsteps" that are simulated from the model as it exists in its current state in order to update either the score functions or the parameter estimates.
 These simulated microsteps are observed instances of the continuous-time Markov chain that is the backbone of the stochastic actor-oriented model.
-In Section~\ref{sec:ModelVis}, we further explore these phases in the SIENA method-of-moments algorithm through visualization, bringing them out of the ``black-box" and into the light.
-
-\subsection{Model Goodness-of-Fit}\label{sec:modgof}
-
-The \texttt{RSiena} software that fits the models to data also includes a goodnes-of-fit function for examining model fit, \texttt{sienaGOF()}.
+In Section~\ref{sec:ModelVis}, we further explore these phases in the SIENA method-of-moments algorithm through visualization, bringing them out of their ``black box" and into the light.
+
+%\subsection{Model Goodness-of-Fit}\label{sec:modgof}
+
+%The \texttt{RSiena} software that fits the models to data also includes a goodnes-of-fit function for examining model fit, \texttt{sienaGOF()}.
 This function ``assess[es] the fit of the model with respect to auxiliary statistics of networks" \citep[p.~53]{RSienamanual}.
@@
 The goodness-of-fit is evaluated as follows:
-\begin{enumerate}
-\item The auxiliary statistics are computed on the observed data and on $N$ simulated observations from the model. Typically, $N=1000$.
-\item The mean vector and covariance matrix of the statistics on the simulations from the model are computed.
-\item The Mahalanobis distance from the observed statistics to the distribution of the simulated statistics is computed using the mean and covariance found in step 2.
-\item The Mahalanobis distance from each of the $N$ simulations to the same distribution is computed, and the Mahalanobis distance of the observed data is compared to this distribution of distances.
-\item An empirical $p$-value is found by computing the proportion of simulated distances found in step 4 that are as large or larger than the Mahalanobis distance from the data.
+% \begin{enumerate}
+% \item The auxiliary statistics are computed on the observed data and $N$ simulated observations from the fitted model. Typically, $N=1000$.
+% \item The mean vector and covariance matrix of the statistics on the simulations from the model are computed.
+% \item The Mahalanobis distance from the observed statistics to the distribution of the simulated statistics is computed using the mean and covariance found in step 2.
+% \item The Mahalanobis distance from each of the $N$ simulations to the same distribution is computed, and the Mahalanobis distance of the observed data is compared to this distribution of distances.
+% \item An empirical $p$-value is found by computing the proportion of simulated distances found in step 4 that are as large or larger than the Mahalanobis distance from the data.
 A SAOM is thus considered a good fit if $p$ is large.
-\end{enumerate}
+% \end{enumerate}
 
@@
 
-The \texttt{plot.sienaGOF()} function allows us to visualize this fit.
+%The \texttt{plot.sienaGOF()} function allows us to visualize this fit.
 %In Figure~\ref{fig:sienagof}, we provide an example of what the goodness-of-fit plot output looks like for indegree and outdegree statistics for a small model.
-This function draws a box plot and a violin plot at each value of the statistic of interest observed in the simulations: \st{on the $x$-axis, the out(in)degrees observed in the data ($0, 1, 2, 3, \dots$), and on the $y$-axis, the cumulative number of times that out(in)degree value appears in the simulations and the data.} In order to compare the distribution of the counts of nodes with the specified degree as calculated on the simulated networks to the counts observed in the true data, red points connected by red lines representing the observed values are superimposed on the boxplots.
+%This function draws a box plot and a violin plot at each value of the statistic of interest observed in the simulations: \st{on the $x$-axis, the out(in)degrees observed in the data ($0, 1, 2, 3, \dots$), and on the $y$-axis, the cumulative number of times that out(in)degree value appears in the simulations and the data.} In order to compare the distribution of the counts of nodes with the specified degree as calculated on the simulated networks to the counts observed in the true data, red points connected by red lines representing the observed values are superimposed on the boxplots.
 If the red points lie "well within" the simulated values, the model is a good fit to the data.
@@
 In order to understand the fit of the model, we should try to understand how well the model captures the\textit{overall} structure of the network, not just one or two summary measurements of that structure.} %The model examined in Figure~\ref{fig:sienagof} appears to do a better job of capturing the outdegree distribution than the indegree distribution of the data.
-In Section~\ref{sec:minds}, we propose a new way of visualizing goodness-of-fit that uses the traditional node-link diagram to visualize the entire network instead of numerical summaries of the the network.
+%In Section~\ref{sec:minds}, we propose a new way of visualizing goodness-of-fit that uses the traditional node-link diagram to visualize the entire network instead of numerical summaries of the the network.
 
@@
        aes(from_id = X1, to_id = X2, color = as.ordered(drink), group = as.ordered(drink))) +
-  geom_net(fiteach = FALSE, directed=TRUE, layout.alg = 'fruchtermanreingold', size = 5, arrowsize = .25, arrowgap = .05, ecolour = 'grey40', labelon=T, labelcolour = 'black', fontsize = 4, vjust = .5) +
+  geom_net(fiteach = FALSE, directed=TRUE, layout.alg = 'fruchtermanreingold', size = 5, arrowsize = .25, arrowgap = .05, ecolour = 'grey40', labelon=T, labelcolour = 'black', fontsize = 4, vjust = .5, linewidth=0.5) +
   scale_color_brewer(palette = "YlOrRd", name = "Drinking\nBehavior", labels = c("never", "1-2 times per year", "once per month", "once per week")) +
@@
 This subset contained actors 20 through 35 and the ties between them, as well as the drinking behavior of each actor at each of the three waves.
-This specific subset was chosen because it showed somewhat higher connectivity than other subsets, as we've emphasized in the visualizations of the three network observations in Figure~\ref{fig:adjmatviz}.
-For model fitting, we condition on wave 1 and estimate the parameters of the models from the second and third waves.
-We will also be working with one actor level categorical covariate, drinking behavior.
+%This specific subset was chosen because it showed somewhat higher connectivity than other subsets, as we've emphasized in the
+The node-link visualizations of the three network observations are provided in Figure~\ref{fig:adjmatviz}.
+For model fitting, we condition on wave 1 and estimate the parameters of the models for the second and third waves.
+We will also be working with an actor level categorical covariate, drinking behavior.
 This variable has five values in the original data: (1) does not drink, (2) drinks once or twice a year, (3) drinks once a month, (4) drinks once a week, and (5) drinks more than once a week.
-The network and the actor covariate values are visualized using a node-link diagram in Figure~\ref{fig:alldataviz}.
-\st{In the node-link diagram, the nodes are colored according to the drinking behavior of that student.
+%The network and the actor covariate values are visualized using a node-link diagram in Figure~\ref{fig:alldataviz}.
+In the node-link diagram in Figure~\ref{fig:alldataviz}, the nodes are colored according to the drinking behavior of that student.
 Over time, we can see that the students tend to drink more and become increasingly isolated into smaller groups.
-An analysis of this type of data with a SAOM should capture these dynamics in a way that allows the researcher to draw conclusions about the nature of the network and behavioral forces at play.}
-
-\st{The second data example we use is a collaboration network in the United States Senate during the $111^{th}$ through $114^{th}$ Congresses.
-These sessions of congress correspond to the years of Barack Obama's presidency, from 2009-2016.\footnote{Details of how this data can be downloaded are provided by François Briatte at \url{https://github.com/briatte/congress}}.
+An analysis of this type of data with a SAOM should capture these dynamics in a way that allows the researcher to draw conclusions about the nature of the network and behavioral forces at play.
+
+The second data example we use is a collaboration network in the United States Senate during the $111^{th}$ through $114^{th}$ Congresses.
+These sessions of congress correspond to the years of Barack Obama's presidency, from 2009-2016.
+(Details of how this data can be downloaded are provided by François Briatte at \url{https://github.com/briatte/congress}).
 In this network, ties are directed from senator $i$ to senator $j$ when senator $i$ signs on as a cosponsor to the bill that senator $j$ authored.
-There are (somewhat surprisingly) many hundreds of ties between senators when they are connected in this way, so we simplify the network by computing a single value for each senator-senator collaboration called the \textit{weighted propensity to cosponsor} (WPC).
+There are many hundreds of ties between senators when they are connected in this way, so we simplify the network by computing a single value for each senator-senator collaboration called the \textit{weighted propensity to cosponsor} (WPC).
 This value is defined in \citet{senate} as
@@
 \]
-The networks we constructed for the four senates during President Obama's administration are shown in Figure~\ref{fig:senateplot}.
-In Section~\ref{sec:ModelVis}, we fit several stochastic actor-oriented models to these data sets use those models to guide our further exploration of SAOMs.
-}
+The networks we constructed for the four senates during the Obama administration are shown in Figure~\ref{fig:senateplot}.
+In Section~\ref{sec:ModelVis}, we fit several stochastic actor-oriented models to these two example data sets and we use those models to guide our exploration of SAOMs.
 
@@
 The concept of model visualization was developed to complement traditional model diagnostic tools.
-Typically, numerical summaries such as $R^2$ are used to assess model fit, and the occasional visualization, like a residual plot, are used to determine how well the model fits the data.
-\citeauthor{modelvis} outline three separate ideas, each of which are can be referred to simply as the "model": the model family, the model form, and the fitted model.
+Typically, numerical summaries such as $R^2$ are computed to assess model fit, and the occasional visualization, like a residual plot, are used to determine how well the model fits the data.
+\citeauthor{modelvis} outline three separate ideas, each of which can be referred to simply as the ``model": the model family, the model form, and the fitted model.
 The latter is primarily what one thinks of first when considering a model in a data analysis, where a specified model is fit to data, and parameter estimates and other numerical summaries, such as $R^2$ are reported.
-In the context of SAOMs, the fitted model contains the form of the rate and objective functions, the estimated rate parameters, and the estimated objective function parameters.
+In the context of SAOMs, the \emph{fitted model} contains the form of the rate and objective functions, the estimated rate parameters, and the estimated objective function parameters.
 The model form describes the the model \textit{before} the fitting process, defining which parameters are in the model within the context of the larger model family.
-In SAOMs, the model form includes description of the rate and objective functions and the variables therein that describe how the network evolves over time.
+In SAOMs, the \emph{model form} includes description of the rate and objective functions and the variables therein that describe how the network evolves over time.
 Finally, the model family is the broadest description of the model.
 This is the type of model that you wish to fit to the data, and is chosen based on the problem, data, and knowledge at hand.
-For example, we chose to use a SAOM to model network data over an exponential random graph model (ERGM) because we believe that actor-level variables effect network structure and formation, and we wanted to model the network changes over time.
+For example, we chose to use a SAO \emph{model family} to model our network data over an exponential random graph model (ERGM) because we believe that actor-level variables effect network structure and formation, and we wanted to model the network changes over time.
 

### 2018-05-14 paper-child.Rnw
getting there :pensive:

@@
 %These three uses of the term ``model"
-\st{The model family, the model form, and the fitted model can each be visualized according to the three principals of model visualization: we can }view the model in the data space, visualize collections of models, and explore the process of fitting the model, not just the end result.
-\st{Since we have already decided on our model family, SAOMs, }we now shift our focus to the fitted model and the model form.
-Specifically, we want to learn more about how the model form we choose affects the fitted model by using our example data sets and our visualization toolbox.
+The model family, the model form, and the fitted model can each be visualized according to the three principals of model visualization: we can view the model in the data space, visualize collections of models, and explore the process of fitting the model, not just the end result.
+Since we have already decided on our model family, SAOMs, we shift our focus to the fitted model and the model form.
+Specifically, we want to learn more about how the \emph{model form} we choose affects the \emph{fitted model} by using our example data sets and our visualization toolbox.
 We begin by introducing the five models that we fit to our example data.
-\st{Next we use the five models to guide our visual explorations of SAOMs.
-We first use novel tools and ideas to view a SAOM in the data space of a dynamic network.
-We then explore collections of the same models fit many times to the example data to see how the simulation processes in \texttt{RSiena} affect the model fits.
-Finally, we look behind the scenes and into the individual steps of the continuous time Markov chain to learn more about how this "hidden data" mechanism works and how it results in a fitted model.
-}
+We then use the five models to guide our visual explorations of SAOMs.
+We begin with novel tools and ideas to view a SAOM in the data space of a dynamic network.
+We then explore collections of the same models fit many times to the example data to see how the simulation processes in the \texttt{RSiena} ``black-box" affect the model fits.
+Finally, we look behind the scenes and into the individual steps of the continuous time Markov chain to learn more about how this ``hidden data" mechanism works and how it results in a fitted model.
 
@@
 
-We first consider the 16 actor subset of the teenage friends and lifestyle data available on the \texttt{RSiena} website \citep{RSiena}.
+We first consider the 16 actor subset of the teenage friends and lifestyle data show in Figures \ref{fig:adjmatviz, fig:alldataviz}.
 To this data, we fit three different SAOMs.
@@
 where $s_{i3} = \sum_{j \neq h} x_{ij}x_{ih}x_{hj} \mathbb{I}(z_i = z_h \neq z_j)$, and $s_{i4} =  |\{j : x_{ij} = 0, \sum\limits_h x_{ih}x_{hj} \geq 2\}|$.
-These statistics are known as the number of jumping transitive triplets and the number of doubly achieved distances two effect, respectively.
+These statistics are known as the \emph{number of jumping transitive triplets} and the \emph{number of doubly achieved distances two effect}, respectively.
 The first statistic emphasizes triad relationships that are formed between actors from different covariate groups, while the other emphasizes indirect ties between actors.
-The covariate groups are determined by the student's drinking behavior, the values of which are numeric and mean-centered.
+The covariate groups are determined by the student's drinking behavior, the values of which are converted to numeric then mean-centered.
 The four values in our data are $\{\text{never, 1-2 times per year, once per month, once per week}\}$, which after being converted to numeric and mean-centered become $\{-0.8125,-1.8125,0.1875,1.1875\}$.
@@
 In Section~\ref{sec:samemoddiffdat}, we also fit models M1, M2, and M3 to the senate collaboration data for comparison.
-\st{For fitting M2 to the senate data, we use the number of bills authored by each senator as the node covariate for the jumping transitive triplet variable.
-In terms of the senate data, then, the value of $\beta_3$ should dictate how willing a senator is to coauthor a bill on which the author of the bill has a different level of authorship, assuming there is an intermediary between the two who has the same authorship level as the first.} %Comparison of the parameter estimates of these models fit to the two different data sets is given in Table~\ref{tab:meanests}.
+To fit M2 to the senate data, we use the number of bills authored by each senator as the node covariate for the jumping transitive triplet variable.
+In terms of the senate data, then, the value of $\beta_3$ should dictate how willing a senator is to coauthor a bill on which the author of the bill has a different level of authorship, assuming there is an intermediary between the two who has the same authorship level as the first.
+%Comparison of the parameter estimates of these models fit to the two different data sets is given in Table~\ref{tab:meanests}.
 %In addition, we fit two more models, M4 and M5 to the senate data that were found to be significant according to the Wald-type test:
@@
 If the value of the parameter $\beta_6$ is high, nodes belonging to the same covariate group are encourage to reciprocate incoming ties.
-\st{Due to the intractability of SAOMs, it is difficult to know for certain how to interpret a fitted value of a parameter.
+Due to the intractability of SAOMs, it is difficult to know for certain how to interpret a fitted value of a parameter.
 We can make educated guesses based on the definition of the effect and the sign of the fitted value, but a direct interpretation is not always possible.
-But,} by exploring these models more visually, we aim to understand these effects, their interpretations, and the model fitting process better.
+By graphically exploring these models, we aim to understand their effects, their interpretations, and the model fitting process better.
 
@@
 The first way we hope to better understand stochastic actor-oriented models is by viewing the model(s) in the data space.
-In \citeauthor{modelvis}, they \st{happen to} define the \textit{data space} as ``the region over which we can reliably make inferences, usually a hypercube containing the data" \citep[p.~206]{modelvis}.
+In \citeauthor{modelvis}, they chose define the \textit{data space} as ``the region over which we can reliably make inferences, usually a hypercube containing the data" \citep[p.~206]{modelvis}.
 But what does this definition mean for network data?
-For dynamic social networks, there are a few different data ``spaces":
-\begin{enumerate}
-\item the actors and their corresponding covariates,
-\item the edges and their correpsonding variables that describe the relationships between the nodes, and
-\item the time, both the continuous unobserved time and the discrete observed time points, over which the network evolves
-\end{enumerate}
-These three data pieces can be visualized together in various ways.
-The traditional node-link visualization uses one of many algorithms to layout the actors as points in 2D space, then draws segments connecting the points in 2D if there is an edge between two nodes, and draws nothing otherwise.
+The multi-level data structure of networks makes answering this question difficult.
+% %For dynamic social networks, there are a few different data ``spaces":
+% \begin{enumerate}
+% \item the actors and their corresponding covariates,
+% \item the edges and their correpsonding variables that describe the relationships between the nodes, and
+% \item the time, both the continuous unobserved time and the discrete observed time points, over which the network evolves
+% \end{enumerate}
+The node, edge, and time data can be visualized together in various ways.
+%The traditional node-link visualization uses one of many algorithms to layout the actors as points in 2D space, then draws segments connecting the points in 2D if there is an edge between two nodes, and draws nothing otherwise.
 The time aspect can be visualized by drawing each network observation in time and placing the observed timepoints side-by-side.
 
-\st{Because longitudinal network data consist of three different ``spaces" of data, viewing the model in the data space can depend on which aspect of the \textit{model} and the \textit{data} we are interested in viewing.
+Because longitudinal network data consist of three different ``spaces" of data, viewing the model in the data space can depend on which aspect of the \textit{model} and the \textit{data} we are interested in viewing.
 %The SIENA software includes a method for modelling covariate change in the network
@@
 In this case, a plot of predicted covariate values over time would put the model into the time and node data space.
-Most likely, however, is that we would want to view all of the data spaces simultaneously.}
+Most likely, however, is that we would want to view all of the data spaces simultaneously.
 

### 2018-05-26 paper-child.Rnw
delete a bunch of unnecessary words. pdf errors on a lot of plots when compiling. no clue what's going on there. i hate this.

@@
 <<pkgs>>=
+# if (packageVersion("ggplot2") > "2.2.0"){
+#   remove.packages("ggplot2")
+#   install.packages("ggplot2")
+# }
+library(ggplot2)
 library(tidyverse)
@@
 library(GGally)
-devtools::install_github("sctyner/netvizinf")
+#devtools::install_github("sctyner/netvizinf")
 library(netvizinf)
@@
         plot.title = element_text(size = 20),
-        axis.title.x.top = element_text(size = 18),
-        axis.title.y.left = element_text(size = 18),
-        axis.text.x.top  = element_text(size = 18),
-        axis.text.y.left  = element_text(size = 18)) +
+        axis.title.x = element_text(size = 18),
+        axis.title.y = element_text(size = 18),
+        axis.text.x  = element_text(size = 18),
+        axis.text.y  = element_text(size = 18)) +
   labs(title = "Adjacency Matrix Visualization", x = "\nto (alter)", y = "from (ego)\n")
@@
 $$ E_{\theta}S = s_{obs}$$
-where $\theta = (\boldsymbol{alpha}, \boldsymbol{\beta})$ is the vector of rate and objective function parameters, and $s_{obs}$ is the observed vector of model statistics.
+where $\theta = (\boldsymbol{\alpha}, \boldsymbol{\beta})$ is the vector of rate and objective function parameters, and $s_{obs}$ is the observed vector of model statistics.
 The entire algorithm is provided in \citet{SienaAlgs}.
@@
 In \citeauthor{modelvis}, they chose define the \textit{data space} as ``the region over which we can reliably make inferences, usually a hypercube containing the data" \citep[p.~206]{modelvis}.
-But what does this definition mean for network data?
-The multi-level data structure of networks makes answering this question difficult.
+But what does this definition mean for network models?
+We want to make inferences on parameters in the model, but the multi-level data structure of networks makes the ``region of inference" difficult to define.
 % %For dynamic social networks, there are a few different data ``spaces":
@@
 % \end{enumerate}
-The node, edge, and time data can be visualized together in various ways.
 %The traditional node-link visualization uses one of many algorithms to layout the actors as points in 2D space, then draws segments connecting the points in 2D if there is an edge between two nodes, and draws nothing otherwise.
@@
 
+The node, edge, and time data can be visualized together in various ways.
 Because longitudinal network data consist of three different ``spaces" of data, viewing the model in the data space can depend on which aspect of the \textit{model} and the \textit{data} we are interested in viewing.
@@
 Incorporating data covariates into the network structure allows us to assess whether the ties between nodes are affected by how nodes behave over time.
-In this instance, we would want to view predictions over time.
-A SAOM can also model behavior change over time, taking both the node and edge information into account.
+In this instance, we would want to view predictions of edge presence over time as node variables change, visualizing all three data spaces at once.
+In addition, a SAOM can model node variable change over time, taking both the node and edge information into account.
 In this case, a plot of predicted covariate values over time would put the model into the time and node data space.
-Most likely, however, is that we would want to view all of the data spaces simultaneously.
-
-One tool that can bring the \st{node, edge, and time} data spaces together in this way is the \texttt{R} package \texttt{geomnet} \citep{geomnet}.
+%It is more intutive to view all of the data spaces simultaneously.
+
+One tool that can bring the node, edge, and time data spaces together is the \texttt{R} package \texttt{geomnet} \citep{geomnet}.
 Different visual features in the node-link diagram can be tied to the underlying node or edge data.
 The color, size, and shape of the points can be used to represent variables in the node data, while the color, linewidth, and linetype of the lines between points can be used to represent the edge variables.
-In a social network, node data might be age, gender, and occupation of the person in the network, and edge data might be length of connection between two people, how the people first met (school, work, church, etc.), and how often they interact, and we can view the network at different timepoints side-by-side to see its evolution.
+In a social network, node data might be age, gender, and/or occupation of the person in the network, and edge data might be length of connection between two people, how the people first met (school, work, church, etc.), and/or how often they interact.
+To view temporal changes, we can view the network at different timepoints side-by-side to see the evolution.
 Pulling all of this information together with \texttt{geomnet} allows the entire data space to be viewed at once.
@@
 In that time, she authored two bills and was a cosponsor on 17 other bills.
-With Clinton included in the node-link diagram, the senate looks much more highly collaborative than it does without her in the diagram.
-We can compare the number of bills authored throughout the senate by mapping the size of the node to the numer of bills authored by that senator.
+There are many Democratic senators who cosponsored both of her bills, giving their edge a $WPC$ of 1.
+So with Clinton included in the node-link diagram, the senate looks much more collaborative than it does without her in the diagram.
+We can compare the Senators' prolificness by mapping the size of the node to the numer of bills authored by the senator.
 We also map shape to sex, and keep color mapped to party of the senator.
-In addition, we can see the strength of the tie by mapping the linewidth of the edge to the $WPC$ value between the two senators.
+In addition, we visualize the strength of the tie by mapping the linewidth of the edge to the $WPC$ value between the two senators.
 In this single visualization, we have viewed node information (number of bills authored by a senator and the sex and party of that senator), edge information (the direction and strength of ties between two senators), and time (before and after Clinton left the senate).
+To view the model in the data space, plot simulations from different models over time with applicable node and edge variables using \texttt{geomnet}.
 % ST: I don't like this paragraph
@@
 This summary network represents 1,000 simulations of wave 2 using the values from the simple fitted model M1.',results='hide', fig.height=2.5>>=
-library(netvizinf)
 friendshipData <- array(c(fd2.w1, fd2.w2, fd2.w3), dim = c(16, 16, 3))
-drink1 <- s50a[20:35,1]
-drink2 <- s50a[20:35,2]
-drink3 <- s50a[20:35,3]
+
+drink1 <- drink[20:35,1]
+drink2 <- drink[20:35,2]
+drink3 <- drink[20:35,3]
 friendship <- sienaDependent(friendshipData)
@@
 Another way to view the model in the data space is through simulation from the model.
-No single network alone simulated from a SAOM is going to look like the data or represent the model, just as no single value simulated from the standard normal distribution will look like a bell-shaped curve.
-\st{Statisticians would prefer to look at a sample, or at least a summary statistic or two, }so it would therefore better to visualize many simulations together.
-From a statistician's point of view, a stumbling block with statistical network models generally is the lack of an ``average network" or ``expected network" value.
-Statisticians frequently rely on averages and expected values in data analyses, but statistical network models, especially those as complex as SAOMs, lack a single, intuitive expected value measure.
+No one network simulated from a SAOM is going to look like the data or represent the model, just as no single value simulated from the standard normal distribution will look like a bell-shaped curve.
+As statisticians, we prefer to look at a sample, or at least a summary statistic or two, so it would therefore better to visualize many simulated networks at once.
+A stumbling block with network models generally is the lack of an ``average network" or ``expected network" value.
+We frequently rely on averages and expected values in data analyses, but network models, especially those as complex as SAOMs, lack a single, intuitive expected value measure.
 We could talk about expected values of parameters, but the parameters can be hard to interpret.
-\st{Expected values of parameters are important, but if they cannot directly tell us anything about our dependent variable (networks in this case), they lose some value to us.} Furthermore, there is no way to talk about the expected value of an observation \st{simulated} from a statistical network model.
+Expected values of parameters can be computed, but if they cannot directly tell us anything about our dependent variable, they lose value.
+Furthermore, there is no way to talk about the expected value of an observation simulated from a statistical network model, though there are with most other model simulations.
 How then, can we arrive at an ``average" network?
 We answer this question through visualization.
-For network data, one way we view an "average" network is through a summary network drawn using the traditional node-link diagram.
+For network data, one way we view an ``average" network is through a \textit{summary network} drawn using the traditional node-link diagram.
 In Figure~\ref{fig:summnetM1}, we show an \textit{average network} created with 1,000 simulations of the second wave of the network from Model 1.
-\st{To make this average network, we first simulated 1,000 wave 2 and wave 3 observations of our small friendship example data from model M1, for which parameters had previously been estimated.
-We then combine the 1,000 instances of wave 2, and count up the number of times each edge appears in a simulation.
+
+To make this average network, we first simulated 1,000 wave 2 and wave 3 observations of our small friendship example data from model M1, for which parameters had previously been estimated.
+We then combine the 1,000 simulated instances of wave 2, and count up the number of times each edge appears in a network simulation.
 Then, we combine these 1,000 networks into a single network with edgeweight equal to the proportion of time that edge appears in our 1,000 simulations.
-This weighted edgelist is the network we draw using the node-link diagram.} An edge is only drawn in the average network if it appears in more than 5\% of the simulations (in at least 51 of the 1000 simulations), with edges that appear more frequently emphasized by thicker linewidths and a darker color, representing the proportion of times appeared in the model simulations.
+This weighted edgelist is the network we draw using the node-link diagram.
+An edge is only drawn in the average network if it appears in at least 51 of the 1000 simulations, with edges that appear with greater frequency emphasized by thicker linewidths and a darker color.
 On either side of the average network in Figure~\ref{fig:summnetM1}, we show the actual data, wave 1 on the left, and wave 2 on the right.
-\st{We can see that the structure of the average network is much more similar to the first wave than to the second wave.
-However, the simulations are supposed to represent the second wave of data, which is shown on the right in Figure~\ref{fig:summnetM1}.
+We can see that the structure of the average network is much more similar to the first wave than to the second wave.
+However, the simulations are supposed to represent the second wave of data on the right of Firgure~|ref{fig:summnetM1}.
 This is an indication that the simple model, M1, is doing a very poor job of capturing the change mechanism from the first to the second wave of observation.
 The average network can thus be used to help determine model goodness-of-fit.
-Because the the average network looks more like the first wave than the second wave, we can use the visualization in Figure \ref{fig:summnetM1} as evidence of poor model fit.}  % XXX ST - I was looking at the wrong one.
+Because the the average network looks more like the first wave than the second wave, we can use the visualization in Figure \ref{fig:summnetM1} as evidence of poor model fit.
+% XXX ST - I was looking at the wrong one.
 I had wave == 2 instead of wave == 1 in the code.
@@
 
-Another potential goodness-of-fit visualization \st{that places the model in the node and edge data space} is a lineup like those proposed in \citet{buja:2009}.
+Another potential goodness-of-fit visualization that places the model in the node and edge data space is a lineup like those proposed in \citet{buja:2009}.
 A \textit{lineup} ``asks the witness to identify the plot of the real data from among a set of decoys, the null plots, under the veil of ignorance" \citep[p.
 4369]{buja:2009}.
-It can be thought of like a police lineup, where the ``suspect" is in a lineup among several innocent lookalike fillers, and a witness picking the suspect out of the lineup is considered evidence against the suspect.
-In data and model visualization, the ``suspect" is a plot of the true data, while the ``filler" is composed of several plots of mock data, simulated from a hypothesized model.
-If the true data stands out among the simulated data, that is taken as evidence of poor model fit, whereas if the true data is difficult to identify among the simulated data, that is taked as evidence of good model fit.
-\st{An application of the lineup protocal can be found in \citet{hofmann2012graphical}, where the authors examine, for instance, the differences between polar and cartesian coordinates for plotting categorical data, and density plots and box plots for determining distributional differences.
-They pose questions to experiment participants such as, ``Which plot is most different from the others?" for the first example and, ``In which plot is the blue group furthest to the right?" for the distributional differences.
-The data visualizations examined in \citet{hofmann2012graphical} are less complex than a node-link network diagram.
+It can be thought of like a police lineup, where the ``suspect" is in a lineup among several innocent similar looking people, and if the witness picks the suspect out of the lineup, that is evidence the suspect is guilty.
+
+In data and model visualization, the ``suspect" is a plot of the true data, while the ``filler" is composed of several plots of ``mock data", simulated from a hypothesized model.
+If the true data stands out among the simulated data, that is taken as evidence agains the hypothesised model.
+If, however, the true data is difficult to identify among the simulated data, that is taken as evidence in favor of the hypothesized model.
+An application of the lineup protocal can be found in \citet{hofmann2012graphical}, where the authors examine, for instance, the differences between polar and cartesian coordinates for plotting categorical data, and density plots and box plots for determining distributional differences.
+They pose questions to experiment participants such as, ``Which plot is most different from the others?" or, ``In which plot is the blue group furthest to the right?"
+
+The data visualizations examined in previous applications such as  \citet{hofmann2012graphical} are less complex than a node-link network diagram.
 What questions should we ask for network data visualizations?
-Asking participants to identify the most different plot may be difficult.
-In the network lineup shown in Figure~\ref{fig:lineup}, the second wave of the small friendship network is shown among five simulated networks from model M1 using parameter values estimated from the data.
-What makes these plots ``different"?
+The most neutral choice, asking participants to identify the most different plot, may be difficult.
+As an example, consider the network lineup shown in Figure~\ref{fig:lineup}.
+In this small network lineup, the second wave of the small friendship network is shown among five simulated networks from model M1 using parameter values estimated from the data.
+What exactly makes these plots ``different" from one another?
+
 It seems possible to argue for any one of the six plots in Figure~\ref{fig:lineup} as most different.
-So, we guide participants to look at the overall structure of the graphs to determine which has the most and least complex structure.
-The least complex plot, number six, has no triplets, while the most complex plot, number three, has three triplets, and in fact, plot three is the data.
-We have found in an experiment where all the node-link diagrams shown are based on simulated data, one observation from an ``alternative" model and the rest from a ``null" model, the triangular shape of the triplets stands out the most to participants.
-We can use the lineup protocol to help better understand SAOMs because it allows us to view the model in the data space by placing observations the model side-by-side with the data and examining the differences.
+So, we suggest guiding participants to look at the overall structure of the graphs to determine which has the most or least connected or complex structure.
+The least complex plot, number six, has no triangles (3-cycles), while the most complex plot, number three, has three 3-cycles.
+The most complex plot, number three, is in fact the data.
+We have found that when shown lineups like that of Figure~\ref{fig:lineup}, online survey participants %where all the node-link diagrams shown are based on simulated data, %one observation from an ``alternative" model and the rest from a ``null" model,
+pick up on the triangular shape of the 3-cycles most often.
+
+In a future paper, we will use the lineup protocol to help better understand SAOMs by allowing us to view the model in the data space.
+Lineups place observations simulated from the model side-by-side with the data and let the viewer examining the differences.
 This can also help us determine the significant structural effects, if any, of the parameters in the model on observations simulated from the model.
-The triangular shape mentioned above becomes more prevalent when a transitive triplet parameter in included in a model, but it does not always cause triplets to form in the simulated data.
-This requires more investigation with larger data sets and more complex models, but is promising for the development of additional goodness-of-fit measures for network models.}
+The triangular shape mentioned above becomes more prevalent when a transitive triplet parameter ($\beta_3$ in Section~\ref{sec:models}) is included in a model, indicating that it may have a significant effect on network structure.
+%, but it does not always cause triplets to form in the simulated data.
+This requires more investigation with larger data sets and more complex models, but is promising for the development of additional goodness-of-fit measures for network models.
 
@@
 
-\st{There are many possible ways to collect models together.
-We could look at the same models fit to different data, different models fit to the same data, or because of the nature of SAOMs, we could fit the same models to the same data many times to see how the simulations change.
-For the SAOMs, we decided that there were four collections that were most important:
+The second method of model visualization we use is visualizing collections of models.
+There are many possible ways to collect SAO models together.
+We could look at the same model fit to different data, different models fit to the same data, or we could fit the same models to the same data many times to see how the simulations change.
+For the SAOMs, we decided that there were four collections of models that would give us the most insight:
 \begin{enumerate}
@@
 We chose these four collections because they each explore something different about SAOMs.
-The first takes the many dozens of parameters available to include in a SAOM into account, which easily translates into the second by showing how those many parameters affect the fitted models.
-Then, we look at how the same model looks when fit to different sets of dynamic network data.
-Finally, we look at the results from fitting the same model to the same data because the MCMCs and CTMCs that make up a SAOM lead to different parameter estimates every time, so it important to see how the results can vary.}
-
-\subsubsection{Exploring the space of all possible models}
-
-<<sigtesting, fig.show = 'hold', out.width='\\textwidth', fig.height=2, fig.cap="Significant effects for the two data sets, at a significance level of 0.10 or lower as calculated by the Wald-type test available in the SIENA software.">>=
-smfr_effs <- read_csv("data/effects_significance_smallFriends.csv")
-smfr_effs$inter1[is.na(smfr_effs$inter1)] <- "none"
-library(ggrepel)
-set.seed(12345678)
-smfr_effs %>% filter(type == "eval" & Waldpval <=0.10) %>%
-ggplot(aes(x = estimate, y = Waldpval, color = inter1)) +
-  #geom_errorbarh(aes(xmin = estimate - se, xmax = estimate + se), color = 'black') +
-  geom_vline(xintercept = 0) +
-  geom_point(size = 3) +
-  #geom_text_repel(aes(label = shortName)) +
-  scale_color_brewer(palette = "Paired", name = "Interaction\nVariable") +
-  ggtitle("Significant effects at level 0.10 for friendship data") +
-  coord_flip() +
-  ThemeNoNet
-sen_effs <- read_csv("data/congress/senateModelSigEffects.csv")
-#sen_effs$Waldpval <- parse_number(sen_effs$Waldpval)
-sen_effs$inter1[is.na(sen_effs$inter1)] <- "none"
-ggplot(data = sen_effs %>% filter(Waldpval <=0.10), aes(x = estimate, y = Waldpval, color = inter1)) +
-  # geom_errorbarh(aes(xmin = estimate - se, xmax = estimate + se), color = 'black') +
-  geom_vline(xintercept = 0) +
-  geom_point(size = 3) +
-  #geom_text_repel(aes(label = shortName)) +
-  scale_color_brewer(palette = "Paired", name = "Interaction\nVariable") +
-  ggtitle("Significant effects at level 0.10 for senate data") +
-  coord_flip() +
-  ThemeNoNet
-@
-
-\st{The \texttt{RSiena} manual contains over eighty possible effects to include in the model.
+The first considers the many dozens of parameters available to include in a SAOM.
+The second collection show how those many parameters affect the parameter values of the fitted models.
+Third, looks at how the same model behaves when fit to different sets of dynamic network data.
+Finally, the results from fitting the same model to the same data may lead to different parameter estimates, so it important to see how the results can vary.
+
+\subsubsection{Exploring the space of all possible models}\label{sec:allpossible}
+
+The \texttt{RSiena} manual contains over eighty possible effects to include in the model.
 In order to select parameters to include in the models for our example data, we searched through the possible effects available to model given the data structure to find \textit{significant} effects.
@@
 We performed this procedure for both the small friendship and the senate collaboration data.
-The results for significant effects are shown in Figure~\ref{fig:sigtesting}.
-We see in both the friendship data and the senate data results that most of the significant effects have absolute value less than ten.
+A visualization of the significant effects for the two example data sets are shown in the appendix.
+%We see in both the friendship data and the senate data results that most of the significant effects have absolute value less than ten.
 In addition, the $p$-values for the effects from the friendship data are more spread out than the $p$-values for the senate data, which are concentrated at about 0.02 or less.
-This may suggest that larger data sets tend to result generally in smaller $p$-values, just like a larger sample size results in smaller $p$-values in a $t$-test.}
+This may suggest that larger data sets tend to result generally in smaller $p$-values, just like a larger sample size results in smaller $p$-values in a $t$-test.
 
@@
 In this process, we condition on the first friendship network observation, and the second and third observations are simulated from the SAOM models with the given parameter values.
-From these simulations, we first create a visualization that represents an average network.
-
-<<averagenet, results='hide', out.width='\\textwidth',fig.cap='The node-link diagrams from the three "average" networks that we calculated are in the top row, and the true wave 1 and wave 2 data are shown in the bottom row above.
-There is some difference between the three models, but overall, these three models cannot capture the structure in the true second wave of data.', fig.height=5>>=
-#friend.data.w1 <- as.matrix(read.table("s50_data/s50-network1.dat"))
-#friend.data.w2 <- as.matrix(read.table("s50_data/s50-network2.dat"))
-#friend.data.w3 <- as.matrix(read.table("s50_data/s50-network3.dat"))
-#drink <- as.matrix(read.table("s50_data/s50-alcohol.dat"))
-# friendshipData <- array(c(fd2.w1, fd2.w2, fd2.w3), dim = c(16, 16, 3))
-# friendship <- sienaDependent(friendshipData)
-# alcohol <- varCovar(drink[20:35,])
-# mydata <- sienaDataCreate(friendship, alcohol)
-# M1eff <- getEffects(mydata)
-# M2eff <- includeEffects(M1eff, jumpXTransTrip, interaction1 = "alcohol")
-# M3eff <- includeEffects(M1eff, nbrDist2twice)
-# data(M1ests_bigfriends)
-# M1parms <- (means %>% filter(Model == "M1"))$means
-# M2parms <- (means %>% filter(Model == "M2"))$means
-# M3parms <- (means %>% filter(Model == "M3"))$means
-# M1sims1000 <- saom_simulate(dat = mydata,
-#                             struct = M1eff,
-#                        parms = as.numeric(M1parms),
-#                        N = 1000)
-# M2sims1000 <- saom_simulate(dat = mydata,
-#                             struct = M2eff,
-#                        parms = as.numeric(M2parms),
-#                        N = 1000)
-# M3sims1000 <- saom_simulate(dat = mydata,
-#                             struct = M3eff,
-#                        parms = as.numeric(M3parms),
-#                        N = 1000)
-# M1simsdf <- sims_to_df(M1sims1000)
-# M2simsdf <- sims_to_df(M2sims1000)
-# M3simsdf <- sims_to_df(M3sims1000)
-# M1avg <- M1simsdf %>% filter(!is.na(to) & wave == 1) %>%
-#   group_by(from, to) %>%
-#   summarise(count = n()) %>%
-#   mutate(weight = ifelse(from == to, 0, count))
-# M1avgW2 <- M1simsdf %>% filter(!is.na(to) & wave == 2) %>%
-#   group_by(from, to) %>%
-#   summarise(count = n()) %>%
-#   mutate(weight = ifelse(from == to, 0, count))
-# M1avg2 <- M1simsdf %>% filter(!is.na(to)) %>%
-#  group_by(from, to, wave) %>%
-#  summarise(count = n()) %>%
-#  mutate(weight = ifelse(from == to, 0, count))
-# M2avg <- M2simsdf %>% filter(!is.na(to) & wave == 1) %>%
-#   group_by(from, to) %>%
-#   summarise(count = n()) %>%
-#   mutate(weight = ifelse(from == to, 0, count))
-# M2avgW2 <- M2simsdf %>% filter(!is.na(to) & wave == 2) %>%
-#   group_by(from, to) %>%
-#   summarise(count = n()) %>%
-#   mutate(weight = ifelse(from == to, 0, count))
-# M2avg2 <- M2simsdf %>% filter(!is.na(to)) %>%
-#  group_by(from, to, wave) %>%
-#  summarise(count = n()) %>%
-#  mutate(weight = ifelse(from == to, 0, count))
-# M3avg <- M3simsdf %>% filter(!is.na(to) & wave == 1) %>%
-#   group_by(from, to) %>%
-#   summarise(count = n()) %>%
-#   mutate(weight = ifelse(from == to, 0, count))
-# M3avgW2 <- M3simsdf %>% filter(!is.na(to) & wave == 2) %>%
-#   group_by(from, to) %>%
-#   summarise(count = n()) %>%
-#   mutate(weight = ifelse(from == to, 0, count))
-# M3avg2 <- M3simsdf %>% filter(!is.na(to)) %>%
-#  group_by(from, to, wave) %>%
-#  summarise(count = n()) %>%
-#  mutate(weight = ifelse(from == to, 0, count))
-
-# M1avg$from <- paste0("V", M1avg$from)
-# M1avg$to <- paste0("V", M1avg$to)
-# M1avg2$from <- paste0("V", M1avg2$from)
-# M1avg2$to <- paste0("V", M1avg2$to)
-# M1avgW2$from <- paste0("V", M1avgW2$from)
-# M1avgW2$to <- paste0("V", M1avgW2$to)
-# M2avg$from <- paste0("V", M2avg$from)
-# M2avg$to <- paste0("V", M2avg$to)
-# M2avgW2$from <- paste0("V", M2avgW2$from)
-# M2avgW2$to <- paste0("V", M2avgW2$to)
-# M2avg2$from <- paste0("V", M2avg2$from)
-# M2avg2$to <- paste0("V", M2avg2$to)
-# M3avg$from <- paste0("V", M3avg$from)
-# M3avg$to <- paste0("V", M3avg$to)
-# M3avgW2$from <- paste0("V", M3avgW2$from)
-# M3avgW2$to <- paste0("V", M3avgW2$to)
-# M3avg2$from <- paste0("V", M3avg2$from)
-# M3avg2$to <- paste0("V", M3avg2$to)
-combinedavgs$linewidth[combinedavgs$cat %in% c("1st Wave", "2nd Wave")] <- .5
-combinedavgs$cat <- as.factor(combinedavgs$cat)
-combinedavgs$cat <- ordered(combinedavgs$cat, levels = c("Model1", "Model2", "Model3", "1st Wave", "2nd Wave"))
-
-ggplot(data = combinedavgs) +
-  geom_net(aes(from_id = from, to_id = to, linewidth= linewidth),
-           directed = T, curvature = .2, labelon=T, size = 5,
-           arrow = arrow(angle = 20, type = "open", length = unit(.2, "cm")), ecolour = "#969696", fiteach = T,
-           color = "#252525", hjust = .5, vjust = .5, labelcolour = 'white', fontsize = 4, arrowgap = .02,
-           singletons = T) +
-  ThemeNet + theme(panel.background = element_rect(color = 'black')) +
-  facet_wrap(~cat)
-
-# wave2drink <- drink2[,2]
-# wave2drink <- as.data.frame(wave2drink)
-# wave2drink$id <- paste0("V", 1:16)
-# names(wave2drink)[1] <- "drink"
-# wave3drink <- drink2[,3]
-# wave3drink <- as.data.frame(wave3drink)
-# wave3drink$id <- paste0("V", 1:16)
-# names(wave3drink)[1] <- "drink"
-# M1avgcov <- merge(M1avg, wave2drink, by.x = "from", by.y = "id", all = T)
-# M2avgcov <- merge(M2avg, wave2drink, by.x = "from", by.y = "id", all = T)
-# M3avgcov <- merge(M3avg, wave2drink, by.x = "from", by.y = "id", all = T)
-# M1avgW2cov <- merge(M1avgW2, wave3drink, by.x = "from", by.y = "id", all = T)
-# M2avgW2cov <- merge(M2avgW2, wave3drink, by.x = "from", by.y = "id", all = T)
-# M3avgW2cov <- merge(M3avgW2, wave3drink, by.x = "from", by.y = "id", all = T)
-#
-# M1avg[,-3] %>% spread(to, weight) %>% data.frame-> distmatM1
-# rownames(distmatM1) <- distmatM1$from
-# distmatM1 <- distmatM1[,-1]
-# distmatM1 <- as.matrix(sqrt(1/distmatM1))
-# distmatM1[is.na(distmatM1)] <- 0
-# M1avgW2[,-3] %>% spread(to, weight) %>% data.frame-> distmatM1W2
-# rownames(distmatM1W2) <- distmatM1W2$from
-# distmatM1W2 <- distmatM1W2[,-1]
-# distmatM1W2 <- as.matrix(sqrt(1/distmatM1W2))
-# distmatM1W2[is.na(distmatM1W2)] <- 0
-# M2avg[,-3] %>% spread(to, weight) %>% data.frame-> distmatM2
-# rownames(distmatM2) <- distmatM2$from
-# distmatM2 <- distmatM2[,-1]
-# distmatM2 <- as.matrix(sqrt(1/distmatM2))
-# distmatM2[is.na(distmatM2)] <- 0
-# M2avgW2[,-3] %>% spread(to, weight) %>% data.frame-> distmatM2W2
-# rownames(distmatM2W2) <- distmatM2W2$from
-# distmatM2W2 <- distmatM2W2[,-1]
-# distmatM2W2 <- as.matrix(sqrt(1/distmatM2W2))
-# distmatM2W2[is.na(distmatM2W2)] <- 0
-# M3avg[,-3] %>% spread(to, weight) %>% data.frame-> distmatM3
-# rownames(distmatM3) <- distmatM3$from
-# distmatM3 <- distmatM3[,-1]
-# distmatM3 <- as.matrix(sqrt(1/distmatM3))
-# distmatM3[is.na(distmatM3)] <- 0
-# M3avgW2[,-3] %>% spread(to, weight) %>% data.frame-> distmatM3W2
-# rownames(distmatM3W2) <- distmatM3W2$from
-# distmatM3W2 <- distmatM3W2[,-1]
-# distmatM3W2 <- as.matrix(sqrt(1/distmatM3W2))
-# distmatM3W2[is.na(distmatM3W2)] <- 0
-
-# cols <- brewer.pal(n = 4, "YlOrRd")
-#
-# plotM1W2 <- ggplot(data = M1avgcov) +
-#   geom_net(aes(from_id = from, to_id = to, linewidth = weight/251, color = as.factor(drink)),
-#            arrowgap = .03, size = 3,arrowsize = .5,
-#            directed = TRUE, curvature = .2,
-#            #layout.alg = "fruchtermanreingold"
-#            layout.alg = "kamadakawai",
-#            layout.par = list(elen = distmatM1)
-#            ) +
-#   scale_color_brewer(palette = "YlOrRd", name = "Drinking\nBehavior") +
-#   #labs(title = "Wave 2, M1") +
-#   scale_size_continuous(name = "Weight") +
-#   theme_net() +
-#   theme(plot.title = element_text(hjust = .5), legend.position = 'none',plot.margin=unit(c(0,0,0,0),"mm")) +
-#   labs(title = "Model 1 Simulated Wave 2")
-# plotM1W3 <- ggplot(data = M1avgW2cov) +
-#   geom_net(aes(from_id = from, to_id = to, linewidth = weight/388.5, color = as.factor(drink)),
-#            arrowgap = .03, size = 3,arrowsize = .5,
-#            directed = TRUE, curvature = .2,
-#            #layout.alg = "fruchtermanreingold"
-#            layout.alg = "kamadakawai",
-#            layout.par = list(elen = distmatM1W2)
-#            ) +
-#   scale_color_manual(values = cols[2:4], name = "Drinking\nBehavior") +
-#   #labs(title = "Wave 3, M1") +
-#   scale_size_continuous(name = "Weight") +
-#   theme_net() +
-#   theme(plot.title = element_text(hjust = .5), legend.position = 'none',plot.margin=unit(c(0,0,0,0),"mm"))
-# plotM2W2 <- ggplot(data = M2avgcov) +
-#   geom_net(aes(from_id = from, to_id = to, linewidth = weight/325, color = as.factor(drink)),
-#            arrowgap = .03, size = 3,arrowsize = .5,
-#            directed = TRUE, curvature = .2,
-#            #layout.alg = "fruchtermanreingold"
-#            layout.alg = "kamadakawai",
-#            layout.par = list(elen = distmatM2)
-#            ) +
-#   scale_color_brewer(palette = "YlOrRd", name = "Drinking\nBehavior") +
-#   #labs(title = "Wave 2, M2") +
-#   scale_size_continuous(name = "Weight") +
-#   theme_net() +
-#   theme(plot.title = element_text(hjust = .5), legend.position = 'none',plot.margin=unit(c(0,0,0,0),"mm")) +
-#   labs(title = "Model 2 Simulated Wave 2")
-# plotM2W3 <- ggplot(data = M2avgW2cov) +
-#   geom_net(aes(from_id = from, to_id = to, linewidth = weight/357, color = as.factor(drink)),
-#            arrowgap = .03, size = 3,arrowsize = .5,
-#            directed = TRUE, curvature = .2,
-#            #layout.alg = "fruchtermanreingold"
-#            layout.alg = "kamadakawai",
-#            layout.par = list(elen = distmatM2W2)
-#            ) +
-#   scale_color_manual(values = cols[2:4], name = "Drinking\nBehavior") +
-#   #labs(title = "Wave 3, M2") +
-#   scale_size_continuous(name = "Weight") +
-#   theme_net() +
-#   theme(plot.title = element_text(hjust = .5), legend.position = 'none',plot.margin=unit(c(0,0,0,0),"mm"))
-# plotM3W2 <- ggplot(data = M3avgcov) +
-#   geom_net(aes(from_id = from, to_id = to, linewidth = weight/258, color = as.factor(drink)),
-#            arrowgap = .03, size = 3,arrowsize = .5,
-#            directed = TRUE, curvature = .2,
-#            #layout.alg = "fruchtermanreingold"
-#            layout.alg = "kamadakawai",
-#            layout.par = list(elen = distmatM3)
-#            ) +
-#   scale_color_brewer(palette = "YlOrRd", name = "Drinking\nBehavior") +
-#   #labs(title = "Wave 2, M3") +
-#   scale_size_continuous(name = "Weight") +
-#   theme_net() +
-#   theme(plot.title = element_text(hjust = .5), legend.position = 'none',plot.margin=unit(c(0,0,0,0),"mm")) +
-#   labs(title = "Model 3 Simulated Wave 2")
-# plotM3W3 <- ggplot(data = M3avgW2cov) +
-#   geom_net(aes(from_id = from, to_id = to, linewidth = weight/372.5, color = as.factor(drink)),
-#            arrowgap = .03, size = 3,arrowsize = .5,
-#            directed = TRUE, curvature = .2,
-#            #layout.alg = "fruchtermanreingold"
-#            layout.alg = "kamadakawai",
-#            layout.par = list(elen = distmatM3W2)
-#            ) +
-#   scale_color_manual(values = cols[2:4], name = "Drinking\nBehavior") +
-#   #labs(title = "Wave 3, M3") +
-#   scale_size_continuous(name = "Weight") +
-#   theme_net() +
-#   theme(plot.title = element_text(hjust = .5), legend.position = 'none',plot.margin=unit(c(0,0,0,0),"mm"))
-#
-# origW1 <- ggplot(data = alldat %>% filter(Wave==1),
-#        aes(from_id = X1, to_id = X2, color = as.ordered(drink))) +
-#   geom_net(fiteach = FALSE, directed=TRUE, layout.alg = "kamadakawai", size = 3, arrowsize = .5, curvature = .2,
-#            arrowgap = .03) +
-#   scale_color_brewer(palette = "YlOrRd", name = "Drinking\nBehavior") +
-#   theme_net() +
-#   theme(legend.position = 'none',plot.margin=unit(c(0,0,0,0),"mm"))
-# origW2 <- ggplot(data = alldat %>% filter(Wave==2),
-#        aes(from_id = X1, to_id = X2, color = as.ordered(drink))) +
-#   geom_net(fiteach = FALSE, directed=TRUE, layout.alg = "kamadakawai", size = 3, arrowsize = .5, curvature = .2,
-#            arrowgap = .03) +
-#   scale_color_brewer(palette = "YlOrRd", name = "Drinking\nBehavior") +
-#   theme_net() +
-#   theme(legend.position = 'none',plot.margin=unit(c(0,0,0,0),"mm")) +
-#   labs(title = "Observed Wave 2")
-# origW3 <- ggplot(data = alldat %>% filter(Wave==3),
-#        aes(from_id = X1, to_id = X2, color = as.ordered(drink))) +
-#   geom_net(fiteach = FALSE, directed=TRUE, layout.alg = "kamadakawai", size = 3, arrowsize = .5, curvature = .2,
-#            arrowgap = .03) +
-#   scale_color_manual(values = cols[2:4], name = "Drinking\nBehavior") +
-#   theme_net() +
-#   theme(legend.position = 'none',plot.margin=unit(c(0,0,0,0),"mm"))
-#
-# grid.arrange(origW2, plotM1W2,
-#               plotM2W2,plotM3W2,
-#               ncol=2, clip = T)
-# alldataviz2 <- alldataviz + theme(legend.position = 'none')
-# ggdraw() +
-#   draw_plot(alldataviz2, x=0, y=.75, width=1, height=.25) +
-#   draw_plot(plotM1W2, .1, .5, .5, .25) +
-#   draw_plot(plotM1W3, .3, .5, .5, .25) +
-#   draw_plot(plotM2W2, .1, .25, .5, .25) +
-#   draw_plot(plotM2W3, .3, .25, .5, .25) +
-#   draw_plot(plotM3W2, .1, 0, .5, .25) +
-#   draw_plot(plotM3W3, .3, 0, .5, .25) +
-#   draw_plot_label(label = c("Original Data", "M1: Wave 2", "M1: Wave 3", "M2: Wave 2", "M2: Wave 3", "M3: Wave 2", "M3: Wave 3"))
-#
-# M1avg2$Model <- "M1"
-# M2avg2$Model <- "M2"
-# M3avg2$Model <- "M3"
-# allavg <- rbind(as.data.frame(M1avg2),
-#                 as.data.frame(M2avg2),
-#                 as.data.frame(M3avg2))
-#
-#
-# ggplot(data = allavg) +
-#   geom_histogram(aes(x = weight), binwidth = 25) +
-#   facet_grid(wave~Model, labeller = "label_both") +
-#   labs(x = "Number of times an edge occurs in 1,000 simulations") +
-#   theme_bw()
-
-# ggplot(data = M3avgcov) +
-#   geom_net(aes(from_id = from, to_id = to, linewidth = weight/251, color = as.factor(drink)),
-#            directed = TRUE, curvature = .2,
-#            ealpha = .75,
-#            layout.alg = "kamadakawai",
-#            layout.par = list(elen = distmatM2)
-#            ) +
-#   scale_color_brewer(palette = "YlOrRd", name = "Drinking\nBehavior") +
-#   scale_size_continuous(name = "Weight") +
-#   theme_net()
-# M1avg2$wave <- M1avg2$wave + 1
-# alldrink <- data.frame(drink2[,2:3])
-# alldrink <- gather(alldrink, wave,drink) %>% mutate(wave = parse_number(wave), id = rep(paste0("V",1:16),2))
-
-# M1avg2cov <- merge(M1avg2, alldrink, by.x = c("from", "wave"), by.y = c("id", "wave"), all = T)
-# ggplot(data = M1avg2cov) + # edge appears > 50% of the time. (is this actually what that is????)
-#   geom_net(aes(from_id = from, to_id = to, linewidth = weight/388, color = as.factor(drink)),
-#            directed = TRUE, curvature = .2, labelon = F,
-#            #layout.alg = "fruchtermanreingold"
-#            layout.alg = "kamadakawai", fiteach = T,
-#            layout.par = list(elen = distmat)
-#            ) +
-#   scale_color_brewer(palette = "YlOrRd", name = "Drinking\nBehavior") +
-#   facet_wrap(~wave, labeller = "label_both") + theme_net() + labs(title = "Average Wave 2 and Wave 3 Networks from Model M1")
-# ggplot(data = M1avg %>% filter(weight >= 51)) +
-#   geom_net(aes(from_id = from, to_id = to, linewidth = weight/500),
-#            #directed = TRUE,
-#            layout.alg = "mds",
-#            layout.par = list(var = 'user', dist = 'none', exp = 3,vm = distmat))
-@
-
-To create the average network visualization shown in Figure~\ref{fig:averagenet}, we follow the same procedure as in Section~\ref{sec:minds}, counting occurrences of each possible edge in the simulations, resulting in a summary network with weighted edges representing the number of times an edge appeared in the simulated wave 2 when simulating from the SAOM 1,000 times.
-As in Figure~\ref{fig:summnetM1}, edges only appear in the average networks if they appear more than 5\% of the time in the simulations.
-In Figure~\ref{fig:averagenet}, we show the ``average'' network from the three models we fit and the first and second waves of data.
-\st{Comparing the three averages to waves 1 and 2, we see that they have very similar structure to wave 1.
-Model 2, which included the transitive triplet parameter, seems to have created a larger connected component overall than models 1 and 3.
-In particular, if we look at the group of nodes $\{10,11,14\}$, we see they are very strongly connected within the three average networks, and they are completely separate from the other nodes in the true wave 2.
-None of the three average networks show node 16 gaining ties as it does in wave two, nor do they show nodes 4 and 7 becoming isolated.
-In Model 2, however, the ties to node 7 appear much weaker than in Model 1 or Model 2, suggesting that of the three, Model 2 may be the best fit for our data.}  %XXXX ST commented out after I realized I had made the same mistake as figure 7.
+From these simulations, we create visualizations that represents an average network using the same method described in Section~\ref{sec:minds}.
+These results are provided in the appendix.
+
+%XXXX ST commented out after I realized I had made the same mistake as figure 7.
 XX This could be an indication that, despite the highly significant nature of the parameter $\beta_3$ according to the $t$-tests in \texttt{RSiena}, the transitive triplet parameter should not be included because it does not result in a structure similar to the data.
@@
 
-\st{As mentioned in Section~\ref{sec:models}, we fit the models M1, M2, and M3 to both the small friendship network and the senate collaboration network.
+As mentioned in Section~\ref{sec:models}, we fit the models M1, M2, and M3 to both the small friendship network and the senate collaboration network.
 We fit each model to the friendship data 1,000 times, and to the senate data 100 times.
-The means and standard deviations of the parameter estimates for each combination of model and data are given in Table~\ref{tab:meanests} }
+The means and standard deviations of the parameter estimates for each combination of model and data are given in Table~\ref{tab:meanests}, while the density plots of each parameter in the model for each data set is given in Figure~\ref{fig:compareModels}.
 
@@
 
-<<compareModels, fig.cap="Density plots of the repeated estimates from fitting models M1, M2, and M3 to the friendship and senate example data.
-Note that the rate parameters are excluded since the two data sets have different numbers of waves.
-The first two parameters, outdegree and reciprocity, have the same relationship for both data sets: the outdegree parameter estimate is strongly negative, while the reciprocity estimate is strongly postitive.
+<<compareModels, fig.cap="Density plots of the objective function estimates from repeatedly fitting models M1, M2, and M3 to the friendship and senate example data.
+The first two parameters, outdegree and reciprocity, have the same inverse relationship for both data sets.
 For the friendship data, the inclusion of the transitive triplet parameter has strong effect on the estimates of the other two parameters, while it does not affect the estimates of the other two parameters for the senate data.", fig.height = 3>>=
@@
 
-\st{Looking at Figure~\ref{fig:compareModels} and Table~\ref{tab:meanests}, we see a few patterns in the estimates from both models.
-First, we can see in both the table and the density of the estimates that the same relationship between the outdegree parameter, $\beta_1$, and the reciprocity parameter, $\beta_2$.
+Looking at Figure~\ref{fig:compareModels} and Table~\ref{tab:meanests}, we see a few patterns in the estimates from both models.
+First, the table and the density plots demonstrate the same relationship between the outdegree parameter, $\beta_1$, and the reciprocity parameter, $\beta_2$.
 In both data sets and across all three models, the estimates of $\beta_1$ are all negative and hover between -5 and -3, while the estimates of $\beta_2$ are all positive and hover between four and five.
-This suggests that in both data sets, nodes are \textit{discouraged} from forming ties that are outgoing without being reciprocated, while also being \textit{encouraged} to form outgoing ties to nodes that tie to them.
-In both data sets, people seem to want to have reciprocated relationships: teenage girls want be friends with other girls that reciprocate their friendship, and senators want to coauthor bills with senators who have also been coauthors on their bills.
+This suggests that in both data sets, nodes are \textit{encouraged} to form outgoing ties that are reciprocated and \textit{discouraged} from forming ties that are not.
+People seem to want to have reciprocated relationships: teenage girls want be friends with other girls that reciprocate their friendship, and senators want to coauthor bills with senators who have also been coauthors on their bills.
 We explore the relationship between $\beta_1$ and $\beta_2$ further in Section~\ref{sec:samemodsamedat}.
@@
 The number of bills authored by a senator varies wildly, from no bills to 114 bills authored in two years, so this effect could simply be nonsensical for the senate data, since senators are less likely to have the same number of bills authors than teenage girls are to have the same drinking behavior.
-Looking at the estimates of $\beta_4$, we see that the estimates for the senate data are near zero, suggesting this effect, which considers indirect ties, is not important for the senate data: indirect relationships do not describe the senate collaboration structure as much as they do the teenage friendship structure.
-}
+Looking at the estimates of $\beta_4$, we see that the estimates for the senate data are near zero, suggesting that this effect, which considers indirect ties, is not important for the senate data.
+It is, however, much stronger in the friendship data, suggesting that indirect ties are discouraged from forming: teenage girls want to have direct friendships.
+This indirect relationship variable does not impact the senate collaboration structure, but it does affect teenage friendship structure.
 
@@
 We can see from these distributions that the inclusion of the jumping transitive triplet parameter, $\beta_3$ is obviously affecting the distributions of the other four parameters included in all models, $\alpha_1$, $\alpha_2$, $\beta_1$, and $\beta_2$.
-\st{When $\beta_3$ is included, its estimate is postitive, meaning that friendships between two girls with different drinking behaviors tend to form when there is an intermediary who is already friends with the two girls.
+When $\beta_3$ is included, its estimate is postitive, meaning that friendships between two girls with different drinking behaviors tend to form when there is an intermediary who is already friends with the two girls.
 The inclusion of this parameter leads to increases in the rate parameters' estimates, suggesting that encouraging the transitive triplet behavior means that the girls would also change friends more frequently.
@@
 This implies the girls in the data prefer to form closer friend groups, as indicated by reciprocated ties and jumping transitive triplet formation, as opposed to being popular and having many friends.
-Having many friends who do not reciprocate is discouraged by M2.} In comparison with models M1 and M3, model M2 typically has higher estimates of the rate parameter, meaning that the inclusion of the covariate statistic in the model leads to higher estimates of the number of times, on average, a node gets to change its ties.
+Having many friends who do not reciprocate is more stronlgy discouraged by M2 than in the other models.
+In comparison with models M1 and M3, model M2 typically has higher estimates of the rate parameter, meaning that the inclusion of the covariate statistic in the model leads to higher estimates of the number of times, on average, a node gets to change its ties.
 It is not clear, however, that the addition of a parameter to the objective function \textit{should} effect the estimation of the rate parameters, so we continue to explore the collection of parameter estimates.
@@
 
-<<corplots, fig.cap="A matrix of plots demonstrating the strong correlations between parameter estimate in our SAOMs.
-The strongest correlation within each model is between $\\beta_1$ and $\\beta_2$.", fig.width=8>>=
-
-#simu2 <- read.csv("data/simulation-1000-M1-M2-M3.csv")
-simu_spread <- simu2 %>% spread(parameter, estimate)
-ggpairs(simu_spread, columns = 3:8, ggplot2::aes(colour=Model, alpha = .5)) + theme(plot.title = element_text(size = 12,
-                                            face = 'plain', angle = 0, family = "Times New Roman"),
-                  axis.title.x = element_text(size = 12,
-                                            face = 'plain', angle = 0, family = "Times New Roman", inherit.blank = T),
-                  axis.title.y = element_text(size = 12,
-                                            face = 'plain', angle = 90, inherit.blank = T),
-                  axis.text.x.top = element_text(size = 12,
-                                            face = 'plain', angle = 90, inherit.blank = TRUE),
-                  axis.text.x.bottom = element_text(size = 12,
-                                            face = 'plain', angle = 90, inherit.blank = TRUE),
-                  axis.text.y.left = element_text(size = 12,
-                                            face = 'plain', angle = 0, family = "Times New Roman", inherit.blank = T),
-                  axis.text.y.right = element_text(size = 12,
-                                            face = 'plain', angle = 0, family = "Times New Roman", inherit.blank = T),
-                  strip.text.x = element_text(size = 14,
-                                            face = 'plain', angle = 0, family = "Times New Roman",
-                                            margin = margin(t = 3, r = 0, b = 3, l = 0, unit = "pt"), inherit.blank = T),
-                  strip.text.y =element_text(size = 14,
-                                            face = 'plain', family = "Times New Roman", angle = 90,margin = margin(t = 3, r = 3, b = 3, l = 3, unit = "pt"),inherit.blank = T),
-                  strip.background = element_rect(colour = "black", fill = "white")
-                    )
-@
-
-To further investigate the odd relationship between the parameter values, we look at correlations between each of the parameter estimates in each model.
-In Figure~\ref{fig:corplots}, we examine correlations between each of pair of parameters within each model and overall.\nocite{ggally} The strongest correlation within each model is between $\beta_1$ and $\beta_2$, with absolute value of correlation between those two parameter values greater than 0.90 in all three models.
-The $\beta_1$ parameter is also highly correlated with the $\beta_3$ parameter within model M2, but it is not as highly correlated with the $\beta_4$ parameter in model M3.
-\hh{It might therefore be advisable to consider only models that either allow $\beta_1$, or $\beta_2$.
-Looking at the high correlation with $\alpha$, we might swiths to a model without $\beta_1$.}
+Finally, to further investigate the odd relationship between the parameter values, we look at correlations between each of the parameter estimates in each model.
+These results are presented in the Appendix.
 
@@
 
-\st{The last principle of model visualization is to explore the process of fitting the model, instead of just focusing on the end result.
-This principle is perhaps the most important for SAOMs because} the model fitting process in \texttt{RSiena} involves several simulation steps that are hidden from the user.
-Hiding the MCMC steps is practical and efficient if a researcher is primarily interested in fitting one model to a set of longitudinal network data, obtaining parameter estimates, \st{and drawing conclusions or making predictions}.
+The last principle of model visualization we use is exploring the process of fitting the model, instead of just focusing on the end result.
+This principle is perhaps the most important for SAOMs because the model fitting process in \texttt{RSiena} involves several simulation steps that are hidden from the user.
+Hiding the MCMC steps is practical and efficient if a researcher is primarily interested in fitting one model to a set of longitudinal network data, obtaining parameter estimates, and drawing conclusions or making predictions.
 We are more interested in \textit{how} the models are fit, so we extracted and explored the different steps of that process instead of allowing them stay hidden.
@@
 The options for change are (1) removing a current tie, (2) adding a new tie, or (3) making no change at all.
-Saving and exploring all of these steps is not computationally efficient if one is only interested in estimating parameter values, but they can be saved and extracted using options in \texttt{RSiena}, which is what we did to create our visualizations.
 Between two network observations $x(t_m)$ and $x(t_{m+1})$, there can be dozens, hundreds, or even thousands of microsteps, depending on the size of the network and the number of changes between two network observations.
@@
 If there are no changes at a particular microstep, no changes are seen.
-Some frames of the animations are shown in Figure~\ref{fig:makinggif}, and the full movie can be viewed at \url{https://vimeo.com/240089108}.
-
-<<makinggif, fig.cap='A selection of images in the microstep animation.
-The selected edges and nodes are emphasized by changing the size and the color, then when edges are removed, they fade out, shrinking in size, while the nodes change color and shrink to blend in with the other nodes.', message = FALSE, warning = FALSE, fig.height=2.5>>=
-# Need old version of dplyr to get this to work:
-# if (!require("dplyr050",character.only = TRUE)) {
-#       devtools::install_github("sctyner/dplyr050")
-# }
-# library(dplyr050)
-friend.data.w1 <- as.matrix(read.table("data/s50_data/s50-network1.dat"))
-friend.data.w2 <- as.matrix(read.table("data/s50_data/s50-network2.dat"))
-friend.data.w3 <- as.matrix(read.table("data/s50_data/s50-network3.dat"))
-drink <- as.matrix(read.table("data/s50_data/s50-alcohol.dat"))
-fd2.w1 <- friend.data.w1[20:35,20:35]
-fd2.w2 <- friend.data.w2[20:35,20:35]
-fd2.w3 <- friend.data.w3[20:35,20:35]
-friendshipData <- array(c(fd2.w1, fd2.w2,fd2.w3), dim = c(16, 16, 3))
-
-# running Siena
-#friendship <- sienaDependent(friendshipData)
-#alcohol <- varCovar(drink[20:35,])
-#mydata <- sienaDataCreate(friendship, alcohol)
-#myeffnull <- getEffects(mydata)
-#myalgorithm <- sienaAlgorithmCreate(projname = 's16_3')
-library(dplyr)
-set.seed(823746)
-load("data/ansnullpaper.rda")
-ansnullchains <- get_chain_info(ansnull)
-ansnullchains %>%
-  dplyr::filter(period == 1) %>%  #only look at chains from wave 1 to wave 2
-  #dplyr::group_by(rep) %>%
-  dplyr::select(rep, from = ego, to = alter) %>%
-  dplyr::mutate(val = as.numeric(!from == to),
-         from = paste0("V", parse_number(from)+1), # make the chains
-         to = paste0("V", parse_number(to)+1)) -> ansnullchainsw1w2
-colnames(fd2.w1) <- paste0("V", 1:16)
-rownames(fd2.w1) <- paste0("V", 1:16)
-wave1friends <- fortify(as.adjmat(fd2.w1))
-ms1 <- netvizinf::listMicrosteps(dat = wave1friends,
-                      microsteps = dplyr::filter(ansnullchainsw1w2, rep == 1))
-microsteps <- ms1[1:20]
-
-microsteps_df <- getMicrostepsDF(microsteps)
-pte <- pretween_edges(microsteps = microsteps)
-ptv <- pretween_vertices(pte = pte, layoutparams = list(n = 16))
-# step 1 is remove, step 9 is add.
-# get tweened data for all microsteps
-pall <- tween_microsteps(ptv, pte, microsteps_df)
-
-static_pall <- dplyr::filter(pall, .frame %in% c(15,20,22,24,26:28, 71:73, 75, 79, 81, 84))
-
-static_pall$type <- NA
-static_pall$type[which(static_pall$.frame %in% c(15,20,22,24,26:28))] <- "Remove Edge"
-static_pall$type[which(static_pall$.frame %in% c(71:73, 75, 79, 81, 84))] <- "Add Edge"
-
-set.seed(34569234)
-froms <- static_pall %>% dplyr::select(id = from, microstep, x = from.x,
-        y = from.y, ms, .frame, size, color) %>% unique
-tos <- static_pall %>% dplyr::select(id = to, microstep, x = to.x, y = to.y,
-        ms, .frame) %>% unique
-tos <- dplyr::left_join(tos, froms)
-nodedata <- rbind(froms, tos) %>% unique
-nodedata <- nodedata %>% tidyr::replace_na(replace = list(size = 1,
-        color = "#333333"))
-to_ani <- ggplot() +
-  geom_curve(data = static_pall, aes(x = from.x,
-        y = from.y, xend = to.x, yend = to.y, frame = .frame,
-        colour = ecolor, size = esize), curvature = 0.1, arrow = arrow(angle = 20,
-        length = unit(0.1, "inches"))) +
-  geom_point(data = nodedata,
-        aes(x = x, y = y, colour = color, size = size, frame = .frame)) +
-  scale_colour_identity() + scale_size_identity() + theme_void() +
-        coord_fixed()
-to_ani +
-  facet_wrap(~.frame, nrow=2, labeller = 'label_both') +
-  ThemeNet +
-  theme(panel.background = element_rect(color = 'black'))
-@
-
-The top row of Figure~\ref{fig:makinggif} shows an edge being removed, and the bottom row shows one being added.
-In both cases, the ego node chosen to act change color from black to red, and they also increase greatly in size.
-In the case of an edge being removed, in the top row, the edge that currently exists is emphasized with the same color and size change that the node gets, and as the animation proceeds the edge shrinks to nothing, as the ego node shrinks and changes color back to its original black.
-If an edge is being added, as in the bottom row of the figure, the ego node's appearance changes in the same way as when the edge is being removed, while the edge appears colored red from nothing, and grows to a large size, then changes color and size to match the rest of the edges, while the node shrinks and changes color to match the other nodes.
-
-\st{In the network animation, we see the possible steps of the unobserved CTMC process that is underlying the SAOM fit to the data.
+Some frames of the animations are shown in Figure~\ref{fig:makinggif} in the Appendix, and the full movie can be viewed at \url{https://vimeo.com/240089108}.
+
+In the network animation, we see the possible steps of the unobserved CTMC process that is underlying the SAOM fit to the data.
 We see each part of the model come into play.
@@
 Then, we see the result of the actor maximizing its objective function by either deleting or adding a node.
-In addition, the layout of the nodes changes as edges are removed or added, which gives us a better sense of how the overall network structure changes with these individual tie changes.}
+In addition, the layout of the nodes changes as edges are removed or added, which gives us a better sense of how the overall network structure changes with these individual tie changes.
 
@@
 # reinstall new dplyr
-numordersf <- data.frame(fd2.w1) %>%
-  mutate(from = rownames(fd2.w1)) %>%
+fd.w12 <- data.frame(fd2.w1)
+numordersf <- fd.w12 %>%
+  mutate(from = names(fd.w12)) %>%
   gather(key = to ,value = value, -from)
+
 pcaordersf <- numordersf
 
-numordersf$from <- factor(numordersf$from, levels = paste0("V", 1:16))
-numordersf$to <- factor(numordersf$to, levels = rev(paste0("V", 1:16)))
-
-p1 <- ggplot(data = numordersf, aes(y = from, x = to)) + geom_tile(color = 'black', aes(fill = as.factor(value))) +
+numordersf$from <- factor(numordersf$from)
+levels(numordersf$from) <-  1:16
+numordersf$to <- factor(numordersf$to)
+levels(numordersf$to) <- 1:16
+numordersf$from <- as.numeric(as.character(numordersf$from))
+numordersf$to <- as.numeric(as.character(numordersf$to))
+
+p1 <- ggplot(data = numordersf, aes(y = -from, x = to)) + geom_tile(color = 'black', aes(fill = as.factor(value))) +
   scale_fill_manual(values = c('white', 'black')) +
-  scale_x_discrete(position = "top", labels = as.character(1:16)) +
-  scale_y_discrete(labels = rev(as.character(1:16)))  +
+  scale_x_continuous(breaks = 1:16) +
+  scale_y_continuous(breaks = (1:16)*-1, labels =1:16)  +
   ThemeNoNet +
-  theme(legend.position = 'none') +
+  theme(legend.position = 'none', panel.grid = element_blank(), 
+        panel.border = element_blank(), 
+        panel.spacing = unit(rep(0,4),"cm")) +
   coord_fixed()
@@
 
-<<adjmatani, fig.cap='A selection of frames from the adjacency matrix visualization animation for one series of microsteps.
-(Ego and alter node labels are removed to declutter the graph.) At the beginning of the animation, shown in the top row, there are two clearly connected components: one in the top left corner, and one in the bottom right corner.
-By the end, the component in the top left has spread out, while the bottom right component has shrunk, but remains closely connected.', warning = FALSE, message = FALSE, results='hide', fig.height=2.5>>=
-adjmatall <- matrix(0, nrow = 16, ncol = 16)
-rownames(adjmatall) <- paste0("V", 1:16)
-colnames(adjmatall) <- paste0("V", 1:16)
-dat2 <- sfmsall %>% filter(rep == 1)
-adjmatall2 <- NULL
-for(i in 0:55){
-  MS <- dat2[dat2$ms == i,c("from", "to")]
-  myadjmat <- adjmatall
-  for(j in 1:nrow(MS)){
-    myadjmat[MS[j,"from"][[1]],MS[j,"to"][[1]]] <- 1
-  }
-  adjmatall2 <- rbind(adjmatall2, myadjmat)
-}
-
-adjmatall2 <- cbind(adjmatall2, rep(0:55, each = 16))
-adjmatall2 <- data.frame(adjmatall2)
-adjmatall2$from <- rep(paste0("V", 1:16), 56)
-names(adjmatall2)[17] <- "ms"
-adjmatall2 %>% gather(to, val, V1:V16) -> adjmatall3
-
-adjmatall3$to <- factor(adjmatall3$to, levels = order1)
-adjmatall3$from <- factor(adjmatall3$from, levels = rev(order1))
-
-adjmatall3$frame <- adjmatall3$ms
-#plots <- list()
-#for(i in 0:55){
-#plots[[i + 1]] <-
-
-dat3 <- adjmatall3 %>% filter(ms %in% c(0,2,4,6,8,10,12,43,45,47,49,51,53,55))
-
-dat4 <- dat3 %>% select(-ms) %>% spread(frame, val)
-dat5 <- dat4
-#p <- Sys.time()
-for(i in 4:16){
-  for(j in 1:nrow(dat5)){
-    if(dat4[j,i] != dat4[j,i-1]){
-      if(dat4[j,i] > dat4[j,i-1]){
-        dat5[j,i] <- 2
-      } else { dat5[j,i] <- 3}
-    }
-  }
-}
-#Sys.time() - p
-
-dat6 <- dat5 %>% gather(frame, val, 3:16) %>% mutate(frame = parse_number(frame))
-
-ggplot() +
-  geom_tile(data = dat6, aes(x = to, y = from, fill = as.factor(val)), color = '#969696') +
-  scale_fill_manual(values = c("white", "black", "red", "white")) +
-  geom_tile(data = dat6 %>% filter(val == 3), aes(x = to, y = from), fill = NA, color = 'red', size = .25) +
-  scale_x_discrete(position = "bottom") +
-  labs(x = "alter node (j)", y = "ego node (i)") +
-  ThemeNoNet +
-  theme(legend.position = 'none', axis.text.x.bottom = element_blank(),
-        axis.text.y.left = element_blank()) +
-  coord_fixed() +
-  facet_wrap(~frame, labeller = 'label_both', nrow = 2)
-#}
-#plots
-@
-
 Using PCA on $\mathbf{A}^{cum}$ to order the rows and columns of the adjacency matrix visualization clearly shows the two distinct connected components in the first wave of the network, which are difficult to find in the arbitrarily ordered visualization.
 We also use the PCA seriated layout to fix the layout in the animation of one of the microstep process simulations.
-This animation, some frames of which are shown in Figure~\ref{fig:adjmatani} is very simple: a square appears or disappears in the animation as that edge appears or disappears in the microstep process.
-Through this animation, we can see edges appearing, and then later on disappearing.
+This animation, some frames of which are shown in Figure~\ref{fig:adjmatani} of the Appendix is very simple: a square appears or disappears in the animation as that edge appears or disappears in the microstep process.
+Through this animation, which can be viewed at \url{https://vimeo.com/240092677}, we can see edges appearing, and then later on disappearing.
 These in-between steps are not shown when we look at the network at our discrete observation points, so by viewing this animation we can gain a better understanding of the underlying dynamics of this model.
-The full movie can be viewed at \url{https://vimeo.com/240092677}.
 
@@
 We also attempt to better understand the microstep process by visualizing the observed transition probabilites for the first microstep in the process.
-\st{We only do the first step of many because the \texttt{RSiena} transition probabilities for any two edges $i,j$ after the first step are only directly comparable for identical ego node states due to the conditioning on the current network state in the model.
-By ego node state, we mean the current set of incoming and outgoing ties to the ego node, $i$.
+We only do the first step of many because the \texttt{RSiena} transition probabilities for any two edges $i,j$ after the first step are only directly comparable for identical microsteps due to the conditioning on the current network state in the model.
+%By ego node state, we mean the current set of incoming and outgoing ties to the ego node, $i$.
 Put more precisely, let the vector $X_i = (\{x_{ik}\}, \{x_{ki}\})_{k \neq i} \in \{0,1\}^{2(n-1)}$ represent the set of all tie variables involving node $i$ in the current network.
@@
 Thus we have 1,000 transition probabilities to examine: one transition probability for the first microstep taken for each of the simulations.
-}
 %Put more precisely, let the vector $X_i = (x_{ij}|_{j \neq i}, x_{ji}|_{j \neq i}$ for $i \neq j \neq k \neq \ell \in \{1, \dots, n\}$, $p_{ij}$ is not necessarily the same \st{ARGH NOT SURE THIS IS TRUE } if the current state of the network achieved from the first microstep, denoted $X(1)$, is $X(1) = x(i \leadsto \ell)$ than is different in the case of $x_{ij}$ changing in the first step from the case of $x_{ij}$ not changing in the first step.
@@
 The empirical probability is calculated by counting all instances of the ego node first, then computing the proportion of each different alter node change.
-In most cases, they are almost identical, which demonstrates that the algorithm is performing about as expected.
-\st{There are, however, many steps that are never taken.
+In most cases, they are almost identical, which demonstrates that the algorithm is performing as expected.
+There are, however, many steps that are never taken.
 Some ego nodes, like 1, 4, and 16, really explore the space of all possible outgoing ties.
 On the other hand, nodes like 13 and 15 explore very few possible outgoing ties.
-It is unclear why this would be happening; it may have something to do with the ties in wave 1, where 13 and 15 have multiple connections, while 1 and 16 have none, but there are several other counterexamples of well connected nodes that explore the space more.
+It is unclear why this would be happening; it may have something to do with the ties in wave 1, where 13 and 15 have multiple connections, while 1 and 16 have none, but there are counterexamples of well connected nodes, such as 12, that explore the space more.
 In addition, we can see that the ties changed most often are removing ties, not adding ties.
-This tracks with the overall pattern of the data: we see the most ties in the first wave of data, and the least ties in the third wave.}
+This matches the overall pattern of the data: we see the most ties in the first wave of data, and the least ties in the third wave.
 
@@
 Finally, SAOMs and \texttt{RSiena} are able to also model behavior change of the actors in the network, which again is a capability we did not explore here.
+
+\section*{Appendix: Additional visualizations}
+
+<<sigtesting, fig.show = 'hold', out.width='\\textwidth', fig.height=2, fig.cap="From Section~\ref{sec:allpossible}.
+Significant effects for the two data sets, at a significance level of 0.10 or lower as calculated by the Wald-type test available in the SIENA software.">>=
+smfr_effs <- read_csv("data/effects_significance_smallFriends.csv")
+smfr_effs$inter1[is.na(smfr_effs$inter1)] <- "none"
+library(ggrepel)
+set.seed(12345678)
+smfr_effs %>% filter(type == "eval" & Waldpval <=0.10) %>%
+ggplot(aes(x = estimate, y = Waldpval, color = inter1)) +
+  #geom_errorbarh(aes(xmin = estimate - se, xmax = estimate + se), color = 'black') +
+  geom_vline(xintercept = 0) +
+  geom_point(size = 3) +
+  #geom_text_repel(aes(label = shortName)) +
+  scale_color_brewer(palette = "Paired", name = "Interaction\nVariable") +
+  ggtitle("Significant effects at level 0.10 for friendship data") +
+  coord_flip() +
+  ThemeNoNet
+sen_effs <- read_csv("data/congress/senateModelSigEffects.csv")
+#sen_effs$Waldpval <- parse_number(sen_effs$Waldpval)
+sen_effs$inter1[is.na(sen_effs$inter1)] <- "none"
+ggplot(data = sen_effs %>% filter(Waldpval <=0.10), aes(x = estimate, y = Waldpval, color = inter1)) +
+  # geom_errorbarh(aes(xmin = estimate - se, xmax = estimate + se), color = 'black') +
+  geom_vline(xintercept = 0) +
+  geom_point(size = 3) +
+  #geom_text_repel(aes(label = shortName)) +
+  scale_color_brewer(palette = "Paired", name = "Interaction\nVariable") +
+  ggtitle("Significant effects at level 0.10 for senate data") +
+  coord_flip() +
+  ThemeNoNet
+@
+
+In Figure~\ref{fig:sigtesting}, we see in both the friendship data and the senate data results that most of the significant effects have absolute value less than ten.
+In addition, the $p$-values for the effects from the friendship data are more spread out than the $p$-values for the senate data, which are concentrated at about 0.02 or less.
+This may suggest that larger data sets tend to result generally in smaller $p$-values, just like a larger sample size results in smaller $p$-values in a $t$-test.
+
+<<averagenet, results='hide', out.width='\\textwidth',fig.cap='The node-link diagrams from the three "average" networks that we calculated are in the top row, and the true wave 1 and wave 2 data are shown in the bottom row above.
+There is some difference between the three models, but overall, these three models cannot capture the structure in the true second wave of data.', fig.height=5>>=
+
+combinedavgs$linewidth[combinedavgs$cat %in% c("1st Wave", "2nd Wave")] <- .5
+combinedavgs$cat <- as.factor(combinedavgs$cat)
+combinedavgs$cat <- ordered(combinedavgs$cat, levels = c("Model1", "Model2", "Model3", "1st Wave", "2nd Wave"))
+
+ggplot(data = combinedavgs) +
+  geom_net(aes(from_id = from, to_id = to, linewidth= linewidth),
+           directed = T, curvature = .2, labelon=T, size = 5,
+           arrow = arrow(angle = 20, type = "open", length = unit(.2, "cm")), ecolour = "#969696", fiteach = T,
+           color = "#252525", hjust = .5, vjust = .5, labelcolour = 'white', fontsize = 4, arrowgap = .02,
+           singletons = T) +
+  ThemeNet + theme(panel.background = element_rect(color = 'black')) +
+  facet_wrap(~cat)
+@
+
+To create the average network visualization shown in Figure~\ref{fig:averagenet}, we follow the same procedure as in Section~\ref{sec:minds}, counting occurrences of each possible edge in the simulations, resulting in a summary network with weighted edges representing the number of times an edge appeared in the simulated wave 2 when simulating from the SAOM 1,000 times.
+As in Figure~\ref{fig:summnetM1}, edges only appear in the average networks if they appear more than 5\% of the time in the simulations.
+In Figure~\ref{fig:averagenet}, we show the ``average'' network from the three models we fit and the first and second waves of data.
+Comparing the three averages to waves 1 and 2, we see that they have very similar structure to wave 1.
+Model 2, which included the transitive triplet parameter, seems to have created a larger connected component overall than models 1 and 3.
+In particular, if we look at the group of nodes $\{10,11,14\}$, we see they are very strongly connected within the three average networks, and they are completely separate from the other nodes in the true wave 2.
+None of the three average networks show node 16 gaining ties as it does in wave two, nor do they show nodes 4 and 7 becoming isolated.
+In Model 2, however, the ties to node 7 appear much weaker than in Model 1 or Model 2, suggesting that of the three, Model 2 may be the best fit for our data.
+
+<<corplots, fig.cap="A matrix of plots demonstrating the strong correlations between parameter estimate in our SAOMs.
+The strongest correlation within each model is between $\\beta_1$ and $\\beta_2$.", fig.width=8>>=
+
+#simu2 <- read.csv("data/simulation-1000-M1-M2-M3.csv")
+simu_spread <- simu2 %>% spread(parameter, estimate)
+ggpairs(simu_spread, columns = 3:8, ggplot2::aes(colour=Model, alpha = .5)) + theme(plot.title = element_text(size = 12,
+                                            face = 'plain', angle = 0, family = "Times New Roman"),
+                  axis.title.x = element_text(size = 12,
+                                            face = 'plain', angle = 0, family = "Times New Roman", inherit.blank = T),
+                  axis.title.y = element_text(size = 12,
+                                            face = 'plain', angle = 90, inherit.blank = T),
+                  axis.text.x= element_text(size = 12,
+                                            face = 'plain', angle = 90, inherit.blank = TRUE),
+                 # axis.text.x.bottom = element_text(size = 12,
+             #                               face = 'plain', angle = 90, inherit.blank = TRUE),
+                  axis.text.y = element_text(size = 12,
+                                            face = 'plain', angle = 0, family = "Times New Roman", inherit.blank = T),
+                #  axis.text.y.right = element_text(size = 12,
+               #                             face = 'plain', angle = 0, family = "Times New Roman", inherit.blank = T),
+                  strip.text.x = element_text(size = 14,
+                                            face = 'plain', angle = 0, family = "Times New Roman",
+                                            margin = margin(t = 3, r = 0, b = 3, l = 0, unit = "pt"), inherit.blank = T),
+                  strip.text.y =element_text(size = 14,
+                                            face = 'plain', family = "Times New Roman", angle = 90,margin = margin(t = 3, r = 3, b = 3, l = 3, unit = "pt"),inherit.blank = T),
+                  strip.background = element_rect(colour = "black", fill = "white")
+                    )
+@
+
+In Figure~\ref{fig:corplots}, we examine correlations between each of pair of parameters within each model and overall.\nocite{ggally} The strongest correlation within each model is between $\beta_1$ and $\beta_2$, with absolute value of correlation between those two parameter values greater than 0.90 in all three models.
+The $\beta_1$ parameter is also highly correlated with the $\beta_3$ parameter within model M2, but it is not as highly correlated with the $\beta_4$ parameter in model M3.
+It might therefore be advisable to consider only models that either allow $\beta_1$, or $\beta_2$.
+Looking at the high correlation with $\alpha$, we might switch to a model without $\beta_1$.
+
+<<makinggif, fig.cap='A selection of images in the microstep animation.
+The selected edges and nodes are emphasized by changing the size and the color, then when edges are removed, they fade out, shrinking in size, while the nodes change color and shrink to blend in with the other nodes.', message = FALSE, warning = FALSE, fig.height=2.5>>=
+# Need old version of dplyr to get this to work:
+# if (!require("dplyr050",character.only = TRUE)) {
+#       devtools::install_github("sctyner/dplyr050")
+# }
+# library(dplyr050)
+friend.data.w1 <- as.matrix(read.table("data/s50_data/s50-network1.dat"))
+friend.data.w2 <- as.matrix(read.table("data/s50_data/s50-network2.dat"))
+friend.data.w3 <- as.matrix(read.table("data/s50_data/s50-network3.dat"))
+drink <- as.matrix(read.table("data/s50_data/s50-alcohol.dat"))
+fd2.w1 <- friend.data.w1[20:35,20:35]
+fd2.w2 <- friend.data.w2[20:35,20:35]
+fd2.w3 <- friend.data.w3[20:35,20:35]
+friendshipData <- array(c(fd2.w1, fd2.w2,fd2.w3), dim = c(16, 16, 3))
+
+# running Siena
+#friendship <- sienaDependent(friendshipData)
+#alcohol <- varCovar(drink[20:35,])
+#mydata <- sienaDataCreate(friendship, alcohol)
+#myeffnull <- getEffects(mydata)
+#myalgorithm <- sienaAlgorithmCreate(projname = 's16_3')
+library(dplyr)
+set.seed(823746)
+load("data/ansnullpaper.rda")
+ansnullchains <- get_chain_info(ansnull)
+ansnullchains %>%
+  dplyr::filter(period == 1) %>%  #only look at chains from wave 1 to wave 2
+  #dplyr::group_by(rep) %>%
+  dplyr::select(rep, from = ego, to = alter) %>%
+  dplyr::mutate(val = as.numeric(!from == to),
+         from = paste0("V", parse_number(from)+1), # make the chains
+         to = paste0("V", parse_number(to)+1)) -> ansnullchainsw1w2
+colnames(fd2.w1) <- paste0("V", 1:16)
+rownames(fd2.w1) <- paste0("V", 1:16)
+wave1friends <- fortify(as.adjmat(fd2.w1))
+ms1 <- netvizinf::listMicrosteps(dat = wave1friends,
+                      microsteps = dplyr::filter(ansnullchainsw1w2, rep == 1))
+microsteps <- ms1[1:20]
+
+microsteps_df <- getMicrostepsDF(microsteps)
+pte <- pretween_edges(microsteps = microsteps)
+ptv <- pretween_vertices(pte = pte, layoutparams = list(n = 16))
+# step 1 is remove, step 9 is add.
+# get tweened data for all microsteps
+pall <- tween_microsteps(ptv, pte, microsteps_df)
+
+static_pall <- dplyr::filter(pall, .frame %in% c(15,20,22,24,26:28, 71:73, 75, 79, 81, 84))
+
+static_pall$type <- NA
+static_pall$type[which(static_pall$.frame %in% c(15,20,22,24,26:28))] <- "Remove Edge"
+static_pall$type[which(static_pall$.frame %in% c(71:73, 75, 79, 81, 84))] <- "Add Edge"
+
+set.seed(34569234)
+froms <- static_pall %>% dplyr::select(id = from, microstep, x = from.x,
+        y = from.y, ms, .frame, size, color) %>% unique
+tos <- static_pall %>% dplyr::select(id = to, microstep, x = to.x, y = to.y,
+        ms, .frame) %>% unique
+tos <- dplyr::left_join(tos, froms)
+nodedata <- rbind(froms, tos) %>% unique
+nodedata <- nodedata %>% tidyr::replace_na(replace = list(size = 1,
+        color = "#333333"))
+to_ani <- ggplot() +
+  geom_curve(data = static_pall, aes(x = from.x,
+        y = from.y, xend = to.x, yend = to.y, frame = .frame,
+        colour = ecolor, size = esize), curvature = 0.1, arrow = arrow(angle = 20,
+        length = unit(0.1, "inches"))) +
+  geom_point(data = nodedata,
+        aes(x = x, y = y, colour = color, size = size, frame = .frame)) +
+  scale_colour_identity() + scale_size_identity() + theme_void() +
+        coord_fixed()
+to_ani +
+  facet_wrap(~.frame, nrow=2, labeller = 'label_both') +
+  ThemeNet +
+  theme(panel.background = element_rect(color = 'black'))
+@
+
+The top row of Figure~\ref{fig:makinggif} shows an edge being removed, and the bottom row shows one being added.
+In both cases, the ego node chosen to act change color from black to red, and they also increase greatly in size.
+In the case of an edge being removed, in the top row, the edge that currently exists is emphasized with the same color and size change that the node gets, and as the animation proceeds the edge shrinks to nothing, as the ego node shrinks and changes color back to its original black.
+If an edge is being added, as in the bottom row of the figure, the ego node's appearance changes in the same way as when the edge is being removed, while the edge appears colored red from nothing, and grows to a large size, then changes color and size to match the rest of the edges, while the node shrinks and changes color to match the other nodes.
+
+<<adjmatani, fig.cap='A selection of frames from the adjacency matrix visualization animation for one series of microsteps.
+(Ego and alter node labels are removed to declutter the graph.) At the beginning of the animation, shown in the top row, there are two clearly connected components: one in the top left corner, and one in the bottom right corner.
+By the end, the component in the top left has spread out, while the bottom right component has shrunk, but remains closely connected.', warning = FALSE, message = FALSE, results='hide', fig.height=2.5>>=
+adjmatall <- matrix(0, nrow = 16, ncol = 16)
+rownames(adjmatall) <- paste0("V", 1:16)
+colnames(adjmatall) <- paste0("V", 1:16)
+dat2 <- sfmsall %>% filter(rep == 1)
+adjmatall2 <- NULL
+for(i in 0:55){
+  MS <- dat2[dat2$ms == i,c("from", "to")]
+  myadjmat <- adjmatall
+  for(j in 1:nrow(MS)){
+    myadjmat[MS[j,"from"][[1]],MS[j,"to"][[1]]] <- 1
+  }
+  adjmatall2 <- rbind(adjmatall2, myadjmat)
+}
+
+adjmatall2 <- cbind(adjmatall2, rep(0:55, each = 16))
+adjmatall2 <- data.frame(adjmatall2)
+adjmatall2$from <- rep(paste0("V", 1:16), 56)
+names(adjmatall2)[17] <- "ms"
+adjmatall2 %>% gather(to, val, V1:V16) -> adjmatall3
+
+adjmatall3$to <- factor(adjmatall3$to, levels = order1)
+adjmatall3$from <- factor(adjmatall3$from, levels = rev(order1))
+
+adjmatall3$frame <- adjmatall3$ms
+#plots <- list()
+#for(i in 0:55){
+#plots[[i + 1]] <-
+
+dat3 <- adjmatall3 %>% filter(ms %in% c(0,2,4,6,8,10,12,43,45,47,49,51,53,55))
+
+dat4 <- dat3 %>% select(-ms) %>% spread(frame, val)
+dat5 <- dat4
+#p <- Sys.time()
+for(i in 4:16){
+  for(j in 1:nrow(dat5)){
+    if(dat4[j,i] != dat4[j,i-1]){
+      if(dat4[j,i] > dat4[j,i-1]){
+        dat5[j,i] <- 2
+      } else { dat5[j,i] <- 3}
+    }
+  }
+}
+#Sys.time() - p
+
+dat6 <- dat5 %>% gather(frame, val, 3:16) %>% mutate(frame = parse_number(frame))
+
+ggplot() +
+  geom_tile(data = dat6, aes(x = to, y = from, fill = as.factor(val)), color = '#969696') +
+  scale_fill_manual(values = c("white", "black", "red", "white")) +
+  geom_tile(data = dat6 %>% filter(val == 3), aes(x = to, y = from), fill = NA, color = 'red', size = .25) +
+  scale_x_discrete(position = "bottom") +
+  labs(x = "alter node (j)", y = "ego node (i)") +
+  ThemeNoNet +
+  theme(legend.position = 'none', axis.text.x.bottom = element_blank(),
+        axis.text.y = element_blank()) +
+  coord_fixed() +
+  facet_wrap(~frame, labeller = 'label_both', nrow = 2)
+#}
+#plots
+@

### 2018-05-29 paper-child.Rnw
it compiles now! :smile:

@@
 
-We first consider the 16 actor subset of the teenage friends and lifestyle data show in Figures \ref{fig:adjmatviz, fig:alldataviz}.
+We first consider the 16 actor subset of the teenage friends and lifestyle data show in Figure~\ref{fig:alldataviz}.
 To this data, we fit three different SAOMs.
@@
 levels(numordersf$to) <- 1:16
-numordersf$from <- as.numeric(as.character(numordersf$from))
-numordersf$to <- as.numeric(as.character(numordersf$to))
-
-p1 <- ggplot(data = numordersf, aes(y = -from, x = to)) + geom_tile(color = 'black', aes(fill = as.factor(value))) +
+
+p1 <- ggplot(data = numordersf, aes(y = rev(from), x = to)) + geom_tile(color = 'black', aes(fill = as.factor(value))) +
   scale_fill_manual(values = c('white', 'black')) +
-  scale_x_continuous(breaks = 1:16) +
-  scale_y_continuous(breaks = (1:16)*-1, labels =1:16)  +
+  scale_x_discrete(labels = 1:16, position = "top") +
+  scale_y_discrete(labels =16:1, name = "from")  +
   ThemeNoNet +
@@
         panel.border = element_blank(), 
-        panel.spacing = unit(rep(0,4),"cm")) +
+        axis.text.y = element_text(margin = margin(r=0))) +
   coord_fixed()
@@
 order1 <- names(sort(pca1$rotation[,1], decreasing = T))
-order2 <- names(sort(abs(pca1$rotation[,1]), decreasing = T))
-
-pcaordersf$from <- factor(pcaordersf$from, levels = rev(order1))
-pcaordersf$to <- factor(pcaordersf$to, levels = order1)
-
-p2 <- ggplot(data = pcaordersf, aes(x = to, y = from)) + geom_tile(color = 'black', aes(fill = as.factor(value))) +
+order1 <- gsub("V", "",order1)
+# order2 <- names(sort(abs(pca1$rotation[,1]), decreasing = T))
+
+pcaordersf$from <- as.factor(pcaordersf$from)
+levels(pcaordersf$from) <- 1:16
+pcaordersf$from <- factor(pcaordersf$from , levels = levels(pcaordersf$from)[match( order1,levels(pcaordersf$from) )])
+pcaordersf$to <- as.factor(pcaordersf$to)
+levels(pcaordersf$to) <- 1:16
+pcaordersf$to <- factor(pcaordersf$to , levels = levels(pcaordersf$to)[match( order1,levels(pcaordersf$to) )])
+
+p2 <- ggplot(data = pcaordersf, aes(x = to, y = from)) +
+  geom_tile(color = 'black', aes(fill = as.factor(value))) +
   scale_fill_manual(values = c('white', 'black')) +
-  scale_x_discrete(position = "top", labels = str_replace(order1, "V", "")) +
-  scale_y_discrete(labels = rev(str_replace(order1, "V", ""))) +
+  scale_x_discrete(position = "top") +
+  #scale_y_discrete(name = "from",labels = rev(str_replace(order1, "V", ""))) +
   ThemeNoNet +
@@
 The color of the line indicates whether the tie from the ego to the alter node is being added, removed, or if there is no change to the network in this step.">>=
+set.seed(823746)
+load("data/ansnullpaper.rda")
+ansnullchains <- get_chain_info(ansnull)
 ansnullchains %>% group_by(period, rep) %>%
@@
 changes <- cumprobs %>% dplyr::select(ego, alter) %>% unique
-
+wave1friends <- fortify(as.adjmat(fd2.w1))
 wave1friends2 <- wave1friends[,-3]
@@
   filter(msno == 1) %>%
-  mutate(ego = paste0("V",as.numeric(as.character(ego))+1),
-         alter = paste0("V",as.numeric(as.character(alter))+1),
+  mutate(ego = as.numeric(as.character(ego))+1,
+         alter = as.numeric(as.character(alter))+1,
          prob = exp(logAlterProb)) -> probsDat
@@
        aes(y = factor(ego, levels = order1),
-             x = factor(alter, levels = rev(order1)),
+             x = factor(alter, levels = order1),
              fill = prob)) +
@@
   labs(y = "ego", x = "alter")+
-  scale_x_discrete(position = 'top', labels = str_replace(order1, "V", "")) +
-  scale_y_discrete(labels = rev(str_replace(order1, "V", ""))) +
-  ThemeNoNet + theme(panel.grid = element_blank()) +
+  scale_x_discrete(position = 'top') +
+  #scale_y_discrete(labels = rev(str_replace(order1, "V", ""))) +
+  ThemeNoNet + 
+  theme(panel.grid = element_blank()) +
   coord_fixed()
@@
 names(smallfriends)[1:2] <- c("from", "to")
+ansnullchains %>%
+  dplyr::filter(period == 1) %>%  #only look at chains from wave 1 to wave 2
+  #dplyr::group_by(rep) %>%
+  dplyr::select(rep, from = ego, to = alter) %>%
+  dplyr::mutate(val = as.numeric(!from == to),
+         from = paste0("V", parse_number(from)+1), # make the chains
+         to = paste0("V", parse_number(to)+1)) -> ansnullchainsw1w2
 msall1 <- netvizinf::listMicrosteps(dat = wave1friends,
@@
 <<sigtesting, fig.show = 'hold', out.width='\\textwidth', fig.height=2, fig.cap="From Section~\ref{sec:allpossible}.
-Significant effects for the two data sets, at a significance level of 0.10 or lower as calculated by the Wald-type test available in the SIENA software.">>=
+Significant effects for the two data sets, at a significance level of 0.10 or lower as calculated by the Wald-type test available in the SIENA software.", error=FALSE>>=
 smfr_effs <- read_csv("data/effects_significance_smallFriends.csv")
@@
 adjmatall2 <- cbind(adjmatall2, rep(0:55, each = 16))
-adjmatall2 <- data.frame(adjmatall2)
+rownames(adjmatall2) <- NULL
+adjmatall2 <- as.data.frame(adjmatall2)
 adjmatall2$from <- rep(paste0("V", 1:16), 56)
@@
 
+adjmatall3$to <- as.factor(adjmatall3$to)
+levels(adjmatall3$to) <- gsub("V", "", levels(adjmatall3$to))
+adjmatall3$from <- as.factor(adjmatall3$from)
+levels(adjmatall3$from) <- gsub("V", "", levels(adjmatall3$from))
 adjmatall3$to <- factor(adjmatall3$to, levels = order1)
-adjmatall3$from <- factor(adjmatall3$from, levels = rev(order1))
+adjmatall3$from <- factor(adjmatall3$from, levels = order1)
 

### 2018-05-29 paper-child.Rnw
still needs some trimming :scissors:

@@
 This simulation process follows the steps of the continuous-time Markov chain.
-\st{Each tie change in the CTMC is referred to as one ``microstep".
-At each microstep}, an ``ego node" is selected to make a change, and the chosen ego node randomly makes one change in its ties according to the probabilities, $\{p_{ij}: i\neq j \in \{1, \dots, n\}\}$ defined in Equation~\ref{eq:pij}, determined by its objective function.
+Each tie change in the CTMC is referred to as one ``microstep".
+At each microstep, an ``ego node" is selected to make a change, and the chosen ego node randomly makes one change in its ties according to the probabilities, $\{p_{ij}: i\neq j \in \{1, \dots, n\}\}$ defined in Equation~\ref{eq:pij}, determined by its objective function.
 The options for change are (1) removing a current tie, (2) adding a new tie, or (3) making no change at all.
 Between two network observations $x(t_m)$ and $x(t_{m+1})$, there can be dozens, hundreds, or even thousands of microsteps, depending on the size of the network and the number of changes between two network observations.
-We wanted to view these in-between steps in order to better understand the behavior of the underlying continuous-time Markov chain.
+We want to view these in-between steps in order to better understand the behavior of the underlying continuous-time Markov chain.
 
@@
 <<histogramalters, fig.width = 7, fig.height=9, fig.cap="Each panel shows the theoretical (as lines) and empirical (as points) probabilities of the chosen ego node changing its tie to each of the other nodes.
-The color of the line indicates whether the tie from the ego to the alter node is being added, removed, or if there is no change to the network in this step.">>=
+The color of the line indicates whether the tie from the ego to the alter node is being added, removed, or if there is no change to the network in this step.", eval = FALSE>>=
 set.seed(823746)
@@
 %Put more precisely, let the vector $X_i = (x_{ij}|_{j \neq i}, x_{ji}|_{j \neq i}$ for $i \neq j \neq k \neq \ell \in \{1, \dots, n\}$, $p_{ij}$ is not necessarily the same \st{ARGH NOT SURE THIS IS TRUE } if the current state of the network achieved from the first microstep, denoted $X(1)$, is $X(1) = x(i \leadsto \ell)$ than is different in the case of $x_{ij}$ changing in the first step from the case of $x_{ij}$ not changing in the first step.
-In Figure~\ref{fig:histogramalters}, we present each ego node and the resulting probabilities of tie changes.
+%In Figure~\ref{fig:histogramalters}, we present each ego node and the resulting probabilities of tie changes.
 The probability shown by the bars is the theoretical probability, according to the objective function, of the ego node changing its tie to the alter node, while the probability shown by the points is the empirical probability of that change being made.
@@
 
-Another way to view these transition probabilities is through the adjacency matrix visualization.
+%Another way to view these transition probabilities is through the adjacency matrix visualization. 
 In Figure~\ref{fig:heatmap}, we build on the concept of the ordered adjacency matrix of Figure~\ref{fig:adjmatorder}.
@@
 of the $16^2 = 256$ possible steps for the CTMC to take, only 103, or about 40\%, are taken in the 1,000 simulated chains.
-\st{This reinforces what we saw in Figure~\ref{fig:histogramalters}, where there are many paths not taken.
-This effect could only be exacerbated as more steps are taken in the CTMC, leading to a very large area of our network space, $\mathcal{X}$, completely unexplored by the SAOM model fitting process.}
+This leaves %reinforces what we saw in Figure~\ref{fig:histogramalters}, where there are many paths not taken.
+This effect could only be exacerbated as more steps are taken in the CTMC, leading to
+a very large area of our network space, $\mathcal{X}$, completely unexplored by the SAOM model fitting process.
 
@@
 The $x$ axis is the microstep number, with step 0 representing the first wave of data and the final step representing the second wave of data.
-We can see that many edges are underrepresented in this process: they are in the second wave, but never appear in the microsteps.", fig.height=2.5>>=
+We can see that many edges are underrepresented in this process: they are in the second wave, but never appear in the microsteps.", fig.height=2.5, eval = FALSE>>=
 smallfriends <- read_csv("data/smallfriends4Geomnet.csv")
@@
 
-We also wanted to better understand the entire microstep process from the first wave, all the way through to the second wave.
+%We also wanted to better understand the entire microstep process from the first wave, all the way through to the second wave.
 The number of steps taken from wave 1 to wave 2 varies.
@@
 
-We also combine 1,000 simulations from model M1 into a visualization like the one shown in Figure~\ref{fig:allsteps}.
-We first assign each possible edge an edge ID number so that we can keep track of it throughout all the microsteps and all the simlations.
+Finally, we combine 1,000 simulations from model M1 into a visualization that displays the entire microstep process for all simulations in Figure ~\ref{fig:alledgests}.
+To make this visualization, we first assign each possible edge an edge ID number so that we can keep track of it throughout all the microsteps and all the simlations.
 Then, we count up the total number of times each edge appears in the microstep process in each of the 1,000 simulations for use as an ordering variable later.
@@
 Next, we compute a proportion, which we call the occurrence percentage, which is the number of times the edge occurred in a microstep divided by 1,000.
-Finally, we visualize all this information together in Figure ~\ref{fig:alledgests}.
+Finally, we visualize all this information together in Figure~\ref{fig:alledgests}.
 In this plot, all possible edges are shown, and we see that every one of the $16 \times 15 = 240$ possible edges in the network occurs at some point in the simulation process.
@@
 But, about half of the edges in wave two are in the bottom half when ordered by number of occurrences, meaning they do not appear as much as they would if the model was truly excellent at capturing the mechanisms of tie change in the network.
-\st{This solidifies what we found in Figures~\ref{fig:histogramalters} and~\ref{fig:heatmap}: the model M1 and the SAOM fitting process do not explore the data space enough to adequately capture the network change mechanism.} %In addition, there is not a distinct pattern or structure that we can see in this particular edge visualization.
+This solidifies what we found in Figure~\ref{fig:heatmap}: the model M1 and the SAOM fitting process do not explore the data space enough to adequately capture the network change mechanism.
+%In addition, there is not a distinct pattern or structure that we can see in this particular edge visualization.
 The edges that the model starts with appear more often in the first 50 or so microsteps than any other set of edges.
@@
 
-\hh{The visualizations should address most elements of structure described before, i.e.
+%\hh{The visualizations should address most elements of structure described before, i.e.
 follow along in the setup (for descriptive and teaching purposes), the fitting (understanding the model and interpreting the results) and then diagnostics (how well does the model fit the data - where are the most differences to the actual data/what are the sensitive parameters?) }
@@
 By looking at the underlying algorithms, visualizing collections of these models, and viewing the model in the data space, we have been able to gain knowledge and appreciation for these complicated models and everything that goes into them.
-
-\st{XX HH help please!}
 

### 2018-05-30 paper-child.Rnw
still need to :scissors: a lot.

@@
  The waiting time to the next change opportunity by \textit{any} actor in the network is also exponentially distributed with expected value $(n\alpha_m)^{-1}$.
-The distribution of waiting time for the whole network to change, $\tau(x|m) = \sum_i \tau_i(m) |x_{i1}(m), \dots, x_{in}(m) $ can then be written as
-
-\begin{equation} \label{eq:c2}
- \tau(x|m) \sim Exp(n\alpha_m)
+%The distribution of waiting time for the whole network to change, $\tau(x|m) = \sum_i \tau_i(m) |x_{i1}(m), \dots, x_{in}(m) $ can then be written as
+%\begin{equation} \label{eq:c2}
+% \tau(x|m) \sim Exp(n\alpha_m)
+%\end{equation}
+The parameter for the wait time for the whole network $n\alpha_m$ is the rate at which any tie change occurs.
+The estimation of this parameter is straightforward: the method of moments is used to estimate the rate with the statistic
+\begin{equation}\label{eq:cdiff}
+C = \sum_i\sum_j |x_{ij}(t_{m+1}) - x_{ij}(t_m)|
 \end{equation}
-
-The parameter for the wait time for the whole network $n\alpha_m$ is the rate at which any tie change occurs.
-The estimation of this parameter is straightforward: a the method of moments is used to estimate the rate with the statistic $$C = \sum_i\sum_j |x_{ij}(t_{m+1}) - x_{ij}(t_m)|$$ which is the total number of changes from observation at time $t_m$ to the observation at time $t_{m+1}$.
-%In order to achieve the memorylessness property of a Markov process,
+which is the total number of changes from observation at time $t_m$ to the observation at time $t_{m+1}$. %In order to achieve the memorylessness property of a Markov process,
 
@@
 When node $i$ is given the chance to change a tie, %\hh{`we assume that they wish' sounds strange - don't assume wishes, state assumptions}
-it attempts to maximize the value of its objective function $f_i(\boldsymbol{\beta}, x)$ %as well as a random element, $U_i(x)$, to account for unknown attraction between nodes.
+it acts to maximize the value of its objective function $f_i(\boldsymbol{\beta}, x)$ %as well as a random element, $U_i(x)$, to account for unknown attraction between nodes.
 The additional random element is included to account for any random, unexplainable change in the network ties.
@@
 %This distribution, which is also known as the log-Weibull distribution, has probability distribution function of:
-
 %\begin{equation} \label{eq:dist1}
@@
 %using $\mu$ for the mean parameter and $\sigma$ for the scale parameter.
-
 <<logweibull, eval = FALSE, fig.align='center', out.width = '.5\\textwidth', fig.height = 4, fig.cap='The probability distribution function for the type 1 extreme value distribution, also known as the log-Weibull or Gumbel distribution with location parameter $ \\mu = 0$ and scale parameter $ \\sigma = 0$.'>>=
@@
 @
-
 %The probability density function for the type 1 extreme value distribution is shown in Figure~\ref{fig:logweibull}.
@@
 The covariate groups are determined by the student's drinking behavior, the values of which are converted to numeric then mean-centered.
-The four values in our data are $\{\text{never, 1-2 times per year, once per month, once per week}\}$, which after being converted to numeric and mean-centered become $\{-0.8125,-1.8125,0.1875,1.1875\}$.
+The four values in our data are $\{\text{never, 1-2 times per year, once per month, once per week}\}$.%, which after being converted to numeric and mean-centered become $\{-0.8125,-1.8125,0.1875,1.1875\}$.
 The two additional effects are visually represented and further described in Figure~\ref{fig:jtt} and Figure~\ref{fig:dad}, respectively.
@@
 A series of microsteps is obtained by simulating from the model in its current state, $x(t_m)$ with current parameter values $\mathbf{\theta}_{0} = \{\alpha_{1_0}, \dots \alpha_{{m-1}_0}, \beta_{1_0}, \dots, \beta_{K_0}\}$, to the next state, $x(t_{m+1})$.
-This microstep process stops when the simulated network has achieved the same number of differences, $C$, from $x(t_m)$ as $x(t_{m+1})$, where
-$$C = \sum_{i \neq j} |x_{ij}(t_{m+1}) - x_{ij}(t_{m})|.$$
+This microstep process stops when the simulated network has achieved the same number of differences, $C$, from $x(t_m)$ as $x(t_{m+1})$, where $C$ is definined in Equation~\ref{eq:cdiff}.
+
 This simulation process follows the steps of the continuous-time Markov chain.
@@
 We also attempt to better understand the microstep process by visualizing the observed transition probabilites for the first microstep in the process.
+These probabilities are defined in Equation~\ref{eq:pij}.
 We only do the first step of many because the \texttt{RSiena} transition probabilities for any two edges $i,j$ after the first step are only directly comparable for identical microsteps due to the conditioning on the current network state in the model.
@@
 There were many ties with empirical probability zero.', fig.height=2.5>>=
+library(dplyr)
+set.seed(823746)
+load("data/ansnullpaper.rda")
+ansnullchains <- get_chain_info(ansnull)
 ansnullchains %>%
@@
 #myalgorithm <- sienaAlgorithmCreate(projname = 's16_3')
-library(dplyr)
-set.seed(823746)
-load("data/ansnullpaper.rda")
-ansnullchains <- get_chain_info(ansnull)
 ansnullchains %>%

### 2018-07-11 paper-child.Rnw
ugh get this done

@@
 library(GGally)
-#devtools::install_github("sctyner/netvizinf")
+# devtools::install_github("sctyner/netvizinf")
 library(netvizinf)
@@
 \section{Introduction}
-% \st{
-% \begin{itemize}
-% %\item What? - model visualization and SAOMs. briefly dsecribe what these things are
-% %\item Why? - bring light to the RSiena blackbox. Need ways to visualize network models, collections of networks.
-% \item How?
-% \end{itemize}
-% %also talk about other dynamic network models briefly to put SAOMs in context.
-% }
-%\hh{XXX - talk a bit about social networks first, why we want to model them, and that usually the models are not dynamic.}
-%XXX - okay, here is what i have done for that:
-Social networks, such as collaboration networks between academic researchers, friendship networks in a school or university, and trade networks between nations have been studied for decades.  
-%beginning with a few foundational works, including the 1967 study, "The Small World Problem" by Stanley Milgram \citep{goldenberg09}. Examples of . 
-In recent years, the study of social networks has grown in popularity due to an increase in the availability and access to social network data.
-%There are many kinds of social networks, but there are not as many statistical models for social network data.
-Some network models that have been applied to social networks include the exponential random graph model (ERGM) and latent space models.
-These models, however, are only for single instance networks.
-If we only have one network observation, or only care about one state of a complex network in time, like a snapshot of the World Wide Web, using the well-established models for single network observating is not a problem.
-If, on the other hand, we have many observations of social network over time, these models may not be appropriate because they are not able to model network change as time passes.
+
+Social networks, such as collaboration networks between academic researchers or friendship networks in a school or university have been studied for decades.
+Statistical models such as exponential random graph models (ERGM) and latent space models (LSM) are often used when studying one observation of a network when trying to model its structure.
+Social networks can also be observed at many points in time, and ERGM or LSM may not be appropriate for modelling the changes because they do not model network change as time passes.
 When studying \textit{dynamic networks}, networks observed at many time points, we need a model that allows for variation in the network over time.
-Models for dynamic social networks have a great deal of potential because of how realistic they can be.
-A social network does not form spontaneously: it \textit{evolves} over time.
-Ties are formed and dissolved, and new actors join the social structure.
+Models for dynamic social networks hold a great deal of potential because of how realistic they can be.
+Social networks do not form spontaneously, but rather they \textit{evolve} over time as edges are formed and dissolved, and new nodes join the social structure.
 Modeling the underlying mechanisms that create network changes in time is very complex but also provides potential to uncover hidden truths.
 
-\par We concentrate on one type of model for dynamic social networks: the stochastic actor-oriented models (SAOMs), introduced by \citet{saompaper}.
-These models are fundamentally different from other social network models because they allow us to incorporate network \textit{and} actor statistics, where other models only rely on the network statistics, to model the changes in the network.
-Allowing the actor-level statistics to directly effect the structure of the network leads to a more practical and relevant approach to model change in a social network: we expect people with common interests to be more likely to form relationships, and SAOMs allow us to incorporate this intuition in the modeling process.
-
-%\par In order to estimate the parameters of  SAOMs the software SIENA, and its \texttt{R} implementation \texttt{RSiena}, was developed \cite{RSiena}.
-This software marks a huge contribution to the field of social network analysis., and when we began working with it we were eager to understand how it works.
-HH: we are still interested in this.
-%\hh{The next sentence is very vague - make it more specific: what aspect of SAOMs is not well understood?
-I think what you want is something like ``SAOMs are not very tractable analytically with traditional methods, e.g.
-likelihood functions turn into very complex objects because of the sheer number of parameters involved.
-Therefore, computationally more tractable solutions are used to fit estimators, in particular, SAOMs are fit using ...
-}
-Unlike other network models, SAOMs are not very well understood.
+\par To model dynamic social networks, we use stochastic actor-oriented models (SAOMs), introduced by \citet{saompaper}.
+These models are different from other network models because they allow us to add node level information into the model with the traditional network structure information.
+Adding node-level information into the model is a more intuitive approach to model changes in a social network.
+For example, we expect people with common interests to be more likely to form relationships, and SAOMs allow us to incorporate this information on shared interests in the modeling process.
+
+\par Unlike other network models, SAOMs are not very well understood.
 They are relatively new, especially compared to the classic ERGMs, and they are not very tractable analytically.
 Likelihood functions quickly become very complex objects to analyze due to the dependency structure inherent in the data.
-Therefore, computationally more tractable solutions are used to fit models, and in particular, % Traditional modeling steps, like determining the likelihood function, are very complex.
-Because of this,
-SAOMs are often fit to data using a series of Markov Chain Monte Carlo (MCMC) phases for finding method of moments estimates of model parameters.
+Therefore, computational methods, such as Markov chain Monte Carlo (MCMC) are used to fit models.
+SAOMs are typically fit to dynamic social network data using a series of MCMC phases for finding method of moments (MoM) estimates of model parameters.
 In order to estimate the parameters of SAOMs, we use the software SIENA, and its \texttt{R} implementation \texttt{RSiena}, which was developed by \citet{RSiena}.
-This software marks a huge contribution to the field of social network analysis, but the many moving pieces involved in parameter estimation are largely ``behind the scenes" and hidden from the software user.
-In order to better understand the model-fitting process, we attempt to bring SAOM fits and the fitting process out into the light.
-%, by combining the principals of network visualization with those of model visualization as discussed in \citet{modelvis}.
-%Here, we explore RSiena and SAOMs with innovative network visualizations, both static and dynamic.
-By visualizing the underlying methods and structures that are behind the scenes when fitting SAOMs to network data, we aim to help researchers working with the models better understand the implications and analyses of these models.
-
-\par To get behind the scenes of the SAOM fitting process, we use model visualization techniques (see \citet{modelvis}) to display the model in the data space, view collections of models instead of single models, and visually explore the process of fitting the SAOMs as opposed to looking at only the final output.
-Stochastic actor-oriented models are an excellent example of a family of models that can benefit greatly from the application of model visualization.
+This software is a considerable contribution to the field of social network analysis, but the many simulations involved in parameter estimation are largely hidden from the software user.
+In order to better understand the model-fitting process of \texttt{RSiena}, we attempt to bring model fits and the fitting process to the foreground by visualizing the underlying methods and structures.
+Utimately, we aim to help researchers working with the models better understand the structure of and results from these models.
+
+\par To learn more about the SAOM fitting process, we use model visualization techniques (see \citet{modelvis}) to display the model in the data space, view collections of models instead of single models, and explore the algorithms fitting the SAOMs instead of studying only the final output.
+SAOMs are an excellent example of a family of models that can reap many benefits from the application of model visualization techniques.
 For instance, the models themselves include a continuous-time Markov chain (CTMC) that is completely hidden from the analyst in the model fitting process.
-Bringing the CTMC out of the black box and into the light through model visualization can provide researchers with insights into the underlying features of the model.
-Furthermore, SAOMs can include a great deal of parameters to be added to the model structure, each of which is attached to a network statistic.
-These statistics are often somewhat, if not highly, correlated, which causes high correlation between the associated parameters in a SAOM.
+Visualizing the CTMC can provide researchers with more insight into the underlying features of the model.
+In addition, SAOMs can include a great deal of parameters in the model structure, each of which is attached to a network statistic.
+These statistics are often somewhat, if not highly, correlated, which causes high correlation between the corresponding parameters in a SAOM.
 By visualizing collections of SAOMs, we gain a better understanding of these correlations and can find ways to account for their effects in the model.
-In addition, the estimation of the parameters in a SAOM relies on a Robbins-Monro algorithm, and the convergence checks for these estimates rely on simulation from the fitted model.
-Again, each of these steps are largely kept in the background of the estimation process.
-With the help of static and dynamic visualizations we bring the hidden model fitting processes into the foreground, eventually leading to a better understanding and higher accessibility of stochastic actor-oriented models for social network analysts.
-
-In Section~\ref{sec:netintro}, we introduce basic concepts of networks and network visualizations.
-In Section~\ref{sec:saoms}, we present the family of stochastic actor-oriented models for social network analysis.
+Furthermore, the estimation of the parameters in a SAOM rely on convergence checks based on simulations from the fitted model.
+These simulations are hidden and unsaved by default for computational efficiency.
+With the help of both static and dynamic visualizations, we bring the hidden model fitting processes into the foreground, eventually leading to a better understanding and higher accessibility of stochastic actor-oriented models for social network analysts.
+
+The remainder of this paper is organized as follows: In Section~\ref{sec:netintro}, we introduce basic concepts of networks and network visualizations.
+In Section~\ref{sec:saoms}, we formally define the family of stochastic actor-oriented models for social network analysis.
 In Section~\ref{sec:ModelVis}, we combine concepts from Sections~\ref{sec:netintro} and \ref{sec:saoms} in an application of the model-vis paradigm, and conclude with a discussion in Section~\ref{sec:discussion}.
@@
 
-Network data is of frequent interest to researchers in a wide array of fields, such as biologists studying neural networks and sociologists studying social networks.% There are technological networks, like power grids or the internet, information networks, such as citation networks or the World Wide Web, biological networks, like neural networks, and social networks, just to name a few \citep{newman}.
-% Each of these examples have one thing in common:
-Network data across fields have a similar data structure. There are always units of observation, such as neurons %the power stations, websites, neurons, 
-and people, which we refer to throughout this paper as \textit{nodes} or \textit{actors}.
-There are also always connections of some kind between those units, such as electrical signals and relationships, which we will call \textit{edges} or \textit{ties}.
-Networks often change over time, such as when there are relationships formed in a friendship network.
-The nodes and edges themselves can also have inherent variables of interest, e.g.\ the age, gender, and political affiliation of people in a friendship network and the length of a friendship.
-%the institution of authors in a co-authorship network, or the number of times two authors have collaborated.
-
-The multiple layers of network data structures pose unique problems to analysts.
-Some questions that researchers may aim to answer are: How does the strength of a tie between two nodes affect the overall structure of the network?
-Do differences in node variables affect the formation or dissolution of edges?
-Which views of the data are most informative for communicating significant model effects?
-What about other results of statistical analyses of the network?
-%These are, of course, just a few broad questions, and we focus here on the latter, which
-We aim to answer some of these questions through visual exploration of network data and models.
+Network data across fields have similar data structures.
+There are always units of observation and connections between those units.
+In a social network, the units of observations, called \textit{nodes} or \textit{actors}, may be people, while the connections, called \textit{edges} or \textit{ties}, are the relationships between people.
+Networks often change over time, like when new relationships are formed in a social network.
+The nodes and edges themselves can also have inherent variables of interest, e.g.\ the age, gender, and political affiliation of people in a social network and the type of relationship (friends, married, etc.).
+
+These two network data structures pose unique problems to analysts, such as, ``How does the strength of a tie between two nodes affect the overall structure of the network?" or ``Do differences in node variables affect the formation of edges?" We propose to answer these questions and more through visual exploration of network data and models.
 
@@
 Most data visualizations rely on well-defined axes inherited from the data, such as Cartesian coordinates or spatial locations.
-%If variables are numerical, histograms, scatterplots, or time series plots are straightforward to construct.
-If the variables are categorical, bar charts and mosaic plots are available to to the researcher.
-If the data are spatial, there is a well-defined region in which to view information.
-Network data, however, are much less cut-and-dried.
+Network data, however, carry no such inherent structure.
 
@@
 
-There are two primary methods used to visualize networks: node-link diagrams and adjacency matrix visualization \citep{knuth2000, adjmatpro}.
-As a toy example, let us assume that we have five nodes, $\{1,2,3,4,5\}$, connected by five directed edges:  $\{2 \to 4, 3 \to 4, 1 \to 5, 3 \to 5, 5 \to 4\}$ We use this toy data set to demonstrate the two visualization methods in Figure~\ref{fig:toyplots}.
-
-The first method, the node-link diagram, represents nodes with points in two dimensions and then represents edges by connecting the points with lines when there is an edge between the two nodes.
-These lines can also have arrows on them indicating the direction of the edge for directed networks.
-But because there is typically no natural placement of the points unless they have inherent spatial locations, a random placement of the points is used, then adjusted via a layout algorithm, of which there are many \citep{drawingalgs}.
-
-Some commonly used layout algorithms, such as  the Kamada-Kawai layout \citep{kamadakawai} and the Fruchterman-Reingold layout \citep{fruchterman-reingold}, are designed to mimic physical systems, drawing the graphs based on the ``forces" connecting them.
-In these algorithms, the edges of the network act as springs pushing and pulling the nodes in a low dimensional (usually two-dimensional) space.
-Another algorithm uses multi-dimensional scaling, relying on distance metric and computing a matrix whose entries represent the ``distance" between every pair of nodes.
-There are also layout algorithms that use properties of the adjacency matrix, like its eigenstructure, to place the nodes in 2D space \citep{drawingalgs}.
-The node-link diagram using the Kamada-Kawai layout algorithm for our toy network is shown in Figure~\ref{fig:toyplots}.
-Unless otherwise stated, all other node-link diagrams in this paper will use the Kamada-Kawai layout.
-%Other algorithms use properties of the network's adjaceny matrix in order to place the nodes.
-%hh{XXX how does the adjacency matrix feature here?
-The adjacency matrix is a different mathematical representation of a network, but describes the same network as a node-link diagram.} \st{XXX I'm taking this to mean I should describe adjacency matrices more, so I've added a discussion below.
-XXX }
-
-The second primary method for network visualization uses the adjacency matrix of the network.
+There are two primary methods used to visualize networks that are very different from most traditional data visualizations.
+They are the node-link diagram and the adjacency matrix visualization \citep{knuth2000, adjmatpro}.
+For example, consider a network with five nodes, $\{1,2,3,4,5\}$, connected by five directed edges:  $\{2 \to 4, 3 \to 4, 1 \to 5, 3 \to 5, 5 \to 4\}$.
+This data is shown using the two visualization methods in Figure~\ref{fig:toyplots}.
+
+The first method, the node-link diagram, represents nodes with points in two dimensions and then represents edges by connecting the points with lines.
+These lines can also have arrows on them indicating the direction of the edge for directed networks, as show in Figure~\ref{fig:toyplots}.
+Because there is usually no natural placement of the points (the exception is nodes with spatial locations), a random placement of the points is used, then adjusted with a layout algorithm, of which there are many \citep{drawingalgs}.
+
+\par Some commonly used layout algorithms, such as  the Kamada-Kawai (KK) layout \citep{kamadakawai} and the Fruchterman-Reingold (FR) layout \citep{fruchterman-reingold}, are designed to mimic physical systems, drawing the graphs based on the ``forces" connecting them.
+Other algorithms use multi-dimensional scaling or properties of the adjacency matrix (e.g.
+its eigenstructure), to place the nodes \citep{drawingalgs}.
+In Figure~\ref{fig:toyplots}, we use the KK algorithm, and unless otherwise stated, all other node-link diagrams in this paper will use the KK layout.
+
+\par The second method for network visualization uses the adjacency matrix of the network.
 The adjaceny matrix of a network, $\mathbf{A}$ is defined as follows: an entry $A_{ij}$ of $\mathbf{A}$, for two nodes $i\neq j$ in the network is defined as
-\[
+\begin{equation}\label{eq:adjmat}
   A_{ij} =
@@
   \end{cases}
-\]
-We only consider the presence or absence of an edge, but if the network has weighted edges, for example an email network where edge weights represent the number of emails sent from one person to another, the entries in the adjacency matrix can be the edge weights instead of zeroes and ones.
-In an undirected network, $A_{ij} = A_{ji}$ for all $i \neq j$, but in a directed graph this is only true if there is an edge from $i$ to $j$ and from $j$ to $i$.
-Thus, $\mathbf{A}$ is always symmetric for undirected networks, and is symmetric for directed networks if and only if every edge between two nodes is reciprocated.
-%\hh{XXX what is an adjacency matrix? - explain at this point in mathematical terms} \st{XXX see above XXX}
+\end{equation}
+We only consider the presence or absence of an edge, but if the network has weighted edges, the entries in the adjacency matrix are the edge weights.
 An adjacency matrix visualization for our toy example is also shown in Figure~\ref{fig:toyplots}.
-
-Each type of visualization comes with its own advantages and disadvantages.
+This visualization shows the network as a grid, with each row and column representing a node.
+The grid space for edge $i\rightarrow j$ is filled in if $A_{ij} = 1$ and is empty if $A_{ij} = 0$.
+
+\par Each visualization comes with its own advantages and disadvantages.
 For instance, paths between two nodes in a network are easier to determine with node-link diagrams than with adjacency matrix visualizations \citep{adjmatviz}.
-In node-link diagrams, node-level information can be incorporated into the visualization by coloring or changing the shape of the points representing the nodes, and edge-level information can be incorporated by coloring the lines, or changing their thickness, linetype, or color.
+In node-link diagrams, node-level information can be incorporated into the visualization by coloring or changing the shape of the points, and edge-level information can be incorporated by coloring the lines, or changing their thickness or linetype.
 Incorporating a node-level variable into an adjacency matrix visualization is not as straightforward or simple, because this type of visualization features the edges.
-Adjacency matrix visualization can be useful when the network is very complex, dense, or large.
-Experimental studies have shown adjacency matrix visualization to be superior to node-link diagrames for large networks.
+Adjacency matrix visualization can, however, be useful when the network is very complex, dense, or large \citep{adjmatviz}.% Experimental studies have shown adjacency matrix visualization to be superior to node-link diagrams for large networks.
 For example, for basic perceptual tasks on networks, including node and edge count, adjacency matrix visualizations outperform node-link diagrams as the size and density of the network increases \citep{adjmatviz}.
-One drawback of the adjacency matrix visualization that \citeauthor{adjmatviz} found was that edges are overrepresented for undirected graphs, due to the symmetry of $\mathbf{A}$: the edge $x_{ij}$ for $i \neq j$ appears in $\mathbf{A}$ twice: in $A_{ij}$ and $A_{ji}$, and so it also appears twice in the adjacency matrix visualization.
-%\hh{XXX What is the exact logic here that Ghoniem used to come to the conclusion that edges in undirected graphs are over-represented?
-Why would directed graphs make that better?
-wouldn't that just add to the number of edges?
-} \st{XXX in undirected nets, the adj mat is symmetric, where one edge ($x_{ij}$) is represented in entry $i,j$ in the matrix and in entry $j,i$.
-So, there are many repetitions in the matrix, which makes the network appear more dense than it really is.}
-This is advantageous for \emph{directed} graphs, where exactly the correct number of edges is represented in a matrix visualization, due to the fact that the edges $x_{ij}$ and $x_{ji}$ are not interchangeable.
+One drawback of the adjacency matrix visualization that \citeauthor{adjmatviz} found was that edges are overrepresented for undirected graphs, due to the symmetry of $\mathbf{A}$: the edge $i \rightarrow j$ for $i \neq j$ appears in $\mathbf{A}$ twice: in $A_{ij}$ and $A_{ji}$, and so it also appears twice in the adjacency matrix visualization.
+This is an advantage, however, for \emph{directed} graphs, where exactly the correct number of edges is represented in a matrix visualization.
 A node-link diagram, however, may underrepresent the edge count if the edges $x_{ij}$ and $x_{ji}$ both exist and are drawn on top of one another.
-Ultimately, however there is not one "correct" way to visualize network information, and we will be using both the node-link and adjacency matrix visualization methods throughout this paper to explore social networks and stochastic actor-oriented models.
-
-\section{Stochastic Actor-Oriented Models for Longitudinal Social Networks} \label{sec:saoms}
-
-% \st{What do I need to communicate in this section?
-% \begin{itemize}
-% \item What the models are for: dynamic social networks with actor-level covariates
-% \item What are the model components?
-rate function, objective function (network statistics, actor covariates, interactions between covariates and network statistics), time points ($t_m = 1, \dots, M$)
-% \item What is RSiena? What methods of estimation, convergence checks does it use? What are the algorithms? What kinds of models can it fit?
-% \end{itemize}}
-
-%The stochastic actor-oriented models are very aptly named: they model dynamic networks that can account for the individual node characteristics when making the edge changes.
-The object of analysis is a social network that has been observed at several discrete time points.
-The nodes in the network can have associated covariate variables that may be incorporated into modeling the tie changes in the network.
-The model itself has two primary pieces: the rate of change of ties, and the objective function of the nodes, which determines the exact change that gets made.
-
-%We start with a set of network observations at $M$ discrete time points: $x(t_1), x(t_2) \dots, x(t_M)$.
-The first observation, $x(t_1)$ is always conditioned on.
-The number of nodes / actors / vertices is constant in all time points and is denoted $n$.
-The tie variables are directed and  binary, and are denoted at $x_{ij}(t_m)$ for time point $m$, and this is sometimes abbreviated to $x_{ij}$ in notation.
-If there is a tie from node $i$ to node $j$, then $x_{ij}=1$, and $x_{ij}=0$ otherwise.
-
-%\hh{mostly lit review plus a detailed description of the model and fitting process.}
-
-A Stochastic Actor-Oriented Model (SAOM) is a model that incorporates all three data levels of dynamic networks: edge, node, and temporal information.
-It models the change of a network over time, allowing for changes in network structure due to actor-level covariates.
+Ultimately, there is no one ``correct" way to visualize network information, and we will be using both the node-link and adjacency matrix visualization methods throughout this paper to explore social networks and stochastic actor-oriented models.
+
+\section{Defining SAOMs} \label{sec:saoms}
+
+A Stochastic Actor-Oriented Model (SAOM) is a model that incorporates all three data levels of dynamic networks: actor, tie, and temporal information.
+It models the change of a network in time, while accounting for changes in network structure due to actor-level covariates.
 This model was first introduced by Snijders in 1996 \citep{saompaper}.
@@
 Most social networks, even holding constant the set of actors over time, are ever-changing as relationships decay or grow in seemingly random ways.
-Most actors in social networks have inherent properties that could affect how they change their role within the network, and ties between actors may affect formation or dissolution of ties.
+Actors in social networks also have inherent properties that could affect how they change their role within the network, and existing ties may affect formation or dissolution of other ties.
 
@@
 We denote the network observation at timepoint $t_k$ by $x(t_k)$.
-In the modelling process, we condition on the first observation, $x(t_1)$.
 The SAOM assumes that this longitudinal network of discrete observations is embedded within a continuous time Markov chain (CTMC), which we will denote $X(T)$.
 This process is almost entirely unobserved: we assume that the beginning of the process, $X(0)$, is equivalent to the first network observation $x(t_1)$, while the end of the process, $X(\infty)$, is equivalent to the last observation $x(t_M)$.
-Nearly all other parts of the process are unseen, with the exception of $x(t_2), \dots, x(t_{M-1})$.
+Nearly all other steps in the chain are unseen, with the exception of $x(t_2), \dots, x(t_{M-1})$.
 Unlike the first and last observations of the network, these ``in-between" observations do not have direct correspondence with steps in the continuous time Markov chain.
@@
 
-In the broadest definition of a SAOM, each actor $i$ in the network $x$ may change its ties, $x_{ij}$, to other nodes $j \neq i$ according to the SAOM's  \emph{rate function}, $\rho(x, \mathbf{z}, \boldsymbol{\alpha})$, where $\boldsymbol{\alpha}$ are the parameters in the function $\rho$, and $x$ is the current network state, with covariates of interest $\mathbf{z}$.
+In general, each actor $i$ in the network $x$ may change its ties, $x_{ij}$, to other nodes $j \neq i$ according to the SAOM's  \emph{rate function}, $\rho(x, \mathbf{z}, \boldsymbol{\alpha})$, where $\boldsymbol{\alpha}$ are the parameters in the function $\rho$, and $x$ is the current network state, with covariates of interest $\mathbf{z}$.
 With this general formulation, it is possible that each node will have a different rate of change.
-We assume a simple rate function, $\rho(x, \mathbf{z}, \boldsymbol{\alpha}) = \alpha_m$ that is constant across all actors between observations at time $t_m$ to $t_{m+1}$, and thus our rate function is just a set of rate parameters in the overall model.
-%\hh{XXX that means that the function is essentially just a rate parameter, right?} \st{XXX yes XXX}
-%Other modelling scenarios allow this rate to be more flexible,  e.g.\ a function that depends on the time period of observation, some actor-level covariates or some actor-level network statistics.
-%\hh{How is the rate parameter different from a rate parameter?} \st{I think you meant how is a rate fn different from a rate parameter?
-I updated the language XXX->}%In our model with the simple rate function,
-In the simple model with a rate parameters instead of a rate function, the rate parameters dictate how quickly an actor $i$ gets an opportunity to change one of its ties to the other nodes in the network, $x_{ij}$, for $j \in \{1, \dots, n\}$ in the time period from $t_{m}$ to $t_{m+1}$ for $m = 1, \dots, M-1$ If $j = i$, no change in the network is made.
-We also assume that the actors $i$ are conditionally independent given their ties, $x_{i1}, \dots, x_{in}$ at the current network state.
-Let $\tau(i|x,m) $ be the wait time until actor $i$ makes its next change from its current state in the network $x$.
-%Note that $m$ indicates the number of the wave that is conditioned on in the SAOM.
-For any time point, $T$, where $t_{m} \leq T < t_{m+1}$, the waiting time to the next change opportunity by actor $i$ is exponentially distributed with expected value $\alpha_m^{-1}$.
-The conditional independence of nodes in the network given their current ties is expressed in Equation~\ref{eq:c1}.
+We assume a simple rate function, $\rho(x, \mathbf{z}, \boldsymbol{\alpha}) = \alpha_m$ that is constant across all actors between time $t_m$ and $t_{m+1}$.
+With this simple rate function, we have just one set of rate parameters in the overall model.
+In this simple model, the rate parameters dictate how often an actor $i$ ``gets an opportunity" to change one of its ties, $x_{ij}$, for $j \in \{1, \dots, n\}$ in the time period from $t_{m}$ to $t_{m+1}$ for $m = 1, \dots, M-1$ (Note that when $j = i$, no change is made.) We also assume that the actors $i$ are conditionally independent given their ties, $x_{i1}, \dots, x_{in}$ at the current network state.
+Let $\tau(i|x,m) $ be the wait time until actor $i$ gets to the opportunity to change its current set of ties.
+For any time point, $T$, where $t_{m} \leq T < t_{m+1}$, the waiting time to the next change made by actor $i$ is exponentially distributed with expected value $\alpha_m^{-1}$.
+Then, we express this in Equation~\ref{eq:c1}.
 
@@
 
- The waiting time to the next change opportunity by \textit{any} actor in the network is also exponentially distributed with expected value $(n\alpha_m)^{-1}$.
-%The distribution of waiting time for the whole network to change, $\tau(x|m) = \sum_i \tau_i(m) |x_{i1}(m), \dots, x_{in}(m) $ can then be written as
-%\begin{equation} \label{eq:c2}
-% \tau(x|m) \sim Exp(n\alpha_m)
-%\end{equation}
-The parameter for the wait time for the whole network $n\alpha_m$ is the rate at which any tie change occurs.
-The estimation of this parameter is straightforward: the method of moments is used to estimate the rate with the statistic
+From Equation~\ref{eq:c1}, we can derive the waiting time to the next change opportunity by \textit{any} actor in the network, $\tau(x)$.
+The value $\tau(x)$ is also exponentially distributed with expected value $(n\alpha_m)^{-1}$, where $n\alpha_m$ is the rate at which any tie change occurs.
+The estimation of this parameter in \texttt{RSiena} is simple: the method of moments is used to estimate the rate with the statistic
 \begin{equation}\label{eq:cdiff}
@@
 \end{equation}
-which is the total number of changes from observation at time $t_m$ to the observation at time $t_{m+1}$. %In order to achieve the memorylessness property of a Markov process,
-
-% \hh{XXX we tried to estimate the rate based on the simulations - you could include a histogram of the number of changes for the 1000 simulations in the viz part XXX}
-% \st{XXX see line 1257 for plot. where should I put it? }
-% \hh{XXX Put a reference to the plot here.}
+which is the total number of changes from observation at time $t_m$ to the observation at time $t_{m+1}$.
 
@@
 
-%\hh{shouldn't that be independence? - make the conditional independence assumption above a labelled formula (e.g. C1) and then call it here }
-Because of the conditional independence assumptions given in Equation~\ref{eq:c1}, we can consider the objective function for each node separately, as only one tie from one node is changing at a time.
-The node $i$, which is the node that is changing at the current time point, is called the \textit{ego} node.
+Because of the conditional independence assumptions given in Equation~\ref{eq:c1}, we can consider the objective function for each node separately, as only one tie from one node is changing at a time in the CTMC.
+The node $i$, which is the node that is given the opportunity to change at the current time point, is called the \textit{ego} node.
 It has the potential to interact with all other nodes in the network, $j \neq i$.
-These nodes $j$, are referred to as \textit{alter} nodes.
-These nodes are acted upon by the ego node, and they only act when they become the ego node at a subsequent time point in the CTMC.
-For the ego node, $i$, in the current network state $x$, its \emph{objective function}, which it tries to maximize, is written as
+These nodes $j$, are referred to as \textit{alter} nodes, and do not change any of their ties when $i$ may change.
+For the ego node $i$ in the current network state $x$, its \emph{objective function}, which it tries to maximize, is written as
 \begin{equation} \label{eq:of1}
@@
 The vector $\boldsymbol{\beta}$ contains the parameters of the model with corresponding network and covariate statistics, $s_{ik}(x, \mathbf{Z})$, for $k = 1,\dots, K$.
-%\hh{XXX ego node isn't defined yet - the description of the charts would be a good place for it.} \st{XX I added it above the obj function equation.
-Is it ok there?}
-Given the ego node, $i$, there are $n$ possible steps for the actor $i$ to take: either one of all current ties $x_{ij} = 1$ will be destroyed, a new tie will be created that is currently $x_{ij}=0$, or no change will occur.
-
-The parameters, $\boldsymbol{\beta}$, correspond to various actor-level statistics, $s_{ik}(x)$.
-%\hh{'There are always' ...
-that seems to come from a Snijder paper?
-Re-phrase this to be less vague: In the current implementation of RSiena (snijder) a model has minimally two parameters...}
+To increase the value of $f_i$, $i$ will destroy one tie where $x_{ij} = 1$, create one tie where $x_{ij}=0$, or make no change if any changes result in lower values of $f_i$.
+
 According to \citet[p.
 371]{snijders01}, there should be at least two parameters included in the model: $\beta_1$ for the outdegree of a node, and $\beta_2$ for the number of reciprocal ties held by a node.
-These effects are also used in ERGMs.
-The outdegree represents the propensity of nodes with a lot of outgoing ties to form more outgoing ties (the "rich get richer" effect), and the reciprocity parameter measures the tendency of outgoing ties to be returned within a network.
-The statistics corresponding to these two parameters are functions of the edge variables, $x_{ij}$.
-In the \texttt{RSiena} software that we use to fit the SAOMs, there are over 80 possible parameters to add to the model.
-The formulas for the effects are provided in \citet{RSienamanual}.
-
-Each of the parameters in the model belongs to one of two groups: structural or covariate.
-The structural parameters, such as the outdegree and reciprocity parameters, correspond to statistics that are functions of the edge variables only.
-They model underlying mechanisms of network change, answering questions such as, ``How does the existing network structure influence change in the network?" %\hh{XXX move the last colored section to the front of this paragraph.
-That sets up the structure nicely, e.g.
-`In determining the covariates of a SAOM, two sources of parameters are considered: ...} \st{XXX as I'm going through and uncommenting your comments (sorry....) this one has lost meaning a little bit.
-Could you give me more detail?
-I'm not sure which section is which at this point.
-(sorry again...)} \hh{I guess you addressed this comment very thoroughly :)}
-% and ``How do the behavior and characteristics of the nodes influence change in the network?" put this sentence below somewhere
-The covariate parameters are functions of the edge variables and additional node-level covariates of interest.
+(These effects are also used in ERGMs.) The outdegree represents the propensity of nodes with a lot of outgoing ties to form more outgoing ties (the "rich get richer" effect), and the reciprocity parameter measures the tendency of outgoing ties to be returned within a network.
+The statistics corresponding to these two parameters, $s_{i1}(x)$ and $s_{i2}(x)$ are given in Figure~\ref{tab:effects}.
+In the \texttt{RSiena} software, there are over 80 possible $\boldsymbol{\beta}$ parameters to add to the model.
+The formulas for the corresponding statistics for all effects are provided in \citet{RSienamanual}.
+
+The $\boldsymbol{\beta}$ parameters are either structural or covariate parameters.
+The structural parameters, such as the outdegree and reciprocity parameters, correspond to statistics that are only functions of the current network state, $x$.
+They model underlying mechanisms of network change, answering questions such as, ``How does the existing network structure influence change in the network?" The covariate parameters are functions of $x$ and additional node-level covariates of interest, $\mathbf{Z}$.
 A table of some possible structural and covariate parameters and their corresponding statistics is given in Table~\ref{tab:effects}.
-For a complete list of the network and covariate statistics that can currently be included in the objective function, see \citet{RSienamanual}.
 
@@
 \begin{tabular}{m{4.5cm}m{4.5cm}m{3cm}}
-\multicolumn{3}{l}{\textbf{Structural Effects}} \\ \hline\hline
+\multicolumn{3}{l}{\textbf{Structural Effects}} \\
+Name & Statistic & Figure \\
+\hline\hline
 outdegree  & $s_{i1}(x) = \sum_j x_{ij}$ & \includegraphics[width=2cm]{img/outdegree}\\ \hline
-reciprocity  & $s_{i2}(x) = \sum_j x_{ij}x_{ji}$ & \includegraphics[width=2cm]{img/reciprocity}\\ \hline
-transitive triplets  & $s_{i3}(x) = \sum_{j,h} x_{ij}x_{jh}x_{ih}$ & \includegraphics[width=1.5cm]{img/transtrip}  \\
-\\
-\multicolumn{3}{l}{\textbf{Covariate Effects}}\\  \hline \hline
+reciprocity  & $s_{i2}(x) = \sum_j x_{ij}x_{ji}$ & \includegraphics[width=2cm]{img/reciprocity}\\ 
+%\hline
+%transitive triplets  & $s_{i3}(x) = \sum_{j,h} x_{ij}x_{jh}x_{ih}$ & \includegraphics[width=1.5cm]{img/transtrip}  \\
+%\\
+\multicolumn{3}{l}{\textbf{Covariate Effects}}\\  
+Name & Statistic & Figure \\
+\hline \hline
 covariate-alter  & $s_{i4}(x) = \sum_j x_{ij}z_j$ & \includegraphics[width=3cm]{img/covaralter} \\  \hline
-covariate-ego  & $s_{i5}(x) = z_i\sum_j x_{ij}$ & \includegraphics[width=3cm]{img/covarego} \\  \hline
-same covariate & $s_{i6}(x) = \sum_j x_{ij} \mathbb{I}(z_i = z_j)$ & \includegraphics[width=3cm]{img/covarsim}%\\
+covariate-ego  & $s_{i5}(x) = z_i\sum_j x_{ij}$ & \includegraphics[width=3cm]{img/covarego} \\ % \hline
+%same covariate & $s_{i6}(x) = \sum_j x_{ij} \mathbb{I}(z_i = z_j)$ & \includegraphics[width=3cm]{img/covarsim}%\\
 %jumping transitive triplets  & $s_{i7}(x) = \sum_{j \neq h} x_{ij}x_{ih}x_{hj} \mathbb{I}(z_i = z_h \neq z_j)$
 \end{tabular}
-\caption{\label{tab:effects} Some of the possible effects to be included in the stochastic actor-oriented models in RSiena.
-There are many more possible effects, but we only consider a select few here.
-For a complete list, see the RSiena manual \citep{RSiena}.
-The darker nodes in the small figures are the ego nodes, while the lighter are the alters.
-The dotted lines are ties being formed when the parameter corresponding to each effect is positive.
-The differently shaped nodes in the covariate effect diagrams represent different levels of the covariate $z$.}
+\caption{\label{tab:effects} Examples of effects included in a SAOM.
+The dark nodes in the Figures are the ego nodes, and all others are alters.
+The dotted lines are ties the ego node is en(dis)couraged to make when the parameter corresponding to each effect is positive (negative).
+Node shapes represent levels of the covariate $z$.}
 \end{table}
 
-When node $i$ is given the chance to change a tie, %\hh{`we assume that they wish' sounds strange - don't assume wishes, state assumptions}
-it acts to maximize the value of its objective function $f_i(\boldsymbol{\beta}, x)$ %as well as a random element, $U_i(x)$, to account for unknown attraction between nodes.
-The additional random element is included to account for any random, unexplainable change in the network ties.
-\citet{modelsSnijders} recommends that the $U_i(x)$ be random draws from a type 1 extreme value distribution.
-% XX What is the thought behind that?
-} ``the type 1 extreme value distribution (or Gumbel distribution) with mean 0 and scale parameter 1" \citep[p.
-368]{snijders01}.
-%This distribution, which is also known as the log-Weibull distribution, has probability distribution function of:
-%\begin{equation} \label{eq:dist1}
-%g(u|\mu, \sigma) = \frac{1}{\sigma}\exp\left\{-\left(\frac{u-\mu}{\sigma} + \exp^{-\frac{u-\mu}{\sigma}}\right)\right\}.
-%\end{equation}
-%using $\mu$ for the mean parameter and $\sigma$ for the scale parameter.
-<<logweibull, eval = FALSE, fig.align='center', out.width = '.5\\textwidth', fig.height = 4, fig.cap='The probability distribution function for the type 1 extreme value distribution, also known as the log-Weibull or Gumbel distribution with location parameter $ \\mu = 0$ and scale parameter $ \\sigma = 0$.'>>=
-library(reliaR)
-x <- seq(-3,10,.01)
-y <- dgumbel(x = seq(-3,10,.01), sigma = 1, mu=0)
-qplot(x = x, y= y, geom = "line") +
-  annotate('text', x = 6, y = .25,
-        label = "g(u) = exp(-(u + exp(-u)))",size=5, family = "Times New Roman")  +
-  theme(text = element_text(size = 20, family = "Times New Roman")) +
-  labs(x = "u", y = "density of g(u)")
-@
-%The probability density function for the type 1 extreme value distribution is shown in Figure~\ref{fig:logweibull}.
-For mean $\mu = 0$ and scale $\sigma = 1$ as \citet{modelsSnijders} suggests %\hh{XXX I'm not following - which distribution do you mean by `this' distribution?
-The density above is the type 1 extreme value density.
-I got lost at the  ``objective function $f_i(\boldsymbol{\beta}, x)$ plus a random element, $U_i(x)$"} \st{XXX I hope it is clearer now above}
-%is convenient because it \st{leads to a simple probability formula for} the probability that actor $i$ chooses to change its tie to actor $j$ \st{that can be written \textit{only}} in terms of the objective function.
-We can use this objective function to define the probability that actor $i$, once selected with the rate function, chooses to change its tie to actor $j$, $p_{ij}(\boldsymbol{\beta}, x)$.
-Let $x(i \leadsto j)$, denote the network identical to $x$ with only one change: the tie $x_{ij}$ has changed to $1-x_{ij}$.
-Then, the probility that the tie $x_{ij}$ changes is
+The objective function also defines the probability that actor $i$, when it is the ego node, chooses to change the tie to actor $j$, $p_{ij}(\boldsymbol{\beta}, x, \mathbf{Z})$.
+Let $x(i \leadsto j)$ denote the network identical to $x$ with the exception of tie $x_{ij}$.
+It has changed to $1-x_{ij}$.
+The probility that the tie $x_{ij}$ changes is defined as:
 
 \begin{equation}\label{eq:pij}
-p_{ij}(\boldsymbol{\beta}, x) = \dfrac{\exp\left\{f_i(\boldsymbol{\beta}, x(i\leadsto j))\right\}}{\sum_{h \neq i} \exp\left\{f_i(\boldsymbol{\beta}, x(i \leadsto h))\right\}}
+p_{ij}(\boldsymbol{\beta}, x, \mathbf{Z}) = \dfrac{\exp\left\{f_i(\boldsymbol{\beta}, x(i\leadsto j), \mathbf{Z})\right\}}{\sum_{h \neq i} \exp\left\{f_i(\boldsymbol{\beta}, x(i \leadsto h), \mathbf{Z})\right\}}
 \end{equation}
-
-When $i = j$ in $p_{ij}$, the numerator represents the exponential of the value of the objective function when evaluated at the current network state.
+When $i = j$, the numerator represents the exponential of the value of the objective function when evaluated at the current network state.
 When the value of the objective function is high at the current state, the probability of not making a change is also high.
-In the CTMC, when actor $i$ may make a change, it chooses which tie $x_{i1, \dots, x_{in}}$ to change at random according to the probabilities $p_{ij}(\boldsymbol{\beta}, x)$.
+In the CTMC, when actor $i$ may make a change, it chooses which tie $x_{i1, \dots, x_{in}}$ to change at random according to the probabilities $p_{ij}(\boldsymbol{\beta}, x, \mathbf{Z})$.
 The objective function and the resulting values of $p_{ij}$ are combined with the rate function to fully describe the CTMC that is used to model network change in a SAOM.
 For more on CTMCs, see \citet{ctmc}.
-
-%\hh{XXX This paragraph ends a bit abruptly - what is the future state probability used for?} \st{XX The future state probability is used within the continuous time markov chain to move through the different network states.
-This needs more fleshing out.
-Adding to the to-do list.}
+We explore these probabilities further in Section~\ref{sec:algorithms}
 
@@
 
-To fit a SAOM to observations of a dynamic network, we use the package \texttt{RSiena} \citep{RSiena}.
-This package uses simulation methods to estimate parameter values using either the method of moments or maximum likelihood estimation.
-We use method of moments estimation throughout.
+To fit a SAOM to dynamic network data, we apply the \texttt{RSiena} software, which uses simulation methods to estimate parameter values using either MoM or maximum likelihood (ML) estimation \citep{RSiena}.
+We chose to use MoM estimation, so we describe that fitting process here, and more information on ML estimation can be found in \citet{SienaAlgs}.
 %because the theory behind it was established in \citet{saompaper}, while the maximum likelihood estimation methods were not fully established until \citet{saomsml2010}, though \texttt{RSiena} contains capabilities to use maximum likelihood estimation.
-We also use the score function method for estimating the derivatives of the expected values, as opposed to the finite differences method, both of which are outlined in detail in \citet{SienaAlgs}.
+We also use the score function method for estimating the derivatives of the expected values, as opposed to the finite differences method, both of which are outlined in detail in .
 
@@
 where $\theta = (\boldsymbol{\alpha}, \boldsymbol{\beta})$ is the vector of rate and objective function parameters, and $s_{obs}$ is the observed vector of model statistics.
-The entire algorithm is provided in \citet{SienaAlgs}.
-
-There are three phases in the the SIENA algorithm, as described in \citet{RSienamanual, SienaAlgs}.
-The first phase performs initial estimation of the score functions for use in the Robbins-Monro procedure for method-of-moments estimation.
-The second phase carries out the Robbins-Monro algorithm and obtains estimates of the parameter values through iterative updates and simulation from the CTMC at current parameter values.
-The third phase uses the parameter vector estimated in phase two to estimate the score functions and covariance matrix of the parameter estimate, and also carries out convergence checks.
+The full SIENA MoM estimation algorithm operates in three phases, as described in \citet{RSienamanual, SienaAlgs}.
+The first phase performs initial estimation of the score functions for use in the Robbins-Monro procedure for MoM estimation.
+The second phase carries out the Robbins-Monro algorithm and obtains estimates of the parameter values through iterative updates and simulation from the CTMC at iterative parameter values.
+The third phase uses the parameter vector estimated in phase two to estimate the score functions and covariance matrix of the vector of parameter estimates and carries out convergence checks.
 
 In each of the the first two phases of the algorithm, the procedure uses ``microsteps" that are simulated from the model as it exists in its current state in order to update either the score functions or the parameter estimates.
-These simulated microsteps are observed instances of the continuous-time Markov chain that is the backbone of the stochastic actor-oriented model.
+These simulated microsteps are observed instances of the CTMC that is the backbone of the SAOM.
 In Section~\ref{sec:ModelVis}, we further explore these phases in the SIENA method-of-moments algorithm through visualization, bringing them out of their ``black box" and into the light.
@@
 <<alldataviz, fig.cap="The smaller friendship network data we will be modelling throughout the paper.
-In the first wave, we can see there are two large, separate friend group.
-By the second wave, three students with heavier drinking behavior have separated from their group, while the others become members of the other group.
-By the third wave, the large group has almost completely broken up.
-We want to capture these changes with our SAOMs.", out.width='.99\\textwidth', fig.height=3>>=
+The nodes change color as their drinking behavior changes, and the network becomes less densely connected over time.
+These are the mechanisms we hope to capture with SAOMs.", out.width='.99\\textwidth', fig.height=3>>=
 actual1 <- merge(data.frame(as.edgelist(as.network(fd2.w1))),
@@
 
-To guide our visual exploration of stochastic actor-oriented models, we use two data sources.
+To guide our visual exploration of SAOMs, we use two example data sources.
 The first is a subset of the 50 actor dataset from the ``Teenage Friends and Lifestyle Study" that is provided on the \texttt{RSiena} webpage.
-These data come from \citet{friendsdata}, and we chose to only work with a subset of the data to make network visualizations less busy and to make any changes in the network more noticeable.
-To determine which subset to select, we visualized all waves of the full network using the adjacency matrix visualization approach, which we show in Figure~\ref{fig:adjmatviz}.
+These data come from \citet{friendsdata}, and we chose to only work with a subset of the data to make node-link diagrams easier to see and to make any changes in the network more noticeable.
+%To determine which subset to select, we visualized all waves of the full network using the adjacency matrix visualization approach, which we show in Figure~\ref{fig:adjmatviz}.
 This adjacency matrix visualization is different from the one in Figure~\ref{fig:toyplots} because it does not show the node IDs on the axes.
@@
 In all three adjacency matrices, the ego and the alter nodes are ordered by node ID, 1-50, which were determined arbitrarily by the relationships in wave 1.
-The subset we selected is outlined in red in the visualization.
-This subset contained actors 20 through 35 and the ties between them, as well as the drinking behavior of each actor at each of the three waves.
-%This specific subset was chosen because it showed somewhat higher connectivity than other subsets, as we've emphasized in the
-The node-link visualizations of the three network observations are provided in Figure~\ref{fig:adjmatviz}.
+In Figure~\ref{fig:adjmatviz}, we show the adjacency matrix visualization of all three observations of this network, with the subset we selected for analysis outlined in red.
+This subset contains 16 actors, their ties and their drinking behavior for three timepoints.
+The node-link visualization of the data is provided in Figure~\ref{fig:alldataviz}.
 For model fitting, we condition on wave 1 and estimate the parameters of the models for the second and third waves.
-We will also be working with an actor level categorical covariate, drinking behavior.
-This variable has five values in the original data: (1) does not drink, (2) drinks once or twice a year, (3) drinks once a month, (4) drinks once a week, and (5) drinks more than once a week.
-%The network and the actor covariate values are visualized using a node-link diagram in Figure~\ref{fig:alldataviz}.
-In the node-link diagram in Figure~\ref{fig:alldataviz}, the nodes are colored according to the drinking behavior of that student.
+The actor-level categorical covariate, drinking behavior, has four values: (1) does not drink, (2) drinks once or twice a year, (3) drinks once a month, and (4) drinks once a week.
+In Figure~\ref{fig:alldataviz}, the nodes are colored according to the drinking behavior of that student.
 Over time, we can see that the students tend to drink more and become increasingly isolated into smaller groups.
@@
 (Details of how this data can be downloaded are provided by François Briatte at \url{https://github.com/briatte/congress}).
-In this network, ties are directed from senator $i$ to senator $j$ when senator $i$ signs on as a cosponsor to the bill that senator $j$ authored.
-There are many hundreds of ties between senators when they are connected in this way, so we simplify the network by computing a single value for each senator-senator collaboration called the \textit{weighted propensity to cosponsor} (WPC).
+In this network, ties are directed from senator $i$ to senator $j$ when senator $i$ cosponsors the bill that senator $j$ authored.
+There are hundreds of ties between senators when they are connected in this way, so we simplify the network by computing a single value for each pair of senators called the \textit{weighted propensity to cosponsor} (WPC).
 This value is defined in \citet{senate} as
@@
 \end{equation}
-
 where $n_j$ is the number of bills in a congressional session authored by senator $j$, $c_{j(k)}$ is the number of cosponsors on senator $j$'s $k^{th}$ bill, where $k \in \{1,\dots, n_j\}$, and $Y_{ij(k)}$ is a binary variable that is 1 if senator $i$ cosponsored senator $j$'s $k^{th}$ bill, and is 0 otherwise.
@@
 <<senateplot, fig.cap='The collaboration network in the four senates during the Obama years, 2009-2016.
-Edges are shown only if the weighted propensity to cosponsor from one senator to another is greater than 0.25.
-We use the Fruchterman-Reingold algorithm to layout the node-link diagram.', fig.width=8, fig.height=8, out.width='\\textwidth'>>=
+Edges are shown only if the WPC between two senators is greater than 0.25.
+We use the FR layout algorithm here.', fig.width=8, fig.height=8, out.width='\\textwidth'>>=
 seobama <- read_csv("data/congress/senateobamapres_gsw_25.csv")
@@
 
-Because we require binary edges for our models, we focus only on very strong collaborations. For our senate collaboration networks, $x$, edges are defined as
-\[
+Because we require binary edges for SAOMs, we focus only on very strong collaborations. For our senate collaboration networks, $x$, edges are defined as
+\begin{equation}\label{eq:wpc2}
   x_{ij} =
@@
   \end{cases}
-\]
-The networks we constructed for the four senates during the Obama administration are shown in Figure~\ref{fig:senateplot}.
-In Section~\ref{sec:ModelVis}, we fit several stochastic actor-oriented models to these two example data sets and we use those models to guide our exploration of SAOMs.
+\end{equation}
+The node-link diagram for the four senates during the Obama administration are shown in Figure~\ref{fig:senateplot}.
+We can see that Senate 111 is much more densely connected than the other three, which have one large, sparsely connected component with a few smaller connected groups throughout.
+Some prominent names are shown to demonstrate their place in the network.
+
+We use these two example newtoks to demonstrate our model visualization methods for SAOMs in section~\ref{sec:ModelVis}
 
@@
 Every good data analysis includes both numerical and visual summaries of the data, so why restrict model description and diagnostics to numerical summaries?
-The concept of model visualization was developed to complement traditional model diagnostic tools.
-Typically, numerical summaries such as $R^2$ are computed to assess model fit, and the occasional visualization, like a residual plot, are used to determine how well the model fits the data.
-\citeauthor{modelvis} outline three separate ideas, each of which can be referred to simply as the ``model": the model family, the model form, and the fitted model.
-The latter is primarily what one thinks of first when considering a model in a data analysis, where a specified model is fit to data, and parameter estimates and other numerical summaries, such as $R^2$ are reported.
-In the context of SAOMs, the \emph{fitted model} contains the form of the rate and objective functions, the estimated rate parameters, and the estimated objective function parameters.
-The model form describes the the model \textit{before} the fitting process, defining which parameters are in the model within the context of the larger model family.
-In SAOMs, the \emph{model form} includes description of the rate and objective functions and the variables therein that describe how the network evolves over time.
-Finally, the model family is the broadest description of the model.
+The concept of model visualization was developed to complement traditional model diagnostic tools such as $R^2$ values and residual plots \citep{modelvis}.
+When performing analyses, there are three different types of ``model": the model family, the model form, and the fitted model.
+The latter is primarily what one thinks of first when considering a model in a data analysis, where a specified model is fit to data, and parameter estimates are reported.% In the context of SAOMs, the \emph{fitted model} contains the form of the rate and objective functions, the estimated rate parameters, and the estimated objective function parameters.
+The \textit{model form} describes the the model \textit{before} the fitting process, defining which parameters are in the model within the context of the larger model family.% In SAOMs, the \emph{model form} includes description of the rate and objective functions and the variables therein that describe how the network evolves over time.
+Finally, the \textit{model family} is the broadest description of the model.
 This is the type of model that you wish to fit to the data, and is chosen based on the problem, data, and knowledge at hand.
-For example, we chose to use a SAO \emph{model family} to model our network data over an exponential random graph model (ERGM) because we believe that actor-level variables effect network structure and formation, and we wanted to model the network changes over time.
-
-%These three uses of the term ``model"
-The model family, the model form, and the fitted model can each be visualized according to the three principals of model visualization: we can view the model in the data space, visualize collections of models, and explore the process of fitting the model, not just the end result.
+%For example, we chose to use a SAO \emph{model family} to model our network data over an exponential random graph model (ERGM) because we believe that actor-level variables effect network structure and formation, and we wanted to model the network changes over time.
+
+The model family, the model form, and the fitted model can each be visualized according to the three principles of model visualization: we can view the model in the data space, visualize collections of models, and explore the process of fitting the model, not just the end result.
 Since we have already decided on our model family, SAOMs, we shift our focus to the fitted model and the model form.
@@
 We begin by introducing the five models that we fit to our example data.
-We then use the five models to guide our visual explorations of SAOMs.
-We begin with novel tools and ideas to view a SAOM in the data space of a dynamic network.
+We then use the five models to guide our visual explorations of SAOMs under the three principles of model visualization.
+%We begin with novel tools and ideas to view a SAOM in the data space of a dynamic network.
 We then explore collections of the same models fit many times to the example data to see how the simulation processes in the \texttt{RSiena} ``black-box" affect the model fits.
@@
 Each SAOM used a simple rate function, $\alpha_m$, and an objective function with two or three parameters.
-The first model, M1, contains the absolute minimum number of parameters in the objective function $f_i(x)$:  $$f_i(x)^{M1} = \beta_1s_{i1} + \beta_2s_{i2},$$ where $s_{i1}$ is the density network statistic and $s_{i2}$ is the reciprocity network statistic for actor $i$ at the current network state $x$.
+The first model, M1, contains the absolute minimum number of parameters in the objective function $f_i(x)$:
+\begin{equation}\label{eq:m1}
+f_i(x)^{M1} = \beta_1s_{i1} + \beta_2s_{i2},
+\end{equation}
+where $s_{i1}$ is the density network statistic and $s_{i2}$ is the reciprocity network statistic for actor $i$ at the current network state $x$.
 The second and third models, M2 and M3, contain one additional parameter each in the objective function which were determined by a Wald-type test provided in the \texttt{RSiena} software to be significant, with $p$-values less than 0.05 \citep{RSienaTest}.
 The M2 model contains an actor-level covariate parameter, and the M3 model contains an additional strutural effect in the objective function.
-\begin{align*}
+\begin{align}\label{eq:m2m3}
 f_i(x)^{M2} & = \beta_1s_{i1} + \beta_2s_{i2} + \beta_3s_{i3} \\
 f_i(x)^{M3} & = \beta_1s_{i1} + \beta_2s_{i2} + \beta_4s_{i4},
-\end{align*}
+\end{align}
 
@@
 @
-\caption{\label{fig:jtt}Realization of a jumping transitive triplet, where $i$ is the focal actor, $j$ is the target actor, and $h$ is the intermediary.
-The group of the actors is represented by the shape of the node.}
+\caption{\label{fig:jtt}Realization of a JTT, where $i$ is the ego, $j$ is the alter, and $h$ is the intermediary. The covariate is represented by the shape of the nodes.}
 \end{subfigure}\hfill
@@
 @
-\caption{\label{fig:dad}Doubly achieved distance between actors $i$ and $k$.}
+\caption{\label{fig:dad}Realization of a N22 between actors $i$ and $k$.}
 \end{subfigure}
-\caption{The additional network effects included in our models fit to the friends data. On the left, a jumping transitive triplet (JTT). On the right, a doubly achieved distance between $i$ and $k$.}
+\caption{The additional network effects included in our models fit to the friends data. On the left, a JTT and on the right, a N22 between $i$ and $k$.}
 \end{figure}
@@
 where $s_{i3} = \sum_{j \neq h} x_{ij}x_{ih}x_{hj} \mathbb{I}(z_i = z_h \neq z_j)$, and $s_{i4} =  |\{j : x_{ij} = 0, \sum\limits_h x_{ih}x_{hj} \geq 2\}|$.
-These statistics are known as the \emph{number of jumping transitive triplets} and the \emph{number of doubly achieved distances two effect}, respectively.
+These statistics are known as the \emph{number of jumping transitive triplets} (JTT) and the \emph{number of doubly achieved distances two effect} (N22), respectively.
 The first statistic emphasizes triad relationships that are formed between actors from different covariate groups, while the other emphasizes indirect ties between actors.
-The covariate groups are determined by the student's drinking behavior, the values of which are converted to numeric then mean-centered.
-The four values in our data are $\{\text{never, 1-2 times per year, once per month, once per week}\}$.%, which after being converted to numeric and mean-centered become $\{-0.8125,-1.8125,0.1875,1.1875\}$.
-The two additional effects are visually represented and further described in Figure~\ref{fig:jtt} and Figure~\ref{fig:dad}, respectively.
+The covariate groups are determined by the student's drinking behavior.
+These additional effects are shown in Figure~\ref{fig:jtt} and Figure~\ref{fig:dad}.
+
+It is ofen difficult to know for certain how to interpret a fitted value of a parameter.
+We can make educated guesses based on the definition of the effect and the sign of the fitted value, but a direct interpretation is not always possible.
+By graphically exploring these models, we aim to understand their effects, their interpretations, and the model fitting process better.
 
@@
 To fit M2 to the senate data, we use the number of bills authored by each senator as the node covariate for the jumping transitive triplet variable.
-In terms of the senate data, then, the value of $\beta_3$ should dictate how willing a senator is to coauthor a bill on which the author of the bill has a different level of authorship, assuming there is an intermediary between the two who has the same authorship level as the first.
+%In terms of the senate data, then, the value of $\beta_3$ should dictate how willing a senator is to coauthor a bill on which the author of the bill has a different level of authorship, assuming there is an intermediary between the two who has the same authorship level as the first.
 %Comparison of the parameter estimates of these models fit to the two different data sets is given in Table~\ref{tab:meanests}.
@@
 If the value of the parameter $\beta_6$ is high, nodes belonging to the same covariate group are encourage to reciprocate incoming ties.
-Due to the intractability of SAOMs, it is difficult to know for certain how to interpret a fitted value of a parameter.
-We can make educated guesses based on the definition of the effect and the sign of the fitted value, but a direct interpretation is not always possible.
-By graphically exploring these models, we aim to understand their effects, their interpretations, and the model fitting process better.
 
@@
 
-The first way we hope to better understand stochastic actor-oriented models is by viewing the model(s) in the data space.
+The first way we hope to better understand SAOMs is by viewing the model(s) in the data space.
 In \citeauthor{modelvis}, they chose define the \textit{data space} as ``the region over which we can reliably make inferences, usually a hypercube containing the data" \citep[p.~206]{modelvis}.
 But what does this definition mean for network models?
-We want to make inferences on parameters in the model, but the multi-level data structure of networks makes the ``region of inference" difficult to define.
+We define the data space as the combined data structures of the network: the node, edge, and time information.
+%We want to make inferences on parameters in the model, but the multi-level data structure of networks makes the ``region of inference" difficult to define.
 % %For dynamic social networks, there are a few different data ``spaces":
@@
 
-The node, edge, and time data can be visualized together in various ways.
-Because longitudinal network data consist of three different ``spaces" of data, viewing the model in the data space can depend on which aspect of the \textit{model} and the \textit{data} we are interested in viewing.
+%The node, edge, and time data can be visualized together in various ways.% Because longitudinal network data consist of three different ``spaces" of data, viewing the model in the data space can depend on which aspect of the \textit{model} and the \textit{data} we are interested in viewing.
 %The SIENA software includes a method for modelling covariate change in the network
-Incorporating data covariates into the network structure allows us to assess whether the ties between nodes are affected by how nodes behave over time.
+%Incorporating data covariates into the network structure allows us to assess whether the ties between nodes are affected by how nodes behave over time.
 In this instance, we would want to view predictions of edge presence over time as node variables change, visualizing all three data spaces at once.
@@
 
-One tool that can bring the node, edge, and time data spaces together is the \texttt{R} package \texttt{geomnet} \citep{geomnet}.
+The node, edge, and time data can be visualized together in various ways, and one tool that can bring the node, edge, and time data spaces together is the \texttt{R} package \texttt{geomnet} \citep{geomnet}.
 Different visual features in the node-link diagram can be tied to the underlying node or edge data.
 The color, size, and shape of the points can be used to represent variables in the node data, while the color, linewidth, and linetype of the lines between points can be used to represent the edge variables.
-In a social network, node data might be age, gender, and/or occupation of the person in the network, and edge data might be length of connection between two people, how the people first met (school, work, church, etc.), and/or how often they interact.
+%In a social network, node data might be age, gender, and/or occupation of the person in the network, and edge data might be length of connection between two people, how the people first met (school, work, church, etc.), and/or how often they interact.
 To view temporal changes, we can view the network at different timepoints side-by-side to see the evolution.
@@
 
-<<funsenatevis, fig.height=3, fig.cap='The 111th Senate at two discrete time points: while Clinton was in the senate in the first few weeks of 2009 (on the left) and after she left the senate to become Secretary of State (on the right).
-We put some potential model covariates, sex and party of the senator and the number of bills they authored in the data space through the shape, color, and size of the nodes, respectively.
-We also map the strength of the tie, the $WPC$, to the width of the edge between two nodes.
-Thicker lines implies higher propensity to collaborate.
-In addition, senators with no ties higher than $WPC = 0.25$ are not shown.'>>=
+<<funsenatevis, fig.height=3, fig.cap='The 111th Senate at two times: while Clinton was in the senate in 2009 (on the left) and after she left (on the right).
+We map sex, party, and bills authored to the shape, color, and size of the nodes, respectively.
+We also map the $WPC$, to the edge linewidth.'>>=
 se111clint <- filter(seobama, senate == 111)
@@
 In that time, she authored two bills and was a cosponsor on 17 other bills.
-There are many Democratic senators who cosponsored both of her bills, giving their edge a $WPC$ of 1.
-So with Clinton included in the node-link diagram, the senate looks much more collaborative than it does without her in the diagram.
-We can compare the Senators' prolificness by mapping the size of the node to the numer of bills authored by the senator.
-We also map shape to sex, and keep color mapped to party of the senator.
+There are many Democratic senators who cosponsored both of her bills, giving their edges a $WPC$ of 1.
+So with Clinton included in the node-link diagram, the senate looks much more collaborative than it does without her.
+We then map the size of the node to the numer of bills authored by the senator, the shape to sex, color to party.
+Bills, sex, and party are all potential covariates in a SAOM.
 In addition, we visualize the strength of the tie by mapping the linewidth of the edge to the $WPC$ value between the two senators.
-In this single visualization, we have viewed node information (number of bills authored by a senator and the sex and party of that senator), edge information (the direction and strength of ties between two senators), and time (before and after Clinton left the senate).
-To view the model in the data space, plot simulations from different models over time with applicable node and edge variables using \texttt{geomnet}.
-% ST: I don't like this paragraph
-% The visual aspects of the node-link diagram connecting the node, edge, and time data spaces are themselves part of the data space, because they can be viewed in different ways.
-Another way to connect the node and the edge data, as discussed in Section~\ref{sec:netvizintro}, is through the network's adjaceny matrix.
-The time dimension can also be incorporated with side-by-side adjacency matrix visualizations.
-
-% below chunk contains some repetitive stuff. Haven't figured out what it's purpose is yet.
+In this single visualization, we have viewed node information, edge information, and time.
+We will use \texttt{geomnet} many ways to view the model in the data space.
+
 <<summnetM1, message=FALSE, fig.cap='On the left, the first wave of observed data that is conditioned on in the model.
@@
 friendshipData <- array(c(fd2.w1, fd2.w2, fd2.w3), dim = c(16, 16, 3))
-
+friendshipData <- array(c(fd2.w1, fd2.w2, fd2.w3), dim = c(16, 16, 3))
 drink1 <- drink[20:35,1]
@@
                        N = 1000)
-M1simsdf <- sims_to_df(M1sims1000)
-M2simsdf <- sims_to_df(M2sims1000)
-M3simsdf <- sims_to_df(M3sims1000)
+M1simsdf <- netvizinf::sims_to_df(M1sims1000)
+M2simsdf <- netvizinf::sims_to_df(M2sims1000)
+M3simsdf <- netvizinf::sims_to_df(M3sims1000)
 M1avgW2 <- M1simsdf %>% filter(!is.na(to) & wave == 1) %>%
@@
 Another way to view the model in the data space is through simulation from the model.
-No one network simulated from a SAOM is going to look like the data or represent the model, just as no single value simulated from the standard normal distribution will look like a bell-shaped curve.
-As statisticians, we prefer to look at a sample, or at least a summary statistic or two, so it would therefore better to visualize many simulated networks at once.
-A stumbling block with network models generally is the lack of an ``average network" or ``expected network" value.
-We frequently rely on averages and expected values in data analyses, but network models, especially those as complex as SAOMs, lack a single, intuitive expected value measure.
+No one network simulated from a SAOM is going to look like the data or represent the model, so we might want to look at an ``average network" or ``expected network" value.
+We frequently rely on averages and expected values in data analyses, but network models, especially those as complex as SAOMs, lack a expected network value measure.
 We could talk about expected values of parameters, but the parameters can be hard to interpret.
-Expected values of parameters can be computed, but if they cannot directly tell us anything about our dependent variable, they lose value.
-Furthermore, there is no way to talk about the expected value of an observation simulated from a statistical network model, though there are with most other model simulations.
+%Expected values of parameters can be computed, but if they cannot directly tell us anything about our dependent variable, they lose value.
+There is no defined expected value of an simulation from a network model, though there are with most other model simulations.
 How then, can we arrive at an ``average" network?
 We answer this question through visualization.
-For network data, one way we view an ``average" network is through a \textit{summary network} drawn using the traditional node-link diagram.
+
+For network data, one way we view an ``average" network is through a sory of summary network drawn using the traditional node-link diagram.
 In Figure~\ref{fig:summnetM1}, we show an \textit{average network} created with 1,000 simulations of the second wave of the network from Model 1.
-
 To make this average network, we first simulated 1,000 wave 2 and wave 3 observations of our small friendship example data from model M1, for which parameters had previously been estimated.
@@
 Then, we combine these 1,000 networks into a single network with edgeweight equal to the proportion of time that edge appears in our 1,000 simulations.
-This weighted edgelist is the network we draw using the node-link diagram.
+This weighted edgelist is the network we draw in Figure~\ref{fig:summnetM1}.
 An edge is only drawn in the average network if it appears in at least 51 of the 1000 simulations, with edges that appear with greater frequency emphasized by thicker linewidths and a darker color.
@@
 We can see that the structure of the average network is much more similar to the first wave than to the second wave.
-However, the simulations are supposed to represent the second wave of data on the right of Firgure~|ref{fig:summnetM1}.
+However, the simulations are supposed to represent the second wave of data on the right of Figure~\ref{fig:summnetM1}.
 This is an indication that the simple model, M1, is doing a very poor job of capturing the change mechanism from the first to the second wave of observation.
 The average network can thus be used to help determine model goodness-of-fit.
-Because the the average network looks more like the first wave than the second wave, we can use the visualization in Figure \ref{fig:summnetM1} as evidence of poor model fit.
-% XXX ST - I was looking at the wrong one.
-I had wave == 2 instead of wave == 1 in the code.
-This means I was looking at the wrong simulations, so they really do look more like the first data obs than the second obs.
-XXX Comparing the three visualizations side-by-side shows us that the summary network of simulations of wave 2 looks more like the true wave 2 than wave 1, which is a good indication that the model is capturing the true data structure.
-It's not ideal, however, because there appear to be too many reciprocated ties in the summary network, far more than there are in the data.
-This could indicate that we should remove the reciprocity parameter from the model.
+Because the the average network looks more like the first wave than the second wave, we can use the visualization in Figure~\ref{fig:summnetM1} as evidence of poor model fit.
 
@@
 
-Another potential goodness-of-fit visualization that places the model in the node and edge data space is a lineup like those proposed in \citet{buja:2009}.
+Another potential goodness-of-fit visualization that places the fitted model in the node and edge data space is a lineup like those proposed in \citet{buja:2009}.
 A \textit{lineup} ``asks the witness to identify the plot of the real data from among a set of decoys, the null plots, under the veil of ignorance" \citep[p.
@@
 If, however, the true data is difficult to identify among the simulated data, that is taken as evidence in favor of the hypothesized model.
-An application of the lineup protocal can be found in \citet{hofmann2012graphical}, where the authors examine, for instance, the differences between polar and cartesian coordinates for plotting categorical data, and density plots and box plots for determining distributional differences.
+% An application of the lineup protocal can be found in \citet{hofmann2012graphical}, where the authors examine, for instance, the differences between polar and cartesian coordinates for plotting categorical data, and density plots and box plots for determining distributional differences.
 They pose questions to experiment participants such as, ``Which plot is most different from the others?" or, ``In which plot is the blue group furthest to the right?"
@@
 The data visualizations examined in previous applications such as  \citet{hofmann2012graphical} are less complex than a node-link network diagram.
-What questions should we ask for network data visualizations?
-The most neutral choice, asking participants to identify the most different plot, may be difficult.
+Typically, the lineup viewers have to pick out the most different plot, which is hard for network data.
 As an example, consider the network lineup shown in Figure~\ref{fig:lineup}.
 In this small network lineup, the second wave of the small friendship network is shown among five simulated networks from model M1 using parameter values estimated from the data.
-What exactly makes these plots ``different" from one another?
-
 It seems possible to argue for any one of the six plots in Figure~\ref{fig:lineup} as most different.
-So, we suggest guiding participants to look at the overall structure of the graphs to determine which has the most or least connected or complex structure.
-The least complex plot, number six, has no triangles (3-cycles), while the most complex plot, number three, has three 3-cycles.
+So, we suggest guiding participants to look at the overall structure of the graphs to determine which has the most or least connected or complex structure.% The least complex plot, number six, has no triangles (3-cycles), while the most complex plot, number three, has three 3-cycles.
 The most complex plot, number three, is in fact the data.
-We have found that when shown lineups like that of Figure~\ref{fig:lineup}, online survey participants %where all the node-link diagrams shown are based on simulated data, %one observation from an ``alternative" model and the rest from a ``null" model,
-pick up on the triangular shape of the 3-cycles most often.
-
+We have found that when shown lineups like that of Figure~\ref{fig:lineup}, online survey participants %where all the node-link diagrams shown are based on simulated data, %one observation from an ``alternative" model and the rest from a ``null" model, %pick up on the triangular shape of the 3-cycles most often.
 In a future paper, we will use the lineup protocol to help better understand SAOMs by allowing us to view the model in the data space.
-Lineups place observations simulated from the model side-by-side with the data and let the viewer examining the differences.
+%Lineups place observations simulated from the model side-by-side with the data and let the viewer examining the differences.
 This can also help us determine the significant structural effects, if any, of the parameters in the model on observations simulated from the model.
@@
 The second method of model visualization we use is visualizing collections of models.
-There are many possible ways to collect SAO models together.
+There are many possible ways to collect SAO models together, so %.
 We could look at the same model fit to different data, different models fit to the same data, or we could fit the same models to the same data many times to see how the simulations change.
-For the SAOMs, we decided that there were four collections of models that would give us the most insight:
+F
+for the SAOMs, we decided that there were four ways to create collections of models that would give us the most insight:
 \begin{enumerate}
-\item the collections resulting from exploring the space of all possible models;
-\item the collections we get when varying model settings;
-\item the results from fitting the same model form to different data;
-\item the results from fitting the same model to the same data many times.
+\item exploring the space of all possible models;
+\item varying model settings;
+\item fitting the same model form to different data;
+\item fitting the same model to the same data many timess
 \end{enumerate}
-
 We chose these four collections because they each explore something different about SAOMs.
-The first considers the many dozens of parameters available to include in a SAOM.
-The second collection show how those many parameters affect the parameter values of the fitted models.
-Third, looks at how the same model behaves when fit to different sets of dynamic network data.
-Finally, the results from fitting the same model to the same data may lead to different parameter estimates, so it important to see how the results can vary.
+The first considers the many parameters to include in a SAOM.
+The second collection shows how those many parameters affect the parameter values of the fitted models.
+The third looks at how the same model behaves when fit to different data.
+Finally, the results from fitting the same model to the same data may lead to different parameter estimates, so we want to see how the parameter estimates can vary.
 
@@
 
-The \texttt{RSiena} manual contains over eighty possible effects to include in the model.
+The \texttt{RSiena} manual contains over eighty effects to include in a SAOM.
 In order to select parameters to include in the models for our example data, we searched through the possible effects available to model given the data structure to find \textit{significant} effects.
 We tested for significance using the Wald-type tests built into \texttt{RSiena} for one-at-a-time effects testing.
-We start with the outdegree and reciprocity measures as the foundation of the models we fit, then add one evaluation effect, fit the model, test the additional effect for significance, and repeat for all possible parameters to add to the model.
+We start with the outdegree and reciprocity measures as the foundation of the models we fit, then add one effect, fit the model, test the additional effect for significance, and repeat for all possible parameters to add to the model.
 We performed this procedure for both the small friendship and the senate collaboration data.
-A visualization of the significant effects for the two example data sets are shown in the appendix.
+A visualization of the significant effects for the two example data sets are shown in Figure~\ref{fig:sigtesting} in the Appendix.
 %We see in both the friendship data and the senate data results that most of the significant effects have absolute value less than ten.
@@
 From each of these three models using the means of parameter estimates as our fixed parameter values, we simulated 1,000 observations from each of the three models.
-In this process, we condition on the first friendship network observation, and the second and third observations are simulated from the SAOM models with the given parameter values.
-From these simulations, we create visualizations that represents an average network using the same method described in Section~\ref{sec:minds}.
-These results are provided in the appendix.
-
-%XXXX ST commented out after I realized I had made the same mistake as figure 7.
-XX This could be an indication that, despite the highly significant nature of the parameter $\beta_3$ according to the $t$-tests in \texttt{RSiena}, the transitive triplet parameter should not be included because it does not result in a structure similar to the data.
-
-% \st{What are some things that could go in here?:
-% \begin{itemize}
-% \item "average" networks -- you've already done this! Write it up, make several examples. How does the procedure for averaging generalize? What is it good for?
-% \item distributions of fitted parameters. again, you've already done this!!! just write it up and put it in here.
-% \item view correlations between model parameters under different model sets. eg 2 betas, 3 betas, 4 betas, etc.
-% \end{itemize}
-%  }
+From these simulations, we create visualizations that represents an average network using the same method described in Section~\ref{sec:minds}, shown in Figure~\ref{fig:averagenet} in the Appendix.
 
@@
 
-As mentioned in Section~\ref{sec:models}, we fit the models M1, M2, and M3 to both the small friendship network and the senate collaboration network.
-We fit each model to the friendship data 1,000 times, and to the senate data 100 times.
+In addition to fitting M1, M2, and M3 to the friendship data 1,000 times, we also fit them to the senate data 100 times.
 The means and standard deviations of the parameter estimates for each combination of model and data are given in Table~\ref{tab:meanests}, while the density plots of each parameter in the model for each data set is given in Figure~\ref{fig:compareModels}.
@@
 }
-\caption{\label{tab:meanests} The means (standard deviations) of parameter values estimated from repeated fittings of $M1, M2, M3$ to the small friendship network and the senate collaboration network.
-Each model was fit 1,000 times to the friend data, while each model was fit 100 times to the senate data.}
+\caption{\label{tab:meanests} The means (std. dev.) of parameter values estimated from repeated fittings of $M1, M2, M3$ to the two example data sets.}
 \end{table}
 
-<<compareModels, fig.cap="Density plots of the objective function estimates from repeatedly fitting models M1, M2, and M3 to the friendship and senate example data.
-The first two parameters, outdegree and reciprocity, have the same inverse relationship for both data sets.
-For the friendship data, the inclusion of the transitive triplet parameter has strong effect on the estimates of the other two parameters, while it does not affect the estimates of the other two parameters for the senate data.", fig.height = 3>>=
+<<compareModels, fig.cap="Density plots of objective function parameter estimates from repeatedly fitting models M1, M2, and M3 to the example data.
+Outdegree and reciprocity have the same inverse relationship for both data sets.
+For the friendship data, the inclusion of $\\beta_3$ has strong effect on the estimates of the other two parameters, but not for the senate data.", fig.height = 3>>=
 simu2$data <- "friends"
@@
 This suggests that in both data sets, nodes are \textit{encouraged} to form outgoing ties that are reciprocated and \textit{discouraged} from forming ties that are not.
-People seem to want to have reciprocated relationships: teenage girls want be friends with other girls that reciprocate their friendship, and senators want to coauthor bills with senators who have also been coauthors on their bills.
+Actors in these networks want reciprocated relationships: the students want to have close mutual friendships, and senators want mutual support for their bills.
 We explore the relationship between $\beta_1$ and $\beta_2$ further in Section~\ref{sec:samemodsamedat}.
 
-The inclusion of $\beta_3$, the jumping transitive triplet parameter, for the friendship data had a noticeable effect on the other parameters in the model.
-The same cannot be said for the inclusion of $\beta_3$ for the senate data.
-The covariate used in the senate data for the jumping transitive triplet calculation was the number of bills authored by the ego node in the given year.
+The inclusion of $\beta_3$, the JTT parameter, for the friendship data had a noticeable effect on the other parameters in the model.
+This is not true in the senate data.
+%The covariate used in the senate data for the JTT calculation was the number of bills authored by the ego node in the given year.
 The number of bills authored by a senator varies wildly, from no bills to 114 bills authored in two years, so this effect could simply be nonsensical for the senate data, since senators are less likely to have the same number of bills authors than teenage girls are to have the same drinking behavior.
@@
 
-To our small friendship network, we fit three models, M1, M2, and M3, using \texttt{RSiena} 1,000 times each.
-We then looked at the distribution of the fitted values, which are shown in Figure~\ref{fig:distests}.
+After fitting models M1, M2, and M3 to the friendship network 1,000 times each, we examine the distribution of the fitted values, which are shown in Figure~\ref{fig:distests}.
 We can see from these distributions that the inclusion of the jumping transitive triplet parameter, $\beta_3$ is obviously affecting the distributions of the other four parameters included in all models, $\alpha_1$, $\alpha_2$, $\beta_1$, and $\beta_2$.
@@
 In comparison with models M1 and M3, model M2 typically has higher estimates of the rate parameter, meaning that the inclusion of the covariate statistic in the model leads to higher estimates of the number of times, on average, a node gets to change its ties.
-It is not clear, however, that the addition of a parameter to the objective function \textit{should} effect the estimation of the rate parameters, so we continue to explore the collection of parameter estimates.
+It is not clear, however, that the addition of a parameter to the objective function \textit{should} effect the estimation of the rate parameters.
+The strong correlations between the estimates of the rate parameters and of the objective function parameters are shown in Figure~\ref{fig:corplots} in the Appendix.
 
@@
 @
-
-Finally, to further investigate the odd relationship between the parameter values, we look at correlations between each of the parameter estimates in each model.
-These results are presented in the Appendix.
 
@@
 
-\subsection{Explore algorithms, not just end result}
+\subsection{Explore algorithms, not just end result}\label{sec:algorithms}
 
@@
 The top row of Figure~\ref{fig:makinggif} shows an edge being removed, and the bottom row shows one being added.
-In both cases, the ego node chosen to act change color from black to red, and they also increase greatly in size.
+In both cases, the ego node that is acting changes color from black to red, and they also increase greatly in size.
 In the case of an edge being removed, in the top row, the edge that currently exists is emphasized with the same color and size change that the node gets, and as the animation proceeds the edge shrinks to nothing, as the ego node shrinks and changes color back to its original black.

### 2018-07-11 paper.Rnw
ugh get this done

@@
 %                 Editing Commands
-\newcommand{\hh}[1]{{\color{magenta} #1}}
-\newcommand{\st}[1]{{\color{orange} #1}}
+%\newcommand{\hh}[1]{{\color{magenta} #1}}
+%\newcommand{\st}[1]{{\color{orange} #1}}
 
@@
 \begin{abstract}
-Social networks have been studied for decades, beginning with a few foundational works, including the 1967 study, ``The Small World Problem" by Stanley Milgram.
-In this paper, we concentrate on one type of model for dynamic social networks: the stochastic actor-oriented models (SAOMs), introduced by Snijders (1996).
+Stochastic actor-oriented models, first introduced by Snijders (1996), are a type of statistical network model for dynamic social networks.
 Unlike other network models, SAOMs are not very well understood.

### 2018-07-12 paper-child.Rnw
most recent version -- just needs discussion section work now.

@@
 
-The inclusion of $\beta_3$, the JTT parameter, for the friendship data had a noticeable effect on the other parameters in the model.
+The inclusion of $\beta_3$ for the friendship data had a noticeable effect on the other parameters in the model.
 This is not true in the senate data.
@@
 The number of bills authored by a senator varies wildly, from no bills to 114 bills authored in two years, so this effect could simply be nonsensical for the senate data, since senators are less likely to have the same number of bills authors than teenage girls are to have the same drinking behavior.
-Looking at the estimates of $\beta_4$, we see that the estimates for the senate data are near zero, suggesting that this effect, which considers indirect ties, is not important for the senate data.
-It is, however, much stronger in the friendship data, suggesting that indirect ties are discouraged from forming: teenage girls want to have direct friendships.
-This indirect relationship variable does not impact the senate collaboration structure, but it does affect teenage friendship structure.
+Looking at the estimates of $\beta_4$, we see that the estimates for the senate data are nearer to zero, suggesting that this effect, which considers indirect ties, is not important for the senate data.
+This effect is, however, very strong in the friendship data with a mean of about -7.5, suggesting that indirect ties are discouraged from forming: teenage girls want to have direct friendships.
+This indirect relationship variable does not help explain the senate collaboration structure, but it does help clarify the teenage friendship structure.
 
@@
 When $\beta_3$ is included, its estimate is postitive, meaning that friendships between two girls with different drinking behaviors tend to form when there is an intermediary who is already friends with the two girls.
-The inclusion of this parameter leads to increases in the rate parameters' estimates, suggesting that encouraging the transitive triplet behavior means that the girls would also change friends more frequently.
-The outegree parameter, $\beta_1$ decreases when $\beta_3$ is included, while the reciprocity parameter, $\beta_2$ increases.
-This implies the girls in the data prefer to form closer friend groups, as indicated by reciprocated ties and jumping transitive triplet formation, as opposed to being popular and having many friends.
-Having many friends who do not reciprocate is more stronlgy discouraged by M2 than in the other models.
-In comparison with models M1 and M3, model M2 typically has higher estimates of the rate parameter, meaning that the inclusion of the covariate statistic in the model leads to higher estimates of the number of times, on average, a node gets to change its ties.
+The inclusion of this parameter leads to increases in the rate parameters' estimates, suggesting that the presence of the transitive triplet behavior means that the girls would also change friends more frequently.
 It is not clear, however, that the addition of a parameter to the objective function \textit{should} effect the estimation of the rate parameters.
 The strong correlations between the estimates of the rate parameters and of the objective function parameters are shown in Figure~\ref{fig:corplots} in the Appendix.
+In addition, the outegree parameter, $\beta_1$ decreases when $\beta_3$ is included, while the reciprocity parameter, $\beta_2$ increases.
+This implies the girls in the data prefer to form small, closely knit friend groups, as indicated by reciprocated ties and jumping transitive triplet formation, as opposed to being popular and having many friends.
+Having many friends who do not reciprocate is more stronlgy discouraged by M2 than in the other models.
+%In comparison with models M1 and M3, model M2 typically has higher estimates of the rate parameter, meaning that the inclusion of the covariate statistic in the model leads to higher estimates of the number of times, on average, a node gets to change its ties.
 
@@
 
-The last principle of model visualization we use is exploring the process of fitting the model, instead of just focusing on the end result.
+The last principle of model visualization we use is to explore the process of fitting the model, instead of just focusing on the end result.
 This principle is perhaps the most important for SAOMs because the model fitting process in \texttt{RSiena} involves several simulation steps that are hidden from the user.
@@
 
-A key component of each step of the SIENA method of moments algorithm is the ``microstep" process.
+One of the hidden steps in the SIENA method of moments algorithm is the ``microstep" process.
 A series of microsteps is obtained by simulating from the model in its current state, $x(t_m)$ with current parameter values $\mathbf{\theta}_{0} = \{\alpha_{1_0}, \dots \alpha_{{m-1}_0}, \beta_{1_0}, \dots, \beta_{K_0}\}$, to the next state, $x(t_{m+1})$.
-This microstep process stops when the simulated network has achieved the same number of differences, $C$, from $x(t_m)$ as $x(t_{m+1})$, where $C$ is definined in Equation~\ref{eq:cdiff}.
+One tie changes at a time until the simulated network has achieved the same number of differences, $C$, from $x(t_m)$ as $x(t_{m+1})$, where $C$ is definined in Equation~\ref{eq:cdiff}.
 
@@
 Between two network observations $x(t_m)$ and $x(t_{m+1})$, there can be dozens, hundreds, or even thousands of microsteps, depending on the size of the network and the number of changes between two network observations.
-We want to view these in-between steps in order to better understand the behavior of the underlying continuous-time Markov chain.
-
-The first vizualization we present here is an animation of the simulated microsteps that form the transition steps of the CTMC from wave 1 to wave 2 of the small friendship network example shown in Figure~\ref{fig:alldataviz} when fitting model M1.
+We want to simulate and view these in-between steps in order to better understand the behavior of the underlying continuous-time Markov chain.
+
+The first vizualization we present here is an animation of the simulated microsteps, when fitting model M1, that form the transition steps of the CTMC from wave 1 to wave 2 of the small friendship network example shown in Figure~\ref{fig:alldataviz}.
+.
 Movies similar to this animation were used to visualize the changes of dynamic networks in \citet{dynnetviz}.
@@
 If there are no changes at a particular microstep, no changes are seen.
-Some frames of the animations are shown in Figure~\ref{fig:makinggif} in the Appendix, and the full movie can be viewed at \url{https://vimeo.com/240089108}.
+Some frames of the animations are shown in Figure~\ref{fig:makinggif} in the Appendix, and the full movie, created with \texttt{tweenr} and \texttt{gganimate} can be viewed at \url{https://vimeo.com/240089108} \citep{tweenr, gganimate}.
 
@@
 To reorder the vertices for the matrix visualization, we first constructed a cumulative adjaceny matrix, $\mathbf{A}^{cum}$, for the series of microsteps simulating the network from $x(t_m)$ to $x(t_{m+1})$.
-A single entry in the cumulative adjacency matrix, $\mathbf{A}^{cum}_{ij}$, is the total number of times the edge from node $i$ to node $j$ appears in the network from the initial observation, $x(t_m) \equiv X(0)$ to the final result of the last microstep, $X(R)$, where $R$ is the total number of microsteps taken:
+A single entry in the cumulative adjacency matrix, $\mathbf{A}^{cum}_{ij}$, is the total number of times the edge $i \righarrow j$ appears in the network from the initial observation, $x(t_m) \equiv X(0)$ to the final result of the last microstep, $X(R)$, where $R$ is the total number of microsteps taken:
 $$ \mathbf{A}^{cum}_{ij} = \sum_{r = 0}^R X_{ij}(r). $$
@@
 
-We also attempt to better understand the microstep process by visualizing the observed transition probabilites for the first microstep in the process.
+We also attempt to better understand the microstep process by visualizing the empirical transition probabilites for the first microstep in the process.
 These probabilities are defined in Equation~\ref{eq:pij}.
@@
 This effect could only be exacerbated as more steps are taken in the CTMC, leading to
-a very large area of our network space, $\mathcal{X}$, completely unexplored by the SAOM model fitting process.
+a very large area of our network space, $\mathcal{X}$, unexplored by the first steps in the SAOM model fitting process.
+We present additional evidence for lack of coverage in Figure~\ref{fig:alledgests} the Appendix.
+The Figures~\ref{fig:heatmap} and \ref{fig:alledgests} show that the model M1 and the SAOM fitting process do not explore the data space enough to adequately capture the network change mechanism, a finding we were only able to discern thanks to model visualization techniques.
 
@@
 
-<<alledgests, fig.height = 4, fig.cap="Visualizing all microsteps taken in 1,000 simulations from the model M1.
-The occurrence percent is split up into groups to correspond with its distribution: only about 10\\% of the edges appear more than 10\\% of the time in the 1,000 simulations, while about 60\\% appear less than 1\\% of the time.
-The first wave network is shown at microstep 0, and the second wave of the network is shown as the last microstep for comparison.
-We see that it is rare for a microstep process to last longer than 150 steps, and also that the edges that appear past the 150th step tend to be in either the first wave or the second wave.">>=
-# get a plot of all microsteps from w1 to w2 by edge
-alledges <- expand.grid(from = paste0("V", 1:16), to = paste0("V", 1:16))
-alledges$eid <- 1:nrow(alledges)
-
-sfmsall %>% left_join(alledges) -> sfmsall2
-
-sfmsall2 %>% group_by(rep) %>% mutate(isWave2 = (ms == max(ms))) -> sfmsall2
-
-sfmsall2 %>% group_by(from, to, eid) %>%
-  summarize(total =n()) %>%
-  arrange(desc(total)) %>%
-  ungroup %>%
-  mutate(plotorder = row_number()) -> sfmsallorder
-
-sfmsall2 %>% group_by(ms, eid) %>%
-  summarise(count = n()) %>%
-  mutate(alpha = count/1000) %>%
-  arrange(ms, eid) -> sfmsall3
-
-left_join(sfmsall3, sfmsallorder, by = "eid") -> sfmsall4
-
-#add true wave 2 to plot
-wave2 <- na.omit(actual2)
-wave2$from <- paste0("V", wave2$from)
-wave2$to <- paste0("V", wave2$to)
-wave2 <- left_join(wave2, alledges)
-wave2 <- left_join(wave2, sfmsallorder)
-
-sfmsall5 <- sfmsall4 %>% ungroup %>% add_row(ms = max(sfmsall4$ms) + 1,
-        eid = wave2$eid,
-        count = NA,
-        alpha = 1,
-        from = wave2$from,
-        to = wave2$to,
-        total = wave2$total,
-        plotorder = wave2$plotorder)
-sfmsall5$alphaFctr <- cut(sfmsall5$alpha, c(0,.01, .05, .10, .25, .5, .75, 1, 1.1))
-ggplot(data = sfmsall5) +
-  geom_tile(aes(x = ms, y = reorder(eid,-plotorder), fill = alphaFctr)) +
-  #scale_fill_gradient(low = 'grey90', high = '#0868ac', name = "occurrence %") +
-  scale_fill_brewer(palette = "Blues", name = "Occurrence %", labels = c("(0-1]%", "(1-5]%", "(5-10]%","(10-25]%", "(25-50]%", "(50-75]%", "(75-100]%")) +
-  labs(x = "Microstep No.", y = "Edges (ordered by total # occurrences)") +
-  ThemeNoNet +
-  theme(axis.ticks.y = element_blank(), axis.text.y = element_blank(), panel.grid = element_blank())
-@
-
-Finally, we combine 1,000 simulations from model M1 into a visualization that displays the entire microstep process for all simulations in Figure ~\ref{fig:alledgests}.
-To make this visualization, we first assign each possible edge an edge ID number so that we can keep track of it throughout all the microsteps and all the simlations.
-Then, we count up the total number of times each edge appears in the microstep process in each of the 1,000 simulations for use as an ordering variable later.
-We also count up the number of times an edge occurs in each microstep number in the 1,000 simulations.
-Since the number of microsteps in the process varies, the number of times an edge occurs decreases as the microstep number increases.
-Next, we compute a proportion, which we call the occurrence percentage, which is the number of times the edge occurred in a microstep divided by 1,000.
-Finally, we visualize all this information together in Figure~\ref{fig:alledgests}.
-In this plot, all possible edges are shown, and we see that every one of the $16 \times 15 = 240$ possible edges in the network occurs at some point in the simulation process.
-We also see, however, that the process struggles to focus in on the edges in the second wave of the data.
-Ideally, we would like to see more occurrences of the edges which appear in the second wave of data.
-But, about half of the edges in wave two are in the bottom half when ordered by number of occurrences, meaning they do not appear as much as they would if the model was truly excellent at capturing the mechanisms of tie change in the network.
-This solidifies what we found in Figure~\ref{fig:heatmap}: the model M1 and the SAOM fitting process do not explore the data space enough to adequately capture the network change mechanism.
  %In addition, there is not a distinct pattern or structure that we can see in this particular edge visualization.
@@
 @
+
+Finally, we combine 1,000 simulations from model M1 into a visualization that displays the entire microstep process for all simulations in Figure ~\ref{fig:alledgests}.
+To make this visualization, we first assign each possible edge an edge ID number so that we can keep track of it throughout all the microsteps and all the simlations.
+Then, we count up the total number of times each edge appears in the microstep process in each of the 1,000 simulations for use as an ordering variable later.
+We also count up the number of times an edge occurs in each microstep number in the 1,000 simulations.
+Since the number of microsteps in the process varies, the number of times an edge occurs decreases as the microstep number increases.
+Next, we compute a proportion, which we call the occurrence percentage, which is the number of times the edge occurred in a microstep divided by 1,000.
+Finally, we visualize all this information together in Figure~\ref{fig:alledgests}.
+In this plot, all possible edges are shown, and we see that every one of the $16 \times 15 = 240$ possible edges in the network occurs at some point in the simulation process.
+We also see, however, that the process struggles to focus in on the edges in the second wave of the data.
+Ideally, we would like to see more occurrences of the edges which appear in the second wave of data.
+But, about half of the edges in wave two are in the bottom half when ordered by number of occurrences, meaning they do not appear as much as they would if the model was truly excellent at capturing the mechanisms of tie change in the network.
+This solidifies what we found in Figure~\ref{fig:heatmap}: the model M1 and the SAOM fitting process do not explore the data space enough to adequately capture the network change mechanism.
+
+<<alledgests, fig.height = 4, fig.cap="Visualizing all microsteps taken in 1,000 simulations from the model M1.
+The occurrence percent is split up into groups to correspond with its distribution: only about 10\\% of the edges appear more than 10\\% of the time in the 1,000 simulations, while about 60\\% appear less than 1\\% of the time.
+The first wave network is shown at microstep 0, and the second wave of the network is shown as the last microstep for comparison.
+We see that it is rare for a microstep process to last longer than 150 steps, and also that the edges that appear past the 150th step tend to be in either the first wave or the second wave.">>=
+# get a plot of all microsteps from w1 to w2 by edge
+alledges <- expand.grid(from = paste0("V", 1:16), to = paste0("V", 1:16))
+alledges$eid <- 1:nrow(alledges)
+
+sfmsall %>% left_join(alledges) -> sfmsall2
+
+sfmsall2 %>% group_by(rep) %>% mutate(isWave2 = (ms == max(ms))) -> sfmsall2
+
+sfmsall2 %>% group_by(from, to, eid) %>%
+  summarize(total =n()) %>%
+  arrange(desc(total)) %>%
+  ungroup %>%
+  mutate(plotorder = row_number()) -> sfmsallorder
+
+sfmsall2 %>% group_by(ms, eid) %>%
+  summarise(count = n()) %>%
+  mutate(alpha = count/1000) %>%
+  arrange(ms, eid) -> sfmsall3
+
+left_join(sfmsall3, sfmsallorder, by = "eid") -> sfmsall4
+
+#add true wave 2 to plot
+wave2 <- na.omit(actual2)
+wave2$from <- paste0("V", wave2$from)
+wave2$to <- paste0("V", wave2$to)
+wave2 <- left_join(wave2, alledges)
+wave2 <- left_join(wave2, sfmsallorder)
+
+sfmsall5 <- sfmsall4 %>% ungroup %>% add_row(ms = max(sfmsall4$ms) + 1,
+        eid = wave2$eid,
+        count = NA,
+        alpha = 1,
+        from = wave2$from,
+        to = wave2$to,
+        total = wave2$total,
+        plotorder = wave2$plotorder)
+sfmsall5$alphaFctr <- cut(sfmsall5$alpha, c(0,.01, .05, .10, .25, .5, .75, 1, 1.1))
+ggplot(data = sfmsall5) +
+  geom_tile(aes(x = ms, y = reorder(eid,-plotorder), fill = alphaFctr)) +
+  #scale_fill_gradient(low = 'grey90', high = '#0868ac', name = "occurrence %") +
+  scale_fill_brewer(palette = "Blues", name = "Occurrence %", labels = c("(0-1]%", "(1-5]%", "(5-10]%","(10-25]%", "(25-50]%", "(50-75]%", "(75-100]%")) +
+  labs(x = "Microstep No.", y = "Edges (ordered by total # occurrences)") +
+  ThemeNoNet +
+  theme(axis.ticks.y = element_blank(), axis.text.y = element_blank(), panel.grid = element_blank())
+@
+
+Much of the code used to create these plots has been packaged an posted online at https://github.com/sctyner/netvizinf and https://github.com/sctyner/geomnet (also on CRAN).
+Additional, non-packaged code can be found at https://github.com/sctyner/SAOM-removing-blindfold.

### 2018-07-17 paper-child.Rnw
ugh forgot to :arrow_up: last night. :anger:

@@
 
-\section{Introduction}
+\section{Introduction \& Background}
 
@@
 
-\section{Networks and their Visualizations} \label{sec:netintro}
+\subsection{Networks and their Visualizations} \label{sec:netintro}
 
@@
 
-\subsection{Introduction to Network Structures}
+\subsubsection{Introduction to Network Structures}
 
@@
 
-\subsection{Visualizing Network Data} \label{sec:netvizintro}
+\subsubsection{Visualizing Network Data} \label{sec:netvizintro}
 
@@
 
-\section{Defining SAOMs} \label{sec:saoms}
+\subsection{Defining SAOMs} \label{sec:saoms}
 
@@
 
-\subsection{Definitions, Terminology, and Notation}
+\subsubsection{Definitions, Terminology, and Notation}
 
@@
 
-The CTMC process $X(T)$ is a series of single tie changes that happen according to some pre-defined rate function (Section \ref{sec:saomratefn}), where one actor at a time is selected and it adds or removes one outgoing tie, or does nothing.
-Once an actor is chosen at random according to the rate function, it is ``given" the chance to change a tie, and it tries to maximize its utility function (Section \ref{sec:saomobjfn}) based on the current and possible future states of the network.
-
-\subsubsection{The Rate Function}\label{sec:saomratefn}
-
-In general, each actor $i$ in the network $x$ may change its ties, $x_{ij}$, to other nodes $j \neq i$ according to the SAOM's  \emph{rate function}, $\rho(x, \mathbf{z}, \boldsymbol{\alpha})$, where $\boldsymbol{\alpha}$ are the parameters in the function $\rho$, and $x$ is the current network state, with covariates of interest $\mathbf{z}$.
+The CTMC process $X(T)$ is a series of single tie changes that happen according to a pre-defined rate function, where one actor at a time is selected and it adds or removes one outgoing tie, or does nothing.
+Once an actor is chosen at random according to the rate function, it is ``given" the chance to change a tie, and it tries to maximize its utility function based on the current and possible future states of the network.
+
+\textbf{The Rate Function} In general, each actor $i$ in the network $x$ may change its ties, $x_{ij}$, to other nodes $j \neq i$ according to the SAOM's  \emph{rate function}, $\rho(x, \mathbf{z}, \boldsymbol{\alpha})$, where $\boldsymbol{\alpha}$ are the parameters in the function $\rho$, and $x$ is the current network state, with covariates of interest $\mathbf{z}$.
 With this general formulation, it is possible that each node will have a different rate of change.
@@
 
-\subsubsection{The Objective Function}\label{sec:saomobjfn}
-
-Because of the conditional independence assumptions given in Equation~\ref{eq:c1}, we can consider the objective function for each node separately, as only one tie from one node is changing at a time in the CTMC.
+\textbf{The Objective Function} Because of the conditional independence assumptions given in Equation~\ref{eq:c1}, we can consider the objective function for each node separately, as only one tie from one node is changing at a time in the CTMC.
 The node $i$, which is the node that is given the opportunity to change at the current time point, is called the \textit{ego} node.
@@
 
-\subsection{Fitting Models to Data}
+\subsubsection{Fitting Models to Data}
 
@@
 
-\subsection{Example Data}
+\section{Example Data and Models}
 
 <<adjmatviz, fig.align='center', fig.cap="A visualization of the adjacency matrices of the three waves of network observations in the ``Teenage Friends and Lifestyle Study'' data.
-The subset we will be using is outlined in red.", results='hide',fig.show='asis', fig.width=8, fig.height=3>>=
+The subset we will be using is outlined in red.", results='hide',fig.show='asis', fig.width=8, fig.height=3, eval = FALSE>>=
 source("code/00e_small_friends.R")
@@
 These are the mechanisms we hope to capture with SAOMs.", out.width='.99\\textwidth', fig.height=3>>=
+source("code/00e_small_friends.R")
+view_rs2$wave <- factor(view_rs2$wave)
+levels(view_rs2$wave) <- c("Wave 1", "Wave 2", "Wave 3")
 actual1 <- merge(data.frame(as.edgelist(as.network(fd2.w1))),
@@
 In all three adjacency matrices, the ego and the alter nodes are ordered by node ID, 1-50, which were determined arbitrarily by the relationships in wave 1.
-In Figure~\ref{fig:adjmatviz}, we show the adjacency matrix visualization of all three observations of this network, with the subset we selected for analysis outlined in red.
+%In Figure~\ref{fig:adjmatviz}, we show the adjacency matrix visualization of all three observations of this network, with the subset we selected for analysis outlined in red.
 This subset contains 16 actors, their ties and their drinking behavior for three timepoints.
-The node-link visualization of the data is provided in Figure~\ref{fig:alldataviz}.
+The node-link visualization of this data is provided in Figure~\ref{fig:alldataviz}.
 For model fitting, we condition on wave 1 and estimate the parameters of the models for the second and third waves.
@@
 
-We use these two example newtoks to demonstrate our model visualization methods for SAOMs in section~\ref{sec:ModelVis}
-
-\section{Model Visualizations} \label{sec:ModelVis}
-
-Every good data analysis includes both numerical and visual summaries of the data, so why restrict model description and diagnostics to numerical summaries?
-The concept of model visualization was developed to complement traditional model diagnostic tools such as $R^2$ values and residual plots \citep{modelvis}.
-When performing analyses, there are three different types of ``model": the model family, the model form, and the fitted model.
-The latter is primarily what one thinks of first when considering a model in a data analysis, where a specified model is fit to data, and parameter estimates are reported.% In the context of SAOMs, the \emph{fitted model} contains the form of the rate and objective functions, the estimated rate parameters, and the estimated objective function parameters.
-The \textit{model form} describes the the model \textit{before} the fitting process, defining which parameters are in the model within the context of the larger model family.% In SAOMs, the \emph{model form} includes description of the rate and objective functions and the variables therein that describe how the network evolves over time.
-Finally, the \textit{model family} is the broadest description of the model.
-This is the type of model that you wish to fit to the data, and is chosen based on the problem, data, and knowledge at hand.
-%For example, we chose to use a SAO \emph{model family} to model our network data over an exponential random graph model (ERGM) because we believe that actor-level variables effect network structure and formation, and we wanted to model the network changes over time.
-
-The model family, the model form, and the fitted model can each be visualized according to the three principles of model visualization: we can view the model in the data space, visualize collections of models, and explore the process of fitting the model, not just the end result.
-Since we have already decided on our model family, SAOMs, we shift our focus to the fitted model and the model form.
-Specifically, we want to learn more about how the \emph{model form} we choose affects the \emph{fitted model} by using our example data sets and our visualization toolbox.
-We begin by introducing the five models that we fit to our example data.
-We then use the five models to guide our visual explorations of SAOMs under the three principles of model visualization.
-%We begin with novel tools and ideas to view a SAOM in the data space of a dynamic network.
-We then explore collections of the same models fit many times to the example data to see how the simulation processes in the \texttt{RSiena} ``black-box" affect the model fits.
-Finally, we look behind the scenes and into the individual steps of the continuous time Markov chain to learn more about how this ``hidden data" mechanism works and how it results in a fitted model.
-
-\subsection{The Models}\label{sec:models}
+\subsection{Models}
 
@@
 
-\subsection{View the model in the data space}\label{sec:minds}
+\section{Model in the data space}\label{sec:minds}
+
+Every good data analysis includes both numerical and visual summaries of the data, so why restrict model description and diagnostics to numerical summaries?
+The concept of model visualization was developed to complement traditional model diagnostic tools such as $R^2$ values and residual plots \citep{modelvis}.
+When performing analyses, there are three different types of ``model": the model family, the model form, and the fitted model.
+The latter is primarily what one thinks of first when considering a model in a data analysis, where a specified model is fit to data, and parameter estimates are reported.% In the context of SAOMs, the \emph{fitted model} contains the form of the rate and objective functions, the estimated rate parameters, and the estimated objective function parameters.
+The \textit{model form} describes the the model \textit{before} the fitting process, defining which parameters are in the model within the context of the larger model family.% In SAOMs, the \emph{model form} includes description of the rate and objective functions and the variables therein that describe how the network evolves over time.
+Finally, the \textit{model family} is the broadest description of the model.
+This is the type of model that you wish to fit to the data, and is chosen based on the problem, data, and knowledge at hand.
+%For example, we chose to use a SAO \emph{model family} to model our network data over an exponential random graph model (ERGM) because we believe that actor-level variables effect network structure and formation, and we wanted to model the network changes over time.
+
+The model family, the model form, and the fitted model can each be visualized according to the three principles of model visualization: we can view the model in the data space, visualize collections of models, and explore the process of fitting the model, not just the end result.
+Since we have already decided on our model family, SAOMs, we shift our focus to the fitted model and the model form.
+Specifically, we want to learn more about how the \emph{model form} we choose affects the \emph{fitted model} by using our example data sets and our visualization toolbox.
+We begin by introducing the five models that we fit to our example data.
+We then use the five models to guide our visual explorations of SAOMs under the three principles of model visualization.
+%We begin with novel tools and ideas to view a SAOM in the data space of a dynamic network.
+We then explore collections of the same models fit many times to the example data to see how the simulation processes in the \texttt{RSiena} ``black-box" affect the model fits.
+Finally, we look behind the scenes and into the individual steps of the continuous time Markov chain to learn more about how this ``hidden data" mechanism works and how it results in a fitted model.
 
@@
 set.seed(4231352)
-M1sims1000 <- netvizinf::saom_simulate(dat = mydata,
-                            struct = M1eff,
-                       parms = as.numeric(M1parms),
-                       N = 1000)
-M2sims1000 <- netvizinf::saom_simulate(dat = mydata,
-                            struct = M2eff,
-                       parms = as.numeric(M2parms),
-                       N = 1000)
-M3sims1000 <- netvizinf::saom_simulate(dat = mydata,
-                            struct = M3eff,
-                       parms = as.numeric(M3parms),
-                       N = 1000)
-M1simsdf <- netvizinf::sims_to_df(M1sims1000)
-M2simsdf <- netvizinf::sims_to_df(M2sims1000)
-M3simsdf <- netvizinf::sims_to_df(M3sims1000)
+#M1sims1000 <- netvizinf::saom_simulate(dat = mydata,
+#                            struct = M1eff,
+#                       parms = as.numeric(M1parms),
+#                       N = 1000)
+#save(M1sims1000, file = "data/M1sims1000.RData")
+load("data/M1sims1000.RData")
+# M2sims1000 <- netvizinf::saom_simulate(dat = mydata,
+#                             struct = M2eff,
+#                        parms = as.numeric(M2parms),
+#                        N = 1000)
+# save(M2sims1000, file = "data/M2sims1000.RData")
+load("data/M1sims1000.RData")
+# M3sims1000 <- netvizinf::saom_simulate(dat = mydata,
+#                             struct = M3eff,
+#                        parms = as.numeric(M3parms),
+#                        N = 1000)
+# save(M3sims1000, file = "data/M3sims1000.RData")
+load("data/M3sims1000.RData")
+
+#M1simsdf <- netvizinf::sims_to_df(M1sims1000)
+# write_csv(M1simsdf, "data/M1simsdf.csv")
+M1simsdf <- read_csv("data/M1simsdf.csv")
+#M2simsdf <- netvizinf::sims_to_df(M2sims1000)
+#write_csv(M2simsdf, "data/M2simsdf.csv")
+M2simsdf <- read_csv("data/M2simsdf.csv")
+# M3simsdf <- netvizinf::sims_to_df(M3sims1000)
+# write_csv(M3simsdf, "data/M3simsdf.csv")
+M3simsdf <- read_csv("data/M3simsdf.csv")
+
 M1avgW2 <- M1simsdf %>% filter(!is.na(to) & wave == 1) %>%
@@
 
-For network data, one way we view an ``average" network is through a sory of summary network drawn using the traditional node-link diagram.
+For network data, we view an ``average" network through a sort of summary network drawn using the traditional node-link diagram.
 In Figure~\ref{fig:summnetM1}, we show an \textit{average network} created with 1,000 simulations of the second wave of the network from Model 1.
@@
 
+<<summary2, fig.height=2.5, fig.cap="Network differences between the average network from M1 and the first and second data waves.
+The green edges represent edges in the average network that are not in the data, while red edges are in the data but not in the average network.
+Black edges are in both the data and the average, and the edgeweight represents the rate at which that edge shows up in simulations.">>=
+am_t1 <- as.matrix(igraph::as_adj(igraph::graph_from_data_frame(arrange(t1, parse_number(from), parse_number(to)))))[-17,-17]
+
+diff1 <- am_t1 - fd2.w1
+diff1[which(diff1==1)] <- 2
+diff1[which(fd2.w1 == 1 & am_t1 == 1)]  <- 1
+diff1 <- diff1 %>% dplyr::as_data_frame() 
+names(diff1) <- paste0("V", 1:16)
+diff1 %>% mutate(from =paste0("V", 1:16)) %>% gather(to, value, -from) %>%
+  mutate(from = as.character(parse_number(from)), to = as.character(parse_number(to)), 
+         var = recode(value, `0` = "Neither", `2` = "InAvgNotW1",  `1`= "InBoth"),
+         color2 = recode(value, `0`="white", `1`= "black", `2` = "green")) %>%
+  filter(value !=0) %>% 
+  full_join(t1, c("from","to")) -> t1_1 
+t1_1 <- arrange(t1_1, from, to)
+p1 <- ggplot(data = t1_1) +
+  geom_net(aes(from_id = from, to_id = to, linewidth= linewidth),
+           directed = T, curvature = .2, labelon=T, size = 5,
+           arrow = arrow(angle = 20, type = "open", length = unit(.2, "cm")), ecolour = t1_1$color2,
+           color = "#252525", hjust = .5, vjust = .5, labelcolour = 'white', fontsize = 4, arrowgap = .02,
+           singletons = T) +
+  labs(title = "Difference between Wave 1 and Average from M1") + ThemeNet +
+  theme(plot.background = element_rect(color = 'black'))
+
+diff2 <- -am_t1 + fd2.w2 
+diff2[which(diff2==1)] <- 2
+diff2[which(fd2.w2 == 1 & am_t1 == 1)]  <- 1
+diff2 <- dplyr::as_data_frame(diff2)
+names(diff2) <- paste0("V", 1:16)
+diff2 %>% mutate(from =paste0("V", 1:16)) %>% gather(to, value, -from) %>% 
+  mutate(from = as.character(parse_number(from)), to = as.character(parse_number(to)), 
+         var = recode(value, `0` = "Neither", `-1` = "InAvgNotW2",  `1`= "InBoth", `2` = "InW2notAvg"),
+         color2 = recode(value, `0`="white", `1`= "black", `2` = "red", `-1` = "green")) %>% 
+  filter(value !=0) %>% 
+  full_join(t1, c("from","to")) -> t1_2 
+t1_2$linewidth[is.na(t1_2$linewidth) & !is.na(t1_2$to)] <- 1
+t1_2 <- arrange(t1_2, from, to)
+t1_2 <- t1_2 %>% filter(!(from == 16 & is.na(to)))
+p2 <- ggplot(data = t1_2) +
+  geom_net(aes(from_id = from, to_id = to, linewidth= linewidth),
+           directed = T, curvature = .2, labelon=T, size = 5,
+           arrow = arrow(angle = 20, type = "open", length = unit(.2, "cm")), ecolour = t1_2$color2,
+           color = "#252525", hjust = .5, vjust = .5, labelcolour = 'white', fontsize = 4, arrowgap = .02,
+           singletons = T) +
+  labs(title = "Difference between Wave 2 and Average from M1") + ThemeNet +
+  theme(plot.background = element_rect(color = 'black'))
+grid.arrange(p1, p2, nrow = 1)
+@
+
+In Figure~\ref{fig:summary2}, we show \textit{difference networks} between the average network from M1 and the first and second waves of data.
+The lack of model fit is very obvious here, with the green edges representing ties that only show up in the average network, and the red edges representing ties in the data that the model misses completely.
+Again, the edgeweights demonstrate the frequency of the ties in the simulations from M1.
+On the left in Figure~\ref{fig:summary2}, we see the difference between the first wave of data and the average network.
+We only see green and black edges, meaning that the simulated networks consist primarily of edges in wave 1, plus some new edges that aren't in the first wave of data that are reciprocating the ties that are in the data.
+On the right in Figure~\ref{fig:summary2}, we see the difference between the second wave of data and the average network.
+Remember, the model is simulating the second wave of data, which is how the average network is created.
+There are six red edges, which are edges in wave 2 that do not appear at all in the average network.
+In addition, there are several green edges that are in the average network, but not in wave 2.
+Finally, there are a few black edges that appear in both.
+For all edges but the red ones, the linewidth represents the frequency of the edge in the simulations.
+
 <<gof2, fig.cap="An example of viewing the distribution of a statistic of interest from a SAOM.
@@
 
-\begin{figure}
-\centering
-\includegraphics[width=.75\linewidth]{img/Model1LUSmallFriends.png}
-\caption{\label{fig:lineup} A small lineup of node-link diagrams showing the second wave of our small friendship network among five networks simulated from model M1.}
-\end{figure}
-
-Another potential goodness-of-fit visualization that places the fitted model in the node and edge data space is a lineup like those proposed in \citet{buja:2009}.
-A \textit{lineup} ``asks the witness to identify the plot of the real data from among a set of decoys, the null plots, under the veil of ignorance" \citep[p.
-4369]{buja:2009}.
-It can be thought of like a police lineup, where the ``suspect" is in a lineup among several innocent similar looking people, and if the witness picks the suspect out of the lineup, that is evidence the suspect is guilty.
-
-In data and model visualization, the ``suspect" is a plot of the true data, while the ``filler" is composed of several plots of ``mock data", simulated from a hypothesized model.
-If the true data stands out among the simulated data, that is taken as evidence agains the hypothesised model.
-If, however, the true data is difficult to identify among the simulated data, that is taken as evidence in favor of the hypothesized model.
-% An application of the lineup protocal can be found in \citet{hofmann2012graphical}, where the authors examine, for instance, the differences between polar and cartesian coordinates for plotting categorical data, and density plots and box plots for determining distributional differences.
-They pose questions to experiment participants such as, ``Which plot is most different from the others?" or, ``In which plot is the blue group furthest to the right?"
-
-The data visualizations examined in previous applications such as  \citet{hofmann2012graphical} are less complex than a node-link network diagram.
-Typically, the lineup viewers have to pick out the most different plot, which is hard for network data.
-As an example, consider the network lineup shown in Figure~\ref{fig:lineup}.
-In this small network lineup, the second wave of the small friendship network is shown among five simulated networks from model M1 using parameter values estimated from the data.
-It seems possible to argue for any one of the six plots in Figure~\ref{fig:lineup} as most different.
-So, we suggest guiding participants to look at the overall structure of the graphs to determine which has the most or least connected or complex structure.% The least complex plot, number six, has no triangles (3-cycles), while the most complex plot, number three, has three 3-cycles.
-The most complex plot, number three, is in fact the data.
-We have found that when shown lineups like that of Figure~\ref{fig:lineup}, online survey participants %where all the node-link diagrams shown are based on simulated data, %one observation from an ``alternative" model and the rest from a ``null" model, %pick up on the triangular shape of the 3-cycles most often.
-In a future paper, we will use the lineup protocol to help better understand SAOMs by allowing us to view the model in the data space.
-%Lineups place observations simulated from the model side-by-side with the data and let the viewer examining the differences.
-This can also help us determine the significant structural effects, if any, of the parameters in the model on observations simulated from the model.
-The triangular shape mentioned above becomes more prevalent when a transitive triplet parameter ($\beta_3$ in Section~\ref{sec:models}) is included in a model, indicating that it may have a significant effect on network structure.
-%, but it does not always cause triplets to form in the simulated data.
-This requires more investigation with larger data sets and more complex models, but is promising for the development of additional goodness-of-fit measures for network models.
-
-% \st{
 % \begin{itemize}
@@
 
-\subsection{Visualizing collections of models}
+\section{Collections of models}
 
@@
 
-\subsubsection{Exploring the space of all possible models}\label{sec:allpossible}
+\subsection{Exploring the space of all possible models}\label{sec:allpossible}
 
@@
 
-\subsubsection{Varying model settings}
+\subsection{Varying model settings}
 
@@
 
-\subsubsection{Fitting the same model to different data}\label{sec:samemoddiffdat}
+\subsection{Fitting the same model to different data}\label{sec:samemoddiffdat}
 
@@
 
-\subsubsection{Fitting the same model to the same data}\label{sec:samemodsamedat}
+\subsection{Fitting the same model to the same data}\label{sec:samemodsamedat}
 
@@
 
-\subsection{Explore algorithms, not just end result}\label{sec:algorithms}
+\section{Explore algorithms, not just end result}\label{sec:algorithms}
 

### 2018-07-17 paper.Rnw
ugh forgot to :arrow_up: last night. :anger:

@@
 \begin{abstract}
-Stochastic actor-oriented models, first introduced by Snijders (1996), are a type of statistical network model for dynamic social networks.
+Stochastic actor-oriented models (SAOMs), first introduced by Snijders (1996), are a type of statistical network model for dynamic social networks.
 Unlike other network models, SAOMs are not very well understood.
-We use model visualization techniques introduced in Wickham et al (2015) in order to make them a little less murky.
-The SAOMs are a prime example of a set of models that can benefit greatly from application of model visualization.
-With the help of static and dynamic visualizations, we bring the hidden model fitting processes into the foreground, eventually leading to a better understanding and higher accessibility of stochastic actor-oriented models for social network analysts.
+We use model visualization techniques introduced in Wickham et al (2015) in order to gain a better understanding of the models and their behavior.
+Viewing the model in the data space places these models into the field of network drawing, while exploring collections of models helps with parameter interpretation.
+Finally, delving into the fitting algorithms provides more insight into the underlying mechanisms.
+With the help of static and dynamic visualizations, we bring the hidden SAOM fitting processes into the foreground, eventually leading to a better understanding and higher accessibility of SAOMs for social network analysts.
 \end{abstract}

### 2018-07-25 paper-child.Rnw
getting there...

@@
 
+<<getdata , echo=FALSE>>=
+set.seed(823746)
+load("data/ansnullpaper.rda")
+ansnullchains <- get_chain_info(ansnull)
+load("data/M1sims1000.RData")
+load("data/M2sims1000.RData")
+load("data/M3sims1000.RData")
+@
+
 \section{Introduction \& Background}
@@
 Once an actor is chosen at random according to the rate function, it is ``given" the chance to change a tie, and it tries to maximize its utility function based on the current and possible future states of the network.
-
-\textbf{The Rate Function} In general, each actor $i$ in the network $x$ may change its ties, $x_{ij}$, to other nodes $j \neq i$ according to the SAOM's  \emph{rate function}, $\rho(x, \mathbf{z}, \boldsymbol{\alpha})$, where $\boldsymbol{\alpha}$ are the parameters in the function $\rho$, and $x$ is the current network state, with covariates of interest $\mathbf{z}$.
+\\
+
+\noindent \textbf{The Rate Function} In general, each actor $i$ in the network $x$ may change its ties, $x_{ij}$, to other nodes $j \neq i$ according to the SAOM's  \emph{rate function}, $\rho(x, \mathbf{z}, \boldsymbol{\alpha})$, where $\boldsymbol{\alpha}$ are the parameters in the function $\rho$, and $x$ is the current network state, with covariates of interest $\mathbf{z}$.
 With this general formulation, it is possible that each node will have a different rate of change.
@@
 \end{equation}
-which is the total number of changes from observation at time $t_m$ to the observation at time $t_{m+1}$.
-
-\textbf{The Objective Function} Because of the conditional independence assumptions given in Equation~\ref{eq:c1}, we can consider the objective function for each node separately, as only one tie from one node is changing at a time in the CTMC.
+which is the total number of changes from observation at time $t_m$ to the observation at time $t_{m+1}$. \\
+
+\noindent \textbf{The Objective Function} Because of the conditional independence assumptions given in Equation~\ref{eq:c1}, we can consider the objective function for each node separately, as only one tie from one node is changing at a time in the CTMC.
 The node $i$, which is the node that is given the opportunity to change at the current time point, is called the \textit{ego} node.
@@
 
+% removed the following plot
 <<adjmatviz, fig.align='center', fig.cap="A visualization of the adjacency matrices of the three waves of network observations in the ``Teenage Friends and Lifestyle Study'' data.
@@
 The nodes change color as their drinking behavior changes, and the network becomes less densely connected over time.
-These are the mechanisms we hope to capture with SAOMs.", out.width='.99\\textwidth', fig.height=3>>=
+These are the mechanisms we hope to capture with SAOMs.", out.width='.99\\textwidth', fig.height=3, results='hide'>>=
 source("code/00e_small_friends.R")
@@
 
-To guide our visual exploration of SAOMs, we use two example data sources.
+\noindent \textbf{Example Data:} To guide our visual exploration of SAOMs, we use two example data sources.
 The first is a subset of the 50 actor dataset from the ``Teenage Friends and Lifestyle Study" that is provided on the \texttt{RSiena} webpage.
@@
 
-\subsection{Models}
-
-We first consider the 16 actor subset of the teenage friends and lifestyle data show in Figure~\ref{fig:alldataviz}.
+\noindent \textbf{Models:} We first consider the 16 actor subset of the teenage friends and lifestyle data show in Figure~\ref{fig:alldataviz}.
 To this data, we fit three different SAOMs.
@@
 
-In Section~\ref{sec:samemoddiffdat}, we also fit models M1, M2, and M3 to the senate collaboration data for comparison.
-To fit M2 to the senate data, we use the number of bills authored by each senator as the node covariate for the jumping transitive triplet variable.
+In this paper, we use three different model visualization techniques on SAOMs.
+First, Section~\ref{sec:minds} places the models in the data space.
+Then, Section~\ref{sec:collectmodels} explores visualizations of various collections of models.
+Finally, in Section~\ref{sec:algorithms} we use model visualization to explore the underlying algorithms in the SAOM fitting process.
  %In terms of the senate data, then, the value of $\beta_3$ should dictate how willing a senator is to coauthor a bill on which the author of the bill has a different level of authorship, assuming there is an intermediary between the two who has the same authorship level as the first.
@@
   geom_net(directed = T, labelon=F, arrowsize = .3, singletons= F, fiteach = T, arrowgap = .01, layout.alg = 'fruchtermanreingold',
-           aes(from_id = source, to_id = target, color = party, linewidth = gsw, shape = sex,size = n_au)) +
-  scale_size_continuous(name="# bills\nauthored") +
+           aes(from_id = source, to_id = target, color = party, linewidth = gsw, shape = sex,size = n_au, group = party)) +
+  scale_size_continuous(name="# bills\nauthored" , guide = guide_legend(nrow= 2)) +
   #theme_net() +
   ThemeNet +
-  scale_shape_manual(values = c(17,16)) +
-  scale_color_manual(values = c("royalblue", "forestgreen","firebrick")) +
-  theme(legend.position = 'bottom') +
-  facet_wrap(~Clinton, nrow = 1, labeller = 'label_both')
+  scale_shape_manual(values = c(17,16), guide = guide_legend(nrow= 2)) +
+  scale_color_manual(values = c("royalblue", "forestgreen","firebrick"), guide = guide_legend(nrow=2)) +
+  facet_wrap(~Clinton, nrow = 1, labeller = 'label_both')  +
+  theme(legend.position = "bottom")
 @
@@
 #save(M1sims1000, file = "data/M1sims1000.RData")
-load("data/M1sims1000.RData")
+
 # M2sims1000 <- netvizinf::saom_simulate(dat = mydata,
@@
 # save(M2sims1000, file = "data/M2sims1000.RData")
-load("data/M1sims1000.RData")
+
 # M3sims1000 <- netvizinf::saom_simulate(dat = mydata,
@@
 # save(M3sims1000, file = "data/M3sims1000.RData")
-load("data/M3sims1000.RData")
 
@@
            singletons = T) +
-  labs(title = "Difference between Wave 1 and Average from M1") + ThemeNet +
+  labs(title = "Difference between\nWave 1 and Average from M1") + ThemeNet +
   theme(plot.background = element_rect(color = 'black'))
@@
            singletons = T) +
-  labs(title = "Difference between Wave 2 and Average from M1") + ThemeNet +
+  labs(title = "Difference between\nWave 2 and Average from M1") + ThemeNet +
   theme(plot.background = element_rect(color = 'black'))
@@
 
+% No longer using this plot
 <<gof2, fig.cap="An example of viewing the distribution of a statistic of interest from a SAOM.
@@
 The distance computed from the data is shown as a vertical line.", fig.height=3, out.width='.6\\textwidth', eval = FALSE>>=
-load("data/ansnullpaper.rda")
+
 library(lattice)
@@
 
-\section{Collections of models}
+\section{Collections of models}\label{sec:collectmodels}
 
@@
 
-<<compareModels, fig.cap="Density plots of objective function parameter estimates from repeatedly fitting models M1, M2, and M3 to the example data.
-Outdegree and reciprocity have the same inverse relationship for both data sets.
-For the friendship data, the inclusion of $\\beta_3$ has strong effect on the estimates of the other two parameters, but not for the senate data.", fig.height = 3>>=
-simu2$data <- "friends"
-senateEsts$data <- "senate"
-allDataEsts <- rbind(senateEsts, simu2)
-allDataEsts$parameter <- as.factor(allDataEsts$parameter)
-allDataEsts$parameter <- ordered(allDataEsts$parameter, levels = c(paste0("alpha", 1:3),paste0("beta", 1:4)))
-ggplot(allDataEsts %>% filter(str_detect(parameter, "beta"))) +
-  geom_density(aes(x = estimate, fill = Model), alpha = .5) +
-  facet_grid(data~parameter, scales = "free", labeller = "label_both") +
-  ThemeNoNet +
-  theme(legend.position = 'bottom')
-@
-
-Looking at Figure~\ref{fig:compareModels} and Table~\ref{tab:meanests}, we see a few patterns in the estimates from both models.
+Looking at Table~\ref{tab:meanests} and Figure~\ref{fig:compareModels} in the Appendix, we see a few patterns in the estimates from both models.
 First, the table and the density plots demonstrate the same relationship between the outdegree parameter, $\beta_1$, and the reciprocity parameter, $\beta_2$.
@@
 
+% taken out.
 <<getdatam, eval = FALSE>>=
@@
 )
-null_mod_sv <- as.numeric(nulls[,-c(1,6,7)] %>% summarise_each(funs(mean)))
-alt_mod_sv <- as.numeric(alt[,-c(1, 7,8)] %>% summarise_each(funs(mean)))
-alt_mod2_sv <- as.numeric(alt2nd[,-c(1,7,8)] %>% summarise_each(funs(mean)))
-@
-
+null_mod_sv <- as.numeric(nulls[,-c(1,6,7)] %>% summarise_all(funs(mean)))
+alt_mod_sv <- as.numeric(alt[,-c(1, 7,8)] %>% summarise_all(funs(mean)))
+alt_mod2_sv <- as.numeric(alt2nd[,-c(1,7,8)] %>% summarise_all(funs(mean)))
+@
+
+% taken out
 <<corplotsSenate, eval = FALSE>>=
@@
 
+<<makinggif, fig.cap='A selection of images in the microstep animation.
+The selected edges and nodes are emphasized by changing the size and the color, then when edges are removed, they fade out, shrinking in size, while the nodes change color and shrink to blend in with the other nodes.', message = FALSE, warning = FALSE, fig.height=2.5>>=
+# Need old version of dplyr to get this to work:
+# if (!require("dplyr050",character.only = TRUE)) {
+#       devtools::install_github("sctyner/dplyr050")
+# }
+# library(dplyr050)
+friend.data.w1 <- as.matrix(read.table("data/s50_data/s50-network1.dat"))
+friend.data.w2 <- as.matrix(read.table("data/s50_data/s50-network2.dat"))
+friend.data.w3 <- as.matrix(read.table("data/s50_data/s50-network3.dat"))
+drink <- as.matrix(read.table("data/s50_data/s50-alcohol.dat"))
+fd2.w1 <- friend.data.w1[20:35,20:35]
+fd2.w2 <- friend.data.w2[20:35,20:35]
+fd2.w3 <- friend.data.w3[20:35,20:35]
+friendshipData <- array(c(fd2.w1, fd2.w2,fd2.w3), dim = c(16, 16, 3))
+
+# running Siena
+#friendship <- sienaDependent(friendshipData)
+#alcohol <- varCovar(drink[20:35,])
+#mydata <- sienaDataCreate(friendship, alcohol)
+#myeffnull <- getEffects(mydata)
+#myalgorithm <- sienaAlgorithmCreate(projname = 's16_3')
+
+ansnullchains %>%
+  dplyr::filter(period == 1) %>%  #only look at chains from wave 1 to wave 2
+  #dplyr::group_by(rep) %>%
+  dplyr::select(rep, from = ego, to = alter) %>%
+  dplyr::mutate(val = as.numeric(!from == to),
+         from = paste0("V", parse_number(from)+1), # make the chains
+         to = paste0("V", parse_number(to)+1)) -> ansnullchainsw1w2
+colnames(fd2.w1) <- paste0("V", 1:16)
+rownames(fd2.w1) <- paste0("V", 1:16)
+wave1friends <- fortify(as.adjmat(fd2.w1))
+ms1 <- netvizinf::listMicrosteps(dat = wave1friends,
+                      microsteps = dplyr::filter(ansnullchainsw1w2, rep == 1))
+microsteps <- ms1[1:20]
+
+microsteps_df <- getMicrostepsDF(microsteps)
+pte <- pretween_edges(microsteps = microsteps)
+ptv <- pretween_vertices(pte = pte, layoutparams = list(n = 16))
+# step 1 is remove, step 9 is add.
+# get tweened data for all microsteps
+pall <- tween_microsteps(ptv, pte, microsteps_df)
+
+static_pall <- dplyr::filter(pall, .frame %in% c(15,20,22,24,26:28, 71:73, 75, 79, 81, 84))
+
+static_pall$type <- NA
+static_pall$type[which(static_pall$.frame %in% c(15,20,22,24,26:28))] <- "Remove Edge"
+static_pall$type[which(static_pall$.frame %in% c(71:73, 75, 79, 81, 84))] <- "Add Edge"
+
+set.seed(34569234)
+froms <- static_pall %>% dplyr::select(id = from, microstep, x = from.x,
+        y = from.y, ms, .frame, size, color) %>% unique
+tos <- static_pall %>% dplyr::select(id = to, microstep, x = to.x, y = to.y,
+        ms, .frame) %>% unique
+tos <- dplyr::left_join(tos, froms)
+nodedata <- rbind(froms, tos) %>% unique
+nodedata <- nodedata %>% tidyr::replace_na(replace = list(size = 1,
+        color = "#333333"))
+to_ani <- ggplot() +
+  geom_curve(data = static_pall, aes(x = from.x,
+        y = from.y, xend = to.x, yend = to.y, frame = .frame,
+        colour = ecolor, size = esize), curvature = 0.1, arrow = arrow(angle = 20,
+        length = unit(0.1, "inches"))) +
+  geom_point(data = nodedata,
+        aes(x = x, y = y, colour = color, size = size, frame = .frame)) +
+  scale_colour_identity() + scale_size_identity() + theme_void() +
+        coord_fixed()
+to_ani +
+  facet_wrap(~.frame, nrow=2, labeller = 'label_both') +
+  ThemeNet +
+  theme(panel.background = element_rect(color = 'black'))
+@
+
 The first vizualization we present here is an animation of the simulated microsteps, when fitting model M1, that form the transition steps of the CTMC from wave 1 to wave 2 of the small friendship network example shown in Figure~\ref{fig:alldataviz}.
-.
 Movies similar to this animation were used to visualize the changes of dynamic networks in \citet{dynnetviz}.
@@
 If there are no changes at a particular microstep, no changes are seen.
-Some frames of the animations are shown in Figure~\ref{fig:makinggif} in the Appendix, and the full movie, created with \texttt{tweenr} and \texttt{gganimate} can be viewed at \url{https://vimeo.com/240089108} \citep{tweenr, gganimate}.
+Some frames of the animations are shown in Figure~\ref{fig:makinggif}, and the full movie, created with \texttt{tweenr} and \texttt{gganimate} can be viewed at \url{https://vimeo.com/240089108} \citep{tweenr, gganimate}.
+
+The top row of Figure~\ref{fig:makinggif} shows an edge being removed, and the bottom row shows one being added.
+In both cases, the ego node that is acting changes color from black to red, and they also increase greatly in size.
+In the case of an edge being removed, in the top row, the edge that currently exists is emphasized with the same color and size change that the node gets, and as the animation proceeds the edge shrinks to nothing, as the ego node shrinks and changes color back to its original black.
+If an edge is being added, as in the bottom row of the figure, the ego node's appearance changes in the same way as when the edge is being removed, while the edge appears colored red from nothing, and grows to a large size, then changes color and size to match the rest of the edges, while the node shrinks and changes color to match the other nodes.
 
@@
 
-We next use animation to view the changing structure of the adjacency matrix the microsteps.
-The adjacency matrices for the three waves of friendship data as shown in Figure~\ref{fig:alldataviz} are ordered by node id.
-There are 16 nodes in the data, numbered 1-16, and that order is used on the $x$ and $y$ axes for the matrix visualization.
+We also use animation to view the changing structure of the adjacency matrix the microsteps.
+Since there are 16 nodes in the data, numbered 1-16, we first use that order on the $x$ and $y$ axes for the matrix visualization.
 Viewing the adjacency matrices with this arbitrary ordering does not provide much information to the viewer about the underlying structure of the network.
@@
 <<adjmatorder, message = FALSE, warning = FALSE, fig.cap="On the left, the starting friendship network represented in adjacency matrix form, ordered by vertex id.
-On the right, the same adjacency matrix is presented after ordering the vertices by one repetition of the microstep simulation process from wave one to wave two.", fig.width=8, fig.height=3>>=
+On the right, the same adjacency matrix is presented after ordering the vertices by one repetition of the microstep simulation process from wave one to wave two.", fig.width=8, fig.height=3, fig.align='center'>>=
 # first present the original wave 1 adj mat, then present the adj mat after being ordered by the 1st pc of the big adjmat A.
-# reinstall new dplyr
 fd.w12 <- data.frame(fd2.w1)
+fd.w12 <- data.frame(fd2.w1)
+
 numordersf <- fd.w12 %>%
@@
   summarise(count = n()) -> sfmsall_summ
-sfmsall_summ$from <- factor(sfmsall_summ$from, levels = paste0('V', 1:16))
-sfmsall_summ$to <- factor(sfmsall_summ$to, levels = paste0('V', 1:16))
+sfmsall_summ <- sfmsall_summ %>% ungroup %>% mutate(from = parse_number(from), to = parse_number(to))
+sfmsall_summ$from <- as.factor(sfmsall_summ$from)
+levels(sfmsall_summ$from) <- paste0('V', 1:16)
+sfmsall_summ$to <- as.factor(sfmsall_summ$to)
+levels(sfmsall_summ$to) <- paste0('V', 1:16)
+#sfmsall_summ <- sfmsall_summ %>% ungroup() %>% mutate(from = as.factor(from), to = as.factor(to))
 sfmsall_summ %>%
@@
 pcaordersf$from <- as.factor(pcaordersf$from)
-levels(pcaordersf$from) <- 1:16
+levels(pcaordersf$from) <- gsub("V", "", levels(pcaordersf$from))
 pcaordersf$from <- factor(pcaordersf$from , levels = levels(pcaordersf$from)[match( order1,levels(pcaordersf$from) )])
 pcaordersf$to <- as.factor(pcaordersf$to)
-levels(pcaordersf$to) <- 1:16
+levels(pcaordersf$to) <-gsub("V", "", levels(pcaordersf$to))
 pcaordersf$to <- factor(pcaordersf$to , levels = levels(pcaordersf$to)[match( order1,levels(pcaordersf$to) )])
@@
 p2 <- ggplot(data = pcaordersf, aes(x = to, y = from)) +
+  geom_tile(color = 'black', aes(fill = as.factor(value))) +
+  scale_fill_manual(values = c('white', 'black'), name = "Tie is...", labels = c("not present","present")) +
+  scale_x_discrete(position = "top") +
+  #scale_y_discrete(name = "from",labels = rev(str_replace(order1, "V", ""))) +
+  ThemeNoNet +
+  coord_fixed()
+p2a <- ggplot(data = pcaordersf, aes(x = to, y = from)) +
   geom_tile(color = 'black', aes(fill = as.factor(value))) +
@@
   #scale_y_discrete(name = "from",labels = rev(str_replace(order1, "V", ""))) +
-  ThemeNoNet +
-  theme(legend.position = 'none') +
+  ThemeNoNet + theme(legend.position = "none") + 
   coord_fixed()
-
-grid.arrange(p1,p2, nrow = 1)
+legendp2 <- cowplot::get_legend(p2)
+ggdraw() + draw_plot(p1, width = .4) +  draw_plot(p2a, width = .4, x=.4) + 
+  draw_grob(legendp2,x=.8, y=.25, width = .2, height = .4)
+# cowplot::plot_grid(p1,p2a,legendp2, ncol=3)
 @
@@
 To reorder the vertices for the matrix visualization, we first constructed a cumulative adjaceny matrix, $\mathbf{A}^{cum}$, for the series of microsteps simulating the network from $x(t_m)$ to $x(t_{m+1})$.
-A single entry in the cumulative adjacency matrix, $\mathbf{A}^{cum}_{ij}$, is the total number of times the edge $i \righarrow j$ appears in the network from the initial observation, $x(t_m) \equiv X(0)$ to the final result of the last microstep, $X(R)$, where $R$ is the total number of microsteps taken:
+A single entry in the cumulative adjacency matrix, $\mathbf{A}^{cum}_{ij}$, is the total number of times the edge $i \rightarrow j$ appears in the network from the initial observation, $x(t_m) \equiv X(0)$ to the final result of the last microstep, $X(R)$, where $R$ is the total number of microsteps taken:
 $$ \mathbf{A}^{cum}_{ij} = \sum_{r = 0}^R X_{ij}(r). $$
@@
 We also use the PCA seriated layout to fix the layout in the animation of one of the microstep process simulations.
-This animation, some frames of which are shown in Figure~\ref{fig:adjmatani} of the Appendix is very simple: a square appears or disappears in the animation as that edge appears or disappears in the microstep process.
+This animation, some frames of which are shown in Figure~\ref{fig:adjmatani}, is very simple: a square appears or disappears in the animation as that edge appears or disappears in the microstep process.
 Through this animation, which can be viewed at \url{https://vimeo.com/240092677}, we can see edges appearing, and then later on disappearing.
@@
 
+<<adjmatani, fig.cap='A selection of frames from the adjacency matrix visualization animation for one series of microsteps.
+(Ego and alter node labels are removed to declutter the graph.) At the beginning of the animation, shown in the top row, there are two clearly connected components: one in the bottom left corner, and one in the top right corner.
+By the end, the component in the bottom left has spread out and become sparsely connected, while the top right component has shrunk, but remains closely connected.', warning = FALSE, message = FALSE, results='hide', fig.height=2.5>>=
+adjmatall <- matrix(0, nrow = 16, ncol = 16)
+rownames(adjmatall) <- paste0("V", 1:16)
+colnames(adjmatall) <- paste0("V", 1:16)
+dat2 <- sfmsall %>% filter(rep == 1)
+adjmatall2 <- NULL
+for(i in 0:55){
+  MS <- dat2[dat2$ms == i,c("from", "to")]
+  myadjmat <- adjmatall
+  for(j in 1:nrow(MS)){
+    myadjmat[MS[j,"from"][[1]],MS[j,"to"][[1]]] <- 1
+  }
+  adjmatall2 <- rbind(adjmatall2, myadjmat)
+}
+
+adjmatall2 <- cbind(adjmatall2, rep(0:55, each = 16))
+rownames(adjmatall2) <- NULL
+adjmatall2 <- as.data.frame(adjmatall2)
+adjmatall2$from <- rep(paste0("V", 1:16), 56)
+names(adjmatall2)[17] <- "ms"
+adjmatall2 %>% gather(to, val, V1:V16) -> adjmatall3
+
+adjmatall3$to <- as.factor(adjmatall3$to)
+levels(adjmatall3$to) <- gsub("V", "", levels(adjmatall3$to))
+adjmatall3$from <- as.factor(adjmatall3$from)
+levels(adjmatall3$from) <- gsub("V", "", levels(adjmatall3$from))
+adjmatall3$to <- factor(adjmatall3$to, levels = order1)
+adjmatall3$from <- factor(adjmatall3$from, levels = order1)
+
+adjmatall3$frame <- adjmatall3$ms
+#plots <- list()
+#for(i in 0:55){
+#plots[[i + 1]] <-
+
+dat3 <- adjmatall3 %>% filter(ms %in% c(0,2,4,6,8,10,12,43,45,47,49,51,53,55))
+
+dat4 <- dat3 %>% select(-ms) %>% spread(frame, val)
+dat5 <- dat4
+#p <- Sys.time()
+for(i in 4:16){
+  for(j in 1:nrow(dat5)){
+    if(dat4[j,i] != dat4[j,i-1]){
+      if(dat4[j,i] > dat4[j,i-1]){
+        dat5[j,i] <- 2
+      } else { dat5[j,i] <- 3}
+    }
+  }
+}
+#Sys.time() - p
+
+dat6 <- dat5 %>% gather(frame, val, 3:16) %>% mutate(frame = parse_number(frame))
+
+ggplot() +
+  geom_tile(data = dat6, aes(x = to, y = from, fill = as.factor(val), color=as.factor(val))) +
+  scale_fill_manual(values = c("white", "black", "red", "white"), 
+                    labels = c("not present", "present", "added", "removed"), 
+                    name = "Tie is...") +
+  scale_color_manual(values = c("#969696", "#969696", "red", "red"), 
+                    labels = c("not present", "present", "added", "removed"), 
+                    name = "Tie is...") + 
+  geom_tile(data = dat6 %>% filter(val == 3), aes(x = to, y = from), fill = NA, color = 'red', size = .25) +
+  scale_x_discrete(position = "bottom") +
+  labs(x = "alter node (j)", y = "ego node (i)") +
+  ThemeNoNet +
+  theme(axis.text.x.bottom = element_blank(),
+        axis.text.y = element_blank(), 
+        strip.text = element_text(size = rel(.75)),
+        strip.text.x = element_text(size = rel(.75)),
+        strip.text.y = element_text(size = rel(.75)),
+        legend.text = element_text(size = rel(.75)), 
+        legend.title = element_text(size = rel(.75))) +
+  coord_fixed() +
+  facet_wrap(~frame, labeller = 'label_both', nrow = 2)
+#}
+#plots
+@
+
+% removed
 <<histogramalters, fig.width = 7, fig.height=9, fig.cap="Each panel shows the theoretical (as lines) and empirical (as points) probabilities of the chosen ego node changing its tie to each of the other nodes.
 The color of the line indicates whether the tie from the ego to the alter node is being added, removed, or if there is no change to the network in this step.", eval = FALSE>>=
-set.seed(823746)
-load("data/ansnullpaper.rda")
-ansnullchains <- get_chain_info(ansnull)
 ansnullchains %>% group_by(period, rep) %>%
@@
 #              ncol = 2, padding = unit(0, 'line'))
+
 @
@@
 There were many ties with empirical probability zero.', fig.height=2.5>>=
-library(dplyr)
-set.seed(823746)
-load("data/ansnullpaper.rda")
-ansnullchains <- get_chain_info(ansnull)
+
 ansnullchains %>%
@@
   geom_tile(color = 'black') +
-  scale_fill_gradient(low = 'white', high = 'blue', limits = c(0,1)) +
+  scale_fill_gradient(low = 'white', high = 'black', limits = c(0,1)) +
   labs(y = "ego", x = "alter")+
@@
 
+% removed 
 <<allsteps, fig.cap="Two simulations (out of 1,000) of the microstep process from wave 1 to wave 2.
@@
 
-<<sigtesting, fig.show = 'hold', out.width='\\textwidth', fig.height=2, fig.cap="From Section~\ref{sec:allpossible}.
+<<sigtesting, fig.show = 'hold', out.width='\\textwidth', fig.height=2, fig.cap="From Section~\\ref{sec:allpossible}.
 Significant effects for the two data sets, at a significance level of 0.10 or lower as calculated by the Wald-type test available in the SIENA software.", error=FALSE>>=
@@
 
+<<compareModels, fig.cap="Density plots of objective function parameter estimates from repeatedly fitting models M1, M2, and M3 to the example data.
+Outdegree and reciprocity have the same inverse relationship for both data sets.
+For the friendship data, the inclusion of $\\beta_3$ has strong effect on the estimates of the other two parameters, but not for the senate data.", fig.height = 3>>=
+simu2$data <- "friends"
+senateEsts$data <- "senate"
+allDataEsts <- rbind(senateEsts, simu2)
+allDataEsts$parameter <- as.factor(allDataEsts$parameter)
+allDataEsts$parameter <- ordered(allDataEsts$parameter, levels = c(paste0("alpha", 1:3),paste0("beta", 1:4)))
+ggplot(allDataEsts %>% filter(str_detect(parameter, "beta"))) +
+  geom_density(aes(x = estimate, fill = Model), alpha = .5) +
+  facet_grid(data~parameter, scales = "free", labeller = "label_both") +
+  ThemeNoNet +
+  theme(legend.position = 'bottom')
+@
+
+The density plots in Figure~\ref{fig:compareModels} show the distribution of the parameter estimates for the objective function parameters in M1, M2, and M3 the two example data sets we used.
+The most notable difference is in the values of $\beta_1, \beta_2$.
+In the senate data, the inclusion of $\beta_3$ has no effect on the estimates of $\beta_1, \beta_2$, while $\beta_3$ does seem to have an effect with the friendship data.
+
 <<averagenet, results='hide', out.width='\\textwidth',fig.cap='The node-link diagrams from the three "average" networks that we calculated are in the top row, and the true wave 1 and wave 2 data are shown in the bottom row above.
@@
 
-<<makinggif, fig.cap='A selection of images in the microstep animation.
-The selected edges and nodes are emphasized by changing the size and the color, then when edges are removed, they fade out, shrinking in size, while the nodes change color and shrink to blend in with the other nodes.', message = FALSE, warning = FALSE, fig.height=2.5>>=
-# Need old version of dplyr to get this to work:
-# if (!require("dplyr050",character.only = TRUE)) {
-#       devtools::install_github("sctyner/dplyr050")
-# }
-# library(dplyr050)
-friend.data.w1 <- as.matrix(read.table("data/s50_data/s50-network1.dat"))
-friend.data.w2 <- as.matrix(read.table("data/s50_data/s50-network2.dat"))
-friend.data.w3 <- as.matrix(read.table("data/s50_data/s50-network3.dat"))
-drink <- as.matrix(read.table("data/s50_data/s50-alcohol.dat"))
-fd2.w1 <- friend.data.w1[20:35,20:35]
-fd2.w2 <- friend.data.w2[20:35,20:35]
-fd2.w3 <- friend.data.w3[20:35,20:35]
-friendshipData <- array(c(fd2.w1, fd2.w2,fd2.w3), dim = c(16, 16, 3))
-
-# running Siena
-#friendship <- sienaDependent(friendshipData)
-#alcohol <- varCovar(drink[20:35,])
-#mydata <- sienaDataCreate(friendship, alcohol)
-#myeffnull <- getEffects(mydata)
-#myalgorithm <- sienaAlgorithmCreate(projname = 's16_3')
-ansnullchains %>%
-  dplyr::filter(period == 1) %>%  #only look at chains from wave 1 to wave 2
-  #dplyr::group_by(rep) %>%
-  dplyr::select(rep, from = ego, to = alter) %>%
-  dplyr::mutate(val = as.numeric(!from == to),
-         from = paste0("V", parse_number(from)+1), # make the chains
-         to = paste0("V", parse_number(to)+1)) -> ansnullchainsw1w2
-colnames(fd2.w1) <- paste0("V", 1:16)
-rownames(fd2.w1) <- paste0("V", 1:16)
-wave1friends <- fortify(as.adjmat(fd2.w1))
-ms1 <- netvizinf::listMicrosteps(dat = wave1friends,
-                      microsteps = dplyr::filter(ansnullchainsw1w2, rep == 1))
-microsteps <- ms1[1:20]
-
-microsteps_df <- getMicrostepsDF(microsteps)
-pte <- pretween_edges(microsteps = microsteps)
-ptv <- pretween_vertices(pte = pte, layoutparams = list(n = 16))
-# step 1 is remove, step 9 is add.
-# get tweened data for all microsteps
-pall <- tween_microsteps(ptv, pte, microsteps_df)
-
-static_pall <- dplyr::filter(pall, .frame %in% c(15,20,22,24,26:28, 71:73, 75, 79, 81, 84))
-
-static_pall$type <- NA
-static_pall$type[which(static_pall$.frame %in% c(15,20,22,24,26:28))] <- "Remove Edge"
-static_pall$type[which(static_pall$.frame %in% c(71:73, 75, 79, 81, 84))] <- "Add Edge"
-
-set.seed(34569234)
-froms <- static_pall %>% dplyr::select(id = from, microstep, x = from.x,
-        y = from.y, ms, .frame, size, color) %>% unique
-tos <- static_pall %>% dplyr::select(id = to, microstep, x = to.x, y = to.y,
-        ms, .frame) %>% unique
-tos <- dplyr::left_join(tos, froms)
-nodedata <- rbind(froms, tos) %>% unique
-nodedata <- nodedata %>% tidyr::replace_na(replace = list(size = 1,
-        color = "#333333"))
-to_ani <- ggplot() +
-  geom_curve(data = static_pall, aes(x = from.x,
-        y = from.y, xend = to.x, yend = to.y, frame = .frame,
-        colour = ecolor, size = esize), curvature = 0.1, arrow = arrow(angle = 20,
-        length = unit(0.1, "inches"))) +
-  geom_point(data = nodedata,
-        aes(x = x, y = y, colour = color, size = size, frame = .frame)) +
-  scale_colour_identity() + scale_size_identity() + theme_void() +
-        coord_fixed()
-to_ani +
-  facet_wrap(~.frame, nrow=2, labeller = 'label_both') +
-  ThemeNet +
-  theme(panel.background = element_rect(color = 'black'))
-@
-
-The top row of Figure~\ref{fig:makinggif} shows an edge being removed, and the bottom row shows one being added.
-In both cases, the ego node that is acting changes color from black to red, and they also increase greatly in size.
-In the case of an edge being removed, in the top row, the edge that currently exists is emphasized with the same color and size change that the node gets, and as the animation proceeds the edge shrinks to nothing, as the ego node shrinks and changes color back to its original black.
-If an edge is being added, as in the bottom row of the figure, the ego node's appearance changes in the same way as when the edge is being removed, while the edge appears colored red from nothing, and grows to a large size, then changes color and size to match the rest of the edges, while the node shrinks and changes color to match the other nodes.
-
-<<adjmatani, fig.cap='A selection of frames from the adjacency matrix visualization animation for one series of microsteps.
-(Ego and alter node labels are removed to declutter the graph.) At the beginning of the animation, shown in the top row, there are two clearly connected components: one in the top left corner, and one in the bottom right corner.
-By the end, the component in the top left has spread out, while the bottom right component has shrunk, but remains closely connected.', warning = FALSE, message = FALSE, results='hide', fig.height=2.5>>=
-adjmatall <- matrix(0, nrow = 16, ncol = 16)
-rownames(adjmatall) <- paste0("V", 1:16)
-colnames(adjmatall) <- paste0("V", 1:16)
-dat2 <- sfmsall %>% filter(rep == 1)
-adjmatall2 <- NULL
-for(i in 0:55){
-  MS <- dat2[dat2$ms == i,c("from", "to")]
-  myadjmat <- adjmatall
-  for(j in 1:nrow(MS)){
-    myadjmat[MS[j,"from"][[1]],MS[j,"to"][[1]]] <- 1
-  }
-  adjmatall2 <- rbind(adjmatall2, myadjmat)
-}
-
-adjmatall2 <- cbind(adjmatall2, rep(0:55, each = 16))
-rownames(adjmatall2) <- NULL
-adjmatall2 <- as.data.frame(adjmatall2)
-adjmatall2$from <- rep(paste0("V", 1:16), 56)
-names(adjmatall2)[17] <- "ms"
-adjmatall2 %>% gather(to, val, V1:V16) -> adjmatall3
-
-adjmatall3$to <- as.factor(adjmatall3$to)
-levels(adjmatall3$to) <- gsub("V", "", levels(adjmatall3$to))
-adjmatall3$from <- as.factor(adjmatall3$from)
-levels(adjmatall3$from) <- gsub("V", "", levels(adjmatall3$from))
-adjmatall3$to <- factor(adjmatall3$to, levels = order1)
-adjmatall3$from <- factor(adjmatall3$from, levels = order1)
-
-adjmatall3$frame <- adjmatall3$ms
-#plots <- list()
-#for(i in 0:55){
-#plots[[i + 1]] <-
-
-dat3 <- adjmatall3 %>% filter(ms %in% c(0,2,4,6,8,10,12,43,45,47,49,51,53,55))
-
-dat4 <- dat3 %>% select(-ms) %>% spread(frame, val)
-dat5 <- dat4
-#p <- Sys.time()
-for(i in 4:16){
-  for(j in 1:nrow(dat5)){
-    if(dat4[j,i] != dat4[j,i-1]){
-      if(dat4[j,i] > dat4[j,i-1]){
-        dat5[j,i] <- 2
-      } else { dat5[j,i] <- 3}
-    }
-  }
-}
-#Sys.time() - p
-
-dat6 <- dat5 %>% gather(frame, val, 3:16) %>% mutate(frame = parse_number(frame))
-
-ggplot() +
-  geom_tile(data = dat6, aes(x = to, y = from, fill = as.factor(val)), color = '#969696') +
-  scale_fill_manual(values = c("white", "black", "red", "white")) +
-  geom_tile(data = dat6 %>% filter(val == 3), aes(x = to, y = from), fill = NA, color = 'red', size = .25) +
-  scale_x_discrete(position = "bottom") +
-  labs(x = "alter node (j)", y = "ego node (i)") +
-  ThemeNoNet +
-  theme(legend.position = 'none', axis.text.x.bottom = element_blank(),
-        axis.text.y = element_blank()) +
-  coord_fixed() +
-  facet_wrap(~frame, labeller = 'label_both', nrow = 2)
-#}
-#plots
-@
-
 Finally, we combine 1,000 simulations from model M1 into a visualization that displays the entire microstep process for all simulations in Figure ~\ref{fig:alledgests}.

### 2018-07-25 paper-child.Rnw
trying to comply with ASA guidelines

@@
 \scalebox{0.9}{
-\begin{tabular}{l|ccc|ccc }
-  \hline
+\begin{tabular}{lcccccc }
+
   & \multicolumn{3}{|c|}{Friendship Data} & \multicolumn{3}{|c|}{Senate Data} \\
@@
   & M1 & M2 & M3 & M1 & M2 & M3 \\
+  \hline
   $\alpha_1$ & \Sexpr{SFE$M1[1]} & \Sexpr{SFE$M2[1]} & \Sexpr{SFE$M3[1]} & \Sexpr{SSE$M1[1]} & \Sexpr{SSE$M2[1]} & \Sexpr{SSE$M3[1]} \\
@@
   $\alpha_3$ & -- & -- & -- & \Sexpr{SSE$M1[3]} & \Sexpr{SSE$M2[3]} & \Sexpr{SSE$M3[3]} \\
+  \hline
   $\beta_1$ & \Sexpr{SFE$M1[3]} & \Sexpr{SFE$M2[3]} & \Sexpr{SFE$M3[3]} & \Sexpr{SSE$M1[4]} & \Sexpr{SSE$M2[4]} & \Sexpr{SSE$M3[4]} \\

### 2018-07-26 paper-child.Rnw
remove some lines
proofing

@@
 
-Social networks, such as collaboration networks between academic researchers or friendship networks in a school or university have been studied for decades.
-Statistical models such as exponential random graph models (ERGM) and latent space models (LSM) are often used when studying one observation of a network when trying to model its structure.
+Social networks, such as collaboration networks between academic researchers or friendship networks have been studied for decades.
+Statistical models such as exponential random graph models (ERGM) and latent space models (LSM) are often used when studying one observation of a network when trying to model its structure \citep{ergm, lsm}.
 Social networks can also be observed at many points in time, and ERGM or LSM may not be appropriate for modelling the changes because they do not model network change as time passes.
-When studying \textit{dynamic networks}, networks observed at many time points, we need a model that allows for variation in the network over time.
-Models for dynamic social networks hold a great deal of potential because of how realistic they can be.
-Social networks do not form spontaneously, but rather they \textit{evolve} over time as edges are formed and dissolved, and new nodes join the social structure.
+When studying \textit{dynamic networks} observed at many time points, we need a model that allows for variation over time.
+Models for dynamic social networks hold a great deal of potential because of how realistic they can be: social networks evolve over time as edges are formed and dissolved and new nodes join.
 Modeling the underlying mechanisms that create network changes in time is very complex but also provides potential to uncover hidden truths.
 
-\par To model dynamic social networks, we use stochastic actor-oriented models (SAOMs), introduced by \citet{saompaper}.
-These models are different from other network models because they allow us to add node level information into the model with the traditional network structure information.
-Adding node-level information into the model is a more intuitive approach to model changes in a social network.
-For example, we expect people with common interests to be more likely to form relationships, and SAOMs allow us to incorporate this information on shared interests in the modeling process.
+\par One set of models for dynamic social networks, are stochastic actor-oriented models (SAOMs), introduced by \citet{saompaper}.
+These models are different from other dynamic network models because they allow us to add node level information into the model with the traditional network structure information.
+Adding node-level information into the model is a more intuitive approach to model changes in a social network because, for example we expect people with common interests to be more likely to form relationships.
+SAOMs allow us to incorporate this information on shared interests in the modeling process.
 
@@
 They are relatively new, especially compared to the classic ERGMs, and they are not very tractable analytically.
-Likelihood functions quickly become very complex objects to analyze due to the dependency structure inherent in the data.
-Therefore, computational methods, such as Markov chain Monte Carlo (MCMC) are used to fit models.
+Likelihood functions quickly become very complex due to the dependency structure inherent in the data, and so computational methods, such as Markov chain Monte Carlo (MCMC) are used to fit models.
 SAOMs are typically fit to dynamic social network data using a series of MCMC phases for finding method of moments (MoM) estimates of model parameters.
-In order to estimate the parameters of SAOMs, we use the software SIENA, and its \texttt{R} implementation \texttt{RSiena}, which was developed by \citet{RSiena}.
-This software is a considerable contribution to the field of social network analysis, but the many simulations involved in parameter estimation are largely hidden from the software user.
-In order to better understand the model-fitting process of \texttt{RSiena}, we attempt to bring model fits and the fitting process to the foreground by visualizing the underlying methods and structures.
-Utimately, we aim to help researchers working with the models better understand the structure of and results from these models.
-
-\par To learn more about the SAOM fitting process, we use model visualization techniques (see \citet{modelvis}) to display the model in the data space, view collections of models instead of single models, and explore the algorithms fitting the SAOMs instead of studying only the final output.
-SAOMs are an excellent example of a family of models that can reap many benefits from the application of model visualization techniques.
+In order to estimate the parameters of SAOMs, we use the software SIENA, and its \texttt{R} implementation \texttt{RSiena} by \citet{RSiena}.
+This software is a considerable contribution to the field of social network analysis, but the parameter estimation process is largely hidden from the user.
+In order to better understand the model-fitting process of \texttt{RSiena}, we visualize the underlying methods and structures, and utimately aim to help researchers working with SAOMs better understand the structure of and results from these models.
+
+\par To learn more about the SAOM fitting process, we use model visualization techniques (see \citet{modelvis}) to display the model in the data space, view collections of models instead of single models, and visually explore the algorithms fitting the SAOMs.
+SAOMs are a model family that can reap many benefits from the application of model visualization techniques.
 For instance, the models themselves include a continuous-time Markov chain (CTMC) that is completely hidden from the analyst in the model fitting process.
@@
 
-The remainder of this paper is organized as follows: In Section~\ref{sec:netintro}, we introduce basic concepts of networks and network visualizations.
+%The remainder of this paper is organized as follows: In Section~\ref{sec:netintro}, we introduce basic concepts of networks and network visualizations.
 In Section~\ref{sec:saoms}, we formally define the family of stochastic actor-oriented models for social network analysis.
@@
 
-We provide a brief introduction to the structure of network data and common network visualization methods.
-
-\subsubsection{Introduction to Network Structures}
-
-Network data across fields have similar data structures.
+We provide a brief introduction to the structure of network data and common network visualization methods. \\
+
+\noindent \textbf{Introduction to Network Structures:} Network data across fields have similar data structures.
 There are always units of observation and connections between those units.
@@
 The nodes and edges themselves can also have inherent variables of interest, e.g.\ the age, gender, and political affiliation of people in a social network and the type of relationship (friends, married, etc.).
-
-These two network data structures pose unique problems to analysts, such as, ``How does the strength of a tie between two nodes affect the overall structure of the network?" or ``Do differences in node variables affect the formation of edges?" We propose to answer these questions and more through visual exploration of network data and models.
-
-\subsubsection{Visualizing Network Data} \label{sec:netvizintro}
-
-Network visualization, also called network mapping, is a prominent subfield of network analysis.
+\\
+
+%These two network data structures pose unique problems to analysts, such as, ``How does the strength of a tie between two nodes affect the overall structure of the network?" or ``Do differences in node variables affect the formation of edges?" We propose to answer these questions and more through visual exploration of network data and models.
+\\
+
+\noindent \textbf{Visualizing Network Data:} Network visualization, also called network mapping, is a prominent subfield of network analysis.
 Visualizing network data is uniquely difficult because of the structure of the data itself.
-Most data visualizations rely on well-defined axes inherited from the data, such as Cartesian coordinates or spatial locations.
-Network data, however, carry no such inherent structure.
+Most data visualizations rely on well-defined axes inherited from the data, such as Cartesian coordinates or spatial locations, but network data typically do not have an inherent structure.
 
@@
 
-There are two primary methods used to visualize networks that are very different from most traditional data visualizations.
-They are the node-link diagram and the adjacency matrix visualization \citep{knuth2000, adjmatpro}.
+There are two primary methods used to visualize networks: the node-link diagram and the adjacency matrix visualization \citep{knuth2000, adjmatpro}.
 For example, consider a network with five nodes, $\{1,2,3,4,5\}$, connected by five directed edges:  $\{2 \to 4, 3 \to 4, 1 \to 5, 3 \to 5, 5 \to 4\}$.
@@
 These lines can also have arrows on them indicating the direction of the edge for directed networks, as show in Figure~\ref{fig:toyplots}.
-Because there is usually no natural placement of the points (the exception is nodes with spatial locations), a random placement of the points is used, then adjusted with a layout algorithm, of which there are many \citep{drawingalgs}.
-
-\par Some commonly used layout algorithms, such as  the Kamada-Kawai (KK) layout \citep{kamadakawai} and the Fruchterman-Reingold (FR) layout \citep{fruchterman-reingold}, are designed to mimic physical systems, drawing the graphs based on the ``forces" connecting them.
-Other algorithms use multi-dimensional scaling or properties of the adjacency matrix (e.g.
+Because there is usually no natural placement of the points, a random placement of the points is used, then adjusted with a layout algorithm, of which there are many \citep{drawingalgs}.
+Some commonly used algorithms, such as the Kamada-Kawai (KK) layout \citep{kamadakawai} and the Fruchterman-Reingold (FR) layout \citep{fruchterman-reingold}, are designed to mimic physical systems, drawing the graphs based on the ``forces" connecting them.
+%Other algorithms use multi-dimensional scaling or properties of the adjacency matrix (e.g.
 its eigenstructure), to place the nodes \citep{drawingalgs}.
@@
 \end{equation}
-We only consider the presence or absence of an edge, but if the network has weighted edges, the entries in the adjacency matrix are the edge weights.
+In this paper, we only consider the presence or absence of an edge, but if the network has weighted edges, $A_{ij}$ is the weight of the edge from $i \to j$.
 An adjacency matrix visualization for our toy example is also shown in Figure~\ref{fig:toyplots}.
@@
 It models the change of a network in time, while accounting for changes in network structure due to actor-level covariates.
-This model was first introduced by Snijders in 1996 \citep{saompaper}.
 The two titular properties of SAOMs, stochasticity and actor-orientation, are key to understanding networks as they exist naturally.
@@
 
-We use the term \textit{dynamic network} to refer to a network, consisting of a fixed set of $n$ nodes, that is changing over time, and is observed at $M$ discrete time points, $t_1, \dots, t_M$ with $t_1 \le t_2 \le \dots \le t_M$.
-We denote the network observation at timepoint $t_k$ by $x(t_k)$.
-The SAOM assumes that this longitudinal network of discrete observations is embedded within a continuous time Markov chain (CTMC), which we will denote $X(T)$.
-This process is almost entirely unobserved: we assume that the beginning of the process, $X(0)$, is equivalent to the first network observation $x(t_1)$, while the end of the process, $X(\infty)$, is equivalent to the last observation $x(t_M)$.
+We use the term \textit{dynamic network} to refer to a network, consisting of a fixed set of $n$ nodes, that is changing over time, and is observed at $M$ discrete time points, $t_1, \dots, t_M$ with $t_1 < t_2 < \dots < t_M$.
+We denote the network observation at timepoint $t_m$ by $x(t_m)$ for $m = 1, \dots M$.
+The SAOM assumes that this longitudinal network of discrete observations is embedded within a CTMC, which we will denote by $X(T)$.
+This CTMC is almost entirely unobserved: we assume that the beginning of the process, $X(0)$, is equivalent to the first network observation $x(t_1)$, while the end of the process, $X(\infty)$, is equivalent to the last observation $x(t_M)$.
 Nearly all other steps in the chain are unseen, with the exception of $x(t_2), \dots, x(t_{M-1})$.
 Unlike the first and last observations of the network, these ``in-between" observations do not have direct correspondence with steps in the continuous time Markov chain.
-Thus, the ``in-between" observations are considered to be ``snapshots" of the network at some point between two steps in the CTMC.
-
-The CTMC process $X(T)$ is a series of single tie changes that happen according to a pre-defined rate function, where one actor at a time is selected and it adds or removes one outgoing tie, or does nothing.
-Once an actor is chosen at random according to the rate function, it is ``given" the chance to change a tie, and it tries to maximize its utility function based on the current and possible future states of the network.
+Thus, the observations $x(t_2), \dots, x(t_{M-1})$ are considered to be ``snapshots" of the network at some point between two steps in the CTMC.
+
+The CTMC process $X(T)$ is a series of single tie changes that happen according to the SAOM's pre-defined rate function, where one actor at a time is selected and it adds or removes one outgoing tie, or does nothing.
+Once an actor is chosen at random according to the rate function, it is ``given" the chance to change a tie, and it tries to maximize its utility function based on the current and currently reachable states of the network.
 \\
 
-\noindent \textbf{The Rate Function} In general, each actor $i$ in the network $x$ may change its ties, $x_{ij}$, to other nodes $j \neq i$ according to the SAOM's  \emph{rate function}, $\rho(x, \mathbf{z}, \boldsymbol{\alpha})$, where $\boldsymbol{\alpha}$ are the parameters in the function $\rho$, and $x$ is the current network state, with covariates of interest $\mathbf{z}$.
+\noindent \textbf{The Rate Function} In general, each actor $i$ in the network $x$ gets an opportunity to changes its ties, $x_{ij}$, to other nodes $j \neq i$ according to the SAOM's \emph{rate function}, $\rho(x, \mathbf{z}, \boldsymbol{\alpha})$, where $\boldsymbol{\alpha}$ are the parameters in the function $\rho$, and $x$ is the current network state, with covariates of interest $\mathbf{z}$.
 With this general formulation, it is possible that each node will have a different rate of change.
@@
 With this simple rate function, we have just one set of rate parameters in the overall model.
-In this simple model, the rate parameters dictate how often an actor $i$ ``gets an opportunity" to change one of its ties, $x_{ij}$, for $j \in \{1, \dots, n\}$ in the time period from $t_{m}$ to $t_{m+1}$ for $m = 1, \dots, M-1$ (Note that when $j = i$, no change is made.) We also assume that the actors $i$ are conditionally independent given their ties, $x_{i1}, \dots, x_{in}$ at the current network state.
+In this simple model, the rate parameters dictate how often an actor $i$ ``gets an opportunity" to change one of its ties, $x_{ij}$, for $j \in \{1, \dots, n\}$ in the time period from $t_{m}$ to $t_{m+1}$ for $m = 1, \dots, M-1$ (Note that if $j = i$, no change is made.) A SAOM also assumes that the actors $i$ are conditionally independent given their ties, $x_{i1}, \dots, x_{in}$ at the current network state.
 Let $\tau(i|x,m) $ be the wait time until actor $i$ gets to the opportunity to change its current set of ties.
-For any time point, $T$, where $t_{m} \leq T < t_{m+1}$, the waiting time to the next change made by actor $i$ is exponentially distributed with expected value $\alpha_m^{-1}$.
-Then, we express this in Equation~\ref{eq:c1}.
+For any time point, $T$, where $t_{m} \leq T < t_{m+1}$, the waiting time to the next change made by actor $i$ is exponentially distributed with expected value $\alpha_m^{-1}$, as shown in Equation~\ref{eq:c1}.
 
@@
 \noindent \textbf{The Objective Function} Because of the conditional independence assumptions given in Equation~\ref{eq:c1}, we can consider the objective function for each node separately, as only one tie from one node is changing at a time in the CTMC.
-The node $i$, which is the node that is given the opportunity to change at the current time point, is called the \textit{ego} node.
+The node $i$, which is given the opportunity to change at the current time point, is called the \textit{ego} node.
 It has the potential to interact with all other nodes in the network, $j \neq i$.
@@
 \begin{equation} \label{eq:of1}
-f_i(\boldsymbol{\beta}, x) = \sum_k \beta_k s_{ik}(x, \mathbf{Z}),
+f_i(\boldsymbol{\beta}, x, \mathbf{Z}) = \sum_{k=1}^K \beta_k s_{ik}(x, \mathbf{Z}),
 \end{equation}
-for $x \in \mathcal{X}$, the space of all possible directed networks with the $n$ nodes, and $\mathbf{Z}$, the matrix of covariates.
-The vector $\boldsymbol{\beta}$ contains the parameters of the model with corresponding network and covariate statistics, $s_{ik}(x, \mathbf{Z})$, for $k = 1,\dots, K$.
-To increase the value of $f_i$, $i$ will destroy one tie where $x_{ij} = 1$, create one tie where $x_{ij}=0$, or make no change if any changes result in lower values of $f_i$.
+for $x \in \mathcal{X}$, the space of all possible directed networks with the $n$ nodes, and $\mathbf{Z}$, the matrix of covariates, and $K$ the total number of parameters in the objective function.
+The vector $\boldsymbol{\beta}$ contains the parameters of the model with corresponding statistics, $s_{ik}(x, \mathbf{Z})$, for $k = 1,\dots, K$.
+To increase the value of $f_i$, $i$ will destroy one tie where $x_{ij} = 1$, create one tie where $x_{ij}=0$, or make no change if all changes will result in lower values of $f_i(\boldsymbol{\beta}, x, \mathbf{Z})$.
 
 According to \citet[p.
-371]{snijders01}, there should be at least two parameters included in the model: $\beta_1$ for the outdegree of a node, and $\beta_2$ for the number of reciprocal ties held by a node.
-(These effects are also used in ERGMs.) The outdegree represents the propensity of nodes with a lot of outgoing ties to form more outgoing ties (the "rich get richer" effect), and the reciprocity parameter measures the tendency of outgoing ties to be returned within a network.
-The statistics corresponding to these two parameters, $s_{i1}(x)$ and $s_{i2}(x)$ are given in Figure~\ref{tab:effects}.
+371]{snijders01}, there should be at least two parameters included in the model: $\beta_1$ for the outdegree of nodes, and $\beta_2$ for the reciprocity of nodes.
+(These effects are also used in ERGMs.) The outdegree represents the propensity of nodes with a lot of outgoing ties to form more outgoing ties (the ``rich get richer" effect), and the reciprocity parameter measures the tendency of outgoing ties to be returned within a network.
+The statistics corresponding to these two parameters, $s_{i1}(x)$ and $s_{i2}(x)$ are given inTable~\ref{tab:effects}.
 In the \texttt{RSiena} software, there are over 80 possible $\boldsymbol{\beta}$ parameters to add to the model.
@@
 
-The $\boldsymbol{\beta}$ parameters are either structural or covariate parameters.
-The structural parameters, such as the outdegree and reciprocity parameters, correspond to statistics that are only functions of the current network state, $x$.
+The parameters of $f_i(\boldsymbol{\beta}, x, \mathbf{Z})$ are either structural or covariate parameters.
+The structural parameters, such as the outdegree and reciprocity parameters, correspond to statistics that are \emph{only} functions of the current network state, $x$.
 They model underlying mechanisms of network change, answering questions such as, ``How does the existing network structure influence change in the network?" The covariate parameters are functions of $x$ and additional node-level covariates of interest, $\mathbf{Z}$.
-A table of some possible structural and covariate parameters and their corresponding statistics is given in Table~\ref{tab:effects}.
+Some possible covariate parameters and their corresponding statistics are discussed in Section~\ref{sec:exdatamods} and shown in Figure~\ref{fig:structures.}
 
@@
 %\\
-\multicolumn{3}{l}{\textbf{Covariate Effects}}\\  
-Name & Statistic & Figure \\
-\hline \hline
-covariate-alter  & $s_{i4}(x) = \sum_j x_{ij}z_j$ & \includegraphics[width=3cm]{img/covaralter} \\  \hline
-covariate-ego  & $s_{i5}(x) = z_i\sum_j x_{ij}$ & \includegraphics[width=3cm]{img/covarego} \\ % \hline
+% \multicolumn{3}{l}{\textbf{Covariate Effects}}\\  
+% Name & Statistic & Figure \\
+% \hline \hline
+% covariate-alter  & $s_{i4}(x) = \sum_j x_{ij}z_j$ & \includegraphics[width=3cm]{img/covaralter} \\  \hline
+% covariate-ego  & $s_{i5}(x) = z_i\sum_j x_{ij}$ & \includegraphics[width=3cm]{img/covarego} \\ % \hline
 %same covariate & $s_{i6}(x) = \sum_j x_{ij} \mathbb{I}(z_i = z_j)$ & \includegraphics[width=3cm]{img/covarsim}%\\
@@
 \end{tabular}
-\caption{\label{tab:effects} Examples of effects included in a SAOM.
+\caption{\label{tab:effects} Examples of structural effects included in a SAOM.
 The dark nodes in the Figures are the ego nodes, and all others are alters.
-The dotted lines are ties the ego node is en(dis)couraged to make when the parameter corresponding to each effect is positive (negative).
-Node shapes represent levels of the covariate $z$.}
+The dotted lines are ties the ego node is en(dis)couraged to make when the parameter corresponding to each effect is positive (negative).}
 \end{table}
 
-The objective function also defines the probability that actor $i$, when it is the ego node, chooses to change the tie to actor $j$, $p_{ij}(\boldsymbol{\beta}, x, \mathbf{Z})$.
-Let $x(i \leadsto j)$ denote the network identical to $x$ with the exception of tie $x_{ij}$.
-It has changed to $1-x_{ij}$.
+The objective function also defines the probability that actor $i$, when it is the ego node, chooses to change the tie to actor $j$, denoted by $p_{ij}(\boldsymbol{\beta}, x, \mathbf{Z})$.
+Let $x(i \leadsto j)$ be the network identical to $x$ with the exception of tie $x_{ij}$, which is equal to $1-x_{ij}$.
 The probility that the tie $x_{ij}$ changes is defined as:
@@
 p_{ij}(\boldsymbol{\beta}, x, \mathbf{Z}) = \dfrac{\exp\left\{f_i(\boldsymbol{\beta}, x(i\leadsto j), \mathbf{Z})\right\}}{\sum_{h \neq i} \exp\left\{f_i(\boldsymbol{\beta}, x(i \leadsto h), \mathbf{Z})\right\}}
-\end{equation}
-When $i = j$, the numerator represents the exponential of the value of the objective function when evaluated at the current network state.
+\end{equation}%When $i = j$, the numerator represents the exponential of the value of the objective function when evaluated at the current network state. 
 When the value of the objective function is high at the current state, the probability of not making a change is also high.
-In the CTMC, when actor $i$ may make a change, it chooses which tie $x_{i1, \dots, x_{in}}$ to change at random according to the probabilities $p_{ij}(\boldsymbol{\beta}, x, \mathbf{Z})$.
+In the CTMC, when actor $i$ may make a change, it chooses which tie $x_{i1}, \dots, x_{in}$ to change at random according to the probabilities $p_{ij}(\boldsymbol{\beta}, x, \mathbf{Z})$.
 The objective function and the resulting values of $p_{ij}$ are combined with the rate function to fully describe the CTMC that is used to model network change in a SAOM.
@@
 \par The underlying SIENA software uses a Robbins-Monro algorithm (see \citet{robbinsmonro}) to estimate the solution of the moment equation
-$$ E_{\theta}S = s_{obs}$$
-where $\theta = (\boldsymbol{\alpha}, \boldsymbol{\beta})$ is the vector of rate and objective function parameters, and $s_{obs}$ is the observed vector of model statistics.
+\begin{equation}\label{eq:mom}
+E_{\theta}S = s_{obs}
+\end{equation}
+where $\theta = (\boldsymbol{\alpha}, \boldsymbol{\beta})$ is the vector of rate and objective function parameters, and $s_{obs}$ is the observed vector of the model statistics over all network observations.
 The full SIENA MoM estimation algorithm operates in three phases, as described in \citet{RSienamanual, SienaAlgs}.
@@
 In each of the the first two phases of the algorithm, the procedure uses ``microsteps" that are simulated from the model as it exists in its current state in order to update either the score functions or the parameter estimates.
-These simulated microsteps are observed instances of the CTMC that is the backbone of the SAOM.
-In Section~\ref{sec:ModelVis}, we further explore these phases in the SIENA method-of-moments algorithm through visualization, bringing them out of their ``black box" and into the light.
+These simulated microsteps are realized steps in the CTMC that is the backbone of the SAOM.
+In Section~\ref{sec:algorithms}, we further explore these phases in the SIENA method-of-moments algorithm through visualization, bringing them out of their ``black box" and into the light.
 
@@
 
-\section{Example Data and Models}
+\section{Example Data and Models}\label{sec:exdatamods}
 
@@
 
-<<alldataviz, fig.cap="The smaller friendship network data we will be modelling throughout the paper.
+<<alldataviz, fig.cap="The small friendship network data we will be modelling throughout this paper.
 The nodes change color as their drinking behavior changes, and the network becomes less densely connected over time.
-These are the mechanisms we hope to capture with SAOMs.", out.width='.99\\textwidth', fig.height=3, results='hide'>>=
+These are the mechanisms we hope to incorporate and model with SAOMs.", out.width='.99\\textwidth', fig.height=3, results='hide'>>=
 source("code/00e_small_friends.R")
@@
 
-Because we require binary edges for SAOMs, we focus only on very strong collaborations. For our senate collaboration networks, $x$, edges are defined as
+Because we require binary edges for SAOMs, we focus only on very strong collaborations, where $WPC$ is high. For a senate collaboration network $x$, edges are defined as
 \begin{equation}\label{eq:wpc2}
@@
 
-  & \multicolumn{3}{|c|}{Friendship Data} & \multicolumn{3}{|c|}{Senate Data} \\
+  & \multicolumn{3}{c}{Friendship Data} & \multicolumn{3}{c}{Senate Data} \\
   \hline

### 2018-07-26 paper.Rnw
proofing

@@
   \title{\bf Model Visualization Techniques for a Social Network Model}
-  \author{Samantha Tyner\thanks{
-    The authors gratefully acknowledge funding from the National Science Foundation Grant \# DMS 1007697.
-All data collection has been conducted with approval from the Institutional Review Board IRB 10-347}\hspace{.2cm}\\
+  \author{Sam Tyner \hspace{.2cm}\\
     Department of Statistics and Statistical Laboratory, Iowa State University\\
@@
 \noindent%
-{\it Keywords:} social network analysis, model visualization, dynamic networks, network visualization, network mapping, animation
+{\it Keywords:} social network analysis, dynamic networks, network mapping, animation, statistical graphics
 \vfill

### 2018-08-15 paper-child.Rnw
ugh what is happening with the fonts of the plots????

@@
 
-\noindent \textbf{Introduction to Network Structures:} Network data across fields have similar data structures.
+\noindent \textbf{Network Data Structure:} Network data across fields have similar data structures.
 There are always units of observation and connections between those units.
@@
 Some prominent names are shown to demonstrate their place in the network.
-
-\noindent \textbf{Models:} We first consider the 16 actor subset of the teenage friends and lifestyle data show in Figure~\ref{fig:alldataviz}.
-To this data, we fit three different SAOMs.
-Each SAOM used a simple rate function, $\alpha_m$, and an objective function with two or three parameters.
+\\
+
+\noindent \textbf{Models:} To our example data, we fit three different SAOMs.
+Each SAOM uses a simple rate function, $\alpha_m$, and an objective function with two or three parameters.
 The first model, M1, contains the absolute minimum number of parameters in the objective function $f_i(x)$:
 \begin{equation}\label{eq:m1}
-f_i(x)^{M1} = \beta_1s_{i1} + \beta_2s_{i2},
+f_i(x)^{M1} = \beta_1s_{i1}(x) + \beta_2s_{i2}(x),
 \end{equation}
-where $s_{i1}$ is the density network statistic and $s_{i2}$ is the reciprocity network statistic for actor $i$ at the current network state $x$.
+where $s_{i1}$ is the density network statistic and $s_{i2}$ is the reciprocity network statistic for actor $i$ at the current network state $x$ (see Table~\ref{tab:effects}.
 The second and third models, M2 and M3, contain one additional parameter each in the objective function which were determined by a Wald-type test provided in the \texttt{RSiena} software to be significant, with $p$-values less than 0.05 \citep{RSienaTest}.
@@
 \begin{align}\label{eq:m2m3}
-f_i(x)^{M2} & = \beta_1s_{i1} + \beta_2s_{i2} + \beta_3s_{i3} \\
-f_i(x)^{M3} & = \beta_1s_{i1} + \beta_2s_{i2} + \beta_4s_{i4},
+f_i(x,z)^{M2} & = \beta_1s_{i1}(x) + \beta_2s_{i2}(x) + \beta_3s_{i3}(x, z) \\
+f_i(x)^{M3} & = \beta_1s_{i1}(x) + \beta_2s_{i2}(x) + \beta_4s_{i4}(x),
 \end{align}
@@
   geom_net(aes(shape = as.factor(group)), directed = T, labelon = T,
-           labelcolour='grey80',vjust = 0.5, hjust =0.5, arrowgap = .15,
-           colour = 'black', size=10, fontsize = 5,
-           ecolour = c("grey40", "grey40","red", "grey40")) +
+           labelcolour='white',vjust = 0.5, hjust =0.5, arrowgap = .15,
+           size=10, fontsize = 5, ecolour="black",
+           linetype = c("solid", "solid","dotted", "solid")) +
   expand_limits(x=c(-0.1,1.1), y=c(-0.1,1.1)) +
   ThemeNet +
-  theme(legend.position = "none")
+  theme(legend.position = "none") + 
+  labs(title = expression(paste(s[i3], "(x,z)", "=",Sigma[j != h], x[ij], x[ih], x[hj] %.% 
+                                I, "(", z[i], "=", z[h]!= z[j], ")")
+                          )
+       )
 @
@@
 <<dab, echo=FALSE, fig.width=2, fig.height=2, message = FALSE, warning = FALSE, fig.align='center', out.width='.49\\textwidth'>>=
-dade <- data.frame(from = c('i', 'i', 'h', 'j', 'i'), to = c('h', 'j', 'k', 'k','k'))
+dade <- data.frame(from = c('i', 'i', 'h', 'k', 'i'), to = c('h', 'k', 'j', 'j','j'))
 dadn <- data.frame(id = letters[8:11], group = c(1,1,1,1))
@@
 ggplot(data = dad, aes(from_id = from, to_id = to)) +
-  geom_net(directed = T, labelon = T,
-           ecolour = c("black", "black", "black","red", "black" ,"black" ),
-           labelcolour='grey80',vjust = 0.5, hjust =0.5, arrowgap = .15, size=10, fontsize = 5) +
+  geom_net(directed = T, labelon = T, ecolour="black",
+           linetype = c("solid", "solid", "dotted", "solid","solid" ,"solid"),
+           labelcolour='white',vjust = 0.5, hjust =0.5, arrowgap = .15, size=10, fontsize = 5) +
   expand_limits(x=c(-0.1,1.1), y=c(-0.1,1.1)) +
   ThemeNet +
-  theme(legend.position = "none")
-@
-\caption{\label{fig:dad}Realization of a N22 between actors $i$ and $k$.}
+  theme(legend.position = "none") + 
+  labs(title = expression(paste(s[i4], "(x)", "=", "|{j:", x[ij],"=0, ", Sigma[h], x[ih], x[hj] >= 2, "}|" )
+                          )
+       )
+@
+\caption{\label{fig:dad}Realization of a N22 between actors $i$ and $j$. The dotted tie is encouraged if $\beta_4$ has a positive value.}
 \end{subfigure}
-\caption{The additional network effects included in our models fit to the friends data. On the left, a JTT and on the right, a N22 between $i$ and $k$.}
+\caption{The additional network effects included in our models fit to the friends data.
+On the left, a JTT between $i$ and $j$, who have different covariate values, and on the right, a N22 between $i$ and $j$.}
 \end{figure}
 
-where $s_{i3} = \sum_{j \neq h} x_{ij}x_{ih}x_{hj} \mathbb{I}(z_i = z_h \neq z_j)$, and $s_{i4} =  |\{j : x_{ij} = 0, \sum\limits_h x_{ih}x_{hj} \geq 2\}|$.
+\noindent where $s_{i3}(x,z) = \sum_{j \neq h} x_{ij}x_{ih}x_{hj} \mathbb{I}(z_i = z_h \neq z_j)$, and $s_{i4}(x) =  |\{j : x_{ij} = 0, \sum\limits_h x_{ih}x_{hj} \geq 2\}|$.
 These statistics are known as the \emph{number of jumping transitive triplets} (JTT) and the \emph{number of doubly achieved distances two effect} (N22), respectively.
-The first statistic emphasizes triad relationships that are formed between actors from different covariate groups, while the other emphasizes indirect ties between actors.
-The covariate groups are determined by the student's drinking behavior.
-These additional effects are shown in Figure~\ref{fig:jtt} and Figure~\ref{fig:dad}.
+The first statistic emphasizes triad relationships formed between actors with different covariate values, while $s_{i4}$ emphasizes indirect ties between actors.
+The covariate for the friendship data is the drinking behavior, and is the number of bills authored for the senate data.
+These additional effects are shown in Figure~\ref{fig:jtt} and Figure~\ref{fig:dad}, where the ties affected by the parameter value are represented by dotted lines.
 
@@
 Finally, in Section~\ref{sec:algorithms} we use model visualization to explore the underlying algorithms in the SAOM fitting process.
+We conclude with a discussion in Section~\ref{sec:discussion}.
  %In terms of the senate data, then, the value of $\beta_3$ should dictate how willing a senator is to coauthor a bill on which the author of the bill has a different level of authorship, assuming there is an intermediary between the two who has the same authorship level as the first.
@@
 Every good data analysis includes both numerical and visual summaries of the data, so why restrict model description and diagnostics to numerical summaries?
-The concept of model visualization was developed to complement traditional model diagnostic tools such as $R^2$ values and residual plots \citep{modelvis}.
-When performing analyses, there are three different types of ``model": the model family, the model form, and the fitted model.
-The latter is primarily what one thinks of first when considering a model in a data analysis, where a specified model is fit to data, and parameter estimates are reported.% In the context of SAOMs, the \emph{fitted model} contains the form of the rate and objective functions, the estimated rate parameters, and the estimated objective function parameters.
-The \textit{model form} describes the the model \textit{before} the fitting process, defining which parameters are in the model within the context of the larger model family.% In SAOMs, the \emph{model form} includes description of the rate and objective functions and the variables therein that describe how the network evolves over time.
+The concept of \emph{model visualization} was developed to complement traditional model diagnostic tools such as $R^2$ values and residual plots \citep{modelvis}.
+In a modeling task, there are three different things called ``model" under consideration: the model \emph{family}, the model \emph{form}, and the \emph{fitted} model.
+The latter is usally what first comes to mind when considering a model, where a specified model is fit, and parameter estimates are reported.% In the context of SAOMs, the \emph{fitted model} contains the form of the rate and objective functions, the estimated rate parameters, and the estimated objective function parameters.
+The \textit{model form} describes the the model \textit{before} the fitting process, defining which parameters are in the model within the context of the larger model family.
+% In SAOMs, the \emph{model form} includes description of the rate and objective functions and the variables therein that describe how the network evolves over time.
 Finally, the \textit{model family} is the broadest description of the model.
@@
 Specifically, we want to learn more about how the \emph{model form} we choose affects the \emph{fitted model} by using our example data sets and our visualization toolbox.
-We begin by introducing the five models that we fit to our example data.
-We then use the five models to guide our visual explorations of SAOMs under the three principles of model visualization.
+%We begin by introducing the five models that we fit to our example data.
+We use the data and models described in Section~\ref{sec:exdatamods} to guide our visual explorations of SAOMs under the three principles of model visualization.
 %We begin with novel tools and ideas to view a SAOM in the data space of a dynamic network.
@@
 
-The first way we hope to better understand SAOMs is by viewing the model(s) in the data space.
+The first way we hope to better understand SAOMs is by viewing the model in the data space.
 In \citeauthor{modelvis}, they chose define the \textit{data space} as ``the region over which we can reliably make inferences, usually a hypercube containing the data" \citep[p.~206]{modelvis}.
-But what does this definition mean for network models?
-We define the data space as the combined data structures of the network: the node, edge, and time information.
+But what does this definition mean for dynamic network models?
+We interpret the data space for our application as the combined data structures of the network: the node, edge, and time information.
 %We want to make inferences on parameters in the model, but the multi-level data structure of networks makes the ``region of inference" difficult to define.
@@
 
-The node, edge, and time data can be visualized together in various ways, and one tool that can bring the node, edge, and time data spaces together is the \texttt{R} package \texttt{geomnet} \citep{geomnet}.
-Different visual features in the node-link diagram can be tied to the underlying node or edge data.
+The node, edge, and time data can be visualized together in various ways, and one tool that can help is the \texttt{R} package \texttt{geomnet} \citep{geomnet}.
+The network's node or edge data can be mapped to different visual features in the node-link diagram.
 The color, size, and shape of the points can be used to represent variables in the node data, while the color, linewidth, and linetype of the lines between points can be used to represent the edge variables.
@@
 To demonstrate, we use \texttt{geomnet} to visualize the connections in the $111^{th}$ United States Senate at two different points: when Hillary Clinton was in the senate, and after she left to become Secretary of State.
-Clinton was only in the $111^{th}$ senate for 17 days, from January 3, 2009, to January 20, 2009, when she was in the middle of her second term as senator from New York.
-In that time, she authored two bills and was a cosponsor on 17 other bills.
-There are many Democratic senators who cosponsored both of her bills, giving their edges a $WPC$ of 1.
-So with Clinton included in the node-link diagram, the senate looks much more collaborative than it does without her.
-We then map the size of the node to the numer of bills authored by the senator, the shape to sex, color to party.
-Bills, sex, and party are all potential covariates in a SAOM.
-In addition, we visualize the strength of the tie by mapping the linewidth of the edge to the $WPC$ value between the two senators.
+Clinton was only in the $111^{th}$ senate from January 3-20, 2009, in the middle of her second term as the junior senator from New York.
+In that time, she authored two bills and was a cosponsor on 17s.
+There are many senators who cosponsored both of her bills, giving their edges to her a $WPC$ of 1.
+So with Clinton included in the node-link diagram in Figure~\ref{fig:funsenatevis}, the senate looks much more collaborative than it does without her.
+We then map our potential covariates to visual features: number of bills authored  maps to the size of the node, gender maps to the shape, and party to the color.
+In addition, we visualize the strength of the tie by mapping the $WPC$ value between the two senators to the linewidth of the edge.
 In this single visualization, we have viewed node information, edge information, and time.
-We will use \texttt{geomnet} many ways to view the model in the data space.
 
@@
            directed = T, curvature = .2, labelon=T, size = 5,
-           arrow = arrow(angle = 20, type = "open", length = unit(.2, "cm")), ecolour = t1$color,
+           arrow = arrow(angle = 20, type = "open", length = unit(.2, "cm")), ecolour = "black",
            color = "#252525", hjust = .5, vjust = .5, labelcolour = 'white', fontsize = 4, arrowgap = .02,
            singletons = T) +
-  labs(title = "Average: M1") + ThemeNet +
-  theme(plot.background = element_rect(color = 'black'))
+  labs(title = "Average: M1") + ThemeNet
 
@@
 
-For network data, we view an ``average" network through a sort of summary network drawn using the traditional node-link diagram.
+%For network data, we view an ``average" network through a sort of summary network drawn using the traditional node-link diagram. 
 In Figure~\ref{fig:summnetM1}, we show an \textit{average network} created with 1,000 simulations of the second wave of the network from Model 1.
 To make this average network, we first simulated 1,000 wave 2 and wave 3 observations of our small friendship example data from model M1, for which parameters had previously been estimated.
-We then combine the 1,000 simulated instances of wave 2, and count up the number of times each edge appears in a network simulation.
-Then, we combine these 1,000 networks into a single network with edgeweight equal to the proportion of time that edge appears in our 1,000 simulations.
-This weighted edgelist is the network we draw in Figure~\ref{fig:summnetM1}.
-An edge is only drawn in the average network if it appears in at least 51 of the 1000 simulations, with edges that appear with greater frequency emphasized by thicker linewidths and a darker color.
+We then aggregate the 1,000 simulated instances of wave 2, and by counting up the number of times each edge $i \rightarrow j$ appears in a network simulation.
+Then, we construct a single network for each time point with edge weights equal to the proportion of time the edge appears in our 1,000 simulations.
+This weighted network is shown in the middle panel of Figure~\ref{fig:summnetM1}, with edges present only in the average network if it appears in at least 51 of the 1,000 simulations.
+The edges that appear most frequently in the model are drawn with darker, thicker lines.
 On either side of the average network in Figure~\ref{fig:summnetM1}, we show the actual data, wave 1 on the left, and wave 2 on the right.
@@
 
-<<summary2, fig.height=2.5, fig.cap="Network differences between the average network from M1 and the first and second data waves.
-The green edges represent edges in the average network that are not in the data, while red edges are in the data but not in the average network.
-Black edges are in both the data and the average, and the edgeweight represents the rate at which that edge shows up in simulations.">>=
+<<summary2, fig.height=2.5, fig.cap="Network differences between the average network from M1 and Wave 1 and Wave 2 of the data.
+Some edges appear in both the data and the average Wave 2 from M1, while others only appear in the data or only in the average.
+Width of edges in the average network represent the proportion of appearances in the 1,000 simulations.
+The FR layout is used.">>=
+
+red <- "#e41a1c"
+green <- "#4daf4a"
+
 am_t1 <- as.matrix(igraph::as_adj(igraph::graph_from_data_frame(arrange(t1, parse_number(from), parse_number(to)))))[-17,-17]
@@
          var = recode(value, `0` = "Neither", `2` = "InAvgNotW1",  `1`= "InBoth"),
-         color2 = recode(value, `0`="white", `1`= "black", `2` = "green")) %>%
+         color2 = recode(value, `0`="white", `1`= "black", `2` = green)) %>%
   filter(value !=0) %>% 
@@
            arrow = arrow(angle = 20, type = "open", length = unit(.2, "cm")), ecolour = t1_1$color2,
-           color = "#252525", hjust = .5, vjust = .5, labelcolour = 'white', fontsize = 4, arrowgap = .02,
+           color = "#252525", hjust = .5, vjust = .5, labelcolour = 'white', fontsize = 4, arrowgap = .02, layout.alg = "fruchtermanreingold",
            singletons = T) +
-  labs(title = "Difference between\nWave 1 and Average from M1") + ThemeNet +
+  labs(title = "Difference between Wave 1 &\nWave 2 average from M1") + ThemeNet +
   theme(plot.background = element_rect(color = 'black'))
+
+pforleg <- ggplot(data = mtcars) + 
+  geom_line(aes(x = mpg, drat, group = as.factor(cyl), color = as.factor(cyl)), size = 1.5) + 
+  scale_color_manual(values = c("black", red, green), name = "Tie is...", labels = c("in data,\nin average", "in data,\nnot in average", "not in data,\nin average")) + 
+  theme_grey() + 
+  theme(legend.key = element_rect(color = "black", fill = "white"), 
+        legend.key.height = unit(25, "pt"))
+#pforleg
 
@@
          var = recode(value, `0` = "Neither", `-1` = "InAvgNotW2",  `1`= "InBoth", `2` = "InW2notAvg"),
-         color2 = recode(value, `0`="white", `1`= "black", `2` = "red", `-1` = "green")) %>% 
+         color2 = recode(value, `0`="white", `1`= "black", `2` = red, `-1` = green)) %>% 
   filter(value !=0) %>% 
@@
            arrow = arrow(angle = 20, type = "open", length = unit(.2, "cm")), ecolour = t1_2$color2,
-           color = "#252525", hjust = .5, vjust = .5, labelcolour = 'white', fontsize = 4, arrowgap = .02,
+           color = "#252525", hjust = .5, vjust = .5, labelcolour = 'white', fontsize = 4, arrowgap = .02, layout.alg = "fruchtermanreingold",
            singletons = T) +
-  labs(title = "Difference between\nWave 2 and Average from M1") + ThemeNet +
+  labs(title = "Difference between Wave 2 &\nWave 2 average from M1") + ThemeNet +
   theme(plot.background = element_rect(color = 'black'))
-grid.arrange(p1, p2, nrow = 1)
+legendsumm <- cowplot::get_legend(pforleg)
+ggdraw() + draw_plot(p1, width = .4) +  draw_plot(p2, width = .4, x=.4) + 
+  draw_grob(legendsumm,x=.8, y=.25, width = .2, height = .4)
 @
@@
 In Figure~\ref{fig:summary2}, we show \textit{difference networks} between the average network from M1 and the first and second waves of data.
+These networks are constructed by computing the difference between two adjacency matrices.
 The lack of model fit is very obvious here, with the green edges representing ties that only show up in the average network, and the red edges representing ties in the data that the model misses completely.
@@
 On the left in Figure~\ref{fig:summary2}, we see the difference between the first wave of data and the average network.
-We only see green and black edges, meaning that the simulated networks consist primarily of edges in wave 1, plus some new edges that aren't in the first wave of data that are reciprocating the ties that are in the data.
+We only see green and black edges, meaning that the simulated networks consist primarily of edges in wave 1, plus some new edges that are not in the first wave of data.
+The additional green edges are all reciprocating black edges, meaning that the model predicts more reciprocity in ties than there actually is.
 On the right in Figure~\ref{fig:summary2}, we see the difference between the second wave of data and the average network.
-Remember, the model is simulating the second wave of data, which is how the average network is created.
-There are six red edges, which are edges in wave 2 that do not appear at all in the average network.
-In addition, there are several green edges that are in the average network, but not in wave 2.
-Finally, there are a few black edges that appear in both.
-For all edges but the red ones, the linewidth represents the frequency of the edge in the simulations.
+The model is simulating the second wave of data, so we would hope to see few differences between the average and wave2.
+However, there are six red edges, which are edges in wave 2 that do not appear at all in the average network, and there are several green edges that are in the average network, but not in the data.
+Although there are some edges that appear in both the average and wave 2, the majority of edges in the average are not in wave 2, and there are several missing edges.
+Thus, the model M1 is an incredibly poor fit, which we are able to see very easily by viewing the model in the data space in this way.
 
@@
 The second method of model visualization we use is visualizing collections of models.
-There are many possible ways to collect SAO models together, so %.
+There are many possible ways to collect SAOMs together, so %.
 We could look at the same model fit to different data, different models fit to the same data, or we could fit the same models to the same data many times to see how the simulations change.
 F
-for the SAOMs, we decided that there were four ways to create collections of models that would give us the most insight:
-\begin{enumerate}
-\item exploring the space of all possible models;
-\item varying model settings;
-\item fitting the same model form to different data;
-\item fitting the same model to the same data many timess
-\end{enumerate}
+we chose four ways to create collections of models that would give us the most insight: (1) exploring the space of all possible models, (2) fitting the same model to the same data many times, (3) varying model settings, and (4) fitting the same model form to different data.
 We chose these four collections because they each explore something different about SAOMs.
-The first considers the many parameters to include in a SAOM.
-The second collection shows how those many parameters affect the parameter values of the fitted models.
-The third looks at how the same model behaves when fit to different data.
-Finally, the results from fitting the same model to the same data may lead to different parameter estimates, so we want to see how the parameter estimates can vary.
+The first considers the many parameters to include in a SAOM, while the second results in different parameter estimates, so we can see how parameter estimates vary.
+The third collection shows how those many possible parameters affect the values of parameters in the fitted models.
+Finally, we explore how the same model behaves under two completely different circumstances.
 
@@
 
-The \texttt{RSiena} manual contains over eighty effects to include in a SAOM.
+The \texttt{RSiena} manual contains over eighty effects to include in a SAOM's objective function.
 In order to select parameters to include in the models for our example data, we searched through the possible effects available to model given the data structure to find \textit{significant} effects.
@@
 We start with the outdegree and reciprocity measures as the foundation of the models we fit, then add one effect, fit the model, test the additional effect for significance, and repeat for all possible parameters to add to the model.
-We performed this procedure for both the small friendship and the senate collaboration data.
+We performed this procedure, which we used to pick M2 and M3, for both the small friendship and the senate collaboration data.
 A visualization of the significant effects for the two example data sets are shown in Figure~\ref{fig:sigtesting} in the Appendix.
@@
 
+\subsection{Fitting the same model to the same data}\label{sec:samemodsamedat}
+
+<<distests, fig.cap="Distribution of fitted parameter values for our three SAOMs.
+The inclusion of $\\beta_3$ or $\\beta_4$ clearly has an effect on the distribution of the rate parameters, $\\alpha_1$ and $\\alpha_2$, and on the other objective function parameters $\\beta_1, \\beta_2$.", fig.height=3>>=
+
+simu2 <- read.csv("data/simulation-1000-M1-M2-M3.csv")
+
+ggplot(data = simu2) +
+  geom_density(aes(x = estimate, fill = Model), alpha = .5) +
+  facet_wrap(~parameter, scales = 'free') +
+  ThemeNoNet +
+  theme(legend.position = 'bottom')
+@
+
+After fitting models M1, M2, and M3 to the friendship network 1,000 times each, we examine the distribution of the fitted values, which are shown in Figure~\ref{fig:distests}.
+We can see from these distributions that the inclusion of the JTT parameter, $\beta_3$ is obviously affecting the distributions of the other four parameters included in all models, $\alpha_1$, $\alpha_2$, $\beta_1$, and $\beta_2$, while these parameters are less affected by the inclusion of $\beta_4$.
+When $\beta_3$ is included, its estimate is postitive, meaning that friendships between two girls with different drinking behaviors tend to form when there is an intermediary who is already friends with the two girls.
+The inclusion of this parameter leads to increases in the rate parameters' estimates, suggesting that the presence of the transitive triplet behavior means that the girls would also change friends more frequently.
+It is not clear, however, that the addition of a parameter to the objective function \textit{should} effect the estimation of the rate parameters.
+The strong correlations between the estimates of the rate parameters and of the objective function parameters are shown in Figure~\ref{fig:corplots} in the Appendix.
+In addition, the outegree parameter, $\beta_1$ decreases when $\beta_3$ is included, while the reciprocity parameter, $\beta_2$ increases.
+This implies the girls in the data prefer to form small, closely knit friend groups, as indicated by reciprocated ties and jumping transitive triplet formation, as opposed to being popular and having many friends.
+Thus, having many friends who do not reciprocate is more stronlgy discouraged by M2 than in the other models.
+%In comparison with models M1 and M3, model M2 typically has higher estimates of the rate parameter, meaning that the inclusion of the covariate statistic in the model leads to higher estimates of the number of times, on average, a node gets to change its ties.
+
 \subsection{Varying model settings}
 
-We have varied model settings already by choosing models M1, M2, M3 to fit to our small friendship network data set.
-In Section~\ref{sec:samemodsamedat}, we fit these models to the data 1,000 times, and in this section, we explore simulations from these three models given the mean values of from the 1,000 fitted parameter values as the parameters in our models.
-From each of these three models using the means of parameter estimates as our fixed parameter values, we simulated 1,000 observations from each of the three models.
-From these simulations, we create visualizations that represents an average network using the same method described in Section~\ref{sec:minds}, shown in Figure~\ref{fig:averagenet} in the Appendix.
+The next method of viewing collections of models is with varying model settings.
+We have varied model settings already by choosing three different models M1, M2, M3 to fit to our example data sets.
+In Section~\ref{sec:samemodsamedat}, we fit these three models to the data 1,000 times, and examined the parameter densities.
+Here, we simulate from these three models.
+For simulation, we used the mean parameter values from the 1,000 fits of the corresponding model.
+We simulated 1,000 observations from each of the three models in this way.
+From these simulations, we created visualizations that represent an average network from models M1, M2, and M3 using the same method described in Section~\ref{sec:minds}.
+These averages are shown in Figure~\ref{fig:averagenet} in the Appendix.
+We see that all models are failing to capture the structure of the second wave.
 
@@
 Actors in these networks want reciprocated relationships: the students want to have close mutual friendships, and senators want mutual support for their bills.
-We explore the relationship between $\beta_1$ and $\beta_2$ further in Section~\ref{sec:samemodsamedat}.
+%We explore the relationship between $\beta_1$ and $\beta_2$ further in Section~\ref{sec:samemodsamedat}.
 
@@
 Because fitting a SAOM requires MCMC methods, we also fit the same model many times, resulting in distributions of fitted parameters for one data set.
-
-\subsection{Fitting the same model to the same data}\label{sec:samemodsamedat}
-
-<<distests, fig.cap="Distribution of fitted parameter values for our three SAOMs.
-The inclusion of $\\beta_3$ or $\\beta_4$ clearly has an effect on the distribution of the rate parameters, $\\alpha_1$ and $\\alpha_2$.", fig.height=3>>=
-
-simu2 <- read.csv("data/simulation-1000-M1-M2-M3.csv")
-
-ggplot(data = simu2) +
-  geom_density(aes(x = estimate, fill = Model), alpha = .5) +
-  facet_wrap(~parameter, scales = 'free') +
-  ThemeNoNet +
-  theme(legend.position = 'bottom')
-@
-
-After fitting models M1, M2, and M3 to the friendship network 1,000 times each, we examine the distribution of the fitted values, which are shown in Figure~\ref{fig:distests}.
-We can see from these distributions that the inclusion of the jumping transitive triplet parameter, $\beta_3$ is obviously affecting the distributions of the other four parameters included in all models, $\alpha_1$, $\alpha_2$, $\beta_1$, and $\beta_2$.
-When $\beta_3$ is included, its estimate is postitive, meaning that friendships between two girls with different drinking behaviors tend to form when there is an intermediary who is already friends with the two girls.
-The inclusion of this parameter leads to increases in the rate parameters' estimates, suggesting that the presence of the transitive triplet behavior means that the girls would also change friends more frequently.
-It is not clear, however, that the addition of a parameter to the objective function \textit{should} effect the estimation of the rate parameters.
-The strong correlations between the estimates of the rate parameters and of the objective function parameters are shown in Figure~\ref{fig:corplots} in the Appendix.
-In addition, the outegree parameter, $\beta_1$ decreases when $\beta_3$ is included, while the reciprocity parameter, $\beta_2$ increases.
-This implies the girls in the data prefer to form small, closely knit friend groups, as indicated by reciprocated ties and jumping transitive triplet formation, as opposed to being popular and having many friends.
-Having many friends who do not reciprocate is more stronlgy discouraged by M2 than in the other models.
-%In comparison with models M1 and M3, model M2 typically has higher estimates of the rate parameter, meaning that the inclusion of the covariate statistic in the model leads to higher estimates of the number of times, on average, a node gets to change its ties.
 
@@
 
-The last principle of model visualization we use is to explore the process of fitting the model, instead of just focusing on the end result.
+The last principle of model visualization we use is exploring the process of fitting the model, instead of just focusing on the end result.
 This principle is perhaps the most important for SAOMs because the model fitting process in \texttt{RSiena} involves several simulation steps that are hidden from the user.
-Hiding the MCMC steps is practical and efficient if a researcher is primarily interested in fitting one model to a set of longitudinal network data, obtaining parameter estimates, and drawing conclusions or making predictions.
+Hiding the MCMC steps is practical and efficient if the goal is to fit a model to a set of longitudinal network data, obtain parameter estimates, and draw conclusions or make predictions.
 We are more interested in \textit{how} the models are fit, so we extracted and explored the different steps of that process instead of allowing them stay hidden.
@@
 One of the hidden steps in the SIENA method of moments algorithm is the ``microstep" process.
-A series of microsteps is obtained by simulating from the model in its current state, $x(t_m)$ with current parameter values $\mathbf{\theta}_{0} = \{\alpha_{1_0}, \dots \alpha_{{m-1}_0}, \beta_{1_0}, \dots, \beta_{K_0}\}$, to the next state, $x(t_{m+1})$.
+A series of microsteps is obtained by simulating from the model in its current state, $x(t_m)$ with current parameter values $\mathbf{\theta}_{0} = \{\alpha_{1_0}, \dots \alpha_{{(m-1)}_0}, \beta_{1_0}, \dots, \beta_{K_0}\}$, to the next state, $x(t_{m+1})$.
 One tie changes at a time until the simulated network has achieved the same number of differences, $C$, from $x(t_m)$ as $x(t_{m+1})$, where $C$ is definined in Equation~\ref{eq:cdiff}.
@@
 At each microstep, an ``ego node" is selected to make a change, and the chosen ego node randomly makes one change in its ties according to the probabilities, $\{p_{ij}: i\neq j \in \{1, \dots, n\}\}$ defined in Equation~\ref{eq:pij}, determined by its objective function.
-The options for change are (1) removing a current tie, (2) adding a new tie, or (3) making no change at all.
-Between two network observations $x(t_m)$ and $x(t_{m+1})$, there can be dozens, hundreds, or even thousands of microsteps, depending on the size of the network and the number of changes between two network observations.
+The options for change are removing a current tie, adding a new tie, or making no change at all.
+Between two network observations $x(t_m)$ and $x(t_{m+1})$, there can be dozens, hundreds, or even thousands of microsteps, depending on the size of the network and the number of tie changes between two network observations.
 We want to simulate and view these in-between steps in order to better understand the behavior of the underlying continuous-time Markov chain.
@@
 <<makinggif, fig.cap='A selection of images in the microstep animation.
-The selected edges and nodes are emphasized by changing the size and the color, then when edges are removed, they fade out, shrinking in size, while the nodes change color and shrink to blend in with the other nodes.', message = FALSE, warning = FALSE, fig.height=2.5>>=
+The selected edges and nodes are emphasized by changing the size and the color, then when edges are removed, they fade out, shrinking in size, while the nodes change color and shrink to blend in with the other nodes.
+Frames 15-28 show a an edge disappearing, while frames 71-84 show an edge appearing.', message = FALSE, warning = FALSE, fig.height=2.5>>=
 # Need old version of dplyr to get this to work:
@@
 The top row of Figure~\ref{fig:makinggif} shows an edge being removed, and the bottom row shows one being added.
-In both cases, the ego node that is acting changes color from black to red, and they also increase greatly in size.
-In the case of an edge being removed, in the top row, the edge that currently exists is emphasized with the same color and size change that the node gets, and as the animation proceeds the edge shrinks to nothing, as the ego node shrinks and changes color back to its original black.
-If an edge is being added, as in the bottom row of the figure, the ego node's appearance changes in the same way as when the edge is being removed, while the edge appears colored red from nothing, and grows to a large size, then changes color and size to match the rest of the edges, while the node shrinks and changes color to match the other nodes.
-
-In the network animation, we see the possible steps of the unobserved CTMC process that is underlying the SAOM fit to the data.
-We see each part of the model come into play.
-First, we see the rate at which the nodes are selected to change.
-Then, we see the result of the actor maximizing its objective function by either deleting or adding a node.
+In both cases, the ego node is emphasized with color and size changes.
+When an edge is removed, the edge that currently exists is emphasized with the same color and size change as the node.
+In the animation, the edge shrinks and disappears as the ego node shrinks and changes color back to its original black.
+If an edge is being added, the red edge appears from nothing, grows large for emphasis, then changes color and size to match the rest of the edges.
+
+In the network animation, we see the modeled steps of the unobserved CTMC process between two obsered networks.
+We see each part of the SAOM come into play: we first see the rate at which the nodes are selected to change, then we see the result of the actor maximizing its objective function by either deleting or adding a node.
 In addition, the layout of the nodes changes as edges are removed or added, which gives us a better sense of how the overall network structure changes with these individual tie changes.
 
-We also use animation to view the changing structure of the adjacency matrix the microsteps.
+We continue to use animation to view the changing structure of the network via adjacency matrix visualizations.
 Since there are 16 nodes in the data, numbered 1-16, we first use that order on the $x$ and $y$ axes for the matrix visualization.
-Viewing the adjacency matrices with this arbitrary ordering does not provide much information to the viewer about the underlying structure of the network.
-This lack of perceived structure would be exacerbated in an animation, so we adjust the ordering so that the viewer can better perceive the structure of the network.
+Viewing the adjacency matrices with this arbitrary ordering does not provide much information to the viewer about the underlying structure of the network, as shown in Figure~\ref{fig:adjmatviz}.
+This lack of perceived structure would contribute to further confusion as the network chages, so we adjust the ordering so that the viewer can better perceive the structure of the network.
 This process is known as matrix seriation \citep{seriation}.
@@
   scale_x_discrete(position = "top") +
-  #scale_y_discrete(name = "from",labels = rev(str_replace(order1, "V", ""))) +
+  scale_y_discrete(name = "from",labels = rev(str_replace(order1, "V", ""))) +
   ThemeNoNet + theme(legend.position = "none") + 
@@
 A single entry in the cumulative adjacency matrix, $\mathbf{A}^{cum}_{ij}$, is the total number of times the edge $i \rightarrow j$ appears in the network from the initial observation, $x(t_m) \equiv X(0)$ to the final result of the last microstep, $X(R)$, where $R$ is the total number of microsteps taken:
-$$ \mathbf{A}^{cum}_{ij} = \sum_{r = 0}^R X_{ij}(r). $$
+\begin{equation}\label{eq:acum}
+\mathbf{A}^{cum}_{ij} = \sum_{r = 0}^R X_{ij}(r). 
+\end{equation}
 We then performed a principal component analysis (PCA) on $\mathbf{A}^{cum}$, and used the values of the first principal component to order the vertices on the $x$ and $y$ axes for the adjacency matrix animation.
-For one such series of microsteps simulated by \texttt{RSiena}, we present the adjacency matrix ordered by the (arbitrary) vertex id alongside the seriated adjacency matrix using the first principal component loading on the cumulative adjacency matrix, $\mathbf{A}^{cum}$, in Figure~\ref{fig:adjmatorder}.
-
+In Figure~\ref{fog:adjmatorder}, we present the first wave of data ordered by the arbitrary node ID order alongside the seriated adjacency matrix ordered by the first principal component loading on the cumulative adjacency matrix, $\mathbf{A}^{cum}$.
 Using PCA on $\mathbf{A}^{cum}$ to order the rows and columns of the adjacency matrix visualization clearly shows the two distinct connected components in the first wave of the network, which are difficult to find in the arbitrarily ordered visualization.
-We also use the PCA seriated layout to fix the layout in the animation of one of the microstep process simulations.
-This animation, some frames of which are shown in Figure~\ref{fig:adjmatani}, is very simple: a square appears or disappears in the animation as that edge appears or disappears in the microstep process.
+
+We also use the PCA seriated layout to fix the layout in the animation of one of the microstep process simulations, some frames of which are shown in Figure~\ref{fig:adjmatani}.
+This animation is very simple: a square appears or disappears in the animation as that edge appears or disappears in the microstep process.
 Through this animation, which can be viewed at \url{https://vimeo.com/240092677}, we can see edges appearing, and then later on disappearing.
@@
 <<adjmatani, fig.cap='A selection of frames from the adjacency matrix visualization animation for one series of microsteps.
-(Ego and alter node labels are removed to declutter the graph.) At the beginning of the animation, shown in the top row, there are two clearly connected components: one in the bottom left corner, and one in the top right corner.
+(Ego and alter node labels are removed to declutter the figure.) Starting at frame 0, there are two clearly connected components: one in the bottom left corner, and one in the top right corner.
 By the end, the component in the bottom left has spread out and become sparsely connected, while the top right component has shrunk, but remains closely connected.', warning = FALSE, message = FALSE, results='hide', fig.height=2.5>>=
@@
 #   geom_hline(yintercept = 1, color = 'red', linetype = 'dashed') +
-#   labs(x = "alter", y = expression(p[ij]), title = paste("Ego node:", names(cprobslist)[i])) + # ylab(expression(p[ij])) +
+#   labs(x = "alter", y = (p[ij]), title = paste("Ego node:", names(cprobslist)[i])) + # ylab((p[ij])) +
 #   theme_gray() +
@@
 
-<<heatmap, fig.align='center', fig.cap='A heatmap showing the empirical transition probabilities for the first microstep in 1,000 simulations.
-The acting ego node is on the y-axis, and the alter node is on the x-axis.
+<<heatmap, fig.align='center', fig.cap='An adjacency matrix visualization showing the empirical transition probabilities for the first microstep in 1,000 simulations.
+The ego node is on the y-axis, and the alter node is on the x-axis.
 There were many ties with empirical probability zero.', fig.height=2.5>>=
@@
 ggplot(data = probsDat,
-       aes(y = factor(ego, levels = order1),
+       aes(y = factor(ego, levels = rev(order1)),
              x = factor(alter, levels = order1),
@@
 In Figure~\ref{fig:heatmap}, we build on the concept of the ordered adjacency matrix of Figure~\ref{fig:adjmatorder}.
-This heatmap shows the transition probabilities of all ties that are changed in the first microstep of the 1,000 simulations.
-The heatmap is noticeably sparse: %showing the lack of coverage in the model.
-of the $16^2 = 256$ possible steps for the CTMC to take, only 103, or about 40\%, are taken in the 1,000 simulated chains.
+This adjacency matrix visualization shows the transition probabilities of all ties that are changed in the first microstep of the 1,000 simulations.
+The figure is noticeably sparse: %showing the lack of coverage in the model.
+of the $16\cdot 15 = 240$ possible changes for the CTMC to make, only 103, or about 43\%, are taken in the 1,000 simulated chains.
 This leaves %reinforces what we saw in Figure~\ref{fig:histogramalters}, where there are many paths not taken.
 This effect could only be exacerbated as more steps are taken in the CTMC, leading to
-a very large area of our network space, $\mathcal{X}$, unexplored by the first steps in the SAOM model fitting process.
+a very large area of our possible network space unexplored by the first steps in the SAOM model fitting process.
 We present additional evidence for lack of coverage in Figure~\ref{fig:alledgests} the Appendix.
@@
 We have used novel visualization methods in order to better understand the family of models known as stochastic actor-oriented models for social network data.
-By looking at the underlying algorithms, visualizing collections of these models, and viewing the model in the data space, we have been able to gain knowledge and appreciation for these complicated models and everything that goes into them.
+By viewing the model in the data space, visualizing collections of  models, and looking at the underlying algorithms, we have been able to gain knowledge and appreciation for these complicated models and everything that goes into them.
+When we looked at the model in the data space by viewing the average network simulated from a model, we were able to see the lack of model fit.
+When visualizing collections of models, we uncovered concerning relationships between parameters.
+Finally, when we explore the algorithms, we gain a better understanding of the model and its behavior.
 
@@
 The \texttt{RSiena} software is incredibly powerful, and can fit a whole slew of much more flexible stochastic actor-oriented models than we have examined here.
-If a researcher thinks the network structure or an actor covariate effects the rate of change of the network, there is a way to incorporate that belief into the rate function of the SAOM.
-More than one actor-level covariate can be included in the model, and way more than three parameters can be included in the objective function itself.
-In addition, \texttt{RSiena} allows the user to tell it which parameters lead to tie creation, and which parameters lead to tie endowment, or dissolution.
-We have used ``evaluation" parameters, which assume that creation and endowment are equal \citep{RSienamanual}.
+If, for example, a researcher thinks the network structure or an actor covariate effects the rate of change of the network, there is a way to incorporate that belief into the rate function of the SAOM.
+In addition, more than one actor-level covariate can be included in the model, and many more than three parameters can be included in the objective function itself.
+The flexibility of \texttt{RSiena} allows the user to specify which parameters lead to tie creation, and which parameters lead to tie dissolution.
+We have used ``evaluation" parameters, which are agnostic, allowing the data to decide which parameters affect creation or dissolution via the sign \citep{RSienamanual}.
 Finally, SAOMs and \texttt{RSiena} are able to also model behavior change of the actors in the network, which again is a capability we did not explore here.
@@
 
-In Figure~\ref{fig:sigtesting}, we see in both the friendship data and the senate data results that most of the significant effects have absolute value less than ten.
+\subsection*{Section 4.1}
+
+In Figure~\ref{fig:sigtesting}, we see for the friendship data and the senate data that most of the significant effects have absolute value less than ten.
 In addition, the $p$-values for the effects from the friendship data are more spread out than the $p$-values for the senate data, which are concentrated at about 0.02 or less.
 This may suggest that larger data sets tend to result generally in smaller $p$-values, just like a larger sample size results in smaller $p$-values in a $t$-test.
+Since the tests are Wald-type tests, this makes sense.
+
+\subsection*{Section 4.2}
+
+<<averagenet, results='hide', out.width='\\textwidth',fig.cap='The node-link diagrams from the three "average" networks that we calculated are in the top row, and the true wave 1 and wave 2 data are shown in the bottom row above.
+There is some difference between the three models, but overall, these three models cannot capture the structure in the true second wave of data.', fig.height=5>>=
+
+combinedavgs$linewidth[combinedavgs$cat %in% c("1st Wave", "2nd Wave")] <- .5
+combinedavgs$cat <- as.factor(combinedavgs$cat)
+combinedavgs$cat <- ordered(combinedavgs$cat, levels = c("Model1", "Model2", "Model3", "1st Wave", "2nd Wave"))
+
+ggplot(data = combinedavgs) +
+  geom_net(aes(from_id = from, to_id = to, linewidth= linewidth),
+           directed = T, curvature = .2, labelon=T, size = 5,
+           arrow = arrow(angle = 20, type = "open", length = unit(.2, "cm")), ecolour = "#969696", fiteach = T,
+           color = "#252525", hjust = .5, vjust = .5, labelcolour = 'white', fontsize = 4, arrowgap = .02,
+           singletons = T) +
+  ThemeNet + theme(panel.background = element_rect(color = 'black')) +
+  facet_wrap(~cat)
+@
+
+To create the average network visualization shown in Figure~\ref{fig:averagenet}, we follow the same procedure as in Section~\ref{sec:minds}, counting occurrences of each possible edge in the simulations, resulting in a summary network with weighted edges representing the number of times an edge appeared in the simulated wave 2 when simulating from the SAOM 1,000 times.
+As in Figure~\ref{fig:summnetM1}, edges only appear in the average networks if they appear more than 5\% of the time in the simulations.
+In Figure~\ref{fig:averagenet}, we show the ``average'' network from the three models we fit and the first and second waves of data.
+Comparing the three averages to waves 1 and 2, we see that they have very similar structure to wave 1.
+Model 2, which included the transitive triplet parameter, seems to have created a larger connected component overall than models 1 and 3.
+In particular, if we look at the group of nodes $\{10,11,14\}$, we see they are very strongly connected within the three average networks, and they are completely separate from the other nodes in the true wave 2.
+None of the three average networks show node 16 gaining ties as it does in wave two, nor do they show nodes 4 and 7 becoming isolated.
+In Model 2, however, the ties to node 7 appear much weaker than in Model 1 or Model 2, suggesting that of the three, Model 2 may be the best fit for our data.
+
+\subsection*{Section 4.3}
 
@@
 
-<<averagenet, results='hide', out.width='\\textwidth',fig.cap='The node-link diagrams from the three "average" networks that we calculated are in the top row, and the true wave 1 and wave 2 data are shown in the bottom row above.
-There is some difference between the three models, but overall, these three models cannot capture the structure in the true second wave of data.', fig.height=5>>=
-
-combinedavgs$linewidth[combinedavgs$cat %in% c("1st Wave", "2nd Wave")] <- .5
-combinedavgs$cat <- as.factor(combinedavgs$cat)
-combinedavgs$cat <- ordered(combinedavgs$cat, levels = c("Model1", "Model2", "Model3", "1st Wave", "2nd Wave"))
-
-ggplot(data = combinedavgs) +
-  geom_net(aes(from_id = from, to_id = to, linewidth= linewidth),
-           directed = T, curvature = .2, labelon=T, size = 5,
-           arrow = arrow(angle = 20, type = "open", length = unit(.2, "cm")), ecolour = "#969696", fiteach = T,
-           color = "#252525", hjust = .5, vjust = .5, labelcolour = 'white', fontsize = 4, arrowgap = .02,
-           singletons = T) +
-  ThemeNet + theme(panel.background = element_rect(color = 'black')) +
-  facet_wrap(~cat)
-@
-
-To create the average network visualization shown in Figure~\ref{fig:averagenet}, we follow the same procedure as in Section~\ref{sec:minds}, counting occurrences of each possible edge in the simulations, resulting in a summary network with weighted edges representing the number of times an edge appeared in the simulated wave 2 when simulating from the SAOM 1,000 times.
-As in Figure~\ref{fig:summnetM1}, edges only appear in the average networks if they appear more than 5\% of the time in the simulations.
-In Figure~\ref{fig:averagenet}, we show the ``average'' network from the three models we fit and the first and second waves of data.
-Comparing the three averages to waves 1 and 2, we see that they have very similar structure to wave 1.
-Model 2, which included the transitive triplet parameter, seems to have created a larger connected component overall than models 1 and 3.
-In particular, if we look at the group of nodes $\{10,11,14\}$, we see they are very strongly connected within the three average networks, and they are completely separate from the other nodes in the true wave 2.
-None of the three average networks show node 16 gaining ties as it does in wave two, nor do they show nodes 4 and 7 becoming isolated.
-In Model 2, however, the ties to node 7 appear much weaker than in Model 1 or Model 2, suggesting that of the three, Model 2 may be the best fit for our data.
+\subsection*{Section 4.4}
 
@@
 
+\subsection*{Section 5}
+
 Finally, we combine 1,000 simulations from model M1 into a visualization that displays the entire microstep process for all simulations in Figure ~\ref{fig:alledgests}.
@@
 <<alledgests, fig.height = 4, fig.cap="Visualizing all microsteps taken in 1,000 simulations from the model M1.
-The occurrence percent is split up into groups to correspond with its distribution: only about 10\\% of the edges appear more than 10\\% of the time in the 1,000 simulations, while about 60\\% appear less than 1\\% of the time.
+The occurrence percent is split up into groups to correspond with its distribution: only about 10\\% of the possible edges appear more than 10\\% of the time in the 1,000 simulations, while about 60\\% appear less than 1\\% of the time.
 The first wave network is shown at microstep 0, and the second wave of the network is shown as the last microstep for comparison.

### 2018-08-15 PaperToDo.Rmd
lots of done things

@@
 - [X] Add another example of a data set for M-in-DS (Sect 4.1) (Congress)
-- [ ] Make node-link gif available online
-- [ ] Insert link to full node-link gif in paper
-- [ ] Make adjmat gif available online
-- [ ] Insert link to full adjmat gif in paper
+- [X] Make node-link gif available online
+- [X] Insert link to full node-link gif in paper
+- [X] Make adjmat gif available online
+- [X] Insert link to full adjmat gif in paper
 - [ ] Beef up transition before Sect 4.4

### 2018-08-15 paper-child.Rnw
fix a fig ref

@@
   ThemeNet +
-  theme(legend.position = "none") + 
-  labs(title = expression(paste(s[i3], "(x,z)", "=",Sigma[j != h], x[ij], x[ih], x[hj] %.% 
+  theme(legend.position = "none", plot.title = element_text(size = rel(0.9))) + 
+  labs(title = expression(paste(s[i3], "(x,z)", "=",Sigma[j != h], x[ij], x[ih], x[hj] %*% 
                                 I, "(", z[i], "=", z[h]!= z[j], ")")
@@
 We then performed a principal component analysis (PCA) on $\mathbf{A}^{cum}$, and used the values of the first principal component to order the vertices on the $x$ and $y$ axes for the adjacency matrix animation.
-In Figure~\ref{fog:adjmatorder}, we present the first wave of data ordered by the arbitrary node ID order alongside the seriated adjacency matrix ordered by the first principal component loading on the cumulative adjacency matrix, $\mathbf{A}^{cum}$.
+In Figure~\ref{fig:adjmatorder}, we present the first wave of data ordered by the arbitrary node ID order alongside the seriated adjacency matrix ordered by the first principal component loading on the cumulative adjacency matrix, $\mathbf{A}^{cum}$.
 Using PCA on $\mathbf{A}^{cum}$ to order the rows and columns of the adjacency matrix visualization clearly shows the two distinct connected components in the first wave of the network, which are difficult to find in the arbitrarily ordered visualization.

### 2018-08-17 paper-child.Rnw

@@
 
-\par One set of models for dynamic social networks, are stochastic actor-oriented models (SAOMs), introduced by \citet{saompaper}.
-These models are different from other dynamic network models because they allow us to add node level information into the model with the traditional network structure information.
-Adding node-level information into the model is a more intuitive approach to model changes in a social network because, for example we expect people with common interests to be more likely to form relationships.
-SAOMs allow us to incorporate this information on shared interests in the modeling process.
+\par One set of models for dynamic social networks, are stochastic actor-oriented models (SAOMs), introduced by \citet{saompaper}. \hh{
+These models allow us to add node-level covariates into the model by incorporating structural information on a single node, such as in-degree or out-degree, and information derived from pairs of nodes by comparing values of covariates.
+For example,
+} 
+%These models are different from other dynamic network models because they allow us to add node level information into the model with the traditional network structure information.
+Adding node-level information into the model is a more intuitive approach to model changes in a social network because, for example
+we expect people with common interests to be more likely to form relationships. SAOMs allow us to incorporate this information on shared interests in the modeling process.
 
@@
 They are relatively new, especially compared to the classic ERGMs, and they are not very tractable analytically.
-Likelihood functions quickly become very complex due to the dependency structure inherent in the data, and so computational methods, such as Markov chain Monte Carlo (MCMC) are used to fit models.
+Likelihood functions quickly become very complex due to the dependency structure inherent in the data, and so computational methods, such as Markov chain Monte Carlo (MCMC), are used to fit models.
 SAOMs are typically fit to dynamic social network data using a series of MCMC phases for finding method of moments (MoM) estimates of model parameters.
@@
 This software is a considerable contribution to the field of social network analysis, but the parameter estimation process is largely hidden from the user.
-In order to better understand the model-fitting process of \texttt{RSiena}, we visualize the underlying methods and structures, and utimately aim to help researchers working with SAOMs better understand the structure of and results from these models.
+In order to better understand the model-fitting process of \texttt{RSiena}, we visualize the underlying methods and structures, and ultimately aim to help researchers working with SAOMs better understand the structure of and results from these models.
 

### 2018-08-23 paper-child.Rnw
re-write of section 5 complete.

@@
 
-The last principle of model visualization we use is exploring the process of fitting the model, instead of just focusing on the end result.
-This principle is perhaps the most important for SAOMs because the model fitting process in \texttt{RSiena} involves several simulation steps that are hidden from the user.
-Hiding the MCMC steps is practical and efficient if the goal is to fit a model to a set of longitudinal network data, obtain parameter estimates, and draw conclusions or make predictions.
-We are more interested in \textit{how} the models are fit, so we extracted and explored the different steps of that process instead of allowing them stay hidden.
-
-One of the hidden steps in the SIENA method of moments algorithm is the ``microstep" process.
-A series of microsteps is obtained by simulating from the model in its current state, $x(t_m)$ with current parameter values $\mathbf{\theta}_{0} = \{\alpha_{1_0}, \dots \alpha_{{(m-1)}_0}, \beta_{1_0}, \dots, \beta_{K_0}\}$, to the next state, $x(t_{m+1})$.
+When fitting a SAOM in \texttt{RSiena}, several simulation steps are hidden from the user.
+One of the hidden steps in the SIENA method of moments algorithm is the \textit{microstep} process.
+At each microstep, an \textit{ego} node is selected to make a change, and the chosen ego node randomly makes one change in its ties to maximize its objective function.
+Between two network observations $x(t_m)$ and $x(t_{m+1})$, there can be dozens, hundreds, or even thousands of microsteps, depending on the size of the network and the number of differences between two network observations.
+We want to simulate and view these in-between steps in order to better understand the behavior of the underlying continuous-time Markov chain.
+Hiding the microsteps is practical and efficient if the goal is to fit a model to data, obtain parameter estimates, and draw conclusions or make predictions.
+But in order to understand \textit{how} the models are fit, we need to extract and explore the different steps of the process.
+We use these steps to visually explore the algorithms, so that we are visualizing the process, and not just the main result.
+
+%A series of microsteps is obtained by simulating from the model in its current state, $x(t_m)$ with current parameter values $\mathbf{\theta}_{0} = \{\alpha_{1_0}, \dots \alpha_{{(m-1)}_0}, \beta_{1_0}, \dots, \beta_{K_0}\}$, to the next state, $x(t_{m+1})$.
 One tie changes at a time until the simulated network has achieved the same number of differences, $C$, from $x(t_m)$ as $x(t_{m+1})$, where $C$ is definined in Equation~\ref{eq:cdiff}.
 
-This simulation process follows the steps of the continuous-time Markov chain.
+%This simulation process follows the steps of the continuous-time Markov chain.
 Each tie change in the CTMC is referred to as one ``microstep".
-At each microstep, an ``ego node" is selected to make a change, and the chosen ego node randomly makes one change in its ties according to the probabilities, $\{p_{ij}: i\neq j \in \{1, \dots, n\}\}$ defined in Equation~\ref{eq:pij}, determined by its objective function.
 The options for change are removing a current tie, adding a new tie, or making no change at all.
-Between two network observations $x(t_m)$ and $x(t_{m+1})$, there can be dozens, hundreds, or even thousands of microsteps, depending on the size of the network and the number of tie changes between two network observations.
-We want to simulate and view these in-between steps in order to better understand the behavior of the underlying continuous-time Markov chain.
 
@@
 
-The first vizualization we present here is an animation of the simulated microsteps, when fitting model M1, that form the transition steps of the CTMC from wave 1 to wave 2 of the small friendship network example shown in Figure~\ref{fig:alldataviz}.
+To visualize the underlying algorithm, we animate the microsteps that form the transition steps of the CTMC from wave 1 to wave 2 of the small friendship network example shown in Figure~\ref{fig:alldataviz}.
+The microstep animation, some frames of which are shown in Figure~\ref{fig:makinggif}, represent one possible path from wave 1 to wave 2 according to the model M1.
+We show each change in the network as follows: we first increase the size and change the color of the ego node to focus attention to its ties.
+Then, the changing tie either fades in or fades out.
+If there are no changes at a particular microstep, the microstep does not appear.
+Some frames of the animations are shown in Figure~\ref{fig:makinggif}, and the full movie, created with \texttt{tweenr} and \texttt{gganimate} can be viewed at \url{https://vimeo.com/240089108} \citep{tweenr, gganimate}.
 Movies similar to this animation were used to visualize the changes of dynamic networks in \citet{dynnetviz}.
-When each ego node is selected in a microstep, it is emphasized in the animation, then the associated edge either appears or disappears.
-If there are no changes at a particular microstep, no changes are seen.
-Some frames of the animations are shown in Figure~\ref{fig:makinggif}, and the full movie, created with \texttt{tweenr} and \texttt{gganimate} can be viewed at \url{https://vimeo.com/240089108} \citep{tweenr, gganimate}.
-
-The top row of Figure~\ref{fig:makinggif} shows an edge being removed, and the bottom row shows one being added.
-In both cases, the ego node is emphasized with color and size changes.
-When an edge is removed, the edge that currently exists is emphasized with the same color and size change as the node.
-In the animation, the edge shrinks and disappears as the ego node shrinks and changes color back to its original black.
-If an edge is being added, the red edge appears from nothing, grows large for emphasis, then changes color and size to match the rest of the edges.
-
-In the network animation, we see the modeled steps of the unobserved CTMC process between two obsered networks.
-We see each part of the SAOM come into play: we first see the rate at which the nodes are selected to change, then we see the result of the actor maximizing its objective function by either deleting or adding a node.
-In addition, the layout of the nodes changes as edges are removed or added, which gives us a better sense of how the overall network structure changes with these individual tie changes.
-
-We continue to use animation to view the changing structure of the network via adjacency matrix visualizations.
-Since there are 16 nodes in the data, numbered 1-16, we first use that order on the $x$ and $y$ axes for the matrix visualization.
-Viewing the adjacency matrices with this arbitrary ordering does not provide much information to the viewer about the underlying structure of the network, as shown in Figure~\ref{fig:adjmatviz}.
-This lack of perceived structure would contribute to further confusion as the network chages, so we adjust the ordering so that the viewer can better perceive the structure of the network.
-This process is known as matrix seriation \citep{seriation}.
+Frames 15-28 of Figure~\ref{fig:makinggif} show a deleted tie, while frames 71-84 show an added tie.
+For both changes, the ego node is emphasized with color and size.
+When a tie is removed, it is emphasized with the same color and size as the node.
+As the animation proceeds, the tie shrinks and disappears as the ego node shrinks and changes color back to black.
+If a tie is being added, the it appears red from nothing, grows large for emphasis, then changes color and size to match the rest of the edges.
+In this network animation, we see the modeled steps of the \textit{unobserved} CTMC process between two observed networks.
+We see each part of the SAOM come into play: we first see the selection of nodes to change via the rate function, then we see the result of the actor maximizing its objective function by either deleting or adding a node.
+Finally, the layout of the nodes changes as edges are removed or added, which demonstrates how the overall network structure changes with each of the individual tie changes.
+Combining the tie changes in this animation makes the CTMC process the focus and demonstrates the underlying change mechanism of the dynamic network according to the chosen model.
 
@@
 
-To reorder the vertices for the matrix visualization, we first constructed a cumulative adjaceny matrix, $\mathbf{A}^{cum}$, for the series of microsteps simulating the network from $x(t_m)$ to $x(t_{m+1})$.
+We continue to use animation to view the changing structure of the network via adjacency matrix visualizations.
+In the first adjacency matrix visualization in Figure~\ref{fig:adjmatorder}, we use the node ID to order the rows and columns.
+This arbitrary ordering does not easily convey the underlying structure of the network.
+We use \textit{matrix seriation} to reorder the rows and columns of the adjacency matrix visualization so that the structure of the network is apparent \citep{seriation}.
+%This lack of perceived structure would contribute to further confusion as the network chages, so we adjust the ordering so that the viewer can better perceive the structure of the network.
+This process is known as matrix seriation
+To seriate the matrix, we first constructed a cumulative adjaceny matrix, $\mathbf{A}^{cum}$, for the series of microsteps simulating the network from $x(t_m)$ to $x(t_{m+1})$.
 A single entry in the cumulative adjacency matrix, $\mathbf{A}^{cum}_{ij}$, is the total number of times the edge $i \rightarrow j$ appears in the network from the initial observation, $x(t_m) \equiv X(0)$ to the final result of the last microstep, $X(R)$, where $R$ is the total number of microsteps taken:
@@
 We then performed a principal component analysis (PCA) on $\mathbf{A}^{cum}$, and used the values of the first principal component to order the vertices on the $x$ and $y$ axes for the adjacency matrix animation.
-In Figure~\ref{fig:adjmatorder}, we present the first wave of data ordered by the arbitrary node ID order alongside the seriated adjacency matrix ordered by the first principal component loading on the cumulative adjacency matrix, $\mathbf{A}^{cum}$.
-Using PCA on $\mathbf{A}^{cum}$ to order the rows and columns of the adjacency matrix visualization clearly shows the two distinct connected components in the first wave of the network, which are difficult to find in the arbitrarily ordered visualization.
-
-We also use the PCA seriated layout to fix the layout in the animation of one of the microstep process simulations, some frames of which are shown in Figure~\ref{fig:adjmatani}.
-This animation is very simple: a square appears or disappears in the animation as that edge appears or disappears in the microstep process.
-Through this animation, which can be viewed at \url{https://vimeo.com/240092677}, we can see edges appearing, and then later on disappearing.
-These in-between steps are not shown when we look at the network at our discrete observation points, so by viewing this animation we can gain a better understanding of the underlying dynamics of this model.
+The resulting visualization for the first wave of data is shown in Figure~\ref{fig:adjmatorder}.
+%we present the first wave of data ordered by the arbitrary node ID order alongside the seriated adjacency matrix ordered by the first principal component loading on the cumulative adjacency matrix, $\mathbf{A}^{cum}$.
+Using PCA on $\mathbf{A}^{cum}$ to order the rows and columns of the adjacency matrix visualization clearly shows the two distinct connected components in the first wave of the network, which are difficult to find in the visualization ordered by node ID.
+We continue to use the seriated matrix in another animation of the microstep process, which is avaiable at \url{https://vimeo.com/240092677}.
+A few frames of the video are shown in Figure~\ref{fig:adjmatani}.
+In the animation, a square appears or disappears as that edge appears or disappears in the microstep process.
+Through this animation, we can see edges appearing, and then later on disappearing.
+This visualization also shows the microsteps where the nodes do not make a change.
+From start to finish, we see network evolve according to the model: sometimes ties are removed right after they are created, and sometimes several ties are removed or created in a row.
+The adjacency matrix animation shows the many ways the algorithm explores the data space.
+By viewing this animation, we can gain a better understanding of the dynamics of this model.
 
 <<adjmatani, fig.cap='A selection of frames from the adjacency matrix visualization animation for one series of microsteps.
-(Ego and alter node labels are removed to declutter the figure.) Starting at frame 0, there are two clearly connected components: one in the bottom left corner, and one in the top right corner.
+The node lables are removed to declutter the figure.
+Starting at frame 0, there are two clearly connected components: one in the bottom left corner, and one in the top right corner.
 By the end, the component in the bottom left has spread out and become sparsely connected, while the top right component has shrunk, but remains closely connected.', warning = FALSE, message = FALSE, results='hide', fig.height=2.5>>=
@@
 
-We also attempt to better understand the microstep process by visualizing the empirical transition probabilites for the first microstep in the process.
+<<heatmap, fig.align='center', fig.cap='Adjacency matrix visualizations showing the theoretical (left) and empirical (right) transition probabilities for the first microstep observed in 1,000 simulations.
+The ego node is on the y-axis, and the alter node is on the x-axis.
+There were many ties with empirical probability zero.', fig.height=2.5>>=
+
+ansnullchains %>%
+  filter(period == 1) %>%
+  group_by(rep) %>%
+  mutate(msno = row_number()) %>%
+  filter(msno == 1) %>%
+  mutate(ego = as.numeric(as.character(ego))+1,
+         alter = as.numeric(as.character(alter))+1,
+         prob = exp(logAlterProb)) -> probsDat
+
+ansnullchains %>%
+  filter(period == 1) %>%
+  group_by(rep) %>%
+  mutate(msno = row_number()) %>%
+  filter(msno == 1) %>%
+  ungroup() %>% group_by(ego, alter) %>% 
+  tally() %>% ungroup() %>% mutate(empprob = n/1000) %>% 
+  mutate(ego = as.numeric(as.character(ego))+1,
+         alter = as.numeric(as.character(alter))+1) ->empprobsDat
+
+# probsDat %>% group_by(ego, alter) %>% summarize(count = n()) %>% dim
+# 103, 3
+p1 <- ggplot(data = probsDat,
+       aes(y = factor(ego, levels = rev(order1)),
+             x = factor(alter, levels = order1),
+             fill = prob)) +
+  geom_tile(color = 'black') +
+  scale_fill_gradient(low = 'white', high = 'black', limits = c(0,1)) +
+  labs(y = "ego", x = "alter")+
+  scale_x_discrete(position = 'top') +
+  #scale_y_discrete(labels = rev(str_replace(order1, "V", ""))) +
+  ThemeNoNet + 
+  theme(panel.grid = element_blank()) +
+  coord_fixed() + 
+  ggtitle("Theoretical Probabilities")
+p2 <- ggplot(data = empprobsDat,
+       aes(y = factor(ego, levels = rev(order1)),
+             x = factor(alter, levels = order1),
+             fill = empprob)) +
+  geom_tile(color = 'black') +
+  scale_fill_gradient(low = 'white', high = 'black', name = "prob") +
+  labs(y = "ego", x = "alter")+
+  scale_x_discrete(position = 'top') +
+  #scale_y_discrete(labels = rev(str_replace(order1, "V", ""))) +
+  ThemeNoNet + 
+  theme(panel.grid = element_blank()) +
+  coord_fixed() + 
+  ggtitle("Empirical Probabilities")
+gridExtra::grid.arrange(p1, p2, nrow = 1)
+@
+
+We also attempt to better understand the microstep process by visualizing the theoretical and observed transition probabilites for the first microstep in the process.
 These probabilities are defined in Equation~\ref{eq:pij}.
-We only do the first step of many because the \texttt{RSiena} transition probabilities for any two edges $i,j$ after the first step are only directly comparable for identical microsteps due to the conditioning on the current network state in the model.
+We only do the first step of many because each microstep is dependent on the previous network state.
 %By ego node state, we mean the current set of incoming and outgoing ties to the ego node, $i$.
@@
 It is possible to incorporate this information into our visualizations, but for now we look at only the first step, because we know without needing any complicated conditioning steps that the previous node state $X_i$ is identical in all cases.
-Thus we have 1,000 transition probabilities to examine: one transition probability for the first microstep taken for each of the simulations.
+There are 1,000 transitions to examine: the first microstep taken in each one of the simulations. %Another way to view these transition probabilities is through the adjacency matrix visualization. 
+We use another seriated adjacency matrix visualization to show theoretical and empirical transition probabilities of the first microstep in Figure~\ref{fig:heatmap} for each tie that was the first change in one of the simulations.
+The figure is noticeably sparse: %showing the lack of coverage in the model.
+of the $16\cdot 16 = 256$ possible choices for the CTMC to make, only 103, or about 43\%, are made in the 1,000 simulated chains.
+This leaves %reinforces what we saw in Figure~\ref{fig:histogramalters}, where there are many paths not taken.
+This effect could only be exacerbated as more steps are taken in the CTMC, leading to
+a large area of data space unexplored by the first step in the SAOM model fitting process.
+We present additional evidence for lack of coverage in Figure~\ref{fig:alledgests} the Appendix.
+The Figures~\ref{fig:heatmap} and \ref{fig:alledgests} show that the model M1 and the the corresponding fitting process do not explore the data space enough to adequately capture the network change mechanism, a finding we were only able to discern thanks to model visualization techniques.
 %Put more precisely, let the vector $X_i = (x_{ij}|_{j \neq i}, x_{ji}|_{j \neq i}$ for $i \neq j \neq k \neq \ell \in \{1, \dots, n\}$, $p_{ij}$ is not necessarily the same \st{ARGH NOT SURE THIS IS TRUE } if the current state of the network achieved from the first microstep, denoted $X(1)$, is $X(1) = x(i \leadsto \ell)$ than is different in the case of $x_{ij}$ changing in the first step from the case of $x_{ij}$ not changing in the first step.
@@
 This matches the overall pattern of the data: we see the most ties in the first wave of data, and the least ties in the third wave.
-
-<<heatmap, fig.align='center', fig.cap='An adjacency matrix visualization showing the empirical transition probabilities for the first microstep in 1,000 simulations.
-The ego node is on the y-axis, and the alter node is on the x-axis.
-There were many ties with empirical probability zero.', fig.height=2.5>>=
-
-ansnullchains %>%
-  filter(period == 1) %>%
-  group_by(rep) %>%
-  mutate(msno = row_number()) %>%
-  filter(msno == 1) %>%
-  mutate(ego = as.numeric(as.character(ego))+1,
-         alter = as.numeric(as.character(alter))+1,
-         prob = exp(logAlterProb)) -> probsDat
-
-# probsDat %>% group_by(ego, alter) %>% summarize(count = n()) %>% dim
-# 103, 3
-ggplot(data = probsDat,
-       aes(y = factor(ego, levels = rev(order1)),
-             x = factor(alter, levels = order1),
-             fill = prob)) +
-  geom_tile(color = 'black') +
-  scale_fill_gradient(low = 'white', high = 'black', limits = c(0,1)) +
-  labs(y = "ego", x = "alter")+
-  scale_x_discrete(position = 'top') +
-  #scale_y_discrete(labels = rev(str_replace(order1, "V", ""))) +
-  ThemeNoNet + 
-  theme(panel.grid = element_blank()) +
-  coord_fixed()
-@
-
-%Another way to view these transition probabilities is through the adjacency matrix visualization. 
-In Figure~\ref{fig:heatmap}, we build on the concept of the ordered adjacency matrix of Figure~\ref{fig:adjmatorder}.
-This adjacency matrix visualization shows the transition probabilities of all ties that are changed in the first microstep of the 1,000 simulations.
-The figure is noticeably sparse: %showing the lack of coverage in the model.
-of the $16\cdot 15 = 240$ possible changes for the CTMC to make, only 103, or about 43\%, are taken in the 1,000 simulated chains.
-This leaves %reinforces what we saw in Figure~\ref{fig:histogramalters}, where there are many paths not taken.
-This effect could only be exacerbated as more steps are taken in the CTMC, leading to
-a very large area of our possible network space unexplored by the first steps in the SAOM model fitting process.
-We present additional evidence for lack of coverage in Figure~\ref{fig:alledgests} the Appendix.
-The Figures~\ref{fig:heatmap} and \ref{fig:alledgests} show that the model M1 and the SAOM fitting process do not explore the data space enough to adequately capture the network change mechanism, a finding we were only able to discern thanks to model visualization techniques.
 

### 2018-08-29 paper-child.Rnw
most recent changes

@@
 Another way to view the model in the data space is through simulation from the model.
-No one network simulated from a SAOM is going to look like the data or represent the model, so we might want to look at an ``average network" or ``expected network" value.
-We frequently rely on averages and expected values in data analyses, but network models, especially those as complex as SAOMs, lack a expected network value measure.
+No one network simulated from a SAOM is going to look like the data or represent the model, so we might want to look at an \textit{expectation} from the network model.
+We frequently rely on expected values in data analyses, but network models, especially those as complex as SAOMs, lack an expected network value measure.
 We could talk about expected values of parameters, but the parameters can be hard to interpret.
 %Expected values of parameters can be computed, but if they cannot directly tell us anything about our dependent variable, they lose value.
-There is no defined expected value of an simulation from a network model, though there are with most other model simulations.
+There is no defined expected value of a network model, though there are with most other statistical models.
 How then, can we arrive at an ``average" network?
@@
 %For network data, we view an ``average" network through a sort of summary network drawn using the traditional node-link diagram. 
-In Figure~\ref{fig:summnetM1}, we show an \textit{average network} created with 1,000 simulations of the second wave of the network from Model 1.
-To make this average network, we first simulated 1,000 wave 2 and wave 3 observations of our small friendship example data from model M1, for which parameters had previously been estimated.
+In Figure~\ref{fig:summnetM1}, we show the \textit{expected network} of M1, created with 1,000 simulations of the second wave of the network from the model.
+To make this expected network, we first simulated 1,000 wave 2 and wave 3 observations of our small friendship example data from model M1, for which parameters had previously been estimated.
 We then aggregate the 1,000 simulated instances of wave 2, and by counting up the number of times each edge $i \rightarrow j$ appears in a network simulation.
@@
 The average network can thus be used to help determine model goodness-of-fit.
-Because the the average network looks more like the first wave than the second wave, we can use the visualization in Figure~\ref{fig:summnetM1} as evidence of poor model fit.
-
-<<summary2, fig.height=2.5, fig.cap="Network differences between the average network from M1 and Wave 1 and Wave 2 of the data.
-Some edges appear in both the data and the average Wave 2 from M1, while others only appear in the data or only in the average.
-Width of edges in the average network represent the proportion of appearances in the 1,000 simulations.
+Because the the expected  network looks more like the first wave than the second wave, we can use the visualization in Figure~\ref{fig:summnetM1} as evidence of poor model fit.
+
+<<summary2, fig.height=2.5, fig.cap="Network differences between the expected network from M1 and Wave 1 and Wave 2 of the data.
+Some edges appear in both the data and the expected Wave 2 from M1, while others only appear in the data or only in the exectation.
+Width of edges in the represent the proportion of appearances in the 1,000 simulations.
 The FR layout is used.">>=
 
-red <- "#e41a1c"
-green <- "#4daf4a"
+red <- "#fd8d3c"
+green <- "#41ab5d"
+
+#762a83
+#af8dc3
+#e7d4e8
+#d9f0d3
+#7fbf7b
+#1b7837
 
@@
            singletons = T) +
-  labs(title = "Difference between Wave 1 &\nWave 2 average from M1") + ThemeNet +
+  labs(title = "Difference between Wave 1 &\nWave 2 expectation from M1") + ThemeNet +
   theme(plot.background = element_rect(color = 'black'))
@@
   geom_line(aes(x = mpg, drat, group = as.factor(cyl), color = as.factor(cyl)), size = 1.5) + 
-  scale_color_manual(values = c("black", red, green), name = "Tie is...", labels = c("in data,\nin average", "in data,\nnot in average", "not in data,\nin average")) + 
+  scale_color_manual(values = c("black", red, green), name = "Tie is...", labels = c("in data,\nin expectation", "in data,\nnot in expectation", "not in data,\nin expectation")) + 
   theme_grey() + 
@@
            singletons = T) +
-  labs(title = "Difference between Wave 2 &\nWave 2 average from M1") + ThemeNet +
+  labs(title = "Difference between Wave 2 &\nWave 2 expectation from M1") + ThemeNet +
   theme(plot.background = element_rect(color = 'black'))
@@
 
-<<adjmatorder, message = FALSE, warning = FALSE, fig.cap="On the left, the starting friendship network represented in adjacency matrix form, ordered by vertex id.
+<<adjmatorder, message = FALSE, warning = FALSE, fig.cap="On the left, the starting friendship network represented in adjacency matrix form, ordered by node ID.
 On the right, the same adjacency matrix is presented after ordering the vertices by one repetition of the microstep simulation process from wave one to wave two.", fig.width=8, fig.height=3, fig.align='center'>>=
@@
 In the animation, a square appears or disappears as that edge appears or disappears in the microstep process.
-Through this animation, we can see edges appearing, and then later on disappearing.
+%Through this animation, we can see edges appearing, and then later on disappearing.
 This visualization also shows the microsteps where the nodes do not make a change.
@@
   ThemeNoNet + 
-  theme(panel.grid = element_blank()) +
+  theme(panel.grid = element_blank(), axis.text = element_text(size=rel(.8))) +
   coord_fixed() + 

### 2018-09-04 paper-child.Rnw
committing just in case

@@
 
-The second method of model visualization we use is visualizing collections of models.
-There are many possible ways to collect SAOMs together, so %.
-We could look at the same model fit to different data, different models fit to the same data, or we could fit the same models to the same data many times to see how the simulations change.
-F
+In addition to assessing model fit, we are also interested in model behavior so that we can choose better fitting models in the future.
+We examine model behavior by visualizing collections of models.
+There are many possible ways to collect SAOMs together, so
 we chose four ways to create collections of models that would give us the most insight: (1) exploring the space of all possible models, (2) fitting the same model to the same data many times, (3) varying model settings, and (4) fitting the same model form to different data.
@@
 
-\subsection{Exploring the space of all possible models}\label{sec:allpossible}
-
-The \texttt{RSiena} manual contains over eighty effects to include in a SAOM's objective function.
+%\subsection{Exploring the space of all possible models}\label{sec:allpossible}
+
+\noindent \textbf{Exploring the space of all possible models:} The \texttt{RSiena} manual contains over eighty effects to include in a SAOM's objective function.
 In order to select parameters to include in the models for our example data, we searched through the possible effects available to model given the data structure to find \textit{significant} effects.
@@
 
-\subsection{Fitting the same model to the same data}\label{sec:samemodsamedat}
+%\subsection{Varying model settings}
+
+\noindent \textbf{Varying model settings:} The next method of viewing collections of models is with varying model settings.
+We have varied model settings already by choosing three different models M1, M2, M3 to fit to our example data sets.
+In Section~\ref{sec:samemodsamedat}, we fit these three models to the data 1,000 times, and examined the parameter densities.
+Here, we simulate from these three models.
+For simulation, we used the mean parameter values from the 1,000 fits of the corresponding model.
+We simulated 1,000 observations from each of the three models in this way.
+From these simulations, we created visualizations that represent an average network from models M1, M2, and M3 using the same method described in Section~\ref{sec:minds}.
+These averages are shown in Figure~\ref{fig:averagenet} in the Appendix.
+We see that all models are failing to capture the structure of the second wave.
+
+%\subsection{Fitting the same model to the same data}\label{sec:samemodsamedat}
 
@@
 
-After fitting models M1, M2, and M3 to the friendship network 1,000 times each, we examine the distribution of the fitted values, which are shown in Figure~\ref{fig:distests}.
+\noindent \textbf{Fitting the same model to the same data:} After fitting models M1, M2, and M3 to the friendship network 1,000 times each, we examine the distribution of the fitted values, which are shown in Figure~\ref{fig:distests}.
 We can see from these distributions that the inclusion of the JTT parameter, $\beta_3$ is obviously affecting the distributions of the other four parameters included in all models, $\alpha_1$, $\alpha_2$, $\beta_1$, and $\beta_2$, while these parameters are less affected by the inclusion of $\beta_4$.
@@
 
-\subsection{Varying model settings}
-
-The next method of viewing collections of models is with varying model settings.
-We have varied model settings already by choosing three different models M1, M2, M3 to fit to our example data sets.
-In Section~\ref{sec:samemodsamedat}, we fit these three models to the data 1,000 times, and examined the parameter densities.
-Here, we simulate from these three models.
-For simulation, we used the mean parameter values from the 1,000 fits of the corresponding model.
-We simulated 1,000 observations from each of the three models in this way.
-From these simulations, we created visualizations that represent an average network from models M1, M2, and M3 using the same method described in Section~\ref{sec:minds}.
-These averages are shown in Figure~\ref{fig:averagenet} in the Appendix.
-We see that all models are failing to capture the structure of the second wave.
-
-\subsection{Fitting the same model to different data}\label{sec:samemoddiffdat}
-
-In addition to fitting M1, M2, and M3 to the friendship data 1,000 times, we also fit them to the senate data 100 times.
+%\subsection{Fitting the same model to different data}\label{sec:samemoddiffdat}
+
+\noindent \textbf{Fitting the same model to different data:} In addition to fitting M1, M2, and M3 to the friendship data 1,000 times, we also fit them to the senate data 100 times.
 The means and standard deviations of the parameter estimates for each combination of model and data are given in Table~\ref{tab:meanests}, while the density plots of each parameter in the model for each data set is given in Figure~\ref{fig:compareModels}.

### 2018-09-12 paper-child.Rnw
forgot commit

@@
 
-In addition to assessing model fit, we are also interested in model behavior so that we can choose better fitting models in the future.
-We examine model behavior by visualizing collections of models.
-There are many possible ways to collect SAOMs together, so
-we chose four ways to create collections of models that would give us the most insight: (1) exploring the space of all possible models, (2) fitting the same model to the same data many times, (3) varying model settings, and (4) fitting the same model form to different data.
-We chose these four collections because they each explore something different about SAOMs.
-The first considers the many parameters to include in a SAOM, while the second results in different parameter estimates, so we can see how parameter estimates vary.
+In addition to assessing model fit, we also want to learn about model behavior so that we can choose better fitting models in the future.
+We examine the behavior of SAOMs by visualizing collections of models.
+We build collections of models in four ways.
+First, we explore the space of all possible models to determine the most significant model parameters.
+Second, we vary model settings to uncover relationships between parameters and their effects on the fitted model.
+Next, we fit the same model to the same data many times to assess the behavior of the underlying simulations.
+Finally, we fit the same model to different data to learn about fitted parameter values under different conditions.
+We chose these four collections because they each explore something different about SAOM behavior: the choice of parameters in the objective function, the effects of underlying simulations, and the resulting fitted values under different circumstances.
+Using the tools in our visualization toolbox, we %The first considers the many parameters to include in a SAOM, while the second results in different parameter estimates, so we can see how parameter estimates vary.
 The third collection shows how those many possible parameters affect the values of parameters in the fitted models.

### 2018-09-12 paper-child.Rnw
just do it.

@@
 
-In addition to assessing model fit, we also want to learn about model behavior so that we can choose better fitting models in the future.
+Now that we have assessed model fit, we next want to learn about model behavior in order to choose better fitting models in the future.
 We examine the behavior of SAOMs by visualizing collections of models.
@@
 
-\noindent \textbf{Exploring the space of all possible models:} The \texttt{RSiena} manual contains over eighty effects to include in a SAOM's objective function.
-In order to select parameters to include in the models for our example data, we searched through the possible effects available to model given the data structure to find \textit{significant} effects.
-We tested for significance using the Wald-type tests built into \texttt{RSiena} for one-at-a-time effects testing.
-We start with the outdegree and reciprocity measures as the foundation of the models we fit, then add one effect, fit the model, test the additional effect for significance, and repeat for all possible parameters to add to the model.
+\noindent \textbf{Exploring the space of all possible models:} The \texttt{RSiena} manual contains over eighty effects to include in the SAOM objective function.
+To select objective function parameters for fitting models to our example data, we searched through all possible effects available to find \textit{significant} effects.
+The \texttt{RSiena} package includes the function \texttt{Wald.Rsiena()} for Wald-type tests of parameter significance.
+We start by including the outdegree and reciprocity parameters in the model then add one effect to the objective function, fit the model to the example data, test the added effect for significance, and repeat for all possible model parameters.
 We performed this procedure, which we used to pick M2 and M3, for both the small friendship and the senate collaboration data.
 A visualization of the significant effects for the two example data sets are shown in Figure~\ref{fig:sigtesting} in the Appendix.
+Using this one-at-a-time testing method results in many possible significant parameters of different type and magnitude, demonstrating the flexibility of these models.
 %We see in both the friendship data and the senate data results that most of the significant effects have absolute value less than ten.
@@
 
-\noindent \textbf{Varying model settings:} The next method of viewing collections of models is with varying model settings.
-We have varied model settings already by choosing three different models M1, M2, M3 to fit to our example data sets.
-In Section~\ref{sec:samemodsamedat}, we fit these three models to the data 1,000 times, and examined the parameter densities.
-Here, we simulate from these three models.
-For simulation, we used the mean parameter values from the 1,000 fits of the corresponding model.
-We simulated 1,000 observations from each of the three models in this way.
-From these simulations, we created visualizations that represent an average network from models M1, M2, and M3 using the same method described in Section~\ref{sec:minds}.
+\noindent \textbf{Varying model settings:} We now view collections of models by varying model settings.
+We have done this already by choosing three different models M1, M2, M3 to fit to our example data sets.
+To explore the behavior of these three models, we simulate from them.
+Using the parameter values given in Table~\ref{tab:meanests} for the friendship data, we simulated 1,000 observations from each of the three models and created expected network visualizations from models M1, M2, and M3 using the same method as in Section~\ref{sec:minds}.
 These averages are shown in Figure~\ref{fig:averagenet} in the Appendix.
 We see that all models are failing to capture the structure of the second wave.
+%In Section~\ref{sec:samemodsamedat}, we fit these three models to the data 1,000 times, and examine the parameter densities.
+Here, .
+For simulation, we used the mean parameter values from the 1,000 fits of the corresponding model.
 

### 2018-09-18 paper-child.Rnw
slowly :tired_face:

@@
 
-\noindent \textbf{Varying model settings:} We now view collections of models by varying model settings.
+\noindent \textbf{Varying model settings:} The second method for viewing collections of models is varying model settings.
 We have done this already by choosing three different models M1, M2, M3 to fit to our example data sets.
 To explore the behavior of these three models, we simulate from them.
-Using the parameter values given in Table~\ref{tab:meanests} for the friendship data, we simulated 1,000 observations from each of the three models and created expected network visualizations from models M1, M2, and M3 using the same method as in Section~\ref{sec:minds}.
-These averages are shown in Figure~\ref{fig:averagenet} in the Appendix.
-We see that all models are failing to capture the structure of the second wave.
+Using the parameter values given in Table~\ref{tab:meanests} for the friendship data, we simulated 1,000 observations from each of the three models.
+From the simulations, we created an expected network for wave 2 from models M1, M2, and M3 using the same method described in Section~\ref{sec:minds}.
+These expected networks are shown in Figure~\ref{fig:averagenet} in the Appendix.
+We see that all models are failing to capture the structure of the second wave, even though the individual parameters were all significant.
 %In Section~\ref{sec:samemodsamedat}, we fit these three models to the data 1,000 times, and examine the parameter densities.
@@
 
-\noindent \textbf{Fitting the same model to the same data:} After fitting models M1, M2, and M3 to the friendship network 1,000 times each, we examine the distribution of the fitted values, which are shown in Figure~\ref{fig:distests}.
-We can see from these distributions that the inclusion of the JTT parameter, $\beta_3$ is obviously affecting the distributions of the other four parameters included in all models, $\alpha_1$, $\alpha_2$, $\beta_1$, and $\beta_2$, while these parameters are less affected by the inclusion of $\beta_4$.
+\noindent \textbf{Fitting the same model to the same data:} The simulations performed to fit a SAOM lead to inherent variability in the fitted model.
+To examine this variability, we fit models M1, M2, and M3 to the friendship network 1,000 times each.
+The distributions of the fitted model parameters are shown in Figure~\ref{fig:distests}.
+Clearly, the inclusion of the JTT parameter, $\beta_3$, results in very different estimates the of the other four parameters in all models, $\alpha_1$, $\alpha_2$, $\beta_1$, and $\beta_2$.
+On the other hand, the inclusion of the indirect tie parameter, $\beta_4$, does not lead to different estimates of other model parameters.
 When $\beta_3$ is included, its estimate is postitive, meaning that friendships between two girls with different drinking behaviors tend to form when there is an intermediary who is already friends with the two girls.
-The inclusion of this parameter leads to increases in the rate parameters' estimates, suggesting that the presence of the transitive triplet behavior means that the girls would also change friends more frequently.
-It is not clear, however, that the addition of a parameter to the objective function \textit{should} effect the estimation of the rate parameters.
-The strong correlations between the estimates of the rate parameters and of the objective function parameters are shown in Figure~\ref{fig:corplots} in the Appendix.
-In addition, the outegree parameter, $\beta_1$ decreases when $\beta_3$ is included, while the reciprocity parameter, $\beta_2$ increases.
-This implies the girls in the data prefer to form small, closely knit friend groups, as indicated by reciprocated ties and jumping transitive triplet formation, as opposed to being popular and having many friends.
-Thus, having many friends who do not reciprocate is more stronlgy discouraged by M2 than in the other models.
+The inclusion of this parameter leads to increases in the rate parameters' estimates.
+Thus, when modeling the transitive triplet behavior in this way, we conclude that girls change friends more frequently.
+It is not clear, however, that the addition of a parameter to the objective function \textit{should} effect the estimates of the rate parameters.
+The strong relationship between rate parameter estimates and objective function parameters is further shown in Figure~\ref{fig:corplots} in the Appendix.
+After visualizing the results of the same models fit many times to the same data, we were able to see potential model behavior problems right away.
 %In comparison with models M1 and M3, model M2 typically has higher estimates of the rate parameter, meaning that the inclusion of the covariate statistic in the model leads to higher estimates of the number of times, on average, a node gets to change its ties.
@@
 %\subsection{Fitting the same model to different data}\label{sec:samemoddiffdat}
-
-\noindent \textbf{Fitting the same model to different data:} In addition to fitting M1, M2, and M3 to the friendship data 1,000 times, we also fit them to the senate data 100 times.
-The means and standard deviations of the parameter estimates for each combination of model and data are given in Table~\ref{tab:meanests}, while the density plots of each parameter in the model for each data set is given in Figure~\ref{fig:compareModels}.
 
@@
 
-Looking at Table~\ref{tab:meanests} and Figure~\ref{fig:compareModels} in the Appendix, we see a few patterns in the estimates from both models.
-First, the table and the density plots demonstrate the same relationship between the outdegree parameter, $\beta_1$, and the reciprocity parameter, $\beta_2$.
+\noindent \textbf{Fitting the same model to different data:} The final procedure for viewing collections of models is to fit the same model to different data to understand how the model behaves in different contexts.
+As we did for the friendship data, we estimated the parameters for M1, M2, and M3, when fit to the senate data 100 times each.
+The means and standard deviations of the parameter estimates for all six model-data combinations are given in Table~\ref{tab:meanests}, while corresponding density plots of objective function parameters are given in Figure~\ref{fig:compareModels} in the Appendix.
+Looking at the estimates for both data sets, we can see a few patterns emerging.
+First, we can see the same relationship between the outdegree parameter, $\beta_1$, and the reciprocity parameter, $\beta_2$.
 In both data sets and across all three models, the estimates of $\beta_1$ are all negative and hover between -5 and -3, while the estimates of $\beta_2$ are all positive and hover between four and five.
-This suggests that in both data sets, nodes are \textit{encouraged} to form outgoing ties that are reciprocated and \textit{discouraged} from forming ties that are not.
-Actors in these networks want reciprocated relationships: the students want to have close mutual friendships, and senators want mutual support for their bills.
+%In addition, the outegree parameter, $\beta_1$, \textit{decreases} when $\beta_3$ is included, while the reciprocity parameter, $\beta_2$ increases.
+This inverse relationship implies that the girls prefer to form small, closely knit friend groups, as opposed to being popular and having many friends.
+Tight-knit friend groups are more encouraged by model M2 than by models M1 and M3.
+This suggests that in both data sets, nodes prefer outgoing ties that are reciprocated to those that are not.
+When we Consider the context, this suddenly seems obvious: the students want to have close mutual friendships, and senators want mutual support for their bills.
+We are also concerned, however, with the extremely high correlation between the estimates of $\beta_1, \beta_2$.
+In all model-data combinations, the absolute correlation between the estimates is above 0.93.
 %We explore the relationship between $\beta_1$ and $\beta_2$ further in Section~\ref{sec:samemodsamedat}.
-
-The inclusion of $\beta_3$ for the friendship data had a noticeable effect on the other parameters in the model.
-This is not true in the senate data.
-%The covariate used in the senate data for the JTT calculation was the number of bills authored by the ego node in the given year.
-The number of bills authored by a senator varies wildly, from no bills to 114 bills authored in two years, so this effect could simply be nonsensical for the senate data, since senators are less likely to have the same number of bills authors than teenage girls are to have the same drinking behavior.
-Looking at the estimates of $\beta_4$, we see that the estimates for the senate data are nearer to zero, suggesting that this effect, which considers indirect ties, is not important for the senate data.
-This effect is, however, very strong in the friendship data with a mean of about -7.5, suggesting that indirect ties are discouraged from forming: teenage girls want to have direct friendships.
-This indirect relationship variable does not help explain the senate collaboration structure, but it does help clarify the teenage friendship structure.
+% We have seen how the same model forms behave in different context, with some  differences and some similarities. 
+XXX NEED CONCLUDING SENTENCES XXX
+
+We saw earlier that the inclusion of $\beta_3$ for the friendship data has a noticeable effect on the other parameters in the model.
+We see in Figure~\ref{fig:compareModels} that this effect is not present in the senate data.
+Furthermore, the estimates of $\beta_4$ for the senate data are near zero, suggesting that this effect, which considers indirect ties, is not informative for the senate data.
+But in the friendship data, the mean estimate of $\beta_4$ is about -7.5, which suggests that indirect ties are strongly discouraged from forming, probably because teenage girls want to have direct friendships with each other.
+Senators, on the other hand, have to wheel and deal to pass their legislation, and indirect relationships may be beneficial at times.
+%Although this indirect relationship variable does not help explain the senate collaboration structure, it does help clarify the teenage friendship structure.
 
@@
 <<corplotsSenate, eval = FALSE>>=
+senM1est <- read.csv("data/congress/senateM1ests.csv")
+senM2est <- read.csv("data/congress/senateM2ests.csv")
+senM3est <- read.csv("data/congress/senateM3ests.csv")
 senM4est <- read.csv("data/congress/senateM4ests.csv")
 senM5est <- read.csv("data/congress/senateM5ests.csv")
+names(senM1est) <- c("alpha1", "alpha2", "alpha3", "beta1", "beta2")
+names(senM2est) <- c("alpha1", "alpha2", "alpha3", "beta1", "beta2", "beta3")
+names(senM3est) <- c("alpha1", "alpha2", "alpha3", "beta1", "beta2", "beta4")
 names(senM4est) <- c("alpha1", "alpha2", "alpha3", "beta1", "beta2", "beta5")
 names(senM5est) <- c("alpha1", "alpha2", "alpha3", "beta1", "beta2", "beta6")
+senM1est$Model <- "M1"
+senM2est$Model <- "M2"
+senM3est$Model <- "M3"
 senM4est$Model <- "M4"
 senM5est$Model <- "M5"
+senM1est$Sim <- 1:nrow(senM1est)
+senM2est$Sim <- 1:nrow(senM2est)
+senM3est$Sim <- 1:nrow(senM3est)
 senM4est$Sim <- 1:nrow(senM4est)
 senM5est$Sim <- 1:nrow(senM5est)
+senM1est2 <- senM1est %>% gather(parameter, estimate, alpha1:beta2)
+senM2est2 <- senM2est %>% gather(parameter, estimate, alpha1:beta3)
+senM3est2 <- senM3est %>% gather(parameter, estimate, alpha1:beta4)
 senM4est2 <- senM4est %>% gather(parameter, estimate, alpha1:beta5)
@@
 senM45est <- rbind(senM4est2, senM5est2)
+senM123est <- rbind(senM1est2, senM2est2, senM3est2)
+senM123est %>% spread(parameter, estimate)-> senM123est2
+
 senM45est %>% spread(parameter, estimate)-> senM45est2
-ggpairs(senM45est2, columns = 3:9, ggplot2::aes(colour=Model, alpha = .5))
+ggpairs(senM123est2, columns = 3:9, ggplot2::aes(colour=Model, alpha = .5))
 @
@@
 
-#simu2 <- read.csv("data/simulation-1000-M1-M2-M3.csv")
+simu2 <- read.csv("data/simulation-1000-M1-M2-M3.csv")
 simu_spread <- simu2 %>% spread(parameter, estimate)

### 2018-09-26 paper-child.Rnw
done with sec 4 edits, onto sec 3 edits. sec 2,1,A,6 still to go.

@@
 
-Every good data analysis includes both numerical and visual summaries of the data, so why restrict model description and diagnostics to numerical summaries?
-The concept of \emph{model visualization} was developed to complement traditional model diagnostic tools such as $R^2$ values and residual plots \citep{modelvis}.
-In a modeling task, there are three different things called ``model" under consideration: the model \emph{family}, the model \emph{form}, and the \emph{fitted} model.
+%Every good \emph{data} analysis includes both numerical and visual summaries of the data, so why restrict \emph{model} analysis, description, and diagnostics to numerical summaries? 
+\emph{Model visualization} complements traditional model diagnostic tools such as $R^2$ values and residual plots by enriching our understanding and interpretation of statistical models.
+Each of the three different meanings of the word ``model", the model \emph{family}, the model \emph{form}, and the \emph{fitted} model \citep[see][pg.
+2]{modelvis}, can benefit from model visualization (model-vis).
+We apply the three principles of model-vis: view the model in the data space, visualize collections of models, and explore the process of fitting the model, to the models and data introduced in Section~\ref{sec:exdatamods}.
+Since we have already decided on our model family (SAOMs), we shift our focus to the fitted model and the model form.
+Specifically, we use model visualization to learn more about how the \emph{model form} affects the \emph{fitted model}.
+%In a modeling task, there are three different things called ``model" under consideration: .
 The latter is usally what first comes to mind when considering a model, where a specified model is fit, and parameter estimates are reported.% In the context of SAOMs, the \emph{fitted model} contains the form of the rate and objective functions, the estimated rate parameters, and the estimated objective function parameters.
-The \textit{model form} describes the the model \textit{before} the fitting process, defining which parameters are in the model within the context of the larger model family.
+%The \textit{model form} describes the the model \textit{before} the fitting process, defining which parameters are in the model within the context of the larger model family.
 % In SAOMs, the \emph{model form} includes description of the rate and objective functions and the variables therein that describe how the network evolves over time.
-Finally, the \textit{model family} is the broadest description of the model.
+%Finally, the \textit{model family} is the broadest description of the model.
 This is the type of model that you wish to fit to the data, and is chosen based on the problem, data, and knowledge at hand.
 %For example, we chose to use a SAO \emph{model family} to model our network data over an exponential random graph model (ERGM) because we believe that actor-level variables effect network structure and formation, and we wanted to model the network changes over time.
-
-The model family, the model form, and the fitted model can each be visualized according to the three principles of model visualization: we can view the model in the data space, visualize collections of models, and explore the process of fitting the model, not just the end result.
-Since we have already decided on our model family, SAOMs, we shift our focus to the fitted model and the model form.
-Specifically, we want to learn more about how the \emph{model form} we choose affects the \emph{fitted model} by using our example data sets and our visualization toolbox.
-%We begin by introducing the five models that we fit to our example data.
-We use the data and models described in Section~\ref{sec:exdatamods} to guide our visual explorations of SAOMs under the three principles of model visualization.
+%The model family, the model form, and the fitted model can each be visualized according to  %We begin by introducing the five models that we fit to our example data. 
+%We use the data and models described in Section~\ref{sec:exdatamods} to guide our visual explorations of SAOMs under the three principles of model visualization.
 %We begin with novel tools and ideas to view a SAOM in the data space of a dynamic network.
@@
 Now that we have assessed model fit, we next want to learn about model behavior in order to choose better fitting models in the future.
+This is the second principle of model visualization: "collections [of models] are more informative" \citep{modelvis}.
 We examine the behavior of SAOMs by visualizing collections of models.
@@
 
-\noindent \textbf{Fitting the same model to different data:} The final procedure for viewing collections of models is to fit the same model to different data to understand how the model behaves in different contexts.
-As we did for the friendship data, we estimated the parameters for M1, M2, and M3, when fit to the senate data 100 times each.
-The means and standard deviations of the parameter estimates for all six model-data combinations are given in Table~\ref{tab:meanests}, while corresponding density plots of objective function parameters are given in Figure~\ref{fig:compareModels} in the Appendix.
-Looking at the estimates for both data sets, we can see a few patterns emerging.
-First, we can see the same relationship between the outdegree parameter, $\beta_1$, and the reciprocity parameter, $\beta_2$.
-In both data sets and across all three models, the estimates of $\beta_1$ are all negative and hover between -5 and -3, while the estimates of $\beta_2$ are all positive and hover between four and five.
+\noindent \textbf{Fitting the same model to different data:} Finally, we view collections of models by fitting the same models to different data to understand how the model behaves in different contexts.
+As we did for the friendship data, we estimated the parameters for M1, M2, and M3, when fit to the senate data (see Figure \ref{fig:senateplot}) 100 times each.
+The means and standard deviations of the repeated parameter estimates for all six model-data combinations are given in Table~\ref{tab:meanests}, while corresponding density plots of the objective function parameters are given in Figure~\ref{fig:compareModels} in the Appendix.
+Looking at the estimates for both data sets, a few patterns emerge.
+First, we can see the same relationship between the outdegree parameter, $\beta_1$, and the reciprocity parameter, $\beta_2$ in both data sets and across all three models.
+The estimates of $\beta_1$ are all negative and range from about -5 and -3, while the estimates of $\beta_2$ are all positive and range from about four to five.
 %In addition, the outegree parameter, $\beta_1$, \textit{decreases} when $\beta_3$ is included, while the reciprocity parameter, $\beta_2$ increases.
@@
 This suggests that in both data sets, nodes prefer outgoing ties that are reciprocated to those that are not.
-When we Consider the context, this suddenly seems obvious: the students want to have close mutual friendships, and senators want mutual support for their bills.
+When we consider the context, this seems obvious: the students want to have close mutual friendships, and senators want mutual support for their bills.
 We are also concerned, however, with the extremely high correlation between the estimates of $\beta_1, \beta_2$.
-In all model-data combinations, the absolute correlation between the estimates is above 0.93.
+In all six model-data combinations, the absolute correlation between $\hat{\beta}_1, \hat{\beta}_2$ is greater than 0.93.
 %We explore the relationship between $\beta_1$ and $\beta_2$ further in Section~\ref{sec:samemodsamedat}.
 % We have seen how the same model forms behave in different context, with some  differences and some similarities. 
-XXX NEED CONCLUDING SENTENCES XXX
-
-We saw earlier that the inclusion of $\beta_3$ for the friendship data has a noticeable effect on the other parameters in the model.
-We see in Figure~\ref{fig:compareModels} that this effect is not present in the senate data.
+After seeing the high correlation between parameter estimates in very different data contexts, we question the guideline that both parameters need to be included in the objective function.
+
+In Figure~\ref{fig:distests}, we saw that the inclusion of $\beta_3$ for the friendship data has a noticeable effect on the other parameters in the model.
+In Figure~\ref{fig:compareModels}, however, we see that this effect is not present in the senate data.
 Furthermore, the estimates of $\beta_4$ for the senate data are near zero, suggesting that this effect, which considers indirect ties, is not informative for the senate data.
-But in the friendship data, the mean estimate of $\beta_4$ is about -7.5, which suggests that indirect ties are strongly discouraged from forming, probably because teenage girls want to have direct friendships with each other.
+In contrast for the friendship data, the mean estimate of $\beta_4$ is about -7.5, which suggests that indirect ties are \textit{strongly} discouraged from forming.
+Again, this is fairly intuitive given the context: teenage girls likely want to have direct friendships with each other.
 Senators, on the other hand, have to wheel and deal to pass their legislation, and indirect relationships may be beneficial at times.
+By examining the results of models fitted to different data, we better understand the interpretation of the objective function parameters, and we have learned of some potential parameter correlation problems affecting the fit of the models.
+All in all, we have used four different ways of viewing collections of models, which have given us insight into the behavior of SAOMs.
+
 %Although this indirect relationship variable does not help explain the senate collaboration structure, it does help clarify the teenage friendship structure.
@@
 
+The last model visualization principle, "explore the process of model fitting," is crucial for understanding SAOMs \citep{modelvis}.
 When fitting a SAOM in \texttt{RSiena}, several simulation steps are hidden from the user.

### 2018-09-27 paper-child.Rnw
done with section 3 edits. plots still need work tho

@@
 Specifically, we use model visualization to learn more about how the \emph{model form} affects the \emph{fitted model}.
-%In a modeling task, there are three different things called ``model" under consideration: .
-The latter is usally what first comes to mind when considering a model, where a specified model is fit, and parameter estimates are reported.% In the context of SAOMs, the \emph{fitted model} contains the form of the rate and objective functions, the estimated rate parameters, and the estimated objective function parameters.
-%The \textit{model form} describes the the model \textit{before} the fitting process, defining which parameters are in the model within the context of the larger model family.
-% In SAOMs, the \emph{model form} includes description of the rate and objective functions and the variables therein that describe how the network evolves over time.
-%Finally, the \textit{model family} is the broadest description of the model.
-This is the type of model that you wish to fit to the data, and is chosen based on the problem, data, and knowledge at hand.
-%For example, we chose to use a SAO \emph{model family} to model our network data over an exponential random graph model (ERGM) because we believe that actor-level variables effect network structure and formation, and we wanted to model the network changes over time.
-%The model family, the model form, and the fitted model can each be visualized according to  %We begin by introducing the five models that we fit to our example data. 
-%We use the data and models described in Section~\ref{sec:exdatamods} to guide our visual explorations of SAOMs under the three principles of model visualization.
-%We begin with novel tools and ideas to view a SAOM in the data space of a dynamic network.
-We then explore collections of the same models fit many times to the example data to see how the simulation processes in the \texttt{RSiena} ``black-box" affect the model fits.
-Finally, we look behind the scenes and into the individual steps of the continuous time Markov chain to learn more about how this ``hidden data" mechanism works and how it results in a fitted model.
-
-The first way we hope to better understand SAOMs is by viewing the model in the data space.
-In \citeauthor{modelvis}, they chose define the \textit{data space} as ``the region over which we can reliably make inferences, usually a hypercube containing the data" \citep[p.~206]{modelvis}.
+
+First, we put models in the data space to assess the model fit.
+%In \citeauthor{modelvis}, they chose define the \textit{data space} as ``the region over which we can reliably make inferences, usually a hypercube containing the data" \citep[p.~206]{modelvis}.
 But what does this definition mean for dynamic network models?
-We interpret the data space for our application as the combined data structures of the network: the node, edge, and time information.
-%We want to make inferences on parameters in the model, but the multi-level data structure of networks makes the ``region of inference" difficult to define.
-% %For dynamic social networks, there are a few different data ``spaces":
-% \begin{enumerate}
-% \item the actors and their corresponding covariates,
-% \item the edges and their correpsonding variables that describe the relationships between the nodes, and
-% \item the time, both the continuous unobserved time and the discrete observed time points, over which the network evolves
-% \end{enumerate}
-%The traditional node-link visualization uses one of many algorithms to layout the actors as points in 2D space, then draws segments connecting the points in 2D if there is an edge between two nodes, and draws nothing otherwise.
-The time aspect can be visualized by drawing each network observation in time and placing the observed timepoints side-by-side.
-
-%The node, edge, and time data can be visualized together in various ways.% Because longitudinal network data consist of three different ``spaces" of data, viewing the model in the data space can depend on which aspect of the \textit{model} and the \textit{data} we are interested in viewing.
-%The SIENA software includes a method for modelling covariate change in the network
-%Incorporating data covariates into the network structure allows us to assess whether the ties between nodes are affected by how nodes behave over time.
-In this instance, we would want to view predictions of edge presence over time as node variables change, visualizing all three data spaces at once.
-In addition, a SAOM can model node variable change over time, taking both the node and edge information into account.
-In this case, a plot of predicted covariate values over time would put the model into the time and node data space.
-%It is more intutive to view all of the data spaces simultaneously.
-
-The node, edge, and time data can be visualized together in various ways, and one tool that can help is the \texttt{R} package \texttt{geomnet} \citep{geomnet}.
-The network's node or edge data can be mapped to different visual features in the node-link diagram.
+We define the data space for SAOMs as the combined data structures of the network: the node, edge, and time information.
+We use the \texttt{R} package \texttt{geomnet} \citep{geomnet} to visualize the three data spaces together.
+The package lays out the network using node-link diagrams, and the network's node or edge data are then mapped to different visual features.
 The color, size, and shape of the points can be used to represent variables in the node data, while the color, linewidth, and linetype of the lines between points can be used to represent the edge variables.
-%In a social network, node data might be age, gender, and/or occupation of the person in the network, and edge data might be length of connection between two people, how the people first met (school, work, church, etc.), and/or how often they interact.
-To view temporal changes, we can view the network at different timepoints side-by-side to see the evolution.
-Pulling all of this information together with \texttt{geomnet} allows the entire data space to be viewed at once.
+To view temporal changes, we place the network at different timepoints side-by-side to see the evolution.
+By pulling all of this information together using \texttt{geomnet}, we can show the entire data space at once.
 
@@
 
-To demonstrate, we use \texttt{geomnet} to visualize the connections in the $111^{th}$ United States Senate at two different points: when Hillary Clinton was in the senate, and after she left to become Secretary of State.
+To demonstrate the functionality of \texttt{geomnet}, we visualize the connections in the $111^{th}$ United States Senate at two different timepoints: when Hillary Clinton was in the senate, and after she left to become Secretary of State.
 Clinton was only in the $111^{th}$ senate from January 3-20, 2009, in the middle of her second term as the junior senator from New York.
-In that time, she authored two bills and was a cosponsor on 17s.
+In that time, she authored two bills and was a cosponsor on 17.
 There are many senators who cosponsored both of her bills, giving their edges to her a $WPC$ of 1.
@@
 In addition, we visualize the strength of the tie by mapping the $WPC$ value between the two senators to the linewidth of the edge.
-In this single visualization, we have viewed node information, edge information, and time.
+By viewing node information, edge information, and time simultaneously, we have learned that Senator Clinton greatly affected the structure of collaboration within the senate.
 
@@
   geom_net(aes(from_id = from, to_id = to, linewidth= linewidth),
-           directed = T, curvature = .2, labelon=T, size = 5,
+           directed = T, curvature = .2, labelon=T, size = 4,
            arrow = arrow(angle = 20, type = "open", length = unit(.2, "cm")), ecolour = "black",
-           color = "#252525", hjust = .5, vjust = .5, labelcolour = 'white', fontsize = 4, arrowgap = .02,
+           color = "#252525", hjust = .5, vjust = .5, labelcolour = 'white', fontsize = 3, arrowgap = .02,
            singletons = T) +
-  labs(title = "Average: M1") + ThemeNet
+  labs(title = "M1 Expected Network") + ThemeNet + 
+  theme(plot.background = element_rect(color = 'black'))
 
@@
   geom_net(aes(from_id = from, to_id = to, linewidth= linewidth/3),
-           directed = T, curvature = .2, labelon=T, size = 5,
+           directed = T, curvature = .2, labelon=T, size = 4,
            arrow = arrow(angle = 20, type = "open", length = unit(.2, "cm")), ecolour = "#969696",
-           color = "#252525", hjust = .5, vjust = .5, labelcolour = 'white', fontsize = 4, arrowgap = .03,
+           color = "#252525", hjust = .5, vjust = .5, labelcolour = 'white', fontsize = 3, arrowgap = .03,
            singletons = T) +
@@
 
-Another way to view the model in the data space is through simulation from the model.
-No one network simulated from a SAOM is going to look like the data or represent the model, so we might want to look at an \textit{expectation} from the network model.
-We frequently rely on expected values in data analyses, but network models, especially those as complex as SAOMs, lack an expected network value measure.
-We could talk about expected values of parameters, but the parameters can be hard to interpret.
+To view a SAOM in the data space, we simulate from the fitted model.
+No single network instance simulated from a SAOM will fully represent the model, so instead we look at an \textit{expectation} from the network model.
+We frequently rely on expected values in statistics, but network models, especially those as complex as SAOMs, lack an expected network measure.
+We could talk about expected values of parameters, but the parameters can be hard to interpret and cannot teach us much about the overall structure of the network.
+How then, can we arrive at an ``expected" network?
+We answer this question with visualization.
 %Expected values of parameters can be computed, but if they cannot directly tell us anything about our dependent variable, they lose value. 
-There is no defined expected value of a network model, though there are with most other statistical models.
-How then, can we arrive at an ``average" network?
-We answer this question through visualization.
+%There is no defined expected value of a network model, though expectations exist for most other statistical models.
 
 %For network data, we view an ``average" network through a sort of summary network drawn using the traditional node-link diagram. 
-In Figure~\ref{fig:summnetM1}, we show the \textit{expected network} of M1, created with 1,000 simulations of the second wave of the network from the model.
-To make this expected network, we first simulated 1,000 wave 2 and wave 3 observations of our small friendship example data from model M1, for which parameters had previously been estimated.
-We then aggregate the 1,000 simulated instances of wave 2, and by counting up the number of times each edge $i \rightarrow j$ appears in a network simulation.
-Then, we construct a single network for each time point with edge weights equal to the proportion of time the edge appears in our 1,000 simulations.
-This weighted network is shown in the middle panel of Figure~\ref{fig:summnetM1}, with edges present only in the average network if it appears in at least 51 of the 1,000 simulations.
-The edges that appear most frequently in the model are drawn with darker, thicker lines.
-On either side of the average network in Figure~\ref{fig:summnetM1}, we show the actual data, wave 1 on the left, and wave 2 on the right.
-We can see that the structure of the average network is much more similar to the first wave than to the second wave.
-However, the simulations are supposed to represent the second wave of data on the right of Figure~\ref{fig:summnetM1}.
-This is an indication that the simple model, M1, is doing a very poor job of capturing the change mechanism from the first to the second wave of observation.
-The average network can thus be used to help determine model goodness-of-fit.
+%In Figure~\ref{fig:summnetM1}, we show the \textit{expected network} of wave 2 from model M1, created with 1,000 simulations from the model. 
+To construct an expected network, we first simulated 1,000 wave 2 and wave 3 instances of our small friendship network from model M1, for which parameters had previously been estimated.
+We then counted the number of times each edge $i \rightarrow j$ appeared in one of the 1,000 network simulations.
+From these counts, we constructed one summary network for each time point, which has edge weights equal to the proportion of the simulated networks in which the edge appeared.
+Finally, we created the \textit{expected network} by including only the edges that appeared in more than 5\% of the simulations.
+The resulting expected network is in the middle panel of Figure~\ref{fig:summnetM1}.
+Edges with higher weights are darker, thicker lines.
+On either side of the average network in Figure~\ref{fig:summnetM1}, we show the data for comparison: wave 1 on the left, and wave 2 on the right.
+We can see that the structure of the average network is much more similar to wave 1 than to wave 2.
+In both wave 1 and the expected network, the edges form two distinct clusters of nodes: $\{2,3,4,5,6,12,13,15\}$ and $\{7,8,9,10,11,14\}$.
+The expected network, however, is supposed to represent the second wave of data, on the right in Figure~\ref{fig:summnetM1}.
+The expected network does not resemble wave 2, which has one small node cluster and on large node cluster, with four unconnected nodes.
+Because the expected network does not resemble wave 2, we take this as evidence that the simple model, M1, does not capture the change in network structure from the first to the second wave of data.
+Visualizing the expected network in this way places the model M1 in the data space, and has helped us assess model fit.
+% We can use the average network to help determine model goodness-of-fit.
 Because the the expected  network looks more like the first wave than the second wave, we can use the visualization in Figure~\ref{fig:summnetM1} as evidence of poor model fit.
@@
 
-In Figure~\ref{fig:summary2}, we show \textit{difference networks} between the average network from M1 and the first and second waves of data.
-These networks are constructed by computing the difference between two adjacency matrices.
-The lack of model fit is very obvious here, with the green edges representing ties that only show up in the average network, and the red edges representing ties in the data that the model misses completely.
-Again, the edgeweights demonstrate the frequency of the ties in the simulations from M1.
-On the left in Figure~\ref{fig:summary2}, we see the difference between the first wave of data and the average network.
+Another way to view a SAOM in the data space is with \textit{difference networks} between the expected network and the data, as shown in Figure~\ref{fig:summary2}.
+We created these networks by computing the difference between the adjacency matrices for the expected network and wave 1, and for wave 2 and the expected network.
+The lack of model fit becomes even more clear in Figure~\ref{fig:summary2}, with the XX NEED TO GO BACK AND REDO LINETYPES green XX edges representing ties that only show up in the expected network, and the red edges representing ties in the data that the model misses completely.
+Again, the widths of the edges demonstrate the frequency of the ties in the simulations.
+On the left in Figure~\ref{fig:summary2}, we see the difference between the first wave of data and the expected network.
 We only see green and black edges, meaning that the simulated networks consist primarily of edges in wave 1, plus some new edges that are not in the first wave of data.
 The additional green edges are all reciprocating black edges, meaning that the model predicts more reciprocity in ties than there actually is.
-On the right in Figure~\ref{fig:summary2}, we see the difference between the second wave of data and the average network.
-The model is simulating the second wave of data, so we would hope to see few differences between the average and wave2.
-However, there are six red edges, which are edges in wave 2 that do not appear at all in the average network, and there are several green edges that are in the average network, but not in the data.
-Although there are some edges that appear in both the average and wave 2, the majority of edges in the average are not in wave 2, and there are several missing edges.
-Thus, the model M1 is an incredibly poor fit, which we are able to see very easily by viewing the model in the data space in this way.
+On the right in Figure~\ref{fig:summary2}, we see the difference between the second wave of data and the expected network.
+As the model is simulating the second wave of data, we would hope to see few differences between the expected network and wave 2.
+However, there are six red edges, which are edges in the data that do not appear at all in the expected network, and there are several green edges that are in the expected network, but not in the data.
+Although there are some edges that appear in both the expected network and wave 2, the majority of edges in the expected network are not in wave 2, and there are several missing edges.
+We conclude that the model M1 is a poor fit to our example data, which have seen very clearly by viewing the model in the data space.
 

### 2018-09-28 paper-child.Rnw
edit section 2

@@
 The node-link visualization of this data is provided in Figure~\ref{fig:alldataviz}.
-For model fitting, we condition on wave 1 and estimate the parameters of the models for the second and third waves.
+%For model fitting, we condition on wave 1 and estimate the parameters of the models for the second and third waves.
 The actor-level categorical covariate, drinking behavior, has four values: (1) does not drink, (2) drinks once or twice a year, (3) drinks once a month, and (4) drinks once a week.
 In Figure~\ref{fig:alldataviz}, the nodes are colored according to the drinking behavior of that student.
-Over time, we can see that the students tend to drink more and become increasingly isolated into smaller groups.
+We can see that, as time passes, the students seem to drink more and become increasingly isolated into smaller groups.
 An analysis of this type of data with a SAOM should capture these dynamics in a way that allows the researcher to draw conclusions about the nature of the network and behavioral forces at play.
 
-The second data example we use is a collaboration network in the United States Senate during the $111^{th}$ through $114^{th}$ Congresses.
+The second data example we use is a much more recent dataset on collaboration in the United States Senate during the $111^{th}$ through $114^{th}$ Congresses.
 These sessions of congress correspond to the years of Barack Obama's presidency, from 2009-2016.
@@
 Some prominent names are shown to demonstrate their place in the network.
+As with the friendship example, analyzing the senate data with a SAOM should capture the underlying dynamics in a way that gives insight into the impact the behavioral factors have on the senate's collaboration structure.
+By using data sets from very different contexts, we hope to assess possible differences in model performance.
 \\
@@
 \end{equation}
-where $s_{i1}$ is the density network statistic and $s_{i2}$ is the reciprocity network statistic for actor $i$ at the current network state $x$ (see Table~\ref{tab:effects}.
-The second and third models, M2 and M3, contain one additional parameter each in the objective function which were determined by a Wald-type test provided in the \texttt{RSiena} software to be significant, with $p$-values less than 0.05 \citep{RSienaTest}.
+where $s_{i1}$ is the density network statistic and $s_{i2}$ is the reciprocity network statistic for actor $i$ at the current network state $x$ (see Table~\ref{tab:effects}).
+The second and third models, M2 and M3, contain one additional parameter each in the objective function.
+We chose the additional parameters because they were significant ($p$-value $<$ 0.05) according to a Wald-type test provided in the \texttt{RSiena} software \citep{RSienaTest}.
 The M2 model contains an actor-level covariate parameter, and the M3 model contains an additional strutural effect in the objective function.
@@
 These statistics are known as the \emph{number of jumping transitive triplets} (JTT) and the \emph{number of doubly achieved distances two effect} (N22), respectively.
-The first statistic emphasizes triad relationships formed between actors with different covariate values, while $s_{i4}$ emphasizes indirect ties between actors.
-The covariate for the friendship data is the drinking behavior, and is the number of bills authored for the senate data.
-These additional effects are shown in Figure~\ref{fig:jtt} and Figure~\ref{fig:dad}, where the ties affected by the parameter value are represented by dotted lines.
-
-It is ofen difficult to know for certain how to interpret a fitted value of a parameter.
+The JTT effect emphasizes triads formed between actors with different covariate values, while the N22 effect emphasizes indirect ties between actors.
+The covariate for the friendship data is the drinking behavior, and the number of bills authored for the senate data.
+These additional effects are visualized in Figure~\ref{fig:jtt} and Figure~\ref{fig:dad}, where the ties affected by the parameter value are represented by dotted lines.
+If the estimates of these parameters $\hat{\beta}_1, \dots, \hat{\beta}_4$ are positive, the model encourages the tie represented by the dotted line to form, while the opposite is true if the parameter estimate is negative.
+
+%It is ofen difficult to know for certain how to interpret a fitted value of a parameter.
 We can make educated guesses based on the definition of the effect and the sign of the fitted value, but a direct interpretation is not always possible.
 By graphically exploring these models, we aim to understand their effects, their interpretations, and the model fitting process better.
-
-In this paper, we use three different model visualization techniques on SAOMs.
+% XXXX MOVE BELOW PAR TO SECTION 1 XXX
+% In this paper, we use three different model visualization techniques on SAOMs.
 First, Section~\ref{sec:minds} places the models in the data space.
@@
 Although there are some edges that appear in both the expected network and wave 2, the majority of edges in the expected network are not in wave 2, and there are several missing edges.
-We conclude that the model M1 is a poor fit to our example data, which have seen very clearly by viewing the model in the data space.
+We conclude that the model M1 is a poor fit to our example data, which we have seen very clearly by viewing the model in the data space.
 

### 2018-09-28 paper-child.Rnw
some section 1 changes made. first paragraph(s) totally re-written.

@@
 \section{Introduction \& Background}
-
-Social networks, such as collaboration networks between academic researchers or friendship networks have been studied for decades.
-Statistical models such as exponential random graph models (ERGM) and latent space models (LSM) are often used when studying one observation of a network when trying to model its structure \citep{ergm, lsm}.
-Social networks can also be observed at many points in time, and ERGM or LSM may not be appropriate for modelling the changes because they do not model network change as time passes.
-When studying \textit{dynamic networks} observed at many time points, we need a model that allows for variation over time.
-Models for dynamic social networks hold a great deal of potential because of how realistic they can be: social networks evolve over time as edges are formed and dissolved and new nodes join.
-Modeling the underlying mechanisms that create network changes in time is very complex but also provides potential to uncover hidden truths.
-
-\par One set of models for dynamic social networks, are stochastic actor-oriented models (SAOMs), introduced by \citet{saompaper}. \hh{
-These models allow us to add node-level covariates into the model by incorporating structural information on a single node, such as in-degree or out-degree, and information derived from pairs of nodes by comparing values of covariates.
-For example,
-} 
+% peices of old intro
+%Social networks, such as collaboration networks between academic researchers or friendship networks 
+% new intro 
+%As humans and researchers, it is our instint to try to learn as much as we can about the world around us. 
+As members of the human family, one of the hardest things we try to do is understand our relationships with one another, whether as friends, family, or colleagues.
+These relationships, or \emph{social networks}, have been studied for decades, in every format from self-help books to publications in academic journals.
+There are even statistical models, such as exponential random graph models (ERGM) and latent space models (LSM), that strive to capture the underlying relationship-generating mechanism \citep{ergm, lsm}.
+These static models, however, are not optimal for modelling ``real-world" social networks because they do not account for changes in the network as time passes.
+\textit{Dynamic social networks} are more realistic because they contain multiple observations of the same network over time, and models for dynamic social networks hold a great deal of promise because of their potential to capture realistic mechanisms.
+Our social networks evolve over time as new relationships are made, and being able to model these evolutionary mechanisms can uncover hidden truths about our world.
+
+\par One set of models for dynamic social networks, are stochastic actor-oriented models (SAOMs), introduced in \citet{saompaper}.
+These models incorporate node-level covariate effects in addition to the traditional structural effects, such as in-degree or out-degree.
 %These models are different from other dynamic network models because they allow us to add node level information into the model with the traditional network structure information.
 Adding node-level information into the model is a more intuitive approach to model changes in a social network because, for example
-we expect people with common interests to be more likely to form relationships. SAOMs allow us to incorporate this information on shared interests in the modeling process.
-
-\par Unlike other network models, SAOMs are not very well understood.
-They are relatively new, especially compared to the classic ERGMs, and they are not very tractable analytically.
+With these models, we can examine our assumptions that people with common interests are more likely to become friends.
+For example, we may assume that high schoolers who drink heavily are more likely to be friends with others who drink heavily.
+SAOMs allow us to incorporate this behavioral information on shared interests into the modeling process.
+
+\par Unlike other network models, SAOMs are not analytically tractable.
 Likelihood functions quickly become very complex due to the dependency structure inherent in the data, and so computational methods, such as Markov chain Monte Carlo (MCMC), are used to fit models.
-SAOMs are typically fit to dynamic social network data using a series of MCMC phases for finding method of moments (MoM) estimates of model parameters.
-In order to estimate the parameters of SAOMs, we use the software SIENA, and its \texttt{R} implementation \texttt{RSiena} by \citet{RSiena}.
-This software is a considerable contribution to the field of social network analysis, but the parameter estimation process is largely hidden from the user.
-In order to better understand the model-fitting process of \texttt{RSiena}, we visualize the underlying methods and structures, and ultimately aim to help researchers working with SAOMs better understand the structure of and results from these models.
-
-\par To learn more about the SAOM fitting process, we use model visualization techniques (see \citet{modelvis}) to display the model in the data space, view collections of models instead of single models, and visually explore the algorithms fitting the SAOMs.
+%SAOMs are typically fit to dynamic social network data using a series of MCMC phases for finding method of moments (MoM) estimates of model parameters.
+In order to estimate the parameters of a SAOM, we use the software SIENA, and its \texttt{R} implementation \texttt{RSiena} by \citet{RSiena}.
+This software is a considerable contribution to the field of social network analysis, but the extensive parameter estimation process is largely hidden from the user.
+In order to better understand the fitting process in \texttt{RSiena}, we use model visualization, or model-vis \citep{modelvis}, to examine the underlying methods, hidden structures, and simulation results.
+Ultimately, we aim to help researchers working with SAOMs better understand the structure, behavior, fit, and results of these models.
+
+In the remainder of this introduction, we first describe the structure of network data and different network visualization techniques.
+Then, we carefully define the many moving pieces in SAOMs.
+Finally, we briefly ouline the SAOM fitting process to a set of dynamic network data.
+
+%\par To learn more about the SAOM fitting process, we use model visualization techniques (see \citet{modelvis}) to display the model in the data space, view collections of models instead of single models, and visually explore the algorithms fitting the SAOMs.
 SAOMs are a model family that can reap many benefits from the application of model visualization techniques.
@@
 
-We provide a brief introduction to the structure of network data and common network visualization methods. \\
-
-\noindent \textbf{Network Data Structure:} Network data across fields have similar data structures.
+%We provide a brief introduction to the structure of network data and common network visualization methods. \\
+
+%\noindent \textbf{Network Data Structure:} 
+Network data across fields have similar data structures.
 There are always units of observation and connections between those units.
-In a social network, the units of observations, called \textit{nodes} or \textit{actors}, may be people, while the connections, called \textit{edges} or \textit{ties}, are the relationships between people.
-Networks often change over time, like when new relationships are formed in a social network.
-The nodes and edges themselves can also have inherent variables of interest, e.g.\ the age, gender, and political affiliation of people in a social network and the type of relationship (friends, married, etc.).
+In a social network, the units of observations, called \textit{nodes} or \textit{actors}, are people, while the connections, called \textit{edges} or \textit{ties}, are their relationships.
+%Networks change over time, like when new relationships are formed in a social network.
+Other data on a network include inherent variables of interest on the nodes and edges themselves, e.g.\ someone's age, gender, and political affiliation, the type of relationship (friendship, coworkers, etc) that an edge represents, in addition to the changing structure of the network in time.
 \\

### 2018-10-01 paper-child.Rnw
revised abstract, finished revising intro section. Quite pleased with it

@@
 \textit{Dynamic social networks} are more realistic because they contain multiple observations of the same network over time, and models for dynamic social networks hold a great deal of promise because of their potential to capture realistic mechanisms.
-Our social networks evolve over time as new relationships are made, and being able to model these evolutionary mechanisms can uncover hidden truths about our world.
-
-\par One set of models for dynamic social networks, are stochastic actor-oriented models (SAOMs), introduced in \citet{saompaper}.
+Our social networks evolve over time as new relationships are made, and being able to model these evolutionary mechanisms can uncover hidden truths about our relationships with each other.
+
+\par One set of models for dynamic social networks are stochastic actor-oriented models (SAOMs), introduced in \citet{saompaper}.
 These models incorporate node-level covariate effects in addition to the traditional structural effects, such as in-degree or out-degree.
@@
 %SAOMs are typically fit to dynamic social network data using a series of MCMC phases for finding method of moments (MoM) estimates of model parameters.
-In order to estimate the parameters of a SAOM, we use the software SIENA, and its \texttt{R} implementation \texttt{RSiena} by \citet{RSiena}.
+For SAOM parameter estimation, we use the software SIENA, and its \texttt{R} implementation \texttt{RSiena} by \citet{RSiena}.
 This software is a considerable contribution to the field of social network analysis, but the extensive parameter estimation process is largely hidden from the user.
-In order to better understand the fitting process in \texttt{RSiena}, we use model visualization, or model-vis \citep{modelvis}, to examine the underlying methods, hidden structures, and simulation results.
+To better understand SAOMs and their fitting process in \texttt{RSiena}, we use model visualization, or model-vis \citep{modelvis}, to examine the underlying methods, hidden structures, and simulation results.
 Ultimately, we aim to help researchers working with SAOMs better understand the structure, behavior, fit, and results of these models.
 
-In the remainder of this introduction, we first describe the structure of network data and different network visualization techniques.
-Then, we carefully define the many moving pieces in SAOMs.
-Finally, we briefly ouline the SAOM fitting process to a set of dynamic network data.
+In the remainder of this section, we introduce the structure of network data and two different network visualization (net-vis) techniques.
+Then, we carefully define the many moving pieces that make up a SAOM.
+Finally, we briefly ouline the SAOM fitting process that we use throughout this paper.
 
@@
 
-\noindent \textbf{Visualizing Network Data:} Network visualization, also called network mapping, is a prominent subfield of network analysis.
-Visualizing network data is uniquely difficult because of the structure of the data itself.
-Most data visualizations rely on well-defined axes inherited from the data, such as Cartesian coordinates or spatial locations, but network data typically do not have an inherent structure.
+\noindent \textbf{Visualizing Network Data:} Network visualization, also called graph drawing, is a prominent subfield of network analysis \citep{graphdrawing}.
+Visualizing network data is inherently challenging because of the structure of the data itself.
+Most data visualizations rely on well-defined axes inherited from the data, such as Cartesian coordinates or spatial locations, but network data typically do not have such built-in structure.
 
@@
 
-There are two primary methods used to visualize networks: the node-link diagram and the adjacency matrix visualization \citep{knuth2000, adjmatpro}.
-For example, consider a network with five nodes, $\{1,2,3,4,5\}$, connected by five directed edges:  $\{2 \to 4, 3 \to 4, 1 \to 5, 3 \to 5, 5 \to 4\}$.
-This data is shown using the two visualization methods in Figure~\ref{fig:toyplots}.
+The lack of a native coordinate system has led to two very different methods to visualize networks: the node-link diagram and the adjacency matrix visualization \citep{knuth2000, adjmatpro}.
+We demonstrate these with a simpe example.
+Consider the network with five nodes, $\{1,2,3,4,5\}$, connected by five directed edges:  $\{2 \to 4, 3 \to 4, 1 \to 5, 3 \to 5, 5 \to 4\}$ shown in Figure~\ref{fig:toyplots}.
 
@@
 These lines can also have arrows on them indicating the direction of the edge for directed networks, as show in Figure~\ref{fig:toyplots}.
-Because there is usually no natural placement of the points, a random placement of the points is used, then adjusted with a layout algorithm, of which there are many \citep{drawingalgs}.
+Because there is often no natural placement of the nodes, they are placed in 2D at random, then adjusted with one of the many layout algorithms available \citep{drawingalgs}.
 Some commonly used algorithms, such as the Kamada-Kawai (KK) layout \citep{kamadakawai} and the Fruchterman-Reingold (FR) layout \citep{fruchterman-reingold}, are designed to mimic physical systems, drawing the graphs based on the ``forces" connecting them.
@@
 its eigenstructure), to place the nodes \citep{drawingalgs}.
-In Figure~\ref{fig:toyplots}, we use the KK algorithm, and unless otherwise stated, all other node-link diagrams in this paper will use the KK layout.
-
-\par The second method for network visualization uses the adjacency matrix of the network.
-The adjaceny matrix of a network, $\mathbf{A}$ is defined as follows: an entry $A_{ij}$ of $\mathbf{A}$, for two nodes $i\neq j$ in the network is defined as
+In Figure~\ref{fig:toyplots}, we use the KK algorithm, and unless otherwise stated, all other node-link diagrams in this paper were drawn with the KK layout algorithm.
+
+\par The second method for net-vis uses the adjacency matrix of the network. The adjaceny matrix of a network, $\mathbf{A}$ is defined as:
 \begin{equation}\label{eq:adjmat}
@@
   \begin{cases}
-                                   1 &  \text{if an edge exists } i\to j \\
+                                   1 &  \text{if an edge exists } i\to j, \quad i \neq j \\
                                    0 & otherwise
@@
 \end{equation}
-In this paper, we only consider the presence or absence of an edge, but if the network has weighted edges, $A_{ij}$ is the weight of the edge from $i \to j$.
+%In this paper, we only consider the presence or absence of an edge, but if the network has weighted edges, $A_{ij}$ is the weight of the edge from $i \to j$. 
 An adjacency matrix visualization for our toy example is also shown in Figure~\ref{fig:toyplots}.
@@
 
-\par Each visualization comes with its own advantages and disadvantages.
-For instance, paths between two nodes in a network are easier to determine with node-link diagrams than with adjacency matrix visualizations \citep{adjmatviz}.
+\par Each net-vis method has its own advantages and disadvantages.
+For instance, \emph{paths} between two nodes in a network are easier to determine with node-link diagrams than with adjacency matrix visualizations \citep{adjmatviz}.
 In node-link diagrams, node-level information can be incorporated into the visualization by coloring or changing the shape of the points, and edge-level information can be incorporated by coloring the lines, or changing their thickness or linetype.
-Incorporating a node-level variable into an adjacency matrix visualization is not as straightforward or simple, because this type of visualization features the edges.
-Adjacency matrix visualization can, however, be useful when the network is very complex, dense, or large \citep{adjmatviz}.% Experimental studies have shown adjacency matrix visualization to be superior to node-link diagrams for large networks.
+Visualizing a node-level variable into an adjacency matrix visualization is not as simple, because this type of visualization features the edges.
+Adjacency matrix visualization can, however, be useful when the network is very complex, dense, or large \citep{adjmatviz}.
+% Experimental studies have shown adjacency matrix visualization to be superior to node-link diagrams for large networks.
 For example, for basic perceptual tasks on networks, including node and edge count, adjacency matrix visualizations outperform node-link diagrams as the size and density of the network increases \citep{adjmatviz}.
@@
 A node-link diagram, however, may underrepresent the edge count if the edges $x_{ij}$ and $x_{ji}$ both exist and are drawn on top of one another.
-Ultimately, there is no one ``correct" way to visualize network information, and we will be using both the node-link and adjacency matrix visualization methods throughout this paper to explore social networks and stochastic actor-oriented models.
+Ultimately, there is no one ``correct" way to visualize network information, and we will be using both the node-link and adjacency matrix visualization methods throughout this paper to explore our example data and SAOMs.
 
@@
 
-A Stochastic Actor-Oriented Model (SAOM) is a model that incorporates all three data levels of dynamic networks: actor, tie, and temporal information.
-It models the change of a network in time, while accounting for changes in network structure due to actor-level covariates.
+Stochastic Actor-Oriented Models (SAOM) are unique because they model \textit{dynamic} networks.
+We use SAOMs to study the change of a network in time while also examining the structural and node covariate efects.
 The two titular properties of SAOMs, stochasticity and actor-orientation, are key to understanding networks as they exist naturally.
-Most social networks, even holding constant the set of actors over time, are ever-changing as relationships decay or grow in seemingly random ways.
+Most social networks, even holding constant the set of actors over time, are ever-changing as relationships form or disappear.
 Actors in social networks also have inherent properties that could affect how they change their role within the network, and existing ties may affect formation or dissolution of other ties.
@@
 
-We use the term \textit{dynamic network} to refer to a network, consisting of a fixed set of $n$ nodes, that is changing over time, and is observed at $M$ discrete time points, $t_1, \dots, t_M$ with $t_1 < t_2 < \dots < t_M$.
+A \textit{dynamic network} is a network, that consists of a fixed set of $n$ nodes, that is changing over time, and is observed at $M$ discrete time points, $t_1, \dots, t_M$ with $t_1 < t_2 < \dots < t_M$.
 We denote the network observation at timepoint $t_m$ by $x(t_m)$ for $m = 1, \dots M$.
-The SAOM assumes that this longitudinal network of discrete observations is embedded within a CTMC, which we will denote by $X(T)$.
-This CTMC is almost entirely unobserved: we assume that the beginning of the process, $X(0)$, is equivalent to the first network observation $x(t_1)$, while the end of the process, $X(\infty)$, is equivalent to the last observation $x(t_M)$.
-Nearly all other steps in the chain are unseen, with the exception of $x(t_2), \dots, x(t_{M-1})$.
-Unlike the first and last observations of the network, these ``in-between" observations do not have direct correspondence with steps in the continuous time Markov chain.
+The SAOM assumes that this longitudinal network of discrete observations is embedded within a continuous time Markov chain, (CTMC), which we will denote by $X(T)$.
+The CTMC is almost entirely unobserved: we assume the first and last network observations are the starting and ending points of the CTMC, i.e.$x(t_1) \equiv X(0)$ and $x(t_M) \equiv X(\infty)$.
+Nearly all other steps in the chain are unobserved, with the exception of $x(t_2), \dots, x(t_{M-1})$.
+Unlike the first and last observations of the network, these observations do not have direct correspondence with steps in the continuous time Markov chain.
 Thus, the observations $x(t_2), \dots, x(t_{M-1})$ are considered to be ``snapshots" of the network at some point between two steps in the CTMC.
 
-The CTMC process $X(T)$ is a series of single tie changes that happen according to the SAOM's pre-defined rate function, where one actor at a time is selected and it adds or removes one outgoing tie, or does nothing.
-Once an actor is chosen at random according to the rate function, it is ``given" the chance to change a tie, and it tries to maximize its utility function based on the current and currently reachable states of the network.
+The CTMC process $X(T)$ is a series of single tie changes that happen according to the SAOM's \textit{rate function}, where one actor at a time is ``selected" to make a change.
+Once selected, the actor tries to maximize its \textit{objective function} based on the current state and possible future states of the network.
 \\
 
-\noindent \textbf{The Rate Function} In general, each actor $i$ in the network $x$ gets an opportunity to changes its ties, $x_{ij}$, to other nodes $j \neq i$ according to the SAOM's \emph{rate function}, $\rho(x, \mathbf{z}, \boldsymbol{\alpha})$, where $\boldsymbol{\alpha}$ are the parameters in the function $\rho$, and $x$ is the current network state, with covariates of interest $\mathbf{z}$.
+\noindent \textbf{The Rate Function} In general, the SAOM's \emph{rate function}, $\rho(x, \mathbf{z}, \boldsymbol{\alpha})$, is a function of the network at its current state $x$, covariates of interest, $\mathbf{z}$, and some parameters,   $\boldsymbol{\alpha}$.
 With this general formulation, it is possible that each node will have a different rate of change.
-We assume a simple rate function, $\rho(x, \mathbf{z}, \boldsymbol{\alpha}) = \alpha_m$ that is constant across all actors between time $t_m$ and $t_{m+1}$.
-With this simple rate function, we have just one set of rate parameters in the overall model.
-In this simple model, the rate parameters dictate how often an actor $i$ ``gets an opportunity" to change one of its ties, $x_{ij}$, for $j \in \{1, \dots, n\}$ in the time period from $t_{m}$ to $t_{m+1}$ for $m = 1, \dots, M-1$ (Note that if $j = i$, no change is made.) A SAOM also assumes that the actors $i$ are conditionally independent given their ties, $x_{i1}, \dots, x_{in}$ at the current network state.
+Throughout this paper, we assume a simple rate function, $\rho(x, \mathbf{z}, \boldsymbol{\alpha}) = \alpha_m$ that is a constant across all actors between time $t_m$ and $t_{m+1}$.
+With this simple rate function, we have $M-1$ rate parameters, one for each transition from $x(t_{m-1})$ to $x(t_m)$ for $m = 2, \dots, M$.
+The rate parameter, $\alpha_m$ dictates how often an actor $i$ ``gets an opportunity" to change one of its ties, $x_{ij}$, for $j \in \{1, \dots, n\}$ in the time period from $t_{m}$ to $t_{m+1}$ for $m = 1, \dots, M-1$.
+A SAOM also assumes that the actors $i$ are conditionally independent given their ties, $x_{i1}, \dots, x_{in}$ at the current network state.
 Let $\tau(i|x,m) $ be the wait time until actor $i$ gets to the opportunity to change its current set of ties.
-For any time point, $T$, where $t_{m} \leq T < t_{m+1}$, the waiting time to the next change made by actor $i$ is exponentially distributed with expected value $\alpha_m^{-1}$, as shown in Equation~\ref{eq:c1}.
-
+For any time point, $T$, where $t_{m} \leq T < t_{m+1}$, the waiting time to the next change made by actor $i$ follows an exponential distribution:
 \begin{equation} \label{eq:c1}
@@
 
-From Equation~\ref{eq:c1}, we can derive the waiting time to the next change opportunity by \textit{any} actor in the network, $\tau(x)$.
-The value $\tau(x)$ is also exponentially distributed with expected value $(n\alpha_m)^{-1}$, where $n\alpha_m$ is the rate at which any tie change occurs.
-The estimation of this parameter in \texttt{RSiena} is simple: the method of moments is used to estimate the rate with the statistic
+\noindent From Equation~\ref{eq:c1}, we can derive the waiting time to the next change opportunity by \textit{any} actor in the network, $\tau(x)$.
+The value $\tau(x)$ is also exponentially distributed, and has expected value $(n\alpha_m)^{-1}$.
+The estimation of $\alpha_m$ in \texttt{RSiena} is simple: the method of moments (MoM) is used to estimate the rate with the statistic
 \begin{equation}\label{eq:cdiff}
-C = \sum_i\sum_j |x_{ij}(t_{m+1}) - x_{ij}(t_m)|
+C_m = \sum_i\sum_j |x_{ij}(t_{m+1}) - x_{ij}(t_m)|
 \end{equation}
@@
 
-\noindent \textbf{The Objective Function} Because of the conditional independence assumptions given in Equation~\ref{eq:c1}, we can consider the objective function for each node separately, as only one tie from one node is changing at a time in the CTMC.
-The node $i$, which is given the opportunity to change at the current time point, is called the \textit{ego} node.
-It has the potential to interact with all other nodes in the network, $j \neq i$.
-These nodes $j$, are referred to as \textit{alter} nodes, and do not change any of their ties when $i$ may change.
-For the ego node $i$ in the current network state $x$, its \emph{objective function}, which it tries to maximize, is written as
+\noindent \textbf{The Objective Function} Because of the conditional independence assumptions given in Equation~\ref{eq:c1}, we consider the objective function for each node separately.
+Suppose node $i$ is selected to change at the next step in the CTMC.
+This node, called the \textit{ego} node, has the potential to interact with all other nodes in the network, $j \neq i$.
+These nodes $j$, are referred to as \textit{alter} nodes, and will not change any of their ties at the next step in the CTMC.
+The ego node $i$ in the current network state $x$, will make a change that maximizes its \emph{objective function}, written as:
 \begin{equation} \label{eq:of1}
@@
 \end{equation}
-for $x \in \mathcal{X}$, the space of all possible directed networks with the $n$ nodes, and $\mathbf{Z}$, the matrix of covariates, and $K$ the total number of parameters in the objective function.
-The vector $\boldsymbol{\beta}$ contains the parameters of the model with corresponding statistics, $s_{ik}(x, \mathbf{Z})$, for $k = 1,\dots, K$.
-To increase the value of $f_i$, $i$ will destroy one tie where $x_{ij} = 1$, create one tie where $x_{ij}=0$, or make no change if all changes will result in lower values of $f_i(\boldsymbol{\beta}, x, \mathbf{Z})$.
-
+for the network $x$, covariates $\mathbf{Z}$, and $K$ the total number of parameters in the objective function.
+The parameters of the model, $\beta_1, \dots, \beta_K$ correspond to network statistics, $s_{ik}(x, \mathbf{Z})$, for $k = 1,\dots, K$.
+To maximize $f_i$, $i$ will either destroy a tie where $x_{ij} = 1$, create a tie where $x_{ij}=0$, or make no change if all changes will result in lower $f_i$.
+
+The parameters of $f_i(\boldsymbol{\beta}, x, \mathbf{Z})$ are either structural or covariate parameters.
+The structural parameters correspond to statistics that are \emph{only} functions of the current network state, $x$, while the covariate parameters are also functions of the covariates, $\mathbf{Z}$.
 According to \citet[p.
 371]{snijders01}, there should be at least two parameters included in the model: $\beta_1$ for the outdegree of nodes, and $\beta_2$ for the reciprocity of nodes.
-(These effects are also used in ERGMs.) The outdegree represents the propensity of nodes with a lot of outgoing ties to form more outgoing ties (the ``rich get richer" effect), and the reciprocity parameter measures the tendency of outgoing ties to be returned within a network.
-The statistics corresponding to these two parameters, $s_{i1}(x)$ and $s_{i2}(x)$ are given inTable~\ref{tab:effects}.
+The former measures the propensity of nodes with a lot of outgoing ties to form more outgoing ties (the ``rich get richer" effect), while the latter measures the tendency of outgoing ties to be returned within a network.
+The statistics corresponding to these two parameters, $s_{i1}(x)$ and $s_{i2}(x)$ are given in Table~\ref{tab:effects}.
 In the \texttt{RSiena} software, there are over 80 possible $\boldsymbol{\beta}$ parameters to add to the model.
-The formulas for the corresponding statistics for all effects are provided in \citet{RSienamanual}.
-
-The parameters of $f_i(\boldsymbol{\beta}, x, \mathbf{Z})$ are either structural or covariate parameters.
-The structural parameters, such as the outdegree and reciprocity parameters, correspond to statistics that are \emph{only} functions of the current network state, $x$.
-They model underlying mechanisms of network change, answering questions such as, ``How does the existing network structure influence change in the network?" The covariate parameters are functions of $x$ and additional node-level covariates of interest, $\mathbf{Z}$.
-Some possible covariate parameters and their corresponding statistics are discussed in Section~\ref{sec:exdatamods} and shown in Figure~\ref{fig:structures.}
+The definition of the statistics for all possible parameters are provided in \citet{RSienamanual}, and two selected covariate parameters and corresponding statistics are discussed in Section~\ref{sec:exdatamods} and shown in Figure~\ref{fig:structures}.
 
@@
 
-The objective function also defines the probability that actor $i$, when it is the ego node, chooses to change the tie to actor $j$, denoted by $p_{ij}(\boldsymbol{\beta}, x, \mathbf{Z})$.
+With the objective function, we can define the probability that the ego node $i$ chooses to change its tie to alter node $j$, denoted by $p_{ij}(\boldsymbol{\beta}, x, \mathbf{Z})$.
 Let $x(i \leadsto j)$ be the network identical to $x$ with the exception of tie $x_{ij}$, which is equal to $1-x_{ij}$.
 The probility that the tie $x_{ij}$ changes is defined as:
-
 \begin{equation}\label{eq:pij}
@@
 \end{equation}%When $i = j$, the numerator represents the exponential of the value of the objective function when evaluated at the current network state. 
-When the value of the objective function is high at the current state, the probability of not making a change is also high.
-In the CTMC, when actor $i$ may make a change, it chooses which tie $x_{i1}, \dots, x_{in}$ to change at random according to the probabilities $p_{ij}(\boldsymbol{\beta}, x, \mathbf{Z})$.
-The objective function and the resulting values of $p_{ij}$ are combined with the rate function to fully describe the CTMC that is used to model network change in a SAOM.
+%When the value of the objective function is high at the current state, the probability of not making a change is also high.
+In the CTMC, when actor $i$ can make a change, it chooses which tie $x_{i1}, \dots, x_{in}$ to change at random according to the probabilities $p_{ij}(\boldsymbol{\beta}, x, \mathbf{Z})$.
+We explore these probabilities further in Section~\ref{sec:algorithms}.
+The probabilities $p_{ij}$ combined with the rate parameters $\alpha_m$ fully characterize the underlying CTMC that models network change.
 For more on CTMCs, see \citet{ctmc}.
-We explore these probabilities further in Section~\ref{sec:algorithms}
-
 % \subsubsection{Continuous Time Markov Chain (CTMC)}
@@
 To fit a SAOM to dynamic network data, we apply the \texttt{RSiena} software, which uses simulation methods to estimate parameter values using either MoM or maximum likelihood (ML) estimation \citep{RSiena}.
-We chose to use MoM estimation, so we describe that fitting process here, and more information on ML estimation can be found in \citet{SienaAlgs}.
+We chose to use MoM estimation, so we describe that process here, while more information on ML estimation can be found in \citet{SienaAlgs}.
 %because the theory behind it was established in \citet{saompaper}, while the maximum likelihood estimation methods were not fully established until \citet{saomsml2010}, though \texttt{RSiena} contains capabilities to use maximum likelihood estimation.
 We also use the score function method for estimating the derivatives of the expected values, as opposed to the finite differences method, both of which are outlined in detail in .
-
-\par The underlying SIENA software uses a Robbins-Monro algorithm (see \citet{robbinsmonro}) to estimate the solution of the moment equation
+The underlying SIENA software uses a Robbins-Monro algorithm (see \citet{robbinsmonro}) to estimate the solution of the moment equation
 \begin{equation}\label{eq:mom}
@@
 where $\theta = (\boldsymbol{\alpha}, \boldsymbol{\beta})$ is the vector of rate and objective function parameters, and $s_{obs}$ is the observed vector of the model statistics over all network observations.
-The full SIENA MoM estimation algorithm operates in three phases, as described in \citet{RSienamanual, SienaAlgs}.
-The first phase performs initial estimation of the score functions for use in the Robbins-Monro procedure for MoM estimation.
-The second phase carries out the Robbins-Monro algorithm and obtains estimates of the parameter values through iterative updates and simulation from the CTMC at iterative parameter values.
-The third phase uses the parameter vector estimated in phase two to estimate the score functions and covariance matrix of the vector of parameter estimates and carries out convergence checks.
-
-In each of the the first two phases of the algorithm, the procedure uses ``microsteps" that are simulated from the model as it exists in its current state in order to update either the score functions or the parameter estimates.
-These simulated microsteps are realized steps in the CTMC that is the backbone of the SAOM.
+The full SIENA MoM estimation algorithm operates in three phases, as described in \citet{RSienamanual} and \citet{SienaAlgs}, and we briefly summarize these phases.
+The first phase gets initial estimates of the score functions for use in the second phase.
+The second phase is the Robbins-Monro algorithm, and results in parameter estimates through iterative updates and simulation from the CTMC at iterations of the parameter values.
+The third phase uses the parameters from phase two to estimate the score functions and covariance matrix of the parameter estimates and carries out convergence checks.
 In Section~\ref{sec:algorithms}, we further explore these phases in the SIENA method-of-moments algorithm through visualization, bringing them out of their ``black box" and into the light.
+
+The rest of the paper is strucured as follows.
+In Section \ref{sec:exdatamods}, we introduce the example data and models we use for model-vis.
+Then, in Section~\ref{sec:minds}, we explore the first step in model-vis by placing the models in the data space.
+Section~\ref{sec:collectmodels} follows, in which we explore collections of models.
+Finally, in Section~\ref{sec:algorithms}, the last principle of model-vis, we visualize the underlying algorithms in the SAOM fitting process.
+We conclude with a discussion in Section~\ref{sec:discussion}.
 
@@
 When fitting a SAOM in \texttt{RSiena}, several simulation steps are hidden from the user.
-One of the hidden steps in the SIENA method of moments algorithm is the \textit{microstep} process.
+One of the hidden steps in the SIENA MoM algorithm is the \textit{microstep} process.
 At each microstep, an \textit{ego} node is selected to make a change, and the chosen ego node randomly makes one change in its ties to maximize its objective function.
@@
 
-To create the average network visualization shown in Figure~\ref{fig:averagenet}, we follow the same procedure as in Section~\ref{sec:minds}, counting occurrences of each possible edge in the simulations, resulting in a summary network with weighted edges representing the number of times an edge appeared in the simulated wave 2 when simulating from the SAOM 1,000 times.
+To create the average net-vis shown in Figure~\ref{fig:averagenet}, we follow the same procedure as in Section~\ref{sec:minds}, counting occurrences of each possible edge in the simulations, resulting in a summary network with weighted edges representing the number of times an edge appeared in the simulated wave 2 when simulating from the SAOM 1,000 times.
 As in Figure~\ref{fig:summnetM1}, edges only appear in the average networks if they appear more than 5\% of the time in the simulations.

### 2018-10-01 paper.Rnw
fix a fig ref
revised abstract, finished revising intro section. Quite pleased with it

@@
 Stochastic actor-oriented models (SAOMs), first introduced by Snijders (1996), are a type of statistical network model for dynamic social networks.
-Unlike other network models, SAOMs are not very well understood.
-We use model visualization techniques introduced in Wickham et al (2015) in order to gain a better understanding of the models and their behavior.
-Viewing the model in the data space places these models into the field of network drawing, while exploring collections of models helps with parameter interpretation.
-Finally, delving into the fitting algorithms provides more insight into the underlying mechanisms.
+Unlike other network models, SAOMs are analytically intractable, and therefore hard to interpret.
+We use model visualization techniques introduced in Wickham et al (2015) better understand these models and their behavior.
+We use a three-pronged attack: (a) a network visualization to view the \emph{model}, (b) exploring collections of models for parameter interpretation, and (c) delving into the fitting algorithms to gain insight into the underlying mechanisms.
 With the help of static and dynamic visualizations, we bring the hidden SAOM fitting processes into the foreground, eventually leading to a better understanding and higher accessibility of SAOMs for social network analysts.

### 2018-10-02 paper-child.Rnw
revise appendix

@@
 
+\subsection*{Section 4.1}
+
 <<sigtesting, fig.show = 'hold', out.width='\\textwidth', fig.height=2, fig.cap="From Section~\\ref{sec:allpossible}.
@@
 
-\subsection*{Section 4.1}
-
 In Figure~\ref{fig:sigtesting}, we see for the friendship data and the senate data that most of the significant effects have absolute value less than ten.
 In addition, the $p$-values for the effects from the friendship data are more spread out than the $p$-values for the senate data, which are concentrated at about 0.02 or less.
-This may suggest that larger data sets tend to result generally in smaller $p$-values, just like a larger sample size results in smaller $p$-values in a $t$-test.
-Since the tests are Wald-type tests, this makes sense.
+This suggests that larger data sets tend to result generally in smaller $p$-values, which is consistent with the construction of Wald-type tests.
 
@@
 
-<<averagenet, results='hide', out.width='\\textwidth',fig.cap='The node-link diagrams from the three "average" networks that we calculated are in the top row, and the true wave 1 and wave 2 data are shown in the bottom row above.
-There is some difference between the three models, but overall, these three models cannot capture the structure in the true second wave of data.', fig.height=5>>=
+<<averagenet, results='hide', out.width='\\textwidth',fig.cap='The node-link diagrams from the three expected networks that we calculated are in the top row, and the true wave 1 and wave 2 data are shown in the bottom row above.
+There is some difference between the three models, but overall, these three models struggle to capture the structure in the true second wave of data.', fig.height=5>>=
 
@@
 
-To create the average net-vis shown in Figure~\ref{fig:averagenet}, we follow the same procedure as in Section~\ref{sec:minds}, counting occurrences of each possible edge in the simulations, resulting in a summary network with weighted edges representing the number of times an edge appeared in the simulated wave 2 when simulating from the SAOM 1,000 times.
-As in Figure~\ref{fig:summnetM1}, edges only appear in the average networks if they appear more than 5\% of the time in the simulations.
-In Figure~\ref{fig:averagenet}, we show the ``average'' network from the three models we fit and the first and second waves of data.
-Comparing the three averages to waves 1 and 2, we see that they have very similar structure to wave 1.
-Model 2, which included the transitive triplet parameter, seems to have created a larger connected component overall than models 1 and 3.
-In particular, if we look at the group of nodes $\{10,11,14\}$, we see they are very strongly connected within the three average networks, and they are completely separate from the other nodes in the true wave 2.
-None of the three average networks show node 16 gaining ties as it does in wave two, nor do they show nodes 4 and 7 becoming isolated.
-In Model 2, however, the ties to node 7 appear much weaker than in Model 1 or Model 2, suggesting that of the three, Model 2 may be the best fit for our data.
+To create the expected network visualization shown in Figure~\ref{fig:averagenet}, we follow the same procedure as in Section~\ref{sec:minds}, counting occurrences of each realized edge in the simulations.
+The resulting summary network has weighted edges representing the proportion of edge appearances in all 1,000 simulations of wave 2.
+As in Figure~\ref{fig:summnetM1}, edges only appear in the expected networks if they appear more than 5\% of the time in the simulations.
+In Figure~\ref{fig:averagenet}, we show the  expected network from the three models we fit and the first and second waves of data.
+Comparing the three averages to waves 1 and 2, we see that they have very similar structure to wave 1 but that model M2, which included the transitive triplet parameter, results in more strongly connected components.
+The lack of fit is clear: none of the three average networks show node 16 gaining ties as it does in wave two, nor do they show nodes 4 and 7 becoming isolated.
+In model M2, however, the ties to node 7 appear much weaker than in model M1 or model M3, and the nodes $\{10,11,14\}$ are also strongly connected as they are in wave 2 of data, suggesting that M2 may be the best model of the three.
 
@@
 <<compareModels, fig.cap="Density plots of objective function parameter estimates from repeatedly fitting models M1, M2, and M3 to the example data.
-Outdegree and reciprocity have the same inverse relationship for both data sets.
+%Outdegree and reciprocity have the same inverse relationship for both data sets.
 For the friendship data, the inclusion of $\\beta_3$ has strong effect on the estimates of the other two parameters, but not for the senate data.", fig.height = 3>>=
@@
 The most notable difference is in the values of $\beta_1, \beta_2$.
-In the senate data, the inclusion of $\beta_3$ has no effect on the estimates of $\beta_1, \beta_2$, while $\beta_3$ does seem to have an effect with the friendship data.
+In the senate data, the inclusion of $\beta_3$ has no effect on the estimates of $\beta_1, \beta_2$, while the opposite is true for the friendship data.
 
@@
 
-In Figure~\ref{fig:corplots}, we examine correlations between each of pair of parameters within each model and overall.\nocite{ggally} The strongest correlation within each model is between $\beta_1$ and $\beta_2$, with absolute value of correlation between those two parameter values greater than 0.90 in all three models.
+In Figure~\ref{fig:corplots}, we examine correlations between each of pair of parameters within each model and overall.\nocite{ggally} The strongest correlation within each model is between $\beta_1$ and $\beta_2$, with absolute correlation between them greater than 0.90 in all three models.
 The $\beta_1$ parameter is also highly correlated with the $\beta_3$ parameter within model M2, but it is not as highly correlated with the $\beta_4$ parameter in model M3.
-It might therefore be advisable to consider only models that either allow $\beta_1$, or $\beta_2$.
+It might therefore be advisable to consider only models that either allow $\beta_1$ or $\beta_2$ but not both.
 Looking at the high correlation with $\alpha$, we might switch to a model without $\beta_1$.
@@
 \subsection*{Section 5}
-
-Finally, we combine 1,000 simulations from model M1 into a visualization that displays the entire microstep process for all simulations in Figure ~\ref{fig:alledgests}.
-To make this visualization, we first assign each possible edge an edge ID number so that we can keep track of it throughout all the microsteps and all the simlations.
-Then, we count up the total number of times each edge appears in the microstep process in each of the 1,000 simulations for use as an ordering variable later.
-We also count up the number of times an edge occurs in each microstep number in the 1,000 simulations.
-Since the number of microsteps in the process varies, the number of times an edge occurs decreases as the microstep number increases.
-Next, we compute a proportion, which we call the occurrence percentage, which is the number of times the edge occurred in a microstep divided by 1,000.
-Finally, we visualize all this information together in Figure~\ref{fig:alledgests}.
-In this plot, all possible edges are shown, and we see that every one of the $16 \times 15 = 240$ possible edges in the network occurs at some point in the simulation process.
-We also see, however, that the process struggles to focus in on the edges in the second wave of the data.
-Ideally, we would like to see more occurrences of the edges which appear in the second wave of data.
-But, about half of the edges in wave two are in the bottom half when ordered by number of occurrences, meaning they do not appear as much as they would if the model was truly excellent at capturing the mechanisms of tie change in the network.
-This solidifies what we found in Figure~\ref{fig:heatmap}: the model M1 and the SAOM fitting process do not explore the data space enough to adequately capture the network change mechanism.
 
@@
 
-Much of the code used to create these plots has been packaged an posted online at https://github.com/sctyner/netvizinf and https://github.com/sctyner/geomnet (also on CRAN).
+Finally, we combine 1,000 simulations from model M1 into a visualization that displays the entire microstep process in Figure ~\ref{fig:alledgests}.
+To make this visualization, we first assign each possible edge in the network an ID number so that we can keep track of it throughout all microsteps and all simlations.
+Then, we count up the number of times each edge appears in the network throughout the microstep process for each of the 1,000 simulations.
+We also count the number of times an edge occurs at each step number.
+Since the number of microsteps in the process varies, the number of times an edge occurs decreases as the step number increases.
+Next, we compute a proportion, which we call the occurrence percentage, which is the number of times the edge was in the network in step 1, step 2, etc.
+divided by 1,000.
+Finally, we visualize all this information together in Figure~\ref{fig:alledgests}.
+In this plot, we see that all possible edges appear at least once at some point in the microsteps of at least one simulated process.
+We also see, however, that the process struggles to focus in on the edges in the second wave of the data.
+Ideally, we would like to see more occurrences the edges in wave 2 that are not in wave 1.
+But, about half of the edges in wave two are in the bottom half of the figure, which means they do not appear as much as they would if the model was actually capturing the mechanisms of tie change in the network.
+More of the darkest areas of the figure should belong to ties in wave 2, but those are often the lightest.
+This solidifies what we found in Figure~\ref{fig:heatmap}: the model M1 and the SAOM fitting process do not explore the data space enough to adequately capture the network change mechanism.
+
+\noindent \textbf{Online supplemental material:} Much of the code used to create these plots has been packaged an posted online at https://github.com/sctyner/netvizinf and https://github.com/sctyner/geomnet (also on CRAN).
 Additional, non-packaged code can be found at https://github.com/sctyner/SAOM-removing-blindfold.

### 2018-10-02 paper-child.Rnw
add a citation in intro
add pdf. Just need to spend time on conclusion, then fix some plots, then ready to submit.

@@
 
-\par Unlike other network models, SAOMs are not analytically tractable.
+\par Unlike other network models, SAOMs are not analytically tractable \citep{saomreview}.
 Likelihood functions quickly become very complex due to the dependency structure inherent in the data, and so computational methods, such as Markov chain Monte Carlo (MCMC), are used to fit models.
@@
 
-\subsection*{Section 4.1}
-
-<<sigtesting, fig.show = 'hold', out.width='\\textwidth', fig.height=2, fig.cap="From Section~\\ref{sec:allpossible}.
-Significant effects for the two data sets, at a significance level of 0.10 or lower as calculated by the Wald-type test available in the SIENA software.", error=FALSE>>=
+\subsection*{Section 4}
+
+<<sigtesting, fig.show = 'hold', out.width='\\textwidth', fig.height=2, fig.cap="Significant effects for the two data sets, at a significance level of 0.10 or lower as calculated by the Wald-type test available in the SIENA software.", error=FALSE>>=
 smfr_effs <- read_csv("data/effects_significance_smallFriends.csv")
@@
 
-\subsection*{Section 4.2}
-
 <<averagenet, results='hide', out.width='\\textwidth',fig.cap='The node-link diagrams from the three expected networks that we calculated are in the top row, and the true wave 1 and wave 2 data are shown in the bottom row above.
@@
 
-\subsection*{Section 4.3}
-
-<<compareModels, fig.cap="Density plots of objective function parameter estimates from repeatedly fitting models M1, M2, and M3 to the example data.
 %Outdegree and reciprocity have the same inverse relationship for both data sets.
-For the friendship data, the inclusion of $\\beta_3$ has strong effect on the estimates of the other two parameters, but not for the senate data.", fig.height = 3>>=
+For the friendship data, the inclusion of $\\beta_3$ has strong effect on the estimates of the other two parameters, but not for the senate data.
+<<compareModels, fig.cap="Density plots of objective function parameter estimates from repeatedly fitting models M1, M2, and M3 to the example data. ", fig.height = 3>>=
 simu2$data <- "friends"
@@
 In the senate data, the inclusion of $\beta_3$ has no effect on the estimates of $\beta_1, \beta_2$, while the opposite is true for the friendship data.
-
-\subsection*{Section 4.4}
 
@@
 
-Finally, we combine 1,000 simulations from model M1 into a visualization that displays the entire microstep process in Figure ~\ref{fig:alledgests}.
+Finally, we combine 1,000 simulations from model M1 into a visualization that displays the entire microstep process in Figure~\ref{fig:alledgests}.
 To make this visualization, we first assign each possible edge in the network an ID number so that we can keep track of it throughout all microsteps and all simlations.

### 2018-10-03 paper-child.Rnw
change title, change plots to be grayscale safe, i hate writing conclusions.

@@
 ThemeNoNet <- theme_bw() %+replace%
-            theme(plot.title = element_text(size = 12,
+            theme(plot.title = element_text(size = 10,
                                             face = 'plain',
@@
                                             family="Times New Roman"),
-                  axis.title.x = element_text(size = 12,
+                  axis.title.x = element_text(size = 10,
                                             face = 'plain',
@@
                                             family="Times New Roman"),
-                  axis.title.y = element_text(size = 12,
+                  axis.title.y = element_text(size = 10,
                                             face = 'plain',
@@
                                             family="Times New Roman"),
-                  axis.text.x = element_text(size = 12,
+                  axis.text.x = element_text(size = 10,
                                             face = 'plain',
@@
                                             family="Times New Roman"),
-                  # axis.text.x.top = element_text(size = 12,
+                  # axis.text.x.top = element_text(size = 10,
                   #                           face = 'plain',
@@
                   #                           family="Times New Roman"),
-                  # axis.text.x.bottom = element_text(size = 12,
+                  # axis.text.x.bottom = element_text(size = 10,
                   #                           face = 'plain',
@@
                   #                           family="Times New Roman"),
-                  axis.text.y = element_text(size = 12,
+                  axis.text.y = element_text(size = 10,
                                             face = 'plain',
@@
                                             family="Times New Roman"),
-                  # axis.text.y.left = element_text(size = 12,
+                  # axis.text.y.left = element_text(size = 10,
                   #                           face = 'plain',
@@
                   #                           family="Times New Roman"),
-                  # axis.text.y.right = element_text(size = 12,
+                  # axis.text.y.right = element_text(size = 10,
                   #                           face = 'plain',
@@
                   #                           family="Times New Roman"),
-                  strip.text.x = element_text(size = 12,
+                  strip.text.x = element_text(size = 10,
                                             face = 'plain',
@@
                                             margin = margin(t = 3, r = 0, b = 3, l = 0, unit = "pt")),
-                  strip.text.y =element_text(size = 12,
+                  strip.text.y =element_text(size = 10,
                                             face = 'plain',
@@
 ThemeNet <- theme_net() %+replace%
-            theme(plot.title = element_text(size = 12,
+            theme(plot.title = element_text(size = 10,
                                             face = 'plain', angle = 0, family = "Times New Roman"),
-                  strip.text.x = element_text(size = 12,
+                  strip.text.x = element_text(size = 10,
                                             face = 'plain', angle = 0, family = "Times New Roman",
                                             margin = margin(t = 3, r = 0, b = 3, l = 0, unit = "pt")),
-                  strip.text.y =element_text(size = 12,
+                  strip.text.y =element_text(size = 10,
                                             face = 'plain', angle = 0, family = "Times New Roman",margin = margin(t = 3, r = 0, b = 3, l = 0, unit = "pt")),
@@
 Other data on a network include inherent variables of interest on the nodes and edges themselves, e.g.\ someone's age, gender, and political affiliation, the type of relationship (friendship, coworkers, etc) that an edge represents, in addition to the changing structure of the network in time.
-\\
-
 %These two network data structures pose unique problems to analysts, such as, ``How does the strength of a tie between two nodes affect the overall structure of the network?" or ``Do differences in node variables affect the formation of edges?" We propose to answer these questions and more through visual exploration of network data and models.
@@
                                    1 &  \text{if an edge exists } i\to j, \quad i \neq j \\
-                                   0 & otherwise
+                                   0 & \text{otherwise}
   \end{cases}
@@
 The SAOM assumes that this longitudinal network of discrete observations is embedded within a continuous time Markov chain, (CTMC), which we will denote by $X(T)$.
-The CTMC is almost entirely unobserved: we assume the first and last network observations are the starting and ending points of the CTMC, i.e.$x(t_1) \equiv X(0)$ and $x(t_M) \equiv X(\infty)$.
+The CTMC is almost entirely unobserved: we assume the first and last network observations are the starting and ending points of the CTMC, i.e.
+$x(t_1) \equiv X(0)$ and $x(t_M) \equiv X(\infty)$.
 Nearly all other steps in the chain are unobserved, with the exception of $x(t_2), \dots, x(t_{M-1})$.
@@
 Once selected, the actor tries to maximize its \textit{objective function} based on the current state and possible future states of the network.
-\\
-
-\noindent \textbf{The Rate Function} In general, the SAOM's \emph{rate function}, $\rho(x, \mathbf{z}, \boldsymbol{\alpha})$, is a function of the network at its current state $x$, covariates of interest, $\mathbf{z}$, and some parameters,   $\boldsymbol{\alpha}$.
+
+\noindent \textbf{The Rate Function} In general, a SAOM's \emph{rate function}, $\rho(x, \mathbf{z}, \boldsymbol{\alpha})$, is a function of the network at its current state $x$, covariates of interest, $\mathbf{z}$, and some parameters,   $\boldsymbol{\alpha}$.
 With this general formulation, it is possible that each node will have a different rate of change.
-Throughout this paper, we assume a simple rate function, $\rho(x, \mathbf{z}, \boldsymbol{\alpha}) = \alpha_m$ that is a constant across all actors between time $t_m$ and $t_{m+1}$.
+Throughout this paper, we assume a simple rate function, $\rho(x, \mathbf{z}, \boldsymbol{\alpha}) \equiv \alpha_m$ that is a constant across all actors between time $t_m$ and $t_{m+1}$.
 With this simple rate function, we have $M-1$ rate parameters, one for each transition from $x(t_{m-1})$ to $x(t_m)$ for $m = 2, \dots, M$.
@@
 \end{equation}
-which is the total number of changes from observation at time $t_m$ to the observation at time $t_{m+1}$. \\
+which is the total number of changes from observation at time $t_m$ to the observation at time $t_{m+1}$.
 
@@
   geom_net(fiteach = FALSE, directed=TRUE, layout.alg = 'fruchtermanreingold', size = 5, arrowsize = .25, arrowgap = .05, ecolour = 'grey40', labelon=T, labelcolour = 'black', fontsize = 4, vjust = .5, linewidth=0.5) +
-  scale_color_brewer(palette = "YlOrRd", name = "Drinking\nBehavior", labels = c("never", "1-2 times per year", "once per month", "once per week")) +
+  scale_color_brewer(palette = "Greys", name = "Drinking\nBehavior", labels = c("never", "1-2 times per year", "once per month", "once per week")) +
   ThemeNet +
@@
         panel.background = element_rect(fill = NA, color = "grey30"),
-        legend.text = element_text(size = 12, family = "Times New Roman"),
-        legend.title = element_text(size = 12, family = "Times New Roman"))
+        legend.text = element_text(size = 10, family = "Times New Roman"),
+        legend.title = element_text(size = 10, family = "Times New Roman"))
 alldataviz
@@
 These data come from \citet{friendsdata}, and we chose to only work with a subset of the data to make node-link diagrams easier to see and to make any changes in the network more noticeable.
-%To determine which subset to select, we visualized all waves of the full network using the adjacency matrix visualization approach, which we show in Figure~\ref{fig:adjmatviz}.
-This adjacency matrix visualization is different from the one in Figure~\ref{fig:toyplots} because it does not show the node IDs on the axes.
-The network in Figure~\ref{fig:adjmatviz} is much larger than our toy data example, so we remove the node labels to remove clutter.
-In all three adjacency matrices, the ego and the alter nodes are ordered by node ID, 1-50, which were determined arbitrarily by the relationships in wave 1.
-%In Figure~\ref{fig:adjmatviz}, we show the adjacency matrix visualization of all three observations of this network, with the subset we selected for analysis outlined in red.
-This subset contains 16 actors, their ties and their drinking behavior for three timepoints.
 The node-link visualization of this data is provided in Figure~\ref{fig:alldataviz}.
@@
   ThemeNet +
-  scale_color_manual(values = c("royalblue", "forestgreen","firebrick")) +
+  scale_color_manual(values = c("black", "grey30","#ef3b2c")) +
   theme(legend.position = 'bottom', strip.text = element_text(size = 20)) +
@@
 By using data sets from very different contexts, we hope to assess possible differences in model performance.
-\\
 
@@
 
-%It is ofen difficult to know for certain how to interpret a fitted value of a parameter.
-We can make educated guesses based on the definition of the effect and the sign of the fitted value, but a direct interpretation is not always possible.
-By graphically exploring these models, we aim to understand their effects, their interpretations, and the model fitting process better.
-% XXXX MOVE BELOW PAR TO SECTION 1 XXX
-% In this paper, we use three different model visualization techniques on SAOMs.
-First, Section~\ref{sec:minds} places the models in the data space.
-Then, Section~\ref{sec:collectmodels} explores visualizations of various collections of models.
-Finally, in Section~\ref{sec:algorithms} we use model visualization to explore the underlying algorithms in the SAOM fitting process.
-We conclude with a discussion in Section~\ref{sec:discussion}.
- %In terms of the senate data, then, the value of $\beta_3$ should dictate how willing a senator is to coauthor a bill on which the author of the bill has a different level of authorship, assuming there is an intermediary between the two who has the same authorship level as the first.
-%Comparison of the parameter estimates of these models fit to the two different data sets is given in Table~\ref{tab:meanests}.
-%In addition, we fit two more models, M4 and M5 to the senate data that were found to be significant according to the Wald-type test:
-
-%\begin{align*}
-%f_i(x)^{M4} & = \beta_1s_{i1} + \beta_2s_{i2} + \beta_5s_{i5} \\
-%f_i(x, z)^{M5} & = \beta_1s_{i1} + \beta_2s_{i2} + \beta_6s_{i6}.
-%\end{align*}
-
-%In M4, $s_{i5} = \mathbb{I}(x_{i+} = x_{+i} = 0)$ is an indicator variable for isolated nodes, called the network-isolate effect.
-This parameter attempts to measure the effect of a node being completely isolated in the network.
-If the value of the parameter $\beta_5$ is high, nodes are encouraged to remain isolated.
-In M5,  $s_{i6} = \sum\limits_j x_{ij}x_{ji}\mathbb{I}(z_i = z_j)$ is a measure of reciprocity between senators in the same party, $z_i, z_j \in \{\text{Democrat, Independent, Republican}\}$, generally called the same covariate reciprocity effect.
-The values of the covariate are converted to numeric, $\{1,2,3\}$, respectively, and mean-centered within \texttt{RSiena}.
-The mean-centered values of $z \in \{\text{Democrat, Independent, Republican}\}$ are $\{-0.96, 0.04, 1.04\}$.
-The parameter $\beta_6$ attempts to measure the propensity for forming reciprocal ties between nodes with the same covariate value.
-If the value of the parameter $\beta_6$ is high, nodes belonging to the same covariate group are encourage to reciprocate incoming ties.
-
 \section{Model in the data space}\label{sec:minds}
@@
 
-<<funsenatevis, fig.height=3, fig.cap='The 111th Senate at two times: while Clinton was in the senate in 2009 (on the left) and after she left (on the right).
+<<funsenatevis, fig.height=4, fig.cap='The 111th Senate at two times: while Clinton was in the senate in 2009 (on the left) and after she left (on the right).
 We map sex, party, and bills authored to the shape, color, and size of the nodes, respectively.
@@
 
+set.seed(12981120)
 clintonSenate %>% filter(!(source %in% c("Roland Burris", "Bernard Sanders", "Frank R. Lautenberg", "Mary L. Landrieu") & is.na(target))) %>%
 ggplot() +
-  geom_net(directed = T, labelon=F, arrowsize = .3, singletons= F, fiteach = T, arrowgap = .01, layout.alg = 'fruchtermanreingold',
-           aes(from_id = source, to_id = target, color = party, linewidth = gsw, shape = sex,size = n_au, group = party)) +
-  scale_size_continuous(name="# bills\nauthored" , guide = guide_legend(nrow= 2)) +
+  geom_net(directed = T, labelon=F, arrowsize = .3, singletons= F, fiteach = T, arrowgap = .01, layout.alg = 'fruchtermanreingold', ecolour = "grey70",
+           aes(from_id = source, to_id = target, color = party, linewidth = gsw, shape = sex,size = n_au/30)) +
+  scale_size_continuous(name="# bills\nauthored" , guide = guide_legend(nrow= 2), breaks = c(0,1,2,3), labels = c(0,30, 60, 90)) +
   #theme_net() +
@@
   scale_shape_manual(values = c(17,16), guide = guide_legend(nrow= 2)) +
-  scale_color_manual(values = c("royalblue", "forestgreen","firebrick"), guide = guide_legend(nrow=2)) +
+  scale_color_manual(values = c("black", "grey30","#ef3b2c"), guide = guide_legend(nrow=2)) +
   facet_wrap(~Clinton, nrow = 1, labeller = 'label_both')  +
@@
   geom_net(aes(from_id = from, to_id = to, linewidth= linewidth),
-           directed = T, curvature = .2, labelon=T, size = 4,
+           directed = T, curvature = .2, labelon=T, size = 3,
            arrow = arrow(angle = 20, type = "open", length = unit(.2, "cm")), ecolour = "black",
-           color = "#252525", hjust = .5, vjust = .5, labelcolour = 'white', fontsize = 3, arrowgap = .02,
+           color = "#252525", hjust = .5, vjust = .5, labelcolour = 'white', fontsize = 2, arrowgap = .02,
            singletons = T) +
@@
   geom_net(aes(from_id = from, to_id = to, linewidth= linewidth/3),
-           directed = T, curvature = .2, labelon=T, size = 4,
+           directed = T, curvature = .2, labelon=T, size = 3,
            arrow = arrow(angle = 20, type = "open", length = unit(.2, "cm")), ecolour = "#969696",
-           color = "#252525", hjust = .5, vjust = .5, labelcolour = 'white', fontsize = 3, arrowgap = .03,
+           color = "#252525", hjust = .5, vjust = .5, labelcolour = 'white', fontsize = 2, arrowgap = .03,
            singletons = T) +
@@
 <<summary2, fig.height=2.5, fig.cap="Network differences between the expected network from M1 and Wave 1 and Wave 2 of the data.
-Some edges appear in both the data and the expected Wave 2 from M1, while others only appear in the data or only in the exectation.
-Width of edges in the represent the proportion of appearances in the 1,000 simulations.
+Some edges appear in both the data and the expected Wave 2, while others only appear in the data or only in the exectation.
+Width of edges represent the proportion of appearances in the 1,000 simulations.
 The FR layout is used.">>=
@@
 p1 <- ggplot(data = t1_1) +
-  geom_net(aes(from_id = from, to_id = to, linewidth= linewidth),
+  geom_net(aes(from_id = from, to_id = to, linewidth= linewidth, linetype = color2),
            directed = T, curvature = .2, labelon=T, size = 5,
-           arrow = arrow(angle = 20, type = "open", length = unit(.2, "cm")), ecolour = t1_1$color2,
+           arrow = arrow(angle = 20, type = "open", length = unit(.2, "cm")), # ecolour = t1_1$color2,
            color = "#252525", hjust = .5, vjust = .5, labelcolour = 'white', fontsize = 4, arrowgap = .02, layout.alg = "fruchtermanreingold",
            singletons = T) +
+  scale_linetype_manual(values =c("dotted", "solid") ) + 
   labs(title = "Difference between Wave 1 &\nWave 2 expectation from M1") + ThemeNet +
-  theme(plot.background = element_rect(color = 'black'))
+  theme(plot.background = element_rect(color = 'black'), legend.position = "none")
 
 pforleg <- ggplot(data = mtcars) + 
-  geom_line(aes(x = mpg, drat, group = as.factor(cyl), color = as.factor(cyl)), size = 1.5) + 
-  scale_color_manual(values = c("black", red, green), name = "Tie is...", labels = c("in data,\nin expectation", "in data,\nnot in expectation", "not in data,\nin expectation")) + 
+  geom_line(aes(x = mpg, drat, group = as.factor(cyl), linetype = as.factor(cyl))) + 
+  scale_linetype_manual(values = c("solid", "longdash", "dotted"), name = "Tie is...", labels = c("in data, in\nexpectation", "in data,not in\nexpectation", "not in data, in\nexpectation")) + 
   theme_grey() + 
@@
         legend.key.height = unit(25, "pt"))
-#pforleg
 
@@
 p2 <- ggplot(data = t1_2) +
-  geom_net(aes(from_id = from, to_id = to, linewidth= linewidth),
+  geom_net(aes(from_id = from, to_id = to, linewidth= linewidth, linetype = color2),
            directed = T, curvature = .2, labelon=T, size = 5,
-           arrow = arrow(angle = 20, type = "open", length = unit(.2, "cm")), ecolour = t1_2$color2,
+           arrow = arrow(angle = 20, type = "open", length = unit(.2, "cm")), #ecolour = t1_2$color2,
            color = "#252525", hjust = .5, vjust = .5, labelcolour = 'white', fontsize = 4, arrowgap = .02, layout.alg = "fruchtermanreingold",
            singletons = T) +
+  scale_linetype_manual(values = c("dotted", "longdash", "solid")) + 
   labs(title = "Difference between Wave 2 &\nWave 2 expectation from M1") + ThemeNet +
-  theme(plot.background = element_rect(color = 'black'))
+  theme(plot.background = element_rect(color = 'black'), legend.position = "none")
 legendsumm <- cowplot::get_legend(pforleg)
@@
 We created these networks by computing the difference between the adjacency matrices for the expected network and wave 1, and for wave 2 and the expected network.
-The lack of model fit becomes even more clear in Figure~\ref{fig:summary2}, with the XX NEED TO GO BACK AND REDO LINETYPES green XX edges representing ties that only show up in the expected network, and the red edges representing ties in the data that the model misses completely.
+The lack of model fit becomes even more clear in Figure~\ref{fig:summary2}, with the dotted edges representing ties that only show up in the expected network, and the broken edges representing ties in the data that the model misses completely.
 Again, the widths of the edges demonstrate the frequency of the ties in the simulations.
 On the left in Figure~\ref{fig:summary2}, we see the difference between the first wave of data and the expected network.
-We only see green and black edges, meaning that the simulated networks consist primarily of edges in wave 1, plus some new edges that are not in the first wave of data.
-The additional green edges are all reciprocating black edges, meaning that the model predicts more reciprocity in ties than there actually is.
+We only see dotted and solid edges, meaning that the simulated networks consist primarily of edges in wave 1, plus some new edges that are not in the first wave of data.
+The additional edges are all reciprocating edges in the data, meaning that the model predicts more reciprocity in ties than there actually is.
 On the right in Figure~\ref{fig:summary2}, we see the difference between the second wave of data and the expected network.
 As the model is simulating the second wave of data, we would hope to see few differences between the expected network and wave 2.
-However, there are six red edges, which are edges in the data that do not appear at all in the expected network, and there are several green edges that are in the expected network, but not in the data.
+However, there are six broken edges, which are edges in the data that do not appear at all in the expected network, and there are a great many dotted edges that are in the expected network, but not in the data.
 Although there are some edges that appear in both the expected network and wave 2, the majority of edges in the expected network are not in wave 2, and there are several missing edges.
@@
 We chose these four collections because they each explore something different about SAOM behavior: the choice of parameters in the objective function, the effects of underlying simulations, and the resulting fitted values under different circumstances.
-Using the tools in our visualization toolbox, we %The first considers the many parameters to include in a SAOM, while the second results in different parameter estimates, so we can see how parameter estimates vary.
+%Using the tools in our visualization toolbox, we view these collections to learn about model behavior.%The first considers the many parameters to include in a SAOM, while the second results in different parameter estimates, so we can see how parameter estimates vary.
 The third collection shows how those many possible parameters affect the values of parameters in the fitted models.
@@
 
-We have used novel visualization methods in order to better understand the family of models known as stochastic actor-oriented models for social network data.
-By viewing the model in the data space, visualizing collections of  models, and looking at the underlying algorithms, we have been able to gain knowledge and appreciation for these complicated models and everything that goes into them.
-When we looked at the model in the data space by viewing the average network simulated from a model, we were able to see the lack of model fit.
-When visualizing collections of models, we uncovered concerning relationships between parameters.
-Finally, when we explore the algorithms, we gain a better understanding of the model and its behavior.
-
-We have only just begun to scratch the surface of these complicated and multi-layered models for social networks.
-The \texttt{RSiena} software is incredibly powerful, and can fit a whole slew of much more flexible stochastic actor-oriented models than we have examined here.
-If, for example, a researcher thinks the network structure or an actor covariate effects the rate of change of the network, there is a way to incorporate that belief into the rate function of the SAOM.
+We have used novel visualization methods to explore the family of models known as stochastic actor-oriented models for dynamic social network data.
+By viewing the model in the data space, visualizing collections of models, and looking at the underlying algorithms, we have been able to gain knowledge and appreciation for these complicated and flexible models.
+When we looked at the model in the data space by viewing the expected network simulated from a model, we were able to easily see the lack of model fit.
+When visualizing collections of models, we quickly uncovered concerning relationships between parameters.
+When we explored the algorithms, we gained a better understanding of the model and its stochastic behavior.
+% Visualizing the CTMC can provide researchers with more insight into the underlying features of the model.
+In addition, SAOMs can include a great deal of parameters in the model structure, each of which is attached to a network statistic.
+These statistics are often somewhat, if not highly, correlated, which causes high correlation between the corresponding parameters in a SAOM.
+By visualizing collections of SAOMs, we gain a better understanding of these correlations and can find ways to account for their effects in the model.
+Furthermore, the estimation of the parameters in a SAOM rely on convergence checks based on simulations from the fitted model.
+These simulations are hidden and unsaved by default for computational efficiency.
+With the help of both static and dynamic visualizations, we bring the hidden model fitting processes into the foreground, eventually leading to a better understanding and higher accessibility of stochastic actor-oriented models for social network analysts.
+We have only just begun to scratch the surface of these multi-layered models for social networks.
+The \texttt{RSiena} software is incredibly powerful, and can fit much more flexible stochastic actor-oriented models than we have examined here.
+If, for example, a researcher thinks the network structure or an actor covariate affects the rate of change of the network, there is a way to incorporate that belief into the rate function of the SAOM.
 In addition, more than one actor-level covariate can be included in the model, and many more than three parameters can be included in the objective function itself.
-The flexibility of \texttt{RSiena} allows the user to specify which parameters lead to tie creation, and which parameters lead to tie dissolution.
+As we have shown, the rate and objective function parameters are often highly correlated, so researchers can now use model-vis techniques to gain a better understanding of these correlations and hopefully find ways to account for their effects in the model.
+In addition, SAOMs can model behavior change of the actors in the network, a capability we did not explore here.
+
+The data sets we have used are relatively small by network data standards, with only 16 and 100 nodes.
+We are aware that some of the visualizations we have presented may not scale up to larger networks.
+
+%The flexibility of \texttt{RSiena} allows the user to specify which parameters lead to tie creation, and which parameters lead to tie dissolution.
 We have used ``evaluation" parameters, which are agnostic, allowing the data to decide which parameters affect creation or dissolution via the sign \citep{RSienamanual}.
@@
 
-\section*{Appendix: Additional visualizations}
+\section*{Appendix: Additional model visualizations}
 
@@
   geom_net(aes(from_id = from, to_id = to, linewidth= linewidth),
-           directed = T, curvature = .2, labelon=T, size = 5,
+           directed = T, curvature = .2, labelon=T, size = 4,
            arrow = arrow(angle = 20, type = "open", length = unit(.2, "cm")), ecolour = "#969696", fiteach = T,
-           color = "#252525", hjust = .5, vjust = .5, labelcolour = 'white', fontsize = 4, arrowgap = .02,
+           color = "#252525", hjust = .5, vjust = .5, labelcolour = 'white', fontsize = 3, arrowgap = .02,
            singletons = T) +
@@
 simu_spread <- simu2 %>% spread(parameter, estimate)
-ggpairs(simu_spread, columns = 3:8, ggplot2::aes(colour=Model, alpha = .5)) + theme(plot.title = element_text(size = 12,
+ggpairs(simu_spread, columns = 3:8, ggplot2::aes(colour=Model, alpha = .5)) + theme(plot.title = element_text(size = 10,
                                             face = 'plain', angle = 0, family = "Times New Roman"),
-                  axis.title.x = element_text(size = 12,
+                  axis.title.x = element_text(size = 10,
                                             face = 'plain', angle = 0, family = "Times New Roman", inherit.blank = T),
-                  axis.title.y = element_text(size = 12,
+                  axis.title.y = element_text(size = 10,
                                             face = 'plain', angle = 90, inherit.blank = T),
-                  axis.text.x= element_text(size = 12,
+                  axis.text.x= element_text(size = 10,
                                             face = 'plain', angle = 90, inherit.blank = TRUE),
-                 # axis.text.x.bottom = element_text(size = 12,
+                 # axis.text.x.bottom = element_text(size = 10,
              #                               face = 'plain', angle = 90, inherit.blank = TRUE),
-                  axis.text.y = element_text(size = 12,
+                  axis.text.y = element_text(size = 10,
                                             face = 'plain', angle = 0, family = "Times New Roman", inherit.blank = T),
@@
                #                             face = 'plain', angle = 0, family = "Times New Roman", inherit.blank = T),
-                  strip.text.x = element_text(size = 14,
+                  strip.text.x = element_text(size = 12,
                                             face = 'plain', angle = 0, family = "Times New Roman",
                                             margin = margin(t = 3, r = 0, b = 3, l = 0, unit = "pt"), inherit.blank = T),
-                  strip.text.y =element_text(size = 14,
+                  strip.text.y =element_text(size = 12,
                                             face = 'plain', family = "Times New Roman", angle = 90,margin = margin(t = 3, r = 3, b = 3, l = 3, unit = "pt"),inherit.blank = T),

### 2018-10-03 paper.Rnw
change title, change plots to be grayscale safe, i hate writing conclusions.

@@
   \begin{center}
-    {\LARGE\bf Model Visualization Techniques for a Social Network Model}
+    {\LARGE\bf Visualizing Dynamic Social Network Models}
 \end{center}
@@
 Stochastic actor-oriented models (SAOMs), first introduced by Snijders (1996), are a type of statistical network model for dynamic social networks.
-Unlike other network models, SAOMs are analytically intractable, and therefore hard to interpret.
+Unlike other network models, SAOMs are analytically intractable, and require extensive simulation procedures for model fitting.
 We use model visualization techniques introduced in Wickham et al (2015) better understand these models and their behavior.

### 2018-10-03 paper-child.Rnw
change all plots to grayscale (minus ggpairs because it won't do it i guess?)

@@
   ThemeNoNet +
+  scale_fill_brewer(palette = "Greys") + 
   theme(legend.position = 'bottom')
@@
 smfr_effs %>% filter(type == "eval" & Waldpval <=0.10) %>%
-ggplot(aes(x = estimate, y = Waldpval, color = inter1)) +
+ggplot(aes(x = estimate, y = Waldpval, shape = inter1)) +
   #geom_errorbarh(aes(xmin = estimate - se, xmax = estimate + se), color = 'black') +
   geom_vline(xintercept = 0) +
-  geom_point(size = 3) +
+  geom_point(size = 3, alpha = .75) +
   #geom_text_repel(aes(label = shortName)) +
-  scale_color_brewer(palette = "Paired", name = "Interaction\nVariable") +
+  scale_shape(name = "Interaction\nVariable") +
   ggtitle("Significant effects at level 0.10 for friendship data") +
@@
 sen_effs$inter1[is.na(sen_effs$inter1)] <- "none"
-ggplot(data = sen_effs %>% filter(Waldpval <=0.10), aes(x = estimate, y = Waldpval, color = inter1)) +
+ggplot(data = sen_effs %>% filter(Waldpval <=0.10), aes(x = estimate, y = Waldpval, shape= inter1)) +
   # geom_errorbarh(aes(xmin = estimate - se, xmax = estimate + se), color = 'black') +
@@
   #geom_text_repel(aes(label = shortName)) +
-  scale_color_brewer(palette = "Paired", name = "Interaction\nVariable") +
+  scale_shape(name = "Interaction\nVariable") +
   ggtitle("Significant effects at level 0.10 for senate data") +
@@
   geom_net(aes(from_id = from, to_id = to, linewidth= linewidth),
-           directed = T, curvature = .2, labelon=T, size = 4,
+           directed = T, curvature = .2, labelon=T, size = 3,
            arrow = arrow(angle = 20, type = "open", length = unit(.2, "cm")), ecolour = "#969696", fiteach = T,
-           color = "#252525", hjust = .5, vjust = .5, labelcolour = 'white', fontsize = 3, arrowgap = .02,
+           color = "#252525", hjust = .5, vjust = .5, labelcolour = 'white', fontsize = 2, arrowgap = .02,
            singletons = T) +
@@
   facet_grid(data~parameter, scales = "free", labeller = "label_both") +
+  scale_fill_brewer(palette = "Greys") + 
   ThemeNoNet +
@@
 simu_spread <- simu2 %>% spread(parameter, estimate)
-ggpairs(simu_spread, columns = 3:8, ggplot2::aes(colour=Model, alpha = .5)) + theme(plot.title = element_text(size = 10,
+ggpairs(simu_spread, columns = 3:8, aes(colour=Model, alpha = .5)) + 
+  theme(plot.title = element_text(size = 10,
                                             face = 'plain', angle = 0, family = "Times New Roman"),
@@
   #scale_fill_gradient(low = 'grey90', high = '#0868ac', name = "occurrence %") +
-  scale_fill_brewer(palette = "Blues", name = "Occurrence %", labels = c("(0-1]%", "(1-5]%", "(5-10]%","(10-25]%", "(25-50]%", "(50-75]%", "(75-100]%")) +
+  scale_fill_brewer(palette = "Greys", name = "Occurrence %", labels = c("(0-1]%", "(1-5]%", "(5-10]%","(10-25]%", "(25-50]%", "(50-75]%", "(75-100]%")) +
   labs(x = "Microstep No.", y = "Edges (ordered by total # occurrences)") +

### 2018-10-08 paper.Rnw
add HH changes to abstract. not sure abt 2 of them.

@@
 {
-  \title{\bf Model Visualization Techniques for a Social Network Model}
+  \title{\bf Visualizing Dynamic Social Network Models}
   \author{Sam Tyner \hspace{.2cm}\\
@@
 Stochastic actor-oriented models (SAOMs), first introduced by Snijders (1996), are a type of statistical network model for dynamic social networks.
-Unlike other network models, SAOMs are analytically intractable, and require extensive simulation procedures for model fitting.
-We use model visualization techniques introduced in Wickham et al (2015) better understand these models and their behavior.
-We use a three-pronged attack: (a) a network visualization to view the \emph{model}, (b) exploring collections of models for parameter interpretation, and (c) delving into the fitting algorithms to gain insight into the underlying mechanisms.
-With the help of static and dynamic visualizations, we bring the hidden SAOM fitting processes into the foreground, eventually leading to a better understanding and higher accessibility of SAOMs for social network analysts.
+Unlike other network models, SAOMs are analytically intractable and require extensive simulation procedures for model fitting.
+We use model visualization techniques introduced in Wickham et al (2015) to better understand these models and their behavior.
+We use a three-pronged attack: (a) network visualization to view the \emph{model}, (b) exploring collections of models for parameter interpretation, and (c) delving into the fitting algorithms to gain insight into the underlying mechanisms.
+With the help of static and dynamic visualizations, we bring the hidden SAOM fitting processes into the foreground, with the goal of a better understanding and higher accessibility of SAOMs by social network analysts.
 \end{abstract}

### 2018-10-08 paper-child.Rnw
make some changes. stop on pg 20.

@@
 These relationships, or \emph{social networks}, have been studied for decades, in every format from self-help books to publications in academic journals.
-There are even statistical models, such as exponential random graph models (ERGM) and latent space models (LSM), that strive to capture the underlying relationship-generating mechanism \citep{ergm, lsm}.
+There are statistical models, such as exponential random graph models (ERGM) and latent space models (LSM), that strive to capture the underlying relationship-generating mechanism \citep{ergm, lsm}.
 These static models, however, are not optimal for modelling ``real-world" social networks because they do not account for changes in the network as time passes.
-\textit{Dynamic social networks} are more realistic because they contain multiple observations of the same network over time, and models for dynamic social networks hold a great deal of promise because of their potential to capture realistic mechanisms.
-Our social networks evolve over time as new relationships are made, and being able to model these evolutionary mechanisms can uncover hidden truths about our relationships with each other.
+Models for \textit{dynamic social networks} are more realistic because they contain multiple observations of the same network over time.
+%., and models for dynamic social networks hold a great deal of promise because of their potential to capture realistic mechanisms.
+%Our social networks evolve over time as new relationships are made, and being able to model these evolutionary mechanisms can uncover hidden truths about our relationships with each other.
 
@@
 Adding node-level information into the model is a more intuitive approach to model changes in a social network because, for example
-With these models, we can examine our assumptions that people with common interests are more likely to become friends.
+With these models, we can examine assumptionsm made about social networks, i.e.
+that people with common interests are more likely to become friends.
 For example, we may assume that high schoolers who drink heavily are more likely to be friends with others who drink heavily.
@@
 There are always units of observation and connections between those units.
-In a social network, the units of observations, called \textit{nodes} or \textit{actors}, are people, while the connections, called \textit{edges} or \textit{ties}, are their relationships.
+In a social network, the units of observations, called \textit{nodes} or \textit{actors}, are people, while the connections, called \textit{edges} or \textit{ties}, are their relationships with each other.
 %Networks change over time, like when new relationships are formed in a social network.
@@
 
-\noindent \textbf{Visualizing Network Data:} Network visualization, also called graph drawing, is a prominent subfield of network analysis \citep{graphdrawing}.
+\noindent \textbf{Visualizing Network Data:} Network visualization, also called graph drawing, is a prominent subfield of network analysis.
+For a review of graph drawing, see \citet{graphdrawing}.
 Visualizing network data is inherently challenging because of the structure of the data itself.
@@
 
-The lack of a native coordinate system has led to two very different methods to visualize networks: the node-link diagram and the adjacency matrix visualization \citep{knuth2000, adjmatpro}.
+The lack of a native coordinate system has led to two very different primary methods to visualize networks: the node-link diagram and the adjacency matrix visualization \citep{knuth2000, adjmatpro}.
 We demonstrate these with a simpe example.
@@
 With the objective function, we can define the probability that the ego node $i$ chooses to change its tie to alter node $j$, denoted by $p_{ij}(\boldsymbol{\beta}, x, \mathbf{Z})$.
-Let $x(i \leadsto j)$ be the network identical to $x$ with the exception of tie $x_{ij}$, which is equal to $1-x_{ij}$.
-The probility that the tie $x_{ij}$ changes is defined as:
+Let $x(i \leadsto j)$ be the network identical to network $x$ with the exception of tie $x_{ij}$, which is equal to $1-x_{ij}$; i.e.
+the presence or absence of node $x_{ij}$ is flipped in the new network.
+The probability that the tie $x_{ij}$ changes is defined as:
 \begin{equation}\label{eq:pij}
@@
 The probabilities $p_{ij}$ combined with the rate parameters $\alpha_m$ fully characterize the underlying CTMC that models network change.
-For more on CTMCs, see \citet{ctmc}.
+For more details on CTMCs, see e.g.
+\citet{ctmc}.
 % \subsubsection{Continuous Time Markov Chain (CTMC)}
@@
 
-To fit a SAOM to dynamic network data, we apply the \texttt{RSiena} software, which uses simulation methods to estimate parameter values using either MoM or maximum likelihood (ML) estimation \citep{RSiena}.
+To fit a SAOM to dynamic network data, we apply the \texttt{RSiena} software, which uses simulation methods to estimate parameter values using either method of moments (MoM) or maximum likelihood (ML) estimation \citep{RSiena}.
 We chose to use MoM estimation, so we describe that process here, while more information on ML estimation can be found in \citet{SienaAlgs}.
@@
 These are the mechanisms we hope to incorporate and model with SAOMs.", out.width='.99\\textwidth', fig.height=3, results='hide'>>=
+cols <- RColorBrewer::brewer.pal(8, "Greys")
+
 source("code/00e_small_friends.R")
@@
        aes(from_id = X1, to_id = X2, color = as.ordered(drink), group = as.ordered(drink))) +
-  geom_net(fiteach = FALSE, directed=TRUE, layout.alg = 'fruchtermanreingold', size = 5, arrowsize = .25, arrowgap = .05, ecolour = 'grey40', labelon=T, labelcolour = 'black', fontsize = 4, vjust = .5, linewidth=0.5) +
-  scale_color_brewer(palette = "Greys", name = "Drinking\nBehavior", labels = c("never", "1-2 times per year", "once per month", "once per week")) +
+  geom_net(fiteach = FALSE, directed=TRUE, layout.alg = 'fruchtermanreingold', size = 4, arrowsize = .25, arrowgap = .05, ecolour = 'grey40', labelon=T, labelcolour = 'white', fontsize = 2, vjust = .5, linewidth=0.5) +
+  scale_color_manual(values = cols[8:5], name = "Drinking\nBehavior", labels = c("never", "1-2 times per year", "once per month", "once per week")) +
   ThemeNet +
@@
 There are hundreds of ties between senators when they are connected in this way, so we simplify the network by computing a single value for each pair of senators called the \textit{weighted propensity to cosponsor} (WPC).
-This value is defined in \citet{senate} as
+This value is defined in \citet{senate} as the total number of times senator $i$ cosponsors senator $j$'s bill, weighted by the number of total cosponsors:
 \begin{equation}\label{eq:sen1}
@@
 \begin{equation}\label{eq:m1}
-f_i(x)^{M1} = \beta_1s_{i1}(x) + \beta_2s_{i2}(x),
+M2: \quad f_i(x) = \beta_1s_{i1}(x) + \beta_2s_{i2}(x),
 \end{equation}
@@
 \begin{align}\label{eq:m2m3}
-f_i(x,z)^{M2} & = \beta_1s_{i1}(x) + \beta_2s_{i2}(x) + \beta_3s_{i3}(x, z) \\
-f_i(x)^{M3} & = \beta_1s_{i1}(x) + \beta_2s_{i2}(x) + \beta_4s_{i4}(x),
+M2: \quad f_i(x,z) & = \beta_1s_{i1}(x) + \beta_2s_{i2}(x) + \beta_3s_{i3}(x, z) \\
+M3: \quad f_i(x) & = \beta_1s_{i1}(x) + \beta_2s_{i2}(x) + \beta_4s_{i4}(x),
 \end{align}
 
-\begin{figure}\label{fig:structures}
+\begin{figure}
 \begin{subfigure}[t]{.45\textwidth}
@@
   ThemeNet +
-  theme(legend.position = "none", plot.title = element_text(size = rel(0.9))) + 
+  theme(legend.position = "none", 
+        plot.title = element_text(size = rel(1.1))) + 
   labs(title = expression(paste(s[i3], "(x,z)", "=",Sigma[j != h], x[ij], x[ih], x[hj] %*% 
@@
   ThemeNet +
-  theme(legend.position = "none") + 
+  theme(legend.position = "none", 
+        plot.title = element_text(size = rel(1.1))) + 
   labs(title = expression(paste(s[i4], "(x)", "=", "|{j:", x[ij],"=0, ", Sigma[h], x[ih], x[hj] >= 2, "}|" )
@@
 \end{subfigure}
-\caption{The additional network effects included in our models fit to the friends data.
+\caption{\label{fig:structures} The additional network effects included in our models fit to the friends data.
 On the left, a JTT between $i$ and $j$, who have different covariate values, and on the right, a N22 between $i$ and $j$.}
@@
 %In Figure~\ref{fig:summnetM1}, we show the \textit{expected network} of wave 2 from model M1, created with 1,000 simulations from the model. 
-To construct an expected network, we first simulated 1,000 wave 2 and wave 3 instances of our small friendship network from model M1, for which parameters had previously been estimated.
-We then counted the number of times each edge $i \rightarrow j$ appeared in one of the 1,000 network simulations.
-From these counts, we constructed one summary network for each time point, which has edge weights equal to the proportion of the simulated networks in which the edge appeared.
-Finally, we created the \textit{expected network} by including only the edges that appeared in more than 5\% of the simulations.
+To construct an expected network, we: 
+\begin{enumerate}
+\item simulate 1,000 wave 2 and wave 3 instances of our small friendship network from model M1 using parameters that have previously been estimated. Then, we 
+\item count the number of times each edge $i \rightarrow j$ appeared in one of the 1,000 network simulations. Then, we
+\item construct one weighted summary network for each time point, which has edge weights equal to the proportion of the simulated networks in which the edge appeared. Finally, we 
+\item creat the \textit{expected network} by including only the edges that appeared in more than 5\% of the simulations. 
+\end{enumerate}
 The resulting expected network is in the middle panel of Figure~\ref{fig:summnetM1}.
@@
 Because the expected network does not resemble wave 2, we take this as evidence that the simple model, M1, does not capture the change in network structure from the first to the second wave of data.
-Visualizing the expected network in this way places the model M1 in the data space, and has helped us assess model fit.
+Visualizing the expected network in this way places the model M1 in the data space, and helps us assess model fit.
 % We can use the average network to help determine model goodness-of-fit.
@@
 To select objective function parameters for fitting models to our example data, we searched through all possible effects available to find \textit{significant} effects.
-The \texttt{RSiena} package includes the function \texttt{Wald.Rsiena()} for Wald-type tests of parameter significance.
+The \texttt{RSiena} package includes the function \texttt{Wald.Rsiena} for Wald-type tests of parameter significance.
 We start by including the outdegree and reciprocity parameters in the model then add one effect to the objective function, fit the model to the example data, test the added effect for significance, and repeat for all possible model parameters.

### 2018-10-10 paper-child.Rnw
add the pdf
incorporate most of hh changes. still need to do fig 7

@@
 \begin{equation}\label{eq:m1}
-M2: \quad f_i(x) = \beta_1s_{i1}(x) + \beta_2s_{i2}(x),
+M1: \quad f_i(x) = \beta_1s_{i1}(x) + \beta_2s_{i2}(x),
 \end{equation}
@@
 simu2 <- read.csv("data/simulation-1000-M1-M2-M3.csv")
-
+levels(simu2$parameter) <- c("alpha[1]","alpha[2]","beta[1]","beta[2]","beta[3]","beta[4]")
 ggplot(data = simu2) +
   geom_density(aes(x = estimate, fill = Model), alpha = .5) +
-  facet_wrap(~parameter, scales = 'free') +
+  facet_wrap(~parameter, scales = 'free', labeller= label_parsed) +
   ThemeNoNet +
   scale_fill_brewer(palette = "Greys") + 
+  xlab("estimate value") + 
   theme(legend.position = 'bottom')
@@
 }
-\caption{\label{tab:meanests} The means (std. dev.) of parameter values estimated from repeated fittings of $M1, M2, M3$ to the two example data sets.}
+\caption{\label{tab:meanests} The means (std.\ dev.) of parameter values estimated from repeated fittings of $M1, M2, M3$ to the two example data sets.}
 \end{table}
@@
 % We have seen how the same model forms behave in different context, with some  differences and some similarities. 
-After seeing the high correlation between parameter estimates in very different data contexts, we question the guideline that both parameters need to be included in the objective function.
+After seeing the high correlation between parameter estimates in very different data contexts, we might question the guideline that both parameters need to be included in the objective function.
 
@@
 This process is known as matrix seriation
-To seriate the matrix, we first constructed a cumulative adjaceny matrix, $\mathbf{A}^{cum}$, for the series of microsteps simulating the network from $x(t_m)$ to $x(t_{m+1})$.
+To seriate the matrix, we first construct a cumulative adjaceny matrix, $\mathbf{A}^{cum}$, for the series of microsteps simulating the network from $x(t_m)$ to $x(t_{m+1})$.
 A single entry in the cumulative adjacency matrix, $\mathbf{A}^{cum}_{ij}$, is the total number of times the edge $i \rightarrow j$ appears in the network from the initial observation, $x(t_m) \equiv X(0)$ to the final result of the last microstep, $X(R)$, where $R$ is the total number of microsteps taken:
@@
 \end{equation}
-We then performed a principal component analysis (PCA) on $\mathbf{A}^{cum}$, and used the values of the first principal component to order the vertices on the $x$ and $y$ axes for the adjacency matrix animation.
+We then perform a principal component analysis (PCA) on $\mathbf{A}^{cum}$, and use the values of the first principal component to order the vertices on the $x$ and $y$ axes for the adjacency matrix animation.
 The resulting visualization for the first wave of data is shown in Figure~\ref{fig:adjmatorder}.
@@
 Using PCA on $\mathbf{A}^{cum}$ to order the rows and columns of the adjacency matrix visualization clearly shows the two distinct connected components in the first wave of the network, which are difficult to find in the visualization ordered by node ID.
-We continue to use the seriated matrix in another animation of the microstep process, which is avaiable at \url{https://vimeo.com/240092677}.
+We continue to use the seriated matrix in another animation of the microstep process, which is available at \url{https://vimeo.com/240092677}.
 A few frames of the video are shown in Figure~\ref{fig:adjmatani}.
@@
 <<adjmatani, fig.cap='A selection of frames from the adjacency matrix visualization animation for one series of microsteps.
-The node lables are removed to declutter the figure.
+The node labels are removed to declutter the figure.
 Starting at frame 0, there are two clearly connected components: one in the bottom left corner, and one in the top right corner.
@@
 adjmatall3$frame <- adjmatall3$ms
-#plots <- list()
-#for(i in 0:55){
-#plots[[i + 1]] <-
 
@@
 dat5 <- dat4
-#p <- Sys.time()
 for(i in 4:16){
@@
         legend.text = element_text(size = rel(.75)), 
-        legend.title = element_text(size = rel(.75))) +
+        legend.title = element_text(size = rel(.75)), 
+        legend.position = "bottom") +
   coord_fixed() +
@@
 The ego node is on the y-axis, and the alter node is on the x-axis.
-There were many ties with empirical probability zero.', fig.height=2.5>>=
+There are many ties with empirical probability zero.', fig.height=2.5>>=
 
@@
   geom_tile(color = 'black') +
-  scale_fill_gradient(low = 'white', high = 'black', limits = c(0,1)) +
+  scale_fill_gradient(low = 'white', high = 'black', limits = c(0,1), name = "theoretical\nprobability") +
   labs(y = "ego", x = "alter")+
@@
   ThemeNoNet + 
-  theme(panel.grid = element_blank(), axis.text = element_text(size=rel(.8))) +
+  theme(panel.grid = element_blank(), axis.text = element_text(size=rel(.7)),
+        legend.position = "bottom", legend.title = element_text(size=rel(.7))) +
   coord_fixed() + 
@@
   geom_tile(color = 'black') +
-  scale_fill_gradient(low = 'white', high = 'black', name = "prob") +
+  scale_fill_gradient(low = 'white', high = 'black', name = "empirical\nprobability") +
   labs(y = "ego", x = "alter")+
@@
   ThemeNoNet + 
-  theme(panel.grid = element_blank(), axis.text = element_text(size=rel(.8))) +
+  theme(panel.grid = element_blank(), axis.text = element_text(size=rel(.7)),
+        legend.position = 'bottom', legend.title = element_text(size=rel(.7))) +
   coord_fixed() + 
@@
 
-We have used novel visualization methods to explore the family of models known as stochastic actor-oriented models for dynamic social network data.
+We have used traditional network visualization methods to explore the family of models known as stochastic actor-oriented models for dynamic social network data in new ways.
 By viewing the model in the data space, visualizing collections of models, and looking at the underlying algorithms, we have been able to gain knowledge and appreciation for these complicated and flexible models.
-When we looked at the model in the data space by viewing the expected network simulated from a model, we were able to easily see the lack of model fit.
-When visualizing collections of models, we quickly uncovered concerning relationships between parameters.
-When we explored the algorithms, we gained a better understanding of the model and its stochastic behavior.
+When we look at the model in the data space by viewing the expected network simulated from a model, we are able to easily see the lack of model fit.
+When visualizing collections of models, we uncover concerning relationships between parameters.
+When we explore the algorithms, we gain a better understanding of the model and its stochastic behavior.
 % Visualizing the CTMC can provide researchers with more insight into the underlying features of the model.

### 2018-10-10 paper.Rnw
incorporate most of hh changes. still need to do fig 7

@@
 We use model visualization techniques introduced in Wickham et al (2015) to better understand these models and their behavior.
-We use a three-pronged attack: (a) network visualization to view the \emph{model}, (b) exploring collections of models for parameter interpretation, and (c) delving into the fitting algorithms to gain insight into the underlying mechanisms.
+We use a three-pronged visualization attack: (a) network visualizations to view the \emph{model}, (b) collections of models to interpret parameters, and (c) visualizations of fitting algorithms to gain insights into the underlying mechanisms.
 With the help of static and dynamic visualizations, we bring the hidden SAOM fitting processes into the foreground, with the goal of a better understanding and higher accessibility of SAOMs by social network analysts.

### 2018-10-11 paper-child.Rnw
add prettier conclusion

@@
 These static models, however, are not optimal for modelling ``real-world" social networks because they do not account for changes in the network as time passes.
-Models for \textit{dynamic social networks} are more realistic because they contain multiple observations of the same network over time.
+Models for \textit{dynamic social networks} are more realistic because they provide mechanisms for analyzing multiple observations of the same network over time.
 %., and models for dynamic social networks hold a great deal of promise because of their potential to capture realistic mechanisms.
@@
        aes(from_id = X1, to_id = X2, color = as.ordered(drink), group = as.ordered(drink))) +
-  geom_net(fiteach = FALSE, directed=TRUE, layout.alg = 'fruchtermanreingold', size = 4, arrowsize = .25, arrowgap = .05, ecolour = 'grey40', labelon=T, labelcolour = 'white', fontsize = 2, vjust = .5, linewidth=0.5) +
+  geom_net(fiteach = FALSE, directed=TRUE, layout.alg = 'fruchtermanreingold', size = 4, arrowsize = .25, arrowgap = .05, ecolour = 'black', labelon=T, labelcolour = 'white', fontsize = 2, vjust = .5, linewidth=0.25) +
   scale_color_manual(values = cols[8:5], name = "Drinking\nBehavior", labels = c("never", "1-2 times per year", "once per month", "once per week")) +
@@
            linetype = c("solid", "solid","dotted", "solid")) +
-  expand_limits(x=c(-0.1,1.1), y=c(-0.1,1.1)) +
+  expand_limits(x=c(-0.2,1.2), y=c(-0.1,1.1)) +
   ThemeNet +
-  theme(legend.position = "none", 
-        plot.title = element_text(size = rel(1.1))) + 
-  labs(title = expression(paste(s[i3], "(x,z)", "=",Sigma[j != h], x[ij], x[ih], x[hj] %*% 
-                                I, "(", z[i], "=", z[h]!= z[j], ")")
-                          )
-       )
-@
-\caption{\label{fig:jtt}Realization of a JTT, where $i$ is the ego, $j$ is the alter, and $h$ is the intermediary. The covariate is represented by the shape of the nodes.}
+  theme(legend.position = "none") + 
+  labs(title = expression(paste(s[i3], "(x,z)")))
+  # labs(title = expression(paste(s[i3], "(x,z)", "=",Sigma[j != h], x[ij], x[ih], x[hj] %*% 
+  #                               I, "(", z[i], "=", z[h]!= z[j], ")")
+  #                         )
+  #      )
+@
+\caption{\label{fig:jtt}Realization of a JTT, where $i$ is the ego, $j$ is the alter, and $h$ is the intermediary.
+The covariate is represented by the shape of the nodes.
+$s_{i3}(x,z) = \sum_{j \neq h} x_{ij}x_{ih}x_{hj} \mathbb{I}(z_i = z_h \neq z_j)$}
 \end{subfigure}\hfill
@@
   ThemeNet +
-  theme(legend.position = "none", 
-        plot.title = element_text(size = rel(1.1))) + 
-  labs(title = expression(paste(s[i4], "(x)", "=", "|{j:", x[ij],"=0, ", Sigma[h], x[ih], x[hj] >= 2, "}|" )
-                          )
-       )
-@
-\caption{\label{fig:dad}Realization of a N22 between actors $i$ and $j$. The dotted tie is encouraged if $\beta_4$ has a positive value.}
+  theme(legend.position = "none") + 
+  labs(title = expression(paste(s[i4], "(x)")))
+  # labs(title = expression(paste(s[i4], "(x)", "=", "|{j:", x[ij],"=0, ", Sigma[h], x[ih], x[hj] >= 2, "}|" )
+  #                         )
+  #      )
+@
+\caption{\label{fig:dad}Realization of a N22 between actors $i$ and $j$.
+The dotted tie is encouraged if $\beta_4$ has a positive value.
+$s_{i4}(x) =  |\{j : x_{ij} = 0, \sum\limits_h x_{ih}x_{hj} \geq 2\}|$}
 \end{subfigure}
@@
   scale_shape_manual(values = c(17,16), guide = guide_legend(nrow= 2)) +
-  scale_color_manual(values = c("black", "grey30","#ef3b2c"), guide = guide_legend(nrow=2)) +
+  scale_color_manual(values = c("black", "grey80","#ef3b2c"), guide = guide_legend(nrow=2)) +
   facet_wrap(~Clinton, nrow = 1, labeller = 'label_both')  +
@@
 %There is no defined expected value of a network model, though expectations exist for most other statistical models. 
-
 %For network data, we view an ``average" network through a sort of summary network drawn using the traditional node-link diagram. 
@@
 \item construct one weighted summary network for each time point, which has edge weights equal to the proportion of the simulated networks in which the edge appeared. Finally, we 
-\item creat the \textit{expected network} by including only the edges that appeared in more than 5\% of the simulations. 
+\item create the \textit{expected network} by including only the edges that appeared in more than 5\% of the simulations. 
 \end{enumerate}
@@
            singletons = T) +
-  scale_linetype_manual(values =c("dotted", "solid") ) + 
+  scale_linetype_manual(values =c("longdash", "dotted") ) + 
   labs(title = "Difference between Wave 1 &\nWave 2 expectation from M1") + ThemeNet +
@@
   geom_line(aes(x = mpg, drat, group = as.factor(cyl), linetype = as.factor(cyl))) + 
-  scale_linetype_manual(values = c("solid", "longdash", "dotted"), name = "Tie is...", labels = c("in data, in\nexpectation", "in data,not in\nexpectation", "not in data, in\nexpectation")) + 
+  scale_linetype_manual(values = c("dotted", "solid", "longdash"), name = "Tie is...", labels = c("in data, in\nexpectation", "in data,not in\nexpectation", "not in data, in\nexpectation")) + 
   theme_grey() + 
@@
            singletons = T) +
-  scale_linetype_manual(values = c("dotted", "longdash", "solid")) + 
+  scale_linetype_manual(values = c("longdash", "solid", "dotted")) + 
   labs(title = "Difference between Wave 2 &\nWave 2 expectation from M1") + ThemeNet +
@@
 We created these networks by computing the difference between the adjacency matrices for the expected network and wave 1, and for wave 2 and the expected network.
-The lack of model fit becomes even more clear in Figure~\ref{fig:summary2}, with the dotted edges representing ties that only show up in the expected network, and the broken edges representing ties in the data that the model misses completely.
+The lack of model fit becomes even more clear in Figure~\ref{fig:summary2}, with the broken edges representing ties that only show up in the expected network, and the solid edges representing ties in the data that the model misses completely.
+Dotted edges are in both the expectation and the data and represent where the model fit the data well.
 Again, the widths of the edges demonstrate the frequency of the ties in the simulations.
 On the left in Figure~\ref{fig:summary2}, we see the difference between the first wave of data and the expected network.
-We only see dotted and solid edges, meaning that the simulated networks consist primarily of edges in wave 1, plus some new edges that are not in the first wave of data.
-The additional edges are all reciprocating edges in the data, meaning that the model predicts more reciprocity in ties than there actually is.
+We only see dotted and broken edges, meaning that the simulated networks consist primarily of edges in wave 1, plus some new edges that are not in the first wave of data.
+The added ties are all reciprocating edges in the data, meaning that the model predicts more reciprocity between actors than there actually is.
 On the right in Figure~\ref{fig:summary2}, we see the difference between the second wave of data and the expected network.
-As the model is simulating the second wave of data, we would hope to see few differences between the expected network and wave 2.
-However, there are six broken edges, which are edges in the data that do not appear at all in the expected network, and there are a great many dotted edges that are in the expected network, but not in the data.
-Although there are some edges that appear in both the expected network and wave 2, the majority of edges in the expected network are not in wave 2, and there are several missing edges.
+Because the model is simulating the second wave of data, we would hope to see very few differences between the expected network and wave 2.
+However, there are six solid edges, which are edges in the data that do not appear at all in the expected network, and there are many broken edges that appear in the expected network but not in the data.
+Although there are some edges that appear in both the expected network and wave 2, the majority of edges in the expected network are not in wave 2, and there are several edges missing from the expectation.
 We conclude that the model M1 is a poor fit to our example data, which we have seen very clearly by viewing the model in the data space.
@@
 }
-\caption{\label{tab:meanests} The means (std.\ dev.) of parameter values estimated from repeated fittings of $M1, M2, M3$ to the two example data sets.}
+\caption{\label{tab:meanests} The means (std.\ dev.)\ of parameter values estimated from repeated fittings of $M1, M2, M3$ to the two example data sets.}
 \end{table}
@@
 Starting at frame 0, there are two clearly connected components: one in the bottom left corner, and one in the top right corner.
-By the end, the component in the bottom left has spread out and become sparsely connected, while the top right component has shrunk, but remains closely connected.', warning = FALSE, message = FALSE, results='hide', fig.height=2.5>>=
+By the end, the component in the bottom left has spread out and become sparsely connected, while the top right component has shrunk, but remains closely connected.', warning = FALSE, message = FALSE, results='hide', fig.height=3, out.width="\\linewidth">>=
 adjmatall <- matrix(0, nrow = 16, ncol = 16)
@@
   geom_tile(color = 'black') +
-  scale_fill_gradient(low = 'white', high = 'black', limits = c(0,1), name = "theoretical\nprobability") +
+  scale_fill_gradient(low = 'white', high = 'black', limits = c(0,1), name = "theoretical\nprobability", labels = c("0","",".5","","1")) +
   labs(y = "ego", x = "alter")+
@@
   ThemeNoNet + 
-  theme(panel.grid = element_blank(), axis.text = element_text(size=rel(.7)),
-        legend.position = "bottom", legend.title = element_text(size=rel(.7))) +
+  theme(panel.grid = element_blank(), 
+          axis.text.x.top = element_text(size=rel(.7)),
+          axis.text.y.left = element_text(size=rel(.7)),
+          legend.position = "bottom", 
+          legend.title = element_text(size=rel(.7)), 
+          legend.text = element_text(size=rel(.7)), 
+          legend.key.height = unit(.5, "cm"), 
+          legend.margin=margin(0,0,0,0),
+          legend.box.margin=margin(-10,-10,-5,-10)) +
   coord_fixed() + 
@@
   geom_tile(color = 'black') +
-  scale_fill_gradient(low = 'white', high = 'black', name = "empirical\nprobability") +
+  scale_fill_gradient(low = 'white', high = 'black', name = "empirical\nprobability", labels = c(".02", ".04", ".06"), breaks = c(.02, .04, .06)) +
   labs(y = "ego", x = "alter")+
@@
   ThemeNoNet + 
-  theme(panel.grid = element_blank(), axis.text = element_text(size=rel(.7)),
-        legend.position = 'bottom', legend.title = element_text(size=rel(.7))) +
+    theme(panel.grid = element_blank(), 
+          axis.text.x.top = element_text(size=rel(.7)),
+          axis.text.y.left = element_text(size=rel(.7)),
+          legend.position = "bottom", 
+          legend.title = element_text(size=rel(.7)), 
+          legend.text = element_text(size=rel(.7)), 
+          legend.key.height = unit(.5, "cm"), 
+          legend.margin=margin(0,0,0,0),
+          legend.box.margin=margin(-10,-10,-5,-10)) +
   coord_fixed() + 
   ggtitle("Empirical Probabilities")
-gridExtra::grid.arrange(p1, p2, nrow = 1)
+plot_grid(p1, p2)
+#gridExtra::grid.arrange(p1, p2, nrow = 1)
 @
@@
 
-We have used traditional network visualization methods to explore the family of models known as stochastic actor-oriented models for dynamic social network data in new ways.
-By viewing the model in the data space, visualizing collections of models, and looking at the underlying algorithms, we have been able to gain knowledge and appreciation for these complicated and flexible models.
+We have used traditional network visualization methods to explore the family of stochastic actor-oriented models for dynamic social network data in new ways.
+The guiding principles of model-vis: viewing the model in the data space, visualizing collections of models, and looking at the underlying algorithms, gave us knowledge and appreciation for these complicated and flexible models.
 When we look at the model in the data space by viewing the expected network simulated from a model, we are able to easily see the lack of model fit.
@@
 When we explore the algorithms, we gain a better understanding of the model and its stochastic behavior.
-% Visualizing the CTMC can provide researchers with more insight into the underlying features of the model.
-In addition, SAOMs can include a great deal of parameters in the model structure, each of which is attached to a network statistic.
-These statistics are often somewhat, if not highly, correlated, which causes high correlation between the corresponding parameters in a SAOM.
-By visualizing collections of SAOMs, we gain a better understanding of these correlations and can find ways to account for their effects in the model.
-Furthermore, the estimation of the parameters in a SAOM rely on convergence checks based on simulations from the fitted model.
-These simulations are hidden and unsaved by default for computational efficiency.
-With the help of both static and dynamic visualizations, we bring the hidden model fitting processes into the foreground, eventually leading to a better understanding and higher accessibility of stochastic actor-oriented models for social network analysts.
+Our three-pronged, model-vis attack has uncovered many properties of SAOMs.
+
 We have only just begun to scratch the surface of these multi-layered models for social networks.
-The \texttt{RSiena} software is incredibly powerful, and can fit much more flexible stochastic actor-oriented models than we have examined here.
+The \texttt{RSiena} software is incredibly powerful, and can fit much more flexible SAOMs than we have examined here.
 If, for example, a researcher thinks the network structure or an actor covariate affects the rate of change of the network, there is a way to incorporate that belief into the rate function of the SAOM.
@@
 
-The data sets we have used are relatively small by network data standards, with only 16 and 100 nodes.
-We are aware that some of the visualizations we have presented may not scale up to larger networks.
+Furthermore, the data sets we have used are relatively small by network data standards, with only 16 and 100 nodes.
+Thus, some of the visualizations we have presented may not scale up to larger networks.
+Future SAOM and other network model visualization research could include interactive and/or three-dimensional graphics.
+Interactivity is the state-of-the-art in data visualization, and network visualization is just starting to catch on \citep[see e.g.\ ][]{refinery}.
+\citet{modelvis} relied on interactive tools like GGobi \citep{STLBC03} for interactivity, and for networks, there is some interactive support with the package \texttt{plotly} \citep{plotly} built into \texttt{geomnet}.
+
+We hope that the principles of model-vis continue to be used for learning more about the fit and behavior of complicated statistical models.
+Our model-vis application has uncovered many properties of SAOMs, and potential abounds for use in other areas of network study.
 
@@
 
-\noindent \textbf{Online supplemental material:} Much of the code used to create these plots has been packaged an posted online at https://github.com/sctyner/netvizinf and https://github.com/sctyner/geomnet (also on CRAN).
-Additional, non-packaged code can be found at https://github.com/sctyner/SAOM-removing-blindfold.
+\noindent \textbf{Online supplemental material:} Much of the code used to create the visualizations in this paper has been packaged and made available online at \url{https://github.com/sctyner/netvizinf} and \url{https://github.com/sctyner/geomnet} (also on CRAN).
+Additional, non-packaged code can be found at \url{https://github.com/sctyner/SAOM-removing-blindfold}.

### 2018-10-15 paper-child.Rnw
final proof-reading tweaks

@@
 These relationships, or \emph{social networks}, have been studied for decades, in every format from self-help books to publications in academic journals.
-There are statistical models, such as exponential random graph models (ERGM) and latent space models (LSM), that strive to capture the underlying relationship-generating mechanism \citep{ergm, lsm}.
-These static models, however, are not optimal for modelling ``real-world" social networks because they do not account for changes in the network as time passes.
+There are also statistical models, such as exponential random graph models (ERGM) and latent space models (LSM), that strive to capture the underlying relationship-generating mechanism \citep{ergm, lsm}.
+These static models, however, are not optimal for modeling ``real-world" social networks because they do not account for changes in the network as time passes.
 Models for \textit{dynamic social networks} are more realistic because they provide mechanisms for analyzing multiple observations of the same network over time.
@@
 \par One set of models for dynamic social networks are stochastic actor-oriented models (SAOMs), introduced in \citet{saompaper}.
-These models incorporate node-level covariate effects in addition to the traditional structural effects, such as in-degree or out-degree.
+These models incorporate node-level covariate effects in addition to the traditional structural effects, such as in-degree or out-degree, common in ERGMs.
 %These models are different from other dynamic network models because they allow us to add node level information into the model with the traditional network structure information.
 Adding node-level information into the model is a more intuitive approach to model changes in a social network because, for example
-With these models, we can examine assumptionsm made about social networks, i.e.
+With these models, we can examine assumptions made about social networks, i.e.
 that people with common interests are more likely to become friends.
-For example, we may assume that high schoolers who drink heavily are more likely to be friends with others who drink heavily.
-SAOMs allow us to incorporate this behavioral information on shared interests into the modeling process.
+For example, we may assume that teenagers who drink heavily are more likely to be friends with other tennagers who drink heavily.
+The SAOM family allows us to incorporate this behavioral information on shared interests into the modeling process.
 
 \par Unlike other network models, SAOMs are not analytically tractable \citep{saomreview}.
-Likelihood functions quickly become very complex due to the dependency structure inherent in the data, and so computational methods, such as Markov chain Monte Carlo (MCMC), are used to fit models.
+Likelihood functions quickly become very complex due to the inherent dependency structure in the data, and therefore computational methods, such as Markov chain Monte Carlo (MCMC), are used to fit models.
 %SAOMs are typically fit to dynamic social network data using a series of MCMC phases for finding method of moments (MoM) estimates of model parameters.
@@
 This software is a considerable contribution to the field of social network analysis, but the extensive parameter estimation process is largely hidden from the user.
-To better understand SAOMs and their fitting process in \texttt{RSiena}, we use model visualization, or model-vis \citep{modelvis}, to examine the underlying methods, hidden structures, and simulation results.
+To better understand SAOMs and the fitting process in \texttt{RSiena}, we use model visualization, or model-vis \citep{modelvis}, to examine the underlying methods, hidden structures, and simulation results.
 Ultimately, we aim to help researchers working with SAOMs better understand the structure, behavior, fit, and results of these models.
@@
 %\noindent \textbf{Network Data Structure:} 
-Network data across fields have similar data structures.
+Network data across fields have similar structure.
 There are always units of observation and connections between those units.
@@
 %Networks change over time, like when new relationships are formed in a social network.
-Other data on a network include inherent variables of interest on the nodes and edges themselves, e.g.\ someone's age, gender, and political affiliation, the type of relationship (friendship, coworkers, etc) that an edge represents, in addition to the changing structure of the network in time.
+Other network data include inherent variables of interest on the nodes and edges themselves, e.g.\ someone's age, gender, and political affiliation, and the type of relationship (friendship, coworkers, etc) that an edge represents.
 %These two network data structures pose unique problems to analysts, such as, ``How does the strength of a tie between two nodes affect the overall structure of the network?" or ``Do differences in node variables affect the formation of edges?" We propose to answer these questions and more through visual exploration of network data and models.
@@
 For a review of graph drawing, see \citet{graphdrawing}.
-Visualizing network data is inherently challenging because of the structure of the data itself.
+Visualizing network data is inherently challenging due to the structure of the data itself.
 Most data visualizations rely on well-defined axes inherited from the data, such as Cartesian coordinates or spatial locations, but network data typically do not have such built-in structure.
@@
 @
-<<toyplots, fig.show='hold', fig.cap="On the left, a node-link diagram of our directed toy network, with nodes placed using the Kamada-Kawai algorithm.
-On the right, the adjacency matrix visualization for that same network.", out.width='.49\\textwidth', fig.height=4>>=
+<<toyplots, fig.show='hold', fig.cap="On the left, a node-link diagram of our toy network, with nodes placed using the Kamada-Kawai algorithm.
+On the right, the corresponding adjacency matrix visualization.", out.width='.49\\textwidth', fig.height=4>>=
 toynetdat <- fortify(as.edgedf(toynet$edges), toynet$nodes)
@@
 
-The lack of a native coordinate system has led to two very different primary methods to visualize networks: the node-link diagram and the adjacency matrix visualization \citep{knuth2000, adjmatpro}.
+The lack of a native coordinate system has given rise two very different primary net-vis methods: the node-link diagram and the adjacency matrix visualization \citep{knuth2000, adjmatpro}.
 We demonstrate these with a simpe example.
@@
 
-The first method, the node-link diagram, represents nodes with points in two dimensions and then represents edges by connecting the points with lines.
+The first net-vis method, the node-link diagram, represents nodes with points in two dimensions and then represents edges by connecting the points with lines.
 These lines can also have arrows on them indicating the direction of the edge for directed networks, as show in Figure~\ref{fig:toyplots}.
@@
 %In this paper, we only consider the presence or absence of an edge, but if the network has weighted edges, $A_{ij}$ is the weight of the edge from $i \to j$. 
-An adjacency matrix visualization for our toy example is also shown in Figure~\ref{fig:toyplots}.
-This visualization shows the network as a grid, with each row and column representing a node.
+The adjacency matrix visualization for our toy example is also shown in Figure~\ref{fig:toyplots}.
+This visualization shows the network as a grid, with each row representing the ``from" node and each column representing the ``to" node of an edge.
 The grid space for edge $i\rightarrow j$ is filled in if $A_{ij} = 1$ and is empty if $A_{ij} = 0$.
+For undirected networks, $\mathbf{A}$ and the grid visualization are symmetric, with each edge represented twice.
 
@@
 For instance, \emph{paths} between two nodes in a network are easier to determine with node-link diagrams than with adjacency matrix visualizations \citep{adjmatviz}.
-In node-link diagrams, node-level information can be incorporated into the visualization by coloring or changing the shape of the points, and edge-level information can be incorporated by coloring the lines, or changing their thickness or linetype.
-Visualizing a node-level variable into an adjacency matrix visualization is not as simple, because this type of visualization features the edges.
-Adjacency matrix visualization can, however, be useful when the network is very complex, dense, or large \citep{adjmatviz}.
+In node-link diagrams, node information can be incorporated into the visualization by coloring or changing the shape of the points, and edge-level information can be incorporated by coloring the lines, or changing their thickness.
+Visualizing a node variable in an adjacency matrix visualization is not as simple, because this type of net-vis features the edges.
+An adjacency matrix visualization can, however, be useful when the network is very complex, dense, or large \citep{adjmatviz}.
 % Experimental studies have shown adjacency matrix visualization to be superior to node-link diagrams for large networks.
@@
 A node-link diagram, however, may underrepresent the edge count if the edges $x_{ij}$ and $x_{ji}$ both exist and are drawn on top of one another.
-Ultimately, there is no one ``correct" way to visualize network information, and we will be using both the node-link and adjacency matrix visualization methods throughout this paper to explore our example data and SAOMs.
+Ultimately, there is no one ``correct" way to visualize network information, and we will be using both the node-link and adjacency matrix net-vis methods throughout this paper to explore our example data and SAOMs.
 
@@
 
-Stochastic Actor-Oriented Models (SAOM) are unique because they model \textit{dynamic} networks.
-We use SAOMs to study the change of a network in time while also examining the structural and node covariate efects.
-The two titular properties of SAOMs, stochasticity and actor-orientation, are key to understanding networks as they exist naturally.
+Stochastic Actor-Oriented Models (SAOMs) model \textit{dynamic} networks, so we can use SAOMs to study the change of a network in time while also examining the structural and node covariate efects.
+The two titular properties of SAOMs, stochasticity and actor-orientation, are key to understanding social networks as they exist naturally.
 Most social networks, even holding constant the set of actors over time, are ever-changing as relationships form or disappear.
@@
 
-A \textit{dynamic network} is a network, that consists of a fixed set of $n$ nodes, that is changing over time, and is observed at $M$ discrete time points, $t_1, \dots, t_M$ with $t_1 < t_2 < \dots < t_M$.
+A \textit{dynamic network} is consists of a fixed set of $n$ nodes, that is changing over time, and is observed at $M$ discrete time points, $t_1, \dots, t_M$ with $t_1 < t_2 < \dots < t_M$.
 We denote the network observation at timepoint $t_m$ by $x(t_m)$ for $m = 1, \dots M$.
-The SAOM assumes that this longitudinal network of discrete observations is embedded within a continuous time Markov chain, (CTMC), which we will denote by $X(T)$.
+The SAOM assumes that this longitudinal network of discrete observations is embedded within a continuous time Markov chain, (CTMC), which we denote by $X(T)$.
 The CTMC is almost entirely unobserved: we assume the first and last network observations are the starting and ending points of the CTMC, i.e.
@@
 The CTMC process $X(T)$ is a series of single tie changes that happen according to the SAOM's \textit{rate function}, where one actor at a time is ``selected" to make a change.
-Once selected, the actor tries to maximize its \textit{objective function} based on the current state and possible future states of the network.
+Once selected, the actor changes its ties in order to maximize its \textit{objective function} based on the current state and possible future states of the network.
 
 \noindent \textbf{The Rate Function} In general, a SAOM's \emph{rate function}, $\rho(x, \mathbf{z}, \boldsymbol{\alpha})$, is a function of the network at its current state $x$, covariates of interest, $\mathbf{z}$, and some parameters,   $\boldsymbol{\alpha}$.
-With this general formulation, it is possible that each node will have a different rate of change.
-Throughout this paper, we assume a simple rate function, $\rho(x, \mathbf{z}, \boldsymbol{\alpha}) \equiv \alpha_m$ that is a constant across all actors between time $t_m$ and $t_{m+1}$.
-With this simple rate function, we have $M-1$ rate parameters, one for each transition from $x(t_{m-1})$ to $x(t_m)$ for $m = 2, \dots, M$.
+With this general formulation, each node can have a different rate of change.
+Throughout this paper, we assume a simple rate function, $\rho(x, \mathbf{z}, \boldsymbol{\alpha}) \equiv \alpha_m$ that is constant across all actors between time $t_m$ and $t_{m+1}$.
+With this simple rate function, we have $M-1$ rate parameters, one for each transition from $x(t_{m})$ to $x(t_{m+1})$ for $m = 1, \dots, M-1$.
 The rate parameter, $\alpha_m$ dictates how often an actor $i$ ``gets an opportunity" to change one of its ties, $x_{ij}$, for $j \in \{1, \dots, n\}$ in the time period from $t_{m}$ to $t_{m+1}$ for $m = 1, \dots, M-1$.
 A SAOM also assumes that the actors $i$ are conditionally independent given their ties, $x_{i1}, \dots, x_{in}$ at the current network state.
-Let $\tau(i|x,m) $ be the wait time until actor $i$ gets to the opportunity to change its current set of ties.
-For any time point, $T$, where $t_{m} \leq T < t_{m+1}$, the waiting time to the next change made by actor $i$ follows an exponential distribution:
+Let $\tau(i|x,m) $ be the wait time until actor $i$ gets to the opportunity to change its ties.
+For any time point $T$, where $t_{m} \leq T < t_{m+1}$, the waiting time to actor $i$'s next change follows an exponential distribution:
 \begin{equation} \label{eq:c1}
@@
 \end{equation}
-
-\noindent From Equation~\ref{eq:c1}, we can derive the waiting time to the next change opportunity by \textit{any} actor in the network, $\tau(x)$.
+\noindent with expected value $\alpha_m^{-1}$.
+From Equation~\ref{eq:c1}, we can derive the waiting time to the next change opportunity by \textit{any} actor in the network, $\tau(x)$.
 The value $\tau(x)$ is also exponentially distributed, and has expected value $(n\alpha_m)^{-1}$.
-The estimation of $\alpha_m$ in \texttt{RSiena} is simple: the method of moments (MoM) is used to estimate the rate with the statistic
+The estimation of $\alpha_m$ in \texttt{RSiena} is simple: method of moments (MoM) is used with the statistic
 \begin{equation}\label{eq:cdiff}
@@
 \end{equation}
-which is the total number of changes from observation at time $t_m$ to the observation at time $t_{m+1}$.
-
-\noindent \textbf{The Objective Function} Because of the conditional independence assumptions given in Equation~\ref{eq:c1}, we consider the objective function for each node separately.
+which is the total number of changes from $x(t_m)$ to $x(t_{m+1})$.
+
+\noindent \textbf{The Objective Function} Because of the conditional independence assumptions given in Equation~\ref{eq:c1}, the objective function is written separately for each node.
 Suppose node $i$ is selected to change at the next step in the CTMC.
@@
 These nodes $j$, are referred to as \textit{alter} nodes, and will not change any of their ties at the next step in the CTMC.
-The ego node $i$ in the current network state $x$, will make a change that maximizes its \emph{objective function}, written as:
+The ego node $i$ in the current network state $x$, wants to make a change that maximizes its \emph{objective function}:
 \begin{equation} \label{eq:of1}
@@
 \end{equation}
-for the network $x$, covariates $\mathbf{Z}$, and $K$ the total number of parameters in the objective function.
+where $x$ is the current network state, $\mathbf{Z}$ are node covariates, and $K$ is the total number of parameters in the objective function.
 The parameters of the model, $\beta_1, \dots, \beta_K$ correspond to network statistics, $s_{ik}(x, \mathbf{Z})$, for $k = 1,\dots, K$.
@@
 The statistics corresponding to these two parameters, $s_{i1}(x)$ and $s_{i2}(x)$ are given in Table~\ref{tab:effects}.
-In the \texttt{RSiena} software, there are over 80 possible $\boldsymbol{\beta}$ parameters to add to the model.
-The definition of the statistics for all possible parameters are provided in \citet{RSienamanual}, and two selected covariate parameters and corresponding statistics are discussed in Section~\ref{sec:exdatamods} and shown in Figure~\ref{fig:structures}.
+In the \texttt{RSiena} software, there are over 80 possible $\boldsymbol{\beta}$ parameters with corresponding statistics $s_{ik}(x, \mathbf{Z})$ to add to the model.
+The definition of the statistics for all possible parameters are provided in \citet{RSienamanual}, and two additional parameters and their corresponding statistics are discussed in Section~\ref{sec:exdatamods} and shown in Figure~\ref{fig:structures}.
 
@@
 \caption{\label{tab:effects} Examples of structural effects included in a SAOM.
-The dark nodes in the Figures are the ego nodes, and all others are alters.
+The darker nodes in the Figure column are the ego nodes ($i$), and all others are alters ($j$).
 The dotted lines are ties the ego node is en(dis)couraged to make when the parameter corresponding to each effect is positive (negative).}
@@
 
-With the objective function, we can define the probability that the ego node $i$ chooses to change its tie to alter node $j$, denoted by $p_{ij}(\boldsymbol{\beta}, x, \mathbf{Z})$.
-Let $x(i \leadsto j)$ be the network identical to network $x$ with the exception of tie $x_{ij}$, which is equal to $1-x_{ij}$; i.e.
-the presence or absence of node $x_{ij}$ is flipped in the new network.
+In the CTMC, when actor $i$ can make a change, it chooses which tie $x_{i1}, \dots, x_{in}$ to change at random according to the probabilities $p_{ij}(\boldsymbol{\beta}, x, \mathbf{Z})$.
+These probabilities are defined via the objective function.
+Let $x(i \leadsto j)$ be the network identical to network $x$ with the exception of tie $x_{ij}$, which is equal to $1-x_{ij}$; i.e.\ the presence or absence of node $x_{ij}$ is flipped in the new network.
 The probability that the tie $x_{ij}$ changes is defined as:
@@
 %When the value of the objective function is high at the current state, the probability of not making a change is also high.
-In the CTMC, when actor $i$ can make a change, it chooses which tie $x_{i1}, \dots, x_{in}$ to change at random according to the probabilities $p_{ij}(\boldsymbol{\beta}, x, \mathbf{Z})$.
 We explore these probabilities further in Section~\ref{sec:algorithms}.
@@
 
-To fit a SAOM to dynamic network data, we apply the \texttt{RSiena} software, which uses simulation methods to estimate parameter values using either method of moments (MoM) or maximum likelihood (ML) estimation \citep{RSiena}.
+To fit a SAOM to dynamic network data, we apply the \texttt{RSiena} software, which uses simulation methods to estimate parameter values using either MoM or maximum likelihood (ML) estimation \citep{RSiena}.
 We chose to use MoM estimation, so we describe that process here, while more information on ML estimation can be found in \citet{SienaAlgs}.
@@
 \end{equation}
-where $\theta = (\boldsymbol{\alpha}, \boldsymbol{\beta})$ is the vector of rate and objective function parameters, and $s_{obs}$ is the observed vector of the model statistics over all network observations.
-The full SIENA MoM estimation algorithm operates in three phases, as described in \citet{RSienamanual} and \citet{SienaAlgs}, and we briefly summarize these phases.
-The first phase gets initial estimates of the score functions for use in the second phase.
-The second phase is the Robbins-Monro algorithm, and results in parameter estimates through iterative updates and simulation from the CTMC at iterations of the parameter values.
+where $\theta = (\boldsymbol{\alpha}, \boldsymbol{\beta})$ is the vector of rate and objective function parameters, and $s_{obs}$ is the observed vector of the corresponding statistics summed over all network observations.
+The full SIENA MoM estimation algorithm operates in three phases, as described in \citet{RSienamanual} and \citet{SienaAlgs}, and we briefly summarize these phases here.
+The first phase obtains initial estimates of the score functions for use in the second phase.
+The second phase is the Robbins-Monro algorithm, and results in parameter estimates through iterative updates and simulation from the CTMC at the iterations of the parameter values.
 The third phase uses the parameters from phase two to estimate the score functions and covariance matrix of the parameter estimates and carries out convergence checks.
-In Section~\ref{sec:algorithms}, we further explore these phases in the SIENA method-of-moments algorithm through visualization, bringing them out of their ``black box" and into the light.
+In Section~\ref{sec:algorithms}, we further explore these phases in the SIENA MoM algorithm through visualization, bringing them out of their ``black box" and into the light.
 
@@
 Section~\ref{sec:collectmodels} follows, in which we explore collections of models.
-Finally, in Section~\ref{sec:algorithms}, the last principle of model-vis, we visualize the underlying algorithms in the SAOM fitting process.
+Finally, in Section~\ref{sec:algorithms}, we visualize the underlying algorithms in the SAOM fitting process.
 We conclude with a discussion in Section~\ref{sec:discussion}.
@@
 
-<<alldataviz, fig.cap="The small friendship network data we will be modelling throughout this paper.
-The nodes change color as their drinking behavior changes, and the network becomes less densely connected over time.
-These are the mechanisms we hope to incorporate and model with SAOMs.", out.width='.99\\textwidth', fig.height=3, results='hide'>>=
+<<alldataviz, fig.cap="The small friendship network data we model throughout this paper.
+The teenagers' drinking behavior changes and the network becomes less densely connected over time.
+These are the mechanisms we hope to model with SAOMs.", out.width='.99\\textwidth', fig.height=3, results='hide'>>=
 cols <- RColorBrewer::brewer.pal(8, "Greys")
@@
 These data come from \citet{friendsdata}, and we chose to only work with a subset of the data to make node-link diagrams easier to see and to make any changes in the network more noticeable.
-The node-link visualization of this data is provided in Figure~\ref{fig:alldataviz}.
-%For model fitting, we condition on wave 1 and estimate the parameters of the models for the second and third waves.
-The actor-level categorical covariate, drinking behavior, has four values: (1) does not drink, (2) drinks once or twice a year, (3) drinks once a month, and (4) drinks once a week.
+We visualize this data with a node-link diagram in Figure~\ref{fig:alldataviz}. %For model fitting, we condition on wave 1 and estimate the parameters of the models for the second and third waves. 
+The actor covariate of interest, drinking behavior, has four values: (1) does not drink, (2) drinks once or twice a year, (3) drinks once a month, and (4) drinks once a week.
 In Figure~\ref{fig:alldataviz}, the nodes are colored according to the drinking behavior of that student.
@@
 
-The second data example we use is a much more recent dataset on collaboration in the United States Senate during the $111^{th}$ through $114^{th}$ Congresses.
+The second data example we use is a much more recent dataset on collaboration in the United States Senate during the $111^{th}$ through $114^{th}$ Congresses, shown in Figure~\ref{fig:senateplot}.
 These sessions of congress correspond to the years of Barack Obama's presidency, from 2009-2016.
-(Details of how this data can be downloaded are provided by François Briatte at \url{https://github.com/briatte/congress}).
-In this network, ties are directed from senator $i$ to senator $j$ when senator $i$ cosponsors the bill that senator $j$ authored.
+(Details of how this data can be downloaded are provided by François Briatte at \url{https://github.com/briatte/congress}.) In this network, ties are directed from senator $i$ to senator $j$ when senator $i$ cosponsors the bill that senator $j$ authored.
 There are hundreds of ties between senators when they are connected in this way, so we simplify the network by computing a single value for each pair of senators called the \textit{weighted propensity to cosponsor} (WPC).
@@
 
-Because we require binary edges for SAOMs, we focus only on very strong collaborations, where $WPC$ is high. For a senate collaboration network $x$, edges are defined as
+Because we require binary edges for SAOMs, we focus only on strong collaborations, where $WPC$ is high. For a senate collaboration network $x$, edges are defined as
 \begin{equation}\label{eq:wpc2}
@@
 Some prominent names are shown to demonstrate their place in the network.
-As with the friendship example, analyzing the senate data with a SAOM should capture the underlying dynamics in a way that gives insight into the impact the behavioral factors have on the senate's collaboration structure.
-By using data sets from very different contexts, we hope to assess possible differences in model performance.
+As with the friendship example, analyzing the senate data with a SAOM should capture the underlying dynamics, giving insight into the impact that the actor covairates have on the senate's collaboration structure.
+By using data sets from very different contexts, we hope to assess some differences in model performance.
 
@@
 @
-\caption{\label{fig:jtt}Realization of a JTT, where $i$ is the ego, $j$ is the alter, and $h$ is the intermediary.
-The covariate is represented by the shape of the nodes.
+\caption{\label{fig:jtt}Realization of a JTT, where the covariate is represented by the shape of the nodes.
+The dotted tie is encouraged if $\beta_3$ is positive.
 $s_{i3}(x,z) = \sum_{j \neq h} x_{ij}x_{ih}x_{hj} \mathbb{I}(z_i = z_h \neq z_j)$}
@@
 \caption{\label{fig:structures} The additional network effects included in our models fit to the friends data.
-On the left, a JTT between $i$ and $j$, who have different covariate values, and on the right, a N22 between $i$ and $j$.}
+On the left, a JTT between $i$ and $j$, who have different covariate values, and on the right, a N22 between $i$ and $j$, who are indirectly connected by nodes $k$, $h$.}
 \end{figure}
@@
 These statistics are known as the \emph{number of jumping transitive triplets} (JTT) and the \emph{number of doubly achieved distances two effect} (N22), respectively.
-The JTT effect emphasizes triads formed between actors with different covariate values, while the N22 effect emphasizes indirect ties between actors.
-The covariate for the friendship data is the drinking behavior, and the number of bills authored for the senate data.
+The JTT effect emphasizes triads formed between actors with \emph{different} covariate values, while the N22 effect emphasizes \emph{indirect} ties between actors.
+The covariate for the friendship data is the drinking behavior, and is the number of bills authored for the senate data.
 These additional effects are visualized in Figure~\ref{fig:jtt} and Figure~\ref{fig:dad}, where the ties affected by the parameter value are represented by dotted lines.

### 2018-10-15 paper.Rnw
final proof-reading tweaks

--- 2018-10-08 paper.Rnw
@@
 \begin{abstract}
-Stochastic actor-oriented models (SAOMs), first introduced by Snijders (1996), are a type of statistical network model for dynamic social networks.
+Stochastic actor-oriented models (SAOMs), first introduced by Snijders (1996), are statistical network models for dynamic social networks.
 Unlike other network models, SAOMs are analytically intractable and require extensive simulation procedures for model fitting.
-We use model visualization techniques introduced in Wickham et al (2015) to better understand these models and their behavior.
-We use a three-pronged attack: (a) network visualization to view the \emph{model}, (b) exploring collections of models for parameter interpretation, and (c) delving into the fitting algorithms to gain insight into the underlying mechanisms.
+In this paper, we use model visualization techniques introduced in Wickham et al (2015) to better understand SAOMs and their behavior.
+We use a three-pronged visualization attack: (a) network visualizations to view the \emph{model}, (b) visualizations of collections of models to interpret parameters, and (c) visualizations of fitting algorithms to gain insights into the underlying mechanisms.
 With the help of static and dynamic visualizations, we bring the hidden SAOM fitting processes into the foreground, with the goal of a better understanding and higher accessibility of SAOMs by social network analysts.

### 2018-10-15 paper-child.Rnw
ready to start submitting....

@@
 %Every good \emph{data} analysis includes both numerical and visual summaries of the data, so why restrict \emph{model} analysis, description, and diagnostics to numerical summaries? 
-\emph{Model visualization} complements traditional model diagnostic tools such as $R^2$ values and residual plots by enriching our understanding and interpretation of statistical models.
+\emph{Model visualization} complements traditional model diagnostic tools such as $R^2$ values and residual plots to enrich our understanding and interpretation of statistical models.
 Each of the three different meanings of the word ``model", the model \emph{family}, the model \emph{form}, and the \emph{fitted} model \citep[see][pg.
@@
 Since we have already decided on our model family (SAOMs), we shift our focus to the fitted model and the model form.
-Specifically, we use model visualization to learn more about how the \emph{model form} affects the \emph{fitted model}.
+Specifically, we use model-vis to learn more about how the \emph{model form} affects the \emph{fitted model}.
 
@@
 But what does this definition mean for dynamic network models?
-We define the data space for SAOMs as the combined data structures of the network: the node, edge, and time information.
+We define the \emph{data space} for SAOMs as the combined data structures of the network: the node, edge, and time information.
 We use the \texttt{R} package \texttt{geomnet} \citep{geomnet} to visualize the three data spaces together.
 The package lays out the network using node-link diagrams, and the network's node or edge data are then mapped to different visual features.
-The color, size, and shape of the points can be used to represent variables in the node data, while the color, linewidth, and linetype of the lines between points can be used to represent the edge variables.
+The color, size, and shape of the points can be used to represent variables in the node data, while the color, linewidth, and linetype of the edges between nodes can be used to represent variables in the edge data.
 To view temporal changes, we place the network at different timepoints side-by-side to see the evolution.
@@
 
-<<funsenatevis, fig.height=4, fig.cap='The 111th Senate at two times: while Clinton was in the senate in 2009 (on the left) and after she left (on the right).
+<<funsenatevis, fig.height=3, fig.cap='The 111$^{th}$ Senate at two times: while Clinton was in the senate in 2009 (left) and after she departed (right).
 We map sex, party, and bills authored to the shape, color, and size of the nodes, respectively.
-We also map the $WPC$, to the edge linewidth.'>>=
+We also map $WPC$ to the edge width.'>>=
 se111clint <- filter(seobama, senate == 111)
@@
            aes(from_id = source, to_id = target, color = party, linewidth = gsw, shape = sex,size = n_au/30)) +
-  scale_size_continuous(name="# bills\nauthored" , guide = guide_legend(nrow= 2), breaks = c(0,1,2,3), labels = c(0,30, 60, 90)) +
+  scale_size_continuous(name="# bills\nauthored" , guide = guide_legend(nrow= 2), breaks = c(0,1,2,3), labels = c(0,30, 60, 90), range = c(.5,3)) +
   #theme_net() +
@@
 
-To demonstrate the functionality of \texttt{geomnet}, we visualize the connections in the $111^{th}$ United States Senate at two different timepoints: when Hillary Clinton was in the senate, and after she left to become Secretary of State.
+To demonstrate the functionality of \texttt{geomnet}, we visualize the connections in the $111^{th}$ U.S.
+Senate at two different timepoints: when Hillary Clinton was in the senate, and after she left to become Secretary of State.
 Clinton was only in the $111^{th}$ senate from January 3-20, 2009, in the middle of her second term as the junior senator from New York.
@@
 So with Clinton included in the node-link diagram in Figure~\ref{fig:funsenatevis}, the senate looks much more collaborative than it does without her.
-We then map our potential covariates to visual features: number of bills authored  maps to the size of the node, gender maps to the shape, and party to the color.
-In addition, we visualize the strength of the tie by mapping the $WPC$ value between the two senators to the linewidth of the edge.
+We then map our potential model covariates to visual features: number of bills authored to size of the node, gender to shape, and party to color.
+In addition, we visualize the strength of the tie by mapping the $WPC$ value between the two senators to the linewidth.
 By viewing node information, edge information, and time simultaneously, we have learned that Senator Clinton greatly affected the structure of collaboration within the senate.
 
-<<summnetM1, message=FALSE, fig.cap='On the left, the first wave of observed data that is conditioned on in the model.
+<<summnetM1, message=FALSE, fig.cap='On the left, the first wave of observed data, conditioned on in the model.
 On the right, the second wave of observed data.
@@
 %In Figure~\ref{fig:summnetM1}, we show the \textit{expected network} of wave 2 from model M1, created with 1,000 simulations from the model. 
-To construct an expected network, we: 
+To construct an expected network, we
 \begin{enumerate}
-\item simulate 1,000 wave 2 and wave 3 instances of our small friendship network from model M1 using parameters that have previously been estimated. Then, we 
+\item simulate 1,000 wave 2 and wave 3 instances of our small friendship network from model M1 using estimated parameter values (see Table~\ref{tab:meanests}) that have previously been estimated.
+Then, we
 \item count the number of times each edge $i \rightarrow j$ appeared in one of the 1,000 network simulations. Then, we
-\item construct one weighted summary network for each time point, which has edge weights equal to the proportion of the simulated networks in which the edge appeared. Finally, we 
+\item construct one weighted summary network for time points 2 and 3. Each summary network has edge weights equal to the proportion of the simulated networks in which the edge appeared. Finally, we 
 \item create the \textit{expected network} by including only the edges that appeared in more than 5\% of the simulations. 
@@
 On either side of the average network in Figure~\ref{fig:summnetM1}, we show the data for comparison: wave 1 on the left, and wave 2 on the right.
-We can see that the structure of the average network is much more similar to wave 1 than to wave 2.
+We can see that the overall structure of the average network is much more similar to wave 1 than to wave 2.
 In both wave 1 and the expected network, the edges form two distinct clusters of nodes: $\{2,3,4,5,6,12,13,15\}$ and $\{7,8,9,10,11,14\}$.
 The expected network, however, is supposed to represent the second wave of data, on the right in Figure~\ref{fig:summnetM1}.
-The expected network does not resemble wave 2, which has one small node cluster and on large node cluster, with four unconnected nodes.
+The expected network does not resemble wave 2, which has one small, strongly connected cluster and one larger, sparsely connected cluster, with four unconnected nodes.
 Because the expected network does not resemble wave 2, we take this as evidence that the simple model, M1, does not capture the change in network structure from the first to the second wave of data.
@@
 We created these networks by computing the difference between the adjacency matrices for the expected network and wave 1, and for wave 2 and the expected network.
-The lack of model fit becomes even more clear in Figure~\ref{fig:summary2}, with the broken edges representing ties that only show up in the expected network, and the solid edges representing ties in the data that the model misses completely.
-Dotted edges are in both the expectation and the data and represent where the model fit the data well.
+The lack of model fit becomes even more clear in Figure~\ref{fig:summary2}, with the broken edges representing ties that only show up in the expected network, and the solid edges representing ties in the data that the model fails to capture.
+Dotted edges are in both the expectation and the data and represent where the model ``got it right".
 Again, the widths of the edges demonstrate the frequency of the ties in the simulations.
@@
 We only see dotted and broken edges, meaning that the simulated networks consist primarily of edges in wave 1, plus some new edges that are not in the first wave of data.
-The added ties are all reciprocating edges in the data, meaning that the model predicts more reciprocity between actors than there actually is.
+The unobserved ties are all reciprocating edges in the data, so the model predicts more reciprocity between actors than there actually is.
 On the right in Figure~\ref{fig:summary2}, we see the difference between the second wave of data and the expected network.
@@
 However, there are six solid edges, which are edges in the data that do not appear at all in the expected network, and there are many broken edges that appear in the expected network but not in the data.
-Although there are some edges that appear in both the expected network and wave 2, the majority of edges in the expected network are not in wave 2, and there are several edges missing from the expectation.
+Although there are some edges that appear in both the expected network and wave 2, the majority of edges in the expected network are not in wave 2, and there are several edges in the data missing from the expectation.
 We conclude that the model M1 is a poor fit to our example data, which we have seen very clearly by viewing the model in the data space.
@@
 
-Now that we have assessed model fit, we next want to learn about model behavior in order to choose better fitting models in the future.
-This is the second principle of model visualization: "collections [of models] are more informative" \citep{modelvis}.
-We examine the behavior of SAOMs by visualizing collections of models.
-We build collections of models in four ways.
+Now that we have assessed model fit, we want to learn about model behavior in order to choose better fitting models in the future.
+This is the second principle of model visualization: ``collections [of models] are more informative" \citep{modelvis}.
+We examine the behavior of SAOMs by visualizing collections of them, which we construct in four ways.
 First, we explore the space of all possible models to determine the most significant model parameters.
@@
 
-\noindent \textbf{Exploring the space of all possible models:} The \texttt{RSiena} manual contains over eighty effects to include in the SAOM objective function.
-To select objective function parameters for fitting models to our example data, we searched through all possible effects available to find \textit{significant} effects.
+\noindent \textbf{Exploring the space of all possible models:} The \texttt{RSiena} manual contains over 80 effects to include in the SAOM objective function.
+To select objective function parameters for modeling our example data, we searched through all effects available to find \textit{significant} effects.
 The \texttt{RSiena} package includes the function \texttt{Wald.Rsiena} for Wald-type tests of parameter significance.
@@
 To explore the behavior of these three models, we simulate from them.
-Using the parameter values given in Table~\ref{tab:meanests} for the friendship data, we simulated 1,000 observations from each of the three models.
+Using the estimated parameter values in Table~\ref{tab:meanests}, we simulated 1,000 observations of waves 2 and 3 ofthe friendship data from each of the three models.
 From the simulations, we created an expected network for wave 2 from models M1, M2, and M3 using the same method described in Section~\ref{sec:minds}.
-These expected networks are shown in Figure~\ref{fig:averagenet} in the Appendix.
-We see that all models are failing to capture the structure of the second wave, even though the individual parameters were all significant.
+These expected networks are shown in Figure~\ref{fig:averagenet} in the Appendix, and we see that all models are failing to capture the structure of the second wave, even though the individual parameters were all significant.
 %In Section~\ref{sec:samemodsamedat}, we fit these three models to the data 1,000 times, and examine the parameter densities.
@@
 <<distests, fig.cap="Distribution of fitted parameter values for our three SAOMs.
-The inclusion of $\\beta_3$ or $\\beta_4$ clearly has an effect on the distribution of the rate parameters, $\\alpha_1$ and $\\alpha_2$, and on the other objective function parameters $\\beta_1, \\beta_2$.", fig.height=3>>=
+The inclusion of $\\beta_3$ or $\\beta_4$ clearly has an effect on the distributions of the rate parameters, $\\alpha_1$ and $\\alpha_2$, and on the other objective function parameters $\\beta_1, \\beta_2$.", fig.height=3>>=
 
@@
 
-\noindent \textbf{Fitting the same model to the same data:} The simulations performed to fit a SAOM lead to inherent variability in the fitted model.
+\noindent \textbf{Fitting the same model to the same data:} The simulations required to fit a SAOM lead to inherent variability in the fitted values.
 To examine this variability, we fit models M1, M2, and M3 to the friendship network 1,000 times each.
@@
 Clearly, the inclusion of the JTT parameter, $\beta_3$, results in very different estimates the of the other four parameters in all models, $\alpha_1$, $\alpha_2$, $\beta_1$, and $\beta_2$.
-On the other hand, the inclusion of the indirect tie parameter, $\beta_4$, does not lead to different estimates of other model parameters.
-When $\beta_3$ is included, its estimate is postitive, meaning that friendships between two girls with different drinking behaviors tend to form when there is an intermediary who is already friends with the two girls.
-The inclusion of this parameter leads to increases in the rate parameters' estimates.
-Thus, when modeling the transitive triplet behavior in this way, we conclude that girls change friends more frequently.
+On the other hand, the inclusion of the indirect tie parameter, $\beta_4$, does not lead to such drastically different estimates of other model parameters.
+When $\beta_3$ is included, its estimate is postitive, meaning that friendships between two girls with different drinking behaviors tend to form when there is third girl who is already friends with the other two.
+The inclusion of $\beta_3$ also leads to increases in the estimates of $\alpha_1, \alpha_2$.
+Thus, when modeling the transitive triplet behavior in this way, we may conclude that girls change friends more frequently.
 It is not clear, however, that the addition of a parameter to the objective function \textit{should} effect the estimates of the rate parameters.
-The strong relationship between rate parameter estimates and objective function parameters is further shown in Figure~\ref{fig:corplots} in the Appendix.
-After visualizing the results of the same models fit many times to the same data, we were able to see potential model behavior problems right away.
+This strong and possibly problematic relationship between rate parameter estimates and objective function parameters is further shown in Figure~\ref{fig:corplots} in the Appendix.
+When visualizing the results of the same models fit many times to the same data, we can to see potential model behavior problems right away.
 %In comparison with models M1 and M3, model M2 typically has higher estimates of the rate parameter, meaning that the inclusion of the covariate statistic in the model leads to higher estimates of the number of times, on average, a node gets to change its ties.
@@
 \scalebox{0.9}{
-\begin{tabular}{lcccccc }
+\begin{tabular}{lccc|ccc }
 
@@
 Looking at the estimates for both data sets, a few patterns emerge.
-First, we can see the same relationship between the outdegree parameter, $\beta_1$, and the reciprocity parameter, $\beta_2$ in both data sets and across all three models.
+First, we can see the same relationship between the outdegree parameter, $\beta_1$ and the reciprocity parameter, $\beta_2$ in both data sets and across all three models.
 The estimates of $\beta_1$ are all negative and range from about -5 and -3, while the estimates of $\beta_2$ are all positive and range from about four to five.
@@
 Furthermore, the estimates of $\beta_4$ for the senate data are near zero, suggesting that this effect, which considers indirect ties, is not informative for the senate data.
-In contrast for the friendship data, the mean estimate of $\beta_4$ is about -7.5, which suggests that indirect ties are \textit{strongly} discouraged from forming.
+In contrast, for the friendship data the mean estimate of $\beta_4$ is about -7.5, which suggests that indirect ties are \textit{strongly} discouraged from forming.
 Again, this is fairly intuitive given the context: teenage girls likely want to have direct friendships with each other.
-Senators, on the other hand, have to wheel and deal to pass their legislation, and indirect relationships may be beneficial at times.
+Senators, on the other hand, have to wheel and deal to pass their legislation, and indirect relationships may be beneficial.
 By examining the results of models fitted to different data, we better understand the interpretation of the objective function parameters, and we have learned of some potential parameter correlation problems affecting the fit of the models.
@@
 
-The last model visualization principle, "explore the process of model fitting," is crucial for understanding SAOMs \citep{modelvis}.
+The last model visualization principle, ``explore the process of model fitting," is crucial to understanding SAOMs \citep{modelvis}.
 When fitting a SAOM in \texttt{RSiena}, several simulation steps are hidden from the user.
-One of the hidden steps in the SIENA MoM algorithm is the \textit{microstep} process.
-At each microstep, an \textit{ego} node is selected to make a change, and the chosen ego node randomly makes one change in its ties to maximize its objective function.
-Between two network observations $x(t_m)$ and $x(t_{m+1})$, there can be dozens, hundreds, or even thousands of microsteps, depending on the size of the network and the number of differences between two network observations.
-We want to simulate and view these in-between steps in order to better understand the behavior of the underlying continuous-time Markov chain.
+One of the hidden steps in the SIENA MoM algorithm is the \textit{microstep} process, which corresponds to the CTMC described in Section~\ref{sec:saoms}.
+At each microstep, an \textit{ego} node is selected to change, and the chosen node randomly makes one change in its ties according to to the probabilities $p_{ij}$ defined in Equation~\ref{eq:pij}.
+Between two network observations $x(t_m)$ and $x(t_{m+1})$, there can be dozens, hundreds, or even thousands of microsteps, depending on the size of the network and the number of tie changes between two network observations.
+We want to simulate and view these in-between steps in order to better understand the behavior of the underlying CTMC.
 Hiding the microsteps is practical and efficient if the goal is to fit a model to data, obtain parameter estimates, and draw conclusions or make predictions.
 But in order to understand \textit{how} the models are fit, we need to extract and explore the different steps of the process.
-We use these steps to visually explore the algorithms, so that we are visualizing the process, and not just the main result.
+We use these steps to visually explore the algorithms, so that we are visualizing the model fitting process, and not just the model fitting result.
 
@@
 <<makinggif, fig.cap='A selection of images in the microstep animation.
-The selected edges and nodes are emphasized by changing the size and the color, then when edges are removed, they fade out, shrinking in size, while the nodes change color and shrink to blend in with the other nodes.
-Frames 15-28 show a an edge disappearing, while frames 71-84 show an edge appearing.', message = FALSE, warning = FALSE, fig.height=2.5>>=
+The selected edges and nodes are emphasized by changing the size and the color, then they fade in or out, while the nodes change color and shrink to blend in with the other nodes.
+Frames 15-28 show a an edge disappearing, while frames 71-84 show an edge appearing.
+A spring layout is used.', message = FALSE, warning = FALSE, fig.height=2.5>>=
 # Need old version of dplyr to get this to work:
@@
 
-To visualize the underlying algorithm, we animate the microsteps that form the transition steps of the CTMC from wave 1 to wave 2 of the small friendship network example shown in Figure~\ref{fig:alldataviz}.
+To visualize the underlying algorithm, we animate the microsteps that form the CTMC from wave 1 to wave 2 of the small friendship network (see Figure~\ref{fig:alldataviz}).
 The microstep animation, some frames of which are shown in Figure~\ref{fig:makinggif}, represent one possible path from wave 1 to wave 2 according to the model M1.
@@
 For both changes, the ego node is emphasized with color and size.
-When a tie is removed, it is emphasized with the same color and size as the node.
-As the animation proceeds, the tie shrinks and disappears as the ego node shrinks and changes color back to black.
-If a tie is being added, the it appears red from nothing, grows large for emphasis, then changes color and size to match the rest of the edges.
-In this network animation, we see the modeled steps of the \textit{unobserved} CTMC process between two observed networks.
-We see each part of the SAOM come into play: we first see the selection of nodes to change via the rate function, then we see the result of the actor maximizing its objective function by either deleting or adding a node.
-Finally, the layout of the nodes changes as edges are removed or added, which demonstrates how the overall network structure changes with each of the individual tie changes.
+When a tie is removed, it is first emphasized with the same color and size as the node.
+As the animation proceeds, the tie shrinks and disappears as the ego node shrinks and changes color to match the others.
+If a tie is being added, it appears red from nothing, grows large for emphasis, then changes color and size to match the rest of the edges.
+In this network animation, we show the modeled steps of the \textit{unobserved} CTMC process between two observed networks.
+We see each part of the SAOM come into play: first, the selection of the ego node via the rate function, then the result of the actor maximizing its objective function by either deleting or adding a node.
+Finally, the layout of the network transitions as edges are removed or added, demonstrating how the overall network structure changes with each of the individual tie changes.
 Combining the tie changes in this animation makes the CTMC process the focus and demonstrates the underlying change mechanism of the dynamic network according to the chosen model.
@@
 <<adjmatorder, message = FALSE, warning = FALSE, fig.cap="On the left, the starting friendship network represented in adjacency matrix form, ordered by node ID.
-On the right, the same adjacency matrix is presented after ordering the vertices by one repetition of the microstep simulation process from wave one to wave two.", fig.width=8, fig.height=3, fig.align='center'>>=
+On the right, the same adjacency matrix is presented after seriation", fig.width=8, fig.height=3, fig.align='center'>>=
 # first present the original wave 1 adj mat, then present the adj mat after being ordered by the 1st pc of the big adjmat A.
@@
 This process is known as matrix seriation
-To seriate the matrix, we first construct a cumulative adjaceny matrix, $\mathbf{A}^{cum}$, for the series of microsteps simulating the network from $x(t_m)$ to $x(t_{m+1})$.
-A single entry in the cumulative adjacency matrix, $\mathbf{A}^{cum}_{ij}$, is the total number of times the edge $i \rightarrow j$ appears in the network from the initial observation, $x(t_m) \equiv X(0)$ to the final result of the last microstep, $X(R)$, where $R$ is the total number of microsteps taken:
+To seriate the matrix, we first construct a cumulative adjaceny matrix, $\mathbf{A}^{cum}$, for the series of microsteps in the CTMC simulated from $x(t_m)$ to $x(t_{m+1})$.
+A single entry in the cumulative adjacency matrix, $\mathbf{A}^{cum}_{ij}$, is the total number of times the edge $i \rightarrow j$ appears in the network from the initial observation, $x(t_m) \equiv X(0)$ to the last microstep, $X(R)$, where $R$ is the total number of microsteps taken:
 \begin{equation}\label{eq:acum}
@@
 We continue to use the seriated matrix in another animation of the microstep process, which is available at \url{https://vimeo.com/240092677}.
-A few frames of the video are shown in Figure~\ref{fig:adjmatani}.
+A few frames of this video are shown in Figure~\ref{fig:adjmatani}.
 In the animation, a square appears or disappears as that edge appears or disappears in the microstep process.
@@
 This visualization also shows the microsteps where the nodes do not make a change.
-From start to finish, we see network evolve according to the model: sometimes ties are removed right after they are created, and sometimes several ties are removed or created in a row.
-The adjacency matrix animation shows the many ways the algorithm explores the data space.
-By viewing this animation, we can gain a better understanding of the dynamics of this model.
+From start to finish, we see the network evolve: sometimes ties are removed right after they are created, and sometimes several ties are removed or created in a row.
+The adjacency matrix animation shows the algorithm exploring the data space.
+In contrast to the previous animation, all possible edges are represented in the adjacency matrix visualization.
+By viewing this animation, we gain a better understanding of the underlying dynamics of this model.
 
@@
 The node labels are removed to declutter the figure.
-Starting at frame 0, there are two clearly connected components: one in the bottom left corner, and one in the top right corner.
-By the end, the component in the bottom left has spread out and become sparsely connected, while the top right component has shrunk, but remains closely connected.', warning = FALSE, message = FALSE, results='hide', fig.height=3, out.width="\\linewidth">>=
+Starting at frame 0 (the data), there are two clearly connected components: one in the bottom left, and one in the top right.
+By the end, the bottom left component has become more sparse, while the top right component has shrunk, but remains closely connected.', warning = FALSE, message = FALSE, results='hide', fig.height=3, out.width="\\linewidth">>=
 adjmatall <- matrix(0, nrow = 16, ncol = 16)
@@
 
-<<heatmap, fig.align='center', fig.cap='Adjacency matrix visualizations showing the theoretical (left) and empirical (right) transition probabilities for the first microstep observed in 1,000 simulations.
+<<heatmap, fig.align='center', fig.cap='Adjacency matrix visualizations showing the theoretical (left) and empirical (right) transition probabilities for the first microstep taken in 1,000 simulations.
 The ego node is on the y-axis, and the alter node is on the x-axis.
@@
 
-We also attempt to better understand the microstep process by visualizing the theoretical and observed transition probabilites for the first microstep in the process.
+We also illuminate the microstep process by visualizing the theoretical and observed transition probabilites for the first microstep in the process.
 These probabilities are defined in Equation~\ref{eq:pij}.
@@
 There are 1,000 transitions to examine: the first microstep taken in each one of the simulations. %Another way to view these transition probabilities is through the adjacency matrix visualization. 
-We use another seriated adjacency matrix visualization to show theoretical and empirical transition probabilities of the first microstep in Figure~\ref{fig:heatmap} for each tie that was the first change in one of the simulations.
+We use another seriated adjacency matrix visualization to show theoretical and empirical transition probabilities of the first microstep in Figure~\ref{fig:heatmap} for each tie that was the first change in a simulation.
 The figure is noticeably sparse: %showing the lack of coverage in the model.
@@
 
-We have used traditional network visualization methods to explore the family of stochastic actor-oriented models for dynamic social network data in new ways.
+We have used traditional graphics and net-vis methods to explore the family of stochastic actor-oriented models for dynamic social network data in new ways.
 The guiding principles of model-vis: viewing the model in the data space, visualizing collections of models, and looking at the underlying algorithms, gave us knowledge and appreciation for these complicated and flexible models.
@@
 When we explore the algorithms, we gain a better understanding of the model and its stochastic behavior.
-Our three-pronged, model-vis attack has uncovered many properties of SAOMs.
+Our three-pronged model-vis attack has uncovered many properties of SAOMs.
 
 We have only just begun to scratch the surface of these multi-layered models for social networks.
-The \texttt{RSiena} software is incredibly powerful, and can fit much more flexible SAOMs than we have examined here.
-If, for example, a researcher thinks the network structure or an actor covariate affects the rate of change of the network, there is a way to incorporate that belief into the rate function of the SAOM.
-In addition, more than one actor-level covariate can be included in the model, and many more than three parameters can be included in the objective function itself.
+The \texttt{RSiena} software is incredibly powerful, and we can fit much more flexible SAOMs than we have examined here.
+If, for example, we think the network structure or an actor covariate affects the rate of change of the network, we can incorporate that belief into the rate function of the SAOM.
+In addition, more than one actor-level covariate can be included in the model, and many more than three parameters can be included in the objective function.
 As we have shown, the rate and objective function parameters are often highly correlated, so researchers can now use model-vis techniques to gain a better understanding of these correlations and hopefully find ways to account for their effects in the model.
@@
 Thus, some of the visualizations we have presented may not scale up to larger networks.
-Future SAOM and other network model visualization research could include interactive and/or three-dimensional graphics.
+Future SAOM and network model-vis research could include interactive and/or three-dimensional graphics, which we did not explore here.
 Interactivity is the state-of-the-art in data visualization, and network visualization is just starting to catch on \citep[see e.g.\ ][]{refinery}.
-\citet{modelvis} relied on interactive tools like GGobi \citep{STLBC03} for interactivity, and for networks, there is some interactive support with the package \texttt{plotly} \citep{plotly} built into \texttt{geomnet}.
+\citet{modelvis} also relied on interactive tools like GGobi \citep{STLBC03} for interactivity, and for networks, there is some interactive support with the package \texttt{plotly} \citep{plotly} built into \texttt{geomnet}.
 
 We hope that the principles of model-vis continue to be used for learning more about the fit and behavior of complicated statistical models.
-Our model-vis application has uncovered many properties of SAOMs, and potential abounds for use in other areas of network study.
+Our model-vis application has uncovered many properties of SAOMs, and potential abounds for application of model-vis in other areas of network analysis.
 

### 2018-11-12 paper-child.Rnw
starting to generalize

@@
 
-<<setup, echo = FALSE, message = FALSE, warning = FALSE>>=
+<<setup, echo = FALSE, message = FALSE, warning = FALSE, purl=FALSE>>=
 options(replace.assign=TRUE,width=70, digits=3)
@@
 
-<<getdata , echo=FALSE>>=
+<<getdata, echo=FALSE>>=
 set.seed(823746)
@@
 These static models, however, are not optimal for modeling ``real-world" social networks because they do not account for changes in the network as time passes.
-Models for \textit{dynamic social networks} are more realistic because they provide mechanisms for analyzing multiple observations of the same network over time.
+\textit{Dynamic social networks}, which are observed at many points in time, are the alternative to static networks.
+
+Models for dynamic social networks, however, are more complicated than those for static networks.
+These models must account for both the structure and the change in time.
+But if we can capture the network change mechanisms in time, % are more realistic because they provide mechanisms for analyzing multiple observations of the same network over time.
+
 %., and models for dynamic social networks hold a great deal of promise because of their potential to capture realistic mechanisms.
@@
 
-\par One set of models for dynamic social networks are stochastic actor-oriented models (SAOMs), introduced in \citet{saompaper}.
-These models incorporate node-level covariate effects in addition to the traditional structural effects, such as in-degree or out-degree, common in ERGMs.
+\par One family of models for dynamic social networks are continuous time Markov chain (CTMC) models, introduced in \citet{ctmcorig}.
+These models condition on the past to model the present state of the network.
+%These models incorporate node-level covariate effects in addition to the traditional structural effects, such as in-degree or out-degree, common in ERGMs.
 %These models are different from other dynamic network models because they allow us to add node level information into the model with the traditional network structure information.
@@
   labs(title = "Node-Link Diagram")
-#ggsave("figure/toynetnld.pdf",plot = nld, width = 4, height = 4)
 
 ggplot(data = toynet$edges) +
-  #geom_tile(aes(x = from, y = -to), color = 'grey20', fill = 'black') +
   geom_tile(aes(x = to, y = -from), color = 'grey20', fill = 'black') +
@@
 @
-%\hh{XXX Explain each of these visualizations and either comment on the differences in the visualizations (colors) or don't show them. }
 
@@
 \citet{ctmc}.
-% \subsubsection{Continuous Time Markov Chain (CTMC)}
-% \st{
-% In the continuous-time Markov chain literature, see for instance \citet{ctmc}, chains are characterized by their \textit{generator} or \textit{intensity} matrix $\mathbf{Q}$.
-This matrix describes the rate of change between two states of the CTMC process, and the rows of this matrix always add to zero.
-For directed networks with binary edge variables like the ones we will be working with, there are a very large number of possible states for a directed network with $n$ nodes.
-We denote the state space as $\mathcal{X}$, a set which contains $2^{n(n-1)}$ states: there are two possible states for an edge, $\{0,1\}$, and there are $n(n-1)$ edge relationships because the network is directed and we exclude self-ties.
-The intensity matrix for a CTMC in a SAOM is then a square matrix of dimension $2^{n(n-1)} \times 2^{n(n-1)}$.
-Only one tie changes at a time in the CTMC, resulting in $n(n-1)$ reachable states from the current network state.
-Thus, the intensity matrix $\mathbf{Q}$ is very sparse, with only $n(n-1) + 1$ non-zero entries in each row.
-Note that $n(n-1)$ of these entries represent the possible states that are one edge different from a given state, while the additional non-zero entry is for the state to remain unchanged.
-All other entries in a row are structural zeroes because those network states cannot be reached from the current state in a single change.
-% 
-% The two pieces of a SAOM, the rate function/parameter and the objective function, each contribute to the entries of the intensity matrix to describe the rate of change between two network states.
-The entries of $\mathbf{Q}$ are defined as follows: let $b \neq c \in \{1, 2, \dots, 2^{n(n-1)} \}$ be indices of two different possible states of the network, $x^b, x^c \in \mathcal{X}$.
-Then the $bc^{th}$ entry of $Q$ is:
-% \[ q_{bc} = \begin{cases}
-%      q_{ij}= \alpha_m p_{ij}(\boldsymbol{\beta}, x^b) & \text{if } x^c \in \{x^b(i \leadsto j) | \text{ any } i \neq j \in \{1, \dots, n\}\} \\
-%       0 & \text{if } \sum_i \sum_j |x_{ij}^c -x_{ij}^b | > 1 \\
-%       -\sum_{i\neq j} q_{ij} & \text{if } x^b = x^c
-%    \end{cases}
-% \]
-% 
-% Thus, the rate of change between any two states, $x^b$ and $x^c$, that differ by only one tie $x_{ij}$, is the product of the rate at which actor $i$ gets to change a tie and the probability that the tie that will change is the tie to node $j$.
-This matrix $\mathbf{Q}$ is the foundation for estimation of a SAOM.
-%Estimating the parameters in these models is difficult, but thanks to the SIENA software \citep{RSiena}, we have an accessible way to fit SAOMs to network observations.
-% 
-% %XXX HH can you help? I need a way better transition than that XXX
-% }
-
-%### A SAOM as a CTMC {#saomctmc}
-
-%In order to fit this model definition back into the original context of the CTMC described in Section \@ref(dynamicnets), it must be written in terms of its intensity matrix, $\mathbf{Q}$.
-This matrix describes the rate of change between states of the process.
-For networks, there are a very large number of possible states, $2^{n(n-1)}$, so the intensity matrix is a square matrix of that dimension.
-But, thanks to the property of SAOMs that the states are allowed to change only one tie at a time, there are only $n$ possible states given the current state, $n-1$ of which are uniquely determined by the node $i$ that is given the opportunity to change.
-Thus, the intensity matrix $\mathbf{Q}$ is very sparse, with only $n(n-1) + 1$ non-zero entries in each row.
-Note that $n(n-1)$ of these represent the possible states that are one edge different from a given state, and the additional non-zero entry is for the state to remain the same.
-All other entries in a row are zero because those column states cannot be reached from the row state by just one change as dictated by the SAOM.
-The entries of $\mathbf{Q}$ are defined as follows: let $b \neq c \in \{1, 2, \dots, 2^{n(n-1)} \}$ be indices of two different possible states of the network, $x^b, x^c \in \mathcal{X}$.
-Then the $bc^{th}$ entry of $Q$ is:
-% \[ q(x^b, x^c) = \begin{cases}
-%       q_{ij}(\alpha, \rho, \boldsymbol{\beta}, x^b) = \lambda_i(\alpha, \rho, x^b, m)p_{ij}(\boldsymbol{\beta}, x^b) & \text{if } x^c \in \{x^b(i \leadsto j) | \text{ any } i \neq j \in \mathcal{N}\} \\
-%       0 & \text{if } x^c \text{ differs from } x^b \text{ by more than 1 tie} \\
-%       -\sum_{i\neq j} q_{ij}(\alpha, \rho, \boldsymbol{\beta}, x^b)  & \text{if } x^b = x^c
-%    \end{cases}
-% \]
-%
-% Thus, the rate of change between any two states that differ by only one tie, $x_{ij}$, is the product of the rate at which actor $i$ gets to change a tie and the probability that the tie that will change is the tie to node $j$.^[Just to be clear, the change is from $x^b_{ij}$ to $x^c_{ij} = 1 - x^b_{ij}$.] Furthermore, the theory of continuous time Markov chains gives that the matrix of transition probabilities between observation times $t_{m-1}$ and $t_{m}$ is dependent only on the difference between timepoints, $t_m - t_{m-1}$.
-Following the same definition for transition probabilities in Section \@ref(dynamicnets), the matrix of transition probabilities is
-% $$e^{(t_m - t_{m-1})\mathbf{Q}},$$
-% where $\mathbf{Q}$ is the matrix defined above and $e^X$ for a real or complex square matrix $X$ is equal to $\sum_{k=0}^{\infty} \frac{1}{k!} X^k$.
 
@@
 
-%\subsection{Model Goodness-of-Fit}\label{sec:modgof}
-
-%The \texttt{RSiena} software that fits the models to data also includes a goodnes-of-fit function for examining model fit, \texttt{sienaGOF()}.
-This function ``assess[es] the fit of the model with respect to auxiliary statistics of networks" \citep[p.~53]{RSienamanual}.
-Examples of auxiliary statistics include the out- or indegree distribution on the nodes, with the option for users to input their own statistics to examine.
-The goodness-of-fit is evaluated as follows:
-% \begin{enumerate}
-% \item The auxiliary statistics are computed on the observed data and $N$ simulated observations from the fitted model. Typically, $N=1000$.
-% \item The mean vector and covariance matrix of the statistics on the simulations from the model are computed.
-% \item The Mahalanobis distance from the observed statistics to the distribution of the simulated statistics is computed using the mean and covariance found in step 2.
-% \item The Mahalanobis distance from each of the $N$ simulations to the same distribution is computed, and the Mahalanobis distance of the observed data is compared to this distribution of distances.
-% \item An empirical $p$-value is found by computing the proportion of simulated distances found in step 4 that are as large or larger than the Mahalanobis distance from the data.
-A SAOM is thus considered a good fit if $p$ is large.
-% \end{enumerate}
-
-% <<sienagof, fig.cap="Goodness of fit measures for two network statistics computed on the first simulated wave of a SAOM: indegree and outdegree cumulative distributions.
-On the x-axis is the degree value, and on the y-axis is the number of times that degree value occurs in the network.
-\\st{The p-value shown on the x-axis is the proportion of observations simulated from the model that have observed statistic values as large or larger than the values observed in the data.}", cache = TRUE, message = FALSE, warning = FALSE, fig.height=5>>=
-% load("data/ansnullpaper.rda")
-% library(lattice)
-% gof1 <- sienaGOF(ansnull, OutdegreeDistribution, varName ="friendship", join = FALSE)
-% gof2 <- sienaGOF(ansnull, IndegreeDistribution, varName ="friendship", join = FALSE)
-%
-% sims1 <- data.frame(gof1$`Period 1`$Simulations)
-% dat1 <- gof1$`Period 1`$Observations
-% dat1 %>% data.frame() %>% gather(outdegree, val) %>%
-%   mutate(outdegree = parse_number(outdegree)-1) -> dat1
-% p1 <- sims1 %>%
-%   mutate(sim = row_number()) %>%
-%   gather(outdegree, val, X1:X9) %>%
-%   mutate(outdegree = parse_number(outdegree) - 1) %>%
-%   filter(outdegree <= 6) %>%
-%   ggplot(aes(x = outdegree, y = val)) +
-%   geom_boxplot(aes(group = outdegree), size = .5, outlier.shape = "x", outlier.size = 5) +
-%   geom_violin(aes(group = outdegree), bw = 2, fill = NA) +
-%   geom_point(data = dat1, color = 'red') +
-%   geom_line(data= dat1, color = 'red') +
-%   geom_text(data = dat1, aes(label = val), hjust = -.5) +
-%   labs(x = "Outdegree (p = 0.154)", y = "Statistic",
-%        title = "Goodness-of-Fit: Outdegree distribution period 1") +
-%   ThemeNoNet
-%
-% sims2 <- data.frame(gof2$`Period 1`$Simulations)
-% dat2 <- gof2$`Period 1`$Observations
-% dat2 %>% data.frame() %>% gather(indegree, val) %>%
-%   mutate(indegree = parse_number(indegree)-1) -> dat2
-% p2 <- sims2 %>%
-%   mutate(sim = row_number()) %>%
-%   gather(indegree, val, X1:X9) %>%
-%   mutate(indegree = parse_number(indegree) - 1) %>%
-%   filter(indegree <= 6) %>%
-%   ggplot(aes(x = indegree, y = val)) +
-%   geom_boxplot(aes(group = indegree), outlier.shape = "x", outlier.size = 5) +
-%   geom_violin(aes(group = indegree), bw = 2, fill = NA) +
-%   geom_point(data = dat2, color = 'red') +
-%   geom_line(data= dat2, color = 'red') +
-%   geom_text(data = dat2, aes(label = val), hjust = -.5) +
-%   labs(x = "Indegree (p = 0.063)", y = "Statistic",
-%        title = "Goodness-of-Fit: Indegree distribution period 1") +
-%   ThemeNoNet
-%
-%
-% gridExtra::grid.arrange(p1, p2)
-% # sum(gof1$`Period 1`$SimulatedTestStat >= gof1$`Period 1`$ObservedTestStat)/1000
-% # [1] 0.154
-% # > sum(gof2$`Period 1`$SimulatedTestStat >= gof2$`Period 1`$ObservedTestStat)/1000
-% # [1] 0.063
-%
-% @
-
-%The \texttt{plot.sienaGOF()} function allows us to visualize this fit.
-%In Figure~\ref{fig:sienagof}, we provide an example of what the goodness-of-fit plot output looks like for indegree and outdegree statistics for a small model.
-%This function draws a box plot and a violin plot at each value of the statistic of interest observed in the simulations: \st{on the $x$-axis, the out(in)degrees observed in the data ($0, 1, 2, 3, \dots$), and on the $y$-axis, the cumulative number of times that out(in)degree value appears in the simulations and the data.} In order to compare the distribution of the counts of nodes with the specified degree as calculated on the simulated networks to the counts observed in the true data, red points connected by red lines representing the observed values are superimposed on the boxplots.
-If the red points lie "well within" the simulated values, the model is a good fit to the data.
-\st{This plot is not shown because it is not intuitive for understanding network data.
-Boxplots separated by the many outdegree values observed do not communicate \textit{how} the nodes are connected, just that some have more connections than others.
-In order to understand the fit of the model, we should try to understand how well the model captures the\textit{overall} structure of the network, not just one or two summary measurements of that structure.} %The model examined in Figure~\ref{fig:sienagof} appears to do a better job of capturing the outdegree distribution than the indegree distribution of the data.
-%In Section~\ref{sec:minds}, we propose a new way of visualizing goodness-of-fit that uses the traditional node-link diagram to visualize the entire network instead of numerical summaries of the the network.
-
 \section{Example Data and Models}\label{sec:exdatamods}
@@
 <<adjmatviz, fig.align='center', fig.cap="A visualization of the adjacency matrices of the three waves of network observations in the ``Teenage Friends and Lifestyle Study'' data.
-The subset we will be using is outlined in red.", results='hide',fig.show='asis', fig.width=8, fig.height=3, eval = FALSE>>=
+The subset we will be using is outlined in red.", results='hide',fig.show='asis', fig.width=8, fig.height=3, eval = FALSE, purl=FALSE>>=
 source("code/00e_small_friends.R")
@@
 alldat <- rbind(actual1, actual2, actual3)
-#alldat$X1 <- paste0("V", alldat$X1)
 alldat$X1 <- as.factor(alldat$X1)
-#alldat$X2 <- ifelse(!is.na(alldat$X2), paste0("V", alldat$X2), NA)
 alldat$X2 <- as.factor(alldat$X2)
-#write_csv(alldat, "smallfriends4Geomnet.csv")
 alldataviz <- ggplot(data = alldat,
@@
   labs(title = expression(paste(s[i3], "(x,z)")))
-  # labs(title = expression(paste(s[i3], "(x,z)", "=",Sigma[j != h], x[ij], x[ih], x[hj] %*% 
-  #                               I, "(", z[i], "=", z[h]!= z[j], ")")
-  #                         )
-  #      )
 @
@@
   labs(title = expression(paste(s[i4], "(x)")))
-  # labs(title = expression(paste(s[i4], "(x)", "=", "|{j:", x[ij],"=0, ", Sigma[h], x[ih], x[hj] >= 2, "}|" )
-  #                         )
-  #      )
 @
@@
   scale_size_continuous(name="# bills\nauthored" , guide = guide_legend(nrow= 2), breaks = c(0,1,2,3), labels = c(0,30, 60, 90), range = c(.5,3)) +
-  #theme_net() +
   ThemeNet +
@@
 set.seed(4231352)
-#M1sims1000 <- netvizinf::saom_simulate(dat = mydata,
-#                            struct = M1eff,
-#                       parms = as.numeric(M1parms),
-#                       N = 1000)
-#save(M1sims1000, file = "data/M1sims1000.RData")
-
-# M2sims1000 <- netvizinf::saom_simulate(dat = mydata,
-#                             struct = M2eff,
-#                        parms = as.numeric(M2parms),
-#                        N = 1000)
-# save(M2sims1000, file = "data/M2sims1000.RData")
-
-# M3sims1000 <- netvizinf::saom_simulate(dat = mydata,
-#                             struct = M3eff,
-#                        parms = as.numeric(M3parms),
-#                        N = 1000)
-# save(M3sims1000, file = "data/M3sims1000.RData")
-
-#M1simsdf <- netvizinf::sims_to_df(M1sims1000)
-# write_csv(M1simsdf, "data/M1simsdf.csv")
 M1simsdf <- read_csv("data/M1simsdf.csv")
-#M2simsdf <- netvizinf::sims_to_df(M2sims1000)
-#write_csv(M2simsdf, "data/M2simsdf.csv")
 M2simsdf <- read_csv("data/M2simsdf.csv")
-# M3simsdf <- netvizinf::sims_to_df(M3sims1000)
-# write_csv(M3simsdf, "data/M3simsdf.csv")
 M3simsdf <- read_csv("data/M3simsdf.csv")
@@
 
-#762a83
-#af8dc3
-#e7d4e8
-#d9f0d3
-#7fbf7b
-#1b7837
-
 am_t1 <- as.matrix(igraph::as_adj(igraph::graph_from_data_frame(arrange(t1, parse_number(from), parse_number(to)))))[-17,-17]
@@
            directed = T, curvature = .2, labelon=T, size = 5,
-           arrow = arrow(angle = 20, type = "open", length = unit(.2, "cm")), # ecolour = t1_1$color2,
+           arrow = arrow(angle = 20, type = "open", length = unit(.2, "cm")), 
            color = "#252525", hjust = .5, vjust = .5, labelcolour = 'white', fontsize = 4, arrowgap = .02, layout.alg = "fruchtermanreingold",
@@
            directed = T, curvature = .2, labelon=T, size = 5,
-           arrow = arrow(angle = 20, type = "open", length = unit(.2, "cm")), #ecolour = t1_2$color2,
+           arrow = arrow(angle = 20, type = "open", length = unit(.2, "cm")), 
            color = "#252525", hjust = .5, vjust = .5, labelcolour = 'white', fontsize = 4, arrowgap = .02, layout.alg = "fruchtermanreingold",
@@
 Here, the Mahalanobis distance of the cumulative outdegree values computed on a simulated network to the mean value simulated is shown.
-The distance computed from the data is shown as a vertical line.", fig.height=3, out.width='.6\\textwidth', eval = FALSE>>=
+The distance computed from the data is shown as a vertical line.", fig.height=3, out.width='.6\\textwidth', eval = FALSE, purl = FALSE>>=
 
@@
 @
-
-%\st{If, as discussed in Section~\ref{sec:modgof},
-%Figure~\ref{fig:sienagof},
-%summary statistics such as outdegree and indegree distribution computed on the network are of interest, we can look at the computed distances between the data and the model simulations.
-distributions of these statistics computed on networks simulated from the model show the model in the edge data space.
-For an example, see Figure~\ref{fig:gof2}, where the distribution of the log of the Mahalanobis distance is shown for the outdegree distribution auxiliary statistics, and the log Mahalanobis distance for the observed data is shown as a vertical line.
-}
-
-% \begin{itemize}
-% \item WHat is the model? the network model (SAOM model is just one of many)
-% \item What is the data? Edge data, node data, observed time points, unobserved timepoints (CTMC)
-% \item Note: specifically dealing with dynamic networks observed at discrete time points.
-% \end{itemize}
-% What are some things that could go in here? need more data examples; not just the friends data.
-% Harry Potter data??? I want to find my own data, too.}
 
@@
 
-simu2 <- read.csv("data/simulation-1000-M1-M2-M3.csv")
 levels(simu2$parameter) <- c("alpha[1]","alpha[2]","beta[1]","beta[2]","beta[3]","beta[4]")
@@
 <<tableofmeans>>=
-simu2 <- read.csv("data/simulation-1000-M1-M2-M3.csv")
 senM1est <- read.csv("data/congress/senateM1ests.csv")
@@
 % taken out.
-<<getdatam, eval = FALSE>>=
+<<getdatam, eval = FALSE, purl=FALSE>>=
 nulls <- read.csv("data/distribution_null_model.csv")
@@
 % taken out
-<<corplotsSenate, eval = FALSE>>=
+<<corplotsSenate, eval = FALSE, purl = FALSE>>=
 senM1est <- read.csv("data/congress/senateM1ests.csv")
@@
 A spring layout is used.', message = FALSE, warning = FALSE, fig.height=2.5>>=
-# Need old version of dplyr to get this to work:
-# if (!require("dplyr050",character.only = TRUE)) {
-#       devtools::install_github("sctyner/dplyr050")
-# }
-# library(dplyr050)
 friend.data.w1 <- as.matrix(read.table("data/s50_data/s50-network1.dat"))
@@
 
-# running Siena
-#friendship <- sienaDependent(friendshipData)
-#alcohol <- varCovar(drink[20:35,])
-#mydata <- sienaDataCreate(friendship, alcohol)
-#myeffnull <- getEffects(mydata)
-#myalgorithm <- sienaAlgorithmCreate(projname = 's16_3')
-
 ansnullchains %>%
   dplyr::filter(period == 1) %>%  #only look at chains from wave 1 to wave 2
-  #dplyr::group_by(rep) %>%
   dplyr::select(rep, from = ego, to = alter) %>%
@@
 # histogram of number of MS in 1000 reps
-# sfmsall %>% group_by(rep) %>% summarize(noMs = max(ms)) %>% ggplot() + geom_histogram(aes(x = noMs), binwidth = 5)
-
 sfmsall <- read_csv("data/smallfriendsmicrosteps.csv")
@@
 levels(sfmsall_summ$to) <- paste0('V', 1:16)
-#sfmsall_summ <- sfmsall_summ %>% ungroup() %>% mutate(from = as.factor(from), to = as.factor(to))
 sfmsall_summ %>%
@@
 order1 <- gsub("V", "",order1)
-# order2 <- names(sort(abs(pca1$rotation[,1]), decreasing = T))
 
@@
   scale_x_discrete(position = "top") +
-  #scale_y_discrete(name = "from",labels = rev(str_replace(order1, "V", ""))) +
   ThemeNoNet +
@@
   draw_grob(legendp2,x=.8, y=.25, width = .2, height = .4)
-# cowplot::plot_grid(p1,p2a,legendp2, ncol=3)
 @
@@
   facet_wrap(~frame, labeller = 'label_both', nrow = 2)
-#}
-#plots
+
 @
@@
 <<histogramalters, fig.width = 7, fig.height=9, fig.cap="Each panel shows the theoretical (as lines) and empirical (as points) probabilities of the chosen ego node changing its tie to each of the other nodes.
-The color of the line indicates whether the tie from the ego to the alter node is being added, removed, or if there is no change to the network in this step.", eval = FALSE>>=
+The color of the line indicates whether the tie from the ego to the alter node is being added, removed, or if there is no change to the network in this step.", eval = FALSE, purl=FALSE>>=
 ansnullchains %>% group_by(period, rep) %>%
@@
   ThemeNoNet
-#
-# cumprobs %>% ggplot() + geom_histogram(aes(x = alter, weight = count), color = 'black', fill = 'white', binwidth = 1) + scale_x_continuous(labels = paste0("V",1:16), breaks = 0:15) + coord_flip() + facet_wrap(~ego, labeller = "label_both") + theme_gray()
-
-# cprobslist <- split(cumprobs, cumprobs$ego)
-# plots <- list()
-# for(i in 1:length(cprobslist)){
-#   cprobslist[[i]] %>%
-#   ggplot() +
-#   geom_histogram(aes(x = as.numeric(as.character(alter)),
-#                      y = alterProb), stat = 'identity', fill = 'white', color = 'black') +
-#   geom_line(aes(x = as.numeric(as.character(alter)),
-#                 y = csum)) +
-#   geom_point(aes(x = as.numeric(as.character(alter)),
-#                 y = csum)) +
-#   scale_x_continuous(limits = c(-.5, 15), breaks = 0:15, labels = 0:15) +
-#   geom_hline(yintercept = 1, color = 'red', linetype = 'dashed') +
-#   labs(x = "alter", y = (p[ij]), title = paste("Ego node:", names(cprobslist)[i])) + # ylab((p[ij])) +
-#   theme_gray() +
-#     theme(plot.title = element_text(size = 10),
-#           axis.title = element_text(size = 8),
-#           plot.margin = unit(rep(0,4),"pt")) -> p
-#   plots[[i]] <- p
-# }
-# grid.arrange(plots[[1]], plots[[2]], plots[[3]], plots[[4]],
-#              plots[[5]], plots[[6]], plots[[7]], plots[[8]],
-#              #plots[[9]], plots[[10]], plots[[11]], plots[[12]],
-#              #plots[[13]], plots[[14]], plots[[15]], plots[[16]],
-#              ncol = 2, padding = unit(0, 'line'))
 
@@
 
-# probsDat %>% group_by(ego, alter) %>% summarize(count = n()) %>% dim
-# 103, 3
 p1 <- ggplot(data = probsDat,
@@
   scale_x_discrete(position = 'top') +
-  #scale_y_discrete(labels = rev(str_replace(order1, "V", ""))) +
   ThemeNoNet + 
@@
 plot_grid(p1, p2)
-#gridExtra::grid.arrange(p1, p2, nrow = 1)
 @
@@
 We only do the first step of many because each microstep is dependent on the previous network state.
-%By ego node state, we mean the current set of incoming and outgoing ties to the ego node, $i$.
-Put more precisely, let the vector $X_i = (\{x_{ik}\}, \{x_{ki}\})_{k \neq i} \in \{0,1\}^{2(n-1)}$ represent the set of all tie variables involving node $i$ in the current network.
-For $p_{ij}$ to be comparable for any two network states $x$ and $x'$, the value of the vector $X_i$ must be the same in both states.
-It is possible to incorporate this information into our visualizations, but for now we look at only the first step, because we know without needing any complicated conditioning steps that the previous node state $X_i$ is identical in all cases.
 There are 1,000 transitions to examine: the first microstep taken in each one of the simulations. %Another way to view these transition probabilities is through the adjacency matrix visualization. 
@@
 The Figures~\ref{fig:heatmap} and \ref{fig:alledgests} show that the model M1 and the the corresponding fitting process do not explore the data space enough to adequately capture the network change mechanism, a finding we were only able to discern thanks to model visualization techniques.
-%Put more precisely, let the vector $X_i = (x_{ij}|_{j \neq i}, x_{ji}|_{j \neq i}$ for $i \neq j \neq k \neq \ell \in \{1, \dots, n\}$, $p_{ij}$ is not necessarily the same \st{ARGH NOT SURE THIS IS TRUE } if the current state of the network achieved from the first microstep, denoted $X(1)$, is $X(1) = x(i \leadsto \ell)$ than is different in the case of $x_{ij}$ changing in the first step from the case of $x_{ij}$ not changing in the first step.
-%In Figure~\ref{fig:histogramalters}, we present each ego node and the resulting probabilities of tie changes.
-The probability shown by the bars is the theoretical probability, according to the objective function, of the ego node changing its tie to the alter node, while the probability shown by the points is the empirical probability of that change being made.
-The empirical probability is calculated by counting all instances of the ego node first, then computing the proportion of each different alter node change.
-In most cases, they are almost identical, which demonstrates that the algorithm is performing as expected.
-There are, however, many steps that are never taken.
-Some ego nodes, like 1, 4, and 16, really explore the space of all possible outgoing ties.
-On the other hand, nodes like 13 and 15 explore very few possible outgoing ties.
-It is unclear why this would be happening; it may have something to do with the ties in wave 1, where 13 and 15 have multiple connections, while 1 and 16 have none, but there are counterexamples of well connected nodes, such as 12, that explore the space more.
-In addition, we can see that the ties changed most often are removing ties, not adding ties.
-This matches the overall pattern of the data: we see the most ties in the first wave of data, and the least ties in the third wave.
 
@@
 The $x$ axis is the microstep number, with step 0 representing the first wave of data and the final step representing the second wave of data.
-We can see that many edges are underrepresented in this process: they are in the second wave, but never appear in the microsteps.", fig.height=2.5, eval = FALSE>>=
+We can see that many edges are underrepresented in this process: they are in the second wave, but never appear in the microsteps.", fig.height=2.5, eval = FALSE, purl=FALSE>>=
 smallfriends <- read_csv("data/smallfriends4Geomnet.csv")
@@
         axis.text.y = element_blank(), axis.title = element_blank()) #+
-  #labs(x = "Microstep", y = "Edges (most frequent at top)")
-
 # rep 3
@@
 
-%We also wanted to better understand the entire microstep process from the first wave, all the way through to the second wave.
-The number of steps taken from wave 1 to wave 2 varies.
-In one set of 1,000 simulations from Model 1, the smallest number of steps taken was 58, and the longest was 248, with a mean of 106 and a median of 103.
-In the 1,000 simulations, the standard deviation of the number of microsteps was 22.8.
-In Figure~\ref{fig:allsteps}, we see two simulations of the process from wave 1 to wave 2, with wave 1 shown on the left, and wave 2 shown on the right.
-In each of the three plots, the $y$-axis contains the edges sorted by how often they appear in the networks along the way.
-We can see that some edges are there in the beginning, but disappear and never come back, while others appear a few steps in, only to dispappear again.
-There are also some edges that were observed in wave 2 that don't appear at all in the microstep process in a given simulation.
-\st{So, even though the CTMC makes about the right \textit{number} of changes as it was designed to do, the changes it is making are not necessarily in the right direction.}
-
- %In addition, there is not a distinct pattern or structure that we can see in this particular edge visualization.
-The edges that the model starts with appear more often in the first 50 or so microsteps than any other set of edges.
-This makes sense, because we condition on the first wave, but it would be better if the model explored the space of possible edges more than it currently does.
-It could be that the simple model M1 is not complicated enough for the friendship dynamics in the data, or it could be that the chains in the MCMCs ``get stuck" when exploring the parameter space, but we cannot know this .}
-
-% \st{The algorithms here are many!
-% \begin{enumerate}
-% %\item First, the microsteps / CTMC between each time point (gifs, other animations)
-% \item Phases of RSiena: phase 1 = initial parameter estimate
-% \item phase 2 = iterative update of parameters - trace plots
-% \item phase 3 = simulations for convergence diagnostics
-% \end{enumerate}
-%  }
-
-%\hh{The visualizations should address most elements of structure described before, i.e.
-follow along in the setup (for descriptive and teaching purposes), the fitting (understanding the model and interpreting the results) and then diagnostics (how well does the model fit the data - where are the most differences to the actual data/what are the sensitive parameters?) }
-
-% \hh{Some of the visualizations we talked about
-% \begin{enumerate}
-% \item visualization of the MCMC process:  movie of individual actors' step-by-step decisions; could be done with the ndtv package, as long as we can get individual steps out of RSiena
-% \item global picture: 1000 simulated networks + alpha blending should give a pretty good idea of the most important edges
-% %\item global structure + principal components (tensor pca)
-% \end{enumerate}}
-
 \section{Discussion}\label{sec:discussion}
@@
 Our model-vis application has uncovered many properties of SAOMs, and potential abounds for application of model-vis in other areas of network analysis.
-
-%The flexibility of \texttt{RSiena} allows the user to specify which parameters lead to tie creation, and which parameters lead to tie dissolution.
-We have used ``evaluation" parameters, which are agnostic, allowing the data to decide which parameters affect creation or dissolution via the sign \citep{RSienamanual}.
-Finally, SAOMs and \texttt{RSiena} are able to also model behavior change of the actors in the network, which again is a capability we did not explore here.
 
@@
 sen_effs <- read_csv("data/congress/senateModelSigEffects.csv")
-#sen_effs$Waldpval <- parse_number(sen_effs$Waldpval)
 sen_effs$inter1[is.na(sen_effs$inter1)] <- "none"
@@
 
-simu2 <- read.csv("data/simulation-1000-M1-M2-M3.csv")
 simu_spread <- simu2 %>% spread(parameter, estimate)

### 2018-11-14 paper-child.Rnw
CTMC converting

@@
 
-\par One family of models for dynamic social networks are continuous time Markov chain (CTMC) models, introduced in \citet{ctmcorig}.
-These models condition on the past to model the present state of the network.
-%These models incorporate node-level covariate effects in addition to the traditional structural effects, such as in-degree or out-degree, common in ERGMs.
+\par One family of models for dynamic social networks are continuous time Markov chain (CTMC) models, first introduced for networks in \citet{ctmcorig}.
+These models represent a hierarchical approach to network modeling: the changes happen at the smallest possible level, and subsequent choices are conditioned on the past.
+The transition probabilities between two adjacent states of a network depend on the structure of the network, the characteristics of the nodes, or both \citep{saompaper}.
+This means that node-level covariate effects, in addition to common ERGM effects such as reciprocity or transitivity, can be incorporated into the model.
 %These models are different from other dynamic network models because they allow us to add node level information into the model with the traditional network structure information.
@@
 For example, we may assume that teenagers who drink heavily are more likely to be friends with other tennagers who drink heavily.
-The SAOM family allows us to incorporate this behavioral information on shared interests into the modeling process.
-
-\par Unlike other network models, SAOMs are not analytically tractable \citep{saomreview}.
+The CTMC family of models allows us to incorporate this behavioral information on shared interests into the modeling process through a particular subfamily of CTMC models known as stochastic actor-oriented models (SAOMs) \citep{saompaper}.
+
+\par With the added flexibility of CTMC models comes increased estimation difficuly \citep{saomreview}.
 Likelihood functions quickly become very complex due to the inherent dependency structure in the data, and therefore computational methods, such as Markov chain Monte Carlo (MCMC), are used to fit models.
 %SAOMs are typically fit to dynamic social network data using a series of MCMC phases for finding method of moments (MoM) estimates of model parameters.
-For SAOM parameter estimation, we use the software SIENA, and its \texttt{R} implementation \texttt{RSiena} by \citet{RSiena}.
+For CTMC parameter estimation, we use the software SIENA, and its \texttt{R} implementation \texttt{RSiena} by \citet{RSiena}.
 This software is a considerable contribution to the field of social network analysis, but the extensive parameter estimation process is largely hidden from the user.
-To better understand SAOMs and the fitting process in \texttt{RSiena}, we use model visualization, or model-vis \citep{modelvis}, to examine the underlying methods, hidden structures, and simulation results.
-Ultimately, we aim to help researchers working with SAOMs better understand the structure, behavior, fit, and results of these models.
-
-In the remainder of this section, we introduce the structure of network data and two different network visualization (net-vis) techniques.
-Then, we carefully define the many moving pieces that make up a SAOM.
-Finally, we briefly ouline the SAOM fitting process that we use throughout this paper.
+To better understand CTMCs and the fitting process in \texttt{RSiena}, we use model visualization, or model-vis \citep{modelvis}, to examine the underlying methods, hidden structures, and simulation results.
+Ultimately, we aim to help researchers better understand the structure, behavior, fit, and results of CTMC models for social network data.
+
+In the remainder of this section, we briefly introduce the structure of network data and two different network visualization (net-vis) techniques.
+Then, we carefully define the hierarchical model structure of a CTMC.
+Finally, we briefly ouline the CTMC fitting process in \texttt{RSiena} that we use throughout this paper \citep{SienaAlgs}.
 
@@
 A node-link diagram, however, may underrepresent the edge count if the edges $x_{ij}$ and $x_{ji}$ both exist and are drawn on top of one another.
-Ultimately, there is no one ``correct" way to visualize network information, and we will be using both the node-link and adjacency matrix net-vis methods throughout this paper to explore our example data and SAOMs.
-
-\subsection{Defining SAOMs} \label{sec:saoms}
-
-Stochastic Actor-Oriented Models (SAOMs) model \textit{dynamic} networks, so we can use SAOMs to study the change of a network in time while also examining the structural and node covariate efects.
-The two titular properties of SAOMs, stochasticity and actor-orientation, are key to understanding social networks as they exist naturally.
+Ultimately, there is no one ``correct" way to visualize network information, and we will be using both the node-link and adjacency matrix net-vis methods throughout this paper to explore our example data and CTMC models.
+
+\subsection{Defining CTMC Network Models} \label{sec:saoms}
+
+The CTMC network model family is for analyzing \textit{dynamic} networks, so we can use them to study the change of a network in time while also examining the structural and node covariate efects.
+To define CTMCs for our purposes, we follow the general hierarchical definition outlined in \citet{saomreview} and elsewhere (see e.g.\ \citet{snijders01} and \citet{snijdersetal2010}).
+The hierarchy reflects our intution about social networks as they exist naturally.
 Most social networks, even holding constant the set of actors over time, are ever-changing as relationships form or disappear.
 Actors in social networks also have inherent properties that could affect how they change their role within the network, and existing ties may affect formation or dissolution of other ties.
+In this section, we define the data structure, the hierarchy, and intensity matrix of the CTMC.
 

### 2018-11-15 paper-child.Rnw
making more generalizing edits

@@
 Actors in social networks also have inherent properties that could affect how they change their role within the network, and existing ties may affect formation or dissolution of other ties.
-In this section, we define the data structure, the hierarchy, and intensity matrix of the CTMC.
+In this section, we define the data structure, the hierarchal model, and intensity matrix of the CTMC.
 
@@
 We denote the network observation at timepoint $t_m$ by $x(t_m)$ for $m = 1, \dots M$.
-The SAOM assumes that this longitudinal network of discrete observations is embedded within a continuous time Markov chain, (CTMC), which we denote by $X(T)$.
+The model assumes that this longitudinal network of discrete observations is embedded within a CTMC, which we denote by $X(T)$.
 The CTMC is almost entirely unobserved: we assume the first and last network observations are the starting and ending points of the CTMC, i.e.
@@
 Nearly all other steps in the chain are unobserved, with the exception of $x(t_2), \dots, x(t_{M-1})$.
-Unlike the first and last observations of the network, these observations do not have direct correspondence with steps in the continuous time Markov chain.
+Unlike the first and last observations of the network, these observations do not have direct correspondence with steps in the CTMC.
 Thus, the observations $x(t_2), \dots, x(t_{M-1})$ are considered to be ``snapshots" of the network at some point between two steps in the CTMC.
 
-The CTMC process $X(T)$ is a series of single tie changes that happen according to the SAOM's \textit{rate function}, where one actor at a time is ``selected" to make a change.
-Once selected, the actor changes its ties in order to maximize its \textit{objective function} based on the current state and possible future states of the network.
-
-\noindent \textbf{The Rate Function} In general, a SAOM's \emph{rate function}, $\rho(x, \mathbf{z}, \boldsymbol{\alpha})$, is a function of the network at its current state $x$, covariates of interest, $\mathbf{z}$, and some parameters,   $\boldsymbol{\alpha}$.
+The CTMC process $X(T)$ is a series of single tie changes that happen according to the model's \textit{rate function}, the first piece of the hierarchy.
+One actor at a time is ``selected" to make a change, then in the second piece of the hierarchy, it randomly changes its ties according to probabilities derived from the \textit{objective function}.
+
+\noindent \textbf{The Rate Function:} In general, the \emph{rate function}, $\rho(x, \mathbf{z}, \boldsymbol{\alpha})$, is a function of the network at its current state $x$, covariates of interest, $\mathbf{z}$, and some parameters, $\boldsymbol{\alpha}$.
 With this general formulation, each node can have a different rate of change.
@@
 With this simple rate function, we have $M-1$ rate parameters, one for each transition from $x(t_{m})$ to $x(t_{m+1})$ for $m = 1, \dots, M-1$.
-The rate parameter, $\alpha_m$ dictates how often an actor $i$ ``gets an opportunity" to change one of its ties, $x_{ij}$, for $j \in \{1, \dots, n\}$ in the time period from $t_{m}$ to $t_{m+1}$ for $m = 1, \dots, M-1$.
-A SAOM also assumes that the actors $i$ are conditionally independent given their ties, $x_{i1}, \dots, x_{in}$ at the current network state.
+The rate parameter, $\alpha_m$ dictates how often an actor $i$ ``gets an opportunity" to change one of its ties, $x_{ij}$, for $j \neq i \in \{1, \dots, n\}$ in the time period from $t_{m}$ to $t_{m+1}$ for $m = 1, \dots, M-1$.
+The CTMC model assumes that the actors $i$ are conditionally independent given their ties, $x_{i1}, \dots, x_{in}$ at the current network state.
 Let $\tau(i|x,m) $ be the wait time until actor $i$ gets to the opportunity to change its ties.
@@
 
-\noindent \textbf{The Objective Function} Because of the conditional independence assumptions given in Equation~\ref{eq:c1}, the objective function is written separately for each node.
+\noindent \textbf{The Objective Function} Because of the conditional independence assumptions given in Equation~\ref{eq:c1}, the objective function can be written separately for each node.
 Suppose node $i$ is selected to change at the next step in the CTMC.
@@
 These nodes $j$, are referred to as \textit{alter} nodes, and will not change any of their ties at the next step in the CTMC.
-The ego node $i$ in the current network state $x$, wants to make a change that maximizes its \emph{objective function}:
+The ego node $i$ in the current network state $x$, prioritizes making changes that maximize its \emph{objective function}:
 \begin{equation} \label{eq:of1}
@@
 The parameters of the model, $\beta_1, \dots, \beta_K$ correspond to network statistics, $s_{ik}(x, \mathbf{Z})$, for $k = 1,\dots, K$.
-To maximize $f_i$, $i$ will either destroy a tie where $x_{ij} = 1$, create a tie where $x_{ij}=0$, or make no change if all changes will result in lower $f_i$.
+To maximize $f_i$, $i$ will either destroy a tie where $x_{ij} = 1$, create a tie where $x_{ij}=0$, or make no change.% if all changes will result in lower $f_i$.
 
@@
 According to \citet[p.
-371]{snijders01}, there should be at least two parameters included in the model: $\beta_1$ for the outdegree of nodes, and $\beta_2$ for the reciprocity of nodes.
+371]{snijders01}, there should be at least two parameters included in the objective function: $\beta_1$ for the outdegree of nodes, and $\beta_2$ for the reciprocity of nodes.
 The former measures the propensity of nodes with a lot of outgoing ties to form more outgoing ties (the ``rich get richer" effect), while the latter measures the tendency of outgoing ties to be returned within a network.
@@
 \end{tabular}
-\caption{\label{tab:effects} Examples of structural effects included in a SAOM.
+\caption{\label{tab:effects} Examples of structural effects included in the objective function.
 The darker nodes in the Figure column are the ego nodes ($i$), and all others are alters ($j$).
@@
 We explore these probabilities further in Section~\ref{sec:algorithms}.
-The probabilities $p_{ij}$ combined with the rate parameters $\alpha_m$ fully characterize the underlying CTMC that models network change.
-For more details on CTMCs, see e.g.
-\citet{ctmc}.
+
+\par \textbf{Intensity Matrix:} The probabilities $p_{ij}$ combined with the rate parameters $\alpha_m$ fully characterize the underlying CTMC that models network change.
+In the CTMC literature, see for instance \citet{ctmc}, chains are characterized by their  \textit{intensity} matrix $\mathbf{Q}$.
+This matrix describes the rate of change between two states of the CTMC process, and the rows of this matrix always add to zero.
+The state space for a network, denoted $\mathcal{X}$, is set which contains $2^{n(n-1)}$ states: two possible states for an edge, $\{0,1\}$, and $n(n-1)$ edge relationships because the network is directed and we exclude self-ties.
+The intensity matrix is then a square matrix of dimension $2^{n(n-1)}$.
+Only one tie changes at a time in the CTMC, resulting in $n(n-1)$ reachable states from the current network state.
+Thus, the intensity matrix $\mathbf{Q}$ is very sparse, with only $n(n-1) + 1$ non-zero entries in each row.
+Note that $n(n-1)$ of these entries represent the possible states that are one edge different from a given state, while the additional non-zero entry is for the state to remain unchanged.% All other entries in a row are structural zeroes because those network states cannot be reached from the current state in a single change.
+
+The two pieces of the model hierarchy each contribute to the entries of the intensity matrix to describe the rate of change between two network states.
+The entries of $\mathbf{Q}$ are defined as follows: let $b \neq c \in \{1, 2, \dots, 2^{n(n-1)} \}$ be indices of two different possible states of the network, $x^b, x^c \in \mathcal{X}$.
+Then the $bc^{th}$ entry of $Q$ is:
+\[ q_{bc} = \begin{cases}
+     q_{ij}= \alpha_m p_{ij}(\boldsymbol{\beta}, x^b) & \text{if } \sum_{i,j} |x_{ij}^c - x_{ij}^b| = 1\\
+      -\sum_{i\neq j} q_{ij} & \text{if } x^b = x^c
+      0 & \text{otherwise} \\
+   \end{cases}
+\]
+The rate of change between any two states, $x^b$ and $x^c$, that differ by only one tie $x_{ij}$, is the product of the rate at which actor $i$ gets to change a tie and the probability that the tie that will change is the tie to node $j$.
+The intensity matrix $\mathbf{Q}$ fully defines the model for network change, and is the foundation required for parameter estimation.
+%Estimating the parameters in these models is difficult, but thanks to the SIENA software \citep{RSiena}, we have an accessible way to fit SAOMs to network observations.
 
@@
 
-To fit a SAOM to dynamic network data, we apply the \texttt{RSiena} software, which uses simulation methods to estimate parameter values using either MoM or maximum likelihood (ML) estimation \citep{RSiena}.
+To fit a CTMC to dynamic network data, we apply the \texttt{RSiena} software, which uses simulation methods to estimate parameter values using either MoM or maximum likelihood (ML) estimation \citep{RSiena}.
 We chose to use MoM estimation, so we describe that process here, while more information on ML estimation can be found in \citet{SienaAlgs}.
@@
 Section~\ref{sec:collectmodels} follows, in which we explore collections of models.
-Finally, in Section~\ref{sec:algorithms}, we visualize the underlying algorithms in the SAOM fitting process.
+Finally, in Section~\ref{sec:algorithms}, we visualize the underlying algorithms in the CTMC fitting process.
 We conclude with a discussion in Section~\ref{sec:discussion}.
@@
 
-\noindent \textbf{Example Data:} To guide our visual exploration of SAOMs, we use two example data sources.
+\noindent \textbf{Example Data:} To guide our visual exploration of CTMC models for dynamic networks, we use two example data sources.
 The first is a subset of the 50 actor dataset from the ``Teenage Friends and Lifestyle Study" that is provided on the \texttt{RSiena} webpage.
@@
 We can see that, as time passes, the students seem to drink more and become increasingly isolated into smaller groups.
-An analysis of this type of data with a SAOM should capture these dynamics in a way that allows the researcher to draw conclusions about the nature of the network and behavioral forces at play.
+An analysis of this type of data with a CTMC model should capture these dynamics in a way that allows the researcher to draw conclusions about the nature of the network and behavioral forces at play.
 
@@
 
-Because we require binary edges for SAOMs, we focus only on strong collaborations, where $WPC$ is high. For a senate collaboration network $x$, edges are defined as
+Because CTMC models require binary edges, we focus only on strong collaborations, where $WPC$ is high. For a senate collaboration network $x$, edges are defined as
 \begin{equation}\label{eq:wpc2}
@@
 Some prominent names are shown to demonstrate their place in the network.
-As with the friendship example, analyzing the senate data with a SAOM should capture the underlying dynamics, giving insight into the impact that the actor covairates have on the senate's collaboration structure.
+As with the friendship example, analyzing the senate data with a CTMC model should capture the underlying dynamics, giving insight into the impact that the actor covairates have on the senate's collaboration structure.
 By using data sets from very different contexts, we hope to assess some differences in model performance.
 
-\noindent \textbf{Models:} To our example data, we fit three different SAOMs.
-Each SAOM uses a simple rate function, $\alpha_m$, and an objective function with two or three parameters.
+\noindent \textbf{Models:} To our example data, we fit three different models.
+Each model uses a simple rate function, $\alpha_m$, and an objective function with two or three parameters.
 The first model, M1, contains the absolute minimum number of parameters in the objective function $f_i(x)$:
@@
 \end{subfigure}
-\caption{\label{fig:structures} The additional network effects included in our models fit to the friends data.
+\caption{\label{fig:structures} The additional network effects included in the models we fit to the friends data.
 On the left, a JTT between $i$ and $j$, who have different covariate values, and on the right, a N22 between $i$ and $j$, who are indirectly connected by nodes $k$, $h$.}
@@
 But what does this definition mean for dynamic network models?
-We define the \emph{data space} for SAOMs as the combined data structures of the network: the node, edge, and time information.
+We define the \emph{data space} for our models as the combined data structures of the network: the node, edge, and time information.
 We use the \texttt{R} package \texttt{geomnet} \citep{geomnet} to visualize the three data spaces together.
@@
 
-To view a SAOM in the data space, we simulate from the fitted model.
-No single network instance simulated from a SAOM will fully represent the model, so instead we look at an \textit{expectation} from the network model.
-We frequently rely on expected values in statistics, but network models, especially those as complex as SAOMs, lack an expected network measure.
+To view our models in the data space, we first simulate from the fitted model.
+No single network instance simulated from a CTMC will fully represent the model, so instead we look at an \textit{expectation} from the network model.
+We frequently rely on expected values in statistics, but network models, especially those as complex as the CTMC models, lack an expected network measure.
 We could talk about expected values of parameters, but the parameters can be hard to interpret and cannot teach us much about the overall structure of the network.
@@
 
-Another way to view a SAOM in the data space is with \textit{difference networks} between the expected network and the data, as shown in Figure~\ref{fig:summary2}.
+Another way to view our models in the data space is with \textit{difference networks} between the expected network and the data, as shown in Figure~\ref{fig:summary2}.
 We created these networks by computing the difference between the adjacency matrices for the expected network and wave 1, and for wave 2 and the expected network.
@@
 We conclude that the model M1 is a poor fit to our example data, which we have seen very clearly by viewing the model in the data space.
-
-% No longer using this plot
-<<gof2, fig.cap="An example of viewing the distribution of a statistic of interest from a SAOM.
-Here, the Mahalanobis distance of the cumulative outdegree values computed on a simulated network to the mean value simulated is shown.
-The distance computed from the data is shown as a vertical line.", fig.height=3, out.width='.6\\textwidth', eval = FALSE, purl = FALSE>>=
-
-library(lattice)
-gof1 <- sienaGOF(ansnull, OutdegreeDistribution, varName ="friendship", join = FALSE)
-histdat <- data.frame(x = log(gof1$`Period 1`$SimulatedTestStat))
-ggplot(histdat) + geom_histogram(aes(x = x), binwidth = .1, fill = 'white', color = "black") +
-  #geom_density(aes(x = x, y = ..density..*250)) +
-  geom_vline(xintercept = log(gof1$`Period 1`$ObservedTestStat), color = 'red') +
-  ThemeNoNet +
-  labs(x = "log(Statistic Value)", title = "Distribution of log Mahalanobis Distance")
-
-@
 

### 2018-11-20 paper-child.Rnw
generalizations made

@@
 
-\par With the added flexibility of CTMC models comes increased estimation difficuly \citep{saomreview}.
+\par With the added flexibility of CTMC models comes increased estimation difficulty \citep{saomreview}.
 Likelihood functions quickly become very complex due to the inherent dependency structure in the data, and therefore computational methods, such as Markov chain Monte Carlo (MCMC), are used to fit models.
@@
 The teenagers' drinking behavior changes and the network becomes less densely connected over time.
-These are the mechanisms we hope to model with SAOMs.", out.width='.99\\textwidth', fig.height=3, results='hide'>>=
+These are the mechanisms we hope to model with our chosen CTMCs.", out.width='.99\\textwidth', fig.height=3, results='hide'>>=
 cols <- RColorBrewer::brewer.pal(8, "Greys")
@@
 We chose the additional parameters because they were significant ($p$-value $<$ 0.05) according to a Wald-type test provided in the \texttt{RSiena} software \citep{RSienaTest}.
-The M2 model contains an actor-level covariate parameter, and the M3 model contains an additional strutural effect in the objective function.
+The model M2 contains an actor-level covariate parameter, and the model M3 contains an additional strutural effect in the objective function.
 \begin{align}\label{eq:m2m3}
@@
 We apply the three principles of model-vis: view the model in the data space, visualize collections of models, and explore the process of fitting the model, to the models and data introduced in Section~\ref{sec:exdatamods}.
-Since we have already decided on our model family (SAOMs), we shift our focus to the fitted model and the model form.
+Since we have already decided on our model family (CTMCs), we shift our focus to the fitted model and the model form.
 Specifically, we use model-vis to learn more about how the \emph{model form} affects the \emph{fitted model}.
@@
 No single network instance simulated from a CTMC will fully represent the model, so instead we look at an \textit{expectation} from the network model.
-We frequently rely on expected values in statistics, but network models, especially those as complex as the CTMC models, lack an expected network measure.
+We frequently rely on expected values in statistics, but network models, especially those as complex as our CTMC models, lack an expected network measure.
 We could talk about expected values of parameters, but the parameters can be hard to interpret and cannot teach us much about the overall structure of the network.
@@
 This is the second principle of model visualization: ``collections [of models] are more informative" \citep{modelvis}.
-We examine the behavior of SAOMs by visualizing collections of them, which we construct in four ways.
+We examine the behavior of the CTMCs by visualizing collections of them, which we construct in four ways.
 First, we explore the space of all possible models to determine the most significant model parameters.
@@
 Finally, we fit the same model to different data to learn about fitted parameter values under different conditions.
-We chose these four collections because they each explore something different about SAOM behavior: the choice of parameters in the objective function, the effects of underlying simulations, and the resulting fitted values under different circumstances.
+We chose these four collections because they each explore something different about model behavior: the choice of parameters in the objective function, the effects of underlying simulations, and the resulting fitted values under different circumstances.
 %Using the tools in our visualization toolbox, we view these collections to learn about model behavior.%The first considers the many parameters to include in a SAOM, while the second results in different parameter estimates, so we can see how parameter estimates vary.
@@
 
-\noindent \textbf{Exploring the space of all possible models:} The \texttt{RSiena} manual contains over 80 effects to include in the SAOM objective function.
+\noindent \textbf{Exploring the space of all possible models:} The \texttt{RSiena} manual contains over 80 effects to include in the objective function.
 To select objective function parameters for modeling our example data, we searched through all effects available to find \textit{significant} effects.
@@
 
-<<distests, fig.cap="Distribution of fitted parameter values for our three SAOMs.
+<<distests, fig.cap="Distribution of fitted parameter values for our three models.
 The inclusion of $\\beta_3$ or $\\beta_4$ clearly has an effect on the distributions of the rate parameters, $\\alpha_1$ and $\\alpha_2$, and on the other objective function parameters $\\beta_1, \\beta_2$.", fig.height=3>>=
@@
 
-\noindent \textbf{Fitting the same model to the same data:} The simulations required to fit a SAOM lead to inherent variability in the fitted values.
+\noindent \textbf{Fitting the same model to the same data:} The simulations required to fit the CTMC network models lead to inherent variability in the fitted values.
 To examine this variability, we fit models M1, M2, and M3 to the friendship network 1,000 times each.
@@
 By examining the results of models fitted to different data, we better understand the interpretation of the objective function parameters, and we have learned of some potential parameter correlation problems affecting the fit of the models.
-All in all, we have used four different ways of viewing collections of models, which have given us insight into the behavior of SAOMs.
+All in all, we have used four different ways of viewing collections of models, which have given us insight into the behavior of CTMC network models.
 
@@
 
-The last model visualization principle, ``explore the process of model fitting," is crucial to understanding SAOMs \citep{modelvis}.
-When fitting a SAOM in \texttt{RSiena}, several simulation steps are hidden from the user.
-One of the hidden steps in the SIENA MoM algorithm is the \textit{microstep} process, which corresponds to the CTMC described in Section~\ref{sec:saoms}.
+The last model visualization principle, ``explore the process of model fitting," is crucial to understanding our models \citep{modelvis}.
+When fitting models in \texttt{RSiena}, several simulation steps are hidden from the user.
+One of the hidden steps in the SIENA MoM algorithm is the \textit{microstep} process, which is an observed instance of the CTMC described in Section~\ref{sec:saoms}.
 At each microstep, an \textit{ego} node is selected to change, and the chosen node randomly makes one change in its ties according to to the probabilities $p_{ij}$ defined in Equation~\ref{eq:pij}.
@@
 In this network animation, we show the modeled steps of the \textit{unobserved} CTMC process between two observed networks.
-We see each part of the SAOM come into play: first, the selection of the ego node via the rate function, then the result of the actor maximizing its objective function by either deleting or adding a node.
+We see each part of the hierarchical come into play: first, the selection of the ego node via the rate function, then the result of the actor maximizing its objective function by either deleting or adding a node.
 Finally, the layout of the network transitions as edges are removed or added, demonstrating how the overall network structure changes with each of the individual tie changes.
@@
 This effect could only be exacerbated as more steps are taken in the CTMC, leading to
-a large area of data space unexplored by the first step in the SAOM model fitting process.
+a large area of data space unexplored by the first step in the model fitting process.
 We present additional evidence for lack of coverage in Figure~\ref{fig:alledgests} the Appendix.
@@
 
-We have used traditional graphics and net-vis methods to explore the family of stochastic actor-oriented models for dynamic social network data in new ways.
+We have used traditional graphics and net-vis methods to explore a set of continuous time markov chain models for dynamic social network data in new ways.
 The guiding principles of model-vis: viewing the model in the data space, visualizing collections of models, and looking at the underlying algorithms, gave us knowledge and appreciation for these complicated and flexible models.
@@
 When we explore the algorithms, we gain a better understanding of the model and its stochastic behavior.
-Our three-pronged model-vis attack has uncovered many properties of SAOMs.
+Our three-pronged model-vis attack has uncovered many properties of these models.
 
 We have only just begun to scratch the surface of these multi-layered models for social networks.
-The \texttt{RSiena} software is incredibly powerful, and we can fit much more flexible SAOMs than we have examined here.
-If, for example, we think the network structure or an actor covariate affects the rate of change of the network, we can incorporate that belief into the rate function of the SAOM.
+The \texttt{RSiena} software is incredibly powerful, and we can fit much more flexible hierarchical models than we have examined here.
+If, for example, we think the network structure or an actor covariate affects the rate of change of the network, we can incorporate that belief into the rate function.
 In addition, more than one actor-level covariate can be included in the model, and many more than three parameters can be included in the objective function.
 As we have shown, the rate and objective function parameters are often highly correlated, so researchers can now use model-vis techniques to gain a better understanding of these correlations and hopefully find ways to account for their effects in the model.
-In addition, SAOMs can model behavior change of the actors in the network, a capability we did not explore here.
+In addition, extensions of the model can predict behavior change of the actors in the network, a capability we did not explore here.
 
@@
 Thus, some of the visualizations we have presented may not scale up to larger networks.
-Future SAOM and network model-vis research could include interactive and/or three-dimensional graphics, which we did not explore here.
+Future network model-vis research could also include interactive and/or three-dimensional graphics, which we did not explore here.
 Interactivity is the state-of-the-art in data visualization, and network visualization is just starting to catch on \citep[see e.g.\ ][]{refinery}.
@@
 We hope that the principles of model-vis continue to be used for learning more about the fit and behavior of complicated statistical models.
-Our model-vis application has uncovered many properties of SAOMs, and potential abounds for application of model-vis in other areas of network analysis.
+Our model-vis application has uncovered many properties of CTMC models for dynamic social networks, and potential abounds for application of model-vis in other areas of network analysis.
 
@@
 
-<<corplots, fig.cap="A matrix of plots demonstrating the strong correlations between parameter estimate in our SAOMs.
+<<corplots, fig.cap="A matrix of plots demonstrating the strong correlations between parameter estimates.
 The strongest correlation within each model is between $\\beta_1$ and $\\beta_2$.", fig.width=8>>=
@@
 More of the darkest areas of the figure should belong to ties in wave 2, but those are often the lightest.
-This solidifies what we found in Figure~\ref{fig:heatmap}: the model M1 and the SAOM fitting process do not explore the data space enough to adequately capture the network change mechanism.
+This solidifies what we found in Figure~\ref{fig:heatmap}: the fitting process does not explore the data space enough to adequately capture the network change mechanism.
 
 \noindent \textbf{Online supplemental material:} Much of the code used to create the visualizations in this paper has been packaged and made available online at \url{https://github.com/sctyner/netvizinf} and \url{https://github.com/sctyner/geomnet} (also on CRAN).
-Additional, non-packaged code can be found at \url{https://github.com/sctyner/SAOM-removing-blindfold}.
+Additional, non-packaged code can be found at \url{http://bit.ly/ctmc-model-viz}.

### 2018-11-20 paper-child.Rnw
revise abstract

@@
   \begin{cases}
-                                   1 &  \text{if an edge exists } i\to j, \quad i \neq j \\
-                                   0 & \text{otherwise}
+             1 &  \quad \text{if an edge exists } i\to j, \quad i \neq j \\
+             0 & \quad \text{otherwise}
   \end{cases}
@@
 \[ q_{bc} = \begin{cases}
-     q_{ij}= \alpha_m p_{ij}(\boldsymbol{\beta}, x^b) & \text{if } \sum_{i,j} |x_{ij}^c - x_{ij}^b| = 1\\
-      -\sum_{i\neq j} q_{ij} & \text{if } x^b = x^c
-      0 & \text{otherwise} \\
+     q_{ij} = \alpha_m p_{ij}(\boldsymbol{\beta}, x^b) & \text{ if } \sum_{i,j} |x_{ij}^c - x_{ij}^b| = 1 \\
+      -\sum_{i\neq j} q_{ij} & \text{ if } x^b = x^c \\
+      0 & \text{ otherwise} 
    \end{cases}

### 2018-11-20 paper.Rnw
revise abstract

@@
 \graphicspath{{figure/}}
-% \usepackage{endfloat} % Figures to the end of the document
+%\usepackage{endfloat} % Figures to the end of the document
 
@@
 
-%\pdfminorversion=4
+\pdfminorversion=4
 % NOTE: To produce blinded version, replace "0" with "1" below.
@@
 {
-  \title{\bf Visualizing Dynamic Social Network Models}
+  \title{\bf Visual Diagnostics of Dynamic Social Network Models}
   \author{Sam Tyner \hspace{.2cm}\\
@@
   \begin{center}
-    {\LARGE\bf Visualizing Dynamic Social Network Models}
+    {\LARGE\bf Visual Diagnostics of Dynamic Social Network Models}
 \end{center}
@@
 \begin{abstract}
-Stochastic actor-oriented models (SAOMs), first introduced by Snijders (1996), are statistical network models for dynamic social networks.
-Unlike other network models, SAOMs are analytically intractable and require extensive simulation procedures for model fitting.
-In this paper, we use model visualization techniques introduced in Wickham et al (2015) to better understand SAOMs and their behavior.
+Statistical models for networks have been studied for decades, and they have gotten more complex as computing power has grown.
+Some network models, such as those for dynamic social networks networks, are analytically intractable and therefore require extensive simulation procedures for model fitting.
+In this paper, we use model visualization techniques introduced in Wickham et al (2015) to better understand some of these models and their behavior.
 We use a three-pronged visualization attack: (a) network visualizations to view the \emph{model}, (b) visualizations of collections of models to interpret parameters, and (c) visualizations of fitting algorithms to gain insights into the underlying mechanisms.
-With the help of static and dynamic visualizations, we bring the hidden SAOM fitting processes into the foreground, with the goal of a better understanding and higher accessibility of SAOMs by social network analysts.
+With the help of static and dynamic visualizations, we bring the hidden algorithmic fitting processes into the foreground, with the goal of a better understanding and higher accessibility of statistical models for social network analysts.
 \end{abstract}

### 2018-11-26 paper-child.Rnw
make some final changes beforehand

@@
 \textit{Dynamic social networks}, which are observed at many points in time, are the alternative to static networks.
-
 Models for dynamic social networks, however, are more complicated than those for static networks.
 These models must account for both the structure and the change in time.
-But if we can capture the network change mechanisms in time, % are more realistic because they provide mechanisms for analyzing multiple observations of the same network over time.
+But if we can capture the network change mechanisms in time, we can gain a better understanding of the underlying network dynamics that drive our relationships.
+% are more realistic because they provide mechanisms for analyzing multiple observations of the same network over time.
 
@@
 that people with common interests are more likely to become friends.
-For example, we may assume that teenagers who drink heavily are more likely to be friends with other tennagers who drink heavily.
-The CTMC family of models allows us to incorporate this behavioral information on shared interests into the modeling process through a particular subfamily of CTMC models known as stochastic actor-oriented models (SAOMs) \citep{saompaper}.
+For example, we may assume that teenagers who drink heavily are more likely to be friends with other teenagers who drink heavily.
+We can assess our assumptions on shared interests into the modeling process through a particular subfamily of CTMC models known as stochastic actor-oriented models (SAOMs) \citep{saompaper}.
 
@@
 %SAOMs are typically fit to dynamic social network data using a series of MCMC phases for finding method of moments (MoM) estimates of model parameters.
-For CTMC parameter estimation, we use the software SIENA, and its \texttt{R} implementation \texttt{RSiena} by \citet{RSiena}.
+For CTMC parameter estimation, we use the software SIENA, and its \texttt{R} implementation \texttt{RSiena} by \citet{RSiena} \citep{rstats}.
 This software is a considerable contribution to the field of social network analysis, but the extensive parameter estimation process is largely hidden from the user.
@@
 
-\noindent \textbf{The Rate Function:} In general, the \emph{rate function}, $\rho(x, \mathbf{z}, \boldsymbol{\alpha})$, is a function of the network at its current state $x$, covariates of interest, $\mathbf{z}$, and some parameters, $\boldsymbol{\alpha}$.
+\noindent \textbf{Rate Function:} In general, the \emph{rate function}, $\rho(x, \mathbf{z}, \boldsymbol{\alpha})$, is a function of the network at its current state $x$, covariates of interest, $\mathbf{z}$, and some parameters, $\boldsymbol{\alpha}$.
 With this general formulation, each node can have a different rate of change.
@@
 
-\noindent \textbf{The Objective Function} Because of the conditional independence assumptions given in Equation~\ref{eq:c1}, the objective function can be written separately for each node.
+\noindent \textbf{Objective Function:} Because of the conditional independence assumptions given in Equation~\ref{eq:c1}, the objective function can be written separately for each node.
 Suppose node $i$ is selected to change at the next step in the CTMC.
@@
 
-\par \textbf{Intensity Matrix:} The probabilities $p_{ij}$ combined with the rate parameters $\alpha_m$ fully characterize the underlying CTMC that models network change.
+\par \noindent \textbf{Intensity Matrix:} The probabilities $p_{ij}$ combined with the rate parameters $\alpha_m$ fully characterize the underlying CTMC that models network change.
 In the CTMC literature, see for instance \citet{ctmc}, chains are characterized by their  \textit{intensity} matrix $\mathbf{Q}$.
 This matrix describes the rate of change between two states of the CTMC process, and the rows of this matrix always add to zero.
-The state space for a network, denoted $\mathcal{X}$, is set which contains $2^{n(n-1)}$ states: two possible states for an edge, $\{0,1\}$, and $n(n-1)$ edge relationships because the network is directed and we exclude self-ties.
+The state space for a network, denoted $\mathcal{X}$, contains $2^{n(n-1)}$ states: two possible states for an edge, $\{0,1\}$, and $n(n-1)$ edge relationships, as the network is directed and we exclude self-ties.
 The intensity matrix is then a square matrix of dimension $2^{n(n-1)}$.
@@
 Thus, the intensity matrix $\mathbf{Q}$ is very sparse, with only $n(n-1) + 1$ non-zero entries in each row.
-Note that $n(n-1)$ of these entries represent the possible states that are one edge different from a given state, while the additional non-zero entry is for the state to remain unchanged.% All other entries in a row are structural zeroes because those network states cannot be reached from the current state in a single change.
+Note that $n(n-1)$ of these entries represent the possible states that are one edge different from a given state, while the additional non-zero entry is diagonal entry for the current state.% All other entries in a row are structural zeroes because those network states cannot be reached from the current state in a single change.
 
@@
 The entries of $\mathbf{Q}$ are defined as follows: let $b \neq c \in \{1, 2, \dots, 2^{n(n-1)} \}$ be indices of two different possible states of the network, $x^b, x^c \in \mathcal{X}$.
-Then the $bc^{th}$ entry of $Q$ is:
+Then the $bc^{th}$ entry of $\mathbf{Q}$ is:
 \[ q_{bc} = \begin{cases}
-     q_{ij} = \alpha_m p_{ij}(\boldsymbol{\beta}, x^b) & \text{ if } \sum_{i,j} |x_{ij}^c - x_{ij}^b| = 1 \\
+     q_{ij} = \alpha_m p_{ij}(\boldsymbol{\beta}, x^b) & \text{ if } x_{ij}^c = 1-x_{ij}^b \text{ for some } i \neq j \in \{1, 2, \dots, n \} \\
       -\sum_{i\neq j} q_{ij} & \text{ if } x^b = x^c \\
@@
 \]
-The rate of change between any two states, $x^b$ and $x^c$, that differ by only one tie $x_{ij}$, is the product of the rate at which actor $i$ gets to change a tie and the probability that the tie that will change is the tie to node $j$.
+Thus, the rate of change between any two states, $x^b$ and $x^c$, that differ by only one tie $x_{ij}$, is the product of the rate at which actor $i$ gets to change a tie and the probability that the tie that will change is the tie to node $j$.
 The intensity matrix $\mathbf{Q}$ fully defines the model for network change, and is the foundation required for parameter estimation.
@@
 This measure ranges in value from 0 to 1, where $WPC_{ij} = 1$ if senator $i$ is a cosponsor on every one of senator $j$'s bills and $WPC_{ij} = 0$ if senator $i$ is never a cosponsor any of senator $j$'s bills.
-
 <<senateplot, fig.cap='The collaboration network in the four senates during the Obama years, 2009-2016.
@@
 @
-
 Because CTMC models require binary edges, we focus only on strong collaborations, where $WPC$ is high. For a senate collaboration network $x$, edges are defined as
@@
 simu_spread <- simu2 %>% spread(parameter, estimate)
-ggpairs(simu_spread, columns = 3:8, aes(colour=Model, alpha = .5)) + 
+ggpairs(simu_spread, columns = 4:9, aes(colour=Model, alpha = .5)) + 
   theme(plot.title = element_text(size = 10,

### 2019-07-01 paper-child.Rnw
split appendix off into supporting information

@@
 We performed this procedure, which we used to pick M2 and M3, for both the small friendship and the senate collaboration data.
-A visualization of the significant effects for the two example data sets are shown in Figure~\ref{fig:sigtesting} in the Appendix.
+A visualization of the significant effects for the two example data sets are available in the Supporing Information.
 Using this one-at-a-time testing method results in many possible significant parameters of different type and magnitude, demonstrating the flexibility of these models.
@@
 To explore the behavior of these three models, we simulate from them.
-Using the estimated parameter values in Table~\ref{tab:meanests}, we simulated 1,000 observations of waves 2 and 3 ofthe friendship data from each of the three models.
+Using the estimated parameter values in Table~\ref{tab:meanests}, we simulated 1,000 observations of waves 2 and 3 of the friendship data from each of the three models.
 From the simulations, we created an expected network for wave 2 from models M1, M2, and M3 using the same method described in Section~\ref{sec:minds}.
-These expected networks are shown in Figure~\ref{fig:averagenet} in the Appendix, and we see that all models are failing to capture the structure of the second wave, even though the individual parameters were all significant.
+These expected networks are presented in the Supporting Information, and we see that all models are failing to capture the structure of the second wave, even though the individual parameters were all significant.
 %In Section~\ref{sec:samemodsamedat}, we fit these three models to the data 1,000 times, and examine the parameter densities.
@@
 It is not clear, however, that the addition of a parameter to the objective function \textit{should} effect the estimates of the rate parameters.
-This strong and possibly problematic relationship between rate parameter estimates and objective function parameters is further shown in Figure~\ref{fig:corplots} in the Appendix.
+This strong and possibly problematic relationship between rate parameter estimates and objective function parameters is further explored in the Supporting Information.
 When visualizing the results of the same models fit many times to the same data, we can to see potential model behavior problems right away.
@@
 As we did for the friendship data, we estimated the parameters for M1, M2, and M3, when fit to the senate data (see Figure \ref{fig:senateplot}) 100 times each.
-The means and standard deviations of the repeated parameter estimates for all six model-data combinations are given in Table~\ref{tab:meanests}, while corresponding density plots of the objective function parameters are given in Figure~\ref{fig:compareModels} in the Appendix.
+The means and standard deviations of the repeated parameter estimates for all six model-data combinations are given in Table~\ref{tab:meanests}, while corresponding density plots of the objective function parameters are provided in the Supporting Information.
 Looking at the estimates for both data sets, a few patterns emerge.
@@
 a large area of data space unexplored by the first step in the model fitting process.
-We present additional evidence for lack of coverage in Figure~\ref{fig:alledgests} the Appendix.
+We present additional evidence for lack of coverage in the Supporting Information.
 The Figures~\ref{fig:heatmap} and \ref{fig:alledgests} show that the model M1 and the the corresponding fitting process do not explore the data space enough to adequately capture the network change mechanism, a finding we were only able to discern thanks to model visualization techniques.
@@
 
-\section*{Appendix: Additional model visualizations}
-
-\subsection*{Section 4}
-
-<<sigtesting, fig.show = 'hold', out.width='\\textwidth', fig.height=2, fig.cap="Significant effects for the two data sets, at a significance level of 0.10 or lower as calculated by the Wald-type test available in the SIENA software.", error=FALSE>>=
-smfr_effs <- read_csv("data/effects_significance_smallFriends.csv")
-smfr_effs$inter1[is.na(smfr_effs$inter1)] <- "none"
-library(ggrepel)
-set.seed(12345678)
-smfr_effs %>% filter(type == "eval" & Waldpval <=0.10) %>%
-ggplot(aes(x = estimate, y = Waldpval, shape = inter1)) +
-  #geom_errorbarh(aes(xmin = estimate - se, xmax = estimate + se), color = 'black') +
-  geom_vline(xintercept = 0) +
-  geom_point(size = 3, alpha = .75) +
-  #geom_text_repel(aes(label = shortName)) +
-  scale_shape(name = "Interaction\nVariable") +
-  ggtitle("Significant effects at level 0.10 for friendship data") +
-  coord_flip() +
-  ThemeNoNet
-sen_effs <- read_csv("data/congress/senateModelSigEffects.csv")
-sen_effs$inter1[is.na(sen_effs$inter1)] <- "none"
-ggplot(data = sen_effs %>% filter(Waldpval <=0.10), aes(x = estimate, y = Waldpval, shape= inter1)) +
-  # geom_errorbarh(aes(xmin = estimate - se, xmax = estimate + se), color = 'black') +
-  geom_vline(xintercept = 0) +
-  geom_point(size = 3) +
-  #geom_text_repel(aes(label = shortName)) +
-  scale_shape(name = "Interaction\nVariable") +
-  ggtitle("Significant effects at level 0.10 for senate data") +
-  coord_flip() +
-  ThemeNoNet
-@
-
-In Figure~\ref{fig:sigtesting}, we see for the friendship data and the senate data that most of the significant effects have absolute value less than ten.
-In addition, the $p$-values for the effects from the friendship data are more spread out than the $p$-values for the senate data, which are concentrated at about 0.02 or less.
-This suggests that larger data sets tend to result generally in smaller $p$-values, which is consistent with the construction of Wald-type tests.
-
-<<averagenet, results='hide', out.width='\\textwidth',fig.cap='The node-link diagrams from the three expected networks that we calculated are in the top row, and the true wave 1 and wave 2 data are shown in the bottom row above.
-There is some difference between the three models, but overall, these three models struggle to capture the structure in the true second wave of data.', fig.height=5>>=
-
-combinedavgs$linewidth[combinedavgs$cat %in% c("1st Wave", "2nd Wave")] <- .5
-combinedavgs$cat <- as.factor(combinedavgs$cat)
-combinedavgs$cat <- ordered(combinedavgs$cat, levels = c("Model1", "Model2", "Model3", "1st Wave", "2nd Wave"))
-
-ggplot(data = combinedavgs) +
-  geom_net(aes(from_id = from, to_id = to, linewidth= linewidth),
-           directed = T, curvature = .2, labelon=T, size = 3,
-           arrow = arrow(angle = 20, type = "open", length = unit(.2, "cm")), ecolour = "#969696", fiteach = T,
-           color = "#252525", hjust = .5, vjust = .5, labelcolour = 'white', fontsize = 2, arrowgap = .02,
-           singletons = T) +
-  ThemeNet + theme(panel.background = element_rect(color = 'black')) +
-  facet_wrap(~cat)
-@
-
-To create the expected network visualization shown in Figure~\ref{fig:averagenet}, we follow the same procedure as in Section~\ref{sec:minds}, counting occurrences of each realized edge in the simulations.
-The resulting summary network has weighted edges representing the proportion of edge appearances in all 1,000 simulations of wave 2.
-As in Figure~\ref{fig:summnetM1}, edges only appear in the expected networks if they appear more than 5\% of the time in the simulations.
-In Figure~\ref{fig:averagenet}, we show the  expected network from the three models we fit and the first and second waves of data.
-Comparing the three averages to waves 1 and 2, we see that they have very similar structure to wave 1 but that model M2, which included the transitive triplet parameter, results in more strongly connected components.
-The lack of fit is clear: none of the three average networks show node 16 gaining ties as it does in wave two, nor do they show nodes 4 and 7 becoming isolated.
-In model M2, however, the ties to node 7 appear much weaker than in model M1 or model M3, and the nodes $\{10,11,14\}$ are also strongly connected as they are in wave 2 of data, suggesting that M2 may be the best model of the three.
-
-%Outdegree and reciprocity have the same inverse relationship for both data sets.
-For the friendship data, the inclusion of $\\beta_3$ has strong effect on the estimates of the other two parameters, but not for the senate data.
-<<compareModels, fig.cap="Density plots of objective function parameter estimates from repeatedly fitting models M1, M2, and M3 to the example data. ", fig.height = 3>>=
-simu2$data <- "friends"
-senateEsts$data <- "senate"
-allDataEsts <- rbind(senateEsts, simu2)
-allDataEsts$parameter <- as.factor(allDataEsts$parameter)
-allDataEsts$parameter <- ordered(allDataEsts$parameter, levels = c(paste0("alpha", 1:3),paste0("beta", 1:4)))
-ggplot(allDataEsts %>% filter(str_detect(parameter, "beta"))) +
-  geom_density(aes(x = estimate, fill = Model), alpha = .5) +
-  facet_grid(data~parameter, scales = "free", labeller = "label_both") +
-  scale_fill_brewer(palette = "Greys") + 
-  ThemeNoNet +
-  theme(legend.position = 'bottom')
-@
-
-The density plots in Figure~\ref{fig:compareModels} show the distribution of the parameter estimates for the objective function parameters in M1, M2, and M3 the two example data sets we used.
-The most notable difference is in the values of $\beta_1, \beta_2$.
-In the senate data, the inclusion of $\beta_3$ has no effect on the estimates of $\beta_1, \beta_2$, while the opposite is true for the friendship data.
-
-<<corplots, fig.cap="A matrix of plots demonstrating the strong correlations between parameter estimates.
-The strongest correlation within each model is between $\\beta_1$ and $\\beta_2$.", fig.width=8>>=
-
-simu_spread <- simu2 %>% spread(parameter, estimate)
-ggpairs(simu_spread, columns = 4:9, aes(colour=Model, alpha = .5)) + 
-  theme(plot.title = element_text(size = 10,
-                                            face = 'plain', angle = 0, family = "Times New Roman"),
-                  axis.title.x = element_text(size = 10,
-                                            face = 'plain', angle = 0, family = "Times New Roman", inherit.blank = T),
-                  axis.title.y = element_text(size = 10,
-                                            face = 'plain', angle = 90, inherit.blank = T),
-                  axis.text.x= element_text(size = 10,
-                                            face = 'plain', angle = 90, inherit.blank = TRUE),
-                 # axis.text.x.bottom = element_text(size = 10,
-             #                               face = 'plain', angle = 90, inherit.blank = TRUE),
-                  axis.text.y = element_text(size = 10,
-                                            face = 'plain', angle = 0, family = "Times New Roman", inherit.blank = T),
-                #  axis.text.y.right = element_text(size = 12,
-               #                             face = 'plain', angle = 0, family = "Times New Roman", inherit.blank = T),
-                  strip.text.x = element_text(size = 12,
-                                            face = 'plain', angle = 0, family = "Times New Roman",
-                                            margin = margin(t = 3, r = 0, b = 3, l = 0, unit = "pt"), inherit.blank = T),
-                  strip.text.y =element_text(size = 12,
-                                            face = 'plain', family = "Times New Roman", angle = 90,margin = margin(t = 3, r = 3, b = 3, l = 3, unit = "pt"),inherit.blank = T),
-                  strip.background = element_rect(colour = "black", fill = "white")
-                    )
-@
-
-In Figure~\ref{fig:corplots}, we examine correlations between each of pair of parameters within each model and overall.\nocite{ggally} The strongest correlation within each model is between $\beta_1$ and $\beta_2$, with absolute correlation between them greater than 0.90 in all three models.
-The $\beta_1$ parameter is also highly correlated with the $\beta_3$ parameter within model M2, but it is not as highly correlated with the $\beta_4$ parameter in model M3.
-It might therefore be advisable to consider only models that either allow $\beta_1$ or $\beta_2$ but not both.
-Looking at the high correlation with $\alpha$, we might switch to a model without $\beta_1$.
-
-\subsection*{Section 5}
-
-<<alledgests, fig.height = 4, fig.cap="Visualizing all microsteps taken in 1,000 simulations from the model M1.
-The occurrence percent is split up into groups to correspond with its distribution: only about 10\\% of the possible edges appear more than 10\\% of the time in the 1,000 simulations, while about 60\\% appear less than 1\\% of the time.
-The first wave network is shown at microstep 0, and the second wave of the network is shown as the last microstep for comparison.
-We see that it is rare for a microstep process to last longer than 150 steps, and also that the edges that appear past the 150th step tend to be in either the first wave or the second wave.">>=
-# get a plot of all microsteps from w1 to w2 by edge
-alledges <- expand.grid(from = paste0("V", 1:16), to = paste0("V", 1:16))
-alledges$eid <- 1:nrow(alledges)
-
-sfmsall %>% left_join(alledges) -> sfmsall2
-
-sfmsall2 %>% group_by(rep) %>% mutate(isWave2 = (ms == max(ms))) -> sfmsall2
-
-sfmsall2 %>% group_by(from, to, eid) %>%
-  summarize(total =n()) %>%
-  arrange(desc(total)) %>%
-  ungroup %>%
-  mutate(plotorder = row_number()) -> sfmsallorder
-
-sfmsall2 %>% group_by(ms, eid) %>%
-  summarise(count = n()) %>%
-  mutate(alpha = count/1000) %>%
-  arrange(ms, eid) -> sfmsall3
-
-left_join(sfmsall3, sfmsallorder, by = "eid") -> sfmsall4
-
-#add true wave 2 to plot
-wave2 <- na.omit(actual2)
-wave2$from <- paste0("V", wave2$from)
-wave2$to <- paste0("V", wave2$to)
-wave2 <- left_join(wave2, alledges)
-wave2 <- left_join(wave2, sfmsallorder)
-
-sfmsall5 <- sfmsall4 %>% ungroup %>% add_row(ms = max(sfmsall4$ms) + 1,
-        eid = wave2$eid,
-        count = NA,
-        alpha = 1,
-        from = wave2$from,
-        to = wave2$to,
-        total = wave2$total,
-        plotorder = wave2$plotorder)
-sfmsall5$alphaFctr <- cut(sfmsall5$alpha, c(0,.01, .05, .10, .25, .5, .75, 1, 1.1))
-ggplot(data = sfmsall5) +
-  geom_tile(aes(x = ms, y = reorder(eid,-plotorder), fill = alphaFctr)) +
-  #scale_fill_gradient(low = 'grey90', high = '#0868ac', name = "occurrence %") +
-  scale_fill_brewer(palette = "Greys", name = "Occurrence %", labels = c("(0-1]%", "(1-5]%", "(5-10]%","(10-25]%", "(25-50]%", "(50-75]%", "(75-100]%")) +
-  labs(x = "Microstep No.", y = "Edges (ordered by total # occurrences)") +
-  ThemeNoNet +
-  theme(axis.ticks.y = element_blank(), axis.text.y = element_blank(), panel.grid = element_blank())
-@
-
-Finally, we combine 1,000 simulations from model M1 into a visualization that displays the entire microstep process in Figure~\ref{fig:alledgests}.
-To make this visualization, we first assign each possible edge in the network an ID number so that we can keep track of it throughout all microsteps and all simlations.
-Then, we count up the number of times each edge appears in the network throughout the microstep process for each of the 1,000 simulations.
-We also count the number of times an edge occurs at each step number.
-Since the number of microsteps in the process varies, the number of times an edge occurs decreases as the step number increases.
-Next, we compute a proportion, which we call the occurrence percentage, which is the number of times the edge was in the network in step 1, step 2, etc.
-divided by 1,000.
-Finally, we visualize all this information together in Figure~\ref{fig:alledgests}.
-In this plot, we see that all possible edges appear at least once at some point in the microsteps of at least one simulated process.
-We also see, however, that the process struggles to focus in on the edges in the second wave of the data.
-Ideally, we would like to see more occurrences the edges in wave 2 that are not in wave 1.
-But, about half of the edges in wave two are in the bottom half of the figure, which means they do not appear as much as they would if the model was actually capturing the mechanisms of tie change in the network.
-More of the darkest areas of the figure should belong to ties in wave 2, but those are often the lightest.
-This solidifies what we found in Figure~\ref{fig:heatmap}: the fitting process does not explore the data space enough to adequately capture the network change mechanism.
-
-\noindent \textbf{Online supplemental material:} Much of the code used to create the visualizations in this paper has been packaged and made available online at \url{https://github.com/sctyner/netvizinf} and \url{https://github.com/sctyner/geomnet} (also on CRAN).
+\noindent \textbf{Online supplemental material:} In addition to the Supporting Information available online from the publisher, much of the code used to create the visualizations in this paper has been packaged and made available online at \url{https://github.com/sctyner/netvizinf} and \url{https://github.com/sctyner/geomnet} (also on CRAN).
 Additional, non-packaged code can be found at \url{http://bit.ly/ctmc-model-viz}.

### 2019-07-01 supporting-info-child.Rnw
split appendix off into supporting information

<<setup, echo = FALSE, message = FALSE, warning = FALSE, purl=FALSE>>=
options(replace.assign=TRUE,width=70, digits=3)
require(knitr)
opts_chunk$set(fig.path='figure/', cache.path='cache/', fig.align='center', fig.pos='h', out.width='.99\\textwidth', par=TRUE, cache=FALSE, concordance=TRUE, autodep=TRUE, message=F, warning=F, echo = FALSE, dev="cairo_pdf", fig.width = 6, fig.height = 6)#, root.dir = "~/Desktop/Dissertation/SAOM-removing-blindfold/")
@

<<pkgs>>=
# if (packageVersion("ggplot2") > "2.2.0"){
#   remove.packages("ggplot2")
#   install.packages("ggplot2")
# }
library(ggplot2)
library(tidyverse)
library(RSiena)
library(network)
library(sna)
library(geomnet)
library(GGally)
# devtools::install_github("sctyner/netvizinf")
library(netvizinf)
library(RColorBrewer)
library(gridExtra)
library(cowplot)
library(extrafont)
loadfonts(quiet = T)

ThemeNoNet <- theme_bw() %+replace%
            theme(plot.title = element_text(size = 10,
                                            face = 'plain',
                                            angle = 0,
                                            family="Times New Roman"),
                  axis.title.x = element_text(size = 10,
                                            face = 'plain',
                                            angle = 0,
                                            family="Times New Roman"),
                  axis.title.y = element_text(size = 10,
                                            face = 'plain',
                                            angle = 90,
                                            family="Times New Roman"),
                  axis.text.x = element_text(size = 10,
                                            face = 'plain',
                                            angle = 0,
                                            family="Times New Roman"),
                  # axis.text.x.top = element_text(size = 10,
                  #                           face = 'plain',
                  #                           angle = 0,
                  #                           family="Times New Roman"),
                  # axis.text.x.bottom = element_text(size = 10,
                  #                           face = 'plain',
                  #                           angle = 0,
                  #                           family="Times New Roman"),
                  axis.text.y = element_text(size = 10,
                                            face = 'plain',
                                            angle = 0,
                                            family="Times New Roman"),
                  # axis.text.y.left = element_text(size = 10,
                  #                           face = 'plain',
                  #                           angle = 0,
                  #                           family="Times New Roman"),
                  # axis.text.y.right = element_text(size = 10,
                  #                           face = 'plain',
                  #                           angle = 0,
                  #                           family="Times New Roman"),
                  strip.text.x = element_text(size = 10,
                                            face = 'plain',
                                            angle = 0,
                                            family="Times New Roman",
                                            margin = margin(t = 3, r = 0, b = 3, l = 0, unit = "pt")),
                  strip.text.y =element_text(size = 10,
                                            face = 'plain',
                                            angle = 90,
                                            family="Times New Roman",
                                            margin = margin(t = 3, r = 3, b = 3, l = 3, unit = "pt")),
                  strip.background = element_rect(colour = "black", fill = "white")
                    )
ThemeNet <- theme_net() %+replace%
            theme(plot.title = element_text(size = 10,
                                            face = 'plain', angle = 0, family = "Times New Roman"),
                  strip.text.x = element_text(size = 10,
                                            face = 'plain', angle = 0, family = "Times New Roman",
                                            margin = margin(t = 3, r = 0, b = 3, l = 0, unit = "pt")),
                  strip.text.y =element_text(size = 10,
                                            face = 'plain', angle = 0, family = "Times New Roman",margin = margin(t = 3, r = 0, b = 3, l = 0, unit = "pt")),
                  strip.background = element_rect(colour = "black", fill = "white"))
@

<<getdata, echo=FALSE, results="hide">>=
set.seed(823746)
load("data/ansnullpaper.rda")
sfmsall <- read_csv("data/smallfriendsmicrosteps.csv")
ansnullchains <- get_chain_info(ansnull)
load("data/M1sims1000.RData")
load("data/M2sims1000.RData")
load("data/M3sims1000.RData")

# from chunk "alldataviz" in paper-child.Rnw
cols <- RColorBrewer::brewer.pal(8, "Greys")

source("code/00e_small_friends.R")
view_rs2$wave <- factor(view_rs2$wave)
levels(view_rs2$wave) <- c("Wave 1", "Wave 2", "Wave 3")
actual1 <- merge(data.frame(as.edgelist(as.network(fd2.w1))),
                 data.frame(id = 1:16, drink = drink2[,1]),
                 by.x = "X1", by.y = "id", all = T)
actual1$Wave <- 1
actual2 <- merge(data.frame(as.edgelist(as.network(fd2.w2))),
                 data.frame(id = 1:16,drink = drink2[,2]),
                 by.x = "X1", by.y = "id", all = T)
actual2$Wave <- 2
actual3 <- merge(data.frame(as.edgelist(as.network(fd2.w3))),
                 data.frame(id = 1:16,drink = drink2[,3]),
                 by.x = "X1", by.y = "id", all = T)
actual3$Wave <- 3
alldat <- rbind(actual1, actual2, actual3)
alldat$X1 <- as.factor(alldat$X1)
alldat$X2 <- as.factor(alldat$X2)

# from chunk "makinggif" in paper-child.Rnw
friend.data.w1 <- as.matrix(read.table("data/s50_data/s50-network1.dat"))
friend.data.w2 <- as.matrix(read.table("data/s50_data/s50-network2.dat"))
friend.data.w3 <- as.matrix(read.table("data/s50_data/s50-network3.dat"))
drink <- as.matrix(read.table("data/s50_data/s50-alcohol.dat"))
fd2.w1 <- friend.data.w1[20:35,20:35]
fd2.w2 <- friend.data.w2[20:35,20:35]
fd2.w3 <- friend.data.w3[20:35,20:35]
friendshipData <- array(c(fd2.w1, fd2.w2,fd2.w3), dim = c(16, 16, 3))

ansnullchains %>%
  dplyr::filter(period == 1) %>%  #only look at chains from wave 1 to wave 2
  dplyr::select(rep, from = ego, to = alter) %>%
  dplyr::mutate(val = as.numeric(!from == to),
         from = paste0("V", parse_number(as.character(from))+1), # make the chains
         to = paste0("V", parse_number(as.character(to))+1)) -> ansnullchainsw1w2
colnames(fd2.w1) <- paste0("V", 1:16)
rownames(fd2.w1) <- paste0("V", 1:16)
wave1friends <- fortify(as.adjmat(fd2.w1))

# from the paper-child.Rnw chunk number summnetM1
friendshipData <- array(c(fd2.w1, fd2.w2, fd2.w3), dim = c(16, 16, 3))
friendshipData <- array(c(fd2.w1, fd2.w2, fd2.w3), dim = c(16, 16, 3))
drink1 <- drink[20:35,1]
drink2 <- drink[20:35,2]
drink3 <- drink[20:35,3]
friendship <- sienaDependent(friendshipData)
alcohol <- varCovar(cbind(drink1,drink2, drink3))
mydata <- sienaDataCreate(friendship, alcohol)
M1eff <- getEffects(mydata)
M2eff <- includeEffects(M1eff, jumpXTransTrip, interaction1 = "alcohol")
M3eff <- includeEffects(M1eff, nbrDist2twice)
simu2 <- read.csv("data/simulation-1000-M1-M2-M3.csv")
means <- simu2 %>% group_by(Model, parameter) %>% summarize(
  means = mean(estimate)
)
M1parms <- (means %>% filter(Model == "M1"))$means
M2parms <- (means %>% filter(Model == "M2"))$means
M3parms <- (means %>% filter(Model == "M3"))$means
set.seed(4231352)
M1simsdf <- read_csv("data/M1simsdf.csv")
M2simsdf <- read_csv("data/M2simsdf.csv")
M3simsdf <- read_csv("data/M3simsdf.csv")

M1avgW2 <- M1simsdf %>% filter(!is.na(to) & wave == 1) %>%
  group_by(from, to) %>%
  summarise(count = n()) %>%
  mutate(weight = ifelse(from == to, 0, count))
M2avgW2 <- M2simsdf %>% filter(!is.na(to) & wave == 1) %>%
  group_by(from, to) %>%
  summarise(count = n()) %>%
  mutate(weight = ifelse(from == to, 0, count))
M3avgW2 <- M3simsdf %>% filter(!is.na(to) & wave == 1) %>%
  group_by(from, to) %>%
  summarise(count = n()) %>%
  mutate(weight = ifelse(from == to, 0, count))

# make a df of wave 2, wave 1, and the three averages and facet.
names(actual1)[1:2] <- c("from", "to")
names(actual2)[1:2] <- c("from", "to")
actual1$count <- 1
actual1$weight <- 1
actual2$count <- 1
actual2$weight <- 1
actual1 <- actual1 %>% dplyr::select(from,to, count, weight)
actual2 <- actual2 %>% dplyr::select(from,to, count, weight)
actual1$cat <- "1st Wave"
actual2$cat <- "2nd Wave"

avgW2M1 <- M1avgW2 %>% ungroup() %>%
  filter(weight > 50) %>%
  mutate(from = as.factor(from), to = as.factor(to),
         cat = "Model1")
add1 <- setdiff(as.character(1:16), unique(c(as.character(avgW2M1$from), as.character(avgW2M1$to))))
avgW2M1 %>% add_row(from = add1, to = NA, count = NA, weight = NA, cat = "Model1") -> avgW2M1
avgW2M2 <- M2avgW2 %>% ungroup() %>%
  filter(weight > 50) %>%
  mutate(from = as.factor(from), to = as.factor(to),
         cat = "Model2")
add2 <- setdiff(as.character(1:16), unique(c(as.character(avgW2M2$from), as.character(avgW2M2$to))))
avgW2M2 %>% add_row(from = add2, to = NA, count = NA, weight = NA, cat = "Model2") -> avgW2M2
avgW2M3 <- M3avgW2 %>% ungroup() %>%
  filter(weight > 50) %>%
  mutate(from = as.factor(from), to = as.factor(to),
         cat = "Model3")
add3 <- setdiff(as.character(1:16), unique(c(as.character(avgW2M3$from), as.character(avgW2M3$to))))
avgW2M3 %>% add_row(from = add3, to = NA, count = NA, weight = NA, cat = "Model3") -> avgW2M3

combinedavgs <- rbind(actual1, actual2, avgW2M1, avgW2M2, avgW2M3)
combinedavgs %>% group_by(cat) %>%
  mutate(linewidth = weight / max(weight,na.rm = T)) -> combinedavgs

combinedavgs %>% filter(cat == "Model1") %>% mutate(logweight = log(weight)) -> t1
colors <- tweenr::tween_color(data = c("#969696", "#d73027"), n = table(t1$cat) %>% as.numeric, ease = 'linear')
t1 %>% arrange(logweight) -> t1
t1$color <- NA
t1$color <- colors[[1]]

# from chunk tableofmeans in paper-child.Rnw
senM1est <- read.csv("data/congress/senateM1ests.csv")
senM2est <- read.csv("data/congress/senateM2ests.csv")
senM3est <- read.csv("data/congress/senateM3ests.csv")
senM1est$Model <- "M1"
senM2est$Model <- "M2"
senM3est$Model <- "M3"
senM1est$Sim <- 1:nrow(senM1est)
senM2est$Sim <- 1:nrow(senM2est)
senM3est$Sim <- 1:nrow(senM3est)
senM1est %>% gather(parameter, estimate, Rate:recip) %>% dplyr::select(Sim, Model, parameter, estimate) -> senM1est2
senM2est %>% gather(parameter, estimate, Rate:jumpXTransTrip) %>% dplyr::select(Sim, Model, parameter, estimate) -> senM2est2
senM3est %>% gather(parameter, estimate, Rate:nbrDist2twice) %>% dplyr::select(Sim, Model, parameter, estimate) -> senM3est2
senateEsts <- rbind(senM1est2, senM2est2, senM3est2)
senateEsts$parameter <- as.factor(senateEsts$parameter)
levels(senateEsts$parameter) <- c("beta1", "beta3", "beta4", "alpha1", "alpha2", "alpha3", "beta2")
@

\section{Supporting Information for Section 4: Collections of models}

<<sigtesting, fig.show = 'hold', out.width='\\textwidth', fig.height=2, fig.cap="Significant effects for the two data sets, at a significance level of 0.10 or lower as calculated by the Wald-type test available in the SIENA software.", error=FALSE>>=
smfr_effs <- read_csv("data/effects_significance_smallFriends.csv")
smfr_effs$inter1[is.na(smfr_effs$inter1)] <- "none"
library(ggrepel)
set.seed(12345678)
smfr_effs %>% filter(type == "eval" & Waldpval <=0.10) %>%
ggplot(aes(x = estimate, y = Waldpval, shape = inter1)) +
  #geom_errorbarh(aes(xmin = estimate - se, xmax = estimate + se), color = 'black') +
  geom_vline(xintercept = 0) +
  geom_point(size = 3, alpha = .75) +
  #geom_text_repel(aes(label = shortName)) +
  scale_shape(name = "Interaction\nVariable") +
  ggtitle("Significant effects at level 0.10 for friendship data") +
  coord_flip() +
  ThemeNoNet
sen_effs <- read_csv("data/congress/senateModelSigEffects.csv")
sen_effs$inter1[is.na(sen_effs$inter1)] <- "none"
ggplot(data = sen_effs %>% filter(Waldpval <=0.10), aes(x = estimate, y = Waldpval, shape= inter1)) +
  # geom_errorbarh(aes(xmin = estimate - se, xmax = estimate + se), color = 'black') +
  geom_vline(xintercept = 0) +
  geom_point(size = 3) +
  #geom_text_repel(aes(label = shortName)) +
  scale_shape(name = "Interaction\nVariable") +
  ggtitle("Significant effects at level 0.10 for senate data") +
  coord_flip() +
  ThemeNoNet
@

In Figure~\ref{fig:sigtesting}, we see for the friendship data and the senate data that most of the significant effects have absolute value less than ten. In addition, the $p$-values for the effects from the friendship data are more spread out than the $p$-values for the senate data, which are concentrated at about 0.02 or less. This suggests that larger data sets tend to result generally in smaller $p$-values, which is consistent with the construction of Wald-type tests.

<<averagenet, results='hide', out.width='\\textwidth',fig.cap='The node-link diagrams from the three expected networks that we calculated are in the top row, and the true wave 1 and wave 2 data are shown in the bottom row above. There is some difference between the three models, but overall, these three models struggle to capture the structure in the true second wave of data.', fig.height=5>>=

combinedavgs$linewidth[combinedavgs$cat %in% c("1st Wave", "2nd Wave")] <- .5
combinedavgs$cat <- as.factor(combinedavgs$cat)
combinedavgs$cat <- ordered(combinedavgs$cat, levels = c("Model1", "Model2", "Model3", "1st Wave", "2nd Wave"))

ggplot(data = combinedavgs) +
  geom_net(aes(from_id = from, to_id = to, linewidth= linewidth),
           directed = T, curvature = .2, labelon=T, size = 3,
           arrow = arrow(angle = 20, type = "open", length = unit(.2, "cm")), ecolour = "#969696", fiteach = T,
           color = "#252525", hjust = .5, vjust = .5, labelcolour = 'white', fontsize = 2, arrowgap = .02,
           singletons = T) +
  ThemeNet + theme(panel.background = element_rect(color = 'black')) +
  facet_wrap(~cat)
@

To create the expected network visualization shown in Figure~\ref{fig:averagenet}, we follow the same procedure as in Section 3: Model in Data Space, counting occurrences of each realized edge in the simulations.  The resulting summary network has weighted edges representing the proportion of edge appearances in all 1,000 simulations of wave 2. As in Figure 6, Section 3, edges only appear in the expected networks if they appear more than 5\% of the time in the simulations. In Figure~\ref{fig:averagenet}, we show the  expected network from the three models we fit and the first and second waves of data. Comparing the three averages to waves 1 and 2, we see that they have very similar structure to wave 1 but that model M2, which included the transitive triplet parameter, results in more strongly connected components. The lack of fit is clear: none of the three average networks show node 16 gaining ties as it does in wave two, nor do they show nodes 4 and 7 becoming isolated. In model M2, however, the ties to node 7 appear much weaker than in model M1 or model M3, and the nodes $\{10,11,14\}$ are also strongly connected as they are in wave 2 of data, suggesting that M2 may be the best model of the three.

%Outdegree and reciprocity have the same inverse relationship for both data sets. For the friendship data, the inclusion of $\\beta_3$ has strong effect on the estimates of the other two parameters, but not for the senate data.
<<compareModels, fig.cap="Density plots of objective function parameter estimates from repeatedly fitting models M1, M2, and M3 to the example data. ", fig.height = 3>>=
simu2$data <- "friends"
senateEsts$data <- "senate"
allDataEsts <- rbind(senateEsts, simu2)
allDataEsts$parameter <- as.factor(allDataEsts$parameter)
allDataEsts$parameter <- ordered(allDataEsts$parameter, levels = c(paste0("alpha", 1:3),paste0("beta", 1:4)))
ggplot(allDataEsts %>% filter(str_detect(parameter, "beta"))) +
  geom_density(aes(x = estimate, fill = Model), alpha = .5) +
  facet_grid(data~parameter, scales = "free", labeller = "label_both") +
  scale_fill_brewer(palette = "Greys") + 
  ThemeNoNet +
  theme(legend.position = 'bottom')
@

The density plots in Figure~\ref{fig:compareModels} show the distribution of the parameter estimates for the objective function parameters in M1, M2, and M3 the two example data sets we used. The most notable difference is in the values of $\beta_1, \beta_2$. In the senate data, the inclusion of $\beta_3$ has no effect on the estimates of $\beta_1, \beta_2$, while the opposite is true for the friendship data.

<<corplots, fig.cap="A matrix of plots demonstrating the strong correlations between parameter estimates. The strongest correlation within each model is between $\\beta_1$ and $\\beta_2$.", fig.width=8>>=

simu_spread <- simu2 %>% spread(parameter, estimate)
ggpairs(simu_spread, columns = 4:9, aes(colour=Model, alpha = .5)) + 
  theme(plot.title = element_text(size = 10,
                                            face = 'plain', angle = 0, family = "Times New Roman"),
                  axis.title.x = element_text(size = 10,
                                            face = 'plain', angle = 0, family = "Times New Roman", inherit.blank = T),
                  axis.title.y = element_text(size = 10,
                                            face = 'plain', angle = 90, inherit.blank = T),
                  axis.text.x= element_text(size = 10,
                                            face = 'plain', angle = 90, inherit.blank = TRUE),
                 # axis.text.x.bottom = element_text(size = 10,
             #                               face = 'plain', angle = 90, inherit.blank = TRUE),
                  axis.text.y = element_text(size = 10,
                                            face = 'plain', angle = 0, family = "Times New Roman", inherit.blank = T),
                #  axis.text.y.right = element_text(size = 12,
               #                             face = 'plain', angle = 0, family = "Times New Roman", inherit.blank = T),
                  strip.text.x = element_text(size = 12,
                                            face = 'plain', angle = 0, family = "Times New Roman",
                                            margin = margin(t = 3, r = 0, b = 3, l = 0, unit = "pt"), inherit.blank = T),
                  strip.text.y =element_text(size = 12,
                                            face = 'plain', family = "Times New Roman", angle = 90,margin = margin(t = 3, r = 3, b = 3, l = 3, unit = "pt"),inherit.blank = T),
                  strip.background = element_rect(colour = "black", fill = "white")
                    )
@

In Figure~\ref{fig:corplots}, we examine correlations between each of pair of parameters within each model and overall.\nocite{ggally} The strongest correlation within each model is between $\beta_1$ and $\beta_2$, with absolute correlation between them greater than 0.90 in all three models. The $\beta_1$ parameter is also highly correlated with the $\beta_3$ parameter within model M2, but it is not as highly correlated with the $\beta_4$ parameter in model M3. It might therefore be advisable to consider only models that either allow $\beta_1$ or $\beta_2$ but not both. Looking at the high correlation with $\alpha$, we might switch to a model without $\beta_1$.

\section{Supporting Information for Section 5: Explore algorithms, not just end result}

<<alledgests, fig.height = 4, fig.cap="Visualizing all microsteps taken in 1,000 simulations from the model M1. The occurrence percent is split up into groups to correspond with its distribution: only about 10\\% of the possible edges appear more than 10\\% of the time in the 1,000 simulations, while about 60\\% appear less than 1\\% of the time. The first wave network is shown at microstep 0, and the second wave of the network is shown as the last microstep for comparison. We see that it is rare for a microstep process to last longer than 150 steps, and also that the edges that appear past the 150th step tend to be in either the first wave or the second wave.">>=
# get a plot of all microsteps from w1 to w2 by edge
alledges <- expand.grid(from = paste0("V", 1:16), to = paste0("V", 1:16))
alledges$eid <- 1:nrow(alledges)

sfmsall %>% left_join(alledges) -> sfmsall2

sfmsall2 %>% group_by(rep) %>% mutate(isWave2 = (ms == max(ms))) -> sfmsall2

sfmsall2 %>% group_by(from, to, eid) %>%
  summarize(total =n()) %>%
  arrange(desc(total)) %>%
  ungroup %>%
  mutate(plotorder = row_number()) -> sfmsallorder

sfmsall2 %>% group_by(ms, eid) %>%
  summarise(count = n()) %>%
  mutate(alpha = count/1000) %>%
  arrange(ms, eid) -> sfmsall3

left_join(sfmsall3, sfmsallorder, by = "eid") -> sfmsall4

#add true wave 2 to plot
wave2 <- na.omit(actual2)
wave2$from <- paste0("V", wave2$from)
wave2$to <- paste0("V", wave2$to)
wave2 <- left_join(wave2, alledges)
wave2 <- left_join(wave2, sfmsallorder)

sfmsall5 <- sfmsall4 %>% ungroup %>% add_row(ms = max(sfmsall4$ms) + 1,
        eid = wave2$eid,
        count = NA,
        alpha = 1,
        from = wave2$from,
        to = wave2$to,
        total = wave2$total,
        plotorder = wave2$plotorder)
sfmsall5$alphaFctr <- cut(sfmsall5$alpha, c(0,.01, .05, .10, .25, .5, .75, 1, 1.1))
ggplot(data = sfmsall5) +
  geom_tile(aes(x = ms, y = reorder(eid,-plotorder), fill = alphaFctr)) +
  #scale_fill_gradient(low = 'grey90', high = '#0868ac', name = "occurrence %") +
  scale_fill_brewer(palette = "Greys", name = "Occurrence %", labels = c("(0-1]%", "(1-5]%", "(5-10]%","(10-25]%", "(25-50]%", "(50-75]%", "(75-100]%")) +
  labs(x = "Microstep No.", y = "Edges (ordered by total # occurrences)") +
  ThemeNoNet +
  theme(axis.ticks.y = element_blank(), axis.text.y = element_blank(), panel.grid = element_blank())
@

Finally, we combine 1,000 simulations from model M1 into a visualization that displays the entire microstep process in Figure~\ref{fig:alledgests}. To make this visualization, we first assign each possible edge in the network an ID number so that we can keep track of it throughout all microsteps and all simlations. Then, we count up the number of times each edge appears in the network throughout the microstep process for each of the 1,000 simulations. We also count the number of times an edge occurs at each step number. Since the number of microsteps in the process varies, the number of times an edge occurs decreases as the step number increases. Next, we compute a proportion, which we call the occurrence percentage, which is the number of times the edge was in the network in step 1, step 2, etc. divided by 1,000. Finally, we visualize all this information together in Figure~\ref{fig:alledgests}. In this plot, we see that all possible edges appear at least once at some point in the microsteps of at least one simulated process. We also see, however, that the process struggles to focus in on the edges in the second wave of the data. Ideally, we would like to see more occurrences the edges in wave 2 that are not in wave 1. But, about half of the edges in wave two are in the bottom half of the figure, which means they do not appear as much as they would if the model was actually capturing the mechanisms of tie change in the network. More of the darkest areas of the figure should belong to ties in wave 2, but those are often the lightest. This solidifies what we found in Figure 12, Section 5: the fitting process does not explore the data space enough to adequately capture the network change mechanism.

### 2019-07-02 paper-child.Rnw
ditch senate data example, clarity in notation, some other minor edits, merge table 1 & figure 4

@@
 %As humans and researchers, it is our instint to try to learn as much as we can about the world around us. 
-As members of the human family, one of the hardest things we try to do is understand our relationships with one another, whether as friends, family, or colleagues.
+One of the hardest things scholars try to do is understand relationships between friends, family, or colleagues.
 These relationships, or \emph{social networks}, have been studied for decades, in every format from self-help books to publications in academic journals.
@@
 \textit{Dynamic social networks}, which are observed at many points in time, are the alternative to static networks.
-Models for dynamic social networks, however, are more complicated than those for static networks.
-These models must account for both the structure and the change in time.
-But if we can capture the network change mechanisms in time, we can gain a better understanding of the underlying network dynamics that drive our relationships.
+The available models for dynamic social networks, however, are more complicated, as these models must account for both the network structure and the way it changes in time.
+But if we can capture the change mechanisms over time, we can gain a better understanding of the underlying network dynamics that drive social connections.
 % are more realistic because they provide mechanisms for analyzing multiple observations of the same network over time.
@@
 These models represent a hierarchical approach to network modeling: the changes happen at the smallest possible level, and subsequent choices are conditioned on the past.
-The transition probabilities between two adjacent states of a network depend on the structure of the network, the characteristics of the nodes, or both \citep{saompaper}.
-This means that node-level covariate effects, in addition to common ERGM effects such as reciprocity or transitivity, can be incorporated into the model.
+The transition probabilities between two adjacent states of a network depend on the current structure of the network, the characteristics of the nodes, or both \citep{saompaper}.
+This means that node-level covariate effects, in addition to well-known ERGM effects such as reciprocity or transitivity, can be incorporated into the model.
 %These models are different from other dynamic network models because they allow us to add node level information into the model with the traditional network structure information.
@@
 %\noindent \textbf{Network Data Structure:} 
-Network data across fields have similar structure.
-There are always units of observation and connections between those units.
+Network data across fields have similar structure: units of observation and connections between those units.
 In a social network, the units of observations, called \textit{nodes} or \textit{actors}, are people, while the connections, called \textit{edges} or \textit{ties}, are their relationships with each other.
 %Networks change over time, like when new relationships are formed in a social network.
-Other network data include inherent variables of interest on the nodes and edges themselves, e.g.\ someone's age, gender, and political affiliation, and the type of relationship (friendship, coworkers, etc) that an edge represents.
+Other network data include inherent variables of interest on the nodes, such as age or gender, and the type of relationship (friendship, coworkers, etc) that an edge represents.
 %These two network data structures pose unique problems to analysts, such as, ``How does the strength of a tie between two nodes affect the overall structure of the network?" or ``Do differences in node variables affect the formation of edges?" We propose to answer these questions and more through visual exploration of network data and models.
@@
 
-The lack of a native coordinate system has given rise two very different primary net-vis methods: the node-link diagram and the adjacency matrix visualization \citep{knuth2000, adjmatpro}.
+The lack of a native coordinate system has given rise two primary net-vis methods: the node-link diagram and the adjacency matrix visualization \citep{knuth2000, adjmatpro}.
 We demonstrate these with a simpe example.
@@
 The first net-vis method, the node-link diagram, represents nodes with points in two dimensions and then represents edges by connecting the points with lines.
-These lines can also have arrows on them indicating the direction of the edge for directed networks, as show in Figure~\ref{fig:toyplots}.
+These lines can also have arrows on them indicating the direction of the edge for directed networks, as shown in Figure~\ref{fig:toyplots}.
 Because there is often no natural placement of the nodes, they are placed in 2D at random, then adjusted with one of the many layout algorithms available \citep{drawingalgs}.
@@
 
-The CTMC network model family is for analyzing \textit{dynamic} networks, so we can use them to study the change of a network in time while also examining the structural and node covariate efects.
+The continuous-time Markov chain model family is for analyzing \textit{dynamic} networks, so we can use them to study the change of a network in time while also examining the structural and node covariate efects.
 To define CTMCs for our purposes, we follow the general hierarchical definition outlined in \citet{saomreview} and elsewhere (see e.g.\ \citet{snijders01} and \citet{snijdersetal2010}).
-The hierarchy reflects our intution about social networks as they exist naturally.
-Most social networks, even holding constant the set of actors over time, are ever-changing as relationships form or disappear.
-Actors in social networks also have inherent properties that could affect how they change their role within the network, and existing ties may affect formation or dissolution of other ties.
+This hierarchy reflects our intution about social networks: most social networks, even holding constant the set of actors over time, are ever-changing as relationships form or disappear.
+Actors in social networks also have properties that could change their role within the network, and existing ties may affect formation or dissolution of other ties.
 In this section, we define the data structure, the hierarchal model, and intensity matrix of the CTMC.
@@
 Let $\tau(i|x,m) $ be the wait time until actor $i$ gets to the opportunity to change its ties.
-For any time point $T$, where $t_{m} \leq T < t_{m+1}$, the waiting time to actor $i$'s next change follows an exponential distribution:
+For any time point $T$, where $t_{m} \leq T < t_{m+1}$, the waiting time to actor $i$'s next change follows an exponential distribution,
 \begin{equation} \label{eq:c1}
@@
 \end{equation}
-where $x$ is the current network state, $\mathbf{Z}$ are node covariates, and $K$ is the total number of parameters in the objective function.
+where $x$ is the current network state, $\mathbf{Z}$ is a matrix of node covariates, and $K$ is the total number of parameters in the objective function.
 The parameters of the model, $\beta_1, \dots, \beta_K$ correspond to network statistics, $s_{ik}(x, \mathbf{Z})$, for $k = 1,\dots, K$.
-To maximize $f_i$, $i$ will either destroy a tie where $x_{ij} = 1$, create a tie where $x_{ij}=0$, or make no change.% if all changes will result in lower $f_i$.
+To maximize $f_i$, $i$ will either destroy a tie, create a tie, or make no change.% if all changes will result in lower $f_i$.
 
@@
 According to \citet[p.
-371]{snijders01}, there should be at least two parameters included in the objective function: $\beta_1$ for the outdegree of nodes, and $\beta_2$ for the reciprocity of nodes.
+371]{snijders01}, there should be at least two parameters in the objective function: $\beta_1$ for the outdegree of nodes, and $\beta_2$ for the reciprocity of nodes.
 The former measures the propensity of nodes with a lot of outgoing ties to form more outgoing ties (the ``rich get richer" effect), while the latter measures the tendency of outgoing ties to be returned within a network.
@@
 
-\begin{table}
-\centering
-\begin{tabular}{m{4.5cm}m{4.5cm}m{3cm}}
-\multicolumn{3}{l}{\textbf{Structural Effects}} \\
-Name & Statistic & Figure \\
-\hline\hline
-outdegree  & $s_{i1}(x) = \sum_j x_{ij}$ & \includegraphics[width=2cm]{img/outdegree}\\ \hline
-reciprocity  & $s_{i2}(x) = \sum_j x_{ij}x_{ji}$ & \includegraphics[width=2cm]{img/reciprocity}\\ 
-%\hline
-%transitive triplets  & $s_{i3}(x) = \sum_{j,h} x_{ij}x_{jh}x_{ih}$ & \includegraphics[width=1.5cm]{img/transtrip}  \\
-%\\
-% \multicolumn{3}{l}{\textbf{Covariate Effects}}\\  
+% \begin{table}
+% \centering
+% \begin{tabular}{m{4.5cm}m{4.5cm}m{3cm}}
+% \multicolumn{3}{l}{\textbf{Structural Effects}} \\
 % Name & Statistic & Figure \\
-% \hline \hline
-% covariate-alter  & $s_{i4}(x) = \sum_j x_{ij}z_j$ & \includegraphics[width=3cm]{img/covaralter} \\  \hline
-% covariate-ego  & $s_{i5}(x) = z_i\sum_j x_{ij}$ & \includegraphics[width=3cm]{img/covarego} \\ % \hline
-%same covariate & $s_{i6}(x) = \sum_j x_{ij} \mathbb{I}(z_i = z_j)$ & \includegraphics[width=3cm]{img/covarsim}%\\
-%jumping transitive triplets  & $s_{i7}(x) = \sum_{j \neq h} x_{ij}x_{ih}x_{hj} \mathbb{I}(z_i = z_h \neq z_j)$
-\end{tabular}
-\caption{\label{tab:effects} Examples of structural effects included in the objective function.
-The darker nodes in the Figure column are the ego nodes ($i$), and all others are alters ($j$).
+% \hline\hline
+% outdegree  & $s_{i1}(x) = \sum_j x_{ij}$ & \includegraphics[width=2cm]{img/outdegree}\\ \hline
+% reciprocity  & $s_{i2}(x) = \sum_j x_{ij}x_{ji}$ & \includegraphics[width=2cm]{img/reciprocity}\\ 
+% %\hline
+% %transitive triplets  & $s_{i3}(x) = \sum_{j,h} x_{ij}x_{jh}x_{ih}$ & \includegraphics[width=1.5cm]{img/transtrip}  \\
+% %\\
+% % \multicolumn{3}{l}{\textbf{Covariate Effects}}\\  
+% % Name & Statistic & Figure \\
+% % \hline \hline
+% % covariate-alter  & $s_{i4}(x) = \sum_j x_{ij}z_j$ & \includegraphics[width=3cm]{img/covaralter} \\  \hline
+% % covariate-ego  & $s_{i5}(x) = z_i\sum_j x_{ij}$ & \includegraphics[width=3cm]{img/covarego} \\ % \hline
+% %same covariate & $s_{i6}(x) = \sum_j x_{ij} \mathbb{I}(z_i = z_j)$ & \includegraphics[width=3cm]{img/covarsim}%\\
+% %jumping transitive triplets  & $s_{i7}(x) = \sum_{j \neq h} x_{ij}x_{ih}x_{hj} \mathbb{I}(z_i = z_h \neq z_j)$
+% \end{tabular}
+% \caption{\label{tab:effects} Examples of structural effects included in the objective function.
+The darker nodes in the figures are the ego nodes ($i$), and all others are alters ($j$).
 The dotted lines are ties the ego node is en(dis)couraged to make when the parameter corresponding to each effect is positive (negative).}
-\end{table}
-
-In the CTMC, when actor $i$ can make a change, it chooses which tie $x_{i1}, \dots, x_{in}$ to change at random according to the probabilities $p_{ij}(\boldsymbol{\beta}, x, \mathbf{Z})$.
-These probabilities are defined via the objective function.
+% \end{table}
+
+In the CTMC, when actor $i$ can make a change, it chooses which tie $x_{i1}, \dots, x_{in}$ to change at random according to the probability $p_{ij}(\boldsymbol{\beta}, x, \mathbf{Z})$, defined via the objective function.
 Let $x(i \leadsto j)$ be the network identical to network $x$ with the exception of tie $x_{ij}$, which is equal to $1-x_{ij}$; i.e.\ the presence or absence of node $x_{ij}$ is flipped in the new network.
@@
 \par \noindent \textbf{Intensity Matrix:} The probabilities $p_{ij}$ combined with the rate parameters $\alpha_m$ fully characterize the underlying CTMC that models network change.
-In the CTMC literature, see for instance \citet{ctmc}, chains are characterized by their  \textit{intensity} matrix $\mathbf{Q}$.
+In the CTMC literature, see for instance \citet{ctmc}, chains are characterized by their  \textit{intensity matrix} $\mathbf{Q}$.
 This matrix describes the rate of change between two states of the CTMC process, and the rows of this matrix always add to zero.
@@
 Thus, the intensity matrix $\mathbf{Q}$ is very sparse, with only $n(n-1) + 1$ non-zero entries in each row.
-Note that $n(n-1)$ of these entries represent the possible states that are one edge different from a given state, while the additional non-zero entry is diagonal entry for the current state.% All other entries in a row are structural zeroes because those network states cannot be reached from the current state in a single change.
-
-The two pieces of the model hierarchy each contribute to the entries of the intensity matrix to describe the rate of change between two network states.
+Note that $n(n-1)$ of these entries represent the possible states that are one edge different from a given state, while the additional non-zero entry is the diagonal entry for the current state.% All other entries in a row are structural zeroes because those network states cannot be reached from the current state in a single change.
+
+The two pieces of the model hierarchy each contribute to the intensity matrix to describe the rate of change between two network states.
 The entries of $\mathbf{Q}$ are defined as follows: let $b \neq c \in \{1, 2, \dots, 2^{n(n-1)} \}$ be indices of two different possible states of the network, $x^b, x^c \in \mathcal{X}$.
@@
 Thus, the rate of change between any two states, $x^b$ and $x^c$, that differ by only one tie $x_{ij}$, is the product of the rate at which actor $i$ gets to change a tie and the probability that the tie that will change is the tie to node $j$.
-The intensity matrix $\mathbf{Q}$ fully defines the model for network change, and is the foundation required for parameter estimation.
+The intensity matrix $\mathbf{Q}$ fully defines the CTMC model for network change.
+%, and is the foundation required for parameter estimation.
 %Estimating the parameters in these models is difficult, but thanks to the SIENA software \citep{RSiena}, we have an accessible way to fit SAOMs to network observations.
@@
 where $\theta = (\boldsymbol{\alpha}, \boldsymbol{\beta})$ is the vector of rate and objective function parameters, and $s_{obs}$ is the observed vector of the corresponding statistics summed over all network observations.
+Estimation conditions on the first observed network, $x(t_1)$, then proceeds to estimate model parameters for all future observations.
+% awkward.
+fix it.
 The full SIENA MoM estimation algorithm operates in three phases, as described in \citet{RSienamanual} and \citet{SienaAlgs}, and we briefly summarize these phases here.
@@
 The third phase uses the parameters from phase two to estimate the score functions and covariance matrix of the parameter estimates and carries out convergence checks.
-In Section~\ref{sec:algorithms}, we further explore these phases in the SIENA MoM algorithm through visualization, bringing them out of their ``black box" and into the light.
+In Section~\ref{sec:algorithms}, we further explore these phases in the SIENA MoM algorithm through visualization.
+%, bringing them out of their ``black box" and into the light.
 
@@
 Then, in Section~\ref{sec:minds}, we explore the first step in model-vis by placing the models in the data space.
-Section~\ref{sec:collectmodels} follows, in which we explore collections of models.
+Section~\ref{sec:collectmodels} follows, in which we visualize collections of models.
 Finally, in Section~\ref{sec:algorithms}, we visualize the underlying algorithms in the CTMC fitting process.
@@
                  by.x = "X1", by.y = "id", all = T)
-actual1$Wave <- 1
+actual1$Time <- 1
 actual2 <- merge(data.frame(as.edgelist(as.network(fd2.w2))),
@@
                  by.x = "X1", by.y = "id", all = T)
-actual2$Wave <- 2
+actual2$Time <- 2
 actual3 <- merge(data.frame(as.edgelist(as.network(fd2.w3))),
@@
                  by.x = "X1", by.y = "id", all = T)
-actual3$Wave <- 3
+actual3$Time <- 3
 alldat <- rbind(actual1, actual2, actual3)
@@
   ThemeNet +
-  facet_wrap(~Wave, labeller = "label_both") +
+  facet_wrap(~Time, labeller = "label_both") +
   theme(legend.position = 'bottom',
@@
 The first is a subset of the 50 actor dataset from the ``Teenage Friends and Lifestyle Study" that is provided on the \texttt{RSiena} webpage.
-These data come from \citet{friendsdata}, and we chose to only work with a subset of the data to make node-link diagrams easier to see and to make any changes in the network more noticeable.
-We visualize this data with a node-link diagram in Figure~\ref{fig:alldataviz}. %For model fitting, we condition on wave 1 and estimate the parameters of the models for the second and third waves. 
-The actor covariate of interest, drinking behavior, has four values: (1) does not drink, (2) drinks once or twice a year, (3) drinks once a month, and (4) drinks once a week.
-In Figure~\ref{fig:alldataviz}, the nodes are colored according to the drinking behavior of that student.
+These data first appeared in \citet{friendsdata}, and we chose to only work with a subset of the data to make node-link diagrams easier to see and to make any changes in the network more noticeable.
+We visualize this data with a node-link diagram in Figure~\ref{fig:alldataviz}, where %For model fitting, we condition on wave 1 and estimate the parameters of the models for the second and third waves.
+nodes are colored according to the girl's reported drinking behavior: does not drink, drinks once or twice a year, drinks once a month, and drinks once a week.
 We can see that, as time passes, the students seem to drink more and become increasingly isolated into smaller groups.
@@
 
-The second data example we use is a much more recent dataset on collaboration in the United States Senate during the $111^{th}$ through $114^{th}$ Congresses, shown in Figure~\ref{fig:senateplot}.
+%The second data example we use is a much more recent dataset on collaboration in the United States Senate during the $111^{th}$ through $114^{th}$ Congresses, shown in Figure~\ref{fig:senateplot}.
 These sessions of congress correspond to the years of Barack Obama's presidency, from 2009-2016.
@@
 This value is defined in \citet{senate} as the total number of times senator $i$ cosponsors senator $j$'s bill, weighted by the number of total cosponsors:
-\begin{equation}\label{eq:sen1}
-WPC_{ij} = \dfrac{\sum\limits_{k=1}^{n_j} \frac{Y_{ij(k)}}{c_{j(k)}}}{\sum\limits_{k=1}^{n_j} \frac{1}{c_{j(k)}}}
-\end{equation}
-where $n_j$ is the number of bills in a congressional session authored by senator $j$, $c_{j(k)}$ is the number of cosponsors on senator $j$'s $k^{th}$ bill, where $k \in \{1,\dots, n_j\}$, and $Y_{ij(k)}$ is a binary variable that is 1 if senator $i$ cosponsored senator $j$'s $k^{th}$ bill, and is 0 otherwise.
+% \begin{equation}\label{eq:sen1}
+% WPC_{ij} = \dfrac{\sum\limits_{k=1}^{n_j} \frac{Y_{ij(k)}}{c_{j(k)}}}{\sum\limits_{k=1}^{n_j} \frac{1}{c_{j(k)}}}
+% \end{equation}
+%where $n_j$ is the number of bills in a congressional session authored by senator $j$, $c_{j(k)}$ is the number of cosponsors on senator $j$'s $k^{th}$ bill, where $k \in \{1,\dots, n_j\}$, and $Y_{ij(k)}$ is a binary variable that is 1 if senator $i$ cosponsored senator $j$'s $k^{th}$ bill, and is 0 otherwise.
 This measure ranges in value from 0 to 1, where $WPC_{ij} = 1$ if senator $i$ is a cosponsor on every one of senator $j$'s bills and $WPC_{ij} = 0$ if senator $i$ is never a cosponsor any of senator $j$'s bills.
@@
 Edges are shown only if the WPC between two senators is greater than 0.25.
-We use the FR layout algorithm here.', fig.width=8, fig.height=8, out.width='\\textwidth'>>=
+We use the FR layout algorithm here.', fig.width=8, fig.height=8, out.width='\\textwidth', eval = FALSE>>=
 seobama <- read_csv("data/congress/senateobamapres_gsw_25.csv")
@@
 @
-Because CTMC models require binary edges, we focus only on strong collaborations, where $WPC$ is high. For a senate collaboration network $x$, edges are defined as
-\begin{equation}\label{eq:wpc2}
-  x_{ij} =
-  \begin{cases}
-                                   1 &  \text{if } WPC_{ij} > 0.25 \\
-                                   0 & \text{if } WPC_{ij} \leq 0.25.
-  \end{cases}
-\end{equation}
-The node-link diagram for the four senates during the Obama administration are shown in Figure~\ref{fig:senateplot}.
+% Because CTMC models require binary edges, we focus only on strong collaborations, where $WPC$ is high. For a senate collaboration network $x$, edges are defined as
+% \begin{equation}\label{eq:wpc2}
+%   x_{ij} =
+%   \begin{cases}
+%                                    1 &  \text{if } WPC_{ij} > 0.25 \\
+%                                    0 & \text{if } WPC_{ij} \leq 0.25.
+%   \end{cases}
+% \end{equation}
+% The node-link diagram for the four senates during the Obama administration are shown in Figure~\ref{fig:senateplot}.
 We can see that Senate 111 is much more densely connected than the other three, which have one large, sparsely connected component with a few smaller connected groups throughout.
@@
 \begin{subfigure}[t]{.45\textwidth}
+<<si1, fig.width=2, fig.height=2, message = FALSE, warning = FALSE, fig.align='center',out.width='.5\\textwidth'>>=
+s1e <- data.frame(from = rep('i',4), to = letters[10:13])
+s1n <- data.frame(id = letters[9:13], group = c(1,1,1,1,2))
+s1net <- merge(s1e,s1n, by.x = "from", by.y = "id", all = T)
+set.seed(1456)
+ggplot(data = s1net, aes(from_id = from, to_id = to)) +
+  geom_net(directed = T, labelon = T, arrowsize = .5,
+           labelcolour='white',vjust = 0.5, hjust =0.5, arrowgap = .15,
+           size=6, fontsize = 5, ecolour="black",
+           linetype = c("solid", "solid","dotted", "solid","solid", "solid","solid", "solid")) +
+  expand_limits(x=c(-0.2,1.2), y=c(-0.1,1.1)) +
+  ThemeNet +
+  theme(legend.position = "none") + 
+  labs(title = expression(paste(s[i1], "(x)")))
+@
+\caption{\label{fig:beta1}Demonstrating the outdegree effect $\beta_1$. The dotted tie is encouraged if $\beta_1$ is positive. $s_{i1}(x) = \sum_{j} x_{ij}$}
+\end{subfigure}
+\begin{subfigure}[t]{.45\textwidth}
+<<s2, fig.width=2, fig.height=2, message = FALSE, warning = FALSE, fig.align='center',out.width='.5\\textwidth'>>=
+s2e <- data.frame(from = letters[9:10], to = letters[10:9])
+s2n <- data.frame(id = letters[9:10], group = c(1,2))
+s2net <- merge(s2e,s2n, by.x = "from", by.y = "id", all = T)
+set.seed(13902)
+ggplot(data = s2net, aes(from_id = from, to_id = to)) +
+  geom_net(directed = T, labelon = T, curvature=.25, arrow = arrow(type = "closed",length = unit(4,"mm")),
+           labelcolour='white',vjust = 0.5, hjust =0.5, arrowgap = .15,
+           size=10, fontsize = 5, ecolour="black",
+           linetype = c("solid","dotted")) +
+  expand_limits(x=c(-0.2,1.2), y=c(-0.1,1.1)) +
+  ThemeNet +
+  theme(legend.position = "none") + 
+  labs(title = expression(paste(s[i2], "(x)")))
+@
+\caption{\label{fig:beta2}Demonstrating the reciprocity effect $\beta_2$. The dotted tie is encouraged if $\beta_2$ is positive. $s_{i2}(x) = \sum_{j} x_{ij}x_{ji}$}
+\end{subfigure}
+\begin{subfigure}[t]{.45\textwidth}
 <<jtt, echo=FALSE, fig.width=2, fig.height=2, message = FALSE, warning = FALSE, fig.align='center',out.width='.49\\textwidth'>>=
@@
 The JTT effect emphasizes triads formed between actors with \emph{different} covariate values, while the N22 effect emphasizes \emph{indirect} ties between actors.
-The covariate for the friendship data is the drinking behavior, and is the number of bills authored for the senate data.
+%The covariate for the friendship data is the drinking behavior, and is the number of bills authored for the senate data.
 These additional effects are visualized in Figure~\ref{fig:jtt} and Figure~\ref{fig:dad}, where the ties affected by the parameter value are represented by dotted lines.
@@
 
+% need a way better transition here.
+
 \section{Model in the data space}\label{sec:minds}
@@
 \emph{Model visualization} complements traditional model diagnostic tools such as $R^2$ values and residual plots to enrich our understanding and interpretation of statistical models.
-Each of the three different meanings of the word ``model", the model \emph{family}, the model \emph{form}, and the \emph{fitted} model \citep[see][pg.
-2]{modelvis}, can benefit from model visualization (model-vis).
+Each of the three different meanings of the word ``model", the model \emph{family}, the model \emph{form}, and the \emph{fitted} model, can benefit from model visualization (model-vis) \citep{modelvis}.
 We apply the three principles of model-vis: view the model in the data space, visualize collections of models, and explore the process of fitting the model, to the models and data introduced in Section~\ref{sec:exdatamods}.
@@
 
-First, we put models in the data space to assess the model fit.
+Placing the model in the data space is crucial to understanding how well the model represents the observed data.
+%First, we put models in the data space to assess the model fit.
 %In \citeauthor{modelvis}, they chose define the \textit{data space} as ``the region over which we can reliably make inferences, usually a hypercube containing the data" \citep[p.~206]{modelvis}.
@@
 We use the \texttt{R} package \texttt{geomnet} \citep{geomnet} to visualize the three data spaces together.
-The package lays out the network using node-link diagrams, and the network's node or edge data are then mapped to different visual features.
+The package draws the network using node-link diagrams, and the network's node or edge data are then mapped to different visual features.
 The color, size, and shape of the points can be used to represent variables in the node data, while the color, linewidth, and linetype of the edges between nodes can be used to represent variables in the edge data.
-To view temporal changes, we place the network at different timepoints side-by-side to see the evolution.
+Finally, the networks observed at each time point are placed side-by-side, showing the network's evolution in time.
+Bringing all network information together in \texttt{geomnet} allows us to show the entire data space at once.
+%To view temporal changes, we place the network at different timepoints side-by-side to see the evolution.
 By pulling all of this information together using \texttt{geomnet}, we can show the entire data space at once.
@@
 We map sex, party, and bills authored to the shape, color, and size of the nodes, respectively.
-We also map $WPC$ to the edge width.'>>=
+We also map $WPC$ to the edge width.', eval = FALSE>>=
 se111clint <- filter(seobama, senate == 111)
@@
 
-To demonstrate the functionality of \texttt{geomnet}, we visualize the connections in the $111^{th}$ U.S.
+%To demonstrate the functionality of \texttt{geomnet}, we visualize the connections in the $111^{th}$ U.S.
 Senate at two different timepoints: when Hillary Clinton was in the senate, and after she left to become Secretary of State.
@@
 
-<<summnetM1, message=FALSE, fig.cap='On the left, the first wave of observed data, conditioned on in the model.
-On the right, the second wave of observed data.
-In the middle, a summary network from the first model fit to the data.
-This summary network represents 1,000 simulations of wave 2 using the values from the simple fitted model M1.',results='hide', fig.height=2.5>>=
+<<summnetM1, message=FALSE, fig.cap='On the left, observed data from the first time point, which is conditioned on in the first model.
+On the right, the data observed at the second time point.
+In the middle, the expected network from the first model fit to the data.
+This expected network represents 1,000 simulations of $x(t_2)$ from model M1.',results='hide', fig.height=2.5>>=
 friendshipData <- array(c(fd2.w1, fd2.w2, fd2.w3), dim = c(16, 16, 3))
@@
            singletons = T) +
-   labs(title = "Wave 1") + ThemeNet +
+   labs(title = "Time 1") + ThemeNet +
     theme(plot.background = element_rect(color = 'black'))
@@
            singletons = T) +
-  labs(title = "Wave 2")  +  ThemeNet +
+  labs(title = "Time 2")  +  ThemeNet +
   theme(plot.background = element_rect(color = 'black'))
@@
 
-To view our models in the data space, we first simulate from the fitted model.
+Then to view our models in the data space, we first simulate from the fitted model.
 No single network instance simulated from a CTMC will fully represent the model, so instead we look at an \textit{expectation} from the network model.
 We frequently rely on expected values in statistics, but network models, especially those as complex as our CTMC models, lack an expected network measure.
-We could talk about expected values of parameters, but the parameters can be hard to interpret and cannot teach us much about the overall structure of the network.
+We could talk about expected values of parameters, but the parameters can be hard to interpret and do not illuminate the overall structure of the network.
 How then, can we arrive at an ``expected" network?
@@
 \begin{enumerate}
-\item simulate 1,000 wave 2 and wave 3 instances of our small friendship network from model M1 using estimated parameter values (see Table~\ref{tab:meanests}) that have previously been estimated.
+\item condition on $x(t_1)$ and simulate 1,000 instances of $x(t_2)$ and $x(t_3)$ from the model of interest using parameter values (see Table~\ref{tab:meanests}) that have previously been estimated.
 Then, we
@@
 \item construct one weighted summary network for time points 2 and 3. Each summary network has edge weights equal to the proportion of the simulated networks in which the edge appeared. Finally, we 
-\item create the \textit{expected network} by including only the edges that appeared in more than 5\% of the simulations. 
+\item create the \textit{expected network} by including only the edges that appeared in more than 5\% of the simulations. Call these expected networks $\bar{x}(t_2)$ and $\bar{x}(t_3)$.
 \end{enumerate}
-The resulting expected network is in the middle panel of Figure~\ref{fig:summnetM1}.
-Edges with higher weights are darker, thicker lines.
-On either side of the average network in Figure~\ref{fig:summnetM1}, we show the data for comparison: wave 1 on the left, and wave 2 on the right.
-We can see that the overall structure of the average network is much more similar to wave 1 than to wave 2.
-In both wave 1 and the expected network, the edges form two distinct clusters of nodes: $\{2,3,4,5,6,12,13,15\}$ and $\{7,8,9,10,11,14\}$.
-The expected network, however, is supposed to represent the second wave of data, on the right in Figure~\ref{fig:summnetM1}.
-The expected network does not resemble wave 2, which has one small, strongly connected cluster and one larger, sparsely connected cluster, with four unconnected nodes.
-Because the expected network does not resemble wave 2, we take this as evidence that the simple model, M1, does not capture the change in network structure from the first to the second wave of data.
+
+\noindent The expected network, $\bar{x}(t_2)$, is in the middle panel of Figure~\ref{fig:summnetM1}.
+Edges that appear with higher frequency are drawn with darker, thicker lines.
+On either side of the average network in Figure~\ref{fig:summnetM1}, we show the data for comparison: $x(t_1)$ on the left, and $x(t_2)$ on the right.
+We can see that the overall structure of the average network is much more similar to $x(t_1)$  than to $x(t_2)$ .
+In both wave 1 and the expected network, the edges form two clusters of nodes: $\{2,3,4,5,6,12,13,15\}$ and $\{7,8,9,10,11,14\}$.
+The expected network, however, is supposed to represent $x(t_2)$ , on the right in Figure~\ref{fig:summnetM1}.
+The expected network does not resemble  $x(t_2)$, which has one small, strongly connected cluster and one larger, sparsely connected cluster, with four unconnected nodes.
+Because the expected network does not resemble  $x(t_2)$ , we take this as evidence that the simple model, M1, does not capture the change in network structure from the first to the second observation of data.
 Visualizing the expected network in this way places the model M1 in the data space, and helps us assess model fit.
@@
 
-<<summary2, fig.height=2.5, fig.cap="Network differences between the expected network from M1 and Wave 1 and Wave 2 of the data.
-Some edges appear in both the data and the expected Wave 2, while others only appear in the data or only in the exectation.
+<<summary2, fig.height=2.5, fig.cap="Network differences between the expected network  $\\bar{x}(t_2)$ and $x(t_1)$  and  $x(t_2)$.
+Some edges appear in both the data and the expected network $\\bar{x}(t_2)$, while others only appear in the data or only in the expectation.
 Width of edges represent the proportion of appearances in the 1,000 simulations.
@@
   scale_linetype_manual(values =c("longdash", "dotted") ) + 
-  labs(title = "Difference between Wave 1 &\nWave 2 expectation from M1") + ThemeNet +
+  labs(title = "Difference between Time 1 Data &\n Expectation from M1") + ThemeNet +
   theme(plot.background = element_rect(color = 'black'), legend.position = "none")
@@
   scale_linetype_manual(values = c("longdash", "solid", "dotted")) + 
-  labs(title = "Difference between Wave 2 &\nWave 2 expectation from M1") + ThemeNet +
+  labs(title = "Difference between Time 2 Data &\n Expectation from M1") + ThemeNet +
   theme(plot.background = element_rect(color = 'black'), legend.position = "none")
@@
 Another way to view our models in the data space is with \textit{difference networks} between the expected network and the data, as shown in Figure~\ref{fig:summary2}.
-We created these networks by computing the difference between the adjacency matrices for the expected network and wave 1, and for wave 2 and the expected network.
-The lack of model fit becomes even more clear in Figure~\ref{fig:summary2}, with the broken edges representing ties that only show up in the expected network, and the solid edges representing ties in the data that the model fails to capture.
+We created these networks by computing the difference between the adjacency matrices $\mathbf{A}(\bar{x}(t_2)) - \mathbf{A}(x(t_1))$ of the expected network $\bar{x}(t_2)$ and $x(t_1)$ respectively, and $\mathbf{A}(x(t_2)) - \mathbf{A}(\bar{x}(t_2))$, the difference between $\bar{x}(t_2)$ and the expected network $\bar{x}(t_2)$.
+The lack of model fit becomes more apparent in Figure~\ref{fig:summary2}, with the broken edges representing ties that only show up in the expected network, and the solid edges representing ties in the data that the model fails to capture.
 Dotted edges are in both the expectation and the data and represent where the model ``got it right".
-Again, the widths of the edges demonstrate the frequency of the ties in the simulations.
-On the left in Figure~\ref{fig:summary2}, we see the difference between the first wave of data and the expected network.
-We only see dotted and broken edges, meaning that the simulated networks consist primarily of edges in wave 1, plus some new edges that are not in the first wave of data.
-The unobserved ties are all reciprocating edges in the data, so the model predicts more reciprocity between actors than there actually is.
-On the right in Figure~\ref{fig:summary2}, we see the difference between the second wave of data and the expected network.
-Because the model is simulating the second wave of data, we would hope to see very few differences between the expected network and wave 2.
-However, there are six solid edges, which are edges in the data that do not appear at all in the expected network, and there are many broken edges that appear in the expected network but not in the data.
-Although there are some edges that appear in both the expected network and wave 2, the majority of edges in the expected network are not in wave 2, and there are several edges in the data missing from the expectation.
+Again, the widths of the edges correspond to the frequency with which the ties appear in the simulations.
+On the left in Figure~\ref{fig:summary2}, we see the difference between $x(t_1)$ and the expected network.
+We only see dotted and broken edges, meaning that the simulated networks consist primarily of edges in $x(t_1)$, plus some new edges that are not in the data.
+The unobserved ties are all reciprocating edges in the data, thus the model predicts more reciprocity between actors than there actually is.
+On the right in Figure~\ref{fig:summary2}, we see the difference between the $x(t_2)$ and the expected network.
+Because the model is simulating  $x(t_2)$, we would hope to see very few differences between the expected network and the data.
+However, there are six solid edges which are in the data observed at the second time point that do not appear at all in the expected network, and there are many broken edges that appear in the expected network but are not in $x(t_2)$.
+Although there are some edges that appear in both the expected network and $x(t_2)$, the majority of edges in the expected network are not in the observed data, and there are several edges in the data missing from the expectation.
 We conclude that the model M1 is a poor fit to our example data, which we have seen very clearly by viewing the model in the data space.
@@
 Now that we have assessed model fit, we want to learn about model behavior in order to choose better fitting models in the future.
-This is the second principle of model visualization: ``collections [of models] are more informative" \citep{modelvis}.
-We examine the behavior of the CTMCs by visualizing collections of them, which we construct in four ways.
-First, we explore the space of all possible models to determine the most significant model parameters.
+To assess model behavior, we employ the second principle of model visualization: ``collections [of models] are more informative" \citep{modelvis}.
+We rely on three methods for constructing collections of CTMC models.
+%We examine the behavior of the CTMCs by visualizing collections of them, which we construct in four ways.
+First, we explore the space of all possible models to determine significant model parameters.
 Second, we vary model settings to uncover relationships between parameters and their effects on the fitted model.
-Next, we fit the same model to the same data many times to assess the behavior of the underlying simulations.
-Finally, we fit the same model to different data to learn about fitted parameter values under different conditions.
-We chose these four collections because they each explore something different about model behavior: the choice of parameters in the objective function, the effects of underlying simulations, and the resulting fitted values under different circumstances.
+Finally,  we fit the model many times to assess the behavior of the underlying simulations.
+%Finally, we fit the same model to different data to learn about fitted parameter values under different conditions.
+We chose these three collections because they each explore something different about model behavior: the choice of and relationships between parameters in the objective functio and the stochastic effects from simulation.%, and the resulting fitted values under different circumstances.
 %Using the tools in our visualization toolbox, we view these collections to learn about model behavior.%The first considers the many parameters to include in a SAOM, while the second results in different parameter estimates, so we can see how parameter estimates vary.
@@
 \noindent \textbf{Exploring the space of all possible models:} The \texttt{RSiena} manual contains over 80 effects to include in the objective function.
-To select objective function parameters for modeling our example data, we searched through all effects available to find \textit{significant} effects.
+To select objective function parameters for modeling our example data, we searched through all effects available to find  significant effects.
 The \texttt{RSiena} package includes the function \texttt{Wald.Rsiena} for Wald-type tests of parameter significance.
 We start by including the outdegree and reciprocity parameters in the model then add one effect to the objective function, fit the model to the example data, test the added effect for significance, and repeat for all possible model parameters.
-We performed this procedure, which we used to pick M2 and M3, for both the small friendship and the senate collaboration data.
+%We performed this procedure, which we used to pick M2 and M3, for both the small friendship and the senate collaboration data.
 A visualization of the significant effects for the two example data sets are available in the Supporing Information.
 Using this one-at-a-time testing method results in many possible significant parameters of different type and magnitude, demonstrating the flexibility of these models.
+We chose models M1, M2, and M3 to examine more closely because they had the smallest $p$-values from the Wald-type tests.
 %We see in both the friendship data and the senate data results that most of the significant effects have absolute value less than ten.
@@
 
-\noindent \textbf{Varying model settings:} The second method for viewing collections of models is varying model settings.
+\noindent \textbf{Varying model settings:} The second method for obtaining collections of models is to vary model settings.
 We have done this already by choosing three different models M1, M2, M3 to fit to our example data sets.
 To explore the behavior of these three models, we simulate from them.
-Using the estimated parameter values in Table~\ref{tab:meanests}, we simulated 1,000 observations of waves 2 and 3 of the friendship data from each of the three models.
-From the simulations, we created an expected network for wave 2 from models M1, M2, and M3 using the same method described in Section~\ref{sec:minds}.
-These expected networks are presented in the Supporting Information, and we see that all models are failing to capture the structure of the second wave, even though the individual parameters were all significant.
+Using the estimated parameter values in Table~\ref{tab:meanests}, we simulated 1,000 observations of $x(t_2)$ and $x(t_3)$ of the friendship data from each of the three models.
+From the simulations, we created an expected network for $x(t_2)$ from models M1, M2, and M3 using the same method described in Section~\ref{sec:minds}.
+These expected networks are presented in the Supporting Information, and we see that all models are failing to capture the structure of $x(t_2)$, even though the individual parameters were all significant according to the statistical tests.
 %In Section~\ref{sec:samemodsamedat}, we fit these three models to the data 1,000 times, and examine the parameter densities.
@@
 
-\noindent \textbf{Fitting the same model to the same data:} The simulations required to fit the CTMC network models lead to inherent variability in the fitted values.
+\noindent \textbf{Fitting the model repeatedly:} The simulations required to fit the CTMC network models lead to inherent variability in the fitted values.
 To examine this variability, we fit models M1, M2, and M3 to the friendship network 1,000 times each.
@@
 The inclusion of $\beta_3$ also leads to increases in the estimates of $\alpha_1, \alpha_2$.
-Thus, when modeling the transitive triplet behavior in this way, we may conclude that girls change friends more frequently.
-It is not clear, however, that the addition of a parameter to the objective function \textit{should} effect the estimates of the rate parameters.
+Thus, when modeling the transitive triplet behavior in this way, the model concludes that girls change friends more frequently.
+It is not clear, however, that the addition of a parameter to the objective function should effect the estimates of the rate parameters.
 This strong and possibly problematic relationship between rate parameter estimates and objective function parameters is further explored in the Supporting Information.
@@
 
+% % table of fitted M1 M2 M3
+% \begin{table}
+% \scalebox{0.9}{
+% \begin{tabular}{lccc|ccc }
+% 
+%   & \multicolumn{3}{c}{Friendship Data} & \multicolumn{3}{c}{Senate Data} \\
+%   \hline
+%   & M1 & M2 & M3 & M1 & M2 & M3 \\
+%   \hline
+%   $\alpha_1$ & \Sexpr{SFE$M1[1]} & \Sexpr{SFE$M2[1]} & \Sexpr{SFE$M3[1]} & \Sexpr{SSE$M1[1]} & \Sexpr{SSE$M2[1]} & \Sexpr{SSE$M3[1]} \\
+%   $\alpha_2$ & \Sexpr{SFE$M1[2]} & \Sexpr{SFE$M2[2]} & \Sexpr{SFE$M3[2]} & \Sexpr{SSE$M1[2]} & \Sexpr{SSE$M2[2]} & \Sexpr{SSE$M3[2]}  \\
+%   $\alpha_3$ & -- & -- & -- & \Sexpr{SSE$M1[3]} & \Sexpr{SSE$M2[3]} & \Sexpr{SSE$M3[3]} \\
+%   \hline
+%   $\beta_1$ & \Sexpr{SFE$M1[3]} & \Sexpr{SFE$M2[3]} & \Sexpr{SFE$M3[3]} & \Sexpr{SSE$M1[4]} & \Sexpr{SSE$M2[4]} & \Sexpr{SSE$M3[4]} \\
+%   $\beta_2$ & \Sexpr{SFE$M1[4]} & \Sexpr{SFE$M2[4]} & \Sexpr{SFE$M3[4]} & \Sexpr{SSE$M1[5]} & \Sexpr{SSE$M2[5]} & \Sexpr{SSE$M3[5]}\\
+% $\beta_3$ & -- & \Sexpr{SFE$M2[5]} & -- & -- & \Sexpr{SSE$M2[6]} & -- \\
+%   $\beta_4$ & -- & --  & \Sexpr{SFE$M3[6]} & -- & -- & \Sexpr{SSE$M3[7]} \\
+%   \hline
+% \end{tabular}
+% }
+% \caption{\label{tab:meanests} The means (std.\ dev.)\ of parameter values estimated from repeated fittings of $M1, M2, M3$ to the two example data sets.}
+% \end{table}
+
 % table of fitted M1 M2 M3
 \begin{table}
-\scalebox{0.9}{
-\begin{tabular}{lccc|ccc }
-
-  & \multicolumn{3}{c}{Friendship Data} & \multicolumn{3}{c}{Senate Data} \\
-  \hline
-  & M1 & M2 & M3 & M1 & M2 & M3 \\
-  \hline
-  $\alpha_1$ & \Sexpr{SFE$M1[1]} & \Sexpr{SFE$M2[1]} & \Sexpr{SFE$M3[1]} & \Sexpr{SSE$M1[1]} & \Sexpr{SSE$M2[1]} & \Sexpr{SSE$M3[1]} \\
-  $\alpha_2$ & \Sexpr{SFE$M1[2]} & \Sexpr{SFE$M2[2]} & \Sexpr{SFE$M3[2]} & \Sexpr{SSE$M1[2]} & \Sexpr{SSE$M2[2]} & \Sexpr{SSE$M3[2]}  \\
-  $\alpha_3$ & -- & -- & -- & \Sexpr{SSE$M1[3]} & \Sexpr{SSE$M2[3]} & \Sexpr{SSE$M3[3]} \\
-  \hline
-  $\beta_1$ & \Sexpr{SFE$M1[3]} & \Sexpr{SFE$M2[3]} & \Sexpr{SFE$M3[3]} & \Sexpr{SSE$M1[4]} & \Sexpr{SSE$M2[4]} & \Sexpr{SSE$M3[4]} \\
-  $\beta_2$ & \Sexpr{SFE$M1[4]} & \Sexpr{SFE$M2[4]} & \Sexpr{SFE$M3[4]} & \Sexpr{SSE$M1[5]} & \Sexpr{SSE$M2[5]} & \Sexpr{SSE$M3[5]}\\
-$\beta_3$ & -- & \Sexpr{SFE$M2[5]} & -- & -- & \Sexpr{SSE$M2[6]} & -- \\
-  $\beta_4$ & -- & --  & \Sexpr{SFE$M3[6]} & -- & -- & \Sexpr{SSE$M3[7]} \\
-  \hline
+\centering
+\begin{tabular}{lccc}
+  \toprule
+  & M1 & M2 & M3 \\
+  \midrule
+  $\alpha_1$ & \Sexpr{SFE$M1[1]} & \Sexpr{SFE$M2[1]} & \Sexpr{SFE$M3[1]}  \\
+  $\alpha_2$ & \Sexpr{SFE$M1[2]} & \Sexpr{SFE$M2[2]} & \Sexpr{SFE$M3[2]}   \\
+  \midrule
+  $\beta_1$ & \Sexpr{SFE$M1[3]} & \Sexpr{SFE$M2[3]} & \Sexpr{SFE$M3[3]}  \\
+  $\beta_2$ & \Sexpr{SFE$M1[4]} & \Sexpr{SFE$M2[4]} & \Sexpr{SFE$M3[4]} \\
+$\beta_3$ & -- & \Sexpr{SFE$M2[5]} & --  \\
+  $\beta_4$ & -- & --  & \Sexpr{SFE$M3[6]} \\
+  \bottomrule
 \end{tabular}
-}
-\caption{\label{tab:meanests} The means (std.\ dev.)\ of parameter values estimated from repeated fittings of $M1, M2, M3$ to the two example data sets.}
+
+\caption{\label{tab:meanests} The means (std.\ dev.)\ of parameter values estimated from repeated fittings of $M1, M2, M3$ to the example data.}
 \end{table}
 
-\noindent \textbf{Fitting the same model to different data:} Finally, we view collections of models by fitting the same models to different data to understand how the model behaves in different contexts.
+% \noindent \textbf{Fitting the same model to different data:} Finally, we view collections of models by fitting the same models to different data to understand how the model behaves in different contexts.
 As we did for the friendship data, we estimated the parameters for M1, M2, and M3, when fit to the senate data (see Figure \ref{fig:senateplot}) 100 times each.
@@
 Tight-knit friend groups are more encouraged by model M2 than by models M1 and M3.
-This suggests that in both data sets, nodes prefer outgoing ties that are reciprocated to those that are not.
+% This suggests that in both data sets, nodes prefer outgoing ties that are reciprocated to those that are not.
 When we consider the context, this seems obvious: the students want to have close mutual friendships, and senators want mutual support for their bills.
@@
 %We explore the relationship between $\beta_1$ and $\beta_2$ further in Section~\ref{sec:samemodsamedat}.
-% We have seen how the same model forms behave in different context, with some  differences and some similarities. 
-After seeing the high correlation between parameter estimates in very different data contexts, we might question the guideline that both parameters need to be included in the objective function.
-
-In Figure~\ref{fig:distests}, we saw that the inclusion of $\beta_3$ for the friendship data has a noticeable effect on the other parameters in the model.
+% % We have seen how the same model forms behave in different context, with some  differences and some similarities. 
+% After seeing the high correlation between parameter estimates in very different data contexts, we might question the guideline that both parameters need to be included in the objective function. 
+% 
+% In Figure~\ref{fig:distests}, we saw that the inclusion of $\beta_3$ for the friendship data has a noticeable effect on the other parameters in the model.
 In Figure~\ref{fig:compareModels}, however, we see that this effect is not present in the senate data.
@@
 By examining the results of models fitted to different data, we better understand the interpretation of the objective function parameters, and we have learned of some potential parameter correlation problems affecting the fit of the models.
-All in all, we have used four different ways of viewing collections of models, which have given us insight into the behavior of CTMC network models.
+
+%All in all, we have used three different ways of viewing collections of models, which have given us insight into the behavior of CTMC network models.
 
@@
 
-The last model visualization principle, ``explore the process of model fitting," is crucial to understanding our models \citep{modelvis}.
-When fitting models in \texttt{RSiena}, several simulation steps are hidden from the user.
-One of the hidden steps in the SIENA MoM algorithm is the \textit{microstep} process, which is an observed instance of the CTMC described in Section~\ref{sec:saoms}.
+The third and final model visualization principle, ``explore the process of model fitting," is crucial to understanding our models \citep{modelvis}.
+When fitting models in \texttt{RSiena}, several simulation processes are hidden from the user.
+We refer to one of these hidden pieces as in the SIENA MoM algorithm is the \textit{microstep process}, which is a series of state changes in the CTMC as described in Section~\ref{sec:saoms}.
 At each microstep, an \textit{ego} node is selected to change, and the chosen node randomly makes one change in its ties according to to the probabilities $p_{ij}$ defined in Equation~\ref{eq:pij}.
+This process continues until the network differs from the $x(t_m)$ by exactly $C_m$, as defined in Equation~\ref{eq:cdiff}.
 Between two network observations $x(t_m)$ and $x(t_{m+1})$, there can be dozens, hundreds, or even thousands of microsteps, depending on the size of the network and the number of tie changes between two network observations.
@@
 But in order to understand \textit{how} the models are fit, we need to extract and explore the different steps of the process.
-We use these steps to visually explore the algorithms, so that we are visualizing the model fitting process, and not just the model fitting result.
+We use these steps to visually explore the algorithms, so that we are visualizing the model fitting process, and not only the model fitting result.
 
@@
   dplyr::mutate(val = as.numeric(!from == to),
-         from = paste0("V", parse_number(from)+1), # make the chains
-         to = paste0("V", parse_number(to)+1)) -> ansnullchainsw1w2
+         from = paste0("V", parse_number(as.character(from))+1), # make the chains
+         to = paste0("V", parse_number(as.character(to))+1)) -> ansnullchainsw1w2
 colnames(fd2.w1) <- paste0("V", 1:16)
@@
 
-To visualize the underlying algorithm, we animate the microsteps that form the CTMC from wave 1 to wave 2 of the small friendship network (see Figure~\ref{fig:alldataviz}).
-The microstep animation, some frames of which are shown in Figure~\ref{fig:makinggif}, represent one possible path from wave 1 to wave 2 according to the model M1.
+To visualize the underlying algorithm, we animate the microsteps that form the CTMC from $x(t_1)$ to $x(t_2)$ for the example data.
+The microstep animation, some frames of which are shown in Figure~\ref{fig:makinggif}, represent one possible path from $x(t_1)$ to $x(t_2)$ according to the model M1.
 We show each change in the network as follows: we first increase the size and change the color of the ego node to focus attention to its ties.
@@
 If a tie is being added, it appears red from nothing, grows large for emphasis, then changes color and size to match the rest of the edges.
-In this network animation, we show the modeled steps of the \textit{unobserved} CTMC process between two observed networks.
+In this network animation, we show the modeled steps of the unobserved CTMC process between two observed networks.
 We see each part of the hierarchical come into play: first, the selection of the ego node via the rate function, then the result of the actor maximizing its objective function by either deleting or adding a node.
 Finally, the layout of the network transitions as edges are removed or added, demonstrating how the overall network structure changes with each of the individual tie changes.
-Combining the tie changes in this animation makes the CTMC process the focus and demonstrates the underlying change mechanism of the dynamic network according to the chosen model.
-
-<<adjmatorder, message = FALSE, warning = FALSE, fig.cap="On the left, the starting friendship network represented in adjacency matrix form, ordered by node ID.
+Combining the tie changes in this animation makes the CTMC process the focu, demonstrating the underlying change mechanism of the dynamic network according to the model.
+
+<<adjmatorder, message = FALSE, warning = FALSE, fig.cap="On the left, the first observation of the friendship network represented in adjacency matrix form, ordered by node ID.
 On the right, the same adjacency matrix is presented after seriation", fig.width=8, fig.height=3, fig.align='center'>>=
@@
         axis.text.y = element_text(margin = margin(r=0))) +
-  coord_fixed()
+  coord_fixed() + 
+  labs(x= "alter (j)", y = "ego (i)")
 
@@
   ThemeNoNet + theme(legend.position = "none") + 
-  coord_fixed()
+  coord_fixed() + 
+  labs(x= "alter (j)", y = "ego (i)")
 legendp2 <- cowplot::get_legend(p2)
@@
 \end{equation}
-We then perform a principal component analysis (PCA) on $\mathbf{A}^{cum}$, and use the values of the first principal component to order the vertices on the $x$ and $y$ axes for the adjacency matrix animation.
-The resulting visualization for the first wave of data is shown in Figure~\ref{fig:adjmatorder}.
+We then perform a principal component analysis (PCA) on $\mathbf{A}^{cum}$, and use the values of the first principal component to order the vertices on the $x$ and $y$ axes for the adjacency matrix net-vis.
+The resulting visualization for the first network observation is shown in Figure~\ref{fig:adjmatorder}.
 %we present the first wave of data ordered by the arbitrary node ID order alongside the seriated adjacency matrix ordered by the first principal component loading on the cumulative adjacency matrix, $\mathbf{A}^{cum}$.
-Using PCA on $\mathbf{A}^{cum}$ to order the rows and columns of the adjacency matrix visualization clearly shows the two distinct connected components in the first wave of the network, which are difficult to find in the visualization ordered by node ID.
+Using PCA on $\mathbf{A}^{cum}$ to order the rows and columns of the adjacency matrix visualization clearly shows the two distinct connected components in the first network, which are difficult to find in the visualization ordered by node ID.
 We continue to use the seriated matrix in another animation of the microstep process, which is available at \url{https://vimeo.com/240092677}.
@@
   scale_x_discrete(position = "bottom") +
-  labs(x = "alter node (j)", y = "ego node (i)") +
+  labs(x = "alter (j)", y = "ego (i)") +
   ThemeNoNet +
@@
   coord_fixed() + 
-  ggtitle("Theoretical Probabilities")
+  ggtitle("Theoretical Probabilities") + 
+  labs(x= "alter (j)", y = "ego (i)")
 p2 <- ggplot(data = empprobsDat,
@@
   coord_fixed() + 
-  ggtitle("Empirical Probabilities")
+  ggtitle("Empirical Probabilities") + 
+  labs(x= "alter (j)", y = "ego (i)")
 plot_grid(p1, p2)
@@
 These probabilities are defined in Equation~\ref{eq:pij}.
-We only do the first step of many because each microstep is dependent on the previous network state.
+We only do the first step of many because each microstep is dependent on the previous network state, and thus only the first steps can be compared across simulations.
 There are 1,000 transitions to examine: the first microstep taken in each one of the simulations. %Another way to view these transition probabilities is through the adjacency matrix visualization. 
@@
 % removed 
-<<allsteps, fig.cap="Two simulations (out of 1,000) of the microstep process from wave 1 to wave 2.
+<<allsteps, fig.cap="Two simulations (out of 1,000) of the microstep process from $x(t_1)$ to $x(t_2)$.
 The $x$ axis is the microstep number, with step 0 representing the first wave of data and the final step representing the second wave of data.
@@
 
-We have used traditional graphics and net-vis methods to explore a set of continuous time markov chain models for dynamic social network data in new ways.
+We have used several visualizations, some old and some new, to explore a set of continuous time markov chain models for dynamic social network data in new ways.
 The guiding principles of model-vis: viewing the model in the data space, visualizing collections of models, and looking at the underlying algorithms, gave us knowledge and appreciation for these complicated and flexible models.
@@
 As we have shown, the rate and objective function parameters are often highly correlated, so researchers can now use model-vis techniques to gain a better understanding of these correlations and hopefully find ways to account for their effects in the model.
-In addition, extensions of the model can predict behavior change of the actors in the network, a capability we did not explore here.
-
-Furthermore, the data sets we have used are relatively small by network data standards, with only 16 and 100 nodes.
+In addition, some extensions of the CTMC model can aslo model behavior change of the actors in the network, which we have just treated as covariates.
+The CTMC family of models for dynamic networks has immense potential for additional model visualization applications.
+
+Furthermore, the data we have used are quite small by network data standards, with only 16 nodes.
 Thus, some of the visualizations we have presented may not scale up to larger networks.
@@
 
-\noindent \textbf{Online supplemental material:} In addition to the Supporting Information available online from the publisher, much of the code used to create the visualizations in this paper has been packaged and made available online at \url{https://github.com/sctyner/netvizinf} and \url{https://github.com/sctyner/geomnet} (also on CRAN).
+\subsection*{Online supplemental material:}
+
+In addition to the Supporting Information available online from the publisher, much of the code used to create the visualizations in this paper has been packaged and made available online at \url{https://github.com/sctyner/netvizinf} and \url{https://github.com/sctyner/geomnet} (also on CRAN).
 Additional, non-packaged code can be found at \url{http://bit.ly/ctmc-model-viz}.

### 2019-07-02 paper.Rnw
ditch senate data example, clarity in notation, some other minor edits, merge table 1 & figure 4

@@
 \usepackage{multirow}
+\usepackage{booktabs}
 \usepackage{rotating}
@@
 \graphicspath{{figure/}}
-%\usepackage{endfloat} % Figures to the end of the document
+\usepackage{endfloat} % Figures to the end of the document
 
@@
 Statistical models for networks have been studied for decades, and they have gotten more complex as computing power has grown.
-Some network models, such as those for dynamic social networks networks, are analytically intractable and therefore require extensive simulation procedures for model fitting.
-In this paper, we use model visualization techniques introduced in Wickham et al (2015) to better understand some of these models and their behavior.
-We use a three-pronged visualization attack: (a) network visualizations to view the \emph{model}, (b) visualizations of collections of models to interpret parameters, and (c) visualizations of fitting algorithms to gain insights into the underlying mechanisms.
+Some network models, such as those for dynamic social networks, are analytically intractable and therefore require extensive simulation procedures for model fitting.
+In this paper, we use model visualization techniques to better understand some of these models and their behavior.
+We use a three-pronged visualization attack: (a) network visualizations to view the model, (b) visualizations of collections of models to interpret parameters, and (c) visualizations of fitting algorithms to gain insights into the underlying mechanisms.
 With the help of static and dynamic visualizations, we bring the hidden algorithmic fitting processes into the foreground, with the goal of a better understanding and higher accessibility of statistical models for social network analysts.

### 2019-07-02 supporting-info-child.Rnw
ditch senate data example, clarity in notation, some other minor edits, merge table 1 & figure 4

@@
 For the friendship data, the inclusion of $\\beta_3$ has strong effect on the estimates of the other two parameters, but not for the senate data.
-<<compareModels, fig.cap="Density plots of objective function parameter estimates from repeatedly fitting models M1, M2, and M3 to the example data. ", fig.height = 3>>=
+<<compareModels, fig.cap="Density plots of objective function parameter estimates from repeatedly fitting models M1, M2, and M3 to the example data. ", fig.height = 3, eval=FALSE>>=
 simu2$data <- "friends"
@@
 
-The density plots in Figure~\ref{fig:compareModels} show the distribution of the parameter estimates for the objective function parameters in M1, M2, and M3 the two example data sets we used.
+% The density plots in Figure~\ref{fig:compareModels} show the distribution of the parameter estimates for the objective function parameters in M1, M2, and M3 the two example data sets we used.
 The most notable difference is in the values of $\beta_1, \beta_2$.

### 2019-07-03 paper-child.Rnw
working on the intro

@@
 %As humans and researchers, it is our instint to try to learn as much as we can about the world around us. 
-One of the hardest things scholars try to do is understand relationships between friends, family, or colleagues.
-These relationships, or \emph{social networks}, have been studied for decades, in every format from self-help books to publications in academic journals.
-There are also statistical models, such as exponential random graph models (ERGM) and latent space models (LSM), that strive to capture the underlying relationship-generating mechanism \citep{ergm, lsm}.
-These static models, however, are not optimal for modeling ``real-world" social networks because they do not account for changes in the network as time passes.
-\textit{Dynamic social networks}, which are observed at many points in time, are the alternative to static networks.
-The available models for dynamic social networks, however, are more complicated, as these models must account for both the network structure and the way it changes in time.
-But if we can capture the change mechanisms over time, we can gain a better understanding of the underlying network dynamics that drive social connections.
-% are more realistic because they provide mechanisms for analyzing multiple observations of the same network over time.
-
-%., and models for dynamic social networks hold a great deal of promise because of their potential to capture realistic mechanisms.
-%Our social networks evolve over time as new relationships are made, and being able to model these evolutionary mechanisms can uncover hidden truths about our relationships with each other.
-
-\par One family of models for dynamic social networks are continuous time Markov chain (CTMC) models, first introduced for networks in \citet{ctmcorig}.
-These models represent a hierarchical approach to network modeling: the changes happen at the smallest possible level, and subsequent choices are conditioned on the past.
-The transition probabilities between two adjacent states of a network depend on the current structure of the network, the characteristics of the nodes, or both \citep{saompaper}.
-This means that node-level covariate effects, in addition to well-known ERGM effects such as reciprocity or transitivity, can be incorporated into the model.
-%These models are different from other dynamic network models because they allow us to add node level information into the model with the traditional network structure information.
-Adding node-level information into the model is a more intuitive approach to model changes in a social network because, for example
-With these models, we can examine assumptions made about social networks, i.e.
-that people with common interests are more likely to become friends.
-For example, we may assume that teenagers who drink heavily are more likely to be friends with other teenagers who drink heavily.
+Scholars have been trying to understand relationships between people, or \emph{social networks}, for decades, especially in the social sciences.
+There are a host of quantitative methods available for analyzing social networks, including statistical models, such as exponential random graph models (ERGM) and latent space models (LSM), that strive to capture the underlying network-generating mechanism \citep{ergm, lsm}.
+The ERGM and LSM families are examples of static network models, which focus on a single ``snapshot" of a network and ``global network statistics" such as outdegree distribution \citep{goldenberg09}.
+Other static network model families include the Erd\"{o}s-R\'{e}nyi random graph model and the exchangeable graph model \citep{goldenberg09}.
+% describe more about the ergms. cite goldenberg. answer the questions: how are they analyzed? What can they do? what can't they do? HOw are they applied? 
+These static models, however, are not optimal for modeling real networks because they do not account for changes in the network as time passes.
+In real networks, edges may be created or destroyed, and nodes can be born or die \citep{goldenberg09}.
+Dynamic models for networks, which model networks observed at many points in time, are the more realistic alternative to static network models.
+
+Dynamic network models do not have as rich of a history as static network models, largely due to the historical lack of data and available analytical tools \citep{goldenberg09}.
+With the advent of social media and ever-increasing computational power, however, more and more researchers are turning toward dynamic network models.
+Some of the classical models, such as the Erd\"{o}s-R\'{e}nyi model and the small-world model of \citet{watts} are pseudo-dynamic because they could be applied in a dynamic way but rarely are \citep{goldenberg09}.
+% talk up dynamic social network models.
+they are much more realistic.
+they don't treat nodes as exchangeable, they allow the nodes to affect the ties.
+They account for both structural and covariate effects.
+They are very flexible.
+They are the ideal model because they are the only ones that have any chance of actually capturing the mechanisms.
+Fully dynamic social network models, however, are more complicated: they incorporate both the network structure and its changes in time.
+If we can capture this change mechanism in time, we can truly begin to model the underlying network dynamics that drive social connections.
+A comprehensive review of dynamic network models is presented in \citet{golenberg09}.
+
+\par We are particularly interested one truly dynamic family of network models: continuous time Markov chain (CTMC) models, first introduced in \citet{ctmcorig}.
+We chose these models because they correspond to our intuition about how social networks are formed, and they are very flexible and widely applicable.
+In these models, network changes are modeled hierarchically at the node level: nodes are given opportunities to change via a rate function, then the node chooses which edge to change according to its utility function.
+%first step is the changes happen at the smallest possible level (one edge between two nodes), and subsequent changes condition on past changes.
+In each step of the hierarchy, many possible effects can be implemented to capture the change mechanism.
+For example,  the rate function could be a simple waiting-time model, or it could depend on a node's indegree or outdegree: we may have reason to believe that the more "popular" nodes change their ties more frequently.
+In addition, we can incorporate node covariate information into the hierarchy, such as gender or education level of people in a social network.
+For instance, the genders of two people could effect whether and how quickly they establish a relationship, and people will likely form more relationships with those who share their education level.
+Thus, the transition probabilities between two states of a network depend on the network's structure, characteristics of the nodes, or both \citep{saompaper}.
+For example, \citet{depression} use a CTMC model to learn about the tendency of depressed adolescents to form friendships and found that depressed students tended to avoid creating new friendships over time.
+The CTMC model family can also be expanded to model node behavior, as in \citet{chyzh2016}, who applies CTMC models to show that more indirect trade relationships in a country decrease its overall respect for human rights.
+% This means that node-level covariate effects, in addition to commonly used effects such as reciprocity or transitivity, can be incorporated into the model.
+% Cite some of olga's work. what can you do with these models? cite other work with social networks. 
+Thus, with CTMC models we can examine assumptions made about social networks, i.e.
+that people with common qualities are more likely to become friends.
+%For example, we may assume that teenagers who drink heavily are more likely to be friends with other teenagers who drink heavily.
 We can assess our assumptions on shared interests into the modeling process through a particular subfamily of CTMC models known as stochastic actor-oriented models (SAOMs) \citep{saompaper}.
 
-\par With the added flexibility of CTMC models comes increased estimation difficulty \citep{saomreview}.
+\par With the flexibility of CTMC models comes increased estimation difficulty \citep{saomreview}.
 Likelihood functions quickly become very complex due to the inherent dependency structure in the data, and therefore computational methods, such as Markov chain Monte Carlo (MCMC), are used to fit models.
+Methods for maximum likelihood estimation (MLE) for these models were not introduced until \citet{saomsml2010}, and method of moments (Mom) estimation is still used \citep{RSiena}.
 %SAOMs are typically fit to dynamic social network data using a series of MCMC phases for finding method of moments (MoM) estimates of model parameters.
-For CTMC parameter estimation, we use the software SIENA, and its \texttt{R} implementation \texttt{RSiena} by \citet{RSiena} \citep{rstats}.
-This software is a considerable contribution to the field of social network analysis, but the extensive parameter estimation process is largely hidden from the user.
-To better understand CTMCs and the fitting process in \texttt{RSiena}, we use model visualization, or model-vis \citep{modelvis}, to examine the underlying methods, hidden structures, and simulation results.
+For CTMC parameter estimation, we use the software SIENA, and its \texttt{R} implementation \texttt{RSiena} \citep{rstats, RSiena}.
+This software is a considerable contribution to the field of social network analysis, allowing for estimation of structural effects, node covariate effects, as well as modeling both network and node changes.
+This extensive parameter estimation process is largely hidden from the user, and the primary outputs are parameter estimates.
+Direct simulation from a given model with fixed parameter values is also made easy with \texttt{RSiena}.
+%Comprehending these underlying algorithms for parameter estimation is difficult without an advanced statistics degree, so we use model visualization, or model-vis \citep{modelvis}, to examine the methods, hidden structures, simulation results, and model behavior.
 Ultimately, we aim to help researchers better understand the structure, behavior, fit, and results of CTMC models for social network data.
@@
 
-% removed the following plot
-<<adjmatviz, fig.align='center', fig.cap="A visualization of the adjacency matrices of the three waves of network observations in the ``Teenage Friends and Lifestyle Study'' data.
-The subset we will be using is outlined in red.", results='hide',fig.show='asis', fig.width=8, fig.height=3, eval = FALSE, purl=FALSE>>=
-source("code/00e_small_friends.R")
-view_rs2$wave <- factor(view_rs2$wave)
-levels(view_rs2$wave) <- c("Wave 1", "Wave 2", "Wave 3")
-ggplot(data= view_rs2, aes(x = y, y=-x)) +
-  geom_tile(aes(fill = as.factor(value))) +
-  scale_fill_manual("i likes j", labels=c("no", "yes"),
-                    values = c('white', 'black')) +
-  geom_rect(data= NULL, inherit.aes = FALSE, color = 'red',
-            aes(xmin = 19.5, xmax = 35.5, ymax = -19.5, ymin = -35.5, fill =NA)) +
-  facet_wrap(~wave) + theme_bw() +
-  labs(x = "node j (alter)", y = "node i (ego)") +
-  ThemeNoNet +
-  theme(aspect.ratio=1, axis.text = element_blank(), axis.ticks = element_blank(), panel.grid = element_blank())
-@
+% This is a canonical example in the social network modeling literature. That's why we chose it. 
+% It's easy to understand: everyone was once a teenager and knows how fragile those relationships are and how 
+% much and how quickly they can change. 
+% data has previously been used to quantify behavior & identify which ones will take risks. @pearson2003drifting
 
@@
 $s_{i3}(x,z) = \sum_{j \neq h} x_{ij}x_{ih}x_{hj} \mathbb{I}(z_i = z_h \neq z_j)$}
-\end{subfigure}\hfill
+\end{subfigure}
 \begin{subfigure}[t]{.45\textwidth}
@@
 @
-
-<<heatmap, fig.align='center', fig.cap='Adjacency matrix visualizations showing the theoretical (left) and empirical (right) transition probabilities for the first microstep taken in 1,000 simulations.
-The ego node is on the y-axis, and the alter node is on the x-axis.
-There are many ties with empirical probability zero.', fig.height=2.5>>=
+\begin{figure}
+\begin{subfigure}[t]{\textwidth}
+<<heatmap, fig.align='center', fig.height=2.5>>=
 
@@
 @
+\caption{\label{fig:heatmap}Adjacency matrix visualizations showing the theoretical (left) and empirical (right) transition probabilities for the first microstep taken in 1,000 simulations.
+There are many ties with empirical probability zero.}
+\end{subfigure}
+\begin{subfigure}[t]{\textwidth}
+<<alledgests, fig.height = 3>>=
+# get a plot of all microsteps from w1 to w2 by edge
+alledges <- expand.grid(from = paste0("V", 1:16), to = paste0("V", 1:16))
+alledges$eid <- 1:nrow(alledges)
+
+sfmsall %>% left_join(alledges) -> sfmsall2
+
+sfmsall2 %>% group_by(rep) %>% mutate(isWave2 = (ms == max(ms))) -> sfmsall2
+
+sfmsall2 %>% group_by(from, to, eid) %>%
+  summarize(total =n()) %>%
+  arrange(desc(total)) %>%
+  ungroup %>%
+  mutate(plotorder = row_number()) -> sfmsallorder
+
+sfmsall2 %>% group_by(ms, eid) %>%
+  summarise(count = n()) %>%
+  mutate(alpha = count/1000) %>%
+  arrange(ms, eid) -> sfmsall3
+
+left_join(sfmsall3, sfmsallorder, by = "eid") -> sfmsall4
+
+#add true wave 2 to plot
+wave2 <- na.omit(actual2)
+wave2$from <- paste0("V", wave2$from)
+wave2$to <- paste0("V", wave2$to)
+wave2 <- left_join(wave2, alledges)
+wave2 <- left_join(wave2, sfmsallorder)
+
+sfmsall5 <- sfmsall4 %>% ungroup %>% add_row(ms = max(sfmsall4$ms) + 1,
+        eid = wave2$eid,
+        count = NA,
+        alpha = 1,
+        from = wave2$from,
+        to = wave2$to,
+        total = wave2$total,
+        plotorder = wave2$plotorder)
+sfmsall5$alphaFctr <- cut(sfmsall5$alpha, c(0,.01, .05, .10, .25, .5, .75, 1, 1.1))
+ggplot(data = sfmsall5) +
+  geom_tile(aes(x = ms, y = reorder(eid,-plotorder), fill = alphaFctr)) +
+  annotate(geom = "rect", xmin=.25, xmax = 1.75, ymin = -Inf, ymax = Inf, color = "red", fill = NA, size = .01, alpha = .15) + 
+  #scale_fill_gradient(low = 'grey90', high = '#0868ac', name = "occurrence %") +
+  scale_fill_brewer(palette = "Greys", name = "Occurrence %", labels = c("(0-1]%", "(1-5]%", "(5-10]%","(10-25]%", "(25-50]%", "(50-75]%", "(75-100]%")) +
+  labs(x = "Microstep No.", y = "Edges (ordered by total # occurrences)") +
+  ThemeNoNet +
+  theme(axis.ticks.y = element_blank(), axis.text.y = element_blank(), panel.grid = element_blank())
+@
+\caption{\label{fig:alledgests}Visualizing all microsteps taken in 1,000 simulations from the model M1.
+All $16\times(16-1)=240$ possible edges are represented on the y-axis, which is ordered by how often edges occur in the simulation process.
+The occurrence percent is split up into groups to correspond with its distribution: only about 10\% of the possible edges appear more than 10\% of the time in the 1,000 simulations, while about 60\% appear less than 1\% of the time.
+The network $x(t_1)$ is shown at microstep 0, and the network $x(t_2)$ is shown as the last microstep for comparison.
+We see that it is rare for a microstep process to last longer than 150 steps.}
+\end{subfigure}
+\caption{\label{fig:probabilities}Visualizing the underlying algorithms when fitting CTMC models.
+The emprical probability plot (on the right in Figure~\ref{fig:heatmap}) is also shown at microstep number 1 in Figure~\ref{fig:alledgests}.}
+\end{figure}
 

### 2019-07-03 supporting-info-child.Rnw
working on the intro

@@
 The first wave network is shown at microstep 0, and the second wave of the network is shown as the last microstep for comparison.
-We see that it is rare for a microstep process to last longer than 150 steps, and also that the edges that appear past the 150th step tend to be in either the first wave or the second wave.">>=
+We see that it is rare for a microstep process to last longer than 150 steps, and also that the edges that appear past the 150th step tend to be in either the first wave or the second wave.", eval=FALSE>>=
 # get a plot of all microsteps from w1 to w2 by edge
@@
 
+% move this into the body eventually. 
 Finally, we combine 1,000 simulations from model M1 into a visualization that displays the entire microstep process in Figure~\ref{fig:alledgests}.

### 2019-07-05 paper-child.Rnw
forgot to commit end of day wed.
intro almost there... still needs some kaboom.

@@
 Likelihood functions quickly become very complex due to the inherent dependency structure in the data, and therefore computational methods, such as Markov chain Monte Carlo (MCMC), are used to fit models.
-Methods for maximum likelihood estimation (MLE) for these models were not introduced until \citet{saomsml2010}, and method of moments (Mom) estimation is still used \citep{RSiena}.
+Methods for maximum likelihood estimation (MLE) for these models were not introduced until \citet{saomsml2010}, and method of moments (Mom) estimation is frequently used \citep{RSiena}.
 %SAOMs are typically fit to dynamic social network data using a series of MCMC phases for finding method of moments (MoM) estimates of model parameters.
@@
 This extensive parameter estimation process is largely hidden from the user, and the primary outputs are parameter estimates.
-Direct simulation from a given model with fixed parameter values is also made easy with \texttt{RSiena}.
-%Comprehending these underlying algorithms for parameter estimation is difficult without an advanced statistics degree, so we use model visualization, or model-vis \citep{modelvis}, to examine the methods, hidden structures, simulation results, and model behavior.
-Ultimately, we aim to help researchers better understand the structure, behavior, fit, and results of CTMC models for social network data.
-
-In the remainder of this section, we briefly introduce the structure of network data and two different network visualization (net-vis) techniques.
-Then, we carefully define the hierarchical model structure of a CTMC.
-Finally, we briefly ouline the CTMC fitting process in \texttt{RSiena} that we use throughout this paper \citep{SienaAlgs}.
-
-%\par To learn more about the SAOM fitting process, we use model visualization techniques (see \citet{modelvis}) to display the model in the data space, view collections of models instead of single models, and visually explore the algorithms fitting the SAOMs.
-SAOMs are a model family that can reap many benefits from the application of model visualization techniques.
-For instance, the models themselves include a continuous-time Markov chain (CTMC) that is completely hidden from the analyst in the model fitting process.
-Visualizing the CTMC can provide researchers with more insight into the underlying features of the model.
-In addition, SAOMs can include a great deal of parameters in the model structure, each of which is attached to a network statistic.
-These statistics are often somewhat, if not highly, correlated, which causes high correlation between the corresponding parameters in a SAOM.
-By visualizing collections of SAOMs, we gain a better understanding of these correlations and can find ways to account for their effects in the model.
-Furthermore, the estimation of the parameters in a SAOM rely on convergence checks based on simulations from the fitted model.
-These simulations are hidden and unsaved by default for computational efficiency.
-With the help of both static and dynamic visualizations, we bring the hidden model fitting processes into the foreground, eventually leading to a better understanding and higher accessibility of stochastic actor-oriented models for social network analysts.
-
-%The remainder of this paper is organized as follows: In Section~\ref{sec:netintro}, we introduce basic concepts of networks and network visualizations.
-In Section~\ref{sec:saoms}, we formally define the family of stochastic actor-oriented models for social network analysis.
-In Section~\ref{sec:ModelVis}, we combine concepts from Sections~\ref{sec:netintro} and \ref{sec:saoms} in an application of the model-vis paradigm, and conclude with a discussion in Section~\ref{sec:discussion}.
+Direct simulation from a given network model with fixed parameter values is also made easy with \texttt{RSiena}.
+This software opens up a great deal of possibilities for network modeling and simulation.% need to emphasize the need for visualization.
+
+% ok. we know why ctmcs but WHY visualize them? where is the motivation?? who cares and why? 
+% talk about lack of goodness of fit diagnostics. 
+After choosing the CTMC models for dynamic social networks, we would like to determine how well they fit data.
+We chose these models because they match our intuition about how social networks form in reality, and after we fit a model, we will want to assess its goodness-of-fit.
+In statistical network models generally, the lack of large-sample theory leads to a lack of goodness-of-fit tests \citep{goldenberg09}.
+Most methods for assessing goodness-of-fit for network models, therefore, involve comparing statistics computed on simulated networks to the same statistics computed on the data, for example in \citet{hunter2008}.
+More recently, \citet{gofschwein} used score-type tests for forward model selection.
+The \texttt{RSiena} software also contains methods for goodness-of-fit based on network statistics, such as outdegree distribution, where the data values are compared to simulated values as in \citet{hunter2008} \citep{RSiena}.
+% need to transition to model vis.
+% say something about how important visualization is in network analysis.
+
+We propose new methods for goodness-of-fit and other diagnostics that use visualization as a supplement to computational methods.
+Network visualization, also called graph drawing, is a prominent subfield of network analysis with a rich history: early drawings of complete graphs date back to 1280 CE.
+We use this visualization foundation to combine modeling with visualization.
+Note that we do not propose new methods of visualizing networks, and for a complete review of the field of graph drawing we direct the reader to \citet{graphdrawing}.
+What is novel in our work is the use of common visualization methods to diagnose network models.
+By translating the model visualization work of \citet{modelvis} to the dynamic network model space, we % do something awesome
+
+In the remainder of this section, we provide an overview of the structure of network data and discuss two common network visualization (net-vis) techniques.
+Then, we detail the hierarchical model structure of a CTMC following the definition in \citet{saompaper} and ouline the CTMC model fitting process in \texttt{RSiena} \citep{SienaAlgs}.
 
@@
 %\noindent \textbf{Network Data Structure:} 
-Network data across fields have similar structure: units of observation and connections between those units.
+Network data across fields have similar structure: the data consist of units of observation and connections between those units.
 In a social network, the units of observations, called \textit{nodes} or \textit{actors}, are people, while the connections, called \textit{edges} or \textit{ties}, are their relationships with each other.
 %Networks change over time, like when new relationships are formed in a social network.
-Other network data include inherent variables of interest on the nodes, such as age or gender, and the type of relationship (friendship, coworkers, etc) that an edge represents.
+Other network data include inherent variables of interest on the nodes, such as age or gender, and the type of relationship (friends, coworkers, etc) that an edge represents.
+Dynamic networks are observed at many discrete points in time, where each observation of a network is often called a "wave" of data.
 %These two network data structures pose unique problems to analysts, such as, ``How does the strength of a tie between two nodes affect the overall structure of the network?" or ``Do differences in node variables affect the formation of edges?" We propose to answer these questions and more through visual exploration of network data and models.
@@
 
-\noindent \textbf{Visualizing Network Data:} Network visualization, also called graph drawing, is a prominent subfield of network analysis.
-For a review of graph drawing, see \citet{graphdrawing}.
-Visualizing network data is inherently challenging due to the structure of the data itself.
+\noindent \textbf{Visualizing Network Data:} Visualizing network data is inherently challenging due to the structure of the data itself.
 Most data visualizations rely on well-defined axes inherited from the data, such as Cartesian coordinates or spatial locations, but network data typically do not have such built-in structure.
@@
 These lines can also have arrows on them indicating the direction of the edge for directed networks, as shown in Figure~\ref{fig:toyplots}.
-Because there is often no natural placement of the nodes, they are placed in 2D at random, then adjusted with one of the many layout algorithms available \citep{drawingalgs}.
+Because there is often no natural placement of the nodes, they are placed in 2D at random, then adjusted with a layout algorithm.
+For a complete overview of available node-link layout algorithms, we direct the reader to \citet{drawingalgs}.
 Some commonly used algorithms, such as the Kamada-Kawai (KK) layout \citep{kamadakawai} and the Fruchterman-Reingold (FR) layout \citep{fruchterman-reingold}, are designed to mimic physical systems, drawing the graphs based on the ``forces" connecting them.
@@
 its eigenstructure), to place the nodes \citep{drawingalgs}.
-In Figure~\ref{fig:toyplots}, we use the KK algorithm, and unless otherwise stated, all other node-link diagrams in this paper were drawn with the KK layout algorithm.
+In Figure~\ref{fig:toyplots}, we use the KK algorithm, and unless otherwise stated, all other node-link diagrams in this paper are laid out using the KK algorithm \citep{kamadakawai}.
 
@@
 The adjacency matrix visualization for our toy example is also shown in Figure~\ref{fig:toyplots}.
-This visualization shows the network as a grid, with each row representing the ``from" node and each column representing the ``to" node of an edge.
+This visualization shows the network as a grid, with each row representing the ``from" or ``ego" node and each column representing the ``to" or ``alter" node of an edge.
 The grid space for edge $i\rightarrow j$ is filled in if $A_{ij} = 1$ and is empty if $A_{ij} = 0$.
-For undirected networks, $\mathbf{A}$ and the grid visualization are symmetric, with each edge represented twice.
+For undirected networks $\mathbf{A}$ is symmetric, so the adjacency matrix visualization is also symmetric, and each edge is drawn twice.
 
 \par Each net-vis method has its own advantages and disadvantages.
-For instance, \emph{paths} between two nodes in a network are easier to determine with node-link diagrams than with adjacency matrix visualizations \citep{adjmatviz}.
-In node-link diagrams, node information can be incorporated into the visualization by coloring or changing the shape of the points, and edge-level information can be incorporated by coloring the lines, or changing their thickness.
+For instance, paths between two nodes in a network are easier to determine with node-link diagrams than with adjacency matrix visualizations \citep{adjmatviz}.
+Furthermore, in node-link diagrams, node information can be incorporated into the visualization by coloring or changing the shape of the points, and edge-level information can be incorporated by coloring the lines, or changing their thickness.
 Visualizing a node variable in an adjacency matrix visualization is not as simple, because this type of net-vis features the edges.
@@
 A node-link diagram, however, may underrepresent the edge count if the edges $x_{ij}$ and $x_{ji}$ both exist and are drawn on top of one another.
-Ultimately, there is no one ``correct" way to visualize network information, and we will be using both the node-link and adjacency matrix net-vis methods throughout this paper to explore our example data and CTMC models.
+Ultimately, there is no one ``correct" way to visualize network information, and we will be using both the node-link and adjacency matrix net-vis methods throughout this paper to visualize our example data and CTMC models.
 

### 2019-07-08 paper-child.Rnw
stuff from last week

@@
 Dynamic network models do not have as rich of a history as static network models, largely due to the historical lack of data and available analytical tools \citep{goldenberg09}.
-With the advent of social media and ever-increasing computational power, however, more and more researchers are turning toward dynamic network models.
+With the advent of social media and ever-increasing computational power, however, more and more researchers are turning their attention toward dynamic network models.
 Some of the classical models, such as the Erd\"{o}s-R\'{e}nyi model and the small-world model of \citet{watts} are pseudo-dynamic because they could be applied in a dynamic way but rarely are \citep{goldenberg09}.
@@
 If we can capture this change mechanism in time, we can truly begin to model the underlying network dynamics that drive social connections.
-A comprehensive review of dynamic network models is presented in \citet{golenberg09}.
+A comprehensive review of dynamic network models is presented in \citet{goldenberg09}.
 
@@
 In each step of the hierarchy, many possible effects can be implemented to capture the change mechanism.
-For example,  the rate function could be a simple waiting-time model, or it could depend on a node's indegree or outdegree: we may have reason to believe that the more "popular" nodes change their ties more frequently.
+For example,  the rate function could be a simple waiting-time model, or it could depend on a node's indegree or outdegree: we may have reason to believe that the more popular nodes change their ties more frequently.
 In addition, we can incorporate node covariate information into the hierarchy, such as gender or education level of people in a social network.
@@
 % Cite some of olga's work. what can you do with these models? cite other work with social networks. 
-Thus, with CTMC models we can examine assumptions made about social networks, i.e.
-that people with common qualities are more likely to become friends.
+Thus, with CTMC models we can examine assumptions made about social networks.
 %For example, we may assume that teenagers who drink heavily are more likely to be friends with other teenagers who drink heavily.
@@
 
-\par With the flexibility of CTMC models comes increased estimation difficulty \citep{saomreview}.
+\par With the flexibility of CTMC models comes increased complexity and estimation difficulty \citep{saomreview}.
 Likelihood functions quickly become very complex due to the inherent dependency structure in the data, and therefore computational methods, such as Markov chain Monte Carlo (MCMC), are used to fit models.
-Methods for maximum likelihood estimation (MLE) for these models were not introduced until \citet{saomsml2010}, and method of moments (Mom) estimation is frequently used \citep{RSiena}.
+Methods for maximum likelihood estimation (MLE) for these models were not introduced until \citet{saomsml2010}, and method of moments (MM) estimation is frequently used \citep{RSiena}.
 %SAOMs are typically fit to dynamic social network data using a series of MCMC phases for finding method of moments (MoM) estimates of model parameters.
 For CTMC parameter estimation, we use the software SIENA, and its \texttt{R} implementation \texttt{RSiena} \citep{rstats, RSiena}.
-This software is a considerable contribution to the field of social network analysis, allowing for estimation of structural effects, node covariate effects, as well as modeling both network and node changes.
+This software is a considerable contribution to the field of social network analysis, allowing for estimation of structural effects, node covariate effects, as well as modeling both network and node changes in a wide variety of CTMC models.
 This extensive parameter estimation process is largely hidden from the user, and the primary outputs are parameter estimates.
@@
 % talk about lack of goodness of fit diagnostics. 
-After choosing the CTMC models for dynamic social networks, we would like to determine how well they fit data.
+Since we chose the CTMC models for dynamic social networks, we would like to determine how well they fit data.
 We chose these models because they match our intuition about how social networks form in reality, and after we fit a model, we will want to assess its goodness-of-fit.
@@
 Most methods for assessing goodness-of-fit for network models, therefore, involve comparing statistics computed on simulated networks to the same statistics computed on the data, for example in \citet{hunter2008}.
-More recently, \citet{gofschwein} used score-type tests for forward model selection.
+More recently, \citet{gofschwein} used score-type tests for forward model selection in the CTMC model context.
 The \texttt{RSiena} software also contains methods for goodness-of-fit based on network statistics, such as outdegree distribution, where the data values are compared to simulated values as in \citet{hunter2008} \citep{RSiena}.
@@
 We propose new methods for goodness-of-fit and other diagnostics that use visualization as a supplement to computational methods.
-Network visualization, also called graph drawing, is a prominent subfield of network analysis with a rich history: early drawings of complete graphs date back to 1280 CE.
+Network visualization, also called graph drawing, is a prominent subfield of network analysis with a rich history: early drawings of complete graphs date back to 1280 CE \citep{knuth2000}.
 We use this visualization foundation to combine modeling with visualization.
 Note that we do not propose new methods of visualizing networks, and for a complete review of the field of graph drawing we direct the reader to \citet{graphdrawing}.
-What is novel in our work is the use of common visualization methods to diagnose network models.
-By translating the model visualization work of \citet{modelvis} to the dynamic network model space, we % do something awesome
+What is novel in our work is the use of common visualization methods to diagnose network model with model visualization
+
+\emph{Model visualization} complements traditional model diagnostic tools such as $R^2$ values and residual plots to enrich our understanding and interpretation of statistical models \citep{modelvis}.
+We apply the three principles of model-vis: view the model in the data space, visualize collections of models, and explore the process of fitting the model, to the models and data introduced in Section~\ref{sec:exdatamods}.
+By translating the model visualization work of \citet{modelvis} to the dynamic network model space, we provide network researchers with a new suite of tools for visualizing and diagnosing performance of their network models.
+Though we apply the tools to CTMCs because we find the models intuitive, the methods we use can be applied to any network model, static or dynamic.% do something awesome
 
@@
 %\noindent \textbf{Network Data Structure:} 
-Network data across fields have similar structure: the data consist of units of observation and connections between those units.
+Network data across fields have similar structure: they consist of units of observation and connections between those units.
 In a social network, the units of observations, called \textit{nodes} or \textit{actors}, are people, while the connections, called \textit{edges} or \textit{ties}, are their relationships with each other.
@@
 Other network data include inherent variables of interest on the nodes, such as age or gender, and the type of relationship (friends, coworkers, etc) that an edge represents.
-Dynamic networks are observed at many discrete points in time, where each observation of a network is often called a "wave" of data.
+Dynamic networks are observed at many discrete points in time, where each observation of a network is often called a ``wave" of data.
 %These two network data structures pose unique problems to analysts, such as, ``How does the strength of a tie between two nodes affect the overall structure of the network?" or ``Do differences in node variables affect the formation of edges?" We propose to answer these questions and more through visual exploration of network data and models.
@@
 
-\par The second method for net-vis uses the adjacency matrix of the network. The adjaceny matrix of a network, $\mathbf{A}$ is defined as:
+\par The second method for net-vis uses the adjacency matrix of the network. The adjaceny matrix of a network, $\mathbf{A}(x)$ is defined as:
 \begin{equation}\label{eq:adjmat}
-  A_{ij} =
+  A_{ij}(x) =
   \begin{cases}
-             1 &  \quad \text{if an edge exists } i\to j, \quad i \neq j \\
+             1 &  \quad \text{if } x_{ij} = 1, \quad i \neq j \\
              0 & \quad \text{otherwise}
@@
 This visualization shows the network as a grid, with each row representing the ``from" or ``ego" node and each column representing the ``to" or ``alter" node of an edge.
-The grid space for edge $i\rightarrow j$ is filled in if $A_{ij} = 1$ and is empty if $A_{ij} = 0$.
-For undirected networks $\mathbf{A}$ is symmetric, so the adjacency matrix visualization is also symmetric, and each edge is drawn twice.
+The grid space for $x_{ij}$ (the edge $i\rightarrow j$) is filled in if $A_{ij} = 1$ and is empty if $A_{ij} = 0$.
+For undirected networks $\mathbf{A}$ is symmetric, so the adjacency matrix visualization is also symmetric, with each edge drawn twice.
 
@@
 For instance, paths between two nodes in a network are easier to determine with node-link diagrams than with adjacency matrix visualizations \citep{adjmatviz}.
-Furthermore, in node-link diagrams, node information can be incorporated into the visualization by coloring or changing the shape of the points, and edge-level information can be incorporated by coloring the lines, or changing their thickness.
+Furthermore, in node-link diagrams, node information can be incorporated into the visualization by coloring or changing the shape of the points, and edge-level information can be incorporated by coloring the lines or changing their thickness.
 Visualizing a node variable in an adjacency matrix visualization is not as simple, because this type of net-vis features the edges.
@@
 
-The continuous-time Markov chain model family is for analyzing \textit{dynamic} networks, so we can use them to study the change of a network in time while also examining the structural and node covariate efects.
+The continuous-time Markov chain (CTMC) model family is for analyzing dynamic networks, so we can use them to study the change of a network in time while also examining the structural and node covariate efects.
 To define CTMCs for our purposes, we follow the general hierarchical definition outlined in \citet{saomreview} and elsewhere (see e.g.\ \citet{snijders01} and \citet{snijdersetal2010}).
@@
 
-A \textit{dynamic network} is consists of a fixed set of $n$ nodes, that is changing over time, and is observed at $M$ discrete time points, $t_1, \dots, t_M$ with $t_1 < t_2 < \dots < t_M$.
+A dynamic network consists of a fixed set of $n$ nodes that is changing over time, and is observed at $M$ discrete time points, $t_1, \dots, t_M$ with $t_1 < t_2 < \dots < t_M$.
 We denote the network observation at timepoint $t_m$ by $x(t_m)$ for $m = 1, \dots M$.
@@
 Nearly all other steps in the chain are unobserved, with the exception of $x(t_2), \dots, x(t_{M-1})$.
-Unlike the first and last observations of the network, these observations do not have direct correspondence with steps in the CTMC.
-Thus, the observations $x(t_2), \dots, x(t_{M-1})$ are considered to be ``snapshots" of the network at some point between two steps in the CTMC.
+Unlike the first and last observations of the network, $x(t_2), \dots, x(t_{M-1})$ do not have direct correspondence with steps in the CTMC.
+Thus, the observations $x(t_2), \dots, x(t_{M-1})$ treated as ``snapshots" of the network at some point between two steps in the CTMC.
 
@@
 With this simple rate function, we have $M-1$ rate parameters, one for each transition from $x(t_{m})$ to $x(t_{m+1})$ for $m = 1, \dots, M-1$.
-The rate parameter, $\alpha_m$ dictates how often an actor $i$ ``gets an opportunity" to change one of its ties, $x_{ij}$, for $j \neq i \in \{1, \dots, n\}$ in the time period from $t_{m}$ to $t_{m+1}$ for $m = 1, \dots, M-1$.
-The CTMC model assumes that the actors $i$ are conditionally independent given their ties, $x_{i1}, \dots, x_{in}$ at the current network state.
-Let $\tau(i|x,m) $ be the wait time until actor $i$ gets to the opportunity to change its ties.
+The rate parameter, $\alpha_m$ dictates how often an actor $i$ is selected to change one of its ties, $x_{ij}$, for $j \neq i \in \{1, \dots, n\}$ in the time period from $t_{m}$ to $t_{m+1}$ for $m = 1, \dots, M-1$.
+The CTMC model assumes that the actors $i$ are conditionally independent given their ties $x_{i1}, \dots, x_{in}$ at the current network state.
+Let $\tau(i|x,m)$ be the wait time until actor $i$ gets to the opportunity to change its ties.
 For any time point $T$, where $t_{m} \leq T < t_{m+1}$, the waiting time to actor $i$'s next change follows an exponential distribution,
 \begin{equation} \label{eq:c1}
- \tau(i|x,m) | x_{i1}(m), \dots, x_{in}(m) \stackrel{\text{iid}}{\sim} Exp(\alpha_m)
+ \tau(i | x_{i1}(m), \dots, x_{in}(m)) \stackrel{\text{iid}}{\sim} Exp(\alpha_m)
 \end{equation}
@@
 \begin{equation}\label{eq:cdiff}
-C_m = \sum_i\sum_j |x_{ij}(t_{m+1}) - x_{ij}(t_m)|
+C_m = \sum_i\sum_j |x_{ij}(t_{m+1}) - x_{ij}(t_m)|,
 \end{equation}
-which is the total number of changes from $x(t_m)$ to $x(t_{m+1})$.
+which is the total number of edge differences between $x(t_m)$ and $x(t_{m+1})$.
 
@@
 
-The parameters of $f_i(\boldsymbol{\beta}, x, \mathbf{Z})$ are either structural or covariate parameters.
-The structural parameters correspond to statistics that are \emph{only} functions of the current network state, $x$, while the covariate parameters are also functions of the covariates, $\mathbf{Z}$.
+The parameters of $f_i(\boldsymbol{\beta}, x, \mathbf{Z})$ are either structure or covariate parameters.
+The structure parameters correspond to statistics that are \emph{only} functions of the current network state, $x$, while the covariate parameters are also functions of the covariates, $\mathbf{Z}$.
 According to \citet[p.
@@
 The former measures the propensity of nodes with a lot of outgoing ties to form more outgoing ties (the ``rich get richer" effect), while the latter measures the tendency of outgoing ties to be returned within a network.
-The statistics corresponding to these two parameters, $s_{i1}(x)$ and $s_{i2}(x)$ are given in Table~\ref{tab:effects}.
+The statistics corresponding to these two parameters, $s_{i1}(x)$ and $s_{i2}(x)$ are given in Figure~\ref{fig:structures} with visualizations of the corresponding effects.
 In the \texttt{RSiena} software, there are over 80 possible $\boldsymbol{\beta}$ parameters with corresponding statistics $s_{ik}(x, \mathbf{Z})$ to add to the model.
@@
 
-In the CTMC, when actor $i$ can make a change, it chooses which tie $x_{i1}, \dots, x_{in}$ to change at random according to the probability $p_{ij}(\boldsymbol{\beta}, x, \mathbf{Z})$, defined via the objective function.
+In the CTMC, when actor $i$ can make a change, it chooses which tie $x_{i1}, \dots, x_{in}$ to change at random according to the probability $p_{ij}(\boldsymbol{\beta}, x, \mathbf{Z})$.
 Let $x(i \leadsto j)$ be the network identical to network $x$ with the exception of tie $x_{ij}$, which is equal to $1-x_{ij}$; i.e.\ the presence or absence of node $x_{ij}$ is flipped in the new network.
@@
 This matrix describes the rate of change between two states of the CTMC process, and the rows of this matrix always add to zero.
-The state space for a network, denoted $\mathcal{X}$, contains $2^{n(n-1)}$ states: two possible states for an edge, $\{0,1\}$, and $n(n-1)$ edge relationships, as the network is directed and we exclude self-ties.
+The state space for a network, denoted $\mathcal{X}$, contains $2^{n(n-1)}$ states: there are two possible states for an edge, $\{0,1\}$, and $n(n-1)$ possible edge relationships, as the network is directed and we exclude self-ties.
 The intensity matrix is then a square matrix of dimension $2^{n(n-1)}$.
@@
 fix it.
-The full SIENA MoM estimation algorithm operates in three phases, as described in \citet{RSienamanual} and \citet{SienaAlgs}, and we briefly summarize these phases here.
+The full SIENA MoM estimation algorithm operates in three phases, as described in \citet{RSienamanual} and \citet{SienaAlgs}, which we briefly summarize here.
 The first phase obtains initial estimates of the score functions for use in the second phase.
@@
 In Section~\ref{sec:algorithms}, we further explore these phases in the SIENA MoM algorithm through visualization.
-%, bringing them out of their ``black box" and into the light.
-
-The rest of the paper is strucured as follows.
-In Section \ref{sec:exdatamods}, we introduce the example data and models we use for model-vis.
-Then, in Section~\ref{sec:minds}, we explore the first step in model-vis by placing the models in the data space.
-Section~\ref{sec:collectmodels} follows, in which we visualize collections of models.
-Finally, in Section~\ref{sec:algorithms}, we visualize the underlying algorithms in the CTMC fitting process.
+For a more detailed description of the fitting process, we direct the reader to \citet{RSienamanual, SienaAlgs}%, bringing them out of their ``black box" and into the light.
+
+The rest of this paper is strucured as follows.
+In Section \ref{sec:exdatamods}, we introduce the motivating data and CTMC models we assess using model-vis.
+Then, in Section~\ref{sec:minds}, we explore the first step in model-vis by placing the models in the data space to determine how well the models fit the data.
+Section~\ref{sec:collectmodels} follows, in which we visualize collections of models to explore their behavior.
+Finally, in Section~\ref{sec:algorithms}, we visualize the underlying algorithms in the CTMC fitting process to gain insight into the fitting process.
 We conclude with a discussion in Section~\ref{sec:discussion}.
@@
 <<alldataviz, fig.cap="The small friendship network data we model throughout this paper.
+The FR layout is used.
 The teenagers' drinking behavior changes and the network becomes less densely connected over time.
@@
 alldat$X2 <- as.factor(alldat$X2)
+set.seed(35291)
 alldataviz <- ggplot(data = alldat,
@@
   geom_net(fiteach = FALSE, directed=TRUE, layout.alg = 'fruchtermanreingold', size = 4, arrowsize = .25, arrowgap = .05, ecolour = 'black', labelon=T, labelcolour = 'white', fontsize = 2, vjust = .5, linewidth=0.25) +
-  scale_color_manual(values = cols[8:5], name = "Drinking\nBehavior", labels = c("never", "1-2 times per year", "once per month", "once per week")) +
+  scale_color_manual(values = cols[5:8], name = "Drinking\nBehavior", labels = c("never", "1-2 times per year", "once per month", "once per week")) +
   ThemeNet +
@@
 
-\noindent \textbf{Example Data:} To guide our visual exploration of CTMC models for dynamic networks, we use two example data sources.
-The first is a subset of the 50 actor dataset from the ``Teenage Friends and Lifestyle Study" that is provided on the \texttt{RSiena} webpage.
-These data first appeared in \citet{friendsdata}, and we chose to only work with a subset of the data to make node-link diagrams easier to see and to make any changes in the network more noticeable.
+\noindent \textbf{Example Data:} To guide our visual exploration of CTMC models for dynamic networks, we use a canonical example in the social network literature: a subset of the 50 actor dataset from the ``Teenage Friends and Lifestyle Study"  \citet{friendsdata}.
+We chose this data because it is a well-known and easy to understand example.
+We were all teenagers once, and we know the volatility and fragility of those relationships, as well as why they might change.
+We chose to only work with a subset of the data to make node-link diagrams easier to see and to make any changes in the network more noticeable.
 We visualize this data with a node-link diagram in Figure~\ref{fig:alldataviz}, where %For model fitting, we condition on wave 1 and estimate the parameters of the models for the second and third waves.
@@
 We can see that, as time passes, the students seem to drink more and become increasingly isolated into smaller groups.
-An analysis of this type of data with a CTMC model should capture these dynamics in a way that allows the researcher to draw conclusions about the nature of the network and behavioral forces at play.
+An analysis of this type of data with a CTMC model should capture these dynamics in a way that allows the researcher to draw conclusions about the nature of the network and the behavioral forces at play.
 
@@
 Each model uses a simple rate function, $\alpha_m$, and an objective function with two or three parameters.
-The first model, M1, contains the absolute minimum number of parameters in the objective function $f_i(x)$:
+Though our models are very simple, the model-vis techniques we present can easily be applied to much more complex models.
+The first model of interest, M1, contains the absolute minimum number of parameters in the objective function $f_i(x)$:
 \begin{equation}\label{eq:m1}
@@
 \end{equation}
-where $s_{i1}$ is the density network statistic and $s_{i2}$ is the reciprocity network statistic for actor $i$ at the current network state $x$ (see Table~\ref{tab:effects}).
+where $s_{i1}(x)$ is the density network statistic and $s_{i2}(x)$ is the reciprocity network statistic for actor $i$ at the current network state $x$ (see Figure~\ref{fig:structures}).
 The second and third models, M2 and M3, contain one additional parameter each in the objective function.
-We chose the additional parameters because they were significant ($p$-value $<$ 0.05) according to a Wald-type test provided in the \texttt{RSiena} software \citep{RSienaTest}.
+We chose the additional parameters because they were significant ($p$-value $<$ 0.05) according to a Wald-type test (\texttt{Wald.RSiena()}) provided in the \texttt{RSiena} software \citep{RSienaTest}.
 The model M2 contains an actor-level covariate parameter, and the model M3 contains an additional strutural effect in the objective function.
@@
 \caption{\label{fig:beta1}Demonstrating the outdegree effect $\beta_1$. The dotted tie is encouraged if $\beta_1$ is positive. $s_{i1}(x) = \sum_{j} x_{ij}$}
-\end{subfigure}
+\end{subfigure}\hfill
 \begin{subfigure}[t]{.45\textwidth}
@@
            labelcolour='white',vjust = 0.5, hjust =0.5, arrowgap = .15,
-           size=10, fontsize = 5, ecolour="black",
+           size=7, fontsize = 4, ecolour="black",
            linetype = c("solid","dotted")) +
@@
 @
-\caption{\label{fig:jtt}Realization of a JTT, where the covariate is represented by the shape of the nodes.
+\caption{\label{fig:jtt}Demonstrating a JTT, where the covariate is represented by the shape of the nodes.
 The dotted tie is encouraged if $\beta_3$ is positive.
 $s_{i3}(x,z) = \sum_{j \neq h} x_{ij}x_{ih}x_{hj} \mathbb{I}(z_i = z_h \neq z_j)$}
-\end{subfigure}
+\end{subfigure}\hfill
 \begin{subfigure}[t]{.45\textwidth}
@@
 @
-\caption{\label{fig:dad}Realization of a N22 between actors $i$ and $j$.
+\caption{\label{fig:dad}Demonstracting a N22 between actors $i$ and $j$.
 The dotted tie is encouraged if $\beta_4$ has a positive value.
@@
 \end{subfigure}
-\caption{\label{fig:structures} The additional network effects included in the models we fit to the friends data.
-On the left, a JTT between $i$ and $j$, who have different covariate values, and on the right, a N22 between $i$ and $j$, who are indirectly connected by nodes $k$, $h$.}
+\caption{\label{fig:structures} The network statistics whose corresponding effects are included in the models we fit to the friends data.}
 \end{figure}
 
-\noindent where $s_{i3}(x,z) = \sum_{j \neq h} x_{ij}x_{ih}x_{hj} \mathbb{I}(z_i = z_h \neq z_j)$, and $s_{i4}(x) =  |\{j : x_{ij} = 0, \sum\limits_h x_{ih}x_{hj} \geq 2\}|$.
+\noindent where $s_{i3}(x,z)$ and $s_{i4}(x)$ are defined in Figure~\ref{fig:structures}.
 These statistics are known as the \emph{number of jumping transitive triplets} (JTT) and the \emph{number of doubly achieved distances two effect} (N22), respectively.
@@
 %The covariate for the friendship data is the drinking behavior, and is the number of bills authored for the senate data.
-These additional effects are visualized in Figure~\ref{fig:jtt} and Figure~\ref{fig:dad}, where the ties affected by the parameter value are represented by dotted lines.
+All model effects are visualized in Figures~\ref{fig:beta1}-\ref{fig:dad}, where the ties affected by the parameter value are represented by dotted lines.
 If the estimates of these parameters $\hat{\beta}_1, \dots, \hat{\beta}_4$ are positive, the model encourages the tie represented by the dotted line to form, while the opposite is true if the parameter estimate is negative.
+Mean estimates from 1,000 repeated fittings are given in Table~\ref{tab:meanests}.
+Any time we simulate from a network model, these are the parameters we use.
+
+% table of fitted M1 M2 M3
+<<parammeans>>=
+simu2 <- read.csv("data/simulation-1000-M1-M2-M3.csv")
+SFE <- simu2 %>% group_by(Model, parameter) %>%
+  summarize(meanEst = mean(estimate), sdEst = sd(estimate)) %>%
+   mutate(val = paste0(sprintf("%.3f",meanEst), " ", "(", sprintf("%.3f", sdEst), ")")) %>%
+  select(Model, parameter, val) %>%
+  dplyr::select(Model, parameter, val) %>%
+  spread(Model, val) %>% arrange(parameter)
+@
+\begin{table}
+\centering
+\begin{tabular}{lccc}
+  \toprule
+  & M1 & M2 & M3 \\
+  \midrule
+  $\alpha_1$ & \Sexpr{SFE$M1[1]} & \Sexpr{SFE$M2[1]} & \Sexpr{SFE$M3[1]}  \\
+  $\alpha_2$ & \Sexpr{SFE$M1[2]} & \Sexpr{SFE$M2[2]} & \Sexpr{SFE$M3[2]}   \\
+  \midrule
+  $\beta_1$ & \Sexpr{SFE$M1[3]} & \Sexpr{SFE$M2[3]} & \Sexpr{SFE$M3[3]}  \\
+  $\beta_2$ & \Sexpr{SFE$M1[4]} & \Sexpr{SFE$M2[4]} & \Sexpr{SFE$M3[4]} \\
+$\beta_3$ & -- & \Sexpr{SFE$M2[5]} & --  \\
+  $\beta_4$ & -- & --  & \Sexpr{SFE$M3[6]} \\
+  \bottomrule
+\end{tabular}
+
+\caption{\label{tab:meanests} The means (std.\ dev.)\ of parameter values estimated from 1,000 repeated fittings of M1, M2, and M3 to the example data.}
+\end{table}
+
+We encourage interested readers to consult the \texttt{RSiena} manual for a complete definition and description of these and all other possible effects available to use in CTMC models \citep{RSienamanual}.
+Our goal here is to assess the fit of these models to our data, not to find the best model for this data, which is why we have kept both the data and the models very small.
+A true analysis of these data would include all of the available data, including additional covariates related to smoking and other behaviors, and would use sociological research to select and test the most pertinent effects at each step in the model hierarchy.
+The exact model structure and effects we choose are not as important to us as is our ability to assess their impact on model fit.
+We keep the data and model simple in order to apply the more complicated model visualization tools that we discuss in Sections~\ref{sec:minds}-\ref{sec:algorithms}.
 
@@
 %Every good \emph{data} analysis includes both numerical and visual summaries of the data, so why restrict \emph{model} analysis, description, and diagnostics to numerical summaries?
-\emph{Model visualization} complements traditional model diagnostic tools such as $R^2$ values and residual plots to enrich our understanding and interpretation of statistical models.
-Each of the three different meanings of the word ``model", the model \emph{family}, the model \emph{form}, and the \emph{fitted} model, can benefit from model visualization (model-vis) \citep{modelvis}.
-We apply the three principles of model-vis: view the model in the data space, visualize collections of models, and explore the process of fitting the model, to the models and data introduced in Section~\ref{sec:exdatamods}.
-Since we have already decided on our model family (CTMCs), we shift our focus to the fitted model and the model form.
-Specifically, we use model-vis to learn more about how the \emph{model form} affects the \emph{fitted model}.
 
@@
 We define the \emph{data space} for our models as the combined data structures of the network: the node, edge, and time information.
-We use the \texttt{R} package \texttt{geomnet} \citep{geomnet} to visualize the three data spaces together.
+We use the \texttt{R} package \texttt{geomnet} to visualize the three data spaces together  \citep{geomnet}.
 The package draws the network using node-link diagrams, and the network's node or edge data are then mapped to different visual features.
@@
 
-%To demonstrate the functionality of \texttt{geomnet}, we visualize the connections in the $111^{th}$ U.S.
-Senate at two different timepoints: when Hillary Clinton was in the senate, and after she left to become Secretary of State.
-Clinton was only in the $111^{th}$ senate from January 3-20, 2009, in the middle of her second term as the junior senator from New York.
-In that time, she authored two bills and was a cosponsor on 17.
-There are many senators who cosponsored both of her bills, giving their edges to her a $WPC$ of 1.
-So with Clinton included in the node-link diagram in Figure~\ref{fig:funsenatevis}, the senate looks much more collaborative than it does without her.
-We then map our potential model covariates to visual features: number of bills authored to size of the node, gender to shape, and party to color.
-In addition, we visualize the strength of the tie by mapping the $WPC$ value between the two senators to the linewidth.
-By viewing node information, edge information, and time simultaneously, we have learned that Senator Clinton greatly affected the structure of collaboration within the senate.
-
-<<summnetM1, message=FALSE, fig.cap='On the left, observed data from the first time point, which is conditioned on in the first model.
-On the right, the data observed at the second time point.
-In the middle, the expected network from the first model fit to the data.
-This expected network represents 1,000 simulations of $x(t_2)$ from model M1.',results='hide', fig.height=2.5>>=
+<<summnetM1, message=FALSE,results='hide'>>=
 friendshipData <- array(c(fd2.w1, fd2.w2, fd2.w3), dim = c(16, 16, 3))
@@
   theme(plot.background = element_rect(color = 'black'))
-grid.arrange(p2, p1, p3, nrow = 1)
-@
-
-Then to view our models in the data space, we first simulate from the fitted model.
+#grid.arrange(p2, p1, p3, nrow = 1)
+@
+
+<<averagenet, results='hide', out.width='\\textwidth', fig.cap='In the bottom left, the network at the first time point ($x(t_1)$), which is conditioned on in the model.
+In the bottom right, the network at the second time point($x(t_2)$).
+In the top wor, the three expected networks for time point 2 ($\\bar{x}(t_2)$) according to model M1, M2, M3.', fig.height=5>>=
+
+combinedavgs$linewidth[combinedavgs$cat %in% c("1st Wave", "2nd Wave")] <- .5
+combinedavgs$cat <- as.factor(combinedavgs$cat)
+combinedavgs$cat <- ordered(combinedavgs$cat, levels = c("Model1", "Model2", "Model3", "1st Wave", "2nd Wave"))
+levels(combinedavgs$cat) <- c("Model 1", "Model 2", "Model 3", "Time 1", "Time 2")
+
+ggplot(data = combinedavgs) +
+  geom_net(aes(from_id = from, to_id = to, linewidth= linewidth),
+           directed = T, curvature = .2, labelon=T, size = 3,
+           arrow = arrow(angle = 20, type = "open", length = unit(.2, "cm")), ecolour = "#969696", fiteach = T,
+           color = "#252525", hjust = .5, vjust = .5, labelcolour = 'white', fontsize = 2, arrowgap = .02,
+           singletons = T) +
+  ThemeNet + theme(panel.background = element_rect(color = 'black')) +
+  facet_wrap(~cat)
+@
+
+To view our models in the data space, we first simulate from the fitted model.
 No single network instance simulated from a CTMC will fully represent the model, so instead we look at an \textit{expectation} from the network model.
@@
 How then, can we arrive at an ``expected" network?
-We answer this question with visualization.
-%Expected values of parameters can be computed, but if they cannot directly tell us anything about our dependent variable, they lose value. 
-%There is no defined expected value of a network model, though expectations exist for most other statistical models. 
-%For network data, we view an ``average" network through a sort of summary network drawn using the traditional node-link diagram. 
-%In Figure~\ref{fig:summnetM1}, we show the \textit{expected network} of wave 2 from model M1, created with 1,000 simulations from the model. 
 To construct an expected network, we
 \begin{enumerate}
-\item condition on $x(t_1)$ and simulate 1,000 instances of $x(t_2)$ and $x(t_3)$ from the model of interest using parameter values (see Table~\ref{tab:meanests}) that have previously been estimated.
+\item condition on $x(t_1)$ and simulate 1,000 instances of $x(t_2)$ and $x(t_3)$ from the model of interest using previously estimated parameter values (see Table~\ref{tab:meanests}).
+Call these simulations $x^{(1)}(t_m), x^{(2)}(t_m),\dots x^{(1000)}(t_m)$ for $m=2,3$.
 Then, we
-\item count the number of times each edge $i \rightarrow j$ appeared in one of the 1,000 network simulations. Then, we
-\item construct one weighted summary network for time points 2 and 3. Each summary network has edge weights equal to the proportion of the simulated networks in which the edge appeared. Finally, we 
-\item create the \textit{expected network} by including only the edges that appeared in more than 5\% of the simulations. Call these expected networks $\bar{x}(t_2)$ and $\bar{x}(t_3)$.
+\item count the number of times each edge $i \rightarrow j$ appeared in one of the 1,000 network simulations.
+i.e.\ compute $x_{ij}^{(B)}(t_m) = \sum_{b=1}^{1000} x_{ij}^{(b)}(t_m)$ for $m=2,3$.
+Then, we
+\item construct one weighted summary network for time points 2 and 3.
+Each summary network has edge weights equal to the proportion of the simulated networks in which the edge appeared.i.e.\ $\bar{x}_{ij}(t_m) = x_{ij}^{(B)}(t_m) /1000$.
+Finally, we
+\item create the \textit{expected network} by including only the edges that appeared in more than 5\% of the simulations.
+i.e.\ $\bar{x}_{ij}(t_m) := \bar{x}_{ij}(t_m) \cdot \mathbb{I}(\bar{x}_{ij}(t_m) > 0.05)$.
+We write the expected networks as $\bar{x}(t_2)$ and $\bar{x}(t_3)$.
 \end{enumerate}
@@
 On either side of the average network in Figure~\ref{fig:summnetM1}, we show the data for comparison: $x(t_1)$ on the left, and $x(t_2)$ on the right.
+Visualizing the expected network in this way places the model M1 in the data space, and helps us assess model fit.
 We can see that the overall structure of the average network is much more similar to $x(t_1)$  than to $x(t_2)$ .
-In both wave 1 and the expected network, the edges form two clusters of nodes: $\{2,3,4,5,6,12,13,15\}$ and $\{7,8,9,10,11,14\}$.
-The expected network, however, is supposed to represent $x(t_2)$ , on the right in Figure~\ref{fig:summnetM1}.
-The expected network does not resemble  $x(t_2)$, which has one small, strongly connected cluster and one larger, sparsely connected cluster, with four unconnected nodes.
-Because the expected network does not resemble  $x(t_2)$ , we take this as evidence that the simple model, M1, does not capture the change in network structure from the first to the second observation of data.
-Visualizing the expected network in this way places the model M1 in the data space, and helps us assess model fit.
+In both $x(t_1)$ and $\bar{x}(t_2)$, there are two clusters of nodes: $\{2,3,4,5,6,12,13,15\}$ and $\{7,8,9,10,11,14\}$.
+The expected network, however, does not resemble  $x(t_2)$, which it is supposed to represent $x(t_2)$.
+In $x(t_2)$, there is one small, strongly connected cluster and one larger, sparsely connected cluster, with four unconnected nodes.
+Because the expected network $\bar{x(t_1)}$ more closely resembles $x(t_1)$ than $x(t_2)$, we take this as evidence that the simple model, M1, is not a good fit to the data.
 % We can use the average network to help determine model goodness-of-fit.
@@
 
-<<summary2, fig.height=2.5, fig.cap="Network differences between the expected network  $\\bar{x}(t_2)$ and $x(t_1)$  and  $x(t_2)$.
-Some edges appear in both the data and the expected network $\\bar{x}(t_2)$, while others only appear in the data or only in the expectation.
-Width of edges represent the proportion of appearances in the 1,000 simulations.
+<<summary2, fig.height=2.5, fig.cap="Differences between the expected network  $\\bar{x}(t_2)$ and $x(t_1)$  and  $x(t_2)$.
+Some edges appear in both the data and the expected network $\\bar{x}(t_2)$ (dotted lines), while others only appear in the data (solid lines) or only in the expectation (dashed lines).
+Width of non-data edges represent the proportion of appearances in the 1,000 simulations.
 The FR layout is used.">>=
@@
 Another way to view our models in the data space is with \textit{difference networks} between the expected network and the data, as shown in Figure~\ref{fig:summary2}.
-We created these networks by computing the difference between the adjacency matrices $\mathbf{A}(\bar{x}(t_2)) - \mathbf{A}(x(t_1))$ of the expected network $\bar{x}(t_2)$ and $x(t_1)$ respectively, and $\mathbf{A}(x(t_2)) - \mathbf{A}(\bar{x}(t_2))$, the difference between $\bar{x}(t_2)$ and the expected network $\bar{x}(t_2)$.
-The lack of model fit becomes more apparent in Figure~\ref{fig:summary2}, with the broken edges representing ties that only show up in the expected network, and the solid edges representing ties in the data that the model fails to capture.
-Dotted edges are in both the expectation and the data and represent where the model ``got it right".
-Again, the widths of the edges correspond to the frequency with which the ties appear in the simulations.
+We created these networks by computing the difference between $\bar{x}(t_2)$ and $x(t_1)$, $\mathbf{A}(\bar{x}(t_2)) - \mathbf{A}(x(t_1))$, and the difference between $\bar{x}(t_2)$ and $\bar{x}(t_2)$, $\mathbf{A}(x(t_2)) - \mathbf{A}(\bar{x}(t_2))$.
+The lack of model fit becomes more apparent in Figure~\ref{fig:summary2}, with the dashed edges representing ties that only show up in the expected network, and the solid edges representing ties in the data that the model fails to capture.
+Dotted edges are in both the expectation and the data and represent where the model is correct.
 On the left in Figure~\ref{fig:summary2}, we see the difference between $x(t_1)$ and the expected network.
-We only see dotted and broken edges, meaning that the simulated networks consist primarily of edges in $x(t_1)$, plus some new edges that are not in the data.
-The unobserved ties are all reciprocating edges in the data, thus the model predicts more reciprocity between actors than there actually is.
-On the right in Figure~\ref{fig:summary2}, we see the difference between the $x(t_2)$ and the expected network.
+We see mostly dotted edges, meaning that the simulated networks consist primarily of edges in $x(t_1)$, plus some new edges represented by the dashed lines that are not in the data.
+The unobserved ties are all reciprocating ties in the data, so the model is predicting more reciprocity between actors than there actually is.
+On the right in Figure~\ref{fig:summary2}, we see the difference between $x(t_2)$ and the expected network.
 Because the model is simulating  $x(t_2)$, we would hope to see very few differences between the expected network and the data.
-However, there are six solid edges which are in the data observed at the second time point that do not appear at all in the expected network, and there are many broken edges that appear in the expected network but are not in $x(t_2)$.
+However, there are six solid edges, which are in the data but do not appear at all in the expected network, and there are many dashed edges that appear in the expected network but are not in $x(t_2)$.
 Although there are some edges that appear in both the expected network and $x(t_2)$, the majority of edges in the expected network are not in the observed data, and there are several edges in the data missing from the expectation.
-We conclude that the model M1 is a poor fit to our example data, which we have seen very clearly by viewing the model in the data space.
+We therefore conclude that the model M1 is a poor fit to our example data, which we have seen very clearly by viewing the model in the data space.
 
@@
 Now that we have assessed model fit, we want to learn about model behavior in order to choose better fitting models in the future.
-To assess model behavior, we employ the second principle of model visualization: ``collections [of models] are more informative" \citep{modelvis}.
+To assess model behavior, we employ the second principle of model visualization: ``collections are more informative" \citep{modelvis}.
 We rely on three methods for constructing collections of CTMC models.
-%We examine the behavior of the CTMCs by visualizing collections of them, which we construct in four ways.
 First, we explore the space of all possible models to determine significant model parameters.
@@
 Finally,  we fit the model many times to assess the behavior of the underlying simulations.
-%Finally, we fit the same model to different data to learn about fitted parameter values under different conditions.
-We chose these three collections because they each explore something different about model behavior: the choice of and relationships between parameters in the objective functio and the stochastic effects from simulation.%, and the resulting fitted values under different circumstances.
+We chose these three collections because they each explore something different about model behavior: the choice of and relationships between parameters in the objective function and the stochastic effects from simulation.
+%, and the resulting fitted values under different circumstances.
 %Using the tools in our visualization toolbox, we view these collections to learn about model behavior.%The first considers the many parameters to include in a SAOM, while the second results in different parameter estimates, so we can see how parameter estimates vary.
@@
 
-%\subsection{Exploring the space of all possible models}\label{sec:allpossible}
-
-\noindent \textbf{Exploring the space of all possible models:} The \texttt{RSiena} manual contains over 80 effects to include in the objective function.
-To select objective function parameters for modeling our example data, we searched through all effects available to find  significant effects.
-The \texttt{RSiena} package includes the function \texttt{Wald.Rsiena} for Wald-type tests of parameter significance.
+\noindent \textbf{Exploring the space of all possible models:} The \texttt{RSiena} manual contains over 80 effects to include in the objective function \citep{RSienamanual}.
+To select objective function parameters for modeling our example data, we searched through all available parameters to find  significant effects.
+The \texttt{RSiena} package includes the function \texttt{Wald.Rsiena} for Wald-type tests of parameter significance \citep{RSiena}.
 We start by including the outdegree and reciprocity parameters in the model then add one effect to the objective function, fit the model to the example data, test the added effect for significance, and repeat for all possible model parameters.
@@
 Using this one-at-a-time testing method results in many possible significant parameters of different type and magnitude, demonstrating the flexibility of these models.
-We chose models M1, M2, and M3 to examine more closely because they had the smallest $p$-values from the Wald-type tests.
+We chose models M2 and M3 in addition to model M1 to examine more closely because their added effects had the smallest $p$-values from the Wald-type tests.
 %We see in both the friendship data and the senate data results that most of the significant effects have absolute value less than ten.
@@
 We have done this already by choosing three different models M1, M2, M3 to fit to our example data sets.
-To explore the behavior of these three models, we simulate from them.
+% took out this next bit just because it doesn't really add much and we can just put the appendix figure into the previous section. % add more later.
+
+%To explore the behavior of these three models, we simulate from them.
 Using the estimated parameter values in Table~\ref{tab:meanests}, we simulated 1,000 observations of $x(t_2)$ and $x(t_3)$ of the friendship data from each of the three models.
@@
 To examine this variability, we fit models M1, M2, and M3 to the friendship network 1,000 times each.
-The distributions of the fitted model parameters are shown in Figure~\ref{fig:distests}.
-Clearly, the inclusion of the JTT parameter, $\beta_3$, results in very different estimates the of the other four parameters in all models, $\alpha_1$, $\alpha_2$, $\beta_1$, and $\beta_2$.
+The resulting mean estimates are given in Table~\ref{tab:meanests}, and the distributions of all parameter estimates are shown in Figure~\ref{fig:distests}.
+Clearly, the inclusion of the JTT parameter, $\beta_3$, results in very different estimates the of the other four parameterss, $\alpha_1$, $\alpha_2$, $\beta_1$, and $\beta_2$.
 On the other hand, the inclusion of the indirect tie parameter, $\beta_4$, does not lead to such drastically different estimates of other model parameters.
@@
 When visualizing the results of the same models fit many times to the same data, we can to see potential model behavior problems right away.
+% maybe talk about dependence of statistics, inherent correlation, etc.
+
 %In comparison with models M1 and M3, model M2 typically has higher estimates of the rate parameter, meaning that the inclusion of the covariate statistic in the model leads to higher estimates of the number of times, on average, a node gets to change its ties.
@@
 
-<<tableofmeans>>=
+<<tableofmeans, eval = FALSE>>=
 senM1est <- read.csv("data/congress/senateM1ests.csv")
@@
 % \end{table}
-
-% table of fitted M1 M2 M3
-\begin{table}
-\centering
-\begin{tabular}{lccc}
-  \toprule
-  & M1 & M2 & M3 \\
-  \midrule
-  $\alpha_1$ & \Sexpr{SFE$M1[1]} & \Sexpr{SFE$M2[1]} & \Sexpr{SFE$M3[1]}  \\
-  $\alpha_2$ & \Sexpr{SFE$M1[2]} & \Sexpr{SFE$M2[2]} & \Sexpr{SFE$M3[2]}   \\
-  \midrule
-  $\beta_1$ & \Sexpr{SFE$M1[3]} & \Sexpr{SFE$M2[3]} & \Sexpr{SFE$M3[3]}  \\
-  $\beta_2$ & \Sexpr{SFE$M1[4]} & \Sexpr{SFE$M2[4]} & \Sexpr{SFE$M3[4]} \\
-$\beta_3$ & -- & \Sexpr{SFE$M2[5]} & --  \\
-  $\beta_4$ & -- & --  & \Sexpr{SFE$M3[6]} \\
-  \bottomrule
-\end{tabular}
-
-\caption{\label{tab:meanests} The means (std.\ dev.)\ of parameter values estimated from repeated fittings of $M1, M2, M3$ to the example data.}
-\end{table}
 
@@
 When fitting models in \texttt{RSiena}, several simulation processes are hidden from the user.
-We refer to one of these hidden pieces as in the SIENA MoM algorithm is the \textit{microstep process}, which is a series of state changes in the CTMC as described in Section~\ref{sec:saoms}.
-At each microstep, an \textit{ego} node is selected to change, and the chosen node randomly makes one change in its ties according to to the probabilities $p_{ij}$ defined in Equation~\ref{eq:pij}.
+We refer to one of these hidden pieces in the SIENA MoM algorithm as the \textit{microstep process}, which is a series of state changes in the CTMC as described in Section~\ref{sec:saoms}.
+At each microstep, an ego node is selected to change, and then randomly makes one change in its ties according to to the probabilities $p_{ij}$ defined in Equation~\ref{eq:pij}.
 This process continues until the network differs from the $x(t_m)$ by exactly $C_m$, as defined in Equation~\ref{eq:cdiff}.
-Between two network observations $x(t_m)$ and $x(t_{m+1})$, there can be dozens, hundreds, or even thousands of microsteps, depending on the size of the network and the number of tie changes between two network observations.
-We want to simulate and view these in-between steps in order to better understand the behavior of the underlying CTMC.
+Between two network observations $x(t_m)$ and $x(t_{m+1})$, there can be dozens, hundreds, or even thousands of microsteps, depending on the size of the network and the value of $C_m$.
+We want to simulate and view these in-between steps in order to better understand the behavior of the  CTMC model.
 Hiding the microsteps is practical and efficient if the goal is to fit a model to data, obtain parameter estimates, and draw conclusions or make predictions.
-But in order to understand \textit{how} the models are fit, we need to extract and explore the different steps of the process.
+But in order to understand how the models are fit, we need to extract and explore the different steps of the process.
 We use these steps to visually explore the algorithms, so that we are visualizing the model fitting process, and not only the model fitting result.
@@
 
-<<makinggif, fig.cap='A selection of images in the microstep animation.
-The selected edges and nodes are emphasized by changing the size and the color, then they fade in or out, while the nodes change color and shrink to blend in with the other nodes.
-Frames 15-28 show a an edge disappearing, while frames 71-84 show an edge appearing.
+<<makinggif, fig.cap='A selection of images in the microstep animation at \\url{https://vimeo.com/240089108}.
+The selected edges and nodes are emphasized by changing size and color, then they fade in or out while the nodes change color and shrink to blend in with the other nodes.
+Frames 15-28 show edge destruction, while frames 71-84 show edge creation.
 A spring layout is used.', message = FALSE, warning = FALSE, fig.height=2.5>>=
@@
 Movies similar to this animation were used to visualize the changes of dynamic networks in \citet{dynnetviz}.
-Frames 15-28 of Figure~\ref{fig:makinggif} show a deleted tie, while frames 71-84 show an added tie.
+Frames 15-28 of Figure~\ref{fig:makinggif} show tie that is destroyed, while frames 71-84 show a tie that is created.
 For both changes, the ego node is emphasized with color and size.
@@
 In this network animation, we show the modeled steps of the unobserved CTMC process between two observed networks.
-We see each part of the hierarchical come into play: first, the selection of the ego node via the rate function, then the result of the actor maximizing its objective function by either deleting or adding a node.
+We see each part of the hierarchical model: first, the selection of the ego node via the rate function, then the result of the actor maximizing its objective function by either deleting or adding a node.
 Finally, the layout of the network transitions as edges are removed or added, demonstrating how the overall network structure changes with each of the individual tie changes.
-Combining the tie changes in this animation makes the CTMC process the focu, demonstrating the underlying change mechanism of the dynamic network according to the model.
+Combining the tie changes in this animation makes the CTMC process the focus, demonstrating the underlying change mechanism of the dynamic network according to the model.
+% need more about what we learned.
 
@@
   scale_fill_manual(values = c('white', 'black')) +
-  scale_x_discrete(labels = 1:16, position = "top") +
-  scale_y_discrete(labels =16:1, name = "from")  +
+  scale_x_discrete(labels = 1:16, position = "top", name = "alter (j)") +
+  scale_y_discrete(labels =16:1, name = "ego (i)")  +
   ThemeNoNet +
@@
         axis.text.y = element_text(margin = margin(r=0))) +
-  coord_fixed() + 
-  labs(x= "alter (j)", y = "ego (i)")
+  coord_fixed()
 
@@
   scale_fill_manual(values = c('white', 'black'), name = "Tie is...", labels = c("not present","present")) +
-  scale_x_discrete(position = "top") +
+  scale_x_discrete(position = "top", name = "alter (j)") +
   ThemeNoNet +
@@
   scale_fill_manual(values = c('white', 'black')) + 
-  scale_x_discrete(position = "top") +
-  scale_y_discrete(name = "from",labels = rev(str_replace(order1, "V", ""))) +
+  scale_x_discrete(position = "top", name = "alter (j)") +
+  scale_y_discrete(name = "ego (i)",labels = rev(str_replace(order1, "V", ""))) +
   ThemeNoNet + theme(legend.position = "none") + 
-  coord_fixed() + 
-  labs(x= "alter (j)", y = "ego (i)")
+  coord_fixed()
 legendp2 <- cowplot::get_legend(p2)
@@
 To seriate the matrix, we first construct a cumulative adjaceny matrix, $\mathbf{A}^{cum}$, for the series of microsteps in the CTMC simulated from $x(t_m)$ to $x(t_{m+1})$.
-A single entry in the cumulative adjacency matrix, $\mathbf{A}^{cum}_{ij}$, is the total number of times the edge $i \rightarrow j$ appears in the network from the initial observation, $x(t_m) \equiv X(0)$ to the last microstep, $X(R)$, where $R$ is the total number of microsteps taken:
+A single entry in the cumulative adjacency matrix, $\mathbf{A}^{cum}_{ij}$, is the total number of times the edge $x_{ij}$ appears in the network from the initial observation, $x(t_m) \equiv X(0)$ to the last microstep, $X(R)$, where $R$ is the total number of microsteps taken:
 \begin{equation}\label{eq:acum}
@@
 We then perform a principal component analysis (PCA) on $\mathbf{A}^{cum}$, and use the values of the first principal component to order the vertices on the $x$ and $y$ axes for the adjacency matrix net-vis.
-The resulting visualization for the first network observation is shown in Figure~\ref{fig:adjmatorder}.
+The resulting visualization for the first network observation is shown on the right in Figure~\ref{fig:adjmatorder}.
 %we present the first wave of data ordered by the arbitrary node ID order alongside the seriated adjacency matrix ordered by the first principal component loading on the cumulative adjacency matrix, $\mathbf{A}^{cum}$.
-Using PCA on $\mathbf{A}^{cum}$ to order the rows and columns of the adjacency matrix visualization clearly shows the two distinct connected components in the first network, which are difficult to find in the visualization ordered by node ID.
+Using matrix seriation on the %PCA on $\mathbf{A}^{cum}$ to order the rows and columns of the 
+adjacency matrix visualization clearly shows the two distinct connected components in the first network, which are difficult to find in the visualization ordered by node ID.
 We continue to use the seriated matrix in another animation of the microstep process, which is available at \url{https://vimeo.com/240092677}.
-A few frames of this video are shown in Figure~\ref{fig:adjmatani}.
+Some frames of this video are shown in Figure~\ref{fig:adjmatani}.
 In the animation, a square appears or disappears as that edge appears or disappears in the microstep process.
 %Through this animation, we can see edges appearing, and then later on disappearing.
-This visualization also shows the microsteps where the nodes do not make a change.
+This vanimation, unlike the node-link visualization, shows the microsteps where the chosen node does not make a change.
 From start to finish, we see the network evolve: sometimes ties are removed right after they are created, and sometimes several ties are removed or created in a row.
@@
 By viewing this animation, we gain a better understanding of the underlying dynamics of this model.
+% need more oomph
 
@@
 
-We have used several visualizations, some old and some new, to explore a set of continuous time markov chain models for dynamic social network data in new ways.
+We have used several visualizations, some old and some new, to explore social network models new ways.
 The guiding principles of model-vis: viewing the model in the data space, visualizing collections of models, and looking at the underlying algorithms, gave us knowledge and appreciation for these complicated and flexible models.

### 2019-07-08 supporting-info-child.Rnw
stuff from last week

@@
 
-<<sigtesting, fig.show = 'hold', out.width='\\textwidth', fig.height=2, fig.cap="Significant effects for the two data sets, at a significance level of 0.10 or lower as calculated by the Wald-type test available in the SIENA software.", error=FALSE>>=
+<<sigtesting, fig.show = 'hold', out.width='\\textwidth', fig.height=2, fig.cap="Significant effects for the two data sets, at a significance level of 0.10 or lower as calculated by the Wald-type test available in the SIENA software.", error=FALSE,eval = FALSE>>=
 smfr_effs <- read_csv("data/effects_significance_smallFriends.csv")
@@
 
-In Figure~\ref{fig:sigtesting}, we see for the friendship data and the senate data that most of the significant effects have absolute value less than ten.
+%In Figure~\ref{fig:sigtesting}, we see for the friendship data and the senate data that most of the significant effects have absolute value less than ten.
 In addition, the $p$-values for the effects from the friendship data are more spread out than the $p$-values for the senate data, which are concentrated at about 0.02 or less.
@@
 simu_spread <- simu2 %>% spread(parameter, estimate)
-ggpairs(simu_spread, columns = 4:9, aes(colour=Model, alpha = .5)) + 
+ggpairs(simu_spread, columns = 3:8, aes(colour=Model, alpha = .5)) + 
   theme(plot.title = element_text(size = 10,

### 2019-07-08 paper-child.Rnw
all plots from SI moved to body. body includes more discussion of applicable methods to other net. models. Still need to wrap up sec. 4, go through sec 5, and majorly add to Sec. 6.

@@
 To define CTMCs for our purposes, we follow the general hierarchical definition outlined in \citet{saomreview} and elsewhere (see e.g.\ \citet{snijders01} and \citet{snijdersetal2010}).
-This hierarchy reflects our intution about social networks: most social networks, even holding constant the set of actors over time, are ever-changing as relationships form or disappear.
+CTMC models reflect our intution about social networks: most social networks, even holding constant the set of actors over time, are ever-changing as relationships form or dissolve.
 Actors in social networks also have properties that could change their role within the network, and existing ties may affect formation or dissolution of other ties.
@@
 
-\noindent \textbf{Rate Function:} In general, the \emph{rate function}, $\rho(x, \mathbf{z}, \boldsymbol{\alpha})$, is a function of the network at its current state $x$, covariates of interest, $\mathbf{z}$, and some parameters, $\boldsymbol{\alpha}$.
+\noindent \textbf{Rate Function:} In general, the \emph{rate function}, $\rho(x, \mathbf{Z}, \boldsymbol{\alpha})$, is a function of the network at its current state $x$, covariates of interest, $\mathbf{Z}$, and some parameters, $\boldsymbol{\alpha}$.
 With this general formulation, each node can have a different rate of change.
-Throughout this paper, we assume a simple rate function, $\rho(x, \mathbf{z}, \boldsymbol{\alpha}) \equiv \alpha_m$ that is constant across all actors between time $t_m$ and $t_{m+1}$.
+Throughout this paper, we assume a simple rate function, $\rho(x, \mathbf{Z}, \boldsymbol{\alpha}) \equiv \alpha_m$ that is constant across all actors between time $t_m$ and $t_{m+1}$.
 With this simple rate function, we have $M-1$ rate parameters, one for each transition from $x(t_{m})$ to $x(t_{m+1})$ for $m = 1, \dots, M-1$.
@@
 A true analysis of these data would include all of the available data, including additional covariates related to smoking and other behaviors, and would use sociological research to select and test the most pertinent effects at each step in the model hierarchy.
-The exact model structure and effects we choose are not as important to us as is our ability to assess their impact on model fit.
+The exact model structure and effects we choose are not as important to us as our ability to assess their impact on model fit.
 We keep the data and model simple in order to apply the more complicated model visualization tools that we discuss in Sections~\ref{sec:minds}-\ref{sec:algorithms}.
@@
 In the bottom right, the network at the second time point($x(t_2)$).
-In the top wor, the three expected networks for time point 2 ($\\bar{x}(t_2)$) according to model M1, M2, M3.', fig.height=5>>=
+In the top row, the three expected networks for time point 2 ($\\bar{x}(t_2)$) according to model M1, M2, M3.
+Legend shows edge weights in the three expected networks.', fig.height=5>>=
 
@@
 combinedavgs$cat <- ordered(combinedavgs$cat, levels = c("Model1", "Model2", "Model3", "1st Wave", "2nd Wave"))
-levels(combinedavgs$cat) <- c("Model 1", "Model 2", "Model 3", "Time 1", "Time 2")
-
-ggplot(data = combinedavgs) +
+levels(combinedavgs$cat) <- c("M1 expectation", "M2 expectation", "M3 expectation", "Time 1", "Time 2")
+
+summary(combinedavgs$linewidth)
+
+combinedavgs %>% filter(!is.na(to)) %>% 
+  ggplot() +
+  geom_point(aes(x = from, y = to)) + 
+  geom_line(aes(x = from, y = to, size = linewidth), color = "#969696") +
+  scale_size_continuous(name = "Edge frequency", range = c(0,1)) + 
+  ThemeNet -> forlineleg  
+legendlines <- cowplot::get_legend(forlineleg)
+set.seed(35291)
+allp <- ggplot(data = combinedavgs) +
   geom_net(aes(from_id = from, to_id = to, linewidth= linewidth),
@@
   facet_wrap(~cat)
+
+ggdraw() + draw_plot(allp, width = 1) + 
+  draw_grob(legendlines,x=.7, y=.1, width = .2, height = .4)
 @
@@
 We could talk about expected values of parameters, but the parameters can be hard to interpret and do not illuminate the overall structure of the network.
-How then, can we arrive at an ``expected" network?
-To construct an expected network, we
+How then, can we arrive at an expected network?
+We propose the following method:
 \begin{enumerate}
-\item condition on $x(t_1)$ and simulate 1,000 instances of $x(t_2)$ and $x(t_3)$ from the model of interest using previously estimated parameter values (see Table~\ref{tab:meanests}).
+\item Condition on $x(t_1)$ and simulate 1,000 instances of $x(t_2)$ and $x(t_3)$ from the model of interest using previously estimated parameter values (see Table~\ref{tab:meanests}).
 Call these simulations $x^{(1)}(t_m), x^{(2)}(t_m),\dots x^{(1000)}(t_m)$ for $m=2,3$.
-Then, we
-\item count the number of times each edge $i \rightarrow j$ appeared in one of the 1,000 network simulations.
-i.e.\ compute $x_{ij}^{(B)}(t_m) = \sum_{b=1}^{1000} x_{ij}^{(b)}(t_m)$ for $m=2,3$.
-Then, we
-\item construct one weighted summary network for time points 2 and 3.
+\item Count the number of times each edge $i \rightarrow j$ appeared in one of the 1,000 network simulations. i.e.\ compute $x_{ij}^{(B)}(t_m) = \sum_{b=1}^{1000} x_{ij}^{(b)}(t_m)$ for $m=2,3$.
+\item Construct one weighted summary network for time points 2 and 3.
 Each summary network has edge weights equal to the proportion of the simulated networks in which the edge appeared.i.e.\ $\bar{x}_{ij}(t_m) = x_{ij}^{(B)}(t_m) /1000$.
-Finally, we
-\item create the \textit{expected network} by including only the edges that appeared in more than 5\% of the simulations.
+\item Create the \textit{expected network} by including only the edges that appeared in more than 5\% of the simulations.
 i.e.\ $\bar{x}_{ij}(t_m) := \bar{x}_{ij}(t_m) \cdot \mathbb{I}(\bar{x}_{ij}(t_m) > 0.05)$.
@@
 
-\noindent The expected network, $\bar{x}(t_2)$, is in the middle panel of Figure~\ref{fig:summnetM1}.
-Edges that appear with higher frequency are drawn with darker, thicker lines.
-On either side of the average network in Figure~\ref{fig:summnetM1}, we show the data for comparison: $x(t_1)$ on the left, and $x(t_2)$ on the right.
-Visualizing the expected network in this way places the model M1 in the data space, and helps us assess model fit.
-We can see that the overall structure of the average network is much more similar to $x(t_1)$  than to $x(t_2)$ .
+\noindent The above method can be generalized and applied to any network model, static or dynamic, from which we can simulate, for any number of simulations from the network.
+The expected networks from our three models of interest are shown in the top row of Figure~\ref{fig:averagenet}.
+Edges that appear with higher frequency are drawn with thicker lines.
+On the bottom row of Figure~\ref{fig:averagenet}, we show the data for comparison: $x(t_1)$ on the left, and $x(t_2)$ on the right.
+Visualizing the expected network in this way places models M1, M2, and M3, in the data space, and helps assess model fit.
+For model M1 we can see that the overall structure of the average network is much more similar to $x(t_1)$  than to $x(t_2)$ .
 In both $x(t_1)$ and $\bar{x}(t_2)$, there are two clusters of nodes: $\{2,3,4,5,6,12,13,15\}$ and $\{7,8,9,10,11,14\}$.
-The expected network, however, does not resemble  $x(t_2)$, which it is supposed to represent $x(t_2)$.
+The expected network, however, does not resemble  $x(t_2)$, which it is supposed to represent.
 In $x(t_2)$, there is one small, strongly connected cluster and one larger, sparsely connected cluster, with four unconnected nodes.
 Because the expected network $\bar{x(t_1)}$ more closely resembles $x(t_1)$ than $x(t_2)$, we take this as evidence that the simple model, M1, is not a good fit to the data.
+Comparing the three expected networks to $x(t_1)$ and $x(t_2)$, we see that they all have nodes that are connected in the same groups as $x(t_1)$.
+Model M2, which included the transitive triplet parameter, results in more strongly connected components, as shown by the increased number of edges and the thicker lines between nodes.
+The lack of fit in all models is clear:  not one of the three model expectations show node 16 gaining ties as it does in $x(t_2)$, nor do they show nodes 4 and 7 becoming isolated.
+In model M2, however, the ties to node 7 appear much weaker than in model M1 or model M3, and the nodes $\{10,11,14\}$ are also more strongly connected, as they are in wave 2 of data.
+This suggests that, despite the fact that neither of the three models is a good fit, M2 may be the best fitting model of the three.
+In a complete data analysis, we can use this information to select more models for analysis that utilize transitivity and covariate effects, as opposed to structural and indirect effects.
 % We can use the average network to help determine model goodness-of-fit.
@@
   geom_net(aes(from_id = from, to_id = to, linewidth= linewidth, linetype = color2),
-           directed = T, curvature = .2, labelon=T, size = 5,
-           arrow = arrow(angle = 20, type = "open", length = unit(.2, "cm")), 
-           color = "#252525", hjust = .5, vjust = .5, labelcolour = 'white', fontsize = 4, arrowgap = .02, layout.alg = "fruchtermanreingold",
+           directed = T, curvature = .2, labelon=T, size = 3,
+           arrow = arrow(angle = 20, type = "open", length = unit(.15, "cm")), 
+           color = "#252525", hjust = .5, vjust = .5, labelcolour = 'white', fontsize = 2, arrowgap = .02, layout.alg = "fruchtermanreingold",
            singletons = T) +
@@
 Another way to view our models in the data space is with \textit{difference networks} between the expected network and the data, as shown in Figure~\ref{fig:summary2}.
-We created these networks by computing the difference between $\bar{x}(t_2)$ and $x(t_1)$, $\mathbf{A}(\bar{x}(t_2)) - \mathbf{A}(x(t_1))$, and the difference between $\bar{x}(t_2)$ and $\bar{x}(t_2)$, $\mathbf{A}(x(t_2)) - \mathbf{A}(\bar{x}(t_2))$.
-The lack of model fit becomes more apparent in Figure~\ref{fig:summary2}, with the dashed edges representing ties that only show up in the expected network, and the solid edges representing ties in the data that the model fails to capture.
+We created these networks by computing the difference between $\bar{x}(t_2)$ and $x(t_1)$, $\mathbf{A}(\bar{x}(t_2)) - \mathbf{A}(x(t_1))$, and the difference between $\bar{x}(t_2)$ and $\bar{x}(t_2)$, $\mathbf{A}(x(t_2)) - \mathbf{A}(\bar{x}(t_2))$ for model M1.
+The lack of model fit is also apparent in Figure~\ref{fig:summary2}, where the dashed edges represent ties that only show up in the expected network, and the solid edges representing ties in the data that the model fails to capture.
 Dotted edges are in both the expectation and the data and represent where the model is correct.
@@
 We therefore conclude that the model M1 is a poor fit to our example data, which we have seen very clearly by viewing the model in the data space.
+As with the expected network, we can apply the difference network computation other network models to determine how closely the model fits the data.
 
@@
 
+<<getergmeffects, eval = FALSE>>=
+url <-"https://cran.r-project.org/web/packages/ergm/vignettes/ergm-term-crossRef.html"
+library(rvest)
+(read_html(url) %>% html_nodes("table"))[[2]] %>% html_table() %>% select(`Term name`) %>% distinct() %>% count()
+
+@
+
 \noindent \textbf{Exploring the space of all possible models:} The \texttt{RSiena} manual contains over 80 effects to include in the objective function \citep{RSienamanual}.
@@
 We chose models M2 and M3 in addition to model M1 to examine more closely because their added effects had the smallest $p$-values from the Wald-type tests.
+This particular type of model exploration is applicable to any network model with an \texttt{R} package or other software that allows the user to identify model parameters, such as the \texttt{ergm} package, which has well over 100 effects for fitting exponential-family models to networks \citep{ergm1, ergm2}.
 %We see in both the friendship data and the senate data results that most of the significant effects have absolute value less than ten.
@@
 We have done this already by choosing three different models M1, M2, M3 to fit to our example data sets.
+CTMC models, however, are much more flexible than the three simple models we have chosen.
+Other models could have included different rate functions for each node, where for example the rate of change could depend on drinking behavior if we have reason to believe that girls who drink more change friends more frequently.
+We could also model the behavior changes, not just the network changes, if we believe that friends influence behavior, via peer pressure.
+And as before, this method of gathering a collection of models can be used for any network model, such as the hierarchichal Bayesian network estimation model in the \texttt{R} package \texttt{sna}, which allows the user to vary prior and model parameters \citep{sna}.
 % took out this next bit just because it doesn't really add much and we can just put the appendix figure into the previous section. % add more later.
@@
 %\subsection{Fitting the same model to the same data}\label{sec:samemodsamedat}
-
-<<distests, fig.cap="Distribution of fitted parameter values for our three models.
-The inclusion of $\\beta_3$ or $\\beta_4$ clearly has an effect on the distributions of the rate parameters, $\\alpha_1$ and $\\alpha_2$, and on the other objective function parameters $\\beta_1, \\beta_2$.", fig.height=3>>=
-
+\begin{figure}
+\begin{subfigure}[t]{.4\textwidth}
+<<distests, fig.height=7.5>>=
 levels(simu2$parameter) <- c("alpha[1]","alpha[2]","beta[1]","beta[2]","beta[3]","beta[4]")
-ggplot(data = simu2) +
-  geom_density(aes(x = estimate, fill = Model), alpha = .5) +
-  facet_wrap(~parameter, scales = 'free', labeller= label_parsed) +
+filter(simu2, parameter %in% c("beta[3]","beta[4]")) %>% 
+ggplot() +
+  geom_density(aes(x = estimate)) +
+  facet_wrap(~parameter, scales = 'free', labeller= label_parsed, nrow = 2) +
   ThemeNoNet +
@@
 @
+\caption{\label{fig:beta34} Distribution of fitted $\beta_3, \beta_4$ values.}
+\end{subfigure}
+\begin{subfigure}[t]{.6\textwidth}
+<<img, out.width = "\\linewidth">>=
+# correlation for other params. 
+simsM2 <- filter(simu2, Model == "M2") %>% spread(parameter, estimate)
+corm2 <- cor(simsM2[, 3:7])
+simsM3 <- filter(simu2, Model == "M3") %>% spread(parameter, estimate)
+corM3 <- cor(simsM3[, 3:7])
+
+knitr::include_graphics("img/gray_corplot-1.png")
+@
+\caption{\label{fig:corplots} Relationships between parameters common to all three models. \nocite{ggally}}
+\end{subfigure}
+\caption{\label{fig:distests} Distribution of fitted parameter values for our three models.
+The inclusion of $\beta_3$ or $\beta_4$ clearly has an effect on the distributions of the rate parameters, $\alpha_1$ and $\alpha_2$, and on the other objective function parameters $\beta_1, \beta_2$.}
+\end{figure}
 
@@
 To examine this variability, we fit models M1, M2, and M3 to the friendship network 1,000 times each.
-The resulting mean estimates are given in Table~\ref{tab:meanests}, and the distributions of all parameter estimates are shown in Figure~\ref{fig:distests}.
-Clearly, the inclusion of the JTT parameter, $\beta_3$, results in very different estimates the of the other four parameterss, $\alpha_1$, $\alpha_2$, $\beta_1$, and $\beta_2$.
+The resulting means and standard deviations of the estimates are given in Table~\ref{tab:meanests}, and the distributions of all parameter estimates are shown in Figure~\ref{fig:distests}.
+In Figure~\ref{fig:corplots}, we can see that the addition of $\beta_3$ to the model results in very different estimates the of the other four parameters, $\alpha_1$, $\alpha_2$, $\beta_1$, and $\beta_2$.
+In fact, the distributions of $\alpha_1$ and $\beta_1$ for M2 do not overlap at all with the distribution of the same parameters from model M1.
 On the other hand, the inclusion of the indirect tie parameter, $\beta_4$, does not lead to such drastically different estimates of other model parameters.
@@
 It is not clear, however, that the addition of a parameter to the objective function should effect the estimates of the rate parameters.
-This strong and possibly problematic relationship between rate parameter estimates and objective function parameters is further explored in the Supporting Information.
+%This strong and possibly problematic relationship between rate parameter estimates and objective function parameters is further explored in the Supporting Information.
 When visualizing the results of the same models fit many times to the same data, we can to see potential model behavior problems right away.
+In Figure~\ref{fig:corplots}, we further examine the relationship between each of pair of parameters common to all three models.
+The strongest correlation within each model is between $\beta_1$ and $\beta_2$, with absolute correlation greater than 0.90 in all three models.
+The $\beta_1$ parameter is also highly correlated with the $\beta_3$ parameter within model M2 ($\rho = $\Sexpr{round(corm2[5, 3])}), but it is not as highly correlated with the $\beta_4$ parameter in model M3 ($\rho = $\Sexpr{round(corM3[5, 3])}).
+The inherent relationship between effects is a known problem in network analysis \citep{goldenberg09}.
+% need a better reference here.
+
+%It might therefore be advisable to consider only models that either allow $\beta_1$ or $\beta_2$ but not both.
+Looking at the high correlation with $\alpha$, we might switch to a model without $\beta_1$.
+
 % maybe talk about dependence of statistics, inherent correlation, etc.
@@
 
-The third and final model visualization principle, ``explore the process of model fitting," is crucial to understanding our models \citep{modelvis}.
+The third and final model visualization principle, ``explore the process of model fitting," is crucial to understanding the CTMC models \citep{modelvis}.
 When fitting models in \texttt{RSiena}, several simulation processes are hidden from the user.
@@
 Additional, non-packaged code can be found at \url{http://bit.ly/ctmc-model-viz}.
+
+\nocite{ggplot2}

### 2019-07-08 supporting-info-child.Rnw
all plots from SI moved to body. body includes more discussion of applicable methods to other net. models. Still need to wrap up sec. 4, go through sec 5, and majorly add to Sec. 6.

@@
 
-<<averagenet, results='hide', out.width='\\textwidth',fig.cap='The node-link diagrams from the three expected networks that we calculated are in the top row, and the true wave 1 and wave 2 data are shown in the bottom row above.
-There is some difference between the three models, but overall, these three models struggle to capture the structure in the true second wave of data.', fig.height=5>>=
-
-combinedavgs$linewidth[combinedavgs$cat %in% c("1st Wave", "2nd Wave")] <- .5
-combinedavgs$cat <- as.factor(combinedavgs$cat)
-combinedavgs$cat <- ordered(combinedavgs$cat, levels = c("Model1", "Model2", "Model3", "1st Wave", "2nd Wave"))
-
-ggplot(data = combinedavgs) +
-  geom_net(aes(from_id = from, to_id = to, linewidth= linewidth),
-           directed = T, curvature = .2, labelon=T, size = 3,
-           arrow = arrow(angle = 20, type = "open", length = unit(.2, "cm")), ecolour = "#969696", fiteach = T,
-           color = "#252525", hjust = .5, vjust = .5, labelcolour = 'white', fontsize = 2, arrowgap = .02,
-           singletons = T) +
-  ThemeNet + theme(panel.background = element_rect(color = 'black')) +
-  facet_wrap(~cat)
-@
-
-To create the expected network visualization shown in Figure~\ref{fig:averagenet}, we follow the same procedure as in Section 3: Model in Data Space, counting occurrences of each realized edge in the simulations.
-The resulting summary network has weighted edges representing the proportion of edge appearances in all 1,000 simulations of wave 2.
-As in Figure 6, Section 3, edges only appear in the expected networks if they appear more than 5\% of the time in the simulations.
-In Figure~\ref{fig:averagenet}, we show the  expected network from the three models we fit and the first and second waves of data.
-Comparing the three averages to waves 1 and 2, we see that they have very similar structure to wave 1 but that model M2, which included the transitive triplet parameter, results in more strongly connected components.
-The lack of fit is clear: none of the three average networks show node 16 gaining ties as it does in wave two, nor do they show nodes 4 and 7 becoming isolated.
-In model M2, however, the ties to node 7 appear much weaker than in model M1 or model M3, and the nodes $\{10,11,14\}$ are also strongly connected as they are in wave 2 of data, suggesting that M2 may be the best model of the three.
-
 %Outdegree and reciprocity have the same inverse relationship for both data sets.
@@
 
+levels(simu2$parameter) <- c("alpha[1]","alpha[2]","beta[1]","beta[2]","beta[3]","beta[4]")
+ggplot(data = (simu2 %>% filter(parameter %in% c("beta[3]","beta[4]")))) +
+  geom_density(aes(x = estimate), alpha = .5) +
+  facet_wrap(~parameter, scales = 'free', labeller= label_parsed) +
+  ThemeNoNet +
+  xlab("estimate value") + 
+  theme(legend.position = 'bottom')
 simu_spread <- simu2 %>% spread(parameter, estimate)
-ggpairs(simu_spread, columns = 3:8, aes(colour=Model, alpha = .5)) + 
+
+simu_spreadM12 <- simu_spread %>% filter(Model == "M1" | Model == "M2") %>% select(-beta4)
+simu_spreadM13 <- simu_spread %>% filter(Model == "M1" | Model == "M3") %>% select(-beta3)
+
+theme_set(theme_classic())
+ggpairs(simu_spread, columns = 3:6, aes(colour=Model, fill=Model, alpha = .5)) 
+        lower = list(continuous = wrap(my_color)),
+        diag = list(continuous = wrap(my_fill))) + 
+  theme(plot.title = element_text(size = 10,
+                                            face = 'plain', angle = 0, family = "Times New Roman"),
+                  axis.title.x = element_text(size = 10,
+                                            face = 'plain', angle = 0, family = "Times New Roman", inherit.blank = T),
+                  axis.title.y = element_text(size = 10,
+                                            face = 'plain', angle = 90, inherit.blank = T),
+                  axis.text.x= element_text(size = 10,
+                                            face = 'plain', angle = 90, inherit.blank = TRUE),
+                 # axis.text.x.bottom = element_text(size = 10,
+             #                               face = 'plain', angle = 90, inherit.blank = TRUE),
+                  axis.text.y = element_text(size = 10,
+                                            face = 'plain', angle = 0, family = "Times New Roman", inherit.blank = T),
+                #  axis.text.y.right = element_text(size = 12,
+               #                             face = 'plain', angle = 0, family = "Times New Roman", inherit.blank = T),
+                  strip.text.x = element_text(size = 12,
+                                            face = 'plain', angle = 0, family = "Times New Roman",
+                                            margin = margin(t = 3, r = 0, b = 3, l = 0, unit = "pt"), inherit.blank = T),
+                  strip.text.y =element_text(size = 12,
+                                            face = 'plain', family = "Times New Roman", angle = 90,margin = margin(t = 3, r = 3, b = 3, l = 3, unit = "pt"),inherit.blank = T),
+                  strip.background = element_rect(colour = "black", fill = "white"), 
+        panel.background = element_rect(color = "black")
+                    )
+
+my_color <- function(data, mapping, ..., values = c("#bdbdbd", "#737373", "#252525")){
+  ggplot(data = data, mapping = mapping) + 
+    geom_point(...) + 
+    scale_color_manual(values = values) + 
+    scale_fill_manual(values = values)
+}
+
+my_fill <- function(data, mapping, ..., values = c("#bdbdbd", "#737373", "#252525")){
+  ggplot(data = data, mapping = mapping) + 
+    geom_density(...) + 
+    scale_fill_manual(values = values) + 
+    scale_color_manual(values = values)
+}
+
+my_text <- function(data, mapping, ..., values = c("#bdbdbd", "#737373", "#252525")){
+  ggplot(data = data, mapping = mapping) + 
+    geom_text(...) + 
+    scale_fill_manual(values = values) + 
+    scale_color_manual(values = values)
+}
+
+ggpairs(simu_spreadM13, columns = 3:7, aes(colour=Model, alpha = .5)) + 
   theme(plot.title = element_text(size = 10,

### 2019-07-10 paper-child.Rnw
just about done but i think it needs some more oomph in the discussion and a bit more connective tissue between sections.

@@
 
-The lack of a native coordinate system has given rise two primary net-vis methods: the node-link diagram and the adjacency matrix visualization \citep{knuth2000, adjmatpro}.
+The lack of a native coordinate system has given rise two primary net-vis methods: the node-link diagram and the adjacency matrix visualization \citep{adjmatpro}.
 We demonstrate these with a simpe example.
@@
 Because there is often no natural placement of the nodes, they are placed in 2D at random, then adjusted with a layout algorithm.
-For a complete overview of available node-link layout algorithms, we direct the reader to \citet{drawingalgs}.
 Some commonly used algorithms, such as the Kamada-Kawai (KK) layout \citep{kamadakawai} and the Fruchterman-Reingold (FR) layout \citep{fruchterman-reingold}, are designed to mimic physical systems, drawing the graphs based on the ``forces" connecting them.
+For a full discussion of available node-link layout algorithms, we direct the reader to \citet{drawingalgs}.
 %Other algorithms use multi-dimensional scaling or properties of the adjacency matrix (e.g.
@@
 
-\subsubsection{Fitting Models to Data}
+\subsubsection{Fitting Models to Data}\label{subsec:fitting}
 
@@
 The second and third models, M2 and M3, contain one additional parameter each in the objective function.
-We chose the additional parameters because they were significant ($p$-value $<$ 0.05) according to a Wald-type test (\texttt{Wald.RSiena()}) provided in the \texttt{RSiena} software \citep{RSienaTest}.
+We chose the additional parameters because they were significant ($p$-value $<$ 0.05) according to a Wald-type test (\texttt{Wald.RSiena()}) provided in the \texttt{RSiena} software \citep{RSiena}.
 The model M2 contains an actor-level covariate parameter, and the model M3 contains an additional strutural effect in the objective function.
@@
 Other models could have included different rate functions for each node, where for example the rate of change could depend on drinking behavior if we have reason to believe that girls who drink more change friends more frequently.
-We could also model the behavior changes, not just the network changes, if we believe that friends influence behavior, via peer pressure.
-And as before, this method of gathering a collection of models can be used for any network model, such as the hierarchichal Bayesian network estimation model in the \texttt{R} package \texttt{sna}, which allows the user to vary prior and model parameters \citep{sna}.
+We could also model the behavior changes, not just the network changes, if we believe that friends influence behavior via peer pressure.
+And as before, this method of gathering a collection of models can be used for any network model, such as the hierarchichal Bayesian network estimation model in the \texttt{R} package \texttt{sna}, which allows the user to vary hyperparameters in the prior and model parameters \citep{sna}.
 % took out this next bit just because it doesn't really add much and we can just put the appendix figure into the previous section. % add more later.
@@
   xlab("estimate value") + 
-  theme(legend.position = 'bottom')
+  theme(legend.position = 'bottom', strip.text.x = element_text(size = 20), 
+        axis.text.x = element_text(size = 20), axis.title.x = element_text(size = 20),
+        axis.text.y = element_text(size = 20), axis.title.y = element_text(size = 20))
 @
 \caption{\label{fig:beta34} Distribution of fitted $\beta_3, \beta_4$ values.}
-\end{subfigure}
+\end{subfigure}\hfill
 \begin{subfigure}[t]{.6\textwidth}
-<<img, out.width = "\\linewidth">>=
+<<img, out.width = "\\linewidth", fig.height=7.5>>=
 # correlation for other params. 
@@
 
-knitr::include_graphics("img/gray_corplot-1.png")
-@
-\caption{\label{fig:corplots} Relationships between parameters common to all three models. \nocite{ggally}}
+#knitr::include_graphics("img/gray_corplot-1.png")
+knitr::include_graphics("img/plot-matrix.png")
+
+@
+\caption{\label{fig:corplots} Relationships between parameters common to all models. Densities and scatterplots are colored by model. \nocite{ggally}}
 \end{subfigure}
@@
 In fact, the distributions of $\alpha_1$ and $\beta_1$ for M2 do not overlap at all with the distribution of the same parameters from model M1.
-On the other hand, the inclusion of the indirect tie parameter, $\beta_4$, does not lead to such drastically different estimates of other model parameters.
 When $\beta_3$ is included, its estimate is postitive, meaning that friendships between two girls with different drinking behaviors tend to form when there is third girl who is already friends with the other two.
@@
 It is not clear, however, that the addition of a parameter to the objective function should effect the estimates of the rate parameters.
+
+On the other hand, the inclusion of the indirect tie parameter, $\beta_4$, does not lead to such drastically different estimates of other model parameters.
+The mean estimate of $\beta_4$ is large and negative, which suggests that indirect ties are strongly discouraged from forming.
+This is fairly intuitive given the context: teenage girls likely want to have direct friendships in tightly knit groups, so indirect relationships are not common.
 %This strong and possibly problematic relationship between rate parameter estimates and objective function parameters is further explored in the Supporting Information.
 When visualizing the results of the same models fit many times to the same data, we can to see potential model behavior problems right away.
-In Figure~\ref{fig:corplots}, we further examine the relationship between each of pair of parameters common to all three models.
+Finally, in Figure~\ref{fig:corplots}, we demonstrate the relationship between each of pair of parameters common to all three models.
 The strongest correlation within each model is between $\beta_1$ and $\beta_2$, with absolute correlation greater than 0.90 in all three models.
-The $\beta_1$ parameter is also highly correlated with the $\beta_3$ parameter within model M2 ($\rho = $\Sexpr{round(corm2[5, 3])}), but it is not as highly correlated with the $\beta_4$ parameter in model M3 ($\rho = $\Sexpr{round(corM3[5, 3])}).
+The $\beta_1$ parameter is also highly correlated with the $\beta_3$ parameter within model M2 ($\rho = $ \Sexpr{round(corm2[5, 3],3)}), but it is not as highly correlated with the $\beta_4$ parameter in model M3 ($\rho = $ \Sexpr{round(corM3[5, 3], 3)}).
 The inherent relationship between effects is a known problem in network analysis \citep{goldenberg09}.
@@
 At each microstep, an ego node is selected to change, and then randomly makes one change in its ties according to to the probabilities $p_{ij}$ defined in Equation~\ref{eq:pij}.
-This process continues until the network differs from the $x(t_m)$ by exactly $C_m$, as defined in Equation~\ref{eq:cdiff}.
+The microstep process is a key piece of the model fitting procedure and is used in the third phase mentioned in Section~\ref{subsec:fitting} \citep{SienaAlgs}.
+During modeling, the process proceeds until the network differs from the $x(t_m)$ by exactly $C_m$, as defined in Equation~\ref{eq:cdiff}.
 Between two network observations $x(t_m)$ and $x(t_{m+1})$, there can be dozens, hundreds, or even thousands of microsteps, depending on the size of the network and the value of $C_m$.
 We want to simulate and view these in-between steps in order to better understand the behavior of the CTMC model.
-Hiding the microsteps is practical and efficient if the goal is to fit a model to data, obtain parameter estimates, and draw conclusions or make predictions.
-But in order to understand how the models are fit, we need to extract and explore the different steps of the process.
+The default in the software is discard the microsteps, which is practical and efficient if the goal is to fit a model to data, obtain parameter estimates, and draw conclusions or make predictions.
+But in order to understand how the models are fit, we need to extract and explore the different steps of the process, which is possible in \texttt{RSiena} with some small changes to the default settings.
 We use these steps to visually explore the algorithms, so that we are visualizing the model fitting process, and not only the model fitting result.
@@
 
-To visualize the underlying algorithm, we animate the microsteps that form the CTMC from $x(t_1)$ to $x(t_2)$ for the example data.
-The microstep animation, some frames of which are shown in Figure~\ref{fig:makinggif}, represent one possible path from $x(t_1)$ to $x(t_2)$ according to the model M1.
+To visualize the underlying process, we animate the microsteps that form the CTMC from $x(t_1)$ to $x(t_2)$ for the example data.
+The microstep animation, some frames of which are shown in Figure~\ref{fig:makinggif}, represents one possible path from $x(t_1)$ to $x(t_2)$ according to the model M1.
 We show each change in the network as follows: we first increase the size and change the color of the ego node to focus attention to its ties.
@@
 Movies similar to this animation were used to visualize the changes of dynamic networks in \citet{dynnetviz}.
-Frames 15-28 of Figure~\ref{fig:makinggif} show tie that is destroyed, while frames 71-84 show a tie that is created.
+Frames 15-28 of Figure~\ref{fig:makinggif} show a tie that is destroyed, while frames 71-84 show a tie that is created.
 For both changes, the ego node is emphasized with color and size.
@@
 Combining the tie changes in this animation makes the CTMC process the focus, demonstrating the underlying change mechanism of the dynamic network according to the model.
+
+With this animation, we can see how the model behaves.
+Some tie changes appear to alter the overall network structure, as shown by the change in the layout of the nodes, while others seem to be inevitable, such as an added reciprocal tie between two nodes in a large connected group that do not change the layout.
+Other edges are created in one step and destroyed the next.
+When watching the animation, we are watching the model in action, which helps us better understand its behavior.
+This same animation method can be applied to any network model which changes one tie at a time.
+For example, the Erd\"{o}s-R\'{e}nyi random graph model is a series of tie additions, and we could animate the creation of a network from this model.
+Unlike CTMCs, however, Erd\"{o}s-R\'{e}nyi random graph models only create ties \citep{goldenberg09}.
+Regardless of the network model under consideration, animating the node-link diagram to view how the network changes gives insight into the model behavior and the process of model fitting.
 % need more about what we learned.
@@
 We continue to use the seriated matrix in another animation of the microstep process, which is available at \url{https://vimeo.com/240092677}.
-Some frames of this video are shown in Figure~\ref{fig:adjmatani}.
-In the animation, a square appears or disappears as that edge appears or disappears in the microstep process.
+Some frames of the video are shown in Figure~\ref{fig:adjmatani}.
+In the animation, a square appears or disappears when the edge it represents appears or disappears in the microstep process.
 %Through this animation, we can see edges appearing, and then later on disappearing.
-This vanimation, unlike the node-link visualization, shows the microsteps where the chosen node does not make a change.
-From start to finish, we see the network evolve: sometimes ties are removed right after they are created, and sometimes several ties are removed or created in a row.
-The adjacency matrix animation shows the algorithm exploring the data space.
-In contrast to the previous animation, all possible edges are represented in the adjacency matrix visualization.
-By viewing this animation, we gain a better understanding of the underlying dynamics of this model.
+This animation, unlike the previous one, shows the microsteps where the chosen node does not make a change.
+Starting at frame 0 (the first network observation $x(t_1)$), there are two clearly connected components: one in the bottom left, and one in the top right.
+With each frame, we see the network evolve: sometimes ties are removed right after they are created, and sometimes several ties are removed or created in a row.
+By the end, the bottom left component has become more sparsely connected, while the top right component has shrunk, but remains closely connected.
+In contrast to the node-link animation, the adjacency matrix animation shows how the algorithm explore the space of all possible networks, because possible edges are represented in each frame of the adjacency matrix animation.
+It is also easier to see the connected components of the graph in this animation because we have clearly grouped them with matrix seriation.
+As with the node-link animation, the adjacency matrix animation could also be used for other dynamic network models.
+Because the adjacency matrix emphasizes edges, this animation method is best for large directed networks whose structure is easier to see in an adjacency matrix visualization than in a node-link visualization \citep{adjmatviz}.
+Both animation methods demonstrate model behavior that would have not been seen with traditional visualizations or computations.
 % need more oomph
@@
 The node labels are removed to declutter the figure.
-Starting at frame 0 (the data), there are two clearly connected components: one in the bottom left, and one in the top right.
-By the end, the bottom left component has become more sparse, while the top right component has shrunk, but remains closely connected.', warning = FALSE, message = FALSE, results='hide', fig.height=3, out.width="\\linewidth">>=
+Complete video available at \\url{https://vimeo.com/240092677}.', warning = FALSE, message = FALSE, results='hide', fig.height=3, out.width="\\linewidth">>=
 adjmatall <- matrix(0, nrow = 16, ncol = 16)
@@
 @
-\begin{figure}
-\begin{subfigure}[t]{\textwidth}
-<<heatmap, fig.align='center', fig.height=2.5>>=
-
-ansnullchains %>%
-  filter(period == 1) %>%
-  group_by(rep) %>%
-  mutate(msno = row_number()) %>%
-  filter(msno == 1) %>%
-  mutate(ego = as.numeric(as.character(ego))+1,
-         alter = as.numeric(as.character(alter))+1,
-         prob = exp(logAlterProb)) -> probsDat
-
-ansnullchains %>%
-  filter(period == 1) %>%
-  group_by(rep) %>%
-  mutate(msno = row_number()) %>%
-  filter(msno == 1) %>%
-  ungroup() %>% group_by(ego, alter) %>% 
-  tally() %>% ungroup() %>% mutate(empprob = n/1000) %>% 
-  mutate(ego = as.numeric(as.character(ego))+1,
-         alter = as.numeric(as.character(alter))+1) ->empprobsDat
-
-p1 <- ggplot(data = probsDat,
-       aes(y = factor(ego, levels = rev(order1)),
-             x = factor(alter, levels = order1),
-             fill = prob)) +
-  geom_tile(color = 'black') +
-  scale_fill_gradient(low = 'white', high = 'black', limits = c(0,1), name = "theoretical\nprobability", labels = c("0","",".5","","1")) +
-  labs(y = "ego", x = "alter")+
-  scale_x_discrete(position = 'top') +
-  ThemeNoNet + 
-  theme(panel.grid = element_blank(), 
-          axis.text.x.top = element_text(size=rel(.7)),
-          axis.text.y.left = element_text(size=rel(.7)),
-          legend.position = "bottom", 
-          legend.title = element_text(size=rel(.7)), 
-          legend.text = element_text(size=rel(.7)), 
-          legend.key.height = unit(.5, "cm"), 
-          legend.margin=margin(0,0,0,0),
-          legend.box.margin=margin(-10,-10,-5,-10)) +
-  coord_fixed() + 
-  ggtitle("Theoretical Probabilities") + 
-  labs(x= "alter (j)", y = "ego (i)")
-p2 <- ggplot(data = empprobsDat,
-       aes(y = factor(ego, levels = rev(order1)),
-             x = factor(alter, levels = order1),
-             fill = empprob)) +
-  geom_tile(color = 'black') +
-  scale_fill_gradient(low = 'white', high = 'black', name = "empirical\nprobability", labels = c(".02", ".04", ".06"), breaks = c(.02, .04, .06)) +
-  labs(y = "ego", x = "alter")+
-  scale_x_discrete(position = 'top') +
-  ThemeNoNet + 
-    theme(panel.grid = element_blank(), 
-          axis.text.x.top = element_text(size=rel(.7)),
-          axis.text.y.left = element_text(size=rel(.7)),
-          legend.position = "bottom", 
-          legend.title = element_text(size=rel(.7)), 
-          legend.text = element_text(size=rel(.7)), 
-          legend.key.height = unit(.5, "cm"), 
-          legend.margin=margin(0,0,0,0),
-          legend.box.margin=margin(-10,-10,-5,-10)) +
-  coord_fixed() + 
-  ggtitle("Empirical Probabilities") + 
-  labs(x= "alter (j)", y = "ego (i)")
-plot_grid(p1, p2)
-@
-\caption{\label{fig:heatmap}Adjacency matrix visualizations showing the theoretical (left) and empirical (right) transition probabilities for the first microstep taken in 1,000 simulations.
+% \begin{figure}
+% \begin{subfigure}[t]{\textwidth}
+% <<heatmap, fig.align='center', fig.height=2.5>>=
+% 
+% ansnullchains %>%
+%   filter(period == 1) %>%
+%   group_by(rep) %>%
+%   mutate(msno = row_number()) %>%
+%   filter(msno == 1) %>%
+%   mutate(ego = as.numeric(as.character(ego))+1,
+%          alter = as.numeric(as.character(alter))+1,
+%          prob = exp(logAlterProb)) -> probsDat
+% 
+% ansnullchains %>%
+%   filter(period == 1) %>%
+%   group_by(rep) %>%
+%   mutate(msno = row_number()) %>%
+%   filter(msno == 1) %>%
+%   ungroup() %>% group_by(ego, alter) %>% 
+%   tally() %>% ungroup() %>% mutate(empprob = n/1000) %>% 
+%   mutate(ego = as.numeric(as.character(ego))+1,
+%          alter = as.numeric(as.character(alter))+1) ->empprobsDat
+% 
+% p1 <- ggplot(data = probsDat,
+%        aes(y = factor(ego, levels = rev(order1)),
+%              x = factor(alter, levels = order1),
+%              fill = prob)) +
+%   geom_tile(color = 'black') +
+%   scale_fill_gradient(low = 'white', high = 'black', limits = c(0,1), name = "theoretical\nprobability", labels = c("0","",".5","","1")) +
+%   labs(y = "ego", x = "alter")+
+%   scale_x_discrete(position = 'top') +
+%   ThemeNoNet + 
+%   theme(panel.grid = element_blank(), 
+%           axis.text.x.top = element_text(size=rel(.7)),
+%           axis.text.y.left = element_text(size=rel(.7)),
+%           legend.position = "bottom", 
+%           legend.title = element_text(size=rel(.7)), 
+%           legend.text = element_text(size=rel(.7)), 
+%           legend.key.height = unit(.5, "cm"), 
+%           legend.margin=margin(0,0,0,0),
+%           legend.box.margin=margin(-10,-10,-5,-10)) +
+%   coord_fixed() + 
+%   ggtitle("Theoretical Probabilities") + 
+%   labs(x= "alter (j)", y = "ego (i)")
+% p2 <- ggplot(data = empprobsDat,
+%        aes(y = factor(ego, levels = rev(order1)),
+%              x = factor(alter, levels = order1),
+%              fill = empprob)) +
+%   geom_tile(color = 'black') +
+%   scale_fill_gradient(low = 'white', high = 'black', name = "empirical\nprobability", labels = c(".02", ".04", ".06"), breaks = c(.02, .04, .06)) +
+%   labs(y = "ego", x = "alter")+
+%   scale_x_discrete(position = 'top') +
+%   ThemeNoNet + 
+%     theme(panel.grid = element_blank(), 
+%           axis.text.x.top = element_text(size=rel(.7)),
+%           axis.text.y.left = element_text(size=rel(.7)),
+%           legend.position = "bottom", 
+%           legend.title = element_text(size=rel(.7)), 
+%           legend.text = element_text(size=rel(.7)), 
+%           legend.key.height = unit(.5, "cm"), 
+%           legend.margin=margin(0,0,0,0),
+%           legend.box.margin=margin(-10,-10,-5,-10)) +
+%   coord_fixed() + 
+%   ggtitle("Empirical Probabilities") + 
+%   labs(x= "alter (j)", y = "ego (i)")
+% plot_grid(p1, p2)
+% @
+% \caption{\label{fig:heatmap}Adjacency matrix visualizations showing the theoretical (left) and empirical (right) transition probabilities for the first microstep taken in 1,000 simulations.
 There are many ties with empirical probability zero.}
-\end{subfigure}
-\begin{subfigure}[t]{\textwidth}
-<<alledgests, fig.height = 3>>=
-# get a plot of all microsteps from w1 to w2 by edge
-alledges <- expand.grid(from = paste0("V", 1:16), to = paste0("V", 1:16))
-alledges$eid <- 1:nrow(alledges)
-
-sfmsall %>% left_join(alledges) -> sfmsall2
-
-sfmsall2 %>% group_by(rep) %>% mutate(isWave2 = (ms == max(ms))) -> sfmsall2
-
-sfmsall2 %>% group_by(from, to, eid) %>%
-  summarize(total =n()) %>%
-  arrange(desc(total)) %>%
-  ungroup %>%
-  mutate(plotorder = row_number()) -> sfmsallorder
-
-sfmsall2 %>% group_by(ms, eid) %>%
-  summarise(count = n()) %>%
-  mutate(alpha = count/1000) %>%
-  arrange(ms, eid) -> sfmsall3
-
-left_join(sfmsall3, sfmsallorder, by = "eid") -> sfmsall4
-
-#add true wave 2 to plot
-wave2 <- na.omit(actual2)
-wave2$from <- paste0("V", wave2$from)
-wave2$to <- paste0("V", wave2$to)
-wave2 <- left_join(wave2, alledges)
-wave2 <- left_join(wave2, sfmsallorder)
-
-sfmsall5 <- sfmsall4 %>% ungroup %>% add_row(ms = max(sfmsall4$ms) + 1,
-        eid = wave2$eid,
-        count = NA,
-        alpha = 1,
-        from = wave2$from,
-        to = wave2$to,
-        total = wave2$total,
-        plotorder = wave2$plotorder)
-sfmsall5$alphaFctr <- cut(sfmsall5$alpha, c(0,.01, .05, .10, .25, .5, .75, 1, 1.1))
-ggplot(data = sfmsall5) +
-  geom_tile(aes(x = ms, y = reorder(eid,-plotorder), fill = alphaFctr)) +
-  annotate(geom = "rect", xmin=.25, xmax = 1.75, ymin = -Inf, ymax = Inf, color = "red", fill = NA, size = .01, alpha = .15) + 
-  #scale_fill_gradient(low = 'grey90', high = '#0868ac', name = "occurrence %") +
-  scale_fill_brewer(palette = "Greys", name = "Occurrence %", labels = c("(0-1]%", "(1-5]%", "(5-10]%","(10-25]%", "(25-50]%", "(50-75]%", "(75-100]%")) +
-  labs(x = "Microstep No.", y = "Edges (ordered by total # occurrences)") +
-  ThemeNoNet +
-  theme(axis.ticks.y = element_blank(), axis.text.y = element_blank(), panel.grid = element_blank())
-@
-\caption{\label{fig:alledgests}Visualizing all microsteps taken in 1,000 simulations from the model M1.
-All $16\times(16-1)=240$ possible edges are represented on the y-axis, which is ordered by how often edges occur in the simulation process.
+% \end{subfigure}
+% \begin{subfigure}[t]{\textwidth}
+% <<alledgests, fig.height = 3>>=
+% # get a plot of all microsteps from w1 to w2 by edge
+% alledges <- expand.grid(from = paste0("V", 1:16), to = paste0("V", 1:16))
+% alledges$eid <- 1:nrow(alledges)
+% 
+% sfmsall %>% left_join(alledges) -> sfmsall2
+% 
+% sfmsall2 %>% group_by(rep) %>% mutate(isWave2 = (ms == max(ms))) -> sfmsall2
+% 
+% sfmsall2 %>% group_by(from, to, eid) %>%
+%   summarize(total =n()) %>%
+%   arrange(desc(total)) %>%
+%   ungroup %>%
+%   mutate(plotorder = row_number()) -> sfmsallorder
+% 
+% sfmsall2 %>% group_by(ms, eid) %>%
+%   summarise(count = n()) %>%
+%   mutate(alpha = count/1000) %>%
+%   arrange(ms, eid) -> sfmsall3
+% 
+% left_join(sfmsall3, sfmsallorder, by = "eid") -> sfmsall4
+% 
+% #add true wave 2 to plot
+% wave2 <- na.omit(actual2)
+% wave2$from <- paste0("V", wave2$from)
+% wave2$to <- paste0("V", wave2$to)
+% wave2 <- left_join(wave2, alledges)
+% wave2 <- left_join(wave2, sfmsallorder)
+% 
+% sfmsall5 <- sfmsall4 %>% ungroup %>% add_row(ms = max(sfmsall4$ms) + 1,
+%         eid = wave2$eid,
+%         count = NA,
+%         alpha = 1,
+%         from = wave2$from,
+%         to = wave2$to,
+%         total = wave2$total,
+%         plotorder = wave2$plotorder)
+% sfmsall5$alphaFctr <- cut(sfmsall5$alpha, c(0,.01, .05, .10, .25, .5, .75, 1, 1.1))
+% ggplot(data = sfmsall5) +
+%   geom_tile(aes(x = ms, y = reorder(eid,-plotorder), fill = alphaFctr)) +
+%   annotate(geom = "rect", xmin=.25, xmax = 1.75, ymin = -Inf, ymax = Inf, color = "red", fill = NA, size = .01, alpha = .15) + 
+%   #scale_fill_gradient(low = 'grey90', high = '#0868ac', name = "occurrence %") +
+%   scale_fill_brewer(palette = "Greys", name = "Occurrence %", labels = c("(0-1]%", "(1-5]%", "(5-10]%","(10-25]%", "(25-50]%", "(50-75]%", "(75-100]%")) +
+%   labs(x = "Microstep No.", y = "Edges (ordered by total # occurrences)") +
+%   ThemeNoNet +
+%   theme(axis.ticks.y = element_blank(), axis.text.y = element_blank(), panel.grid = element_blank())
+% @
+% \caption{\label{fig:alledgests}Visualizing all microsteps taken in 1,000 simulations from the model M1.
+All $16\times(16-1)=240$ possible edges are represented on the y-axis, which is in descending order according to how often edges occur in the simulation process.
 The occurrence percent is split up into groups to correspond with its distribution: only about 10\% of the possible edges appear more than 10\% of the time in the 1,000 simulations, while about 60\% appear less than 1\% of the time.
@@
 We see that it is rare for a microstep process to last longer than 150 steps.}
-\end{subfigure}
-\caption{\label{fig:probabilities}Visualizing the underlying algorithms when fitting CTMC models.
+% \end{subfigure}
+% \caption{\label{fig:probabilities}Visualizing the underlying algorithms when fitting CTMC models.
 The emprical probability plot (on the right in Figure~\ref{fig:heatmap}) is also shown at microstep number 1 in Figure~\ref{fig:alledgests}.}
-\end{figure}
-
-We also illuminate the microstep process by visualizing the theoretical and observed transition probabilites for the first microstep in the process.
+% \end{figure}
+% 
+% Finally, we illuminate the microstep process by visualizing the theoretical and observed transition probabilites for the first microstep in the process.
 These probabilities are defined in Equation~\ref{eq:pij}.
-We only do the first step of many because each microstep is dependent on the previous network state, and thus only the first steps can be compared across simulations.
-There are 1,000 transitions to examine: the first microstep taken in each one of the simulations. %Another way to view these transition probabilities is through the adjacency matrix visualization. 
-We use another seriated adjacency matrix visualization to show theoretical and empirical transition probabilities of the first microstep in Figure~\ref{fig:heatmap} for each tie that was the first change in a simulation.
+We examine the first step because each microstep is dependent on its predecessor, and thus only the first steps can be compared across simulations.
+% There are 1,000 transitions to examine: the first microstep taken in each one of the simulations. %Another way to view these transition probabilities is through the adjacency matrix visualization. 
+% We use another seriated adjacency matrix visualization to show theoretical and empirical transition probabilities of the first microstep in Figure~\ref{fig:heatmap} for each tie that was the first change in a simulation.
 The figure is noticeably sparse: %showing the lack of coverage in the model.
-of the $16\cdot 16 = 256$ possible choices for the CTMC to make, only 103, or about 43\%, are made in the 1,000 simulated chains.
+% of the $16\cdot 16 = 256$ possible choices for the CTMC to make, only 103, or about 43\%, are made in the 1,000 simulated chains.
 This leaves %reinforces what we saw in Figure~\ref{fig:histogramalters}, where there are many paths not taken.
 This effect could only be exacerbated as more steps are taken in the CTMC, leading to
-a large area of data space unexplored by the first step in the model fitting process.
-We present additional evidence for lack of coverage in the Supporting Information.
+% a large area of data space unexplored by the first step in the model fitting process.
 The Figures~\ref{fig:heatmap} and \ref{fig:alledgests} show that the model M1 and the the corresponding fitting process do not explore the data space enough to adequately capture the network change mechanism, a finding we were only able to discern thanks to model visualization techniques.
@@
 
-We have used several visualizations, some old and some new, to explore social network models new ways.
-The guiding principles of model-vis: viewing the model in the data space, visualizing collections of models, and looking at the underlying algorithms, gave us knowledge and appreciation for these complicated and flexible models.
+We have used several visualization methods, some old and some new, to diagnose and understand network models in new ways.
+Though we have applied these methods to CTMC models for dynamic social networks, all of our visualizations can be applied to other dynamic network models and most can also be applied to any static network model.
+We gained insight into and appreciation for complicated and flexible network models by following the guiding principles of model visualization: viewing the model in the data space, visualizing collections of models, and looking at the underlying algorithms \citep{modelvis}.
 When we look at the model in the data space by viewing the expected network simulated from a model, we are able to easily see the lack of model fit.
-When visualizing collections of models, we uncover concerning relationships between parameters.
-When we explore the algorithms, we gain a better understanding of the model and its stochastic behavior.
-Our three-pronged model-vis attack has uncovered many properties of these models.
+Computing and visualizing the difference network between the data and the expected network also helps assess model fit by showing us where the model is completely missing the data.
+When exploring collections of models, we learn more about the large number of parameters possible in network models, and we can also uncover relationships between these parameters.
+Finally, when we explore the algorithms fitting the models, we have a better understanding of the model and its stochastic behavior.
+All in all, this three-pronged model-vis attack has uncovered many properties of CTMC models for social networks and shows great potential for application to other network models.
 
@@
 Furthermore, the data we have used are quite small by network data standards, with only 16 nodes.
-Thus, some of the visualizations we have presented may not scale up to larger networks.
+Thus, some of the visualizations we have presented, such as the small node-link diagram animation, may not scale up to larger networks.
 Future network model-vis research could also include interactive and/or three-dimensional graphics, which we did not explore here.
 Interactivity is the state-of-the-art in data visualization, and network visualization is just starting to catch on \citep[see e.g.\ ][]{refinery}.
-\citet{modelvis} also relied on interactive tools like GGobi \citep{STLBC03} for interactivity, and for networks, there is some interactive support with the package \texttt{plotly} \citep{plotly} built into \texttt{geomnet}.
-
-We hope that the principles of model-vis continue to be used for learning more about the fit and behavior of complicated statistical models.
+\citet{modelvis} also relied on interactive tools like GGobi \citep{STLBC03} for interactivity, and for networks, there is some interactive support with the package \texttt{plotly}  built into \texttt{geomnet} \citep{plotly, geomnet}.
+
+We hope that the principles of model-vis continue to be used for learning more about the fit and behavior of statistical models for network data.
 Our model-vis application has uncovered many properties of CTMC models for dynamic social networks, and potential abounds for application of model-vis in other areas of network analysis.
 
-\subsection*{Online supplemental material:}
-
-In addition to the Supporting Information available online from the publisher, much of the code used to create the visualizations in this paper has been packaged and made available online at \url{https://github.com/sctyner/netvizinf} and \url{https://github.com/sctyner/geomnet} (also on CRAN).
+{\centering \subsection*{Online supplemental material:}}
+
+Much of the code used to create the visualizations in this paper has been packaged and made available online at \url{https://github.com/sctyner/netvizinf} and \url{https://github.com/sctyner/geomnet} (also on CRAN).
 Additional, non-packaged code can be found at \url{http://bit.ly/ctmc-model-viz}.

### 2019-07-10 paper.Rnw
working on the intro
stuff from last week
just about done but i think it needs some more oomph in the discussion and a bit more connective tissue between sections.

@@
 \graphicspath{{figure/}}
-\usepackage{endfloat} % Figures to the end of the document
+%\usepackage{endfloat} % Figures to the end of the document
 
@@
 \if0\blind
-{
-  \title{\bf Visual Diagnostics of Dynamic Social Network Models}
+{ % new title ideas: Dynamic Social Network models and their Visualizations
+  %                  Visualizations for Diagnosing Network Models  
+  \title{\bf Visualizations for Diagnosing Network Models}
   \author{Sam Tyner \hspace{.2cm}\\
@@
   \begin{center}
-    {\LARGE\bf Visual Diagnostics of Dynamic Social Network Models}
+    {\LARGE\bf Visualizations for Diagnosing Network Models}
 \end{center}

### 2019-07-12 paper-child.Rnw
some edits

@@
 
+\hh{Sam, it might be better to chop section 1 into two - 1.
+introduction and 2.
+Background (or Network Models and Visualization).
+12 pages of intro is too much.
+Make sure to move the last sentence of 2.
+Background to the end of 1.
+Introduction.
+That allows people familiar with network models to skim over section 2 and go directly to the meat of the paper.}
 \section{Introduction \& Background}
@@
 
-\par We are particularly interested one truly dynamic family of network models: continuous time Markov chain (CTMC) models, first introduced in \citet{ctmcorig}.
+\par We are particularly interested in one truly dynamic family of network models: continuous time Markov chain (CTMC) models, first introduced in \citet{ctmcorig}.
 We chose these models because they correspond to our intuition about how social networks are formed, and they are very flexible and widely applicable.
@@
 In addition, we can incorporate node covariate information into the hierarchy, such as gender or education level of people in a social network.
-For instance, the genders of two people could effect whether and how quickly they establish a relationship, and people will likely form more relationships with those who share their education level.
-Thus, the transition probabilities between two states of a network depend on the network's structure, characteristics of the nodes, or both \citep{saompaper}.
+For instance, the genders of two people could affect whether and how quickly they establish a relationship.
+People also will likely form more relationships with those who share their education level.
+Thus, the transition probabilities between two states of a network depend on the network's structure, the characteristics of its nodes, or both \citep{saompaper}.
 For example, \citet{depression} use a CTMC model to learn about the tendency of depressed adolescents to form friendships and found that depressed students tended to avoid creating new friendships over time.
-The CTMC model family can also be expanded to model node behavior, as in \citet{chyzh2016}, who applies CTMC models to show that more indirect trade relationships in a country decrease its overall respect for human rights.
+The CTMC model family can also be expanded to model node behavior, as in \citet{chyzh2016}, who apply CTMC models to show that more indirect trade relationships in a country decrease its overall respect for human rights.
 % This means that node-level covariate effects, in addition to commonly used effects such as reciprocity or transitivity, can be incorporated into the model.
@@
 More recently, \citet{gofschwein} used score-type tests for forward model selection in the CTMC model context.
-The \texttt{RSiena} software also contains methods for goodness-of-fit based on network statistics, such as outdegree distribution, where the data values are compared to simulated values as in \citet{hunter2008} \citep{RSiena}.
+The \texttt{RSiena} software also contains methods for goodness-of-fit based on network statistics, such as outdegree distribution, where the data values are compared to simulated values as in \citet{hunter2008} and in \citet{RSiena}.
 % need to transition to model vis.
 % say something about how important visualization is in network analysis.
+HH: I think this transition is fine.
 
@@
 Note that we do not propose new methods of visualizing networks, and for a complete review of the field of graph drawing we direct the reader to \citet{graphdrawing}.
-What is novel in our work is the use of common visualization methods to diagnose network model with model visualization
+What is novel in our work is the use of existing visualization methods to diagnose network models with model visualization.
 
@@
 By translating the model visualization work of \citet{modelvis} to the dynamic network model space, we provide network researchers with a new suite of tools for visualizing and diagnosing performance of their network models.
-Though we apply the tools to CTMCs because we find the models intuitive, the methods we use can be applied to any network model, static or dynamic.% do something awesome
+Though we apply the tools to CTMCs because we find the models intuitive, the methods we use can be applied to any network model, static or dynamic.% do something awesome :)
 
@@
 We also use the score function method for estimating the derivatives of the expected values, as opposed to the finite differences method, both of which are outlined in detail in .
-The underlying SIENA software uses a Robbins-Monro algorithm (see \citet{robbinsmonro}) to estimate the solution of the moment equation
+The underlying SIENA software uses a Robbins-Monro algorithm (see \citet[see][]{robbinsmonro}) to estimate the solution of the moment equation
 \begin{equation}\label{eq:mom}
@@
 <<alldataviz, fig.cap="The small friendship network data we model throughout this paper.
-The FR layout is used.
-The teenagers' drinking behavior changes and the network becomes less densely connected over time.
+Each node represents a teenager, an edge is drawn if XXX.
+A Fruchterman-Reingold layout is used.
+By comparing the node-link diagrams we see  that teenagers' drinking behavior changes -- drinking generally increases; darker nodes indicate more occurrences -- and the network becomes less densely connected over time.
 These are the mechanisms we hope to model with our chosen CTMCs.", out.width='.99\\textwidth', fig.height=3, results='hide'>>=
@@
 
-\noindent \textbf{Example Data:} To guide our visual exploration of CTMC models for dynamic networks, we use a canonical example in the social network literature: a subset of the 50 actor dataset from the ``Teenage Friends and Lifestyle Study"  \citet{friendsdata}.
+\noindent \textbf{Example Data:} To guide our visual exploration of CTMC models for dynamic networks, we use a canonical example in the social network literature: a subset of the 50 actor dataset from the ``Teenage Friends and Lifestyle Study"  \citep{friendsdata}.
 We chose this data because it is a well-known and easy to understand example.
@@
 We chose to only work with a subset of the data to make node-link diagrams easier to see and to make any changes in the network more noticeable.
+\hh{HH: ground the figure - this is the first time you are showing this example. A node is ... an edge is ... , colour is ...}
 We visualize this data with a node-link diagram in Figure~\ref{fig:alldataviz}, where %For model fitting, we condition on wave 1 and estimate the parameters of the models for the second and third waves.
@@
 We chose the additional parameters because they were significant ($p$-value $<$ 0.05) according to a Wald-type test (\texttt{Wald.RSiena()}) provided in the \texttt{RSiena} software \citep{RSiena}.
-The model M2 contains an actor-level covariate parameter, and the model M3 contains an additional strutural effect in the objective function.
+The model M2 contains an actor-level covariate parameter, and the model M3 contains an additional structural effect in the objective function.
 \begin{align}\label{eq:m2m3}
@@
 \noindent where $s_{i3}(x,z)$ and $s_{i4}(x)$ are defined in Figure~\ref{fig:structures}.
-These statistics are known as the \emph{number of jumping transitive triplets} (JTT) and the \emph{number of doubly achieved distances two effect} (N22), respectively.
+These statistics are known as the \emph{number of jumping transitive triplets} (JTT) and the \emph{number of doubly achieved distances two effect} \hh{HH do you need the `two` in there?} (N22), respectively.
 The JTT effect emphasizes triads formed between actors with \emph{different} covariate values, while the N22 effect emphasizes \emph{indirect} ties between actors.
@@
 
+\hh{HH: include the names of the effects in the table -- otherwise it's really hard to make any sense of the numbers. }
 % table of fitted M1 M2 M3
@@
 
-We encourage interested readers to consult the \texttt{RSiena} manual for a complete definition and description of these and all other possible effects available to use in CTMC models \citep{RSienamanual}.
+\hh{The values in the table show us that all effects are very significant and relatively stable across the different models. }
+\hh{HH: sentence of discussion - do we see the expected increase/decrease in the parameters?}
+\hh{What we do not learn from the values in the table is anything about the goodness of fit of each of the models, or dependencies between the parameters and their estimates. }
+
+%We encourage interested readers to consult the \texttt{RSiena} manual for a complete definition and description of these and all other possible effects available to use in CTMC models \citep{RSienamanual}.
+\hh{HH The next paragraph could be shortened some more to emphasize what we (want to) do.}
 Our goal here is to assess the fit of these models to our data, not to find the best model for this data, which is why we have kept both the data and the models very small.
@@
 
+\subsection{\hh{Expectation Networks}}
 <<funsenatevis, fig.height=3, fig.cap='The 111$^{th}$ Senate at two times: while Clinton was in the senate in 2009 (left) and after she departed (right).
@@
 
+\hh{HH: we inconvenient question - could we use the same layout for all the networks (by using facetting instead of drawing individual plots).
+That would highlight the structure and changes to the structure better.
+I think, highliting the edges by making them darker and downweighing the nodes by making them lighter would also help.
+Something like (code in comment):
+% ggplot(data = combinedavgs) +
+%   geom_net(aes(from_id = from, to_id = to, linewidth= linewidth/3),
+%            directed = T, curvature = .2, labelon=T, size = 3,
+%            arrow = arrow(angle = 20, type = "open", length = unit(.2, "cm")), ecolour = "black",
+%            color = "grey60", hjust = .5, vjust = .5, labelcolour = 'white', fontsize = 2, arrowgap = .03,
+%            singletons = T) +
+%   labs(title = "Time 1") + ThemeNet +
+%   theme(plot.background = element_rect(color = 'black')) + facet_wrap(~cat)
+Hopefully this helps with the description below, because it's quite hard to read with all the numbers of nodes.
+}
+
 \noindent The above method can be generalized and applied to any network model, static or dynamic, from which we can simulate, for any number of simulations from the network.
@@
 Edges that appear with higher frequency are drawn with thicker lines.
+\hh{HH Give a definition of how edge frequency is defined in the plot - frequency of 1 means that this edge was present in all of the simulations, right?}
 On the bottom row of Figure~\ref{fig:averagenet}, we show the data for comparison: $x(t_1)$ on the left, and $x(t_2)$ on the right.
@@
 
+\subsection{\hh{Difference Networks}}
+
 Another way to view our models in the data space is with \textit{difference networks} between the expected network and the data, as shown in Figure~\ref{fig:summary2}.
@@
 Thus, when modeling the transitive triplet behavior in this way, the model concludes that girls change friends more frequently.
-It is not clear, however, that the addition of a parameter to the objective function should effect the estimates of the rate parameters.
+It is not clear, however, that the addition of a parameter to the objective function should affect the estimates of the rate parameters.
 
@@
 
+\hh{It would make the discussion easier, if all of the main plots we introduce had a name - like expectation networks, difference (or residual?) networks, ...
+maybe 'parameter diagnostics' for the second approach?
+and change networks for the last one?
+}
+
+\hh{We started with the three pronged approach of visualization ... let's bring it back to that.}
+\hh{Another point is a call for action: along the lines of now we need to use it!
+Visualization of network models should become second nature, similar to linear model diagnostics, we should have a suite of visualizations next to diagnostic values ready to launch. Ease of use?}
+
 We have used several visualization methods, some old and some new, to diagnose and understand network models in new ways.

### 2019-07-12 paper-child.Rnw
run spell check, incorporate some of hh's changes

@@
-\newcommand{\st}[1]{{\color{orange} #1}}
-\newcommand{\hh}[1]{{\color{magenta} #1}}
+%\newcommand{\st}[1]{{\color{orange} #1}}
+%\newcommand{\hh}[1]{{\color{magenta} #1}}
 
@@
 
-\hh{Sam, it might be better to chop section 1 into two - 1.
+%\hh{Sam, it might be better to chop section 1 into two - 1.
 introduction and 2.
@@
 That allows people familiar with network models to skim over section 2 and go directly to the meat of the paper.}
-\section{Introduction \& Background}
-% peices of old intro
-%Social networks, such as collaboration networks between academic researchers or friendship networks 
-% new intro 
-%As humans and researchers, it is our instint to try to learn as much as we can about the world around us. 
+\section{Introduction}
 Scholars have been trying to understand relationships between people, or \emph{social networks}, for decades, especially in the social sciences.
@@
 Other static network model families include the Erd\"{o}s-R\'{e}nyi random graph model and the exchangeable graph model \citep{goldenberg09}.
-% describe more about the ergms. cite goldenberg. answer the questions: how are they analyzed? What can they do? what can't they do? HOw are they applied? 
 These static models, however, are not optimal for modeling real networks because they do not account for changes in the network as time passes.
@@
 Some of the classical models, such as the Erd\"{o}s-R\'{e}nyi model and the small-world model of \citet{watts} are pseudo-dynamic because they could be applied in a dynamic way but rarely are \citep{goldenberg09}.
-% talk up dynamic social network models.
-they are much more realistic.
-they don't treat nodes as exchangeable, they allow the nodes to affect the ties.
-They account for both structural and covariate effects.
-They are very flexible.
-They are the ideal model because they are the only ones that have any chance of actually capturing the mechanisms.
 Fully dynamic social network models, however, are more complicated: they incorporate both the network structure and its changes in time.
@@
 In these models, network changes are modeled hierarchically at the node level: nodes are given opportunities to change via a rate function, then the node chooses which edge to change according to its utility function.
-%first step is the changes happen at the smallest possible level (one edge between two nodes), and subsequent changes condition on past changes.
 In each step of the hierarchy, many possible effects can be implemented to capture the change mechanism.
@@
 The CTMC model family can also be expanded to model node behavior, as in \citet{chyzh2016}, who apply CTMC models to show that more indirect trade relationships in a country decrease its overall respect for human rights.
-% This means that node-level covariate effects, in addition to commonly used effects such as reciprocity or transitivity, can be incorporated into the model.
-% Cite some of olga's work. what can you do with these models? cite other work with social networks. 
 Thus, with CTMC models we can examine assumptions made about social networks.
+
 %For example, we may assume that teenagers who drink heavily are more likely to be friends with other teenagers who drink heavily.
@@
 Methods for maximum likelihood estimation (MLE) for these models were not introduced until \citet{saomsml2010}, and method of moments (MM) estimation is frequently used \citep{RSiena}.
-%SAOMs are typically fit to dynamic social network data using a series of MCMC phases for finding method of moments (MoM) estimates of model parameters.
 For CTMC parameter estimation, we use the software SIENA, and its \texttt{R} implementation \texttt{RSiena} \citep{rstats, RSiena}.
@@
 
-\emph{Model visualization} complements traditional model diagnostic tools such as $R^2$ values and residual plots to enrich our understanding and interpretation of statistical models \citep{modelvis}.
+Model visualization or model-vis complements traditional model diagnostic tools such as $R^2$ values and residual plots to enrich our understanding and interpretation of statistical models \citep{modelvis}.
 We apply the three principles of model-vis: view the model in the data space, visualize collections of models, and explore the process of fitting the model, to the models and data introduced in Section~\ref{sec:exdatamods}.
 By translating the model visualization work of \citet{modelvis} to the dynamic network model space, we provide network researchers with a new suite of tools for visualizing and diagnosing performance of their network models.
-Though we apply the tools to CTMCs because we find the models intuitive, the methods we use can be applied to any network model, static or dynamic.% do something awesome :)
-
-In the remainder of this section, we provide an overview of the structure of network data and discuss two common network visualization (net-vis) techniques.
-Then, we detail the hierarchical model structure of a CTMC following the definition in \citet{saompaper} and ouline the CTMC model fitting process in \texttt{RSiena} \citep{SienaAlgs}.
-
-\subsection{Networks and their Visualizations} \label{sec:netintro}
+Though we apply the tools to CTMCs because we find the models intuitive, the methods we use can be applied to any network model, static or dynamic.
+
+The rest of this paper is structured as follows.
+In Section~\ref{sec:netintro}, we provide an overview of network data and visualization and give a more detailed description of CTMC models and how we fit them with \texttt{RSiena} \citep{RSiena}.
+In Section \ref{sec:exdatamods}, we introduce the motivating data and CTMC models we assess using model-vis.
+Then, in Section~\ref{sec:minds}, we explore the first step in model-vis by placing the models in the data space to determine how well the models fit the data.
+Section~\ref{sec:collectmodels} follows, in which we visualize collections of models to explore their behavior.
+Finally, in Section~\ref{sec:algorithms}, we visualize the underlying algorithms in the CTMC fitting process to gain insight into the fitting process.
+We conclude with a discussion in Section~\ref{sec:discussion}.
+
+\section{Network Models and Visualizations} \label{sec:netintro}
 
@@
 
-%\noindent \textbf{Network Data Structure:} 
 Network data across fields have similar structure: they consist of units of observation and connections between those units.
 In a social network, the units of observations, called \textit{nodes} or \textit{actors}, are people, while the connections, called \textit{edges} or \textit{ties}, are their relationships with each other.
-%Networks change over time, like when new relationships are formed in a social network.
 Other network data include inherent variables of interest on the nodes, such as age or gender, and the type of relationship (friends, coworkers, etc) that an edge represents.
@@
 The lack of a native coordinate system has given rise two primary net-vis methods: the node-link diagram and the adjacency matrix visualization \citep{adjmatpro}.
-We demonstrate these with a simpe example.
+We demonstrate these with a simple example.
 Consider the network with five nodes, $\{1,2,3,4,5\}$, connected by five directed edges:  $\{2 \to 4, 3 \to 4, 1 \to 5, 3 \to 5, 5 \to 4\}$ shown in Figure~\ref{fig:toyplots}.
@@
 For a full discussion of available node-link layout algorithms, we direct the reader to \citet{drawingalgs}.
-%Other algorithms use multi-dimensional scaling or properties of the adjacency matrix (e.g.
-its eigenstructure), to place the nodes \citep{drawingalgs}.
 In Figure~\ref{fig:toyplots}, we use the KK algorithm, and unless otherwise stated, all other node-link diagrams in this paper are laid out using the KK algorithm \citep{kamadakawai}.
 
-\par The second method for net-vis uses the adjacency matrix of the network. The adjaceny matrix of a network, $\mathbf{A}(x)$ is defined as:
+\par The second method for net-vis uses the adjacency matrix of the network. The adjacency matrix of a network, $\mathbf{A}(x)$ is defined as:
 \begin{equation}\label{eq:adjmat}
@@
 \end{equation}
-%In this paper, we only consider the presence or absence of an edge, but if the network has weighted edges, $A_{ij}$ is the weight of the edge from $i \to j$. 
 The adjacency matrix visualization for our toy example is also shown in Figure~\ref{fig:toyplots}.
@@
 An adjacency matrix visualization can, however, be useful when the network is very complex, dense, or large \citep{adjmatviz}.
-% Experimental studies have shown adjacency matrix visualization to be superior to node-link diagrams for large networks.
-For example, for basic perceptual tasks on networks, including node and edge count, adjacency matrix visualizations outperform node-link diagrams as the size and density of the network increases \citep{adjmatviz}.
-One drawback of the adjacency matrix visualization that \citeauthor{adjmatviz} found was that edges are overrepresented for undirected graphs, due to the symmetry of $\mathbf{A}$: the edge $i \rightarrow j$ for $i \neq j$ appears in $\mathbf{A}$ twice: in $A_{ij}$ and $A_{ji}$, and so it also appears twice in the adjacency matrix visualization.
-This is an advantage, however, for \emph{directed} graphs, where exactly the correct number of edges is represented in a matrix visualization.
-A node-link diagram, however, may underrepresent the edge count if the edges $x_{ij}$ and $x_{ji}$ both exist and are drawn on top of one another.
 Ultimately, there is no one ``correct" way to visualize network information, and we will be using both the node-link and adjacency matrix net-vis methods throughout this paper to visualize our example data and CTMC models.
@@
 
-The continuous-time Markov chain (CTMC) model family is for analyzing dynamic networks, so we can use them to study the change of a network in time while also examining the structural and node covariate efects.
+The continuous-time Markov chain (CTMC) model family is for analyzing dynamic networks, so we can use them to study the change of a network in time while also examining the structural and node covariate effects.
 To define CTMCs for our purposes, we follow the general hierarchical definition outlined in \citet{saomreview} and elsewhere (see e.g.\ \citet{snijders01} and \citet{snijdersetal2010}).
-CTMC models reflect our intution about social networks: most social networks, even holding constant the set of actors over time, are ever-changing as relationships form or dissolve.
+CTMC models reflect our intuition about social networks: most social networks, even holding constant the set of actors over time, are ever-changing as relationships form or dissolve.
 Actors in social networks also have properties that could change their role within the network, and existing ties may affect formation or dissolution of other ties.
-In this section, we define the data structure, the hierarchal model, and intensity matrix of the CTMC.
+In this section, we define the data structure, the hierarchical model, and intensity matrix of the CTMC.
 
@@
 The parameters of the model, $\beta_1, \dots, \beta_K$ correspond to network statistics, $s_{ik}(x, \mathbf{Z})$, for $k = 1,\dots, K$.
-To maximize $f_i$, $i$ will either destroy a tie, create a tie, or make no change.% if all changes will result in lower $f_i$.
+To maximize $f_i$, $i$ will either destroy a tie, create a tie, or make no change.
 
@@
 p_{ij}(\boldsymbol{\beta}, x, \mathbf{Z}) = \dfrac{\exp\left\{f_i(\boldsymbol{\beta}, x(i\leadsto j), \mathbf{Z})\right\}}{\sum_{h \neq i} \exp\left\{f_i(\boldsymbol{\beta}, x(i \leadsto h), \mathbf{Z})\right\}}
-\end{equation}%When $i = j$, the numerator represents the exponential of the value of the objective function when evaluated at the current network state. 
-%When the value of the objective function is high at the current state, the probability of not making a change is also high.
+\end{equation}
 We explore these probabilities further in Section~\ref{sec:algorithms}.
@@
 Thus, the intensity matrix $\mathbf{Q}$ is very sparse, with only $n(n-1) + 1$ non-zero entries in each row.
-Note that $n(n-1)$ of these entries represent the possible states that are one edge different from a given state, while the additional non-zero entry is the diagonal entry for the current state.% All other entries in a row are structural zeroes because those network states cannot be reached from the current state in a single change.
+Note that $n(n-1)$ of these entries represent the possible states that are one edge different from a given state, while the additional non-zero entry is the diagonal entry for the current state.
 
@@
 The intensity matrix $\mathbf{Q}$ fully defines the CTMC model for network change.
-%, and is the foundation required for parameter estimation.
-%Estimating the parameters in these models is difficult, but thanks to the SIENA software \citep{RSiena}, we have an accessible way to fit SAOMs to network observations.
 
@@
 In Section~\ref{sec:algorithms}, we further explore these phases in the SIENA MoM algorithm through visualization.
-For a more detailed description of the fitting process, we direct the reader to \citet{RSienamanual, SienaAlgs}%, bringing them out of their ``black box" and into the light.
-
-The rest of this paper is strucured as follows.
-In Section \ref{sec:exdatamods}, we introduce the motivating data and CTMC models we assess using model-vis.
-Then, in Section~\ref{sec:minds}, we explore the first step in model-vis by placing the models in the data space to determine how well the models fit the data.
-Section~\ref{sec:collectmodels} follows, in which we visualize collections of models to explore their behavior.
-Finally, in Section~\ref{sec:algorithms}, we visualize the underlying algorithms in the CTMC fitting process to gain insight into the fitting process.
-We conclude with a discussion in Section~\ref{sec:discussion}.
+For a more detailed description of the fitting process, we direct the reader to \citet{RSienamanual, SienaAlgs}.
 
@@
 
-\noindent \textbf{Example Data:} To guide our visual exploration of CTMC models for dynamic networks, we use a canonical example in the social network literature: a subset of the 50 actor dataset from the ``Teenage Friends and Lifestyle Study"  \citep{friendsdata}.
+\noindent \textbf{Example Data:} To guide our visual exploration of CTMC models for dynamic networks, we use a canonical example in the social network literature: a subset of the 50 actor data set from the ``Teenage Friends and Lifestyle Study"  \citep{friendsdata}.
 We chose this data because it is a well-known and easy to understand example.
@@
 We chose to only work with a subset of the data to make node-link diagrams easier to see and to make any changes in the network more noticeable.
-\hh{HH: ground the figure - this is the first time you are showing this example. A node is ... an edge is ... , colour is ...}
-We visualize this data with a node-link diagram in Figure~\ref{fig:alldataviz}, where %For model fitting, we condition on wave 1 and estimate the parameters of the models for the second and third waves.
-nodes are colored according to the girl's reported drinking behavior: does not drink, drinks once or twice a year, drinks once a month, and drinks once a week.
+In the small data set, there are 16 teenage girls sampled over three years, who were asked about their friends and their behaviors, such as how often they smoke cigarettes or drink alcohol  \citep{friendsdata}.
+Each girl was asked to identify her friends in the school to form a directed network.
+We visualize this data with a node-link diagram in Figure~\ref{fig:alldataviz}.
+Each girl is represented by a node and an arrow is drawn from her to each of the friends she has reported.
+Sometimes, two girls identify each other as friends, as shown by nodes 8 and 9 in the first sample on the left in Figure~\ref{fig:alldataviz}.
+Often, the ties are not reciprocated, as shown by the edges from node 5 to nodes 2 and 3 in the first sample.
+%\hh{HH: ground the figure - this is the first time you are showing this example. A node is ... an edge is ... , colour is ...}
+We also color nodes in Figure~\ref{fig:alldataviz} according to each girl's reported drinking behavior: does not drink, drinks once or twice a year, drinks once a month, and drinks once a week.
 We can see that, as time passes, the students seem to drink more and become increasingly isolated into smaller groups.
 An analysis of this type of data with a CTMC model should capture these dynamics in a way that allows the researcher to draw conclusions about the nature of the network and the behavioral forces at play.
-
-%The second data example we use is a much more recent dataset on collaboration in the United States Senate during the $111^{th}$ through $114^{th}$ Congresses, shown in Figure~\ref{fig:senateplot}.
-These sessions of congress correspond to the years of Barack Obama's presidency, from 2009-2016.
-(Details of how this data can be downloaded are provided by François Briatte at \url{https://github.com/briatte/congress}.) In this network, ties are directed from senator $i$ to senator $j$ when senator $i$ cosponsors the bill that senator $j$ authored.
-There are hundreds of ties between senators when they are connected in this way, so we simplify the network by computing a single value for each pair of senators called the \textit{weighted propensity to cosponsor} (WPC).
-This value is defined in \citet{senate} as the total number of times senator $i$ cosponsors senator $j$'s bill, weighted by the number of total cosponsors:
-% \begin{equation}\label{eq:sen1}
-% WPC_{ij} = \dfrac{\sum\limits_{k=1}^{n_j} \frac{Y_{ij(k)}}{c_{j(k)}}}{\sum\limits_{k=1}^{n_j} \frac{1}{c_{j(k)}}}
-% \end{equation}
-%where $n_j$ is the number of bills in a congressional session authored by senator $j$, $c_{j(k)}$ is the number of cosponsors on senator $j$'s $k^{th}$ bill, where $k \in \{1,\dots, n_j\}$, and $Y_{ij(k)}$ is a binary variable that is 1 if senator $i$ cosponsored senator $j$'s $k^{th}$ bill, and is 0 otherwise.
-This measure ranges in value from 0 to 1, where $WPC_{ij} = 1$ if senator $i$ is a cosponsor on every one of senator $j$'s bills and $WPC_{ij} = 0$ if senator $i$ is never a cosponsor any of senator $j$'s bills.
-<<senateplot, fig.cap='The collaboration network in the four senates during the Obama years, 2009-2016.
-Edges are shown only if the WPC between two senators is greater than 0.25.
-We use the FR layout algorithm here.', fig.width=8, fig.height=8, out.width='\\textwidth', eval = FALSE>>=
-seobama <- read_csv("data/congress/senateobamapres_gsw_25.csv")
-to_label <- c("Joseph R. Biden Jr.", "John S. McCain", "Ted Cruz","Marco Rubio",
-  "Lindsey O. Graham","Rand Paul", "Bernard Sanders", "Jim Webb", "Mitch McConnell",
-  "Harry M. Reid", "Hillary Rodham Clinton", "Amy Jean Klobuchar")
-seobama$label <- as.factor(ifelse(seobama$source %in% to_label, seobama$source, ""))
-levels(seobama$label) <- c("", "Amy Klobuchar", "Bernie Sanders", "Harry Reid",
-                           "Hillary Clinton", "Jim Webb", "John McCain", "Joe Biden",
-                           "Lindsey Graham", "Marco Rubio", "Mitch McConnell",
-                           "Rand Paul", "Ted Cruz")
-set.seed(56049382)
-ggplot(data = seobama) +
-  geom_net(directed = T, labelon=T, arrowsize = .25, singletons= T, fiteach = T, linewidth = .25, layout.alg = 'fruchtermanreingold', fontsize = 3, repel=T,
-           aes(from_id = source, to_id = target, color = party, label = label),
-           size = 1) +
-  ThemeNet +
-  scale_color_manual(values = c("black", "grey30","#ef3b2c")) +
-  theme(legend.position = 'bottom', strip.text = element_text(size = 20)) +
-  facet_wrap(~senate, nrow = 2, labeller = "label_both") +
-  xlim(c(0,1.05)) +
-  ylim(c(0,1.05))
-@
-% Because CTMC models require binary edges, we focus only on strong collaborations, where $WPC$ is high. For a senate collaboration network $x$, edges are defined as
-% \begin{equation}\label{eq:wpc2}
-%   x_{ij} =
-%   \begin{cases}
-%                                    1 &  \text{if } WPC_{ij} > 0.25 \\
-%                                    0 & \text{if } WPC_{ij} \leq 0.25.
-%   \end{cases}
-% \end{equation}
-% The node-link diagram for the four senates during the Obama administration are shown in Figure~\ref{fig:senateplot}.
-We can see that Senate 111 is much more densely connected than the other three, which have one large, sparsely connected component with a few smaller connected groups throughout.
-Some prominent names are shown to demonstrate their place in the network.
-As with the friendship example, analyzing the senate data with a CTMC model should capture the underlying dynamics, giving insight into the impact that the actor covairates have on the senate's collaboration structure.
-By using data sets from very different contexts, we hope to assess some differences in model performance.
 
@@
 @
-\caption{\label{fig:dad}Demonstracting a N22 between actors $i$ and $j$.
+\caption{\label{fig:dad}Demonstrating a N22 between actors $i$ and $j$.
 The dotted tie is encouraged if $\beta_4$ has a positive value.
@@
 \noindent where $s_{i3}(x,z)$ and $s_{i4}(x)$ are defined in Figure~\ref{fig:structures}.
-These statistics are known as the \emph{number of jumping transitive triplets} (JTT) and the \emph{number of doubly achieved distances two effect} \hh{HH do you need the `two` in there?} (N22), respectively.
+These statistics are called the "number of jumping transitive triplets" (JTT) and the"number of doubly achieved distances two effect" %\hh{HH do you need the `two` in there?
+ST: Yes its in the documentation}
+(N22), respectively in \citet{RSienamanual}.
 The JTT effect emphasizes triads formed between actors with \emph{different} covariate values, while the N22 effect emphasizes \emph{indirect} ties between actors.
-%The covariate for the friendship data is the drinking behavior, and is the number of bills authored for the senate data.
 All model effects are visualized in Figures~\ref{fig:beta1}-\ref{fig:dad}, where the ties affected by the parameter value are represented by dotted lines.
@@
 
-\hh{HH: include the names of the effects in the table -- otherwise it's really hard to make any sense of the numbers. }
+%\hh{HH: include the names of the effects in the table -- otherwise it's really hard to make any sense of the numbers. ST: done. }
 % table of fitted M1 M2 M3
@@
 \centering
-\begin{tabular}{lccc}
+\begin{tabular}{llccc}
   \toprule
-  & M1 & M2 & M3 \\
+  Effect Name & Parameter & M1 & M2 & M3 \\
   \midrule
-  $\alpha_1$ & \Sexpr{SFE$M1[1]} & \Sexpr{SFE$M2[1]} & \Sexpr{SFE$M3[1]}  \\
-  $\alpha_2$ & \Sexpr{SFE$M1[2]} & \Sexpr{SFE$M2[2]} & \Sexpr{SFE$M3[2]}   \\
-  \midrule
-  $\beta_1$ & \Sexpr{SFE$M1[3]} & \Sexpr{SFE$M2[3]} & \Sexpr{SFE$M3[3]}  \\
-  $\beta_2$ & \Sexpr{SFE$M1[4]} & \Sexpr{SFE$M2[4]} & \Sexpr{SFE$M3[4]} \\
-$\beta_3$ & -- & \Sexpr{SFE$M2[5]} & --  \\
-  $\beta_4$ & -- & --  & \Sexpr{SFE$M3[6]} \\
+  Rate (time 1 $\rightarrow$ 2) & $\alpha_1$ & \Sexpr{SFE$M1[1]} & \Sexpr{SFE$M2[1]} & \Sexpr{SFE$M3[1]}  \\
+   Rate (time 2 $\rightarrow$ 3) & $\alpha_2$ & \Sexpr{SFE$M1[2]} & \Sexpr{SFE$M2[2]} & \Sexpr{SFE$M3[2]}   \\
+  Outdegree & $\beta_1$ & \Sexpr{SFE$M1[3]} & \Sexpr{SFE$M2[3]} & \Sexpr{SFE$M3[3]}  \\
+  Reciprocity & $\beta_2$ & \Sexpr{SFE$M1[4]} & \Sexpr{SFE$M2[4]} & \Sexpr{SFE$M3[4]} \\
+  JTT & $\beta_3$ & -- & \Sexpr{SFE$M2[5]} & --  \\
+  N22 & $\beta_4$ & -- & --  & \Sexpr{SFE$M3[6]} \\
   \bottomrule
@@
 
-\hh{The values in the table show us that all effects are very significant and relatively stable across the different models. }
-\hh{HH: sentence of discussion - do we see the expected increase/decrease in the parameters?}
-\hh{What we do not learn from the values in the table is anything about the goodness of fit of each of the models, or dependencies between the parameters and their estimates. }
+The values in Table~\ref{tab:meanests} show us that all effects are very significant and relatively stable across the different models.
+\hh{HH: sentence of discussion - do we see the expected increase/decrease in the parameters?} \st{I do not understand.}
+What we do not learn from the values in Table~\ref{tab:meanests} is anything about the goodness of fit of each of the models, or dependencies between the parameters and their estimates.
 
 %We encourage interested readers to consult the \texttt{RSiena} manual for a complete definition and description of these and all other possible effects available to use in CTMC models \citep{RSienamanual}.
-\hh{HH The next paragraph could be shortened some more to emphasize what we (want to) do.}
+% \hh{HH The next paragraph could be shortened some more to emphasize what we (want to) do. ST Done. I think!}
 Our goal here is to assess the fit of these models to our data, not to find the best model for this data, which is why we have kept both the data and the models very small.
-A true analysis of these data would include all of the available data, including additional covariates related to smoking and other behaviors, and would use sociological research to select and test the most pertinent effects at each step in the model hierarchy.
-The exact model structure and effects we choose are not as important to us as our ability to assess their impact on model fit.
+A true analysis of these data would include all of the available data, including additional covariates related to smoking and other behaviors.
+The exact model structure and effects we choose are not as important to us as our goal of assessing their impact on model fit.
 We keep the data and model simple in order to apply the more complicated model visualization tools that we discuss in Sections~\ref{sec:minds}-\ref{sec:algorithms}.
@@
 The package draws the network using node-link diagrams, and the network's node or edge data are then mapped to different visual features.
-The color, size, and shape of the points can be used to represent variables in the node data, while the color, linewidth, and linetype of the edges between nodes can be used to represent variables in the edge data.
+The color, size, and shape of the points can be used to represent variables in the node data, while the color, linewidth, and line type of the edges between nodes can be used to represent variables in the edge data.
 Finally, the networks observed at each time point are placed side-by-side, showing the network's evolution in time.
@@
 
-\subsection{\hh{Expectation Networks}}
-<<funsenatevis, fig.height=3, fig.cap='The 111$^{th}$ Senate at two times: while Clinton was in the senate in 2009 (left) and after she departed (right).
-We map sex, party, and bills authored to the shape, color, and size of the nodes, respectively.
-We also map $WPC$ to the edge width.', eval = FALSE>>=
-se111clint <- filter(seobama, senate == 111)
-se111noclint <- filter(seobama, senate == 111)
-se111noclint$target[which(se111noclint$target == "Hillary Rodham Clinton")] <- NA
-se111clint$Clinton <- 'Yes'
-se111noclint$Clinton <- 'No'
-clintonSenate <- rbind(se111clint, se111noclint)
-clintonSenate$Clinton <- as.factor(clintonSenate$Clinton)
-clintonSenate$Clinton <- ordered(clintonSenate$Clinton, levels = c("Yes", "No"))
-
-set.seed(12981120)
-clintonSenate %>% filter(!(source %in% c("Roland Burris", "Bernard Sanders", "Frank R. Lautenberg", "Mary L. Landrieu") & is.na(target))) %>%
-ggplot() +
-  geom_net(directed = T, labelon=F, arrowsize = .3, singletons= F, fiteach = T, arrowgap = .01, layout.alg = 'fruchtermanreingold', ecolour = "grey70",
-           aes(from_id = source, to_id = target, color = party, linewidth = gsw, shape = sex,size = n_au/30)) +
-  scale_size_continuous(name="# bills\nauthored" , guide = guide_legend(nrow= 2), breaks = c(0,1,2,3), labels = c(0,30, 60, 90), range = c(.5,3)) +
-  ThemeNet +
-  scale_shape_manual(values = c(17,16), guide = guide_legend(nrow= 2)) +
-  scale_color_manual(values = c("black", "grey80","#ef3b2c"), guide = guide_legend(nrow=2)) +
-  facet_wrap(~Clinton, nrow = 1, labeller = 'label_both')  +
-  theme(legend.position = "bottom")
-@
+\subsection{Expectation Networks}
 
@@
 legendlines <- cowplot::get_legend(forlineleg)
-set.seed(35291)
+#set.seed(1356)
+#set.seed(5)
+set.seed(10) # this was my favorite seed
 allp <- ggplot(data = combinedavgs) +
   geom_net(aes(from_id = from, to_id = to, linewidth= linewidth),
-           directed = T, curvature = .2, labelon=T, size = 3,
-           arrow = arrow(angle = 20, type = "open", length = unit(.2, "cm")), ecolour = "#969696", fiteach = T,
-           color = "#252525", hjust = .5, vjust = .5, labelcolour = 'white', fontsize = 2, arrowgap = .02,
+           directed = T, curvature = .2, labelon=T, size = 3, layout.alg = "fruchtermanreingold", 
+           arrow = arrow(angle = 20, type = "open", length = unit(.2, "cm")), ecolour = "black", fiteach = F,
+           color = "grey60", hjust = .5, vjust = .5, labelcolour = 'white', fontsize = 2, arrowgap = .02,
            singletons = T) +
@@
 
-\hh{HH: we inconvenient question - could we use the same layout for all the networks (by using facetting instead of drawing individual plots).
+% \hh{HH: we inconvenient question - could we use the same layout for all the networks (by using facetting instead of drawing individual plots).
 That would highlight the structure and changes to the structure better.
@@
 Something like (code in comment):
-% ggplot(data = combinedavgs) +
-%   geom_net(aes(from_id = from, to_id = to, linewidth= linewidth/3),
-%            directed = T, curvature = .2, labelon=T, size = 3,
-%            arrow = arrow(angle = 20, type = "open", length = unit(.2, "cm")), ecolour = "black",
-%            color = "grey60", hjust = .5, vjust = .5, labelcolour = 'white', fontsize = 2, arrowgap = .03,
-%            singletons = T) +
-%   labs(title = "Time 1") + ThemeNet +
-%   theme(plot.background = element_rect(color = 'black')) + facet_wrap(~cat)
-Hopefully this helps with the description below, because it's quite hard to read with all the numbers of nodes.
-}
+% % ggplot(data = combinedavgs) +
+% %   geom_net(aes(from_id = from, to_id = to, linewidth= linewidth/3),
+% %            directed = T, curvature = .2, labelon=T, size = 3,
+% %            arrow = arrow(angle = 20, type = "open", length = unit(.2, "cm")), ecolour = "black",
+% %            color = "grey60", hjust = .5, vjust = .5, labelcolour = 'white', fontsize = 2, arrowgap = .03,
+% %            singletons = T) +
+% %   labs(title = "Time 1") + ThemeNet +
+% %   theme(plot.background = element_rect(color = 'black')) + facet_wrap(~cat)
+% Hopefully this helps with the description below, because it's quite hard to read with all the numbers of nodes.
+% }
 
@@
 Edges that appear with higher frequency are drawn with thicker lines.
-\hh{HH Give a definition of how edge frequency is defined in the plot - frequency of 1 means that this edge was present in all of the simulations, right?}
+An edge frequency value of one indicates that the edge was present in all simulations used to create the expected network.
+%\hh{HH Give a definition of how edge frequency is defined in the plot - frequency of 1 means that this edge was present in all of the simulations, right?ST DONE.
+}
 On the bottom row of Figure~\ref{fig:averagenet}, we show the data for comparison: $x(t_1)$ on the left, and $x(t_2)$ on the right.
@@
 
-\subsection{\hh{Difference Networks}}
+\subsection{Difference Networks}
 
@@
 We could also model the behavior changes, not just the network changes, if we believe that friends influence behavior via peer pressure.
-And as before, this method of gathering a collection of models can be used for any network model, such as the hierarchichal Bayesian network estimation model in the \texttt{R} package \texttt{sna}, which allows the user to vary hyperparameters in the prior and model parameters \citep{sna}.
+And as before, this method of gathering a collection of models can be used for any network model, such as the hierarchical Bayesian network estimation model in the \texttt{R} package \texttt{sna}, which allows the user to vary hyperparameters in the prior and model parameters \citep{sna}.
 % took out this next bit just because it doesn't really add much and we can just put the appendix figure into the previous section. % add more later.
@@
 @
-\caption{\label{fig:corplots} Relationships between parameters common to all models. Densities and scatterplots are colored by model. \nocite{ggally}}
+\caption{\label{fig:corplots} Relationships between parameters common to all models. Densities and scatter plots are colored by model. \nocite{ggally}}
 \end{subfigure}
@@
 In fact, the distributions of $\alpha_1$ and $\beta_1$ for M2 do not overlap at all with the distribution of the same parameters from model M1.
-When $\beta_3$ is included, its estimate is postitive, meaning that friendships between two girls with different drinking behaviors tend to form when there is third girl who is already friends with the other two.
+When $\beta_3$ is included, its estimate is positive, meaning that friendships between two girls with different drinking behaviors tend to form when there is third girl who is already friends with the other two.
 The inclusion of $\beta_3$ also leads to increases in the estimates of $\alpha_1, \alpha_2$.
@@
 This process is known as matrix seriation
-To seriate the matrix, we first construct a cumulative adjaceny matrix, $\mathbf{A}^{cum}$, for the series of microsteps in the CTMC simulated from $x(t_m)$ to $x(t_{m+1})$.
+To seriate the matrix, we first construct a cumulative adjacency matrix, $\mathbf{A}^{cum}$, for the series of microsteps in the CTMC simulated from $x(t_m)$ to $x(t_{m+1})$.
 A single entry in the cumulative adjacency matrix, $\mathbf{A}^{cum}_{ij}$, is the total number of times the edge $x_{ij}$ appears in the network from the initial observation, $x(t_m) \equiv X(0)$ to the last microstep, $X(R)$, where $R$ is the total number of microsteps taken:
@@
 
-We have only just begun to scratch the surface of these multi-layered models for social networks.
+We have only just begun to scratch the surface of these multilayered models for social networks.
 The \texttt{RSiena} software is incredibly powerful, and we can fit much more flexible hierarchical models than we have examined here.
@@
 As we have shown, the rate and objective function parameters are often highly correlated, so researchers can now use model-vis techniques to gain a better understanding of these correlations and hopefully find ways to account for their effects in the model.
-In addition, some extensions of the CTMC model can aslo model behavior change of the actors in the network, which we have just treated as covariates.
+In addition, some extensions of the CTMC model can also model behavior change of the actors in the network, which we have just treated as covariates.
 The CTMC family of models for dynamic networks has immense potential for additional model visualization applications.

### 2019-07-12 paper-child.Rnw
*mic drop*

@@
 
-Model visualization or model-vis complements traditional model diagnostic tools such as $R^2$ values and residual plots to enrich our understanding and interpretation of statistical models \citep{modelvis}.
+Model visualization or ``model-vis" complements traditional numerical model diagnostic tools to enrich our understanding and interpretation of statistical models \citep{modelvis}.
 We apply the three principles of model-vis: view the model in the data space, visualize collections of models, and explore the process of fitting the model, to the models and data introduced in Section~\ref{sec:exdatamods}.
@@
 Though we apply the tools to CTMCs because we find the models intuitive, the methods we use can be applied to any network model, static or dynamic.
+The visual diagnostics proposed here serve the same purpose for network models as residual plots for assessing goodness-of-fit in linear models.
 
@@
 For a full discussion of available node-link layout algorithms, we direct the reader to \citet{drawingalgs}.
-In Figure~\ref{fig:toyplots}, we use the KK algorithm, and unless otherwise stated, all other node-link diagrams in this paper are laid out using the KK algorithm \citep{kamadakawai}.
+In Figure~\ref{fig:toyplots}, we use the KK algorithm, and unless otherwise stated, all other node-link diagrams in this paper are laid out using the FR algorithm \citep{fruchtermanreingold}.
 
@@
 <<alldataviz, fig.cap="The small friendship network data we model throughout this paper.
-Each node represents a teenager, an edge is drawn if XXX.
-A Fruchterman-Reingold layout is used.
+Each node represents a teenager, an edge is drawn when a girl identifies another as her friend.
 By comparing the node-link diagrams we see  that teenagers' drinking behavior changes -- drinking generally increases; darker nodes indicate more occurrences -- and the network becomes less densely connected over time.
@@
 % \hh{HH The next paragraph could be shortened some more to emphasize what we (want to) do. ST Done. I think!}
-Our goal here is to assess the fit of these models to our data, not to find the best model for this data, which is why we have kept both the data and the models very small.
+Our goal here is to assess the fit of these models to the data, not to find the best model for this data, which is why we have kept both the data and the models very small.
 A true analysis of these data would include all of the available data, including additional covariates related to smoking and other behaviors.
@@
 
+% this chunk doesn't produce any results. see next chunk.
 <<summnetM1, message=FALSE,results='hide'>>=
@@
 Some edges appear in both the data and the expected network $\\bar{x}(t_2)$ (dotted lines), while others only appear in the data (solid lines) or only in the expectation (dashed lines).
-Width of non-data edges represent the proportion of appearances in the 1,000 simulations.
-The FR layout is used.">>=
+Width of non-data edges represent the proportion of appearances in the 1,000 simulations.">>=
 
@@
   geom_net(aes(from_id = from, to_id = to, linewidth= linewidth, linetype = color2),
-           directed = T, curvature = .2, labelon=T, size = 3,
+           directed = T, curvature = .2, labelon=T, size = 3, ecolour = "black",
            arrow = arrow(angle = 20, type = "open", length = unit(.15, "cm")), 
-           color = "#252525", hjust = .5, vjust = .5, labelcolour = 'white', fontsize = 2, arrowgap = .02, layout.alg = "fruchtermanreingold",
+           color = "gray60", hjust = .5, vjust = .5, labelcolour = 'white', fontsize = 2, arrowgap = .02, layout.alg = "fruchtermanreingold",
            singletons = T) +
@@
 
-\subsection{Difference Networks}
-
-Another way to view our models in the data space is with \textit{difference networks} between the expected network and the data, as shown in Figure~\ref{fig:summary2}.
+\subsection{Residual Networks}
+
+Another way to view our models in the data space is with \textit{residual networks} between the expected network and the data, as shown in Figure~\ref{fig:summary2}.
 We created these networks by computing the difference between $\bar{x}(t_2)$ and $x(t_1)$, $\mathbf{A}(\bar{x}(t_2)) - \mathbf{A}(x(t_1))$, and the difference between $\bar{x}(t_2)$ and $\bar{x}(t_2)$, $\mathbf{A}(x(t_2)) - \mathbf{A}(\bar{x}(t_2))$ for model M1.
-The lack of model fit is also apparent in Figure~\ref{fig:summary2}, where the dashed edges represent ties that only show up in the expected network, and the solid edges representing ties in the data that the model fails to capture.
+The lack of model fit is apparent in Figure~\ref{fig:summary2}, where the dashed edges represent ties that only show up in the expected network, and the solid edges representing ties in the data that the model fails to capture.
 Dotted edges are in both the expectation and the data and represent where the model is correct.
-On the left in Figure~\ref{fig:summary2}, we see the difference between $x(t_1)$ and the expected network.
+On the left in Figure~\ref{fig:summary2}, we see the residual network between $x(t_1)$ and the expected network.
 We see mostly dotted edges, meaning that the simulated networks consist primarily of edges in $x(t_1)$, plus some new edges represented by the dashed lines that are not in the data.
 The unobserved ties are all reciprocating ties in the data, so the model is predicting more reciprocity between actors than there actually is.
-On the right in Figure~\ref{fig:summary2}, we see the difference between $x(t_2)$ and the expected network.
+On the right in Figure~\ref{fig:summary2}, we see the residual network between $x(t_2)$ and the expected network.
 Because the model is simulating  $x(t_2)$, we would hope to see very few differences between the expected network and the data.
 However, there are six solid edges, which are in the data but do not appear at all in the expected network, and there are many dashed edges that appear in the expected network but are not in $x(t_2)$.
-Although there are some edges that appear in both the expected network and $x(t_2)$, the majority of edges in the expected network are not in the observed data, and there are several edges in the data missing from the expectation.
+Although the residual network shows that there some edges that appear in both the expected network and $x(t_2)$, the majority of edges in the expected network are not in the observed data, and there are several edges in the data missing from the expectation.
 We therefore conclude that the model M1 is a poor fit to our example data, which we have seen very clearly by viewing the model in the data space.
-As with the expected network, we can apply the difference network computation other network models to determine how closely the model fits the data.
+As with the expected network, we can apply the residual network visualization other network models to determine how closely the model fits the data.
 
@@
 @
-\caption{\label{fig:corplots} Relationships between parameters common to all models. Densities and scatter plots are colored by model. \nocite{ggally}}
+\caption{\label{fig:corplots} Diagnostic pairs plot. Relationships between parameters common to all models. Densities and scatter plots are colored by model. \nocite{ggally}}
 \end{subfigure}
@@
 The resulting means and standard deviations of the estimates are given in Table~\ref{tab:meanests}, and the distributions of all parameter estimates are shown in Figure~\ref{fig:distests}.
-In Figure~\ref{fig:corplots}, we can see that the addition of $\beta_3$ to the model results in very different estimates the of the other four parameters, $\alpha_1$, $\alpha_2$, $\beta_1$, and $\beta_2$.
-In fact, the distributions of $\alpha_1$ and $\beta_1$ for M2 do not overlap at all with the distribution of the same parameters from model M1.
+Figure~\ref{fig:corplots} shows a \textbf{diagnostic pairs plot}, in which we can see that the addition of $\beta_3$ to the model results in very different estimates the of the other four parameters, $\alpha_1$, $\alpha_2$, $\beta_1$, and $\beta_2$.
+In fact, the diagnostic pairs plot reveals that the distributions of $\alpha_1$ and $\beta_1$ for M2 do not overlap at all with the distribution of the same parameters from model M1.
 When $\beta_3$ is included, its estimate is positive, meaning that friendships between two girls with different drinking behaviors tend to form when there is third girl who is already friends with the other two.
@@
 
-\hh{It would make the discussion easier, if all of the main plots we introduce had a name - like expectation networks, difference (or residual?) networks, ...
+%\hh{It would make the discussion easier, if all of the main plots we introduce had a name - like expectation networks, difference (or residual?) networks, ...
 maybe 'parameter diagnostics' for the second approach?
@@
 
-\hh{We started with the three pronged approach of visualization ... let's bring it back to that.}
-\hh{Another point is a call for action: along the lines of now we need to use it!
-Visualization of network models should become second nature, similar to linear model diagnostics, we should have a suite of visualizations next to diagnostic values ready to launch. Ease of use?}
-
-We have used several visualization methods, some old and some new, to diagnose and understand network models in new ways.
-Though we have applied these methods to CTMC models for dynamic social networks, all of our visualizations can be applied to other dynamic network models and most can also be applied to any static network model.
-We gained insight into and appreciation for complicated and flexible network models by following the guiding principles of model visualization: viewing the model in the data space, visualizing collections of models, and looking at the underlying algorithms \citep{modelvis}.
-When we look at the model in the data space by viewing the expected network simulated from a model, we are able to easily see the lack of model fit.
-Computing and visualizing the difference network between the data and the expected network also helps assess model fit by showing us where the model is completely missing the data.
-When exploring collections of models, we learn more about the large number of parameters possible in network models, and we can also uncover relationships between these parameters.
-Finally, when we explore the algorithms fitting the models, we have a better understanding of the model and its stochastic behavior.
+%\hh{We started with the three pronged approach of visualization ... let's bring it back to that.}
+%\hh{Another point is a call for action: along the lines of now we need to use it!
+%Visualization of network models should become second nature, similar to linear model diagnostics, we should have a suite of visualizations next to diagnostic values ready to launch. Ease of use?}
+
+We have introduced diagnostic visualizations, the expected network, the residual network, the node-link animation, and the adjacency matrix animation, to complement existing numerical diagnostics for network models.
+These visual diagnostics give insights into the network model fit that purely numerical methods do not.
+%We have used several visualization methods, some old and some new, to diagnose and understand network models in new ways. 
+Though we have applied these methods to CTMC models for dynamic social networks, all of the diagnostics presented can be applied to other dynamic network models and most can also be applied to any static network model.
+By following the guiding principles of model visualization, viewing the model in the data space, visualizing collections of models, and looking at the underlying algorithms, we gained insight into and appreciation for a complicated and flexible family of network models \citep{modelvis}.
+The expected network and the residual network let us view the model in the data space, and we are then able to easily assess model goodness-of-fit.
+Visualizing the residual network between the data and the expected network also helps assess model fit by clearly showing us exactly where the model fits the network well and where it lacks fit.
+Exploring collections of models give us the opportunity to learn more about the large number of parameters possible in CTMC models.
+The diagnostic pairs plot visualizes collections of models giving us insight into the marginal distribution of parameters and their covariance structure.
+%and we also visualized relationships between parameters in the model that require further investigation.
+Finally, we used animation of node-link diagrams and the adjacency matrix visualizations to explore the underlying algorithms and expand our understanding of the model's stochastic behavior.
 All in all, this three-pronged model-vis attack has uncovered many properties of CTMC models for social networks and shows great potential for application to other network models.
 
-We have only just begun to scratch the surface of these multilayered models for social networks.
-The \texttt{RSiena} software is incredibly powerful, and we can fit much more flexible hierarchical models than we have examined here.
-If, for example, we think the network structure or an actor covariate affects the rate of change of the network, we can incorporate that belief into the rate function.
-In addition, more than one actor-level covariate can be included in the model, and many more than three parameters can be included in the objective function.
-As we have shown, the rate and objective function parameters are often highly correlated, so researchers can now use model-vis techniques to gain a better understanding of these correlations and hopefully find ways to account for their effects in the model.
+Using existing \texttt{R} packages, we have demonstrated how to apply model-vis to one family of models for social network analysis.
+The models and data we have used are multilayered and multidimensional, making them more complex than common statistical models, such as linear models.
+As models become more complicated, visualization is more important to model diagnosis because increasing complexity makes it harder to adequately summarize model fit only using numerical methods.
+Even more complex models, such as those for machine learning, can also benefit greatly from model visualization techniques without a lot of additional effort.
+The visualization methods we have used, such as the node-link diagram, are not new but we have used them in conjuction with model simulation to learn a great deal about our models of interest.
+
+We only explored a few simple models from the complicated and multilayered CTMC model family for social networks.
+%The \texttt{RSiena} software is incredibly powerful, and we can fit much more flexible hierarchical models than we have examined here.
+Many more than three parameters can be included in the objective function, and more than one actor-level covariate can be included at either level in the hierarchical model.
+We can also model the rate of change differently for each actor if, for example, we think an actor's covariate affects its rate of change.
+% <- reverse this section
+With increased model complexity, there is more potential for adverse effects, such as the correlation between parameters that we saw in Section~\ref{sec:collectmodels}.
+Armed with model-vis techniques, researchers can gain a better understanding of these and other adverse effects and find ways to alleviate them the model.
 In addition, some extensions of the CTMC model can also model behavior change of the actors in the network, which we have just treated as covariates.
@@
 
-Furthermore, the data we have used are quite small by network data standards, with only 16 nodes.
-Thus, some of the visualizations we have presented, such as the small node-link diagram animation, may not scale up to larger networks.
-Future network model-vis research could also include interactive and/or three-dimensional graphics, which we did not explore here.
+While the data we have used are quite small by network data standards, which favors node-link views, we also discuss adjacency matrix visualization, which scale up to large networks.
+%Thus, some of the visualizations we have presented, such as the small node-link diagram animation, may not scale up to larger networks.
+Additionally, interactive and/or three-dimensional graphics, which we did not explore here, could be used for larger networks.
 Interactivity is the state-of-the-art in data visualization, and network visualization is just starting to catch on \citep[see e.g.\ ][]{refinery}.
@@
 
-We hope that the principles of model-vis continue to be used for learning more about the fit and behavior of statistical models for network data.
+The principles of model-vis provide a solid foundation on how to use visualization to gain insight into more complex models, such as network models.
 Our model-vis application has uncovered many properties of CTMC models for dynamic social networks, and potential abounds for application of model-vis in other areas of network analysis.

### 2019-07-12 paper.Rnw
some edits
run spell check, incorporate some of hh's changes
*mic drop*

@@
 %                 Editing Commands
-%\newcommand{\hh}[1]{{\color{magenta} #1}}
-%\newcommand{\st}[1]{{\color{orange} #1}}
+\newcommand{\hh}[1]{{\color{magenta} #1}}
+\newcommand{\st}[1]{{\color{orange} #1}}
 
@@
 Statistical models for networks have been studied for decades, and they have gotten more complex as computing power has grown.
-Some network models, such as those for dynamic social networks, are analytically intractable and therefore require extensive simulation procedures for model fitting.
-In this paper, we use model visualization techniques to better understand some of these models and their behavior.
-We use a three-pronged visualization attack: (a) network visualizations to view the model, (b) visualizations of collections of models to interpret parameters, and (c) visualizations of fitting algorithms to gain insights into the underlying mechanisms.
-With the help of static and dynamic visualizations, we bring the hidden algorithmic fitting processes into the foreground, with the goal of a better understanding and higher accessibility of statistical models for social network analysts.
+Because of their complexity, traditional model diagnostic methods are unavailable.
+In this paper, we propose new methods for diagnosing network models.
+We use a three-pronged visualization attack: (a) network visualization of the fitted model, (b) visualizations of collections of models to reveal relationships between parameters, and (c) visualizations of algorithms to gain insights into the underlying network mechanisms.
+The visual diagnostics we propose serve the same purpose for network models as residual plots for assessing goodness-of-fit in linear models.%we bring the hidden algorithmic fitting processes into the foreground, with the goal of a better understanding and higher accessibility of statistical models for social network analysts.
 \end{abstract}

### 2019-07-15 paper-child.Rnw
ready to sub.

@@
 Some of the classical models, such as the Erd\"{o}s-R\'{e}nyi model and the small-world model of \citet{watts} are pseudo-dynamic because they could be applied in a dynamic way but rarely are \citep{goldenberg09}.
-Fully dynamic social network models, however, are more complicated: they incorporate both the network structure and its changes in time.
-If we can capture this change mechanism in time, we can truly begin to model the underlying network dynamics that drive social connections.
+Fully dynamic social network models, however, are more complicated: they consider both the network structure and its changes in time.
+If we can capture this change mechanism over time, we can truly begin to model the underlying network dynamics that drive social connections.
 A comprehensive review of dynamic network models is presented in \citet{goldenberg09}.
@@
 In each step of the hierarchy, many possible effects can be implemented to capture the change mechanism.
-For example,  the rate function could be a simple waiting-time model, or it could depend on a node's indegree or outdegree: we may have reason to believe that the more popular nodes change their ties more frequently.
+For example, the rate function could be a simple waiting-time model, or it could depend on a node's indegree or outdegree, if we have reason to believe that the more popular nodes change their ties more frequently.
 In addition, we can incorporate node covariate information into the hierarchy, such as gender or education level of people in a social network.
@@
 For example, \citet{depression} use a CTMC model to learn about the tendency of depressed adolescents to form friendships and found that depressed students tended to avoid creating new friendships over time.
-The CTMC model family can also be expanded to model node behavior, as in \citet{chyzh2016}, who apply CTMC models to show that more indirect trade relationships in a country decrease its overall respect for human rights.
+The CTMC model family can also be expanded to model node behavior, as in \citet{chyzh2016}, who applies CTMC models to show that more indirect trade relationships in a country decrease its overall respect for human rights.
 Thus, with CTMC models we can examine assumptions made about social networks.
@@
 Section~\ref{sec:collectmodels} follows, in which we visualize collections of models to explore their behavior.
-Finally, in Section~\ref{sec:algorithms}, we visualize the underlying algorithms in the CTMC fitting process to gain insight into the fitting process.
+Finally, in Section~\ref{sec:algorithms}, we visualize the underlying algorithms in the CTMC fitting process to gain insight into model fits.
 We conclude with a discussion in Section~\ref{sec:discussion}.
@@
 Other network data include inherent variables of interest on the nodes, such as age or gender, and the type of relationship (friends, coworkers, etc) that an edge represents.
-Dynamic networks are observed at many discrete points in time, where each observation of a network is often called a ``wave" of data.
+Dynamic networks are observed at two or more discrete points in time, where each observation of a network is often called a ``wave" of data.
 %These two network data structures pose unique problems to analysts, such as, ``How does the strength of a tie between two nodes affect the overall structure of the network?" or ``Do differences in node variables affect the formation of edges?" We propose to answer these questions and more through visual exploration of network data and models.
@@
 Most data visualizations rely on well-defined axes inherited from the data, such as Cartesian coordinates or spatial locations, but network data typically do not have such built-in structure.
+The lack of a native coordinate system has given rise two primary network visualization or ``net-vis" methods: the node-link diagram and the adjacency matrix visualization \citep{adjmatpro}.
+We demonstrate these with a simple example.
+Consider the network with five nodes, $\{1,2,3,4,5\}$, connected by five directed edges:  $\{2 \to 4, 3 \to 4, 1 \to 5, 3 \to 5, 5 \to 4\}$ shown in Figure~\ref{fig:toyplots}.
 
@@
 
-The lack of a native coordinate system has given rise two primary net-vis methods: the node-link diagram and the adjacency matrix visualization \citep{adjmatpro}.
-We demonstrate these with a simple example.
-Consider the network with five nodes, $\{1,2,3,4,5\}$, connected by five directed edges:  $\{2 \to 4, 3 \to 4, 1 \to 5, 3 \to 5, 5 \to 4\}$ shown in Figure~\ref{fig:toyplots}.
-
 The first net-vis method, the node-link diagram, represents nodes with points in two dimensions and then represents edges by connecting the points with lines.
-These lines can also have arrows on them indicating the direction of the edge for directed networks, as shown in Figure~\ref{fig:toyplots}.
-Because there is often no natural placement of the nodes, they are placed in 2D at random, then adjusted with a layout algorithm.
+These lines can also have arrows on them indicating the direction of the edge for directed networks, as shown at left in Figure~\ref{fig:toyplots}.
+Because there is often no natural placement of the nodes, they are placed in 2D space at random, then adjusted with a layout algorithm.
 Some commonly used algorithms, such as the Kamada-Kawai (KK) layout \citep{kamadakawai} and the Fruchterman-Reingold (FR) layout \citep{fruchterman-reingold}, are designed to mimic physical systems, drawing the graphs based on the ``forces" connecting them.
 For a full discussion of available node-link layout algorithms, we direct the reader to \citet{drawingalgs}.
-In Figure~\ref{fig:toyplots}, we use the KK algorithm, and unless otherwise stated, all other node-link diagrams in this paper are laid out using the FR algorithm \citep{fruchtermanreingold}.
-
-\par The second method for net-vis uses the adjacency matrix of the network. The adjacency matrix of a network, $\mathbf{A}(x)$ is defined as:
+In Figure~\ref{fig:toyplots}, we use the KK algorithm, and unless otherwise stated, all other node-link diagrams in this paper are laid out using the FR algorithm \citep{fruchterman-reingold}.
+
+\par The second method for net-vis uses the adjacency matrix of the network. The adjacency matrix of network $x$, $\mathbf{A}(x)$, is defined as:
 \begin{equation}\label{eq:adjmat}
@@
 The adjacency matrix visualization for our toy example is also shown in Figure~\ref{fig:toyplots}.
-This visualization shows the network as a grid, with each row representing the ``from" or ``ego" node and each column representing the ``to" or ``alter" node of an edge.
+This visualization shows the network as a grid, with each row representing the ``from" node and each column representing the ``to" node of an edge.
 The grid space for $x_{ij}$ (the edge $i\rightarrow j$) is filled in if $A_{ij} = 1$ and is empty if $A_{ij} = 0$.
@@
 The continuous-time Markov chain (CTMC) model family is for analyzing dynamic networks, so we can use them to study the change of a network in time while also examining the structural and node covariate effects.
-To define CTMCs for our purposes, we follow the general hierarchical definition outlined in \citet{saomreview} and elsewhere (see e.g.\ \citet{snijders01} and \citet{snijdersetal2010}).
+To define CTMCs for our purposes, we follow the general hierarchical definition outlined in \citet{saomreview} and elsewhere (see e.g.\ \citep{snijders01} and \citep{snijdersetal2010}).
 CTMC models reflect our intuition about social networks: most social networks, even holding constant the set of actors over time, are ever-changing as relationships form or dissolve.
@@
 A dynamic network consists of a fixed set of $n$ nodes that is changing over time, and is observed at $M$ discrete time points, $t_1, \dots, t_M$ with $t_1 < t_2 < \dots < t_M$.
-We denote the network observation at timepoint $t_m$ by $x(t_m)$ for $m = 1, \dots M$.
+We denote the network observated at timepoint $t_m$ by $x(t_m)$ for $m = 1, \dots, M$.
 The model assumes that this longitudinal network of discrete observations is embedded within a CTMC, which we denote by $X(T)$.
@@
 Unlike the first and last observations of the network, $x(t_2), \dots, x(t_{M-1})$ do not have direct correspondence with steps in the CTMC.
-Thus, the observations $x(t_2), \dots, x(t_{M-1})$ treated as ``snapshots" of the network at some point between two steps in the CTMC.
+Thus, the observations $x(t_2), \dots, x(t_{M-1})$ are treated as ``snapshots" of the network at some point between two steps in the CTMC.
 
 The CTMC process $X(T)$ is a series of single tie changes that happen according to the model's \textit{rate function}, the first piece of the hierarchy.
-One actor at a time is ``selected" to make a change, then in the second piece of the hierarchy, it randomly changes its ties according to probabilities derived from the \textit{objective function}.
-
-\noindent \textbf{Rate Function:} In general, the \emph{rate function}, $\rho(x, \mathbf{Z}, \boldsymbol{\alpha})$, is a function of the network at its current state $x$, covariates of interest, $\mathbf{Z}$, and some parameters, $\boldsymbol{\alpha}$.
+One actor at a time is ``selected" to make a change, then in the second piece of the hierarchy the selected node randomly changes its ties according to probabilities derived from the \textit{objective function}.
+
+\noindent \textbf{Rate Function:} In general, the \emph{rate function}, $\rho(x, \mathbf{Z}, \boldsymbol{\alpha})$, is a function of the network at its current state, $x$, covariates of interest, $\mathbf{Z}$, and some parameters, $\boldsymbol{\alpha}$.
 With this general formulation, each node can have a different rate of change.
@@
 With this simple rate function, we have $M-1$ rate parameters, one for each transition from $x(t_{m})$ to $x(t_{m+1})$ for $m = 1, \dots, M-1$.
-The rate parameter, $\alpha_m$ dictates how often an actor $i$ is selected to change one of its ties, $x_{ij}$, for $j \neq i \in \{1, \dots, n\}$ in the time period from $t_{m}$ to $t_{m+1}$ for $m = 1, \dots, M-1$.
+The rate parameter, $\alpha_m$, dictates how often an actor $i$ is selected to change one of its ties, $x_{ij}$, for $j \neq i \in \{1, \dots, n\}$ in the time period from $t_{m}$ to $t_{m+1}$.
 The CTMC model assumes that the actors $i$ are conditionally independent given their ties $x_{i1}, \dots, x_{in}$ at the current network state.
-Let $\tau(i|x,m)$ be the wait time until actor $i$ gets to the opportunity to change its ties.
+Let $\tau(i | x_{i1}(m), \dots, x_{in}(m))$ be the wait time until actor $i$ gets to the opportunity to change its ties.
 For any time point $T$, where $t_{m} \leq T < t_{m+1}$, the waiting time to actor $i$'s next change follows an exponential distribution,
@@
 The value $\tau(x)$ is also exponentially distributed, and has expected value $(n\alpha_m)^{-1}$.
-The estimation of $\alpha_m$ in \texttt{RSiena} is simple: method of moments (MoM) is used with the statistic
+The estimation of $\alpha_m$ in \texttt{RSiena} is simple: method of moments (MM) is used with the statistic
 \begin{equation}\label{eq:cdiff}
@@
 The parameters of the model, $\beta_1, \dots, \beta_K$ correspond to network statistics, $s_{ik}(x, \mathbf{Z})$, for $k = 1,\dots, K$.
-To maximize $f_i$, $i$ will either destroy a tie, create a tie, or make no change.
-
-The parameters of $f_i(\boldsymbol{\beta}, x, \mathbf{Z})$ are either structure or covariate parameters.
-The structure parameters correspond to statistics that are \emph{only} functions of the current network state, $x$, while the covariate parameters are also functions of the covariates, $\mathbf{Z}$.
+To maximize $f_i(\boldsymbol{\beta}, x, \mathbf{Z})$, $i$ will either destroy a tie, create a tie, or make no change.
+
+The parameters of $f_i(\boldsymbol{\beta}, x, \mathbf{Z})$ are either structure or covariate parameters: structure parameters correspond to statistics that are \emph{only} functions of the current network state, $x$, while covariate parameters are also functions of the covariates, $\mathbf{Z}$.
 According to \citet[p.
@@
 The statistics corresponding to these two parameters, $s_{i1}(x)$ and $s_{i2}(x)$ are given in Figure~\ref{fig:structures} with visualizations of the corresponding effects.
-In the \texttt{RSiena} software, there are over 80 possible $\boldsymbol{\beta}$ parameters with corresponding statistics $s_{ik}(x, \mathbf{Z})$ to add to the model.
-The definition of the statistics for all possible parameters are provided in \citet{RSienamanual}, and two additional parameters and their corresponding statistics are discussed in Section~\ref{sec:exdatamods} and shown in Figure~\ref{fig:structures}.
+In the \texttt{RSiena} software, there are over 80 possible parameters $\beta_k$ with corresponding statistics $s_{ik}(x, \mathbf{Z})$ to add to the model, the definitions of which are provided in \citet{RSienamanual}.
+We define two additional parameters and their corresponding statistics in Figure~\ref{fig:structures}, and  we discuss all parameters more in Section~\ref{sec:exdatamods}.
 
@@
 \end{equation}
-We explore these probabilities further in Section~\ref{sec:algorithms}.
+These probabilities are used to define the intensity matrix.
 
@@
 This matrix describes the rate of change between two states of the CTMC process, and the rows of this matrix always add to zero.
-The state space for a network, denoted $\mathcal{X}$, contains $2^{n(n-1)}$ states: there are two possible states for an edge, $\{0,1\}$, and $n(n-1)$ possible edge relationships, as the network is directed and we exclude self-ties.
+The state space for a directed network, denoted $\mathcal{X}$, contains $2^{n(n-1)}$ states: there are two possible states for an edge, $\{0,1\}$, and $n(n-1)$ possible edge relationships.
 The intensity matrix is then a square matrix of dimension $2^{n(n-1)}$.
@@
 Thus, the rate of change between any two states, $x^b$ and $x^c$, that differ by only one tie $x_{ij}$, is the product of the rate at which actor $i$ gets to change a tie and the probability that the tie that will change is the tie to node $j$.
-The intensity matrix $\mathbf{Q}$ fully defines the CTMC model for network change.
+The intensity matrix $\mathbf{Q}$ completely defines the CTMC model for network change.
 
@@
 
-To fit a CTMC to dynamic network data, we apply the \texttt{RSiena} software, which uses simulation methods to estimate parameter values using either MoM or maximum likelihood (ML) estimation \citep{RSiena}.
-We chose to use MoM estimation, so we describe that process here, while more information on ML estimation can be found in \citet{SienaAlgs}.
+To fit a CTMC to dynamic network data, we apply the \texttt{RSiena} software, which uses simulation methods to estimate parameter values using either MM or maximum likelihood (ML) estimation \citep{RSiena}.
+We chose to use MM estimation, so we describe that process here, while more information on ML estimation can be found in \citet{saomsml2010, SienaAlgs}.
 %because the theory behind it was established in \citet{saompaper}, while the maximum likelihood estimation methods were not fully established until \citet{saomsml2010}, though \texttt{RSiena} contains capabilities to use maximum likelihood estimation.
 We also use the score function method for estimating the derivatives of the expected values, as opposed to the finite differences method, both of which are outlined in detail in .
-The underlying SIENA software uses a Robbins-Monro algorithm (see \citet[see][]{robbinsmonro}) to estimate the solution of the moment equation
+The underlying SIENA software uses a Robbins-Monro algorithm (see \citep{robbinsmonro}) to estimate the solution of the moment equation,
 \begin{equation}\label{eq:mom}
@@
 fix it.
-The full SIENA MoM estimation algorithm operates in three phases, as described in \citet{RSienamanual} and \citet{SienaAlgs}, which we briefly summarize here.
+The full SIENA MM estimation algorithm operates in three phases which we briefly summarize here.
+Full description of the algorithmc can be found in \citet{RSienamanual} and \citet{SienaAlgs}.
 The first phase obtains initial estimates of the score functions for use in the second phase.
-The second phase is the Robbins-Monro algorithm, and results in parameter estimates through iterative updates and simulation from the CTMC at the iterations of the parameter values.
-The third phase uses the parameters from phase two to estimate the score functions and covariance matrix of the parameter estimates and carries out convergence checks.
-In Section~\ref{sec:algorithms}, we further explore these phases in the SIENA MoM algorithm through visualization.
-For a more detailed description of the fitting process, we direct the reader to \citet{RSienamanual, SienaAlgs}.
+The second phase is the Robbins-Monro algorithm, which results in parameter estimates via iterative parameter updates and subsequent simulation from the CTMC at each set of parameter values.
+The third phase uses the parameters estimated in phase two to estimate the score functions and covariance matrix of the parameters and carries out convergence checks.
+In Section~\ref{sec:algorithms}, we further explore these phases in the SIENA MM algorithm through visualization.
 
@@
 Each girl is represented by a node and an arrow is drawn from her to each of the friends she has reported.
-Sometimes, two girls identify each other as friends, as shown by nodes 8 and 9 in the first sample on the left in Figure~\ref{fig:alldataviz}.
+Sometimes, two girls identify each other as friends, as shown by nodes 8 and 9 in the first sample, on the left in Figure~\ref{fig:alldataviz}.
 Often, the ties are not reciprocated, as shown by the edges from node 5 to nodes 2 and 3 in the first sample.
 %\hh{HH: ground the figure - this is the first time you are showing this example. A node is ... an edge is ... , colour is ...}
-We also color nodes in Figure~\ref{fig:alldataviz} according to each girl's reported drinking behavior: does not drink, drinks once or twice a year, drinks once a month, and drinks once a week.
+We also color nodes in Figure~\ref{fig:alldataviz} according to each girl's self-reported drinking behavior: does not drink, drinks once or twice per year, drinks once per month, and drinks once per week.
 We can see that, as time passes, the students seem to drink more and become increasingly isolated into smaller groups.
@@
 \end{equation}
-where $s_{i1}(x)$ is the density network statistic and $s_{i2}(x)$ is the reciprocity network statistic for actor $i$ at the current network state $x$ (see Figure~\ref{fig:structures}).
+where $s_{i1}(x)$ is the density network statistic and $s_{i2}(x)$ is the reciprocity network statistic for actor $i$ at the current network state $x$ (see Figures~\ref{fig:beta1} \& \ref{fig:beta2}).
 The second and third models, M2 and M3, contain one additional parameter each in the objective function.
@@
 ggplot(data = s1net, aes(from_id = from, to_id = to)) +
-  geom_net(directed = T, labelon = T, arrowsize = .5,
+  geom_net(directed = T, labelon = T, arrow = arrow(type = "closed", length = unit(2,"mm")),
            labelcolour='white',vjust = 0.5, hjust =0.5, arrowgap = .15,
            size=6, fontsize = 5, ecolour="black",
-           linetype = c("solid", "solid","dotted", "solid","solid", "solid","solid", "solid")) +
+           linetype = c("dotted", "solid", "solid","solid","solid", "solid","solid", "solid")) +
   expand_limits(x=c(-0.2,1.2), y=c(-0.1,1.1)) +
@@
 @
-\caption{\label{fig:beta1}Demonstrating the outdegree effect $\beta_1$. The dotted tie is encouraged if $\beta_1$ is positive. $s_{i1}(x) = \sum_{j} x_{ij}$}
+\caption{\label{fig:beta1}Demonstrating the outdegree effect $\beta_1$. $s_{i1}(x) = \sum_{j} x_{ij}$}
 \end{subfigure}\hfill
@@
            size=7, fontsize = 4, ecolour="black",
-           linetype = c("solid","dotted")) +
+           linetype = c("dotted","solid")) +
   expand_limits(x=c(-0.2,1.2), y=c(-0.1,1.1)) +
@@
 @
-\caption{\label{fig:beta2}Demonstrating the reciprocity effect $\beta_2$. The dotted tie is encouraged if $\beta_2$ is positive. $s_{i2}(x) = \sum_{j} x_{ij}x_{ji}$}
+\caption{\label{fig:beta2}Demonstrating the reciprocity effect $\beta_2$. $s_{i2}(x) = \sum_{j} x_{ij}x_{ji}$}
 \end{subfigure}
@@
 @
-\caption{\label{fig:jtt}Demonstrating a JTT, where the covariate is represented by the shape of the nodes.
-The dotted tie is encouraged if $\beta_3$ is positive.
-$s_{i3}(x,z) = \sum_{j \neq h} x_{ij}x_{ih}x_{hj} \mathbb{I}(z_i = z_h \neq z_j)$}
+\caption{\label{fig:jtt}Demonstrating a JTT, where the covariate is represented by the shape of the nodes.  $s_{i3}(x,z) = \sum_{j \neq h} x_{ij}x_{ih}x_{hj} \mathbb{I}(z_i = z_h \neq z_j)$}
 \end{subfigure}\hfill
@@
 @
-\caption{\label{fig:dad}Demonstrating a N22 between actors $i$ and $j$.
-The dotted tie is encouraged if $\beta_4$ has a positive value.
-$s_{i4}(x) =  |\{j : x_{ij} = 0, \sum\limits_h x_{ih}x_{hj} \geq 2\}|$}
+\caption{\label{fig:dad}Demonstrating a N22 between actors $i$ and $j$.  $s_{i4}(x) =  |\{j : x_{ij} = 0, \sum\limits_h x_{ih}x_{hj} \geq 2\}|$}
 \end{subfigure}
-\caption{\label{fig:structures} The network statistics whose corresponding effects are included in the models we fit to the friends data.}
+\caption{\label{fig:structures} The network statistics whose corresponding effects are included in the models we fit to the friends data.
+In each drawing, the dotted tie is encouraged to for if the parameter is positive, and is discouraged otherwise.}
 \end{figure}
 
-\noindent where $s_{i3}(x,z)$ and $s_{i4}(x)$ are defined in Figure~\ref{fig:structures}.
-These statistics are called the "number of jumping transitive triplets" (JTT) and the"number of doubly achieved distances two effect" %\hh{HH do you need the `two` in there?
+\noindent where $s_{i3}(x,z)$ and $s_{i4}(x)$ are defined in Figures~\ref{fig:jtt} \& \ref{fig:dad}.
+These statistics are called the ``number of jumping transitive triplets" (JTT) and the ``number of doubly achieved distances two effect" %\hh{HH do you need the `two` in there?
 ST: Yes its in the documentation}
@@
 The JTT effect emphasizes triads formed between actors with \emph{different} covariate values, while the N22 effect emphasizes \emph{indirect} ties between actors.
-All model effects are visualized in Figures~\ref{fig:beta1}-\ref{fig:dad}, where the ties affected by the parameter value are represented by dotted lines.
-If the estimates of these parameters $\hat{\beta}_1, \dots, \hat{\beta}_4$ are positive, the model encourages the tie represented by the dotted line to form, while the opposite is true if the parameter estimate is negative.
-Mean estimates from 1,000 repeated fittings are given in Table~\ref{tab:meanests}.
-Any time we simulate from a network model, these are the parameters we use.
+The four model parameters are visualized in Figure~\ref{fig:structures}, where the ties affected by the parameter value are represented by dotted lines.
+If the estimates of the parameters $\hat{\beta}_1, \dots, \hat{\beta}_4$ are positive, the model encourages the tie represented by the dotted line to form, while the opposite is true if the parameter estimate is negative.
+Mean estimates from 1,000 repeated fittings are given in Table~\ref{tab:meanests}, which we use when we simulate from the network models.
 
@@
 
-\caption{\label{tab:meanests} The means (std.\ dev.)\ of parameter values estimated from 1,000 repeated fittings of M1, M2, and M3 to the example data.}
+\caption{\label{tab:meanests} The means (standard deviations)\ of parameter values estimated from 1,000 repeated fittings of M1, M2, and M3 to the example data.}
 \end{table}
@@
 The values in Table~\ref{tab:meanests} show us that all effects are very significant and relatively stable across the different models.
-\hh{HH: sentence of discussion - do we see the expected increase/decrease in the parameters?} \st{I do not understand.}
+Estimates of $\beta_1$ are negative in all three models, meaning that creating new outgoing ties is discouraged, while $\beta_2$ estimates are positive in all three models, meaning that creating reciprocated ties is encouraged.
+Thus in the context of teenage girls, we interpret this to mean they do not want to form new friends unless they are becoming friends with a girl who already thinks of them as a friend.
+The estimate of $\beta_3$ is positive, meaning that girls want to be close friends with their friend's friends even if they have different drinking behaviors.
+And finally, the estimate of $\beta_4$ is negative, so indirect relationships are discouraged.
+%\hh{HH: sentence of discussion - do we see the expected increase/decrease in the parameters?} \st{I do not understand.}
 What we do not learn from the values in Table~\ref{tab:meanests} is anything about the goodness of fit of each of the models, or dependencies between the parameters and their estimates.
@@
 % \hh{HH The next paragraph could be shortened some more to emphasize what we (want to) do. ST Done. I think!}
-Our goal here is to assess the fit of these models to the data, not to find the best model for this data, which is why we have kept both the data and the models very small.
-A true analysis of these data would include all of the available data, including additional covariates related to smoking and other behaviors.
+Our goal here is to assess the fit of models M1, M2, and M3, to the data, not to find the best model for this data, which is why we have kept both the data and the models very small.
+A complete analysis of the network would include all of the available data, including additional covariates related to smoking and other behaviors.
 The exact model structure and effects we choose are not as important to us as our goal of assessing their impact on model fit.
@@
 We use the \texttt{R} package \texttt{geomnet} to visualize the three data spaces together  \citep{geomnet}.
-The package draws the network using node-link diagrams, and the network's node or edge data are then mapped to different visual features.
+The package draws the network using node-link diagrams, and node or edge data are then mapped to different visual features.
 The color, size, and shape of the points can be used to represent variables in the node data, while the color, linewidth, and line type of the edges between nodes can be used to represent variables in the edge data.
-Finally, the networks observed at each time point are placed side-by-side, showing the network's evolution in time.
+Finally, the networks observed at each time point are placed side-by-side, showing the network evolve over time.
 Bringing all network information together in \texttt{geomnet} allows us to show the entire data space at once.
@@
            directed = T, curvature = .2, labelon=T, size = 3,
-           arrow = arrow(angle = 20, type = "open", length = unit(.2, "cm")), ecolour = "#969696",
+           arrow = arrow(angle = 20, type = "open", length = unit(.05, "cm")), ecolour = "#969696",
            color = "#252525", hjust = .5, vjust = .5, labelcolour = 'white', fontsize = 2, arrowgap = .03,
@@
 <<averagenet, results='hide', out.width='\\textwidth', fig.cap='In the bottom left, the network at the first time point ($x(t_1)$), which is conditioned on in the model.
-In the bottom right, the network at the second time point($x(t_2)$).
+In the bottom middle, the network at the second time point ($x(t_2)$).
 In the top row, the three expected networks for time point 2 ($\\bar{x}(t_2)$) according to model M1, M2, M3.
-Legend shows edge weights in the three expected networks.', fig.height=5>>=
+The legend shows how edge width corresponds to frequency in the three expected networks.', fig.height=5>>=
 
@@
   geom_point(aes(x = from, y = to)) + 
-  geom_line(aes(x = from, y = to, size = linewidth), color = "#969696") +
+  geom_line(aes(x = from, y = to, size = linewidth), color = "black") +
   scale_size_continuous(name = "Edge frequency", range = c(0,1)) + 
@@
            directed = T, curvature = .2, labelon=T, size = 3, layout.alg = "fruchtermanreingold", 
-           arrow = arrow(angle = 20, type = "open", length = unit(.2, "cm")), ecolour = "black", fiteach = F,
-           color = "grey60", hjust = .5, vjust = .5, labelcolour = 'white', fontsize = 2, arrowgap = .02,
+           arrow = arrow(angle = 20, type = "open", length = unit(.1, "cm")), ecolour = "black", fiteach = F,
+           color = "grey60", hjust = .5, vjust = .5, labelcolour = 'white', fontsize = 2, arrowgap = .04,
            singletons = T) +
@@
 \item Construct one weighted summary network for time points 2 and 3.
-Each summary network has edge weights equal to the proportion of the simulated networks in which the edge appeared.i.e.\ $\bar{x}_{ij}(t_m) = x_{ij}^{(B)}(t_m) /1000$.
+Each summary network has edge weights equal to the proportion of the simulated networks in which the edge appeared.
+i.e.\ $\bar{x}_{ij}(t_m) = x_{ij}^{(B)}(t_m) /1000$.
 \item Create the \textit{expected network} by including only the edges that appeared in more than 5\% of the simulations.
@@
 \noindent The above method can be generalized and applied to any network model, static or dynamic, from which we can simulate, for any number of simulations from the network.
+In step 4, we chose 5\% because that is the typical significance level in statistical tests, and other values could be chosen.
 The expected networks from our three models of interest are shown in the top row of Figure~\ref{fig:averagenet}.
 Edges that appear with higher frequency are drawn with thicker lines.
-An edge frequency value of one indicates that the edge was present in all simulations used to create the expected network.
+Edge frequency equal to one indicates that the edge was present in all 1,000 simulations used to create the expected network.
 %\hh{HH Give a definition of how edge frequency is defined in the plot - frequency of 1 means that this edge was present in all of the simulations, right?ST DONE.
@@
 For model M1 we can see that the overall structure of the average network is much more similar to $x(t_1)$  than to $x(t_2)$ .
-In both $x(t_1)$ and $\bar{x}(t_2)$, there are two clusters of nodes: $\{2,3,4,5,6,12,13,15\}$ and $\{7,8,9,10,11,14\}$.
-The expected network, however, does not resemble  $x(t_2)$, which it is supposed to represent.
+In  $x(t_1)$ and each of the three $\bar{x}(t_2)$, there are two clusters of nodes: $\{2,3,4,5,6,12,13,15\}$ and $\{7,8,9,10,11,14\}$.
+The expected networks, however, do not resemble  $x(t_2)$, which they are supposed to represent.
 In $x(t_2)$, there is one small, strongly connected cluster and one larger, sparsely connected cluster, with four unconnected nodes.
-Because the expected network $\bar{x(t_1)}$ more closely resembles $x(t_1)$ than $x(t_2)$, we take this as evidence that the simple model, M1, is not a good fit to the data.
-Comparing the three expected networks to $x(t_1)$ and $x(t_2)$, we see that they all have nodes that are connected in the same groups as $x(t_1)$.
+Because the expected networks $\bar{x}(t_2)$ more closely resemble $x(t_1)$ than $x(t_2)$, we take this as evidence that none of the models are a good fit to the data.
 Model M2, which included the transitive triplet parameter, results in more strongly connected components, as shown by the increased number of edges and the thicker lines between nodes.
-The lack of fit in all models is clear:  not one of the three model expectations show node 16 gaining ties as it does in $x(t_2)$, nor do they show nodes 4 and 7 becoming isolated.
-In model M2, however, the ties to node 7 appear much weaker than in model M1 or model M3, and the nodes $\{10,11,14\}$ are also more strongly connected, as they are in wave 2 of data.
+The lack of fit in all models is clear:  not one of the three model expectations show node 16 gaining ties as it does in $x(t_2)$, nor do they show nodes 4, 7, and 15, becoming isolated.
+In model M2, however, the ties to node 4, 7, and 15, appearless frequently than in model M1 or model M3, and the nodes $\{10,11,14\}$ are also more often connected, as they are in wave 2 of data.
 This suggests that, despite the fact that neither of the three models is a good fit, M2 may be the best fitting model of the three.
-In a complete data analysis, we can use this information to select more models for analysis that utilize transitivity and covariate effects, as opposed to structural and indirect effects.
+In a complete data analysis, we can use this information to select more models for analysis that utilize transitivity and covariate effects like $\beta_3$, as opposed to structural and indirect effects like $\beta_4$.
 % We can use the average network to help determine model goodness-of-fit.
@@
 
-<<summary2, fig.height=2.5, fig.cap="Differences between the expected network  $\\bar{x}(t_2)$ and $x(t_1)$  and  $x(t_2)$.
+\subsection{Residual Networks}
+
+<<summary2, fig.height=2.5, fig.cap="Residual networks between the expected network from model M1 $\\bar{x}(t_2)$ and $x(t_1)$  and  $x(t_2)$.
 Some edges appear in both the data and the expected network $\\bar{x}(t_2)$ (dotted lines), while others only appear in the data (solid lines) or only in the expectation (dashed lines).
-Width of non-data edges represent the proportion of appearances in the 1,000 simulations.">>=
+Width of dotted and dashed edges represent the frequency of appearances in the 1,000 simulations.">>=
 
@@
 t1_1 <- arrange(t1_1, from, to)
+set.seed(25)
 p1 <- ggplot(data = t1_1) +
@@
            directed = T, curvature = .2, labelon=T, size = 3, ecolour = "black",
-           arrow = arrow(angle = 20, type = "open", length = unit(.15, "cm")), 
-           color = "gray60", hjust = .5, vjust = .5, labelcolour = 'white', fontsize = 2, arrowgap = .02, layout.alg = "fruchtermanreingold",
+           arrow = arrow(angle = 20, type = "closed", length = unit(.15, "cm")), 
+           color = "gray60", hjust = .5, vjust = .5, labelcolour = 'white', fontsize = 2, arrowgap = .03, layout.alg = "fruchtermanreingold",
            singletons = T) +
   scale_linetype_manual(values =c("longdash", "dotted") ) + 
-  labs(title = "Difference between Time 1 Data &\n Expectation from M1") + ThemeNet +
+  labs(title = "Residual between Time 1 Data &\n M1 Expectation") + ThemeNet +
   theme(plot.background = element_rect(color = 'black'), legend.position = "none")
@@
 pforleg <- ggplot(data = mtcars) + 
-  geom_line(aes(x = mpg, drat, group = as.factor(cyl), linetype = as.factor(cyl))) + 
+  geom_line(aes(x = mpg, drat, group = as.factor(cyl),
+                linetype = as.factor(cyl))) + 
+  geom_line(aes(x = mpg, drat, size = vs)) + 
   scale_linetype_manual(values = c("dotted", "solid", "longdash"), name = "Tie is...", labels = c("in data, in\nexpectation", "in data,not in\nexpectation", "not in data, in\nexpectation")) +
+   scale_size_continuous(name = "Edge Frequency", breaks = 1:4/4, range = c(.25, 1)) + 
   theme_grey() + 
-  theme(legend.key = element_rect(color = "black", fill = "white"), 
-        legend.key.height = unit(25, "pt"))
+  theme(legend.key = element_rect(color = "black", fill = "white"))
 
@@
 t1_2 <- t1_2 %>% filter(!(from == 16 & is.na(to)))
+set.seed(25)
 p2 <- ggplot(data = t1_2) +
@@
            directed = T, curvature = .2, labelon=T, size = 3, ecolour = "black",
-           arrow = arrow(angle = 20, type = "open", length = unit(.15, "cm")), 
-           color = "gray60", hjust = .5, vjust = .5, labelcolour = 'white', fontsize = 2, arrowgap = .02, layout.alg = "fruchtermanreingold",
+           arrow = arrow(angle = 20, type = "closed", length = unit(.15, "cm")), 
+           color = "gray60", hjust = .5, vjust = .5, labelcolour = 'white', fontsize = 2, arrowgap = .03, layout.alg = "fruchtermanreingold",
            singletons = T) +
   scale_linetype_manual(values = c("longdash", "solid", "dotted")) + 
-  labs(title = "Difference between Time 2 Data &\n Expectation from M1") + ThemeNet +
+  labs(title = "Residual between Time 2 Data &\n M1 Expectation") + ThemeNet +
   theme(plot.background = element_rect(color = 'black'), legend.position = "none")
 legendsumm <- cowplot::get_legend(pforleg)
-ggdraw() + draw_plot(p1, width = .4) +  draw_plot(p2, width = .4, x=.4) + 
-  draw_grob(legendsumm,x=.8, y=.25, width = .2, height = .4)
-@
-
-\subsection{Residual Networks}
+ggdraw() + draw_plot(p1, x = 0, y = 0, height = 1, width = .38) +  
+  draw_plot(p2, width = .38, x = .39, y = 0, height = 1) + 
+  draw_grob(legendsumm,x=.78, y=.4, width = .2 , height = .2,  scale = .1, clip = "on") 
+@
 
 Another way to view our models in the data space is with \textit{residual networks} between the expected network and the data, as shown in Figure~\ref{fig:summary2}.
-We created these networks by computing the difference between $\bar{x}(t_2)$ and $x(t_1)$, $\mathbf{A}(\bar{x}(t_2)) - \mathbf{A}(x(t_1))$, and the difference between $\bar{x}(t_2)$ and $\bar{x}(t_2)$, $\mathbf{A}(x(t_2)) - \mathbf{A}(\bar{x}(t_2))$ for model M1.
-The lack of model fit is apparent in Figure~\ref{fig:summary2}, where the dashed edges represent ties that only show up in the expected network, and the solid edges representing ties in the data that the model fails to capture.
+We created these networks by computing the residual between $\bar{x}(t_2)$ and $x(t_1)$, $\mathbf{A}(x(t_1))-\mathbf{A}(\bar{x}(t_2))$, and the residual between $\bar{x}(t_2)$ and $\bar{x}(t_2)$, $\mathbf{A}(x(t_2)) - \mathbf{A}(\bar{x}(t_2))$ for model M1.
+The lack of model fit is apparent on the right in Figure~\ref{fig:summary2}, where the dashed edges represent ties that only show up in the expected network, and the solid edges represent ties in the data that the model fails to capture.
 Dotted edges are in both the expectation and the data and represent where the model is correct.
@@
 However, there are six solid edges, which are in the data but do not appear at all in the expected network, and there are many dashed edges that appear in the expected network but are not in $x(t_2)$.
-Although the residual network shows that there some edges that appear in both the expected network and $x(t_2)$, the majority of edges in the expected network are not in the observed data, and there are several edges in the data missing from the expectation.
+Although the residual network shows some edges that appear in both the expected network and $x(t_2)$, the majority of edges in the expected network are not in the observed data, and there are several edges in the data missing from the expectation.
 We therefore conclude that the model M1 is a poor fit to our example data, which we have seen very clearly by viewing the model in the data space.
@@
 Finally,  we fit the model many times to assess the behavior of the underlying simulations.
-We chose these three collections because they each explore something different about model behavior: the choice of and relationships between parameters in the objective function and the stochastic effects from simulation.
+We chose these three collections because they each explore something different about model behavior: the choice of and relationships between parameters in the objective function, and the stochastic effects from simulation.
 %, and the resulting fitted values under different circumstances.
@@
 To select objective function parameters for modeling our example data, we searched through all available parameters to find  significant effects.
-The \texttt{RSiena} package includes the function \texttt{Wald.Rsiena} for Wald-type tests of parameter significance \citep{RSiena}.
+The \texttt{RSiena} package includes the function \texttt{Wald.Rsiena()} for Wald-type tests of parameter significance \citep{RSiena}.
 We start by including the outdegree and reciprocity parameters in the model then add one effect to the objective function, fit the model to the example data, test the added effect for significance, and repeat for all possible model parameters.
@@
 We could also model the behavior changes, not just the network changes, if we believe that friends influence behavior via peer pressure.
-And as before, this method of gathering a collection of models can be used for any network model, such as the hierarchical Bayesian network estimation model in the \texttt{R} package \texttt{sna}, which allows the user to vary hyperparameters in the prior and model parameters \citep{sna}.
+This method of gathering a collection of models can be used for any network model which allows the user to vary hyperparameters in the prior and model parameters, such as the hierarchical Bayesian network estimation model in the \texttt{R} package \texttt{sna} \citep{sna}.
 % took out this next bit just because it doesn't really add much and we can just put the appendix figure into the previous section. % add more later.
@@
 \caption{\label{fig:distests} Distribution of fitted parameter values for our three models.
-The inclusion of $\beta_3$ or $\beta_4$ clearly has an effect on the distributions of the rate parameters, $\alpha_1$ and $\alpha_2$, and on the other objective function parameters $\beta_1, \beta_2$.}
+The inclusion of $\beta_3$ clearly has an effect on the distributions of the rate parameters, $\alpha_1$ and $\alpha_2$, and on the other objective function parameters $\beta_1, \beta_2$, while the effect of $\beta_4$ is not as noticeable.}
 \end{figure}
@@
 The $\beta_1$ parameter is also highly correlated with the $\beta_3$ parameter within model M2 ($\rho = $ \Sexpr{round(corm2[5, 3],3)}), but it is not as highly correlated with the $\beta_4$ parameter in model M3 ($\rho = $ \Sexpr{round(corM3[5, 3], 3)}).
-The inherent relationship between effects is a known problem in network analysis \citep{goldenberg09}.
+The inherent relationship between effects is a known problem in network analysis \citep{goldenberg09}, and the diagnostic pairs plot can be used for any network model where parameters are estimated with MCMC methods.
 % need a better reference here.
@@
 When fitting models in \texttt{RSiena}, several simulation processes are hidden from the user.
-We refer to one of these hidden pieces in the SIENA MoM algorithm as the \textit{microstep process}, which is a series of state changes in the CTMC as described in Section~\ref{sec:saoms}.
+We refer to one of these hidden pieces in the SIENA MM algorithm as the \textit{microstep process}, also called the ministep process in \citet{SienaAlgs}.
+The microstep process is a series of state changes in the CTMC as described in Section~\ref{sec:saoms}.
 At each microstep, an ego node is selected to change, and then randomly makes one change in its ties according to to the probabilities $p_{ij}$ defined in Equation~\ref{eq:pij}.
-The microstep process is a key piece of the model fitting procedure and is used in the third phase mentioned in Section~\ref{subsec:fitting} \citep{SienaAlgs}.
-During modeling, the process proceeds until the network differs from the $x(t_m)$ by exactly $C_m$, as defined in Equation~\ref{eq:cdiff}.
+The microstep process is a key piece of the model fitting procedure.
+During modeling, the process proceeds until the network differs from the $x(t_m)$ by exactly $C_m$, defined in Equation~\ref{eq:cdiff}.
 Between two network observations $x(t_m)$ and $x(t_{m+1})$, there can be dozens, hundreds, or even thousands of microsteps, depending on the size of the network and the value of $C_m$.
-We want to simulate and view these in-between steps in order to better understand the behavior of the CTMC model.
-The default in the software is discard the microsteps, which is practical and efficient if the goal is to fit a model to data, obtain parameter estimates, and draw conclusions or make predictions.
+We want to simulate and view these steps in order to better understand the behavior of the CTMC model.
+By default, the software discards the microsteps, which is practical and efficient if the goal is to fit a model to data, obtain parameter estimates, and draw conclusions or make predictions.
 But in order to understand how the models are fit, we need to extract and explore the different steps of the process, which is possible in \texttt{RSiena} with some small changes to the default settings.
-We use these steps to visually explore the algorithms, so that we are visualizing the model fitting process, and not only the model fitting result.
+We visualize the microsteps to visually explore the algorithms, so that we are visualizing the model fitting process, and not only the model fitting result.
 
@@
 
-<<makinggif, fig.cap='A selection of images in the microstep animation at \\url{https://vimeo.com/240089108}.
+<<makinggif, fig.cap='Some frams from the animation at \\url{https://vimeo.com/240089108}.
 The selected edges and nodes are emphasized by changing size and color, then they fade in or out while the nodes change color and shrink to blend in with the other nodes.
 Frames 15-28 show edge destruction, while frames 71-84 show edge creation.
-A spring layout is used.', message = FALSE, warning = FALSE, fig.height=2.5>>=
+A spring layout is used \\citep{sna}.', message = FALSE, warning = FALSE, fig.height=2.5>>=
 friend.data.w1 <- as.matrix(read.table("data/s50_data/s50-network1.dat"))
@@
         colour = ecolor, size = esize), curvature = 0.1, arrow = arrow(angle = 20,
-        length = unit(0.1, "inches"))) +
+        length = unit(0.15, "cm"), type = "closed" )) +
   geom_point(data = nodedata,
@@
 If there are no changes at a particular microstep, the microstep does not appear.
-Some frames of the animations are shown in Figure~\ref{fig:makinggif}, and the full movie, created with \texttt{tweenr} and \texttt{gganimate} can be viewed at \url{https://vimeo.com/240089108} \citep{tweenr, gganimate}.
+Some frames of the animations are shown in Figure~\ref{fig:makinggif}, and the full movie, which we created using \texttt{tweenr} and \texttt{gganimate} can be viewed at \url{https://vimeo.com/240089108} \citep{tweenr, gganimate}.
 Movies similar to this animation were used to visualize the changes of dynamic networks in \citet{dynnetviz}.
@@
 We see each part of the hierarchical model: first, the selection of the ego node via the rate function, then the result of the actor maximizing its objective function by either deleting or adding a node.
-Finally, the layout of the network transitions as edges are removed or added, demonstrating how the overall network structure changes with each of the individual tie changes.
-Combining the tie changes in this animation makes the CTMC process the focus, demonstrating the underlying change mechanism of the dynamic network according to the model.
-
-With this animation, we can see how the model behaves.
-Some tie changes appear to alter the overall network structure, as shown by the change in the layout of the nodes, while others seem to be inevitable, such as an added reciprocal tie between two nodes in a large connected group that do not change the layout.
+The layout of the network also adjusts as edges are removed or added, demonstrating how the overall network structure changes with each of the individual tie changes.
+Combining the microstreps in this animation makes the CTMC process the focus, demonstrating the underlying change mechanism of the dynamic network according to the model.
+
+With the node-link animation, we can see how the model behaves.
+Some tie changes appear to alter the overall network structure, as shown by the change in the layout of the nodes, while others seem to be inevitable, such as an added reciprocal tie between two nodes in a large connected group that does not change the layout.
 Other edges are created in one step and destroyed the next.
@@
 We continue to use animation to view the changing structure of the network via adjacency matrix visualizations.
-In the first adjacency matrix visualization in Figure~\ref{fig:adjmatorder}, we use the node ID to order the rows and columns.
-This arbitrary ordering does not easily convey the underlying structure of the network.
-We use \textit{matrix seriation} to reorder the rows and columns of the adjacency matrix visualization so that the structure of the network is apparent \citep{seriation}.
+Two adjacency matrix visualizations are shown in Figure~\ref{fig:adjmatorder}.
+At left in Figure~\ref{fig:adjmatorder}, we use the node ID to order the rows and columns.
+This arbitrary ordering does not easily convey the underlying structure of the network, so we use \textit{matrix seriation} to reorder the rows and columns of the adjacency matrix visualization to make the structure of the network more obvious \citep{seriation}.
 %This lack of perceived structure would contribute to further confusion as the network chages, so we adjust the ordering so that the viewer can better perceive the structure of the network.
@@
 Using matrix seriation on the %PCA on $\mathbf{A}^{cum}$ to order the rows and columns of the 
-adjacency matrix visualization clearly shows the two distinct connected components in the first network, which are difficult to find in the visualization ordered by node ID.
-We continue to use the seriated matrix in another animation of the microstep process, which is available at \url{https://vimeo.com/240092677}.
+adjacency matrix visualization clearly shows the two distinct connected components in the first network, which are difficult to find in the visualization ordered by node ID at left.
+We use the seriated matrix in an adjacency matrix animation of the microstep process, which is available at \url{https://vimeo.com/240092677}.
 Some frames of the video are shown in Figure~\ref{fig:adjmatani}.
@@
 By the end, the bottom left component has become more sparsely connected, while the top right component has shrunk, but remains closely connected.
-In contrast to the node-link animation, the adjacency matrix animation shows how the algorithm explore the space of all possible networks, because possible edges are represented in each frame of the adjacency matrix animation.
+In contrast to the node-link animation, the adjacency matrix animation shows how the algorithm explores the space of all possible networks, because all possible edges are represented in each frame of the adjacency matrix animation.
 It is also easier to see the connected components of the graph in this animation because we have clearly grouped them with matrix seriation.
@@
 Because the adjacency matrix emphasizes edges, this animation method is best for large directed networks whose structure is easier to see in an adjacency matrix visualization than in a node-link visualization \citep{adjmatviz}.
-Both animation methods demonstrate model behavior that would have not been seen with traditional visualizations or computations.
+Both animation methods demonstrate model behavior that we would not have seen with traditional visualizations or computations.
 % need more oomph
@@
 
-We have introduced diagnostic visualizations, the expected network, the residual network, the node-link animation, and the adjacency matrix animation, to complement existing numerical diagnostics for network models.
+We have introduced four diagnostic visualizations, the expected network, the residual network, the node-link animation, and the adjacency matrix animation, to complement existing numerical diagnostics for network models.
 These visual diagnostics give insights into the network model fit that purely numerical methods do not.
@@
 Exploring collections of models give us the opportunity to learn more about the large number of parameters possible in CTMC models.
-The diagnostic pairs plot visualizes collections of models giving us insight into the marginal distribution of parameters and their covariance structure.
+The diagnostic pairs plot visualizes collections of models, giving us insight into the marginal distribution of parameters and their covariance structure.
 %and we also visualized relationships between parameters in the model that require further investigation.
@@
 The models and data we have used are multilayered and multidimensional, making them more complex than common statistical models, such as linear models.
-As models become more complicated, visualization is more important to model diagnosis because increasing complexity makes it harder to adequately summarize model fit only using numerical methods.
+As models become more complicated, visualization becomes more important to model diagnosis because increasing complexity makes it harder to adequately summarize model fit only using numerical methods.
 Even more complex models, such as those for machine learning, can also benefit greatly from model visualization techniques without a lot of additional effort.
@@
 
-We only explored a few simple models from the complicated and multilayered CTMC model family for social networks.
-%The \texttt{RSiena} software is incredibly powerful, and we can fit much more flexible hierarchical models than we have examined here.
-Many more than three parameters can be included in the objective function, and more than one actor-level covariate can be included at either level in the hierarchical model.
+We  explored a few simple models from the complicated and multilayered CTMC model family for social networks, %The \texttt{RSiena} software is incredibly powerful, and we can fit much more flexible hierarchical models than we have examined here.
+but many more than three parameters can be included in the objective function, and more than one actor-level covariate can be included at either level in the hierarchical model.
 We can also model the rate of change differently for each actor if, for example, we think an actor's covariate affects its rate of change.
@@
 With increased model complexity, there is more potential for adverse effects, such as the correlation between parameters that we saw in Section~\ref{sec:collectmodels}.
-Armed with model-vis techniques, researchers can gain a better understanding of these and other adverse effects and find ways to alleviate them the model.
-In addition, some extensions of the CTMC model can also model behavior change of the actors in the network, which we have just treated as covariates.
-The CTMC family of models for dynamic networks has immense potential for additional model visualization applications.
+Armed with model-vis techniques, researchers can gain a better understanding of these and other adverse effects and can then find ways to alleviate them the model.
+In addition, extensions of the CTMC model can also model behavior change of the actors in the network, which we haveused as model covariates.
+Because of its flexibility, the CTMC family of models for dynamic networks has immense potential for additional model visualization applications.
 

### 2019-07-15 code/00e_small_friends.R

@@ -1,10 +1,10 @@
 # New model with a smaller collection of girls from the friendship data
 #deciding how to subset the girls:
-library(ggplot2)
-library(tidyr)
-library(dplyr)
+#library(ggplot2)
+#library(tidyr)
+#library(dplyr)
 
-library(RSiena) # or RSienaTest
+#library(RSiena) # or RSienaTest
 # read in network data
 friend.data.w1 <- as.matrix(read.table("data/s50_data/s50-network1.dat"))
 friend.data.w2 <- as.matrix(read.table("data/s50_data/s50-network2.dat"))
